### 2022-02-14 acks/acks.tex
initial

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
\setheader{Preface}

Without a doubt, the acknowledgments are the most widely and most
eagerly read part of any thesis.

\begin{flushright}
{\makeatletter\itshape
    Petar \\
    Amsterdam, April 2022
\makeatother}
\end{flushright}

### 2022-02-14 introduction/introduction.tex
initial

\chapter{Introduction}
\label{introduction}

\dropcap{H}alf sleeping, half playing Fruit Ninja on my smartphone at one of the
philosophy lectures in my high school I heard my professor exclaim with
excitement: “Mathematical truth is the highest and absolute form of truth!”.
Unbothered and uninterested, I continued playing Fruit Ninja.

Many years later I started studying various formal methods in computer science.
There I learned the value of formal, mathematical languages and finally
understood him. Not only does using a rigorous mathematical language avoid the
ambiguity of natural language, it allows us to use formal rules of reasoning to
make conclusions from initial arguments in a trustworthy manner.

Though the interest for rigorous and formal reasoning dates back to Aristotle,
it intensified at the end of 19th and the beginning of 20th century
\cite{jf-01-modern-logic}. Researchers in that time were mostly interested in
finding a formal language that is expressive enough to describe complex
mathematical theories, yet intuitive and understandable enough to allow
reasonably simple formal reasoning systems. First-order logic seemed to satisfy
both requirements.  This formalism does not only allow one to model simple
logical relations such as ``if it rains, the road is muddy'' (or formally
$\cst{rains} \imp \cst{muddy}$), but also to quantify over objects as in ``all
people are mortal'' ($\forall x.\, \cst{human}(x) \imp \cst{mortal}(x)$).

The initial study of this formalism and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules such that for any true (valid) statement a
proof of validity can be constructed in this system. Around the same time, the
first computers were leaving the imagination of the engineers and began to
automate many complicated processes.  Naturally, the question as to  whether a
computer can decide if a first-order statement is valid (i.e., a theorem) arose.
Church \cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered
this question negatively in 1936.

% \begin{abstract}
% Sample Abstract.
% \end{abstract}

% \blfootnote{This chapter is partly based on \faFileTextO~\emph{M. Beller. Toward an
%     Empirical Theory of Feedback-Driven Development, ICSE'18 (Student Research Competition)}~\cite{BellerSRC2018}.
% }

% \newpage

% \dropcap{T}his is a introductory page.

% \section{Background \& Context}
% In this thesis, you can reference pictures~\Cref{fig:devmodel} using Cleverref and circles \circled{5}.

% \begin{figure}[htb]
% 	\centering
% 	\includegraphics[width=0.65\columnwidth]{development_model_without_papers}
% 	\caption{The stages of the FDD model and their relationship to other
%           Software Engineering concepts.}
% 	\label{fig:devmodel}
% \end{figure}

% We also have lists:

% \begin{enumerate}
%   \item Static Analysis~\circled{3} examines program artifacts or
%     their source code without executing them~\cite{wichmann1995industrial}, while
%  \item Dynamic Analysis~\circled{4} relies on information gathered from their
%    execution~\cite{cornelissen2009systematic}.
% \end{enumerate}

% Or boxes:

% \begin{framed}
% This thesis is concerned with the empirical assessment of the state of the art of how developers
% drive software development with the help of feedback loops.
% \end{framed}

% Or code:
% \begin{lstlisting}[caption={\textsc{TrinityCore}},label={lst:e1}]
%  x += other.x;
%  y += other.y;
%  z += other.y;
% \end{lstlisting}

% I hope this helps you get started!
% Moritz

### 2022-02-14 title/title.tex
initial

% For VSCode
% !TEX root = ../dissertation.tex
\begin{titlepage}

\begin{center}

%% Extra whitespace at the top.
\vspace*{2\bigskipamount}

%% Print the title.
{\makeatletter
\titlestyle\bfseries\LARGE\@title
\makeatother}

%% Print the optional subtitle.
{\makeatletter
\ifx\@subtitle\undefined\else
    \bigskip
    \titlefont\titleshape\Large\@subtitle
\fi
\makeatother}

\end{center}

\cleardoublepage
\thispagestyle{empty}

\begin{center}

%% The following lines repeat the previous page exactly.

\vspace*{2\bigskipamount}

%% Print the title.
{\makeatletter
\titlestyle\bfseries\LARGE\@title
\makeatother}

%% Print the optional subtitle.
{\makeatletter
\ifx\@subtitle\undefined\else
    \bigskip
    \titlefont\titleshape\Large\@subtitle
\fi
\makeatother}

%% Uncomment the following lines to insert a vertically centered picture into
%% the title page.
%\vfill
%\includegraphics{title}
\vfill

%% Apart from the names and dates, the following text is dictated by the
%% promotieregelement.

{\Large\titlefont\bfseries Proefschrift}

\bigskip
\bigskip

ter verkrijging van de graad van doctor

aan Vrije Universiteit Amsterdam,

op gezag van de Rector Magnificus prof.~dr.~ir.~ [onbekend],

voorzitter van het College voor Promoties,

in het openbaar te verdedigen

op [onbekend]

\bigskip
\bigskip

door

\bigskip
\bigskip

%% Print the full name of the author.
\makeatletter
{\Large\titlefont\bfseries\@firstname\ \titleshape{\MakeUppercase{\@lastname}}}
\makeatother

\bigskip
\bigskip

Master of Science in Computer Science, \\
Vrije Universiteit Amsterdam, Nederland,

geboren te Belgrado, Servi\"e.

%% Extra whitespace at the bottom.
\vspace*{2\bigskipamount}

\end{center}

\clearpage
\thispagestyle{empty}

%% The following line is dictated by the promotieregelement.
\noindent Dit proefschrift is goedgekeurd door de

%% List the promotors (supervisors).
\medskip\noindent
\begin{tabular}{l}
    promotoren: Dr.\ A.E.\ Zaidman, Prof.\ dr.\ A.\ van Deursen \\
    copromotor: Dr.\ ir.\ G.\ Gousios
\end{tabular}

\bigskip
\noindent Samenstelling promotiecommissie:

%% List the committee members, starting with the Rector Magnificus and the
%% promotor(s) and ending with the reserve members.
\medskip\noindent
\begin{tabular}{p{4.5cm}l}
    Rector Magnificus, & voorzitter \\
    Prof.\ dr.\ A.\ van Deursen, & Technische Universiteit Delft \\
    Dr.\ A.E.\ Zaidman, & Technische Universiteit Delft \\
    Dr.\ ir.\ G.\ Gousios, & Technische Universiteit Delft \\

    \medskip
    \mbox{\emph{Onafhankelijke leden:}} & \\
    Prof.\ dr.\ ir.\ G.J.P.M.\ Houben, & Technische Universiteit Delft \\
    Prof. dr. P. Runeson, & Lund Universitet, Sweden \\
    Dr.\ Th.\ Zimmermann, & Microsoft Research, \\ &United States of America \\
    Prof. dr. D. Spinellis, & Athens University of Economics and Business, \\&
    Greece \\

    Prof.\ dr.\ ir.\ E.\ Visser, & Technische Universiteit Delft, reservelid \\ \\

    \multicolumn{2}{l}{Prof. dr. D. Spinellis has contributed to the end
    phase of writing \ldots} \\
\end{tabular}

%% Include the following disclaimer for committee members who have contributed
%% to this dissertation. Its formulation is again dictated by the
%% promotieregelement.
%\medskip
%\noindent  %Prof.\ Dr.\ D.\ Spinellis has contributed to the creation of this thesis.

\medskip
\medskip
% TODO Include http://www.win.tue.nl/ipa/?page_id=309
\noindent The work in the thesis has been carried out under the auspices of the research school IPA
(Institute for Programming research and Algorithmics).

\medskip
%% Here you can include the logos of any institute that contributed financially
%% to this dissertation.
\vfill
\begin{center}
    \includegraphics[height=0.5in]{title/logos/tudelft}
    \hspace{2em}
    %\includegraphics[height=0.5in]{title/logos/casimir} \\
    \includegraphics[height=0.5in]{title/logos/nwo}
    \\ \vspace{0.5cm}
    \includegraphics[height=0.5in]{title/logos/ipa}
\end{center}
\vfill

\noindent
\begin{tabular}{@{}p{0.2\textwidth}@{}p{0.8\textwidth}}
  \textit{Keywords:} &  \\[\medskipamount]
      \textit{Printed by:} &  \\[\medskipamount]
      \textit{Cover:} &  \\[\medskipamount]
      \textit{Style:} & TU Delft House Style, with modifications by Moritz Beller \\& \url{https://github.com/Inventitech/phd-thesis-template} \\[\medskipamount]
\end{tabular}

\medskip
\medskip
\noindent The author set this thesis in \LaTeX\xspace using the Libertinus and Inconsolata fonts.

\vspace{\bigskipamount}

% Copyrighting this is stupid, questionable, and probably illegal, because large parts of the
% thesis have already been published with the copyright resigning with the publisher.
%\noindent Copyright \textcopyright\ 2015 by A.~Einstein

%% Uncomment the following lines if this dissertation is part of the Casimir PhD
%% Series, or a similar research school.
%\medskip
%\noindent Casimir PhD Series, Delft-Leiden 2015-01

%\medskip
\noindent ISBN \ldots

\medskip
\noindent An electronic version of this dissertation is available at \\
\url{http://repository.tudelft.nl/}.

\end{titlepage}

### 2022-02-14 introduction/introduction.tex
more

@@
 this question negatively in 1936.
+
+Despite this negative result, the prospect of automatically proving theorems of
+first-order logic remained too enticing. In 1960 Davis and Putnam
+\cite{dp-1960-dpll} described an algorithm for checking validity of a
+first-order formula. As first-order logic is undecidable this algorithm
+terminated only on valid formulas. Even though it was efficient enough to prove
+only simple formulas, it was an impressive achievement for the time.
+
+A bigger breakthrough happened in 1965 when Robinson introduced a formal system
+(calculus) that would shape the future of automatic reasoning --- the resolution
+calculus \cite{ar-65-resolution}. Consisting of a single inference rule it was
+simple and elegant. It was also \emph{complete}: each theorem could be proven
+using this system. Furthermore, it was less explosive that Davis and Putnam's
+algorithm since it introduced unification as an inference filter.
+
+Since the introduction of resolution calculus it was improved in many ways.
+Strategies and heuristics \cite{lw-65-sos} to curb the explosion of the search
+space were introduced, as well as variants of resolution calculus that are less
+explosive than the original calculus but still complete .
 

### 2022-02-14 introduction/introduction.tex
more

--- 2022-02-14 introduction/introduction.tex
@@
 
-\dropcap{H}alf sleeping, half playing Fruit Ninja on my smartphone at one of the
+Half sleeping, half playing Fruit Ninja on my smartphone at one of the
 philosophy lectures in my high school I heard my professor exclaim with
@@
 Many years later I started studying various formal methods in computer science.
-There I learned the value of formal, mathematical languages and finally
+There I learned the value of formal, mathematical language and finally
 understood him. Not only does using a rigorous mathematical language avoid the
-ambiguity of natural language, it allows us to use formal rules of reasoning to
+ambiguity of natural language, it allows us to use the formal rules of reasoning to
 make conclusions from initial arguments in a trustworthy manner.
@@
 it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
-that there is a system of inference rules such that for any true (valid) statement a
+that there is a system of inference rules (calculus) such that for any valid statement a
 proof of validity can be constructed in this system. Around the same time, the
@@
 this question negatively in 1936.
+
+Despite this negative result, the prospect of automatically proving theorems of
+first-order logic remained too enticing. In 1960 Davis and Putnam
+\cite{dp-1960-dpll} described an algorithm for checking validity of a
+first-order formula. As first-order logic is undecidable this algorithm
+terminated only on valid formulas. Even though it was efficient enough to prove
+only simple formulas, it was an impressive achievement for the time.
+
+A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
+would shape the future of automatic reasoning --- the resolution calculus
+\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
+elegant. It was also \emph{complete}: each theorem could be proven using this
+system. Furthermore, it was less explosive than Davis and Putnam's algorithm
+since it introduced unification as an inference filter.
+
+Since its introduction, resolution was improved in many ways. Strategies and
+heuristics~\cite{lw-65-sos} to curb the explosion of the search space were
+introduced, as well as complete, but less explosive variants of the calculus \cite{cc-73-resolution-book}. Despite this
+progress, reasoning about equality of objects was still hard for these methods. This changed with introduction of
+superposition in 199
 

### 2022-02-14 introduction/introduction.tex
more

@@
 understood him. Not only does using a rigorous mathematical language avoid the
-ambiguity of natural language, it allows us to use the formal rules of reasoning to
-make conclusions from initial arguments in a trustworthy manner.
+ambiguity of natural language, it allows us to use the formal rules of reasoning
+to make conclusions from initial arguments in a trustworthy manner.
 
@@
 it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
-that there is a system of inference rules (calculus) such that for any valid statement a
-proof of validity can be constructed in this system. Around the same time, the
-first computers were leaving the imagination of the engineers and began to
-automate many complicated processes.  Naturally, the question as to  whether a
-computer can decide if a first-order statement is valid (i.e., a theorem) arose.
-Church \cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered
-this question negatively in 1936.
+that there is a system of inference rules (calculus) such that for any valid
+statement a proof of validity can be constructed in this system. Around the same
+time, the first computers were leaving the imagination of the engineers and
+began to automate many complicated processes.  Naturally, the question as to
+whether a computer can decide if a first-order statement is valid (i.e., a
+theorem) arose. Church \cite{ac-36-fol-undecidable} and Turing
+\cite{tm-37-undecidable} answered this question negatively in 1936.
 
@@
 heuristics~\cite{lw-65-sos} to curb the explosion of the search space were
-introduced, as well as complete, but less explosive variants of the calculus \cite{cc-73-resolution-book}. Despite this
-progress, reasoning about equality of objects was still hard for these methods. This changed with introduction of
-superposition in 199
+introduced, as well as complete, but less explosive variants of the calculus
+\cite{cc-73-resolution-book}. Despite this progress, reasoning about equality of
+objects was still hard for these methods. This changed with introduction of
+superposition \cite{bg-94-superposition} in early 1990s.
+
+Superposition is a complete calculus for first-order logic loosely based on
+resolution, featuring support for equality on the calculus level. This support
+is further optimized by use of term orders and selection functions to prune the
+search space. Thanks to efficient implementation, this calculus was used to
+prove long-standing open problem known as ``Robbins conjecture''
+\cite{mccune-97-robbins}. Furthermore, provers based on this calculus have been
+winning the first-order division of CASC theorem proving competition since its
+inception in the 1990s \cite{ss-96-casc}.
+
+Even though the successes such proving Robbins conjecture show that it is
+possible, automatically proving hard mathematical problems is still out of reach
+for theorem provers. However, computers can still be used to verify human-written proofs.
 

### 2022-02-14 introduction/introduction.tex
more

@@
 Despite this negative result, the prospect of automatically proving theorems of
-first-order logic remained too enticing. In 1960 Davis and Putnam
+first-order logic remained too enticing. In 1960, Davis and Putnam
 \cite{dp-1960-dpll} described an algorithm for checking validity of a
-first-order formula. As first-order logic is undecidable this algorithm
+first-order formula. As first-order logic is undecidable, this algorithm
 terminated only on valid formulas. Even though it was efficient enough to prove
@@
 objects was still hard for these methods. This changed with introduction of
-superposition \cite{bg-94-superposition} in early 1990s.
+superposition~\cite{bg-94-superposition} in the early 1990s.
 
@@
 prove long-standing open problem known as ``Robbins conjecture''
-\cite{mccune-97-robbins}. Furthermore, provers based on this calculus have been
+\cite{mccune-97-robbins}. Provers based on this calculus have been
 winning the first-order division of CASC theorem proving competition since its
@@
 
-Even though the successes such proving Robbins conjecture show that it is
+Even though the successes of proving Robbins conjecture showed that it is
 possible, automatically proving hard mathematical problems is still out of reach
-for theorem provers. However, computers can still be used to verify human-written proofs.
+for theorem provers. However, computers can still be of help to construct formal
+proofs.
+
+Proof assistants (or interactive theorem provers) are programs which allow users
+to describe a theory in a formal way and to prove the statements about this
+theory using rules of the inference system the proof assistant is based on.
+These programs can then check if the given proof is correct. As their core is usually
+small and well-tested, proofs that pass the checking phase are considered trustworthy.
 

### 2022-02-14 introduction/introduction.tex
more

@@
 Proof assistants (or interactive theorem provers) are programs which allow users
-to describe a theory in a formal way and to prove the statements about this
-theory using rules of the inference system the proof assistant is based on.
-These programs can then check if the given proof is correct. As their core is usually
-small and well-tested, proofs that pass the checking phase are considered trustworthy.
+to describe a theory formally and to prove the statements about this theory
+using rules of the inference system the proof assistant is based on. These
+programs can then check if the given proof is correct. As their core is usually
+small and well-tested, proofs that pass the checking phase are considered
+trustworthy.
+
+They proved indispensible in proving very complex mathematical theorems in a
+trustworthy manner. For example, four-color theorem \cite{gg-2008-four-color}
+and Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
+ending the need to trust hundreds of pages of hard mathematical proofs. Proof
+assistants are also used to show correctness of complex software such as
+compilers \cite{xl-09-compcert} or operating systems kernels \cite{gk-09-sel4}.
+This verified software can be used in safety-critical scenarios such as nuclear
+plants or autonomous vehicles.
+
+The proofs need to be spelled out in minute details to be accepted by a proof assistant.
+Even though many proof assistants offer some sort of automation, seemingly simple
+statements cannot be automatically proved. This is one of the reasons why large
+verification projects can take years to finish.
+
+Integrating proof automation right into the proof assistant is not an easy task.
+Most proof assistants are based on variants of higher-order logic, which is more
+expressive than first-order logic, but in general does not even allow complete
+proof calculi. To provide some level of automation, proof assistant developers
+use the efficient first-order provers as follows. First, the proof goal is
+translated from higher-order to first-order logic. Then the first-order prover
+is run on the translated problem. If the first-order prover finds the proof, the
+proof is reconstructed in the proof assistant. This approach is implemented in
+modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
+and Isabelle \cite{lc-88-isabelle} using \emph{hammers} CoqHammer \cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer}
+and Sledgehammer \cite{bn-10-sh}, respectively.
 

### 2022-02-14 introduction/introduction.tex
more

@@
 Proof assistants (or interactive theorem provers) are programs which allow users
-to describe a theory formally and to prove the statements about this theory
-using rules of the inference system the proof assistant is based on. These
+to describe a theory formally and to write proofs of the statements about this theory
+using inference system rules of the proof assistant. These
 programs can then check if the given proof is correct. As their core is usually
@@
 
-They proved indispensible in proving very complex mathematical theorems in a
+Proof assistants showed indispensible in proving very complex mathematical theorems in a
 trustworthy manner. For example, four-color theorem \cite{gg-2008-four-color}
@@
 modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
-and Isabelle \cite{lc-88-isabelle} using \emph{hammers} CoqHammer \cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer}
-and Sledgehammer \cite{bn-10-sh}, respectively.
+and Isabelle \cite{lc-88-isabelle} using \emph{hammers} CoqHammer
+\cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
+\cite{bn-10-sh}, respectively.
+
+This approach proved useful: around 69\% of the simple proof goals can be
+discharged automatically using Sledgehammer \cite{kb-13-mash}. However, there is
+clearly some untapped potential in the approach as it communicates with
+automatic theorem provers through a translation. To bridge this translation gap,
+attempts were made to use higher-order automatic theorem provers instead of
+first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
+higher-order of the time theorem provers were primarily designed for small and tricky
+problems. On the other hand, problems coming from hammers consist of mostly first-order formulas, are rather large,
+and mostly require simple higher-order proof steps.
+
+Thus, from a standpoint of proof assistant ideal automatic theorem prover fulfills
+the following wishlist:
+\begin{itemize}
+  \item Performs as best first-order provers
+  \item Supports higher-order logic
+  \item Works well on large problems
+\end{itemize}
+This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}
 

### 2022-02-15 introduction/introduction.tex
separation?

@@
 
-The initial study of this formalism and the rules to reason about statements in
+\paragraph{Automatic Theorem Proving}The initial study of this formalism and the rules to reason about statements in
 it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
@@
 
-Proof assistants (or interactive theorem provers) are programs which allow users
+\paragraph{Proof Assistants and Hammers} Proof assistants (or interactive theorem provers) are programs which allow users
 to describe a theory formally and to write proofs of the statements about this theory

### 2022-02-15 introduction/introduction.tex
more

--- 2022-02-14 introduction/introduction.tex
@@
 
-The initial study of this formalism and the rules to reason about statements in
+\paragraph{Automatic Theorem Proving}The initial study of first-order logic and the rules to reason about statements in
 it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
@@
 time, the first computers were leaving the imagination of the engineers and
-began to automate many complicated processes.  Naturally, the question as to
+began automating many complicated processes.  Naturally, the question as to
 whether a computer can decide if a first-order statement is valid (i.e., a
@@
 possible, automatically proving hard mathematical problems is still out of reach
-for theorem provers. However, computers can still be of help to construct formal
+for theorem provers. However, there are other ways in which humans and computers can work together to construct formal
 proofs.
 
-Proof assistants (or interactive theorem provers) are programs which allow users
-to describe a theory formally and to write proofs of the statements about this theory
-using inference system rules of the proof assistant. These
-programs can then check if the given proof is correct. As their core is usually
-small and well-tested, proofs that pass the checking phase are considered
-trustworthy.
+\paragraph{Proof Assistants and Hammers} Proof assistants (or interactive
+theorem provers) are programs which allow users to describe a theory formally
+and to write proofs of the statements about this theory using inference 
+rules of the proof assistant. These programs can then check if the given proof
+is correct. As their core is usually small and well-tested, proofs that pass the
+checking phase are considered trustworthy.
 
@@
 first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
-higher-order of the time theorem provers were primarily designed for small and tricky
-problems. On the other hand, problems coming from hammers consist of mostly first-order formulas, are rather large,
-and mostly require simple higher-order proof steps.
+higher-order theorem provers of the time were primarily designed for small and
+tricky problems. On the other hand, problems coming from hammers consist of
+mostly first-order formulas, are rather large, and mostly require simple
+higher-order proof steps. Thus, from a standpoint of proof assistant ideal
+automatic theorem prover fulfills the following wishlist:
+\begin{enumerate}
+  \item\label{it:fo} Performs as best first-order provers on first-order problems
+  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
+  \item\label{it:size} Scales well with the size of the problem
+\end{enumerate}
+This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}
 
-Thus, from a standpoint of proof assistant ideal automatic theorem prover fulfills
-the following wishlist:
-\begin{itemize}
-  \item Performs as best first-order provers
-  \item Supports higher-order logic
-  \item Works well on large problems
-\end{itemize}
-This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}
+\paragraph{Our Approach} To improve the automatic reasoning for higher-order
+logic we decided not to start from a blank slate and build a new theorem prover.
+Rather, we start from a position of strength --- from state-of-the-art
+first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
+prover to higher-order logic. In this way, the points \ref{it:fo} and
+\ref{it:size} from the above wishlist  is fulfilled for free, as E is very
+efficient and deals well with large problems.
+
+We do not extend E to higher-order logic in one atomic step. Instead, we extend
+it by adding support for three distinguished features of higher-order logic,
+which gradually increase its support for higher-order logic. With the layered
+approach we make sure that after every extension to higher-order logic, E's
+efficiency on first-order problems was not affect. In other words, while trying
+to fulfill the point \ref{it:ho} from the wishlist we make sure that points
+\ref{it:fo} and \ref{it:size} are not affected.
+
+We have split the road to full higher-order logic in three stages. A
+distinguished feature of higher-order logic is that it not only supports
+quantification over objects, but also over functions that manipulate these
+objects. Thus, higher-order logic allows us to make a statement ``all functions
+with arguments $0$ and $1$ result in $2$" ($\forall x. x (\cst{0}, \cst{1}) \eq
+\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
+disprove this statement. The first stage is to extend E to support
+quantification over functions that are already present in the theory.
+
+Higher-order logic also supports expressing functions that are not explicitly
+stated in the problem. For example, suppose we want to prove that there is a
+function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
+x(\cst{2}) \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such
+function is the function that adds 2 to its argument. In higher-order logic such
+unnamed function can be represented using $\lambda$ abstractions as $\lambda
+x.\, x \, \cst{+} \, 2$. Replacing $x$ with this term suffices to prove the
+original statement. The second stage is to support automatically synthesizing
+such functions.
+
+In higher-order logic we can make statements about statements. For example, we
+can naturally represent statements such as ``whatever Jasmin says is true'' as
+$\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
+feature of first-order logic.
 

### 2022-02-15 introduction/introduction.tex
more

@@
 philosophy lectures in my high school I heard my professor exclaim with
-excitement: “Mathematical truth is the highest and absolute form of truth!”.
+excitement: “Mathematical truth is the highest form of truth!”.
 Unbothered and uninterested, I continued playing Fruit Ninja.
@@
 
-The proofs need to be spelled out in minute details to be accepted by a proof assistant.
+The proofs need to be spelled out in minute detail to be accepted by a proof assistant.
 Even though many proof assistants offer some sort of automation, seemingly simple
@@
 
-This approach proved useful: around 69\% of the simple proof goals can be
-discharged automatically using Sledgehammer \cite{kb-13-mash}. However, there is
+This approach proved useful: around 76\% of the simple proof goals can be
+discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
 clearly some untapped potential in the approach as it communicates with
@@
 prover to higher-order logic. In this way, the points \ref{it:fo} and
-\ref{it:size} from the above wishlist  is fulfilled for free, as E is very
+\ref{it:size} from the above wishlist are fulfilled for free, as E is very
 efficient and deals well with large problems.
@@
 approach we make sure that after every extension to higher-order logic, E's
-efficiency on first-order problems was not affect. In other words, while trying
+efficiency on first-order problems did not decrease. In other words, while trying
 to fulfill the point \ref{it:ho} from the wishlist we make sure that points
@@
 objects. Thus, higher-order logic allows us to make a statement ``all functions
-with arguments $0$ and $1$ result in $2$" ($\forall x. x (\cst{0}, \cst{1}) \eq
+with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
 \cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
@@
 Higher-order logic also supports expressing functions that are not explicitly
-stated in the problem. For example, suppose we want to prove that there is a
+present in the theory. For example, suppose we want to prove that there is a
 function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
@@
 In higher-order logic we can make statements about statements. For example, we
-can naturally represent statements such as ``whatever Jasmin says is true'' as
+can naturally represent statements such as ``what Jasmin says is true'' as
 $\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
 feature of first-order logic.
+
+\section{Contribution}
+\section{Publications and Thesis Structure}
+\section{Related Work}
 

### 2022-02-15 introduction/introduction.tex
more

@@
 
-In higher-order logic we can make statements about statements. For example, we
-can naturally represent statements such as ``what Jasmin says is true'' as
-$\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
-feature of first-order logic.
+In higher-order logic we can make statements about statements. More precisely,
+we it treats Boolean terms (formulas) as first-class citizens, enabling us to
+model some concepts in more natural manner. For example, we can naturally
+represent statements such as ``what Jasmin says is true'' as $\forall x. \,
+\cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support Boolean terms
+as first-class citizens.
 
 \section{Contribution}
+
+The biggest contribution of this thesis is the extension of two theorem provers,
+E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
+extend those provers, we faced many challenges. Some of them were of engineering
+nature, some of them were algorithmic, while others concerned fine-tuning the
+heuristics to achieve best possible performance. We discuss those challenges in more detail:
+\begin{itemize}
+  \item We implemented the complete superposition calculus for $\lambda$-free
+  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
+  We had to modify internal term structure of Ehoh to support $\lambda$-free
+  higher-order terms, implement new unification and matching algorithms, extend
+  indexing data structures to work with higher-order terms and to improve the
+  performance of heuristics on higher-order problems.
+  \item Together with Alexander Bentkamp and other colleagues we developed
+  complete superposition calculus for the second stage described above
+  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
+  decisions that had to be evaluated in short amount of time we decided to
+  implement the calculus in Zipperposition. This prover is less efficient than
+  E, but it is implemented in OCaml and it was designed to make prototyping
+  ideas and experimenting with calculi much easier. In 2019 we entered the
+  higher-order division of CASC theorem proving competition with Zipperposition
+  \cite{gs-19-casc27}. It took third place, closely behind Leo-III (1 percent
+  point) \cite{sb-21-leo3} and far behind (12 percent points) Satallax
+  \cite{cb-12-satallax}.
+  \item Through this experience we identified one of the bottlenecks of
+  Zipperposition: its unification algorithm. To remove this bottleneck we
+  developed a complete higher-order unification algorithm that was more
+  efficient and less redundant than current state of the art. 
+  \item We also noticed that Zipperposition's support for first-class Booleans
+  was lacking compared to competition. We have closely studied the problems on
+  which Zipperposition fails, but are easily solved by other provers and created
+  a set of incomplete rules that enhance Zipperposition's Boolean support.
+  Zipperposition 2 implemented these rules together with the new unification
+  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
+  of CASC. This time it was 20 percent points better than the second best
+  prover, Satallax.
+\end{itemize}
+
 \section{Publications and Thesis Structure}

### 2022-02-15 introduction/introduction.tex
contrib

@@
   prover, Satallax.
+  \item Inspired by this success we implemented the complete superposition
+  calculus for full higher-order logic that we developed by Bentkamp and other
+  colleagues. The development of complete calculus helped us understand in which
+  ways our pragmatic extension was too weak to prove hard theorems of
+  higher-order logic. We further tuned the heuristics and implemented new ways
+  to tame the search space explosion. In 2021 we released Zipperposition 2.1
+  which implemented the new calculus and the described tweaks. It again won
+  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
+  runner-up, Vampire 4.5.
+
+  \item Taking inspiration from our approach of lifting efficient provers from
+  first-order to higher-order logic, we went a step back in an effort to improve
+  first-order provers. In particular, we looked at the most efficient approaches
+  to simplify problems in propositional logic, a logic that disallows quantification,
+  but it is decidable. We lifted some of these approaches to first-order logic
+  and implemented them in Zipperposition.
+
 \end{itemize}

### 2022-02-15 introduction/introduction.tex
more

@@
 
-\section{Contribution}
+\section{Contributions}
 
@@
   runner-up, Vampire 4.5.
+
+  \item Taking all the experience we gathered while working on Zipperposition,
+  we further extended Ehoh to full higher-order logic. Here again modified term
+  structure, implemented new unification and matching algorithms, extended indexing data
+  structures, and tweaked heuristics. Even though this extension has larger scope than 
+  the original extension from E to Ehoh, strong basis that we have earlier built
+  made it manageable.
 

### 2022-02-15 introduction/introduction.tex
more

@@
   higher-order terms, implement new unification and matching algorithms, extend
-  indexing data structures to work with higher-order terms and to improve the
+  indexing data structures to work with higher-order terms, and to improve the
   performance of heuristics on higher-order problems.
   \item Together with Alexander Bentkamp and other colleagues we developed
-  complete superposition calculus for the second stage described above
+  a complete superposition calculus for the second stage described above
   \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
@@
   algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
-  of CASC. This time it was 20 percent points better than the second best
+  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
   prover, Satallax.
   \item Inspired by this success we implemented the complete superposition
-  calculus for full higher-order logic that we developed by Bentkamp and other
+  calculus for full higher-order logic (i.e., the third stage) that we developed with Bentkamp and other
   colleagues. The development of complete calculus helped us understand in which
@@
   CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
-  runner-up, Vampire 4.5.
+  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.
 
   \item Taking all the experience we gathered while working on Zipperposition,
-  we further extended Ehoh to full higher-order logic. Here again modified term
+  we further extended Ehoh to full higher-order logic. We again modified term
   structure, implemented new unification and matching algorithms, extended indexing data
@@
 
-  \item Taking inspiration from our approach of lifting efficient provers from
+  \item Following our approach of lifting efficient provers from
   first-order to higher-order logic, we went a step back in an effort to improve

### 2022-02-15 conclusion/conclusion.tex
toc

\chapter{Conclusion}
\setheader{Conclusion}
\label{ch:conclusion}

### 2022-02-15 ehoh/ehoh.tex
toc

\chapter{Extending a Brainiac Prover to Lambda-Free Higher-Order Logic}
\setheader{Extending a Brainiac Prover to Lambda-Free Higher-Order Logic}
\label{ch:ehoh}

### 2022-02-15 ehoh2/ehoh2.tex
toc

\chapter{Extending a High-Performance Prover to Higher-Order Logic}
\setheader{Extending a High-Performance Prover to Higher-Order Logic}
\label{ch:ehoh2}

### 2022-02-15 ho-techniques/ho-techniques.tex
toc

\chapter{Making Higher-Order Superposition Work}
\setheader{Making Higher-Order Superposition Work}
\label{ch:ho-techniques}

### 2022-02-15 introduction/introduction.tex
toc

@@
 \chapter{Introduction}
-\label{introduction}
+\setheader{Introduction}
+\label{ch:intro}
 
@@
 \end{enumerate}
-This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}
+This brings us to the topic of this thesis: \emph{How to develop a prover fulfilling this wishlist?}
 
@@
 as first-class citizens.
+\pagebreak[2]
 
@@
   and implemented them in Zipperposition. 
-
 \end{itemize}
@@
 \section{Publications and Thesis Structure}
+
+Starting from Chapter ????, every chapter of this thesis corresponds to a work that
+was previously published. Work is presented in this thesis with the approval of coauthors.
+
 \section{Related Work}
-
-% \begin{abstract}
-% Sample Abstract.
-% \end{abstract}
-
-% \blfootnote{This chapter is partly based on \faFileTextO~\emph{M. Beller. Toward an
-%     Empirical Theory of Feedback-Driven Development, ICSE'18 (Student Research Competition)}~\cite{BellerSRC2018}.
-% }
-
-% \newpage
-
-% \dropcap{T}his is a introductory page.
-
-% \section{Background \& Context}
-% In this thesis, you can reference pictures~\Cref{fig:devmodel} using Cleverref and circles \circled{5}.
-
-% \begin{figure}[htb]
-% 	\centering
-% 	\includegraphics[width=0.65\columnwidth]{development_model_without_papers}
-% 	\caption{The stages of the FDD model and their relationship to other
-%           Software Engineering concepts.}
-% 	\label{fig:devmodel}
-% \end{figure}
-
-% We also have lists:
-
-% \begin{enumerate}
-%   \item Static Analysis~\circled{3} examines program artifacts or
-%     their source code without executing them~\cite{wichmann1995industrial}, while
-%  \item Dynamic Analysis~\circled{4} relies on information gathered from their
-%    execution~\cite{cornelissen2009systematic}.
-% \end{enumerate}
-
-% Or boxes:
-
-% \begin{framed}
-% This thesis is concerned with the empirical assessment of the state of the art of how developers
-% drive software development with the help of feedback loops.
-% \end{framed}
-
-% Or code:
-% \begin{lstlisting}[caption={\textsc{TrinityCore}},label={lst:e1}]
-%  x += other.x;
-%  y += other.y;
-%  z += other.y;
-% \end{lstlisting}
-
-% I hope this helps you get started!
-% Moritz

### 2022-02-15 pre/pre.tex
toc

\chapter{Preliminaries}
\setheader{Preliminaries}
\label{ch:preliminatires}

### 2022-02-15 satfol/satfol.tex
toc

\chapter{SAT-Inspired Eliminations for Superposition}
\setheader{SAT-Inspired Eliminations for Superposition}
\label{ch:satfol}

### 2022-02-15 unif/unif.tex
toc

\chapter{Efficient Full Higher-Order Unification}
\setheader{Efficient Full Higher-Order Unification}
\label{ch:unif}

### 2022-02-15 introduction/introduction.tex
more

@@
 
-\section{Publications and Thesis Structure}
+\section{Thesis Structure and Publications}
 
-Starting from Chapter ????, every chapter of this thesis corresponds to a work that
-was previously published. Work is presented in this thesis with the approval of coauthors.
+This thesis is cumulative: chapters~\ref{ch:ehoh}~-~\ref{ch:satfol}  are
+directly taken from previous publications (or drafts in case of
+Chapter~\ref{ch:ehoh2}) where I was the main author. Use
+of publications in this thesis was authorized by the coauthors.
+
+Chapter~\ref{ch:pre} introduces the background material necessary for following this thesis,
+while Chapter~\ref{ch:conclusion} gives a brief summary of the work and describes alleys for future work.
+Other six chapters discuss previously described contributions in detail:
+
+\vspace{0.5em}
+\noindent\textbf{Chapter~\ref{ch:ehoh}} discusses the steps we have taken to extend first-order
+  prover E to $\lambda$-free higher-order logic, obtaining Ehoh. It is mostly based on the journal publication
+  \begin{quote}
+    Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
+    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
+    In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021.  
+  \end{quote}
+\vspace{-0.3em}
+while parts of the chapter were previously published in the conference publication
+\vspace{-0.3em}
+  \begin{quote}
+    Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
+    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
+    In  In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
+  \end{quote}
+\noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification algorithm we designed to be used
+inside efficient theorem provers. It is mostly based on the journal publication
+  \begin{quote}
+    Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
+    In \emph{Logical Methods in Computer Science} 17(4): 18:1--18:31, 2021.
+  \end{quote}
+\vspace{-0.3em}
+while parts of the chapter were previously published in the conference
+publication, which won the best paper by junior researcher award:
+\vspace{-0.3em}
+\begin{quote}
+  Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
+  In Ariola, Z.M. (ed.) \emph{FSCD 2020} LIPIcs 167, pp. 5:1--5:17, Schloss Dagstuhl--Leibniz-Zentrum für Informatik, 2020. 
+\end{quote}
+\noindent\textbf{Chapter~\ref{ch:bools}} discusses the pragmatic extensions of
+the complete calculus for superposition with lambdas to support first-class
+Booleans. It is based on the publication
+    \begin{quote}
+      Petar Vukmirović and Visa Nummelin. Boolean Reasoning in a Higher-Order
+      Superposition Prover. In P. Fontaine, K. Korovin, I.S. Kotsireas, P.
+      R\"ummer, S. Tourret (eds.) \emph{PAAR-2020}, CEUR Workshop Proceedings, vol.
+      2752, pp. 148--166. CEUR-WS.org (2020)
+    \end{quote}
+\noindent\textbf{Chapter~\ref{ch:ho-techniques}} discusses the approach we took to
+control explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
+  \begin{quote}
+    Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
+    Making Higher-Order Superposition Work. In \emph{Journal of Automated Reasoning}. Springer, 2022.
+  \end{quote}
+\vspace{-0.3em}
+  while parts of the chapter were previously published in the conference publication, which won the best paper by a student award:
+\vspace{-0.3em}
+\begin{quote}
+  Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
+  Making Higher-Order Superposition Work. In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28},
+  LNCS 12699, pp. 415--432, Springer, 2021.
+\end{quote}
+\noindent\textbf{Chapter~\ref{ch:ehoh2}} discusses the extension of Ehoh to
+support full higher-order logic. We call the obtained prover \newehoh.
+It is based on the draft of the joint work with Jasmin Blanchette and
+Stephan Schulz, planned for submission at TACAS 2022.
+
+\vspace{0.5em}
+\noindent\textbf{Chapter~\ref{ch:satfol}} discusses lifting some of the techniques used to
+simplify formulas in propositional logic to first-order logic. It is based on the extended, report version of the publication:
+  \begin{quote}
+    Petar Vukmirović, Jasmin Blanchette, and Marijn J.H. Heule. 
+    SAT-Inspired Eliminations for Superposition. In Piskac, R., Whalen, M. (eds.) \emph{FMCAD 2021}, pp. 231--240, IEEE, 2021. 
+  \end{quote}
+
+As a PhD candidate, I coauthored the following articles and papers, which are not included in this thesis:
+
+\begin{itemize}
+  \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
+  Draft paper, submitted to ITP 2022.
+  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
+  Draft article, submitted to Journal of automated reasoning.
+  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
+  \emph{Journal of Automated Reasoning} 65(7): 893--940, 2021. 
+  \item Visa Nummelin, Alexander Bentkamp, Sophie Tourret, and Petar Vukmirović. Superposition with First-Class Booleans and Inprocessing Clausification.
+  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 378--395, Springer, 2021.
+  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Full Higher-Order Logic.
+  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 396--412, Springer, 2021.
+  \item Stephan Schulz, Simon Cruanes, and Petar Vukmirović. Faster, Higher, Stronger: E 2.3.
+  In Fontaine, P. (ed.) \emph{CADE-27},  LNCS 11716, pp. 495--507, Springer, 2019.
+  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
+  In Fontaine, P. (ed.) \emph{CADE-27}, LNCS 11716, pp. 55--73, Springer, 2019. 
+\end{itemize}
 

### 2022-02-15 bools/bools.tex
more

\chapter{Boolean Reasoning in a Higher-Order Superposition Prover}
\setheader{Boolean Reasoning in a Higher-Order Superposition Prover}
\label{ch:bools}

### 2022-02-16 introduction/introduction.tex
more

@@
 understood him. Not only does using a rigorous mathematical language avoid the
-ambiguity of natural language, it allows us to use the formal rules of reasoning
+ambiguity of the natural one, it allows us to use the formal rules of reasoning
 to make conclusions from initial arguments in a trustworthy manner.
@@
 first-order logic remained too enticing. In 1960, Davis and Putnam
-\cite{dp-1960-dpll} described an algorithm for checking validity of a
-first-order formula. As first-order logic is undecidable, this algorithm
+\cite{dp-1960-dpll} described a procedure for checking validity of a
+first-order formula. As first-order logic is undecidable, this procedure
 terminated only on valid formulas. Even though it was efficient enough to prove
@@
 Most proof assistants are based on variants of higher-order logic, which is more
-expressive than first-order logic, but in general does not even allow complete
+expressive than first-order logic, but harder to automate as in general it does not even allow complete
 proof calculi. To provide some level of automation, proof assistant developers
@@
 
-This approach proved useful: around 76\% of the simple proof goals can be
+The approach proved useful: around 76\% of the simple proof goals can be
 discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
@@
   \item\label{it:fo} Performs as best first-order provers on first-order problems
+  \item\label{it:size} Scales well with the size of the problem
   \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
-  \item\label{it:size} Scales well with the size of the problem
 \end{enumerate}
@@
 prover to higher-order logic. In this way, the points \ref{it:fo} and
-\ref{it:size} from the above wishlist are fulfilled for free, as E is very
+\ref{it:size} from the above wishlist are already fulfilled, as E is very
 efficient and deals well with large problems.
@@
 We do not extend E to higher-order logic in one atomic step. Instead, we extend
-it by adding support for three distinguished features of higher-order logic,
-which gradually increase its support for higher-order logic. With the layered
-approach we make sure that after every extension to higher-order logic, E's
-efficiency on first-order problems did not decrease. In other words, while trying
+it by adding support for features of higher-order logic one by one,
+which gradually increase its reasoning capabilities. With this gradual
+approach we make sure that after every extension, E's
+efficiency on first-order problems did not decrease. Thus, while trying
 to fulfill the point \ref{it:ho} from the wishlist we make sure that points
@@
 present in the theory. For example, suppose we want to prove that there is a
-function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
-x(\cst{2}) \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such
-function is the function that adds 2 to its argument. In higher-order logic such
-unnamed function can be represented using $\lambda$ abstractions as $\lambda
-x.\, x \, \cst{+} \, 2$. Replacing $x$ with this term suffices to prove the
-original statement. The second stage is to support automatically synthesizing
-such functions.
+function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
+\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
+the function that adds 2 to its argument. In higher-order logic such unnamed
+function can be represented using $\lambda$ abstractions as $\lambda y.\, y \,
+\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
+second stage is to support automatically synthesizing such functions.
 
 In higher-order logic we can make statements about statements. More precisely,
-we it treats Boolean terms (formulas) as first-class citizens, enabling us to
+it treats Boolean terms (formulas) as first-class citizens, enabling us to
 model some concepts in more natural manner. For example, we can naturally

### 2022-02-16 introduction/open_vs_merged_prs.r

@@ -1,35 +1,0 @@
-library(readr)
-library(ggplot2)
-library(ggthemes)
-library(scales)
-
-### ggplot configuration
-ggplot.defaults <- 
-  theme_few() +
-  theme(axis.text.x = element_text(size = 22)) + 
-  theme(axis.text.y = element_text(size = 22)) +
-  theme(axis.title.x = element_text(size = 22, vjust = 1)) +
-  theme(axis.title.y = element_text(size = 22, vjust = 1)) +
-  theme(legend.text = element_text(size = 22))  +
-  theme(legend.title = element_text(size = 22)) + 
-  theme(axis.title.x=element_text(margin=margin(20,0,0,0))) + 
-  theme(axis.title.y=element_text(margin=margin(0,25,0,0)))
-
-ggplot.small.defaults <-
-  ggplot.defaults + theme(axis.title.x = element_text(size = 22, vjust = -0.5)) +
-  theme(axis.title.y = element_text(size = 22, hjust = 0.5, vjust = 1.5), panel.grid = element_line(colour = "grey", size=0.3))
-
-gh_history_prs <- read_csv("~/projects/phd-thesis/dissertation-tudelft-latex/introduction/results-20171204-171053.csv")
-gh_history_prs$aggregated_date <- as.Date(paste0(gh_history_prs$pr_yr, '-', gh_history_prs$pr_mt, '-1'))
-gh_history_prs <- gh_history_prs[gh_history_prs$action %in% c("opened","merged"),]
-gh_history_prs <- gh_history_prs[gh_history_prs$aggregated_date < as.Date("2014-02-01"),]
-
-ggplot(gh_history_prs, aes(aggregated_date, freq, shape=action, group=action, colour=action)) + geom_point(size=2.5) + geom_line() + geom_smooth(se = FALSE) +
-  scale_y_continuous(labels = scales::comma) +
-  ggplot.small.defaults +
-  ylab("#Events") + xlab("") + labs(shape='Pullrequest Events',colour='Pullrequest Events') + theme(legend.position = "bottom")
-
-total <- aggregate(freq ~ action, gh_history_prs, sum)
-# Number of openend and not closed pull requests
-total[2,]$freq - total[1,]$freq
-

### 2022-02-16 introduction/introduction.tex
more

@@
   higher-order division of CASC theorem proving competition with Zipperposition
-  \cite{gs-19-casc27}. It took third place, closely behind Leo-III (1 percent
-  point) \cite{sb-21-leo3} and far behind (12 percent points) Satallax
-  \cite{cb-12-satallax}.
+  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
+  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
   \item Through this experience we identified one of the bottlenecks of
@@
   \item Inspired by this success we implemented the complete superposition
-  calculus for full higher-order logic (i.e., the third stage) that we developed with Bentkamp and other
-  colleagues. The development of complete calculus helped us understand in which
+  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
+  colleagues. Developing this calculus helped us understand in which
   ways our pragmatic extension was too weak to prove hard theorems of
@@
   \item Taking all the experience we gathered while working on Zipperposition,
-  we further extended Ehoh to full higher-order logic. We again modified term
-  structure, implemented new unification and matching algorithms, extended indexing data
-  structures, and tweaked heuristics. Even though this extension has larger scope than 
-  the original extension from E to Ehoh, strong basis that we have earlier built
-  made it manageable.
+  we further extended Ehoh to full higher-order logic, obtaining prover we call
+  \newehoh{}. We again modified term structure, implemented new unification and
+  matching algorithms, extended indexing data structures, and tweaked
+  heuristics. Even though this extension has larger scope than the original
+  extension from E to Ehoh, strong basis that we have earlier built made it
+  manageable.
 
@@
   first-order provers. In particular, we looked at the most efficient approaches
-  to simplify problems in propositional logic, a logic that disallows quantification,
+  to simplify problems in propositional logic, a simpler logic that disallows quantification,
   but it is decidable. We lifted some of these approaches to first-order logic
@@
   \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
-  Draft paper, submitted to ITP 2022.
+  Draft paper.
   \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
-  Draft article, submitted to Journal of automated reasoning.
+  Draft article.
   \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.

### 2022-02-16 introduction/introduction.tex
more

@@
 \section{Related Work}
+
+Earliest attempts to automate higher-order logic can be traced back to Robinson
+[ehoh 23]. His approach was based on translating higher-order logic to
+first-order logic using combinators, similar to the approach taken by hammers.
+Huet [ehoh 16] and Andrews [ehoh 2] designed calculi that are directly applied
+on higher-order formulas.
+
+These attempts resulted in early higher-order theorem provers. Andrews and
+collegaues developed TPS: a semi-automatic prover which allows user to specify
+proof outlines, based on expansion proofs. Benzm\"uller and colleageues
+developed LEO, a prover based on higher-order resolution which introduced
+cooperative paradigm. Provers of this paradigm invoke first-order back-ends in
+regular intervals to finish the proof. Latest iteration in LEO family of provers
+is Leo-III.

### 2022-02-16 introduction/introduction.tex
related

--- 2022-02-16 introduction/introduction.tex
@@
   prover E to $\lambda$-free higher-order logic, obtaining Ehoh. It is mostly based on the journal publication
-  \begin{quote}
-    Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
+  \begin{enumerate}
+    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
     Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
     In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021.  
-  \end{quote}
+  \end{enumerate}
 \vspace{-0.3em}
@@
 \vspace{-0.3em}
-  \begin{quote}
-    Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
+  \begin{enumerate}[resume]
+    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
     Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
     In  In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
-  \end{quote}
+  \end{enumerate}
 \noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification algorithm we designed to be used
 inside efficient theorem provers. It is mostly based on the journal publication
-  \begin{quote}
-    Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
+  \begin{enumerate}[resume]
+    \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
     In \emph{Logical Methods in Computer Science} 17(4): 18:1--18:31, 2021.
-  \end{quote}
+  \end{enumerate}
 \vspace{-0.3em}
@@
 \vspace{-0.3em}
-\begin{quote}
-  Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
+\begin{enumerate}[resume]
+  \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
   In Ariola, Z.M. (ed.) \emph{FSCD 2020} LIPIcs 167, pp. 5:1--5:17, Schloss Dagstuhl--Leibniz-Zentrum für Informatik, 2020. 
-\end{quote}
+\end{enumerate}
 \noindent\textbf{Chapter~\ref{ch:bools}} discusses the pragmatic extensions of
@@
 Booleans. It is based on the publication
-    \begin{quote}
-      Petar Vukmirović and Visa Nummelin. Boolean Reasoning in a Higher-Order
+    \begin{enumerate}[resume]
+      \item Petar Vukmirović and Visa Nummelin. Boolean Reasoning in a Higher-Order
       Superposition Prover. In P. Fontaine, K. Korovin, I.S. Kotsireas, P.
@@
       2752, pp. 148--166. CEUR-WS.org (2020)
-    \end{quote}
+    \end{enumerate}
 \noindent\textbf{Chapter~\ref{ch:ho-techniques}} discusses the approach we took to
 control explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
-  \begin{quote}
-    Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
+  \begin{enumerate}[resume]
+    \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
     Making Higher-Order Superposition Work. In \emph{Journal of Automated Reasoning}. Springer, 2022.
-  \end{quote}
+  \end{enumerate}
 \vspace{-0.3em}
@@
 \vspace{-0.3em}
-\begin{quote}
-  Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
+\begin{enumerate}[resume]
+  \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
   Making Higher-Order Superposition Work. In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28},
   LNCS 12699, pp. 415--432, Springer, 2021.
-\end{quote}
+\end{enumerate}
 \noindent\textbf{Chapter~\ref{ch:ehoh2}} discusses the extension of Ehoh to
@@
 simplify formulas in propositional logic to first-order logic. It is based on the extended, report version of the publication:
-  \begin{quote}
-    Petar Vukmirović, Jasmin Blanchette, and Marijn J.H. Heule. 
+  \begin{enumerate}[resume]
+    \item Petar Vukmirović, Jasmin Blanchette, and Marijn J.H. Heule. 
     SAT-Inspired Eliminations for Superposition. In Piskac, R., Whalen, M. (eds.) \emph{FMCAD 2021}, pp. 231--240, IEEE, 2021. 
-  \end{quote}
+  \end{enumerate}
 
@@
 
-\begin{itemize}
+\begin{enumerate}[resume]
   \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
@@
   In Fontaine, P. (ed.) \emph{CADE-27}, LNCS 11716, pp. 55--73, Springer, 2019. 
-\end{itemize}
+\end{enumerate}
 
 \section{Related Work}
+
+Earliest attempts to automate higher-order logic can be traced back to
+Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
+logic to first-order logic using combinators, similar to the approach taken by
+hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
+calculi that are directly applied on higher-order formulas.
+
+These attempts resulted in early higher-order theorem provers. Andrews and
+collegaues developed TPS~\cite{abinpx-96-tps}: a semi-automatic prover which
+allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
+and colleageues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
+resolution which introduced cooperative paradigm. Provers of this paradigm
+invoke first-order back-ends in regular intervals to finish the proof. Latest
+iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
+\cite{sb-21-leo3} is based on higher-order analytic tableu and uses SAT solver
+to close the branches. It won of higher-order division of CASC on 7 ocassions:
+in 2011 and every year from 2013 to 2019 (and once in 2011).
+
+Various reserachers also experimented with the idea of extending a first-order
+prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second
+order unification and added $\lambda$-terms to one of the best superposition
+provers of the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
+\cite{sc-15-simon-phd} with arithmetic, induction. and rudimentary support for
+higher-order logic. Bhayat and Reger implemented a complete superposition
+calculus  for higher-order logic, based on combinators  in Vampire
+\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
+they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
+support higher-order logic \cite{brotb-19-ho-smt}.
+
+Higher-order unification is one of the central procedures in a higher-order
+prover. Jensen and Pietrzykowski introduced a complete algorithm for enumerating
+elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
+some calculi remaing complete if solving a difficult subprobem of higher-order
+unification (`flex-flex'' pairs) is delayed. He described algorithm to solve
+this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
+desgined an algorithm to enumerate higher-order unifiers using higher-order
+combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat
+and Reger implemented efficient term indexing in their extension of Vampire
+\cite{br-20-full-sup-w-combs}. Libal and Steen designed efficient term indexing
+based on substitution trees \cite{ls-16-indexing}.
+
+There have been many studies on the performance of algorithms or success rate of
+heuristics in first-order theorem proving. For example, Hoder and Voronkov
+evaluate different unification algorithms \cite{hv-09-unifalgs}, while Reger and
+Voronkov evaluate different parameters of AVATAR architecture
+\cite{rsv-15-playing-with-avatar}. Other authors evaluated different proof
+search heuristics for superposition provers \cite{gs-20-clausesel,
+hrsv-16-selsel, sm-16-clausesel}. Evaluation of many prover parameters has been
+done for higher-order provers as well
+\cite{benzmueller-et-al-05-can-ho-fo-coop,fb-2016-internal-guidance-satallax,sb-15-beta,wskb-16-effective-norm}.

### 2022-02-16 introduction/introduction.tex
more

@@
 support full higher-order logic. We call the obtained prover \newehoh.
-It is based on the draft of the joint work with Jasmin Blanchette and
-Stephan Schulz, planned for submission at TACAS 2022.
+It is based on the draft 
+\begin{enumerate}[resume]
+  \item Petar Vukmirović, Jasmin Blanchette, Stephan Schulz.  Extending a
+  High-Performance Prover to Higher-Order Logic. 2022.
+\end{enumerate}
 
-\vspace{0.5em}
 \noindent\textbf{Chapter~\ref{ch:satfol}} discusses lifting some of the techniques used to
@@
 
-Earliest attempts to automate higher-order logic can be traced back to
+The e
+arliest attempts to automate higher-order logic can be traced back to
 Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
@@
 hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
-calculi that are directly applied on higher-order formulas.
+calculi that are directly applied to higher-order formulas.
 
 These attempts resulted in early higher-order theorem provers. Andrews and
-collegaues developed TPS~\cite{abinpx-96-tps}: a semi-automatic prover which
+collegaues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
 allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
@@
 iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
-\cite{sb-21-leo3} is based on higher-order analytic tableu and uses SAT solver
+\cite{sb-21-leo3} is based on higher-order analytic tableau and uses SAT solver
 to close the branches. It won of higher-order division of CASC on 7 ocassions:
@@
 Various reserachers also experimented with the idea of extending a first-order
-prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second
-order unification and added $\lambda$-terms to one of the best superposition
+prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
+unification and added $\lambda$-terms to one of the best superposition
 provers of the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
-\cite{sc-15-simon-phd} with arithmetic, induction. and rudimentary support for
+\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
 higher-order logic. Bhayat and Reger implemented a complete superposition
@@
 Higher-order unification is one of the central procedures in a higher-order
-prover. Jensen and Pietrzykowski introduced a complete algorithm for enumerating
+prover. Jensen and Pietrzykowski introduced a complete procedure for enumerating
 elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
-some calculi remaing complete if solving a difficult subprobem of higher-order
-unification (`flex-flex'' pairs) is delayed. He described algorithm to solve
+some calculi remaing complete if solving a difficult subproblem of higher-order
+unification (`flex-flex'' pairs) is delayed. He described procedure to solve
 this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
-desgined an algorithm to enumerate higher-order unifiers using higher-order
+desgined a procedure to enumerate higher-order unifiers using higher-order
 combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat

### 2022-02-16 introduction/introduction.tex
more

@@
   higher-order terms, implement new unification and matching algorithms, extend
-  indexing data structures to work with higher-order terms, and to improve the
+  term indexing data structures to work with higher-order terms, and to improve the
   performance of heuristics on higher-order problems.

### 2022-02-16 introduction/introduction.tex
more

--- 2022-02-16 introduction/introduction.tex
@@
 Higher-order logic also supports expressing functions that are not explicitly
-present in the theory. For example, suppose we want to prove that there is a
+present in the theory. Suppose we want to prove that there is a
 function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
@@
 the function that adds 2 to its argument. In higher-order logic such unnamed
-function can be represented using $\lambda$ abstractions as $\lambda y.\, y \,
+function can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
 \cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
@@
   higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
-  We had to modify internal term structure of Ehoh to support $\lambda$-free
+  We had to modify internal term representation of Ehoh to support $\lambda$-free
   higher-order terms, implement new unification and matching algorithms, extend
-  indexing data structures to work with higher-order terms, and to improve the
+  term indexing data structures to work with higher-order terms, and to improve the
   performance of heuristics on higher-order problems.
@@
   calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
-  colleagues. Developing this calculus helped us understand in which
+  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
   ways our pragmatic extension was too weak to prove hard theorems of
@@
   to tame the search space explosion. In 2021 we released Zipperposition 2.1
-  which implemented the new calculus and the described tweaks. It again won
+  which implemented these new features. It again won
   CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
   runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
-
   \item Taking all the experience we gathered while working on Zipperposition,
-  we further extended Ehoh to full higher-order logic, obtaining prover we call
-  \newehoh{}. We again modified term structure, implemented new unification and
+  we further extended Ehoh to full higher-order logic, obtaining a prover we call
+  \newehoh{}. We again modified the term representation, implemented new unification and
   matching algorithms, extended indexing data structures, and tweaked
@@
   first-order provers. In particular, we looked at the most efficient approaches
-  to simplify problems in propositional logic, a simpler logic that disallows quantification,
-  but it is decidable. We lifted some of these approaches to first-order logic
+  to simplify problems in propositional logic, a simpler, decidable logic that disallows quantification. 
+  We lifted some of these approaches to first-order logic
   and implemented them in Zipperposition. 
@@
 
-The e
-arliest attempts to automate higher-order logic can be traced back to
+The earliest attempts to automate higher-order logic can be traced back to
 Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
@@
 iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
-\cite{sb-21-leo3} is based on higher-order analytic tableau and uses SAT solver
-to close the branches. It won of higher-order division of CASC on 7 ocassions:
-in 2011 and every year from 2013 to 2019 (and once in 2011).
+\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
+to close tableau branches. It won higher-order division of CASC on 7 ocassions:
+in 2011 and every year from 2013 to 2019.
 
@@
 unification and added $\lambda$-terms to one of the best superposition
-provers of the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
+provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
 \cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
@@
 some calculi remaing complete if solving a difficult subproblem of higher-order
-unification (`flex-flex'' pairs) is delayed. He described procedure to solve
+unification (``flex-flex'' pairs) is delayed. He described procedure to solve
 this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty

### 2022-02-17 pre/pre.tex
more

@@
 \setheader{Preliminaries}
-\label{ch:preliminatires}
+\label{ch:pre}
+
+\begin{abstract}
+    In this chapter we lay out the basic prerequisites for the remaining
+    chapters. We begin by defining three logics that we work with in this
+    thesis: propositional logic, monomorphic first-order logic and simply typed
+    higher-order logic. Then, we explain the clausal structure which is the
+    backbone of many calculi for automated provers. We further describe the
+    superposition calculus and finish with description of the process superposition
+    provers use to actually perform the inferences.
+\end{abstract}
+
+\newpage
+
+Our work mostly concerns retrofitting first-order provers to support
+higher-order concepts. But part of our work also concerns integrating
+successful approaches for propositional logic in first-order
+provers. We begin by defining all three logics, in the increasing
+order of expressivity.
+
+\section{Propositional Logic}
+
+Basic formulas of propositional logic syntax are propositional variables
+$\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
+formulas are built inductively using logical connectives $\neg, \land, \lor,
+\imp, \lequiv$: if $\phi$ and $\psi$ are formulas, then $\neg \phi, \phi \land
+\psi, \phi \lor \psi, \phi \imp \psi, \phi \lequiv \psi$ are formulas as well.
+
+To interpret the formulas, propositional variables are assigned 0 (false) or 1
+(true), constants $\top$ and $\bot$ are assigned 1 and 0 respectively and the
+formula is interpreted using the rules for each connective
+\cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
+variables in a formula, trying all (finitely many) possible assignments
+describes an algorithm to decide satisfiability (existence of satisfiable
+assignment) or validity (if all assignments are satisfiable).
+
+However, this simple approach is prohibitively expensive and modern tools that
+decide propositional satisfiability problem (SAT solvers) use more advanced
+approaches such as CDCL calculus \cite{mss-96-cdcl}. Modern SAT solvers also
+heavily preprocess the problem and continue simplifying it during proving
+process.
+
+\section{First-Order Logic}
+
+First-order logic increases the expressivity by allowing quantification over
+objects and has more complicated formula structure. There are many flavors of
+first-order logic, but in this thesis we consider monomorphic first-order logic
+with equality.
+
+We distinguish a set of base types $T$ which is required to have the Boolean
+type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
+tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
+$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
+\ldots, \tau_{n})$ are argument types, $\tau$ is the return type, and that $n$
+is the arity of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
+$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. It is
+required that argument types are not Boolean.
+
+Terms are the basic building blocks of first-order logic. Variables
+$x,y,z,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
+terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
+\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
+of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
+abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$.
+
+Terms are used to build atoms. Atom is either a term $t : o$ or an equation $s
+\eq t$ where terms $t$ and and $s$ are of the same type. Atoms are combined
+using logical connectives just like in propositional logic. Additionally,
+first-order formulas are built using quantifiers: if $\phi$ is a formula and $x$
+is a variable then $\forall x.\, \phi$, as well as $\exists x.\, \phi$ are
+formulas. The first formula is interpreted as ``$\phi$ must hold for all terms
+replaced for $x$'' and the second one as ``there is a replacement for $x$ such
+that $\phi$ holds''. As soon as there is a single functional symbol with arity
+greater than 0, there are infinitely many terms that can be substituted for a
+free variable. It is therefore obvious that the propositional technique for
+deciding satisfiability does not work in the first-order case.
+
+Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
+terms of the same type. Substitutions map only finitely many variables to terms
+other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
+x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself. 
+Applying a substitution $\sigma$ to a term $t$ results in replacing all mapped variables 
+by the corresponding terms and it is denoted.
+\section{Higher-Order Logic}
+\section{Clausal Forms}
+\section{Superposition}
+\subsection{Term Order and Selection}
+\subsection{Unification}
+\subsection{The Calculus}
+\subsection{Redundancy Criterion}
+\subsection{Saturation Procedure}

### 2022-02-17 pre/pre.tex
ho

@@
 
-Basic formulas of propositional logic syntax are propositional variables
+Basic formulas of propositional logic are propositional variables
 $\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
@@
 is the arity of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
-$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. It is
-required that argument types are not Boolean.
+$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. Argument types may not be Boolean.
 
 Terms are the basic building blocks of first-order logic. Variables
-$x,y,z,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
+$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
 terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
@@
 of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
-abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$.
+abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar notation,
+we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$ or simply $\tuplen{t}$.
 
-Terms are used to build atoms. Atom is either a term $t : o$ or an equation $s
-\eq t$ where terms $t$ and and $s$ are of the same type. Atoms are combined
-using logical connectives just like in propositional logic. Additionally,
-first-order formulas are built using quantifiers: if $\phi$ is a formula and $x$
-is a variable then $\forall x.\, \phi$, as well as $\exists x.\, \phi$ are
-formulas. The first formula is interpreted as ``$\phi$ must hold for all terms
-replaced for $x$'' and the second one as ``there is a replacement for $x$ such
-that $\phi$ holds''. As soon as there is a single functional symbol with arity
-greater than 0, there are infinitely many terms that can be substituted for a
-free variable. It is therefore obvious that the propositional technique for
-deciding satisfiability does not work in the first-order case.
+Terms are used to build atoms. Atom is either a term $t$ of type $o$ or an
+equation $s \eq t$ where terms $t$ and and $s$ are of the same type. Atoms are
+combined using logical connectives just like in propositional logic.
+Additionally, first-order formulas are built using quantifiers: if $\phi$ is a
+formula and $x$ is a variable then $\forall x.\, \phi$, as well as $\exists x.\,
+\phi$ are formulas. The first formula requires that $\phi$ holds for any value
+of $x$, while second requires that there is some $x$ for which $\phi$ holds. As
+soon as there is a single functional symbol with arity greater than 0, there are
+infinitely many terms that can be substituted for a free variable. It is
+therefore obvious that the propositional technique for deciding satisfiability
+does not work in the first-order case.
 
@@
 x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
-Applying a substitution $\sigma$ to a term $t$ results in replacing all mapped variables 
-by the corresponding terms and it is denoted.
+Applying a substitution $\sigma$ to a term $t$ results in replacing all mapped
+variables by the corresponding terms and it is denoted.
+
 \section{Higher-Order Logic}
+
+We assume classic simply typed monomorphic higher-order logic with the choice
+operator. Assuming a set of base types $T$ and a set of symbols $\Sigma$ as in
+first-order logic, let us define terms and types. Types are either base types
+$\tau in T$, or function types $\tau_1 \rightarrow \tau_2$ where both $\tau_1$
+and $\tau_2$ are types. Each symbol $\cst{f} \in \Sigma$ is assigned a type.
+
+Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
+\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
+\dotsc$. Furthermore is $s$ and $t$ are terms of type $\tau_1 \rightarrow
+\tau_2$, and $\tau_1$ respectively, then $s \, t$ is a term of type $\tau_2$.
+Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
+$\tau_2$, then $\lambda$-abstractions $\lamx{s}$ is a term of type $\tau_1
+\rightarrow tau_2$. The syntactic distinction between free and bound variables
+gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
+\cite{tn-93-patterns}. In Chapter \ref{ch:ehoh} we write free variables in
+lowercase as the logic described in that chapter disallows
+$\lambda$-abstraction, and consequently bound variables.
+
+We assume the standard notions of $\alpha$, $\beta$ and $\eta$ conversions. We
+let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
+$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
+$\beta$-normal term can be written as $\lam{\tuple{x}{m}}{s \, \tuplen{t}}$,
+where $s$ is not an application; we call $s$ the \emph{head} of the term. If $s$
+is a free variable, we call the term \emph{flex}; otherwise, the term is
+\emph{rigid}.
+
+Throughout this thesis, we consider completeness of higher-order calculi only
+with respect to Henkin semantics \cite{bm-14-automation-ho}. Note that no
+complete calculus exists for higher-order calculi with standard (full) semantics.
+
 \section{Clausal Forms}
@@
 \subsection{Unification}
-\subsection{The Calculus}
-\subsection{Redundancy Criterion}
-\subsection{Saturation Procedure}
+\subsection{Inference Rules}
+\subsection{The Redundancy Criterion}
+\subsection{The Saturation Procedure}

### 2022-02-17 pre/pre.tex
pre until super

@@
     In this chapter we lay out the basic prerequisites for the remaining
-    chapters. We begin by defining three logics that we work with in this
+    chapters. We begin by describing the three logics that we work with in this
     thesis: propositional logic, monomorphic first-order logic and simply typed
     higher-order logic. Then, we explain the clausal structure which is the
-    backbone of many calculi for automated provers. We further describe the
-    superposition calculus and finish with description of the process superposition
-    provers use to actually perform the inferences.
+    backbone of many calculi for automated provers. We finish with the description of the
+    superposition calculus.
 \end{abstract}
@@
 
-Basic formulas of propositional logic are propositional variables
+Atomic formulas of propositional logic are propositional variables
 $\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
-formulas are built inductively using logical connectives $\neg, \land, \lor,
-\imp, \lequiv$: if $\phi$ and $\psi$ are formulas, then $\neg \phi, \phi \land
-\psi, \phi \lor \psi, \phi \imp \psi, \phi \lequiv \psi$ are formulas as well.
+formulas are built inductively using logical connectives $\inot, \iand, \ior,
+\iimplies, \allowbreak \iequiv$: if $\phi$ and $\psi$ are formulas, then $\inot \phi, \phi \iand
+\psi, \allowbreak \phi \ior \psi, \phi \iimplies \psi, \phi \iequiv \psi$ are formulas as well.
 
@@
 \cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
-variables in a formula, trying all (finitely many) possible assignments
+variables in a formula, trying all (finitely many) possible variable assignments
 describes an algorithm to decide satisfiability (existence of satisfiable
-assignment) or validity (if all assignments are satisfiable).
+assignment for a formula) or validity (if all assignments are satisfy the formula).
 
@@
 equation $s \eq t$ where terms $t$ and and $s$ are of the same type. Atoms are
-combined using logical connectives just like in propositional logic.
-Additionally, first-order formulas are built using quantifiers: if $\phi$ is a
-formula and $x$ is a variable then $\forall x.\, \phi$, as well as $\exists x.\,
-\phi$ are formulas. The first formula requires that $\phi$ holds for any value
-of $x$, while second requires that there is some $x$ for which $\phi$ holds. As
-soon as there is a single functional symbol with arity greater than 0, there are
-infinitely many terms that can be substituted for a free variable. It is
-therefore obvious that the propositional technique for deciding satisfiability
-does not work in the first-order case.
+combined using logical connectives to build formulas just like in propositional
+logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
+is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
+$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
+for any value of $x$, while the second one requires that there is some $x$ for
+which $\phi$ holds. As soon as there is a single functional symbol with arity
+greater than 0, there are infinitely many terms that can be substituted for a
+free variable. Furthermore, in first-order logic to determine the truth value of
+the formula, it does not suffice to assign values to variables. It is also
+necessary to interpret symbols. Therefore, it is obvious that the propositional
+technique for deciding satisfiability does not work in the first-order case.
+
+A position in a term is a tuple of natural numbers, with $\varepsilon$ denoting
+the empty tuple. Subterms and positions are inductively defined as follows.
+Term $t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm
+of $t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
+position $i.p$. A context is a term with zero or more subterms replaced by a
+hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
+in the holes of a context $C$ with the terms $\overline{u}_n$, left to
+right.
 
@@
 x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
-Applying a substitution $\sigma$ to a term $t$ results in replacing all mapped
-variables by the corresponding terms and it is denoted.
+Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
+replacing all mapped variables by the corresponding terms.
+The composition $\varrho\sigma$ of substitutions is defined by
+$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$.
 
@@
 
-We assume classic simply typed monomorphic higher-order logic with the choice
-operator. Assuming a set of base types $T$ and a set of symbols $\Sigma$ as in
-first-order logic, let us define terms and types. Types are either base types
-$\tau in T$, or function types $\tau_1 \rightarrow \tau_2$ where both $\tau_1$
-and $\tau_2$ are types. Each symbol $\cst{f} \in \Sigma$ is assigned a type.
+In this thesis we use classic simply typed monomorphic higher-order logic with
+the choice operator (and some variations of this logic). Assuming a set of
+base types $T$ and a set of symbols $\Sigma$, let us
+define terms and types. Types are either base types $\tau \in T$, or function
+types $\tau_1 \rightarrow \tau_2$ where both $\tau_1$ and $\tau_2$ are types.
+Each symbol $\cst{f} \in \Sigma$ is assigned a type.
 
@@
 \dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
-\dotsc$. Furthermore is $s$ and $t$ are terms of type $\tau_1 \rightarrow
-\tau_2$, and $\tau_1$ respectively, then $s \, t$ is a term of type $\tau_2$.
+\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
+\tau_2$, and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
 Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
 $\tau_2$, then $\lambda$-abstractions $\lamx{s}$ is a term of type $\tau_1
-\rightarrow tau_2$. The syntactic distinction between free and bound variables
+\rightarrow \tau_2$. The syntactic distinction between free and bound variables
 gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
-\cite{tn-93-patterns}. In Chapter \ref{ch:ehoh} we write free variables in
-lowercase as the logic described in that chapter disallows
-$\lambda$-abstraction, and consequently bound variables.
+\cite{tn-93-patterns}.
 
-We assume the standard notions of $\alpha$, $\beta$ and $\eta$ conversions. We
-let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
-$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
-$\beta$-normal term can be written as $\lam{\tuple{x}{m}}{s \, \tuplen{t}}$,
-where $s$ is not an application; we call $s$ the \emph{head} of the term. If $s$
-is a free variable, we call the term \emph{flex}; otherwise, the term is
-\emph{rigid}.
+We assume the standard notions of $\alpha$, $\beta$ and $\eta$ conversions and
+write $s \equi t$ if terms $s$ and $t$ are equal modulo $\alpha\beta\eta$
+converison. We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
+and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
+$\beta$-reduced term can be written as $\lam{\tuple{x}{m}}{a \, \tuplen{t}}$,
+where $a$ is not an application; we call $a$ the \emph{head} of the term. By
+convention, $a$ and $b$ denote heads. If $a$ is a free variable, we call the term
+\emph{flex}; otherwise, the term is \emph{rigid}.
+
+Deviating from the standard notion of higher-order subterm, we define subterms
+on $\beta$-reduced terms as follows: a term $t$ is a subterm of $t$ at position
+$\varepsilon$. If $s$ is a subterm of $u_i$ at position $p$, then $s$ is a
+subterm of $a\;\overline{u}_n$ at position $i.p$. If $s$ is a subterm of $t$ at
+position $p$, then $s$ is a subterm of $\lambda x.\, t$ at position $1.p$. Our
+definition of subterm gracefully generalizes the corresponding first-order
+notion: $\cst{a}$ is a subterm of $\cst{f} \, \cst{a} \, \cst{b}$ at position 1,
+but $\cst{f}$ and $\cst{f} \, \cst{a}$ are not subterms of $\cst{f} \, \cst{a}
+\, \cst{b}$.
+
+Higher-order substitutions ($\sigma,\varrho,\ldots$) are functions from free and bound
+variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma F \not\equi
+F$. Each substitution maps only finitely many variables, and it is denoted as in
+first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
+this application $\alpha$-renames $t$ to avoid variable capture. For example,
+$\{X \mapsto x\}(\lamx{X \, a})$ results in $\lam{y}{x \, a}$.  Given a substitution
+$\varrho$, which maps $F$ to $s$, we write $\varrho\setminus\{F \mapsto s\}$ to
+denote a substitution that does not map $F$ and otherwise coincides with
+$\varrho$. Given substitutions $\varrho$ and $\sigma$, which map disjoint sets
+of variables, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.
 
@@
 with respect to Henkin semantics \cite{bm-14-automation-ho}. Note that no
-complete calculus exists for higher-order calculi with standard (full) semantics.
-
+complete calculus exists for higher-order calculi with standard (full)
+semantics.
 \section{Clausal Forms}

### 2022-02-17 introduction/introduction.tex
jasmin

@@
 system. Furthermore, it was less explosive than Davis and Putnam's algorithm
-since it introduced unification as an inference filter.
+since it provided better guided search space exploration.
 
@@
 resolution, featuring support for equality on the calculus level. This support
-is further optimized by use of term orders and selection functions to prune the
+is further optimized by use of various heuristics built to help prune the
 search space. Thanks to efficient implementation, this calculus was used to
-prove long-standing open problem known as ``Robbins conjecture''
+prove long-standing open problem known as the Robbins conjecture
 \cite{mccune-97-robbins}. Provers based on this calculus have been
-winning the first-order division of CASC theorem proving competition since its
+winning the first-order division of the CASC theorem proving competition since its
 inception in the 1990s \cite{ss-96-casc}.
@@
 
-\paragraph{Proof Assistants and Hammers} Proof assistants (or interactive
+\paragraph{Proof Assistants and Hammers} Proof assistants (also called interactive
 theorem provers) are programs which allow users to describe a theory formally
@@
 
-Proof assistants showed indispensible in proving very complex mathematical theorems in a
-trustworthy manner. For example, four-color theorem \cite{gg-2008-four-color}
-and Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
-ending the need to trust hundreds of pages of hard mathematical proofs. Proof
+Proof assistants were essential in proving very complex mathematical theorems in a
+trustworthy manner. For example, the four-color theorem \cite{gg-2008-four-color}
+and the Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
+ending the need to trust hundreds of pages of mathematical proofs. Proof
 assistants are also used to show correctness of complex software such as
@@
 
-Integrating proof automation right into the proof assistant is not an easy task.
+Integrating proof automation directly into the proof assistant is not an easy task.
 Most proof assistants are based on variants of higher-order logic, which is more
@@
 modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
-and Isabelle \cite{lc-88-isabelle} using \emph{hammers} CoqHammer
+and Isabelle \cite{lc-88-isabelle} using the \emph{hammers} CoqHammer
 \cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
@@
 
-The approach proved useful: around 76\% of the simple proof goals can be
+The approach proved useful: on some benchmarks, around 77\% of the proof goals can be
 discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
@@
 mostly first-order formulas, are rather large, and mostly require simple
-higher-order proof steps. Thus, from a standpoint of proof assistant ideal
+higher-order proof steps. Thus, from the standpoint of proof assistants the ideal
 automatic theorem prover fulfills the following wishlist:
 \begin{enumerate}
-  \item\label{it:fo} Performs as best first-order provers on first-order problems
+  \item\label{it:fo} Performs as the best first-order provers on first-order problems
   \item\label{it:size} Scales well with the size of the problem
@@
 \paragraph{Our Approach} To improve the automatic reasoning for higher-order
-logic we decided not to start from a blank slate and build a new theorem prover.
-Rather, we start from a position of strength --- from state-of-the-art
+logic, we decided not to start from a blank slate and build a new theorem prover.
+Rather, we start from a position of strength --- from the state-of-the-art
 first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
@@
 
-We have split the road to full higher-order logic in three stages. A
+We identified three stops on the road to full higher-order logic. A
 distinguished feature of higher-order logic is that it not only supports
@@
 with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
-\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
-disprove this statement. The first stage is to extend E to support
+\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
+disprove this statement. The first stop is to extend E to support
 quantification over functions that are already present in the theory.
@@
 \cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
-second stage is to support automatically synthesizing such functions.
+second stop is to support automatically synthesizing such functions.
 
@@
 represent statements such as ``what Jasmin says is true'' as $\forall x. \,
-\cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support Boolean terms
+\cst{says}(\cst{jasmin}, x) \imp x$. The last stop is to support Boolean terms
 as first-class citizens.
@@
 
-The biggest contribution of this thesis is the extension of two theorem provers,
+The main contribution of this thesis is the extension of two theorem provers,
 E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
-extend those provers, we faced many challenges. Some of them were of engineering
+extend these provers, we faced many challenges. Some of them were of engineering
 nature, some of them were algorithmic, while others concerned fine-tuning the
-heuristics to achieve best possible performance. We discuss those challenges in more detail:
+heuristics to achieve the best possible performance. We discuss those challenges in more detail:
 \begin{itemize}
@@
   higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
-  We had to modify internal term representation of Ehoh to support $\lambda$-free
+  We had to modify internal term representation of E to support $\lambda$-free
   higher-order terms, implement new unification and matching algorithms, extend
@@
   higher-order division of CASC theorem proving competition with Zipperposition
-  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
-  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
+  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
+  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
+  \cite{cb-12-satallax}.
   \item Through this experience we identified one of the bottlenecks of
@@
   developed a complete higher-order unification algorithm that was more
-  efficient and less redundant than current state of the art. 
+  efficient and generated less redundant unifiers than the state of the art. 
   \item We also noticed that Zipperposition's support for first-class Booleans
@@
   Zipperposition 2 implemented these rules together with the new unification
-  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
-  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
+  algorithm. In 2020, Zipperposition 2 won the higher-order division
+  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
   prover, Satallax.
@@
   ways our pragmatic extension was too weak to prove hard theorems of
-  higher-order logic. We further tuned the heuristics and implemented new ways
+  higher-order logic. We further fine-tuned the heuristics and implemented new ways
   to tame the search space explosion. In 2021 we released Zipperposition 2.1
@@
   heuristics. Even though this extension has larger scope than the original
-  extension from E to Ehoh, strong basis that we have earlier built made it
+  extension from E to Ehoh, the strong basis that we have earlier built made it
   manageable.
 
-  \item Following our approach of lifting efficient provers from
-  first-order to higher-order logic, we went a step back in an effort to improve
-  first-order provers. In particular, we looked at the most efficient approaches
-  to simplify problems in propositional logic, a simpler, decidable logic that disallows quantification. 
-  We lifted some of these approaches to first-order logic
-  and implemented them in Zipperposition. 
+  \item After spending so much time and energy on developing higher-order
+  provers, we took a step back and worked on improving first-order provers. In
+  particular, we looked at the most efficient approaches to simplify problems in
+  propositional logic, a simpler, decidable logic that disallows quantification.
+  We lifted some of these approaches to first-order logic and implemented them
+  in Zipperposition. 
 \end{itemize}
@@
 
-This thesis is cumulative: chapters~\ref{ch:ehoh}~-~\ref{ch:satfol}  are
-directly taken from previous publications (or drafts in case of
+This thesis is cumulative: chapters~\ref{ch:ehoh}--\ref{ch:satfol}  are
+taken from previous publications (or drafts in case of
 Chapter~\ref{ch:ehoh2}) where I was the main author. Use
-of publications in this thesis was authorized by the coauthors.
+of the publications in this thesis was authorized by the coauthors.
 
@@
 while Chapter~\ref{ch:conclusion} gives a brief summary of the work and describes alleys for future work.
-Other six chapters discuss previously described contributions in detail:
+The other six chapters discuss previously described contributions in detail:
 
@@
     Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
-    In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021.  
+    In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021. Published online only.
   \end{enumerate}
@@
     Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
-    In  In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
+    In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
   \end{enumerate}
-\noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification algorithm we designed to be used
+\noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification procedure we designed to be used
 inside efficient theorem provers. It is mostly based on the journal publication
@@
 while parts of the chapter were previously published in the conference
-publication, which won the best paper by junior researcher award:
+publication, which won the best paper by a junior researcher award:
 \vspace{-0.3em}
@@
 \noindent\textbf{Chapter~\ref{ch:bools}} discusses the pragmatic extensions of
-the complete calculus for superposition with lambdas to support first-class
+the complete calculus for superposition with $\lambda$-abstractions to support first-class
 Booleans. It is based on the publication
@@
 \noindent\textbf{Chapter~\ref{ch:ho-techniques}} discusses the approach we took to
-control explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
+control the explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
   \begin{enumerate}[resume]
@@
 \noindent\textbf{Chapter~\ref{ch:ehoh2}} discusses the extension of Ehoh to
-support full higher-order logic. We call the obtained prover \newehoh.
+support full higher-order logic, resulting in \newehoh.
 It is based on the draft 
 \begin{enumerate}[resume]
-  \item Petar Vukmirović, Jasmin Blanchette, Stephan Schulz.  Extending a
+  \item Petar Vukmirović, Jasmin Blanchette, and Stephan Schulz.  Extending a
   High-Performance Prover to Higher-Order Logic. 2022.
@@
 These attempts resulted in early higher-order theorem provers. Andrews and
-collegaues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
+colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
 allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
-and colleageues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
-resolution which introduced cooperative paradigm. Provers of this paradigm
-invoke first-order back-ends in regular intervals to finish the proof. Latest
+and colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
+resolution which introduced the cooperative paradigm. Provers implementing this paradigm
+invoke first-order backends in regular intervals to finish the proof. Latest
 iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
 \cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
-to close tableau branches. It won higher-order division of CASC on 7 ocassions:
+to close tableau branches. It won higher-order division of CASC on eight occasions:
 in 2011 and every year from 2013 to 2019.
 
-Various reserachers also experimented with the idea of extending a first-order
+Various researchers also experimented with the idea of extending a first-order
 prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
@@
 desgined a procedure to enumerate higher-order unifiers using higher-order
-combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat
+combinators~\cite{dd-93-comb-unif}.
+
+To make unification more efficient, Bhayat
 and Reger implemented efficient term indexing in their extension of Vampire
@@
 evaluate different unification algorithms \cite{hv-09-unifalgs}, while Reger and
-Voronkov evaluate different parameters of AVATAR architecture
+Voronkov evaluate different parameters of a novel prover architecture
 \cite{rsv-15-playing-with-avatar}. Other authors evaluated different proof

### 2022-02-17 introduction/introduction.tex
more

@@
 is run on the translated problem. If the first-order prover finds the proof, the
-proof is reconstructed in the proof assistant. This approach is implemented in
-modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
-and Isabelle \cite{lc-88-isabelle} using the \emph{hammers} CoqHammer
+proof is reconstructed in the proof assistant. Modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
+and Isabelle \cite{lc-88-isabelle} implement the approach using the \emph{hammers} CoqHammer
 \cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
@@
 Rather, we start from a position of strength --- from the state-of-the-art
-first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
-prover to higher-order logic. In this way, the points \ref{it:fo} and
+first-order superposition theorem prover E \cite{scv-19-e23} --- and extend it to higher-order logic. In this way, the points \ref{it:fo} and
 \ref{it:size} from the above wishlist are already fulfilled, as E is very
@@
   \item Petar Vukmirović, Jasmin Blanchette, and Stephan Schulz.  Extending a
-  High-Performance Prover to Higher-Order Logic. 2022.
+  High-Per\-for\-mance Prover to Higher-Order Logic. 2022.
 \end{enumerate}

### 2022-02-17 introduction/introduction.tex
more

@@
 system. Furthermore, it was less explosive than Davis and Putnam's algorithm
-since it provided better guided search space exploration.
+since it provided better guidance for search space exploration.
 
@@
 
-Even though the successes of proving Robbins conjecture showed that it is
+Even though the successes of proving the Robbins conjecture showed that it is
 possible, automatically proving hard mathematical problems is still out of reach
@@
 
-We identified three stops on the road to full higher-order logic. A
+We have identified three stops on the road to full higher-order logic. A
 distinguished feature of higher-order logic is that it not only supports
@@
   higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
-  We had to modify internal term representation of E to support $\lambda$-free
+  We had to modify the internal term representation of E to support $\lambda$-free
   higher-order terms, implement new unification and matching algorithms, extend

### 2022-02-18 pre/pre.tex
more

@@
 of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
-abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar notation,
-we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$ or simply $\tuplen{t}$.
+abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar
+notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
+or simply $\tuplen{t}$.
 
@@
 A position in a term is a tuple of natural numbers, with $\varepsilon$ denoting
-the empty tuple. Subterms and positions are inductively defined as follows.
-Term $t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm
-of $t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
+the empty tuple. Subterms and positions are inductively defined as follows. Term
+$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
+$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
 position $i.p$. A context is a term with zero or more subterms replaced by a
 hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
-in the holes of a context $C$ with the terms $\overline{u}_n$, left to
-right.
+in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
+We say a term is ground if it has no variables.
 
@@
 gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
-\cite{tn-93-patterns}.
+\cite{tn-93-patterns}. Out of terms, atoms and formulas are built in the same
+way as in first-order logic.
 
@@
 but $\cst{f}$ and $\cst{f} \, \cst{a}$ are not subterms of $\cst{f} \, \cst{a}
-\, \cst{b}$.
+\, \cst{b}$. We say a term is ground if it has no variables, and closed
+if it has no loosely bound variables.
 
@@
 semantics.
+
 \section{Clausal Forms}
+
+The resolution and superposition calculi do not work directly on formulas, but
+on their simplified forms, clauses. Thus, the initial problem, expressed as a
+set of formulas, must be transformed in a set of clauses. For first-order and
+higher-order logic, there exists such a transformation that does not affect
+satisfiability of the problem \cite{nw-01-small-cnf}.
+
+A literal $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
+finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
+\cdots \llor l_n$. All free variables occurring in the clause are implicitly
+universally quantified. Note that $\llor$ used to denote disjunction in clauses
+is different than formula-building $\ior$.
+
+In standard, non-clausal, first-order and higher-order logic, an atom  is either
+a predicate atom (term of Boolean type) or an equation. For uniformity, and to
+stay close to the implementation, we encode predicate atoms as equations with
+$\itrue$. Negative literals are encoded as disequations. For example,
+$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
+$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.
+
+Applying substitution to a literal is reduced to applying it to both sides of
+the (dis)equation and extended pointwise to clauses. It is written in postfix notation
+as $\sigmacl{l}$ and $\sigmacl{C}$.
+
 \section{Superposition}
+
+Superposition is one of the most successful calculi for first-order logic with
+equality. It ows its success to careful treatment of equality and built-in
+heuristics to prune the search space such as term order and selection functions.
+Before we introduce the rules of the calculus, we provide definitions of these and other
+background concepts.
+
 \subsection{Term Order and Selection}
+
+Superposition calculus is parametrized by a reduction ordering $\succ$, an
+ordering on terms that has the following properties:
+
+\noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
+    \textit{Irreflexive} & For all terms $s$, $s \not\succ s$ \\
+    \textit{Transitive} & For all terms $s, t, u$,  if $s\succ t$ and $t \succ u$ then $s \succ u$ \\ 
+    \textit{Subterm property} & For all terms $s$ and contexts $C$, $C[s] \succ s$ \\
+    \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
+    \textit{Respects contexts} & For all terms $s,t$ and contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
+    \textit{Ground total} & For any two
+    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$. \\
+    \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
+\end{tabular}
+
+The term order is lifted to literals and clauses using the multiset extension of
+$\succ$. The order is extended to multisets as follows
+\cite[Sect.~2.5]{bg-01-resolution}. We write $S_1 \succ S_2$ if $S_1 \not= S_2$
+and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
+> S_2(y)$. We use notation $S(x)$ to denote the number of occurrences of $x$ in
+$S$.
+
+Two orders that are commonly used in superposition theorem proving are
+Knuth-Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and
+lexicographic path order (LPO) \cite[Sect.~5.4.2]{bn-98-tr-and-all-that}.
+KBO assigns integer weights to symbols and uses precedence between symbols to
+break eventual ties. LPO completely relies on precedence and inspects term
+structure more closely to deterine the order.
+
 \subsection{Unification}
 \subsection{Inference Rules}
-\subsection{The Redundancy Criterion}
+\subsection{The Redundancy Criterion and Simplification Rules}
 \subsection{The Saturation Procedure}

### 2022-02-18 pre/pre.tex
more

@@
 Applying substitution to a literal is reduced to applying it to both sides of
-the (dis)equation and extended pointwise to clauses. It is written in postfix notation
+the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
 as $\sigmacl{l}$ and $\sigmacl{C}$.
@@
 KBO assigns integer weights to symbols and uses precedence between symbols to
-break eventual ties. LPO completely relies on precedence and inspects term
-structure more closely to deterine the order.
+break eventual ties. LPO entirely relies on precedence and inspects term
+structure more closely to determine the order.
 

### 2022-02-18 pre/pre.tex
unif

@@
 \subsection{Unification}
+
+Calculi from the resolution family, including superposition, perform inferences
+only on unifiable terms. We say terms $s$ and $t$ are unifiable if there is a
+substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
+$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an unordered pair of two terms of
+the same type. The most general unifier is a unifier $\sigma$, such that for any
+other unifier $\theta$, there is a substitution $\varrho$ such that $\sigma =
+\varrho\theta$. When there is such $\varrho$, we say $\sigma$ is more general
+than $\theta$ In first-order logic, the most general unifier is unique up to
+variable renaming. Furthermore, first-order logic admits efficient algorithms
+for computing the most general unifier \cite{hv-09-unifalgs}.
+
+In higher-order logic, unification is performed modulo rules of
+$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
+general unifier is not guaranteed. Consider the unification of terms $X \,
+(\cst{f} \, \cst{a})$ and $\cst{f} \, (X \, \cst{a})$. Any substitution of the
+form $X \mapsto \lamx{\cst{f}^i \, x}$ is a unifier, with $\cst{f}^i \, x$
+denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general unifiers
+to higher-order logic, the concept of complete sets of unifiers was introduced.
+
+A (higher-order) \emph{unifier} of a multiset of unification constraints $E$ is
+a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
+\in E$. A \emph{complete set of unifiers} (\emph{CSU}) of $E$ is defined as a
+set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
+such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
+$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
+that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general
+unifier corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a
+unifier of the singleton multiset $\{ s \unif t \}$.
+
 \subsection{Inference Rules}

### 2022-02-18 pre/pre.tex
more

@@
 substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
-$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an unordered pair of two terms of
-the same type. The most general unifier is a unifier $\sigma$, such that for any
-other unifier $\theta$, there is a substitution $\varrho$ such that $\sigma =
-\varrho\theta$. When there is such $\varrho$, we say $\sigma$ is more general
-than $\theta$ In first-order logic, the most general unifier is unique up to
-variable renaming. Furthermore, first-order logic admits efficient algorithms
-for computing the most general unifier \cite{hv-09-unifalgs}.
+$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
+unordered pair of two terms of the same type. The most general unifier is a
+unifier $\sigma$, such that for any other unifier $\theta$, there is a
+substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
+$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
+the most general unifier is unique up to variable renaming. Furthermore,
+first-order logic admits efficient algorithms for computing the most general
+unifier \cite{hv-09-unifalgs}.
 
@@
 $\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
-general unifier is not guaranteed. Consider the unification of terms $X \,
-(\cst{f} \, \cst{a})$ and $\cst{f} \, (X \, \cst{a})$. Any substitution of the
+general unifier is not guaranteed. Consider the constraint $X \,
+(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
 form $X \mapsto \lamx{\cst{f}^i \, x}$ is a unifier, with $\cst{f}^i \, x$
-denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general unifiers
-to higher-order logic, the concept of complete sets of unifiers was introduced.
+denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
+unifiers to higher-order logic, the concept of complete sets of unifiers was
+introduced.
 
@@
 $\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
-that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general
-unifier corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a
-unifier of the singleton multiset $\{ s \unif t \}$.
+that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general unifier
+corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
+the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
+if two higher-order terms are unifiable, but there exist algorithms
+that enumerate all elements of CSU for unifiable terms.
+
+These higher-order concepts gracefully generalize the corresponding first-order
+ones. For example, if $s$ at $t$ are first-order there must exist a singleton
+CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
+unifiers as first-order or higher-order and let the context provide enough
+information.
 

### 2022-02-18 pre/pre.tex
inf rules

@@
 > S_2(y)$. We use notation $S(x)$ to denote the number of occurrences of $x$ in
-$S$.
+$S$. To use multiset extension, positive literals are represented as $ \{ \{s\},
+\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
+multisets of such literals.
 
@@
 structure more closely to determine the order.
+
+\newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
+Literal selection function is a function that selects a (multi)subset of literals.
+Superposition requires that at least one of the selected literals is negative.
 
@@
 \subsection{Inference Rules}
+\newcommand{\mgu}{\ensuremath{\mathit{mgu}}}
+
+Before we spell out the inference rules of standard, first-order superposition
+let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
+With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \rightarrow
+t']$ replacement of this subterm with $t'$. Let \selfun be a selection function
+and $\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
+substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
+is selected by \selfun and \sigmacl{l} is $\succ$-maximal within literals in
+\sigmacl{C} or (2) \selfun selected some literals and \sigmacl{l} is the maximal
+within either positive or negative selected literals. We say \sigmacl{l} is
+eligible for paramodulation if it is positive, nothing is selected and
+\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.
+With $\mgu(s,t)$ we denote the most general unifier of $s$ and $t$.
+Using
+horizontal line to separate premises and the conclusion, four inference rules of superposition are as
+follows:
+
+\[
+\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
+\]
+where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
+for resolution; \infname{ER} stands for equality resolution
+\medskip
+
+\[
+\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
+\]
+where $\sigma = \mgu(u,v), \sigma(t) \not\succ \sigma(s)$ and 
+$(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
+\medskip
+
+\[
+\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
+\]
+where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+$(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
+\infname{SP} stands for positive superposition 
+\medskip
+
+\[
+\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
+\]
+where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+$(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
+\infname{SN} stands for negative superposition
+\medskip
+
 \subsection{The Redundancy Criterion and Simplification Rules}
+
 \subsection{The Saturation Procedure}

### 2022-02-18 pre/pre.tex
pres

@@
     higher-order logic. Then, we explain the clausal structure which is the
-    backbone of many calculi for automated provers. We finish with the description of the
-    superposition calculus.
+    backbone of many calculi for automated provers. We finish with the
+    description of the superposition calculus. The text of this chapter is
+    partially based on the preliminaries sections of the publications listed in
+    Chapter \ref{ch:intro}.
+
 \end{abstract}
@@
 \newpage
-
-Our work mostly concerns retrofitting first-order provers to support
-higher-order concepts. But part of our work also concerns integrating
-successful approaches for propositional logic in first-order
-provers. We begin by defining all three logics, in the increasing
-order of expressivity.
 
@@
 
+Even though we do not formally define semantics of logic, we assume the natural
+extensions of domain, valuation, interpretation and model (as defined by Fitting
+\cite{mf-1996-fol}) from unsorted to many-sorted logic. Models we consider are
+\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
+of (un)satisfiability and (in)validity are assumed. We write
+$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
+set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
+(i.e., it satisfies it under every variable assignment), we simply write
+$\ourmodel \models N.$  Abusing notation, we write $M \models N$ to denote that
+$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$
+
 A position in a term is a tuple of natural numbers, with $\varepsilon$ denoting
@@
 The composition $\varrho\sigma$ of substitutions is defined by
-$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$.
+$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is grounding
+if it maps all variables to ground terms.
 
@@
 the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
-as $\sigmacl{l}$ and $\sigmacl{C}$.
+as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a ground instance of $C$ is $\sigma$ is a grounding
+substitution.
 
@@
 eligible for paramodulation if it is positive, nothing is selected and
-\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.
-With $\mgu(s,t)$ we denote the most general unifier of $s$ and $t$.
-Using
-horizontal line to separate premises and the conclusion, four inference rules of superposition are as
-follows:
+\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
+$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
+line to separate premises and the conclusion, four inference rules of
+superposition are as follows:
 
@@
 
+Above described rules are used to generate new clauses from already derived
+ones. Even though orderings and selection are used to reduce the search space,
+it still grows rather fast. To identify unnecessary clauses from the search
+space, superposition features \emph{redundancy criterion}
+\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
+redundant in a ground set of clauses $N$ if it is entailed by a subset of
+clauses in $N$ such that each clause in this subset is smaller than $C$. More
+generally, a clause (ground or not) is redundant in set $N$ if every ground
+instance of $C$ is redundant for some grounding of $N$. Removing redundant
+clauses does not affect completeness of superposition.
+
+Other than the four generating rules, superposition provers also use
+simplification rules, which are justified using redundancy criterion. These
+rules replace the premises with conclusions, and we denote them using double
+bars. As an example of such rule, let us introduce rewriting (demodulation) of
+negative literals (\infname{RN}):
+
+\[
+\namedinference{RN}{s \eq t \qquad u \noteq v \llor D}{(u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
+\]
+where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.
+
+Simple rules such as removing duplicate literals, literals of the form $s \noteq
+t$ or tautological clauses are clearly justified by the redundancy criterion.
+Schulz gives an extensive list of such rules\cite{ss-02-brainiac}.
+
+We say that a clause $C$ subsumes clause $D$ is there is a substitution $\sigma$
+such that $\sigmacl{C} \subseteq D$. Subsumption is one of the most important
+operations within a theorem prover. This rule does not adhere to redundancy
+criterion, but can be justified using other mechanisms \cite{wtrb-20-sat-framework}.
+
 \subsection{The Saturation Procedure}
+
+Superposition provers saturate the input problem with respect to the calculus's
+inference rules using the \emph{given clause procedure}
+\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
+passive set $\mathcal{P}$ and an active set $\mathcal{A}$. All clauses start in
+$\mathcal{P}$. At each iteration of the procedure's main loop, the prover
+chooses a clause $C$ from $\mathcal{P}$, simplifies it, and moves it to
+$\mathcal{A}$. Then all inferences between $C$ and active clauses are performed.
+The resulting clauses are again simplified and put in $\mathcal{P}$.
+The provers differ in which clauses are used for simplification: Otter-loop
+\cite{mcw-1997-otter} provers use both active and passive clauses whereas
+DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.
+The provers we discuss in this thesis are both DISCOUNT-loop provers.

### 2022-02-18 pre/pre.tex
more

@@
 
-Atomic formulas of propositional logic are propositional variables
-$\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
+\emph{Atomic formulas} of propositional logic are propositional variables
+$\cst{p}, \cst{q}, \ldots$ and constants $\itrue$ and $\ifalse$. More complex
 formulas are built inductively using logical connectives $\inot, \iand, \ior,
@@
 To interpret the formulas, propositional variables are assigned 0 (false) or 1
-(true), constants $\top$ and $\bot$ are assigned 1 and 0 respectively and the
+(true), constants $\itrue$ and $\ifalse$ are assigned 1 and 0, respectively; the
 formula is interpreted using the rules for each connective
@@
 $\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
-\ldots, \tau_{n})$ are argument types, $\tau$ is the return type, and that $n$
-is the arity of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
-$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. Argument types may not be Boolean.
+\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and that $n$
+is the \emph{arity} of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
+$\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.
 
-Terms are the basic building blocks of first-order logic. Variables
+\emph{Terms} are the basic building blocks of first-order logic. Variables
 $x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
@@
 
-Terms are used to build atoms. Atom is either a term $t$ of type $o$ or an
-equation $s \eq t$ where terms $t$ and and $s$ are of the same type. Atoms are
+Terms are used to build \emph{atoms}. Atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
+equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
 combined using logical connectives to build formulas just like in propositional
@@
 greater than 0, there are infinitely many terms that can be substituted for a
-free variable. Furthermore, in first-order logic to determine the truth value of
-the formula, it does not suffice to assign values to variables. It is also
+free variable. Furthermore, in first-order logic, it does not suffice to assign
+values to variables to determine the truth value of the formula. It is also
 necessary to interpret symbols. Therefore, it is obvious that the propositional
@@
 
-A position in a term is a tuple of natural numbers, with $\varepsilon$ denoting
-the empty tuple. Subterms and positions are inductively defined as follows. Term
+A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
+the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
 $t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
 $t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
-position $i.p$. A context is a term with zero or more subterms replaced by a
+position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
 hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
 in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
-We say a term is ground if it has no variables.
+We say a term is {\em ground} if it has no variables.
 
-Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
+{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
 terms of the same type. Substitutions map only finitely many variables to terms
@@
 The composition $\varrho\sigma$ of substitutions is defined by
-$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is grounding
+$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is {\em grounding}
 if it maps all variables to ground terms.
@@
 Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
-$\tau_2$, then $\lambda$-abstractions $\lamx{s}$ is a term of type $\tau_1
+$\tau_2$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau_1
 \rightarrow \tau_2$. The syntactic distinction between free and bound variables
@@
 Higher-order substitutions ($\sigma,\varrho,\ldots$) are functions from free and bound
-variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma F \not\equi
+variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F) \not\equi
 F$. Each substitution maps only finitely many variables, and it is denoted as in
-first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
+the first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
 this application $\alpha$-renames $t$ to avoid variable capture. For example,
-$\{X \mapsto x\}(\lamx{X \, a})$ results in $\lam{y}{x \, a}$.  Given a substitution
+$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$.  Given a substitution
 $\varrho$, which maps $F$ to $s$, we write $\varrho\setminus\{F \mapsto s\}$ to
 denote a substitution that does not map $F$ and otherwise coincides with
-$\varrho$. Given substitutions $\varrho$ and $\sigma$, which map disjoint sets
-of variables, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.
+$\varrho$. Given substitutions $\varrho$ and $\sigma$, mapping disjoint variable
+sets, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.
 
@@
 The resolution and superposition calculi do not work directly on formulas, but
-on their simplified forms, clauses. Thus, the initial problem, expressed as a
-set of formulas, must be transformed in a set of clauses. For first-order and
-higher-order logic, there exists such a transformation that does not affect
+on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
+set of formulas, must be transformed in a set of clauses. For all three
+discussed logic, there exists such a transformation that does not affect
 satisfiability of the problem \cite{nw-01-small-cnf}.
 
-A literal $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
+A {\em literal} $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
 finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
@@
 In standard, non-clausal, first-order and higher-order logic, an atom is either
-a predicate atom (term of Boolean type) or an equation. For uniformity, and to
+predicate or equational. For uniformity, and to
 stay close to the implementation, we encode predicate atoms as equations with
@@
 Applying substitution to a literal is reduced to applying it to both sides of
-the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
-as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a ground instance of $C$ is $\sigma$ is a grounding
-substitution.
+the (dis)equa\-tion and it is extended pointwise to clauses. It is written in
+postfix notation as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a
+ground instance of $C$ if $\sigma$ is a grounding substitution.
 
@@
 
-Superposition calculus is parametrized by a reduction ordering $\succ$, an
+Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an
 ordering on terms that has the following properties:
@@
 
-The term order is lifted to literals and clauses using the multiset extension of
+The term order is lifted to literals and clauses using the {\em multiset extension} of
 $\succ$. The order is extended to multisets as follows
-\cite[Sect.~2.5]{bg-01-resolution}. We write $S_1 \succ S_2$ if $S_1 \not= S_2$
+\cite[Sect.~2.5]{bg-01-resolution}. For two multisets $S_1$ and $S_2$, we write $S_1 \succ S_2$ if $S_1 \not= S_2$
 and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
@@
 \newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
-Literal selection function is a function that selects a (multi)subset of literals.
+{\em Literal selection function} is a function that selects a (multi)subset of literals.
 Superposition requires that at least one of the selected literals is negative.
@@
 Calculi from the resolution family, including superposition, perform inferences
-only on unifiable terms. We say terms $s$ and $t$ are unifiable if there is a
+only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
 substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
 $\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
-unordered pair of two terms of the same type. The most general unifier is a
+unordered pair of two terms of the same type. A \emph{most general unifier} is a
 unifier $\sigma$, such that for any other unifier $\theta$, there is a
@@
 (\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
-form $X \mapsto \lamx{\cst{f}^i \, x}$ is a unifier, with $\cst{f}^i \, x$
+form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
 denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
-unifiers to higher-order logic, the concept of complete sets of unifiers was
+unifiers to higher-order logic, the concept of {\em complete sets of unifiers} was
 introduced.
 
-A (higher-order) \emph{unifier} of a multiset of unification constraints $E$ is
+A (higher-order) unifier of a multiset of unification constraints $E$ is
 a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
-\in E$. A \emph{complete set of unifiers} (\emph{CSU}) of $E$ is defined as a
+\in E$. A complete set of unifiers (\emph{CSU}) of $E$ is defined as a
 set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
@@
 CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
-unifiers as first-order or higher-order and let the context provide enough
+unifiers as first-order or higher-order and ensure the context provides enough
 information.
@@
 let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
-With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \rightarrow
-t']$ replacement of this subterm with $t'$. Let \selfun be a selection function
-and $\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
+With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \leftarrow
+t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
+$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
 substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
-is selected by \selfun and \sigmacl{l} is $\succ$-maximal within literals in
-\sigmacl{C} or (2) \selfun selected some literals and \sigmacl{l} is the maximal
+is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within literals in
+\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
 within either positive or negative selected literals. We say \sigmacl{l} is
@@
 for resolution; \infname{ER} stands for equality resolution
-\medskip
 
@@
 \]
-where $\sigma = \mgu(u,v), \sigma(t) \not\succ \sigma(s)$ and 
+where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
 $(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
-\medskip
 
@@
 \infname{SP} stands for positive superposition
-\medskip
 
@@
 ones. Even though orderings and selection are used to reduce the search space,
-it still grows rather fast. To identify unnecessary clauses from the search
+it still grows rather fast. To identify unnecessary clauses in the search
 space, superposition features \emph{redundancy criterion}
 \cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
-redundant in a ground set of clauses $N$ if it is entailed by a subset of
+{\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
 clauses in $N$ such that each clause in this subset is smaller than $C$. More
@@
 \[
-\namedinference{RN}{s \eq t \qquad u \noteq v \llor D}{(u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
+\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
+              {s \eq t \qquad (u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
 \]
@@
 Simple rules such as removing duplicate literals, literals of the form $s \noteq
-t$ or tautological clauses are clearly justified by the redundancy criterion.
-Schulz gives an extensive list of such rules\cite{ss-02-brainiac}.
+s$ or tautological clauses are clearly justified by the redundancy criterion.
+Schulz gives an extensive list of such rules \cite{ss-02-brainiac}.
 

### 2022-02-19 introduction/introduction.tex
more

@@
 nature, some of them were algorithmic, while others concerned fine-tuning the
-heuristics to achieve the best possible performance. We discuss those challenges in more detail:
-\begin{itemize}
-  \item We implemented the complete superposition calculus for $\lambda$-free
+heuristics to achieve the best possible performance. We discuss those challenges in more detail.
+
+  First, we implemented the complete superposition calculus for $\lambda$-free
   higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
@@
   performance of heuristics on higher-order problems.
-  \item Together with Alexander Bentkamp and other colleagues we developed
-  a complete superposition calculus for the second stage described above
+
+  Together with Alexander Bentkamp and other colleagues we developed
+  a complete superposition calculus for the second phase described above
   \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
@@
   \cite{cb-12-satallax}.
-  \item Through this experience we identified one of the bottlenecks of
+
+  Through this experience we identified one of the bottlenecks of
   Zipperposition: its unification algorithm. To remove this bottleneck we
@@
   efficient and generated less redundant unifiers than the state of the art.
-  \item We also noticed that Zipperposition's support for first-class Booleans
+
+  We also noticed that Zipperposition's support for first-class Booleans
   was lacking compared to competition. We have closely studied the problems on
@@
   prover, Satallax.
-  \item Inspired by this success we implemented the complete superposition
-  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
+
+  Inspired by this success, we implemented the complete superposition
+  calculus for higher-order logic with Booleans (i.e., the third phase) that we developed with Bentkamp and other
   colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
@@
   runner-up, Vampire 4.5 \cite{lkav-13-vampire}.
-  \item Taking all the experience we gathered while working on Zipperposition,
+
+  Taking all the experience we gathered while working on Zipperposition,
   we further extended Ehoh to full higher-order logic, obtaining a prover we call
@@
 
-  \item After spending so much time and energy on developing higher-order
+  After spending so much time and energy on developing higher-order
   provers, we took a step back and worked on improving first-order provers. In
@@
   in Zipperposition.
-\end{itemize}
 
@@
 The earliest attempts to automate higher-order logic can be traced back to
-Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
-logic to first-order logic using combinators, similar to the approach taken by
-hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
-calculi that are directly applied to higher-order formulas.
+Robinson, who provided two approaches. The first one \cite{ar-69-hol} operates
+directly on higher-order formulas. The second approach \cite{ar-70-hol} is based
+on translating higher-order logic to first-order logic using combinators,
+similar to the approach taken by hammers. Huet~\cite{gh-73-hol} and
+Andrews~\cite{pa-71-type-theory} designed calculi that are directly applied to
+higher-order formulas.
 

### 2022-02-19 pre/pre.tex
more

@@
     backbone of many calculi for automated provers. We finish with the
-    description of the superposition calculus. The text of this chapter is
+    description of the superposition calculus. As this thesis discusses practical
+    aspects of theorem proving, we define only the fundamental notions, while
+    more advanced notions are intuitively described with references to rigorous
+    definitions. The text of this chapter is
     partially based on the preliminaries sections of the publications listed in
@@
 
-\emph{Terms} are the basic building blocks of first-order logic. Variables
+\emph{Terms} are the basic building blocks of first-order logic, built inductively as follows. Variables
 $x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
@@
 \ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
-of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
-abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar
+of type $\tau$. If $n=0$, we drop the parentheses and write $\cst{f}$. We also
+abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using a similar
 notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
@@
 values to variables to determine the truth value of the formula. It is also
-necessary to interpret symbols. Therefore, it is obvious that the propositional
+necessary to interpret the symbols. Therefore, it is obvious that the propositional
 technique for deciding satisfiability does not work in the first-order case.
@@
 The composition $\varrho\sigma$ of substitutions is defined by
-$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is {\em grounding}
-if it maps all variables to ground terms.
+$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution $\sigma$ is {\em grounding}
+if $\sigma(x) \not= x$ implies that $\sigma(x)$ is ground.
 
@@
 \dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
-\tau_2$, and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
+\tau_2$ and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
 Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
@@
 gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
-\cite{tn-93-patterns}. Out of terms, atoms and formulas are built in the same
-way as in first-order logic.
+\cite{tn-93-patterns}. Note that there is no requirement that a symbol is not
+applied to terms of Boolean type as in first-order logic. Furthermore, all
+logical connectives are symbols present in the set $\Sigma$. This means that
+higher-order logic does not distinguish between terms and formulas. For example,
+$\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand \cst{q})$ are well-formed
+terms of higher-order logic. For convenience, we still call terms of Boolean
+types \emph{formulas}.
 
@@
 set of formulas, must be transformed in a set of clauses. For all three
-discussed logic, there exists such a transformation that does not affect
+discussed logics, there exists such a transformation that does not affect
 satisfiability of the problem \cite{nw-01-small-cnf}.
@@
 
-In standard, non-clausal, first-order and higher-order logic, an atom is either
+In standard, non-clausal first-order logic, an atom is either
 predicate or equational. For uniformity, and to
@@
 form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
-denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
-unifiers to higher-order logic, the concept of {\em complete sets of unifiers} was
+denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
+is more general than the other. To generalize most general
+unifiers to higher-order logic, the concept of tje {\em complete sets of unifiers} (CSU) was
 introduced.
@@
 A (higher-order) unifier of a multiset of unification constraints $E$ is
-a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
-\in E$. A complete set of unifiers (\emph{CSU}) of $E$ is defined as a
+a substitution $\sigma$, such that $\sigma(s) \equi \sigma(t)$, for all $s \unif t
+\in E$. A complete set of unifiers of $E$ is defined as a
 set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
@@
 $\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
-that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general unifier
-corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
+that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier
+corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
 the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
@@
 substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
-is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within literals in
+is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within the literals in
 \sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
@@
 
-\[
-\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
-\]
-where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
-for resolution; \infname{ER} stands for equality resolution
+\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
+    \multicolumn{2}{l}{{\bf Equality resolution} (ER)} \\[\jot]
+    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible for resolution
+\end{tabular}
 
-\[
-\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
-\]
-where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
-$(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
+\medskip
 
-\[
-\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
-\]
-where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-$(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
-\infname{SP} stands for positive superposition
+\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
+    \multicolumn{2}{l}{{\bf Equality factoring} (EF)} \\[\jot]
+    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}$ 
+        & where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
+        $(s \eq t)\sigma$ is eligible for paramodulation
+\end{tabular}
 
-\[
-\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
-\]
-where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-$(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
-\infname{SN} stands for negative superposition
+\medskip
+
+\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
+    \multicolumn{2}{l}{{\bf Superposition into positive literals} (SP)} \\[\jot]
+    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}$ 
+        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+        $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable
+\end{tabular}
+
+\medskip
+
+\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
+    \multicolumn{2}{l}{{\bf Superposition into negative literals} (SN)} \\[\jot]
+    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}$ 
+        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+         $(s \eq t)\sigma$ is eligible for paramodulation, $(u \noteq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable
+\end{tabular}
+
+% \[
+% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
+% \]
+% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
+% for resolution; \infname{ER} stands for equality resolution
+
+% \[
+% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
+% \]
+% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
+% for resolution; \infname{ER} stands for equality resolution
+
+% \[
+% \namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
+% \]
+% where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
+% $(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
+
+% \[
+% \namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
+% \]
+% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+% $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
+% \infname{SP} stands for positive superposition
+
+% \[
+% \namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
+% \]
+% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
+% $(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
+% \infname{SN} stands for negative superposition
 \medskip
@@
 s$ or tautological clauses are clearly justified by the redundancy criterion.
-Schulz gives an extensive list of such rules \cite{ss-02-brainiac}.
+Schulz gives an extensive list of such rules \cite{ss-02-brainiac}, including rewriting of positive
+literals which has more side conditions.
 
@@
 DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.
-The provers we discuss in this thesis are both DISCOUNT-loop provers.
+The provers we discuss in this thesis, E and Zipperposition, are both DISCOUNT-loop provers.

### 2022-02-20 introduction/introduction.tex
new fonts

@@
 $\cst{rains} \imp \cst{muddy}$), but also to quantify over objects as in ``all
-people are mortal'' ($\forall x.\, \cst{human}(x) \imp \cst{mortal}(x)$).
+people are mortal'' ($\iforall x.\, \cst{human}(x) \imp \cst{mortal}(x)$).
 
@@
 model some concepts in more natural manner. For example, we can naturally
-represent statements such as ``what Jasmin says is true'' as $\forall x. \,
+represent statements such as ``what Jasmin says is true'' as $\iforall x. \,
 \cst{says}(\cst{jasmin}, x) \imp x$. The last stop is to support Boolean terms

### 2022-02-21 title/title.tex
logos, front matter and such
more

@@
 
-op gezag van de Rector Magnificus prof.~dr.~ir.~ [onbekend],
+op gezag van de Rector Magnificus prof.dr.~V.~Subramaniam,
 
@@
 \medskip\noindent
-\begin{tabular}{l}
-    promotoren: Dr.\ A.E.\ Zaidman, Prof.\ dr.\ A.\ van Deursen \\
-    copromotor: Dr.\ ir.\ G.\ Gousios
+\begin{tabular}{ll}
+    promotor: & prof.dr. W. J. Fokkink \\
+    copromotoren: & dr. J. C. Blanchette \\
+                  & dr. Stephan Schulz
 \end{tabular}
@@
 \noindent Samenstelling promotiecommissie:
+
+\textbf{Should this be filled or skipped as in the case of Alex?}
 
@@
 \begin{center}
-    \includegraphics[height=0.5in]{title/logos/tudelft}
+    \includegraphics[height=0.5in]{title/logos/vulogo}
     \hspace{2em}
     %\includegraphics[height=0.5in]{title/logos/casimir} \\
-    \includegraphics[height=0.5in]{title/logos/nwo}
+    \includegraphics[height=0.5in]{title/logos/erclogo}
     \\ \vspace{0.5cm}
@@
 \medskip
-\noindent The author set this thesis in \LaTeX\xspace using the Libertinus and Inconsolata fonts.
+\noindent The author set this thesis in \LaTeX\xspace using the Minion Pro, Nimbus Sans and Inconsolata fonts.
 

### 2022-02-21 ehoh/ehoh.tex
logic

@@
 \label{ch:ehoh}
+
+\renewcommand{\confrep}[2]{#2}
+\includeversion{rep}
+\excludeversion{conf}
+
+\blfootnote{In this work I desgined, implemented and evaluated all changes to term representation, algorithms and indexing data structures.}
+
+\authors{Joint work with\\
+Jasmin~Blanchette, Simon~Cruanes and Stephan~Schulz}
+
+\begin{abstract}
+Decades of work have gone into developing efficient proof calculi, data
+structures, algorithms, and heuristics for first-order automatic theorem
+proving. Higher-order provers lag behind in terms of efficiency. Instead of
+developing a new higher-order prover from the ground up, we propose to start
+with the state-of-the-art superposition prover E and gradually enrich it with
+higher-order features. We explain how to extend the prover's data structures,
+algorithms, and heuristics to $\lambda$-free higher-order logic, a formalism
+that supports partial application and applied variables. Our extension
+outperforms the traditional encoding and appears promising as a stepping stone
+toward full higher-order logic.    
+\end{abstract}
+
+\newpage
+
+\section{Introduction}
+\label{sec:ehoh:introduction}
+
+\begin{sloppypar}
+Superposition provers such as E~\cite{scv-19-e23}, SPASS \cite{wdfksw-09-spass},
+and Vampire \cite{lkav-13-vampire} are among the most successful first-order
+reasoning systems. They serve as backends in various frameworks, including
+software verifiers (e.g., Why3 \cite{fp-13-why3}),
+%
+% \pagebreak[2]%
+automatic higher-order theorem provers (e.g., \hbox{Leo-III} \cite{sb-21-leo3},
+Satallax \cite{cb-12-satallax}), and \begin{rep}one-click \end{rep}``hammers'' in proof assistants
+(e.g., HOLyHammer in HOL Light \cite{ku-15-holyhammer}, Sledgehammer in
+Isabelle \cite{pb-12-sh}).
+
+Dec\-ades of research have gone
+into refining calculi, devising efficient data structures and algorithms,
+and developing heuristics to guide proof search\begin{rep}
+  \cite{ss-17-arcade}\end{rep}.
+This work has mostly focused on first-order logic with equality.
+\end{sloppypar}
+
+\looseness=-1
+Research on higher-order automatic provers has resulted
+in systems such as LEO \cite{cbmk-98-leo}, \textsc{Leo}-II
+\cite{bspt-15-leo2}, and Leo-III \cite{sb-21-leo3},
+based on resolution and paramodulation, and Satallax \cite{cb-12-satallax},
+based on \confrep{tableaux}{tableaux and SAT solving}. They
+feature a ``cooperative'' architecture, pioneered by LEO: They
+are full-fledged higher-order provers that regularly invoke an
+external first-order prover \confrep{}{with a low time limit as a terminal
+procedure, }in an attempt to finish the proof quickly using only
+first-order reasoning. However, the first-order backend will succeed
+only
+if all the necessary higher-order reasoning has been performed,
+meaning that much of the first-order reasoning is carried out by the
+slower higher-order prover. As a result, this architecture leads to
+suboptimal performance on largely first-order problems,
+such as those that often arise
+in interactive verification \cite{ns-13-leo2sh}. For example,
+at the 2017 installment of the CADE ATP System Competition (CASC)
+\cite{gs-17-casc}, Leo-III, which uses E as a backend, proved
+652 out of 2000 first-order problems in the Sledgehammer division, compared
+with 1185 for E on its own and 1433 for Vampire.
+
+To obtain better performance, we propose to start with a competitive
+first-order prover and extend it to full higher-order logic one
+feature at a time.  Our goal is a \emph{graceful} extension, so that
+the system behaves as before on first-order problems, performs mostly
+like a first-order prover on typical, mildly higher-order problems,
+and scales up to arbitrary higher-order problems, in keeping with the
+zero-overhead principle: \emph{What you don't use, you
+  don't pay for.}
+
+As a stepping stone toward full higher-order logic, we initially restrict our
+focus to a higher-order logic without \hbox{$\lambda$-expressions}
+(Sect.~\ref{sec:ehoh:logic}). Compared with first-order logic, its distinguishing
+features are partial application and applied variables. It
+is rich enough to express the recursive equations of higher-order
+combinators, such as $\cst{map}$ on lists:
+%
+\begin{align*}
+\cst{map} \> f \> \cst{nil} & \eq \cst{nil}
+&
+\cst{map} \> f \> (\cst{cons} \; x \; \mathit{xs}) & \eq
+\cst{cons} \; (f \> x) \; (\cst{map} \> f \> \mathit{xs})
+\end{align*}
+
+Our vehicle is E\begin{rep} \cite{ss-02-brainiac,scv-19-e23}\end{rep},
+a prover developed primarily by Schulz. It is written in C and offers
+good performance, with more emphasis on ``brainiac''
+heuristics than on raw speed. E regularly scores among the top
+systems at CASC and is usually the strongest open-source prover in
+the relevant divisions. It also serves as a backend for
+competitive higher-order provers. We refer to our extended version of
+E as Ehoh. It corresponds to a prerelease version of E~2.5 configured with the
+option \verb|--enable-ho|.%
+\footnote{\url{https://github.com/eprover/eprover/commit/80946ac}}
+%We plan to include the new higher-order inferences in the E 2.5 release.
+
+The main challenges we faced concerned the
+representation of types and terms
+(Sect.~\ref{sec:ehoh:types-and-terms}), the unification
+\confrep{algorithm}{and matching algorithms}
+(Sect.~\ref{sec:ehoh:unif-match}), and the indexing data structures
+(Sect.~\ref{sec:ehoh:indexing}). We also adapted the
+inference rules (Sect.~\ref{sec:ehoh:inferences})\confrep{ and}{,} the
+heuristics (Sect.~\ref{sec:ehoh:heuristics})\confrep{}{, and the
+  preprocessor (Sect.~\ref{sec:ehoh:preprocessing})}.
+
+\looseness=-1
+A central aspect of our work is a set of techniques we call
+\emph{prefix optimization}. Taking a traditional look at higher-order terms, they contain twice as many proper
+subterms as first-order terms; for example,
+$\cst{f}\;(\cst{g}\;\cst{a})\;\cst{b}$ contains not only the ``argument'' subterms
+$\cst{g}\;\cst{a}$, $\cst{a}$, $\cst{b}$ but also the ``prefix'' subterms
+$\cst{f}$, $\cst{f}\;(\cst{g}\;\cst{a})$, $\cst{g}$.
+\begin{rep}Many operations, including superposition and rewriting, require
+traversing all subterms of a term.\end{rep}
+Using the optimization, the prover traverses subterms recursively in a
+first-order fashion, considering all the prefixes of a given subterm
+together. %, at little additional cost.
+%
+Our experiments (Sect.~\ref{sec:ehoh:evaluation}) show that Ehoh is
+almost as fast as E on first-order problems and can also prove
+higher-order problems that do not require synthesizing
+$\lambda$-terms. As next steps, we plan to add support for
+$\lambda$-terms and higher-order unification.
+
+\section{Logic}
+\label{sec:ehoh:logic}
+
+\looseness=-1
+Our logic is a variant of the intensional $\lambda$-free Boolean-free higher-order
+logic (\lfhol{}) described by Bentkamp et al.\ %, Blanchette, Cruanes, and Waldmann
+\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
+first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
+%% kotelnikov-15-fool
+extend the syntax of this logic by erasing the distinction between terms and
+formulas, and its semantics by interpreting the Boolean type $o$ as a domain
+of cardinality~2. Functional extensionality can be obtained by adding
+suitable axioms \cite[Sect.~3.1]{bbcw-21-lfho}.
+
+\looseness=-1
+
+This logic differs from the higher-order logic described in
+Sect.~\ref{sec:pre:hol} three ways. First, $\lambda$-abstraction is disallowed.
+Second, logical connectives are not part of the set of symbols. Instead, there
+is a special inductive case in the definition of terms that defines formulas.
+Third, subterms are defined in a more traditional way.
+
+\looseness=-1
+For completeness, we provide the definition of terms. Terms, ranged over by $s,t,u,v$, are either
+\emph{variables} $x, y, z, \dots$, (\emph{function}) \emph{symbols} $\cst{a},
+\cst{b}, \cst{c}, \cst{d},\allowbreak \cst{f},\allowbreak \cst{g}, \ldots{}$
+(often called ``constants'' in the higher-order literature), binary applications
+$s \; t$, or Boolean terms $\itrue$, $\ifalse$, $\inot s$, $s \mathbin{\iand}
+t$, $s \mathbin{\ior} t$, $s \mathbin{\iimplies} t$, $s \mathbin{\iequiv} t$,
+$\iforall x.\> s$, $\iexists x.\> s$, $s \mathrel{\ieq} t$. E and Ehoh clausify
+the input as a preprocessing step, producing a clause set in which the only
+proper Boolean subterms are variables, $\itrue$, and $\ifalse$. Note that we use
+lowercase letters for free variables as bound variables do not appear in
+clauses. A term's \emph{arity} is the number of extra arguments it can take. If
+$\iota$ is a base type, $\cst{f}$ has type $\iota \to \iota \to \iota$, and
+$\cst{a}$ has type $\iota$, then $\cst{f}$ is binary, $\cst{f}\;\cst{a}$ is
+unary, and $\cst{f}\;\cst{a}\;\cst{a}$ is nullary. Subterms are defined in the
+traditional higher-order way; for example, $s \; t$ has all subterms of $s$ and
+$t$ as subterms, in addition to $s \; t$ itself; as a consequence
+$\cst{f} \; \cst{a}$ is a subterm of $\cst{f} \; \cst{a} \;\cst{a}$.
+
+In this chapter, substitutions $\sigma$ are partial
+functions of finite domain from variables to terms, written $\{ x_1 \mapsto s_1,
+\ldots, x_m \mapsto s_m \}$, where each $s_i$ has the same type as $x_i$. The
+substitution $\sigma[x \mapsto s]$ maps $x$ to $s$ and otherwise coincides with
+$\sigma$. Applying $\sigma$ to a variable beyond $\sigma$'s domain is the
+identity. We deviated from the view of substitutions in Section \ref{sec:pre:hol}
+as it made proofs in Sections \ref{sec:ehoh:unif-match} and \ref{sec:ehoh:indexing}
+easier. It is easy to check that both views are equivalent.
+%Applying a substitution to a term~$t$ applies it homomorphically to $t$'s
+%variables.
+
+A well-known technique to support \lfhol{} %in first-order reasoning systems
+is to use the \emph{applicative encoding}:
+Every $n$-ary symbol is mapped to a nullary symbol, and
+application is represented by a distinguished binary symbol $\cst{@}.$ Thus,
+the \lfhol{} term $\cst{f} \; (x\; \cst{a}) \; \cst{b}$ is
+encoded as the first-order term $\cst{@}(\cst{@}(\cst{f}, \cst{@}(x,
+\cst{a})), \cst{b}).$ However, this representation is not graceful, since it
+also introduces $\cst{@}$'s for terms within \lfhol's first-order fragment. By
+doubling the size and depth of terms, the encoding clutters data
+structures and slows down term traversals.
+%By changing the root symbol of terms,
+%it impacts proof search heuristics in subtle ways.
+In our empirical evaluation, we find that the applicative encoding can
+decrease the success rate by up to \NumberOK{15\%} (Sect.~\ref{sec:ehoh:evaluation}).
+For these and further reasons, it
+is not ideal (Sect.~\ref{sec:ehoh:discussion-and-related-work}).
+
+\section{Types and Terms}
+\label{sec:ehoh:types-and-terms}
+
+\section{Unification and Matching}
+\label{sec:ehoh:unif-match}
+
+\section{Indexing Data Structures}
+\label{sec:ehoh:indexing}
+
+\section{Inference Rules}
+\label{sec:ehoh:inferences}
+
+\section{Heuristics}
+\label{sec:ehoh:heuristics}
+
+\section{Preprocessing}
+\label{sec:ehoh:preprocessing}
+
+\section{Evaluation}
+\label{sec:ehoh:evaluation}
+
+\section{Discussion and Related Work}
+\label{sec:ehoh:discussion-and-related-work}
+
+\section{Conclusion}
+\label{sec:ehoh:conclusion}

### 2022-02-21 pre/pre.tex
logic

@@
 \section{Higher-Order Logic}
+\label{sec:pre:hol}
 
@@
 define terms and types. Types are either base types $\tau \in T$, or function
-types $\tau_1 \rightarrow \tau_2$ where both $\tau_1$ and $\tau_2$ are types.
+types $\tau \rightarrow \upsilon$ where both $\tau$ and $\upsilon$ are types.
 Each symbol $\cst{f} \in \Sigma$ is assigned a type.
@@
 \dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
-\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
-\tau_2$ and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
-Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
-$\tau_2$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau_1
-\rightarrow \tau_2$. The syntactic distinction between free and bound variables
+\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau \rightarrow
+\upsilon$ and $\tau$, respectively, then $s \, t$ is a term of type $\upsilon$.
+Lastly, if $x$ is a bound variable of type $\tau$ and $s$ is a term of type
+$\upsilon$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau
+\rightarrow \upsilon$. The syntactic distinction between free and bound variables
 gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)

### 2022-02-21 ehoh/ehoh.tex
more

@@
 
-\looseness=-1
 Research on higher-order automatic provers has resulted
@@
 
+\looseness=-1
 To obtain better performance, we propose to start with a competitive

### 2022-02-22 pre/pre.tex
more

@@
 where $a$ is not an application; we call $a$ the \emph{head} of the term. By
-convention, $a$ and $b$ denote heads. If $a$ is a free variable, we call the term
+convention, $a, b$ or $\zeta$ denote heads. If $a$ is a free variable, we call the term
 \emph{flex}; otherwise, the term is \emph{rigid}.
@@
 
-A {\em literal} $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
+\looseness=-1
+An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq t$. A clause is a
 finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
-\cdots \llor l_n$. All free variables occurring in the clause are implicitly
-universally quantified. Note that $\llor$ used to denote disjunction in clauses
+\cdots \llor l_n$. Free variables of the clauses are implicitly
+universally quantified. Note that $\llor$ denoting disjunction in clauses
 is different than formula-building $\ior$.
@@
 \subsection{Term Order and Selection}
-
 Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an

### 2022-02-23 ehoh/ehoh.tex
unif rules

@@
 
+\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}
 \renewcommand{\confrep}[2]{#2}
@@
 $t$ as subterms, in addition to $s \; t$ itself; as a consequence
-$\cst{f} \; \cst{a}$ is a subterm of $\cst{f} \; \cst{a} \;\cst{a}$.
+$\cst{f} \; \cst{a}$ is a subterm of $\cst{f} \; \cst{a} \;\cst{a}$. With $\Var(x)$
+we denote the set of free variables of $x$ where $x$ ranges over terms, clauses,
+set of clauses or any other objects that contain terms.
 
@@
 as it made proofs in Sections \ref{sec:ehoh:unif-match} and \ref{sec:ehoh:indexing}
-easier. It is easy to check that both views are equivalent.
+easier. It is easy to check that both views are equivalent. We also consider unification 
+constraints $s \unif t$ as \emph{ordered} pairs.
 %Applying a substitution to a term~$t$ applies it homomorphically to $t$'s
@@
 
+\looseness=-1
+The term representation is a central concern when building a theorem
+prover. Delicate changes to E's representation were needed to support
+partial application and especially applied variables. In contrast, the
+introduction of a higher-order type system had a less dramatic impact on the
+prover's code.
+%In this section we
+%describe changes we made to generalize E's term representation and how
+%we completely replaced FOL many-sorted type system with simple types
+%needed for \lfhol{}.
+
+\ourpara{Types} For most of its history, E supported only untyped first-order logic. Cruanes
+implemented support for atomic types for E 2.0
+\cite[p.\,117]{sc-15-simon-phd}. Symbols are declared
+with a type signature:
+$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau.$
+Atomic types are represented by integers, leading to efficient type comparisons.
+
+\looseness=-1
+In \lfhol{}, a type signature is simply a type~$\tau$, in which the
+type constructor $\to$ can be nested---e.g., $(\iota \to \iota)\allowbreak
+\to \iota.$
+%
+A natural way to represent such types is to mimic their recursive
+structure using a tagged union. However, this leads to memory fragmentation;
+a simple operation such as querying the type of a function's $i$th
+argument would require dereferencing $i$ pointers. We prefer a
+flattened representation, in
+which a type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow \iota$
+is represented by a single node labeled with ${\rightarrow}$ and pointing to
+the array $(\tau_1,\dots,\tau_n,\iota).$
+
+Ehoh stores all types in a shared bank and
+implements perfect sharing, ensuring that types that are structurally the same
+are represented by the same object in memory. Type equality can then be
+implemented as a pointer comparison.
+
+\ourpara{Terms}
+
+In E, terms are stored as perfectly shared directed acyclic graphs\begin{rep}
+\cite{ls-01-shared}\end{rep}.
+%
+%\begin{rep}
+%-- or kill completely
+%Even though we could implement applicative encoding as a preprocessor outside of E
+%and feature full support for \lfhol{} in that way, we decided to generalize existing
+%data structures and algorithms and relax existing invariants that are valid
+%only for FOL terms. These generalizations and relaxations are largely dependent
+%on the way E deals with term representation.
+%\end{rep}
+%
+Each node, or \emph{cell}, contains 11~fields, including
+\verb|f_code|, an integer that identifies the term's head symbol (if ${\ge}\;0$)
+or variable (if ${<}\;0$); \verb|arity|, an integer corresponding to the number
+of arguments passed to the head; \verb|args|, an array of size
+\verb|arity| consisting of pointers to arguments; and \verb|binding|,
+which may store a substitution for a variable\begin{rep} (if
+\verb|f_code|$\;{<}\;0$)\end{rep}, used for unification and matching.
+
+In first-order logic, the arity of a variable is always 0, and the arity of a
+symbol is given by its type signature.
+In higher-order logic, variables may have function type and be applied, and
+symbols can be applied to fewer arguments than specified by their type
+signatures. A natural representation of \lfhol{} terms as tagged unions
+would distinguish between variables~$x$, symbols~$\cst{f}$, and binary
+applications $s \; t.$ However, this scheme suffers from memory
+fragmentation and linear-time access, as with the representation of types,
+affecting performance on purely or mostly first-order problems. Instead, we
+propose a flattened representation, as a generalization of E's existing data
+structures: Allow arguments to variables, for symbols let \verb|arity| be
+the number of \begin{conf}\emph{actual} arguments\end{conf}\begin{rep}actual
+arguments, %as opposed to the declared arity,
+and rename the field \verb|num_args|\end{rep}. This representation, often called
+``spine notation,'' % \cite{cervesato-pfenning-2003},
+is isomorphic to the standard definition of higher-order terms with binary
+application. It is employed in various higher-order reasoning systems,
+including Leo-III \cite{sb-21-leo3} and
+Zipperposition \cite{bbtvw-21-sup-lam}.
+
+%\begin{rep} This approach parallels our representation of types.\end{rep}
+
+A side effect of the flattened representation is that prefix subterms are not
+shared. For example, the terms $\cst{f}\;\cst{a}$ and $\cst{f}\;\cst{a}\;\cst{b}$
+correspond to the flattened cells $\cst{f}(\cst{a})$ and
+$\cst{f}(\cst{a}, \cst{b}).$ The argument subterm $\cst{a}$ is shared, but not
+the prefix $\cst{f}\;\cst{a}.$ Similarly,
+$x$ and $x\;\cst{b}$ are represented by two distinct cells, $x()$ and
+$x(\cst{b})$, and there is no connection between the two occurrences of~$x$.
+%
+In particular,
+despite perfect sharing, their \verb|binding| fields are unconnected, leading
+to inconsistencies.
+
+A potential solution would be to systematically traverse a
+clause and set the \verb|binding| fields of all cells of the form $x(\overline{s})$
+whenever a variable~$x$ is bound, but this would be inefficient and inelegant.
+Instead, we implemented a hybrid approach: Variables are applied by an
+explicit application operator
+\appvar{}, to ensure that they are always perfectly shared. Thus, $x\;\cst{b}\;\cst{c}$
+is represented by the cell $\appvar(x, \cst{b}, \cst{c})$, where $x$ is a shared
+subcell. This is graceful, since variables never occur applied in first-order
+terms.
+%
+The main drawback is that some normalization is necessary
+after substitution: Whenever a variable is instantiated by a symbol-headed term,
+the $\appvar$ symbol must be eliminated.
+Applying the substitution $\{x \mapsto \cst{f} \; \cst{a}\}$ to the cell
+$\appvar(x, \cst{b}, \cst{c})$ must produce $\cst{f}(\cst{a}, \cst{b}, \cst{c})$ and
+not $\appvar(\cst{f}(\cst{a}), \cst{b}, \cst{c})$, for consistency with other
+occurrences of $\cst{f} \; \cst{a} \; \cst{b} \; \cst{c}.$
+
+There is one more complication related to the \verb|binding| field. In E, it
+is easy and useful to traverse a term as if a substitution has been applied,
+by following all set \verb|binding| fields. In Ehoh, this is not enough,
+because cells must also be normalized. To avoid repeatedly creating the
+same normalized cells, we introduced a \verb|binding_cache| field
+that connects a $\appvar(x, \overline{s})$ cell with its substitution.
+%
+However, this cache can easily become stale when $x$'s \verb|binding| pointer is
+updated. To detect this situation, we store $x$'s \verb|binding|
+value in the $\appvar(x, \overline{s})$ cell's \verb|binding|
+field (which is otherwise unused).
+To find out whether the cache is valid, it suffices to check that the
+\verb|binding| fields of $x$ and $\appvar(x, \overline{s})$ are equal.
+
+\ourpara{Term Orders} Superposition provers rely on term orders to prune the search
+space. The order must be a simplification order that
+%can be extended to a simplification order that
+% That's true but pointless. It's equivalent to saying that you need a
+% simplification order and the prover may underapproximate it. -JB
+is total on variable-free
+terms. E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
+(LPO). KBO is widely regarded as the more
+robust option for superposition.
+%
+In earlier work, Blanchette and colleagues have shown that only KBO can be
+generalized gracefully while preserving the necessary properties for
+superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
+reason, we focus on KBO.
+
+E implements L\"ochner's linear-time algorithm for KBO
+\cite{bl-06-kbo}, which relies on the \relax{tupling} method to store
+intermediate results. % , avoiding repeated computations.
+It is straightforward to generalize the algorithm to compute the
+graceful \lfhol{} version of KBO \cite{bbww-17-kbo}.
+The main difference is that when comparing two terms $\cst{f} \; \tuple{s}{m}$
+and $\cst{f} \; \tuple{t}{n}$, because of partial application we may now
+have $m \not= n$; this required changing the implementation to
+perform a length-lexicographic comparison of the tuples $\tuple{s}{m}$ and
+$\tuple{t}{n}.$
+
+\ourpara{Input and Output Syntax} 
+E implements the TPTP \cite{gs-17-tptp} formats FOF and TF0, corresponding to
+untyped and mono\-morphic first-order logic, for both input and output. In
+Ehoh, we added support for the \lfhol{} fragment of TPTP TH0, which provides
+mono\-morphic higher-order logic. Thanks to the use of a standard format,
+Ehoh's proofs can immediately be parsed by Sledgehammer
+\cite{pb-12-sh}, which reconstructs them using a variety of
+techniques.
+There is ongoing work on increasing the level of detail of E's proofs, to
+facilitate proof interchange and independent proof checking
+\cite{rs-17-checkable};
+this will also benefit Ehoh.
+
 \section{Unification and Matching}
 \label{sec:ehoh:unif-match}
+
+Syntactic unification of \lfhol{} terms has a first-order flavor.
+It is decidable, and most general unifiers (MGUs) are unique up to variable
+renaming. For example, the unification constraint $\cst{f} \; (x \;
+\cst{a}) \unif x \; (\cst{f} \; \cst{a})$, used to illustrate an infinite set of 
+independent unifiers in full higher-order logic (Sect.~\ref{sec:pre:unif}), has the MGU 
+$\{x \mapsto \cst{f}\}$ in \lfhol{}.
+% in \lfhol{} the unification constraint $\cst{f} \; (y \;
+% \cst{a}) \unif y \; (\cst{f} \; \cst{a})$ has the MGU $\{y \mapsto \cst{f}\}$,
+% while in full higher-order logic infinitely many independent solutions exist as
+% described in Sect.~\ref{sec:pre:unif}. 
+Matching is a special case of unification
+where only the variables on the left-hand side can be instantiated.
+
+An easy but inefficient way to implement unification and matching for \lfhol{}
+is to apply the applicative encoding (Sect.~\ref{sec:ehoh:logic}), perform
+first-order unification or matching, and decode the resulting
+substitution. To avoid the overhead, we generalize the first-order unification
+and matching procedures to operate directly on \lfhol{} terms.
+
+\ourpara{Unification}
+
+We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
+that generalizes Baader and Nipkow \cite{bn-98-tr-and-all-that}.
+A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
+\unif t_i$, where $s_i$ and $t_i$ are of the same type.
+A problem is in \emph{solved form} if it has the form $\{ x_1 \unif t_1,\allowbreak
+\ldots, x_n \unif t_n \}$, where the $x_i$'s are distinct and do not occur in
+the $t_j$'s. The corresponding unifier is
+$\{ x_1 \mapsto t_1,\allowbreak \ldots, x_n \mapsto t_n \}.$
+%
+The transition rules attempt to bring the input constraints into solved form.
+\begin{rep}They can be applied in any order and eventually reach a normal form, which is
+either an idempotent MGU expressed in solved form or the special
+value $\bot$, denoting unsatisfiability of the constraints.\end{rep}
+
+The first group of rules\confrep{ }{---the \emph{positive} rules---}consists
+of operations that focus on a single constraint and replace it with a new
+(possibly empty) set of constraints:
+
+\noindent
+\begin{tabular}{ll}
+  \textsf{Delete} & $\{ t \unif t \} \uplus \mathit{S} \Unifarrow \mathit{S}$ \\[\jot]
+  \textsf{Decompose} & $\{ \cst{f} \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{m} \} \uplus \mathit{S} \Unifarrow {} 
+                        \mathit{S} \cup \{ s_1 \unif t_1, \ldots, s_m \unif t_m \}$ \\[\jot]
+  \textsf{DecomposeX} & $\{ x \; \tuple{s}{{m}} \unif u \; \tuple{t}{{m}} \} \uplus \mathit{S} \Unifarrow {} 
+                         \mathit{S} \cup \{ x \unif u{,}\; s_1 \unif t_1, \ldots, s_{m} \unif t_{m} \}$  \\
+                      & if $x$ and $u$ have the same type and $m > 0$ \\[\jot]
+  \textsf{Orient}     & $\{ \cst{f} \; \overline{s} \unif x \; \overline{t} \} \allowbreak \uplus \mathit{S}
+                         \Unifarrow \mathit{S} \cup \{ x \; \overline{t} \unif \cst{f} \; \overline{s} \}$ \\[\jot]
+  \textsf{OrientXY}   &  $\{ x \; \tuple{s}{m} \unif y \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S}
+                         \Unifarrow \mathit{S} \cup \{ y \; \tuple{t}{n} \unif x \; \tuple{s}{m} \}$  \\
+                      &   if $m > n$ \\[\jot]
+  \textsf{Eliminate}  & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \{ x \unif t \} \cup \{ x \mapsto t \}(\mathit{S})$ \\
+                      &   if $x \in \Var(\mathit{S}) \setminus \Var(t)$
+\end{tabular}
+
+The \textsf{Delete}, \textsf{Decompose}, and \textsf{Eliminate} rules are
+essentially as for first-order terms. The \textsf{Orient} rule is generalized
+to allow applied variables and complemented by a new \textsf{Orient\-XY} rule.
+\textsf{DecomposeX}, also a new rule, can be seen as a variant of
+\textsf{Decompose} that analyzes applied variables; the term $u$ may be an
+application.
+
+The rules of the second group\confrep{ }{---the \emph{negative}
+rules---}detect unsolvable constraints:
+%
+% \begin{description}[labelwidth=\widthof{\rm\textsf{OccursCheck}}]
+% \item[\rm\textsf{Clash}]
+%   $\{ \cst{f} \; \overline{s} \unif \cst{g} \; \overline{t} \} \uplus \mathit{S} \Unifarrow \bot$%
+%   \quad if $\cst{f} \not= \cst{g}$
+
+% \smallskip
+% \item[\rm\textsf{ClashTypeX}]
+%   \begin{tabular}[t]{@{}l@{}}
+%     $\{ x \; \tuple{s}{m} \unif u \; \tuple{t}{m} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$ \\
+%     if $x$ and $u$ have different types
+%   \end{tabular}
+
+% \smallskip
+% \item[\rm\textsf{ClashLenXF}]
+%   $\{ x \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$%
+%   \quad if $m > n$
+
+% \smallskip
+% \item[\rm\textsf{OccursCheck}]
+%   $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \bot$%
+%   \quad if $x \in \Var(t)$ and $x \not= t$
+% \end{description}
+%
+
+\noindent
+\begin{tabular}{ll}
+  \textsf{Clash} &   $\{ \cst{f} \; \overline{s} \unif \cst{g} \; \overline{t} \} \uplus \mathit{S} \Unifarrow \bot$; if $\cst{f} \not= \cst{g}$ \\[\jot]
+  \textsf{ClashTypeX} & $\{ x \; \tuple{s}{m} \unif u \; \tuple{t}{m} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$ ; if $x$ and $u$ have different types \\[\jot]
+  \textsf{ClashLenXF} & $\{ x \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$; if $m > n$ \\[\jot]
+  \textsf{OccursCheck}     & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \bot$;  if $x \in \Var(t)$ and $x \not= t$
+\end{tabular}
+
+\begin{rep}\textsf{Clash} and \textsf{OccursCheck} are essentially as in
+Baader and Nipkow. \textsf{ClashTypeX} and \textsf{ClashLenXF}
+are variants of \textsf{Clash} for applied variables.\end{rep}
+
+\newcommand\unifARROW[1]{\rlap{\ensuremath{\Unifarrow_{#1}\;}}\phantom{\Unifarrow_\textsf{DecomposeX}\;}}
+
+The derivation below demonstrates the computation of MGUs for the
+unification problem
+$\{x \> (z \> \cst{b} \> \cst{c}) \unif \cst{g} \> \cst{a} \> (y \> \cst{c})\}$:
+%
+\begin{align*}
+& \{ \relax{x \> (z \> \cst{b} \> \cst{c})  \unif  \cst{g} \>  \cst{a} \>  (y \> \cst{c})}\}
+\\[-1\jot]
+\unifARROW{\textsf{DecomposeX}}
+& \{ \relax{x \unif \cst{g} \> \cst{a}}{,}\; z \> \cst{b} \> \cst{c} \unif y \> \cst{c} \}
+\\[-1\jot]
+\unifARROW{\textsf{OrientXY}}
+& \{ x \unif \cst{g} \> \cst{a}{,}\; \relax{y \> \cst{c} \unif z \> \cst{b} \> \cst{c}} \}
+\\[-1\jot]
+\unifARROW{\textsf{DecomposeX}}
+& \{ x \unif \cst{g} \> \cst{a}{,}\; \relax{y  \unif z \> \cst{b}}{,}\;  \cst{c} \unif  \cst{c} \}
+\\[-1\jot]
+\unifARROW{\textsf{Delete}}
+& \{ x \unif \cst{g} \> \cst{a}{,}\; y  \unif z \> \cst{b} \}
+\end{align*}
+
+E stores open constraints in a double-ended
+queue. Constraints are processed from the front. New constraints are
+added at the front if they involve complex
+terms that can be dealt with swiftly by \textsf{Decompose} or
+\textsf{Clash}, or to the back if one side is a variable. \begin{rep}This
+delays instantiation of variables %(with a possible increase in term size)
+and allows E to detect structural clashes early. \end{rep}%
+
+During proof search, E repeatedly needs to test a term~$s$ for unifiability
+not only with some other term $t$ but also with $t$'s subterms.
+Prefix optimization speeds up this test: The subterms of $t$ are
+traversed in a first-order fashion; for each such subterm $\zeta \;
+\tuple{t}{n}$, at most one prefix $\zeta \;
+\tuple{t}{k}$, with $k \le n$, is possibly unifiable with $s$, by virtue
+of their having the same arity. \begin{rep}For first-order problems, we can only have $k =
+n$, since all functions are fully applied. \end{rep}Using this technique, Ehoh
+is virtually as efficient as E on first-order terms.
 

### 2022-02-23 ehoh/ehoh.tex
more

@@
 
-\ourpara{Unification}
-
+\subsection{Unification}
 We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
@@
 
+\looseness=-1
+  The transition system introduced above always terminates with a correct
+  answer. Our proofs follow the lines of Baader and Nipkow.
+  The metavariable $\mathcalx{R}$ is used to range over constraint sets $\mathit{S}$
+  and the special value $\bot.$
+  The set of all unifiers of $\mathit{S}$ is denoted by $\Unifier(\mathit{S}).$
+  Note that $\Unifier(\mathit{S} \cup \mathit{S}') = \Unifier(\mathit{S}) \cap \Unifier(\mathit{S}').$
+  We let $\Unifier(\bot) = \emptyset.$
+  The notation $\mathit{S} \Unifarrow^{!} \mathit{S}'$ indicates that $\mathit{S}
+  \Unifarrow^{*} \mathit{S}'$ and $\mathit{S}'$ is a normal form (i.e., there exists no
+  $\mathit{S}''$ such that $\mathit{S}' \Unifarrow \mathit{S}''$).
+  A variable $x$ is \emph{solved} in $\mathit{S}$
+  if it occurs exactly once in $\mathit{S}$, in a constraint of the form $x \unif t.$
+
+  \begin{lemma}\label{lemma:ehoh:unif-preserve}
+    If $\mathit{S} \Unifarrow \mathcalx{R}$, then $\Unifier(\mathit{S}) = \Unifier(\mathcalx{R}).$
+  \end{lemma}
+  \begin{proof}
+    The rules \textsf{Delete}, \textsf{Decompose}, \textsf{Orient}, and
+    \textsf{Eliminate} are proved as in Baader and Nipkow.
+    \textsf{OrientXY} trivially preserves unifiers.
+    For \textsf{DecomposeX}, the core of the argument is as follows:
+    %
+    \begin{align*}
+    & \sigma \in \Unifier(\{ x \; \tuple{s}{{m}} \unif u \; \tuple{t}{{m}} \}) \\[-.5\jot]
+      \text{iff}\quad
+    & \sigma(x \; \tuple{s}{{m}}) = \sigma(u \; \tuple{t}{{m}}) \\[-.5\jot]
+      \text{iff}\quad
+    & \sigma(x) \; \sigma(s_1) \, \ldots \, \sigma(s_m)
+        = \sigma(u) \; \sigma(t_1) \, \ldots \, \sigma(t_m) \\[-.5\jot]
+      \text{iff}\quad
+    & \sigma(x) = \sigma(u){,}~ \sigma(s_1) = \sigma(t_1){,}\; \ldots{,}~ \text{and}~ \sigma(s_m) = \sigma(t_m) \\[-.5\jot]
+      \text{iff}\quad
+    & \sigma \in \Unifier(\{ x \unif u{,}\; s_1 \unif t_1, \ldots, s_{m} \unif t_{m}\})
+    \end{align*}
+
+    The proof of the problem's unsolvability if rule \textsf{Clash} or
+    \textsf{OccursCheck} is applicable carries over from Baader and Nipkow.
+    For \textsf{ClashTypeX}, the justification is that $\sigma(x \;
+    \tuple{s}{m}) = \sigma(u \; \tuple{t}{m})$ is possible only if
+    $\sigma(x) = \sigma(u)$, which requires $x$ and $u$ to have the same
+    type. Similarly, for \textsf{ClashLenXF}, if
+    $\sigma(x \; \tuple{s}{m}) = \sigma(\cst{f} \; \tuple{t}{n})$ with
+    $m > n$, we must have $\sigma(x \; \tuple{s}{{m-n}}) =
+    \sigma(x) \; \sigma(s_1) \, \dots \, \sigma(s_{m-n}) = \cst{f}$,
+    which is impossible.
+  \end{proof}
+\newpage
+  \begin{lemma}\label{lemma:ehoh:unif-normal-imp-solved}
+    If $\mathit{S}$ is a normal form, then $\mathit{S}$ is in solved form.
+  \end{lemma}
+  \begin{proof}
+    Consider an arbitrary unification constraint $s \unif t \in \mathit{S}.$ We show
+    that in all but one cases, a rule is applicable, contradicting the
+    hypothesis that $\mathit{S}$ is a normal form. In the remaining case, $s$ is a
+    solved variable in $\mathit{S}$.
+
+    \medskip
+    \noindent
+    \textsc{Case} $s = x$:
+    \begin{itemize}
+    \item \textsc{Subcase} $t = x$:
+      \textsf{Delete} is applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t \neq x$ and $x \in \Var(t)$:
+      \textsf{OccursCheck} is applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t \neq x$, $x \not \in \Var(t)$, and $x \in
+      \Var(\mathit{S} \setminus \{ s \unif t \})$:
+      \textsf{Eliminate} is applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t \neq x$, $x \not \in \Var(t)$,
+          and $x \notin \Var(\mathit{S} \setminus \{ s \unif t \})$:
+      The variable $x$ is solved in $\mathit{S}$.
+    \end{itemize}
+
+    \noindent
+    \textsc{Case} $s = x \; \tuple{s}{m}$ for $m > 0$:
+    \begin{itemize}
+    \item \textsc{Subcase} $t = \eta \; \tuple{t}{n}$ for $n \ge m$:
+      \textsf{DecomposeX} or \textsf{Clash\-TypeX} is applicable, depending on
+      whether $x$ and $\eta \; \tuple{t}{{n-m}}$ have the same type.
+
+%      \smallskip
+    \item \textsc{Subcase} $t = y \; \tuple{t}{n}$ for $n < m$:
+      \textsf{OrientXY} is applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t = \cst{f} \; \tuple{t}{n}$ for $n < m$:
+      \textsf{ClashLenXF} is applicable.
+    \end{itemize}
+
+    \noindent
+    \textsc{Case} $s = \cst{f} \; \tuple{s}{m}$:
+    \begin{itemize}
+    \item \textsc{Subcase} $t = x \; \tuple{t}{n}$:
+      \textsf{Orient} is applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t = \cst{f} \; \tuple{t}{n}$:
+      Due to well-typedness, $m = n$. \textsf{Decompose} is
+      applicable.
+
+%      \smallskip
+    \item \textsc{Subcase} $t = \cst{g} \; \tuple{t}{n}$:
+      \textsf{Clash} is applicable.
+    \end{itemize}
+
+    \noindent
+    Since each constraint is of the form $x \unif t$ where $x$ is solved in
+    $\mathit{S}$, the problem $\mathit{S}$ is in solved form.
+  \end{proof}
+
+  \begin{lemma}\label{lemma:ehoh:unif-idempotent-mgu}
+    If the constraint set $\mathit{S}$ is in solved form, then the associated
+    substitution is an idempotent MGU of $\mathit{S}.$
+  \end{lemma}
+  \begin{proof}
+    This lemma corresponds to Lemma~4.6.3 of Baader and Nipkow. Their proof
+    carries over to \lfhol{}.
+  \end{proof}
+
+  \begin{theorem}[Partial Correctness]\label{thm:ehoh:unif-partial-correctness}\,%
+    If $\mathit{S} \Unifarrow^{!} \bot$, then $\mathit{S}$ has no solutions.
+    If $\mathit{S} \Unifarrow^{!} \mathit{S}'\!$, then $\mathit{S}'$ is in solved form and the
+    associated substitution is an idempotent MGU of $\mathit{S}$.
+    \end{theorem}
+
+    \begin{proof}
+    The first part follows from Lemma~\ref{lemma:ehoh:unif-preserve}.
+    %applied transitively
+    The second part follows from Lemma \ref{lemma:ehoh:unif-preserve}
+    %applied transitively
+    and Lemmas
+    \ref{lemma:ehoh:unif-normal-imp-solved}~and~\ref{lemma:ehoh:unif-idempotent-mgu}.
+    \end{proof}
+
+      \newpage
+      \begin{theorem}[Termination]\label{thm:ehoh:unif-termination}\,%
+      The relation $\Unifarrow$ is well founded.
+      \end{theorem}
+
+      \begin{proof}
+      We define an auxiliary notion of weight:
+      $\mathcalx{W}(\zeta \; \tuple{s}{m}) =
+        \allowbreak m + 1 + \sum_{i=1}^{m} \mathcalx{W}(s_i)$.
+      Well-foundedness is proved by exhibiting a measure function from
+      constraint sets to quadruples of natural numbers $(n_1, n_2, n_3, n_4)$,
+      where
+      $n_1$ is the number of unsolved variables in $\mathit{S}$;
+      $n_2$ is the sum of all term weights,
+      $\sum_{s \unif t \in \mathit{S}} \mathcalx{W}(s) + \mathcalx{W}(t)$;
+      $n_3$ is the number of right-hand sides with variable heads,
+      $|\{ s \unif x \; \overline{t} \in \mathit{S} \} |$; and
+      $n_4$ is the number of arguments to left-hand side variable heads,
+        $\sum_{x \> \tuple{s}{m} \unif t  \in \mathit{S}} m$.
+
+      The following table shows that the application of each positive rule
+      lexicographically decreases the quadruple:
+
+        \begin{center}
+          \begin{tabular}{l@{\qquad}l@{\quad}l@{\quad}l@{\quad}l@{\quad}}
+                              & $n_1$  & $n_2$ & $n_3$ & $n_4$ \\[\jot]
+          \textsf{Delete}     & $\geq$ & $>$   &       &       \\
+          \textsf{Decompose}  & $\geq$ & $>$   &       &       \\
+          \textsf{DecomposeX} & $\geq$ & $>$   &       &       \\
+          \textsf{Orient}     & $\geq$ & $=$   & $>$   &       \\
+          \textsf{OrientXY}   & $\geq$ & $=$   & $=$   & $>$   \\
+          \textsf{Eliminate}  & $>$    &       &       &
+          \end{tabular}
+        \end{center}
+  %
+  %    \noindent
+  %    \textsf{Eliminate}, by solving one variable, decreases $n_1.$
+  %    \textsf{Delete} removes a constraint from~$\mathit{S}$, decreasing the
+  %    weight.
+  %    \textsf{Decompose} reduces the weight by at least~2.
+  %    \textsf{Prefix} reduces the set weight by at least $m>0$, thanks to the
+  %    presence of $m$ in the definition of $\mathcalx{W}$.
+      The negative rules, which produce the special value $\bot$, cannot
+      contribute to an infinite $\Unifarrow$ chain.
+      %For example, the quadruples corresponding to the derivation
+      %starting from $\{x \> (z \> \cst{b} \> \cst{c}) \unif \cst{g} \> \cst{a} \> (y \> \cst{c})\}$
+      %given above are $(3,14,0,1) > (2,12,1,2) > (2, 12, 1, 1) > (1, 10, 1, 0) > (1, 8, 1, 0)$.
+  %    where $>$ denotes the lexicographic extension of $>$ on natural numbers.
+      \end{proof}
+
+  A unification algorithm for \lfhol{} can be derived from the above transition
+  system, by committing to a strategy for applying the rules.
+  %
+  This algorithm closely follows the Ehoh implementation, abstracting away from
+  complications such as prefix optimization. We assume a flattened
+  representation of terms; as in Ehoh, each variable stores the term it is bound
+  to in its \textit{binding} field (Sect.~\ref{sec:ehoh:types-and-terms}).
+  We also rely on a \textsc{ApplySubst} function, which applies the
+  binding to the top-level variable. The algorithm assumes that the terms
+  to be unified have the same type. The pseudocode is as follows:
+
+  \begin{quotex}
+    \MyFunction{SwapNeeded}{Term $\mathit{s}$, Term $\mathit{t}$}
+    \q \MyReturn $\begin{aligned}[t]
+      & t.\mathit{head}.\mathit{isVar}() \\[-\jot]
+      & {\mathrel\land}\; (\lnot\, s.\mathit{head}.\mathit{isVar}()
+      \mathrel\lor s.\mathit{num\_args} > t.\mathit{num\_args})
+      \end{aligned}$
+
+    \vskip 1.5\jot
+
+    \MyFunction{Deref}{Term $\mathit{s}$}
+    \q \MyWhile{$s.\mathit{head}.\mathit{isVar}() \mathrel\land s.\mathit{head}.\mathit{binding} \neq \mathit{Null}$}
+    \qq $s \gets \textsc{ApplySubst}(s{,}\; s.\mathit{head}.\mathit{binding})$ \\
+    \q \MyReturn $s$
+
+    \vskip 1.5\jot
+
+    \MyFunction{GobblePrefix}{Term $x$, Term $t$}
+    \q $\mathit{res} \gets \mathit{Null}$ \\
+    \q \MyIf{$x.\mathit{type}.\mathit{args}$ is suffix of
+          $\mathit{t}.\mathit{head}.\mathit{type}.\mathit{args}$}
+    \qq $\mathit{pref\_len} \gets \mathit{t}.\mathit{head}.\mathit{type}.\mathit{arity} - x.\mathit{type}.\mathit{arity}$ \\
+    \qq \MyIf{$\mathit{pref\_len} \leq t.\mathit{num\_args}$}
+    \qqq $\mathit{res} \gets $
+                \textsc{Term}$(t.\mathit{head}$, $t.\mathit{args}[1 \,.\,.\, \mathit{pref\_len}])$ \\
+    \q \MyReturn $\mathit{res}$
+
+    \vskip 1.5\jot
+
+    \MyFunction{Unify}{Term $\mathit{s}$, Term $\mathit{t}$}
+    \q $\mathit{constraints} \gets \textsc{DoubleEndedQueue}()$ \\
+    \q $\mathit{constraints}.\mathit{prepend(s)}$ \\
+    \q $\mathit{constraints}.\mathit{prepend(t)}$
+    \\[\jot]
+    \q \MyWhile{$\lnot\, \mathit{constraints}.\mathit{isEmpty}()$}
+    \qq $t \gets \textsc{Deref}(\mathit{constraints}.\mathit{dequeue}())$ \\
+    \qq $s \gets \textsc{Deref}(\mathit{constraints}.\mathit{dequeue}())$
+    \\[\jot]
+    \qq \MyIf{$s \neq t$}
+    \qqq \MyIf{\textsc{SwapNeeded}$(s,t)$}
+    \qqqq $(t, s) \gets (s, t)$
+    \\[\jot]
+    \qqq \MyIf{$s.\mathit{head}.\mathit{isVar}()$}
+    \qqqq $x \gets s.\mathit{head}$ \\
+    \qqqq $\mathit{prefix} \gets \textsc{GobblePrefix}(x,t)$ \\
+    \qqqq \MyIf{$\mathit{prefix} \neq \mathit{Null}$}
+    \qqqqq $\mathit{start\_idx} \gets \mathit{prefix}.\mathit{num\_args} + 1$ \\
+    \qqqqq \MyIf{$x$ occurs in $\mathit{prefix}$}
+    \qqqqqq \MyReturn $\textit{False}$ \\
+    \qqqqq \MyElse
+    \qqqqqq $x.\mathit{binding} \gets \mathit{prefix}$ \\
+    \qqqq \MyElse
+    \qqqqq \MyReturn $\textit{False}$ \\
+    \qqq \MyElsIf{$s.\mathit{head} = t.\mathit{head}$}
+    \qqqq  $\mathit{start\_idx} \gets 1$ \\
+    \qqq \MyElse
+    \qqqq \MyReturn $\textit{False}$
+    \\[\jot]
+    \qqq \MyForTo{$i \gets \mathit{start\_idx}$}{$t.\mathit{num\_args}$}
+    \qqqq $\mathit{s\_arg} \gets s.\mathit{args}[i-\mathit{start\_idx}+1]$ \\
+    \qqqq $\mathit{t\_arg} \gets t.\mathit{args}[i]$ \\
+    \qqqq \MyIf{$(\mathit{s\_arg}.\mathit{head}.\mathit{isVar}() \lor \mathit{t\_arg}.\mathit{head}.\mathit{isVar}())$}
+    \qqqqq $\mathit{constraints}.\mathit{append}(t\_\mathit{arg})$ \\
+    \qqqqq $\mathit{constraints}.\mathit{append}(s\_\mathit{arg})$ \\
+    \qqqq \MyElse
+    \qqqqq $\mathit{constraints}.\mathit{prepend}(s\_\mathit{arg})$ \\
+    \qqqqq $\mathit{constraints}.\mathit{prepend}(t\_\mathit{arg})$
+    \\[\jot]
+    \q \MyReturn $\textit{True}$
+    \end{quotex}
+
 \section{Indexing Data Structures}

### 2022-02-23 ehoh/ehoh.tex
index

@@
 
+\subsection{Matching}
+
+Given $s$ and $t$, the matching problem consists of finding a
+substitution $\sigma$ such that $\sigma(s) = t.$
+We then write that ``$t$ is an instance of $s$'' or
+``$s$~generalizes~$t$.'' %and call $\sigma$ a \emph{generalizing substitution}.
+We are interested in most general generalizations (MGGs).
+Matching can be reduced to unification by treating variables in $t$ as
+nullary symbols \cite{bn-98-tr-and-all-that}, but E implements it
+separately.
+
+Matching can be specified abstractly as a transition system on matching constraints
+$s_i \MATCH t_i$ consisting of the
+unification rules \textsf{Decompose},
+\textsf{DecomposeX},
+\textsf{Clash},
+\textsf{ClashTypeX},
+\textsf{ClashLenXF}
+(with $\MATCH$ instead of $\unif$)
+and augmented with
+
+\noindent
+\begin{tabular}{ll}
+  \textsf{Double} &   $\{ x \MATCH t{,}\; x \MATCH t' \} \uplus \mathit{S} \Matcharrow \bot$; if $t \not= t'$  \\[\jot]
+  \textsf{ClashLenXY} & $\{ x \; \tuple{s}{m} \MATCH y \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Matcharrow \bot$; if $x \not= y$ and $m > n$ \\[\jot]
+  \textsf{ClashFX} & $\{ \cst{f} \; \overline{s} \MATCH x \; \overline{t} \} \uplus \mathit{S} \Matcharrow \bot$
+\end{tabular}
+
+The matching relation is sound, complete, and well founded. Interestingly, a
+\textsf{Delete} rule would be unsound for matching. Consider the problem
+$\{ x \MATCH x, x \MATCH \cst{g} \; x \}$. Applying \textsf{Delete} to the
+first constraint would yield the solution $\{ x \MATCH \cst{g} \; x \}$, even
+though the original problem is clearly unsolvable.
+
 \section{Indexing Data Structures}
@@
 
+Superposition provers like E work by saturation. Their main loop
+heuristically selects a clause and searches for potential inference
+partners among a possibly large set of other clauses. Mechanisms such
+as simplification and subsumption also require locating terms in a
+large clause set. For example, when E derives a new equation
+$s \eq t$, if $s$ is larger than $t$ according to the term order, it
+will rewrite all instances $\sigma(s)$ of $s$ to $\sigma(t)$ in
+existing clauses.
+
+To avoid iterating over all terms (including subterms) in large clause
+sets, superposition provers store the potential inference partners in
+indexing data structures. A term index stores a set of terms $\mathit{S}.$
+Given a \emph{query term}
+%
+$t$, a query returns all terms $s \in \mathit{S}$ that satisfy a given
+\emph{retrieval condition}:\ $\sigma(s) = \sigma(t)$ ($s$~and $t$ are
+unifiable), $\sigma(s) = t$ ($s$ generalizes~$t$), or $s = \sigma(t)$
+($s$ is an instance of $t$), for some substitution~$\sigma.$
+\emph{Perfect} indices return exactly the subset of terms satisfying
+the retrieval condition. In contrast, \emph{imperfect} indices return
+a superset of eligible terms, and the retrieval condition needs to be
+checked for each candidate.
+
+E relies on two term indexing data structures, perfect discrimination
+trees \cite{mcc-92-pdts} and fingerprint indices
+\cite{ss-12-fp-indexing}, that needed to be generalized to
+\lfhol. It also uses feature vector indices
+\cite{ss-2013-feature-vector} to speed up subsumption and
+related techniques, but these require no changes to work
+with \lfhol{}.
+
+\subsection{Discrimination Trees}
+
+\looseness=-1
+Discrimination trees \cite{mcc-92-pdts} are tries in which every
+node is labeled with a symbol or a variable. A path from the root to
+a leaf corresponds to a ``serialized term''---a term expressed without
+parentheses and commas. Consider the following discrimination trees $D_1$ and
+$D_2$:
+%
+\begin{align*}
+ & \vcenter{\hbox{\begin{tikzpicture}[level distance=2.25em, sibling distance=5em,
+%  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
+%   -| (\tikzchildnode.north)},
+  every node/.style = {shape=circle, minimum size=1.75em, draw, align=center}]
+  \node {\phantom{x}}
+    child { node {$\cst{f}$}
+      child { node %[fill=verylightgray]
+        {$\cst{a}$} }
+    }
+    child { node % [fill=verylightgray]
+      {$\cst{g}$}
+      child { node {$\cst{a}$}
+        child { node % [fill=verylightgray]
+          {$\cst{a}$} }
+      }
+      child[sibling distance=2.5em] { node {$\cst{b}$}
+        child { node %[fill=verylightgray]
+          {$\cst{a}$} }
+        child { node %[fill=verylightgray]
+          {$\cst{b}$} }
+      }
+    }
+    ;
+\end{tikzpicture}
+}}
+&&
+  \vcenter{\hbox{\begin{tikzpicture}[level distance=2.25em, sibling distance=5em,
+%  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
+%   -| (\tikzchildnode.north)},
+  every node/.style = {shape=circle,
+    draw, align=center}]
+  \node {\phantom{x}}
+    child { node {$\cst{f}$}
+      child { node %[fill=verylightgray]
+        {$x$} }
+    }
+    child { node % [fill=verylightgray]
+      {$\cst{g}$}
+      child { node {$\cst{a}$}
+        child { node % [fill=verylightgray]
+          {$\cst{a}$} }
+      }
+      child[sibling distance=2.5em] { node {$y$}
+        child { node %[fill=verylightgray]
+          {$\cst{a}$} }
+        child { node %[fill=verylightgray]
+          {$x$} }
+      }
+    }
+    child[sibling distance=3.75em] { node {$x$} }
+    ;
+\end{tikzpicture}
+}}
+\end{align*}
+%
+Assuming $\cst{a}, \cst{b}, x, y : \iota$, $\cst{f} : \iota \to \iota$,
+and $\cst{g} : \iota^2 \to \iota$,
+$D_1$ represents the term set
+$\{
+\cst{f}(\cst{a})
+{,}\allowbreak\;
+\cst{g}(\cst{a}, \cst{a})
+{,}\allowbreak\;
+\cst{g}(\cst{b}, \cst{a})
+{,}\allowbreak\;
+\cst{g}(\cst{b}, \cst{b})
+\}$,
+and $D_2$ represents the term set
+$\{
+\cst{f}(x)
+{,}\allowbreak\;
+\cst{g}(\cst{a}, \cst{a})
+{,}\allowbreak\;
+\cst{g}(y, \cst{a})
+{,}\allowbreak\;
+\cst{g}(y, x)
+{,}\;
+x
+\}$.
+%
+E uses perfect discrimination trees for finding generalizations of
+query terms. Thus, if the query term is
+$\cst{g}(\cst{a}, \cst{a})$, it would follow the path $\cst{g}.\cst{a}.\cst{a}$
+in~$D_1$ and return $\{\cst{g}(\cst{a}, \cst{a})\}.$
+For~$D_2$, it would also explore paths labeled with
+variables, binding them as it proceeds, and return $\{
+\cst{g}(\cst{a}, \cst{a}){,}\allowbreak\; \cst{g}(y, \cst{a}){,}\allowbreak\; \cst{g}(y, x){,}\; x \}.$
+
+It is crucial for this data structure that distinct terms always give rise to
+distinct serialized terms. Conveniently, this property also holds for \lfhol{}
+terms. Suppose that two distinct \lfhol{} terms yield the same serialization.
+Clearly, they must disagree on parentheses; one will have the subterm $s\; t\;
+u$ where the other has $s\; (t\; u).$ However, these two subterms cannot both
+be well typed.
+
+\looseness=-1
+When generalizing the data structure to \lfhol, we face a complication
+due to partial application. First-order terms can only be stored in leaf nodes,
+but in Ehoh we must also be able to represent partially applied terms,
+such as $\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that
+$\cst{f}$ is unary and $\cst{g}$ is binary). Conceptually, this can be solved
+by storing a Boolean on each node indicating whether it is an accepting state.
+In the implementation, the change is more subtle, because several parts of E's
+code implicitly assume that only leaf nodes are accepting.
+
+\newcommand\TermStack{\ensuremath{T}}
+\newcommand\BTStack{\ensuremath{P}}
+
+The main difficulty specific to \lfhol{} concerns applied variables. To
+enumerate all generalizing terms, E needs to backtrack from child to
+parent nodes. This is achieved using two stacks that store subterms of
+the query term:
+\TermStack{} %(\texttt{term\_\allowbreak stack})
+stores the terms that
+must be matched in turn against the
+current subtree, and
+\BTStack{} %(\texttt{term\_\allowbreak proc})
+stores, for each node from
+the root to the current subtree, the corresponding processed term.
+
+Let $[a_1, \dots, a_n]$ denote an $n$-item stack with $a_1$ on top. Given a
+query term $t$, the matching procedure starts at the root with $\sigma =
+\emptyset$, $\TermStack{} = [t]$, and $\BTStack = []$.
+%
+The procedure advances by repeatedly moving to a suitable child node:
+%
+\begin{enumerate}
+\item[A.] If the node is labeled with a symbol $\cst{f}$ and the top item
+  $t$ of \TermStack{} is \confrep{}{of the form }$\cst{f}(\tuple{t}{n})$,
+  replace $t$ by $n$~new items $t_1,\dots,t_n$, and push $t$ onto \BTStack.
+
+\smallskip
+\item[B.] If the node is labeled with a variable $x$, there are two subcases.
+  If $x$ is already bound, check that $\sigma(x) = t$;
+  otherwise, extend $\sigma$ so that $\sigma(x) = t.$ Next, pop the term $t$
+  from \TermStack{} and push it onto \BTStack.
+\end{enumerate}
+
+%
+The goal is to reach an accepting node. If the query term and all the terms
+stored in the tree are first-order, \TermStack{} will then be empty, and the
+entire query term will have been matched.
+%
+Backtracking works in reverse: Pop a term $t$ from \BTStack; if the current
+node is labeled with an $n$-ary symbol, discard \TermStack{}'s topmost
+$n$~items; push $t$ onto \TermStack. Undo any variable bindings.
+
+As an example, looking up $\cst{g}(\cst{b}, \cst{a})$ in the tree $D_1$
+would result in the following succession of stack states, starting from the
+root $\varepsilon$ along the path $\cst{g}.\cst{b}.\cst{a}$:
+%
+\[\begin{array}{@{}l@{\kern1.5em}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
+& \varepsilon & \cst{g} & \cst{g}.\cst{b} & \cst{g}.\cst{b}.\cst{a} \\[.5\jot]
+\sigma{:} & \emptyset & \emptyset & \emptyset & \emptyset \\
+\TermStack{:} & [\cst{g}(\cst{b}, \cst{a})] & [\cst{b}, \cst{a}] & [\cst{a}] & [\,] \\
+\BTStack{:} & [\,] & [\cst{g}(\cst{b}{,}\; \cst{a})] & [\cst{b}{,}\; \cst{g}(\cst{b}{,}\; \cst{a})] & [\cst{a}{,}\; \cst{b}{,}\; \cst{g}(\cst{b}, \cst{a})]
+\end{array}\]
+%
+Backtracking amounts to moving leftward:
+To get back from $\cst{g}$ to the root, we pop $\cst{g}(\cst{b}, \cst{a})$
+from \BTStack, we discard two items from \TermStack{}, and we push
+$\cst{g}(\cst{b}, \cst{a})$ onto \TermStack{}.
+
+To adapt the procedure to \lfhol{}, the key idea is that an applied variable
+is not very different from an applied symbol. A node labeled with an $n$-ary
+head~$\zeta$ matches a prefix~$t'$ of the $k$-ary term $t$
+popped from \TermStack{} and leaves $n - k$ arguments $\overline{u}$ to be
+pushed back, with $t = t' \; \overline{u}.$ If $\zeta$ is a variable, it must
+be bound to the prefix~$t'$ assuming $\zeta$ and $t'$ are of same type.
+%
+Backtracking works analogously: Given the arity $n$ of the node label $\zeta$
+and the arity $k$ of the term $t$ popped from \BTStack{}, we discard the
+topmost $n - k$ items $\overline{u}$ from \BTStack.
+
+\looseness=-1
+To illustrate the procedure, we consider the tree $D_2$ but change $y$'s type
+to $\iota \to \iota.$ This tree stores
+$\{
+\cst{f}\; x
+{,}\allowbreak\;
+\cst{g}\; \cst{a}\; \cst{a}
+{,}\allowbreak\;
+\cst{g}\; (y\; \cst{a})
+{,}\allowbreak\;
+\cst{g}\; (y\; x)
+{,}\allowbreak\;
+x
+\}.$
+Let $\cst{g}\; (\cst{g}\;\cst{a}\;\cst{b})$ be the query term.
+We have the following sequence of substitutions $\sigma$ and stacks
+$\TermStack, \BTStack$:
+%
+\[\begin{array}{@{}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
+\varepsilon & \cst{g} & \cst{g}.y & \cst{g}.y.x \\[.5\jot]
+\emptyset & \emptyset & \{y \mapsto \cst{g}\> \cst{a}\} & \{y \mapsto \cst{g}\> \cst{a}{,}\; x \mapsto \cst{b}\} \\[0pt]
+[\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{g}\> \cst{a}\> \cst{b}] & [\cst{b}] & [\,] \\[0pt]
+[\,] & [\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{g}\> \cst{a}\> \cst{b}{,}\;\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})] & [\cst{b}{,}\;\cst{g}\> \cst{a}\> \cst{b}{,}\;\cst{g}\> (\cst{g}\> \cst{a}\> \cst{b})]
+\end{array}\]
+
+\confrep{}{When backtracking from $\cst{g}.y$ to $\cst{g}$, by comparing
+$y$'s arity of $n = 1$ with $\cst{g}\; \cst{a}\; \cst{b}$'s arity of $k = 0$,
+we determine that one~item must be discarded from \TermStack{}. }%
+Finally, to avoid traversing twice as many subterms as in the first-order
+case, we can optimize prefixes: Given a
+query term $\zeta\;\tuple{t}{n}$, we can also match prefixes
+$\zeta\;\tuple{t}{k}$, where $k < n$, by allowing \TermStack{} to be nonempty when we reach an accepting node.
+%If accepting nodes correspond to legal prefixes, the stack will then contain
+%exactly the remaining arguments: $[t_{k+1},\dots,t_n].$
+
+\newcommand\quadruple{\mathcalx{Q}}
+Similarly to unification and matching, we present finding generalizations
+in a perfect discrimination tree as a transition system. States
+are quadruples $\quadruple = (\overline{t}, \overline{b}, \mathit{D}, \sigma)$, where
+$\overline{t}$ is a list of terms, $\overline{b}$ is a list of tuples storing
+backtracking information, $\mathit{D}$ is a discrimination
+(sub)tree, and $\sigma$ is a substitution.
+
+%With $+$ we denote list append operation, and with
+%$::$ we denote operation of prepending single element to list.
+
+Let $D$ be a perfect discrimination tree. $\Term(D)$ denotes the set of terms
+stored in $D$. The function $D|_\zeta$ returns the child of $D$ labeled with
+$\zeta$, if it exists. Child nodes are themselves perfect discrimination
+(sub)trees. Given any node $D$, if the node is accepting, then the value stored
+on that node is defined as $\mathit{val}(D) = (s, d)$, where $s$ is the
+accepted term and $d$ is some arbitrary data;
+%(typically, an equation whose one side is~$s$)
+otherwise, $\mathit{val}(D)$ is undefined.
+
+%We label the tree nodes using BFS, starting with 0 as label for
+%root node.
+
+%With $\mathcalx{R}$ we denote $(\overline{t}, \overline{b}, D, \sigma)$ or
+%$(\mathit{v}, \sigma)$ or $\bot$.
+
+Starting from an initial state $([t], [\,], D, \emptyset)$, where $t$ is the query
+term and $D$ is an entire discrimination tree, the following transitions are
+possible:
+% \begin{description}[labelwidth=\widthof{\rm\textsf{AdvanceX}}]
+%   \item[\rm\textsf{AdvanceF}]
+%     $ ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+%         \Pdtarrow
+%         (\tuple{s}{m} \cdot \overline{t}{,}\;
+%         (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
+%       $\quad
+%       if $D{|}_\cst{f}$ is defined
+
+%   \smallskip
+%   \item[\rm\textsf{AdvanceX}]
+%     $ ( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+%         \Pdtarrow {}$ \\
+%     $(\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma[x \mapsto s])
+%       $\\
+%       if $D{|}_x$ is defined, $x$ and $s$ have the same type, and
+%       $\sigma(x)$ is either undefined or equal to $s$
+
+%   \smallskip
+%   \item[\rm\textsf{Backtrack}]
+%     $ ( \tuple{s}{m} \cdot \overline{t}{,}\; (s, D_0, \sigma_0) \cdot \overline{b}{,}\; D{,}\; \sigma)
+%         \Pdtarrow %\\\hphantom{\rm\textsf{AdvanceX}}%
+%       (s \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D_0{,}\; \sigma_0)$ \\
+%       if $D_0{|}_\zeta = D$ and $m = \textit{arity}(\zeta) - \textit{arity}(s)$
+
+%   \smallskip
+%   \item[\rm\textsf{Success}]
+%     $ ( [\,]{,}\; \overline{b}{,}\; D{,}\; \sigma )
+%         \Pdtarrow (\mathit{val}(D){,}\; \sigma)$ \\
+%     if~$\mathit{val}(D)$ is defined
+% \end{description}
+
+\noindent
+\begin{tabular}{ll}
+  \textsf{AdvanceF} & $ ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+           \Pdtarrow
+           (\tuple{s}{m} \cdot \overline{t}{,}\;
+           (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
+         $ \\
+                   & if $D{|}_\cst{f}$ is defined \\[\jot]
+  \textsf{AdvanceX} & $( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma ) 
+                      \Pdtarrow {} 
+                      (\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma[x \mapsto s])$ \\
+                    & if $D{|}_x$ is defined, $x$ and $s$ have the same type,\\ & and
+                    $\sigma(x)$ is either undefined or equal to $s$ \\[\jot]
+  \textsf{Backtrack} & $ ( \tuple{s}{m} \cdot \overline{t}{,}\; (s, D_0, \sigma_0) \cdot \overline{b}{,}\; D{,}\; \sigma)
+                      \Pdtarrow
+                      (s \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D_0{,}\; \sigma_0)$  \\
+                      & if $D_0{|}_\zeta = D$ and $m = \textit{arity}(\zeta) - \textit{arity}(s)$ \\[\jot]
+  \textsf{Success}    & $ ( [\,]{,}\; \overline{b}{,}\; D{,}\; \sigma ) \Pdtarrow (\mathit{val}(D){,}\; \sigma)$ \\
+                      & if~$\mathit{val}(D)$ is defined
+\end{tabular}
+
+Above, $\cdot$ denotes prepending an element or a list to a list.
+%
+Intuitively, \textsf{AdvanceF} and \textsf{AdvanceX} move deeper in
+the tree, generalizing cases A and B above to \lfhol{} terms.
+\textsf{Backtrack} can be used to return to a previous state.
+\textsf{Success} extracts the term~$t$ and data~$d$ stored in
+an accepting node.
+
+\newcommand\BDTARROW[1]{\rlap{\ensuremath{\Pdtarrow_{#1}\;}}\phantom{\Pdtarrow_\textsf{AdvanceX}\;}}
+
+The following derivation illustrates how to locate a generalization of
+$\cst{g}\; (\cst{g}\;\cst{a}\;\cst{b})$ in the tree $D_2$:\pagebreak[2]
+%
+\begin{align*}
+ & ([\cst{g} \; (\cst{g} \; \cst{a} \; \cst{b})]{,}\;  [\,]{,}\;  D{,}\;  \emptyset) \\[-1\jot]
+ \BDTARROW{\textsf{AdvanceF}}&
+ ([\cst{g} \; \cst{a} \; \cst{b}]{,}\;  [(\cst{g} \; (\cst{g} \; \cst{a} \; \cst{b}), D, \emptyset)]{,}\;  D{|}_{\cst{g}}{,}\;  \emptyset)\\[-1\jot]
+ \BDTARROW{\textsf{AdvanceX}}&
+ ([\cst{b}]{,}\;  [(\cst{g} \; \cst{a} \; \cst{b}, D{|}_{\cst{g}}, \emptyset), \ldots]{,}\;  D{|}_{\cst{g}.y}{,}\;  \{y \mapsto \cst{g} \; \cst{a} \}) \\[-1\jot]
+ \BDTARROW{\textsf{AdvanceX}}&
+ ([\,]{,}\;  [(\cst{b}, D{|}_{\cst{g}.y}, \{y \mapsto \cst{g} \; \cst{a} \}), \ldots]{,}\; D{|}_{\cst{g}.y.x}{,}\; \{ y \mapsto \cst{g} \; \cst{a}{,}\; x \mapsto \cst{b} \}) \\[-1\jot]
+ \BDTARROW{\textsf{Success}}& ((\cst{g} \; (y \; x), d){,}\; \{ y \mapsto \cst{g} \; \cst{a}{,}\; x \mapsto \cst{b} \})
+\end{align*}
+
+Let ${\Pdtarrow_{\mathsf{Advance}}} = {\Pdtarrow_{\mathsf{AdvanceF}}} \cup
+{\Pdtarrow_{\mathsf{AdvanceX}}}$.
+%and ${\Pdtarrow_{\mathsf{NoBacktrack}}} = {\Pdtarrow_{\mathsf{Advance}}} \cup
+%{\Pdtarrow_{\mathsf{Success}}}$.
+It is easy to show that \textsf{Backtrack} undoes an \textsf{Advance}
+transition: % and can be avoided when starting from an initial state.
+
+\begin{lemma}\label{lemma:ehoh:backtrack-inverses-advance} If $\quadruple \Pdtarrow_{\mathsf{Advance}} \quadruple'$, then
+  $\quadruple' \Pdtarrow_{\mathsf{Backtrack}} \quadruple$.
+  \end{lemma}
+\begin{proof}
+For both \textsf{Advance} steps, we show that \textsf{Backtrack} restores the state
+properly. If \textsf{AdvanceF} was applied, we have
+%
+\begin{align*}
+  ( \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+& \Pdtarrow_{\mathsf{AdvanceF}}
+  (\tuple{s}{m} \cdot \overline{t}{,}\;
+      (\cst{f} \> \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_\cst{f}{,}\; \sigma)
+\\[-\jot]
+& \Pdtarrow_{\mathsf{Backtrack}}
+  ( \overline{t'}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+\end{align*}
+We must show that $\overline{t'} = \cst{f} \> \tuple{s}{m} \mathrel{\cdot} \overline{t}$.
+Let $k = \textit{arity}(\cst{f})$ and $l = \textit{arity}(\cst{f} \> \tuple{s}{m})$.
+By definition of $k$, we have $m = k - l$, as in \textsf{Backtrack}'s side
+condition. Thus, $\overline{t'} = \cst{f} \> \tuple{s}{m} \mathrel{\cdot}
+\overline{t}$. The other case is
+\begin{align*}
+  ( s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+& \Pdtarrow_{\mathsf{AdvanceX}}
+  (\tuple{s}{m} \cdot \overline{t}{,}\; (s \; \tuple{s}{m}, D, \sigma) \cdot \overline{b}{,}\; D{|}_x{,}\; \sigma')
+\\[-\jot]
+& \Pdtarrow_{\mathsf{Backtrack}}
+  ( \overline{t'}{,}\; \overline{b}{,}\; D{,}\; \sigma )
+\end{align*}
+where $\sigma' = \sigma[x \mapsto s]$.
+Again, we must show that $\overline{t'} = s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}$. Terms
+$x$ and $s$ must have the same type for \textsf{AdvanceX} to be applicable; therefore, they have the same arity. Then, we conclude $m=
+\textit{arity}(s) - \textit{arity}(s \; \tuple{s}{m} ) =  \textit{arity}(x) - \textit{arity}(s \; \tuple{s}{m} )$, as in \textsf{Backtrack}'s
+side condition. Thus, $\overline{t'} = s \; \tuple{s}{m} \mathrel{\cdot} \overline{t}$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:advance-inverses-backtrack}
+  If $\quadruple \Pdtarrow_{\mathsf{Advance}} \quadruple'
+  \Pdtarrow_{\mathsf{Backtrack}} \quadruple''$, then
+  $\quadruple'' = \quadruple$.
+  \end{lemma}
+\begin{proof}
+  By Lemma~\ref{lemma:ehoh:backtrack-inverses-advance}, $\quadruple'
+  \Pdtarrow_{\mathsf{Backtrack}} \quadruple$. Furthermore, \textsf{Backtrack} is
+  clearly functional. Thus, $\quadruple'' = \quadruple$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:backtrack-remove}
+  Let $\quadruple = ([t], [\,], D, \emptyset)$. If $\quadruple
+  \Pdtarrow^* \quadruple'$, then $\quadruple \Pdtarrow_{\mathsf{Advance}}^*\allowbreak
+  \quadruple'$.
+\end{lemma}
+
+\begin{proof}
+  \looseness=-1
+  Let $\quadruple = \quadruple{}_{\,\,0} \Pdtarrow \cdots \Pdtarrow \quadruple{}_{\,\,n} = \quadruple'$.
+  Let $i$ be the index of the first transition of the form $\quadruple{}_{\,\,i} \Pdtarrow_{\mathsf{Backtrack}}
+  \quadruple{}_{\,\,i+1}$. Since $\quadruple{}_{\,\,0}$'s backtracking stack is empty, we must
+  have $i \not= 0$. Hence, we have $\quadruple{}_{\,\,i-1} \Pdtarrow_{\mathsf{Advance}}
+  \quadruple{}_{\,\,i} \Pdtarrow_{\mathsf{Backtrack}} \quadruple{}_{\,\,i+1}$. By
+  Lemma~\ref{lemma:ehoh:advance-inverses-backtrack}, $\quadruple{}_{\,\,i-1} =
+  \quadruple{}_{\,\,i+1}$. Thus, we can shorten the derivation to $\quadruple{}_{\,\,0}
+  \Pdtarrow \cdots \Pdtarrow \quadruple{}_{\,\,i-1} = \quadruple{}_{\,\,i+1} \Pdtarrow
+  \cdots \Pdtarrow \quadruple{}_{\,\,n}$,
+  thereby eliminating one \textsf{Backtrack} transition. By repeating this
+  process, we eliminate all \textsf{Backtrack} transitions.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:termination-pdt}
+  There exist no infinite chains of the form
+  $\quadruple{}_{\,\,0} \Pdtarrow_\mathsf{Advance}\allowbreak \quadruple{}_{\,\,1} \Pdtarrow_\mathsf{Advance}
+  \cdots$.
+\end{lemma}
+\begin{proof}
+  With each \textsf{Advance} transition, the height of the discrimination tree
+  decreases by at least one.
+\end{proof}
+
+Perfect discrimination trees match a single term against a set of terms. To
+prove them correct, we will connect them to the transition system
+$\Matcharrow$ for matching (Sect.~\ref{sec:ehoh:unif-match}). This connection\pagebreak[2] will
+help us show that whenever a discrimination tree stores a generalization of a
+query term, this generalization can be found.
+%
+To express the refinement, we introduce an intermediate transition
+system, $\Matchiiarrow$, that focuses on a single pair of terms (like
+$\Matcharrow$) but that solves the constraints in a depth-first, left-to-right
+fashion and builds the substitution incrementally (like $\Pdtarrow$).
+Its initial states are of the form $([s \MATCH t], \emptyset)$. Its transitions
+are as follows:
+
+\noindent
+\begin{tabular}{ll}
+  \textsf{Decompose} & $ (\cst{f} \; \tuple{s}{m} \MATCH \cst{f} \; \tuple{t}{m} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow {} ( (s_1 \MATCH t_1, \ldots, s_m \MATCH t_m) \cdot \overline{c}, \sigma ) $ \\[\jot]
+  \textsf{DecomposeX} & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow {} ( (s_1 \MATCH t_1, \ldots, s_m \MATCH t_m) \cdot \overline{c}, \sigma[x \mapsto u] )$ \\
+                    & if $x$ and $u$ have the same type and either $\sigma(x)$ is undefined or $\sigma(x)=u$ \\[\jot]
+  \textsf{Success} & $ ([\,], \, \sigma) \Matchiiarrow  \sigma $ \\[\jot]
+  \textsf{Clash}    & $ (\cst{f} \; \tuple{s}{m} \MATCH \cst{g} \; \tuple{t}{n} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot $ \\[\jot]
+  \textsf{ClashTypeX}    & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
+                         & if~$x$ and $u$ have different types  \\[\jot]
+  \textsf{ClashLenXF}    & $(x \; \tuple{s}{{m}} \MATCH \cst{f} \; \tuple{t}{{n}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
+                         & if~$m > n$  \\[\jot]
+  \textsf{ClashLenXY}    & $(x \; \tuple{s}{{m}} \MATCH y \; \tuple{t}{{n}} \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
+                         & if~$x \neq y$ and $m > n$ \\[\jot]
+  \textsf{ClashFX}    & $(\cst{f} \; \overline{s} \MATCH x \; \overline{t}  \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\[\jot]
+  \textsf{Double}    & $(x \; \tuple{s}{{m}} \MATCH u \; \tuple{t}{{m}}  \mathrel{\cdot} \overline{c}, \, \sigma) \Matchiiarrow \bot$ \\
+                     & if $x$ and $u$ have the same type, $\sigma(x)$ is defined, and $\sigma(x) \neq u$ \\[\jot]
+\end{tabular}
+
+\newcommand\AlphaMatch[1]{\alpha(#1)}
+
+We need an auxiliary function to convert $\Matchiiarrow$ states to
+$\Matcharrow$ states.
+Let
+$\AlphaMatch{\{ x_1 \mapsto s_1,\allowbreak \ldots, x_m \mapsto s_m \}}
+ = \{ x_1 \MATCH s_1,\ldots,\,\allowbreak x_m \MATCH s_m \}$,
+$\AlphaMatch{\overline{c}, \sigma} =
+\{ c \mid c \in \overline{c} \} \mathrel\cup \AlphaMatch{\sigma}$, and
+$\AlphaMatch{\bot} = \bot$.
+%
+Moreover, let $\mathcal{S}$ range over states of the form $(\overline{c}, \sigma)$ and
+$\mathcalx{R}$ additionally range over special states of the form $\sigma$ or
+$\bot$.
+
+\begin{lemma}\label{lemma:ehoh:match-refinement}
+  If $\mathcal{S} \Matchiiarrow \mathcalx{R}$,
+  then $\AlphaMatch{\mathcal{S}} \Matcharrow^* \AlphaMatch{\mathcalx{R}}$.
+\end{lemma}
+\begin{proof}
+By case distinction on $\mathcalx{R}$. Let $\mathcal{S} = (\overline{c}, \sigma)$.
+
+\medskip
+
+\begin{sloppypar}%
+  \noindent
+  \textsc{Case} $\mathcalx{R} = (\overline{c'}, \sigma')$:\enskip
+  Only $\Matchiiarrow_{\mathsf{Decompose}}$ and
+  $\Matchiiarrow_{\mathsf{DecomposeX}}$ are possible.
+  If $\Matchiiarrow_{\mathsf{Decompose}}$ is applied, then $\Matcharrow_{\mathsf{Decompose}}$
+  is applicable and results in $\AlphaMatch{\mathcalx{R}}$.
+  If $\Matchiiarrow_{\mathsf{DecomposeX}}$ is applied,
+  we have either $m>0$, and $\Matcharrow_{\mathsf{DecomposeX}}$ is applicable,
+  or $m = 0$, and $\AlphaMatch{\overline{c'},\sigma'} = \AlphaMatch{\mathcal{S}}$,
+  which implies that the two states are connected by an idle transition of
+  $\Matcharrow^*$.
+  \end{sloppypar}
+
+  \medskip
+
+  \noindent
+  \textsc{Case} $\mathcalx{R} = \bot$:\enskip
+  All the $\Matchiiarrow$ rules resulting in $\bot$ except for \textsf{Double}
+  have the same side conditions as the corresponding $\Matcharrow$ rules.
+  $\Matchiiarrow_{\mathsf{Double}}$ corresponds to
+  $\Matcharrow_{\mathsf{Double}}$ if $m=0$. If $m \neq 0$, we need
+  an intermediate $\Matcharrow_{\mathsf{DecomposeX}}$ step before
+  $\Matcharrow_{\mathsf{Double}}$ can be applied to derive $\bot$. Since
+  $\Matchiiarrow_{\mathsf{Double}}$ is applicable, $\sigma(x) = u'
+  \neq u$. Hence, $x \MATCH u'$ must be present in $\AlphaMatch{\overline{c},
+  \sigma}$. $\Matcharrow_{\mathsf{DecomposeX}}$ will augment this set with
+  $x \MATCH u$, enabling $\Matcharrow_{\mathsf{Double}}$.
+
+  \medskip
+
+  \noindent
+  \textsc{Case} $\mathcalx{R} = \sigma$:\enskip
+  The only possible rule is $\Matchiiarrow_{\mathsf{Success}}$,
+  with $\overline{c}=[\,]$. Since $\AlphaMatch{\mathcal{S}} = \AlphaMatch{\sigma}$,
+  this transition corresponds to an idle transition of $\Matcharrow^*$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:refined-match-partial-correctness}\,%
+  If $\mathcal{S} \Matchiiarrow^{!} \mathcalx{R}$, then $\mathcalx{R}$ is either
+  some substitution $\sigma'$ or $\bot$.
+  If $\mathcal{S} \Matchiiarrow^{!} \sigma'$, then $\sigma'$ is the MGG
+  of~$\AlphaMatch{\mathcal{S}}$.
+  If $\mathcal{S} \Matchiiarrow^{!} \bot$, then $\AlphaMatch{\mathcal{S}}$ has
+  no solutions.
+  \end{lemma}
+\begin{proof}
+  %Let $\mathcal{S} = (\overline{c}, \sigma)$.
+
+  First, we show that states $\mathcal{S}' = (\overline{c'}, \sigma')$ cannot be
+  normal forms, by exhibiting transitions from such states. If $\overline{c'} =
+  [\,]$, the $\Matchiiarrow_{\mathsf{Success}}$ rule would apply. Otherwise, let
+  $\overline{c'} = c_1 \cdot \overline{c''}$ and consider the matching problem
+  $\{c_1\} \cup \AlphaMatch{\sigma'}$. If this problem is in solved form, $c_1$
+  is a constraint corresponding to a solved variable, and we can apply
+  $\Matchiiarrow_{\mathsf{DecomposeX}}$ to move the constraint into the
+  substitution. Otherwise, some $\Matcharrow$ rule can be applied. It
+  necessarily focuses on $c_1$, since the constraints from
+  $\AlphaMatch{\sigma'}$ correspond to solved variables. In all cases except for
+  $\Matcharrow_{\mathsf{DecomposeX}}$, a homologous
+  $\Matchiiarrow$ rule  can be applied to $\mathcal{S}'$. If $\Matcharrow_{\mathsf{DecomposeX}}$
+  would make $\Matcharrow_{\mathsf{Double}}$ applicable, then we can apply $\Matchiiarrow_{\mathsf{Double}}$ to $\mathcal{S}'$;
+  otherwise, $\Matchiiarrow_{\mathsf{DecomposeX}}$ is applicable.
+
+  Second, by Lemma \ref{lemma:ehoh:match-refinement},
+  if $\mathcal{S} \Matchiiarrow^{!} \sigma'$, then
+  $\AlphaMatch{\mathcal{S}} \Matcharrow^* \AlphaMatch{\sigma'}$.
+  By construction, $\AlphaMatch{\sigma'}$ is in solved form.
+  Therefore, $\AlphaMatch{\mathcal{S}} \Matcharrow^{!} \AlphaMatch{\sigma'}$.
+  By completeness of $\Matcharrow$, the substitution corresponding to
+  $\AlphaMatch{\sigma'}$---that is, $\sigma'$---is the MGG of~$\AlphaMatch{\mathcal{S}}$.
+
+  Third, by Lemma~\ref{lemma:ehoh:match-refinement}, if $\mathcal{S}
+  \Matchiiarrow^{!} \bot$, then $\AlphaMatch{\mathcal{S}} \Matcharrow^{!} \bot$.
+  By soundness of $\Matcharrow$, $\AlphaMatch{\mathcal{S}}$ has
+  no solutions.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:refined-match-termination}\,%
+  The relation $\Matchiiarrow$ is well founded.
+  \end{lemma}
+\begin{proof}
+  By Lemma \ref{lemma:ehoh:match-refinement}, every $\Matchiiarrow$ transition
+  corresponds to zero or more $\Matcharrow$ transitions.
+  Since $\Matcharrow$ is well founded,
+  the only transitions that can violate well-foundedness of $\Matchiiarrow$ are
+  the ones that take idle $\Matcharrow^*$ transitions:
+  $\Matchiiarrow_{\mathsf{DecomposeX}}$ for $m=0$ and
+  $\Matchiiarrow_{\mathsf{Success}}$. The latter is terminal, so it cannot
+  contribute to infinite chains.
+  As for $\Matchiiarrow_{\mathsf{DecomposeX}}$, with $m=0$,
+  it decreases the following measure~$\mu$, which the other rules nonstrictly
+  decrease, with respect to the multiset extension of $<$ on natural numbers:
+  $\mu([s_1 \MATCH t_1, \ldots, s_m \MATCH t_m], \sigma) =
+    \{\left|s_1\right|, \ldots \left|s_m\right|\}$,
+  where $\left|s\right|$ denotes the syntactic size of $s$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:refined-matching-solution}
+  If term $s$ generalizes $t$, then $([s \MATCH t], \emptyset) \Matchiiarrow^! \sigma$,
+  where $\sigma$ is the MGG of $s \MATCH t$.
+\end{lemma}
+\begin{proof}
+By Lemma \ref{lemma:ehoh:refined-match-termination}, there exists a normal form
+$\mathcalx{R}$ starting from $\mathcal{S} = ([s \MATCH t],\allowbreak \emptyset)$. Since $s
+\MATCH t$ is solvable, by Lemma~\ref{lemma:ehoh:refined-match-partial-correctness},
+and soundness of $\Matchiiarrow$ (a consequence of Lemma \ref{lemma:ehoh:match-refinement}
+and soundness of $\Matcharrow$),
+$\mathcalx{R}$ must be the MGG for $s$ and~$t$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:pdt-complete-help}
+  If there exists a term $s \in \Term(D)$ that generalizes the query term~$t$,
+  then there exists a derivation $([t], [\,],\allowbreak D,\allowbreak \emptyset) \Pdtarrow^! ((s, d),
+  \sigma)$. %, where $\sigma$ is the MGG for $s$ and $t$.
+  \end{lemma}
+  \begin{proof}
+    By Lemma \ref{lemma:ehoh:refined-matching-solution}, we know that $(s \MATCH t,
+    \emptyset) \Matchiiarrow^! \sigma$ for each $s \in \Term(D)$ generalizing $t$.
+    This means that there exists a derivation
+    $([s \MATCH t], \emptyset) = (\overline{c_0}, \sigma_0) \Matchiiarrow \cdots \Matchiiarrow (\overline{c_n}, \sigma_n)\allowbreak \Matchiiarrow \sigma$.
+    The $n$ first transitions must be \textsf{Decompose} or \textsf{DecomposeX},
+    and the last transition must be \textsf{Success}.
+
+    We show that there exists a derivation of the form
+    $([t], [\,],\allowbreak D, \emptyset) = \quadruple{}_{\,\,0}\allowbreak \Pdtarrow \cdots
+    \Pdtarrow \quadruple{}_{\,\,n} \Pdtarrow ((s, d),\sigma)$, where $\quadruple{}_{\,\,i} =
+    (\overline{t_i}, \overline{b_i}, \allowbreak D_i, \sigma_i)$ for each $i$.
+  %
+    We define $\overline{t_i}$, $\overline{b_i}$, and $D_i$ as follows, for $i > 0$.
+    The list $\overline{t_i}$ consists of the right-hand sides of the
+    constraints~$\overline{c_i}$, in the same order.
+    %todo:
+    Let $\mathit{hd}$ be the function that extracts the head of a list. We set
+          $b_i = (\mathit{hd}(\overline{t_{i-1}}),\allowbreak D_{i-1}, \sigma_{i-1})$. We know that $\overline{c_{i-1}}$ is
+          nonempty, since there exists a transition $(\overline{c_{i-1}}, \sigma_{i-1}) \Matchiiarrow (\overline{c_i}, \sigma_i)$%
+          ; thus, $\overline{t_{i-1}}$ is nonempty.
+    If an accepting node storing~$s$ was reached in $n$~steps, the
+      serialization of $s$ must be of the form $\zeta_1.\cdots.\zeta_n$. Take $D_i
+      = D_{i-1}{|}_{\zeta_i}$.
+
+    The sequence of states $\quadruple{}_{\,\,i}$ forms a
+    derivation:
+    If $(\overline{c_i},\allowbreak \sigma_i)\allowbreak \Matchiiarrow_\textsf{Decompose}
+    (\tuple{c}{{i+1}},\allowbreak \sigma_{i+1})$, then $\quadruple{}_{\,\,i}
+    \Pdtarrow_\textsf{AdvanceF} \allowbreak \quadruple{}_{\,\,i+1}$.
+    If $(\overline{c_i},\allowbreak \sigma_i) \Matchiiarrow_\textsf{DecomposeX}
+    (\tuple{c}{{i+1}},\allowbreak \sigma_{i+1})$, then $\quadruple{}_{\,\,i}
+    \Pdtarrow_\textsf{AdvanceX} \quadruple{}_{\,\,i+1}$.
+    If $(\overline{c_n},\allowbreak \sigma_n) \Matchiiarrow_\textsf{Success} \sigma$, then
+    $\quadruple{}_{\,\,n} \Pdtarrow_\textsf{Success} ((s, d),\allowbreak \sigma)$.
+\end{proof}
+
+\begin{lemma}\label{lemma:ehoh:pdt-soundi}
+  If $([t], [\,], D, \emptyset) \Pdtarrow^+ ((s, d), \sigma)$, then $s \in \Term(D)$
+  and $\sigma$ is the MGG of $s \MATCH t$.
+  \end{lemma}
+
+  \begin{proof}
+  Let $([t], [\,], D, \emptyset) =
+  \quadruple{}_{\,\,0} \Pdtarrow \cdots \Pdtarrow \quadruple{}_{\,\,n}
+  \Pdtarrow ((s, d), \sigma)$ be a derivation, where $\quadruple{}_{\,\,i} =
+  (\overline{t_i}, \overline{b_i},\allowbreak D_i, \sigma_i)$ for each $i$. Without loss of
+  generality, by Lemma \ref{lemma:ehoh:backtrack-remove}, we can assume that the
+  derivation contains no \textsf{Backtrack} transitions.
+
+  The first conjunct, $s \in \Term(D)$, clearly holds
+  %by definition --- some invariant must be needed here
+  for any term found
+  from an initial state. To prove the second conjunct, we first introduce a
+  function $\mathit{preord}$ that defines the preorder decomposition of a list
+  of terms:
+  $\mathit{preord}([\,]) = [\,]$ and
+  $\mathit{preord}(\zeta \; \tuple{s}{n} \cdot \overline{xs}) =
+  (\zeta{,}\; \tuple{s}{n} \cdot \overline{xs}) \cdot \mathit{preord}(\tuple{s}{n} \cdot \overline{xs})$.
+  %
+  Given a term~$s$, $\mathit{preord}([s])$ gives a sequence
+  $(\zeta_1, \tuple{u}{1}), \dots, \allowbreak
+  (\zeta_n, \tuple{u}{n})$.
+  Since $s \in \Term(D)$,
+  the sequence $D_0, \ldots, D_n$ follows the preorder serialization of $s$:
+  $D_i = D_{i-1}|_{\zeta_i}$ for $i > 0$.
+
+  Next, we show that there exists a derivation of the form
+  $([s \MATCH t], \emptyset) = \mathcal{S}_0 \Matchiiarrow
+  \cdots \Matchiiarrow \mathcal{S}_n \Matchiiarrow \sigma$, where
+  $\mathcal{S}_i = (\overline{c_i}, \sigma_i)$. We define $\overline{c_i}$, for $i>0$,
+  as the list of constraints whose left-hand sides are the elements of $\overline{u_i}$
+  and right-hand sides are the elements of $\overline{t_i}$, in the order they appear in the respective lists.
+  By inspecting the definition of $\mathit{preord}$
+  and the changes each \textsf{Advance} step makes
+  to the head of $\overline{t_i}$, we can see that $\overline{u_i}$ and $\overline{t_i}$  have the same length.
+  The sequence of states $\mathcal{S}_i$ forms a derivation:
+  %
+  If $\quadruple{}_{\,\,i} \Pdtarrow_\textsf{AdvanceF} \quadruple{}_{\,\,i+1}$, then $\mathcal{S}_i \Matchiiarrow_\textsf{Decompose} \mathcal{S}_{i+1}$.
+  If $\quadruple{}_{\,\,i} \Pdtarrow_\textsf{AdvanceX} \quadruple{}_{\,\,i+1}$, then $\mathcal{S}_i\allowbreak \Matchiiarrow_\textsf{DecomposeX} \mathcal{S}_{i+1}$.
+  If $\quadruple{}_{\,\,n} \Pdtarrow_\textsf{Success} \sigma$, then
+  $\mathcal{S}_n\allowbreak \Matchiiarrow_\textsf{Success} \sigma$.
+\end{proof}
+
+\begin{theorem}[Total Correctness]\label{theorem:ehoh:pdt-complete}\,%
+  Let $D$ be a perfect discrimination tree and $t$ be a term.
+  The sets $\{ s \in \Term(D) \mid \exists \sigma.\; \sigma(s) = t \}$
+  and $\{ s \mid \exists d, \sigma.\; ([t], [\,], D, \emptyset) \allowbreak
+  \Pdtarrow^! ((s, d), \sigma) \}$ are equal.
+\end{theorem}
+\begin{proof}
+  This follows from Lemmas \ref{lemma:ehoh:pdt-complete-help} and \ref{lemma:ehoh:pdt-soundi}.
+\end{proof}
+
+The theorem tells us that given a term~$t$, all generalizations $s$ stored
+in the perfect discrimination tree can be found, but it does not exclude
+nondeterminism. Often, both \textsf{AdvanceF} and \textsf{AdvanceX} are
+applicable. To find all generalizations, we need to follow both transitions.
+But for some applications, it is enough to find a single generalization.
+
+To cater for both types of applications, E provides iterators that store the
+state of a traversal. After an iterator is initialized
+with the root node $D$ and the query term $t$, each call to
+\textsc{FindNextVal} will move the iterator to the next node that generalizes
+the query term and stores a value, indicating an accepting node. After all
+such nodes have been traversed, the iterator is set to point to
+$\mathit{Null}$.
+
+The following definitions constitute the high-level interface for
+iterating through values incrementally or for obtaining all values of nodes that
+store generalizations of the query term in $D$.
+
+\begin{quotex}
+  \MyFunction{InitIter}{PDTNode $D$, Term $t$}
+  \q $i \gets \textsc{Iterator}()$ \\
+  \q $(i.\mathit{node}, i.\mathit{t\_stack}, i.\mathit{t\_proc}, i.\mathit{c\_iter}) \gets (D, [t], [\,], \textit{Start})$ \\
+  \q \MyReturn $i$
+
+  \vskip 1.5\jot
+
+  \MyProcedure{FindNextVal}{Iterator $i$}
+  \q \MyDo
+  \qq \textsc{FindNextNode}($i$) \\
+  \q \MyWhileOfDo{$i.\mathit{node} \not= \mathit{Null} \mathrel\land
+    \\ \qqq  (\lnot\, i.\mathit{t\_stack}.\mathit{isEmpty}()
+    \mathrel\lor \lnot\, i.\mathit{node}.\mathit{has\_val}())$}
+
+  \vskip 1.5\jot
+
+  \MyFunction{AllVals}{PDTNode $D$, Term $t$}
+  \q $\mathit{i} \gets \textsc{InitIter}(D, t)$ \\
+  \q $\textsc{FindNextVal}(i)$ \\
+  \q $\mathit{res} \gets \emptyset$ \\
+  \q \MyWhile{$\mathit{i.node} \neq \mathit{Null}$}
+  \qq $\mathit{res} \gets \mathit{res} \cup \{ \mathit{i.node.val()} \}$ \\
+  \qq $\textsc{FindNextVal}(i)$
+  \\[\jot]
+  \q \MyReturn $\mathit{res}$
+  \end{quotex}
+
+  The core functionality is implemented in \textsc{FindNext\-Node}, presented below. This
+  procedure moves the iterator to the next node that has not been explored in the
+  search for generalization, or \textit{Null} if the entire tree has been
+  traversed. It first goes through all child nodes labeled with a variable before
+  possibly visiting the child node labeled with a function symbol. We assume that
+  we can iterate through the children of a node using a function
+  \textsc{NextVarChild} that, given a tree node and iterator through children,
+  advances the iterator to the child representing the next variable. Furthermore, we
+  assume that the iterator can also be in the distinguished states \textit{Start}
+  and \textit{End}.
+  \textit{Start} indicates that no child has been visited yet; \textit{End}
+  indicates that we have visited all children. Finally, the expression
+  $n.\mathit{child}(\zeta)$ returns a child of the node $n$ labeled $\zeta$ if
+  such a child exists or $\mathit{Null}$ otherwise.
+
+  \begin{quotex}
+  \MyProcedure{FindNextNode}{Iterator $i$}
+  \q \MyIf{$i.\mathit{t\_stack.isEmpty}()$}
+  \qq \textsc{BacktrackToVar}($i$)
+  \\[\jot]
+  \q $\mathit{advanced} \gets \mathit{False}$ \\
+  \q \MyWhile{$i.\mathit{node} \neq \mathit{Null} \mathrel\land \lnot\, \mathit{advanced}$}
+  %\qq \MyIf{$i.\mathit{c\_iter} < \textit{End}$}
+  \qq \MyWhile{$i.\mathit{c\_iter} \neq \textit{End} \mathrel\land \lnot\, \mathit{advanced}$}
+  \qqq $i.\mathit{c\_iter} \gets \textsc{NextVarChild}(i.\mathit{node}, i.\mathit{c\_iter})$ \\
+  \qqq \MyIf{$i.\mathit{c\_iter} \neq \textit{End}$}
+  \qqqq $x \gets i.\mathit{c\_iter}.\mathit{var}()$ \\
+  \qqqq $t \gets i.\mathit{t\_stack}.\mathit{top}()$ \\
+  \qqqq $\mathit{s} \gets \textsc{GobblePrefix}(x, t)$ \\
+  \qqqq \MyIf{$\mathit{s} \neq \mathit{Null} \mathrel\land \\ \qqqqqq
+                (x.\mathit{binding} = \mathit{Null} \mathrel\lor x.\mathit{binding} = \mathit{s})$}
+  \qqqqq $i.\mathit{t\_stack}.\mathit{pop}()$ \\
+  \qqqqq \MyForDownto{$j \gets t.\mathit{num\_args}$\\\qqqqqqq\ignorespaces}{$\mathit{s}.\mathit{num\_args}+1$}
+  \qqqqqq $i.\mathit{t\_stack}.\mathit{push}(t.\mathit{args}[j])$ \\
+  \qqqqq \MyIf{$x.\mathit{binding} = \mathit{Null}$}
+  \qqqqqq $x.\mathit{binding} \gets \mathit{s}$ \\
+  \qqqqqq $i.\mathit{t\_proc}.\mathit{push}
+                      ( (t, i.\mathit{node}, i.\mathit{c\_iter}, \mathit{True} ) )$ \\
+  \qqqqq \MyElse
+  \qqqqqq $i.\mathit{t\_proc}.\mathit{push}
+                      ( (t, i.\mathit{node}, i.\mathit{c\_iter}, \mathit{False}) )$
+  \\[\jot]
+  \qqqqq $ \mathit{i.\mathit{node}} \gets i.\mathit{node}.child(x)$ \\
+  \qqqqq $ \mathit{advanced} \gets \mathit{True}$
+  \\[\jot]
+  \qq $t \gets i.\mathit{t\_stack}.\mathit{top}()$ \\
+  \qq \MyIf{$i.\mathit{c\_iter} = \textit{End} \mathrel\land
+          \lnot\, t.\mathit{head}.\mathit{isVar}()  \\ \qqq \mathrel\land
+          D.\mathit{child}(t.\mathit{head}) \neq \mathit{Null}$}
+  \qqq $i.\mathit{t\_stack}.\mathit{pop}()$ \\
+  \qqq \MyForDownto{$j \gets t.\mathit{num\_args}$}{$1$}
+  \qqqq $i.\mathit{t\_stack}.\mathit{push}(t.\mathit{args}[j])$ \\
+  \qqq $i.\mathit{t\_proc}.\mathit{push}
+              ( (t, i.\mathit{node}, \mathit{End}, \mathit{False}) )$ \\
+  \qqq $\mathit{i.\mathit{node}} \gets i.\mathit{node}.\mathit{child}(t.\mathit{head})$ \\
+  \qqq $\mathit{advanced} \gets \mathit{True}$
+  \\[\jot]
+  \qq \MyIf{$\lnot\, \mathit{advanced}$}
+  \qqq \textsc{BacktrackToVar}$(i)$ \\
+  \qq \MyElse
+  \qqq $i.\mathit{c\_iter} \gets \textit{Start}$
+
+  \vskip 2\jot
+
+  \MyProcedure{BacktrackToVar}{Iterator $i$}
+  \q \MyForever
+  \qq \MyIf{$i.\mathit{t\_proc}.\mathit{isEmpty}()$}
+  \qqq $i.\mathit{node} \gets \mathit{Null}$ \\
+  \qqq \MyReturn \\
+  \qq \MyElse
+  \qqq $(t, D, \mathit{c\_iter}, \mathit{var\_unbound}) \gets i.\mathit{t\_proc}.\mathit{pop}()$ \\
+  \qqq $\mathit{label\_arity} \gets i.\mathit{node}.\mathit{label}.\mathit{type}.\mathit{arity}$ \\
+  \qqq $\mathit{t\_arity} \gets t.\mathit{type}.\mathit{arity}$
+  \\[\jot]
+  \qqq \MyForTo{$i \gets 1$}{$\mathit{label\_arity} - \mathit{t\_arity}$}
+  \qqqq $i.\mathit{t\_stack}.\mathit{pop}()$
+  \\[\jot]
+  \qqq $i.\mathit{t\_stack}.\mathit{push}(t)$ \\
+  \qqq $i.\mathit{node} \gets D$ \\
+  \qqq $i.\mathit{c\_iter} \gets \mathit{c\_iter}$
+  \\[\jot]
+  \qqq \MyIf{$\mathit{var\_unbound}$}
+  \qqqq $i.\mathit{node}.\mathit{label}.\mathit{binding} \gets \mathit{Null}$
+  \\[\jot]
+  \qqq \MyIf{$\mathit{c\_iter} \neq\textit{End}$}
+  \qqqq \MyReturn
+  \end{quotex}
+
+The pseudocode uses a slightly different representation of backtracking tuples
+than ${\Pdtarrow}$. In the \textsf{AdvanceX} rule,
+$\sigma$ changes only if the variable~$x$ was previously not bound.
+Instead of creating and storing substitutions explicitly, we simply remember
+whether the variable was bound in this step or not, in the
+$\mathit{var\_unbound}$ tuple component. Then we rely on the label~$x$ of the
+current node and its \textit{binding} field to carry the substitutions.
+Similarly, since our strategy is to traverse the tree by first visiting the
+variable-labeled child nodes, we need to remember how far we have come with
+this traversal. We store this information in the $\mathit{c\_iter}$ tuple
+component.
+
+\subsection{Fingerprint Indices}
+\looseness=-1
+Fingerprint indices \cite{ss-12-fp-indexing} trade perfect indexing for
+a compact memory representation and more flexible retrieval conditions.
+The basic idea is to compare terms by looking only at a few
+predefined sample positions. If we know that term $s$ has
+symbol $\cst{f}$ at the head of the subterm at~2.1 and term $t$ has
+$\cst{g}$ at the same position, we can immediately conclude that $s$ and $t$
+are not unifiable.
+
+% They are
+% based on the idea that when one samples a term position only the following
+% situations may happen: position corresponds to
+% a variable or a complex term; position is invalid---but under some substitution it
+% can become valid; position is invalid and no substitution makes it valid.
+
+Let $\AA$ (``at a variable''),
+$\BB$ (``below a variable''), and
+$\NN$ (``nonexistent'') be distinguished symbols\begin{rep}
+not present in the signature, and let $q < p$ denote that position $q$ is a proper prefix of $p$\end{rep}
+(e.g., $\varepsilon < 2 < 2.1$). Given a term $t$ and a position~$p$, the
+\emph{fingerprint function} $\Gfpf$ is defined as
+\[
+  \Gfpf(t,p) =
+  \begin{cases}
+    \cst{f} & \text{if $t|_p$ has a symbol head $\cst{f}$} \\[-\jot]
+    \AA & \text{if $t|_p$ is a variable} \\[-\jot]
+    \BB & \text{if $t|_q$ is a variable for some $q < p$} \\[-\jot]
+    \NN & \text{otherwise}
+  \end{cases}
+\]
+
+Based on a fixed tuple of positions $\tuple{p}{n}$,
+the \emph{fingerprint} of a term $t$ is defined as
+$\mathit\Fp(t) = \bigl(\Gfpf(t, p_1), \ldots, \Gfpf(t, p_n)\bigr).$ To compare
+two terms $s$ and $t$, it suffices to check that their fingerprints are
+componentwise compatible using the following unification and matching
+matrices, respectively:
+%
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\confrep{
+\begin{align*}
+& \hbox{\begin{tabular}{@{}C{0.75em}|C{0.75em}C{0.75em}C{0.75em}C{0.75em}C{0.75em}@{}}
+& $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+\hline
+$\cst{f}_1\vphantom{^{()}}$ & & \xmark & & & \xmark \\
+\AA & & & & & \xmark \\
+\BB & & & & & \\
+\NN & \xmark & \xmark & \xmark & &
+\end{tabular}}
+&
+& \hbox{\begin{tabular}{@{}C{0.75em}|C{0.75em}C{0.75em}C{0.75em}C{0.75em}C{0.75em}@{}}
+& $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+\hline
+$\cst{f}_1\vphantom{^{()}}$ & & \xmark & \xmark & \xmark & \xmark \\
+\AA & & & & \xmark & \xmark \\
+\BB & & & & & \\
+\NN & \xmark & \xmark & \xmark & \xmark &
+\end{tabular}}
+\end{align*}
+}{
+\begin{align*}
+& \hbox{\begin{tabular}{@{}C{0.75em}|C{0.75em}C{0.75em}C{0.75em}C{0.75em}C{0.75em}@{}}
+& $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+\hline
+$\cst{f}_1\vphantom{^{()}}$ & & \xmark & & & \xmark \\
+$\cst{f}_2$ & \xmark & & & & \xmark \\
+\AA & & & & & \xmark \\
+\BB & & & & & \\
+\NN & \xmark & \xmark & \xmark & &
+\end{tabular}}
+&
+& \hbox{\begin{tabular}{@{}C{0.75em}|C{0.75em}C{0.75em}C{0.75em}C{0.75em}C{0.75em}@{}}
+& $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+\hline
+$\cst{f}_1\vphantom{^{()}}$ & & \xmark & \xmark & \xmark & \xmark \\
+$\cst{f}_2$ & \xmark & & \xmark & \xmark & \xmark \\
+\AA & & & & \xmark & \xmark \\
+\BB & & & & & \\
+\NN & \xmark & \xmark & \xmark & \xmark &
+\end{tabular}}
+\end{align*}
+}%
+%
+The rows and columns correspond to $s$ and $t$, respectively. The
+metavariables $\cst{f}_1$~and~$\cst{f}_2$ represent arbitrary distinct symbols.
+Incompatibility is indicated by $\xmark{}$.
+
+\newcommand\fpok{\text{--}}
+
+As an example,
+let $(\varepsilon, 1,\allowbreak 2,\allowbreak 1.1,\allowbreak 1.2,\allowbreak 2.1, 2.2)$
+be the sample positions,
+and let $s = \cst{f}(\cst{a}, x)$ and $t = \cst{f}(\cst{g}(x), \cst{g}(\cst{a}))$
+be the terms to unify. Their fingerprints are
+$\Fp(s) = (\cst{f}, \cst{a}, \AA, \NN, \NN,\allowbreak \BB, \BB)$
+and
+$\Fp(t) = (\cst{f}, \cst{g}, \cst{g}, \AA, \NN, \cst{a}, \NN)$.
+%
+Using the left matrix, we compute the compatibility vector
+%
+$(\fpok, \xmark, \fpok, \xmark, \fpok, \fpok, \fpok).$
+%
+The mismatches at positions 1 and 1.1 indicate that $s$ and $t$ are not unifiable.
+
+A fingerprint index is a trie that stores a term set $T$ keyed by fingerprint.
+The term $\cst{f}(\cst{g}(x), \cst{g}(\cst{a}))$ above would
+be stored in the node addressed by
+$\cst{f}.\cst{g}.\cst{g}.\AA.\NN.\cst{a}.\NN$, together with other
+terms that share the same fingerprint.
+This scheme makes it possible to unify or match a query term $s$
+against all the terms $T$ in one traversal. Once a node storing the terms $U
+\subseteq T$ has been reached, due to overapproximation we must apply
+unification or matching on $s$ and each $u \in U.$ % to obtain correct results.
+
+When adapting this data structure to \lfhol{}, we must first choose a
+suitable notion of position in a term. Conventionally, higher-order
+positions are strings over $\{1, 2\}$, but this is not graceful.
+% indicating, for each binary
+% application $t_1\;t_2$, which term $t_i$ to follow.
+Instead, it is preferable to generalize the first-order notion to
+flattened \lfhol{} terms---e.g.,
+$x\> \cst{a} \> \cst{b} \> |_{1} = \cst{a}$
+and $x\> \cst{a} \> \cst{b} \> |_{2} = \cst{b}.$
+%
+However, this approach fails on applied variables. For example,
+although $x \> \cst{b}$ and $\cst{f} \> \cst{a} \> \cst{b}$ are
+unifiable (using $\{ x \mapsto \cst{f} \> \cst{a} \}$), sampling position 1
+would yield a clash between
+$\cst{b}$~and~$\cst{a}.$ To ensure that positions remain stable under
+substitution, we propose to number arguments in reverse:
+$t {|}^\varepsilon = t$ and
+$\zeta \> t_n \, \ldots \, t_1  |^{i.p} = t_{i} |^p$
+if $1 \le i \le n$.
+We use a nonstandard notation, $t|^p$, for this nonstandard
+notion. The operation is undefined for out-of-bound indices.
+
+\begin{lemma}\label{lemma:fpindex-aux}
+  Let $s$ and $t$ be unifiable terms, and let $p$ be a position such that the
+  subterms $s|^{p}$ and $t|^{p}$ are defined. Then $s|^{p}$ and $t|^{p}$ are
+  unifiable.
+\end{lemma}
+\begin{proof}
+  By structural induction on $p$. The case $p = \varepsilon$ is trivial.
+
+  \medskip
+
+  \noindent
+  \textsc{Case $p = q.i$}:\enskip
+  Let $s|^{q} = \zeta \; s_m \, \ldots \, s_1$ and $t|^{q} = \eta \; t_n \, \ldots \, t_1$.
+  Since $p$ is defined in both $s$ and $t$, we have
+   $s|^{p} = s_i$ and $t|^{p} = t_i$.
+  By the induction hypothesis, $s|^{q}$ and $t|^{q}$ are
+  unifiable, meaning that there exists a substitution $\sigma$
+  such that $\sigma(\zeta \; s_m \, \ldots \, s_1) = \sigma(\eta \; t_n \, \ldots \, t_1)$.
+  Hence, $\sigma(s_1) = \sigma(t_1)$, \ldots, $\sigma(s_i) = \sigma(t_i)$---i.e.,
+  $\sigma(s|^{p}) = \sigma(t|^{p})$.
+\end{proof}
+
+Let $t{\langle}{}^p$ denote the subterm $t{|}^q$ such that $q$ is the longest
+prefix of $p$ for which $t{|}^q$ is defined.
+%
+The \lfhol{} version of the fingerprint function is defined as follows:
+%
+\[
+  \GfpfRTL(t,p) =
+  \begin{cases}
+    \cst{f} & \text{if $t|^p$ has a symbol head $\cst{f}$} \\
+    \AA & \text{if $t|^p$ has a variable head} \\
+    \BB & \text{if $t|^p$ is undefined} \\[-\jot]
+        & \text{but $t{\langle}{}^p$ has a variable head} \\
+    \NN & \text{otherwise}
+  \end{cases}
+\]
+%
+Except for the reversed numbering scheme,
+$\GfpfRTL$ coincides with $\Gfpf$ on first-order terms.
+The fingerprint $\FpRTL(t)$ of a term $t$ is defined analogously as before,
+and the same compatibility matrices can be used.
+
+The key difference between $\Gfpf$ and $\GfpfRTL$ concerns applied variable.
+Given the sample positions $(\varepsilon, 2, 1)$,
+the fingerprint of $x$ is $(\AA, \BB, \BB)$ as before, whereas
+the fingerprint of $x \; \cst{c}$ is $(\AA, \BB, \cst{c}).$
+%
+As another example,
+let $(\varepsilon, 2,\allowbreak 1,\allowbreak 2.2,\allowbreak 2.1,\allowbreak 1.2, 1.1)$
+be the sample positions,
+and let $s = x \; (\cst{f} \; \cst{b} \; \cst{c})$ and $t = \cst{g}\; \cst{a}\; (y \; \cst{d}).$
+Their fingerprints are
+$\FpRTL(s) = (\AA, \BB, \cst{f}, \BB, \BB, \cst{b}, \cst{c})$ and
+$\FpRTL(t) = (\cst{g}, \cst{a}, \AA, \NN, \NN, \BB, \cst{d})$.
+%
+The terms are not unifiable due to the incompatibility at position 1.1
+($\cst{c}$ vs.~$\cst{d}$).
+
+We can easily support prefix optimization for both terms $s$ and $t$ being
+compared: We simply add enough fresh variables as arguments to ensure that $s$
+and $t$ are fully applied before computing their fingerprints.
+
+\begin{lemma}\label{lemma:ehoh:fpindex-gfpf-sound-unif}
+  If terms $s$ and $t$ are unifiable, then $\GfpfRTL(s, p)$ and $\GfpfRTL(t,
+  p)$ are compatible according to the unification matrix.
+  If $s$ generalizes $t$, then $\GfpfRTL(s, p)$ and $\GfpfRTL(t, p)$ are
+  compatible according to the matching matrix.
+\end{lemma}
+\begin{proof}
+  We focus on the case of unification. By contraposition, it suffices to
+  consider the eight blank cells in the unification matrix, where the rows
+  correspond to $\GfpfRTL(s, p)$ and the columns correspond to $\GfpfRTL(t,
+  p)$. Since unifiability is a symmetric relation, we can rule out four cases.
+
+\medskip
+
+\noindent
+\textsc{Case} $\cst{f}_1$--$\cst{f}_2$:\enskip
+By definition of $\GfpfRTL$, $s|^{p}$ and $t|^{p}$ must be of the forms
+$\cst{f}_1 \; \overline{s}$ and $\cst{f}_2 \; \overline{t}$, respectively.
+Clearly, $s|^{p}$ and $t|^{p}$ are not unifiable. By Lemma
+\ref{lemma:fpindex-aux}, $s$ and $t$ are not unifiable.
+
+\medskip
+
+\noindent
+\textsc{Case} $\cst{f}_1$--$\NN$, $\cst{f}_2$--$\NN$, \textsc{or} $\AA$--$\NN$:\enskip
+From $\GfpfRTL(t, p) = \NN$, we deduce that
+$p \neq \varepsilon$.
+Let $p = q.i.r$, where $q$ is the longest prefix such that $\GfpfRTL(t, q) \neq \NN$.
+Since $\GfpfRTL(t, q.i) = \NN$, the head of $t|^{q}$ must be some symbol $\cst{g}$.
+(For a variable head, we would have $\GfpfRTL(t, q.i) = \BB$.)
+Hence, $t|^{q}$ has the form $\cst{g} \; t_n \, \ldots \, t_1$, for $n < i$.
+Since $q.i$ is a legal position in $s$, $s|^{q}$ has the form
+$\zeta \; s_m \, \ldots \, s_1$, with $i \le m$. A necessary condition for
+$\sigma(s|^{q}) = \sigma(t|^{q})$ is that
+$\sigma(\zeta \; s_m \, \ldots \, s_{n+1}) = \sigma(\cst{g})$,
+but this is impossible because the left-hand side is an application (since
+$n < m$), whereas the right-hand side is the symbol $\cst{g}$.
+By Lemma \ref{lemma:fpindex-aux}, $s$ and $t$ are not unifiable.
+\end{proof}
+
+\begin{corollary}[Overapproximation]\,%
+  If $s$ and $t$ are unifiable terms, then $\FpRTL(s)$ and $\FpRTL(t)$ are
+  compatible according to the unification matrix.
+  If $s$ generalizes $t$, then $\FpRTL(s)$ and $\FpRTL(t)$ are
+  compatible according to the matching matrix.
+\end{corollary}
+%\begin{proof}
+%This follows from Lemma~\ref{lemma:ehoh:fpindex-gfpf-sound-unif}.\qed
+%\end{proof}
+
+% \subsection{Feature Vector Indices}
+
+% A clause $C$
+% subsumes a clause $D$ if there exists a substitution $\sigma$ such that
+% $\sigma(C) \subseteq D.$
+% Subsumption is a crucial operation to prune the search space.
+%  Feature-vector indices
+% \cite{ss-2013-feature-vector} are an imperfect indexing data structure
+% that can be used to retrieve clauses that subsume a query clause
+% % (forward subsumption)
+% or that are subsumed by the query clause.\
+% %
+% Unlike for discrimination trees and fingerprint indices, no changes were
+% necessary to adapt feature vectors indices to \lfhol{}. All the predefined
+% features make sense in \lfhol{}. % and are compatible with subsumption.
+
 \section{Inference Rules}

### 2022-02-23 ehoh/ehoh.tex
more

@@
 
+Saturating provers show the unsatisfiability of a clause set
+by systematically adding logical consequences,
+% (up to simplification and redundancy),
+eventually deriving the empty clause
+as a witness of unsatisfiability. They implement two kinds of
+inference rules: \emph{Generating rules} produce new clauses and are
+needed for completeness, whereas \emph{simplification rules} delete
+existing clauses or replace them by simpler clauses. This
+simplification is crucial for success, and most modern provers spend a
+large part of their time on simplification. E's main loop, which applies the
+rules, implements the \relax{given clause} procedure, as described in
+Sect.~\ref{sec:pre:saturation}.
+
+% The proof state is represented by two disjoint
+% subsets of clauses, the set of \emph{processed} clauses $P$ and the set of
+% \emph{unprocessed} clauses $U$. Initially, all clauses are unprocessed. At each
+% iteration of the loop, the prover heuristically selects a \emph{given clause}
+% from $U$, adds it to $P$, and performs all generating inferences between this
+% clause and all clauses in $P$. Resulting new clauses are added to $U$. This
+% maintains the invariant that all direct consequences between clauses in $P$ have
+% been performed. Simplification is performed on the given clause (using clauses
+% in $P$ as side premises), on clauses in $P$ (using the given clause), and on
+% newly generated clauses (again, using $P$).
+
+Ehoh is based on the same logical calculus as E, except that it is
+generalized to \lfhol{} terms. The standard inference rules and completeness
+proof of superposition with respect to intensional Boolean-free \lfhol{} fragment of our
+logic can be reused verbatim; the only changes concern the basic definitions of
+terms and substitutions \cite[Sect.~1]{bbcw-21-lfho}.
+\begin{rep}Refutational completeness of superposition for \lfhol{} terms has
+been formally proved by Peltier \cite{np-2016-sup-formalisation} using Isabelle.\end{rep}
+%
+We introduced support for first-class Boolean terms in Ehoh by extending the
+preprocessor, as explained in Sect.~\ref{sec:ehoh:preprocessing}.
+
+\ourpara{The Generating Rules}
+
+The superposition rules were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
+we repeat them with slightly simplified notation, as we do not repeat side conditions:
+
+\hfill
+\begin{tabular}{ll}
+  \namedinference{SN}
+  {s \eq t \mathrel\vee C \qquad u[s'] \noteq v \mathrel\vee D}
+  {\sigma(u[t] \mathbin{\noteq} v \mathrel\vee C \mathrel\vee D)}
+  &
+  \namedinference{ER}{s \noteq s' \mathrel\vee C}{\sigma(C)} \\[2\jot]
+  \namedinference{SP}
+    {s \eq t \mathrel\vee C \qquad u[s'] \eq v \mathrel\vee D}
+    {\sigma(u[t] \mathbin{\eq} v \mathrel\vee C \mathrel\vee D)}
+  &
+  \namedinference{EF}{s \eq t \mathrel\vee  s' \eq u \mathrel\vee C}{\sigma(t \noteq u \mathrel\vee  s \eq u \mathrel\vee C)}
+\end{tabular}
+\hfill
+\kern\belowdisplayskip
+
+\noindent
+In each rule, $\sigma$ denotes the MGU of $s$ and $s'$.
+
+Equality resolution (ER) and equality factoring (EF) \begin{rep}are single-premise rules
+that \end{rep}work on the entire left- or right-hand side of a literal of
+the given clause. To generalize them, it suffices to disable prefix
+optimization for unification.
+\confrep{By contrast, t}{\par T}%
+he rules for superposition into negative and positive literals (SN~and~SP) are more
+complex. As two-premise rules, they require the prover to find a partner
+for the given clause. There are two cases to consider\confrep{.}{, depending on whether
+the given clause acts as the first or second premise in an inference.
+Moreover, since the rules operate on subterms~$s'$ of a clause, the prover
+must be able to efficiently locate all relevant subterms, including
+\lfhol{} prefix subterms.}
+%
+To cover the case where the given clause acts as the left premise, the prover
+relies on a fingerprint index to compute a set of clauses containing terms
+possibly unifiable with a side $s$ of a positive literal of the given clause.
+Thanks to our generalization of fingerprints, in Ehoh this candidate set is
+guaranteed to overapproximate the set of all possible inference partners. The
+unification algorithm is then applied to filter out unsuitable candidates.
+Thanks to prefix optimization, we can avoid polluting the
+index with all prefix subterms.
+
+When the given clause is the right premise, the prover traverses
+its subterms $s'$ looking for inference partners in another fingerprint index,
+which contains only entire left-~and right-hand sides of equalities. Like E,
+Ehoh traverses subterms in a first-order fashion. If prefix unification
+succeeds, Ehoh determines the unified prefix and applies the appropriate
+inference instance.
+
+\ourpara{The Simplifying Rules}
+
+Unlike generating rules, simplifying rules do not necessarily add
+conclusions to the proof state---they can also remove premises. E
+implements over a dozen simplifying rules, with unconditional
+rewriting and clause subsumption as the most significant
+examples. Here, we restrict our attention to a single rule, which best illustrates
+the challenges of supporting \lfhol:
+%
+\[\namedsimp{ES}{s \eq t \qquad u[\sigma(s)] \eq u[\sigma(t)] \mathrel\vee C}{s \eq t}\]
+%
+Given an equation $s \eq t$, equality subsumption (\infname{ES}) removes a clause
+containing a literal whose two sides are equal except that an instance of $s$
+appears on one side where the corresponding instance of $t$ appears on the
+other side.
+
+E maintains a perfect discrimination tree storing clauses of the form $s
+\eq t$ indexed by $s$ and $t$.
+%Each node corresponding to an indexed term
+%contains the pointer to unit clause it appears in.
+When applying \infname{ES},
+E considers each \begin{rep}positive \end{rep}literal $u \eq v$ of the given
+clause in turn. It starts by taking the left-hand side~$u$ as a query term.
+If an equation $s \eq t$ (or $t \eq s$) is found in the tree, with $\sigma(s)
+= u$, the prover checks whether $\sigma'(t) = v$ for some \confrep{}{(possibly
+nonstrict) }extension $\sigma'$ of~$\sigma$. If so, ES is
+applicable\confrep{}{, with a second premise of the form $\sigma(s) \eq
+\sigma(t) \mathrel\vee C$}.
+\begin{rep}\par\end{rep}
+To consider nonempty contexts, the prover traverses the subterms $u'$ and $v'$ of
+$u$ and $v$ in lockstep, as long as they appear under identical
+contexts. Thanks to prefix optimization, when Ehoh is given a subterm $u'$,
+it can find an equation $s \eq t$ in the tree such that
+$\sigma(s)$ is equal to some prefix of $u'$, with some arguments
+$\overline{u}$ remaining as unmatched. Checking for equality subsumption
+%
+then amounts to checking that $v' = \sigma'(t) \; \overline{u}$,
+for some extension $\sigma'$ of $\sigma$.
+
+For example, let
+$\cst{f} \; (\cst{g} \; \cst{a} \; \cst{b}) \eq \cst{f} \; (\cst{h} \; \cst{g} \; \cst{b})$
+be the given clause,
+and suppose that $x \; \cst{a} \eq \cst{h} \; x$ is indexed.
+Under context $\cst{f}\>[\phantom{.}]$, Ehoh considers the subterms
+$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; x \; \cst{b}$. It finds the
+prefix $\cst{g} \; \cst{a}$ of $\cst{g} \; \cst{a} \; \cst{b}$ in the
+%
+tree, with $\sigma = \{x \mapsto\nobreak \cst{g}\}$.
+The prefix $\cst{h} \; \cst{g}$ of $\cst{h} \; \cst{g} \; \cst{b}$ matches the
+indexed equation's right-hand side $\cst{h} \; x$ using the same substitution,
+and the remaining argument in both subterms,~$\cst{b}$, is identical.
+\begin{rep}Ehoh concludes that the given clause is redundant.\end{rep}
+
+\ourpara{Pragmatic Extensions}
+
+Since Ehoh is based on a mono\-morphic logic, the only way to support
+extensionality without changing the calculus is to add a set of extensionality
+axioms for every function type occurring in the problem
+\cite[Sect.~3.1]{bbcw-21-lfho}. The evaluation by Bentkamp et al.\ of such
+an approach was discouraging \cite[Sect.~6]{bbcw-21-lfho}, so we decided to
+support extensionality via inference rules in Ehoh. We implemented two well-known
+incomplete rules we had experimented with in the context of Zipperposition.
+
+The negative and positive extensionality (\infname{NE} and \infname{PE}) rules
+are defined as
+
+\vskip\abovedisplayskip
+\noindent\hfill
+  \namedinference{NE}{s \noteq t \mathrel\vee C}{s \; (\cst{sk} \; \overline{x}) \noteq t \; (\cst{sk} \; \overline{x}) \mathrel\vee C}%
+\qquad
+  \namedinference{PE}{s \; x \eq t \; x \mathrel\vee C}{s \eq t \mathrel\vee C}%
+\hfill\hbox{}
+\vskip\belowdisplayskip
+
+\noindent
+For \infname{NE}, $\overline{x}$ contains all the variables occurring in $s$ and
+$t$, the terms $s$ and $t$ are of function type, $\cst{sk}$ is a
+fresh Skolem symbol, % of appropriate type,
+and the literal $s \noteq t$ is eligible for
+resolution \cite[Sect.~5]{bbtvw-21-sup-lam}. For \infname{PE}, variable $x$
+does not occur in any of the $s$, $t$, or $C$, no literals are selected in $C$, and
+$s \; x \eq t \; x$ is a maximal literal.
+
+\begin{sloppypar}
+Finally, we introduced an injectivity recognition (IR) rule, which detects
+injectivity axioms and asserts the existence of the inverse function for
+injective function symbols:
+\end{sloppypar}
+
+\vskip\abovedisplayskip
+
+\noindent\hbox{}\hfill
+\namedinference{IR}{\cst{f} \; \tuple{x}{n} \noteq \cst{f} \; \tuple{y}{n} \mathrel\vee x_i \eq y_i} {\cst{sk} \; (\cst{f} \;\tuple{x}{n}) \; \overline{x}_J \, \eq\, x_i}\hfill\hbox{}
+
+\vskip\belowdisplayskip
+
+\noindent
+where $\textsf{sk}$ is a fresh Skolem symbol,
+$J$ is the largest subset of $\{1, \ldots, n$\} such that $x_j = y_j$ for every $j \in J$.
+We denote the subsequence of $\tuple{x}{n}$ indexed by $J$ by $\tuple{x}{J}$.
+Moreover, we require that
+$x_i \neq y_i$, all variables in $\tuple{x}{K} \cdot \tuple{y}{K}$ are distinct,
+where $K = \{1, \ldots, n\} \setminus J$,
+and neither $\tuple{x}{K}$ nor $\tuple{y}{K}$ shares variables with $\tuple{x}{J}$.
+For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor a \eq b'$,
+\infname{IR} can derive the existence of the inverse
+$\cst{sk}_1$ characterized by $\cst{sk}_{1\!} \; (\cst{add} \; a \; b) \; a \eq b$.
+%As a second example, from the clause
+%$\cst{cons} \; x \; \mathit{xs} \noteq \cst{cons} \; y \; \mathit{ys}
+%\mathrel\lor x \eq y$, we can derive
+%$\cst{sk}_2 \; (\cst{cons} \; x \; \mathit{xs}) \eq x$.
+
 \section{Heuristics}

### 2022-02-23 pre/pre.tex
unif rules
more

@@
 in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
-We say a term is {\em ground} if it has no variables.
+We say a term is {\em ground} if it has no variables. By $u[s]$
+we denote that $s$ is a subterm of $u$.
 
@@
 Applying substitution to a literal is reduced to applying it to both sides of
-the (dis)equa\-tion and it is extended pointwise to clauses. It is written in
-postfix notation as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a
-ground instance of $C$ if $\sigma$ is a grounding substitution.
+the (dis)equa\-tion and it is extended pointwise to clauses. It is sometimes
+also written in postfix notation as $l\sigma$ and $C\sigma$. We say
+$\sigmacl{C}$ is a ground instance of $C$ if $\sigma$ is a grounding
+substitution.
 
@@
 \subsection{Unification}
+\label{sec:pre:unif}
 
@@
 substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
-$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
-unordered pair of two terms of the same type. A \emph{most general unifier} is a
+$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
+We always specify if the constraint should be interpreted as ordered or unordered.
+A \emph{most general unifier} is a
 unifier $\sigma$, such that for any other unifier $\theta$, there is a
@@
 is more general than the other. To generalize most general
-unifiers to higher-order logic, the concept of tje {\em complete sets of unifiers} (CSU) was
+unifiers to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
 introduced.
@@
 \subsection{Inference Rules}
+\label{sec:pre:rules}
 \newcommand{\mgu}{\ensuremath{\mathit{mgu}}}
@@
     \multicolumn{2}{l}{{\bf Equality resolution} (ER)} \\[\jot]
-    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible for resolution
+    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $\sigmacl{s \noteq t}$ is eligible for resolution
 \end{tabular}
@@
     \multicolumn{2}{l}{{\bf Equality factoring} (EF)} \\[\jot]
-    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}$ 
+    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{t \noteq v \llor u \eq v \llor C}}$ 
         & where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
-        $(s \eq t)\sigma$ is eligible for paramodulation
+        $\sigmacl{s \eq t}$ is eligible for paramodulation
 \end{tabular}
@@
     \multicolumn{2}{l}{{\bf Superposition into positive literals} (SP)} \\[\jot]
-    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}$ 
+    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{\sigmacl{u[p \leftarrow t] \eq v \llor C \llor D}}$ 
         & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-        $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable
+        $\sigmacl{s \eq t}$ is eligible for paramodulation, just like $\sigmacl{u \eq v}$ and $u|_p$ is not a variable
 \end{tabular}
@@
     \multicolumn{2}{l}{{\bf Superposition into negative literals} (SN)} \\[\jot]
-    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}$ 
+    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{\sigmacl{u[p \leftarrow t] \noteq v \llor C \llor D}}$ 
         & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-         $(s \eq t)\sigma$ is eligible for paramodulation, $(u \noteq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable
+         $\sigmacl{s \eq t}$ is eligible for paramodulation, $\sigmacl{u \noteq v}$ is eligible for resolution, and $u|_p$ is not a variable
 \end{tabular}
@@
 \namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
-              {s \eq t \qquad (u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
+              {s \eq t \qquad \sigmacl{u[p \leftarrow \sigma(t)] \eq v \llor C \llor D}}
 \]
@@
 \subsection{The Saturation Procedure}
+\label{sec:pre:saturation}
 

### 2022-02-23 ehoh/ehoh.tex
inference rules
finished

@@
 
+\looseness=-1
 A side effect of the flattened representation is that prefix subterms are not
@@
 
+\vskip2\jot
 \noindent
@@
 \end{tabular}
+\vskip2\jot
 
@@
 
-\hfill
+% \vskip\abovedisplayskip
+\hbox{}\hfill
 \begin{tabular}{ll}
@@
 \end{tabular}
-\hfill
-\kern\belowdisplayskip
+\hfill\hbox{}
+\vskip\belowdisplayskip
 
@@
 
+E's heuristics are largely independent of the logic used and work
+unchanged for Ehoh. Yet, in preliminary experiments, we noticed that E proved
+some \lfhol{} benchmarks quickly using the applicative encoding
+(Sect.~\ref{sec:ehoh:introduction}), whereas Ehoh timed out.
+\begin{rep}There were enough such problems to prompt us to take a
+closer look. \end{rep}%
+Based on these observations, we extended the heuristics\confrep{}{ to exploit
+\lfhol{}-specific features}.
+
+\ourpara{Term Order Generation}
+The superposition calculus is parameterized by a term
+order\begin{rep}---typically an instance of KBO or LPO\end{rep} (Sect.~\ref{sec:pre:order}).
+E can generate a \emph{symbol weight} function (for KBO) and a \emph{symbol
+precedence} (for KBO and LPO) based on criteria such as the symbols'
+frequencies\confrep{}{, their arities,} and whether they appear in the conjecture.
+
+In preliminary experiments, we discovered that the presence of an explicit
+application operator \appvar{} can be beneficial for some problems.
+\begin{rep}
+%A small example will help illustrate this behavior.
+Let $\cst{a} : \iota_1$,
+$\cst{b} : \iota_2$, $\cst{c} : \iota_3$, $\cst{f} : \iota_1 \to \iota_2 \to
+\iota_3$, $x : \iota_2 \to \iota_3$, $y : \iota_2$, and $z : \iota_3$, and
+consider the clauses
+%
+$\cst{f} \; \cst{a} \; y \noteq \cst{c}$ and
+$x \; \cst{b} \eq z$,
+%
+where the first one is the negated conjecture. Their applicative encoding
+is
+%
+$\cst{@}_{\iota_2,\iota_3}(\cst{@}_{\iota_1,\iota_2\to\iota_3}(\cst{f}, \cst{a}), y) \noteq \cst{c}$ and
+$\cst{@}_{\iota_2,\iota_3}(x, \cst{b}) \eq z$,
+%
+where $\cst{@}_{\tau,\upsilon}$ is a type-indexed family of
+symbols representing the application of a function of type $\tau
+\to \upsilon$.
+\end{rep}%
+With the applicative encoding, generation schemes can take the symbols
+$\cst{@}_{\tau,\upsilon}$ into account, thereby exploiting the type
+information carried by such symbols.\begin{rep} Since $\cst{@}_{\iota_2,\iota_3}$ is a
+conjecture symbol, some weight generation scheme could give it a low weight,
+which would also impact the second clause. By contrast, the native \lfhol{}
+clauses share no symbols; the connection between them is hidden in the types
+of variables and symbols, which are ignored by the heuristics.
+
+\end{rep}
+
+To simulate the behavior observed on applicative problems,
+we introduced four generation schemes that extend E's existing
+symbol-frequency-based schemes by partitioning the symbols by type. To each
+symbol, the new schemes assign a frequency equal to the sum of all
+symbol frequencies for its class.\confrep{}{ Each new scheme is inspired by a similarly
+named type-agnostic scheme in E, without \texttt{type} in its name:
+%
+\begin{sloppypar}
+\begin{itemize}
+\item \texttt{typefreqcount} assigns as each symbol's weight the
+  number of occurrences of symbols of the same type.
+
+\smallskip
+\item \texttt{typefreqrank} sorts the frequencies calculated by the function
+  \texttt{typefreqcount} in increasing order and assigns each symbol a weight
+  corresponding to its rank.
+
+\smallskip
+\item \texttt{invtypefreqcount} is \texttt{typefreqcount}'s inverse. If
+  \texttt{typefreqcount} would assign a weight $w$ to a symbol, it assigns $M
+  - w + 1$, where $M$ is the maximum symbol weight according to
+  \texttt{typefreqcount}.
+
+\smallskip
+\item \texttt{invtypefreqrank} is \texttt{typefreqrank}'s inverse.
+  It sorts the frequencies in decreasing order.
+\end{itemize}
+\end{sloppypar}}
+%
+\looseness=-1
+We designed four more schemes (whose names begin with \verb|comb| instead of \verb|type|) that combine E's type-agnostic
+and Ehoh's type-aware approaches\confrep{.}{ using a linear equation.}
+
+To generate symbol precedences, E can sort symbols by weight
+%(breaking ties by arity or the first occurrence in the input problem, for example)
+and use the symbol's position in the sorted array as the basis for precedence.
+To reflect the type information introduced by the applicative encoding, we
+implemented four type-aware precedence generation schemes.
+%\begin{rep}, called
+%\texttt{typefreq}, \texttt{invtypefreq}, \texttt{combfreq}, and
+%\texttt{invcombfreq}, that sort the symbols by weight according to
+%\texttt{typefreqcount}, \texttt{invtypefreqcount}, \texttt{combfreqcount},
+%and \texttt{invcombfreqcount}, respectively\end{rep}.
+%
+Ties are broken by comparing the symbols' number of occurrences and, if
+necessary, the position of their first occurrence in the input.
+
+\ourpara{Literal Selection}
+
+The side conditions of the superposition rules SN and SP
+(Sect.~\ref{sec:pre:rules}) rely on a literal
+selection function to restrict the set of \emph{inference literals},
+thereby reducing the search space. Given a clause, a literal selection
+function returns a (possibly empty) subset of its literals. For
+completeness, any nonempty subset selected must contain at least
+one negative literal. If no literal is selected, all \emph{maximal}
+literals become inference literals.
+%Given a clause $L_1 \lor \cdots \lor L_n$, a literal selection function
+%returns either $\{L_i\}$ for some $i$ or $\emptyset$.
+The most widely used function is probably
+\texttt{SelectMaxLComplexAvoidPosPred}, which we abbreviate to
+\texttt{SelectMLCAPP}. It selects at most one negative literal, based
+on size, absence of variables, and maximality of the literal in the clause.
+%It
+%also avoids negative literals that share a predicate symbol with a
+%positive literal in the same clause.
+
+\begin{rep}
+  Intuitively, applied variables can potentially be unified with more terms than
+  terms with rigid heads. This makes them prolific in terms of possible
+  inference partners, a behavior we might want to avoid. On the other hand, shorter proofs
+  might be found if we prefer selecting applied variables. To cover both
+  scenarios, we implemented selection functions that prefer or defer selecting
+  applied variables. % and that break ties using \texttt{SelectMLCAPP}.
+
+  Let $\mathit{max}(L)=1$ if $L$ is a maximal literal of the clause it appears
+  in; otherwise, $\mathit{max}(L) = 0$. Let $\mathit{appvar(L)} = 1$ if $L$ is a
+  literal where either side is an applied variable; otherwise,
+  $\mathit{appvar(L)} = 0$. Based on these definitions, we devised the following
+  selection functions, both of which rely on \texttt{SelectMLCAPP} to break
+  ties:
+  %
+  \begin{itemize}
+  \item \texttt{SelectMLCAPPAvoidAppVar} selects a negative literal $L$ with
+    the maximal value of $(\mathit{max}(L){,}\; 1-\mathit{appvar}(L))$
+    according to the lexicographic order.
+
+  \smallskip
+  \item \texttt{SelectMLCAPPPreferAppVar} selects a negative literal $L$ with the
+    maximal value of $(\mathit{max}(L){,}\; \mathit{appvar}(L))$
+    according to the lexicographic order.
+  \end{itemize}
+  %
+  %The presence of $\mathit{max}(L)$ as the first criterion is motivated by
+  %initial experiments.
+  \end{rep}
+
+  \ourpara{Clause Selection}
+
+  Selection of the given clause is a critical choice point.
+  %To ensure completeness every candidate clause should be selected
+  %eventually\confrep{}{, unless it is identified as redundant and
+  %discarded before its turn has come}. Ignoring a candidate forever
+  %would compromise E's completeness guarantee.
+  E heuristically assigns \emph{clause priorities} and \emph{clause weights} to
+  the candidates.
+  The priorities provide a crude partition, whereas the weights order the clauses
+  within a partition.
+  E's main loop visits, in round-robin fashion, a set of priority queues. From a given
+  queue, the clause with the highest priority and the smallest weight is selected.
+  \begin{rep}%
+  Typically, one of the queues will use the clauses' age as priority, to ensure
+  fairness.
+  \end{rep}
+
+  E provides template weight functions that allow users to fine-tune parameters
+  such as weights assigned to variables or function symbols. The most widely
+  used template is \texttt{ConjectureRelativeSymbolWeight}\confrep{}{, which we abbreviate
+  to \texttt{CRSWeight}}. It computes term and clause weights according to
+  eight parameters, notably \textit{conj\_mul}, a multiplier
+  applied to the weight of conjecture symbols.
+  %
+  % conj_multiplier, fweight, cweight, pweight, vweight, max_term_multiplier, max_literal_multiplier, pos_multiplier
+  % [, app_var_multiplier]
+  %
+  %\begin{description}[labelwidth=\widthof{\rm\textit{appv\_mul}\,}]
+  %  \item[\rm\textit{conj\_mul}]
+  %    multiplier applied to the weight of conjecture symbols;
+  %  \item[\rm\textit{fweight}] weight of a nonnullary function symbol;
+  %  \item[\rm\textit{cweight}] weight of a nullary function symbol;
+  %  \item[\rm\textit{pweight}] weight of a predicate symbol;
+  %  \item[\rm\textit{vweight}] weight of a variable;
+  %  \item[\rm\textit{maxt\_mul}]
+  %    multiplier applied to the weight of maximal terms in a literal;
+  %  \item[\rm\textit{maxl\_mul}]
+  %    multiplier applied to the weight of maximal literals in a clause;
+  %  \item[\rm\textit{pos\_mul}]
+  %    multiplier applied to the weight of positive literals in a clause.
+  %\end{description}
+  %
+  This template works well for some applicatively encoded problems.
+  Let $\cst{a} : \iota$, $\cst{f} : \iota \to \iota$, $x : \iota$, and $y :
+  \iota \to \iota$, and consider the clauses
+  $y \; x \noteq x$ and
+  $\cst{f} \; \cst{a} \eq \cst{a}$,
+  where the first one is the negated conjecture. Their encoding
+  is
+  $\cst{@}_{\iota,\iota}(y, x) \noteq x$ and
+  $\cst{@}_{\iota,\iota}(\cst{f}, \cst{a}) \eq \cst{a}$.
+  %
+  The encoded clauses share $\cst{@}_{\iota,\iota}$, whose weight
+  will be multiplied by \textit{conj\_mul}---usually a factor in the
+  interval $(0, 1)$. By contrast, the native \lfhol{} clauses share no symbols,
+  and the heuristic would fail to notice that $\cst{f}$ and $y$ have the same
+  type, giving a higher weight to the second clause.
+  To mitigate this, we coded a new type-aware template,
+  \CRS\texttt{TypeWeight}, that
+  %\confrep{}{behaves like \CRS\texttt{Weight} except that it }
+  applies the \textit{conj\_mul} multiplier to all symbols whose type
+  occurs in the conjecture. \begin{rep}For the example above, since $\iota \to
+  \iota$ appears in the conjecture, it would notice the relation between the
+  conjecture variable $y$ and the symbol $\cst{f}$ and multiply $\cst{f}$'s
+  weight by \textit{conj\_mul}.
+
+  Natively supporting \lfhol{} allows the prover to recognize
+  applied variables. It may make sense to extend clause weight templates to either
+  penalize or promote clauses with such variables. To
+  support this extension, we added the following parameter to \texttt{CRSTypeWeight},
+  as well as to some other E's weight
+  function templates:
+  \textit{appv\_mul} is a multiplier applied to terms $s = x \; \tuple{t}{n}$, where $s$
+    is either side of the literal and $n>0$. %, and $x$ is a variable.
+  %
+  In addition, we implemented a new clause priority scheme,
+  \texttt{ByAppVarNum}, that separates the clauses by the number of top-level
+  applied variables occurring in the clause, favoring those containing fewer
+  such variables.
+  \end{rep}
+
+\ourpara{Configurations and Modes}
+A combination of parameters, including term order, literal
+selection, and clause selection, is called a \emph{configuration}.
+%
+%It is generally acknowledged that most provable problems arising in practice
+%can be proved quickly in at least one configuration; but the configuration
+%space being infinite, there is no easy way to find this configuration.
+%
+%determines whether a proof can be found in reasonable time. It is very
+%laborious and difficult to manually choose correct parameters.
+%
+For years, E has provided an \emph{auto} mode that analyzes the input
+problem and chooses a configuration known to perform well on similar problems.
+More recently, E has been extended with an \emph{autoschedule} mode that
+applies a portfolio of configurations in sequence on the given problem\confrep{.}{,
+restarting the prover for each configuration.
+%Time slicing approaches tend to perform better in practice, even if each
+%configuration is given a shorter time slice.
+
+}
+Configurations that are suitable for a wide range of problems have emerged over
+time. One of them is the configuration that is most often chosen by E's \emph{auto}
+mode. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
+%
+%%% TYPESETTING: TEMPORARILY DISABLED LEFTMARGIN (?)
+\begin{description}[labelwidth=\widthof{Precedence generation:}]
+  \item[\rm Term order:] KBO
+  \item[\rm Weight generation:] \texttt{invfreqrank}
+  \item[\rm Precedence generation:] \texttt{invfreq}
+  \item[\rm Literal selection:] \texttt{SelectMLCAPP}
+  \item[\rm Clause selection:] ~ \\
+\verb|1.CRSWeight(SimulateSOS, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
+\verb|4.CRSWeight(ConstPrio, 0.1, 100, 100, 100, 100, 1.5, 1.5, 1.5),| \\
+\verb|1.FIFOWeight(PreferProcessed),| \\
+\verb|1.CRSWeight(PreferNonGoals, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
+\verb|4.Refinedweight(SimulateSOS, 3, 2, 2, 1.5, 2)|
+\end{description}
+%
+The clause selection scheme consists of five queues, each of which is
+specified by a weight function template. The prefixes $n$\texttt{.} next to
+the template names indicate that the queue will be visited $n$ times in the round-robin scheme
+before moving to the next one. The first argument to each
+template is the clause priority scheme.
+
 \section{Preprocessing}
@@
 
+E's preprocessor transforms first-order formulas into clausal
+normal form, before the main loop is started. % and inference rules are applied.
+As explained in Sect.~\ref{sec:pre:clauses}, E also encodes all literals
+as equations.
+Beyond turning the problem into a
+conjunction of disjunctive clauses, the preprocessor eliminates quantifiers,
+introducing Skolem symbols for essentially existential quantifiers.
+
+For first-order logic, skolemization preserves both satisfiability
+(unprovability) and unsatisfiability (provability). In contrast, for
+higher-order logics without the axiom of choice, naive skolemization is
+unsound, because it introduces symbols that can be used to instantiate
+higher-order variables.
+%
+One solution proposed by Miller \cite[Sect.~6]{dm-87-compact-proofs} is to ensure that
+Skolem symbols are always applied to a minimum number of arguments.
+%Accordingly, the version of \lfhol{} introduced by Bentkamp et al.\
+%\cite{bbcw-21-lfho} distinguishes between mandatory and optional
+%arguments.
+However, to keep the implementation simple, we have decided to
+ignore this issue and consider all arguments as optional, including those to
+Skolem symbols. In Chapter \ref{ch:ehoh2} we extend Ehoh's logic to full higher-order logic
+with the axiom of choice, which addresses the issue.
+%TODO_PHD: Once Ehoh 2 paper is done, add reference to the section about choice here.
+
+There is another transformation performed by preprocessing that is
+problematic, but for a different reason. \emph{Definition unfolding} is the
+process of replacing equationally defined symbols with their definitions and
+removing the defining equations. A \relax{definition} is a clause of the form
+$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
+$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
+\{x_1,\dots,x_m\}$.
+This transformation preserves unsatisfiability (provability) for first-~and
+higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
+is that by removing the definitional clause, we also remove a symbol $\cst{f}$
+that otherwise could be used to instantiate a higher-order quantifier.
+For example, the clause set
+$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
+whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
+higher-order logic, the second clause set would be unsatisfiable thanks to
+the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-conversion.)
+%
+For the moment, we have simply disabled definition unfolding in Ehoh. We will
+enable it again once we have added support for $\lambda$-terms.
+
+Higher-order logic treats formulas as terms of Boolean type, erasing
+the distinction between terms and formulas. As a consequence, formulas might appear
+as arguments not only to logical connectives but also to function symbols
+or applied variables---e.g., $\cst{p} \> (\cst{a} \iand \cst{b})$,
+$y \> (\inot\, \cst{a})$.
+We call such formulas \emph{nested}. Kotelnikov et al.\
+\cite{kotelnikov-16-fool} describe a modification to
+Vampire's clausification algorithm to support nested formulas. We adapt
+their approach to the clausification algorithm
+\cite{nw-01-small-cnf} used by E. Given a formula $\varphi$ to
+clausify, the following procedure removes nested formulas:
+%
+\begin{enumerate}
+  \item\label{item:rec}
+    Let $\chi = \varphi|_{p}$ be the leftmost outermost nested formula that is different from
+    $\itrue$, $\ifalse$, or a variable $x$, if one exists;
+    otherwise, skip to step~\ref{item:clausif}.
+    Let $p = q.r$ where $q$ is the longest strict prefix of $p$ such that $\psi = \varphi|_q$ is a formula.
+    Let $\psi' = (\chi \mathrel{\iimplies} \psi[\itrue]_r)
+    \mathrel{\iand} (\inot\chi \mathrel{\iimplies} \psi[\ifalse]_r)$.
+    Replace $\varphi$ by $\varphi[\psi']_q$ and repeat this step.
+  \item\label{item:clausif}
+    Apply all the steps of E's clausification algorithm up to and
+    including skolemization.
+  \item Skolemization might replace Boolean variables by new terms with
+    predicate symbol heads. To remove them, follow step \ref{item:rec}.
+  \item Perform the remaining steps of E's clausification algorithm,
+    resulting in a set of clauses.
+  \item Let $C$ be a clause that contains a literal $L$ of the form $x \eq \itrue$ or $x \noteq \itrue$, where $x$
+    is a Boolean variable, if one exists; otherwise, terminate.
+    Delete $C$ if it also contains the complement of $L$.
+    Otherwise, replace $C$ with the clause $C[x \mapsto \ifalse]$
+    if $L$ is of the form $x \eq \itrue$ and else $C[x \mapsto \itrue]$.
+    Trivial literals $\ifalse \eq \itrue$ and $\itrue \noteq \itrue$ are removed from
+    the resulting clause. Repeat this step.
+\end{enumerate}
+
+%Like the approach by Kotelnikov et al., our approach effectively interprets
+%all by hoisting them to the level above the atom in
+%which they appear. This then reduces the problem to one that first-order
+%provers can understand.
+
+As an example, consider the formula $\cst{f} \; x \mathrel{\ieq} x \mathrel{\iimplies} \cst{p} \;
+(\cst{a} \iand \cst{b})$. Step~\ref{item:rec} moves
+the subterm $\cst{a} \iand \cst{b}$ outward, yielding
+  $\cst{f} \; x \mathrel{\ieq} x \mathrel{\iimplies}
+    ((\cst{a} \iand \cst{b}) \mathrel{\iimplies} \cst{p} \>\itrue)
+    \mathrel{\iand}
+    ({\inot}\,(\cst{a} \iand \cst{b}) \mathrel{\iimplies} \cst{p} \>\ifalse)$.
+This formula can be clausified further as usual.
+
+\begin{sloppypar}
+\begin{theorem}[Total Correctness]\,%
+The above procedure always terminates and produces a set of clauses that is
+equisatisfiable with the original formula $\varphi$ in \lfhol{} with
+interpreted Booleans and that contains no nested formulas other than
+$\itrue$, $\ifalse$, and variables.
+\end{theorem}
+\end{sloppypar}
+
+\newcommand{\tmeasure}{\ensuremath{\mathcalx{W}}}
+
+\begin{proof}
+It is easy to see that steps 1,~3, and~5
+produce equivalent formulas or clauses. Moreover, steps 1~and~3 remove all
+offending nested formulas (i.e., other than $\itrue$,
+$\ifalse$, and variables). In conjunction with the standard
+clausification algorithm, which preserves and reflects satisfiability, our
+procedure gives correct results when it terminates.
+
+To prove termination, we will use a measure function~$\tmeasure$ to natural
+numbers that decreases with each application of step 1~or~3.
+Steps 2~and~4 rely on a terminating algorithm, whereas each
+application of step~5 decreases the size of a clause. We define $\tmeasure$ by
+$\tmeasure(\forall x. \> s) = \tmeasure(\exists x. \> s) = \tmeasure(s)$;
+$\tmeasure(\zeta \, \tuple{s}{n}) = \sum_{i=1}^{n}{\tmeasure(s_i)}$ if
+$\zeta$ is a logical connective (including $\itrue$ and
+$\ifalse$); and $\tmeasure(\zeta \, \tuple{s}{n}) = 3^k (1 +
+\sum_{i=1}^{n}{\tmeasure(s_i)})$ otherwise, where
+$k$ is the number of offending outermost nested formulas in $\zeta \, \tuple{s}{n}$.
+We must show $\tmeasure(\psi) > \tmeasure(\psi')$. By definition,
+$\psi$ is of the form $\zeta \,
+\tuple{s}{n}$, where $\zeta$ is not a logical connective.
+Thus $\tmeasure(\psi) = 3^k (1 +
+\sum_{i=1}^{n}{\tmeasure(s_i)})$. Steps 1~and~3 substitute
+$\itrue$ or $\ifalse$, of measure 0, for a nested formula $\chi$
+(including $\chi$'s own nested formulas) in $\psi$.
+Clearly, the longer $r$ is, the more $\tmeasure(\psi')$
+decreases. Taking $\left|r\right|=1$, we get the upper bound $2 \tmeasure(\chi)
++ 2 \cdot 3^{k-1} (1 + \sum_{i=1}^{n}\tmeasure(s_i) - \tmeasure(\chi))$ for
+$\tmeasure(\psi')$, which is less than $\tmeasure(\psi) = 3^k (1 +
+\sum_{i=1}^{n}{\tmeasure(s_i)})$.
+\end{proof}
+
+The output may contain $\itrue$, $\ifalse$, or Bool\-ean
+variables as nested formulas. Since E was first developed as an untyped prover,
+unification of a variable with a Boolean constant was disallowed to avoid
+unsoundness. We needed to undo this in Ehoh. Ehoh must also remove trivial
+literals $\ifalse \eq \itrue$ and $\itrue \noteq
+\itrue$ that emerge during proof search.
+\end{rep}
+
 \section{Evaluation}
@@
 
+%In this section, we consider the following questions:
+How useful are Ehoh's new heuristics? And how does Ehoh perform compared with
+E, used directly or in tandem with the
+applicative encoding, and compared with other provers?
+To answer the first question, we evaluated each new parameter independently. From
+the empirical results, we derived a new configuration optimized for \lfhol{}.
+For the second question, we compared Ehoh's success
+rate\begin{rep} and speed\end{rep} on \lfhol{} problems with
+native higher-order provers and
+on applicatively encoded problems with E. We also included first-order
+benchmarks to measure Ehoh's overhead.
+
+We set a CPU time limit of 60\,s per problem.
+\begin{rep}%
+This is more than allotted by interactive proof tools such as Sledgehammer, or
+by cooperative provers such as Leo-III and Satallax, but less than the 300\,s
+of~CASC \cite{gs-17-casc}.
+\end{rep}%
+The experiments were performed on StarExec \cite{sst-14-starexec}
+nodes equipped with Intel Xeon E5-2609\,0 CPUs clocked at 2.40\,GHz.
+
+\ourpara{Heuristics Tuning}
+
+We used the \emph{boa} configuration as the basis to evaluate the new
+heuristic schemes. For each heuristic parameter we tuned, we changed only its
+value while keeping the other parameters the same as for \emph{boa}.
+\begin{rep}This gives an idea of how each parameter affects overall
+performance.\end{rep}
+All heuristic parameters were tested on a 5012~problem suite generated
+using Sledgehammer, consisting of four variants of the Judgment Day
+\cite{bn-10-sh} suite.\begin{rep} The problems were given in native
+\lfhol{} syntax.\end{rep} The experiments described in this subsection were
+carried out using an earlier E version (2.3).\begin{conf}
+%, exploiting its support for partial application and applied variables.
+Our main findings are as follows:
+%
+\begin{itemize}
+\item The combination of the weight generation scheme \texttt{invtypefreqrank}
+  and the precedence generation scheme \texttt{invtypefreq} performs best.
+
+\smallskip
+\item The literal selection heuristics \texttt{SelectMLCAPP},
+  \texttt{SelectMLCAPPPreferAppVar}, and \texttt{SelectMLCAPPAvoidAppVar} give
+  virtually the same results.
+
+\smallskip
+\item The clause selection function
+  \texttt{ConjectureRelativeSymbolTypeWeight} with \texttt{ConstPrio} priority
+  and an \textit{appv\_mul} factor of 1.41 performs best.
+\end{itemize}
+
+\looseness=-1
+We derived a new configuration from \textit{boa}, called \textit{hoboa},
+by enabling the features identified in the first and third points. Below, we
+present a more detailed evaluation of \textit{hoboa}, along with other
+configurations, on a larger benchmark suite. T%
+\end{conf}%
+
+%invtypefreqrank and invtypefreq weight and precedence generation schemes
+%as the best. For clause evaluation, we picked CRSTypeWeight with ConstPrio priority and applied variable multiplier of 1.41. Lastly, for literal selection we picked SelectMLCAPPAvoidAppVar.
+
+%\ourpara{Term Order Generation}
+
+Evaluating the new weight and pre\-cedence generation heuristics amounted to testing
+each possible combination of frequency-based schemes, including E's original
+type-agnostic schemes. Figure~\ref{fig:ehoh:term-eval} shows the number of
+solved (i.e., proved or disproved) problems for each combination. In this and
+the following figures, the underlined number is for \emph{boa}, whereas bold
+singles out the best value. In the names of the generation schemes, we abbreviated
+\texttt{inv} to \texttt{i}, \texttt{type} to \texttt{t}, \texttt{freq} to
+\texttt{f}, \texttt{comb} to \texttt{cm}, \texttt{count} to \texttt{cn}, and
+\texttt{rank} to~\texttt{r}.
+
+\newcommand\EHOHHLINE{\\[.5\jot]\hline\noalign{\vskip.75\jot}}
+\begin{figure}[t]
+\center
+    \begin{tabular}{@{}l@{\kern1em}c@{\kern.625em}c@{\kern.625em}c@{\kern.625em}c@{\kern.625em}c@{\kern.625em}c@{}}
+    ~                         & \texttt{f} & \texttt{if} & \texttt{tf} & \texttt{itf} & \texttt{cmf} & \texttt{icmf}
+\EHOHHLINE
+    \texttt{fcn}        & 2294             & 2288                & 2287                 & 2297                    & 2290                 & 2287                    \\
+    \texttt{ifcn}     & 2371             & 2373               & 2374                 & 2370                    & 2369                 & 2377                    \\
+    \texttt{fr}         & 2326             & 2317                & 2323                 & 2329                    & 2322                 & 2318                    \\
+    \texttt{ifr}      & 2383             & \underline{2379}                & 2376                 & 2380                    & 2381                 & 2381                    \\
+    \texttt{tfcn}    & 2305             & 2314                & 2301                 & 2306                    & 2302                 & 2311                    \\
+    \texttt{itfcn} & 2386             & 2381                & 2389                 & 2388                    & 2384                 & 2379                    \\
+    \texttt{tfr}     & 2326             & 2334                & 2322                 & 2334                    & 2321                 & 2336                    \\
+    \texttt{itfr}  & 2390             & 2382                & 2390                 & \phantom{2394}\llap{\textbf{2394}}                    & 2387                 & 2386                    \\
+    \texttt{cmfcn}    & 2273             & 2281                & 2271                 & 2285                    & 2269                 & 2280                    \\
+    \texttt{icmfcn} & 2380             & 2375                & 2382                 & 2379                    & 2380                 &  2375                   \\
+    \texttt{cmfr}     & 2321             & 2313                & 2319                 & 2321                    & 2318                 & 2312                    \\
+    \texttt{icmfr}  & 2368             & 2378                & 2371                 & 2378                    & 2368                 & 2380                    \\
+    \end{tabular}
+    \caption{Evaluation of weight and precedence generation schemes}
+    \label{fig:ehoh:term-eval}
+\end{figure}
+
+Figure~\ref{fig:ehoh:term-eval} indicates that including type information in
+the generation schemes results in a somewhat higher number of solved problems
+compared with E's type-agnostic schemes. Against our expectations, Ehoh's
+combined schemes appear to be less efficient than the type-aware schemes.
+
+% \begin{sloppypar}
+  The literal selection function has little impact on performance: Ehoh
+  solves 2379 problems with \texttt{SelectMLCAPP} or
+  \texttt{SelectMLCAPPAvoid}\-\texttt{AppVar}, and one less with
+  \texttt{SelectMLCAPPPreferAppVar}.
+% \end{sloppypar}
+
+Clause selection is the heuristic component that we extended the most. We must
+assess the effect of a new heuristic weight function, a multiplier for the
+occurrence of top-level applied variables, and clause priority based on the
+number of top-level applied variables.
+
+\begin{figure}[t]
+  \center
+  \begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}}
+  ~ & 0.25 & 0.35 & 0.5 & 0.7 & 1 & 1.41 & 2 & 2.82 & 4
+\EHOHHLINE
+\texttt{W} &
+  2311 & 2341 & 2363 & 2374 & \phantom{2379}\llap{\underline{\textbf{2379}}} & 2376 & 2377 & 2376 & 2377 \\
+\texttt{TW} &
+  2331 & 2331 & 2360 & 2371 & 2372 & 2374 & 2373 & 2373 & 2372
+  \end{tabular}
+  \caption{Evaluation of weight function and \textit{appv\_mult} factor}
+  \label{fig:ehoh:clauseweight-fun-eval}
+\end{figure}
+
+To test the effect of the new type-based weight function, we replaced
+\emph{boa}'s queue, which uses
+\texttt{4.CRS\-Weight(}\ldots\texttt{)}, with the queue ordered by
+\texttt{4.CRSTypeWeight(}\ldots\texttt{)}. We call the original
+heuristic \texttt{W} and the type-aware alternative \texttt{TW}.
+%
+We chose nine values for testing the effect of the applied
+variable multiplier \textit{appv\_mult}.
+%
+Figure \ref{fig:ehoh:clauseweight-fun-eval} summarizes the results of combining
+\texttt{W} or \texttt{TW} with the different
+\textit{appv\_\allowbreak mult} values. Applying a multiplier smaller than
+1, which corresponds to preferring literals containing applied variables, can
+lose dozens of solutions. Overall, using the type-aware heuristic seems
+slightly detrimental.
+
+\begin{sloppypar}
+Finally, we evaluated the new clause priority function \texttt{ByAppVarNum}, by replacing
+\verb|4.CRSWeight(ConstPrio,|\allowbreak\ldots\verb|)| with \verb|4.CRSWeight(ByAppVarNum,|\ldots\verb|)| in
+\emph{boa}'s specification. \texttt{ConstPrio} assigns each clause the
+same priority. The results are inconclusive.
+\end{sloppypar}
+
+The results presented above give an idea of how each parameter influences
+performance. We also evaluated their performance in combination, to derive an
+alternative to \emph{boa} for \lfhol{}.
+% to find out which combination is best as design an efficient higher-order heuristic,
+%
+For each category of parameters, we chose either \emph{boa}'s value of the
+parameter in \emph{boa} (``Old'') or the best performing newly implemented
+parameter (``New'').
+Based on the results above, for term orders, we chose the combination of
+\texttt{invtypefreqrank} and \texttt{invtypefreq}; for clause selection, we
+chose \texttt{CRSType}\-\texttt{Weight} with \texttt{ConstPrio} priority and
+an \emph{appv\_mult} factor of 1.41; for literal selection, we chose
+\texttt{SelectMLC}\-\texttt{APP}\-\texttt{AvoidAppVar}.
+
+\begin{figure}[t]
+  \center
+  \begin{tabular}{@{}l@{\hskip 1em}l@{\hskip 1em}l@{\hskip 1em}c@{}}
+  Term order & Literal selection & Clause weight & Solved
+\EHOHHLINE
+  Old   & Old        & Old   & \underline{2379}                      \\
+  Old   & Old        & New   & 2374                      \\
+  Old   & New        & Old   & 2379                      \\
+  Old   & New        & New   & 2373                      \\
+  New   & Old        & Old   & 2394                      \\
+  New   & Old        & New   & \phantom{2397}\llap{\textbf{2397}}             \\
+  New   & New        & Old   & 2395                      \\
+  New   & New        & New   & \phantom{2397}\llap{\textbf{2397}}
+\end{tabular}
+  \caption{Evaluation of combinations of new parameters}
+  \label{fig:ehoh:new-modes}
+\end{figure}
+
+Figure \ref{fig:ehoh:new-modes} shows the number of solved problems for all
+combinations of these parameters. From the two configurations that solve 2397
+problems, we selected the ``New Old New'' combination as our suggested
+``higher-order best of \emph{auto},'' or \emph{hoboa}, configuration.
+
+\ourpara{Main Evaluation}
+
+We now present a more detailed evaluation of \textit{hoboa},
+along with other configurations, on a larger benchmark suite.
+Our raw data are publicly
+available.%
+\footnote{\url{https://doi.org/10.5281/zenodo.4045452}}
+
+The benchmarks are divided into four sets:
+%
+(1)~1147 first-order TPTP \cite{gs-17-tptp} problems belonging
+  to the FOF (untyped) and TF0 (monomorphic) categories, excluding arithmetic;
+(2)~5012 Sledgehammer-generated problems from the Judgment Day
+  \cite{bn-10-sh} suite, targeting the mono\-morphic first-order
+  logic embodied by TPTP TF0;
+(3)~all 955 monomorphic higher-order problems from the TH0 category of the
+  TPTP belonging to our extension of \lfhol{};
+(4)~5012 Judgment Day problems targeting the \lfhol{} fragment of TPTP TH0.
+
+\begin{rep}The TPTP includes benchmarks from various areas of
+computer science and mathematics. It is the de facto standard for evaluating
+automatic provers, but it has few higher-order problems.
+%
+\end{rep}%
+For the first group of benchmarks, we randomly selected 1000 FOF problems
+(out of 8172)
+and all monomorphic TFF problems that are parsable by E\confrep{}{ within 60\,s
+(amounting to 147 out of 231 monomorphic TFF problems)}.
+%
+Both groups of Sledgehammer problems include two subgroups of 2506 problems,
+generated to include 32 or 512 Isabelle lemmas (SH32 and SH512), to
+represent both small and large problems.
+% arising in interactive
+%verification.
+Each subgroup consists of two sub-sub\-groups of 1253
+problems, generated by using either $\lambda$-lifting or SK-style combinators
+to encode $\lambda$-expressions.
+
+\looseness=-1
+To ascertain the effectiveness of our approach, we evaluated Ehoh against
+E used on applicative encodings of problems (denoted by \appE{}). For reference,
+we also evaluated the following versions of higher-order provers that competed in the
+THF division of the 2019 edition of CASC \cite{gs-19-casc27}:\
+CVC4
+1.8 prerelease \cite{brotb-19-ho-smt}, Leo-III 1.4 \cite{sb-21-leo3},
+Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
+and Zipperposition 1.6 \cite{bbtvw-21-sup-lam}.
+Like at CASC, we used different versions of Vampire for first-order
+and higher-order problems. Similarly, Zipperposition does not use E as backend when
+it is run on first-order problems
+and uses different heuristics on first-~and higher-order problems.
+The genuine higher-order provers have
+the unfair advantage that they can instantiate higher-order
+variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
+Thus, some formulas that are
+provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
+they may require tedious reasoning about $\lambda$-lifted functions or SK-style
+combinators}.
+An example is the conjecture $\exists f.\>\forall x\> y.\; f\>x\;y \eq
+\cst{g}\;y\;x$, whose proof requires taking $\lambda x\> y.\; \cst{g}\;y\;x$
+as the witness for~$f$.
+
+We ran all provers except Satallax (which only supports THF) on first-order
+benchmarks to measure the overhead introduced by our extensions, as
+well as that entailed by the applicative encoding.
+Figure~\ref{fig:main-eval} gives the number of problems each system proved.
+In each column, bold highlights the best E value and the best value
+overall.
+We considered the E modes \emph{auto} (a) and
+\emph{autoschedule} (as) and the configurations \emph{boa} (b) and
+\emph{hoboa} (hb).
+
+\begin{figure}[t]
+\center
+\begin{tabular}{@{}l@{\hskip 1.5em}r@{\hskip 1em}r@{\hskip 1em}r@{\hskip 1.5em}r@{\hskip 1em}r@{\hskip 1em}r@{}}
+                            & \multicolumn{3}{@{}c@{\hskip 1.5em}}{First-order} & \multicolumn{3}{@{}c@{}}{Higher-order} \\[.3\jot]
+                      & TPTP              & SH32              & SH512        & TPTP            & SH32             & SH512
+\EHOHHLINE
+E a              			& 624                & 938              & 1237        &                 &               &          \\
+E as             			& \textbf{665}                & \textbf{957}              & \textbf{1298}        &                 &               &          \\
+E b              			& 550                & 943              & 1242        &                 &               &          \\[\jot]
+\appE{} a  		 			  & 531                & 932              & 1111        &  686            & 952           & 1125   \\
+\appE{} as 		 			  & 571                & 949              & 1148        &  692            & 969           & 1164   \\
+\appE{} b  		 			  & 536                & 943              & 1227        &  690            & 959           & 1267   \\[\jot]
+Ehoh a           			& 624                & 939              & 1236        &  694            & 966           & 1235   \\
+Ehoh as          			& \textbf{665}                & \textbf{957}              & 1296        &  \textbf{699}            & \textbf{988}  & \textbf{1309}   \\
+Ehoh b           			& 550                & 943              & 1242        &  697            & 967           & 1262   \\
+Ehoh hb          			& 504                & 947              & 1231        &  693            & 975           & 1267
+\EHOHHLINE
+CVC4            			& 567                & 956              & 1361        &  745            & 973           & \textbf{1351} \\%[\jot]
+Leo-III          			& 548                & 960              & 1239        &  \textbf{834}   & 967           & 1266 \\%[\jot]
+Vampire          			& \textbf{728}       & \textbf{968}     & \textbf{1401} &  805          & 979           & 1214 \\%[\jot]
+Satallax         			&                    &                  &             &  827            & 871           & 1019 \\%[\jot]
+Zipperposition  			& 496                & 933              & 1187        &  815            & 976           & 1069
+\end{tabular}
+\caption{Number of proved problems}
+\label{fig:main-eval}
+\end{figure}
+
+We observe the following.
+%
+First, comparing the Ehoh row with the E row, we see that Ehoh's
+  overhead is barely noticeable---the difference is at most \NumberOK{two}
+  problems.
+  %The raw evaluation data reveal that Ehoh's time overhead is about \NumberNOK{3.7\%}.
+%
+Second, Ehoh outperforms the applicative encoding on both first-order
+  and higher-order problems. Nevertheless, the raw evaluation data reveal that
+  there are quite a few higher-order problems that \appE{}
+  proves faster than Ehoh. %\confrep{}{, because the applicative encoding impacts the
+%  heuristics in subtle ways}.
+%
+Third, it is advantageous to use the higher-order versions of the
+  Sledgehammer problems, although the difference in success rate is
+  small, especially for SH512.
+%
+Fourth, the new \emph{hoboa} %configuration
+  outperforms \emph{boa} on
+  higher-order problems, suggesting that it could be worthwhile to re-train
+  \emph{auto} and \emph{autoschedule} based on \lfhol{} benchmarks
+  and to design further heuristics.
+%
+Fifth, Ehoh cannot compete against the best higher-order systems, but this
+  is no surprise given that it does not yet support $\lambda$-expressions and
+  higher-order unification.
+
+\begin{rep}
+Next to the success rate, the time in which a prover gives an
+answer is also an important consideration. Figure \ref{fig:timing} compares
+the average running times, in seconds, of the various systems on the problems
+that all of the applicable systems proved.
+%We can conclude that classical higher-order provers, Leo-III and
+%Satallax, take considerably more time than Ehoh on average.
+Clearly, Ehoh incurs little overhead on first-order problems. The raw
+evaluation data reveal that for \emph{boa}, it takes Ehoh \NumberOK{2747}\,s
+to prove all first-order problems that E, @+E, and Ehoh can all prove using this
+configuration, compared with \NumberOK{2728}\,s for E, amounting to a
+\NumberOK{0.7}\% overhead. For comparison, @+E needs \NumberOK{3939}\,s---%
+a \NumberOK{44}\% overhead.
+
+\begin{figure}[ht!]
+\center
+\begin{tabular}{@{}l@{\hskip 1.5em}r@{\hskip 1em}r@{\hskip 1em}r@{\hskip 1.5em}r@{\hskip 1em}r@{\hskip 1em}r@{}}
+                            & \multicolumn{3}{@{}c@{\hskip 1.5em}}{First-order} & \multicolumn{3}{@{}c@{}}{Higher-order} \\[.3\jot]
+                            & TPTP             & SH32            & SH512          & TPTP             & SH32             & SH512
+\EHOHHLINE
+E a                         & 0.22          & 0.15             & 0.54          &                  &                  &             \\
+E as                        & 0.38          & 0.20             & 0.74          &                  &                  &             \\
+E b                         & 0.43          & \textbf{0.07}    & 0.56          &                  &                  &             \\[\jot]
+\appE{} a                   & 0.61          & 0.18             & \textbf{0.38}          &  0.03            & 0.21             & \textbf{0.32}       \\
+\appE{} as                  & 0.91          & 0.18             & 0.39          &  0.06            & 0.25             & 0.33        \\
+\appE{} b                   & 0.53          & 0.12             & 0.81          &  0.09            & 0.20             & 0.54        \\[\jot]
+Ehoh a                      & \textbf{0.21}          & 0.15             & 0.54          &  0.03            & 0.08             & 0.51        \\
+Ehoh as                     & 0.38          & 0.20             & 0.73          &  0.07            & 0.14             & 0.60        \\
+Ehoh b                      & 0.42          & \textbf{0.07}             & 0.58          &  \textbf{0.02}            & \textbf{0.07}    & 0.37        \\
+Ehoh hb                     & 0.69          & 0.12             & 1.06          &  0.10            & 0.13             & 0.56
+\EHOHHLINE
+CVC4                        & 3.02          & 1.58             & 1.75          &  1.22            & 2.44             & 1.65 \\%[\jot]
+Leo-III                     & 1.33          & 0.52             & 5.63          &  0.49            & 0.89             & 6.54 \\%[\jot]
+Vampire                     & 0.67          & 0.43             & 1.50          &  0.76            & 1.89             & 4.84 \\%[\jot]
+Satallax                    &               &                  &               &  2.45            & 5.22             & 10.12 \\%[\jot]
+Zipperposition              & 3.81          & 1.60             & 5.09          &  0.76            & 2.21             & 6.31
+\end{tabular}
+\caption{Average running times on the problems proved by all systems}
+\label{fig:timing}
+\end{figure}
+\end{rep}
+
 \section{Discussion and Related Work}
@@
 
+\begin{rep}
+  Our working hypothesis is that it is possible to extend %existing
+  first-order provers to higher-order logic without slowing them down unduly.
+  Our research program is two-pronged:
+  On the theoretical side, we are
+  investigating higher-order extensions of superposition
+  \cite{bbcw-21-lfho,bbtvw-21-sup-lam,vn-20-bools};
+  on the practical side, we are implementing such extensions in a
+  state-of-the-art prover.
+  %TODO_PHD: another reference
+
+  The work described in this chapter required modifying many parts of the E
+  prover. The invariant that variables cannot be applied and that symbols are always
+  passed the same number of arguments were entrenched in E's code,
+  requiring hundreds of modifications. Nonetheless,
+  %with a good understanding of the prover's inner working,
+  %% Obvious, since Stephan is a coauthor. --JB
+  we found the generalization manageable and are now in a position to add
+  support for $\lambda$-terms and higher-order unification.
+  \end{rep}
+
+  \begin{sloppypar}
+  Traditionally, most higher-order provers were designed from the ground up
+  to target higher-order logic. Two
+  exceptions are Otter-$\lambda$ by Beeson \cite{mb-04-lam-logic} and
+  Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
+  \cite{sc-15-simon-phd,bbtvw-21-sup-lam}. \hbox{Otter-$\lambda$}
+  adds $\lambda$-terms and second-order unification to
+  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
+  % The approach is pragmatic. , with little emphasis on completeness.
+  Zipperposition, also based on superposition, was extended
+  to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
+  \cite{bbtvw-21-sup-lam}. Its performance is a far cry from E's, but
+  it is easier to modify.
+  Vukmirovi\'c et al.\ also used it to test and evaluate higher-order unification
+  procedures \cite{vbn-21-unif} and Boolean reasoning
+  \cite{vn-20-bools}.
+  Zipperposition now includes Ehoh as a backend in a cooperative
+  architecture.
+  Finally, there is recent work by the developers of Vampire
+  % br-20-full-sup-w-combs
+  \cite{br-19-restricted-unif} and of the SMT
+  (satisfiability modulo theories) solvers CVC4 and veriT
+  \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
+  \end{sloppypar}
+
+  %Automation of higher-order logic moved in two main directions---building native higher-order provers and
+  %translations into first-order logic.
+  Native higher-order reasoning was pioneered by Rob\-in\-son
+  \cite{ar-69-hol}, Andrews \cite{pa-71-type-theory}, and Huet
+  \cite{gh-73-hol}.\begin{rep} Andrews \cite{pa-01-classical-ty-thy} and Benz\-m\"uller and Miller
+  \cite{bm-14-automation-ho} provide excellent surveys.\end{rep}
+  TPS, by Andrews et al.~\cite{abinpx-96-tps}, was
+  based on expansion proofs and lets users specify proof outlines.
+  The Leo \confrep{}{family of }systems, developed by Benzm\"uller and his colleagues,
+  \confrep{are}{is} based on resolution and paramodulation.
+  LEO~\cite{cbmk-98-leo} \begin{rep}supported extensionality
+  on the calculus level and \end{rep}introduced the cooperative paradigm to
+  integrate first-order provers.
+  Leo-III \cite{sb-21-leo3} expands the cooperation with
+  SMT solvers
+  and introduces term orders\confrep{}{ in a pragmatic, incomplete way}.
+  Brown's Sat\-al\-lax \cite{cb-12-satallax} is based on a
+  complete higher-order tableau calculus,
+  guided by a SAT solver; later versions also cooperate with E and Ehoh.
+  Another noteworthy sys\-tem is Lindblad's agsy\-HOL \cite{fl-14-agsyhol}.
+  It is based on a focused sequent calculus driven by a generic narrowing engine.
+
+  An alternative to all of the above is to reduce higher-order logic to first-order
+  logic via a translation. Robinson \cite{ar-70-hol}
+  outlined this approach decades before tools such as
+  \begin{conf}Sledgehammer \cite{pb-12-sh} and
+  HOLyHammer \cite{ku-15-holyhammer}\end{conf}%
+  \begin{rep}Miz$\mathbb{AR}$ \cite{urs-13-atp},
+  Sledgehammer \cite{pb-12-sh},
+  HOLyHammer \cite{ku-15-holyhammer}, and
+  CoqHammer \cite{ck-18-coqhammer}\end{rep}
+  popularized it in proof assistants.
+  In addition to performing an applicative encoding, such translations must
+  eliminate the \hbox{$\lambda$-expressions}\begin{rep}
+  \cite{mp-08-trans,lc-16-lam-trans}\end{rep} and encode the type
+  information\begin{rep} \cite{bkpu-16-hammering-types}\end{rep}.
+  %
+  \begin{rep}%
+  In practice, on problems with a large first-order component, translations
+  perform very well compared with the existing native provers
+  \cite{ns-13-leo2sh}. Largely thanks to Sledgehammer, Isabelle often came
+  in close second at CASC, even defeating Satallax in 2012
+  \cite{gs-13-cascj6}.
+  \end{rep}
+
+  By removing the need for the applicative encoding, our work reduces
+  the translation gap. The encoding buries the \lfhol{} terms' heads under
+  layers of $\cst{@}$ symbols. %\confrep{}{, which impacts the heuristics that inspect terms}.
+  Terms double in size, cluttering the data structures, and twice as many
+  subterm positions must be considered for inferences. Moreover, the encoding is
+  incompatible with interpreted operators, notably for arithmetic.
+  \begin{rep}A common remedy is to introduce proxies to connect an
+  uninterpreted nullary symbol with its interpreted counterpart
+  (e.g., %$\cst{add}$ and $+$ in
+  $\cst{@}(\cst{@}(\cst{add}, x), y) \eq x + y$), but this is clumsy. \end{rep}A further
+  complication is that in a monomorphic logic, $\cst{@}$ is not a single symbol
+  but a family of symbols $\cst{@}_{\tau,\upsilon}$, which must be
+  correctly introduced and recognized. Finally, the encoding must be undone in
+  the proofs. While it should be possible to base a higher-order
+  prover on such an encoding, the prospect is aesthetically and
+  technically unappealing, and performance would likely suffer.
+
 \section{Conclusion}
 \label{sec:ehoh:conclusion}
+
+Despite considerable progress since the 1970s, higher-order automated
+reasoning has not yet assimilated some of the most successful methods for
+first-order logic with equality, such as superposition. We presented a
+graceful extension of a state-of-the-art first-order theorem prover to a
+fragment of higher-order logic devoid of $\lambda$-terms. Our work
+covers both theoretical and practical aspects. Experiments show promising
+results on $\lambda$-free higher-order problems and very little overhead for
+first-order problems, as we would expect from a graceful generalization.
+
+\begin{sloppypar}
+Despite its lack of support for $\lambda$-terms, Ehoh is already deployed
+as a backend in the leading higher-order provers Satallax and
+Zipperposition. Ehoh also forms the basis of our work toward stronger
+higher-order automation. Our aim is to turn it into a prover that excels on
+proof obligations arising in interactive verification, which tend to be
+large but \relax{only mildly higher-order} \cite{ns-13-leo2sh}.
+In Chapter \ref{ch:ehoh2} we describe the extension of Ehoh to full higher-order logic.
+It was heavily inspired by the techniques implemented in Zipperposition, which made it
+dominate the CASC competition in 2020 and 2021.
+\end{sloppypar}

### 2022-02-23 pre/pre.tex
finished

@@
 \section{Clausal Forms}
+\label{sec:pre:clauses}
 
@@
 \subsection{Term Order and Selection}
+\label{sec:pre:order}
 Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an

### 2022-02-24 ehoh/ehoh.tex
more

@@
 
-The main challenges we faced concerned the
+The main challenges we faced involved the
 representation of types and terms
@@
 (Sect.~\ref{sec:ehoh:unif-match}), and the indexing data structures
-(Sect.~\ref{sec:ehoh:indexing}). We also adapted the
+(Sect. \ref{sec:ehoh:indexing}). We also adapted the
 inference rules (Sect.~\ref{sec:ehoh:inferences})\confrep{ and}{,} the
 heuristics (Sect.~\ref{sec:ehoh:heuristics})\confrep{}{, and the
-  preprocessor (Sect.~\ref{sec:ehoh:preprocessing})}.
+  preprocessor (Sect. \ref{sec:ehoh:preprocessing})}.
 
@@
 
+\vskip\abovedisplayskip
 \noindent
@@
 \end{tabular}
-
+\vskip\belowdisplayskip
 \newcommand\AlphaMatch[1]{\alpha(#1)}
-
 We need an auxiliary function to convert $\Matchiiarrow$ states to

### 2022-02-25 introduction/introduction.tex
more

@@
 nature, some of them were algorithmic, while others concerned fine-tuning the
-heuristics to achieve the best possible performance. We discuss those challenges in more detail.
+heuristics to optimize the performance. We discuss those challenges in more detail.
 
@@
 
-  After spending so much time and energy on developing higher-order
-  provers, we took a step back and worked on improving first-order provers. In
-  particular, we looked at the most efficient approaches to simplify problems in
-  propositional logic, a simpler, decidable logic that disallows quantification.
-  We lifted some of these approaches to first-order logic and implemented them
-  in Zipperposition.
+  After spending so much time and energy on developing higher-order provers, we
+  took a step back and worked on improving first-order provers. As they form the
+  basis of higher-order reasoning in our approach, this should improve
+  higher-order performance as well. In particular, we looked at the most
+  efficient approaches to simplify problems in propositional logic, which is
+  simpler than first-order logic and even decidable. We lifted some of these
+  propositional approaches to first-order logic and implemented them in Zipperposition.
 

### 2022-02-25 pre/pre.tex
more

@@
 \looseness=-1
-An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq t$. A clause is a
+An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq t$. A clause $C$ is a
 finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
@@
     \textit{Respects contexts} & For all terms $s,t$ and contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
-    \textit{Ground total} & For any two
+    \textit{Ground total} & For all distinct
     ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$. \\
@@
 We always specify if the constraint should be interpreted as ordered or unordered.
+We say a substitution $\sigma$ is \emph{more general} than $\varrho$ if there exists a substitution $\theta$
+such that $\varrho = \theta\sigma$.
 A \emph{most general unifier} is a
 unifier $\sigma$, such that for any other unifier $\theta$, there is a
-substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
-$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
+substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
 the most general unifier is unique up to variable renaming. Furthermore,

### 2022-02-28 introduction/introduction.tex
wan main comments

@@
   basis of higher-order reasoning in our approach, this should improve
-  higher-order performance as well. In particular, we looked at the most
+  higher-order performance as well. In particular, we looked at
   efficient approaches to simplify problems in propositional logic, which is

### 2022-02-28 pre/pre.tex
wan main comments

@@
 
-We assume the standard notions of $\alpha$, $\beta$ and $\eta$ conversions and
-write $s \equi t$ if terms $s$ and $t$ are equal modulo $\alpha\beta\eta$
-converison. We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
+\looseness=-1
+With $\alpha$-renaming we assume the rule $(\lambda x. \, s) \longrightarrow \lambda
+y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$ and 
+$y$ is not bound in $s$. Application of substitution $\sigma$ to term
+$t$, implicitly $\alpha$-renames bound variables to avoid capture.  For example,
+$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
+$\beta$-reduction rule corresponds to applying arguments to function: $(\lambda
+x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where the bound variables in $s$
+are renamed to avoid capture. Lastly, $\eta$-reduction is defined as $\lambda
+x.\, s \, x \longrightarrow s$ under the condition that $x$ is not loosely
+bound in $x$. We write $s \equi t$ if terms $s$ and $t$ are equal modulo $\alpha\beta\eta$
+converison.
+
+\looseness=-1
+As hinted in previous definitions, higher-order substitutions
+($\sigma,\varrho,\ldots$) are functions from both free and bound variables to
+terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F) \not\equi F$.
+Substitutions map only finitely many variables. Given a substitution $\varrho$,
+which maps $F$ to $s$, with $\varrho\setminus\{F \mapsto s\}$ we denote a
+substitution that does not map $F$ and otherwise coincides with $\varrho$. Given
+substitutions $\varrho$ and $\sigma$, mapping disjoint variable sets, we write
+$\varrho \cup \sigma$ to denote $\varrho\sigma$.
+
+We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
 and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
@@
 if it has no loosely bound variables.
-
-Higher-order substitutions ($\sigma,\varrho,\ldots$) are functions from free and bound
-variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F) \not\equi
-F$. Each substitution maps only finitely many variables, and it is denoted as in
-the first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
-this application $\alpha$-renames $t$ to avoid variable capture. For example,
-$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$.  Given a substitution
-$\varrho$, which maps $F$ to $s$, we write $\varrho\setminus\{F \mapsto s\}$ to
-denote a substitution that does not map $F$ and otherwise coincides with
-$\varrho$. Given substitutions $\varrho$ and $\sigma$, mapping disjoint variable
-sets, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.
 
@@
 $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
-$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.
+$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the notation,
+we sometimes write the predicate literals in non-encoded form.
 
@@
 
+\medskip
+Let us give some intuition behind how these rules are applied in practice. The
+\infname{ER} rule is used to replace equality with equality within a clause.
+Consider the formula $\iforall x.\,  x \eq \cst{2} \iimplies \cst{even}(x)$. In
+clausal form it is $X \noteq \cst{2} \llor \cst{even}(X)$. Applying the
+\infname{ER} rule effectively applies the precondition $X \eq \cst{2}$ to the
+conclusion $\cst{even}(X)$ to obtain $\cst{even}(\cst{2})$. The \infname{EF}
+rule can be used to find instances of the clause that can further be simplified.
+Consider the clause $\cst{f}(X) \eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor
+\cst{p}(X, Y)$. Applying \infname{EF} to its first two literals yields $\cst{a}
+\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$. This clause
+then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ using \infname{ER}.
+
+Lastly, superposition rules are used to simulate what mathematicians do when
+they ``replace equals by equals''. Left premise intuitively states that an
+equation $s \eq t$ holds under condition $C$ (or more precisely, its negation).
+Similarly, the right premise asserts that a (dis)equation holds under a
+condition. Superposition rule simply concatenates the conditions for both
+premises and replaces equals by equals in two main inference terms. For example
+given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and $\neg
+\cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition between their
+last literals results in $\neg
+\cst{even}(X) \llor \neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq
+\cst{b}$.
+
 % \[

### 2022-02-28 introduction/introduction.tex
more

@@
 reasonably simple formal reasoning systems. First-order logic seemed to satisfy
-both requirements.  This formalism does not only allow one to model simple
+both requirements.  This formalism not only allows one to model simple
 logical relations such as ``if it rains, the road is muddy'' (or formally
@@
 Higher-order logic also supports expressing functions that are not explicitly
-present in the theory. Suppose we want to prove that there is a
+present in the theory. Suppose we want to prove there is a
 function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
@@
 the function that adds 2 to its argument. In higher-order logic such unnamed
-function can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
+functions can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
 \cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
@@
 it treats Boolean terms (formulas) as first-class citizens, enabling us to
-model some concepts in more natural manner. For example, we can naturally
+model some concepts in a more natural manner. For example, we can naturally
 represent statements such as ``what Jasmin says is true'' as $\iforall x. \,
@@
   higher-order terms, implement new unification and matching algorithms, extend
-  term indexing data structures to work with higher-order terms, and to improve the
+  term indexing data structures to work with higher-order terms, and improve the
   performance of heuristics on higher-order problems.
@@
   \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
-  decisions that had to be evaluated in short amount of time we decided to
+  decisions that had to be evaluated in a short amount of time we decided to
   implement the calculus in Zipperposition. This prover is less efficient than
@@
   Zipperposition: its unification algorithm. To remove this bottleneck we
-  developed a complete higher-order unification algorithm that was more
-  efficient and generated less redundant unifiers than the state of the art.
+  developed a complete higher-order unification algorithm that is more
+  efficient and generates fewer redundant unifiers than the state of the art.
 
@@
   was lacking compared to competition. We have closely studied the problems on
-  which Zipperposition fails, but are easily solved by other provers and created
+  which Zipperposition fails but are easily solved by other provers, and created
   a set of incomplete rules that enhance Zipperposition's Boolean support.
@@
   Inspired by this success, we implemented the complete superposition
-  calculus for higher-order logic with Booleans (i.e., the third phase) that we developed with Bentkamp and other
+  calculus for higher-order logic with Booleans (i.e., the third phase), that we developed with Bentkamp and other
   colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
@@
 
-  Taking all the experience we gathered while working on Zipperposition,
+  Building on all the experience we gathered while working on Zipperposition,
   we further extended Ehoh to full higher-order logic, obtaining a prover we call
@@
   matching algorithms, extended indexing data structures, and tweaked
-  heuristics. Even though this extension has larger scope than the original
-  extension from E to Ehoh, the strong basis that we have earlier built made it
+  heuristics. Even though this extension has a larger scope than the original
+  extension from E to Ehoh, the strong basis that we had built earlier made it
   manageable.
@@
   basis of higher-order reasoning in our approach, this should improve
-  higher-order performance as well. In particular, we looked at
-  efficient approaches to simplify problems in propositional logic, which is
-  simpler than first-order logic and even decidable. We lifted some of these
-  propositional approaches to first-order logic and implemented them in Zipperposition.
+  higher-order performance as well. In particular, we looked at efficient
+  approaches to simplify problems in propositional logic, which is simpler than
+  first-order logic and even decidable. We lifted some of these propositional
+  approaches to first-order logic and implemented them in Zipperposition.
 
@@
 \noindent\textbf{Chapter~\ref{ch:satfol}} discusses lifting some of the techniques used to
-simplify formulas in propositional logic to first-order logic. It is based on the extended, report version of the publication:
+simplify formulas in propositional logic to first-order logic. It is based on the technical report version of the publication:
   \begin{enumerate}[resume]
@@
 
-As a PhD candidate, I coauthored the following articles and papers, which are not included in this thesis:
+As a PhD candidate, I coauthored the following articles and papers, that not included in this thesis:
 

### 2022-02-28 introduction/introduction.tex
full wan feedback

@@
 colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
-allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
+allows users to specify proof outlines, based on expansion proofs. Benzm\"uller
 and colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
 resolution which introduced the cooperative paradigm. Provers implementing this paradigm
-invoke first-order backends in regular intervals to finish the proof. Latest
-iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
+invoke first-order backends in regular intervals to finish the proof. The latest
+iteration in the LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
 \cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
-to close tableau branches. It won higher-order division of CASC on eight occasions:
+to close tableau branches. It won the higher-order division of CASC on eight occasions:
 in 2011 and every year from 2013 to 2019.
@@
 elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
-some calculi remaing complete if solving a difficult subproblem of higher-order
-unification (``flex-flex'' pairs) is delayed. He described procedure to solve
+some calculi remain complete if solving a difficult subproblem of higher-order
+unification (``flex-flex'' pairs) is delayed. He described a procedure to solve
 this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
@@
 heuristics in first-order theorem proving. For example, Hoder and Voronkov
-evaluate different unification algorithms \cite{hv-09-unifalgs}, while Reger and
-Voronkov evaluate different parameters of a novel prover architecture
+evaluated different unification algorithms \cite{hv-09-unifalgs}, while Reger and
+Voronkov evaluated different parameters of a novel prover architecture
 \cite{rsv-15-playing-with-avatar}. Other authors evaluated different proof

### 2022-02-28 pre/pre.tex
more
full wan feedback

@@
 To interpret the formulas, propositional variables are assigned 0 (false) or 1
-(true), constants $\itrue$ and $\ifalse$ are assigned 1 and 0, respectively; the
-formula is interpreted using the rules for each connective
+(true), and constants $\itrue$ and $\ifalse$ are assigned 1 and 0, respectively; formulas are interpreted using the rules for each connective
 \cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
@@
 describes an algorithm to decide satisfiability (existence of satisfiable
-assignment for a formula) or validity (if all assignments are satisfy the formula).
+assignment for a formula) or validity (if all assignments satisfy the formula).
 
@@
 decide propositional satisfiability problem (SAT solvers) use more advanced
-approaches such as CDCL calculus \cite{mss-96-cdcl}. Modern SAT solvers also
-heavily preprocess the problem and continue simplifying it during proving
+approaches such as the CDCL calculus \cite{mss-96-cdcl}. Modern SAT solvers also
+heavily preprocess the problem and continue simplifying it during the proving
 process.
@@
 First-order logic increases the expressivity by allowing quantification over
-objects and has more complicated formula structure. There are many flavors of
+objects and has a more complicated formula structure. There are many flavors of
 first-order logic, but in this thesis we consider monomorphic first-order logic
@@
 $\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
-\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and that $n$
+\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and $n$
 is the \emph{arity} of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
@@
 
-Terms are used to build \emph{atoms}. Atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
+Terms are used to build \emph{atoms}. An atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
 equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
@@
 logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
-is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
+is a formula and $x$ is a variable, then $\iforall x.\, \phi$, as well as
 $\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
@@
 
-Even though we do not formally define semantics of logic, we assume the natural
+Even though we do not formally define the semantics of logic, we assume the natural
 extensions of domain, valuation, interpretation and model (as defined by Fitting
-\cite{mf-1996-fol}) from unsorted to many-sorted logic. Models we consider are
+\cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
 \emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
@@
 set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
-(i.e., it satisfies it under every variable assignment), we simply write
+(i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
 $\ourmodel \models N.$  Abusing notation, we write $M \models N$ to denote that
@@
 other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
-x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
-Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
+x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to themselves.
+Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$, results in
 replacing all mapped variables by the corresponding terms.
@@
 terms of higher-order logic. For convenience, we still call terms of Boolean
-types \emph{formulas}.
+type \emph{formulas}.
 
@@
 ($\sigma,\varrho,\ldots$) are functions from both free and bound variables to
-terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F) \not\equi F$.
+terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F)\,\not\equi\,F$.
 Substitutions map only finitely many variables. Given a substitution $\varrho$,
@@
 where $a$ is not an application; we call $a$ the \emph{head} of the term. By
-convention, $a, b$ or $\zeta$ denote heads. If $a$ is a free variable, we call the term
+convention, $a, b$ and $\zeta$ denote heads. If $a$ is a free variable, we call the term
 \emph{flex}; otherwise, the term is \emph{rigid}.
@@
 on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
-set of formulas, must be transformed in a set of clauses. For all three
+set of formulas, must be transformed into a set of clauses. For all three
 discussed logics, there exists such a transformation that does not affect
@@
 finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
-\cdots \llor l_n$. Free variables of the clauses are implicitly
+\cdots \llor l_n$. Free variables of clauses are implicitly
 universally quantified. Note that $\llor$ denoting disjunction in clauses
-is different than formula-building $\ior$.
+is different from the formula-building $\ior$.
 
@@
 Superposition is one of the most successful calculi for first-order logic with
-equality. It ows its success to careful treatment of equality and built-in
+equality. It owes its success to a careful treatment of equality and built-in
 heuristics to prune the search space such as term order and selection functions.
-Before we introduce the rules of the calculus, we provide definitions of these and other
+Before introducing the rules of the calculus, we provide definitions of these and other
 background concepts.
@@
 and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
-> S_2(y)$. We use notation $S(x)$ to denote the number of occurrences of $x$ in
-$S$. To use multiset extension, positive literals are represented as $ \{ \{s\},
+> S_2(y)$. We use the notation $S(x)$ to denote the number of occurrences of $x$ in
+$S$. To use the  multiset extension, positive literals are represented as $ \{ \{s\},
 \{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
@@
 
-Two orders that are commonly used in superposition theorem proving are
-Knuth-Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and
+Two orders that are commonly used in superposition theorem proving are the
+Knuth-Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and the
 lexicographic path order (LPO) \cite[Sect.~5.4.2]{bn-98-tr-and-all-that}.
@@
 break eventual ties. LPO entirely relies on precedence and inspects term
-structure more closely to determine the order.
+structures more closely to determine the order.
 
 \newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
-{\em Literal selection function} is a function that selects a (multi)subset of literals.
-Superposition requires that at least one of the selected literals is negative.
+The {\em literal selection function} is a function that selects a (multi)subset of literals
+from a given (multi)set of literals. Superposition requires that at least one of the selected literals is negative.
 
@@
 if two higher-order terms are unifiable, but there exist algorithms
-that enumerate all elements of CSU for unifiable terms.
+that enumerate all elements of the CSU for unifiable terms.
 
 These higher-order concepts gracefully generalize the corresponding first-order
-ones. For example, if $s$ at $t$ are first-order there must exist a singleton
+ones. For example, if $s$ at $t$ are first-order, then there must exist a singleton
 CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
@@
 
-Before we spell out the inference rules of standard, first-order superposition
+Before spelling out the inference rules of standard, first-order superposition
 let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
@@
 \sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
-$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
+$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using a horizontal
 line to separate premises and the conclusion, four inference rules of
@@
 
+\medskip
 \begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
@@
 conclusion $\cst{even}(X)$ to obtain $\cst{even}(\cst{2})$. The \infname{EF}
-rule can be used to find instances of the clause that can further be simplified.
+rule can be used to find instances of the clause that can be further simplified.
 Consider the clause $\cst{f}(X) \eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor
@@
 
-Above described rules are used to generate new clauses from already derived
+The rules described above are used to generate new clauses from already derived
 ones. Even though orderings and selection are used to reduce the search space,
 it still grows rather fast. To identify unnecessary clauses in the search
-space, superposition features \emph{redundancy criterion}
+space, superposition features a \emph{redundancy criterion}
 \cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
@@
 Other than the four generating rules, superposition provers also use
-simplification rules, which are justified using redundancy criterion. These
+simplification rules, which are justified using the redundancy criterion. These
 rules replace the premises with conclusions, and we denote them using double
-bars. As an example of such rule, let us introduce rewriting (demodulation) of
+bars. As an example of such a rule, let us introduce rewriting (demodulation) of
 negative literals (\infname{RN}):
@@
 
-We say that a clause $C$ subsumes clause $D$ is there is a substitution $\sigma$
+We say that a clause $C$ subsumes clause $D$ if there is a substitution $\sigma$
 such that $\sigmacl{C} \subseteq D$. Subsumption is one of the most important
-operations within a theorem prover. This rule does not adhere to redundancy
+operations within a theorem prover. This rule does not adhere to the redundancy
 criterion, but can be justified using other mechanisms \cite{wtrb-20-sat-framework}.

### 2022-02-28 unif/unif.tex
unif

@@
 \label{ch:unif}
+
+\blfootnote{In this work I designed the main algorithm, solid oracle and
+fingerprint indexing. I also implemented and evaluated the algorithms. Visa
+Nummelin proved the completeness of the main algorithm with Alexander Bentkamp's help.
+The completeness and termination of solid oracle was proven by me with Alexander
+Bentkamp's help.}
+
+\authors{Joint work with\\
+Alexander Betnkamp and Visa Nummelin}
+
+\begin{abstract}
+  We developed a procedure to enumerate complete sets of higher-order unifiers based on work
+  by Jensen and Pietrzykowski. Our procedure removes many redundant unifiers by
+  carefully restricting the search space and tightly integrating decision
+  procedures for fragments that admit a finite complete set of unifiers. We
+  identify a new such fragment and describe a procedure for computing its unifiers.
+  Our unification procedure, together with new higher-order term indexing data structures, 
+  is implemented in the Zipperposition theorem prover.
+  Experimental evaluation shows a clear advantage over Jensen and Pietrzykowski's
+  procedure.    
+\end{abstract}
+
+\newpage
+
+\section{Introduction}
+\label{sec:unif:intro}
+
+\section{Background}
+\label{sec:unif:background}
+
+\section{The Unification Procedure}
+\label{sec:unif:the-unification-procedure}
+
+\section{Proof of Completeness}
+\label{sec:unif:proof-of-completeness}
+
+\section{A New Decidable Fragment}
+\label{sec:unif:solid-oracle}
+
+\section{An Extension of Fingerprint Indexing}
+\label{sec:unif:indexing}
+
+\section{Implementation}
+\label{sec:unif:implementation}
+
+\section{Evaluation}
+\label{sec:unif:evaluation}
+
+\section{Discussion and Related Work}
+
+\section{Conclusion}

### 2022-02-28 pre/pre.tex
background

@@
 types $\tau \rightarrow \upsilon$ where both $\tau$ and $\upsilon$ are types.
+When we write $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$
+we assume $\upsilon$ to be a base type ($\upsilon \in T$), by convention.
 Each symbol $\cst{f} \in \Sigma$ is assigned a type.
@@
 We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
-and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
+and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$,
+omitting the length $n \geq 0$ when it is not important or it can be inferred. Every
 $\beta$-reduced term can be written as $\lam{\tuple{x}{m}}{a \, \tuplen{t}}$,
@@
 but $\cst{f}$ and $\cst{f} \, \cst{a}$ are not subterms of $\cst{f} \, \cst{a}
-\, \cst{b}$. We say a term is ground if it has no variables, and closed
+\, \cst{b}$. We say a term is ground if it has no free variables, and closed
 if it has no loosely bound variables.

### 2022-02-28 unif/unif.tex
background

@@
 
+Unification is concerned with finding a substitution that makes two terms
+equal, for some notion of syntactic equality. Since the invention of Robinson's first-order unification
+algorithm \cite{ar-65-resolution}, it has become an indispensable
+tool in theorem proving, logic
+programming, natural language processing, programming language compilation and other areas of computer science.
+
+Many of these applications are based on higher-order formalisms and require
+higher-order unification. Due to its undecidability and explosiveness,
+the higher-order unification problem is considered one of the main
+obstacles on the road to efficient higher-order tools.
+
+One of the reasons for higher-order unification's explosiveness lies in
+\emph{flex-flex pairs}, which consist of two variable-headed terms,
+e.g., $F \, X \unif G \, \cst{a}$. Even this seemingly simple
+problem has infinitely many incomparable unifiers.
+%
+One of the first methods designed to combat this explosion is Huet's
+preunification~\cite{gh-75-unification}. Huet noticed that some logical calculi
+would remain complete if flex-flex pairs are not eagerly solved but postponed as
+constraints. If only flex-flex constraints remain, we know that a unifier must
+exist and we do not need to solve them.
+%
+Huet's preunification has been used in many reasoning tools including Isabelle
+\cite{tn-02-isabelle}, Leo-III \cite{sb-21-leo3}, and Satallax
+\cite{cb-12-satallax}. However, recent developments in higher-order theorem
+proving~\cite{bbtvw-21-sup-lam,br-19-restricted-unif} require full unification---i.e., enumeration of unifiers even for
+flex-flex pairs, which is the focus of this chapter.
+
+Jensen and Pietrzykowski's (JP) procedure \cite{jp-76-unif} is the best known
+procedure for this purpose (Section \ref{sec:unif:background}). Given two terms to
+unify, it first identifies a position where the terms disagree.
+Then, in parallel branches of the search tree, it applies suitable substitutions, involving a
+variable either at the position of disagreement or above, and repeats this process on
+the resulting terms until they are equal or trivially nonunifiable.
+
+Building on the JP procedure, we designed a new procedure (Section
+\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Section \ref{sec:unif:proof-of-completeness}).
+The new procedure addresses many of the issues that are
+detrimental to the performance of the JP procedure.
+%
+First, the JP procedure does not terminate in many cases of obvious
+nonunifiability, e.g., for $X \unif \cst{f} \, X$, where $X$ is a non-functional
+variable and $\cst{f}$ is a function constant. This example also shows that
+the JP procedure does not generalize Robinson's first-order procedure gracefully. To address
+this issue, our procedure detects whether a unification problem belongs to a
+fragment for which unification is decidable and finite complete sets of unifiers (CSUs)
+exist.
+% Another approach for treating explosiveness of higher-order unification is to
+% restrict our attention to some fragment $L$ of $\lambda$-terms that has
+% decidable, unary unification problem. We say that the unification problem is
+% unary if any two terms belonging to $L$ are either not unifiable or have a most
+% general unifier (MGU). 
+We call algorithms that enumerate elements of the CSU for such fragments
+\emph{oracles}. Noteworthy fragments with oracles are first-order terms,
+patterns \cite{tn-93-patterns}, functions-as-constructors
+\cite{tl-16-facunif}, and a new fragment 
+we present in Section~\ref{sec:unif:solid-oracle}.
+%
+The unification procedures of Isabelle and Leo-III check whether the unification
+problem belongs to a decidable fragment, but we take this idea a step further by
+checking this more efficiently and for every subproblem arising during
+unification.
+
+Second, the JP procedure computes many redundant unifiers. Consider the
+example $F \, (G \, \cst{a}) \unif F \, \cst{b}$, where it produces, in addition
+to the desired unifiers $\{F \mapsto \lambda x. \, H\}$ and $\{G \mapsto \lambda
+x. \, \cst{b}\}$, the redundant unifier $\{F \mapsto \lambda x. \, H,\; G \mapsto
+\lambda x. \, x\}$. 
+The design of our procedure avoids computing many redundant unifiers, including
+this one. Additionally, as oracles usually return a small CSU, 
+their integration reduces the number of redundant unifiers.
+
+\newpage
+Third, the JP procedure applies more explosive rules than Huet's preunification procedure to
+flex-rigid pairs. To gracefully generalize Huet's procedure, 
+we show that his rules for flex-rigid pairs suffice 
+to enumerate CSUs
+if combined with appropriate rules for flex-flex pairs.
+
+Fourth, the JP procedure repeatedly traverses the parts of the unification
+problem that have already been unified. Consider the problem $\cst{f}^{100} \,
+(G \, \cst{a}) \unif \cst{f}^{100} \, (H \, \cst{b})$, where the exponents
+denote repeated application. It is easy to see that this problem can be reduced
+to $G \, \cst{a} \unif H \, \cst{b}$. However, the JP procedure will wastefully
+retraverse the common context $ \cst{f}^{100} [\;] $ after applying each new
+substitution. Since the JP procedure must apply substitutions to the variables occurring in the
+common context above the position of disagreement, it cannot be easily adapted to
+eagerly decompose unification pairs. By contrast, our procedure is designed to decompose
+the pairs eagerly, never traversing a common context twice.
+
+Last, the JP procedure does not allow to apply substitutions and
+$\beta$-reduce lazily.
+The rules of simpler procedures (e.g., first-order
+\cite{hv-09-unifalgs} and pattern unification \cite{tn-93-patterns}) depend only on
+the heads of the unification pair.
+Thus, to
+determine the next step, implementations of these procedures need to
+substitute and
+$\beta$-reduce only until the heads of the current unification
+pair are not mapped by the substitution and are not $\lambda$-abstractions. 
+Since the JP procedure is not based on the decomposition of unification pairs, 
+it is unfit for
+optimizations of this kind.
+We designed our procedure to allow for this optimization.
+
+% In Section \ref{sec:unif:indexing} we discuss a higher-order extension of fingerprint indexing,
+% a technique to filter out terms that are not unifiable with a given
+% query term from a set of terms. 
+To more efficiently find terms (in a large term set) that are unifiable with a given query term,
+we developed a higher-order extension of fingerprint indexing \cite{ss-12-fp-indexing}
+(Section \ref{sec:unif:indexing}).
+We implemented our procedure, several oracles,
+and the fingerprint index in
+the Zipperposition prover (Section \ref{sec:unif:implementation}). Since a
+straightforward implementation of the JP procedure already existed in
+Zipperposition, we used it as a baseline to evaluate the performance of our
+procedure (Section \ref{sec:unif:evaluation}). The results show substantial
+performance improvements.
+
+% This invited article is an extended version of our FSCD-2020 paper \cite{vbn-21-unif}.
+% Most notable extension is the Section \ref{sec:unif:proof-of-completeness}, which gives
+% the detailed proof of completeness of our new procedure. In addition, 
+% we give proofs for all the unproved statements from the paper, expand the examples 
+% and provide more detailed explanations.
+
 \section{Background}
 \label{sec:unif:background}
+
+Our setting is the simply typed $\lambda$-calculus. Unless mentioned otherwise,
+we use the same notation as laid out in Chapter~\ref{ch:pre}. Additional notions
+are introduced as follows.
+
+\emph{Parameters} and \emph{body} for any term $\param[\seq x]s$ are defined to
+be $\seq x$ and $s$ respectively, where $s$ is not a $\lambda$-abstraction. The
+\emph{size} of a term is inductively defined as $\text{size}(F) = 1$;
+$\text{size}(x) = 1$; $\text{size}(\cst{f}) = 1$; $\text{size}(s\,t) =
+\text{size}(s) + \text{size}(t)$; $\text{size}(\lambda x.\, s) = \text{size}(s)
++ 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
+$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, bound
+variable, or a constant. In this case, $a$ is called the \emph{head} of the
+term. Note that this relaxes the condition that the term needs to be in
+$\beta$-normal form to determine its head. A term is called flex or rigid if its
+head is flex or rigid, respectively. By $\nf{s}{\textsf{h}}$ we denote the term
+obtained from a term $s$ by repeated $\beta$-reduction of the leftmost outermost
+redex until it is in hnf. Unless stated otherwise, we view terms syntactically,
+as opposed to $\alpha\beta\eta$-equivalence classes. The common context
+$\mathcal{C}(s,t)$ of two $\eta$-long $\beta$-reduced terms $s$~and~$t$ of the
+same type is defined inductively as follows, assuming that $a \not= b$:
+$\mathcal{C}(\lambda x.\, s, \lambda y.\, t) = \lambda x.\,
+\mathcal{C}(s,\{y\mapsto x\}t)$; $\mathcal{C}(a\,\overline{s}_m,
+b\,\overline{t}_n) = \square$; $\mathcal{C}(a\,\overline{s}_m,
+a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
+Unless otherwise stated, we take unification constraint $s \unif t$ to be an
+unordered pair of two terms of the same type. To ease notation, we do not write
+parentheses around application of substitutions to terms (or other objects containing terms);
+in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$
+
+\ourpara{Remark}
+We use the definition of a CSU from Sect.~\ref{sec:pre:unif} because 
+JP's definition of a CSU, which we have adopted in our earlier work \cite{vbn-21-unif}, is flawed.
+JP's definition does not employ the notion of auxiliary variables,
+but instead requires $\rho X = \theta\sigma X$ for all variables mapped by $\rho$.
+This is problematic because nothing prevents $\rho$ from mapping the auxiliary variables.
+For example, $\sigma = \{F \mapsto \lambda x y.\> G\>y\}$ is supposed
+to be an MGU for $F\>\cst{a}\>\cst{c} \unif F\>\cst{b}\>\cst{c}$.
+But for the unifier $\rho = \{F \mapsto \lambda x y.\> y,\> G \mapsto \lambda x.\> \cst{d}\}$,
+without the notion of auxiliary variables,
+there exists no appropriate substitution $\theta$
+because $\rho G = \theta\sigma G$ requires $\theta G = \lambda x.\> \cst{d}$
+and $\rho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$.
 

### 2022-03-01 unif/unif.tex
more

@@
 parentheses around application of substitutions to terms (or other objects containing terms);
-in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$
+in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$.
 

### 2022-03-01 unif/unif.tex
procedure section;

@@
 
+To unify two terms $s$ and $t$, our procedure builds a tree as follows. The
+nodes of the tree have the form $(E,\sigma)$, where $E$ is a multiset of
+unification constraints $\{(s_1 \unif t_1), \ldots, (s_n~\unif~t_n)\}$ and
+$\sigma$ is the substitution constructed up to that point.  The root node is $(\{s~\unif~t\}, \makeop{id})$, where $\makeop{id}$ is the identity
+substitution. The tree is then constructed applying the transitions listed
+below. The leaves of the tree are either failure nodes  $\bot$ or substitutions
+$\sigma$. Ignoring failure nodes, the set of all substitutions in the leaves
+forms a complete set of unifiers for $s$ and $t$. More generally, our procedure can be used to
+unify a multiset $E$ of constraints by making the root of the unification tree
+$(E, \makeop{id})$.
+
+The procedure requires an infinite supply of fresh free variables. These fresh
+variables must be disjoint from the variables occurring in the initial multiset
+$E$. Whenever a transition $(E,\sigma) \newunifarrow (E',\sigma')$ is made, all
+fresh variables used in $\sigma'$ are removed from the supply and cannot be used
+again as fresh variables.
+
+\looseness=-1
+The transitions are parametrized by a mapping~$\mathcal{P}$ that assigns a set
+of substitutions to a unification pair; this mapping abstracts the
+concept of unification rules present in other unification procedures. Moreover,
+the transitions are parametrized by a selection function~$S$ mapping a multiset
+$E$ of unification constraints to one of those constraints $S(E) \in E$, the
+\emph{selected} constraint in $E$. The transitions, defined as follows, are only
+applied if the $\selected{\text{grayed}}$ constraint is selected.
+\begin{description}[labelwidth=\widthof{\rm\textsf{Normalize$_{\alpha\eta}$}}]
+  \item[\rm\unifrulename{Succeed}]
+      $(\varnothing, \sigma) \newunifarrow \sigma$ 
+  % \smallskip
+  \item[\rm\unifrulename{Normalize$_{\alpha\eta}$}]
+      $(\{\selected{\lambda \overline{x}_m. \, s \unif \lambda \overline{y}_n. \, t}\}\uplus E, \sigma) 
+      \newunifarrow 
+      (\{\lambda \overline{x}_m. \, s \unif \lambda \overline{x}_m. \, t'\, x_{n+1} \ldots x_m\}\uplus E, \sigma)$\\
+      % where $n \geq m$ and $t'=\{y_1\mapsto x_1,\ldots,y_m\mapsto x_m\}t$
+      where $m \geq n$, $\overline{x}_m \not= \overline{y}_n$, and $t'=\{y_1\mapsto x_1,\ldots,y_n\mapsto x_n\}t$
+  % \smallskip
+  \item[\rm\unifrulename{Normalize$_\beta$}]
+      $(\{\selected{\lambda \overline{x}. \, s \unif \lambda \overline{x}. \, t}\}\uplus E, \sigma) 
+      \newunifarrow 
+      (\{\lambda \overline{x}. \, \nf{s}{\textsf{h}} \unif \lambda \overline{x}. \, \nf{t}{\textsf{h}}\}\uplus E, \sigma)$\\
+      where $s$ or $t$ is not in hnf
+      % \smallskip
+  \item[\rm\unifrulename{Dereference}]
+      $(\{\selected{\lambda \overline{x}. \, F \, \overline{s} \unif \lambda \overline{x}. \, t}\}\uplus E, \sigma)
+      \newunifarrow 
+      (\{\lambda \overline{x}. \, (\sigma F) \,\overline{s} \unif \lambda \overline{x}. \, t\}\uplus E, \sigma)$\\
+      where none of the previous transitions apply and $F$ is mapped by $\sigma$
+  % \smallskip
+      \item[\rm\unifrulename{Fail}]
+      $(\{\selected{\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda \overline{x}. \, b \, \overline{t}_n}\}
+      \uplus E, \sigma) \newunifarrow \bot$\\
+      where none of the previous transitions apply, and
+      $a$ and $b$ are different rigid heads
+  % \smallskip
+  \item[\rm\unifrulename{Delete}]  $(\{\selected{s \unif s}\}\uplus E, \sigma) \newunifarrow (E, \sigma)$\\
+      where none of the previous transitions apply
+  % \smallskip 
+  \item[\rm\unifrulename{OracleSucc}]
+  $(\{\selected{s \unif t}\}\uplus E, \sigma) \newunifarrow (E, \varrho \sigma)$\\
+      where none of the previous transitions apply, 
+      some oracle found a finite CSU $U$ for $\sigma s \unif \sigma t $ using fresh auxiliary variables,
+      and $\varrho \in U$; if multiple oracles found a CSU, only one of them is considered
+  % \smallskip
+  \item[\rm\unifrulename{OracleFail}]
+  $(\{\selected{s \unif t}\}\uplus E, \sigma) \newunifarrow \bot$\\
+  where none of the previous transitions apply, 
+  and some oracle determined $\sigma s \unif \sigma t$ has no solutions
+  % \smallskip
+  \item[\rm\unifrulename{Decompose}]
+      $(\{\selected{\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda \overline{x}. \, a \, \overline{t}_m}\}
+      \uplus E, \sigma) 
+      \newunifarrow 
+      ( \{s_1 \unif t_1, \ldots ,s_m \unif t_m\} \uplus E,
+      \sigma)$\\
+      where none of the transitions \unifrulename{Succeed} to \unifrulename{OracleFail} apply
+  % \smallskip
+  % \item [\rm\textsf{Eliminate}] 
+  %     $(\{\selected{\lambda \overline{x}_n. \, F \, \overline{s}_m \unif \lambda \overline{x}_n. \, F \, \overline{t}_m}\}
+  %     \uplus E, \sigma)
+  %     \newunifarrow
+  %     (\{s_{j_{1}}\unif t_{j_{1}},\ldots,s_{j_{k}}\unif t_{j_{k}}\} \uplus E,\varrho \sigma)$\\
+  %     where none of the transitions \textsf{Succeed} to \textsf{OracleFail} apply,
+  %     $ 1 \leq j_1 < \cdots < j_k \leq n$, and
+  %     % $(j_{i})_{i=1}^{k}$ is a subsequence
+  %     % of $\left(i\right)_{i=1}^{m}$, and
+  %     $\varrho = \{ F \mapsto \param G\, x_{j_{1}} \ldots x_{j_{k}} \}$ with $G$ a fresh variable
+  % \smallskip
+  \item[\rm\textsf{Bind}]
+      $(\{\selected{s \unif t}\} 
+      \uplus E, \sigma) 
+      \newunifarrow 
+      (\{s \unif t\}
+      \uplus E, \rho \sigma)$\\
+      where none of the transitions \textsf{Succeed} to \textsf{OracleFail} apply,
+      and $\rho\in\mathcal{P}(s \unif t)$.
+\end{description}
+
+The transitions are designed so that only \unifrulename{OracleSucc},
+\unifrulename{Decompose}, and \unifrulename{Bind} can introduce parallel branches in the
+constructed tree. \unifrulename{OracleSucc} can introduce branches
+using different unifiers of the CSU, \unifrulename{Bind} can introduce branches 
+using different substitutions in $\mathcal{P}$, and \unifrulename{Decompose}
+can be applied in parallel with \unifrulename{Bind}.
+
+The form of the rules \unifrulename{OracleSucc} and \unifrulename{Bind} is similar: both
+extend the current substitution. However,
+they are designed following different principles. \unifrulename{OracleSucc} solves
+the selected unification constraint using an efficient algorithm applicable only
+to certain classes of terms. On the other hand, \unifrulename{Bind} is applied to
+explore the whole search space for any given constraint. These rules are separated
+in two to make \unifrulename{Bind} applicable only if \unifrulename{OracleSucc}
+(or \unifrulename{OracleFail}) is not, so that possible solutions (or failures) are
+detected early.
+
+Our approach is to apply substitutions and $\alpha\beta\eta$-normalize terms
+lazily. In this context, laziness means that the transitions
+\unifrulename{Normalize$_{\alpha\eta}$}, \unifrulename{Normalize$_{\beta}$}, and
+\unifrulename{Dereference} partially normalize and partially apply the constructed
+substitution just enough to ensure that the heads are the ones we would get if
+the substitution was fully applied and the term was fully normalized. 
+Additionally, the transitions that modify the constructed substitution,
+\unifrulename{OracleSucc} and \unifrulename{Bind}, do not apply that substitution to the
+unification pairs directly, but only extend it with a new binding. To support
+lazy dereferencing, these rules must maintain the
+invariant that all substitutions are idempotent. The invariant is easily
+preserved if the substitution $\varrho$ from the definition of
+\unifrulename{OracleSucc} and \unifrulename{Bind} is itself idempotent and no variable
+mapped by $\sigma$ occurs in $\varrho F$, for any variable $F$ mapped by
+$\varrho$.
+
+The \textsf{OracleSucc} and \textsf{OracleFail} transitions
+invoke oracles, such as pattern unification,
+to compute a CSU faster,
+produce fewer redundant unifiers, and
+discover nonunifiability earlier.
+In some cases, addition of oracles lets the procedure terminate more often.
+
+In the literature, oracles are usually stated under the assumption that their
+input belongs to the appropriate fragment. To check whether a unification
+constraint is inside the fragment, we need to fully apply the substitution and
+$\beta$-normalize the constraint.
+% To use an off-the-shelf oracle
+% in our procedure, we need to wrap it inside an algorithm that applies the currently
+% built substitution $\sigma$, normalizes terms and checks whether terms belong to the
+% fragment the oracle handles, before invoking the oracle. Applying substitution and
+% normalizing eagerly to invoke an oracle can inadvertently cancel the effects of
+% lazy normalization built into our procedure.
+To avoid these expensive operations and enable efficient oracle integration, 
+oracles must be redesigned to lazily discover whether the terms
+belong to their fragment.  
+% Usually, the operations
+% oracles perform can clearly be separated into those that do and those that do
+% not depend on whether the terms belong to the fragment oracle manages. 
+Most oracles contain a decomposition operation which requires only a partial application of
+the substitution and only partial $\beta$-normalization. If one of the constraints resulting from decomposition is not in
+the fragment, the original problem is not in the fragment. 
+This allows us to detect that the problem is not in the fragment without fully applying
+the substitution and $\beta$-normalizing.
+% Thus, without checking if all subterms are in the fragment, the oracle
+% can check if the term heads are equal after getting terms in hnf. This
+% effectively delegates the fragment check to the
+% other operations of the oracle.
+
+%  If some
+% other transitions applies, it is the only applicable transition. Transitions
+% \textsf{Bind} and \textsf{OracleSucc} may be applicable with different partial
+% bindings or unifiers.
+
+%
+%Mention log on variables?
+%
+The core of the procedure lies in the \textsf{Bind} step, parameterized
+by the mapping~$\mathcal{P}$ that determines which substitutions (\emph{bindings}) 
+to create. The bindings are defined as follows:
+%
+% \begin{addmargin}[0.5em]{0.5em}
+%\vspace{-0.2em}
+\begin{description}[itemsep=1\jot]
+    \item[JP-style projection for $F$] Let $F$ be a free variable of
+    type $\alpha_1 \rightarrow \cdots \rightarrow \alpha_n \rightarrow \beta$, where
+    some $\alpha_i$ is equal to $\beta$ and $n > 0$. Then the JP-style projection binding is
+        \[F \mapsto \lambda \overline{x}_n.\,x_i\]
+    \item[Huet-style projection for $F$] Let $F$ be a free variable of type $\alpha_1
+    \rightarrow \cdots \rightarrow \alpha_n \rightarrow \beta$, where some $\alpha_i
+    = \gamma_1 \rightarrow \cdots \rightarrow \gamma_m \rightarrow \beta$, $n > 0$ and $m \geq 0$. Huet-style projection is
+    \[F \mapsto \lambda \overline{x}_n. \, x_i \, (F_1 \, \overline{x}_n) \, \ldots \, (F_m \, \overline{x}_n)\]
+    where the fresh free variables $\overline{F}_m$ and bound variables $\overline{x}_n$ are of appropriate types.
+    \item[Imitation of $a$ for $F$] 
+    \looseness=-1 Let $F$ be a free variable of type $\alpha_1 \rightarrow
+    \cdots \rightarrow \alpha_n \rightarrow \beta$ and $a$ be a free variable or a constant
+    of type $\gamma_1 \rightarrow \cdots
+    \rightarrow \gamma_m \rightarrow \beta$ where $n,m \geq 0$. The imitation binding is
+    \[F \mapsto \lambda \overline{x}_n. \, a \, (F_1 \, \overline{x}_n) \ldots
+    (F_m \, \overline{x}_n)\] where the fresh free variables $\overline{F}_m $ and
+    bound variables $\overline{x}_n$ are of appropriate types.
+    \item[Elimination for $F$] Let $F$ be a free variable of type $\alpha_1 \rightarrow
+    \cdots \rightarrow \alpha_n \rightarrow \beta$, where $n >0$. In addition, let $1 \leq j_1 < \cdots < j_i \leq n$ and $i<n$. Elimination
+    for the sequence $(j_k)_{k=1}^i$ is
+    \[ F \mapsto \lambda \overline{x}_n. \, G \, x_{j_1} \, \ldots \, x_{j_i}\]
+    where the fresh free variable $G$ as well as all $x_{j_k}$ are of appropriate type.
+    We call fresh variables emerging from this binding in the role of $G$ 
+    \emph{elimination variables}.
+    \item[Identification for $F$ and $G$] Let $F$ and $G$ be different free variables. Furthermore, let
+    the type of $F$ be $\alpha_1 \rightarrow \cdots \rightarrow \alpha_n
+    \rightarrow \beta$ and the type of $G$ be $\gamma_1 \rightarrow \cdots
+    \rightarrow \gamma_m \rightarrow \beta$, where $n,m\geq 0$. Then, the identification binding binds
+    $F$ and $G$ with
+    \begin{equation*}
+        F \mapsto \lambda \overline{x}_n. \, H \, \overline{x}_n \, (F_1 \, \overline{x}_n)
+        \ldots (F_m \, \overline{x}_n) \quad
+        G \mapsto \lambda \overline{y}_m. \,
+        H \, (G_1 \, \overline{y}_m) \ldots (G_n \, \overline{y}_m) \, \overline{y}_m
+    \end{equation*}
+    \looseness=-1
+    where the fresh free variables $H,\overline{F}_m,\overline{G}_n$ and bound
+    variables $\overline{x}_n$,$\overline{y}_m$ are of appropriate types. 
+    Fresh variables from this binding with the role of $H$ are called \emph{identification variables}.
+    \item[Iteration for $F$]
+    \looseness=-1 
+    Let $F$ be a free variable of the type $\alpha_1 \rightarrow \cdots
+    \rightarrow \alpha_n \rightarrow \beta_1$ and let some $\alpha_i$ be the
+    type $\gamma_1 \rightarrow \cdots \rightarrow \gamma_m \rightarrow
+    \beta_2$, where $n>0$ and $m\geq 0$. Iteration for $F$ at $i$ is  
+    \[ F \mapsto \lambda \overline{x}_n.\,H\,\overline{x}_n \, (\lambda
+    \overline{y}.\,x_i \, (G_1 \, \overline{x}_n \, \overline{y}) \ldots (G_m \,
+    \overline{x}_n \, \overline{y}) )\]
+    The free variables $H$ and $G_1, \ldots,
+    G_m$ are fresh, and $\overline{y}$ is an arbitrary-length sequence of bound
+    variables of arbitrary types. All new variables are of
+    appropriate type. Due to indeterminacy of $\overline{y}$, this step is
+    infinitely branching.
+
+\end{description}
+
+\noindent
+The following mapping $\mathcal{P}_\mathsf{c}(\lambda \overline{x}.
+\, s \unif \lambda \overline{x}. \, t)$ is used as the parameter $\mathcal{P}$ of the procedure:
+\begin{itemize}
+  \setlength\itemsep{1\jot}
+
+    \item If the constraint is rigid-rigid, $\mathcal{P}_\mathsf{c}(\lambda \overline{x}.
+    \, s \unif \lambda \overline{x}. \, t) = \varnothing$.
+
+    \item If the constraint is flex-rigid,
+    let 
+    $\mathcal{P}_\mathsf{c}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, a\, \overline{t})$ 
+    be 
+    \begin{itemize}
+        \item an imitation of $a$ for $F$, if $a$ is a constant, and
+        \item all Huet-style projections for $F$, if $F$ is not an identification variable.
+    \end{itemize}
+
+    \item If the constraint is flex-flex and the heads are different,
+    let 
+    $\mathcal{P}_\mathsf{c}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, G\, \overline{t})$ 
+    be 
+    \begin{itemize}
+        \item all identifications and iterations for both $F$ and $G$, and
+        \item all JP-style projections for non-identification variables among $F$ and $G$.
+    \end{itemize}
+
+    \item If the constraint is flex-flex and the heads are identical,
+    we distinguish two cases:
+    \begin{itemize}
+        \item if the head is an elimination variable, 
+        $\mathcal{P}_\mathsf{c}(\lambda \overline{x}.\, s \unif \lambda \overline{x}.\, t ) = \varnothing$;
+        \item otherwise, let 
+        $\mathcal{P}_\mathsf{c}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, F\, \overline{t})$ 
+        be all iterations for $F$ at arguments of functional type and
+        all eliminations for~$F$.
+    \end{itemize}
+\end{itemize}
+
+\ourpara{Comparison with the JP Procedure}
+%
+% Our procedure is based on the JP procedure, which is the best known procedures that
+% enumerates a CSU \cite{jp-76-unif}. Unlike other complete unification procedures
+% \cite{sg-89-unif}, it does not try to blindly guess the solution of flex-flex
+% unification constraints, but enumerates them systematically. 
+%
+\looseness=-1
+The JP procedure enumerates unifiers by constructing a search tree with nodes of the form $(s \unif t, \sigma)$, where $s \unif t$ is the current unification
+problem and $\sigma$ is the substitution built so far. 
+The initial node consists of the input problem and the
+identity substitution. Success nodes are nodes of the form $(s \unif s, \sigma)$.
+The set of all substitutions contained in the success nodes form a CSU.
+
+To determine the
+child nodes of a node $(s \unif t, \sigma)$, the procedure  computes the common context $C$ of $s$ and $t$,
+yielding term pairs $(s_1, t_1), \ldots, (s_n, t_n)$, called \emph{disagreement
+pairs}, such that $s = C[s_1,\ldots,s_n]$ and $t = C[t_1,\ldots,t_n]$. It
+chooses one of the disagreement pairs $(s_i, t_i)$.
+Depending on the context $C$ and the chosen disagreement pair $(s_i, t_i)$,
+it determines a set of bindings $\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$.
+For each of the bindings $\rho\in\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$, it creates a child node
+$(\nf{(\varrho{s})}{\beta\eta} \unif \nf{(\varrho{t})}{\beta\eta}, \varrho\sigma)$,
+where $\nf{u}{\beta\eta}$ denotes a
+$\beta\eta$-normal form of a term $u$.
+
+The set of bindings $\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$
+is based on the heads of $s_i$ and $t_i$, and the free variables occurring above
+$s_i$ and $t_i$ in $C$. The set $\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$ contains
+\begin{itemize}
+  \item all JP-style projections for free variables that are heads of $s_i$ or $t_i$;%
+\footnote{In JP's formulation of projection, they explicitly mention that the projected argument must be of base type. 
+In our presentation, this follows from $\beta$ being of base type by the convention introduced in Sect.~\ref{sec:pre:hol}.}
+
+  \item an imitation of $a$ for $F$ if a free variable $F$ is the head of $s_i$ and a free
+  variable or constant $a$ is the head of $t_i$ (or vice versa);
+  \item all eliminations for free variables occurring above the
+  chosen disagreement pair
+  eliminating only the argument containing the disagreement pair;
+  \item an identification for the heads of $s_i$ and $t_i$ if they are both free variables; and
+  \item all iterations for the heads of $s_i$ and
+  $t_i$ if they are free variables, and for all free variables occurring above the disagreement pair.%
+\footnote{In JP's formulation of iteration, it is not immediately obvious whether they intend to require iteration of arguments of base type.
+However, their Definition 2.4 \cite{jp-76-unif} shows that they do.}
+\end{itemize}
+
+Architecturally, the most noticeable difference between the JP procedure and ours is
+the representation of the problem: The JP procedure works on a single constraint,
+while our procedure maintains a multiset of constraints. At a first glance, this
+is a merely presentational change. However, it has consequences for termination,
+performance, and redundancy of the procedure.
+
+\looseness=-1
+Since the JP procedure never decomposes the common context of its only constraint, it allows
+iteration or elimination to be applied at a free variable above the disagreement
+pair, even if  bindings were already applied below that free variable. This can
+lead to many different paths to the same unifier. In contrast, our procedure
+makes the decision which binding to apply to a flex-flex pair with the same
+head as soon as it is observed. Also, it explores the possibility of not
+applying a binding and decomposing the pair. In either way, the flex-flex pair
+is never revisited, which improves the performance and returns fewer redundant
+unifiers. We show that this restriction prunes the search space without
+influencing the completeness.
+
+Our procedure makes the choice of child nodes based only on the heads of the
+chosen unification constraint. In contrast, the JP procedure tracks all the
+variables occurring in the common context. Thus, lazy normalization and lazy
+variable substitution cannot be integrated in the JP procedure a straightforward
+fashion. Moreover, as it does not feature a rule similar to
+\unifrulename{Decompose}, it always retraverses the already unified part of the
+problem, resulting in poor performance on deep terms.
+
+\looseness=-1
+% The JP procedure can be modified to solve the preunification problem by making
+% it never choose flex-flex disagreement pairs, ignore the disagreement pairs
+% below a free variable, and terminate with a preunifier when no flex-rigid pair
+% remains. However, such a procedure would be less efficient than Huet's procedure
+% because it would use the iteration binding instead of Huet-style projection to
+% solve flex-rigid pairs. Our procedure applies Huet-style projections on
+% flex-rigid pairs, which results in two important improvements over the JP
+% procedure. First, our procedure terminates more often than the JP procedure
+% because Huet-style projections cause only a finite branching, whereas iteration
+% causes an infinite branching. Second, when our procedure is modified to solve
+% the preunification problem by never selecting flex-flex pairs and stopping when
+% only flex-flex pairs remain, it becomes an optimized variant of Huet's procedure
+% that supports oracles as well as lazy substitution and $\beta$-reduction.
+One of the main drawbacks of the JP procedure is that it features a highly
+explosive, infinitely branching iteration rule. This rule is a more general
+version of Huet-style projection. Its universality enables finding elements of
+CSU for flex-flex pairs, for which Huet-style projection does not suffice.
+However, the JP procedure applies iteration indiscriminately on both flex-flex and
+flex-rigid pairs. We discovered that our procedure remains complete if iteration
+is applied only on flex-flex pairs, and Huet-style projection only on
+flex-rigid ones. This helps our procedure terminate more often than the JP
+procedure. As a side-effect, the restriction of our procedure to
+the preunification problem is a graceful generalization of Huet procedure, with
+additional improvements such as oracles, lazy substitution, and lazy
+$\beta$-reduction.
+
+\looseness=-1
+The bindings of our procedure contain further optimizations that are absent in
+the JP procedure: The JP procedure applies eliminations for one parameter
+at a time, yielding multiple paths to the same unifier. It applies imitations to
+flex-flex pairs, which we found to be unnecessary. Similarly, we found out that
+tracking which rules introduced which variables can avoid computing redundant
+unifiers: It is not necessary to apply iterations and eliminations on
+elimination variables, and projections on identification variables.
+
+\ourpara{Examples}
+We present some examples that demonstrate advantages of our procedure.
+The displayed branches of the constructed trees are not necessarily exhaustive.
+We abbreviate 
+JP-style projection as \textsf{JP\,Proj},
+imitation as \textsf{Imit},
+identification as \textsf{Id},
+\unifrulename{Decompose} as \textsf{Dc}, 
+\unifrulename{Dereference} as \textsf{Dr},
+$\textsf{Normalize}_\beta$ as $\textsf{N}_\beta$,
+and
+\textsf{Bind} of a binding $x$ as $\textsf{B(}x\textsf{)}$.
+Transitions of the JP procedure are denoted by $\Longrightarrow$.
+For the JP transitions we implicitly apply the generated bindings and fully
+normalize terms, which significantly shortens JP derivations.
+
+\begin{exa}
+The JP procedure does not terminate on the problem $G \unif \cst{f} \, G$:
+\[(G \unif \cst{f} \, G, \makeop{id}) 
+\overset{\textsf{Imit}}{\jpunifarrow}
+(\cst{f}\,G' \unif \cst{f}^2 \, G', \sigma_1) 
+\overset{\textsf{Imit}}{\jpunifarrow}
+(\cst{f}^2 \, G'' \unif \cst{f}^3 \, G'', \sigma_2) 
+\overset{\textsf{Imit}}{\jpunifarrow}\cdots\]
+where $\sigma_1 = \{G \mapsto  \lambda x. \, \cst{f} \, G' \}$
+and $\sigma_2 = \{G' \mapsto  \lambda x. \, \cst{f} \, G'' \}\sigma_1$.
+By including any oracle that supports the first-order occurs check, such as the
+pattern oracle or the fixpoint oracle described in Section \ref{sec:unif:implementation},
+our procedure gracefully generalizes first-order unification: 
+\[(\{G \unif \cst{f} \, G\}, \makeop{id}) 
+\overset{\textsf{OracleFail}}{\newunifarrow}
+\bot
+\]
+\end{exa}
+
+\begin{exa}
+The following derivation illustrates the advantage of the \textsf{Decompose} rule.
+\begin{align*}
+&
+(\{\cst{h}^{100} \, (F \, \cst{a}) \unif \cst{h}^{100} \, (G \,\cst{b})\}, \makeop{id}) 
+\overset{\textsf{Dc}^{100}}{\newunifarrow}
+(\{F \, \cst{a} \unif G \,\cst{b}\}, \makeop{id}) 
+\overset{\textsf{B(Id)}}{\newunifarrow}
+(\{F \, \cst{a} \unif G \,\cst{b}\}, \sigma_1) 
+\\ &
+\overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+(\{H \, \cst{a}\, (F' \, \cst{a}) \unif H \, (G'\,\cst{b}) \,\cst{b}\}, \sigma_1)
+\overset{\textsf{Dc}}{\newunifarrow}
+(\{\cst{a} \unif G'\,\cst{b}, F' \, \cst{a} \unif \cst{b}\}, \sigma_1) 
+\\ &
+\overset{\textsf{B(Imit)}}{\newunifarrow}
+(\{\cst{a} \unif G'\,\cst{b}, F' \, \cst{a} \unif \cst{b}\}, \sigma_2) 
+\overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+(\{\cst{a} \unif \cst{a}, F' \, \cst{a} \unif \cst{b}\}, \sigma_2) 
+\overset{\textsf{Delete}}{\newunifarrow}
+(\{F' \, \cst{a} \unif \cst{b}\}, \sigma_2) \\
+&\overset{\textsf{B(Imit)}}{\newunifarrow}
+(\{F' \, \cst{a} \unif \cst{b}\}, \sigma_3) 
+\overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+(\{\cst{b} \unif \cst{b}\}, \sigma_3) 
+\overset{\textsf{Delete}}{\newunifarrow}
+(\varnothing, \sigma_3) 
+\overset{\textsf{Succeed}}{\newunifarrow}
+\sigma_3 
+\end{align*}
+where 
+$\sigma_1 = \{
+    F\mapsto \lambda x. \, H \, x\, (F' \, x), 
+    G\mapsto \lambda y. \, H \, (G' \, y) \, y
+    \}$;
+$\sigma_2 = \{
+  G' \mapsto \lambda x. \, \cst{a}
+  \}\sigma_1$; and
+$\sigma_3 = \{
+  F' \mapsto \lambda x. \, \cst{b}
+  \}\sigma_2$.
+%
+The JP procedure produces the same intermediate substitutions $\sigma_1$ to $\sigma_3$,
+but since it does not decompose the terms, it retraverses the common context
+$\cst{h}^{100}\,[\;]$ at every step to identify the contained disagreement pair:
+\begin{align*}
+    &
+    (\cst{h}^{100} \, (F \, \cst{a}) \unif \cst{h}^{100} \, (G \,\cst{b}), \makeop{id}) 
+    \overset{\textsf{Id}}{\jpunifarrow}
+    (\cst{h}^{100} \, (H \, \cst{a}\, (F' \, \cst{a})) \unif \cst{h}^{100} \, (H \, (G' \,\cst{b}) \,\cst{b}), \sigma_1) 
+    \\ &
+    \overset{\textsf{Imit}}{\jpunifarrow}
+    (\cst{h}^{100} \, (H \, \cst{a}\, (F' \, \cst{a})) \unif \cst{h}^{100} \, (H \, \cst{a} \,\cst{b}), \sigma_2) 
+    \overset{\textsf{Imit}}{\jpunifarrow}
+    (\cst{h}^{100} \, (H \, \cst{a}\, \cst{b}) \unif \cst{h}^{100} \, (H \, \cst{a} \,\cst{b}), \sigma_3) 
+    \overset{\textsf{Succeed}}{\jpunifarrow}
+    \sigma_3
+\end{align*}
+
+\end{exa}
+
+\begin{exa}
+  Even when no oracles are used, our procedure performs better than the JP procedure on small, simple
+  problems. Consider the problem $F \, \cst{a} \unif \cst{a}$, which has a two
+  element CSU: $\{ F \mapsto \lambda x. \, x, F \mapsto \lambda x. \, \cst{a}
+  \}$. Our procedure terminates, finding both unifiers:
+\begin{align*}
+    &
+    (\{ F \, \cst{a} \unif \cst{a} \}, \makeop{id})
+    \overset{\textsf{B(JP Proj)}}{\newunifarrow}
+    (\{ F \, \cst{a} \unif \cst{a} \},  \{ F \mapsto \lambda x. \, x \} )
+    \overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+    (\{ \cst{a} \unif \cst{a} \}, \{  F \mapsto \lambda x. \, x \})
+    \\ &
+    \overset{\textsf{Delete}}{\newunifarrow}
+    (\varnothing, \{  F \mapsto \lambda x. \, x \})
+    \overset{\textsf{Succeed}}{\newunifarrow}
+    \{  F \mapsto \lambda x. \, x \}
+    \\[1\jot]
+    &
+    (\{ F \, \cst{a} \unif \cst{a} \}, \makeop{id})
+    \overset{\textsf{B(Imit)}}{\newunifarrow}
+    (\{ F \, \cst{a} \unif \cst{a} \},  \{ F \mapsto \lambda x. \, \cst{a} \} )
+    \overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+    (\{ \cst{a} \unif \cst{a} \}, \{  F \mapsto \lambda x. \, \cst{a} \})
+    \\ &
+    \overset{\textsf{Delete}}{\newunifarrow}
+    (\varnothing, \{  F \mapsto \lambda x. \, \cst{a} \})
+    \overset{\textsf{Succeed}}{\newunifarrow}
+    \{  F \mapsto \lambda x. \, \cst{a} \}
+\end{align*}
+
+The JP procedure finds those two unifiers as well, but it does not terminate as
+it applies iterations to $F$.
+
+\end{exa}
+
+\begin{exa}
+The search space restrictions also allow us to prune some redundant unifiers. Consider the problem $F \, (G\,\cst{a}) \unif F \,
+\cst{b}$, where $\cst{a}$ and $\cst{b}$ are of base type. 
+Our procedure produces only one failing branch and the following two successful branches:
+%
+\begin{align*}
+&
+(\{ F\,(G\,\cst{a}) \unif F \, \cst{b} \}, \makeop{id})
+\overset{\textsf{Dc}}{\newunifarrow}
+(\{ G\,\cst{a} \unif \cst{b} \}, \makeop{id})
+\overset{\textsf{B(Imit)}}{\newunifarrow}
+(\{ G\,\cst{a} \unif \cst{b} \}, \{ G \mapsto \lambda x.\, \cst{b} \})
+\\ &
+\overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+(\{ \cst{b} \unif \cst{b} \}, \{ G \mapsto \lambda x.\, \cst{b} \})
+\overset{\textsf{Delete}}{\newunifarrow}
+(\varnothing, \{ G \mapsto \lambda x.\, \cst{b} \})
+\overset{\textsf{Succeed}}{\newunifarrow}
+\{ G \mapsto \lambda x.\, \cst{b} \}
+\\[1\jot]
+&
+(\{ F\,(G\,\cst{a}) \unif F \, \cst{b} \}, \makeop{id})
+\overset{\textsf{B(Elim)}}{\newunifarrow}
+(\{ F\,(G\,\cst{a}) \unif F \, \cst{b} \}, \{ F \mapsto \lambda x.\, F' \})
+\\ &
+\overset{\textsf{Dr}+\textsf{N}_\beta}{\newunifarrow}
+(\{ F' \unif F' \}, \{ F \mapsto \lambda x.\, F' \})
+\overset{\textsf{Delete}}{\newunifarrow}
+(\varnothing, \{ F \mapsto \lambda x.\, F' \})
+\overset{\textsf{Succeed}}{\newunifarrow}
+\{ F \mapsto \lambda x.\, F' \}
+\end{align*}
+%
+The JP procedure additionally produces the following redundant unifier:
+%
+\begin{align*}
+    &
+    ( F\,(G\,\cst{a}) \unif F \, \cst{b} , \makeop{id})
+    \overset{\textsf{JP Proj}}{\jpunifarrow}
+    ( F \, \cst{a} = F \, \cst{b} , \{ G \mapsto \lambda x. \, x \})
+    \\ &
+    \overset{\textsf{Elim}}{\jpunifarrow}
+    (F' = F' , \{ G \mapsto \lambda x. \, x, F \mapsto \lambda x. \, F' \})
+    \overset{\textsf{Succeed}}{\jpunifarrow}
+    \{ G \mapsto \lambda x. \, x, F \mapsto \lambda x. \, F' \}
+\end{align*}
+Moreover, the JP procedure does not terminate because an infinite number of iterations is 
+applicable at the root. Our procedure terminates in this case since we only apply iteration
+binding for non base-type arguments, which $F$ does not have.
+\end{exa}
+
+\ourpara{Pragmatic Variant} We structured our procedure so that most of
+the unification machinery is contained in the \unifrulename{Bind} step. Modifying
+$\mathcal{P}$, we can sacrifice completeness and obtain a pragmatic variant of
+the procedure that often performs better in practice. 
+%The simplest such modification
+%is to limit the number or kind of bindings applied to a single constraint. 
+Our preliminary experiments showed that using mapping $\mathcal{P}_\mathsf{p}$ defined as follows is a reasonable compromise between completeness and
+performance:
+\begin{itemize}
+    \setlength\itemsep{1\jot}
+    \item If the constraint is rigid-rigid, $\mathcal{P}_\mathsf{p}(\lambda \overline{x}.
+    \, s \unif \lambda \overline{x}. \, t) = \varnothing$.
+    \item If the constraint is flex-rigid,
+    let 
+    $\mathcal{P}_\mathsf{p}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, a\, \overline{t})$ 
+    be 
+    \begin{itemize}
+        \item an imitation of $a$ for $F$, if $a$ is a constant, and
+        \item all Huet-style projections for $F$ if $F$ is not an identification variable.
+    \end{itemize}
+
+    \item If the constraint is flex-flex and the heads are different,
+    let 
+    $\mathcal{P}_\mathsf{p}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, G\, \overline{t})$ 
+    be
+    \begin{itemize}
+      \item an identification binding for $F$ and $G$, and
+      \item all Huet-style projections for $F$ if $F$ is not an identification variable
+  \end{itemize}
+
+  \item If the constraint is flex-flex and the heads are identical, we distinguish two cases:
+      \begin{itemize}
+      \item if the head is an elimination variable, 
+      $\mathcal{P}_\mathsf{p}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, F\, \overline{t}) = \varnothing$;
+      \item otherwise, let $\mathcal{P}_\mathsf{p}(\lambda \overline{x}. \, F \,\overline{s} \unif \lambda \overline{x}. \, F\, \overline{t})$
+            be the set of all eliminations bindings for $F$. 
+  \end{itemize}
+\end{itemize}
+
+The pragmatic variant of our procedure removes all iteration bindings to enforce finite branching. 
+Moreover, it
+imposes limits 
+on the number of bindings applied,
+counting the applications of bindings locally, per
+constraint. 
+It is useful to distinguish the Huet-style projection cases where
+$\alpha_i$ is a base type (called \emph{simple projection}), which always
+reduces the problem size, and the cases where $\alpha_i$ is a functional type (called \emph{functional
+projection}). We limit the number applications of the following bindings:
+functional projections, eliminations, imitations and identifications.
+In addition,
+a limit on the total number of applied bindings can be set. 
+An elimination binding that removes $k$ arguments counts as $k$
+elimination steps.
+Due to these limits, the pragmatic variant 
+terminates.
+
+To fail as soon as any of the limits
+is reached, the pragmatic variant employs an additional oracle.
+If this oracle determines that the limits are reached and the constraint is of the form
+$\lambda \overline{x}.\, F \, \overline{s}_m \unif \lambda \overline{x}.\, G \, \overline{t}_n$, it returns a \emph{trivial unifier} --
+a substitution $\{ F \mapsto \lambda \overline{x}_m.\, H, G \mapsto \lambda \overline{x}_n.\, H \}$, where $H$ is a fresh variable; if the limits are reached and
+the constraint is flex-rigid, the oracle fails; if the limits are not reached, 
+it reports that terms are outside its fragment. The trivial unifier prevents the procedure from failing on
+easily unifiable flex-flex pairs.
+
+Careful tuning of each limit optimizes the procedure for a specific class of problems. 
+For problems originating from proof assistants, shallow unification depth usually suffices. However, hard
+hand-crafted problems often need
+deeper unification.
+
 \section{Proof of Completeness}

### 2022-03-02 unif/unif.tex
solid

@@
 
-\ourpara{Examples}
+\ourpara{unif:examples}
 We present some examples that demonstrate advantages of our procedure.
@@
 
+Like the JP procedure, our procedure misses no unifiers:
+
+\begin{theorem}
+\label{thm:completeness}
+% \looseness=-1
+The procedure described in Section \ref{sec:unif:the-unification-procedure}
+parametrized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
+substitutions on the leaves of the constructed tree form a CSU.
+More precisely, let $E$ be a multiset of constraints
+and let $V$ be the supply of fresh variables provided to the procedure. 
+Then for any unifier $\varrho$ of $E$ there exists
+a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
+substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
+\end{theorem}
+
+Taking a high-level view, this theorem is proved by incrementally defining states
+$(E_{j},\sigma_{j})$ and \emph{remainder} \emph{substitutions}
+$\varrho_{j}$ starting with
+$\left(E_{0},\sigma_{0}\right)=\left(E,\makeop{id}\right)$ and
+$\varrho_{0}=\varrho$. The substitution $\varrho_{j}$ is what remains to be added
+to $\sigma_{j}$ to reach $\varrho_{0}$. States are defined
+so that the shape of the selected constraint from $E_{j}$ and the remainder
+substitution guide the choice of applicable transition rule.
+We employ a measure based on values of $E_j$ and $\varrho_j$ that decreases with 
+each application of the rules. Therefore, eventually, we will reach the target substitution
+$\sigma$.
+
+In the remaining of this section, we view terms as
+$\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-normal
+form as their canonical representative. Moreover, we consider all
+substitutions to be fully applied. These assumptions are justified because
+all bindings depend only on the head of terms and hence replacing the lazy
+transitions $\unifrulename{Normalize}_{\alpha\eta}$,
+$\unifrulename{Normalize}_{\beta}$, and $\unifrulename{Dereference}$ by eager
+counterparts only affects the efficiency but not the overall behavior of our
+procedure.
+
+We now give the detailed completeness proof of Theorem \ref{thm:completeness}.
+Our proof is an adaptation of the proof given by Jensen and Pietrzykowski
+\cite{jp-76-unif}.
+% %Intimate knowledge of that proof is required from the reader.
+Definitions and lemmas are reused, but are combined together differently to
+suit our procedure. We start by listing all reused definitions and lemmas
+from the original JP proof. The ``JP'' labels in
+their statements refer to the corresponding lemmas and definitions from
+the original proof.
+
+\begin{defi}[{$\jp[1]D6$}] \label{unif:def:disagreement-pairs}
+\looseness=-1
+Given two terms $t$ and $s$ and their common context $C$, 
+we can write $t$ as $C[\seq t]$ and $s$ as $C[\seq s]$
+for some $\seq t$ and $\seq s$.
+The pairs $(s_{j},t_{j})$ are called \emph{disagreement pairs.}
+\end{defi}
+    %
+\begin{defi}
+[$\jp D1$]\label{unif:def:opponents}
+Given two terms $t$ and $s$,
+let $\param[\seq x] t'$ and $\param[\seq y] s'$ 
+be respective $\alpha$-equivalent terms such that
+their parameters $\seq x$ and $\seq y$ are disjoint.
+Then the disagreement pairs of $t'$ and $s'$ are called 
+\emph{opponent pairs} in $t$ and $s$.
+\end{defi}
+
+\begin{lemma}[$\jp L{3\text{ (1)}}$]\label{lem:principal-lemma}Let $\varrho$ be a substitution
+    and $X$, $Y$ be free variables such that $\varrho\left(X\,\seq s\right)=\varrho\left(Y\,\seq t\right)$
+    for some term tuples $\seq s$ and $\seq t$. Then for every opponent
+    pair $u$, $v$ in $\varrho X$ and $\varrho Y$
+    (Definition \ref{unif:def:opponents}), the head of $u$ or $v$ is a parameter
+    of $\varrho X$ or $\varrho Y$.
+    \end{lemma}
+  %
+
+  In contrast to applied constants, applied variables should not be eagerly decomposed.
+  For a constant $\cst{f}$, if $\cst{f}\,\seq s \unif \cst{f}\,\seq t$ has a unifier,
+  that unifier must clearly also unify $s_{i} \unif t_{i}$ for each~$i$. 
+  For a free variable $X$, a unifier of $X\,\seq s \unif X\,\seq t$ does not necessarily
+  unify $s_{i} \unif t_{i}$.
+  The concept of $\omega$-simplicity is
+  a criterion on unifiers that captures some of the cases where eager decomposition is possible.
+  Non-$\omega$-simplicity
+  on the other hand is the main trigger of $\unifrulename{Iteration}$\textemdash the
+  most explosive binding of our procedure.
+
+  \begin{defi}
+  [$\jp D2$]\label{unif:def:omega-simple}An occurrence of a parameter $x$ of term $t$
+  in the body of $t$ is $\omega$\emph{-simple }if both
+  \begin{enumerate}
+  \item the arguments of $x$ are distinct and are exactly (the $\eta$-long
+  forms of) all of the variables bound in the body of $t$, and
+  \item this occurrence of $x$ is not in an argument of any parameter
+  of $t$.
+  \end{enumerate}
+  \end{defi}
+  This definition is slightly too restrictive for our purposes.
+  It is unfortunate that condition~1 requires $x$ to be applied to
+  \emph{all} instead of just some of the bound variables. The JP
+  proof would probably work with such a relaxation, and the definition
+  would then cover all cases where eager decomposition is possible. 
+  However, to reuse the JP lemmas, we stick to the original
+  notion of $\omega$-simplicity and introduce the following relaxation:
+  \begin{defi}
+  An occurrence of a parameter $x$ of term $t$ in the body of $t$
+  is \emph{base-simple} if it is $\omega$-simple or both
+  \begin{enumerate}
+  \item $x$ is of base type, and
+  \item this occurrence of $x$ is not in an argument of any parameter
+  of $t$.
+  \end{enumerate}
+  \end{defi}
+  %
+  \begin{lemma}\label{lem:base-simple->inj.}
+  Let $s$ have parameters $\seq x$ and a subterm $x_{j}\,\seq v$ 
+  where this occurrence of $x_{j}$ is base-simple.
+  Then for any sequence $\seq t$ of (at least $j$) terms, the body of
+  $t_{j}$ is a subterm of $s\,\seq t$ (after normalization) at the position
+  of $x_{j}\,\seq v$ up to renaming of the parameters of $t_{j}$. 
+  To compare positions of $s$ and $s\,\seq t$, ignore
+  the parameter count mismatch.
+  % ASK(ALEX), ASK(VISA): I am no expert, but isn't it unusual to use these
+  % kind of instructions in the statement of the theorem? Maybe it would be good
+  % to locally redefine the concept of position.
+  \end{lemma}
+  \begin{proof}
+  Consider the process of $\beta$-normalizing $s\,\seq t$. After substituting
+  terms $\seq t$ into the body of~$s$, a further reduction can only
+  take place when some $t_{k}$ is an abstraction that gets arguments in
+  $s$. The arguments $\seq v$ to the $x_{j}$ are distinct variables
+  bound in the body of $s$. This follows easily from either case of the
+  definition of base-simplicity. So $t_{j}$ is applied to the unmodified
+  $\seq v$ after substituting
+  terms $\seq t$ into the body of~$s$. Base-simplicity also implies
+  that $t_{j}\,\seq v$ does not occur in an argument to another
+  $t_{k}$. Hence only the reduction of $t_{j}\,\seq v$ itself affects this
+  subterm. The variables $\seq v$ match the parameter count of $t_{j}$
+  because we consider the $\eta$-long form of $t_{j}$; so $t_{j}\,\seq v$ reduces to the body
+  of $t_{j}$ (modulo renaming). The position is obviously that of $x_{j}\,\seq v$.
+  \end{proof}
+  \begin{lemma}[$\jp C{4 strengthened}$]\label{lem:eq-if-omega-simple}Let $\varrho$ be a substitution and $X$ a free
+    variable. If $\varrho\left(\param X\,\seq s\right)=\varrho\left(\param X\,\seq t\right)$
+    and some occurrence of the $i^{\text{th}}$ parameter of $\varrho X$
+    is base-simple, then $\varrho s_{i}=\varrho t_{i}$.
+    \end{lemma}
+    \begin{proof}
+    By Lemma \ref{lem:base-simple->inj.}, $\varrho s_{i}$ occurs in
+    $\varrho X\left(\varrho\seq s\right)$ at certain position that depends
+    only on $\varrho X$. Similarly $\varrho t_{i}$ occurs in $\varrho X\left(\varrho\seq t\right)=\varrho X\left(\varrho\seq s\right)$
+    at the same position, and hence $\varrho s_{i}=\varrho t_{i}$.
+    \end{proof}
+    We define more properties to determine which binding to apply to a given constraint.
+    Roughly speaking, the simple comparison form will trigger identification bindings,
+    projectivity will trigger Huet-style projections, and
+    simple projectivity will trigger JP-style projections.
+    %
+    \begin{defi}[$\jp D4$]
+    \looseness=-1
+    \label{unif:def:simple-comparison-form}We say that $s$ and $t$ are in \emph{simple
+    comparison form }if all $\omega$-simple heads of opponent pairs in $s$ and $t$ are distinct, and
+    each opponent pair has an $\omega$-simple head.
+    \end{defi}
+    %
+    \begin{defi}[$\jp D5$]
+    \label{unif:def:projective} A term $t$ is called \emph{projective }if the head
+    of $t$ is a parameter of $t$. If the whole body is just
+    the parameter, then $t$ is called \emph{simply projective.}
+    \end{defi}
+    A central part of the proof is to find a suitable measure for the
+    remaining problem size. Showing that the measure is strictly decreasing
+    and well-founded guarantees that the procedure finds 
+    a suitable substitution in finitely many steps. We reuse the measure for remainder substitutions
+    from JP \cite{jp-76-unif}, but embed it into a lexicographic measure
+    to handle the decomposition steps and oracles of our procedure.
+    \begin{defi}[$\jp D7$] \label{unif:def:weights}
+    The \emph{free weight} of a term $t$ is the total number of occurrences of
+    free variables and constants in $t$.
+    The \emph{bound weight} of $t$ is the total number of occurrences 
+    (excluding occurrences $\lambda x$) of bound variables in $t$,
+    but with the particular exemption: if a prefix variable $u$ 
+    has one or more $\omega$-simple occurrences in the body,
+    then one such occurrence and its arguments are not counted.
+    It does not matter which occurrence is not counted because
+    in $\eta$-long form the bound weight of the arguments of an 
+    $\omega$-simple variable is the same for all occurrences
+    of that variable.
+    \end{defi}
+    \begin{defi}[$\jp D8$] \label{unif:def:ord}
+    For multisets $E$ of unification constraints and
+    substitutions $\rho$, our measure on pairs $(E, \rho)$
+    is the lexicographic comparison of
+    \begin{description}
+      \item[A] the sum of the sizes of the terms in $\rho E$
+      \item[B] the sum over the free weight of $\rho F$, 
+        for all variables $F$ mapped by $\rho$
+      \item[C] the sum over the bound weight of $\rho F$, 
+      for all variables $F$ mapped by $\rho$
+      \item[D] the sum over the number of parameters of $\rho F$, 
+      for all variables $F$ mapped by $\rho$
+    \end{description}
+    We denote the quadruple containing these numbers as $\ord(E,\rho)$.
+    We denote the triple containing only the last three components of $\ord(E,\rho)$ as $\ord \rho$.
+    %ASK(VISA): Why is ord(E,\rho) with parentheses and ord \rho without them?
+    We write $<$ for the lexicographic comparison of these tuples.
+    \end{defi}
+
+    \looseness=-1
+    The next six lemmas correspond to the bindings of our procedure and sufficient
+    conditions for the binding to bring us closer to a given solution. 
+    This is expressed as a decrease of the $\ord$ measure
+    of the remainder. In each of these lemmas, 
+    let $u$ be a term with a variable head $a$ and
+    $v$ a term with an arbitrary head $b$.
+    Let $\varrho$ be a unifier of $u$ and $v$.
+    The conclusion, let us call it \textbf{\emph{C}},
+    is always the same: there exists a binding $\delta$ applicable to the problem $ u\unif v$, and
+    there exists a substitution $\varrho'$ such that $\ord\varrho'<\ord\varrho$
+    and for all variables $X$ except the fresh variables introduced by the binding $\varrho X= \varrho'\, \delta X$.
+    For most of these lemmas, we refer to JP \cite{jp-76-unif} for proofs.
+    Although JP only claim $\varrho X= \varrho'\, \delta X$
+    for variables $X$ mapped by $\varrho$,
+    inspection of their proofs shows that the equality holds for 
+    all $X$ except the fresh variables introduced by the binding.
+    Moreover, some of our bindings have more preconditions, 
+    yielding additional orthogonal hypotheses in our lemmas,
+    which we address below.
+
+    \begin{lemma}
+    [$\jp L9$]\label{lem:elimination}If $a=b$ is not an elimination variable and $\varrho a$
+    discards any of its parameters, then \textbf{C} by elimination.
+    Moreover, for the elimination variable $G$ introduced by this elimination,
+    $\varrho'G$ discards none of its parameters and has the body of $\varrho a$.
+    \end{lemma}
+    \begin{proof}
+    Let $\varrho a=\param t$ and let $\left(x_{j_{k}}\right)_{k=1}^{i}$
+    be the subsequence of $\seq x$ consisting of those variables which
+    occur in the body $t$. It is a strict subsequence,\emph{ }since $\varrho a$
+    is assumed to discard some parameter. Since the equal heads $a=b$ of the
+    constraint $u\unif v$ are not elimination variables, elimination
+    for $(j_{k})_{k=1}^{i}$ can be applied. Let $\delta=\{ a\mapsto\param G\,x_{j_{1}}\,\ldots\,x_{j_{i}}\} $
+    be the corresponding binding. Define $\varrho'$ to be like $\varrho$
+    except
+    \[
+    \varrho'a=a\quad\text{and}\quad\varrho'G=\param[x_{j_{1}}...x_{j_{i}}]t.
+    \]
+    Obviously $\varrho'G$ is a closed term and $\varrho X = \varrho'\,\delta X$
+    holds for all $X \not= G$. Moreover $\ord\varrho'<\ord\varrho$, because
+    free and bound weights stay the same ($\varrho a$ and $\varrho'G$ have the
+    same body $t$) whereas the number of parameters strictly decreases. The definition
+    of $\left(j_{k}\right)_{k=1}^{i}$ implies that $\varrho'G$ discards
+    none of its parameters.
+    \end{proof}
+
+    %
+    \begin{lemma}
+    [$\jp L{10}$]\label{lem:iteration}
+    Assume that
+    there exists a parameter $x$ of $\varrho a$ such that
+    $x$ has a non-$\omega$-simple (Definition \ref{unif:def:omega-simple}) 
+    occurrence in $\varrho a$, which is not below another parameter, or
+    such that $x$ has at least two $\omega$-simple occurrences in $\varrho a$.
+    Moreover, 
+    if $a = b$, to make iteration applicable,
+    $a$~must not be an elimination variable, and
+    $x$ must be of functional type.
+    Then \textbf{C} is achieved by iteration.
+    \end{lemma}
+    %
+    \begin{lemma}
+    [$\jp L{11}$]\label{lem:jp-projection}
+    \looseness=-1
+    Assume that $a$ and $b$ are different free variables.
+    If $\varrho a$ is simply
+    projective (Definition \ref{unif:def:projective})
+    and $a$ is not an identification variable, 
+    then \textbf{C} by JP-style projection.
+    \end{lemma}
+    %\begin{proof}
+    %This is Lemma 3.11 from \cite{jp-76-unif}, plus the assumption 
+    %that $a$ is not an identification variable.
+    %This extra assumption is necessary because we restrict 
+    %the application of JP-style projection in this way.
+    %\end{proof}
+    %
+    \begin{lemma}
+    [$\jp L{12}$]\label{lem:jp-imitation}If $\varrho a$ is not projective and $b$ is rigid,
+    then \textbf{C} by imitation.
+    \end{lemma}
+    %
+    \begin{lemma}
+    [$\jp L{13}$]\label{lem:identification}
+    Let $a \not= b$.
+    Assume that $\varrho a \not= a$ and $\varrho b \not= b$
+    are in simple comparison form (Definition \ref{unif:def:simple-comparison-form}) and neither is
+    projective. 
+    Then \textbf{C} by identification.
+    Moreover, $\varrho' H$ is not projective, where $H$ is the identification variable
+    introduced by this application of the identification binding.
+    \end{lemma}
+    \begin{proof}
+    This is JP's Lemma~3.13, plus the claim that $\varrho' H$ is not projective.
+    Inspecting the proof of that lemma, it is obvious that $\varrho' H$ cannot be projective
+    because $\varrho a$ and $\varrho b$ are not projective.
+    \end{proof}
+    %
+    %Additionally, we need one more of these lemmas for the Huet-style projection binding:
+    %
+    \begin{lemma}
+    \label{lem:huet-projection}
+    Assume that $\varrho a$ is
+    projective (Definition \ref{unif:def:projective}),
+    $a$ is not an identification variable, and $b$ is rigid.
+    Then \textbf{C} by Huet-style projection.
+    \end{lemma}
+    \begin{proof}
+    Since $\varrho a$ is
+    projective, we have $\varrho a = \lambda \overline{x}_n.\, x_k\,\overline{t}_m$
+    for some $k$ and some terms $\overline{t}_m$.
+    If $\varrho a$ is also simply projective,
+    then $x_k$ must be non-functional
+    and since Huet-style projection and JP-style projection coincide
+    in that case, Lemma~\ref{lem:jp-projection} applies.
+    Hence, in the following we may assume that $\varrho a$ is not simply projective,
+    i.e., that $m>0$.
+
+    Let $\delta$ be
+    the Huet-style projection binding:
+    \[ \delta = \{a \mapsto \lambda \overline{x}_n. \, x_i \, (F_1 \, \overline{x}_n) \, \ldots \, (F_m \, \overline{x}_n)\} \]
+    for fresh variables $F_1,\dots,F_m$.
+    This binding is applicable because $b$ is rigid.
+    Let $\varrho'$ be the same as $\varrho$ except that we set $\varrho' a = a$
+    and for each $1\leq j\leq m$ we set
+    \[
+      \varrho' F_j = \lambda \overline{x}_n. \, t_j
+    \]
+    It remains to show that $\ord\varrho'<\ord\varrho$.
+    The free weight of $\varrho a$ is the same as 
+    the sum of the free weights of $\varrho' F_j$ for $1\leq j\leq m$.
+    Thus, the free weight is the same for $\varrho$ and $\varrho'$.
+    The bound weight of $\varrho a$ however is exactly $1$ larger than
+    the sum of the bound weights of $\varrho' F_j$ for $1\leq j\leq m$
+    because of the additional occurrence of $x_k$ in $\varrho a$.
+    The exemption for $\omega$-simple occurrences in the definition
+    of the bound weight cannot be triggered by this occurrence of $x_k$
+    because $m > 0$ and thus $x_k$ is not $\omega$-simple.
+    It follows that $\ord\varrho'<\ord\varrho$.
+    \end{proof}
+    %
+    We are now ready to prove the completeness theorem (Theorem \ref{thm:completeness}).
+    \begin{proof}
+    Let $E$ be a multiset of constraints
+    and let $V$ be the supply of fresh variables provided to our procedure. 
+    Let $\varrho$ be a unifier of $E$.
+    We must show that there exists
+    a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
+    substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
+
+    Let $E_{0}=E$ and $\sigma_{0}=\makeop{id}$. 
+    Let $\varrho_{0}=\tau\varrho$ for some renaming $\tau$, such that 
+    every free variable occurring in $\varrho_{0}E_{0}$ does not occur in $E_{0}$ and is not contained in $V$.
+    Then $\varrho_{0}$
+    unifies $E_{0}$ because $\varrho$ unifies $E$ by assumption.
+    Moreover, $\varrho_{0}=\varrho_{0}\sigma_{0}$. We proceed
+    to inductively define $E_{j}$, $\sigma_{j}$ and $\varrho_{j}$
+    until we reach some $j$ such that $E_j = \varnothing$.
+    To guarantee well-foundedness, we ensure that the measure 
+    $\ord\left(\varrho_{j},E_{j}\right)$ decreases with each step.
+    We maintain the following invariants for all $j$:
+    \begin{itemize}
+    \item $(E_{j},\sigma_{j})\longrightarrow(E_{j+1},\sigma_{j+1})$;
+    \item $\ord(\varrho_{j},E_{j}) > \ord(\varrho_{j+1},E_{j+1})$;
+    \item $\varrho_{j}$ unifies $E_{j}$;
+    \item $\varrho_{0} X = \varrho_{j}\sigma_{j} X$ for all free variables $X\not\in V$;
+    \item every free variable occurring in $\varrho_{j}E_{j}$ does not occur in $E_{j}$ and is not contained in $V$;
+    \item for every identification variable $X$, $\rho_j X$ is not projective; and
+    \item for every elimination variable $X$, each parameter of $\rho_j X$ 
+          has occurrences in $\rho_j X$,
+          all of which are base-simple.
+    \end{itemize}
+    If $E_j \not= \varnothing$, let $u\unif v$ be the selected constraint $S(E_{j})$ in $E_{j}$.
+
+    First assume that an oracle is able to find a CSU for the constraint $u\unif v$.
+    Since $\varrho_{j}$ unifies $u$ and~$v$, by the definition of a CSU,
+    the CSU discovered by the oracle 
+    contains a
+    unifier $\delta$ of $u$ and $v$ such that there exists $\varrho_{j+1}$
+    and for all free variables $X$ except for the auxiliary variables of the CSU
+    we have $\varrho_{j} X=\varrho_{j+1}\,\delta X$. Thus, an \unifrulename{OracleSucc}
+    transition is applicable and yields the node 
+    $(E_{j+1},\sigma_{j+1})=(\delta(E_{j}\setminus\{ u\unif v\} ),\delta\,\sigma_{j})$.
+    %Since the auxiliary variables introduced by \unifrulename{OracleSucc} are fresh, they cannot occur in $E_{j}$. 
+    Therefore we have a strict containment $\varrho_{j+1}E_{j+1}\subset\varrho_{j+1}\,\delta\,E_{j}=\varrho_{j}E_{j}$.
+    This implies $\ord(E_{j+1},\varrho_{j+1})<\ord(E_{j},\varrho_{j})$.
+    It also shows that the constraints $\varrho_{j+1}E_{j+1}$ are unified
+    when $\varrho_{j}E_{j}$ are. 
+    Since the auxiliary variables introduced by \unifrulename{OracleSucc} are fresh,
+    they cannot occur in $E_j$ nor in $\sigma_{j} X$ for any $X\not\in V$. Hence,
+    we have $\varrho_{0} X = \varrho_{j}\sigma_{j} X = \varrho_{j+1}\,\delta\,\sigma_{j} X =\varrho_{j+1}\sigma_{j+1} X$
+    for all free variables $X\not\in V$.
+    Any free variable occurring in $\varrho_{j+1}E_{j+1}$ cannot not occur in $E_{j+1}$ and is not contained in $V$
+    because  $\varrho_{j+1}E_{j+1}\subset\varrho_{j}E_{j}$
+    and the variables in $E_{j+1} = \delta(E_{j}\setminus\{ u\unif v\} )$
+    are either variables already present in $E_{j}$ or fresh variables introduced by \unifrulename{OracleSucc}.
+    New identification or elimination variables are not introduced;
+    so their properties are preserved. Hence all invariants are preserved.
+
+    Otherwise we proceed by a case distinction on the form of $u\unif v$.
+    Typically, one of the Lemmas \ref{lem:elimination}\textendash{}\ref{lem:huet-projection}
+    is going to apply. Any one of them gives substitutions $\varrho'$
+    and $\delta$ with properties that let us define $E_{j+1}=\delta E_{j}$,
+    $\sigma_{j+1}=\delta\,\sigma_{j}$ and $\varrho_{j+1}=\varrho'$.
+    The problem size always strictly decreases, because 
+    these lemmas imply
+    $\varrho_{j+1}E_{j+1}=\varrho_{j+1}\delta E_{j}=\varrho_{j}E_{j}$
+    and $\ord\varrho_{j+1}=\ord\varrho'<\ord\varrho_{j}$.
+    Regarding the other invariants, the former equation guarantees that $\varrho_{j+1}$
+    unifies $E_{j+1}$, and 
+    $\varrho_{0} X = \varrho_{j}\,\sigma_{j} X = \varrho_{j+1}\,\delta\,\sigma_{j} X =\varrho_{j+1}\,\sigma_{j+1} X$
+    for all $X \not\in V$ because the fresh variables introduced by the binding cannot occur in $\sigma_{j} X$ for any $X\not\in V$. The conditions on variables must be checked
+    separately when new ones are introduced. Let $a$ be
+    the head of $u=\param a\,\seq u$ and $b$ be the head of $v = \param b\,\seq v$.
+    Consider the following cases:
+
+    \begin{description}
+    \item[$u$ and $v$ have the same head symbol $a=b$] \hfill
+    \begin{enumerate}[beginpenalty=1000]
+    \item Suppose that $\varrho_{j}a$ has a parameter with non-base-simple
+    occurrence. By one of the induction invariants, $a$ is not an elimination
+    variable. 
+    Among all non-base-simple occurrences of parameters in $\varrho_{j}a$, 
+    choose the leftmost one, which we call $x$.
+    This occurrence of $x$ cannot be below another parameter, 
+    because having $x$ occur in one of its arguments would make
+    that other parameter non-base-simple, contradicting the occurrence of $x$ being leftmost.
+    Thus $x$ is neither base-simple nor below another parameter; so $x$ is of functional type.
+    Moreover, non-base-simplicity implies non-$\omega$-simplicity.
+    Hence, we can apply Lemma~\ref{lem:iteration} (iteration).
+
+    \item Otherwise suppose that $\varrho_{j}a$ discards some of its parameters.
+    By one of the induction invariants, $\varrho_{j}a$ is not an elimination variable.
+    Hence Lemma \ref{lem:elimination} (elimination) applies. The newly introduced elimination
+    variable $G$ satisfies the required invariants, because Lemma \ref{lem:elimination}
+    guarantees that $\varrho_{j+1}G$ uses its parameters and shares the
+    body with $\varrho_{j}a$ which by assumption of this case contains
+    only base-simple occurrences.
+
+    \item Otherwise every parameter of $\varrho_{j}a$ has occurrences and
+    all of them are base-simple. We are going to show that $\unifrulename{Decompose}$
+    is a valid transition and decreases $\varrho_{j}E_{j}$. By Lemma~\ref{lem:eq-if-omega-simple}
+    we conclude from $\varrho_{j}u=\varrho_{j}v$ that $\varrho_{j}u_{i}=\varrho_{j}v_{i}$
+    for every $i$. Hence the new constraints 
+    $E_{j+1}=E_{j}\setminus\left\{ u\unif v\right\} \cup\set{u_{i}\unif v_{i}}{\text{for all }i}$
+    after $\unifrulename{Decompose}$ are unified by $\varrho_{j}$.
+    This allows us to define $\varrho_{j+1}=\varrho_{j}$ and
+    $\sigma_{j+1}=\sigma_{j}$.
+    To check that $\varrho_{j+1}E_{j+1}=\varrho_{j}E_{j+1}$ is smaller
+    than $\varrho_{j}E_{j}$ it suffices to check that constraints $\varrho_{j}u_{i}\unif\varrho_{j}v_{i}$
+    together are smaller than $\varrho_{j}u\unif\varrho_{j}v$. Since
+    all parameters of $\varrho_{j}a$ have base-simple occurrences, $\varrho_{j}u_{i}$
+    is a subterm of $\varrho_{j}u=\param\varrho_{j}a\,\left(\varrho_{j}\seq u\right)$
+    by Lemma \ref{lem:base-simple->inj.}. Similarly for $\varrho_{j}v$.
+    It follows that $\varrho_{j+1}E_{j+1}$ is smaller
+    than $\varrho_{j}E_{j}$.
+    Since $\varrho_{j+1}=\varrho_{j}$ and
+    $\sigma_{j+1}=\sigma_{j}$, the other
+    invariants are obviously preserved.
+
+    \end{enumerate}
+
+    \item[\mbox{$u \unif v$} is a flex-flex pair with different heads] \hfill
+    \begin{enumerate}
+      \setcounter{enumi}{4}
+      \item First, suppose that
+      $\varrho_j a$ or $\varrho_j b$ is simply projective
+      (Definition \ref{unif:def:projective}). 
+      By the induction hypothesis, the simply projective head 
+      cannot be an identification variable.
+      Thus Lemma \ref{lem:jp-projection} (JP-style projection) applies.
+
+      \item \looseness=-1 
+      Otherwise suppose that 
+      $\varrho_j a$ is projective but not simply.
+      Then the head of $\varrho_j a$ is some parameter $x_{k}$. But
+      this occurrence cannot be $\omega$-simple because it has arguments
+      which cannot be bound above the head $x_{k}$. Thus Lemma \ref{lem:iteration}
+      (iteration) applies.
+      If $\varrho_j b$ is projective but not simply, the same argument applies.
+
+      \item Otherwise suppose that $\varrho_j a$, $\varrho_j b$ are in simple
+      comparison form (Definition \ref{unif:def:simple-comparison-form}).
+      By one of the the induction invariants,
+      the free variables occurring in $\varrho_{j}E_{j}$ do not occur in $E_{j}$.
+      Thus $\varrho_{j} a \not= a$ and $\varrho_{j} b \not= b$.
+      Then Lemma \ref{lem:identification} (identification) applies.
+
+      \item Otherwise $\varrho_j a$, $\varrho_j b$ are not in simple comparison
+      form. By Lemma \ref{lem:principal-lemma} and by the definition of simple comparison form,
+      there is some opponent pair $x_{k}\,\seq r$, $ b$
+      in $\varrho_j a$ and $\varrho_j b$
+      (after possibly swapping $u$ and $v$) where either the occurrence
+      of $x_{k}$ is not $\omega$-simple (Definition~\ref{unif:def:omega-simple}) or else
+      $x_{k}$ has another $\omega$-simple occurrence in the body of $\varrho_j a$.
+      Then Lemma \ref{lem:iteration} 
+      (iteration) applies.
+    \end{enumerate}
+
+    \looseness=-1
+    \item[$u \unif v$ is a flex-rigid pair]
+    Without loss of generality, assume that $a$ is flex and $b$ is rigid.\hfill
+
+    \begin{enumerate}
+    \setcounter{enumi}{8}
+    \item Suppose first that $\varrho_j a$ is projective.
+    By one of the induction invariants, $a$ cannot be an identification variable.
+    Thus Lemma \ref{lem:huet-projection} (Huet-style projection) applies.
+
+    \item Otherwise $\varrho_j a$ is not projective. 
+    The head of $\varrho_j a$ must be $b$ because $b$ is rigid, and $\varrho_j$ unifies $u$ and $v$. 
+    Since $\varrho_j a$ is not projective, that means that $b$ is not a bound variable. 
+    Therefore, $b$ must be a constant.
+    Then Lemma \ref{lem:jp-imitation} (imitation) applies.
+    \end{enumerate}
+
+    We have now constructed a run 
+    $(E_{0},\sigma_{0})\longrightarrow(E_{1},\sigma_{1})\longrightarrow(E_{2},\sigma_{2})\longrightarrow\cdots$
+    of the procedure. 
+    This run cannot be infinite because because the measure $\ord(E_j,\varrho_{j})$
+    strictly decreases as $j$ increases.
+    Hence, at some point we reach a $j$ such
+    that $E_{j}=\varnothing$ and $\varrho_{0} X = \varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$.
+    Therefore, 
+    $(E,\makeop{id})\longrightarrow^*(\varnothing,\sigma_{j})\longrightarrow\sigma_j$,
+    and $\rho X = \tau^{-1}\,\varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$,
+    completing the proof. \qedhere
+    \end{description}
+    \end{proof}
+
+\pagebreak[2]    
 \section{A New Decidable Fragment}
@@
 
+We discovered a new fragment that admits a finite CSU and a simple oracle. The
+oracle is based on work by Prehofer and the PT procedure \cite{cp-95-unifphd},
+an adaptation of the preunification procedure by Snyder and Gallier
+\cite{sg-89-unif} (which itself is an adaptation of Huet's procedure). PT
+transforms an initial multiset of constraints $E_0$ by applying bindings
+$\varrho$. If there is a sequence
+$E_0\ptarrow{\varrho_1}\cdots\ptarrow{\varrho_n}E_n$ such that $E_n$ has only
+flex-flex constraints, we say that PT produces a preunifier $\sigma =
+\varrho_n\ldots\varrho_1$ with constraints $E_n$. A sequence fails if
+$E_n=\bot$. As in the previous section, we consider all terms to be
+$\alpha\beta\eta$-equivalence classes with the $\eta$-long $\beta$-reduced form
+as their canonical representative. Unlike previously, in this section we view
+unification constraints $s \unif t$ as ordered pairs.
+
+The following rules, however, are stated modulo orientation. The PT transition
+rules, adapted for our presentation style, are as follows:
+
+\begin{description}[labelwidth=\widthof{\rm\unifrulename{\bf Decomposition}}]
+    \item[\rm\textsf{Deletion}] 
+      $\{\selected{s \unif s}\} \uplus E \ptarrowid E$
+    \item[\rm\textsf{Decomposition}]
+      $\{\selected{\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda \overline{x}. \, a \, \overline{t}_m} \} 
+        \uplus \allowbreak
+        E \ptarrowid \{ s_1 \unif  t_1, \ldots, s_m \unif t_m \} \uplus E$
+        \\ where $a$ is rigid
+    \item[\rm\textsf{Failure}]
+        $\{\selected{\lambda \overline{x}. \, a \, \overline{s} \unif \lambda \overline{x}. \, b \, \overline{t}} \} 
+          \uplus \allowbreak
+          E \ptarrowid \bot$
+    \\ where $a$ and $b$ are different rigid heads
+    \item[\rm\textsf{Solution}] 
+      $\{\selected{\lambda \overline{x}.\, F\,\overline{x} \unif \lambda \overline{x}.\,t} \} \uplus E \ptarrow{\varrho} \varrho E$
+      \\where $F$ does not occur in $t$, $t$ does not have a flex head, and $\varrho = \{ F \mapsto \lambda \overline{x}.\,t \}$ 
+    \item[\rm\textsf{Imitation}]
+      $\{\selected{\lambda \overline{x}. \, F \, \overline{s}_m \unif \lambda
+      \overline{x}. \, \cst{f} \, \overline{t}_n }\} \uplus E \ptarrow{\varrho}
+      \varrho(\{ G_1 \,
+      \overline{s}_m  \unif  t_1, \ldots,  G_n \, \overline{s}_m \allowbreak \unif t_n\} \uplus E)$ \\
+      where $\varrho = \{ F \mapsto \lambda \overline{x}_m. \, \cst{f} \, (G_1
+      \, \overline{x}_m) \ldots (G_n \, \overline{x}_m) \}$,
+      $\overline{G}_n$ are fresh variables of appropriate types
+    \item[\rm\textsf{Projection}]
+      $\{\selected{\lambda \overline{x}. \, F \, \overline{s}_m \unif \lambda
+      \overline{x}. \, a \, \overline{t} }\} \uplus E \ptarrow{\varrho} 
+      \varrho(\{ s_i \, (G_1 \,
+      \overline{s}_m) \ldots (G_j \, \overline{s}_m) \unif a \, \overline{t}\} \uplus E)$ \\
+      where $\varrho =\{F \mapsto \lambda \overline{x}_m. \, x_i \, (G_1 \,
+      \overline{x}_m) \ldots (G_j \, \overline{x}_m)\}$, $\overline{G}_j$ are
+      fresh variables of appropriate types
+\end{description}
+\looseness=-1
+The \selected{\text{grayed}} constraints are required to be selected by a given
+selection function $S$. We call $S$ \emph{admissible} if it selects only flex-rigid constraints, prioritizes
+selection of constraints applicable for \textsf{Failure} and \textsf{Decomposition}, 
+and of descendant constraints of
+\textsf{Projection} transitions with $j=0$ (i.e., for $x_i$ of base type), in that order of priority. In the
+remainder of this section we consider only admissible selection functions,
+an assumption that Prehofer also makes implicitly in his thesis. Additionally,
+whenever we compare multisets, we use the multiset ordering defined by
+Dershowitz and Manna \cite{dm-79-multisetordering}.
+As above, we assume that the fresh variables are taken from an infinite supply $V$ of fresh variables
+that are different from the variables in the initial problem
+and never reused.
+
+The following lemma states that PT is \emph{complete for preunification}:
+\begin{lemma}
+  \label{unif:lemma:pt-complete}
+  \looseness=-1
+  Let $\varrho$ be a unifier of a multiset of constraints $E_0$. Then PT produces a
+  preunifier $\sigma$ with constraints $E_n$, and there exists a
+  unifier $\theta$ of $E_n$ such that
+  $\varrho X = \theta \sigma X$ for all $X$ that are not contained in the supply $V$ of fresh variables.
+\end{lemma}
+\begin{proof}
+  This lemma is a refinement of Lemma 4.1.7 from Prehofer's PhD thesis
+  \cite{cp-95-unifphd}, and this proof closely follows the proof of that lemma.
+  Compared to the lemma from Prehofer's thesis, our lemma additionally establishes
+  the relationship of unifier $\theta$ of the resulting flex-flex constraint
+  set $E_n$ with preunifier $\sigma$ and the target unifier $\varrho$.
+
+  We prove the lemma by induction 
+  using a well-founded measure on $(\varrho,
+  E_0)$. The ordering
+  is the lexicographic comparison of the following properties:
+
+  \begin{description}
+    \item[A] sum of the abstraction-free sizes of the terms $\varrho F$ for each variable $F$ mapped by $\varrho$
+    %TODO: What are the "bindings in \rho"? Maybe write "the terms $\varrho x$ for each variable $x$ mapped by $\rho$"? --Alex
+    %TODO: Done
+    \item[B] multiset of the sizes of constraints in $E_0$
+  \end{description}
+Here, the \emph{abstraction-free size} is inductively defined by
+$\text{afsize}(F) = 1$; $\text{afsize}(x) = 1$; $\text{afsize}(\cst{f}) = 1$;
+$\text{afsize}(s\,t) = \text{afsize}(s) + \text{afsize}(t)$;
+$\text{afsize}(\lambda x.\, s) = \text{afsize}(s)$.
+
+  If $E_0$ consists only of flex-flex constraints, then we can take an empty
+  transition sequence (i.e., $\sigma = \text{id}$) and $\theta = \varrho$.
+  Otherwise, there exists a constraint that is selected by an
+  admissible selection function. We show that for each such constraint, there is
+  going to be a PT transition bringing us closer to the desired preunifier.
+
+  Let $E_0 = \{\selected{s \unif t}\} \uplus E'_0$. Since $s \unif t$ is selected, at least
+  one of $s$ and $t$ must have a rigid head. We distinguish several cases based on the
+  form of $s \unif t$ (modulo the constraint order):
+
+  \begin{itemize}
+    \item $s \unif s$: in this case, \textsf{Deletion} applies. This transition does not
+    alter $\varrho$, but removes an equation obtaining $E_1$ which is smaller
+    than $E_0$ and still unifiable by $\varrho$. By induction hypothesis, the preunifier is reachable.
+    \item 
+    \looseness=-1
+    $\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda
+    \overline{x}. \, b \, \overline{t}_n$, where $a$ and $b$ are rigid: since
+    $\varrho$ is a unifier, then $a=b$, $n=m$, and \textsf{Decomposition}
+    applies. Similarly to the above case, we conclude that the preunifier is reachable.
+    \item $\lambda \overline{x}.\, F\,\overline{x} \unif \lambda \overline{x}.\,t$ where $t$
+    has a rigid head and $F$ does not occur in $t$: \textsf{Solution} applies. Huet showed \cite[proof of L5.1]{gh-75-unification}
+    that in this case $\varrho = \varrho\sigma_1$, where $\sigma_1 = \{ F \mapsto \lambda\overline{x}.\,t \}$.
+    Let $\varrho_1 = \varrho\setminus\{F \mapsto \varrho F\}$.
+    Then $\varrho = \varrho_1 \sigma_1$.
+    \looseness=-1
+    Since $\varrho$ unifies $E'_0$, $\varrho_1$ unifies $E_1 = \sigma_1 E'_0$. Clearly, $\varrho_1$ is smaller
+    than $\varrho$. Now we can apply induction hypothesis on $\varrho_1$ and $E_1$, from which 
+    we obtain a sequence $E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$, $\theta$ which
+    unifies $E_n$ and $\sigma' = \sigma_n \ldots \sigma_2$, such that $\varrho_1 X = \theta\sigma' X$ for all $X \not\in V$.
+    Finally, it is clear that sequence $E_0 \ptarrow{\sigma_1} E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$,
+    $\theta$, and $\sigma = \sigma' \sigma_1$ are as wanted in the lemma statement.
+    \item \looseness=-1 
+    $\lambda
+    \overline{x}.\,F\,\overline{s}_m~\unif~\lambda \overline{x}.\,a \,
+    \overline{t}$ where $a$ is rigid: depending on the value of $\varrho$, we can
+    either take an \textsf{Imitation} or \textsf{Projection} step. In either case, we show that
+    the measure reduces. Let $\varrho F = \lambda \overline{x}_m. \, b \, \overline{u}_n$ where $b$
+    is either a constant or a bound variable. If $b$ is a constant, we choose
+    the \textsf{Imitation} step; otherwise we choose the \textsf{Projection} step. In either case,
+    $\sigma_1 = \{ F \mapsto \lambda \overline{x}_m. \, b \, (G_1 \, \overline{x}_m) \ldots
+    (G_n \, \overline{x}_m) \}$. Then it is easy to check that $\varrho_1 = \varrho
+    \setminus \{F \mapsto \lambda \overline{x}_m. \, b \, \overline{u}_n\} \cup \{ G_1~\mapsto~\lambda\overline{x}_m.\, u_1, \allowbreak
+    \ldots, G_n \mapsto \lambda\overline{x}_m.\, u_n \}$ unifies $E_1$ created as the
+    result of imitation or projection. Clearly, $\varrho_1$ has smaller abstraction-free size
+    than $\varrho$. Therefore, by the induction hypothesis we obtain the transitions
+    $E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$ and substitutions
+    $\theta$ and  $\sigma' = \sigma_n\ldots\sigma_2$, where $\varrho_1 X =
+    \theta \sigma' X$ for all $X \not\in V\setminus\{G_1,\dots,G_n\}$ and $\theta$ is a unifier of $E_n$. Our goal is to prove
+    $\varrho X = \theta \sigma' \sigma_1 X$ for all $X \not\in V$. 
+    For variables $X \not\in V$ that are not $F$, we have $\sigma_1 X = X$
+    and $\varrho X = \varrho_1 X$, which implies $\varrho X = \theta \sigma' \sigma_1 X$.
+    For $X = F$, since $\varrho_1$ and $\theta \sigma'$ agree on $G_1, \ldots, G_n$, we
+    conclude that $\varrho F = \theta \sigma' \sigma_1 F$.
+    Therefore, by taking $\sigma = \sigma'\sigma_1$, and prepending $E_0$ to the
+    sequence from the induction hypothesis, we show that the preunifier is reachable. 
+    \qedhere
+  \end{itemize}
+\end{proof}
+
+\looseness=-1
+Prehofer showed that PT terminates for some classes of constraints. Those
+classes impose requirements on the free variables occurring in the constraints.
+In particular, he identified a class of terms we call \emph{strictly
+solid}\footnote{In Prehofer's thesis this class is described in the statement of
+Theorem 5.2.6.}, which requires that all arguments of free variables are either
+bound variables (of arbitrary type) or ground second-order terms of base type.
+Another important constraint is linearity: a term is called
+\emph{linear} if it contains no repeated occurrences of free variables.
+% We relax Prehofer's conditions by lifting order requirements: we require all
+% arguments of free variables to be either bound variables or ground base-type
+% terms. We call this new class of terms \emph{solid}.
+
+\begin{exa}
+\label{unif:ex:solid-terms}
+Let $G$, $\cst{a}$, and $x$ be of base type, and $F$, $H$, $\cst{g}$, and $y$ be
+binary. Then, the term $F \, G \, \cst{a}$ is not strictly solid, since $F$ is
+applied to a free variable $G$; similarly $H \, (\lambda x. \, x) \, \cst{a}$ is
+not strictly solid as the first argument is of functional type, but it is not
+a bound variable; $\lambda x. \, F \, x \, \cst{a}$ is strictly solid, but $F \, \cst{a} \, (\cst{g} \, (\lambda y. \,
+y\,\cst{a}\,\cst{a})  \, \cst{a})$ is not, as the second
+argument of $F$ is a ground term of third order. 
+\end{exa}
+% \pagebreak[4]
+
+Prehofer's thesis states that PT terminates on $\{s \unif t\}$
+if $s$ is linear, $s$ shares no free variables with~$t$, $s$ is strictly solid,
+and $t$ is second-order. Together with completeness of PT for preunification,
+this result implies that PT procedure can be used to enumerate finitely many
+elements of complete set of preunifiers for this class of terms.
+
+Prehofer focused on the preunification problem as he remarks that the resulting
+flex-flex pairs are ``intricate" \cite[Sect.~5.2.2]{cp-95-unifphd}. We discovered that these flex-flex
+pairs actually have an MGU, allowing us to solve the full unification problem,
+rather than preunification problem. Moreover, we lift the order restriction
+imposed by Prehofer: We identify a class of terms called \emph{solid} which
+requires that all arguments of free variables are either bound variables (of
+arbitrary type) or ground (arbitrary-order) terms of base type.
+
+\begin{exa}
+  Consider the setting of Example \ref{unif:ex:solid-terms}. The terms $F \, G \,
+  \cst{a}$ and $H \, (\lambda x. \, x) \, \cst{a}$ are not solid, for the same reasons they are not strictly solid.
+  However, since the order restriction is lifted, $F \, \cst{a} \, (\cst{g} \, (\lambda y. \,
+  y\,\cst{a}\,\cst{a})  \, \cst{a})$ is solid.
+\end{exa}
+
+\looseness=-1 
+In other words, we extend the above preunification decidability result
+for linear strictly solid terms along two axes: we create an oracle
+for the full unification problem and we lift the
+order constraints.
+%
+To enumerate a CSU for a problem $E_0 = \{s \unif t\}$, where $s$ and $t$ are solid,
+$s$ is linear and shares no free variables with~$t$, our oracle applies the following two steps:
+\begin{enumerate}
+  \item Apply the PT procedure on $E_0$ to obtain a preunifier $\sigma$ with
+  flex-flex constraints $E_n$.
+  \item\label{step:ff}\looseness=-1 If $E_n$ is empty, return $\sigma$. Otherwise, choose a flex-flex constraint $u \unif v$ from $E_n$, and let the MGU of $u \unif v$
+  be $\varrho$. Then, set $E_n := \varrho(E_n \setminus \{ u \unif v \})$ and
+   $\sigma := \varrho\sigma$, and repeat step \ref{step:ff}.
+\end{enumerate}
+
+To show that this oracle terminates and yields a CSU, we must prove that the PT
+procedure terminates on above described class of problems (Lemma
+\ref{unif:lemma:solid-termination-preunif}). Moreover, we must show how to compute MGUs for
+the remaining flex-flex pairs (Lemma \ref{unif:lemma:solid-var-same-mgu} and
+\ref{unif:lemma:solid-vars-diff-mgu}). Our results are combined together in Theorem
+\ref{unif:thm:oracle}.
+
+Towards proving that the PT
+procedure terminates on above described class of problems, we first consider the corresponding matching problem.
+A \emph{matching problem} is a unification problem
+$\{ s \unif t \}$ where $t$ is ground. In what
+follows, we establish some useful properties of matching problems in which both
+$s$ and $t$ are solid (\emph{solid matching problems}). We call the unifier
+of a matching problem a \emph{matcher}.
+%TODO: @PETAR: This reference should be moved up because the previous lemma also uses multisets.
+%Done.
+
+\begin{lemma}
+    \label{unif:lemma:pt-matching-termination}
+    PT terminates on a solid matching problem $\{s \unif t\}$.
+\end{lemma}
+\begin{proof}
+    We show termination by designing a measure on a matching problem $E$ that
+    decreases with each application of a PT transition (possibly followed by
+    \textsf{Decomposition} or \textsf{Failure} steps). 
+    Our measure function
+    compares the following properties lexicographically:
+    \begin{description}
+        \item[A] multiset of the sizes of right-hand sides of the constraints in $E$
+        \item[B] number of free variables in $E$
+    \end{description}
+
+    Clearly, when applying PT transitions, the problem stays a matching problem
+    because the applied bindings do not introduce free variables on the
+    right-hand sides. It is also easy to check that each applied binding keeps
+    the terms in the solid fragment. Namely, bindings for both
+    \textsf{Imitation} and \textsf{Projection} transitions are patterns, which means
+    that, after applying the binding, fresh free variables are applied only to
+    bound variables or ground base-type terms. The \textsf{Solution} transition effectively
+    replaces the variable with a ground term, which is obviously solid. We show each transition 
+    either trivially terminates or reduces the measure:
+
+    \begin{description}
+      \item[Deletion] A decreases.
+      \item[Decomposition] A decreases.
+      \item[Failure] Trivial -- represents a terminal node.
+      \item[Solution] This transition applies on 
+        constraints of the form
+        $F \unif t$. This reduces A, since the constraint $F \unif t$ is removed
+        and $F$ cannot appear on the right-hand side.
+      \item[Imitation] The rule is applicable only on 
+        $\lambda \overline{x}.\, F \, \overline{s}_m  \unif \lambda
+        \overline{x}.\, \cst{f} \, \overline{t}_n$. The \textsf{Imitation} transition
+        replaces the constraint with $\{H_i \,
+        \overline{s}_m  \unif \overline{t}_i \mid 1\le i\le n\}$. This reduces A, as $F$ does not appear on any right-hand side.
+    \item[Projection] The rule is applicable only on 
+        $\lambda \overline{x}.\, F \, \overline{s}_m  \unif \lambda
+        \overline{x}.\, a \, \overline{t}_n$. If $a$ is a bound variable,
+        and we project $F$ to argument $s_i$ different than $a$, \textsf{Failure} applies and
+        PT trivially terminates. If we project $F$ to $a$, we apply \textsf{Decomposition},
+        which reduces $A$. If $a$ is a constant, for \textsf{Failure} not to apply, we
+        have to project to a base-type ground term $s_i$. This does not increase A,
+        since no variables appear on right-hand sides, but removes the variable $F$
+        from $E$, reducing B by one. \qedhere
+    \end{description}
+\end{proof}
+
+Recall that we say $\sigma$ is a grounding substitution if for every variable $F$ mapped
+by $\sigma$, $\sigma F$ is ground (Sect.~\ref{sec:pre:fol}).
+\begin{lemma}
+    \label{unif:lemma:pt-matching-ground}
+  All unifiers produced by PT for the solid matching problem $\{s \unif t\}$ are grounding substitutions.
+\end{lemma}
+\begin{proof}
+  Closely following the proof of Lemma 5.2.5 in Prehofer's PhD thesis
+  \cite{cp-95-unifphd}, we prove our claim by induction on the length of the
+  PT transition sequence that leads to the unifier. We know this
+  sequence is finite by Lemma \ref{unif:lemma:pt-matching-termination}. The base case of
+  induction, for the empty sequence, is trivial. The induction step is made using
+  one of the following transitions:
+
+  \begin{description}
+    \item[Deletion] Trivial.
+    \item[Decomposition] Trivial.
+    \item[Failure] This rule is not relevant, since it will not lead to the unifier.
+    \item[Solution] This rule applies the substitution $\{F \mapsto t\}$, which 
+    is grounding since $t$ is ground.
+    \item[Imitation] This transition applies on 
+    constraints of the form
+    \[ \lambda \overline{x}. \, F \, \overline{s}_k \unif \lambda
+    \overline{x}. \, \cst{f} \, \overline{t}_l \]
+    The binding for \textsf{Imitation} is
+    $\varrho = \{F \mapsto \lambda \overline{x}_k.\, \cst{f}\, (G_1 \,
+    \overline{s}_k ) \ldots (G_l \, \overline{s}_k ) \}$, and reduces the
+    problem to $\{G_i \,  \overline{s}_k \unif t_i \mid 1\le i\le l\}$. Since the right-hand sides are ground, 
+    any unifier $\sigma$ produced by PT must map all of the variables $G_1, \ldots, G_l$. By induction hypothesis,
+    $\sigma G_i$ is ground. Therefore, $\sigma\varrho$ must map $F$ to a ground term.
+    \item[Projection] This transition applies on 
+    constraints of the form 
+    \[\lambda \overline{x}. \, F \, \overline{s}_k \unif \lambda
+    \overline{x}. \, t\]
+    The binding for \textsf{Projection} is
+    $\varrho = \{F \mapsto \lambda \overline{x}_k. \, x_i \, (G_1 \,
+      \overline{x}_k) \ldots (G_j \, \overline{x}_k)\}$,
+    and reduces the problem to
+    $\{s_i \, (G_1 \,
+      \overline{s}_k) \ldots (G_j \, \overline{s}_k) \unif a \, \overline{t}_l\}$.
+      Since the right-hand side is ground, 
+      any unifier $\sigma$ produced by PT must map all of the variables $G_1, \ldots, G_j$. 
+    By induction hypothesis,
+    $\sigma G_i$ is ground. Therefore, $\sigma\varrho$ must map $F$ to a ground term. \qedhere
+  \end{description}
+\end{proof}
+
+\begin{lemma}
+  \label{unif:lemma:solid-termination-preunif}
+  \looseness=-1
+  If $s$ and $t$ are solid, $s$ is linear and shares no free variables with $t$, then
+  PT terminates for the preunification problem $\{ s \unif t \}$, and
+  all remaining flex-flex constraints are solid.
+\end{lemma}
+\begin{proof}
+  This lemma is a modification of Lemma 5.2.1 and Theorem 5.2.6 from Prehofer's
+  PhD thesis \cite{cp-95-unifphd}. Correspondingly, the following proof closely
+  follows the proofs for these two lemmas. We also adopt the notion of an
+  \emph{isolated} variable from Prehofer's thesis: a variable is isolated in a
+  multiset $E$ of constraints if it appears exactly once in $E$.
+
+  \looseness=-1
+  First, similarly to the proof of previous lemma we conclude each transition
+  maintains the condition that the terms remain solid. Second, we have to show
+  that variables on left-hand sides remain isolated. For \textsf{Imitation} and
+  \textsf{Projection} rules, the preservation of this invariant is obvious.
+  Later, we also show that \textsf{Solution} preserves it. Third, since $s$ and
+  $t$ share no variables, and $s$ is linear, no rule can introduce a variable
+  from the right-hand side to the left-hand side.
+
+  To prove termination of PT, we devise a measure that
+  decreases with each application of a PT transition. The measure
+  lexicographically compares the following properties:
+  \begin{description}
+    \item[A] number of occurrences of constant symbols and bound variables on left-hand sides that
+    are not below free variables
+    \item[B] number of free variables on right-hand sides
+    \item[C] multiset of the sizes of right-hand sides
+  \end{description}
+
+  We show that each transition either trivially terminates or reduces the measure:
+  \begin{description}
+    \item[Deletion] A does not increase and at least one of A or B reduces.
+    \item[Decomposition] A reduces.
+    \item[Failure] Trivial. 
+    \item[Solution] This rule applies in two cases:
+
+    \begin{itemize}
+      \item $F \unif t$: since $F$ is isolated, A is unchanged, B is not
+      increased, and C reduces. All variables on left-hand sides remain isolated
+      since they are unaffected by this substitution.
+      \item $t \unif F $: since $F$ does not occur on any left-hand side and the
+      head of $t$ must be a constant or bound variable (otherwise the rule would
+      not be applicable), $A$ reduces. Even though $t$ might contain some free
+      variables and $F$ can have multiple occurrences on right-hand sides, all
+      free variables on left-hand sides remain isolated. Namely, all free variables
+      in $t$ occur exactly once in the multiset, and since $t \unif F$ is
+      removed, all of them either disappear or end up on right-hand sides.
+    \end{itemize}
+    \item[Imitation] We distinguish the following two forms of the selected constraint:
+    \begin{itemize}
+      \item $\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda \overline{x}. \,
+      \cst{f} \, \overline{u}_m$: 
+      We replace this constraint by 
+      $\{H_i \, \overline{v}_n \unif u_i \mid 1\le i\le m\}$
+      and apply the \textsf{Imitation} binding
+      $F \mapsto \lambda \overline{x}_n.\, \cst{f} \, (H_1 \, \overline{x}_n) \allowbreak
+      \ldots (H_m \, \overline{x}_n)$.
+      As $F$ is isolated, A and B do not increase.
+      Since $F$ is isolated, it does not occur on any right-hand side, and hence C decreases.
+      \item $\lambda \overline{x}. \, \cst{f} \, \overline{u}_m \unif \lambda \overline{x}. \, F \, \overline{v}_n $:
+      applying the \textsf{Imitation} transition 
+      as above will reduce A since $F$ cannot appear on any
+      left-hand side.
+    \end{itemize}
+    \item[Projection] Similarly to the previous transition rule, we have two cases:
+      \begin{itemize}
+        \item$\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda
+        \overline{x}. \, a \, \overline{u}_m$: if $a$ is a bound variable,
+        then for \textsf{Failure} not to be applicable afterwards, we have to
+        project $F$ onto argument $v_i$ equal to $a$. Then we proceed like for
+        \textsf{Imitation}. If $a$ is not a bound variable, then we have to
+        project to some base-type term $v_j$ (otherwise \textsf{Failure} would
+        be applicable afterwards). This reduces the problem to $\lambda
+        \overline{x}. \, v_j \unif \lambda \overline{x}. \, a \,
+        \overline{u}_m$. This is a solid matching problem, whose solutions
+        computed by PT are grounding substitutions (see Lemma
+        \ref{unif:lemma:pt-matching-ground}). Applying one of those solutions will
+        eliminate all the free variables in $\lambda \overline{x}.\, a \,
+        \overline{u}_m$. Since PT is parametrized by an admissible
+        selection function, we know that there are no constraints descending
+        from a simple projection in $E$ since the constraint $\lambda \overline{x}. \, F
+        \, \overline{v}_n \unif \lambda \overline{x}. \, a \, \overline{u}_m$ was
+        chosen,
+        which, due to solidity restrictions, cannot be such a descendant. 
+        Therefore, we know that PT will transform the descendants
+        of the matching problem $\lambda \overline{x}. \, v_j \unif \lambda
+        \overline{x}.\, a \, \overline{u}_m$ until either \textsf{Failure} is
+        observed (making PT trivially terminating) or until no descendant exists
+        and the grounding matcher is computed (see Lemmas
+        \ref{unif:lemma:pt-matching-ground} and \ref{unif:lemma:pt-matching-termination}).
+        This results in removal of the original constraint $\lambda
+        \overline{x}. \, F \, \overline{v}_n \unif \lambda \overline{x}. \,
+        a \, \overline{u}_m$ and application of the computed grounding
+        matcher, which will either remove all the free variables in the right-hand
+        side of the constraint (not increasing A and reducing B)
+        or not increasing A and B, reduce C if no free variables occur in the right-hand side.
+        \item $\lambda \overline{x}. \, a \, \overline{v}_n \unif \lambda
+        \overline{x}. \, F \, \overline{u}_m$: 
+        if $a$ is a bound variable, projecting $F$ onto argument $u_i$ will either enable
+        application of \textsf{Decomposition} as the next step reducing A, or it will result in \textsf{Failure},
+        trivially terminating. If $a$ is a constant, then projecting $F$ onto some $u_j$
+        will either yield  \textsf{Failure} or enable \textsf{Decomposition}, reducing A. \qedhere
+      \end{itemize} 
+    \end{description}
+\end{proof}
+
+Enumerating a CSU for a solid flex-flex pair may seem as hard as 
+for any other flex-flex pair; however,
+the following two lemmas show that solid pairs admit an MGU:
+
+\begin{lemma}
+    \label{unif:lemma:solid-var-same-mgu}
+    The unification problem 
+    $\{\lambda \overline{z}.\, F\,\overline{s}_m \unif   \lambda \overline{z}.\, F\,\overline{s}'_{m}\}$, where both terms are
+    solid, has an MGU of the form $\sigma=\{F \mapsto \lambda \overline{x}_m.\,
+    G \, x_{j_1} \ldots x_{j_r}\}$ where $G$ is an auxiliary variable, and $1 \leq
+    j_1 < \cdots < j_r \leq m$ are exactly those indices $j_i$ for which
+    $s_{j_i} = s'_{j_i}$.
+\end{lemma}
+\begin{proof}
+    Let $\varrho$ be a unifier for the given unification problem.
+    Let $\lambda \overline{x}.\, u = \varrho F$. Take an arbitrary subterm of $u$ whose head is a bound
+    variable $x_i$. If $x_i$ is of function type, it corresponds to either $s_i$
+    or $s'_i$ which, due to solidity restrictions, has to be a bound variable.
+    Furthermore, since $\varrho$ is a unifier, $s_i$ and $s'_i$ have to be the
+    syntactically equal. Similarly, if $x_i$ is of base type, it corresponds
+    to two ground terms $s_i$ and $s'_i$ which have to be syntactically equal.
+    We conclude that $\varrho$ can use variables from $\overline{x}_n$ only
+    if they correspond to syntactically equal terms. Therefore, 
+    there is a substitution $\theta$ such that $\rho X = \theta\sigma X$ for all $X \not = G$. 
+    Due to arbitrary choice of $\varrho$, we conclude that $\sigma$ is an MGU.
+\end{proof}
+
+\begin{lemma}
+  \label{unif:lemma:solid-vars-diff-mgu}
+  Let 
+  $\{\lambda \overline{x}.\, F \, \overline{s}_m \unif \lambda\overline{x}.\, F' \, \overline{s}'_{m'}\}$
+  be a solid unification problem
+  where $F \neq F'$.
+  By Lemma~\ref{unif:lemma:pt-matching-termination}, there exists a finite CSU $\{\sigma_i^1, \dots, \sigma_i^{k_i} \}$
+  of the problem $\{s_i \unif H_i \, \overline{s}'_{m'}\}$, 
+  where $H_i$ is a fresh free variable.
+  Let 
+  $\lambda\overline{y}_{m'}. \, s_i^j = \lambda\overline{y}_{m'}. \, \sigma_i^j(H_i) \,\overline{y}_{m'}$.
+  Similarly, also by Lemma~\ref{unif:lemma:pt-matching-termination}, 
+  there exists a finite CSU $\{ \tilde\sigma_i^1, \dots, \tilde\sigma_i^{l_i} \}$
+  of the problem $\{s'_i \unif \tilde H_i \, \overline{s}_m\}$, where $\tilde H_i$
+  is a fresh free variable. 
+  Let 
+  $\lambda \overline{x}_m. \, {s'}_i^j = \lambda \overline{x}_m. \, \tilde\sigma_i^j(\tilde H_i) \, \overline{x}_m$.
+  Let $Z$ be a fresh free variable.
+  An MGU $\sigma$ for the given problem is
+    \begin{align*} 
+      F &\mapsto \lambda \overline{x}_m.\, Z
+          \, \underbrace{x_1 \, \ldots \, x_1}_{k_1 \text{ times}} \, \ldots \,
+             \underbrace{x_m \, \ldots \, x_m}_{k_m \text{ times}} 
+             \, {s'}_1^1 \ldots {s'}_1^{l_1} \, \ldots \, {s'}_{m'}^1 \ldots {s'}_{m'}^{l_{m'}} \, \\
+      F' &\mapsto \lambda \overline{y}_{m'}. \, Z
+          \, s_1^1 \ldots s_1^{k_1} \ldots s_m^1 \ldots s_m^{k_m} \,
+      \underbrace{y_1 \, \ldots y_1}_{l_1 \text{ times}} \, \ldots \,
+      \underbrace{y_{m'} \, \ldots y_{m'}}_{l_{m'} \text{ times}}
+    \end{align*}
+  where the auxiliary variables are $H_1, \dots, H_m$, $\tilde H_1, \dots \tilde H_{m'}$, $Z$ and all auxiliary variables associated with the above CSUs.
+\end{lemma}
+\begin{proof}
+  Let $\varrho$ be an arbitrary unifier for the problem $\lambda
+  \overline{x}.\, F \, \overline{s}_m \unif \lambda\overline{x}.\, F' \,
+  \overline{s}'_{m'}$. We prove $\sigma$ is an MGU by showing that there exists a substitution
+  $\theta$ such that $\varrho X = \theta\sigma X$ for all non-auxiliary variables $X$. 
+  We can focus only on $X \in \{F, F'\}$
+  because all other non-auxiliary variables
+  do neither appear in the original problem nor in $\sigma F$ or $\sigma F'$; so we can simply define $\theta X = \varrho X$.
+
+  %In what follows we consider all terms
+  %in their $\beta$-reduced $\eta$-long form.
+
+  Let $\lambda \overline{x}_m. \, u = \varrho F$ and  $\lambda \overline{y}_{m'}. \, u' = \varrho{F'}$, where
+  the bound variables $\overline{x}_m$ and $\overline{y}_{m'}$ have been $\alpha$-renamed apart. We also assume
+  that the names of variables bound inside $u$ and $u'$ are $\alpha$-renamed so that they are different from $\overline{x}_m$ and
+  $\overline{y}_{m'}$. Finally, bound variables
+  from the definition of $\sigma$ have been $\alpha$-renamed to match $\overline{x}_m$ and $\overline{y}_{m'}$.
+  We
+  define $\theta$ to be the substitution
+  \[\theta = \{ Z \mapsto \lambda z_1^1 \ldots z_1^{k_1} \ldots z_m^1 \ldots z_m^{k_m} \, w_1^{1} \ldots w_1^{l_1} \ldots w_{m'}^1 \ldots  w_{m'}^{l_{m'}}. \, \textsf{diff}(u, u') \}\]
+  where \textsf{diff}$(v,v')$ is defined recursively as
+  \begin{alignat}{3}
+    &\textsf{diff}(\lambda x.\, v, \lambda y.\, v') &&{}={}&& \lambda x. \, \textsf{diff}(v, \{y \mapsto x\}v') \label{diff:alpha} \\
+    &\textsf{diff}(a \, \overline{v}_n, a \, \overline{v}'_n) &&=&&  a \, \textsf{diff}(v_1, v'_1) \ldots \textsf{diff}(v_n, v'_n) \label{diff:samehd} \\
+    &\textsf{diff}(x_i, v') &&=&&  z_i^k \text{, if $v' = {s}_i^k$ } \label{diff:xi}  \\
+    &\textsf{diff}(v, y_i) &&=&&  w_i^l \text{, if $v = {s'}_i^l$ } \label{diff:yi}  \\
+    &\textsf{diff}(x_i \, \overline{v}_n, y_j \, \overline{v}'_n) &&=&&  z_i^k \, \textsf{diff}(v_1, v'_1) \ldots \textsf{diff}(v_n, v'_n) \text{, if $y_j = s_i^k $ } \label{diff:xiyj} 
+  \end{alignat}
+
+  \looseness=-1
+  From \textsf{diff}'s definition it is clear that there are terms $v,v'$ for which it is undefined.
+  However, we will show that for each $u$ and $u'$ that are bodies of bindings from a unifier $\varrho$, 
+  \textsf{diff} is defined and has the desired property.
+  In equations \ref{diff:xi}, \ref{diff:yi}, and \ref{diff:xiyj}, if there are multiple numbers $k$ or $l$
+  that fulfill the condition, choose an arbitrary one.
+  We need to show that $\varrho F = \theta\sigma F$ and $\varrho F' = \theta\sigma F'$.
+  By the definitions of $u$, $u'$, $\theta$ and $\sigma$ and $\beta$-reduction, this is equivalent to
+  \begin{align*}
+    \lambda \overline{x}_m.\, u &= \lambda \overline{x}_m.\,
+    \{z_i^k \mapsto x_i, w_i^l \mapsto {s'}_i^l \text{ for all }k,l,i\}\,\textsf{diff}(u,u')\\
+    \lambda \overline{y}_{m'}.\, u' &= \lambda \overline{y}_{m'}.\,
+    \{z_i^k \mapsto s_i^k, w_i^l \mapsto y_i \text{ for all }k,l,i\}\,\textsf{diff}(u,u')
+  \end{align*}
+  %
+  We will show by induction that for any $\lambda \overline{x}_m.\, v$, $\lambda \overline{y}_m.\, v'$ such that 
+  \[\{x_1 \mapsto s_1, \ldots, x_m \mapsto  s_m\} v = \{y_1 \mapsto s'_1, \ldots, y_{m'} \mapsto  s'_{m'}\} v' \tag{$\star$}\]
+  we have
+  \begin{align*}
+    \tag{\textdagger}
+    \renewcommand{\arraystretch}{1.2}% Spread out array rows
+    \begin{array}{c@{\qquad}l}
+      v &=
+      \{z_i^k \mapsto x_i, w_i^l \mapsto {s'}_i^l \text{ for all }k,l,i\}\,\textsf{diff}(v,v')\\
+      v' &=
+      \{z_i^k \mapsto s_i^k, w_i^l \mapsto y_i \text{ for all }k,l,i\}\,\textsf{diff}(v,v')
+    \end{array}
+  \end{align*}
+  The equation ($\star$) holds for $v = u$ and $v' = u'$ because
+  $\varrho$ is a unifier of
+  $\lambda
+  \overline{x}.\, F \, \overline{s}_m \unif \lambda\overline{x}.\, F' \,
+  \overline{s}'_{m'}$.
+  Therefore, once we have shown that ($\star$) implies ($\dagger$),
+  we know that ($\dagger$) holds for $v = u$ and $v' = u'$ and we are done.
+
+  We prove that ($\star$) implies ($\dagger$) by induction on the size of $v$ and $v'$. We consider
+  the following cases:
+
+  \begin{description}
+    \item[$v = \lambda x. \, v_1$]  For ($\star$) to hold, $v$ and $v'$ must be of the
+    same type. Therefore, the $\lambda$-prefixes of their $\eta$-long representatives
+    must have the same length and we can apply equation
+    $\ref{diff:alpha}$. By the induction hypothesis, ($\dagger$) holds.
+    \item[$v = x_i$] In this case, $\{x_1 \mapsto s_1, \ldots, x_m \mapsto
+    s_m\} v = s_i$. Since ($\star$) holds, $v'$ must be an instance of a unifier from
+    the CSU of $s_i = H_i \, \overline{s}'_{m'}$. However, since $s_i$ and all
+    terms in $\overline{s}'_{m'}$ are ground, $\lambda \overline{y}_m. \, v' =
+    \sigma_i^k(H_i)$, for some $k$. Then, $\textsf{diff}(x_i, v') = z_i^k$, and
+    it is easy to check that ($\dagger$) holds.
+    \item[$v = x_i \, \overline{v}_n, n > 0$] In this case, $x_i$ is mapped to $s_i$
+    which, due to solidity restrictions, has to be a functional bound variable.
+    Since ($\star$) holds, we conclude that the head of $\{y_1 \mapsto s'_1, \ldots,
+    y_{m'} \mapsto  s'_{m'}\}v'$ must be $s'_{j}$, such that $s'_{j} = s_i$;
+    this also means that $v' = y_j \, \overline{v'}_n$. Therefore, it is easy to
+    check that some $\tau = \{ H_i \mapsto \lambda \overline{y}_{m'}. \, y_j\}$ is
+    a matcher for the problem $s_i = H_i \, \overline{s}'_{m'}$. For some $k$,
+    $\sigma_i^k = \tau$, i.e., $\textsf{diff}(v,v') = z_i^k \, \textsf{diff}(v_1, v'_1)
+    \ldots \textsf{diff}(v_n, v'_n) $. By induction hypothesis, we get that $(\dagger)$ holds.
+    \item[$v = a \, \overline{v}_n$] In the remaining cases $a$ is either a free variable, a loose bound variable
+    different than $x_1, \ldots, x_m$, or a constant. If $a$ is a free variable or a loose bound variable
+    different than $x_1, \ldots, x_m$, then $v' = a \, \overline{v'}_n$, since
+    ($\star$) holds, all of $\overline{s}'_{m'}$ are ground, and bound variables different than
+    $x_1, \ldots, x_m$ and $y_1, \ldots, y_{m'}$ are renamed to match by equation \ref{diff:alpha}. 
+    By the induction
+    hypothesis and by equation \ref{diff:samehd}, we obtain $(\dagger)$. If $a$ is a constant, we consider two
+    cases: either $v' = a \, \overline{v'}_n$, which allows us to apply the induction
+    hypothesis and obtain $(\dagger)$ as above, or $v' = y_j \, \overline{v'}_m$.
+    Since $a$ is a constant, $y_j$ cannot be a functional bound variable, since
+    then it would be mapped to $s'_j$, which due to solidity restrictions 
+    also has to be a functional bound
+    variable and $(\star)$ would not hold. Therefore $v' = y_j$. In this case, we
+    proceed as in the case $v = x_i$ with the roles of $v$ and $v'$ swapped. \qedhere
+  \end{description}
+
+  % It is easy to check that, in the cases in which it is defined, \textsf{diff} specifies
+  % $\theta$ so that $\varrho \subseteq \theta \sigma$, as desired. Now let's show that $\textsf{diff}(u,u')$
+  % is defined for every unifier $\varrho$. This function computes the
+  % needed instantiation by inspecting all positions $p$ defined for both $u$ and
+  % $u'$. We reduce the problem of definability of $\textsf{diff}(u,u')$ to definability
+  % of $\textsf{diff}(u|_p,u'|_p)$ for all such $p$. Notably, \textsf{diff} is undefined for $u|_p$ and $u'|_p$ with heads that are
+  % different free variables, constants or variables bound in $u$ and $u'$
+  % (assuming they are $\alpha$-renamed as in equation \ref{diff:alpha}). However,
+  % if $u|_p$ and $u'|_p$ have such heads, $\varrho$ cannot be a
+  % unifier, due to solidity restrictions, justifying equation \ref{diff:samehd}. Similarly, since $u$ and $u'$ have the same type 
+  % then $u|_p$ and $u'|_p$ (i.e., one of the elements of their $\eta$-equivalence class) have $\lambda$-prefixes of the same
+  % size, justifying equation \ref{diff:alpha}.
+  % The remaining case is that $u|_p$ and $u'|_p$ have different heads and at least one of $u|_p$
+  % and $u'|_p$ has a head that is a variable bound in $\overline{x}_m$ or
+  % $\overline{y}_{m'}$. First, assume that this bound variable appears unapplied. 
+  % This corresponds to equations \ref{diff:xi} and \ref{diff:yi}. For brevity, we will consider the case that the head of
+  % $u|_p = x_i$. The other case is symmetrical. Bound variable
+  % $x_i$ corresponds to $s_i$ which is a ground term, due to solidity
+  % restrictions. Since $\varrho$ is a unifier, $(\lambda \overline{y}_{m'}. \,
+  % u'|_p) \, \overline{s}'_{m'} = s_i$. In other words, $u'|_p$ is the body of
+  % $\sigma_i^k H_i$ where $\sigma_i^k$ is one of the finitely many (see Lemma
+  % \ref{unif:lemma:pt-matching-termination}) matchers for the matching problem $s_i
+  % \unif H_i \, \overline{s}'_{m'}$. By definition of $\sigma$, one of those
+  % bodies has to be $s_i^j$. In this case, the needed instantiation is described
+  % by equation \ref{diff:xi}. Now let's consider the case $u|_p = x_i \, \overline{u}_n$. Clearly,
+  % $x_i$ corresponds to $s_i$ which must be a bound variable of functional type,
+  % due to solidity restrictions. Since $\varrho$ is a unifier, $u'|_p$ must have
+  % some $y_j$ (a variable bound in $\overline{y}_{m'}$) for a head. Bound
+  % variable $y_j$ corresponds to $s'_j$ which must be syntactically equal to
+  % $s_i$ for $\varrho$ to be a unifier. It is easy to see that $\tau = \{ H_i
+  % \mapsto \lambda \overline{y}_{m'}. \, y_j \}$ is a matcher for the problem
+  % $s_i = H_i \, \overline{s'}_{m'}$. Therefore, for some $k$, $\sigma_i^k = \tau$
+  % and by definition of $\sigma$, $s_i^k = y_j$. Equation \ref{diff:xiyj} shows
+  % how to instantiate $Z$ in that case.
+\end{proof}
+
+\begin{theorem}
+    \label{unif:thm:oracle}
+    Let $s$ and $t$ be
+    solid terms that share no free variables,
+    and let $s$ be linear.
+    Then the unification problem $\{s \unif t\}$ has a finite CSU.
+\end{theorem}
+\begin{proof}
+    \looseness=-1
+    By Lemma~\ref{unif:lemma:solid-termination-preunif}, 
+    PT terminates on $\{s \unif t\}$ with a finite set of preunifiers $\sigma$,
+    each associated with a multiset $E$ of solid flex-flex pairs.
+
+    An MGU $\delta_E$ of the remaining multiset $E$
+    of solid flex-flex constraints can be found as follows.
+    Choose one constraint $(u \unif v) \in E$ and determine an MGU $\varrho$ for it
+    using Lemma \ref{unif:lemma:solid-var-same-mgu} or
+    \ref{unif:lemma:solid-vars-diff-mgu}.
+    Then the set $\varrho(E\setminus\{u \unif v\})$ also contains only solid flex-flex constraints,
+    and we iterate this process by choosing a constraint from $\varrho(E\setminus\{u \unif v\})$ next
+    until there are no constraints left, eventually
+    yielding an MGU $\varrho'$ of $\varrho(E\setminus\{u \unif v\})$. Finally, we obtain the MGU
+    $\delta_E = \varrho'\varrho$ of $E$.
+
+    Let $U = \{ \delta_E\sigma
+    \mid \text{PT produces preunifier $\sigma$ with constraints }E\}$. 
+    By termination of PT, $U$ is finite.
+    We show that $U$ is a CSU. 
+    Let $\varrho$ be an arbitrary unifier for $\{ s \unif t\}$. By Lemma~\ref{unif:lemma:pt-complete}, 
+    PT produces a preunifier $\sigma$ with flex-flex constraints $E$
+    such that there is a unifier $\theta$ of $E$ and
+    $\varrho X = \theta\sigma X$ for all $X$ not contained in the supply of fresh variables $V$. 
+    Since
+    $\delta_E$ is an MGU of $E$,
+    assuming that we use variables from $V$ in the role of the auxiliary variables,
+    there exists a
+    substitution $\theta'$ such that $\theta X = \theta' \delta_E X$ for all $X \not\in V$.
+    If we make sure that we never reuse fresh variables and that the supply $V$ does not contain any variables from the initial problem,
+    it follows that $\varrho X = \theta' \delta_E\sigma X$ for all $X \not\in V$.
+    Therefore, $U$ is a CSU.
+\end{proof}
+
+The proof of Theorem \ref{unif:thm:oracle}
+provides an effective way to calculate a CSU using PT and the results
+of Lemmas \ref{unif:lemma:solid-var-same-mgu} and \ref{unif:lemma:solid-vars-diff-mgu}.
+
+%@PETAR: I moved this example up by one paragraph. 
+% Unfortunately, I had to inline the unifier \sigma_3 to make this fit.
+% Maybe it should also use the same names for substitutions as in the previous para?
+ \begin{exa}
+   \label{unif:example:solid}
+   \looseness=-1
+  Let $\{F \, (\cst{f} \, \cst{a}) \unif \cst{g} \, \cst{a} \, (G \, \cst{a})\}$ be
+  the unification problem to solve. Projecting~$F$ onto the first argument will
+  lead to a nonunifiable problem, so we perform imitation of~$\cst{g}$ building a
+  binding $\sigma_1 = \{ F \mapsto \lambda x.\, \cst{g} \, (F_1 \, x) \, (F_2 \, x)
+  \}$. This yields the  problem $\{F_1
+  \, (\cst{f} \, \cst{a}) \unif \cst{a}, \allowbreak F_2 \, (\cst{f} \, \cst{a}) \unif G \,
+  \cst{a}\}$. Again, we can only imitate $\cst{a}$ for $F_1$ -- building a new 
+  binding $\sigma_2 = \{F_1 \mapsto \lambda x.\, \cst{a} \}$. Finally, this
+  yields the problem $\{ F_2 \, (\cst{f} \, \cst{a}) \unif G \, \cst{a} \}$.
+  According to Lemma \ref{unif:lemma:solid-vars-diff-mgu}, we find CSUs for the
+  problems $J_1 \, \cst{a} = \cst{f} \, \cst{a}$ and $I_1 \, (\cst{f} \, \cst{a})
+  \unif \cst{a}$ using PT. The latter problem has a singleton CSU $\{I_1 \mapsto \lambda x.\,
+  \cst{a}\}$, whereas the former has a CSU containing $\{ J_1 \mapsto \lambda x.\,
+  \cst{f} \, x \}$ and $\{ J_1 \mapsto \lambda x.\, \cst{f} \, \cst{a} \}$.
+  Combining these solutions, we obtain an MGU $\sigma_3 =
+  \{ F_2 \mapsto \lambda x.\, H \, x \, x \, \cst{a},\,
+  G   \mapsto \lambda x.\, H \, (\cst{f} \, \cst{a}) \, (\cst{f} \, x) \, x \}$ 
+  for $F_2 \, (\cst{f}\, \cst{a}) \unif G \, \cst{a}$.
+  %\begin{equation*}
+  %    F_2 \mapsto \lambda x.\, H \, x \, x \, \cst{a} \\
+  %    G   \mapsto \lambda x.\, H \, (\cst{f} \, \cst{a}) \, (\cst{f} \, x) \, x
+  %\end{equation*}
+  Finally, we get to the MGU of the original problem $\sigma = \sigma_3 \sigma_2 \sigma_1 = \{ F
+  \mapsto \lambda x.\, \cst{g} \, \cst{a} \, (H \, x \, x \, \cst{a}),\allowbreak G \mapsto
+  \lambda x.\, H \, (\cst{f} \, \cst{a}) \, (\cst{f} \, x) \, x  \}$.
+  For brevity, we omitted the intermediate bindings of auxiliary variables
+  in $\sigma$.
+\end{exa}
+
+The solid fragment is useful for automatic theorem provers based on
+$\lambda$-su\-per\-po\-si\-tion~\cite{bbtvw-21-sup-lam}. As Example \ref{unif:example:solid} shows, when the solid oracle
+is used, superposing from $F \, (\cst{f} \, \cst{a})$ into $\cst{g} \, \cst{a}
+\, (G \, \cst{a})$ yields a single clause; without it, our procedure does not
+terminate.
+
+Small examples that violate conditions of Theorem \ref{unif:thm:oracle} 
+and admit only infinite CSUs can be found easily. The problem $\{ \lambda x.\, F \, (\cst{f} \, x) \unif \lambda x.\, \cst{f} \,
+(F \, x) \}$ violates variable distinctness and is a well-known example of a problem with only infinite CSUs.
+%TODO: @PETAR: Do we have a citation, if this is well-known? 
+%It is really really well-known :)
+Similarly,
+$\lambda x.\, \cst{g} \, (F \, (\cst{f} \, x)) \, F  \unif \lambda x.\, \cst{g}
+\, (\cst{f} \, (G \, x)) \, G $, which violates linearity, reduces to the previous problem. 
+Only ground arguments to free variables are allowed because  $\{F \, X \unif G
+\, \cst{a}\}$ has only infinite CSUs. Finally, it is crucial that
+functional arguments to free variables are only bound variables: the problem
+$\{ \lambda y. \, X (\lambda x.\, x) \, y \unif \lambda y. \, y \}$ has only infinite CSUs.
+
 \section{An Extension of Fingerprint Indexing}

### 2022-03-02 unif/unif.tex
concl

@@
 
+\looseness=-1
+A fundamental building block for almost all automated reasoning tools is the
+operation of retrieving term pairs that satisfy certain conditions, e.g.,
+unifiable terms, instances or generalizations. Indexing data structures are used
+to implement this operation efficiently. If the data structure retrieves
+precisely the terms that satisfy the condition, it is called \emph{perfect}.
+
+Higher-order indexing has received little attention compared to its
+first-order counterpart. However, recent research in higher-order theorem
+proving increased the interest in higher-order indexing
+\cite{ls-16-indexing,br-19-restricted-unif}.
+A \emph{fingerprint index} \cite{ss-12-fp-indexing,vbss-19-ehoh1} is an imperfect index based on
+the idea that the skeleton of the term consisting of all non-variable positions
+is not affected by substitutions. Therefore, we can easily determine that the terms
+are not unifiable (or matchable) if they disagree on a fixed set of sample
+positions.
+
+More formally, when we sample an untyped first-order term $t$ on a sample position $p$, 
+the \emph{generic fingerprinting function} $\text{gfpf}$
+distinguishes four possibilities:
+\[
+  \kern5em
+  \text{gfpf}(t,p) =
+  \begin{cases}
+    \cst{f} & \text{if $t|_p$ has a symbol head $\cst{f}$} \\[-\jot]
+    \AA & \text{if $t|_p$ is a variable} \\[-\jot]
+    \BB & \text{if $t|_q$ is a variable for some proper prefix $q$ of $p$} \\[-\jot]
+    \NN & \text{otherwise}
+  \end{cases}
+  \kern5em
+\]
+We define the \emph{fingerprinting function} $\text{fp}(t) = (\text{gfpf}(t,
+p_1), \ldots, \text{gfpf}(t, p_n))$, based on a fixed tuple of positions $\overline{p}_n$.
+Determining whether two terms are compatible for a given retrieval operation reduces
+to checking their fingerprints' componentwise compatibility. The following matrices
+determine the compatibility for retrieval operations:
+\begin{align*}
+    && \hbox{\begin{tabular}{@{}C{1em}|C{1em}C{1em}C{1em}C{1em}C{1em}@{}}
+    & $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+    \hline
+    $\cst{f}_1\vphantom{^{()}}$ & & \xmark & & & \xmark \\
+    \AA & & & & & \xmark \\
+    \BB & & & & & \\
+    \NN & \xmark & \xmark & \xmark & &
+    \end{tabular}}
+    &&
+    \hbox{\begin{tabular}{@{}C{1em}|C{1em}C{1em}C{1em}C{1em}C{1em}@{}}
+    & $\cst{f}_1$ & $\cst{f}_2$ & \AA & \BB & \NN \\
+    \hline
+    $\cst{f}_1\vphantom{^{()}}$ & & \xmark & \xmark & \xmark & \xmark \\
+    \AA & & & & \xmark & \xmark \\
+    \BB & & & & & \\
+    \NN & \xmark & \xmark & \xmark & \xmark &
+    \end{tabular}}
+\end{align*}
+The left matrix determines unification compatibility, while the right matrix
+determines compatibility for matching term $s$ (rows) onto term $t$ (columns). Symbols
+$\cst{f}_1$ and $\cst{f}_2$ stand for arbitrary distinct constants. Incompatible
+features are marked with \xmark. For example, given a tuple of term positions $(1, 1.1.1, 2)$, and
+terms $\cst{f}(\cst{g}(X), \cst{b})$ and $\cst{f}(\cst{f}(\cst{a},\cst{a}),\cst{b})$, 
+their fingerprints are $(\cst{g}, \BB, \cst{b})$ and $(\cst{f},
+\NN, \cst{b})$, respectively. Since the first fingerprint component is incompatible,
+terms are not unifiable.
+
+Fingerprints for the terms in the index are stored in a trie data structure.
+This allows us to efficiently filter out terms that are not compatible with a given
+retrieval condition. For the remaining terms, a unification or matching
+procedure must be invoked to determine whether they satisfy the condition or not.
+
+The fundamental idea of first-order fingerprint indexing carries over to
+higher-order terms -- 
+application of a substitution does not change the rigid skeleton of a term.
+However, to extend fingerprint indexing to
+higher-order terms, we must address the issues of $\alpha\beta\eta$-normalization and figure how to cope with $\lambda$-abstractions and
+bound variables. To that end, we define a function $\hotofo{t}$, defined on
+%TODO: @PETAR \eta-long?
+% Yes
+$\beta$-reduced $\eta$-long terms in De Bruijn \cite{db-75-dbindices} notation:
+%TODO: @PETAR: also mention that this is an encoding into FOL before the definition?
+% @ALEX: There is a sentence about that later. I do not see how it can easily fit in this description :(
+\begin{align*}
+& \hotofo{F \, \overline{s}} = F
+&& \hotofo{x_i \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
+&& \hotofo{\cst{f} \, \overline{s}_n } = \cst{f}(\hotofo{s_1}, \ldots, \hotofo{s_n}) &&
+\hotofo{\lambda \overline{x}. \, s  } = \hotofo{s}
+\end{align*}
+\looseness=-1
+We let $x_i$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
+and $\cstdba{i}$ be a fresh constant corresponding to this variable. All constants
+$\cstdba{i}$ must be fresh. 
+Effectively, $\hotofo{\;}$ transforms a $\eta$-long $\beta$-reduced higher-order term to an untyped first-order term. Let $\nf{t}{\beta\eta}$ be
+the $\eta$-long $\beta$-reduced form of $t$; the higher-order generic
+fingerprinting function $\text{gfpf}_\textsf{ho}$, which relies on conversion $\nametodb{t}$ from
+named to De Bruijn representation, is defined as
+\[
+  \text{gfpf}_\textsf{ho}(t,p) =
+    \text{gfpf}(\hotofo{\nametodb{\nf{t}{\beta\eta}}}, p)
+\]
+
+If we define $\text{fp}_\textsf{ho}(t) =
+\text{fp}(\hotofo{\nametodb{\nf{t}{\beta\eta}}})$, we can support fingerprint indexing
+for higher-order terms with no changes to the compatibility matrices. For example,
+consider the terms $s = (\lambda x y. \, x \, y) \, \textsf{g}$ and $t = \cst{f}$,
+where $\cst{g}$ has the type $\alpha \rightarrow \beta$ and $\cst{f}$ has the
+type $\alpha \rightarrow \alpha \rightarrow \beta$. 
+For the tuple of
+positions $(1, 1.1.1, 2)$ we get
+\begin{align*}
+  & \text{fp}_\textsf{ho}(s) =
+    \text{fp}(\hotofo{\nametodb{\nf{s}{\beta\eta}}}) =
+    \text{fp}(\cst{g}(\cstdba{0})) = 
+    (\cstdba{0}, \NN, \NN) \\
+  & \text{fp}_\textsf{ho}(t) =
+   \text{fp}(\hotofo{\nametodb{\nf{t}{\beta\eta}}}) =
+    \text{fp}(\cst{f}(\cstdba{1}, \cstdba{0})) = 
+    (\cstdba{1}, \NN, \cstdba{0})
+\end{align*}
+As the first and third fingerprint component are incompatible, the terms are not unifiable.
+
+\looseness=-1
+Other first-order indexing techniques such as feature vector indexing and substitution trees
+can probably be extended to higher-order terms using the method described here.
+
 \section{Implementation}
 \label{sec:unif:implementation}
+Zipperposition \cite{sc-15-simon-phd, sc-supind-17} is an
+open-source\footnote{\url{https://github.com/sneeuwballen/zipperposition}}
+theorem prover written in OCaml.
+It is a versatile testbed for prototyping extensions to
+superposition-based theorem provers.
+It was initially
+designed as a prover for polymorphic first-order logic
+and then extended to higher-order logic.
+A recent addition is a complete mode for
+Boolean-free higher-order logic \cite{bbtvw-21-sup-lam},
+which depends on a unification procedure that can enumerate a CSU. 
+We implemented our procedure in Zipperposition.
+
+\looseness=-1
+We used OCaml's functors to create a modular implementation. The core of our
+procedure is implemented in a module which is parametrized by another module
+providing oracles and implementing the \textsf{Bind} step. In this way we can
+obtain the complete or pragmatic procedure and seamlessly integrate
+oracles while reusing as much common code as possible.
+
+To enumerate all elements of a possibly infinite CSU, we rely on lazy lists whose
+elements are subsingletons of unifiers (either one-element sets containing a unifier
+or empty sets). The search space must be explored in a {\em fair} manner,
+meaning that no branch of the constructed tree is indefinitely postponed.
+
+Each {\sf Bind} step will give rise to new unification problems $E_1, E_2,
+\ldots$ to be solved. Solutions to each of those problems are lazy lists $p_1,
+p_2, \ldots$ containing subsingletons of unifiers. To avoid postponing some
+unifier indefinitely, we use the dovetailing technique: we first take one
+subsingleton from $p_1$, then one from each of $p_1$ and $p_2$. We continue with
+one subsingleton from each of $p_1,p_2$ and $p_3$, and so on. Empty lazy lists are
+ignored in the traversal. To ensure we do not remain stuck waiting for a unifier
+from a particular lazy list, the procedure will periodically return an empty
+set, indicating that the next lazy list should be probed.
+
+\looseness=-1
+The implemented selection function for our procedure prioritizes selection of
+rigid-rigid over flex-rigid pairs, and flex-rigid over flex-flex pairs. However,
+since the constructed substitution $\sigma$ is not applied eagerly, heads can
+appear to be flex, even if they become rigid after dereferencing and
+normalization. To mitigate this issue to some degree, we
+dereference the heads with $\sigma$, but do not normalize, and use the resulting
+heads for prioritization.
+
+We implemented oracles for the pattern, solid, and fixpoint fragment. 
+Fixpoint unification~\cite{gh-75-unification} is concerned with problems of the form
+$\{F \unif t\}$. If $F$ does not occur in $t$, $\{ F \mapsto t \}$
+is an MGU for the problem. If there is a position $p$ in $t$ such that 
+$t|_p = F \, \overline{u}_m$
+and for each prefix $q\neq p$ of $p$, $t|_q$ has a rigid head and either $m=0$ or $t$ is not a $\lambda$-abstraction,
+then we can conclude that $F \unif t$ has no solutions. Otherwise,
+the fixpoint oracle is not applicable.
+
+For second-order logic with only unary constants, it is decidable whether
+a unifier for a problem in this class (called
+\emph{monadic second-order}) exists \cite{wf-88-monadicunif}. As this class of terms admits a
+possibly infinite CSU, this oracle cannot be used for \textsf{OracleSucc} step,
+but it can be used for \textsf{OracleFail}. Similarly the fragment of
+second-order terms with no repeated occurrences of free variables has decidable
+unifier existence but possibly infinite CSUs
+\cite{gd-01-unif-chapter}. Due to their limited applicability and high complexity we
+decided not to implement these oracles.
 
 \section{Evaluation}
-\label{sec:unif:evaluation}
+\label{sec:unif:evaluation}We evaluated the implementation of our unification procedure in Zipperposition,
+%TODO: @PETAR: "the complete and the pragmatic variant"? 
+%Done.
+% Also, the complete variant is sometimes called "full". Maybe we should stick to one of the two?
+% @ALEX: Investingating.
+assessing the complete variant and the pragmatic variant, the latter with several
+different combinations of limits for number of bindings. As part of the
+implementation of the complete mode for Boolean-free higher-order logic in
+Zipperposition \cite{bbtvw-21-sup-lam}, Bentkamp implemented a straightforward
+version of the JP procedure. This version is faithful to the original description,
+with a check as to whether a (sub)problem can be solved using a first-order oracle
+as the only optimization. Our evaluations were performed on StarExec Miami
+\cite{sst-14-starexec} servers with Intel Xeon E5-{2620 v4} CPUs clocked at {2.10}\,GHz with 60\,s CPU limit.
+
+\looseness=-1
+Contrary to first-order unification, there is no widely available corpus of
+benchmarks designed solely for evaluating performance of higher-order
+unification algorithms. Thus, we used all 2606 monomorphic higher-order theorems
+from the TPTP library \cite{gs-17-tptp} and 832 monomorphic higher-order
+Sledgehammer (SH) generated problems \cite{ns-13-leo2sh} as our benchmarks\footnote{An archive with raw
+results, all used problems, and scripts for running each configuration is
+available at \url{http://doi.org/10.5281/zenodo.4269591}}. 
+Many TPTP problems require synthesis of complicated unifiers,
+whereas Sledgehammer problems are only mildly higher-order -- many of them are
+solved with first-order unifiers.
+
+\looseness=-1
+We used the naive implementation of the JP procedure (\textbf{jp}) as a baseline
+to evaluate the performance of our procedure. We compare it with the complete
+variant of our procedure (\textbf{cv}) and pragmatic variants
+(\textbf{pv}) with several different configurations of limits for
+applied bindings. All other Zipperposition parameters have been fixed to the values of
+a variant of a well-performing configuration we used for the CASC-27 theorem proving competition \cite{gs-19-casc27}.
+The cv configuration and all of the pv
+configurations use only pattern unification as an underlying oracle. To test
+the effect of oracle choice, we evaluated the complete variant
+in 8 combinations: with no oracles
+(\textbf{n}), with only fixpoint (\textbf{f}), pattern (\textbf{p}), or solid (\textbf{s}) oracle,
+and with their combinations: \textbf{fp}, \textbf{fs}, \textbf{ps}, \textbf{fps}.
+
+\looseness=-1
+Figure~\ref{fig:new-old} compares different variants of the procedure with the
+naive JP implementation. Each pv configuration is denoted by pv$^{a}_{bcde}$
+where $a$ is the limit on the total number of applied bindings, and $b$, $c$,
+$d$, and $e$ are the limits of functional projections, eliminations, imitations,
+and identifications, respectively. Figure~\ref{fig:oracles} summarizes the
+effects of using different oracles.
+
+The configuration of our procedure with no oracles outperforms the JP procedure with the first-order oracle. This
+suggests that the design of the procedure, in particular lazy normalization and
+lazy application of the substitution, already reduces the effects of the JP
+procedure's main bottlenecks. Raw evaluation data shows that on TPTP benchmarks, complete and pragmatic configurations
+differ in the set of problems they solve -- cv solves 19
+problems not solved by pv$^{4}_{2222}$, whereas pv$^{4}_{2222}$ solves 34
+problems cv does not solve. Similarly, comparing the pragmatic configurations with each other,
+pv$^{6}_{3333}$ and pv$^{4}_{2222}$ each solve 13 problems that the other one does not.  The overall higher success rate of pv$^2_{1020}$ compared to
+pv$^2_{1222}$ suggests that solving flex-flex pairs by trivial unifiers often
+suffices for superposition-based theorem proving.
+
+\newbox\mybox
+\setbox\mybox=\hbox{\small pv$^{12}_{6666}$}
+
+%%% TYPESETTING: hbox to
+\newcommand\HEAD[1]{\hbox to \wd\mybox{\hfill\hbox{#1}\hfill}}
+\newcommand\Z{\phantom{0}}
+\newcommand\MIDLINE{\\[.25ex]\hline\rule{0pt}{3ex}}
+
+\begin{figure}[t]
+	\def\arraystretch{1.1}%
+%\setlength\tabcolsep{.75ex}
+  \centerline{\begin{tabular}{@{}lcccccccc}
+    \strut                  &  \HEAD{jp}               & \HEAD{cv}              & \HEAD{pv$^{12}_{6666}$}    & \HEAD{pv$^{6}_{3333}$}     & \HEAD{pv$^{4}_{2222}$}    & \HEAD{pv$^{2}_{1222}$}   & \HEAD{pv$^{2}_{1121}$}         & \HEAD{pv$^{2}_{1020}$}
+    \MIDLINE
+    TPTP                    &  \eqmakebox[jp][r]{1551} &  \eqmakebox[nc][r]{1717} &  \eqmakebox[np6][r]{1722} &  \eqmakebox[np3][r]{\bf 1732} & \eqmakebox[np2][r]{\bf 1732} & \eqmakebox[np1][r]{1715} &  \eqmakebox[np10][r]{1712}   & \eqmakebox[np0][r]{1719}  \\
+    SH                      &  \eqmakebox[jp][r]{242}  &  \eqmakebox[nc][r]{\bf 260}  &  \eqmakebox[np6][r]{253}  &  \eqmakebox[np3][r]{255}  & \eqmakebox[np2][r]{255}  & \eqmakebox[np1][r]{254}  &  \eqmakebox[np10][r]{259}    & \eqmakebox[np0][r]{257} \\
+    \end{tabular}}
+	\caption{Proved problems, per configuration}
+	\label{fig:new-old}
+\end{figure}
+\begin{figure}[t]
+	\def\arraystretch{1.1}%
+%\setlength\tabcolsep{.75ex}
+  \centerline{\begin{tabular}{@{}lcccccccc}
+    \strut                  &  \HEAD{n}                     & \HEAD{f}                      & \HEAD{p}                       & \HEAD{s}                       & \HEAD{fp}                     & \HEAD{fs}                     & \HEAD{ps}                      & \HEAD{fps}
+    \MIDLINE
+    TPTP                    &  \eqmakebox[n][r]{1658} & \eqmakebox[s][r]{1717} &  \eqmakebox[p][r]{1717} & \eqmakebox[sp][r]{1720} & \eqmakebox[n][r]{1719} & \eqmakebox[s][r]{\bf 1724} &  \eqmakebox[p][r]{1720} & \eqmakebox[sp][r]{1723}  \\
+    SH                      &  \eqmakebox[n][r]{245}  & \eqmakebox[s][r]{255}  &  \eqmakebox[p][r]{\bf 260}  & \eqmakebox[sp][r]{259}  & \eqmakebox[n][r]{255}  & \eqmakebox[s][r]{254}  &  \eqmakebox[p][r]{258}  & \eqmakebox[sp][r]{254} \\
+    \end{tabular}}
+	\caption{Proved problems, per used oracle}
+	\label{fig:oracles}
+\end{figure}
+
+Counterintuitively, in some cases, using oracles can hurt the performance of
+Zipperposition. Using oracles typically results in generating smaller
+CSUs, whose elements are more general substitutions than the ones we obtain without oracles. 
+These more general substitutions usually
+contain more applied variables, which Zipperposition's heuristics avoid due to
+their explosive nature. This can make Zipperposition postpone necessary inferences for too long. 
+Configuration n benefits from this effect and therefore solves 18 TPTP problems that no other configuration in Figure
+\ref{fig:oracles} solves. The same effect also gives configurations with only one oracle 
+an advantage over configurations with multiple oracles on some problems.
+
+The evaluation sheds some light on how often solid unification problems appear in practice. 
+The raw data show that
+configuration s solves 5 TPTP problems that neither f nor p solve. Configuration
+f solves 8 TPTP problems that neither s nor p solve, while p solves 9 TPTP
+problems that two other configurations do not. This suggests that the solid oracle
+is slightly less beneficial than the fixpoint or pattern oracles, but still presents a
+useful addition the set of available oracles.
+
+% TODO: Maybe investigate at some point in future.
+% An interesting observation is that including the solid oracle has a different
+% effect on TPTP and Sledgehammer problems. However, solid fragment does not
+% subsume the first-order fragment -- first-order terms that share variables or
+% are not linear are not solid. This can also explain why solid unification helps
+% on TPTP problems that usually include hard hand-crafted higher-order problems,
+% but hurts on mostly first-order Sledgehammer problems.
+\looseness=-1 
+A subset of TPTP benchmarks, concerning operations on Church numerals, is designed to test the efficiency of
+higher-order unification. Our procedure performs exceptionally well
+on these problems -- it solves all of them, usually
+faster than other competitive higher-order provers.
+  There are 11 benchmarks in \texttt{NUM} category of TPTP that contain conjectures about
+  Church numerals: \texttt{NUM020\^{}1}, \texttt{NUM021\^{}1}, \texttt{NUM415\^{}1},
+  \texttt{NUM416\^{}1}, \texttt{NUM417\^{}1}, \texttt{NUM418\^{}1}, \texttt{NUM419\^{}1}, \texttt{NUM798\^{}1},
+  \texttt{NUM799\^{}1}, \texttt{NUM800\^{}1}, and \texttt{NUM801\^{}1}.
+  We evaluated those problems using the same CPU nodes and the same time limits 
+  as above. In addition to Zipperposition, we used
+  all higher-order provers that took part in the 2019 edition of CASC \cite{gs-19-casc27} (in the THF category) for this evaluation: CVC4
+  1.7 \cite{cbetal-11-cvc4}, Leo-III 1.4 \cite{sb-21-leo3}, Satallax 3.4
+  \cite{cb-12-satallax}, Vampire 4.4 THF \cite{lkav-13-vampire}. Figure~\ref{fig:numerals}
+  shows the CPU time needed to solve a problem or ``--'' if the prover timed out.
+
+  \begin{figure}[tb]
+    \def\arraystretch{1.1}%
+  %\setlength\tabcolsep{.75ex}
+    \centerline{\begin{tabular}{@{}l@{\hskip 1.25em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
+      \strut                  &  CVC4               & Leo-III                & Satallax             & Vampire               & Zipperposition (cv)
+      \MIDLINE
+      \texttt{NUM020\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{0.46} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{\bf 0.03} \\
+      \texttt{NUM021\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{\bf 4.10} \\    
+      \texttt{NUM415\^{}1}  &  \eqmakebox[n][r]{45.80} & \eqmakebox[n][r]{0.34} & \eqmakebox[n][r]{0.21} & \eqmakebox[n][r]{0.42} & \eqmakebox[n][r]{\bf 0.03} \\    
+      \texttt{NUM416\^{}1}  &  \eqmakebox[n][r]{47.37} & \eqmakebox[n][r]{0.92} & \eqmakebox[n][r]{0.21} & \eqmakebox[n][r]{0.41} & \eqmakebox[n][r]{\bf 0.07} \\    
+      \texttt{NUM417\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{49.73} & \eqmakebox[n][r]{\bf 0.30} & \eqmakebox[n][r]{0.40} & \eqmakebox[n][r]{0.45} \\    
+      \texttt{NUM418\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{0.40} & \eqmakebox[n][r]{1.29} & \eqmakebox[n][r]{0.38} & \eqmakebox[n][r]{\bf 0.03} \\    
+      \texttt{NUM419\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{0.42} & \eqmakebox[n][r]{23.33} & \eqmakebox[n][r]{0.37} & \eqmakebox[n][r]{\bf 0.03} \\    
+      \texttt{NUM798\^{}1}  &  \eqmakebox[n][r]{46.29} & \eqmakebox[n][r]{0.35} & \eqmakebox[n][r]{4.01} & \eqmakebox[n][r]{0.38} & \eqmakebox[n][r]{\bf 0.03} \\    
+      \texttt{NUM799\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{5.05} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{\bf 0.03} \\    
+      \texttt{NUM800\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{\bf 0.37} & \eqmakebox[n][r]{3.15} \\
+      \texttt{NUM801\^{}1}  &  \eqmakebox[n][r]{--} & \eqmakebox[n][r]{0.73} & \eqmakebox[n][r]{38.77} & \eqmakebox[n][r]{--} & \eqmakebox[n][r]{\bf 0.50} \\
+      \end{tabular}}
+    \caption{Time needed to prove a problem, in seconds.}
+    \label{fig:numerals}
+  \end{figure}
 
 \section{Discussion and Related Work}
+\looseness=-1
+The problem addressed in this chapter is that of finding
+a complete and efficient higher-order unification procedure. Three main lines of
+research dominated the research field of higher-order unification over the last forty years.
+
+\looseness=-1
+The first line of research went in the direction of finding procedures that
+enumerate CSUs. The most prominent procedure designed for
+this purpose is the JP procedure \cite{jp-76-unif}. Snyder and Gallier
+\cite{sg-89-unif} also provide such a procedure,
+but instead of solving
+flex-flex pairs systematically, their procedure blindly guesses the head of the necessary
+binding by considering all constants in the signature and fresh variables of all
+possible types. Another approach, based on higher-order combinators, is given by
+Dougherty \cite{dd-93-comb-unif}. This approach blindly creates (partially
+applied) \textsf{S}-, \textsf{K}-, and \textsf{I}-combinator bindings for applied
+variables, which results in returning many redundant unifiers, as well as in
+nonterminating behavior even for simple problems such as $X \, \cst{a} =
+\cst{a}$.
+
+The second line of research is concerned with enumerating
+preunifiers. The most prominent procedure in this line of research is Huet's 
+\cite{gh-75-unification}. The Snyder--Gallier procedure restricted not to solve flex-flex pairs
+is a version of the PT procedure presented in Section \ref{sec:unif:solid-oracle}. It improves 
+Huet's procedure by featuring a Solution rule.
+
+The third line of research gives up the expressiveness of the full
+$\lambda$-calculus and focuses on decidable fragments. Patterns
+\cite{tn-93-patterns} are arguably the most important such fragment in practice,
+with implementations
+in Isabelle \cite{tn-02-isabelle}, Leo-III \cite{sb-21-leo3}, Satallax
+\cite{cb-12-satallax}, $\lambda$Prolog \cite{dm-12-lprolog}, and other systems.
+Functions-as-constructors \cite{tl-16-facunif} unification subsumes pattern
+unification but is significantly more complex to implement. Prehofer
+\cite{cp-95-unifphd} lists many other decidable fragments, not only for
+unification but also preunification and unifier existence problems. Most of
+these algorithms are given for second-order terms with various constraints on
+their variables. Finally, one of the first decidability results is Farmer's discovery \cite{wf-88-monadicunif} that
+higher-order unification of terms with unary function symbols is decidable.
+
+Our procedure draws inspiration from and contributes to all three lines of research.
+Accordingly, its advantages over previously known procedures can be laid out
+along those three lines. First, our procedure mitigates many issues of the 
+JP procedure. Second, it can be modified not to solve flex-flex pairs,
+and become a version of Huet's procedure with important built-in optimizations.
+Third, it can integrate any oracle for problems with finite CSUs, including the one we discovered.
+
+% The disadvantages of our algorithm include the need to slightly redesign oracles
+% to perform lazy normalization and substitution application if one wants to use
+% them efficiently. Furthermore, properties like the graceful extension of first-order fragment fully
+% depend on oracles.
+
+The implementation of our procedure in Zipperposition was one of the reasons
+this prover evolved from proof-of-concept prover for higher-order logic to
+competitive higher-order prover. Shortly after the procedure was implemented, in the 2020 edition of CASC, Zipperposition won
+the higher-order division.
 
 \section{Conclusion}
+We presented a procedure for enumerating a complete set of higher-order unifiers
+that is designed for efficiency. Due to a design that restricts 
+the search space and a tight integration of oracles,
+ it reduces the number of redundant unifiers returned and
+gives up early in cases of nonunifiability. In addition, we presented a new
+fragment of higher-order terms that admits finite CSUs. Our evaluation shows
+a clear improvement over previously known procedures.
+
+The procedure was a very prolific playground for tuning various heuristics
+to improve the success rate. In Chapter \ref{ch:ho-techniques} we discuss
+some of the Zipperposition parameters we tuned to make better use of the procedure.
+It also makes the basis of the unification procedure implemented in $\lambda$E,
+the full higher-order extension of Ehoh, described in Chapter \ref{ch:ehoh2}.

### 2022-03-02 unif/unif.tex
fixes

@@
 \end{description}
-
+%
 The transitions are designed so that only \unifrulename{OracleSucc},
@@
 
-\ourpara{unif:examples}
+\ourpara{Examples}
 We present some examples that demonstrate advantages of our procedure.
@@
   \label{unif:lemma:solid-vars-diff-mgu}
+  \looseness=-1
   Let 
@@
 \cite{ls-16-indexing,br-19-restricted-unif}.
-A \emph{fingerprint index} \cite{ss-12-fp-indexing,vbss-19-ehoh1} is an imperfect index based on
+A \emph{fingerprint index} \cite{ss-12-fp-indexing} is an imperfect index based on
 the idea that the skeleton of the term consisting of all non-variable positions
@@
 determine the compatibility for retrieval operations:
+\pagebreak[1]
 \begin{align*}
@@
 can probably be extended to higher-order terms using the method described here.
+\pagebreak[2]
 

### 2022-03-03 ehoh/ehoh.tex
background
jasmin unif

@@
 
-\blfootnote{In this work I desgined, implemented and evaluated all changes to term representation, algorithms and indexing data structures.}
+\blfootnote{In this work I designed, implemented and evaluated all changes to
+term representation, algorithms and indexing data structures. Jasmin Blanchette
+came up with the extension of fingerprinting indexing. Stephan Schulz provided
+the necessary E expertise.}
 
@@
 that supports partial application and applied variables. Our extension
-outperforms the traditional encoding and appears promising as a stepping stone
+outperforms the traditional encoding and forms a stepping stone
 toward full higher-order logic.    
@@
 higher-order problems that do not require synthesizing
-$\lambda$-terms. As next steps, we plan to add support for
+$\lambda$-terms. In Chapter~\ref{ch:ehoh2} we show how to add support for
 $\lambda$-terms and higher-order unification.
@@
 \looseness=-1
-Our logic is a variant of the intensional $\lambda$-free Boolean-free higher-order
-logic (\lfhol{}) described by Bentkamp et al.\ %, Blanchette, Cruanes, and Waldmann
+Our logic is a variant of the intensional $\lambda$-free Boolean-free
+higher-order logic (\lfhol{}) described by Bentkamp et al.\
+%, Blanchette, Cruanes, and Waldmann
 \cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
@@
 %% kotelnikov-15-fool
-extend the syntax of this logic by erasing the distinction between terms and
-formulas, and its semantics by interpreting the Boolean type $o$ as a domain
-of cardinality~2. Functional extensionality can be obtained by adding
-suitable axioms \cite[Sect.~3.1]{bbcw-21-lfho}.
+extend the syntax of this by making formulas a special case of terms,
+and its semantics by interpreting the Boolean type $o$ as a domain of
+cardinality~2. Functional extensionality can be obtained by adding suitable
+axioms \cite[Sect.~3.1]{bbcw-21-lfho}.
 
@@
 This logic differs from the higher-order logic described in
-Sect.~\ref{sec:pre:hol} three ways. First, $\lambda$-abstraction is disallowed.
+Sect.~\ref{sec:pre:hol} in three ways. First, $\lambda$-abstraction is disallowed.
 Second, logical connectives are not part of the set of symbols. Instead, there
 is a special inductive case in the definition of terms that defines formulas.
-Third, subterms are defined in a more traditional way.
+Third, subterms are defined in a more traditional way, as defined below.
 
 \looseness=-1
-For completeness, we provide the definition of terms. Terms, ranged over by $s,t,u,v$, are either
+For reference, we provide the definition of terms. Terms, ranged over by $s,t,u,v$, are either
 \emph{variables} $x, y, z, \dots$, (\emph{function}) \emph{symbols} $\cst{a},
@@
 we denote the set of free variables of $x$ where $x$ ranges over terms, clauses,
-set of clauses or any other objects that contain terms.
+set of clauses, or any other objects that contain terms.
 
@@
 Each subgroup consists of two sub-sub\-groups of 1253
-problems, generated by using either $\lambda$-lifting or SK-style combinators
+problems, generated by using either $\lambda$-lifting or SKBCI-style combinators
 to encode $\lambda$-expressions.
@@
 provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
-they may require tedious reasoning about $\lambda$-lifted functions or SK-style
+they may require tedious reasoning about $\lambda$-lifted functions or SKBCI-style
 combinators}.
@@
 Fifth, Ehoh cannot compete against the best higher-order systems, but this
-  is no surprise given that it does not yet support $\lambda$-expressions and
-  higher-order unification.
+  is no surprise given that it does not support $\lambda$-expressions and
+  higher-order unification. An extension of Ehoh that supports these features
+  is described in Chapter~\ref{ch:ehoh2}.
 
@@
   on the practical side, we are implementing such extensions in a
-  state-of-the-art prover.
+  state-of-the-art prover. In this thesis, the focus is on the second aspect.
   %TODO_PHD: another reference
@@
   we found the generalization manageable and are now in a position to add
-  support for $\lambda$-terms and higher-order unification.
+  support for $\lambda$-terms and higher-order unification, as discussed in Chapter~\ref{ch:ehoh2}.
   \end{rep}
@@
   architecture.
-  Finally, there is recent work by the developers of Vampire
+  Finally, there is work by the developers of Vampire
   % br-20-full-sup-w-combs
@@
 
-Despite considerable progress since the 1970s, higher-order automated
-reasoning has not yet assimilated some of the most successful methods for
+Despite considerable progress since the 1970s, until a few years ago higher-order automated
+reasoning has not assimilated some of the most successful methods for
 first-order logic with equality, such as superposition. We presented a

### 2022-03-03 unif/unif.tex
jasmin unif

@@
 \authors{Joint work with\\
-Alexander Betnkamp and Visa Nummelin}
+Alexander Bentkamp and Visa Nummelin}
 
@@
 tool in theorem proving, logic
-programming, natural language processing, programming language compilation and other areas of computer science.
+programming, natural language processing, programming language compilation, and other areas of computer science.
 
@@
 First, the JP procedure does not terminate in many cases of obvious
-nonunifiability, e.g., for $X \unif \cst{f} \, X$, where $X$ is a non-functional
+nonunifiability, e.g., for $X \unif \cst{f} \, X$, where $X$ is a nonfunctional
 variable and $\cst{f}$ is a function constant. This example also shows that
@@
 JP's definition does not employ the notion of auxiliary variables,
-but instead requires $\rho X = \theta\sigma X$ for all variables mapped by $\rho$.
-This is problematic because nothing prevents $\rho$ from mapping the auxiliary variables.
+but instead requires $\varrho X = \theta\sigma X$ for all variables mapped by $\varrho$.
+This is problematic because nothing prevents $\varrho$ from mapping the auxiliary variables.
 For example, $\sigma = \{F \mapsto \lambda x y.\> G\>y\}$ is supposed
 to be an MGU for $F\>\cst{a}\>\cst{c} \unif F\>\cst{b}\>\cst{c}$.
-But for the unifier $\rho = \{F \mapsto \lambda x y.\> y,\> G \mapsto \lambda x.\> \cst{d}\}$,
+But for the unifier $\varrho = \{F \mapsto \lambda x y.\> y,\> G \mapsto \lambda x.\> \cst{d}\}$,
 without the notion of auxiliary variables,
 there exists no appropriate substitution $\theta$
-because $\rho G = \theta\sigma G$ requires $\theta G = \lambda x.\> \cst{d}$
-and $\rho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$.
+because $\varrho G = \theta\sigma G$ requires $\theta G = \lambda x.\> \cst{d}$
+and $\varrho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$. By declaring $G$ as an auxiliary variable,
+we focus only the important variables from the intitial problem (in this case only $F$),
+which allows $\sigma$ to be an MGU.
 
@@
 $E$ of unification constraints to one of those constraints $S(E) \in E$, the
-\emph{selected} constraint in $E$. The transitions, defined as follows, are only
+\emph{selected} constraint in $E$. In this chapter, we consider \emph{freshness}
+of the variables with respect to the free variables ocurring in $E$. The transitions, defined as follows, are only
 applied if the $\selected{\text{grayed}}$ constraint is selected. 
@@
       \newunifarrow 
-      (\{\lambda \overline{x}_m. \, s \unif \lambda \overline{x}_m. \, t'\, x_{n+1} \ldots x_m\}\uplus E, \sigma)$\\
+      (\{\lambda \overline{z}_m. \, s' \unif \lambda \overline{z}_m. \, t'\, z_{n+1} \ldots z_m\}\uplus E, \sigma)$\\
       % where $n \geq m$ and $t'=\{y_1\mapsto x_1,\ldots,y_m\mapsto x_m\}t$
-      where $m \geq n$, $\overline{x}_m \not= \overline{y}_n$, and $t'=\{y_1\mapsto x_1,\ldots,y_n\mapsto x_n\}t$
+      where $m \geq n$, $\overline{x}_m \not= \overline{y}_n$, $s$ and $t$ are not $\lambda$-abstractions,
+      $\overline{z}_m$ are bound variables fresh with respect to $s$ and $t$, $s' = \{x_1 \mapsto z_1, \ldots, x_m \mapsto z_m \}$,
+      and $t' = \{y_1 \mapsto z_1, \ldots, y_n \mapsto z_n \}$
   % \smallskip
@@
       (\{\lambda \overline{x}. \, (\sigma F) \,\overline{s} \unif \lambda \overline{x}. \, t\}\uplus E, \sigma)$\\
-      where none of the previous transitions apply and $F$ is mapped by $\sigma$
+      where none of the previous transitions applies and $F$ is mapped by $\sigma$
   % \smallskip
@@
       \uplus E, \sigma) \newunifarrow \bot$\\
-      where none of the previous transitions apply, and
+      where none of the previous transitions applies, and
       $a$ and $b$ are different rigid heads
@@
   \item[\rm\unifrulename{Delete}]  $(\{\selected{s \unif s}\}\uplus E, \sigma) \newunifarrow (E, \sigma)$\\
-      where none of the previous transitions apply
+      where none of the previous transitions applies
   % \smallskip 
@@
   $(\{\selected{s \unif t}\}\uplus E, \sigma) \newunifarrow (E, \varrho \sigma)$\\
-      where none of the previous transitions apply, 
+      where none of the previous transitions applies, 
       some oracle found a finite CSU $U$ for $\sigma s \unif \sigma t $ using fresh auxiliary variables,
@@
   $(\{\selected{s \unif t}\}\uplus E, \sigma) \newunifarrow \bot$\\
-  where none of the previous transitions apply, 
+  where none of the previous transitions applies, 
   and some oracle determined $\sigma s \unif \sigma t$ has no solutions
@@
       \sigma)$\\
-      where none of the transitions \unifrulename{Succeed} to \unifrulename{OracleFail} apply
+      where none of the transitions \unifrulename{Succeed} to \unifrulename{OracleFail} applies
   % \smallskip
@@
   %     (\{s_{j_{1}}\unif t_{j_{1}},\ldots,s_{j_{k}}\unif t_{j_{k}}\} \uplus E,\varrho \sigma)$\\
-  %     where none of the transitions \textsf{Succeed} to \textsf{OracleFail} apply,
+  %     where none of the transitions \textsf{Succeed} to \textsf{OracleFail} applies,
   %     $ 1 \leq j_1 < \cdots < j_k \leq n$, and
@@
       (\{s \unif t\}
-      \uplus E, \rho \sigma)$\\
-      where none of the transitions \textsf{Succeed} to \textsf{OracleFail} apply,
-      and $\rho\in\mathcal{P}(s \unif t)$.
+      \uplus E, \varrho \sigma)$\\
+      where none of the transitions \textsf{Succeed} to \textsf{OracleFail} applies,
+      and $\varrho\in\mathcal{P}(s \unif t)$.
 \end{description}
@@
 to certain classes of terms. On the other hand, \unifrulename{Bind} is applied to
-explore the whole search space for any given constraint. These rules are separated
-in two to make \unifrulename{Bind} applicable only if \unifrulename{OracleSucc}
-(or \unifrulename{OracleFail}) is not, so that possible solutions (or failures) are
-detected early.
+explore the whole search space for any given constraint. To compute CSU using efficient oracles and to discover
+failures early, \unifrulename{Bind} is applicable only if \unifrulename{OracleSucc}
+(or \unifrulename{OracleFail}) is not.
 
@@
 discover nonunifiability earlier.
-In some cases, addition of oracles lets the procedure terminate more often.
+In some cases, the addition of oracles lets the procedure terminate more often.
 
@@
 The following mapping $\mathcal{P}_\mathsf{c}(\lambda \overline{x}.
-\, s \unif \lambda \overline{x}. \, t)$ is used as the parameter $\mathcal{P}$ of the procedure:
+\, s \unif \lambda \overline{x}. \, t)$ is used as the parameter $\mathcal{P}$ of the procedure.
+It provides all the bindings that the procedures needs to be complete, and it is defined as follows:
 \begin{itemize}
@@
 it determines a set of bindings $\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$.
-For each of the bindings $\rho\in\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$, it creates a child node
+For each of the bindings $\varrho\in\mathcal{P}_\mathsf{JP}(C,s_i, t_i)$, it creates a child node
 $(\nf{(\varrho{s})}{\beta\eta} \unif \nf{(\varrho{t})}{\beta\eta}, \varrho\sigma)$,
@@
 lead to many different paths to the same unifier. In contrast, our procedure
-makes the decision which binding to apply to a flex-flex pair with the same
+decides which binding to apply to a flex-flex pair with the same
 head as soon as it is observed. Also, it explores the possibility of not
-applying a binding and decomposing the pair. In either way, the flex-flex pair
+applying a binding and decomposing the pair. In either case, the flex-flex pair
 is never revisited, which improves the performance and returns fewer redundant
 unifiers. We show that this restriction prunes the search space without
-influencing the completeness.
-
-Our procedure makes the choice of child nodes based only on the heads of the
-chosen unification constraint. In contrast, the JP procedure tracks all the
+losing completeness.
+
+Our procedure chooses child nodes based only on the heads of the
+selected unification constraint. In contrast, the JP procedure tracks all the
 variables occurring in the common context. Thus, lazy normalization and lazy
@@
 version of Huet-style projection. Its universality enables finding elements of
-CSU for flex-flex pairs, for which Huet-style projection does not suffice.
+the CSU for flex-flex pairs, for which Huet-style projection does not suffice.
 However, the JP procedure applies iteration indiscriminately on both flex-flex and
@@
 \looseness=-1
-The bindings of our procedure contain further optimizations that are absent in
+The bindings of our procedure contain further optimizations that are absent from
 the JP procedure: The JP procedure applies eliminations for one parameter
@@
 and
-\textsf{Bind} of a binding $x$ as $\textsf{B(}x\textsf{)}$.
+\textsf{Bind} of a binding $b$ as $\textsf{B(}b\textsf{)}$.
 Transitions of the JP procedure are denoted by $\Longrightarrow$.
@@
 %is to limit the number or kind of bindings applied to a single constraint. 
-Our preliminary experiments showed that using mapping $\mathcal{P}_\mathsf{p}$ defined as follows is a reasonable compromise between completeness and
-performance:
+Our preliminary experiments showed that using the mapping $\mathcal{P}_\mathsf{p}$ defined as follows is a reasonable compromise between performance and
+completeness:
 \begin{itemize}
@@
 The pragmatic variant of our procedure removes all iteration bindings to enforce finite branching. 
-Moreover, it
+In addition, it
 imposes limits 
@@
 counting the applications of bindings locally, per
-constraint. 
+constraint, as follows. 
 It is useful to distinguish the Huet-style projection cases where
@@
 projection}). We limit the number applications of the following bindings:
-functional projections, eliminations, imitations and identifications.
+functional projections, eliminations, imitations, and identifications.
 In addition,
@@
 elimination steps.
-Due to these limits, the pragmatic variant 
+Thanks to these limits, the pragmatic variant 
 terminates.
@@
 Careful tuning of each limit optimizes the procedure for a specific class of problems. 
-For problems originating from proof assistants, shallow unification depth usually suffices. However, hard
+For problems originating from proof assistants, low unification depth usually suffices. However, hard
 hand-crafted problems often need
@@
 
-Like the JP procedure, our procedure misses no unifiers:
+Like the JP procedure, our procedure, parametrized with $\mathcal{P}_\mathsf{c}$,  misses no unifiers:
 
@@
   For a free variable $X$, a unifier of $X\,\seq s \unif X\,\seq t$ does not necessarily
-  unify $s_{i} \unif t_{i}$.
+  unify $s_{i} \unif t_{i}$. For example, $\{X \mapsto \cst{a}\}$ is a unifier that does
+  refer to any of the arguments of $X$.
   The concept of $\omega$-simplicity is
@@
     variable. If $\varrho\left(\param X\,\seq s\right)=\varrho\left(\param X\,\seq t\right)$
-    and some occurrence of the $i^{\text{th}}$ parameter of $\varrho X$
+    and some occurrence of the $i$th parameter of $\varrho X$
     is base-simple, then $\varrho s_{i}=\varrho t_{i}$.
@@
     remaining problem size. Showing that the measure is strictly decreasing
-    and well-founded guarantees that the procedure finds 
+    and well founded guarantees that the procedure finds 
     a suitable substitution in finitely many steps. We reuse the measure for remainder substitutions
@@
     For multisets $E$ of unification constraints and
-    substitutions $\rho$, our measure on pairs $(E, \rho)$
+    substitutions $\varrho$, our measure on pairs $(E, \varrho)$
     is the lexicographic comparison of
     \begin{description}
-      \item[A] the sum of the sizes of the terms in $\rho E$
-      \item[B] the sum over the free weight of $\rho F$, 
-        for all variables $F$ mapped by $\rho$
-      \item[C] the sum over the bound weight of $\rho F$, 
-      for all variables $F$ mapped by $\rho$
-      \item[D] the sum over the number of parameters of $\rho F$, 
-      for all variables $F$ mapped by $\rho$
+      \item[A] the sum of the sizes of the terms in $\varrho E$
+      \item[B] the sum over the free weight of $\varrho F$, 
+        for all variables $F$ mapped by $\varrho$
+      \item[C] the sum over the bound weight of $\varrho F$, 
+      for all variables $F$ mapped by $\varrho$
+      \item[D] the sum over the number of parameters of $\varrho F$, 
+      for all variables $F$ mapped by $\varrho$
     \end{description}
-    We denote the quadruple containing these numbers as $\ord(E,\rho)$.
-    We denote the triple containing only the last three components of $\ord(E,\rho)$ as $\ord \rho$.
-    %ASK(VISA): Why is ord(E,\rho) with parentheses and ord \rho without them?
+    We denote the quadruple containing these numbers as $\ord(E,\varrho)$.
+    We denote the triple containing only the last three components of $\ord(E,\varrho)$ as $\ord \varrho$.
+    %ASK(VISA): Why is ord(E,\varrho) with parentheses and ord \varrho without them?
     We write $<$ for the lexicographic comparison of these tuples.
@@
     there exists a substitution $\varrho'$ such that $\ord\varrho'<\ord\varrho$
-    and for all variables $X$ except the fresh variables introduced by the binding $\varrho X= \varrho'\, \delta X$.
+    and for all variables $X$ except the fresh variables introduced by the binding we have  $\varrho X= \varrho'\, \delta X$.
     For most of these lemmas, we refer to JP \cite{jp-76-unif} for proofs.
@@
     \begin{proof}
-    Let $\varrho a=\param t$ and let $\left(x_{j_{k}}\right)_{k=1}^{i}$
+    Let $\varrho a=\param t$ and let $(x_{j_{k}})_{k=1}^{i}$
     be the subsequence of $\seq x$ consisting of those variables which
@@
     If $\varrho a$ is also simply projective,
-    then $x_k$ must be non-functional
+    then $x_k$ must be nonfunctional
     and since Huet-style projection and JP-style projection coincide
@@
     We must show that there exists
-    a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
+    a derivation $(E,\makeop{id})\longrightarrow^{*}\sigma$ and a
     substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
@@
     To guarantee well-foundedness, we ensure that the measure 
-    $\ord\left(\varrho_{j},E_{j}\right)$ decreases with each step.
+    $\ord(\varrho_{j},E_{j})$ decreases with each step.
     We maintain the following invariants for all $j$:
@@
     \item $(E_{j},\sigma_{j})\longrightarrow(E_{j+1},\sigma_{j+1})$;
-    \item $\ord(\varrho_{j},E_{j}) > \ord(\varrho_{j+1},E_{j+1})$;
+    \item $\ord(E_{j},\varrho_{j}) > \ord(E_{j+1}, \varrho_{j+1})$;
     \item $\varrho_{j}$ unifies $E_{j}$;
@@
     \item every free variable occurring in $\varrho_{j}E_{j}$ does not occur in $E_{j}$ and is not contained in $V$;
-    \item for every identification variable $X$, $\rho_j X$ is not projective; and
-    \item for every elimination variable $X$, each parameter of $\rho_j X$ 
-          has occurrences in $\rho_j X$,
+    \item for every identification variable $X$, $\varrho_j X$ is not projective; and
+    \item for every elimination variable $X$, each parameter of $\varrho_j X$ 
+          has occurrences in $\varrho_j X$,
           all of which are base-simple.
@@
     Typically, one of the Lemmas \ref{lem:elimination}\textendash{}\ref{lem:huet-projection}
-    is going to apply. Any one of them gives substitutions $\varrho'$
+    will be applicable. Any one of them gives substitutions $\varrho'$
     and $\delta$ with properties that let us define $E_{j+1}=\delta E_{j}$,
@@
     all parameters of $\varrho_{j}a$ have base-simple occurrences, $\varrho_{j}u_{i}$
-    is a subterm of $\varrho_{j}u=\param\varrho_{j}a\,\left(\varrho_{j}\seq u\right)$
+    is a subterm of $\varrho_{j}u=\param\varrho_{j}a\,(\varrho_{j}\seq u)$
     by Lemma \ref{lem:base-simple->inj.}. Similarly for $\varrho_{j}v$.
@@
     $(E,\makeop{id})\longrightarrow^*(\varnothing,\sigma_{j})\longrightarrow\sigma_j$,
-    and $\rho X = \tau^{-1}\,\varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$,
+    and $\varrho X = \tau^{-1}\,\varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$,
     completing the proof. \qedhere
@@
 
-The following rules, however, are stated modulo orientation. The PT transition
-rules, adapted for our presentation style, are as follows:
+The following rules, however, are stated modulo orientation. This means that 
+if a rule applies to a constraint $s \unif t$, then it also applies to a constraint $t
+\unif s$. The PT transition rules, adapted for our presentation style, are as
+follows:
 
@@
     \item[A] sum of the abstraction-free sizes of the terms $\varrho F$ for each variable $F$ mapped by $\varrho$
-    %TODO: What are the "bindings in \rho"? Maybe write "the terms $\varrho x$ for each variable $x$ mapped by $\rho$"? --Alex
+    %TODO: What are the "bindings in \varrho"? Maybe write "the terms $\varrho x$ for each variable $x$ mapped by $\varrho$"? --Alex
     %TODO: Done
@@
   one of $s$ and $t$ must have a rigid head. We distinguish several cases based on the
-  form of $s \unif t$ (modulo the constraint order):
+  form of $s \unif t$ (modulo constraint orientation):
 
@@
 and $t$ is second-order. Together with completeness of PT for preunification,
-this result implies that PT procedure can be used to enumerate finitely many
+this result implies that the PT procedure can be used to enumerate finitely many
 elements of complete set of preunifiers for this class of terms.
@@
 for the full unification problem and we lift the
-order constraints.
+order restrictions.
 %
@@
 
-Towards proving that the PT
+Toward proving that the PT
 procedure terminates on above described class of problems, we first consider the corresponding matching problem.
@@
     \label{unif:lemma:pt-matching-termination}
-    PT terminates on a solid matching problem $\{s \unif t\}$.
+    PT terminates on every solid matching problem $\{s \unif t\}$.
 \end{lemma}
@@
   \looseness=-1
-  If $s$ and $t$ are solid, $s$ is linear and shares no free variables with $t$, then
+  If $s$ and $t$ are solid, and $s$ is linear and shares no free variables with $t$, then
   PT terminates for the preunification problem $\{ s \unif t \}$, and
@@
         \overline{x}. \, a \, \overline{u}_m$: if $a$ is a bound variable,
-        then for \textsf{Failure} not to be applicable afterwards, we have to
+        then for \textsf{Failure} not to be applicable afterward, we have to
         project $F$ onto argument $v_i$ equal to $a$. Then we proceed like for
@@
         project to some base-type term $v_j$ (otherwise \textsf{Failure} would
-        be applicable afterwards). This reduces the problem to $\lambda
+        be applicable afterward). This reduces the problem to $\lambda
         \overline{x}. \, v_j \unif \lambda \overline{x}. \, a \,
@@
     if they correspond to syntactically equal terms. Therefore, 
-    there is a substitution $\theta$ such that $\rho X = \theta\sigma X$ for all $X \not = G$. 
+    there is a substitution $\theta$ such that $\varrho X = \theta\sigma X$ for all $X \not = G$. 
     Due to arbitrary choice of $\varrho$, we conclude that $\sigma$ is an MGU.
@@
   \overline{s}'_{m'}$. We prove $\sigma$ is an MGU by showing that there exists a substitution
-  $\theta$ such that $\varrho X = \theta\sigma X$ for all non-auxiliary variables $X$. 
+  $\theta$ such that $\varrho X = \theta\sigma X$ for all nonauxiliary variables $X$. 
   We can focus only on $X \in \{F, F'\}$
-  because all other non-auxiliary variables
+  because all other nonauxiliary variables
   do neither appear in the original problem nor in $\sigma F$ or $\sigma F'$; so we can simply define $\theta X = \varrho X$.
@@
   \textsf{diff} is defined and has the desired property.
-  In equations \ref{diff:xi}, \ref{diff:yi}, and \ref{diff:xiyj}, if there are multiple numbers $k$ or $l$
+  In equations (\ref{diff:xi}), (\ref{diff:yi}), and (\ref{diff:xiyj}), if there are multiple numbers $k$ or $l$
   that fulfill the condition, choose an arbitrary one.
@@
     \item[$v = a \, \overline{v}_n$] In the remaining cases $a$ is either a free variable, a loose bound variable
-    different than $x_1, \ldots, x_m$, or a constant. If $a$ is a free variable or a loose bound variable
-    different than $x_1, \ldots, x_m$, then $v' = a \, \overline{v'}_n$, since
-    ($\star$) holds, all of $\overline{s}'_{m'}$ are ground, and bound variables different than
+    different from $x_1, \ldots, x_m$, or a constant. If $a$ is a free variable or a loose bound variable
+    different from $x_1, \ldots, x_m$, then $v' = a \, \overline{v'}_n$, since
+    ($\star$) holds, all of $\overline{s}'_{m'}$ are ground, and bound variables different from
     $x_1, \ldots, x_m$ and $y_1, \ldots, y_{m'}$ are renamed to match by equation \ref{diff:alpha}. 
@@
   binding $\sigma_1 = \{ F \mapsto \lambda x.\, \cst{g} \, (F_1 \, x) \, (F_2 \, x)
-  \}$. This yields the  problem $\{F_1
+  \}$. After decomposition, this yields the  problem $\{F_1
   \, (\cst{f} \, \cst{a}) \unif \cst{a}, \allowbreak F_2 \, (\cst{f} \, \cst{a}) \unif G \,
@@
 \, \cst{a}\}$ has only infinite CSUs. Finally, it is crucial that
-functional arguments to free variables are only bound variables: the problem
+functional arguments to free variables are only bound variables: The problem
 $\{ \lambda y. \, X (\lambda x.\, x) \, y \unif \lambda y. \, y \}$ has only infinite CSUs.
@@
 \cite{ls-16-indexing,br-19-restricted-unif}.
-A \emph{fingerprint index} \cite{ss-12-fp-indexing} is an imperfect index based on
-the idea that the skeleton of the term consisting of all non-variable positions
+Recall from Chapter \ref{ch:ehoh} that a \emph{fingerprint index} \cite{ss-12-fp-indexing} is an imperfect index based on
+the idea that the skeleton of the term consisting of all nonvariable positions
 is not affected by substitutions. Therefore, we can easily determine that the terms
@@
     $\cst{f}_1\vphantom{^{()}}$ & & \xmark & & & \xmark \\
+    $\cst{f}_2\vphantom{^{()}}$ & \xmark &  & & & \xmark \\
     \AA & & & & & \xmark \\
@@
     $\cst{f}_1\vphantom{^{()}}$ & & \xmark & \xmark & \xmark & \xmark \\
+    $\cst{f}_2\vphantom{^{()}}$ & \xmark &  & \xmark & \xmark & \xmark \\
     \AA & & & & \xmark & \xmark \\
@@
 % Yes
-$\beta$-reduced $\eta$-long terms in De Bruijn \cite{db-75-dbindices} notation:
+$\eta$-long $\beta$-reduced terms in De Bruijn \cite{db-75-dbindices} notation:
 %TODO: @PETAR: also mention that this is an encoding into FOL before the definition?
@@
 subsingleton from $p_1$, then one from each of $p_1$ and $p_2$. We continue with
-one subsingleton from each of $p_1,p_2$ and $p_3$, and so on. Empty lazy lists are
+one subsingleton from each of $p_1,p_2$, and $p_3$, and so on. Empty lazy lists are
 ignored in the traversal. To ensure we do not remain stuck waiting for a unifier
@@
 appear to be flex, even if they become rigid after dereferencing and
-normalization. To mitigate this issue to some degree, we
+normalization. To mitigate this issue, we
 dereference the heads with $\sigma$, but do not normalize, and use the resulting
@@
 from the TPTP library \cite{gs-17-tptp} and 832 monomorphic higher-order
-Sledgehammer (SH) generated problems \cite{ns-13-leo2sh} as our benchmarks\footnote{An archive with raw
+Sledgehammer (SH) generated problems \cite{ns-13-leo2sh} as our benchmarks.\footnote{An archive with raw
 results, all used problems, and scripts for running each configuration is
-available at \url{http://doi.org/10.5281/zenodo.4269591}}. 
+available at \url{http://doi.org/10.5281/zenodo.4269591}}
 Many TPTP problems require synthesis of complicated unifiers,
@@
 the effect of oracle choice, we evaluated the complete variant
-in 8 combinations: with no oracles
+in eight combinations: with no oracles
 (\textbf{n}), with only fixpoint (\textbf{f}), pattern (\textbf{p}), or solid (\textbf{s}) oracle,
@@
 lazy application of the substitution, already reduces the effects of the JP
-procedure's main bottlenecks. Raw evaluation data shows that on TPTP benchmarks, complete and pragmatic configurations
+procedure's main bottlenecks. The raw evaluation data show that on TPTP benchmarks, complete and pragmatic configurations
 differ in the set of problems they solve -- cv solves 19
@@
 \looseness=-1
-The problem addressed in this chapter is that of finding
+The problem addressed in this chapter is that of designing
 a complete and efficient higher-order unification procedure. Three main lines of
@@
 is a version of the PT procedure presented in Section \ref{sec:unif:solid-oracle}. It improves 
-Huet's procedure by featuring a Solution rule.
+Huet's procedure by featuring a \textsf{Solution} rule.
 
@@
 some of the Zipperposition parameters we tuned to make better use of the procedure.
-It also makes the basis of the unification procedure implemented in $\lambda$E,
+The procedure also forms the basis of the unification procedure implemented in $\lambda$E,
 the full higher-order extension of Ehoh, described in Chapter \ref{ch:ehoh2}.

### 2022-03-04 unif/unif.tex
examples

@@
 
+\begin{exa} To determine the opponent pairs of $\lamx{\cst{g} \, (\cst{g} x)}$
+  and $\lamx{\cst{g} \, x}$ we first $\alpha$-rename the second term to 
+  $\lam{y}{\cst{g}\,{y}}$. Then, we identify the common context of $\cst{g} \, (\cst{g} x)$ and $\cst{g} \, y$
+  as $\cst{g} \, \square $, which leads us to the opponent pair $(\cst{g} \, x, y)$.
+\end{exa}
+
 \begin{lemma}[$\jp L{3\text{ (1)}}$]\label{lem:principal-lemma}Let $\varrho$ be a substitution
@@
   %
+\begin{exa}
+  Let $\varrho$ be the substitution $\{ X \mapsto \lamx{\cst{f} \, x}, Y
+  \mapsto \lamx{x} \}$. Clearly, $\varrho$ unifies the problem $X \, \cst{a}
+  \unif Y \, (\cst{f} \, \cst{a})$. The opponent pair for $\varrho\,X$ and $\varrho\,Y$
+  is $(\cst{f} \, x, y)$ which clearly adheres to the conclusion of Lemma \ref{lem:principal-lemma}.
+%
+  Intuitively, one of the consequences of this lemma is that any solution for
+  flex-flex problems that does not mention parameters will not have opponent pairs. 
+  For example, one solution for the 
+  problem $X \, \cst{a}
+  \unif Y \, (\cst{f} \, \cst{a})$ is $\varrho = \{ X \mapsto \cst{c}, Y \mapsto \cst{c} \}$
+  which has no opponent pairs.
+\end{exa}
 
@@
   \end{defi}
+
+  \begin{exa}
+    % \newcommand{\boldboundvar}[1]{\ensuremath{\mathit{\mathbf{#1}}}}
+    Following Example 1 in JP's work \cite{jp-76-unif}, we denote
+    $\omega$-simple occurrences of parameters using boldface. For example,
+    $\lambda xy. \, \cst{f} \, \mathbfit{x} \, \mathbfit{y}$ and $\lamx{\cst{g}
+    \, (\lam{y}{\mathbfit{x}\, y})}$ have several $\omega$-simple occurrences.
+    In the following two examples, boldface occurrences are \emph{not}
+    $\omega$-simple: $\lamx{\cst{g} \, (\lam{y}{\cst{f} \, y \, \mathbfit{x}})},\;
+    \lam{xy}{\cst{h} \, (x \, \mathbfit{y})}$. In the first case $x$ is not applied to parameter $y$ and in the second $y$
+    occurs as an argument to $x$.
+  \end{exa}
   This definition is slightly too restrictive for our purposes.
@@
   \end{defi}
+
   %

### 2022-03-04 ehoh/ehoh.tex
more

@@
 Each subgroup consists of two sub-sub\-groups of 1253
-problems, generated by using either $\lambda$-lifting or SKBCI-style combinators
+problems, generated by using either $\lambda$-lifting or {\sf SKBCI}-style combinators
 to encode $\lambda$-expressions.
@@
 provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
-they may require tedious reasoning about $\lambda$-lifted functions or SKBCI-style
+they may require tedious reasoning about $\lambda$-lifted functions or {\sf SKBCI}-style
 combinators}.

### 2022-03-04 pre/pre.tex
solid
more

@@
 \section{First-Order Logic}
+\label{sec:pre:fol}
 
@@
 \label{sec:pre:order}
-Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an
+Superposition calculus is parameterized by a {\em reduction ordering} $\succ$, an
 ordering on terms that has the following properties:

### 2022-03-04 unif/unif.tex
more

@@
 \looseness=-1
-The transitions are parametrized by a mapping~$\mathcal{P}$ that assigns a set
+The transitions are parameterized by a mapping~$\mathcal{P}$ that assigns a set
 of substitutions to a unification pair; this mapping abstracts the
 concept of unification rules present in other unification procedures. Moreover,
-the transitions are parametrized by a selection function~$S$ mapping a multiset
+the transitions are parameterized by a selection function~$S$ mapping a multiset
 $E$ of unification constraints to one of those constraints $S(E) \in E$, the
@@
 
-Like the JP procedure, our procedure, parametrized with $\mathcal{P}_\mathsf{c}$,  misses no unifiers:
+Like the JP procedure, our procedure, parameterized with $\mathcal{P}_\mathsf{c}$,  misses no unifiers:
 
@@
 The procedure described in Section \ref{sec:unif:the-unification-procedure}
-parametrized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
+parameterized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
 substitutions on the leaves of the constructed tree form a CSU.
@@
 \begin{lemma}[$\jp L{3\text{ (1)}}$]\label{lem:principal-lemma}Let $\varrho$ be a substitution
-    and $X$, $Y$ be free variables such that $\varrho\left(X\,\seq s\right)=\varrho\left(Y\,\seq t\right)$
+    and $X$, $Y$ be free variables such that $\varrho(X\,\seq s)=\varrho(Y\,\seq t)$
     for some term tuples $\seq s$ and $\seq t$. Then for every opponent
@@
   problem $X \, \cst{a}
-  \unif Y \, (\cst{f} \, \cst{a})$ is $\varrho = \{ X \mapsto \cst{c}, Y \mapsto \cst{c} \}$
+  \unif Y \, (\cst{f} \, \cst{a})$ is $\varrho = \{ X \mapsto \lamx{\cst{c}}, Y \mapsto \lamx{\cst{c}} \}$
   which has no opponent pairs.
@@
   For a free variable $X$, a unifier of $X\,\seq s \unif X\,\seq t$ does not necessarily
-  unify $s_{i} \unif t_{i}$. For example, $\{X \mapsto \cst{a}\}$ is a unifier that does
+  unify $s_{i} \unif t_{i}$. For example, $\{X \mapsto \lam{\overline{x}}{\cst{a}}\}$ is a unifier that does
   refer to any of the arguments of $X$.
@@
   of $x_{j}\,\seq v$ up to renaming of the parameters of $t_{j}$. 
-  To compare positions of $s$ and $s\,\seq t$, ignore
-  the parameter count mismatch.
-  % ASK(ALEX), ASK(VISA): I am no expert, but isn't it unusual to use these
-  % kind of instructions in the statement of the theorem? Maybe it would be good
-  % to locally redefine the concept of position.
+  To compare positions of $s$ and $s\,\seq t$, we ignore
+  the outermost $\lambda$-binders of s.
   \end{lemma}
@@
         eliminate all the free variables in $\lambda \overline{x}.\, a \,
-        \overline{u}_m$. Since PT is parametrized by an admissible
+        \overline{u}_m$. Since PT is parameterized by an admissible
         selection function, we know that there are no constraints descending
@@
 We used OCaml's functors to create a modular implementation. The core of our
-procedure is implemented in a module which is parametrized by another module
+procedure is implemented in a module which is parameterized by another module
 providing oracles and implementing the \textsf{Bind} step. In this way we can

### 2022-03-04 ehoh/ehoh.tex
more

@@
 %% kotelnikov-15-fool
-extend the syntax of this by making formulas a special case of terms,
+extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
 and its semantics by interpreting the Boolean type $o$ as a domain of

### 2022-03-04 unif/unif.tex
more

@@
 %is to limit the number or kind of bindings applied to a single constraint. 
-Our preliminary experiments showed that using the mapping $\mathcal{P}_\mathsf{p}$ defined as follows is a reasonable compromise between performance and
-completeness:
+Our preliminary experiments showed that using the mapping $\mathcal{P}_\mathsf{p}$ defined as follows gives good
+performance, while finding most useful unifiers:
 \begin{itemize}

### 2022-03-04 unif/unif.tex
more

@@
 & \hotofo{F \, \overline{s}} = F
-&& \hotofo{x_i \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
+&& \hotofo{\dbvar{i} \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
 && \hotofo{\cst{f} \, \overline{s}_n } = \cst{f}(\hotofo{s_1}, \ldots, \hotofo{s_n}) &&
@@
 \looseness=-1
-We let $x_i$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
+We let $\dbvar{i}$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
 and $\cstdba{i}$ be a fresh constant corresponding to this variable. All constants

### 2022-03-04 bools/bools.tex
bools

@@
 \label{ch:bools}
+
+\blfootnote{In this work I implemented and evalauted most of the described approaches.
+Visa Nummelin implemented the FOOL preprocessing module.}
+
+\authors{Joint work with Visa Nummelin}
+
+\begin{abstract}
+    We present a pragmatic approach to extending a Boolean-free higher-order
+    superposition calculus to support Boolean reasoning. Our approach extends
+    inference rules that have been used only in a first-order setting, uses some
+    well-known rules previously implemented in higher-order provers, as well as new rules.
+    We have implemented the approach in the Zipperposition
+    theorem prover. The evaluation shows highly competitive performance of our approach
+    and clear improvement over previous techniques.
+\end{abstract}
+
+\newpage
+
+\section{Introduction} 
+\label{sect:bool:introduction}
+
+In the last decades, automatic theorem provers have been used
+successfully as backends to ``hammers'' in proof assistants
+\cite{ku-15-holyhammer, pb-12-sh} and to software
+verifiers \cite{fp-13-why3}. Most advanced provers, such as CVC4
+\cite{cbetal-11-cvc4}, E \cite{scv-19-e23}, and Vampire
+\cite{lkav-13-vampire}, are based on first-order logic, whereas
+most frontends that use them are based on versions of higher-order logic. Thus,
+there is a large gap in expressiveness between front- and backends.
+%
+\looseness=-1
+This gap is bridged using well-known translations from higher-order to
+first-order logic \cite{ar-70-hol,
+mp-08-trans}. However, as shown in Chapter \ref{ch:ehoh}, translations are usually less efficient than native support. The
+distinguishing features of higher-order logic used by proof assistants that the
+translation must eliminate include
+$\lambda$-binders, function extensionality---the property that functions are
+equal if they agree on every argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x \, z \ieq y \,z)
+\iimplies x \ieq y$, and formulas occurring as arguments of function symbols \cite{mp-08-trans}.
+
+\looseness=-1
+A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
+extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate the
+first two above mentioned features of higher-order logic. Kotelnikov et al.\
+\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
+first-order logic to support the third feature of higher-order logic that
+requires translation. They described two approaches: One based on calculus-level
+treatment of Booleans and the other, which requires no changes to the calculus,
+based on preprocessing.
+
+\looseness=-1 To fully bridge the gap between higher-order and first-order tools,
+we combine the two approaches: we use the efficient higher-order superposition
+calculus and extend it with inference rules that reason with
+Boolean terms. In early work, Kotelnikov et al.\ 
+\cite{kotelnikov-15-fool} have described a \emph{FOOL paramodulation}
+rule that, under some order requirements, removes the need for the axiom
+describing the Boolean domain---$\iforall p. \; p \ieq \itrue \llor p
+\ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
+arguments of symbols is translated to first-order logic.
+
+\looseness=-1
+The backbone of our approach is based on an extension of this rule to higher-order
+logic: We do not translate away any Boolean structure that is nested inside
+non-Boolean terms and allow our rule to hoist the nested Booleans to the
+literal level. Then, we clausify the resulting formula (i.e., a clause containing formulas in literals) using a new rule.
+
+\looseness=-1 An important feature that we inherit by building on top of Bentkamp
+et al. \cite{bbtvw-21-sup-lam} is support for (function) extensionality. Moving to higher-order
+logic with Booleans also means that we need to consider \emph{Boolean extensionality}: $\iforall pq. \, (p
+\iequiv q) \iimplies p \ieq q$. We extend the rules of Bentkamp et al.\ 
+that treat function extensionality to also treat Boolean extensionality.
+
+\looseness=-1
+Rules that extend the two orthogonal approaches form the basis of our support
+for Boolean reasoning (Section \ref{sect:bool:native}). In addition, we have
+implemented rules that are inspired by the ones used in the
+higher-order provers Leo-III \cite{sb-21-leo3} and Satallax
+\cite{cb-12-satallax}, such as elimination of Leibniz equality, primitive
+instantiation and treatment of choice operator \cite{pa-01-classical-ty-thy}. We
+have also designed new rules that use higher-order unification to resolve
+Boolean formulas that are hoisted to literal level, delay clausification of non-atomic
+literals, reason about formulas under $\lambda$-binders, and many others. Even
+though the rules we use are inspired by the ones of refutationally complete
+higher-order provers, we do not guarantee completeness of our extension of
+$\lambda$-superposition.
+
+\looseness=-1
+We compare our native approach with two alternatives based on
+preprocessing (Section \ref{sect:bool:alternative}). First, we compare it to  an
+axiomatization of the theory of Booleans. Second, inspired by work of
+Kotelnikov et al.\ \cite{kotelnikov-16-fool}, we implemented the preprocessing approach that does not
+require introduction of Boolean axioms.
+We also discuss some examples, coming from TPTP \cite{gs-17-tptp}, that
+illustrate advantages and disadvantages of our approach (Section \ref{sect:bool:examples}).
+
+Our approach is implemented in the Zipperposition theorem prover
+\cite{sc-15-simon-phd,sc-supind-17}. Zipperposition is an easily
+extensible open source prover that Bentkamp et al.\ used to implement their
+higher-order superposition calculus. We further extend their implementation.
+
+\looseness=-1
+We performed an extensive evaluation of our approach (Section \ref{sect:bool:eval}).
+In addition to evaluating different configurations of our new rules, we have
+compared them to full higher-order provers CVC4, Leo-III,  Satallax and Vampire.
+The results suggest that it is beneficial to natively support Boolean reasoning
+-- the approach outperforms preprocessing-based approaches. Furthermore, it is
+very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
+approaches we base on, as well as related approaches (Section
+\ref{sect:bool:discussion}).
+
+\section{Background} 
+\label{sect:bool:background}
+
+We base our work (and parts of the following text) on Bentkamp et al.'s \cite{bbtvw-21-sup-lam} extensional
+polymorphic clausal higher-order logic. We extend the syntax of this logic by
+adding logical connectives to the signature. The semantic of the logic
+is extended by interpreting Boolean type $o$ as a two-element domain. This
+amounts to extending Bentkamp et al's fragment of higher-order logic to
+full-higher order logic (HOL). Taking a different perspective, this logic is an extension of the one
+presented in Chapter \ref{ch:ehoh} with $\lambda$ abstraction, polymorphic
+types, and native logical connectives. For reference, we provide precise definition of this polymorphic logic.
+
+\looseness=-1
+A signature is a quadruple $(\Sigmaty, \Vty, \Sigma, \VV)$ where $\Sigmaty$ is a
+set of type constructors, $\Vty$ is a set of type variables and $\Sigma$ and
+$\VV$ are sets of constants and term variables, respectively. We require nullary
+type constructor $o$, as well as binary constructor $\rightarrow$
+to be in $\Sigmaty$. A type $\tau, \upsilon$ is either a type variable $\alpha \in
+\Vty$ or of the form $\kappa(\tau_1, \ldots \tau_n)$ where $\kappa$ is an
+$n$-ary type constructor. We write $\kappa$ for $\kappa()$, $\tau \rightarrow
+\upsilon$ for $\rightarrow(\tau, \upsilon)$, and we drop parentheses to shorten 
+$\tau_1 \rightarrow (\cdots \rightarrow (\tau_{n-1} \rightarrow \tau_n) \cdots)$ into $\tau_1 \rightarrow \cdots \rightarrow
+\tau_n$. Each symbol in $\Sigma$ is
+assigned a type declaration of the form $\forallty{\tuplen{\alpha}} \tau$ where all variables
+occurring in $\tau$ are among $\tuplen{\alpha}$.
+
+\looseness=-1
+Function symbols $\cst{a}, \cst{b}, \cst{f}, \cst{g}, \ldots$ are elements of
+$\Sigma$; their type declarations are written as $\cst{f} :
+\forallty{\tuplen{\alpha}} \tau$. Set $\VV$ contains both free and bound variables, following
+the distinction introduced in Sect.~\ref{sec:pre:hol}. Term variables from the set $\VV$ are written
+$F,G,X,Y \ldots$ and we denote their types as $X : \tau$. When the type is not
+important, we omit type declarations. We assume that symbols $\itrue,
+\ifalse,\inot,\iand, \ior,\iimplies,\iequiv$ with their standard meanings and type declarations are elements of
+$\Sigma$. Furthermore, we assume that polymorphic symbols $\iforall$ and $\iexists$
+with type declarations $\forallty{\alpha} (\alpha \rightarrow o) \rightarrow o$
+and $\ieq \; : \forallty{\alpha} {\alpha \rightarrow \alpha \rightarrow o}$ are
+in $\Sigma$, with their standard meanings. All these symbols are called \emph{logical
+symbols}. We use infix notation for binary logical symbols.
+
+\looseness=-1 Terms are defined inductively as follows. Free ($X : \tau$)  and bound ($x : \tau$) variables   are
+terms of type $\tau$. If $\cst{f} : \forallty{\tuplen{\alpha}} \tau$ is in
+$\Sigma$ and $\tuplen{\upsilon}$ is a tuple of types, called type arguments, then
+$\cst{f}\typeargs{\tuplen{\upsilon}}$ (written as $\cst{f}$ if $n=0$, or if type
+arguments can be inferred from the context) is a term of type $\tau \{
+\tuplen{\alpha} \mapsto \tuplen{\upsilon} \}$, called constant. If $x$ is a bound variable
+of type $\tau$ and $s$ is a term of type $\upsilon$ then $\lamx{s}$ is a term of type
+$\tau \rightarrow \upsilon$. If $s$ and $t$ are of type $\tau \rightarrow \upsilon$ and
+$\tau$, respectively, then $s \, t$ is a term of type $\upsilon$. We call terms of
+Boolean type ($o$) \emph{formulas} and denote them by $f,g,h, \ldots$; we use
+$p,q,r, \ldots$ for free variables whose result type is $o$ and
+$\cst{p},\cst{q},\cst{r}$ for constants with the same result type.
+%
+Unless stated otherwise, we view terms as
+$\alpha\beta\eta$-equivalence classes, with $\eta$-long $\beta$-reduced form as
+the representative.
+
+\looseness=-1
+Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
+Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
+constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the formula
+$f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \llor
+\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is
+a nested Boolean, as well as its subterms. First-order logic allows only top-level Booleans, whereas nested Booleans are characteristic for
+higher-order logic.

### 2022-03-04 bools/bools.tex
more

@@
 %
+Formulas whose top-level symbol is not logical are called \emph{atoms}.
 Unless stated otherwise, we view terms as
@@
 Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
-constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the formula
-$f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \llor
-\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is
-a nested Boolean, as well as its subterms. First-order logic allows only top-level Booleans, whereas nested Booleans are characteristic for
+constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
+formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q})
+\llor \inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while
+$f|_{1.2.1}$ is a nested Boolean, as well as its subterms. First-order logic
+allows only top-level Booleans, whereas nested Booleans are characteristic for
 higher-order logic.
+
+\section{The Native Approach} 
+\label{sect:bool:native}
+
+\looseness=-1
+Zipperposition already had some support for Boolean reasoning before we
+started extending the calculus of Bentkamp et al. In this section, we first
+describe the internals of Zipperposition responsible for reasoning with
+Booleans. We continue by describing \NumberOK{15} rules that we have
+implemented. For ease of presentation we divide them in three categories.
+\subsection{Support for Booleans in Zipperposition}
+\label{subsect:bool:zip-bools}
+
+As mentioned in Chapter~\ref{ch:unif}, Zipperposition is an open source prover
+written in OCaml. From its inception, it was designed as a prover that supports
+easy extension of its base superposition calculus to various theories, including
+arithmetic, induction and limited support for higher-order logic
+\cite{sc-15-simon-phd}.
+
+In Zipperposition, applications are represented in flattened, spine notation. In
+addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
+the nested applications of these symbols. For example, terms $\cst{p} \iand
+(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
+internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. 
+Zipperposition's support for $\lambda$-terms is used to represent quantified nested
+Booleans: Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
+$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of the
+input problem, no nested Booleans will be modified or renamed using fresh
+predicate symbols.
+
+The version of Zipperposition preceding our modifications distinguished between non\-equational
+and equational literals. Following E \cite{scv-19-e23}, we
+modified Zipperposition to represent all literals equationally: a nonequational
+literal $f$ is stored as $f \eq \itrue$, whereas $\neg f$ is stored as $f
+\noteq \itrue$. Equations of the form $f \eq \ifalse $ and $f \noteq \ifalse$ are
+transformed into $f \noteq \itrue$ and $f \eq \itrue$, respectively.
+
+\subsection{Core Rules}
+\label{subsect:bool:core}
+
+\looseness=-1
+Kotelnikov et al.\ \cite{kotelnikov-15-fool}, to the best of our
+knowledge, pioneered the approach of extending a first-order superposition prover to support
+nested Booleans. They call effects of including the axiom $\iforall
+p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
+explosive behavior of the axiom, they imposed the following two requirements to
+the simplification order $\succ$: $\itrue \succ \ifalse$ and
+$\itrue$ and $\ifalse$ are two smallest ground terms with respect to $\succ$. If
+these requirements are met, there is no self-paramodulation of the clause
+and only paramodulation possible is from literal $p \eq \itrue$ of the mentioned axiom
+into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
+the axiom with the inference rule \emph{FOOL Paramodulation} (\infname{FP}):
+%
+$$ \namedinference{FP}{C[f]}{C[\itrue] \vee f \eq \ifalse} $$
+%
+\looseness=-1
+where $f$ is a nested non-variable Boolean subterm of clause $C$, different from
+$\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
+Booleans  to first-order logic without interpreted Booleans; thus, symbols $\itrue$ and $\ifalse$, and type $o$
+correspond to proxy symbols and types introduced during the translation.
+
+We created two rules that are syntactically similar to \infname{FP} but
+are adapted for higher-order logic with one key distinction -- we do not perform any translation:
+%
+\begin{align*}
+  & \namedinference{Cases}{C[f]}{C[\ifalse] \vee f \eq \itrue}
+  && \namedsimp{CasesSimp}{C[f]}{C[\ifalse] \vee f \eq \itrue \quad C[\itrue] \vee f \noteq \itrue}
+\end{align*}
+%
+\looseness=-1
+Clearly, the prover that uses the rules should not include them both at the same time.
+In addition, since
+literals $f \eq \ifalse$ are represented as negative equations $f \noteq \itrue$, which cannot be used to paramodulate from,
+we change the first requirement on the order to $\ifalse \succ
+\itrue$.
+
+\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
+These two rules hoist Boolean subterms $f$ to the literal level; therefore,
+some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
+$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
+called eager clausification (\infname{EC}):
+%
+$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
+%
+We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
+where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
+is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{EC}
+is applicable if clause $C = L_1 \llor
+\cdots \llor L_n$ is not standard.
+The resulting clauses $\tuple{D}{m}$ represent
+the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
+\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
+Using Boolean extensionality, Zipperposition's clausification
+algorithm treats Boolean equality as equivalence (i.e., it replaces
+$\mathop{\eq}\typeargs{o}$ with $\iimplies$).
+
+\looseness=-1
+An advantage of leaving nested Booleans unmodified is that the prover will be able
+to prove some problems containing them without using the prolific rules described
+above. For example, given two clauses $\cst{f} \, (\cst{p} \, x
+\iimplies \cst{p} \, y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
+\cst{a} \iimplies \cst{p} \, \cst{b}) \noteq \cst{a}$, the empty clause can
+easily be derived without the above rules. A disadvantage of this approach
+is that the proving process will periodically be interrupted by expensive calls
+to the clausifier.
+\pagebreak[2]
+
+\looseness=-1
+If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result in many redundant clauses. Consider the following
+example: let $\cst{p} : o \rightarrow o$, $\cst{a} : o$ and consider a clause
+set containing $\cst{p} \, (\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq
+\itrue$. Then, the clause $C = \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
+derived in eight ways using the rules, depending on which nested Boolean subterm was
+chosen for the inference. In general, if a clause has a subterm occurrence of the form $\cst{p}^n \, \cst{a}$,
+where both $\cst{p}$ and $\cst{a}$ have result type $o$, the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in $2^{n-1}$ ways.
+% Boolean subterms occurring in different
+% eligible literals cause similar explosive behavior.
+To combat these issues we
+implemented pragmatic restrictions of the rule: only $f$ which is
+the leftmost outermost (or innermost) eligible subterm will be considered. With
+this modification $C$ can be derived in only one way. Furthermore,
+some intermediate conclusions of the rules will not be derived, pruning the search space.
+
+The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
+aggressively simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
+the formula $\cst p\land\itrue$ will be replaced by $\cst p$. To simplify nested Booleans we implemented the rule
+\pagebreak[2]
+\[
+\namedsimp{BoolSimp}{C[f\sigma]}{C[g\sigma]}
+\]
+where $f \longrightarrow g \in E$ runs over fixed set of rewrite rules $E$,
+and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
+described by Nonnengart and Weidenbach \cite[Section~3]{nw-01-small-cnf}. This set contains
+the rules describing how each logical symbol behaves when given arguments $\itrue$ or $\ifalse$: for example, 
+it includes $(\itrue \iimplies p)~\longrightarrow~p $ and $(p \iimplies \itrue) \longrightarrow \itrue$. %full list visa report Sect 4.2
+Leo-III implements a similar rule, called \textsf{simp} \cite[Section 4.2.1.]{as-18-phd}.
+
+Our decision to represent negative atoms as negative equations was motivated by
+the need to alter Zipperposition's earlier behavior as little as possible. 
+Namely, negative atoms were not used as literals that can be used
+to paramodulate from, and as such added to the laziness of the superposition calculus.
+However, it might be useful to consider unit clauses of the form $f \noteq \itrue$
+as $f \eq \ifalse$ to strengthen rewriting. To that end, we have introduced the following
+rule:
+%
+$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[f\sigma]}{f \noteq \itrue \qquad C[\ifalse]} $$

### 2022-03-05 bools/bools.tex
more

@@
 Rules that extend the two orthogonal approaches form the basis of our support
-for Boolean reasoning (Section \ref{sect:bool:native}). In addition, we have
+for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
 implemented rules that are inspired by the ones used in the
@@
 \looseness=-1
-We compare our native approach with two alternatives based on
-preprocessing (Section \ref{sect:bool:alternative}). First, we compare it to  an
+We compare our native approach with two alternatives which are based on
+preprocessing (Sect.~\ref{sect:bool:alternative}). First, we compare it to  an
 axiomatization of the theory of Booleans. Second, inspired by work of
@@
 require introduction of Boolean axioms.
-We also discuss some examples, coming from TPTP \cite{gs-17-tptp}, that
-illustrate advantages and disadvantages of our approach (Section \ref{sect:bool:examples}).
+We discuss some examples, coming from TPTP \cite{gs-17-tptp}, that
+illustrate advantages and disadvantages of our approach (Sect.~\ref{sect:bool:examples}).
 
 Our approach is implemented in the Zipperposition theorem prover
-\cite{sc-15-simon-phd,sc-supind-17}. Zipperposition is an easily
-extensible open source prover that Bentkamp et al.\ used to implement their
-higher-order superposition calculus. We further extend their implementation.
+\cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
+to implement their higher-order superposition calculus. We further extend
+their implementation.
 
 \looseness=-1
-We performed an extensive evaluation of our approach (Section \ref{sect:bool:eval}).
+We performed an extensive evaluation of our approach (Sect.~\ref{sect:bool:eval}).
 In addition to evaluating different configurations of our new rules, we have
@@
 very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
-approaches we base on, as well as related approaches (Section
+approaches we base on, as well as related approaches (Sect.~
 \ref{sect:bool:discussion}).
@@
 \forallty{\tuplen{\alpha}} \tau$. Set $\VV$ contains both free and bound variables, following
-the distinction introduced in Sect.~\ref{sec:pre:hol}. Term variables from the set $\VV$ are written
-$F,G,X,Y \ldots$ and we denote their types as $X : \tau$. When the type is not
+the distinction introduced in Sect.~\ref{sec:pre:hol}. Free term variables from the set $\VV$ are written
+$F,G,X,Y \ldots$ and we denote their types as $X : \tau$. Bound term variables are written $x,y,z,\ldots$,
+and their types are similarly denoted. When the type is not
 important, we omit type declarations. We assume that symbols $\itrue,
@@
 Boolean type ($o$) \emph{formulas} and denote them by $f,g,h, \ldots$; we use
-$p,q,r, \ldots$ for free variables whose result type is $o$ and
+$P,Q,R, \ldots$ for free variables whose result type is $o$ and
 $\cst{p},\cst{q},\cst{r}$ for constants with the same result type.
@@
 the axiom with the inference rule \emph{FOOL Paramodulation} (\infname{FP}):
+\pagebreak[2]
 %
@@
 $f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
-called eager clausification (\infname{EC}):
+called immediate clausification (\infname{IC}):
 %
@@
 where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
-is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{EC}
+is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
 is applicable if clause $C = L_1 \llor
@@
 The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
-aggressively simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
-the formula $\cst p\land\itrue$ will be replaced by $\cst p$. To simplify nested Booleans we implemented the rule
-\pagebreak[2]
+eagerly simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
+the formula $\cst p\land\itrue$ is replaced by $\cst p$. To simplify nested Booleans we implemented the rule
+
 \[
-\namedsimp{BoolSimp}{C[f\sigma]}{C[g\sigma]}
+\namedsimp{BoolSimp}{C[\sigmaterm{f}]}{C[\sigmaterm{g}]}
 \]
@@
 and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
-described by Nonnengart and Weidenbach \cite[Section~3]{nw-01-small-cnf}. This set contains
+described by Nonnengart and Weidenbach \cite[Sect.~3]{nw-01-small-cnf}. This set contains
 the rules describing how each logical symbol behaves when given arguments $\itrue$ or $\ifalse$: for example, 
 it includes $(\itrue \iimplies p)~\longrightarrow~p $ and $(p \iimplies \itrue) \longrightarrow \itrue$. %full list visa report Sect 4.2
-Leo-III implements a similar rule, called \textsf{simp} \cite[Section 4.2.1.]{as-18-phd}.
+Leo-III implements a similar rule, called \textsf{simp} \cite[Sect.~4.2.1.]{as-18-phd}.
 
@@
 %
-$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[f\sigma]}{f \noteq \itrue \qquad C[\ifalse]} $$
+$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[\sigmaterm{f}]}{f \noteq \itrue \qquad C[\ifalse]} $$
+
+\subsection{Higher-Order Considerations}
+\label{subsect:bool:core}
+\looseness=-1
+To achieve refutational completeness of higher-order resolution and similar
+calculi it is necessary to instantiate variables with result type $o$,
+\emph{predicate variables}, with arbitrary formulas
+\cite{as-18-phd,pa-01-classical-ty-thy}. Fortunately, we can approximate the
+formulas using a complete set of logical symbols (e.g., $\inot$, $\iforall$, and
+$\iand$). Since such an approximation is not only necessary for completeness of
+some calculi, but very useful in practice, we implemented the \emph{primitive
+instantiation} (PI) rule:
+%
+$$ \namedinference{PI} {C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq}
+\, t} {\substcl{\{ P
+\mapsto f \}}{C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq} \, t}  } $$
+%
+where $P$ is a free variable of
+the type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow o$. 
+Choosing a different $f$ that instantiates $P$, we can balance between
+explosiveness of approximating a complete set of logical symbols and
+incompleteness of pragmatic approaches. We borrow the notion of imitation from
+higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}) %\cite{OUR UNIFICATION PAPER}:
+, and we say
+that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
+\, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
+\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $z$ of type $\nu_1
+\rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
+are fresh free variables, where each $Y_i$ has the type $\nu_1 \rightarrow
+\cdots \rightarrow \nu_m \rightarrow \tau_i$; variable $x_i$ is of type $\nu_i$.
+\pagebreak[2]
+
+\looseness=-1
+Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
+before we started our modifications. The rule has different modes that generate
+sets of possible terms $f$ for $p: \tau_1 \rightarrow \cdots \rightarrow \tau_n
+\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
+$\star$ is an element of a set of logical constants $P = \{ \iand, \ior,
+\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
+contains imitations (for $p$) of all elements of $P$. Mode $\emph{Pragmatic}$
+contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
+such that $i\ineq j$ and  $\tau_i = \tau_j$, it contains $\lam{\tuplen{x}}{x_i
+\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
+\tau_j = o$, then it contains $\lam{\tuplen{x}}{x_i \iand x_j}$ and
+$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
+contains $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ contains
+imitations of $\itrue$, $\ifalse$ and $\star$ (except for $\mathit{Imit_{\iforall\iexists}}$ which contains imitations
+of both $\iforall$ and $\iexists$).
+
+\looseness=-1
+While experimenting with our implementation we have noticed some proof patterns
+that led us to come up with the following modifications. First, it often
+suffices to perform \infname{PI} only on initial clauses -- which is why we
+allow the rule to be applied only to the clauses created using at most $k$
+generating inferences. Second, if the rule was used in the proof, its premise
+is usually only used as part of that inference -- which is why we implemented
+a version of \infname{PI} that removes the clause after all possible \infname{PI}
+inferences have been performed. We observed that the mode \emph{Imit}$_\star$ is useful in practice since often only a single
+approximation of a logical symbol is necessary.
+
+\looseness=-1
+Efficiently treating axiom of choice is notoriously difficult for higher-order provers. Andrews formulates
+this axiom as $\iforall p.\, (\iexists
+x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
+\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
+operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
+\, (\varepsilon \, P) \eq \itrue$. Since term $P \, X$ matches any Boolean
+term in the proof state, this axiom is very explosive. Therefore, Leo-III
+\cite{sb-21-leo3} deals with the choice operator on the calculus
+level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
+P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
+added to set of choice functions $\mathit{CF}$ (which initially contains just
+$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
+instantiate the axiom of choice. We reused the method of recognizing choice
+functions, but generalized the rule for creating the instance of
+the axiom (assuming $\xi \in \mathit{CF}$):
+%
+$$\namedinference{Choice}
+               {C[\xi \, t]}
+               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
+%
+
+Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
+arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
+the variable $X$ can later be replaced by imitation of logical symbols to create more
+complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
+\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, based on
+Zipperposition parameters, $D$ will either be deleted or kept. Note that $D$
+will not subsume its instances, since the matching algorithm of Zipperposition is
+too weak for this.
+\looseness=-1
+
+Most provers natively support extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
+first-order unification to return unification constraints consisting of pairs of
+terms of functional type, whereas Steen relies on the unification rules of
+Leo-III's calculus \cite[Section 4.3.3.]{as-18-phd} to deal with extensionality.
+Bentkamp et al \cite{bbtvw-21-sup-lam} altered core generating inference rules of the superposition calculus
+to support extensionality. Instead of requiring that terms
+involved in the inference are unifiable, it is required that they can be
+decomposed into \emph{disagreement pairs} such that at least one of the
+disagreement pairs is of functional type. Disagreement pairs of terms $s$ and
+$t$ of the same type are defined inductively using function $\textsf{dp}$:
+$\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal; $\textsf{dp}(a \,
+\tuplen{s}, b \, \tuple{t}{m}) = \{(a \,
+\tuplen{s}, b \, \tuple{t}{m})\}$ if $a$ and $b$ are different heads;
+$\textsf{dp}(\lam{x}{s}, \lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$;
+$\textsf{dp}(a \, \tuplen{s}, a \, \tuplen{t}) = \bigcup_{i=1}^{n}
+\textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
+$$
+\namedinference{AbsSup}
+  {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
+  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} $$
+\pagebreak[2]
+\begin{align*}
+  &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
+  &\namedinference{AbsEF}{s \eq t \llor  s' \eq u \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor t \noteq u \llor  s' \eq u \llor C}}
+\end{align*}
+
+\noindent\looseness=-1
+Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are extensional
+versions of superposition, equality resolution and equality factoring
+(Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of these three
+rules. In each of the rules, $\sigma$ is a most general unifier of the types of
+$s$ and $s'$, and $\textsf{dp}(s\sigma ,s' \sigma) = \{ (s_1, s'_1), \ldots,
+(s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
+the standard rules, except that condition that $s$ and $s'$ are unifiable is
+replaced by the condition that at least one $s_i$ is of functional type and that
+$n>0$. This rule is easily extended to support Boolean extensionality by
+requiring that at least one $s_i$ is of functional or type $o$, and adding the
+condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
+formulas'' to the definition of $\textsf{dp}$.
+
+Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
+(\inot (\cst{p} \iand \cst{q}))$. This problem is obviously unsatisfiable, since
+arguments of $\cst{f}$ on different sides of the disequation are extensionally
+equal; however, without \infname{Abs} rules Zipperposition will rely on
+\infname{Cases}(\infname{Simp}) and \infname{EC} rules to derive the empty
+clause. Rule \infname{AbsER} will generate~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
+\inot (\cst{p} \iand \cst{q})$. Then, $C$ will get clausified using
+\infname{EC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
+\inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.
+
+\looseness=-1
+Zipperposition restricts \infname{AbsSup} by
+requiring that $s$ and $s'$ are not of function or Boolean types. If the terms are of function type, our experience is
+that better treatment of function extensionality is to apply fresh free
+variables (or Skolem terms, depending on the sign
+\cite{bbtvw-21-sup-lam}) to both sides of a (dis)equation to reduce it to
+a first-order literal; Boolean extensionality is usually better supported by
+applying \infname{EC} on the top-level Boolean term. Thus, for the following
+discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
+and $s'$ have the same head, and a functional or Boolean subterm. To speed up
+retrieval of such terms, we added an index that maps symbols to positions in
+clauses where they appear as a head of a term that has a functional or Boolean
+subterm. This index will be empty for first-order problems, incurring no
+overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply \infname{Abs} rules if all
+disagreement pairs have at least one side whose head is a variable; those will
+be dealt with more efficiently using standard, non-extensional, versions of the
+rules. We also eagerly resolve literals $s_i \not\eq s'_i$ using at most one
+unifier returned by terminating, pragmatic variant of unification algorithm described in 
+Sect.~\ref{sec:unif:the-unification-procedure}.
+
+Expressiveness of higher-order logic allows users to define equality using a single axiom,
+called Leibniz equality \cite{pa-01-classical-ty-thy}:
+%
+$ \iforall xy. \,
+   (\iforall p. \, p \, x \iimplies p \, y) \iimplies x \ieq y$.
+%
+Leibniz equality often appears in TPTP problems. Since modern provers have the native support
+for equality, it is usually beneficial to recognize and replace occurrences of Leibniz equality.
+
+Before we began our modifications, Zipperposition had a powerful rule that
+recognizes clauses that contain variations of Leibniz equality and instantiates
+them with native equality. This rule was designed by Simon Cruanes, and to the
+best of our knowledge, it has not been documented so far. With his permission
+we describe this rule as follows:
+\pagebreak[2]
+%
+$$\namedinference{ElimPredVar}{P \, \overline{s}_n^1 \eq \itrue \llor \cdots \llor
+    P \, \overline{s}_n^i \eq \itrue \llor P \, \overline{t}_n^1 \noteq \itrue \llor
+    \cdots \llor P \, \overline{t}_n^j \noteq \itrue \llor C} 
+    {\sigmacl{P \, \overline{s}_n^1
+    \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue  \llor C}} $$
+%
+where $P$ is a free variable, $P$ does not occur in any $s_k^l$ or $t_k^l$, or
+in $C$; $\sigma$ is defined as $\{ P \mapsto \lam{\tuplen{x}}{\pmb{\bigvee}_{k=1}^{j}
+(\pmb{\bigwedge}_{l=1}^{n} x_l \eq t^k_l ) } \}$.
+
+\looseness=-1
+To better understand how this rule removes variable-headed negative literals,
+consider the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor P \,
+\cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
+\itrue$. Since all side conditions are fulfilled, the rule \infname{ElimPredVar}
+will generate $\sigma = \{ P \mapsto \lam{xy}{(x \ieq \cst{b}_1
+\iand y \eq \cst{b}_2) \ior (x \ieq \cst{c}_1 \iand y \ieq \cst{c}_2)  } \}$.
+After applying $\sigma$ to $C$ and subsequent $\beta$-reduction, negative literal
+$  P \, \cst{b}_1 \, \cst{b}_2 \not\eq \itrue$ will reduce to 
+$ (\cst{b}_1 \ieq \cst{b}_1 \iand \cst{b}_2 \ieq \cst{b}_2) \ior (\cst{b}_1 \ieq \cst{c}_1 \iand \cst{b}_2 \eq \cst{c}_2) \noteq \itrue $,
+which is equivalent to $\ifalse$. Thus, we can remove this literal and all negative literals
+of the form $P \, \overline{t}_n \noteq \itrue$ from $C$ and apply $\sigma$ to the remaining ones.
+
+The previous rule removes all variables occurring in disequations in one
+attempt. We implemented two rules that behave more lazily, inspired by the ones present in Leo-III and
+Satallax:
+%
+\begin{align*}
+&\namedinference{ElimLeibniz$+$}
+{P \, \tuplen{s} \eq \itrue \llor P \, \tuplen{t} \noteq \itrue \llor C}
+{\sigmacl{s_i \eq t_i \llor C}}
+&&
+\namedinference{ElimLeibniz$-$}
+{P \, \tuplen{s} \noteq \itrue \llor P \, \tuplen{t} \eq \itrue \llor C}
+{\substcl{\sigma'}{s_i \eq t_i \llor C}}&
+\end{align*}
+%
+\looseness=-1
+where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma =
+\{ P \mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$ and
+$\sigma' = \{ P \mapsto \lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. This
+rule differs from \infname{ElimPredVar} in three ways. First, it acts on
+occurrences of variables in both positive and negative literals. Second,
+due to its simplicity, it usually does not require \infname{IC} as the following step.
+Third, it imposes much weaker conditions on $P$. However, removing all negative
+variables in one step might improve performance.  Coming back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor P \,
+\cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
+\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b_1}} \}$
+to obtain the clause $C' = \cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$.
+
+% \subsection{Additional Rules}
+% \label{subsect:bool:core}
+
+% Zipperposition's unification algorithm 
+% %\cite{OUR UNIFICATION PAPER}
+% uses flattened representation of terms with logical operators $\land$ and $\lor$
+% for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
+% are unifiable modulo associativity and commutativity of $\land$ and $\lor$. Let
+% $\diamond$ denote either $\land$ or $\lor$. When the unification algorithm is given
+% two terms $\diamond \, \tuplen{s}$ and $\diamond \, \tuplen{t}$, where neither
+% of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
+% following steps: First, it removes all terms that appear in both
+% $\tuplen{s}$ and $\tuplen{t}$ from the two argument tuples.
+% %Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
+% Next, the remaining terms are sorted first by their head term and then their weight. Finally,
+% an attempt is made to unify sorted lists pairwise.
+% %
+% As an example, consider the problem of unifying the pair $\big({\land} \, (\cst{p} \,
+% \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\land} \; (\cst{q} \, (\cst{f} \,
+% \cst{a})) \; (r \, (\cst{f} \, (\cst{f} \, \cst{a})))\bigr)$ where $r$ is a free
+% variable. If the arguments of $\land$ are simply sorted as described above, we would
+% try to unify $\cst{p} \, \cst{a}$ with $\cst{q} \, (\cst{f} \, \cst{a})$, and
+% fail to find a unifier. However, by removing term  $\cst{q} \, (\cst{f} \,
+% \cst{a})$ from the argument lists, we will be left with the problem
+% $(\cst{p} \, \cst{a},  r \, (\cst{f} \, (\cst{f} \, \cst{a})))$ which has a unifier.

### 2022-03-07 bools/bools.tex
more

@@
 A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
-extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate the
-first two above mentioned features of higher-order logic. Kotelnikov et al.\
+extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate
+$\lambda$-binders and functional extensionality. Kotelnikov et al.\
 \cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
@@
 have also designed new rules that use higher-order unification to resolve
-Boolean formulas that are hoisted to literal level, delay clausification of non-atomic
+Boolean formulas that are hoisted to literal level, delay clausification of nonatomic
 literals, reason about formulas under $\lambda$-binders, and many others. Even
@@
 \looseness=-1
-A signature is a quadruple $(\Sigmaty, \Vty, \Sigma, \VV)$ where $\Sigmaty$ is a
+A {\em signature} is a quadruple $(\Sigmaty, \Vty, \Sigma, \VV)$ where $\Sigmaty$ is a
 set of type constructors, $\Vty$ is a set of type variables and $\Sigma$ and
@@
 algorithm treats Boolean equality as equivalence (i.e., it replaces
-$\mathop{\eq}\typeargs{o}$ with $\iimplies$).
+$\mathop{\eq}\typeargs{o}$ with $\iequiv$).
 
@@
 to prove some problems containing them without using the prolific rules described
-above. For example, given two clauses $\cst{f} \, (\cst{p} \, x
-\iimplies \cst{p} \, y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
+above. For example, given two clauses $\cst{f} \, (\cst{p} \, X
+\iimplies \cst{p} \, Y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
 \cst{a} \iimplies \cst{p} \, \cst{b}) \noteq \cst{a}$, the empty clause can
@@
 incompleteness of pragmatic approaches. We borrow the notion of imitation from
-higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}) %\cite{OUR UNIFICATION PAPER}:
-, and we say
+higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}), and we say
 that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
 \, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
-\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $z$ of type $\nu_1
+\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $Z$ of type $\nu_1
 \rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
@@
 before we started our modifications. The rule has different modes that generate
-sets of possible terms $f$ for $p: \tau_1 \rightarrow \cdots \rightarrow \tau_n
+sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
 \rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
@@
 \mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
-contains imitations (for $p$) of all elements of $P$. Mode $\emph{Pragmatic}$
+contains imitations (for $P$) of all elements of $P$. Mode $\emph{Pragmatic}$
 contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
@@
 rules. In each of the rules, $\sigma$ is a most general unifier of the types of
-$s$ and $s'$, and $\textsf{dp}(s\sigma ,s' \sigma) = \{ (s_1, s'_1), \ldots,
+$s$ and $s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
 (s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
@@
     \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue  \llor C}} $$
-%
+% \pagebreak[2]
 where $P$ is a free variable, $P$ does not occur in any $s_k^l$ or $t_k^l$, or
@@
 \cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
-\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b_1}} \}$
+\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b}_1} \}$
 to obtain the clause $C' = \cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$.
 
-% \subsection{Additional Rules}
-% \label{subsect:bool:core}
+\subsection{Additional Rules}
+\label{subsect:bool:core}
 
-% Zipperposition's unification algorithm 
-% %\cite{OUR UNIFICATION PAPER}
-% uses flattened representation of terms with logical operators $\land$ and $\lor$
-% for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
-% are unifiable modulo associativity and commutativity of $\land$ and $\lor$. Let
-% $\diamond$ denote either $\land$ or $\lor$. When the unification algorithm is given
-% two terms $\diamond \, \tuplen{s}$ and $\diamond \, \tuplen{t}$, where neither
-% of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
-% following steps: First, it removes all terms that appear in both
-% $\tuplen{s}$ and $\tuplen{t}$ from the two argument tuples.
-% %Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
-% Next, the remaining terms are sorted first by their head term and then their weight. Finally,
-% an attempt is made to unify sorted lists pairwise.
-% %
-% As an example, consider the problem of unifying the pair $\big({\land} \, (\cst{p} \,
-% \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\land} \; (\cst{q} \, (\cst{f} \,
-% \cst{a})) \; (r \, (\cst{f} \, (\cst{f} \, \cst{a})))\bigr)$ where $r$ is a free
-% variable. If the arguments of $\land$ are simply sorted as described above, we would
-% try to unify $\cst{p} \, \cst{a}$ with $\cst{q} \, (\cst{f} \, \cst{a})$, and
-% fail to find a unifier. However, by removing term  $\cst{q} \, (\cst{f} \,
-% \cst{a})$ from the argument lists, we will be left with the problem
-% $(\cst{p} \, \cst{a},  r \, (\cst{f} \, (\cst{f} \, \cst{a})))$ which has a unifier.
+Zipperposition's unification algorithm 
+%\cite{OUR UNIFICATION PAPER}
+uses flattened representation of terms with logical operators $\iand$ and $\ior$
+for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
+are unifiable modulo associativity and commutativity of $\iand$ and $\ior$. Let
+$\Diamond$ denote either $\iand$ or $\ior$. When the unification algorithm is given
+two terms $\Diamond \, \tuplen{s}$ and $\Diamond \, \tuplen{t}$, where neither
+of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
+following steps: First, it removes all terms that appear in both
+$\tuplen{s}$ and $\tuplen{t}$ from the two argument tuples.
+%Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
+Next, the remaining terms are sorted first by their head term and then their weight. Finally,
+an attempt is made to unify sorted lists pairwise.
+%
+As an example, consider the problem of unifying the pair $\big({\iand} \,
+(\cst{p} \, \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\iand} \;
+(\cst{q} \, (\cst{f} \, \cst{a})) \; (R \, (\cst{f} \, (\cst{f} \,
+\cst{a})))\bigr)$ where $R$ is a free variable. If the arguments of $\iand$ are
+simply sorted as described above, we would try to unify $\cst{p} \, \cst{a}$
+with $\cst{q} \, (\cst{f} \, \cst{a})$, and fail to find a unifier. However, by
+removing term  $\cst{q} \, (\cst{f} \, \cst{a})$ from the argument lists, we
+will be left with the problem $(\cst{p} \, \cst{a},  R \, (\cst{f} \, (\cst{f}
+\, \cst{a})))$ which has a unifier.
+
+The winner of THF division of
+CASC-27 \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
+crucial advantage over Zipperposition: it is based on higher-order
+tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
+term, this instantiation naturally propagates through the whole formula. In
+Zipperposition, which is based on higher-order superposition, the original formula
+is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
+it in all clauses that are results of clausification of the same formula as $C$.
+To mitigate this issue, we have created extensions of equality resolution and equality factoring
+that take Boolean extensionality into account:
+%
+\begin{align*}
+  & \namedinference{BoolER}
+  { s \eq s' \llor C}
+  {\sigmacl{C}}
+  &&
+  \namedinference{BoolEF$+-$}
+  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
+  {\sigmacl{s' \not\eq \itrue \llor C}} \\
+  & \namedinference{BoolEF$-+$}
+  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
+  {\sigmacl{s' \eq \itrue \llor C}} &&
+  \namedinference{BoolEF$-$$-$}
+  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
+  {\sigmacl{s' \not\eq \itrue \llor C}}
+\end{align*}
+%
+All side conditions except for the ones concerning the unifiability of terms are
+as in the original equality resolution and equality factoring rules. In rule
+\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \neg s' $. In the $+-$
+and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ x \, \tuplen{s}
+$ and $\neg s'$, and in the remaining version it unifies $ x \, \tuplen{s} $ and
+$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
+form for application of the corresponding base rules. It suffices to consider 
+literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
+$s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
+rewriting into $ s \not\eq t $).
+
+Another approach to mitigate harmful effects of immediate clausification
+is to delay it as long as possible. Following the approach by 
+Ganzinger and Stuber \cite{gs-05-boolsup}, we represent every input formula $f$ as a unit clause $f \eq \itrue$ and use the following
+delayed clausification (\infname{DC}) rules:
+
+\vspace{2\jot}
+\begin{tabular}{ccc}
+  $\namedsimp{DC$_{\iand}$}
+  { (f \iand g) \eq \itrue \llor C}
+  { f \eq \itrue \llor C \quad g \eq \itrue \llor C }$ &
+  $\namedsimp{DC$_{\ior}$}
+  { (f \ior g) \eq \itrue \llor C}
+  { f \eq \itrue \llor g \eq \itrue  \llor C }$ &
+  $\namedsimp{DC$_{\iimplies}$}
+  { (f \iimplies g) \eq \itrue \llor C}
+  { f \not\eq \itrue \llor g \eq \itrue  \llor C } $ \\[2\jot]
+  $\namedsimp{DC$_{\inot}$}
+  { (\inot f) \eq \itrue \llor C}
+  { f \not\eq \itrue \llor C}$ &
+  $\namedsimp{DC$_{\iforall}$}
+  { (\iforall x. \, f) \eq \itrue \llor C}
+  { \{x \mapsto Y\}f \llor C }$ &
+  $\namedsimp{DC$_{\iexists}$}
+  { (\iexists x.\, f) \eq \itrue \llor C}
+  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}f \llor C }$ 
+  \\[2\jot]
+  \multicolumn{3}{c}{$\namedsimp{DC$_\eq$}
+  { f \eq g \llor C}
+  { f \not\eq \itrue \lor g\eq\itrue \llor C \quad f \eq \itrue \lor g\not\eq\itrue \llor C}$}
+\end{tabular}
+\vspace{2\jot}
+
+\looseness=-1
+The rules described above are as given by Ganzinger and Stuber (adapted to
+our setting), with the omission of rules for negative literals ($ f \not\eq
+\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.
+In \infname{DC}$_\eq$ we require both $f$ and $g$ to be formulas and at least one of
+them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
+\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
+freely in $\iexists x.\, f$.
+
+\looseness=-1
+Naive application of the \infname{DC} rules can result in exponential blowup in
+problem size. To avoid this, we rename formulas that have repeated
+occurrences. We keep the count of all nonatomic formulas occurring as
+either side of a literal. Before applying the \infname{DC} rule on a
+clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the number of $f$'s
+occurrences exceeds the threshold $k$. If it does, based on the
+polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause $\cst{p} \, \tuplen{Y}
+\not\eq \itrue \llor f \eq \itrue$ (if the literal is positive) or $\cst{p} \, \tuplen{Y}
+\eq \itrue \llor f \not\eq \itrue$ (if the literal is negative), where $\tuplen{Y}$ are
+all free variables of $f$ and $\cst{p}$ is a fresh symbol. Then, we replace the
+clause $f \mathrel{\dot\eq} \itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor
+C$.
+
+Before the number of occurrences of $f$ is checked, we first check (using a fast, incomplete matching algorithm) if there is a
+formula $g$, for which definition was already introduced, such that $\sigmaterm{g} =
+f$, for some substitution $\sigma$. This check can have three outcomes. First, if
+the definition $\textsf{q} \, \overline{X}_n$ is already introduced for $g$ with the
+polarity matching that of $f \, \dot{\eq} \, \itrue$, then $f$ is
+replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if the definition was introduced,
+but with different polarity, we create the clause defining $g$ with the missing
+polarity, and replace $f$ with $\sigmaterm{\cst{q} \, \tuplen{X}}$. Last, if the there is no renamed
+formula $g$ generalizing $f$, then we perform the previously described check.
+
+\looseness=-1
+In addition to reusing names for formula definitions, we reuse the Skolem
+symbols introduced by the \infname{DC}$_{\iexists}$ rule. When
+\infname{DC}$_{\iexists}$ is applied to $f = \iexists x.\, f'$ we check if there is
+a Skolem $\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y}$ introduced for a
+formula $g = \iexists x.\, g'$, such that $\sigmaterm{g} = f$. If so, the symbol
+$\cst{sk}$ is reused and $\iexists x.\, f'$ is replaced by $\{x \mapsto
+(\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y})\sigma \}f'$. Renaming and
+name reusing techniques are inspired by the VCNF algorithm described by Reger et
+al.\cite{rsv-16-vcnf}.
+
+% \namedinference{LC$_\eq$}
+% { f \eq g \llor C}
+% { f \not\eq \ifalse \llor g \eq \ifalse \llor C \quad f \eq \ifalse \llor g \not\eq \ifalse \llor C } &&
+% \namedinference{LC$_\Rightarrow$}
+
+% Rule \infname{BoolSimp} proved to be very useful in our initial experimentation
+% as it normalizes many equivalent terms to the same Boolean term. Stronger normalization can be achieved if nested formulas
+% are transformed into \emph{negation normal form} (\emph{NNF}). Formula is in negation normal form if it is built from atoms and their
+% negations using only $\wedge$ and $\vee$ \cite{nw-01-small-cnf}. By
+% adding
+% \kern\abovedisplayskip
+
+% \noindent
+% \begin{center}
+% \begin{minipage}[t]{.25\textwidth}
+%   \begin{center}
+%     $\neg\left(p \land q \right) \longrightarrow \neg p \vee \neg q$ \\
+%     $\neg\left(p \lor q \right)  \longrightarrow \neg p \land \neg q$
+%   \end{center}
+%   \end{minipage}%
+%   \begin{minipage}[t]{.43\textwidth}
+%   \begin{center}
+%     $ \neg \left( p \Leftrightarrow q \right) \longrightarrow \left(p \lor q \right) \land \left(\neg p \lor \neg q \right) $ \\
+%     $ p \Leftrightarrow q \longrightarrow \left(\neg p \lor q\right) \land \left(p\lor \neg q\right) $
+%   \end{center}
+%   \end{minipage}%
+%   \begin{minipage}[t]{.27\textwidth}
+%     \begin{center}
+%       $p \Rightarrow q  \longrightarrow \neg p\lor q$
+%     \end{center}
+%   \end{minipage}
+% \end{center}
+% \noindent to the set $E$ of rewrite rules of \infname{BoolSimp} we transform a formula to NNF. We call \infname{BoolSimp}
+% with set $E$ extended in this way \infname{NnfSimp}. Coming back to our previous example
+% of the clause $C = \cst{f} \, (\neg\cst{p} \lor \neg\cst{q}) \noteq
+% \cst{f} \, (\neg (\cst{p} \lland \cst{q}))$, \infname{NnfSimp} simplifies this problem greatly:
+% after normalizing terms, a simple equality resolution with an empty substitution
+% will derive the empty clause.
+
+\looseness=-1
+Rules \infname{Cases} and \infname{CasesSimp} deal with Boolean terms,
+but we need to rely on extensionality reasoning to deal with $\lambda$-abstractions
+whose body has type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
+implies that $\lam{\tuplen{x}}{f}$ is extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
+$\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \eq \lam{\tuplen{x}}{\ifalse})$, we
+designed the following rule (where all loosely bound variables of $f$ are $\tuplen{x}$):
+%\pagebreak[2]
+%
+$$ \namedinference{Interpret$\lambda$}{C[\lam{\tuplen{x}}{f}]}
+{(\forall {\tuplen{x}}.\, f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\itrue}] \qquad
+(\forall {\tuplen{x}}.\, \inot f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\ifalse}]} $$
+%
+
+\section{Alternative Approaches} 
+\label{sect:bool:alternative}
+
+\looseness=-1
+An alternative to heavy modifications of the prover needed to support the rules
+described above is to treat Booleans as yet another theory. Since the theory of Booleans
+is finitely axiomatizable, stating those axioms instead of creating special
+rules might seem appealing. Another approach
+is to preprocess nested Booleans by hoisting them to the top level.
+
+\ourpara{Axiomatization}
+\looseness=-1
+A simple axiomatization of the theory of Booleans is given by Bentkamp et al.\ \cite{bbtvw-21-sup-lam}.
+Following their approach, we introduce the proxy type $\typ{bool}$, which corresponds to $o$, to the
+signature. We define proxy symbols $\cst{t}, \cst{f}, \cst{not}, \cst{and}, \cst{or},
+\cst{impl}, \cst{equiv},\allowbreak \cst{forall}, \cst{exists}, \cst{choice}$, and $\cst{eq}$ which 
+correspond to the homologous logical constants from Section \ref{sect:bool:background}. In their
+type declarations $o$ is replaced by $\typ{bool}$.
+
+\looseness=-1
+To make this paper self-contained we include the axioms from Bentkamp et al. \cite{bbtvw-21-sup-lam}.
+Definitions of symbols are computational in nature: symbols are characterized by their behavior on $\cst{t}$ and $\cst{f}$.
+This also reduces interferences between different axioms. Axioms are listed as follows:
+
+\kern\abovedisplayskip
+
+\noindent
+\begin{minipage}[t]{.2\textwidth}
+\begin{center}
+$\cst{t} \noteq \cst{f}$ \\
+$X \eq \cst{t} \llor X \eq \cst{f}$ \\
+$\cst{not} \, \cst{t} \eq \cst{f}$ \\
+$\cst{not} \, \cst{f} \eq \cst{t}$ \\
+$\cst{and} \, \cst{t} \, X \eq X $ \\
+$\cst{and} \, \cst{f} \, X \eq \cst{f} $ \\ 
+\end{center}
+\end{minipage}%
+\begin{minipage}[t]{.35\textwidth}
+\begin{center}
+$\cst{or} \, \cst{t} \, X \eq \cst{t} $ \\
+$\cst{or} \, \cst{f} \, X \eq X $ \\
+$\cst{impl} \, \cst{t} \, X \eq X $ \\
+$\cst{impl} \, \cst{f} \, X \eq \cst{t} $ \\
+$X \noteq Y \llor \cst{eq}\typeargs{\alpha}\;X\;Y \eq \cst{t}$ \\
+$X \eq Y \llor \cst{eq}\typeargs{\alpha}\;X\;Y \eq \cst{f}$
+\end{center}
+\end{minipage}%
+\begin{minipage}[t]{.45\textwidth}
+  \begin{center}
+    $\cst{equiv} \, X\, Y \eq \cst{and} \, (\cst{impl} \, X\, Y) \, (\cst{impl} \, Y \, X) $ \\
+    $\cst{forall}\typeargs{\alpha}\,(\lambda x.\; \cst{t}) \eq \cst{t}$ \\
+    $Y \eq (\lambda x.\; \cst{t}) \llor \cst{forall}\typeargs{\alpha}\;Y \eq \cst{f}$ \\
+    $\cst{exists}\typeargs{\alpha}\;Y \eq \cst{not} \, (\cst{forall}\typeargs{\alpha}\;(\lambda x. \, \cst{not} \, (Y \, x)))$ \\
+    $Y \, X \eq \cst{f} \llor Y \, (\textsf{choice}\typeargs{\alpha} \, Y) \eq \cst{t}$ \\
+
+  \end{center}
+\end{minipage}
+
+\vspace{0.5em}
+
+\ourpara{Preprocessing Booleans}
+\looseness=-1
+Kotelnikov et al.\ extended VCNF, Vampire's algorithm for clausification, to support
+nested Booleans \cite{kotelnikov-16-fool}. As explained in Sect.~\ref{sec:ehoh:preprocessing}, we extended
+the clausification algorithm of Ehoh to support nested Booleans inspired by VCNF extension.
+Zipperposition and Ehoh share the
+same clausification algorithm, enabling us to reuse the extension with one
+notable difference:
+unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
+will be removed. Namely, Booleans that are below $\lambda$-abstraction
+and contain $\lambda$-bound variables will not be preprocessed. They cannot be easily hoisted to the level of an atom in which
+they appear, since this process might leak any variables bound in the context
+in which the nested Boolean appears. Similar preprocessing techniques are used in other higher-order provers
+\cite{wskb-16-effective-norm}.
+
+\section{Examples}
+\label{sect:bool:examples}
+
+The TPTP library contains thousands of higher-order benchmarks, many of them hand-crafted
+to point out subtle interferences of functional and Boolean properties of higher order logic. In this section
+we discuss some problems from the TPTP library that illustrate the advantages and disadvantages
+of our approach.
+
+During most of 2010s, the core calculus of the best performing
+higher-order prover at CASC was tableaux---a striking contrast from first-order part
+of the competition dominated by superposition-based provers. TPTP
+problem \verb|SET557^1| might shed some light on why tableaux-based provers
+excel on higher-order problems.
+This problem conjectures that there is
+no surjection from a set to its power set: 
+$$ \neg (\iexists x. \, \iforall y. \,
+\iexists z. \, x \, z \eq y) $$ 
+After negating the conjecture and
+clausification this problem becomes $ \cst{sk}_1 \, (\cst{sk}_2 \, Y) \eq Y   $
+where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, we can use
+\infname{ArgCong} rule \cite{bbtvw-21-sup-lam} which applies
+fresh variable $W$ to both sides of the equation, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
+\eq Y \, W$. Most superposition-based higher-order theorem
+provers (such as Leo-III, Vampire and Zipperposition) will split this clause
+into two clauses $C_1 = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \not\eq \top \llor Y
+\, W \eq \top $ and $ C_2 =  \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \eq \top \llor
+Y \, W \not\eq \top $. This clausification step makes the problem
+considerably harder. Namely, the clause $C$ instantiated with the substitution $
+\{ Y \mapsto \lamx{\inot (\cst{sk}_1 \, x \, x)}, W \mapsto \cst{sk}_2 \,
+(\lamx{\inot (\cst{sk}_1 \, x \, x))} \} $ yields the empty clause. However,
+if the original clause is split into two as described above, Zipperposition will
+rely on \infname{PI} rule to instantiate $Y$ with imitation of $\inot$ and on
+equality factoring to further instantiate this approximation. These desired
+inferences need to be applied on both new clauses and represent only a fraction
+of inferences that can be done with $C_1$ and $C_2$, reducing the chance of
+successful proof attempt. Rule \infname{BoolER} imitates the behavior of
+tableaux prover: it essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
+(\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $ which makes finding the necessary substitution easy and
+does not require a clausification step.
+
+\looseness=-1
+Combining rule \infname{(Bool)ER} with dynamic clausification is very fruitful as
+the benchmark \texttt{SYO033\^{}1} illustrates. This problem also contains the
+single conjecture
+$$ \iexists x. \, \iforall y.  \, x \, y \iequiv (\iforall z. \, y \,z)$$ The
+problem is easily solved if we replace variable $x$ with the constant $\forall$.
+Moreover, the prover does not have to blindly guess this instantiation. Instead,
+pretending that bound variables are free, it can obtain it by unifying $  X \, Y
+$ with $ \iforall \, Y$ (which is the $\eta$-short form of $\iforall z. \, Y
+\,z$). However, when the problem is clausified, all quantifiers are removed.
+Then, Zipperposition only finds the proof if appropriate instantiation mode of
+\infname{PI} is used, and if both clauses resulting from clausifying the negated
+conjecture are appropriately instantiated. In contrast, dynamic clausification will
+derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, ) $
+from the negated conjecture in three steps. Then, equality resolution results in
+an empty clause, swiftly finishing the proof without any explosive inferences.
+This effect is even more pronounced on problems \verb|SYO287^5| and
+\verb|SYO288^5|, in which critical proof step is instantiation of a variable
+with imitation of $\ior$ and $\iand$. In configurations that do not use dynamic
+clausification and \infname{BoolER}, Zipperposition times out in any reasonable
+time limit; with those two options it solves mentioned problems in less than
+100\,ms.
+
+In some cases, it is better to preprocess the problem. For example, TPTP problem
+\verb|SYO500^1.005| contains many nested Boolean terms:
+$$ \cst{f}_0 \, ( \cst{f}_1 \, ( \cst{f}_1 \, ( \cst{f}_1 \, ( \cst{f}_2 \, ( \cst{f}_3 \, ( \cst{f}_3 \, ( \cst{f}_3 \, ( \cst{f}_4 \, \cst{a} ) ) ) ) ) ) ) 
+    \eq \cst{f}_0 \, ( \cst{f}_0 \, ( \cst{f}_0 \, ( \cst{f}_1 \, ( \cst{f}_2 \, ( \cst{f}_2 \, ( \cst{f}_2 \, ( \cst{f}_3 \, ( \cst{f}_4 \, ( \cst{f}_4 \, ( \cst{f}_4 \, \cst{a} ) ) ) ) ) ) ) ) ) ) )$$
+In this problem, all functions $\cst{f}_i$ are of type $o \rightarrow o$, and constant $\cst{a}$ is of type $o$.
+FOOL unfolding of nested Boolean terms will result in exponential blowup in the problem size. However,
+superposition-based theorem provers are well-equipped for this issue: Their CNF algorithms use smart simplifications and
+formula renaming to mitigate these effects. Moreover, when the problem is preprocessed, the prover is aware of the problem
+size before the proving process starts and can adjust its heuristics properly. E, Zipperposition and Vampire, instructed to perform FOOL unfolding,
+solve the problem swiftly, using their default modes. However, if problem is not preprocessed, Zipperposition struggles to prove it using
+\infname{Cases(Simp)} and due to the large number of (redundant) clauses it creates, succeeds only if specific heuristic choices are made.

### 2022-03-07 bools/bools.tex
more

@@
 \infname{Cases(Simp)} and due to the large number of (redundant) clauses it creates, succeeds only if specific heuristic choices are made.
+
+\section{Evaluation} 
+\label{sect:bool:eval}
+
+We performed extensive evaluation to determine usefulness of our approach. As
+our benchmark set, we used all 2606 monomorphic theorems from the TPTP library,
+given in THF format. All of the experiments described in this section were
+performed on StarExec \cite{sst-14-starexec} servers with Intel Xeon E5-2609 0 CPUs clocked at 2.40 GHz. The evaluation is separated in two parts that
+answer different questions: How useful are the new rules? How 
+does our approach compare with state-of-the-art higher-order provers?
+
+\subsection{Evaluation of the Rules}
+
+\looseness=-1
+For this part of the evaluation, we fixed a single well-performing Zipperposition configuration
+called \emph{base} (\emph{b}). Since we are testing a single configuration, we
+used the CPU time limit of \NumberOK{15} s -- roughly the time a single configuration is
+given in a portfolio mode. Configuration \emph{b} uses the pragmatic variant
+pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
+It enables \infname{BoolSimp} rule,
+\infname{EC} rule, \infname{PI} rule in \emph{Pragmatic} mode with $k=2$,
+\infname{ElimLeibniz} and \infname{ElimPredVar} rules, \infname{BoolER} rule, and \infname{BoolEF} rules. 
+To evaluate the usefulness of
+all rules we described above, we enable, disable or change the
+parameters of a single rule, while keeping all other parameters of $b$ intact. 
+In figures that contain sufficiently different configurations, cells are of the form $n (m)$ where $n$ is the total
+number of proved problems by a particular configuration and $m$ is the number of
+unique problems that a given configuration solved, compared to the other
+configurations in the same figure. Intersections of rows and columns
+denote corresponding combination of parameters.
+Result for the base configuration is written in \emph{cursive}; the best result
+is written in \textbf{bold}.
+
+First, we tested different parameters of \infname{Cases} and \infname{CasesSimp}
+rules. In Figure \ref{fig:cases} we report the results. The columns correspond to
+three possible options to choose subterm on which the inference is performed:
+\textbf{a} stands for any eligible subterm, \textbf{lo} and \textbf{li} stands
+for leftmost outermost and leftmost innermost subterms, respectively. The rows
+correspond to two different rules: 
+\textbf{b} is the base configuration, which uses \infname{CasesSimp}, and \textbf{b$_\text{c}$}
+swaps this rule for \infname{Cases}. Although the margin is slim, the results show it is usually preferable to select leftmost-outermost
+subterm.
+
+% \newcommand\HEAD[1]{\hbox to \wd\mybox{\hfill\hbox{#1}\hfill}}
+% \newcommand\Z{\phantom{0}}
+% \newcommand\MIDLINE{\\[.25ex]\hline\rule{0pt}{3ex}}
+
+\begin{figure}[t]
+  \begin{center}
+    \def\arraystretch{1.1}%
+    \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
+      \strut                & a           & lo     & li  
+      \MIDLINE
+      b                     &  \emph{1646}   &  \textbf{1648}   &  1640  \\
+      b$_\text{c}$          &  1644   &  1645   &  1644 
+    \end{tabular}}
+    \caption{Effect of the \infname{Cases}(\infname{Simp}) rule on success rate }
+    \label{fig:cases}
+  \end{center}
+\end{figure}
+
+\looseness=-1
+Second, we evaluated all the modes of \infname{PI} rule with 3 values for
+parameter $k$: 1, 2, and~8 (Figure \ref{fig:pi}). The columns denote, from left to
+right: disabling the \infname{PI} rule, \emph{Pragmatic} mode, \emph{Full} mode,
+and \emph{Imit}$_\star$ modes with appropriate logical symbols. The rows denote
+different values of $k$. The results show that different values for $k$ have a
+modest effect on success rate. The raw data reveal that when we focus our
+attention to configurations with $k=2$, mode \emph{Full} can solve 10 problems
+no other mode (including disabling \infname{PI} rule) can. Modes
+\emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve 2  problems, whereas
+\emph{Imit}$_{\ior}$ solves one problem uniquely. This result suggests that, even
+though this is not evident from Figure \ref{fig:pi}, sets of problems solved 
+solved by different modes somewhat differ.
+
+Figure \ref{fig:elim-leibniz} gives results of evaluating rules that treat
+Leibniz equality on the calculus level: EL stands for \infname{ElimLeibniz},
+whereas EPV denotes \infname{ElimPredVar}; signs $-$ and $+$ denote that rule is
+removed from or added to configuration b, respectively. Disabling both rules
+severely lowers the success rate. The results suggest that including
+\infname{ElimLeibniz} is beneficial to performance.
+
+\begin{figure}[t]
+  \begin{center}
+    \def\arraystretch{1.1}%
+    \relax{\begin{tabular}{@{}l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1.5em}c}
+      \strut                    & $-$\infname{PI}                 & b$_p$       & b$_f$       & b$_{\iand}$ & b$_{\ior}$  & b$_{\ieq}$  & b$_{\inot}$ & b$_{\iforall\iexists}$ 
+      \MIDLINE
+      $k=1$                     &  \multirow{3}{*}{1636}      &  \textbf{1648}   & 1628    & 1637  & 1634  & 1630 & 1641 & 1637 \\
+      $k=2$                     &                                 &  \emph{1646}   &  1629   & 1636  & 1631  & 1627 & 1638 & 1634 \\
+      $k=8$                     &                                 &  1643   &  1625   & 1633  & 1631  & 1623 & 1637 & 1635
+    \end{tabular}}
+    \caption{Effect of \infname{PI} rule on success rate}
+    \label{fig:pi}
+  \end{center}
+\end{figure}
+\begin{figure}[t]
+  \centering
+  \begin{minipage}{.48\textwidth}
+    \begin{center}
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
+        \strut        & $-$EL       & $+$EL  
+        \MIDLINE
+        $-$EPV        &  1584 (0)   &  1644 (0)   \\
+        $+$EPV        &  1612 (0)   &  \emph{\textbf{1646 (0)}} \\
+      \end{tabular}}
+      \caption{Effect of Leibniz equality elimination rules }
+      \label{fig:elim-leibniz}
+    \end{center}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{.45\textwidth}
+    \begin{center}
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
+        \strut        & $-$BEF       & $+$BEF  
+        \MIDLINE
+        $-$BER        &  1644 (2)   &  1643 (0)   \\
+        $+$BER        &  1645 (0)   &  \emph{\textbf{1646 (0)}}   \\
+      \end{tabular}}
+      \caption{Effect of \infname{BoolER} and \infname{BoolEF} rules}
+      \label{fig:bool-rules}
+    \end{center}
+  \end{minipage}
+\end{figure}
+
+Similarly, Figure \ref{fig:bool-rules} discusses merits of including ($+$) or
+excluding ($-$) \infname{BoolER} (BER) and \infname{BoolEF} (BEF) rules. Our
+expectations were that inclusion of those two rules would make bigger impact on
+success rate. It turned out that, in practice, most of the effects of these
+rules could be achieved using a combination of the \infname{PI} rule and
+superposition rules.
+\pagebreak[2]
+
+\looseness=-1
+Combining these two rules with lazy clausification is more useful: when the rule
+\infname{IC} is replaced by the rule \infname{DC}, the success rate increases
+(compared to 1646 problems solved by $b$) to 1660 problems. We also discovered
+that reasoning with choice is useful: when rule \infname{Choice} is enabled, the
+success rate increases to 1653. We determined that including or excluding the
+conclusion $D$ of \infname{Choice}, after it is simplified, makes no difference.
+Counterintuitively, disabling \infname{BoolSimp} rule results in 1640 problems,
+which is only 6 problems short of configuration $b$. Disabling \infname{Abs} and
+\infname{Interpret-$\lambda$} rules results in solving 25 and 31 problems less,
+respectively. Raw data show that in total, using configurations from Figure
+\ref{fig:cases} to Figure \ref{fig:bool-rules}, 1682 problems can be solved.
+
+Last, we compare our approach to alternatives. Axiomatizing Booleans brings
+Zipperposition down to a grinding halt: only 1106 problems can be solved using
+this mode. On the other hand, preprocessing is fairly competitive: it solves
+only 8 problems less than the $b$ configuration.
+
+\subsection{Comparison with Other Higher-Order Provers}
+
+\looseness=-1
+We compared Zipperposition with all higher-order theorem provers that took part
+in THF division of CASC-27 \cite{gs-19-casc27}: CVC4 1.8 prerelease \cite{cbetal-11-cvc4},
+Leo-III 1.4 \cite{sb-21-leo3}, Satallax 3.4
+\cite{cb-12-satallax}, and Vampire-THF 4.4
+\cite{lkav-13-vampire}. In this part of the evaluation, Zipperposition is ran
+in portfolio mode that runs configurations in different time slices. 
+We set the CPU time limit to 180 s, the time allotted to each prover at CASC-27.
+
+Leo-III and Satallax are cooperative theorem provers -- they periodically invoke
+first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
+iProver \cite{kk-08-iprover} as backends, while Satallax uses Ehoh
+% \cite{EHOH CHAPTER}
+ as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we run
+the cooperative provers in two versions: \emph{pure}, which disables backends,
+and \emph{coop} which uses all supported backends.
+
+\begin{figure}[t]
+  \label{fig:higher-order-provers}
+  \begin{center}
+    \def\arraystretch{1.1}%
+    \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
+      & CVC4             & Leo-III & Satallax              & Vampire & Zipperposition
+      \MIDLINE
+      pure    & 1806 (5) &  1627 (0) & 2067 (0)            & 1924 (7) & 1980 (\phantom{0}0)                \\
+      coop    & --       &  2085 (3) & \textbf{2214 (9)}   & --       & 2190 (17)
+    \end{tabular}}
+    \caption{Comparison with other higher-order provers}
+  \end{center}
+\end{figure}
+
+\looseness=-1
+In both pure and cooperative mode, Satallax is the winner.
+Zipperposition comes in close second, showing that our approach is a promising
+basis for further extensions. Leo-III uses SMT solver CVC4, which features
+native support for Booleans, as a backend. It is possible that the use of CVC4 is
+one of the reasons for massive improvement in success rate of cooperative
+configuration of Leo-III, compared with the pure version. Therefore, we conjecture
+that including support for SMT backends in Zipperposition might be beneficial.

### 2022-03-07 bools/bools.tex
more

@@
 configuration of Leo-III, compared with the pure version. Therefore, we conjecture
-that including support for SMT backends in Zipperposition might be beneficial.
+that including support for SMT backends in Zipperposition might be beneficial.\section{Discussion and Related Work} 
+\label{sect:bool:discussion}
+
+Our work is primarily motivated by the goal of closing the gap between
+higher-order ``hammer'' or software verifier frontends and first-order backends.
+Considerable amount of research effort has gone into making the translations of
+higher-order logic as efficient as possible. Descriptions of hammers like
+HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer
+\cite{pb-12-sh} for Isabelle contain details of these
+translations. Software verifiers Boogie \cite{lr-10-boogie} and Why3
+\cite{bfcp-11-why3} use similar translations.
+
+\looseness=-1
+Established higher-order provers like Leo-III and Satallax perform very
+well on TPTP benchmarks; however, recent evaluations (including the one in Sect.~\ref{sec:ehoh:evaluation}) show that on Sledgehammer
+problems they are outperformed by translations to first-order logic
+\cite{bbtvw-21-sup-lam, %EHOH CHAPTER,
+cbetal-11-cvc4}. Those two provers are built from the ground up as
+higher-order provers -- treatment of exclusively higher-order issues such as
+extensionality or choice is built into them usually using explosive rules. Those
+explosive rules might contribute to their suboptimal performance on mostly
+first-order Sledgehammer problems.
+
+In contrast, the approach taken in this thesis is to start with a first-order prover and gradually extend it
+with higher-order features. The work performed in the context of Matryoshka project~\cite{matryoshka}, 
+in which I participate, resulted in adding support for
+$\lambda$-free higher-order logic with Booleans to E (Chapter~\ref{ch:ehoh})
+and veriT \cite{cbetal-11-cvc4}, and adding support for Boolean-free higher-order
+logic to Zipperposition. Authors of many all state-of-the-art
+first-order provers have implemented some form of support for higher-order reasoning. This
+is true both for SMT solvers, witnessed by the recent extension of CVC4 and veriT~\cite{cbetal-11-cvc4},
+ and for superposition provers, witnessed
+by the extension of Vampire \cite{br-19-restricted-unif}. All of those approaches were
+arguably more focused on functional aspects of higher-order logic, such as
+$\lambda$-binders and function extensionality, than on Boolean aspects such as
+Boolean subterms and Boolean extensionality. A notable exception is work by
+Kotelnikov et al.\ that introduced support for Boolean subterms to first-order
+Vampire \cite{kotelnikov-15-fool,kotelnikov-16-fool}.
+
+The main merit of our approach is that it combines two successful complementary
+approaches to support features of higher-order logic that have not been combined
+before in a modular way. It is based on a higher-order superposition calculus
+that incurs around 1\% of overhead on first-order problems compared with classic
+superposition \cite{bbtvw-21-sup-lam}. We conjecture that it is
+this efficient reasoning base on which the approach is based that contributes to
+its competitive performance.
+
+\section{Conclusion} 
+\label{sect:bool:conclusion}
+
+We presented a pragmatic approach to support Booleans in a modern automatic prover
+for clausal higher-order logic. Our approach combines
+previous research efforts that extended first-order provers with complementary
+features of higher-order logic. It also proposes some solutions for the issues that
+emerge with this combination. The implementation shows clear improvement over previous
+techniques and competitive performance.
+
+What our work misses is an overview of heuristics that can be used to curb the
+explosion incurred by some of the rules described in this paper. In the next chapter,
+we explore exactly this topic.

### 2022-03-08 ehoh/ehoh.tex
more

@@
 first-order unification or matching, and decode the resulting
-substitution. To avoid the overhead, we generalize the first-order unification
+substitution. To avoid overhead, we generalize the first-order unification
 and matching procedures to operate directly on \lfhol{} terms.
@@
 
-The key difference between $\Gfpf$ and $\GfpfRTL$ concerns applied variable.
+The key difference between $\Gfpf$ and $\GfpfRTL$ concerns applied variables.
 Given the sample positions $(\varepsilon, 2, 1)$,
@@
 \noindent
-where $\textsf{sk}$ is a fresh Skolem symbol,
+where $\textsf{sk}$ is a fresh Skolem symbol, and
 $J$ is the largest subset of $\{1, \ldots, n$\} such that $x_j = y_j$ for every $j \in J$.
@@
 The clause selection scheme consists of five queues, each of which is
-specified by a weight function template. The prefixes $n$\texttt{.} next to
+specified by a weight function template. The prefixes $n$\texttt{.}\ next to
 the template names indicate that the queue will be visited $n$ times in the round-robin scheme
@@
 %
-For the moment, we have simply disabled definition unfolding in Ehoh. We will
-enable it again once we have added support for $\lambda$-terms.
+For the moment, we have simply disabled definition unfolding in Ehoh. We have
+not measured the effect of this choice, but we conjecture it is not substantial.
+Namely, appropriately chosen term order will result in a similar outcome without
+removing the defined symbol. We will enable definition unfolding again once we
+have added support for $\lambda$-terms.
 
@@
   % br-20-full-sup-w-combs
-  \cite{br-19-restricted-unif} and of the SMT
-  (satisfiability modulo theories) solvers CVC4 and veriT
+  \cite{br-19-restricted-unif} and of the SMT solvers CVC4 and veriT
   \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
@@
   \cite{ns-13-leo2sh}. Largely thanks to Sledgehammer, Isabelle often came
-  in close second at CASC, even defeating Satallax in 2012
+  in a close second at CASC, even defeating Satallax in 2012
   \cite{gs-13-cascj6}.

### 2022-03-08 introduction/introduction.tex
more

@@
 calculus  for higher-order logic, based on combinators  in Vampire
-\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
+\cite{br-20-full-sup-w-combs}. Authors of SMT (satisfiability modulo theories) solvers followed a similar path:
 they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to

### 2022-03-08 pre/pre.tex
more

@@
 let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
-With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \leftarrow
+With $t|_p$ we denote the subterm of $t$ at position $p$ and with $t[p \leftarrow
 t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
@@
         & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-        $\sigmacl{s \eq t}$ is eligible for paramodulation, just like $\sigmacl{u \eq v}$ and $u|_p$ is not a variable
+        $\sigmacl{s \eq t}$ and $\sigmacl{u \eq v}$ are eligible for paramodulation, and $u|_p$ is not a variable
 \end{tabular}
@@
 \cst{p}(X, Y)$. Applying \infname{EF} to its first two literals yields $\cst{a}
-\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$. This clause
+\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ (with $\sigma = \{ Y \mapsto X \}$). This clause
 then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ using \infname{ER}.
 
-Lastly, superposition rules are used to simulate what mathematicians do when
-they ``replace equals by equals''. Left premise intuitively states that an
+Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what mathematicians do when
+they ``replace equals by equals''. The left premise intuitively states that an
 equation $s \eq t$ holds under condition $C$ (or more precisely, its negation).
 Similarly, the right premise asserts that a (dis)equation holds under a
-condition. Superposition rule simply concatenates the conditions for both
+condition. The superposition rule simply concatenates the conditions for both
 premises and replaces equals by equals in two main inference terms. For example

### 2022-03-09 unif/unif.tex
more

@@
 
-\blfootnote{In this work I designed the main algorithm, solid oracle and
+\blfootnote{In this work I designed the main algorithm, solid oracle, and
 fingerprint indexing. I also implemented and evaluated the algorithms. Visa
@@
 
-Unification is concerned with finding a substitution that makes two terms
-equal, for some notion of syntactic equality. Since the invention of Robinson's first-order unification
-algorithm \cite{ar-65-resolution}, it has become an indispensable
-tool in theorem proving, logic
-programming, natural language processing, programming language compilation, and other areas of computer science.
-
-Many of these applications are based on higher-order formalisms and require
-higher-order unification. Due to its undecidability and explosiveness,
-the higher-order unification problem is considered one of the main
-obstacles on the road to efficient higher-order tools.
+% Unification is concerned with finding a substitution that makes two terms
+% equal, for some notion of syntactic equality. Since the invention of Robinson's first-order unification
+% algorithm \cite{ar-65-resolution}, it has become an indispensable
+% tool in theorem proving, logic
+% programming, natural language processing, programming language compilation, and other areas of computer science.
+
+% Many of these applications are based on higher-order formalisms and require
+% higher-order unification. Due to its undecidability and explosiveness,
+% the higher-order unification problem is considered one of the main
+% obstacles on the road to efficient higher-order tools.
+
+As mentioned in Chapter \ref{ch:intro}, many of the reasoning systems deployed
+in various areas of computer science and matemathics are based on variants of
+higher-order logic. More often than not, they also require unification of
+higher-order terms. Due to its undecidability and explosiveness, the
+higher-order unification problem is considered one of the main obstacles on the
+road to efficient higher-order tools.
 
@@
 Jensen and Pietrzykowski's (JP) procedure \cite{jp-76-unif} is the best known
-procedure for this purpose (Section \ref{sec:unif:background}). Given two terms to
+procedure for this purpose (Sect.~\ref{sec:unif:background}). Given two terms to
 unify, it first identifies a position where the terms disagree.
@@
 
-Building on the JP procedure, we designed a new procedure (Section
-\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Section \ref{sec:unif:proof-of-completeness}).
+Building on the JP procedure, we designed a new procedure (Section\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Sect.~\ref{sec:unif:proof-of-completeness}).
 The new procedure addresses many of the issues that are
@@
 \cite{tl-16-facunif}, and a new fragment 
-we present in Section~\ref{sec:unif:solid-oracle}.
+we present in Sect.~\ref{sec:unif:solid-oracle}.
 %
@@
 
-% In Section \ref{sec:unif:indexing} we discuss a higher-order extension of fingerprint indexing,
+% In Sect.~\ref{sec:unif:indexing} we discuss a higher-order extension of fingerprint indexing,
 % a technique to filter out terms that are not unifiable with a given
@@
 we developed a higher-order extension of fingerprint indexing \cite{ss-12-fp-indexing}
-(Section \ref{sec:unif:indexing}).
+(Sect.~\ref{sec:unif:indexing}).
 We implemented our procedure, several oracles,
 and the fingerprint index in
-the Zipperposition prover (Section \ref{sec:unif:implementation}). Since a
+the Zipperposition prover (Sect.~\ref{sec:unif:implementation}). Since a
 straightforward implementation of the JP procedure already existed in
 Zipperposition, we used it as a baseline to evaluate the performance of our
-procedure (Section \ref{sec:unif:evaluation}). The results show substantial
+procedure (Sect.~\ref{sec:unif:evaluation}). The results show substantial
 performance improvements.
@@
 % This invited article is an extended version of our FSCD-2020 paper \cite{vbn-21-unif}.
-% Most notable extension is the Section \ref{sec:unif:proof-of-completeness}, which gives
+% Most notable extension is the Sect.~\ref{sec:unif:proof-of-completeness}, which gives
 % the detailed proof of completeness of our new procedure. In addition, 
@@
 + 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
-$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, bound
+$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, a bound
 variable, or a constant. In this case, $a$ is called the \emph{head} of the
@@
 and $\varrho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$. By declaring $G$ as an auxiliary variable,
-we focus only the important variables from the intitial problem (in this case only $F$),
+we focus only on the important variables from the initial problem (in this case only $F$),
 which allows $\sigma$ to be an MGU.
@@
 invariant that all substitutions are idempotent. The invariant is easily
-preserved if the substitution $\varrho$ from the definition of
+preserved if the substitution $\varrho$ from the definitions of
 \unifrulename{OracleSucc} and \unifrulename{Bind} is itself idempotent and no variable
@@
     \[ F \mapsto \lambda \overline{x}_n. \, G \, x_{j_1} \, \ldots \, x_{j_i}\]
-    where the fresh free variable $G$ as well as all $x_{j_k}$ are of appropriate type.
+    where the fresh free variable $G$ as well as all $x_{j_k}$ are of appropriate types.
     We call fresh variables emerging from this binding in the role of $G$ 
@@
     variables of arbitrary types. All new variables are of
-    appropriate type. Due to indeterminacy of $\overline{y}$, this step is
+    appropriate types. Due to indeterminacy of $\overline{y}$, this step is
     infinitely branching.
@@
 By including any oracle that supports the first-order occurs check, such as the
-pattern oracle or the fixpoint oracle described in Section \ref{sec:unif:implementation},
+pattern oracle or the fixpoint oracle described in Sect.~\ref{sec:unif:implementation},
 our procedure gracefully generalizes first-order unification: 
@@
 reduces the problem size, and the cases where $\alpha_i$ is a functional type (called \emph{functional
-projection}). We limit the number applications of the following bindings:
+projection}). We limit the number of applications of the following bindings:
 functional projections, eliminations, imitations, and identifications.
@@
 $\lambda \overline{x}.\, F \, \overline{s}_m \unif \lambda \overline{x}.\, G \, \overline{t}_n$, it returns a \emph{trivial unifier} --
-a substitution $\{ F \mapsto \lambda \overline{x}_m.\, H, G \mapsto \lambda \overline{x}_n.\, H \}$, where $H$ is a fresh variable; if the limits are reached and
+a substitution $\{ F \mapsto \lambda \overline{x}_m.\, H, G \mapsto \lambda \overline{x}_n.\, H \}$, where $H$ is a fresh variable; if a limit is reached and
 the constraint is flex-rigid, the oracle fails; if the limits are not reached, 
@@
 Careful tuning of each limit optimizes the procedure for a specific class of problems. 
-For problems originating from proof assistants, low unification depth usually suffices. However, hard
+For problems originating from proof assistants, a low unification depth usually suffices. However, hard
 hand-crafted problems often need
@@
 % \looseness=-1
-The procedure described in Section \ref{sec:unif:the-unification-procedure}
+The procedure described in Sect.~\ref{sec:unif:the-unification-procedure}
 parameterized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
@@
 
-In the remaining of this section, we view terms as
+In the remainder of this section, we view terms as
 $\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-normal
@@
 substitutions to be fully applied. These assumptions are justified because
-all bindings depend only on the head of terms and hence replacing the lazy
+all bindings depend only on the heads of terms and hence replacing the lazy
 transitions $\unifrulename{Normalize}_{\alpha\eta}$,
@@
 \begin{exa} To determine the opponent pairs of $\lamx{\cst{g} \, (\cst{g} x)}$
-  and $\lamx{\cst{g} \, x}$ we first $\alpha$-rename the second term to 
+  and $\lamx{\cst{g} \, x}$, we first $\alpha$-rename the second term to 
   $\lam{y}{\cst{g}\,{y}}$. Then, we identify the common context of $\cst{g} \, (\cst{g} x)$ and $\cst{g} \, y$
@@
 
-  \begin{exa}
+  \begin{exa}[JP E1]
     % \newcommand{\boldboundvar}[1]{\ensuremath{\mathit{\mathbf{#1}}}}
-    Following Example 1 in JP's work \cite{jp-76-unif}, we denote
-    $\omega$-simple occurrences of parameters using boldface. For example,
-    $\lambda xy. \, \cst{f} \, \mathbfit{x} \, \mathbfit{y}$ and $\lamx{\cst{g}
-    \, (\lam{y}{\mathbfit{x}\, y})}$ have several $\omega$-simple occurrences.
-    In the following two examples, boldface occurrences are \emph{not}
-    $\omega$-simple: $\lamx{\cst{g} \, (\lam{y}{\cst{f} \, y \, \mathbfit{x}})},\;
-    \lam{xy}{\cst{h} \, (x \, \mathbfit{y})}$. In the first case $x$ is not applied to parameter $y$ and in the second $y$
-    occurs as an argument to $x$.
+    Term
+    $\lambda xy. \, \cst{f} \, \mathbfit{x} \, \mathbfit{y}$ has two and term
+    $\lamx{\cst{g} \, (\lam{y}{\mathbfit{x}\, y})}$ has one $\omega$-simple
+    occurrence, denoted in boldface. In the following two examples, boldface
+    occurrences are \emph{not} $\omega$-simple: $\lamx{\cst{g} \,
+    (\lam{y}{\cst{f} \, y \, \mathbfit{x}})},\; \lam{xy}{\cst{h} \, (x \,
+    \mathbfit{y})}$. In the first case $x$ is not applied to parameter $y$ and
+    in the second $y$ occurs as an argument to $x$. 
   \end{exa}
@@
     By Lemma \ref{lem:base-simple->inj.}, $\varrho s_{i}$ occurs in
-    $\varrho X\left(\varrho\seq s\right)$ at certain position that depends
+    $\varrho X\left(\varrho\seq s\right)$ at a certain position that depends
     only on $\varrho X$. Similarly $\varrho t_{i}$ occurs in $\varrho X\left(\varrho\seq t\right)=\varrho X\left(\varrho\seq s\right)$
@@
     (excluding occurrences $\lambda x$) of bound variables in $t$,
-    but with the particular exemption: if a prefix variable $u$ 
+    but with a particular exemption: if a prefix variable $u$ 
     has one or more $\omega$-simple occurrences in the body,
@@
     %ASK(VISA): Why is ord(E,\varrho) with parentheses and ord \varrho without them?
-    We write $<$ for the lexicographic comparison of these tuples.
+    We write $<$ for the lexicographic comparison of these triples.
     \end{defi}
@@
     Let $\varrho$ be a unifier of $E$.
-    We must show that there exists
+    We must show that there exist
     a derivation $(E,\makeop{id})\longrightarrow^{*}\sigma$ and a
@@
     contains a
-    unifier $\delta$ of $u$ and $v$ such that there exists $\varrho_{j+1}$
+    unifier $\delta$ of $u$ and $v$ such that there exists a $\varrho_{j+1}$
     and for all free variables $X$ except for the auxiliary variables of the CSU
@@
     Since the auxiliary variables introduced by \unifrulename{OracleSucc} are fresh,
-    they cannot occur in $E_j$ nor in $\sigma_{j} X$ for any $X\not\in V$. Hence,
+    they can occur neither in $E_j$ nor in $\sigma_{j} X$ for any $X\not\in V$. Hence,
     we have $\varrho_{0} X = \varrho_{j}\sigma_{j} X = \varrho_{j+1}\,\delta\,\sigma_{j} X =\varrho_{j+1}\sigma_{j+1} X$
@@
     \begin{enumerate}[beginpenalty=1000]
-    \item Suppose that $\varrho_{j}a$ has a parameter with non-base-simple
+    \item Suppose that $\varrho_{j}a$ has a parameter with a non-base-simple
     occurrence. By one of the induction invariants, $a$ is not an elimination
@@
       comparison form (Definition \ref{unif:def:simple-comparison-form}).
-      By one of the the induction invariants,
+      By one of the induction invariants,
       the free variables occurring in $\varrho_{j}E_{j}$ do not occur in $E_{j}$.
@@
     of the procedure. 
-    This run cannot be infinite because because the measure $\ord(E_j,\varrho_{j})$
+    This run cannot be infinite because the measure $\ord(E_j,\varrho_{j})$
     strictly decreases as $j$ increases.
@@
 The \selected{\text{grayed}} constraints are required to be selected by a given
-selection function $S$. We call $S$ \emph{admissible} if it selects only flex-rigid constraints, prioritizes
-selection of constraints applicable for \textsf{Failure} and \textsf{Decomposition}, 
+selection function $S$. We call $S$ \emph{admissible} if it selects only flex-rigid constraints, and prioritizes
+selection of constraints applicable for \textsf{Failure} and \textsf{Decomposition} 
 and of descendant constraints of
@@
   Compared to the lemma from Prehofer's thesis, our lemma additionally establishes
-  the relationship of unifier $\theta$ of the resulting flex-flex constraint
-  set $E_n$ with preunifier $\sigma$ and the target unifier $\varrho$.
+  the relationship of the unifier $\theta$ of the resulting flex-flex constraint
+  set $E_n$ with the preunifier $\sigma$ and the target unifier $\varrho$.
 
@@
     alter $\varrho$, but removes an equation obtaining $E_1$ which is smaller
-    than $E_0$ and still unifiable by $\varrho$. By induction hypothesis, the preunifier is reachable.
+    than $E_0$ and still unifiable by $\varrho$. By the induction hypothesis, the preunifier is reachable.
     \item 
@@
     Since $\varrho$ unifies $E'_0$, $\varrho_1$ unifies $E_1 = \sigma_1 E'_0$. Clearly, $\varrho_1$ is smaller
-    than $\varrho$. Now we can apply induction hypothesis on $\varrho_1$ and $E_1$, from which 
+    than $\varrho$. Now we can apply the induction hypothesis on $\varrho_1$ and $E_1$, from which 
     we obtain a sequence $E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$, $\theta$ which
     unifies $E_n$ and $\sigma' = \sigma_n \ldots \sigma_2$, such that $\varrho_1 X = \theta\sigma' X$ for all $X \not\in V$.
-    Finally, it is clear that sequence $E_0 \ptarrow{\sigma_1} E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$,
+    Finally, it is clear that the sequence $E_0 \ptarrow{\sigma_1} E_1 \ptarrow{\sigma_2} \cdots \ptarrow{\sigma_n} E_n$,
     $\theta$, and $\sigma = \sigma' \sigma_1$ are as wanted in the lemma statement.
@@
     Therefore, by taking $\sigma = \sigma'\sigma_1$, and prepending $E_0$ to the
-    sequence from the induction hypothesis, we show that the preunifier is reachable. 
+    sequence from the induction hypothesis, we can conclude that the preunifier is reachable. 
     \qedhere
@@
 this result implies that the PT procedure can be used to enumerate finitely many
-elements of complete set of preunifiers for this class of terms.
+elements of a complete set of preunifiers for this class of terms.
 
@@
 pairs actually have an MGU, allowing us to solve the full unification problem,
-rather than preunification problem. Moreover, we lift the order restriction
+rather than the preunification problem. Moreover, we lift the order restriction
 imposed by Prehofer: We identify a class of terms called \emph{solid} which
@@
 To show that this oracle terminates and yields a CSU, we must prove that the PT
-procedure terminates on above described class of problems (Lemma
+procedure terminates on the above described class of problems (Lemma
 \ref{unif:lemma:solid-termination-preunif}). Moreover, we must show how to compute MGUs for
@@
 Toward proving that the PT
-procedure terminates on above described class of problems, we first consider the corresponding matching problem.
+procedure terminates on the above described class of problems, we first consider the corresponding matching problem.
 A \emph{matching problem} is a unification problem
@@
     \textsf{Imitation} and \textsf{Projection} transitions are patterns, which means
-    that, after applying the binding, fresh free variables are applied only to
+    that, after applying a binding, fresh free variables are applied only to
     bound variables or ground base-type terms. The \textsf{Solution} transition effectively
-    replaces the variable with a ground term, which is obviously solid. We show each transition 
+    replaces a variable with a ground term, which is obviously solid. We show each transition 
     either trivially terminates or reduces the measure:
@@
         \overline{x}.\, a \, \overline{t}_n$. If $a$ is a bound variable,
-        and we project $F$ to argument $s_i$ different than $a$, \textsf{Failure} applies and
+        and we project $F$ to argument $s_i$ different from $a$, \textsf{Failure} applies and
         PT trivially terminates. If we project $F$ to $a$, we apply \textsf{Decomposition},
@@
     problem to $\{G_i \,  \overline{s}_k \unif t_i \mid 1\le i\le l\}$. Since the right-hand sides are ground, 
-    any unifier $\sigma$ produced by PT must map all of the variables $G_1, \ldots, G_l$. By induction hypothesis,
+    any unifier $\sigma$ produced by PT must map all of the variables $G_1, \ldots, G_l$. By the induction hypothesis,
     $\sigma G_i$ is ground. Therefore, $\sigma\varrho$ must map $F$ to a ground term.
@@
       any unifier $\sigma$ produced by PT must map all of the variables $G_1, \ldots, G_j$. 
-    By induction hypothesis,
+    By the induction hypothesis,
     $\sigma G_i$ is ground. Therefore, $\sigma\varrho$ must map $F$ to a ground term. \qedhere
@@
   \looseness=-1
-  First, similarly to the proof of previous lemma we conclude each transition
+  First, similar to the proof of previous lemma we conclude each transition
   maintains the condition that the terms remain solid. Second, we have to show
@@
     \end{itemize}
-    \item[Projection] Similarly to the previous transition rule, we have two cases:
+    \item[Projection] Similar to the previous transition rule, we have two cases:
       \begin{itemize}
@@
         \overline{x}.\, a \, \overline{u}_m$ until either \textsf{Failure} is
-        observed (making PT trivially terminating) or until no descendant exists
+        observed (making PT trivially terminating) or no descendant exists
         and the grounding matcher is computed (see Lemmas
@@
         matcher, which will either remove all the free variables in the right-hand
-        side of the constraint (not increasing A and reducing B)
-        or not increasing A and B, reduce C if no free variables occur in the right-hand side.
+        side of the constraint, not increasing A and reducing B,
+        or, if no free variables occur in the right-hand side, reduce C while not increasing A and B.
         \item $\lambda \overline{x}. \, a \, \overline{v}_n \unif \lambda
@@
         if $a$ is a bound variable, projecting $F$ onto argument $u_i$ will either enable
-        application of \textsf{Decomposition} as the next step reducing A, or it will result in \textsf{Failure},
+        application of \textsf{Decomposition} as the next step, reducing A, or it will result in \textsf{Failure},
         trivially terminating. If $a$ is a constant, then projecting $F$ onto some $u_j$
@@
     there is a substitution $\theta$ such that $\varrho X = \theta\sigma X$ for all $X \not = G$. 
-    Due to arbitrary choice of $\varrho$, we conclude that $\sigma$ is an MGU.
+    Due to the arbitrary choice of $\varrho$, we conclude that $\sigma$ is an MGU.
 \end{proof}
@@
   because all other nonauxiliary variables
-  do neither appear in the original problem nor in $\sigma F$ or $\sigma F'$; so we can simply define $\theta X = \varrho X$.
+  appear neither in the original problem nor in $\sigma F$ or $\sigma F'$; so we can simply define $\theta X = \varrho X$.
 
@@
 
-  We prove that ($\star$) implies ($\dagger$) by induction on the size of $v$ and $v'$. We consider
+  We prove that ($\star$) implies ($\dagger$) by induction on the sizes of $v$ and $v'$. We consider
   the following cases:
@@
     $\sigma_i^k = \tau$, i.e., $\textsf{diff}(v,v') = z_i^k \, \textsf{diff}(v_1, v'_1)
-    \ldots \textsf{diff}(v_n, v'_n) $. By induction hypothesis, we get that $(\dagger)$ holds.
+    \ldots \textsf{diff}(v_n, v'_n) $. By the induction hypothesis, we get that $(\dagger)$ holds.
     \item[$v = a \, \overline{v}_n$] In the remaining cases $a$ is either a free variable, a loose bound variable
@@
 However, to extend fingerprint indexing to
-higher-order terms, we must address the issues of $\alpha\beta\eta$-normalization and figure how to cope with $\lambda$-abstractions and
+higher-order terms, we must address the issues of $\alpha\beta\eta$-normalization and figure out how to cope with $\lambda$-abstractions and
 bound variables. To that end, we define a function $\hotofo{t}$, defined on
@@
 $\cstdba{i}$ must be fresh. 
-Effectively, $\hotofo{\;}$ transforms a $\eta$-long $\beta$-reduced higher-order term to an untyped first-order term. Let $\nf{t}{\beta\eta}$ be
+Effectively, $\hotofo{\;}$ transforms an $\eta$-long $\beta$-reduced higher-order term to an untyped first-order term. Let $\nf{t}{\beta\eta}$ be
 the $\eta$-long $\beta$-reduced form of $t$; the higher-order generic
@@
 
+\looseness=-1
 For second-order logic with only unary constants, it is decidable whether
@@
 \emph{monadic second-order}) exists \cite{wf-88-monadicunif}. As this class of terms admits a
-possibly infinite CSU, this oracle cannot be used for \textsf{OracleSucc} step,
+possibly infinite CSU, this oracle cannot be used for \textsf{OracleSucc},
 but it can be used for \textsf{OracleFail}. Similarly the fragment of
@@
 \looseness=-1
-Contrary to first-order unification, there is no widely available corpus of
-benchmarks designed solely for evaluating performance of higher-order
+Contrary to first-order unification, there is no widely available benchmark set 
+designed solely for evaluating performance of higher-order
 unification algorithms. Thus, we used all 2606 monomorphic higher-order theorems
@@
 Many TPTP problems require synthesis of complicated unifiers,
-whereas Sledgehammer problems are only mildly higher-order -- many of them are
+whereas SH problems are only mildly higher-order---many of them are
 solved with first-order unifiers.
@@
 
-Counterintuitively, in some cases, using oracles can hurt the performance of
+In some cases, using oracles can hurt the performance of
 Zipperposition. Using oracles typically results in generating smaller
@@
 is slightly less beneficial than the fixpoint or pattern oracles, but still presents a
-useful addition the set of available oracles.
+useful addition to the set of available oracles.
 
@@
 faster than other competitive higher-order provers.
-  There are 11 benchmarks in \texttt{NUM} category of TPTP that contain conjectures about
+  There are 11 benchmarks in the \texttt{NUM} category of TPTP that contain conjectures about
   Church numerals: \texttt{NUM020\^{}1}, \texttt{NUM021\^{}1}, \texttt{NUM415\^{}1},
@@
   shows the CPU time needed to solve a problem or ``--'' if the prover timed out.
+  \pagebreak[2]
 
@@
 preunifiers. The most prominent procedure in this line of research is Huet's 
-\cite{gh-75-unification}. The Snyder--Gallier procedure restricted not to solve flex-flex pairs
-is a version of the PT procedure presented in Section \ref{sec:unif:solid-oracle}. It improves 
+\cite{gh-75-unification}. The Snyder--Gallier procedure restricted to not solving flex-flex pairs
+is a version of the PT procedure presented in Sect.~\ref{sec:unif:solid-oracle}. It improves 
 Huet's procedure by featuring a \textsf{Solution} rule.

### 2022-03-09 unif/unif.tex
MORE

@@
 As mentioned in Chapter \ref{ch:intro}, many of the reasoning systems deployed
-in various areas of computer science and matemathics are based on variants of
+in various areas of computer science and mathematics are based on variants of
 higher-order logic. More often than not, they also require unification of

### 2022-03-09 scratch/paper.tex
more

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required

%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
\documentclass[smallcondensed,draft]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage[utf8]{inputenc}
\usepackage[misc]{ifsym}
\usepackage[final]{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{cite}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{prftree}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumerate}
\usepackage{stmaryrd}
\usepackage{booktabs}
\usepackage{algorithmicx}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage[
   a4paper,
   pdftex,
   pdftitle={Making Higher-Order Superposition Work},
   pdfauthor={Petar Vukmirovi\'c, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

\usepackage{mathptmx}      % use Times fonts if available on your TeX system

\DeclareSymbolFont{letters}{OML}{txmi}{m}{it} %% for Greek

% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{Journal of Automated Reasoning}

\makeatletter

\def\vthinspace{\kern+0.083333em}
\newcommand\CORR{{(\vthinspace\Letter\vthinspace)}}

\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}

% mathcal
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}
%\DeclareMathAlphabet{\mathcalx}{OMS}{zplm}{m}{n}

%%% HACK Springer's style to get bold subsection headings --JB
\def\subsection{\@startsection{subsection}{2}{\z@}%
    {-21dd plus-8pt minus-4pt}{10.5dd}
     {\normalsize\bfseries\boldmath}}

\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
\spnewtheorem{examplex}[theorem]{Example}{\bfseries}{\rmfamily}

\begin{document}

\title{Making Higher-Order Superposition Work}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Petar~Vukmirović~\orcid{0000-0001-7049-6847}\and
Alexander~Bentkamp~\orcid{0000-0002-7158-3595}\and
Jasmin~Blanchette~\orcid{0000-0002-8367-0936}\and
Simon~Cruanes~\orcid{0000-0003-3969-5850}\and
Visa~Nummelin~\orcid{0000-0003-0078-790X} \and
Sophie~Tourret~\orcid{0000-0002-6070-796X}}

\authorrunning{Petar Vukmirović et al.} % if too long for running head

\institute{
  Petar~Vukmirovi\'c \CORR \and
  Alexander Bentkamp\and
  Jasmin Blanchette\and
  Visa Nummelin\at%
    Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\
    \email{\{p.vukmirovic,a.bentkamp,j.c.blanchette,visa.nummelin\}@vu.nl}
  \and
  Jasmin Blanchette\and Sophie Tourret \at%
    Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France
    \\
    \email{\{jasmin.blanchette,sophie.tourret\}@inria.fr}
  \and
  Jasmin Blanchette\and Sophie Tourret \at  Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken,~Germany \\
  \email{\{jblanche,stourret\}@mpi-inf.mpg.de}
  \and
  Simon Cruanes \at Imandra, Austin, Texas, USA \\
  \email{simon@imandra.ai}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Superposition is among the most successful calculi for first-order logic. Its
extension to higher-order logic introduces new challenges such as infinitely
branching inference rules, new possibilities such as reasoning about
Booleans, and the need to curb the explosion of specific higher-order rules. We
describe techniques that address these issues and extensively evaluate their
implementation in the Zipperposition theorem prover. Largely thanks to their use,
Zipperposition won the higher-order division of the CASC-J10 competition.
\end{abstract}

\newcommand{\ifalse}{\pmb\bot}
\newcommand{\itrue}{\pmb\top}
\newcommand{\inot}{\pmb{\neg\,}}
\newcommand{\iand}{\pmb\land}
\newcommand{\ior}{\pmb\lor}
\newcommand{\iimplies}{\pmb\rightarrow}
\newcommand{\iequiv}{\pmb\leftrightarrow}
\newcommand{\iforall}{\pmb\forall}
\newcommand{\iexists}{\pmb\exists}
\newcommand{\ieq}{\pmb\approx}
\newcommand{\ineq}{\pmb{\not\approx}}

\newcommand{\comb}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\neglit}[1]{\ensuremath{#1 \eq \ifalse}}
\newcommand{\poslit}[1]{\ensuremath{#1 \eq \itrue}}
\newcommand{\skbci}{\textsf{SKBCI}}
\def\negvthinspace{\kern-0.083333em}

\newcommand{\cst}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\var}[1]{{\mathit{#1}}}
\newcommand{\typ}[1]{{\mathit{#1}}}
\newcommand\foralltynospace[1]{\mathsf{\Pi}#1.}
\newcommand\forallty[1]{\foralltynospace{#1}\;}
\newcommand{\typeargs}[1]{{\negvthinspace\langle#1\rangle\negvthinspace}}
\let\oldDelta=\Delta
\let\oldSigma=\Sigma
\renewcommand\Sigma{\mathrm{\oldSigma}}
\newcommand{\Sigmaty}{\Sigma_\mathsf{ty}}
\newcommand{\VV}{\mathscr{V}}
\newcommand{\Vty}{\VV_\mathsf{ty}}

% Equality predicate
\newcommand{\eq}{\approx}
\newcommand{\noteq}{\not\eq}
\newcommand{\lland}{\mathrel\land}
\newcommand{\llor}{\mathrel\lor}
\newcommand{\ccup}{\mathrel\cup}
\newcommand{\ccap}{\mathrel\cap}

\newcommand{\tuple}[2]{\overline{#1}_{#2}}
\newcommand{\tuplen}[1]{\overline{#1}_{n}}

\newcommand{\lam}[2]{\ensuremath{\lambda #1.\> #2}}
\newcommand{\lamx}[1]{\lam{x}{#1}}

% Names of inference rules
\newcommand{\infname}[1]{\textsc{#1}}

% Inference rule
\newcommand{\namedinference}[3]{\prftree[r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\inference}[2]{\namedinference{}{\strut#1}{\strut#2}}

% Simplification rule
\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\simp}[2]{\namedinference{}{\strut#1}{\strut#2}}
\newcommand{\lamsupraw}{$\lambda$-super\-po\-si\-tion}
\newcommand{\lamsup}{{\color{black}\lamsupraw}}
\newcommand{\ofreelamsup}{{\color{black}Boolean-free \lamsupraw}}
\newcommand{\olamsup}{{\color{black}Boolean \lamsupraw}}

\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
\newcommand{\ourparagraph}[1]{\paragraph{\upshape\bfseries#1}}

\hyphenation{su-per-po-si-tion Zip-per-posi-tion tab-leau tab-leaux Bent-kamp ab-strac-tion
  Bool-ean}

\newcommand{\confrep}[2]{#2}

\section{Introduction}
\label{sec:intro}

%In recent decades,
Superposition-based first-order automatic theorem provers
have emerged as useful reasoning tools. They dominate at the annual CASC
\cite{gs-2016-casc} theorem prover competition, having always won the
first-order theorem division. They are also used as backends to proof assistants
\cite{ck-18-coqhammer,ku-15-holyhammer,pb-12-sh}, automatic
higher-order theorem provers \cite{sb-21-leo3}, and software verifiers
\cite{fp-13-why3}.

The superposition calculus has only recently been extended
to higher-order logic (more precisely, extensional simple type theory
\cite{henkin-1950-completeness}), resulting in
\emph{\ofreelamsup} \cite{bbtvw-21-sup-lam}, which we developed
together with Waldmann, as well as \emph{combinatory superposition}
\cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
calculi do not support an interpreted Boolean type,
they can be extended by ad hoc rules \cite{our-bool-paper} that support
most of the Boolean reasoning necessary in practice.

Both higher-order superposition calculi were designed to gracefully
extend first-order reasoning. As most steps in higher-order
proofs tend to be essentially first-order, extending the most successful first-order
calculus to higher-order logic seemed worth trying.
Our first attempt at testing this idea was in 2019:
Zipperposition~1.5, based on \ofreelamsup, finished third
in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
12~percentage points behind the winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

Studying the competition results, we found that higher-order tableaux have
some advantages over higher-order superposition. To bridge the gap, we developed
techniques and heuristics that simulate tableaux in the
context of saturation. We implemented them in Zipperposition~2, which took part
in CASC-J10 \cite{gs-21-cascj10} in 2020. This time, our prover won the division, proving 84\% of
the problems, a whole 20~percentage points ahead of the runner-up, Satallax
3.4.

In this article, we describe the main techniques that explain this reversal
of fortunes. They cover most parts of a modern higher-order theorem prover, from
preprocessing to additional calculus rules to heuristics to backend
integration. We use a newer version of Zipperposition, based on a newer
calculus:\ Instead of \ofreelamsup{} augmented with ad hoc Boolean rules,
we work with \emph{\olamsup} \cite{bbtv-21-full-ho-sup}, a principled extension of superposition to full
higher-order logic, including an interpreted Boolean type.

Many higher-order problems extensively use symbol definitions to simplify
their representation. We describe several ways to exploit the definitions,
%of which the most successful is
such as turning them into rewrite rules (Sect.~\ref{sec:preprocessing}).
%Interesting patterns can be observed in various higher-order problem encodings.
%We show how we can exploit these to simplify problems (Sect.~\ref{sec:preprocessing}).
%
By working on formulas rather than clauses, tableau techniques take a more
holistic view of a higher-order problem.
Through its support for delayed clausification and, more generally,
calculus-level formula manipulation, \olamsup{} enables us to
simulate most successful tableau techniques in a saturating prover
(Sect.~\ref{sec:formulas}). This calculus also supports \emph{Boolean selection
functions}, a mechanism that allows us to choose on which Boolean subterms
to perform inferences first.
We implemented some Boolean selection functions and
evaluated them (Sect.~\ref{sec:bool-select}).

The main drawback of both \lamsup{} variants compared with combinatory
superposition is that they rely on rules that enumerate possibly infinite sets
of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:infinite-branching}). The prover
retains the same behavior as
before on first-order problems, smoothly scaling with
increasing numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:explosiveness}).

Using first-order backends to finish the proof is common practice in
higher-order reasoning. Since \lamsup{} coincides with standard
superposition on first-order clauses, invoking backends may
seem redundant; yet Zipperposition is nowhere as efficient as E
\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
efficient backend does make sense. We describe how to achieve a balance
between allowing native higher-order reasoning and
delegating reasoning to a backend (Sect.~\ref{sec:backends}).
%
Finally, we compare Zipperposition~2 with other provers on all monomorphic
higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
evaluation than at CASC (Sect.~\ref{sec:comparison}). Our evaluation
corroborates the competition results.

This article is an extended version of a paper accepted at CADE-28
\cite{making-ho-work}. Compared with the conference paper, it
describes a new preprocessing technique, explores the effects of Boolean
selection functions, evaluates %some previously unevaluated
more techniques,
introduces new benchmark sets, and presents more examples.

\section{Background and Setting}
\label{sec:background}

We focus on monomorphic higher-order logic, without the axiom of infinity 
or the axiom of at least two individuals. However, the techniques can easily be
extended with polymorphism. Indeed, Zipperposition already supports some
of them polymorphically.

\ourparagraph{Higher-Order Logic}
We define terms $s, t, u, v$ inductively as free variables $F, X$, bound
variables $x, y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak
\cst{a},\allowbreak \cst{b}, \dotsc$, term applications $s \, t$, and
$\lambda$-abstractions $\lamx{s}$. The syntactic distinction between free and
bound variables yields \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
\cst{a}}$) \cite{tn-93-patterns}.
%We shorten iterated application $s \, t_1 \, \cdots \, t_n$ to $s \,
%\overline{t}_n$ and $\lambda$-abstraction $\lambda x_1. \, \cdots \, \lambda
%x_n. \> s$ to $\lam{\overline{x}_n}{s}$.
We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. The $n$-fold application of
a unary term $s$ to a term $t$ is denoted by $s^n \, t$. Every
$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If the type of a term $t$ is of the form $\tau_1 \to \cdots \to
\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
call $t$ a \emph{predicate}. A term of type $o$ is called a \emph{formula}.

A literal $l$ is an equation $s \eq t$ or a disequation $s \not\eq t$. A clause is
a finite multiset of literals, interpreted and written disjunctively $l_1 \llor
\cdots \llor l_n$. Logical symbols that may occur within terms are written in
boldface: $\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Quantified
formulas are expressed using (a type-indexed
family of) constants $\iforall$ and $\iexists$ as $\iforall \,
(\lambda x. \, t)$ and $\iexists \, (\lambda x. \, t)$, usually abbreviated to
$\iforall x.\,t$ and $\iexists x.\,t$. Following %the convention of
\olamsup{}, predicate literals are encoded as equations with $\itrue$ or
$\ifalse$: for example, $\cst{even}(x)$ becomes $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ becomes $\neglit{\cst{even}(x)}$.

\ourparagraph{Higher-Order Calculi}
\looseness=-1
The \olamsup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
complete inference system and redundancy criterion for higher-order logic with
rank-1 polymorphism, Hilbert choice, and functional and Boolean extensionality.
The calculus relies on
\emph{complete sets of unifiers}
(\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
$V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
\in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
for all variables $X \in V$. The set $V$ is used to distinguish
important variables from auxiliary variables (which may arise in intermediary
states of the unification procedure). We usually omit it.

Unlike \ofreelamsup, this calculus 
does not require axioms defining the logical symbols to cope with formulas.
Instead, it includes Boolean inference rules that mimic
superposition from such axioms into Boolean subterms,
while avoiding the explosion incurred by adding these axioms to the proof state. It
also includes rules that simulate Boolean inferences below applied variables.
Both sets of rules are disabled or replaced with incomplete, ad hoc rules
described by Vukmirović and Nummelin \cite{our-bool-paper} in most configurations
of the CASC portfolio.
A new feature of the calculus that we explore in detail is
the ability to select Boolean subterms
to restrict Boolean and superposition inferences.

In contrast to both \lamsup{} variants, combinatory superposition can
avoid enumerating CSUs by
using a form of first-order unification.
Essentially, it enumerates higher-order terms
using rules that instantiate applied variables with partially applied
combinators from the complete combinator set $\{\cst{S}, \cst{K}, \cst{B},
\cst{C}, \cst{I}\}$. This calculus is the basis of Vampire~4.5
\cite{br-20-full-sup-w-combs}, which finished
closely behind Satallax 3.4 %~and~3.5
at CASC-J10.

\looseness=-1
A different, very successful calculus is Satallax's SAT-guided tableaux
\cite{backes-brown-2011}. Satallax was the leading higher-order prover of the
2010s. Its simple and elegant tableaux avoid deep superposition-style rewriting
inferences.
Nevertheless, our working hypothesis for the past six years has been
that superposition would likely provide a stronger basis for higher-order
reasoning.
Other competing higher-order calculi include SMT (implemented in CVC4
\cite{brotb-19-ho-smt, cbetal-11-cvc4}) and extensional paramodulation (implemented in Leo-III \cite{sb-21-leo3}).

\ourparagraph{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
theorem prover that implements both \lamsup{} variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.

Several of our techniques extend the \emph{given clause procedure}, the standard
saturation procedure pioneered by McCune and Wos \cite[Sect.~2.3]{mcw-1997-otter}. It partitions the proof
state into a set $P$ of \emph{passive} clauses and a set $A$ of \emph{active}
clauses. Initially, $P$ contains all input clauses, and $A$ is empty. At each
iteration, a \emph{given} clause is moved from $P$ to $A$ (i.e., it is
\emph{activated}), all inferences between it and clauses in $A$ are performed,
and the conclusions are added to $P$. Because Zipperposition fully simplifies
clauses only when they are activated, it implements a DISCOUNT-style loop
\cite{adf-1995-discount}.

\ourparagraph{Experimental Setup}
To assess our techniques, we carried out experiments with Zipperposition~2. We
used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
and 1253 Sledgehammer-generated
monomorphic higher-order problems (labeled \emph{SH}).
Although some techniques support polymorphism, we
uniformly used monomorphic benchmarks.

We fixed a \emph{base} configuration
of Zipperposition parameters as a baseline for all comparisons. This is an
incomplete, pragmatic configuration of \olamsup{} using heuristics expected to perform
well on a wide range of problems.
%Note that
%we use a different baseline configuration than in our earlier paper \cite{making-ho-work}.
%%% That goes without saying. We use a different calculus! Now clarified in intro. --JB
In each experiment, we varied
the parameters associated with a specific technique to evaluate it. The
experiments were run on StarExec Miami \cite{sst-14-starexec} servers, equipped with
Intel Xeon E5-2620 v4 CPUs clocked at 2.10 GHz. Unless otherwise stated, we used a
CPU time limit of 15~s, roughly the time each configuration is given in the
CASC portfolio mode. The raw evaluation results are available online.%
\footnote{\url{http://doi.org/10.5281/zenodo.5007440}}

\section{Preprocessing Higher-Order Problems}
\label{sec:preprocessing}

The TPTP library contains thousands of higher-order problems. Despite their
diversity, they have a markedly different flavor from the TPTP first-order
problems. Notably, they extensively use the \verb|definition| role to identify
universally quantified equations (and equivalences) that define symbols.
%
Definitions $s \eq t$ (or $(s \iequiv t) \eq \itrue$) can be replaced by rewrite
rules $s \rightarrow t$,
using the orientation given in the input problem. If there are multiple
definitions for the same symbol, only the first one is replaced by a rewrite rule.
Then, whenever a clause is picked in the given clause procedure, it will be rewritten
using the collected rules.
Alternatively, we can rewrite
the input formulas as a preprocessing step. This ensures that the input
clauses will be fully simplified when the proving process starts and no
defined symbols will occur in clauses, which usually helps the heuristics.

Since the TPTP format enforces no constraints on
definitions, rewriting might diverge. To ensure
termination, we limit the number of applied rewrite steps. In
practice, most TPTP problems are well behaved: Only one
definition is given for each symbol, and the definitions are acyclic.

Turning the defining equations into rewrite rules, unfolding the definitions, and
$\beta$-reduc\-ing the result can eliminate all of a problem's higher-order features, making
it amenable to first-order methods. However, this can inflate the problem
beyond recognition and compromise the refutational completeness of
superposition.

\begin{examplex} 
  Removing higher-order features of a problem can have adverse effects.
  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
  successor constructor. When this definition is kept as is, the
  prover can superpose from either $\cst{m}$ or its definition into the
  (clausified) induction axiom, which is also given in the problem, and quickly prove
  the conjecture, without using any advanced inductive reasoning. In contrast,
  when the definition is
  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
  correct instantiation for the induction axiom.
\end{examplex}

We describe two techniques to mitigate these issues. The first one is based on the observation that in practice,
the explosion associated with definition unfolding mostly
manifests itself on definitions of nonpredicate symbols. In some cases, it is
preferable to rely on superposition's term order and powerful simplification
engine to rewrite the proof state rather than to blindly rewrite definitions. On
the other hand, superposition's reasoning with equivalences is often inadequate
\cite{bbtv-21-full-ho-sup, gs-05-boolsup}. Thus, it makes sense to treat only
predicate definitions as rewrite rules.

The second technique aims at preserving completeness: We can try to force the term order that
parameterizes superposition to orient as many definitions as possible and rely on
demodulation to simplify the proof state. Usually, the Knuth--Bendix order (KBO)
\cite{db-1970-kbo} is used. It compares terms by first comparing their weights,
which is the sum of all the weights assigned to the symbols it contains. Given a
symbol weight assignment $\mathcal{W}$, we can update it so that it orients
acyclic definitions from left to right assuming that they are of the form $
\cst{f} \, \overline{X}_m \eq \lambda \overline{Y}_n. \, t$, where the only free
variables in $t$ are $\overline{X}_m$, no free variable is repeated or appears
applied in $t$, and $\cst{f}$ does not occur in $t$. Then we traverse the
symbols $\cst{f}$ that are defined by such equations following the dependency
relation, starting with a symbol $\cst{f}$ that does not depend on any other
defined symbol. For each $\cst{f}$, we set $\mathcal{W}(\cst{f})$ to $w + 1$,
where $w$ is the maximum weight of the right-hand sides of $\cst{f}$'s
definitions, computed using $\mathcal{W}$. By construction, for each equation
the left-hand side is heavier. Thus, the equations are orientable from left to
right.

\begin{examplex} 
  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
  in higher-order logic. More complex modal operators (such as
  implication and equivalence) are defined in terms of basic connectives (such as negation
  and disjunction). Some of the definitions present in the problems are
  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
  variables, and basic connectives is 2, we can orient equations using
  the above described approach as follows. Starting from symbols that do not
  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
  enable us to orient all definitions from left to right.
\end{examplex}

% Many higher-order problems, especially those generated from proof assistants,
% contain hundreds of needless axioms.
% To filter out axioms that are unlikely to be useful in a proof attempt, many
% theorem provers rely on the SInE algorithm \cite{hv-2011-sine}. SInE starts with
% the set of symbols occurring in the conjecture and tries to find axioms that
% define the properties of these symbols. Then it looks for the definitions of
% newly found symbols until it reaches a fixpoint. Axioms annotated with
% \verb|definition| ease this search because they explicitly record the
% dependency between a symbol and its characterization. To exploit
% this information, we modified SInE to optionally include the definitions of
% symbols in the conjecture, regardless of whether they are filtered out or not.
% Similarly, we implemented mode of SInE which selects only conjecture and axioms
% annotated with \verb|definition|.

\ourparagraph{Evaluation and Discussion}

% The \textit{base} configuration treats all axioms annotated with
% \texttt{definition} as rewrite rules applied as preprocessing.
% In addition, we tested
We designed and evaluated the following strategies for handling
\texttt{definition} axioms:

\begin{enumerate}[\rm no-RW$+$KBO~]
  \item[\rm pre-RW~] rewrite all definitions as a preprocessing step;
  \item[\rm in-RW~] rewrite all definitions during the saturation, as an inprocessing step;
  \item[\rm $o$-RW~] rewrite only predicate definitions, during preprocessing;
  \item[\rm $o$-RW$+$KBO~] like $o$-RW but with adjusted KBO weights for the remaining
    definitions;
  \item[\rm no-RW~] no special treatment of definitions;
  \item[\rm no-RW$+$KBO~] like $no$-RW but adjusting KBO weights for all definitions.
\end{enumerate}

The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
cell gives the number of proved problems, and cells marked with $\star$
correspond to the base configuration. The highest number in a category is typeset in
\relax{bold}. SH benchmarks are not
included because they do not contain the \texttt{definition} role.

\newcommand{\unknownres}{\ensuremath{{\varnothing}}}
\newcommand{\colalign}{\phantom{0}}

\begin{figure}[t]
  \centering
  \def\arraystretch{1.1}%
  \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
                & pre-RW                 & in-RW        & $o$-RW               & $o$-RW$+$KBO    & no-RW        & no-RW$ + $KBO  \\ \midrule
           TPTP & {\bf 1635}$^\star$     & 1619         & 1620                 & 1621            & 1298         & 1296
           \\ \bottomrule
         \end{tabular}}
       \captionof{figure}{Impact of the definition rewriting method}
       \label{fig:rewrite}
     \end{figure}

% {\tt Points to discuss:
% \begin{itemize}
%   \item On TPTP problems, there are only 2 problem proved by no-RW$+$KBO but not no-RW --- both after 11s: probably because of some strange effects
%   \item On TPTP problems, there are 4 problems proved by o-RW$+$KBO but not o-RW --- 2 of them under 1 second: suppressing superposition into rhs of definitions
%   \item 13 problems (some of them of rating 1) proved by oRW but not regular one
%   %  zip-jar-hot_defs_oRW(13) : NUM/NUM638^4.p,SEU/SEU800^1.p,NUM/NUM701^4.p,NUM/NUM699^4.p,SEU/SEU603^1.p,NUM/NUM647^4.p,SEU/SEU695^2.p,SEU/SEU630^2.p,NUM/NUM648^4.p,SEU/SEU540^1.p,SYO/SYO531^1.p,SEU/SEU594^1.p,SEU/SEU648^1.p
%   \item SH problems do not use definition tag, so this is just noise.
% \end{itemize}
% }

%%% @PETAR: I find this para very confusing. I don't know when you're talking
%%% about the old paper's numbers and the present one's. Lert's keep it simple
%%% and avoid mentions to the old paper. It's subsumed.
%%%
%%% Also, careful with words like "counterintuitively". What's counterintuitive
%%% to you might be intuitive to the reader. E.g. my experience is that hardly
%%% any well-meant change to a prover makes a big difference on performance,
%%% so I certainly wasn't surprised.
%
%In our earlier paper \cite{making-ho-work}, only an
%approximation of KBO weight adjustment was implemented: no checks for loops and
%no topological sorting was performed. For this article, we implemented the
%feature as described above. Counterintuitively, this did not substantially
%influence the performance of Zipperposition:

The four configurations in which definitions are treated as rewrite rules
perform much better than the other two. In contrast, adjusting KBO weights gives
no substantial improvement: Looking at raw data, we found only
\NumberOK{2}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
feature was used in the proof. For no-RW and no-RW$+$KBO, the
\NumberOK{2}-problem difference may be just noise. Even though it proves fewer
problems, the configuration $o$-RW has some advantages over pre-RW: It proves
\NumberOK{16} problems that pre-RW does not, \NumberOK{3} of which have a TPTP
difficulty rating (the ratio of eligible %TPTP
provers that cannot prove the problem) of~1.

% \looseness=-1
Rewriting after clausification avoids getting stuck rewriting parts of the
proof state that might not contribute to the proof. In practice, we noticed that
rewriting can be so expensive that the prover can spend all
allotted CPU time in the preprocessing phase. The evaluation results confirm this
observation: There are \NumberOK{64} problems proved by in-RW but not by
pre-RW. Moreover, there are \NumberOK{41} problems that can be proved only
by in-RW but not by any other above described configuration. % from Fig.~\ref{fig:rewrite}.

\section{Reasoning about Formulas}
\label{sec:formulas}

Higher-order logic identifies formulas with terms of Boolean type. To prove a problem, we often
need to instantiate a variable with the right predicate.
Finding this predicate can be easier if the problem is not clausified.
Consider the conjecture $\iexists f. \, f \, \cst{p} \, \cst{q} \iequiv \cst{p}
\iand \cst{q}$. Expressed in this form, the formula is easy to prove by taking
$f := \lambda x \, y. \> x  \iand y$. By contrast, guessing the right
instantiation for the negated, clausified form $ \neglit{F \, \cst{p} \,
\cst{q}} \llor \neglit{\cst{p}} \llor \neglit{\cst{q}},
\poslit{F \, \cst{p} \, \cst{q}} \llor \poslit{\cst{p}}$, $\poslit{F \, \cst{p}
\, \cst{q}} \llor \poslit{\cst{q}}$ is more challenging.
One of the strengths of higher-order tableau provers is that they do not clausify the input
problem. This might partly explain Satallax's dominance in the THF division of CASC
competitions until CASC-J10.

% Traditionally, superposition provers clausify the input problem, simplify the
% initial clause set, and start the saturation loop. We propose to integrate
% clausification tightly in the saturation loop.
% %
% Instead of clausifying the input, we represent the input formula $\varphi$
% as a higher-order clause $\poslit{\varphi}$. Then we use Vukmirović and Nummelin's
% \emph{lazy clausification} rules \cite[Sect.~3.4]{our-bool-paper},
% which extend \ofreelamsup{}. These incrementally clausify top-level
% logical symbols; for example, a clause $C' \llor \neglit{(\cst{p} \iand \cst{q})}$ yields a
% new clause $C' \llor \neglit{\cst p} \llor \neglit{\cst q}$.
% Lazy clausification rules admit advanced
% forms of formula renaming and Skolem symbol reusing that are not available in
% standard clausification algorithms.
% \looseness=-1
The \olamsup{} calculus supports \emph{delayed clausification rules}
that insert problems into the proof state in their original,
nonclausified form, and clausify them gradually. 
Delayed clausification
allows the prover to analyze the syntactic structure of formulas during saturation,
whereas the more traditional approach of \emph{immediate clausification}
applies a standard clausification algorithm \cite{nw-01-small-cnf}
both as a preprocessing step and whenever predicate variables are instantiated.

An earlier evaluation of the \olamsup{} calculus \cite{bbtv-21-full-ho-sup} showed
that the \emph{outer} variant of delayed clausification substantially increases this calculus's performance.
The outer variant clausifies
top-level logical symbols, proceeding from the outside inwards;
for example, a clause $C \llor \neglit{(\cst{p} \iand \cst{q})}$ is transformed into $C \llor
\neglit{\cst p} \llor \neglit{\cst q}$. The calculus also supports \emph{inner} delayed
clausification, which uses only the core calculus rules to clausify problems.
Even though this is the laziest approach to clausification, the earlier
evaluation showed that this approach is inefficient. Thus, we focus only on the
outer rules.

\looseness=-1
Delayed clausification rules can be used as inference rules (which add conclusions
to the passive set) or as simplification rules (which delete premises and add
conclusions to the passive set).
%
Inferences give more flexibility, since all
intermediate clausification states will be stored in the proof state, at the
cost of producing many clauses. Simplifications produce fewer clauses,
but risk destroying informative syntactic structure.
Since clausifying equivalences can destroy a lot of syntactic structure
\cite{gs-05-boolsup}, we never apply simplifying rules
on them.

Delayed clausification can interfere with clause splitting techniques.
Zipperposition supports a lightweight variant of AVATAR \cite{av-2014-avatar},
an architecture that partitions the search space by
splitting clauses into variable-disjoint subclauses. This lightweight AVATAR is described
by Ebner et al.\ \cite[Sect.~7]{2021-ebt-unifying-splitting}. Combining it
with delayed clausification makes it possible to split a %higher-order
clause $(\varphi_1 \ior \cdots \ior \varphi_n) \eq \itrue$, where
the $\varphi_i$'s are arbitrarily complex formulas that share no free
variables with each other, into clauses $\varphi_i \eq \itrue$.
%
To finish the proof, it suffices to derive the empty clause under each assumption
$\varphi_i \eq \itrue$. Since the split is performed at the formula level, this
technique resembles tableaux, but it exploits the strengths of superposition,
such as its powerful redundancy criterion and simplification machinery, to
close the branches.

\newcommand{\instset}{\ensuremath{\mathit{Inst}}}
% When delayed clausification is used, the prover can make
% heuristic choices based on information obtained from the formulas on which
% clausification rules are applied.
% In particular, we look for $\lambda$-abstractions whose bodies are formulas
% in each active clause and store them in a set
% $\mathit{Inst}$. Optionally, $\mathit{Inst}$ can contain \emph{primitive
% instantiations} \cite{our-bool-paper}---that is, imitations (in the sense
% of higher-order unification) of logical symbols that approximate the shape of
% a formula that can instantiate a predicate variable. These instantiations resemble
% the ones tableau provers perform during the proof search.

Beyond splitting, interleaving clausification and saturation allows us to simulate another tableau-inspired
technique. Whenever dynamic clausification substitutes a fresh variable $X$ for
a predicate variable $x$ in a clause of the form $(\iforall x.\, \varphi) \eq
\itrue \llor C$, yielding $\varphi\{x \mapsto\nobreak X\}
\eq \itrue \llor C$, we can create additional clauses in which $x$ is replaced
with $t \in \instset$, where $\instset$ is a set of heuristically chosen terms.
This set contains $\lambda$-abstractions whose bodies are formulas and that
occur in activated clauses, and \emph{primitive instantiations}
\cite{our-bool-paper}---that is, imitations (in the sense of higher-order
unification) of logical symbols that approximate the shape of a predicate that
can instantiate a predicate variable.

Since a new term $t$ can be added to $\mathit{Inst}$ after a clause with a
quantified variable of $t$'s type has been activated, we 
remember the clauses $\varphi\{x \mapsto X\} \eq \itrue\allowbreak \llor C$ and instantiate
them when $\mathit{Inst}$ is extended.
Conveniently, these instantiated clauses are not recognized as subsumed by
Zipperposition, which uses an optimized, incomplete higher-order subsumption
algorithm.

Given a disequation $\cst{f}\,\tuple{s}{n} \not\eq \cst{f}\,\tuple{t}{n}$, the
\emph{abstraction} of $s_i$ is $\lambda x.\, u \ieq v$, where $u$ is obtained by
replacing all occurrences of $s_i$ in $\cst{f}\,\tuple{s}{n}$ with $x$ and $v$ is
obtained by replacing all occurrences of $s_i$ in
$\cst{f}\,\tuple{t}{n}$ with $x$. For \confrep{}{an equation }$\cst{f}\,\tuple{s}{n} \eq
\cst{f}\,\tuple{t}{n}$, the analogous abstraction is $\lambda x.\, \inot (u \ieq
v)$.
%
%
Adding abstractions of the literals occurring in the conjecture to $\mathit{Inst}$ can
provide useful instantiations for formulas such as induction principles of
datatypes. As the conjecture is negated\confrep{}{ in refutational theorem proving},
the equation's polarity is inverted in the
abstraction.

\begin{examplex}
\label{ex:dat056-2}
The clausified conjecture of the problem \texttt{DAT056\^{}2}
\cite{ns-13-leo2sh} from the TPTP library is $\cst{ap} \, \cst{xs}
\, (\cst{ap} \, \cst{ys} \, \cst{zs}) \not\eq \cst{ap} \, (\cst{ap} \, \cst{xs} \,
\cst{ys}) \, \cst{zs}$, where $\cst{ap}$ is the \confrep{}{list }append operator defined
recursively on its first argument and $\cst{xs}$, $\cst{ys}$, and $\cst{zs}$ are
of list type. Abstracting $\cst{xs}$ from the disequation yields $t = \lambda \mathit{xs}.\, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\allowbreak\ieq\allowbreak \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$, which is added
to $\mathit{Inst}$.
Included in the problem is the induction axiom
for the list datatype: $\iforall p. \, p \, \cst{nil} \iand (\iforall x \,
\mathit{xs}. \, p \, \mathit{xs} \iimplies\allowbreak p \, (\cst{cons} \, x \,
\mathit{xs})) \iimplies\allowbreak \iforall \mathit{xs}. \, p \, \mathit{xs}$, where
$\cst{nil}$ and $\cst{cons}$ have the usual meanings.
Instantiating $p$ with $t$ and
using the $\cst{ap}$ definition, we can prove
$\iforall \mathit{xs}. \, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
\ieq \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$,
from which we easily derive a contradiction.
\end{examplex}

\ourparagraph{Evaluation and Discussion}

\begin{figure}
\centering
  \def\arraystretch{1.1}%
  \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 0.5em}@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
    & & $+$\relax{LA} & $-$\relax{LA} \\ \midrule
    TPTP &
%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{TPTP}}}&
      \relax{IC}  & 1616  & 1635$^\star$    \\
    & \relax{DCI} & 1507  & 1532\phantom{$^\star$}    \\
    & \relax{DCS} & 1668  & {\bf 1703}\phantom{$^\star$} \\ \midrule
    SH &
%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
      \relax{IC}  & \colalign425 & \colalign452$^\star$    \\
    & \relax{DCI} & \colalign362 & \colalign385\phantom{$^\star$}    \\
    & \relax{DCS} & \colalign441 & \colalign\textbf{457}\phantom{$^\star$} \\ \bottomrule
  \end{tabular}
  \caption{Impact of clausification and~lightweight AVATAR}
  \label{fig:avatar-clause}
\end{figure}

The base configuration (\emph{base}) uses immediate clausification (\relax{IC}) and 
disables lightweight AVATAR ($-$\relax{LA}). To test
the merits of delayed clausification, we vary \emph{base}'s parameters along two axes: We
choose immediate clausification (\relax{IC}), delayed clausification as inference
(\relax{DCI}), or delayed clausification as simplification (\relax{DCS}), and we
either enable ($+$\relax{LA}) or disable ($-$\relax{LA}) lightweight AVATAR.
Neither of the configurations uses instantiation with terms from $\instset$.

Figure~\ref{fig:avatar-clause} shows that using delayed clausification as
simplification greatly increases the success rate, regardless of whether
lightweight AVATAR is used. Using delayed clausification as inference has the
opposite effect on both problem sets, presumably due to the large number of
clauses it creates. By manually inspecting the  proofs found by the \relax{DCS}
configuration, we noticed that a main reason for its success is that it does
not simplify away equivalences.
%
Overall, the lightweight AVATAR harms performance, but the sets of
problems proved with and without it are vastly different. For example,
the \relax{IC}$+$\relax{LA} configuration proves \NumberOK{38} problems not
proved by \relax{IC}$-$\relax{LA} (i.e., \emph{base}) on TPTP benchmarks and
\NumberOK{14} such problems on SH benchmarks.

\looseness=-1
The Boolean instantiation technique presented above requires delayed
clausification.
We assessed it in the best configuration from
%To assess it, we enabled it in the best configuration from
Fig.~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
Zipperposition proves \NumberOK{1700} TPTP problems and \NumberOK{456} SH
problems.
On TPTP, even though $+$\relax{BI} solves \NumberOK{3} problems less
than \relax{DCS}$-$LA, it is very useful: \NumberOK{41}
problems can be proved with $+$\relax{BI} but not with \relax{DCS}$-$\relax{LA}. 
Conversely, \NumberOK{44} problems are solved with \relax{DCS}$-$\relax{LA},
but not with $+$\relax{BI}, which suggests that Boolean instantiation can be 
explosive.
%
One of the problems Boolean instantiation helps solve is \texttt{NUM636\^{}2} (a
re-encoding of \texttt{NUM636\^{}3}).
It conjectures that $\iforall x.\, \cst{s} \, x \ineq x$, where $x$ ranges over
Peano-style numbers specified by $\cst{z}$ and $\cst{s}$. The given axioms are
the induction principle $\forall p.\, p \, \cst{z} \,\iand\, \forall x. \, (p \,
x \iimplies p \, (\cst{s} \, x)) \iimplies \forall x. \, p\, x$, injectivity
$\forall x y. \, \cst{s}\,x \ieq \cst{s}\,y \iimplies x \ieq y$, and
distinctness $\forall x. \, \cst{s}\,x \ineq \cst{z}$. The conjecture is easily
proved if Boolean instantiation is enabled: Even though the conjecture literal
cannot be abstracted, instantiating $p$ with the term $\lambda x.\, \cst{s}
\, x \ineq x$ used in the encoding of the (nonclausified) conjecture leads to a proof in just
\NumberOK{22} given clause loop iterations. Zipperposition also finds a
proof using the \relax{DCI}$-$\relax{LA} configuration, but this requires
\NumberOK{294} iterations.

%Raw data reveals that Boolean instantiation
The $+$\relax{BI} configuration proves \NumberOK{18} TPTP problems no other
configuration from Fig.~\ref{fig:avatar-clause} can prove. Among these is
\texttt{DAT056\^{}2} (Example~\ref{ex:dat056-2}). In contrast, on SH benchmarks, only
\NumberOK{6} problems are proved using $+$\relax{BI} and not
\relax{DCS}$-$\relax{LA}. For all these problems, Boolean instantiation does not
appear in the proof, suggesting that this result is due to the randomness in the
evaluation environment. The fact that \relax{BI} has no effect on SH benchmarks
is to be expected because Sledgehammer does
not include lemmas whose name contains the substring \texttt{.induct} and that
contain predicate variables. Therefore, \relax{BI} applies to fewer clauses.

\section{Exploring Boolean Selection Functions}
\label{sec:bool-select}

Superposition calculi are parameterized by a literal selection function and a
term order that help prune considerable swaths of the search space without
jeopardizing completeness. The core inferences apply only to a clause's
\emph{eligible} literals, defined as either the clause's selected
literals or, if none are selected, the clause's literals that are maximal with
respect to the term order. To further restrict which terms can be targeted by
an inference, the \olamsup{} calculus introduces \emph{Boolean selection
functions}.

A Boolean selection function chooses \emph{green subterms} of Boolean type
(different than $\top$ or $\bot$ and not occurring at either side of a positive
literal) in a clause and gives rise to a notion of eligibility that considers
the formula structure. Green subterms correspond to the first-order skeleton of
a higher-order term; that is, they do not occur in positions under applied
variables, quantifiers, or $\lambda$-abstractions.

%In particular, green
%subterms are subterms arising inside first-order-like contexts, and for first
%other problems, all subterms are green.

\begin{definitionx}[Green subterms and green positions]
  \,Green subterms and green positions are defined inductively as
  follows: $t$ is a green subterm of $t$ at green position $\varepsilon$; if $t$ is a green subterm of $u_i$ at green position $p$
  and $\cst{f}$ is a constant different from $\iforall$ and $\iexists$, then
  $t$ is a green subterm of $\cst{f} \, \tuplen{u}$ at green position $i.p$,
  assuming $i \leq n$.
\end{definitionx}

\begin{examplex} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
(\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
\, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
\end{examplex} 
%
Green positions are lifted to clauses as follows: If $p$ is the green position
of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
\olamsup{} mandates additional restrictions on the Boolean selection function:
$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
maximal term of the clause, $s$ cannot be selected.

\begin{definitionx}[Eligibility]
  \,Given a substitution $\sigma$ and term order $\succ$, we say a literal $l$
  is (strictly) eligible with respect to $\sigma$ in $C$ if it is selected in
  $C$ or there are no selected literals and no selected Boolean subterms in
  $C$ and $l\sigma$ is (strictly) maximal in $C\sigma$ with respect to the
  term order.
%
  The eligible subterms of a clause $C$ with respect to a substitution
  $\sigma$ are inductively defined as follows:
  Any subterm selected by the Boolean selection function is eligible.
  %Any selected subterm is eligible.
  For a strictly eligible literal $s \eq t$ with $t\sigma \not\succ
  s\sigma$, $s$ is eligible. For an eligible literal $s \not\eq t$ with
  $t\sigma \not\succ s\sigma$, $s$ is eligible. If a subterm $t$ is eligible
  and the head of $t$ is not $\ieq$ or $\ineq$, all direct green subterms of
  $t$ are eligible. If a subterm $t$ is eligible and $t$ is of the form $u
  \ieq v$ or $u \ineq v$, then $u$ is eligible if $v\sigma \not\succ u\sigma$
  and $v$ is eligible if $u\sigma \not\succ v\sigma$.
\end{definitionx}

The above definitions of green subterms and eligibility were originally introduced
with Boolean $\lambda$-superposition \cite{bbtv-21-full-ho-sup}.
The Boolean selection function plays a similar role as the literal
selection function in standard superposition.
Literal selection functions eliminate some of the nondeterminism present in the superposition
calculus by focusing on selected parts of the search space. Boolean selection functions achieve
the same goal, but in a different context: They eliminate nondeterminism that is not
present in standard superposition, namely, the choice of subformula on which 
the Boolean calculus rules are to be applied. As with literal selection functions,
selecting few (and smaller) subterms can give rise to fewer possible inferences
and reduce clause proliferation.

%
This notion of eligibility opens up possibilities for reasoning with
formulas that are hard to simulate with the existing superposition machinery.
For example, given a formula $\varphi \iimplies \psi$, selecting the antecedent
simulates forward reasoning, whereas selecting the consequent simulates backward
reasoning. The new eligibility also makes it possible to restrict the proof
search to a small, promising part of a formula. Note that
literal selection can override Boolean selection: Selecting a literal 
might make some of its green subterms eligible, regardless
of Boolean selection.

In our previous work \cite{nbtv-2021-foboolsup}, we left this area of new
possibilities largely unexplored. We designed simple functions that selected
smallest, largest, innermost, or outermost terms, but they did not impact
performance much. Here, we propose alternatives.
%
Intuitively, a well-performing literal selection
function might succeed at taming the combinatorial explosion if the
selected literal can take part in few inferences
\cite{hrsv-16-selsel}. However, Boolean selection functions
introduce another factor to consider:\ the context in which the selected
subterm occurs. This suggests the following definition:

\begin{definitionx}[Contextualized Boolean selection function]
  \label{def:context-bool-sel}
  \,Let $\mathit{ctx}(C)$ be a function that maps a clause $C$ to a set of green positions
  $p$ such that $C|_p$ is a selectable Boolean subterm, and let
  $\vartriangleright$ be a partial order on pairs of terms and green positions.
  The \emph{context
  Boolean selection function} $\mathit{Sel\/}_{\mathit{ctx}}^{\,\vartriangleright}(C)$
  selects all terms $t$ such that $t=C|_p$, $p \in \mathit{ctx}(C)$, and
  $(t,p)$ is maximal with respect to $\vartriangleright$.
\end{definitionx}

In the above definition, the function $\mathit{ctx}$ lets
us choose the context in which the Boolean subterm appears. Then, among
the terms in the chosen context, we choose the ones that are maximal with
respect to $\vartriangleright$.

Ganzinger and Stuber considered Boolean subterm selection for their extension of
first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus requires
selection of subterms occurring in negative green positions, defined below.

\begin{definitionx}[Polarity of green positions]
 \,Negative and positive green positions in a clause $C = l_1 \llor \cdots \llor
 l_n$ are defined inductively as follows: For each
 $1 \leq i \leq n$, the green position $l_i.s$ is positive if $l_i = \poslit{s}$ and negative if $l_i
 = \neglit{s}$. If $p$ is positive (negative) and $C|_p = s \,
 \tuplen{t}$ where $s$ is either $\iand$ or $\ior$, then each
 $p.i, 1 \leq i \leq n$, is positive (negative). If $p$ is positive and $C|_p =
 \inot \, s$, then $p.1$ is negative; if $p$ is negative and $C|_p = \inot \, s$,
 then $p.1$ is positive. If $p$ is positive and $C|_p = s \iimplies t$, then
 $p.1$ is negative and $p.2$ is positive; if $p$ is negative and $C|_p = s
 \iimplies t$, then $p.1$ is positive and $p.2$ is negative.
\end{definitionx}
Note that the polarity of $p$ is undefined whenever $C|_p$ is not a green Boolean subterm
or it occurs under a (dis)equivalence or an uninterpreted symbol.
To assess how the function $\mathit{ctx}$ affects performance, we use the following
selection functions that consider green positions of selectable Boolean terms:
%
\begin{enumerate}[\rm\texttt{Backward}~]
  \item[\rm\texttt{Any}~] select all green positions;
  \item[\rm\texttt{Pos}~] select all positive green positions;
  \item[\rm\texttt{Neg}~] select all negative green positions;
  \item[\rm\texttt{Forward}~] select all green positions $p = q.1$ such that $C|_q = s \iimplies t$;
  \item[\rm\texttt{Backward}~] select all green positions $p = q.2$ such that $C|_q = s \iimplies t$;
  \item[\rm\texttt{Deep}~] select all green positions of maximal length;
  \item[\rm\texttt{Shallow}~] select all green positions of minimal length.
\end{enumerate}

\looseness=-1
We also introduce three partial orders for selecting subterms from a given
context. For all three orders, if exactly one of the subterms has a logical head,
then the subterm with the nonlogical head is larger, because logical symbols
are more explosive. Otherwise, the orders use the following criteria:
%
\begin{enumerate}[$\vartriangleright_\text{ground}$~]
  \item[$\vartriangleright_\text{ground}$~] If exactly one of the subterms is ground, make the
  ground subterm larger; otherwise, if exactly one of the subterms is of the form $s \ieq
  t$, make this subterm larger.

\smallskip

  \item[$\vartriangleright_\text{depth}$~] If one of the subterms has larger subterm depth
  (longest valid green subterm position), make this subterm larger; otherwise, if one of the
  subterms has less distinct variables, make this subterm larger.

\smallskip

  \item[$\vartriangleright_\text{def}$~] If exactly one of the subterms is of the form
  $\cst{p}\,\tuplen{X}$ where $\tuplen{X}$ is a tuple of free variables, make
  the other subterm larger; otherwise, if exactly one of the subterms is of the form $X \,
  \tuplen{s}$, make the other subterm larger.
\end{enumerate}

In case of a tie, the subterm with the smaller syntactic weight is made larger,
and if both subterms have the same weight, the term that occurs in a position further
to the left (i.e., that has a lexicographically smaller position) is made larger.

These orders follow the design principle enunciated by Hoder et
al.~\cite{hrsv-16-selsel} that ground or deep terms and terms with repeated
variables are ``less unifiable'' with the similar observation for higher-order
logic that reasoning about interpreted symbols or applied variables is usually
explosive.

\begin{examplex}
  Selecting the right Boolean subterm can help avoid elaborating
  high\-er-order inferences. Consider the unsatisfiable clause set consisting of $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a}) \iimplies\allowbreak \inot(\cst{p} \, (\lambda y. \allowbreak\, X
  \, y \, \cst{a}))$, $\cst{p} \, (\lambda y.\, \cst{a})$, and
  $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$.
  Note that $\cst{p} \, (\lambda y.\, X
  \, (\lambda x.\, x) \, \cst{a})$ and $\cst{p} \,
  (\lambda y.\, \cst{a})$ have infinitely many unifiers of the form $\{ X \mapsto \lambda fx.
  \, f^i\, (x) \}, i \geq 0$, whereas $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have only one unifier. 
  If \texttt{Forward} context selection is enabled,
  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
  forcing computation of at least 100 unifiers (under the assumption that
  unifiers are returned in order of increasing $i$) before we get to refute
  $\inot(\cst{p} \, (\lambda y. y^{100} \, \cst{b}))$.
  In contrast, \texttt{Backward} context selection allows us to
  superpose from $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$ into $\cst{p} \, (\lambda y. \, X
  \, y \, \cst{a})$, avoiding this explosion. 
\end{examplex}

\ourparagraph{Evaluation and Discussion}

\begin{figure}
\centering
\def\arraystretch{1.1}%
 \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
  &                                    & \texttt{Any} & \texttt{Pos} & \texttt{Neg}        & \texttt{Forward} & \texttt{Backward} & \texttt{Deep}     & \texttt{Shallow}  \\ \midrule
  TPTP &
    $\vartriangleright_\text{ground}$  & 1538         & 1550         & 1547                & 1534             & {\bf 1554}        & 1539              & 1538     \\
  & $\vartriangleright_\text{depth}$   & 1542         & 1550         & 1528                & 1542             & 1550              & 1547              & 1535     \\
  & $\vartriangleright_\text{def}$     & 1543         & 1551         & 1540                & 1540             & 1551              & 1545              & 1537  \\ \midrule

%  \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
  SH &
    $\vartriangleright_\text{ground}$  & \colalign386 & \colalign379  & \colalign386       & \colalign386     & \colalign379      & \colalign{\bf387} & \colalign{\bf387}     \\
  & $\vartriangleright_\text{depth}$   & \colalign377 & \colalign376  & \colalign384       & \colalign378     & \colalign376      & \colalign379      & \colalign376      \\
  & $\vartriangleright_\text{def}$     & \colalign379 & \colalign374  & \colalign{\bf387}  & \colalign379     & \colalign380      & \colalign377      & \colalign381  \\ \bottomrule
 \end{tabular}
 \caption{Impact of the Boolean selection function}
 \label{fig:bool-sel}
\end{figure}

When the input problem is clausified using immediate clausification, almost
all Boolean structure is lost. In this case, we expect Boolean selection to
have a modest effect. To better assess this feature, in this evaluation we use
\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:formulas} as the baseline
configuration. To avoid interference of literal and Boolean selection, we
additionally forbid the literal selection function from selecting a literal if
it contains a selectable Boolean subterm.

The results of evaluating 21 concrete selection functions obtained
by instantiating the contextualized Boolean selection function are shown in
Fig.~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
while columns denote the function $\mathit{ctx}$.

\looseness=-1
On TPTP benchmarks, Boolean selection helps tame the explosion caused by
dynamic clausification used as inference: \NumberOK{All but one} selection functions
outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
\texttt{NUM636\^{}2} from Sect.~\ref{sec:formulas}, using Boolean selection
can reduce the number of given clause loop iterations from
\NumberOK{294} to \NumberOK{71}.

The results suggest that selection of term context has more impact
than the partial term order. % The number of proved problems fluctuates more when
%changing the function $\mathit{ctx}$ than when changing the partial order.
Also, the best results are obtained when a context more specific than \texttt{Any}
is chosen. Remarkably, functions using \texttt{Pos} context
perform better than the ones using \texttt{Neg} context
on TPTP but the opposite is observed on SH.

Using different Boolean selection functions yields vastly different sets of
proved problems on TPTP benchmarks: In total, there are \NumberOK{103} problems
proved by some configuration from Fig.~\ref{fig:bool-sel} but not by
\relax{DCI}$-$\relax{LA}. However, there are only \NumberOK{16} such SH problems.
It would seem that the advanced formula reasoning facilitated by the Boolean
selection formulas is usually not required by Sledgehammer problems.

% {\tt
%   \begin{itemize}
%      \item To evaluate bool selection functions we used DCI from above, but
%            disabled literal selection function from selecting a literal if it
%            has a selectable Boolean subterm.
%      \item For TPTP it seems that selection is a good idea: all configurations
%      are better than naked DCI (only 6 for SH).
%      \item Overall, it seems that position has bigger influence that the
%      selection function.
%      \item On SH there are 14 problems provable by some of the Boolean
%            selection functions but not naked CSI. Sel3\_deep proves 12 problems
%            not proved by DCI. Interestingly, the best function Sel3\_neg proves
%            only 7 problems not proved by DCI.
%      \item On TPTP much better: 110 problems not proved by DCI but proved by
%      some of the selection functions. sel3\_pos proves 74 problems not proved
%      by DCI, sel1 backward 73.
%      \item TODO: Add any context and maybe $\vartriangleright_4$ which uses weight only
%      \item What we also observed is that it on TPTP problem it is reasonable to have any selection
%            function and that already stops the explosion. On SH quality is more important.
%      \item
%   \end{itemize}
% }

\section{Enumerating Infinitely Branching Inferences}
\label{sec:infinite-branching}

\looseness=-1
As an optimization and to simplify the code, Leo-III
\cite{as-18-phd} and Vampire 4.4 \cite{br-19-restricted-unif} (which uses
\confrep{}{\emph{restricted combinatory unification}, }a predecessor of combinatory
superposition) compute only a finite subset of the possible conclusions for
inferences that require enumerating a CSU. Not only is this a source of
incompleteness, but choosing the cardinality of the computed subset is a
difficult heuristic choice. Small sets can result in missing the unifier
necessary for the proof, whereas large sets make the prover spend too long in
the unification procedure, generate useless clauses, and possibly get sidetracked into the wrong parts of the search space.

We propose a modification to the given clause procedure to seamlessly
interleave unifier computation and proof state exploration. Given a complete
unification procedure, which may yield infinite streams of unifiers, our
modification fairly enumerates all conclusions of inferences relying on
elements of a CSU. Under some reasonable assumptions, it behaves exactly like
the standard given clause procedure on purely first-order problems.
We also describe heuristics that help achieve a similar
performance as when using incomplete, terminating unification procedures
without sacrificing completeness.

Given that we cannot decide whether there exists a next CSU
element in a stream of unifiers, the request for the next conclusion might not
terminate,
effectively bringing the theorem prover to a halt. Our modified given clause procedure
expects the unification procedure to return a lazily computed stream
\cite[Sect.~4.2]{co-1999-funds}, where each element is either $\emptyset$ or
a singleton set containing a unifier. To avoid getting stuck waiting for a unifier
that may not exist, the unification procedure should return
$\emptyset$ after it performs some number of operations without finding a unifier.

The complete unification procedure by Vukmirović et al.\
\cite{our-unif-paper} returns such a stream. Other procedures such
as Huet's \cite{gh-75-unification} and Jensen and
Pietrzykowski's \cite{jp-76-unif} can easily be adapted to meet this
requirement. Based on the stream of unifiers interspersed with $\emptyset$, we
can construct a stream of inferences similarly interspersed with $\emptyset$.
Any finite prefixes of this stream can be computed in finite time.

To support such streams in the given clause procedure, we extend it to
represent the proof state not only by the active ($A$) and passive ($P$) clause
sets, but also by a priority queue $Q$ containing the inference streams,
similar to the ``to do'' set $T$ present in the abstract Zipperposition loop of 
Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
Each stream is associated with a weight, and $Q$ is sorted in order of
increasing weight\confrep{}{, a low weight corresponding to a high priority}.
When they introduced \lamsup, Bentkamp et al.\
\cite{bbtvw-21-sup-lam} described an older version of this extension. Here
we present a newer version in more detail, including heuristics to postpone
unpromising streams. The pseudocode of the modified procedure is as follows:
\vspace{2.5\jot}
%
\newcommand\ParamMode{\ensuremath{K_{\mathrm{fair}}}}
\newcommand\ParamMaxStreams{\ensuremath{K_{\mathrm{best}}}}
\newcommand\ParamRetry{\ensuremath{K_{\mathrm{retry}}}}
\newcommand{\assign}[2]{\State \ensuremath{\mathit{#1} \gets #2}}
\newcommand{\assignSameLine}[2]{\ensuremath{\mathit{#1} \gets #2}}
\algrenewcommand\algorithmicindent{1em}
\begin{algorithmic}[0]
    \Function{ExtractClause}{$Q$, $\mathit{stream}$}
      \assign{maybe\_clause}{\text{pop and compute the first element of } \mathit{stream} }
      \If{$\mathit{stream}$ is not empty}
        \State add $\mathit{stream}$ to $Q$ with an increased weight
      \EndIf
      \State \Return $\mathit{maybe\_clause}$
    \EndFunction

    \vspace{2\jot}

    \Function{HeuristicProbe}{$Q$}
      \assign{i}{0}
      \assign{collected\_clauses}{\emptyset}
      \While{$i < \ParamMaxStreams$ and $Q \not= \emptyset$}
        \assign{j}{0}
        \assign{maybe\_clause}{\emptyset}
        \While{$j < \ParamRetry$ and $Q \not= \emptyset$ and $\mathit{maybe\_clause} = \emptyset$}
          \assign{stream}{\text{pop the lowest-weight stream in } Q}
          \assign{maybe\_clause}{\textsc{ExtractClause}(Q, \mathit{stream})}
          \assign{j}{j+1}
        \EndWhile
        \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \mathit{maybe\_clause}}
        \assign{i}{i+1}
      \EndWhile
      \State \Return $\mathit{collected\_clauses}$
    \EndFunction

    \vspace{2\jot}

  \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}
    \assign{collected\_clauses}{\emptyset}
    \assign{oldest\_streams}{\text{pop } \mathit{num\_oldest} \text{ oldest streams from } Q}
    \For{$\mathit{stream}$ in $\mathit{oldest\_streams}$}
      \assign{collected\_clauses}{collected\_clauses \mathrel\cup \textsc{ExtractClause}(Q, \mathit{stream})}
    \EndFor
    \State \Return $\mathit{collected\_clauses}$
  \EndFunction

  \vspace{2\jot}

  \Function{ForceProbe}{$Q$}
    \assign{collected\_clauses}{\emptyset}
    \While {$Q \not= \emptyset$ and $\mathit{collected\_clauses} = \emptyset$}
      \assign{collected\_clauses}{\textsc{FairProbe}(Q, |Q|)}
    \EndWhile

    \If{$Q = \mathit{collected\_clauses} = \emptyset$}
      \assign{status}{\textsf{Satisfiable}}
      % \Return (\textsf{Satisfiable}, $\mathit{collected\_clauses}$)
    \Else{}
      \assign{status}{\textsf{Unknown}}
      % \Return (\textsf{Unknown}, $\mathit{collected\_clauses}$)
    \EndIf

    \State \Return $(\mathit{status}, \mathit{collected\_clauses})$
  \EndFunction

  \vspace{2\jot}

  \Function{GivenClause}{$P$, $A$, $Q$}

  \assign{i}{0}
  \assign{status}{\textsf{Unknown}}
  \While{$\mathit{status} = \textsf{Unknown}$}
    \If{$P = \emptyset$}
      \assign{(status, \,forced\_clauses)}{\textsc{ForceProbe}(Q)}
      \assign{P}{P \mathrel\cup \mathit{forced\_clauses}}
    \Else
      \assign{given}{\text{pop a chosen clause from }P \text{ and simplify it}}
      \If{$\mathit{given}$ is the empty clause}
        \assign{status}{\textsf{Unsatisfiable}}
      \Else
        \assign{A}{A \mathrel\cup \{\mathit{given}\}}
        \For{$\mathit{stream}$ in streams of inferences between $\mathit{given}$ and $\mathit{other} \in A$}
          \If{$\mathit{stream}$ is not empty} \assign{P}{P \mathrel\cup \textsc{ExtractClause}($Q$, \mathit{stream})}
          \EndIf
        \EndFor
        \assign{i}{i+1}
        \If{$i \; \mathrm{mod} \; \ParamMode = 0$} \assign{P}{P \mathrel\cup \textsc{FairProbe}(Q, i/\ParamMode )}
        \Else{} \assign{P}{P \mathrel\cup \textsc{HeuristicProbe}(Q)}
        \EndIf
      \EndIf
    \EndIf
  \EndWhile

  \State \Return $\mathit{status}$

  \EndFunction
\end{algorithmic}
% \caption{Our modified given clause procedure}
% \label{fig:gc-pseudocode}
%
% \end{figure}
\vspace{2.5\jot}

Initially, all input clauses are put into $P$, and $A$ and $Q$ are empty. Unlike
in the standard given clause procedure, inference results are represented as clause
streams. The first element is inserted into $P$, and the rest of the stream is
stored in $Q$ with some positive integer weight computed from the inference rule.

% The functions \textsc{Probe} and \textsc{ForceProbe} extract some conclusions
% from the inference streams and store them in $P$. \textsc{Probe} heuristically
% chooses some clauses, whereas \textsc{ForceProbe} is used when $P$ is empty to
% find one clause if previous probes failed to produce any clauses, as a fallback.
% By choosing clauses from several streams at the same time,
% we give more freedom to the prover's existing clause selection heuristics,
% which select a clause from $P$.

% \looseness=-1
% \textsc{Probe} has two modes of operation, controlled by a parameter \ParamMode~(by
% default, $\ParamMode = 70$). In every $\ParamMode$th invocation, \textsc{Probe} extracts
% conclusions from an increasing number of oldest streams. This amounts to
% dovetailing, which achieves fairness, analogously to the
% pick--given ratio \cite{ss-02-brainiac,mcw-1997-otter} in the
% given clause procedure. For the other
% invocations, \textsc{Probe} selects the $\ParamMaxStreams$ highest-priority streams (by
% default, $\ParamMaxStreams = 10$) and extracts elements from, or \emph{probes}, each stream,
% discarding $\emptyset$ values, until a clause is found or until
% $\ParamRetry$ (by default, $\ParamRetry=20$) $\emptyset$ values have been discarded.
% Setting $\ParamRetry>1$ ensures that promising streams are given a fair chance to
% produce a clause, even if they rely on a complicated unifier. Each
% probed stream~$S$ is put back in $Q$ with a new priority.%

\looseness=-1
To eventually consider inference conclusions from streams in $Q$ as given
clauses, we extract elements from, or \emph{probe}, streams and move any obtained
clauses to $P$. Analogously to the traditional pick--given ratio
\cite{ss-02-brainiac,mcw-1997-otter}, we use a parameter
\ParamMode{} (by default, $\ParamMode = 70$) to ensure fairness: Every \ParamMode{}th iteration,
\textsc{FairProbe} probes an increasing number of oldest streams, which achieves
dovetailing. In all other iterations, \textsc{HeuristicProbe} attempts to
extract up to \ParamMaxStreams{}~clauses from the most promising streams (by default,
$\ParamMaxStreams = 7$).
In each attempt, the most promising stream in $Q$ is chosen. If its first
element is $\emptyset$, the rest of the stream is inserted into $Q$ and a new stream is
chosen. This is repeated until either \ParamRetry{} occurrences of $\emptyset$ have  been
met (by default, $\ParamRetry = 20$) or the stream yields a singleton. Setting $\ParamRetry > 0$ increases
the chance that \textsc{HeuristicProbe} will return $\ParamMaxStreams$ clauses, as desired. Finally, if $P$ becomes empty, \textsc{ForceProbe}
searches relentlessly for a clause in $Q$, as a fallback.

\looseness=-1
The function \textsc{ExtractClause} extracts an element from a nonempty stream
not in $Q$ and inserts the remaining
stream into $Q$ with an increased weight, calculated as follows.
Let $n$ be the number of times the stream was chosen for
probing. If probing results in $\emptyset$, the stream's weight is increased by
$\max\,\{2{,}\; n-16\}$. If probing results in a clause $C$ whose penalty is
$p$, the stream's weight is increased by $p \cdot \max\,\{1{,}\; n-64\}$. The
penalty of a clause is a number assigned by Zipperposition based on
features such as the depth of its derivation and the rules used in it.
The constants $16$ and $64$ increase the chance that newer clause-producing streams are picked,
which is desirable because their first clauses are expected to be useful.

\looseness=-1
All three probing functions are invoked by
\textsc{GivenClause}, which contains the saturation loop. It differs
from the standard given clause procedure in three ways:
First, the proof state includes $Q$ in addition to $P$ and $A$. Second,
new inferences involving the given clause are added to $Q$ instead of being
performed immediately. Third, inferences in $Q$ are periodically performed
lazily to fill $P$.

\newcommand\infstream[1]{[#1]}

\begin{examplex} 
  \begin{sloppypar}
  Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
  \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}),\allowbreak
  \neg \cst{p} \, (\cst{f}^{100} \, \cst{a})  \}$ and a selection function which
  selects negative literals.
  Let $P \mid A \mid Q$ denote
  the state of the given clause loop (i.e., the contents of the passive and active set
  and of the stream queue), and let $\infstream{ a_1, a_2, \ldots }$
  denote an infinite stream of elements.
  \end{sloppypar}

  The given clause loop begins in the state $X \, (\cst{f} \, \cst{a}) \not\eq
  \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}), \neg \cst{p} \,
  (\cst{f}^{100} \, \cst{a}) \mid \emptyset \mid \emptyset$. If the clause $\neg
  \cst{p} \, (\cst{f}^{100} \, \cst{a})$ is chosen for processing, since $Q$ is empty
  and no inferences with the chosen clause are possible, the state becomes $X \,
  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
  \cst{a}) \mid \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \emptyset$. When
  the clause $X \, (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor
  \cst{p} \, (X \, \cst{a})$ is chosen, a new stream which enumerates results of
  equality resolution (on its first literal) is created. There are infinitely many
  conclusions of this inference, since there are infinitely many unifiers for the
  first literal of the form $\{ X \mapsto \lambda x. \, \cst{f}^i \, x \}$, for
  $i \geq 0$. Thus, the stream is $\infstream{ \{\cst{p} \, \cst{a}\}, \{\cst{p} \, (\cst{f}\,\cst{a})\}, \ldots }$,
  possibly with $\emptyset$s interspersed. With the standard given clause procedure,
  there would have been no way to represent this infinitary result.
  %The creation of the
  %inference result stream is the most striking difference between our
  %and standard given clause procedure.

  When the stream is created, its first element is popped and put into $P$. Then, based on the
  parameters that control inference stream probing, some number of clauses from
  the stream are computed and moved to $P$. After two iterations, the state might be
  $ \cst{p} \,
  \cst{a}, \cst{p} \, (\cst{f} \, \cst{a}), \cst{p} \, (\cst{f} \, (\cst{f} \, \cst{a}))   \mid X \,
  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
  \cst{a}), \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \infstream{ \{\cst{p} \, (\cst{f}^3\,\cst{a})\}, \ldots }$.

  In the next iterations, some clause of the form $\cst{p} \, (\cst{f}^{i} \,
  \cst{a})$, where $i < 100$, is chosen, but no inferences with it can be
  performed. Then, the stream created in the second iteration is probed, and its
  results fill the set $P$. Ultimately, the clause $\cst{p} \, (\cst{f}^{100} \,
  \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
\end{examplex}

\textsc{GivenClause} eagerly stores the first element of a new inference stream
in $P$ to imitate the standard given clause procedure. If the underlying
unification procedure behaves like the standard first-order unification
algorithm on higher-order logic's first-order fragment, our given clause
procedure coincides with the standard one. The unification procedure by
Vukmirović et al.\ terminates on the first-order and other fragments
\cite{tn-93-patterns}. To avoid computing
complicated unifiers eagerly, it immediately returns $\emptyset$ for a problem that does not
belong to one of the fragments that admit efficient unifier computation.

The design of our given clause procedure was guided by folklore knowledge about
higher-order theorem proving. First, in our experience most steps in
long higher-order proofs involve first-order literals. The unification
procedure and inference scheduling ensure that first-order inference
conclusions are put in the proof state as early as possible. Second, some
inference rules are expected to be largely useless. We initialize the stream
penalty differently for each rule, allowing old streams of more useful
inferences to be queried before newly added, but potentially less useful
streams. Finally, if we use a unification procedure that has aggressive
redundancy elimination, we will often find the necessary unifier within the
first few unifiers returned. Similarly, if a stream keeps returning
$\emptyset$, it is likely that it is blocked in a nonterminating computation
and should be ignored. Our heuristics to increase the stream penalties take
both observations into account.

\ourparagraph{Evaluation and Discussion}

When the unification procedure of Vukmirović et al.\ was implemented  in
Zipperposition, it was observed that this is the only competing
higher-order prover that proves all Church numeral problems from the TPTP,
never spending more than 5~s on a problem \cite{our-unif-paper}. On these hard
unification problems, the stream system allows the prover to explore the proof
state lazily.

Consider the TPTP problem \verb|NUM800^1|, which requires finding
a function $F$ such that $F \, \cst{c_1} \, \cst{c_2}
\ieq \cst{c_2} \iand F \, \cst{c_2} \, \cst{c_3} \ieq \cst{c_6}$, where
$\cst{c}_n$ abbreviates the Church numeral for~$n$, $\lambda s\, z. \>
s^n \, z$. To prove
\confrep{it}{the problem}, it suffices to take $F$ to be the multiplication operator
$\lambda x \, y \, s \, z. \> x \, (y \, s) \, z$.
However, this unifier is only one out of many available for each occurrence of
$F$.

% TODO: If I add a new benchmark set, make sure this sentence is fixed
In an independent evaluation setup on a set of 2606 TPTP version 7.2.0
problems almost identical to the one we use, Vukmirović et al.\
\cite[Sect.~7]{our-unif-paper}
compared a complete, nonterminating variant of the unification procedure and a
pragmatic, terminating variant. The
pragmatic variant was used directly---all the inference conclusions were put
immediately in $P$, bypassing $Q$. The complete variant, which relies on
possibly infinite streams and is much more prolific, proved only 15  problems
less than the most competitive pragmatic variant. Furthermore, it proved 19
problems not proved by the pragmatic variant.
%
This shows that our given clause procedure, with its heuristics, allows the
prover to defer exploring less promising branches of the unification and uses
the full power of a complete higher-order unifier search to solve unification
problems that cannot be proved by a restricted procedure.

The parameters \ParamMode{}, \ParamRetry{}, and \ParamMaxStreams{} can greatly
influence the behavior of the given clause procedure, even when the same
unification procedure is used. Figure~\ref{fig:streams}
presents the effects of these parameters on TPTP and SH.
Selecting a low number of best clauses seems to
perform well on both benchmark sets. However, on SH benchmarks, which require
overwhelmingly first-order unifiers, visiting older streams should be delayed
a lot.

As with Boolean selection functions, changing these three parameters causes
a substantial difference in the set of proved problems. For example, the
configuration that performs the worst on TPTP benchmarks proves \NumberOK{12}
problems that the configuration performing the best on TPTP cannot prove; moreover, there
are \NumberOK{29} TPTP problems that are proved by some set of parameters
other than $\ParamMode=\ParamMaxStreams=16, \ParamRetry=2$. On SH, these
effects are much weaker; most reasonable combinations
of parameters perform similarly.

Among the competing higher-order provers, only Satallax uses infinitely
branching calculus rules. It maintains a queue of ``commands'' that contain
instructions on how to create a successor state in the tableau. One
command describes infinite enumeration of all closed terms of a given function
type. Each execution of this command makes progress in the enumeration. Unlike
evaluation of streams representing elements of CSU, each command execution
is guaranteed to make progress in enumerating the next closed functional
term, so there is no need to ever return $\emptyset$.

\begin{figure}
\centering
\begin{subfigure}[b]{1\textwidth}
  \centering
  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
  &&&&&&& \ParamMode \\[.5\jot]
  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
  \cline{3-5}\cline{7-9}\cline{11-13}
  \\[-1.5\jot]
  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
  %                                     2                        16                     256
  &                         & 2    & 16   & 128  & & 2         & 16   & 128  & & 2    & 16   & 128 \\\midrule
  & $2$                     & 1643 & 1645 & 1645 & & 1661      & 1661 & 1658 & & 1669 & 1664 & 1664 \\[0.5\jot]
  $\ParamMaxStreams$ & $16$ & 1647 & 1646 & 1609 & & {\bf1670} & 1654 & 1602 & & 1665 & 1659 & 1597 \\[0.5\jot]
  & $128$                   & 1646 & 1644 & 1583 & & 1661      & 1656 & 1577 & & 1665 & 1658 & 1576 \\ \bottomrule
  \end{tabular}
  \caption{TPTP benchmarks}
  \label{fig:streams-tptp}
\end{subfigure}
\par\bigskip
\begin{subfigure}[b]{1\textwidth}
  \centering
  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
  &&&&&&& \ParamMode \\[.5\jot]
  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
  \cline{3-5}\cline{7-9}\cline{11-13}
  \\[-1.5\jot]
  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
  %                                          2                                               16                                      128
  &                         & 2            & 16            & 128          & & 2            & 16            & 128          & & 2                  & 16            & 128 \\\midrule
  & $2$                     & \colalign460 & \colalign455  & \colalign454 & & \colalign465 & \colalign463  & \colalign458 & & \colalign466       & \colalign461  & \colalign461 \\[0.5\jot]
  $\ParamMaxStreams$ & $16$ & \colalign458 & \colalign453  & \colalign445 & & \colalign464 & \colalign459  & \colalign441 & & \colalign{\bf468}  & \colalign459  & \colalign442 \\[0.5\jot]
  & $128$                   & \colalign456 & \colalign452  & \colalign430 & & \colalign465 & \colalign458  & \colalign428 & & \colalign{\bf468}  & \colalign459  & \colalign425 \\ \bottomrule
  \end{tabular}
  \caption{SH benchmarks}
  \label{fig:streams-sh}
\end{subfigure}
\caption{Impact of the stream enumeration parameter}
\label{fig:streams}
\end{figure}

\section{Controlling Prolific Rules}
\label{sec:explosiveness}

To support higher-order features
such as function extensionality and quantification over functions,
many refutationally complete calculi employ highly prolific rules.
For example, \lamsup{} includes a
\infname{FluidSup} rule \cite{bbtvw-21-sup-lam} that very often applies to two
clauses if one of them contains a term of the form $F \, \overline{s}_n$,
where $n > 0$.
We describe three mechanisms to keep rules like these under control.

% \begin{enumerate}
%   \item Penalize the streams of expensive inferences.
%   \item Defer selecting the resulting clauses for processing.
%   \item Restrict the applicability of the rules.
% \end{enumerate}

% Complete provers that implement \lamsup{} %, such as Zipperposition,
% have two main heuristic choices: which inference stream to query for clauses
% and which clauses  from the passive set to choose for processing. The first two mechanisms
% control these two choice points. The third mechanism
% sacrifices completeness by applying the inference rules only to a subset of chosen clauses.

First, \emph{we limit applicability of the prolific rules}. In practice, it
often suffices to apply prolific higher-order rules only to initial or shallow
clauses---clauses with a shallow derivation depth. Thus, we added an option to
forbid the application of a rule if the derivation depth of any premise exceeds a
limit.

\newcommand\ParamPenaltyIncrease{\ensuremath{K_\mathrm{incr}}}
Second, \emph{we penalize the streams of expensive inferences}. The weight of
each stream is given an initial value based on characteristics of the inference
premises such as their derivation depth. For prolific rules such as
\infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
less prolific variants of this rule, such as \infname{DupSup} \cite{bbtvw-21-sup-lam}, are increased by a
fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).

Third, \emph{we defer the selection of prolific clauses}. To select the given
clause, most saturating provers evaluate clauses according to some criteria and
choose the clause with the lowest evaluation. To make this choice efficient,
passive clauses are organized into a priority queue ordered by their
evaluations. Like E, Zipper\-position maintains multiple
queues, ordered by different evaluations, that are visited in a round-robin
fashion. It also uses E's two-layer evaluation functions, a variant of which
has recently been implemented in Vampire \cite{gs-20-clausesel}.
%
The two layers are \emph{clause priority} and \emph{clause weight}. Clauses
with higher priority are preferred, and the weight is used for tie-breaking.
Intuitively, the first layer crudely separates clauses into priority classes,
whereas the second one uses heuristic weights to prefer clauses within a
priority class. To control the selection of prolific clauses, we introduce new
clause priority functions that take into account features specific to
higher-order clauses.

The first new priority function, \verb|PreferHOSteps| (\verb|PHOS|), assigns a
higher priority if rules specific to higher-order superposition calculi
were used in the clause derivation. Since most of the other clause priority
functions tend to defer higher-order clauses, having a clause queue that prefers
them might be useful to find some proof more
efficiently. A simpler function, which prefers clauses containing
$\lambda$-abstractions, is \verb|PreferLambda| (\verb|PL|).

\verb|PreferHOSteps| separates clauses created using first-
and higher-order inference rules crudely. However, within higher-order
inference rules there are the ones which make clauses simpler and are thus more
preferable. An example of such a rule is
%
\[\namedinference{\infname{ArgCong}}{C \lor s \eq t}{C \lor s \, \tuplen{X}
\eq t \, \tuplen{X}}\]
%
where $s$ is of the type $\alpha_1 \rightarrow \cdots \rightarrow \alpha_k
\rightarrow \beta$, $\beta$ is a base type, $n \leq k$, free variables
$\tuplen{X}$ are fresh, and literal $s \eq t$ is strictly eligible. When $n =
k$, in most cases, the resulting clause has a first-order literal $s \, \tuplen{X}
\eq t \, \tuplen{X}$ in place of the literal $s \eq t$ of functional type, which usually makes the clause more useful. To
prefer clauses that are only mildly higher-order, we designed the function
\verb|PreferEasyHO| (\verb|PEHO|). It prefers clauses that are the result of
\infname{ArgCong}, have equations between terms of functional type or
between higher-order patterns, or have literals containing logical symbols, in that
order of priority.

A higher-order inference that applies a complicated substitution to a clause is
usually followed by a $\beta\eta$-normalization step. If
$\beta\eta$-normalization greatly reduces the size of a clause, it is likely
that this substitution simplifies the clause (e.g., by removing a variable's
arguments). The new priority function \verb|ByNormalizationFactor| (\verb|BNF|)
is designed to exploit this observation. It prefers clauses that were produced
by $\beta\eta$-normalization, and among those it prefers the ones with larger
size reductions.

Another new priority function is \verb|PreferShallowAppVars| (\verb|PSAV|). This
prefers clauses with lower depths of the deepest occurrence of an applied
variable---that is, $C[X \, \cst{a}]$ is preferred over $C[\cst{f}\,(X \,
\cst{a})]$. The intuition is that applying a substitution to an applied
variable often reduces the variable to a term with a constant head,
yielding a less explosive clause, and the gain is greater for variables closer
to the top level.  Among the functions that rely
on properties of applied variables, we implemented \verb|PreferDeepAppVars|
(\verb|PDAV|), which returns the priority opposite of \verb|PSAV|, and
\verb|ByAppVarNum| (\verb|BAVN|), which prefers clauses with fewer occurrences
of applied variables.

\ourparagraph{Evaluation and Discussion}

\begin{figure}[t]
  \centering
  % \def\arraystretch{1.1}%
  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
            & \texttt{CP}                & \texttt{BAVN}  & \texttt{PL}  & \texttt{PSAV}       & \texttt{PHOS} & \texttt{PEHO}  & \texttt{BNF} & \texttt{PDAV}      \\ \midrule
   TPTP     & {1635}$^\star$             & {\bf 1640}     & 1604         & {1635}              & 1609          & 1617           & 1575         & 1533 \\[0.5\jot]
   SH       & {\bf \colalign452}$^\star$ & \colalign451   & \colalign417 & {\colalign450}      & \colalign439  & \colalign407   & \colalign411 & \colalign302 \\ \bottomrule
  \end{tabular}}
  \caption{Impact of the priority function}
  \label{fig:priorities}
\end{figure}
\begin{figure}[t]
  \centering
  \def\arraystretch{1.1}%
  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
              & $\infty$                     & $16$          & $8$           & $4$             & $2$             & $1$      \\ \midrule
   TPTP       & {\bf 1635}$^\star$           & 1625          & 1632          & 1629            & 1628            & 1618 \\[0.5\jot]
   SH         & {\bf \colalign452}$^\star$   & \colalign438  & \colalign435  & \colalign439    & \colalign435    & \colalign440 \\ \bottomrule

  \end{tabular}}
  \caption{Impact of the \infname{FluidSup} weight increment \ParamPenaltyIncrease}
  \label{fig:penalties}
\end{figure}

In the base configuration (\emph{base}), Zipperposition visits several clause
queues. The configuration uses queues that prefer the clauses that stem from
the conjecture, the ones that have at least one positive literal, the ones that
have been moved from active to passive set, and so on. One of the queues uses the constant
priority function \texttt{ConstPrio} (\texttt{CP}), meaning that it assigns the same
priority to every clause. As this queue is the most often visited one in \emph{base},
changing its priority function should affect the result noticeably.
To evaluate the new priority functions, we replaced
\texttt{CP} with one of the new functions in this queue,
%the queue ordered by \texttt{CP} with the queue ordered by one of the new
%functions,
leaving the clause weight intact. The results are shown in Fig.~\ref{fig:priorities}.

Even though constant priority function achieves remarkable performance, the new priority functions are
useful additions to the prover's repertoire: \NumberOK{37} additional
TPTP problems and \NumberOK{17} additional SH problems can be proved when some
nonconstant priority is used. The generally average-performing \verb|PEHO|
function can prove \NumberOK{9} problems not proved with any other priority function on TPTP
(and \NumberOK{1} on SH). Globally, \NumberOK{24} TPTP problems and \NumberOK{6} SH problems
can be proved exclusively using one particular priority function.

% It
% shows that the expensive priority functions \texttt{PHOS} and \texttt{BNF},
% which require inspecting the proof of clauses, hardly help. Simple functions
% such as \texttt{PL} are more effective: Compared with \textit{base}, \texttt{PL}
% loses \NumberNOK{one} problem overall but proves \NumberNOK{22} new problems.

% {\tt
%  \begin{itemize}

%   \item Another result which is hard to analyze without looking into differences.
%   \item On TPTP 32 solutions by using different priorities. On SH 20. Interestingly,
%   generally bad performing PEHO finds the largest number of solutions not found by CP : 15 (TPTP).
%   On SH the largest number of unique solutions is by PHOS and PEHO.
%   \item Another illustration of orthogonality: 25 problems can be proved by only one prio fun (TPTP), 9 on SH.

%  \end{itemize}
% }

Although it is necessary for refutational completeness, the \infname{FluidSup}
rule is disabled in \emph{base} because it is so explosive and so seldom useful.
To test whether increasing inference stream weights makes a difference on the
success rate, we tried enabling \infname{FluidSup} with different weight
increments~$\ParamPenaltyIncrease$ for \infname{FluidSup} inference queues. The
results are shown in Fig.~\ref{fig:penalties}. As expected, using a low
increment with \infname{FluidSup} is detrimental on TPTP. On this
benchmark set, \NumberOK{16} additional problems can be proved when
\infname{FluidSup} is enabled. The penalty mostly affects only proving time: All
but \NumberOK{2} of these problems were proved by using at least three different
values of $\ParamPenaltyIncrease$. On SH problems, the best result is obtained
when the rule is disabled as well. Unexpectedly, the next best result is obtained when
$\ParamPenaltyIncrease=1$.
% However, as the column for
% $\ParamPenaltyIncrease=16$ shows, nor should we use too high an increment, since that
% delays useful \infname{FluidSup} inferences. Interestingly, even
% though the configuration with $\ParamPenaltyIncrease=1$ proves the least problems overall, it proves
% \NumberNOK{7} problems not proved by \emph{base}, which is more than any other
% configuration we tried.

% {\tt
% \begin{itemize}
%   \item Not a single solution found using fluidsup on SH
%   \item 11 solutions found on TPTP, almost all (9) with at least 3 configurations.
%         Configurations usually only influence the time necessary to prove.
%   \item Even though it proves the least problems p1 is necessary in portfolio: with p=1 it takes 0.13 s seconds to prove NUM636\^{}2
%   and 5.95s with p16 (provable by all ps but not base).
%   \end{itemize}
% }

%%% @PETAR: Too little, not informative. What's a "shallow clause"? Not defined yet. --JB
%In informal experiments, we also tested limiting the applicability
%of some rules to shallow clauses, but this had no clear impact.

\section{Controlling the Use of Backends}
\label{sec:backends}

\newcommand{\ParamNumClauses}{\ensuremath{K_\mathrm{size}}}
\newcommand{\ParamTime}{\ensuremath{K_\mathrm{time}}}

Cooperation with efficient off-the-shelf first-order theorem provers is an
essential feature of higher-order theorem provers such as Leo-III
\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-2013-satallax}.
% but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
Those provers invoke first-order backends repeatedly
during a proof attempt and spend a substantial amount of time in backend
collaboration. Since \lamsup{} generalizes a highly efficient
first-order calculus, we expect that future efficient \lamsup{}
implementations will not benefit much from backends.
Nevertheless, experimental provers such
as Zipperposition can still gain a lot. We present some
techniques for controlling the use of backends.

In his thesis \cite[Sect.~6.1]{as-18-phd}, Steen extensively studies
the effects of using different first-order backends on the performance of
Leo-III. His results suggest that adding only one backend already substantially
improves the performance. To reduce the effort required for integrating multiple backends, we chose Ehoh \cite{ehoh-section} as our single
backend. Ehoh is an extension of the highly optimized superposition prover E 2.5
%%% PETAR: Check 2.5. Later we have 2.6 but that's for Ehoh++.
with support for higher-order features such as partial
application, applied variables, and interpreted Booleans.
%formulas occurring as arguments of function symbols.
On the one hand, Ehoh provides the efficiency of E while easing the translation from full
higher-order logic---the only missing syntactic feature is
$\lambda$-abstraction. On the other hand, Ehoh's higher-order reasoning
capabilities are limited. Its unification algorithm is essentially first-order,
and it cannot synthesize $\lambda$-abstractions.
% or instantiate predicate variables.

In a departure from Leo-III and other cooperative provers,
\confrep{}{instead of regularly invoking the backend, }%
we invoke \confrep{the backend}{it} at most once during a run of Zipperposition.
This is because most competitive higher-order provers, including Zipperposition, use a portfolio
mode in which many configurations are run for a short time, and we want to
leave enough time for native higher-order reasoning. Moreover, multiple
backend invocations tend to be wasteful, because currently each invocation
starts with no knowledge of the previous ones.

\looseness=-1
Only a carefully chosen subset of the available clauses are translated and sent
to Ehoh. Let $I$ be the set of \confrep{input clauses}{clauses representing the
input problem}. Given a proof state, let \confrep{$M = P \cup A$}{$M$ denote the
union of the current active and passive sets}, and let $M_\mathrm{ho}$ denote
the subset of $M$ that contains only clauses that were derived using at least
one \lamsup{} rule not present in regular superposition. We order the clauses in
$M_\mathrm{ho}$ by increasing derivation depth, using syntactic weight to break
ties. Then we choose all clauses in $I$ and the first \ParamNumClauses~clauses from
$M_\mathrm{ho}$ for use with the backend reasoner.
%We include all the clauses in $I$ since they constitute the backbone of the
%initial problem. -- Tautology. --JB
We leave out clauses in $M \setminus (I \cup M_\mathrm{ho})$ because Ehoh can rederive them.
We also expect large clauses with deep derivations to be less useful.

The remaining step is the translation of $\lambda$-abstractions. We implemented two
translation methods:\ $\lambda$-lifting \cite{tj-1985-lambdalift} and
$\cst{SKBCI}$ combinators \cite{da-1979-combtrans}. For
$\cst{SKBCI}$, we omit the combinator definition axioms, because they are
very explosive \cite{br-20-full-sup-w-combs}. A third mode simply omits clauses
containing $\lambda$-abstractions.

\ourparagraph{Evaluation and Discussion}
  %
  % \begin{enumerate}
  %   \item Different points when E is called, for different time periods
  %   \item Number of clauses from $M_\mathrm{ho}$ that is translated
  %   \item Different kinds of liftings
  % \end{enumerate}

  % We evaluated the following parameters of backend collaboration: the moment of
  % backend invocation, size of $M_\mathrm{ho}$, and $\lambda$-abstraction
  % translation method.
  In Zipperposition, we can adjust the CPU time allotted to Ehoh, Ehoh's own
  parameters, the point when Ehoh is invoked, the number \ParamNumClauses~of selected clauses from
  $M_\mathrm{ho}$, and the $\lambda$ translation method. We fix the time
  limit to 3~s, use Ehoh in \emph{autoschedule} mode, and focus on the last three
  parameters. In \emph{base}, collaboration with Ehoh is
  disabled (labeled $-$Ehoh).
  %Parameters are evaluated by enabling the collaboration with E and
  %varying parameter values.  -- Uninformative. --JB

  \begin{figure}[t]
    \noindent\hbox{}\hfill
    \begin{minipage}[t]{.46\linewidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
            & $-$Ehoh                   & $0.1$              & $0.25$          & $0.5$               & $0.75$      \\ \midrule
      TPTP & 1635$^\star$               & {\bf 1981}         & 1980            & 1979                & 1972 \\[0.5\jot]
      SH   & \colalign452$^\star$       & \colalign606       & {\bf \colalign608}    & \colalign600        & \colalign592 \\ \bottomrule
      \end{tabular}}
      \caption{Impact of the backend invocation point $\ParamTime$}
      \label{fig:invocation}
    \end{minipage}\hfill\hfill
    \begin{minipage}[t]{.46\textwidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
           & $-$Ehoh              & lifting              & \textsf{SKBCI}    & omitted \\ \midrule
      TPTP & 1635$^\star$         & {\bf 1980}           & 1877              & 1866 \\[0.5\jot]
      SH   & \colalign452$^\star$ & \colalign{\bf 608}   & \colalign577      & \colalign566 \\ \bottomrule
      \end{tabular}}
      \caption{Impact of the method used to translate $\lambda$-abstractions}
      \label{fig:translation}
    \end{minipage}%
    \hfill\hbox{}
  \end{figure}

  \begin{figure}[t]
    \centering
    \def\arraystretch{1.1}%
    \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}}\toprule
          & $-$Ehoh              & $16$               & $32$               & $64$           & $128$             & $256$         & $512$     \\ \midrule
     TPTP & 1635$^\star$         & {\bf 1985}         & 1980               & 1978           & 1968              & 1968          & 1919 \\[0.5\jot]
     SH   & \colalign452$^\star$ & \colalign606       & {\bf \colalign608} & \colalign600   & \colalign598      & \colalign596  & \colalign589 \\ \bottomrule
    \end{tabular}}
    \caption{Impact of the number of selected clauses~$\ParamNumClauses$}
    \label{fig:m-ho-size}
  \end{figure}

  \looseness=-1
  Ehoh is invoked after $ \ParamTime \cdot t $ CPU seconds, where $0 \le \ParamTime < 1$ and $t$
  is the total CPU time allotted to Zipperposition. Figure~\ref{fig:invocation}
  shows the effect of varying $\ParamTime$ when $\ParamNumClauses = 32$
  and $\lambda$-lifting is used. The evaluation confirms that using a highly
  optimized backend such as Ehoh greatly improves the performance of a
  less optimized prover such as Zipperposition.
  The figure indicates that it is preferable to invoke the backend early. We
  have indeed observed that if the backend is invoked late, small clauses with
  deep derivations tend to be present by then. These clauses might have been
  used to delete important shallow clauses already. But due to their derivation
  depth, they will not be translated. In such situations, it is better to
  invoke the backend before the important clauses are deleted.

  Figure~\ref{fig:translation} quantifies the effects of the three
  $\lambda$-abstraction translation methods. We fixed $\ParamTime = 0.25$ and
  $\ParamNumClauses=32$. The clear winner is $\lambda$-lifting.
  $\cst{SKBCI}$ combinators perform slightly better than omitting clauses
  containing $\lambda$-abstractions.

  \looseness=-1
  Figure~\ref{fig:m-ho-size} shows the effect of $\ParamNumClauses$ on
  performance, with $\ParamTime = 0.25$ and $\lambda$-lifting. Including a
  small number of higher-order clauses with the lowest weight performs better
  than including a large number of such clauses.

\section{Comparison with Other Provers}
\label{sec:comparison}

%The raw evaluation data for the previous experiments show that
Different choices of
parameters lead to noticeably different sets of proved problems. In an attempt
to use Zipperposition~2 to its full potential, we have created a portfolio mode
that runs up to 50 configurations in parallel during the allotted time. The portfolio
was designed to solve as many problems as possible from the TPTP benchmark set. To
provide some context, we compare Zipperposition~2 with the latest versions of
all higher-order provers that competed at CASC-J10:\ CVC4 1.9
\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
The provers were run using the same parameters as in CASC, but with updated
executables.
Note that
Vampire's higher-order schedule is optimized for running on a single core.
We also include E~2.7 (more precisely, Ehoh, its higher-order configuration), the first version of this prover to syntactically support
full higher-order logic, including $\lambda$-abstractions.
Semantically, E~2.7 is arguably the weakest among the listed provers:
It simply performs $o$-RW rewriting described in Sect.~\ref{sec:preprocessing} followed by
$\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
(a precursor of all three higher-order superposition calculi) on the preprocessed problem.

We use the same benchmark sets as elsewhere in this article. To imitate the
CASC-J10 setup, we use a 120~s wall-clock limit and a 960~s CPU limit.
We even carried out our evaluation on the 8-core CPU nodes that were used for
CASC-J10. We also ran Zipperposition in uncooperative mode, in which its
collaboration with a backend is disabled. Figure~\ref{fig:other-provers}
summarizes the results.

\begin{figure}[t]
  \centering
  \def\arraystretch{1.1}%
  \relax{\begin{tabular}{@{}l@{\hskip 2.5em}c@{\hskip 2em}c@{}} \toprule
                  & TPTP    & SH  \\ \midrule
   CVC4           & 1816    & 587   \\
   E              & 1980    & 676   \\
   Leo-III        & 2122    & 616  \\
   Satallax       & 2175    & 588  \\
   Vampire        & 2072    & 660   \\
   Zipperposition-uncoop & 2311 & 652 \\
   Zipperposition & {\bf 2412}    & {\bf 715}  \\ \bottomrule
  \end{tabular}}
  \caption{Comparison of competing higher-order theorem provers}
  \label{fig:other-provers}
\end{figure}

% Among the cooperative provers, Zipperposition is the one that depends the least
% on its backend, and its \emph{uncooperative} mode is only \NumberNOK{one}~problem
% behind Satallax's \emph{cooperative} mode. This confirms our hypothesis that
% \olamsup{} is a suitable basis for automatic higher-order reasoning.
% \confrep{This also}{The increase in performance due to the addition of an efficient backend}
% suggests that the implementation of this calculus in a modern first-order
% superposition prover such as E or Vampire would achieve markedly better results.
% Moreover, we believe that there are still techniques inspired by tableaux, SAT
% solving, and SMT solving that could be adapted and integrated in saturation
% provers.

The evaluation results corroborate the CASC results. They also
show that Zipperposition outperforms all other provers on SH benchmarks. This
confirms our hypothesis that \lamsup{} is a suitable basis for automatic
higher-order reasoning. Further confirmation is provided by the success rate of
Zipperposition's uncooperative version: Even without backend,
Zipperposition is substantially better than all other provers on TPTP
benchmarks, and it matches the performance of the top contenders on SH.
On the other hand, the increase in performance due to the addition
of an efficient backend suggests that the implementation of this calculus in a
modern first-order superposition prover such as E or Vampire would achieve
even better results.

We believe that there are still techniques inspired by
tableaux, SAT solving, and SMT solving that could be adapted and integrated in
saturation provers. In particular, there are still \NumberOK{25} TPTP problems
and \NumberOK{17} SH problems that can be proved by other provers but not by
Zipperposition.

\section{Discussion and Conclusion}
\label{sec:discussion}

%  In this paper, we proposed solutions to issues that arise when implementing a
%  higher-order prover based on \lamsup. Some of these also apply
%  to combinatory superposition. We also discussed many choices that can be made
%  during proof search and evaluated some of those choices.

\begin{sloppypar}
Back in 1994, Kohlhase \cite[Sect.~1.3]{mk-94-hores} was optimistic about the
future of higher-order automated reasoning:
%
\begin{quote}
  %The author believes that
  The obstacles to proof search intrinsic
  to higher-order logic may well be compensated by the greater expressive power
  of higher-order logic and by the existence of shorter proofs. Thus
  higher-order automated theorem proving will be practically as feasible as
  first-order theorem proving is now as soon as the technological backlog is made up.
\end{quote}
%
For higher-order superposition, the backlog consisted of designing calculus
extensions, heuristics, and algorithms that mitigate its weaknesses. In
this article, we presented such enhancements, justified their design, and
evaluated them. We explained how each weak point in the
higher-order proving pipeline could be improved, from preprocessing to reasoning
about formulas, to delaying unpromising or explosive inferences, to invoking a
backend. Our evaluation indicates that higher-order superposition is now the
state of the art in higher-order reasoning.
\end{sloppypar}

\looseness=-1
Higher-order extensions of first-order superposition have been
considered %from a theoretical perspective    -- The point is made later in the same para. --JB
by Bentkamp et al.\
\cite{bbtvw-21-sup-lam, bbcw-21-lfho} and Bhayat and Reger
\cite{br-19-restricted-unif, br-20-full-sup-w-combs}. They introduced proof calculi,
proved them refutationally complete, and suggested optional rules, but
they hardly discussed the practical aspects of higher-order superposition. Extensions
of SMT are discussed by Barbosa et al. \cite{brotb-19-ho-smt}.
Bachmair and Ganzinger \cite{bg-1992-nonclausal}, Manna and Waldinger
\cite{mw-1979-nonclausal}, and Murray \cite{nm-1982-nonclausal} have studied
nonclausal resolution calculi.

In contrast, there is a vast literature on practical aspects of first-order
reasoning using superposition and related calculi.
The literature evaluates various procedures and techniques
\cite{hv-09-unifalgs,rsv-15-playing-with-avatar}, literal and term order selection
functions \cite{hrsv-16-selsel}, and clause evaluation functions
\cite{sm-2016-clausesel, gs-20-clausesel}, among others. Our work joins the
select club of papers devoted to practical aspects of higher-order
reasoning
\cite{sb-15-beta,wskb-16-effective-norm,fb-2016-internal-guidance-satallax,benzmueller-et-al-05-can-ho-fo-coop}.

%In contrast, higher-order calculi implemented in competitive provers are usually discussed from a theoretical viewpoint.

%\section{Conclusion and Future Work}
%\label{sec:conclusion}

% We filled in this gap: we describe how
% We described some ways in which the explosion incurred by
% \lamsup{} can be kept under control. We showed how native
% support for a Boolean type can dramatically improve a higher-order prover's
% performance. For many choices that can be made during the higher-order proof
% search, we described heuristics and evaluated them. We also designed a
% saturation loop that enumerates infinite sets of higher-order inference
% conclusions, while maintaining the same behavior as the standard given clause
% procedure on first-order clauses.

As a next step, we plan to implement the described techniques in E.
%%% @PETAR: In E 2.6, basically. --JB
% Ehoh \cite{ehoh-section}, the $\lambda$-free higher-order extension of E.
We
expect the resulting prover to be substantially more efficient than
Zipperposition. Moreover, we want to investigate the proofs found by provers
such as CVC4 and Satallax but missed by Zipperposition. Finding the reason
behind why Zipperposition fails to prove specific problems will likely result in useful new techniques.

\def\ackname{Acknowledgment} % American English
\begin{acknowledgements}
We are grateful to the maintainers of StarExec for letting us use their service.
%Mathias Fleury helped us setup the ORCID icons.
Ahmed Bhayat and Giles Reger guided us
through details of Vampire 4.5. Ahmed Bhayat, Michael F\"arber,
Mathias Fleury, Predrag Jani\v ci\'c,
Mark Summerfield, and the anonymous reviewers suggested content, textual, and
typesetting improvements. We thank them~all.

Vukmirovi\'c, Bentkamp, and Blanchette's research has received funding from
the European Research Council (ERC) under the European Union's Horizon 2020
research and innovation program (grant agreement No.\ 713999, Matry\-osh\-ka).
%
Blanchette and Nummelin's research has
received funding from the Netherlands Organization for Scientific Research (NWO)
under the Vidi program (project No.\ 016.Vidi.189.037, Lean Forward).
\end{acknowledgements}

% Authors must disclose all relationships or interests that
% could have direct or potential influence or impart bias on
% the work:
%
% \section*{Conflict of interest}
%
% The authors declare that they have no conflict of interest.

% BibTeX users please use one of
\bibliographystyle{spmpsci}

\bibliography{bib}

\end{document}
% end of file template.tex

### 2022-03-09 bools/bools.tex
more

@@
 
-In the last decades, automatic theorem provers have been used
-successfully as backends to ``hammers'' in proof assistants
-\cite{ku-15-holyhammer, pb-12-sh} and to software
-verifiers \cite{fp-13-why3}. Most advanced provers, such as CVC4
-\cite{cbetal-11-cvc4}, E \cite{scv-19-e23}, and Vampire
-\cite{lkav-13-vampire}, are based on first-order logic, whereas
-most frontends that use them are based on versions of higher-order logic. Thus,
-there is a large gap in expressiveness between front- and backends.
-%
-\looseness=-1
-This gap is bridged using well-known translations from higher-order to
-first-order logic \cite{ar-70-hol,
-mp-08-trans}. However, as shown in Chapter \ref{ch:ehoh}, translations are usually less efficient than native support. The
+In Chapter~\ref{ch:intro} we motivated our work by the need to bridge the gap
+between higher-order frontends and first-order backends. This gap is
+traditionally bridged using well-known translations from higher-order to
+first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
+\ref{ch:ehoh}, translations are usually less efficient than native support. The
 distinguishing features of higher-order logic used by proof assistants that the
-translation must eliminate include
-$\lambda$-binders, function extensionality---the property that functions are
-equal if they agree on every argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x \, z \ieq y \,z)
-\iimplies x \ieq y$, and formulas occurring as arguments of function symbols \cite{mp-08-trans}.
+translation must eliminate include $\lambda$-binders, function
+extensionality---the property that functions are equal if they agree on every
+argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
+\, z \ieq y \,z) \iimplies x \ieq y$, and formulas occurring as arguments of
+function symbols \cite{mp-08-trans}.
 
@@
 We created two rules that are syntactically similar to \infname{FP} but
-are adapted for higher-order logic with one key distinction -- we do not perform any translation:
+are adapted for higher-order logic with one key distinction---we do not perform any translation:
 %
@@
 that led us to come up with the following modifications. First, it often
-suffices to perform \infname{PI} only on initial clauses -- which is why we
+suffices to perform \infname{PI} only on initial clauses---which is why we
 allow the rule to be applied only to the clauses created using at most $k$
 generating inferences. Second, if the rule was used in the proof, its premise
-is usually only used as part of that inference -- which is why we implemented
+is usually only used as part of that inference---which is why we implemented
 a version of \infname{PI} that removes the clause after all possible \infname{PI}
@@
 \textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
-$$
-\namedinference{AbsSup}
+\begin{align*}
+  & \namedinference{AbsSup}
   {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
-  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} $$
-\pagebreak[2]
-\begin{align*}
+  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} \\
   &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
@@
 $$ \namedinference{Interpret$\lambda$}{C[\lam{\tuplen{x}}{f}]}
-{(\forall {\tuplen{x}}.\, f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\itrue}] \qquad
-(\forall {\tuplen{x}}.\, \inot f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\ifalse}]} $$
+{(\iforall {\tuplen{x}}.\, f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\itrue}] \qquad
+(\iforall {\tuplen{x}}.\, \inot f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\ifalse}]} $$
 %
@@
 
-\subsection{Evaluation of the Rules}
-
+\ourpara{Evaluation of the Rules}
 \looseness=-1
@@
 called \emph{base} (\emph{b}). Since we are testing a single configuration, we
-used the CPU time limit of \NumberOK{15} s -- roughly the time a single configuration is
+used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
 given in a portfolio mode. Configuration \emph{b} uses the pragmatic variant
@@
 
-\subsection{Comparison with Other Higher-Order Provers}
+\ourpara{Comparison with Other Higher-Order Provers}
 
@@
 
-Leo-III and Satallax are cooperative theorem provers -- they periodically invoke
+Leo-III and Satallax are cooperative theorem provers---they periodically invoke
 first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
@@
 cbetal-11-cvc4}. Those two provers are built from the ground up as
-higher-order provers -- treatment of exclusively higher-order issues such as
+higher-order provers---treatment of exclusively higher-order issues such as
 extensionality or choice is built into them usually using explosive rules. Those

### 2022-03-09 unif/unif.tex
more

@@
       \item[Decomposition] A decreases.
-      \item[Failure] Trivial -- represents a terminal node.
+      \item[Failure] Trivial---represents a terminal node.
       \item[Solution] This transition applies on 
@@
   \, (\cst{f} \, \cst{a}) \unif \cst{a}, \allowbreak F_2 \, (\cst{f} \, \cst{a}) \unif G \,
-  \cst{a}\}$. Again, we can only imitate $\cst{a}$ for $F_1$ -- building a new 
+  \cst{a}\}$. Again, we can only imitate $\cst{a}$ for $F_1$---building a new 
   binding $\sigma_2 = \{F_1 \mapsto \lambda x.\, \cst{a} \}$. Finally, this
@@
 The fundamental idea of first-order fingerprint indexing carries over to
-higher-order terms -- 
-application of a substitution does not change the rigid skeleton of a term.
+higher-order terms---application of a substitution does not change the rigid skeleton of a term.
 However, to extend fingerprint indexing to
@@
 procedure's main bottlenecks. The raw evaluation data show that on TPTP benchmarks, complete and pragmatic configurations
-differ in the set of problems they solve -- cv solves 19
+differ in the set of problems they solve---cv solves 19
 problems not solved by pv$^{4}_{2222}$, whereas pv$^{4}_{2222}$ solves 34
@@
 higher-order unification. Our procedure performs exceptionally well
-on these problems -- it solves all of them, usually
+on these problems---it solves all of them, usually
 faster than other competitive higher-order provers.

### 2022-03-10 unif/unif.tex
more

@@
     The head of $\varrho_j a$ must be $b$ because $b$ is rigid, and $\varrho_j$ unifies $u$ and $v$. 
-    Since $\varrho_j a$ is not projective, that means that $b$ is not a bound variable. 
-    Therefore, $b$ must be a constant.
+    As $\varrho_j a$ is not projective, that means that $b$ is not a bound variable. 
+    Thus, $b$ must be a constant.
     Then Lemma \ref{lem:jp-imitation} (imitation) applies.
@@
 
+  \end{description}
 We have now constructed a run 
@@
 completing the proof. \qedhere
-    \end{description}
     \end{proof}
 
-\pagebreak[2]    
 \section{A New Decidable Fragment}
@@
 as their canonical representative. Unlike previously, in this section we view
-unification constraints $s \unif t$ as ordered pairs.
-
-The following rules, however, are stated modulo orientation. This means that 
-if a rule applies to a constraint $s \unif t$, then it also applies to a constraint $t
-\unif s$. The PT transition rules, adapted for our presentation style, are as
+unification constraints $s \unif t$ as {\bf ordered} pairs. The PT transition rules, adapted for our presentation style, are as
 follows:
-
-\begin{description}[labelwidth=\widthof{\rm\unifrulename{\bf Decomposition}}]
+\begin{description}[labelwidth=\widthof{\rm\unifrulename{Decomposition}}]
     \item[\rm\textsf{Deletion}] 
@@
     \\ where $a$ and $b$ are different rigid heads
-    \item[\rm\textsf{Solution}] 
+    \item[\rm\textsf{SolutionL}] 
       $\{\selected{\lambda \overline{x}.\, F\,\overline{x} \unif \lambda \overline{x}.\,t} \} \uplus E \ptarrow{\varrho} \varrho E$
       \\where $F$ does not occur in $t$, $t$ does not have a flex head, and $\varrho = \{ F \mapsto \lambda \overline{x}.\,t \}$
-    \item[\rm\textsf{Imitation}]
+      \item[\rm\textsf{SolutionR}] 
+      $\{\selected{\lambda \overline{x}.\,t \unif \lambda \overline{x}.\, F\,\overline{x}} \} \uplus E \ptarrow{\varrho} \varrho E$
+      \\where $F$ does not occur in $t$, $t$ does not have a flex head, and $\varrho = \{ F \mapsto \lambda \overline{x}.\,t \}$ 
+    \item[\rm\textsf{ImitationL}]
       $\{\selected{\lambda \overline{x}. \, F \, \overline{s}_m \unif \lambda
@@
       \overline{s}_m  \unif  t_1, \ldots,  G_n \, \overline{s}_m \allowbreak \unif t_n\} \uplus E)$ \\
+      \looseness=-1 where $\varrho = \{ F \mapsto \lambda \overline{x}_m. \, \cst{f} \, (G_1
+      \, \overline{x}_m) \ldots (G_n \, \overline{x}_m) \}$,
+      $\overline{G}_n$ are correctly typed fresh variables
+    \item[\rm\textsf{ImitationR}]
+      $\{\selected{\lambda
+      \overline{x}. \, \cst{f} \, \overline{t}_n \unif \lambda \overline{x}. \, F \, \overline{s}_m  }\} \uplus E \ptarrow{\varrho}
+      \varrho(\{  t_1 \unif  G_1 \,
+      \overline{s}_m, \ldots,  t_n \allowbreak \unif G_n \, \overline{s}_m\} \uplus E)$ \\
       where $\varrho = \{ F \mapsto \lambda \overline{x}_m. \, \cst{f} \, (G_1
       \, \overline{x}_m) \ldots (G_n \, \overline{x}_m) \}$,
-      $\overline{G}_n$ are fresh variables of appropriate types
-    \item[\rm\textsf{Projection}]
+      $\overline{G}_n$ are correctly typed fresh variables
+    \item[\rm\textsf{ProjectionL}]
       $\{\selected{\lambda \overline{x}. \, F \, \overline{s}_m \unif \lambda
@@
       \overline{x}_m) \ldots (G_j \, \overline{x}_m)\}$, $\overline{G}_j$ are
-      fresh variables of appropriate types
+      correctly typed fresh variables
+    \item[\rm\textsf{ProjectionR}]
+      $\{\selected{\lambda \overline{x}. \, a \, \overline{t} \unif \lambda \overline{x}. \, F \, \overline{s}_m }\} \uplus E \ptarrow{\varrho} 
+      \varrho(\{ a \, \overline{t} \unif s_i \, (G_1 \,
+      \overline{s}_m) \ldots (G_j \, \overline{s}_m)\} \uplus E)$ \\
+      where $\varrho =\{F \mapsto \lambda \overline{x}_m. \, x_i \, (G_1 \,
+      \overline{x}_m) \ldots (G_j \, \overline{x}_m)\}$, $\overline{G}_j$ are
+      correctly typed fresh variables
 \end{description}
+\pagebreak[2]
 \looseness=-1
@@
 and of descendant constraints of
-\textsf{Projection} transitions with $j=0$ (i.e., for $x_i$ of base type), in that order of priority. In the
+\textsf{ProjectionL} and \textsf{ProjectionR} transitions with $j=0$ (i.e., for $x_i$ of base type), in that order of priority. In the
 remainder of this section we consider only admissible selection functions,
@@
 Here, the \emph{abstraction-free size} is inductively defined by
-$\text{afsize}(F) = 1$; $\text{afsize}(x) = 1$; $\text{afsize}(\cst{f}) = 1$;
+$\text{afsize}(F) = 1$; $\text{afsize}(x) = 1$; $\text{afsize}(\cst{f})\allowbreak= 1$;
 $\text{afsize}(s\,t) = \text{afsize}(s) + \text{afsize}(t)$;
@@
   one of $s$ and $t$ must have a rigid head. We distinguish several cases based on the
-  form of $s \unif t$ (modulo constraint orientation):
+  form of $s \unif t$:
 
@@
     than $E_0$ and still unifiable by $\varrho$. By the induction hypothesis, the preunifier is reachable.
-    \item 
-    \looseness=-1
-    $\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda
+    \item $\lambda \overline{x}. \, a \, \overline{s}_m \unif \lambda
     \overline{x}. \, b \, \overline{t}_n$, where $a$ and $b$ are rigid: since
@@
     \item $\lambda \overline{x}.\, F\,\overline{x} \unif \lambda \overline{x}.\,t$ where $t$
-    has a rigid head and $F$ does not occur in $t$: \textsf{Solution} applies. Huet showed \cite[proof of L5.1]{gh-75-unification}
+    has a rigid head and $F$ does not occur in $t$: \textsf{SolutionL} applies. Huet showed \cite[proof of L5.1]{gh-75-unification}
     that in this case $\varrho = \varrho\sigma_1$, where $\sigma_1 = \{ F \mapsto \lambda\overline{x}.\,t \}$.
@@
     Then $\varrho = \varrho_1 \sigma_1$.
-    \looseness=-1
     Since $\varrho$ unifies $E'_0$, $\varrho_1$ unifies $E_1 = \sigma_1 E'_0$. Clearly, $\varrho_1$ is smaller
@@
     $\theta$, and $\sigma = \sigma' \sigma_1$ are as wanted in the lemma statement.
-    \item \looseness=-1 
-    $\lambda
+
+    \item $\lambda \overline{x}.\,t \unif \lambda \overline{x}.\, F\,\overline{x}$ where $t$
+    has a rigid head and $F$ does not occur in $t$: \textsf{SolutionR} applies, and the case is analogous to the previous one.
+
+    \item $\lambda
     \overline{x}.\,F\,\overline{s}_m~\unif~\lambda \overline{x}.\,a \,
     \overline{t}$ where $a$ is rigid: depending on the value of $\varrho$, we can
-    either take an \textsf{Imitation} or \textsf{Projection} step. In either case, we show that
+    either take an \textsf{ImitationL} or \textsf{ProjectionL} step. In either case, we show that
     the measure reduces. Let $\varrho F = \lambda \overline{x}_m. \, b \, \overline{u}_n$ where $b$
     is either a constant or a bound variable. If $b$ is a constant, we choose
-    the \textsf{Imitation} step; otherwise we choose the \textsf{Projection} step. In either case,
+    the \textsf{ImitationL} step; otherwise we choose the \textsf{ProjectionL} step. In either case,
     $\sigma_1 = \{ F \mapsto \lambda \overline{x}_m. \, b \, (G_1 \, \overline{x}_m) \ldots
@@
     sequence from the induction hypothesis, we can conclude that the preunifier is reachable.
-    \qedhere
+
+    \item $\lambda \overline{x}.\,a \, \overline{t} \unif \lambda
+    \overline{x}.\,F\,\overline{s}_m$ where $a$ is rigid:  either \textsf{ImitationR} or \textsf{ProjectionR} applies
+    and the case is analogous to the previous one. \qedhere
   \end{itemize}
@@
 \looseness=-1 
-In other words, we extend the above preunification decidability result
-for linear strictly solid terms along two axes: we create an oracle
+Put differently, we extend the preunification decidability result
+for linear strictly solid terms along two axes: We create an oracle
 for the full unification problem and we lift the
@@
 \ref{unif:lemma:solid-vars-diff-mgu}). Our results are combined together in Theorem
-\ref{unif:thm:oracle}.
+\ref{unif:thm:oracle}. Note that we use names \textsf{Solution}, \textsf{Imitation}, and
+\textsf{Projection} when only one orientation of the corresponding rule is applicable.
 
@@
   maintains the condition that the terms remain solid. Second, we have to show
-  that variables on left-hand sides remain isolated. For \textsf{Imitation} and
-  \textsf{Projection} rules, the preservation of this invariant is obvious.
-  Later, we also show that \textsf{Solution} preserves it. Third, since $s$ and
+  that variables on left-hand sides remain isolated. For \textsf{ImitationL(R)} and
+  \textsf{ProjectionL(R)} rules, the preservation of this invariant is obvious.
+  Later, we also show that \textsf{SolutionL(R)} preserves it. Third, since $s$ and
   $t$ share no variables, and $s$ is linear, no rule can introduce a variable
@@
     \item[Failure] Trivial. 
-    \item[Solution] This rule applies in two cases:
-
-    \begin{itemize}
-      \item $F \unif t$: since $F$ is isolated, A is unchanged, B is not
+    \item[SolutionL]
+    $F \unif t$: since $F$ is isolated, A is unchanged, B is not
     increased, and C reduces. All variables on left-hand sides remain isolated
     since they are unaffected by this substitution.
-      \item $t \unif F $: since $F$ does not occur on any left-hand side and the
+    \item[SolutionR] $t \unif F $: since $F$ does not occur on any left-hand side and the
     head of $t$ must be a constant or bound variable (otherwise the rule would
-      not be applicable), $A$ reduces. Even though $t$ might contain some free
+    not be applicable), A reduces. Even though $t$ might contain some free
     variables and $F$ can have multiple occurrences on right-hand sides, all
@@
     removed, all of them either disappear or end up on right-hand sides.
-    \end{itemize}
-    \item[Imitation] We distinguish the following two forms of the selected constraint:
-    \begin{itemize}
-      \item $\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda \overline{x}. \,
+    We distinguish the following two forms of the selected constraint:
+      \item[ImitationL] $\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda \overline{x}. \,
       \cst{f} \, \overline{u}_m$: 
@@
       $\{H_i \, \overline{v}_n \unif u_i \mid 1\le i\le m\}$
-      and apply the \textsf{Imitation} binding
+      and apply the \textsf{ImitationL} binding
       $F \mapsto \lambda \overline{x}_n.\, \cst{f} \, (H_1 \, \overline{x}_n) \allowbreak
@@
       Since $F$ is isolated, it does not occur on any right-hand side, and hence C decreases.
-      \item $\lambda \overline{x}. \, \cst{f} \, \overline{u}_m \unif \lambda \overline{x}. \, F \, \overline{v}_n $:
-      applying the \textsf{Imitation} transition 
+      \item[ImitationR] $\lambda \overline{x}. \, \cst{f} \, \overline{u}_m \unif \lambda \overline{x}. \, F \, \overline{v}_n $:
+      applying the \textsf{ImitationR} transition 
       as above will reduce A since $F$ cannot appear on any
       left-hand side.
-    \end{itemize}
-    \item[Projection] Similar to the previous transition rule, we have two cases:
-      \begin{itemize}
-        \item$\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda
+    \item[ProjectionL]
+        $\lambda \overline{x}. \, F \, \overline{v}_n \unif \lambda
         \overline{x}. \, a \, \overline{u}_m$: if $a$ is a bound variable,
@@
         project $F$ onto argument $v_i$ equal to $a$. Then we proceed like for
-        \textsf{Imitation}. If $a$ is not a bound variable, then we have to
+        \textsf{ImitationL}. If $a$ is not a bound variable, then we have to
         project to some base-type term $v_j$ (otherwise \textsf{Failure} would
@@
         or, if no free variables occur in the right-hand side, reduce C while not increasing A and B.
-        \item $\lambda \overline{x}. \, a \, \overline{v}_n \unif \lambda
+    \item[ProjectionR] $\lambda \overline{x}. \, a \, \overline{v}_n \unif \lambda
         \overline{x}. \, F \, \overline{u}_m$: 
@@
         will either yield  \textsf{Failure} or enable \textsf{Decomposition}, reducing A. \qedhere
-      \end{itemize} 
     \end{description}
@@
   An MGU $\sigma$ for the given problem is
+  \pagebreak[2]
     \begin{align*} 

### 2022-03-10 bools/bools.tex
more

@@
     inference rules that have been used only in a first-order setting, uses some
-    well-known rules previously implemented in higher-order provers, as well as new rules.
+    rules previously implemented in higher-order provers, as well as new rules.
     We have implemented the approach in the Zipperposition
@@
 between higher-order frontends and first-order backends. This gap is
-traditionally bridged using well-known translations from higher-order to
+traditionally bridged using translations from higher-order to
 first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
@@
 distinguishing features of higher-order logic used by proof assistants that the
-translation must eliminate include $\lambda$-binders, function
+translation must eliminate include $\lambda$ binders, function
 extensionality---the property that functions are equal if they agree on every
 argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
-\, z \ieq y \,z) \iimplies x \ieq y$, and formulas occurring as arguments of
+\, z \ieq y \,z) \iimplies x \ieq y$---and formulas occurring as arguments of
 function symbols \cite{mp-08-trans}.
@@
 \looseness=-1
-A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
-extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate
-$\lambda$-binders and functional extensionality. Kotelnikov et al.\
+
+% A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
+% extension of superposition for extensional Boolean-free higher-order logic. The
+% extension removes the need to translate $\lambda$-binders and functional
+% extensionality. 
+As mentioned in Sect.~\ref{sec:pre:ho-sup-calculi}, Bentkamp et al. developed
+\lsup{}, a complete higher-order calculus that alleviates the need to translate
+$\lambda$-terms and function extensionality. Kotelnikov et al.\
 \cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
@@
 \looseness=-1 To fully bridge the gap between higher-order and first-order tools,
-we combine the two approaches: we use the efficient higher-order superposition
-calculus and extend it with inference rules that reason with
+we combine the two approaches: We take \lsup{} as our basis and  extend it with inference rules that reason with
 Boolean terms. In early work, Kotelnikov et al.\ 
@@
 
-\looseness=-1 An important feature that we inherit by building on top of Bentkamp
-et al. \cite{bbtvw-21-sup-lam} is support for (function) extensionality. Moving to higher-order
-logic with Booleans also means that we need to consider \emph{Boolean extensionality}: $\iforall pq. \, (p
-\iequiv q) \iimplies p \ieq q$. We extend the rules of Bentkamp et al.\ 
-that treat function extensionality to also treat Boolean extensionality.
+\looseness=-1 An important feature that we inherit by building on top of \lsup{}
+is support for function extensionality. Moving to higher-order logic with
+Booleans also means that we need to consider \emph{Boolean extensionality}:
+$\iforall pq. \, (p \iequiv q) \iimplies p \ieq q$. We extend the rules of
+\lsup{} that treat function extensionality to also treat Boolean
+extensionality.
 
@@
 for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
-implemented rules that are inspired by the ones used in the
-higher-order provers Leo-III \cite{sb-21-leo3} and Satallax
-\cite{cb-12-satallax}, such as elimination of Leibniz equality, primitive
-instantiation and treatment of choice operator \cite{pa-01-classical-ty-thy}. We
-have also designed new rules that use higher-order unification to resolve
-Boolean formulas that are hoisted to literal level, delay clausification of nonatomic
-literals, reason about formulas under $\lambda$-binders, and many others. Even
-though the rules we use are inspired by the ones of refutationally complete
-higher-order provers, we do not guarantee completeness of our extension of
-$\lambda$-superposition.
+implemented rules that are inspired by the ones used in the higher-order provers
+Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
+elimination of Leibniz equality, primitive instantiation, and treatment of choice
+operator \cite{pa-01-classical-ty-thy}. We have also designed new rules that use
+higher-order unification to resolve Boolean formulas that are hoisted to literal
+level, delay clausification of nonatomic literals, reason about formulas under
+$\lambda$-binders, and many others. Even though the rules we use are inspired by
+the ones of refutationally complete higher-order provers, we do not guarantee
+completeness of our extension of \lsup{}. Using the insights that we gained with
+this incomplete extension of \lsup{} to full higher-order logic, Bentkamp
+designed a complete calculus for full higher-order logic---\osup{}
+\cite{bbtv-21-full-ho-sup}.
 
@@
 \cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
-to implement their higher-order superposition calculus. We further extend
+to implement \lsup{}. We further extend
 their implementation.
@@
 compared them to full higher-order provers CVC4, Leo-III,  Satallax and Vampire.
-The results suggest that it is beneficial to natively support Boolean reasoning
--- the approach outperforms preprocessing-based approaches. Furthermore, it is
+The results suggest that it is beneficial to natively support Boolean reasoning---the approach outperforms preprocessing-based approaches. Furthermore, it is
 very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
@@
 
-We base our work (and parts of the following text) on Bentkamp et al.'s \cite{bbtvw-21-sup-lam} extensional
-polymorphic clausal higher-order logic. We extend the syntax of this logic by
-adding logical connectives to the signature. The semantic of the logic
-is extended by interpreting Boolean type $o$ as a two-element domain. This
-amounts to extending Bentkamp et al's fragment of higher-order logic to
-full-higher order logic (HOL). Taking a different perspective, this logic is an extension of the one
-presented in Chapter \ref{ch:ehoh} with $\lambda$ abstraction, polymorphic
-types, and native logical connectives. For reference, we provide precise definition of this polymorphic logic.
+We base our work (and parts of the following text) on Bentkamp et al.'s
+\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. We
+extend the syntax of this logic by adding logical connectives to the signature.
+The semantic of the logic is extended by interpreting Boolean type $o$ as a
+two-element domain. This amounts to extending the logic to full higher-order
+logic. Taking a different perspective, this logic is an extension of the one
+presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
+types, and native logical connectives. For reference, we provide a
+definition of this polymorphic logic.
 
@@
 the distinction introduced in Sect.~\ref{sec:pre:hol}. Free term variables from the set $\VV$ are written
-$F,G,X,Y \ldots$ and we denote their types as $X : \tau$. Bound term variables are written $x,y,z,\ldots$,
+$F,G,X,Y, \ldots$ and we denote their types as $X : \tau$. Bound term variables are written $x,y,z,\ldots$,
 and their types are similarly denoted. When the type is not
@@
 constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
-formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q})
-\llor \inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while
-$f|_{1.2.1}$ is a nested Boolean, as well as its subterms. First-order logic
-allows only top-level Booleans, whereas nested Booleans are characteristic for
-higher-order logic.
+formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \ior
+\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is a
+nested Boolean, as well as its subterms. First-order logic allows only top-level
+Booleans, whereas nested Booleans are characteristic for higher-order logic. 
+Note that while our calculus works with clausal representation of the problem,
+its higher-order nature allows representing formula $f$ as a singleton clause
+containing only literal $f$.
 
@@
 Zipperposition already had some support for Boolean reasoning before we
-started extending the calculus of Bentkamp et al. In this section, we first
+started extending \lsup{}. In this section, we first
 describe the internals of Zipperposition responsible for reasoning with
-Booleans. We continue by describing \NumberOK{15} rules that we have
+Booleans. We continue by describing the rules that we have
 implemented. For ease of presentation we divide them in three categories.
@@
 
-In Zipperposition, applications are represented in flattened, spine notation. In
+In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
+This means that higher-order term $(\cst{f} \, \cst{a}) \, cst{b}$ is
+represented the same as first-order term $\cst{f}(\cst{a}, \cst{b})$. In
 addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
@@
 (\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
-internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. 
-Zipperposition's support for $\lambda$-terms is used to represent quantified nested
-Booleans: Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
-$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of the
-input problem, no nested Booleans will be modified or renamed using fresh
+internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
+support for $\lambda$-terms is used to represent quantified nested Booleans:
+Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
+$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
+the input problem, no nested Booleans will be modified or renamed using fresh
 predicate symbols.
@@
 knowledge, pioneered the approach of extending a first-order superposition prover to support
-nested Booleans. They call effects of including the axiom $\iforall
+nested Booleans. They call including the axiom $\iforall
 p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
@@
 into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
-the axiom with the inference rule \emph{FOOL Paramodulation} (\infname{FP}):
+the axiom with the inference rule \emph{FOOL paramodulation} (\infname{FP}):
 \pagebreak[2]
@@
 $\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
-Booleans  to first-order logic without interpreted Booleans; thus, symbols $\itrue$ and $\ifalse$, and type $o$
+Booleans  to first-order logic without interpreted Booleans; thus, the symbols $\itrue$ and $\ifalse$, and the type $o$
 correspond to proxy symbols and types introduced during the translation.
@@
 \looseness=-1
-Clearly, the prover that uses the rules should not include them both at the same time.
+The prover that uses the rules should not include them both at the same time.
 In addition, since
@@
 we change the first requirement on the order to $\ifalse \succ
-\itrue$.
+\itrue$, making $\itrue$ the smallest symbol.
 
@@
 $f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
-called immediate clausification (\infname{IC}):
-%
-$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
+called immediate clausification (\infname{IC}), where $\tuple{D}{m}$ are defined below:
+%
+$$ \namedsimp{IC}{C}{D_1 \; \cdots \; D_m} $$
 %
@@
 \looseness=-1
-If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result in many redundant clauses. Consider the following
-example: let $\cst{p} : o \rightarrow o$, $\cst{a} : o$ and consider a clause
-set containing $\cst{p} \, (\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq
-\itrue$. Then, the clause $C = \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
-derived in eight ways using the rules, depending on which nested Boolean subterm was
-chosen for the inference. In general, if a clause has a subterm occurrence of the form $\cst{p}^n \, \cst{a}$,
-where both $\cst{p}$ and $\cst{a}$ have result type $o$, the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in $2^{n-1}$ ways.
+If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result
+in many redundant clauses. Consider the following example: let $\cst{p} : o
+\rightarrow o$, $\cst{a} : o$ and consider a clause set containing $\cst{p} \,
+(\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq \itrue$. Then, the clause $C
+= \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in
+eight ways using the rules, depending on which nested Boolean subterm was chosen
+for the inference. In general, if a clause has a subterm occurrence of the form
+$\cst{p}^n \, \cst{a}$, where both $\cst{p}$ and $\cst{a}$ have result type $o$,
+the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
+derived in at least $2^{n-1}$ ways.
 % Boolean subterms occurring in different
@@
 The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
-eagerly simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
-the formula $\cst p\land\itrue$ is replaced by $\cst p$. To simplify nested Booleans we implemented the rule
+eagerly simplifies the input problem using Boolean equivalences before clausifying it. For example,
+the formula $\cst p\iand\itrue$ is replaced by $\cst p$. To simplify nested Booleans we implemented the rule \infname{BoolSimp}
+parameterized by a set of rewrite rules E:
 
@@
 \]
-where $f \longrightarrow g \in E$ runs over fixed set of rewrite rules $E$,
-and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
+where $f \longrightarrow g \in E$ and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
 described by Nonnengart and Weidenbach \cite[Sect.~3]{nw-01-small-cnf}. This set contains
@@
 To achieve refutational completeness of higher-order resolution and similar
-calculi it is necessary to instantiate variables with result type $o$,
+calculi, it is necessary to instantiate variables with result type $o$,
 \emph{predicate variables}, with arbitrary formulas
@@
 $\iand$). Since such an approximation is not only necessary for completeness of
-some calculi, but very useful in practice, we implemented the \emph{primitive
+some calculi, but possibly useful in practice, we implemented the \emph{primitive
 instantiation} (PI) rule:
 %
-$$ \namedinference{PI} {C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq}
+$$ \namedinference{PI} {C \llor (\lam{\tuple{x}{}}{P \, \tuplen{s}}) \, \dot{\eq}
 \, t} {\substcl{\{ P
-\mapsto f \}}{C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq} \, t}  } $$
+\mapsto f \}}{C \llor (\lam{\tuple{x}{}}{P \, \tuplen{s}}) \, \dot{\eq} \, t}  } $$
 %
@@
 Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
-before we started our modifications. The rule has different modes that generate
+before we started our modifications. Its implementation contains the following modes that generate
 sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
-\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
-$\star$ is an element of a set of logical constants $P = \{ \iand, \ior,
+\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$. In  $\mathit{Imit}_{\star}$,
+$\star$ is an element of a set of logical constants $S = \{ \iand, \ior,
 \mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
-contains imitations (for $P$) of all elements of $P$. Mode $\emph{Pragmatic}$
-contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
-such that $i\ineq j$ and  $\tau_i = \tau_j$, it contains $\lam{\tuplen{x}}{x_i
+generates imitations (for $P$) of all elements of $S$. Mode $\emph{Pragmatic}$
+generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
+such that $i\ineq j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
 \ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
-\tau_j = o$, then it contains $\lam{\tuplen{x}}{x_i \iand x_j}$ and
+\tau_j = o$, then it generates $\lam{\tuplen{x}}{x_i \iand x_j}$ and
 $\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
-contains $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ contains
-imitations of $\itrue$, $\ifalse$ and $\star$ (except for $\mathit{Imit_{\iforall\iexists}}$ which contains imitations
-of both $\iforall$ and $\iexists$).
-
-\looseness=-1
-While experimenting with our implementation we have noticed some proof patterns
-that led us to come up with the following modifications. First, it often
-suffices to perform \infname{PI} only on initial clauses---which is why we
-allow the rule to be applied only to the clauses created using at most $k$
-generating inferences. Second, if the rule was used in the proof, its premise
-is usually only used as part of that inference---which is why we implemented
-a version of \infname{PI} that removes the clause after all possible \infname{PI}
-inferences have been performed. We observed that the mode \emph{Imit}$_\star$ is useful in practice since often only a single
-approximation of a logical symbol is necessary.
-
-\looseness=-1
-Efficiently treating axiom of choice is notoriously difficult for higher-order provers. Andrews formulates
+generates $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ generates
+imitations of $\itrue$, $\ifalse$ and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ which generates imitations
+of both $\iforall$ and $\iexists$.
+
+\looseness=-1
+While experimenting with our implementation we noticed some proof patterns that
+led us to come up with the following modifications. First, it often suffices to
+perform \infname{PI} only on initial clauses---which is why we allow the rule to
+be applied only to the clauses created using at most $k$ generating inferences.
+Second, if the rule was used in the proof, its premise is usually only used as
+part of that inference---which is why we implemented a version of \infname{PI}
+that removes the clause after all possible \infname{PI} inferences have been
+performed. We observed that the mode \emph{Imit}$_\star$ is useful in practice
+since often approximation of a single logical symbol suffices.
+
+\looseness=-1
+The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
 this axiom as $\iforall p.\, (\iexists
@@
 operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
-\, (\varepsilon \, P) \eq \itrue$. Since term $P \, X$ matches any Boolean
+\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
 term in the proof state, this axiom is very explosive. Therefore, Leo-III
-\cite{sb-21-leo3} deals with the choice operator on the calculus
-level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
+\cite{sb-21-leo3} recognizes definitions of all symbols that correspond to choice and deals with
+them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
 P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
@@
 functions, but generalized the rule for creating the instance of
-the axiom (assuming $\xi \in \mathit{CF}$):
+the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
 %
@@
 %
-
 Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
 arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
-the variable $X$ can later be replaced by imitation of logical symbols to create more
+the variable $X$ can later be replaced by imitations of logical symbols to create more
 complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
-\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, based on
-Zipperposition parameters, $D$ will either be deleted or kept. Note that $D$
-will not subsume its instances, since the matching algorithm of Zipperposition is
-too weak for this.
-\looseness=-1
-
-Most provers natively support extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
+\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
+Zipperposition options, $D$ will either be deleted or kept. Note that $D$
+will not subsume its instances, since the matching algorithm Zipperposition uses is
+too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
+\looseness=-1
+
+Most provers natively support both functional and Boolean extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
 first-order unification to return unification constraints consisting of pairs of
 terms of functional type, whereas Steen relies on the unification rules of
-Leo-III's calculus \cite[Section 4.3.3.]{as-18-phd} to deal with extensionality.
-Bentkamp et al \cite{bbtvw-21-sup-lam} altered core generating inference rules of the superposition calculus
-to support extensionality. Instead of requiring that terms
+Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
+As a pragmatic extension of the lambda-superposition calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
+to support functional extensionality. Instead of requiring that terms
 involved in the inference are unifiable, it is required that they can be
@@
 \textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
+\pagebreak[2]
 \begin{align*}
@@
 \noindent\looseness=-1
-Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are extensional
-versions of superposition, equality resolution and equality factoring
-(Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of these three
-rules. In each of the rules, $\sigma$ is a most general unifier of the types of
-$s$ and $s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
+In each of the rules, $\sigma$ is a most general unifier of the types of $s$ and
+$s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
 (s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
@@
 replaced by the condition that at least one $s_i$ is of functional type and that
-$n>0$. This rule is easily extended to support Boolean extensionality by
+$n>0$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
+extensional versions of superposition, equality resolution and equality
+factoring (Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of
+these three rules. We extend this rule to support Boolean extensionality by
 requiring that at least one $s_i$ is of functional or type $o$, and adding the
 condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
-formulas'' to the definition of $\textsf{dp}$.
+formulas'' to the definition of $\textsf{dp}$. If another condition of \textsf{dp}'s 
+definition is also applicable, the newly added one is preferred.
 
@@
 (\inot (\cst{p} \iand \cst{q}))$. This problem is obviously unsatisfiable, since
-arguments of $\cst{f}$ on different sides of the disequation are extensionally
-equal; however, without \infname{Abs} rules Zipperposition will rely on
-\infname{Cases}(\infname{Simp}) and \infname{EC} rules to derive the empty
-clause. Rule \infname{AbsER} will generate~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
+the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
+equal. Without the \infname{Abs} rules Zipperposition will rely on
+\infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
+clause. On the other hand, \infname{AbsER} will generate~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
 \inot (\cst{p} \iand \cst{q})$. Then, $C$ will get clausified using
-\infname{EC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
+\infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
 \inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.
 
-\looseness=-1
+\begin{sloppypar}
   Zipperposition restricts \infname{AbsSup} by
@@
   a first-order literal; Boolean extensionality is usually better supported by
-applying \infname{EC} on the top-level Boolean term. Thus, for the following
+  applying \infname{IC} on the top-level Boolean term. Thus, for the following
   discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
@@
   subterm. This index will be empty for first-order problems, incurring no
-overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply \infname{Abs} rules if all
+  overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
   disagreement pairs have at least one side whose head is a variable; those will
-be dealt with more efficiently using standard, non-extensional, versions of the
-rules. We also eagerly resolve literals $s_i \not\eq s'_i$ using at most one
+  be dealt with more efficiently using the core
+  rules of the superposition calculus. To simplify the resulting clauses of \infname{Abs} rules,
+  we also eagerly remove literals $s_i \not\eq s'_i$ by applying the first
   unifier returned by terminating, pragmatic variant of unification algorithm described in 
   Sect.~\ref{sec:unif:the-unification-procedure}.
+\end{sloppypar}
 
@@
 %
-Leibniz equality often appears in TPTP problems. Since modern provers have the native support
+Leibniz equality often appears in TPTP problems  \cite{gs-17-tptp}. Since modern provers have native support
 for equality, it is usually beneficial to recognize and replace occurrences of Leibniz equality.
 
-Before we began our modifications, Zipperposition had a powerful rule that
+Before we did our modifications, Zipperposition had a powerful rule that
 recognizes clauses that contain variations of Leibniz equality and instantiates
@@
     {\sigmacl{P \, \overline{s}_n^1
-    \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue  \llor C}} $$
+    \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue} \llor C} $$
 % \pagebreak[2]
@@
 \cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
-\itrue$. Since all side conditions are fulfilled, the rule \infname{ElimPredVar}
+\itrue$. The rule \infname{ElimPredVar}
 will generate $\sigma = \{ P \mapsto \lam{xy}{(x \ieq \cst{b}_1
 \iand y \eq \cst{b}_2) \ior (x \ieq \cst{c}_1 \iand y \ieq \cst{c}_2)  } \}$.
-After applying $\sigma$ to $C$ and subsequent $\beta$-reduction, negative literal
+After applying $\sigma$ to $C$ and subsequent $\beta$-reduction, the negative literal
 $  P \, \cst{b}_1 \, \cst{b}_2 \not\eq \itrue$ will reduce to 
@@
 \looseness=-1
-where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma =
-\{ P \mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$ and
-$\sigma' = \{ P \mapsto \lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. This
-rule differs from \infname{ElimPredVar} in three ways. First, it acts on
-occurrences of variables in both positive and negative literals. Second,
-due to its simplicity, it usually does not require \infname{IC} as the following step.
-Third, it imposes much weaker conditions on $P$. However, removing all negative
-variables in one step might improve performance.  Coming back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor P \,
-\cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
-\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b}_1} \}$
-to obtain the clause $C' = \cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$.
+where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma = \{ P
+\mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$, and $\sigma' = \{ P \mapsto
+\lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. \infname{ElimLeibniz} directly applies
+Leibniz equality: When the premise of the Leibniz equality is clausified it
+becomes $P \, X \not\eq \itrue \llor P \, Y \eq \itrue$. The rule then tries to find
+a similar subclause in a clause and then instantiate the clause with a
+substitution that asserts equality of $X$ and $Y$.
+
+This rule differs from
+\infname{ElimPredVar} in three ways. First, it can replace a predicate variable
+both with equality and disequality. Second, due to its simplicity, it usually
+does not require \infname{IC} as the following step. Third, it imposes much
+weaker conditions on $P$. However, removing all negative variables in one step
+might improve performance  by generating fewer intermediate clauses.  Coming
+back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor
+P \, \cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2
+\not\eq \itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution
+$\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b}_1} \}$ to obtain the clause $C' =
+\cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$, which is considerably simpler than
+the one obtained by \infname{ElimPredVar}.
 
@@
 
-Zipperposition's unification algorithm 
+Zipperposition's pragmatic, incomplete unification algorithm 
 %\cite{OUR UNIFICATION PAPER}
-uses flattened representation of terms with logical operators $\iand$ and $\ior$
-for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
+uses a flattened representation of terms with logical operators $\iand$ and $\ior$
+as heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
 are unifiable modulo associativity and commutativity of $\iand$ and $\ior$. Let
@@
 %Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
-Next, the remaining terms are sorted first by their head term and then their weight. Finally,
+Next, the remaining terms are sorted first by their head term and then their syntactic weight. Finally,
 an attempt is made to unify sorted lists pairwise.
 %
-As an example, consider the problem of unifying the pair $\big({\iand} \,
-(\cst{p} \, \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\iand} \;
+As an example, consider the problem of unifying the pair ${\iand} \,
+(\cst{p} \, \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})) \unif \; {\iand} \;
 (\cst{q} \, (\cst{f} \, \cst{a})) \; (R \, (\cst{f} \, (\cst{f} \,
-\cst{a})))\bigr)$ where $R$ is a free variable. If the arguments of $\iand$ are
+\cst{a})))$ where $R$ is a free variable. If the arguments of $\iand$ are
 simply sorted as described above, we would try to unify $\cst{p} \, \cst{a}$
@@
 removing term  $\cst{q} \, (\cst{f} \, \cst{a})$ from the argument lists, we
-will be left with the problem $(\cst{p} \, \cst{a},  R \, (\cst{f} \, (\cst{f}
-\, \cst{a})))$ which has a unifier.
-
-The winner of THF division of
-CASC-27 \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
+will be left with the problem $\cst{p} \, \cst{a} \unif  R \, (\cst{f} \, (\cst{f}
+\, \cst{a}))$ which has a unifier. This approach enables us to find more unifiers
+than by simple syntactic unification.
+
+The winner of the higher-order theorem division of the 2019 edition of CASC \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
 crucial advantage over Zipperposition: it is based on higher-order
@@
 term, this instantiation naturally propagates through the whole formula. In
-Zipperposition, which is based on higher-order superposition, the original formula
+Zipperposition, which is based on \lsup{}, the original formula
 is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
@@
 as in the original equality resolution and equality factoring rules. In rule
-\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \neg s' $. In the $+-$
-and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ x \, \tuplen{s}
-$ and $\neg s'$, and in the remaining version it unifies $ x \, \tuplen{s} $ and
+\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \inot s' $. In the $+-$
+and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ P \, \tuplen{s}
+$ and $\inot s'$, and in the remaining version it unifies $ P \, \tuplen{s} $ and
 $s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
-form for application of the corresponding base rules. It suffices to consider 
-literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
-$s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
-rewriting into $ s \not\eq t $).
+form for application of the corresponding base rules. In particular,
+for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
+allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
+For \infname{BoolEF} family of rules, we convert disequations into equations
+using the trick of negating one side to simulate \infname{EF}.
+% It suffices to consider 
+% literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
+% $s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
+% rewriting into $ s \not\eq t $).
+The rule \infname{BoolER} only considers literals of the form $s \eq t$. 
+Equivalences $s \iequiv t = \itrue$ and disequivalences $(s \iequiv t) \not\eq \itrue$ are automatically simplified to $s \eq t$ or $s \not\eq t$.
+The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
+\infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.
 
@@
   { (\iforall x. \, f) \eq \itrue \llor C}
-  { \{x \mapsto Y\}f \llor C }$ &
+  { \{x \mapsto Y\}(f) \llor C }$ &
   $\namedsimp{DC$_{\iexists}$}
   { (\iexists x.\, f) \eq \itrue \llor C}
-  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}f \llor C }$ 
+  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}(f) \llor C }$ 
   \\[2\jot]
@@
 \looseness=-1
-The rules described above are as given by Ganzinger and Stuber (adapted to
-our setting), with the omission of rules for negative literals ($ f \not\eq
-\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.
 In \infname{DC}$_\eq$ we require both $f$ and $g$ to be formulas and at least one of
@@
 \infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
-freely in $\iexists x.\, f$.
+freely in $\iexists x.\, f$. The rules described above are as given by Ganzinger and Stuber (adapted to
+our setting), with the omission of rules for negative literals (of the form $ f \not\eq
+\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.
 
@@
 Naive application of the \infname{DC} rules can result in exponential blowup in
-problem size. To avoid this, we rename formulas that have repeated
-occurrences. We keep the count of all nonatomic formulas occurring as
-either side of a literal. Before applying the \infname{DC} rule on a
-clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the number of $f$'s
-occurrences exceeds the threshold $k$. If it does, based on the
-polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause $\cst{p} \, \tuplen{Y}
-\not\eq \itrue \llor f \eq \itrue$ (if the literal is positive) or $\cst{p} \, \tuplen{Y}
-\eq \itrue \llor f \not\eq \itrue$ (if the literal is negative), where $\tuplen{Y}$ are
-all free variables of $f$ and $\cst{p}$ is a fresh symbol. Then, we replace the
-clause $f \mathrel{\dot\eq} \itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor
-C$.
-
-Before the number of occurrences of $f$ is checked, we first check (using a fast, incomplete matching algorithm) if there is a
-formula $g$, for which definition was already introduced, such that $\sigmaterm{g} =
-f$, for some substitution $\sigma$. This check can have three outcomes. First, if
-the definition $\textsf{q} \, \overline{X}_n$ is already introduced for $g$ with the
-polarity matching that of $f \, \dot{\eq} \, \itrue$, then $f$ is
-replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if the definition was introduced,
-but with different polarity, we create the clause defining $g$ with the missing
-polarity, and replace $f$ with $\sigmaterm{\cst{q} \, \tuplen{X}}$. Last, if the there is no renamed
-formula $g$ generalizing $f$, then we perform the previously described check.
+problem size. To avoid this, we rename formulas $f$ that have repeated
+occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ for them, where
+$\tuplen{X}$ are all free variables of $f$. We keep the count of all nonatomic
+formulas occurring as either side of a literal. Before applying the \infname{DC}
+rules on a clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the
+number of occurrences of $f$  exceeds the threshold $k$. If it does, based on the
+polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause
+$\cst{p} \, \tuplen{Y} \not\eq \itrue \llor f \eq \itrue$ (if the literal is
+positive) or $\cst{p} \, \tuplen{Y} \eq \itrue \llor f \not\eq \itrue$ (if the
+literal is negative), where $\tuplen{Y}$ are all free variables of $f$ and
+$\cst{p}$ is a fresh symbol. Then, we replace the clause $f \mathrel{\dot\eq}
+\itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor C$.
+
+Before the number of occurrences of $f$ is checked, we first check (using a
+fast, incomplete matching algorithm, used for simplification and subsumption) if there is a
+formula $g$, for which definition was already introduced, such that
+$\sigmaterm{g} = f$, for some substitution $\sigma$. This check can have three
+outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ is already
+introduced for $g$ with the polarity matching that of $f \, \dot{\eq} \,
+\itrue$, then $f$ is replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if
+the definition was introduced, but with different polarity, we create the clause
+defining $g$ with the missing polarity, and replace $f$ with $\sigmaterm{\cst{q}
+\, \tuplen{X}}$. Last, if the there is no renamed formula $g$ generalizing $f$,
+then we check if the number of occurrences of $f$ exceeds the threshold.
 
@@
 symbols introduced by the \infname{DC}$_{\iexists}$ rule. When
-\infname{DC}$_{\iexists}$ is applied to $f = \iexists x.\, f'$ we check if there is
+\infname{DC}$_{\iexists}$ is applied to $f = \iexists x.\, f'$, we check if there is
 a Skolem $\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y}$ introduced for a
@@
 $\cst{sk}$ is reused and $\iexists x.\, f'$ is replaced by $\{x \mapsto
-(\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y})\sigma \}f'$. Renaming and
+\sigma(\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y}) \}(f')$. Renaming and
 name reusing techniques are inspired by the VCNF algorithm described by Reger et
-al.\cite{rsv-16-vcnf}.
+al. \cite{rsv-16-vcnf}.
 
@@
 whose body has type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
-implies that $\lam{\tuplen{x}}{f}$ is extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
+implies that $\lam{\tuplen{x}}{f}$ is functional-extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
 $\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \eq \lam{\tuplen{x}}{\ifalse})$, we
-designed the following rule (where all loosely bound variables of $f$ are $\tuplen{x}$):
+designed the following rule (where $\tuplen{x}$ consists of all loosely bound variables of $f$ ):
 %\pagebreak[2]
@@
 \looseness=-1
-An alternative to heavy modifications of the prover needed to support the rules
+An alternative to modifications of the prover needed to support the rules
 described above is to treat Booleans as yet another theory. Since the theory of Booleans
@@
 correspond to the homologous logical constants from Section \ref{sect:bool:background}. In their
-type declarations $o$ is replaced by $\typ{bool}$.
-
-\looseness=-1
-To make this paper self-contained we include the axioms from Bentkamp et al. \cite{bbtvw-21-sup-lam}.
-Definitions of symbols are computational in nature: symbols are characterized by their behavior on $\cst{t}$ and $\cst{f}$.
-This also reduces interferences between different axioms. Axioms are listed as follows:
+type declarations, $o$ is replaced by $\typ{bool}$.
+
+\looseness=-1
+To make this chapter self-contained we include the axioms from Bentkamp et al. \cite{bbtvw-21-sup-lam}.
+Definitions of symbols are computational in nature: Symbols are characterized by their behavior on $\cst{t}$ and $\cst{f}$.
+This also reduces interferences between different axioms. The axioms are listed as follows:
 
@@
 nested Booleans \cite{kotelnikov-16-fool}. As explained in Sect.~\ref{sec:ehoh:preprocessing}, we extended
-the clausification algorithm of Ehoh to support nested Booleans inspired by VCNF extension.
+the clausification algorithm of Ehoh to support nested Booleans inspired by this VCNF extension.
 Zipperposition and Ehoh share the
-same clausification algorithm, enabling us to reuse the extension with one
+same clausification algorithm, enabling us to reuse the extension, with one
 notable difference:
-unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
+Unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
 will be removed. Namely, Booleans that are below $\lambda$-abstraction
@@
 The TPTP library contains thousands of higher-order benchmarks, many of them hand-crafted
-to point out subtle interferences of functional and Boolean properties of higher order logic. In this section
-we discuss some problems from the TPTP library that illustrate the advantages and disadvantages
+to point out subtle interferences of functional and Boolean properties of higher-order logic. In this section
+we discuss some problems from the TPTP library \cite{gs-17-tptp} that illustrate the advantages and disadvantages
 of our approach.
@@
 fresh variable $W$ to both sides of the equation, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
-\eq Y \, W$. Most superposition-based higher-order theorem
+\eq Y \, W$. Superposition-based higher-order theorem
 provers (such as Leo-III, Vampire and Zipperposition) will split this clause
@@
 successful proof attempt. Rule \infname{BoolER} imitates the behavior of
-tableaux prover: it essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
+a tableaux prover: It essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
 (\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $ which makes finding the necessary substitution easy and
@@
 Combining rule \infname{(Bool)ER} with dynamic clausification is very fruitful as
-the benchmark \texttt{SYO033\^{}1} illustrates. This problem also contains the
+the benchmark \texttt{SYO033\^{}1} illustrates. This problem contains the
 single conjecture
 $$ \iexists x. \, \iforall y.  \, x \, y \iequiv (\iforall z. \, y \,z)$$ The
-problem is easily solved if we replace variable $x$ with the constant $\forall$.
+problem is easily solved if we replace the variable $x$ with the constant $\forall$.
 Moreover, the prover does not have to blindly guess this instantiation. Instead,
@@
 \,z$). However, when the problem is clausified, all quantifiers are removed.
-Then, Zipperposition only finds the proof if appropriate instantiation mode of
+Then, Zipperposition finds the proof only if an appropriate instantiation mode of
 \infname{PI} is used, and if both clauses resulting from clausifying the negated
 conjecture are appropriately instantiated. In contrast, dynamic clausification will
-derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, ) $
+derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, \cst{sk} $
 from the negated conjecture in three steps. Then, equality resolution results in
@@
 This effect is even more pronounced on problems \verb|SYO287^5| and
-\verb|SYO288^5|, in which critical proof step is instantiation of a variable
-with imitation of $\ior$ and $\iand$. In configurations that do not use dynamic
+\verb|SYO288^5|, in which a critical proof step consists of instantiating a variable
+with imitations of $\ior$ and $\iand$. In configurations that do not use dynamic
 clausification and \infname{BoolER}, Zipperposition times out in any reasonable
@@
 FOOL unfolding of nested Boolean terms will result in exponential blowup in the problem size. However,
-superposition-based theorem provers are well-equipped for this issue: Their CNF algorithms use smart simplifications and
+superposition-based theorem provers are well equipped for this issue: Their CNF algorithms use smart simplifications and
 formula renaming to mitigate these effects. Moreover, when the problem is preprocessed, the prover is aware of the problem
-size before the proving process starts and can adjust its heuristics properly. E, Zipperposition and Vampire, instructed to perform FOOL unfolding,
-solve the problem swiftly, using their default modes. However, if problem is not preprocessed, Zipperposition struggles to prove it using
+size before the proving process starts and can adjust its heuristics properly. E, Zipperposition, and Vampire, instructed to perform FOOL unfolding,
+solve the problem swiftly, using their default modes. However, if the problem is not preprocessed, Zipperposition struggles to prove it using
 \infname{Cases(Simp)} and due to the large number of (redundant) clauses it creates, succeeds only if specific heuristic choices are made.
@@
 
-We performed extensive evaluation to determine usefulness of our approach. As
-our benchmark set, we used all 2606 monomorphic theorems from the TPTP library,
+We performed an extensive evaluation to determine the  usefulness of our approach. As
+our benchmark set, we used all 2606 monomorphic theorems from the TPTP 7.2.0 library,
 given in THF format. All of the experiments described in this section were
@@
 \looseness=-1
-For this part of the evaluation, we fixed a single well-performing Zipperposition configuration
+For this part of the evaluation, we fixed a single well-perform\-ing Zipperposition configuration
 called \emph{base} (\emph{b}). Since we are testing a single configuration, we
 used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
-given in a portfolio mode. Configuration \emph{b} uses the pragmatic variant
+given in a portfolio mode when participating in CASC. Configuration \emph{b} uses the pragmatic variant
 pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
-It enables \infname{BoolSimp} rule,
-\infname{EC} rule, \infname{PI} rule in \emph{Pragmatic} mode with $k=2$,
-\infname{ElimLeibniz} and \infname{ElimPredVar} rules, \infname{BoolER} rule, and \infname{BoolEF} rules. 
+It enables the \infname{BoolSimp} rule,
+the \infname{EC} rule, the \infname{PI} rule in \emph{Pragmatic} mode with $k=2$,
+rules \infname{ElimLeibniz} and \infname{ElimPredVar}, the \infname{BoolER} rule, and the \infname{BoolEF} rules. 
 To evaluate the usefulness of
-all rules we described above, we enable, disable or change the
+all rules we described above, we enable, disable, or change the
 parameters of a single rule, while keeping all other parameters of $b$ intact. 
@@
 denote corresponding combination of parameters.
-Result for the base configuration is written in \emph{cursive}; the best result
+Result for the base configuration is written in \emph{italics}; the best result
 is written in \textbf{bold}.
@@
 \textbf{b} is the base configuration, which uses \infname{CasesSimp}, and \textbf{b$_\text{c}$}
-swaps this rule for \infname{Cases}. Although the margin is slim, the results show it is usually preferable to select leftmost-outermost
+swaps this rule for \infname{Cases}. Although the margin is slim, the results show it is usually preferable to select the leftmost-outermost
 subterm.
@@
     \end{tabular}}
-    \caption{Effect of the \infname{Cases}(\infname{Simp}) rule on success rate }
+    \caption{Effect of the \infname{Cases}(\infname{Simp}) rule on the success rate }
     \label{fig:cases}
@@
 attention to configurations with $k=2$, mode \emph{Full} can solve 10 problems
-no other mode (including disabling \infname{PI} rule) can. Modes
+no other mode (including disabling the \infname{PI} rule) can. Modes
 \emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve 2  problems, whereas
 \emph{Imit}$_{\ior}$ solves one problem uniquely. This result suggests that, even
-though this is not evident from Figure \ref{fig:pi}, sets of problems solved 
+though this is not evident from Figure \ref{fig:pi}, the sets of problems solved 
 solved by different modes somewhat differ.
@@
     \end{tabular}}
-    \caption{Effect of \infname{PI} rule on success rate}
+    \caption{Effect of \infname{PI} rule on the success rate}
     \label{fig:pi}
@@
 
-Similarly, Figure \ref{fig:bool-rules} discusses merits of including ($+$) or
+Similarly, Figure \ref{fig:bool-rules} shows the merits of including ($+$) or
 excluding ($-$) \infname{BoolER} (BER) and \infname{BoolEF} (BEF) rules. Our
-expectations were that inclusion of those two rules would make bigger impact on
-success rate. It turned out that, in practice, most of the effects of these
+expectations were that inclusion of those two rules would make a larger impact on
+success rate. It turns out that, in practice, most of the effects of these
 rules could be achieved using a combination of the \infname{PI} rule and
@@
 \looseness=-1
-Combining these two rules with lazy clausification is more useful: when the rule
-\infname{IC} is replaced by the rule \infname{DC}, the success rate increases
-(compared to 1646 problems solved by $b$) to 1660 problems. We also discovered
-that reasoning with choice is useful: when rule \infname{Choice} is enabled, the
+Combining these two rules with dynamic clausification is more useful: When rule
+\infname{IC} is replaced by rule \infname{DC}, the success rate increases
+to 1660 problems, compared to 1646 problems solved by $b$. We also discovered
+that reasoning with choice is useful: When rule \infname{Choice} is enabled, the
 success rate increases to 1653. We determined that including or excluding the
@@
 which is only 6 problems short of configuration $b$. Disabling \infname{Abs} and
-\infname{Interpret-$\lambda$} rules results in solving 25 and 31 problems less,
-respectively. Raw data show that in total, using configurations from Figure
+\infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
+respectively. The raw data show that in total, using configurations from Figure
 \ref{fig:cases} to Figure \ref{fig:bool-rules}, 1682 problems can be solved.
@@
 We compared Zipperposition with all higher-order theorem provers that took part
-in THF division of CASC-27 \cite{gs-19-casc27}: CVC4 1.8 prerelease \cite{cbetal-11-cvc4},
+in higher-order division of 2019 edition of CASC \cite{gs-19-casc27}: CVC4 1.8 prerelease \cite{cbetal-11-cvc4},
 Leo-III 1.4 \cite{sb-21-leo3}, Satallax 3.4
@@
 in portfolio mode that runs configurations in different time slices. 
-We set the CPU time limit to 180 s, the time allotted to each prover at CASC-27.
+We set the CPU time limit to 180 s, the time allotted to each prover at 2019 edition of CASC.
 
@@
 first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
-iProver \cite{kk-08-iprover} as backends, while Satallax uses Ehoh
+iProver \cite{kk-08-iprover} as backends, and Satallax uses Ehoh
 % \cite{EHOH CHAPTER}
@@
 \looseness=-1
-In both pure and cooperative mode, Satallax is the winner.
-Zipperposition comes in close second, showing that our approach is a promising
-basis for further extensions. Leo-III uses SMT solver CVC4, which features
-native support for Booleans, as a backend. It is possible that the use of CVC4 is
-one of the reasons for massive improvement in success rate of cooperative
-configuration of Leo-III, compared with the pure version. Therefore, we conjecture
-that including support for SMT backends in Zipperposition might be beneficial.\section{Discussion and Related Work} 
+In both pure and cooperative mode, Satallax is the winner. Zipperposition comes
+in close second, showing that our approach is a promising basis for further
+extensions. Indeed, as we show in Chapter \ref{ch:ho-techniques}, with
+fine-tuning of the heuristics and further extensions to the calculus, this
+approach can outperform all other competitive provers with a large margin.
+
+\subsection{Discussion}
 \label{sect:bool:discussion}
@@
 Established higher-order provers like Leo-III and Satallax perform very
-well on TPTP benchmarks; however, recent evaluations (including the one in Sect.~\ref{sec:ehoh:evaluation}) show that on Sledgehammer
+have been optimized to perform well on TPTP; however, recent evaluations (including the one in Sect.~\ref{sec:ehoh:evaluation}) show that on Sledgehammer
 problems they are outperformed by translations to first-order logic
@@
 The main merit of our approach is that it combines two successful complementary
-approaches to support features of higher-order logic that have not been combined
-before in a modular way. It is based on a higher-order superposition calculus
+approaches, \lsup{} and FOOL paramodulation, to support features of higher-order logic that have not been combined
+before in a modular way. It is based on \lsup{}, a calculus that generalizes the highly successful first-order superposition calculus.
 that incurs around 1\% of overhead on first-order problems compared with classic
-superposition \cite{bbtvw-21-sup-lam}. We conjecture that it is
-this efficient reasoning base on which the approach is based that contributes to
-its competitive performance.
+superposition \cite{bbtvw-21-sup-lam}.
 

### 2022-03-10 introduction/introduction.tex
more

@@
 logic, we decided not to start from a blank slate and build a new theorem prover.
-Rather, we start from a position of strength --- from the state-of-the-art
-first-order superposition theorem prover E \cite{scv-19-e23} --- and extend it to higher-order logic. In this way, the points \ref{it:fo} and
+Rather, we start from a position of strength---from the state-of-the-art
+first-order superposition theorem prover E~\cite{scv-19-e23}---and extend it to higher-order logic. In this way, the points \ref{it:fo} and
 \ref{it:size} from the above wishlist are already fulfilled, as E is very

### 2022-03-10 pre/pre.tex
more

@@
 The provers we discuss in this thesis, E and Zipperposition, are both DISCOUNT-loop provers.
+
+\subsection{Higher-Order Superposition Calculi}
+\label{sec:pre:ho-sup-calculi}
+
+\looseness=-1
+In Chapter \ref{ch:intro} we mention that we identified three steps on the road
+to extending efficient superposition provers to higher-order logic. Together with
+our colleagues we did the necessary theoretical work by extending the superposition
+calculus using the same roadmap:
+
+\begin{enumerate}
+    \item Bentkamp, Blanchette, Cruanes, and Waldmann designed a complete
+    superposition calculus for higher-order logic devoid of
+    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. We call
+    this calculus \lfsup{}.
+    \item Alexander Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
+    extended this calculus to support $\lambda$-abstraction
+    \cite{bbtvw-21-sup-lam}. We call it \lsup{}.
+    \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
+    support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
+    designing the complete calculus for full higher-order logic. We call this
+    calculus \osup{}.
+\end{enumerate}

### 2022-03-11 bools/bools.tex
more

@@
 completeness of our extension of \lsup{}. Using the insights that we gained with
-this incomplete extension of \lsup{} to full higher-order logic, Bentkamp
+this incomplete extension of \lsup{} to full higher-order logic, Bentkamp et al.
 designed a complete calculus for full higher-order logic---\osup{}
@@
 types, and native logical connectives. For reference, we provide a
-definition of this polymorphic logic.
+description of this polymorphic logic. For the notions that are not defined here,
+we assume the definitions introduced in Sect.~\ref{ch:pre}.
 
@@
 Booleans, whereas nested Booleans are characteristic for higher-order logic.
-Note that while our calculus works with clausal representation of the problem,
-its higher-order nature allows representing formula $f$ as a singleton clause
+
+% Note that while our calculus works with clausal representation of the problem,
+% its higher-order nature allows representing formula $f$ as a singleton clause
+% containing only literal $f$.
+Our calculus works with the clausal structure described in Sect.~\ref{sec:pre:clauses}.
+However, its higher-order nature allows representing formula $f$ as a singleton clause
 containing only literal $f$.
@@
 term in the proof state, this axiom is very explosive. Therefore, Leo-III
-\cite{sb-21-leo3} recognizes definitions of all symbols that correspond to choice and deals with
+\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
 them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
@@
 Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
-As a pragmatic extension of the lambda-superposition calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
+As a pragmatic extension of the \lsup{} calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
 to support functional extensionality. Instead of requiring that terms
@@
   disagreement pairs have at least one side whose head is a variable; those will
-  be dealt with more efficiently using the core
-  rules of the superposition calculus. To simplify the resulting clauses of \infname{Abs} rules,
-  we also eagerly remove literals $s_i \not\eq s'_i$ by applying the first
-  unifier returned by terminating, pragmatic variant of unification algorithm described in 
-  Sect.~\ref{sec:unif:the-unification-procedure}.
+  be dealt with more efficiently using the core rules of the superposition
+  calculus. To simplify the resulting clauses of \infname{Abs} rules, we also
+  eagerly remove literals $s_i \not\eq s'_i$ using a simplifying version of
+  \infname{ER} (Sect.~\ref{sec:pre:rules}). In particular, we apply the first
+  unifier returned by the terminating, pragmatic variant of unification
+  algorithm described in Sect.~\ref{sec:unif:the-unification-procedure}, and
+  remove the premise after conclusion of \infname{ER} is computed.
 \end{sloppypar}
@@
 allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
-For \infname{BoolEF} family of rules, we convert disequations into equations
+For the \infname{BoolEF} family of rules, we convert disequations into equations
 using the trick of negating one side to simulate \infname{EF}.
@@
   $\namedsimp{DC$_{\iand}$}
-  { (f \iand g) \eq \itrue \llor C}
-  { f \eq \itrue \llor C \quad g \eq \itrue \llor C }$ &
+  { (g \iand h) \eq \itrue \llor C}
+  { g \eq \itrue \llor C \quad h \eq \itrue \llor C }$ &
   $\namedsimp{DC$_{\ior}$}
-  { (f \ior g) \eq \itrue \llor C}
-  { f \eq \itrue \llor g \eq \itrue  \llor C }$ &
+  { (g \ior h) \eq \itrue \llor C}
+  { g \eq \itrue \llor h \eq \itrue  \llor C }$ &
   $\namedsimp{DC$_{\iimplies}$}
-  { (f \iimplies g) \eq \itrue \llor C}
-  { f \not\eq \itrue \llor g \eq \itrue  \llor C } $ \\[2\jot]
+  { (g\iimplies h) \eq \itrue \llor C}
+  { g\not\eq \itrue \llor h \eq \itrue  \llor C } $ \\[2\jot]
   $\namedsimp{DC$_{\inot}$}
-  { (\inot f) \eq \itrue \llor C}
-  { f \not\eq \itrue \llor C}$ &
+  { (\inot g) \eq \itrue \llor C}
+  { g\not\eq \itrue \llor C}$ &
   $\namedsimp{DC$_{\iforall}$}
-  { (\iforall x. \, f) \eq \itrue \llor C}
-  { \{x \mapsto Y\}(f) \llor C }$ &
+  { (\iforall x. \, g) \eq \itrue \llor C}
+  { \{x \mapsto Y\}(g) \llor C }$ &
   $\namedsimp{DC$_{\iexists}$}
-  { (\iexists x.\, f) \eq \itrue \llor C}
-  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}(f) \llor C }$ 
+  { (\iexists x.\, g) \eq \itrue \llor C}
+  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}(g) \llor C }$ 
   \\[2\jot]
   \multicolumn{3}{c}{$\namedsimp{DC$_\eq$}
-  { f \eq g \llor C}
-  { f \not\eq \itrue \lor g\eq\itrue \llor C \quad f \eq \itrue \lor g\not\eq\itrue \llor C}$}
+  { g\eq h \llor C}
+  { g\not\eq \itrue \lor h\eq\itrue \llor C \quad g \eq \itrue \lor h\not\eq\itrue \llor C}$}
 \end{tabular}
@@
  as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we run
-the cooperative provers in two versions: \emph{pure}, which disables backends,
+the cooperative provers in two versions: \emph{uncoop}, which disables backends,
 and \emph{coop} which uses all supported backends.
@@
       \MIDLINE
-      pure    & 1806 (5) &  1627 (0) & 2067 (0)            & 1924 (7) & 1980 (\phantom{0}0)                \\
+      uncoop    & 1806 (5) &  1627 (0) & 2067 (0)            & 1924 (7) & 1980 (\phantom{0}0)                \\
       coop    & --       &  2085 (3) & \textbf{2214 (9)}   & --       & 2190 (17)
@@
 \looseness=-1
-In both pure and cooperative mode, Satallax is the winner. Zipperposition comes
+In both uncooperative and cooperative mode, Satallax is the winner. Zipperposition comes
 in close second, showing that our approach is a promising basis for further
@@
 
-\subsection{Discussion}
+\section{Discussion}
 \label{sect:bool:discussion}

### 2022-03-11 pre/pre.tex
more

@@
     superposition calculus for higher-order logic devoid of
-    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. We call
-    this calculus \lfsup{}.
-    \item Alexander Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
+    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
+    and we shortly refer to it as \lfsup{}.
+    \item \begin{sloppypar}
+    Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
     extended this calculus to support $\lambda$-abstraction
-    \cite{bbtvw-21-sup-lam}. We call it \lsup{}.
+    \cite{bbtvw-21-sup-lam}. They call this new calculus \emph{Boolean-free $\lambda$-superposition calculus},
+    and we shortly refer to it as \lsup{}.
+    \end{sloppypar}        
     \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
     support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
-    designing the complete calculus for full higher-order logic. We call this
-    calculus \osup{}.
+    designing the complete calculus for full higher-order logic. This
+    calculus is called Boolean $\lambda$-superposition, or shortly \osup{}.
 \end{enumerate}
+%
+Note that we use the name $\lambda$-superposition interchangeably for \lsup{}
+and \osup{} where the context gives enough information to disambiguate this name.

### 2022-03-11 bools/bools.tex
intro

@@
 
-Our approach is implemented in the Zipperposition theorem prover
-\cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
+Our approach is implemented in Zipperposition
+\cite{sc-15-simon-phd,sc-supind-17}. This theorem prover was used by Bentkamp et al.\
 to implement \lsup{}. We further extend
@@
 arithmetic, induction and limited support for higher-order logic
-\cite{sc-15-simon-phd}.
+\cite{sc-15-simon-phd,sc-supind-17}.
 
 In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
-This means that higher-order term $(\cst{f} \, \cst{a}) \, cst{b}$ is
-represented the same as first-order term $\cst{f}(\cst{a}, \cst{b})$. In
+This means that the higher-order term $(\cst{f} \, \cst{a}) \, \cst{b}$ is
+represented the same as the first-order term $\cst{f}(\cst{a}, \cst{b})$. In
 addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
@@
 modified Zipperposition to represent all literals equationally: a nonequational
-literal $f$ is stored as $f \eq \itrue$, whereas $\neg f$ is stored as $f
+literal $f$ is stored as $f \eq \itrue$, whereas $\inot f$ is stored as $f
 \noteq \itrue$. Equations of the form $f \eq \ifalse $ and $f \noteq \ifalse$ are
@@
 generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
-such that $i\ineq j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
-\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
+such that $i\not= j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
+\ieq x_j}$; if there exist indices $i,j$ such that $i \not= j$, and $\tau_i =
 \tau_j = o$, then it generates $\lam{\tuplen{x}}{x_i \iand x_j}$ and
@@
 Zipperposition options, $D$ will either be deleted or kept. Note that $D$
-will not subsume its instances, since the matching algorithm Zipperposition uses is
-too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
-\looseness=-1
-
-Most provers natively support both functional and Boolean extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
-first-order unification to return unification constraints consisting of pairs of
-terms of functional type, whereas Steen relies on the unification rules of
-Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
-As a pragmatic extension of the \lsup{} calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
-to support functional extensionality. Instead of requiring that terms
-involved in the inference are unifiable, it is required that they can be
-decomposed into \emph{disagreement pairs} such that at least one of the
-disagreement pairs is of functional type. Disagreement pairs of terms $s$ and
-$t$ of the same type are defined inductively using function $\textsf{dp}$:
-$\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal; $\textsf{dp}(a \,
-\tuplen{s}, b \, \tuple{t}{m}) = \{(a \,
-\tuplen{s}, b \, \tuple{t}{m})\}$ if $a$ and $b$ are different heads;
-$\textsf{dp}(\lam{x}{s}, \lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$;
-$\textsf{dp}(a \, \tuplen{s}, a \, \tuplen{t}) = \bigcup_{i=1}^{n}
-\textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
+will not subsume its instances, since the matching algorithm Zipperposition uses favors performance over completeness and is
+thus too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
+\looseness=-1
+
+Most provers natively support both functional and Boolean extensionality
+reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify first-order
+unification to return unification constraints consisting of pairs of terms of
+functional type, whereas Steen relies on the unification rules of Leo-III's
+calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality. As a
+pragmatic extension of the \lsup{} calculus, Bentkamp et al.
+\cite{bbtvw-21-sup-lam} propose to alter the core generating inference rules of
+superposition (Sect.~\ref{sec:pre:rules}) to support functional extensionality.
+Instead of requiring that terms involved in the inference are unifiable, it is
+required that they can be decomposed into \emph{disagreement pairs} such that at
+least one of the disagreement pairs is of functional type. Disagreement pairs of
+terms $s$ and $t$ of the same type are defined inductively using function
+$\textsf{dp}$: $\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal;
+$\textsf{dp}(a \, \tuplen{s}, b \, \tuple{t}{m}) = \{(a \, \tuplen{s}, b \,
+\tuple{t}{m})\}$ if $a$ and $b$ are different heads; $\textsf{dp}(\lam{x}{s},
+\lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$; $\textsf{dp}(a \, \tuplen{s}, a
+\, \tuplen{t}) = \bigcup_{i=1}^{n} \textsf{dp}(s_i, t_i)$. Then the
+extensionality rules are stated as follows:
 \pagebreak[2]
@@
 $s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
-(s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
+(s_n, s'_n)\}$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
+extensional versions of superposition, equality resolution and equality
+factoring (Sect.~\ref{sec:pre:rules}). All side conditions for extensional rules are the same as for
 the standard rules, except that condition that $s$ and $s'$ are unifiable is
 replaced by the condition that at least one $s_i$ is of functional type and that
-$n>0$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
-extensional versions of superposition, equality resolution and equality
-factoring (Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of
-these three rules. We extend this rule to support Boolean extensionality by
+$n>0$. By \infname{Abs} we denote the union of
+these three rules. We extend \infname{Abs} to support Boolean extensionality by
 requiring that at least one $s_i$ is of functional or type $o$, and adding the
@@
 formulas'' to the definition of $\textsf{dp}$. If another condition of \textsf{dp}'s 
-definition is also applicable, the newly added one is preferred.
-
+definition is also applicable, the newly added one is preferred. In the rest of the
+text with \infname{Abs} we assume the version that supports Boolean extensionality.
+
+\looseness=-1
 Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
-(\inot (\cst{p} \iand \cst{q}))$. This problem is obviously unsatisfiable, since
+(\inot (\cst{p} \iand \cst{q}))$. This problem is unsatisfiable, as
 the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
-equal. Without the \infname{Abs} rules Zipperposition will rely on
+equal. Without the \infname{Abs} rules Zipperposition relies on
 \infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
-clause. On the other hand, \infname{AbsER} will generate~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
-\inot (\cst{p} \iand \cst{q})$. Then, $C$ will get clausified using
+clause. On the other hand, \infname{AbsER} generates~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
+\inot (\cst{p} \iand \cst{q})$. Then, $C$ gets clausified using
 \infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
@@
 The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
-\infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.
+the \infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.
 
@@
 \looseness=-1
-In \infname{DC}$_\eq$ we require both $f$ and $g$ to be formulas and at least one of
+In \infname{DC}$_\eq$ we require both $g$ and $h$ to be formulas and at least one of
 them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
 \infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
-freely in $\iexists x.\, f$. The rules described above are as given by Ganzinger and Stuber (adapted to
+freely in $\iexists x.\, g$. The rules described above are as given by Ganzinger and Stuber (adapted to
 our setting), with the omission of rules for negative literals (of the form $ f \not\eq
@@
 problem size. To avoid this, we rename formulas $f$ that have repeated
-occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ for them, where
+occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ replacing them, where
 $\tuplen{X}$ are all free variables of $f$. We keep the count of all nonatomic
@@
 outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ is already
-introduced for $g$ with the polarity matching that of $f \, \dot{\eq} \,
+introduced for $g$ with the same polarity as $f \, \dot{\eq} \,
 \itrue$, then $f$ is replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if
@@
 defining $g$ with the missing polarity, and replace $f$ with $\sigmaterm{\cst{q}
-\, \tuplen{X}}$. Last, if the there is no renamed formula $g$ generalizing $f$,
-then we check if the number of occurrences of $f$ exceeds the threshold.
+\, \tuplen{X}}$. Last, if there is no renamed formula $g$ generalizing $f$,
+we check if the number of occurrences of $f$ exceeds the threshold $k$.
 
@@
 implies that $\lam{\tuplen{x}}{f}$ is functional-extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
-$\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \eq \lam{\tuplen{x}}{\ifalse})$, we
+$\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \ieq \lam{\tuplen{x}}{\ifalse})$, we
 designed the following rule (where $\tuplen{x}$ consists of all loosely bound variables of $f$ ):
@@
 
-The TPTP library contains thousands of higher-order benchmarks, many of them hand-crafted
+The TPTP library \cite{gs-17-tptp} contains thousands of higher-order benchmarks, many of them hand-crafted
 to point out subtle interferences of functional and Boolean properties of higher-order logic. In this section
-we discuss some problems from the TPTP library \cite{gs-17-tptp} that illustrate the advantages and disadvantages
+we discuss some problems from the TPTP library that illustrate the advantages and disadvantages
 of our approach.
@@
 conjecture are appropriately instantiated. In contrast, dynamic clausification will
-derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, \cst{sk} $
+derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, X) $
 from the negated conjecture in three steps. Then, equality resolution results in
@@
 FOOL unfolding of nested Boolean terms will result in exponential blowup in the problem size. However,
-superposition-based theorem provers are well equipped for this issue: Their CNF algorithms use smart simplifications and
+superposition-based theorem provers are well equipped for this issue: Their clausification algorithms use smart simplifications and
 formula renaming to mitigate these effects. Moreover, when the problem is preprocessed, the prover is aware of the problem
@@
 For this part of the evaluation, we fixed a single well-perform\-ing Zipperposition configuration
-called \emph{base} (\emph{b}). Since we are testing a single configuration, we
+called \emph{base} (b). Since we are testing a single configuration, we
 used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
-given in a portfolio mode when participating in CASC. Configuration \emph{b} uses the pragmatic variant
+given in a portfolio mode when participating in CASC. Configuration {b} uses the pragmatic variant
 pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
@@
 all rules we described above, we enable, disable, or change the
-parameters of a single rule, while keeping all other parameters of $b$ intact. 
+parameters of a single rule, while keeping all other parameters of b intact. 
 In figures that contain sufficiently different configurations, cells are of the form $n (m)$ where $n$ is the total
@@
 three possible options to choose subterm on which the inference is performed:
-\textbf{a} stands for any eligible subterm, \textbf{lo} and \textbf{li} stands
+{a} stands for any eligible subterm, {lo} and {li} stands
 for leftmost outermost and leftmost innermost subterms, respectively. The rows
 correspond to two different rules: 
-\textbf{b} is the base configuration, which uses \infname{CasesSimp}, and \textbf{b$_\text{c}$}
+{b} is the base configuration, which uses \infname{CasesSimp}, and {b$_\text{c}$}
 swaps this rule for \infname{Cases}. Although the margin is slim, the results show it is usually preferable to select the leftmost-outermost
@@
 \infname{IC} is replaced by rule \infname{DC}, the success rate increases
-to 1660 problems, compared to 1646 problems solved by $b$. We also discovered
+to 1660 problems, compared to 1646 problems solved by b. We also discovered
 that reasoning with choice is useful: When rule \infname{Choice} is enabled, the
@@
 Counterintuitively, disabling \infname{BoolSimp} rule results in 1640 problems,
-which is only 6 problems short of configuration $b$. Disabling \infname{Abs} and
+which is only 6 problems short of configuration b. Disabling \infname{Abs} and
 \infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
@@
 this mode. On the other hand, preprocessing is fairly competitive: it solves
-only 8 problems less than the $b$ configuration.
+only 8 problems less than the b configuration.
 

### 2022-03-11 ehoh/ehoh.tex
intro

@@
 including Leo-III \cite{sb-21-leo3} and
-Zipperposition \cite{bbtvw-21-sup-lam}.
+Zipperposition \cite{sc-15-simon-phd,sc-supind-17}.
 
@@
 Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
-and Zipperposition 1.6 \cite{bbtvw-21-sup-lam}.
+and Zipperposition 1.6 \cite{sc-supind-17,sc-15-simon-phd}.
 Like at CASC, we used different versions of Vampire for first-order
@@
   Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
-  \cite{sc-15-simon-phd,bbtvw-21-sup-lam}. \hbox{Otter-$\lambda$}
+  \cite{sc-15-simon-phd,sc-supind-17}. \hbox{Otter-$\lambda$}
   adds $\lambda$-terms and second-order unification to

### 2022-03-11 ho-techniques/ho-techniques.tex
intro

@@
 \label{ch:ho-techniques}
+
+\authors{
+    Joint work with\\
+    Alexander Bentkamp,
+    Jasmin Blanchette,
+    Simon Cruanes,
+    Visa Nummelin, and Sophie Tourret
+}
+
+\blfootnote{In this work I was the main designer of all presented techniques,
+with the exception of inference streams which were designed by Sophie Tourret
+and Alexander Bentkamp. Alexander Bentkamp and Jasmin Blanchette also discussed many
+of the techniques with me and suggested important updates. Visa Nummelin worked on
+implementation of FOOL preprocessing. Simon Cruanes is the original developer
+of Zipperposition and provided us with invaluable knowledge.
+}
+
+\begin{abstract}
+    Superposition is among the most successful calculi for first-order logic. Its
+    extension to higher-order logic introduces new challenges such as infinitely
+    branching inference rules, new possibilities such as reasoning about
+    Booleans, and the need to curb the explosion of specific higher-order rules. We
+    describe techniques that address these issues and extensively evaluate their
+    implementation in the Zipperposition theorem prover. Largely thanks to their use,
+    Zipperposition won the higher-order division of the CASC competition in 2020.  
+  \end{abstract}
+\newpage
+
+\section{Introduction}
+\label{sec:ho-tech:intro}
+
+% %In recent decades,
+% Superposition-based first-order automatic theorem provers
+% have emerged as useful reasoning tools. They dominate at the annual CASC
+% \cite{gs-2016-casc} theorem prover competition, having always won the
+% first-order theorem division. They are also used as backends to proof assistants
+% \cite{ck-18-coqhammer,ku-15-holyhammer,pb-12-sh}, automatic
+% higher-order theorem provers \cite{sb-21-leo3}, and software verifiers
+% \cite{fp-13-why3}.
+
+% The superposition calculus has only recently been extended
+% to higher-order logic (more precisely, extensional simple type theory
+% \cite{henkin-1950-completeness}), resulting in
+% \emph{\lsup} \cite{bbtvw-21-sup-lam}, which we developed
+% together with Waldmann, as well as \emph{combinatory superposition}
+% \cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
+% calculi do not support an interpreted Boolean type,
+% they can be extended by ad hoc rules \cite{our-bool-paper} that support
+% most of the Boolean reasoning necessary in practice.
+
+% Both higher-order superposition calculi were designed to gracefully
+% extend first-order reasoning. As most steps in higher-order
+% proofs tend to be essentially first-order, extending the most successful first-order
+% calculus to higher-order logic seemed worth trying.
+% Our first attempt at testing this idea was in 2019:
+% Zipperposition~1.5, based on \lsup, finished third
+% in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
+% 12~percentage points behind the winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.
+
+The landscape of higher-order proving techniques based on extension of efficient
+first-order ones has tremendously expanded in the late 2010s and early 2020s. As
+mentioned is Sect.~\ref{sec:pre:ho-sup-calculi} we have implemented three
+higher-order calculi---\lfsup{}, \lsup{}, and \osup{}---which extend first-order superposition in a graceful way.
+Bhayat and Reger also gracefully extended superposition to higher-order logic using
+\textsf{SKBCI} combinators \cite{br-20-full-sup-w-combs}. Significant progress has been
+made on SMT front as well \cite{brotb-19-ho-smt}.
+
+In 2019 we tested if the idea of gracefully extending first-order provers to
+higher-order logic really improves state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
+Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
+calculus to deal with Booleans. It finished third at that year's higher-order
+division of CASC competition \cite{gs-19-casc27}, 12~percentage points behind the
+winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.
+
+Studying the competition results, we found that higher-order tableaux have some
+advantages over higher-order superposition. To bridge the gap, we developed
+techniques and heuristics that simulate tableaux in the context of saturation.
+We implemented them in Zipperposition~2, which took part at the higher-order
+division of CASC \cite{gs-21-cascj10} in 2020. This time, our prover won the
+division, proving 84\% of the problems, a whole 20~percentage points ahead of
+the runner-up, Satallax 3.4.
+
+In this chapter, we describe the main techniques that explain this reversal of
+fortunes. They cover most parts of a modern higher-order theorem prover, from
+preprocessing to additional calculus rules to heuristics to backend integration.
+Compared to the previous chapter, in which we discussed rules used to treat
+Boolean terms, in this chapter we use a newer version of Zipperposition, based
+on a newer calculus. Instead of \lsup{} augmented with ad hoc Boolean rules, we
+work with {\osup} \cite{bbtv-21-full-ho-sup}, a principled extension of
+superposition to full higher-order logic, including an interpreted Boolean type.
+
+Many higher-order problems extensively use symbol definitions to simplify
+their representation. We describe several ways to exploit the definitions,
+%of which the most successful is
+such as turning them into rewrite rules (Sect.~\ref{sec:ho-tech:preprocessing}).
+%Interesting patterns can be observed in various higher-order problem encodings.
+%We show how we can exploit these to simplify problems (Sect.~\ref{sec:ho-tech:preprocessing}).
+%
+By working on formulas rather than clauses, tableau techniques take a more
+holistic view of a higher-order problem.
+Through its support for delayed clausification and, more generally,
+calculus-level formula manipulation, \osup{} enables us to
+simulate most successful tableau techniques in a saturating prover
+(Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
+functions}, a mechanism that allows us to choose on which Boolean subterms
+to perform inferences first.
+We implemented some Boolean selection functions and
+evaluated them (Sect.~\ref{sec:ho-tech:bool-select}).
+
+The main drawback of both $\lambda$-superposition variants compared with
+combinatory superposition is that they rely on rules that enumerate possibly
+infinite sets of unifiers. We describe a mechanism that interleaves infinitely
+branching inferences with the standard saturation process
+(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover retains the same
+behavior as before on first-order problems, smoothly scaling with increasing
+numbers of higher-order clauses.
+%
+We also propose some heuristics to curb the explosion induced by highly
+prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).
+
+Using first-order backends to finish the proof is common practice in
+higher-order reasoning. Since \osup{} coincides with standard
+superposition on first-order clauses, invoking backends may
+seem redundant; yet Zipperposition is nowhere as efficient as E
+\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
+efficient backend does make sense. We describe how to achieve a balance
+between allowing native higher-order reasoning and
+delegating reasoning to a backend (Sect.~\ref{sec:ho-tech:backends}).
+%
+Finally, we compare Zipperposition~2 with other provers on all monomorphic
+higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
+evaluation than at CASC (Sect.~\ref{sec:ho-tech:comparison}). Our evaluation
+corroborates the competition results.

### 2022-03-11 introduction/introduction.tex
intro

@@
   decisions that had to be evaluated in a short amount of time we decided to
-  implement the calculus in Zipperposition. This prover is less efficient than
+  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
   E, but it is implemented in OCaml and it was designed to make prototyping
@@
 provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
-\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
+\cite{sc-15-simon-phd,sc-supind-17} with arithmetic, induction, and rudimentary support for
 higher-order logic. Bhayat and Reger implemented a complete superposition

### 2022-03-11 scratch/paper.tex
intro

@@
 \newcommand{\simp}[2]{\namedinference{}{\strut#1}{\strut#2}}
-\newcommand{\lamsupraw}{$\lambda$-super\-po\-si\-tion}
-\newcommand{\lamsup}{{\color{black}\lamsupraw}}
-\newcommand{\ofreelamsup}{{\color{black}Boolean-free \lamsupraw}}
-\newcommand{\olamsup}{{\color{black}Boolean \lamsupraw}}
+\newcommand{\lsupraw}{$\lambda$-super\-po\-si\-tion}
+\newcommand{\lsup}{{\color{black}\lsupraw}}
+\newcommand{\lsup}{{\color{black}Boolean-free \lsupraw}}
+\newcommand{\osup}{{\color{black}Boolean \lsupraw}}
 
@@
 \section{Introduction}
-\label{sec:intro}
+\label{sec:ho-tech:intro}
 
@@
 \cite{henkin-1950-completeness}), resulting in
-\emph{\ofreelamsup} \cite{bbtvw-21-sup-lam}, which we developed
+\emph{\lsup} \cite{bbtvw-21-sup-lam}, which we developed
 together with Waldmann, as well as \emph{combinatory superposition}
@@
 Our first attempt at testing this idea was in 2019:
-Zipperposition~1.5, based on \ofreelamsup, finished third
+Zipperposition~1.5, based on \lsup, finished third
 in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
@@
 integration. We use a newer version of Zipperposition, based on a newer
-calculus:\ Instead of \ofreelamsup{} augmented with ad hoc Boolean rules,
-we work with \emph{\olamsup} \cite{bbtv-21-full-ho-sup}, a principled extension of superposition to full
+calculus:\ Instead of \lsup{} augmented with ad hoc Boolean rules,
+we work with \emph{\osup} \cite{bbtv-21-full-ho-sup}, a principled extension of superposition to full
 higher-order logic, including an interpreted Boolean type.
@@
 %of which the most successful is
-such as turning them into rewrite rules (Sect.~\ref{sec:preprocessing}).
+such as turning them into rewrite rules (Sect.~\ref{sec:ho-tech:preprocessing}).
 %Interesting patterns can be observed in various higher-order problem encodings.
-%We show how we can exploit these to simplify problems (Sect.~\ref{sec:preprocessing}).
+%We show how we can exploit these to simplify problems (Sect.~\ref{sec:ho-tech:preprocessing}).
 %
@@
 Through its support for delayed clausification and, more generally,
-calculus-level formula manipulation, \olamsup{} enables us to
+calculus-level formula manipulation, \osup{} enables us to
 simulate most successful tableau techniques in a saturating prover
-(Sect.~\ref{sec:formulas}). This calculus also supports \emph{Boolean selection
+(Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
 functions}, a mechanism that allows us to choose on which Boolean subterms
@@
 We implemented some Boolean selection functions and
-evaluated them (Sect.~\ref{sec:bool-select}).
-
-The main drawback of both \lamsup{} variants compared with combinatory
+evaluated them (Sect.~\ref{sec:ho-tech:bool-select}).
+
+The main drawback of both \lsup{} variants compared with combinatory
 superposition is that they rely on rules that enumerate possibly infinite sets
@@
 branching inferences with the standard saturation process
-(Sect.~\ref{sec:infinite-branching}). The prover
+(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover
 retains the same behavior as
@@
 We also propose some heuristics to curb the explosion induced by highly
-prolific calculus rules (Sect.~\ref{sec:explosiveness}).
+prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).
 
 Using first-order backends to finish the proof is common practice in
-higher-order reasoning. Since \lamsup{} coincides with standard
+higher-order reasoning. Since \lsup{} coincides with standard
 superposition on first-order clauses, invoking backends may
@@
 between allowing native higher-order reasoning and
-delegating reasoning to a backend (Sect.~\ref{sec:backends}).
+delegating reasoning to a backend (Sect.~\ref{sec:ho-tech:backends}).
 %
@@
 higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
-evaluation than at CASC (Sect.~\ref{sec:comparison}). Our evaluation
+evaluation than at CASC (Sect.~\ref{sec:ho-tech:comparison}). Our evaluation
 corroborates the competition results.
@@
 \section{Background and Setting}
-\label{sec:background}
+\label{sec:ho-tech:background}
 
@@
 $\iforall x.\,t$ and $\iexists x.\,t$. Following %the convention of
-\olamsup{}, predicate literals are encoded as equations with $\itrue$ or
+\osup{}, predicate literals are encoded as equations with $\itrue$ or
 $\ifalse$: for example, $\cst{even}(x)$ becomes $\poslit{\cst{even}(x)}$, and
@@
 \looseness=-1
-The \olamsup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
+The \osup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
 complete inference system and redundancy criterion for higher-order logic with
@@
 
-Unlike \ofreelamsup, this calculus 
+Unlike \lsup, this calculus 
 does not require axioms defining the logical symbols to cope with formulas.
@@
 
-In contrast to both \lamsup{} variants, combinatory superposition can
+In contrast to both \lsup{} variants, combinatory superposition can
 avoid enumerating CSUs by
@@
 Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
-theorem prover that implements both \lamsup{} variants, combinatory
+theorem prover that implements both \lsup{} variants, combinatory
 superposition, and other superposition-like calculi.
@@
 of Zipperposition parameters as a baseline for all comparisons. This is an
-incomplete, pragmatic configuration of \olamsup{} using heuristics expected to perform
+incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
 well on a wide range of problems.
@@
 \section{Preprocessing Higher-Order Problems}
-\label{sec:preprocessing}
+\label{sec:ho-tech:preprocessing}
 
@@
 \section{Reasoning about Formulas}
-\label{sec:formulas}
+\label{sec:ho-tech:formulas}
 
@@
 % \emph{lazy clausification} rules \cite[Sect.~3.4]{our-bool-paper},
-% which extend \ofreelamsup{}. These incrementally clausify top-level
+% which extend \lsup{}. These incrementally clausify top-level
 % logical symbols; for example, a clause $C' \llor \neglit{(\cst{p} \iand \cst{q})}$ yields a
@@
 % \looseness=-1
-The \olamsup{} calculus supports \emph{delayed clausification rules}
+The \osup{} calculus supports \emph{delayed clausification rules}
 that insert problems into the proof state in their original,
@@
 
-An earlier evaluation of the \olamsup{} calculus \cite{bbtv-21-full-ho-sup} showed
+An earlier evaluation of the \osup{} calculus \cite{bbtv-21-full-ho-sup} showed
 that the \emph{outer} variant of delayed clausification substantially increases this calculus's performance.
@@
 \section{Exploring Boolean Selection Functions}
-\label{sec:bool-select}
+\label{sec:ho-tech:bool-select}
 
@@
 respect to the term order. To further restrict which terms can be targeted by
-an inference, the \olamsup{} calculus introduces \emph{Boolean selection
+an inference, the \osup{} calculus introduces \emph{Boolean selection
 functions}.
@@
 t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
-\olamsup{} mandates additional restrictions on the Boolean selection function:
+\osup{} mandates additional restrictions on the Boolean selection function:
 $\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
@@
 have a modest effect. To better assess this feature, in this evaluation we use
-\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:formulas} as the baseline
+\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:ho-tech:formulas} as the baseline
 configuration. To avoid interference of literal and Boolean selection, we
@@
 outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
-\texttt{NUM636\^{}2} from Sect.~\ref{sec:formulas}, using Boolean selection
+\texttt{NUM636\^{}2} from Sect.~\ref{sec:ho-tech:formulas}, using Boolean selection
 can reduce the number of given clause loop iterations from
@@
 \section{Enumerating Infinitely Branching Inferences}
-\label{sec:infinite-branching}
+\label{sec:ho-tech:infinite-branching}
 
@@
 increasing weight\confrep{}{, a low weight corresponding to a high priority}.
-When they introduced \lamsup, Bentkamp et al.\
+When they introduced \lsup, Bentkamp et al.\
 \cite{bbtvw-21-sup-lam} described an older version of this extension. Here
@@
 \section{Controlling Prolific Rules}
-\label{sec:explosiveness}
+\label{sec:ho-tech:explosiveness}
 
@@
 many refutationally complete calculi employ highly prolific rules.
-For example, \lamsup{} includes a
+For example, \lsup{} includes a
 \infname{FluidSup} rule \cite{bbtvw-21-sup-lam} that very often applies to two
@@
 
-% Complete provers that implement \lamsup{} %, such as Zipperposition,
+% Complete provers that implement \lsup{} %, such as Zipperposition,
 % have two main heuristic choices: which inference stream to query for clauses
@@
 \section{Controlling the Use of Backends}
-\label{sec:backends}
+\label{sec:ho-tech:backends}
 
@@
 during a proof attempt and spend a substantial amount of time in backend
-collaboration. Since \lamsup{} generalizes a highly efficient
-first-order calculus, we expect that future efficient \lamsup{}
+collaboration. Since \lsup{} generalizes a highly efficient
+first-order calculus, we expect that future efficient \lsup{}
 implementations will not benefit much from backends.
@@
 the subset of $M$ that contains only clauses that were derived using at least
-one \lamsup{} rule not present in regular superposition. We order the clauses in
+one \lsup{} rule not present in regular superposition. We order the clauses in
 $M_\mathrm{ho}$ by increasing derivation depth, using syntactic weight to break
@@
 \section{Comparison with Other Provers}
-\label{sec:comparison}
+\label{sec:ho-tech:comparison}
 
@@
 Semantically, E~2.7 is arguably the weakest among the listed provers:
-It simply performs $o$-RW rewriting described in Sect.~\ref{sec:preprocessing} followed by
+It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
 $\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
@@
 % behind Satallax's \emph{cooperative} mode. This confirms our hypothesis that
-% \olamsup{} is a suitable basis for automatic higher-order reasoning.
+% \osup{} is a suitable basis for automatic higher-order reasoning.
 % \confrep{This also}{The increase in performance due to the addition of an efficient backend}
@@
 show that Zipperposition outperforms all other provers on SH benchmarks. This
-confirms our hypothesis that \lamsup{} is a suitable basis for automatic
+confirms our hypothesis that \lsup{} is a suitable basis for automatic
 higher-order reasoning. Further confirmation is provided by the success rate of
@@
 \section{Discussion and Conclusion}
-\label{sec:discussion}
+\label{sec:ho-tech:discussion}
 
 %  In this paper, we proposed solutions to issues that arise when implementing a
-%  higher-order prover based on \lamsup. Some of these also apply
+%  higher-order prover based on \lsup. Some of these also apply
 %  to combinatory superposition. We also discussed many choices that can be made
@@
 %\section{Conclusion and Future Work}
-%\label{sec:conclusion}
+%\label{sec:ho-tech:conclusion}
 
@@
 % We described some ways in which the explosion incurred by
-% \lamsup{} can be kept under control. We showed how native
+% \lsup{} can be kept under control. We showed how native
 % support for a Boolean type can dramatically improve a higher-order prover's

### 2022-03-11 unif/unif.tex
intro

@@
 We implemented our procedure, several oracles,
-and the fingerprint index in
-the Zipperposition prover (Sect.~\ref{sec:unif:implementation}). Since a
+and the fingerprint index (Sect.~\ref{sec:unif:implementation}) in
+the Zipperposition prover \cite{sc-supind-17,sc-15-simon-phd}. Since a
 straightforward implementation of the JP procedure already existed in
@@
 unification algorithms. Thus, we used all 2606 monomorphic higher-order theorems
-from the TPTP library \cite{gs-17-tptp} and 832 monomorphic higher-order
+from the TPTP 7.2.0 library \cite{gs-17-tptp} and 832 monomorphic higher-order
 Sledgehammer (SH) generated problems \cite{ns-13-leo2sh} as our benchmarks.\footnote{An archive with raw
@@
 to evaluate the performance of our procedure. We compare it with the complete
-variant of our procedure (\textbf{cv}) and pragmatic variants
-(\textbf{pv}) with several different configurations of limits for
-applied bindings. All other Zipperposition parameters have been fixed to the values of
-a variant of a well-performing configuration we used for the CASC-27 theorem proving competition \cite{gs-19-casc27}.
-The cv configuration and all of the pv
-configurations use only pattern unification as an underlying oracle. To test
-the effect of oracle choice, we evaluated the complete variant
-in eight combinations: with no oracles
-(\textbf{n}), with only fixpoint (\textbf{f}), pattern (\textbf{p}), or solid (\textbf{s}) oracle,
-and with their combinations: \textbf{fp}, \textbf{fs}, \textbf{ps}, \textbf{fps}.
+variant of our procedure (\textbf{cv}) and pragmatic variants (\textbf{pv}) with
+several different configurations of limits for applied bindings. All other
+Zipperposition parameters have been fixed to the values of a variant of a
+well-performing configuration we used for the 2019 installment of the higher-order
+division of CASC theorem proving competition \cite{gs-19-casc27}. The cv
+configuration and all of the pv configurations use only pattern unification as
+an underlying oracle. To test the effect of oracle choice, we evaluated the
+complete variant in eight combinations: with no oracles (\textbf{n}), with only
+fixpoint (\textbf{f}), pattern (\textbf{p}), or solid (\textbf{s}) oracle, and
+with their combinations: \textbf{fp}, \textbf{fs}, \textbf{ps}, \textbf{fps}.
 

### 2022-03-12 ho-techniques/ho-techniques.tex
pre

@@
 In 2019 we tested if the idea of gracefully extending first-order provers to
-higher-order logic really improves state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
+higher-order logic really improves the state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
 Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
@@
 
+\looseness=-1
 The main drawback of both $\lambda$-superposition variants compared with
@@
 %
-We also propose some heuristics to curb the explosion induced by highly
+We also propose heuristics to curb the explosion induced by highly
 prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).
@@
 corroborates the competition results.
+
+\section{Background and Setting}
+\label{sec:ho-tech:background}
+
+We focus on monomorphic higher-order logic, without the axiom of infinity or the
+axiom of at least two individuals \cite{bm-14-automation-ho}. However, the
+techniques can easily be extended with polymorphism. Indeed, Zipperposition
+already supports some of them polymorphically. Further, we use exactly the same
+notation for this logic as introduced in Chapter \ref{ch:pre}. Since we are
+working with extensions of superposition, we assume the clausal structure
+(Sect.~\ref{sec:pre:clauses}). Like in the previous chapter, literals
+of clauses can contain arbitrary higher-order terms, including formulas.
+
+\ourpara{Higher-Order Calculi}
+\looseness=-1
+We briefly introduced the \osup{} calculus \cite{bbtv-21-full-ho-sup} in
+Sect.~\ref{sec:pre:ho-sup-calculi}. It is a refutationally complete inference
+system and redundancy criterion for higher-order logic with rank-1 polymorphism,
+Hilbert choice, and functional and Boolean extensionality.
+% The calculus relies on
+% \emph{complete sets of unifiers}
+% (\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
+% $V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
+% that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
+% \in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
+% for all variables $X \in V$. The set $V$ is used to distinguish
+% important variables from auxiliary variables (which may arise in intermediary
+% states of the unification procedure). We usually omit it.
+Unlike \lsup{}, this calculus 
+does not require axioms defining the logical symbols to cope with formulas.
+Instead, it includes Boolean inference rules that mimic
+superposition from such axioms into Boolean subterms,
+while avoiding the explosion incurred by adding these axioms to the proof state. It
+also includes rules that simulate Boolean inferences below applied variables.
+Both sets of rules are disabled or replaced with incomplete, ad hoc rules
+described in the previus chapter in most configurations
+of the CASC portfolio.
+A new feature of the calculus that we explore in detail is
+the ability to select Boolean subterms
+to restrict Boolean and superposition inferences.
+
+In contrast to both $\lambda$-superposition variants, combinatory superposition
+does not require enumerating elements of CSU to compute results of inferences.
+Instead, it avoids computing CSUs by using a form of first-order unification.
+Essentially, it enumerates higher-order terms using rules that instantiate
+applied variables with partially applied combinators from the complete
+combinator set $\{\cst{S}, \cst{K}, \cst{B}, \cst{C}, \cst{I}\}$. This calculus
+is the basis of Vampire~4.5 \cite{br-20-full-sup-w-combs}, which finished
+closely behind Satallax 3.4 %~and~3.5
+at higher-order division of CASC in 2020.
+
+\looseness=-1
+A different, very successful calculus is Satallax's SAT-guided tableaux
+\cite{backes-brown-2011}. Satallax was the leading higher-order prover of the
+2010s. Its simple and elegant tableaux avoid deep superposition-style rewriting
+inferences.
+Nevertheless, our working hypothesis for the past years has been
+that superposition would likely provide a stronger basis for higher-order
+reasoning.
+Other competing higher-order calculi include SMT (implemented in CVC4
+\cite{brotb-19-ho-smt, cbetal-11-cvc4}) and extensional paramodulation (implemented in Leo-III \cite{sb-21-leo3}).
+
+\ourpara{Zipperposition}
+Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
+theorem prover that implements both $\lambda$-superposition variants, combinatory
+superposition, and other superposition-like calculi.
+The prover was conceived as a testbed for rapidly
+experimenting with extensions of first-order superposition, but over time it
+has assimilated many of E's techniques and heuristics and become quite powerful.
+Several of our techniques extend the \emph{given clause procedure}, the standard
+saturation procedure described in Sect.~\ref{sec:pre:saturation}.
+
+\ourpara{Experimental Setup}
+To assess our techniques, we carried out experiments with Zipperposition~2. We
+used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
+TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
+and 1253 Sledgehammer-generated
+monomorphic higher-order problems (labeled \emph{SH}).
+Although some techniques support polymorphism, we
+uniformly used monomorphic benchmarks.
+
+We fixed a \emph{base} configuration
+of Zipperposition parameters as a baseline for all comparisons. This is an
+incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
+well on a wide range of problems.
+%Note that
+%we use a different baseline configuration than in our earlier paper \cite{making-ho-work}.
+%%% That goes without saying. We use a different calculus! Now clarified in intro. --JB
+In each experiment, we varied
+the parameters associated with a specific technique to evaluate it. The
+experiments were run on StarExec Miami \cite{sst-14-starexec} servers, equipped with
+Intel Xeon E5-2620 v4 CPUs clocked at 2.10 GHz. Unless otherwise stated, we used a
+CPU time limit of 15~s, roughly the time each configuration is given in the
+CASC portfolio mode. The raw evaluation results are available online.%
+\footnote{\url{http://doi.org/10.5281/zenodo.5007440}}

### 2022-03-12 scratch/paper.tex
pre

@@
 \newcommand\NumberNOK[1]{\colorbox{red}{#1}}
-\newcommand{\ourparagraph}[1]{\paragraph{\upshape\bfseries#1}}
+\newcommand{\ourpara}[1]{\paragraph{\upshape\bfseries#1}}
 
@@
 
-\ourparagraph{Higher-Order Logic}
+\ourpara{Higher-Order Logic}
 We define terms $s, t, u, v$ inductively as free variables $F, X$, bound
@@
 
-\ourparagraph{Higher-Order Calculi}
+\ourpara{Higher-Order Calculi}
 \looseness=-1
@@
 
-\ourparagraph{Zipperposition}
+\ourpara{Zipperposition}
 Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
@@
 
-\ourparagraph{Experimental Setup}
+\ourpara{Experimental Setup}
 To assess our techniques, we carried out experiments with Zipperposition~2. We
@@
 
-\ourparagraph{Evaluation and Discussion}
+\ourpara{Evaluation and Discussion}
 
@@
 
-\ourparagraph{Evaluation and Discussion}
+\ourpara{Evaluation and Discussion}
   %

### 2022-03-12 ho-techniques/ho-techniques.tex
more

@@
 \footnote{\url{http://doi.org/10.5281/zenodo.5007440}}
+
+\section{Preprocessing Higher-Order Problems}
+\label{sec:ho-tech:preprocessing}
+
+The TPTP library contains thousands of higher-order problems. Despite their
+diversity, they have a markedly different flavor from the TPTP first-order
+problems. Notably, they extensively use the \verb|definition| role to identify
+universally quantified equations (and equivalences) that define symbols.
+%
+Definitions $s \eq t$ (or $(s \iequiv t) \eq \itrue$) can be replaced by rewrite
+rules $s \longrightarrow t$,
+using the orientation given in the input problem. If there are multiple
+definitions for the same symbol, only the first one is replaced by a rewrite rule.
+Then, whenever a clause is picked in the given clause procedure, it will be rewritten
+using the collected rules.
+Alternatively, we can rewrite
+the input formulas as a preprocessing step. This ensures that the input
+clauses will be fully simplified when the proving process starts and no
+defined symbols will occur in clauses, which usually helps the heuristics.
+
+Since the TPTP format enforces no constraints on
+definitions, rewriting might diverge. To ensure
+termination, we limit the number of applied rewrite steps. In
+practice, most TPTP problems are well behaved: Only one
+definition is given for each symbol, and the definitions are acyclic.
+
+Turning the defining equations into rewrite rules, unfolding the definitions, and
+$\beta$-reduc\-ing the result can eliminate all of a problem's higher-order features, making
+it susceptible to first-order methods. However, this can inflate the problem
+beyond recognition and compromise the refutational completeness of
+superposition.
+
+\begin{exa} 
+  Removing higher-order features of a problem can have adverse effects.
+  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
+  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
+  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
+  successor constructor. When this definition is kept as is, the
+  prover can superpose from either $\cst{m}$ or its definition into the
+  (clausified) induction axiom, which is also given in the problem, and quickly prove
+  the conjecture, without using any advanced inductive reasoning. In contrast,
+  when the definition is
+  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
+  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
+  correct instantiation for the induction axiom.
+\end{exa}
+
+We describe two techniques to mitigate these issues. The first one is based on the observation that in practice,
+the explosion associated with definition unfolding mostly
+manifests itself on definitions of nonpredicate symbols. In some cases, it is
+preferable to rely on superposition's term order and powerful simplification
+engine to rewrite the proof state rather than to blindly rewrite definitions. On
+the other hand, superposition's reasoning with equivalences is often inadequate
+\cite{bbtv-21-full-ho-sup, gs-05-boolsup}. Thus, it makes sense to treat only
+predicate definitions as rewrite rules.
+
+The second technique aims at preserving completeness: We can try to force the term order that
+parameterizes superposition to orient as many definitions as possible and rely on
+demodulation to simplify the proof state. Usually, the Knuth--Bendix order (KBO)
+\cite{db-1970-kbo} is used. It compares terms by first comparing their weights,
+which is the sum of all the weights assigned to the symbols it contains. Given a
+symbol weight assignment $\mathcal{W}$, we can update it so that it orients
+acyclic definitions from left to right assuming that they are of the form $
+\cst{f} \, \overline{X}_m \eq \lambda \overline{Y}_n. \, t$, where the only free
+variables in $t$ are $\overline{X}_m$, no free variable is repeated or appears
+applied in $t$, and $\cst{f}$ does not occur in $t$. Then we traverse the
+symbols $\cst{f}$ that are defined by such equations following the dependency
+relation, starting with a symbol $\cst{f}$ that does not depend on any other
+defined symbol. For each $\cst{f}$, we set $\mathcal{W}(\cst{f})$ to $w + 1$,
+where $w$ is the maximum weight of the right-hand sides of $\cst{f}$'s
+definitions, computed using $\mathcal{W}$. By construction, for each equation
+the left-hand side is heavier. Thus, the equations are orientable from left to
+right.
+
+\begin{exa} 
+  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
+  in higher-order logic. More complex modal operators (such as
+  implication and equivalence) are defined in terms of basic connectives (such as negation
+  and disjunction). Some of the definitions present in the problems are
+  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
+  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
+  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
+  variables, and basic connectives is 2, we can orient equations using
+  the above described approach as follows. Starting from symbols that do not
+  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
+  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
+  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
+  enable us to orient all definitions from left to right.
+\end{exa}
+
+% Many higher-order problems, especially those generated from proof assistants,
+% contain hundreds of needless axioms.
+% To filter out axioms that are unlikely to be useful in a proof attempt, many
+% theorem provers rely on the SInE algorithm \cite{hv-2011-sine}. SInE starts with
+% the set of symbols occurring in the conjecture and tries to find axioms that
+% define the properties of these symbols. Then it looks for the definitions of
+% newly found symbols until it reaches a fixpoint. Axioms annotated with
+% \verb|definition| ease this search because they explicitly record the
+% dependency between a symbol and its characterization. To exploit
+% this information, we modified SInE to optionally include the definitions of
+% symbols in the conjecture, regardless of whether they are filtered out or not.
+% Similarly, we implemented mode of SInE which selects only conjecture and axioms
+% annotated with \verb|definition|.
+
+\ourpara{Evaluation and Discussion}
+
+% The \textit{base} configuration treats all axioms annotated with
+% \texttt{definition} as rewrite rules applied as preprocessing.
+% In addition, we tested
+We designed and evaluated the following strategies for handling
+\texttt{definition} axioms:
+
+\begin{description}[labelwidth=\widthof{\rm no-RW$+$KBO~}]
+  \item[\rm pre-RW~] rewrite all definitions as a preprocessing step;
+  \item[\rm in-RW~] rewrite all definitions during the saturation, as an inprocessing step;
+  \item[\rm $o$-RW~] rewrite only predicate definitions, during preprocessing;
+  \item[\rm $o$-RW$+$KBO~] like $o$-RW but with adjusted KBO weights for the remaining
+    definitions;
+  \item[\rm no-RW~] no special treatment of definitions;
+  \item[\rm no-RW$+$KBO~] like $no$-RW but adjusting KBO weights for all definitions.
+\end{description}
+
+The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
+cell gives the number of proved problems, and cells marked with $\star$
+correspond to the base configuration. The highest number in a category is typeset in
+\relax{bold}. SH benchmarks are not
+included because they do not contain the \texttt{definition} role.
+
+\newcommand{\unknownres}{\ensuremath{{\varnothing}}}
+\newcommand{\colalign}{\phantom{0}}
+
+\begin{figure}[t]
+  \centering
+  \def\arraystretch{1.1}%
+  \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
+                & pre-RW                 & in-RW        & $o$-RW               & $o$-RW$+$KBO    & no-RW        & no-RW$ + $KBO  \\ \midrule
+           TPTP & {\bf 1635}$^\star$     & 1619         & 1620                 & 1621            & 1298         & 1296
+           \\ \bottomrule
+         \end{tabular}}
+       \captionof{figure}{Impact of the definition rewriting method}
+       \label{fig:rewrite}
+     \end{figure}
+
+The four configurations in which definitions are treated as rewrite rules
+perform much better than the other two. In contrast, adjusting KBO weights gives
+no substantial improvement: Looking at raw data, we found only
+\NumberOK{2}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
+feature was used in the proof. For no-RW and no-RW$+$KBO, the
+\NumberOK{2}-problem difference may be just noise. Even though it proves fewer
+problems, the configuration $o$-RW has some advantages over pre-RW: It proves
+\NumberOK{16} problems that pre-RW does not, \NumberOK{3} of which have a TPTP
+difficulty rating (the ratio of eligible %TPTP
+provers that cannot prove the problem) of~1.
+
+% \looseness=-1
+Rewriting after clausification avoids getting stuck rewriting parts of the
+proof state that might not contribute to the proof. In practice, we noticed that
+rewriting can be so expensive that the prover can spend all
+allotted CPU time in the preprocessing phase. The evaluation results confirm this
+observation: There are \NumberOK{64} problems proved by in-RW but not by
+pre-RW. Moreover, there are \NumberOK{41} problems that can be proved only
+by in-RW but not by any other above described configuration. % from Fig.~\ref{fig:rewrite}.
+
+\section{Reasoning about Formulas}
+\label{sec:ho-tech:formulas}
+
+Higher-order logic identifies formulas with terms of Boolean type. To prove a problem, we often
+need to instantiate a variable with the right predicate.
+Finding this predicate can be easier if the problem is not clausified.
+Consider the conjecture $\iexists f. \, f \, \cst{p} \, \cst{q} \iequiv \cst{p}
+\iand \cst{q}$. Expressed in this form, the formula is easy to prove by taking
+$f := \lambda x \, y. \> x  \iand y$. By contrast, guessing the right
+instantiation for the negated, clausified form $ \neglit{F \, \cst{p} \,
+\cst{q}} \llor \neglit{\cst{p}} \llor \neglit{\cst{q}},
+\poslit{F \, \cst{p} \, \cst{q}} \llor \poslit{\cst{p}}$, $\poslit{F \, \cst{p}
+\, \cst{q}} \llor \poslit{\cst{q}}$ is more challenging.
+One of the strengths of higher-order tableau provers is that they do not clausify the input
+problem. This might partly explain Satallax's dominance in the THF division of CASC
+competitions until the 2020 edition of CASC.
+
+The \osup{} calculus supports \emph{delayed clausification rules} that insert
+problems into the proof state in their original, nonclausified form, and
+clausify them gradually. Delayed clausification allows the prover to analyze the
+syntactic structure of formulas during saturation, whereas the more traditional
+approach of \emph{immediate clausification} applies a standard clausification
+algorithm \cite{nw-01-small-cnf} both as a preprocessing step and whenever
+predicate variables are instantiated.
+
+An earlier evaluation of the \osup{} calculus \cite{bbtv-21-full-ho-sup} showed
+that the \emph{outer} variant of delayed clausification substantially increases
+this calculus's performance. The outer variant clausifies top-level logical
+symbols, proceeding from the outside inwards; this method corresponds to the one
+described in the previous chapter. For example, a clause $C \llor
+\neglit{(\cst{p} \iand \cst{q})}$ is transformed into $C \llor \neglit{\cst p}
+\llor \neglit{\cst q}$. The calculus also supports \emph{inner} delayed
+clausification, which uses only the core calculus rules to clausify problems.
+Even though this is the laziest approach to clausification, the earlier
+evaluation by Bentkamp et al. \cite{bbtv-21-full-ho-sup} showed that this
+approach is inefficient. Thus, we focus only on the outer rules.
+
+\looseness=-1
+Delayed clausification rules can be used as inference rules (which add conclusions
+to the passive set) or as simplification rules (which delete premises and add
+conclusions to the passive set).
+%
+Inferences are more flexible, as all
+intermediate clausification states will be stored in the proof state, at the
+cost of producing many clauses. Simplifications produce fewer clauses,
+but risk destroying informative syntactic structure.
+Since clausifying equivalences can destroy a lot of syntactic structure
+\cite{gs-05-boolsup}, we never apply simplifying rules
+on them.
+
+Delayed clausification can interfere with clause splitting techniques.
+Zipperposition supports a lightweight variant of AVATAR \cite{av-2014-avatar},
+an architecture that partitions the search space by
+splitting clauses into variable-disjoint subclauses. This lightweight AVATAR is described
+by Ebner et al.\ \cite[Sect.~7]{2021-ebt-unifying-splitting}. Combining it
+with delayed clausification makes it possible to split a %higher-order
+clause $(\varphi_1 \ior \cdots \ior \varphi_n) \eq \itrue$, where
+the $\varphi_i$'s are arbitrarily complex formulas that share no free
+variables with each other, into clauses $\varphi_i \eq \itrue$.
+%
+To finish the proof, it suffices to derive the empty clause under each assumption
+$\varphi_i \eq \itrue$. Since the split is performed at the formula level, this
+technique resembles tableaux, but it exploits the strengths of superposition,
+such as its powerful redundancy criterion and simplification machinery, to
+close the branches.

### 2022-03-12 scratch/paper.tex
more

@@
 \spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
-\spnewtheorem{examplex}[theorem]{Example}{\bfseries}{\rmfamily}
+\spnewtheorem{exa}[theorem]{Example}{\bfseries}{\rmfamily}
 
@@
 
-\begin{examplex} 
+\begin{exa} 
   Removing higher-order features of a problem can have adverse effects.
@@
   correct instantiation for the induction axiom.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex} 
+\begin{exa} 
   Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
@@
   enable us to orient all definitions from left to right.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex}
+\begin{exa}
 \label{ex:dat056-2}
@@
 from which we easily derive a contradiction.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
+\begin{exa} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
 (\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
 \, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
-\end{examplex} 
+\end{exa} 
 %
@@
 
-\begin{examplex}
+\begin{exa}
   Selecting the right Boolean subterm can help avoid elaborating
@@
   \, y \, \cst{a})$, avoiding this explosion. 
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex} 
+\begin{exa} 
   \begin{sloppypar}
@@
   \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
-\end{examplex}
+\end{exa}
 

### 2022-03-13 ho-techniques/ho-techniques.tex
bools

@@
 \begin{exa}
+  \label{hot:exa:num636}
   Removing higher-order features of a problem can have adverse effects.
@@
 close the branches.
+
+Beyond splitting, interleaving clausification and saturation allows us to
+simulate another tableau-inspired technique. Whenever dynamic clausification
+substitutes a fresh variable $X$ for a predicate variable $x$ in a clause of the
+form $(\iforall x.\, \varphi) \eq \itrue \llor C$, yielding $\substterm{\{x
+\mapsto\nobreak X\}}{\varphi} \eq \itrue \llor C$, we can create additional
+clauses in which $x$ is replaced with $t \in \instset$, where $\instset$ is a
+set of heuristically chosen terms. This set contains $\lambda$-abstractions
+whose bodies are formulas and that occur in activated clauses; it also containts
+\emph{primitive instantiations}---that is, imitations (in the sense of
+higher-order unification) of logical symbols that approximate the shape of a
+predicate that can instantiate a predicate variable. Primitive instantiations
+are described in Sect.~\ref{sect:bool:native}.
+
+Since a new term $t$ can be added to $\mathit{Inst}$ after a clause with a
+quantified variable of $t$'s type has been activated, we remember the clauses
+$\sigmaterm{\{x \mapsto X\}}{\varphi} \eq \itrue\allowbreak \llor C$ and instantiate them when
+$\mathit{Inst}$ is extended. Conveniently, these instantiated clauses are not
+recognized as subsumed by Zipperposition, which uses an optimized, incomplete
+higher-order subsumption algorithm.
+
+\looseness=-1
+Given a disequation $\cst{f}\,\tuple{s}{n} \not\eq \cst{f}\,\tuple{t}{n}$, the
+\emph{abstraction} of $s_i$ is $\lambda x.\, u \ieq v$, where $u$ is obtained by
+replacing all occurrences of $s_i$ in $\cst{f}\,\tuple{s}{n}$ with $x$ and $v$ is
+obtained by replacing all occurrences of $s_i$ in
+$\cst{f}\,\tuple{t}{n}$ with $x$. For \confrep{}{an equation }$\cst{f}\,\tuple{s}{n} \eq
+\cst{f}\,\tuple{t}{n}$, the analogous abstraction is $\lambda x.\, \inot (u \ieq
+v)$.
+%
+%
+Adding abstractions of the literals occurring in the conjecture to $\mathit{Inst}$
+provides useful instantiations for formulas such as induction principles of
+datatypes. As the conjecture is negated\confrep{}{ in refutational theorem proving},
+the equation's polarity is inverted in the
+abstraction.
+
+\begin{exa}
+\label{ex:dat056-2}
+The clausified conjecture of the problem \texttt{DAT056\^{}2}
+\cite{ns-13-leo2sh} from the TPTP library is $\cst{ap} \, \cst{xs}
+\, (\cst{ap} \, \cst{ys} \, \cst{zs}) \not\eq \cst{ap} \, (\cst{ap} \, \cst{xs} \,
+\cst{ys}) \, \cst{zs}$, where $\cst{ap}$ is the \confrep{}{list }append operator defined
+recursively on its first argument and $\cst{xs}$, $\cst{ys}$, and $\cst{zs}$ are
+of list type. Abstracting $\cst{xs}$ from the disequation yields $t = \lambda \mathit{xs}.\, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
+\allowbreak\ieq\allowbreak \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$, which is added
+to $\mathit{Inst}$.
+Included in the problem is the induction axiom
+for the list datatype: $\iforall p. \, p \, \cst{nil} \iand (\iforall x \,
+\mathit{xs}. \, p \, \mathit{xs} \iimplies\allowbreak p \, (\cst{cons} \, x \,
+\mathit{xs})) \iimplies\allowbreak \iforall \mathit{xs}. \, p \, \mathit{xs}$, where
+$\cst{nil}$ and $\cst{cons}$ have the usual meanings.
+Instantiating $p$ with $t$ and
+using the $\cst{ap}$ definition, we can prove
+$\iforall \mathit{xs}. \, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
+\ieq \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$,
+from which we easily derive a contradiction.
+\end{exa}
+
+\ourpara{Evaluation and Discussion}
+
+\begin{figure}
+\centering
+  \def\arraystretch{1.1}%
+  \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 0.5em}@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
+    & & $+$\relax{LA} & $-$\relax{LA} \\ \midrule
+    TPTP &
+%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{TPTP}}}&
+      \relax{IC}  & 1616  & 1635$^\star$    \\
+    & \relax{DCI} & 1507  & 1532\phantom{$^\star$}    \\
+    & \relax{DCS} & 1668  & {\bf 1703}\phantom{$^\star$} \\ \midrule
+    SH &
+%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
+      \relax{IC}  & \colalign425 & \colalign452$^\star$    \\
+    & \relax{DCI} & \colalign362 & \colalign385\phantom{$^\star$}    \\
+    & \relax{DCS} & \colalign441 & \colalign\textbf{457}\phantom{$^\star$} \\ \bottomrule
+  \end{tabular}
+  \caption{Impact of clausification and~lightweight AVATAR}
+  \label{fig:avatar-clause}
+\end{figure}
+
+The base configuration (\emph{base}) uses immediate clausification (\relax{IC}) and 
+disables lightweight AVATAR ($-$\relax{LA}). To test
+the merits of delayed clausification, we vary \emph{base}'s parameters along two axes: We
+choose immediate clausification (\relax{IC}), delayed clausification as inference
+(\relax{DCI}), or delayed clausification as simplification (\relax{DCS}), and we
+either enable ($+$\relax{LA}) or disable ($-$\relax{LA}) lightweight AVATAR.
+Neither of the configurations uses instantiation with terms from $\instset$.
+
+Figure~\ref{fig:avatar-clause} shows that using delayed clausification as
+simplification greatly increases the success rate, regardless of whether
+lightweight AVATAR is used. Using delayed clausification as inference has the
+opposite effect on both problem sets, presumably due to the large number of
+clauses it creates. By manually inspecting the  proofs found by the \relax{DCS}
+configuration, we noticed that a main reason for its success is that it does
+not simplify away equivalences.
+%
+Overall, the lightweight AVATAR harms performance, but the sets of
+problems proved with and without it are vastly different. For example,
+the \relax{IC}$+$\relax{LA} configuration proves \NumberOK{38} problems not
+proved by \relax{IC}$-$\relax{LA} (i.e., \emph{base}) on TPTP benchmarks and
+\NumberOK{14} such problems on SH benchmarks.
+
+\looseness=-1
+The Boolean instantiation technique presented above requires delayed
+clausification.
+We assessed it in the best configuration from
+%To assess it, we enabled it in the best configuration from
+Fig.~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
+Zipperposition proves \NumberOK{1700} TPTP problems and \NumberOK{456} SH
+problems.
+On TPTP, even though $+$\relax{BI} solves \NumberOK{3} problems less
+than \relax{DCS}$-$LA, it is very useful: \NumberOK{41}
+problems can be proved with $+$\relax{BI} but not with \relax{DCS}$-$\relax{LA}. 
+Conversely, \NumberOK{44} problems are solved with \relax{DCS}$-$\relax{LA},
+but not with $+$\relax{BI}, which suggests that Boolean instantiation can be 
+explosive.
+%
+One of the problems Boolean instantiation helps solve is \texttt{NUM636\^{}2} (a
+re-encoding of \texttt{NUM636\^{}3}, which is disucssed in Example \ref{hot:exa:num636}).
+It conjectures that $\iforall x.\, \cst{s} \, x \ineq x$, where $x$ ranges over
+Peano-style numbers specified by $\cst{z}$ and $\cst{s}$. The given axioms are
+the induction principle $\forall p.\, p \, \cst{z} \,\iand\, \forall x. \, (p \,
+x \iimplies p \, (\cst{s} \, x)) \iimplies \forall x. \, p\, x$, injectivity
+$\forall x y. \, \cst{s}\,x \ieq \cst{s}\,y \iimplies x \ieq y$, and
+distinctness $\forall x. \, \cst{s}\,x \ineq \cst{z}$. The conjecture is easily
+proved if Boolean instantiation is enabled: Even though the conjecture literal
+cannot be abstracted, instantiating $p$ with the term $\lambda x.\, \cst{s}
+\, x \ineq x$ used in the encoding of the (nonclausified) conjecture leads to a proof in just
+\NumberOK{22} given clause loop iterations. Zipperposition also finds a
+proof using the \relax{DCI}$-$\relax{LA} configuration, but this requires
+\NumberOK{294} iterations.
+
+%Raw data reveals that Boolean instantiation
+The $+$\relax{BI} configuration proves \NumberOK{18} TPTP problems no other
+configuration from Fig.~\ref{fig:avatar-clause} can prove. Among these is
+\texttt{DAT056\^{}2} (Example~\ref{ex:dat056-2}). In contrast, on SH benchmarks, only
+\NumberOK{6} problems are proved using $+$\relax{BI} and not
+\relax{DCS}$-$\relax{LA}. For all these problems, Boolean instantiation does not
+appear in the proof, suggesting that this result is due to the randomness in the
+evaluation environment. The fact that \relax{BI} has no effect on SH benchmarks
+is to be expected because Sledgehammer does
+not include lemmas whose name contains the substring \texttt{.induct} and that
+contain predicate variables. Therefore, \relax{BI} applies to fewer clauses.

### 2022-03-13 pre/pre.tex
intro
bools

@@
 \looseness=-1
-An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq t$. A clause $C$ is a
-finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
-\cdots \llor l_n$. Free variables of clauses are implicitly
-universally quantified. Note that $\llor$ denoting disjunction in clauses
-is different from the formula-building $\ior$.
+An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em
+literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq
+t$. A clause $C$ is a finite multiset of literals, interpreted and written
+disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
+implicitly universally quantified. Note that $\llor$ denoting disjunction in
+clauses is different from the formula-building $\ior$. In general,
+formula-building logical connectives are written in boldface.
 
@@
 chooses a clause $C$ from $\mathcal{P}$, simplifies it, and moves it to
-$\mathcal{A}$. Then all inferences between $C$ and active clauses are performed.
+$\mathcal{A}$ (i.e., it \emph{activates} it). Then all inferences between $C$ and active clauses are performed.
 The resulting clauses are again simplified and put in $\mathcal{P}$.

### 2022-03-14 bools/bools.tex
past tense

@@
     superposition calculus to support Boolean reasoning. Our approach extends
-    inference rules that have been used only in a first-order setting, uses some
+    inference rules that have been used only in a first-order setting and uses some
     rules previously implemented in higher-order provers, as well as new rules.
@@
     theorem prover. The evaluation shows highly competitive performance of our approach
-    and clear improvement over previous techniques.
+    and a clear improvement over previous techniques.
 \end{abstract}
@@
 first-order logic to support the third feature of higher-order logic that
-requires translation. They described two approaches: One based on calculus-level
-treatment of Booleans and the other, which requires no changes to the calculus,
+requires translation. They described two approaches: One based on a calculus-level
+treatment of Booleans and other, which requires no changes to the calculus,
 based on preprocessing.
@@
 Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
-elimination of Leibniz equality, primitive instantiation, and treatment of choice
-operator \cite{pa-01-classical-ty-thy}. We have also designed new rules that use
-higher-order unification to resolve Boolean formulas that are hoisted to literal
-level, delay clausification of nonatomic literals, reason about formulas under
-$\lambda$-binders, and many others. Even though the rules we use are inspired by
+elimination of Leibniz equality, primitive instantiation, and treatment of the choice
+operator \cite{pa-01-classical-ty-thy}. We have also designed new rules including those that use
+higher-order unification to resolve Boolean formulas that are hoisted to the literal
+level, delay clausification of nonatomic literals, and reason about formulas under
+$\lambda$-binders. Even though the rules we use are inspired by
 the ones of refutationally complete higher-order provers, we do not guarantee
@@
 axiomatization of the theory of Booleans. Second, inspired by work of
-Kotelnikov et al.\ \cite{kotelnikov-16-fool}, we implemented the preprocessing approach that does not
+Kotelnikov et al.\ \cite{kotelnikov-16-fool}, we implement the preprocessing approach that does not
 require introduction of Boolean axioms.
@@
 
-Our approach is implemented in Zipperposition
-\cite{sc-15-simon-phd,sc-supind-17}. This theorem prover was used by Bentkamp et al.\
-to implement \lsup{}. We further extend
-their implementation.
+The theorem prover Zipperposition \cite{sc-15-simon-phd,sc-supind-17} was used
+by Bentkamp et al.\ to implement \lsup{}. We further extend their implementation
+based on the approach presented in this chapter.
 
@@
 In addition to evaluating different configurations of our new rules, we have
-compared them to full higher-order provers CVC4, Leo-III,  Satallax and Vampire.
+compared them to full higher-order provers CVC4, Leo-III,  Satallax, and Vampire.
 The results suggest that it is beneficial to natively support Boolean reasoning---the approach outperforms preprocessing-based approaches. Furthermore, it is
-very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
-approaches we base on, as well as related approaches (Sect.~
+very competitive in comparison to state-of-the-art higher-order provers. We discuss the differences between our approach and the
+approaches our work is based on, as well as related approaches (Sect.~
 \ref{sect:bool:discussion}).
@@
 extend the syntax of this logic by adding logical connectives to the signature.
-The semantic of the logic is extended by interpreting Boolean type $o$ as a
+The semantics of the logic is extended by interpreting Boolean type $o$ as a
 two-element domain. This amounts to extending the logic to full higher-order
@@
 description of this polymorphic logic. For the notions that are not defined here,
-we assume the definitions introduced in Sect.~\ref{ch:pre}.
+we assume the definitions introduced in Chapter~\ref{ch:pre}.
 
@@
 $\VV$ are sets of constants and term variables, respectively. We require nullary
-type constructor $o$, as well as binary constructor $\rightarrow$
+type constructor $o$ as well as binary constructor $\rightarrow$
 to be in $\Sigmaty$. A type $\tau, \upsilon$ is either a type variable $\alpha \in
@@
 \tuplen{\alpha} \mapsto \tuplen{\upsilon} \}$, called constant. If $x$ is a bound variable
-of type $\tau$ and $s$ is a term of type $\upsilon$ then $\lamx{s}$ is a term of type
+of type $\tau$ and $s$ is a term of type $\upsilon$, then $\lamx{s}$ is a term of type
 $\tau \rightarrow \upsilon$. If $s$ and $t$ are of type $\tau \rightarrow \upsilon$ and
@@
 Unless stated otherwise, we view terms as
-$\alpha\beta\eta$-equivalence classes, with $\eta$-long $\beta$-reduced form as
+$\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-reduced form as
 the representative.
@@
 \looseness=-1
-Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
+Given a formula $f$, we call its Boolean subterm $f|_p$ a \emph{top-level
 Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
@@
 Booleans. We continue by describing the rules that we have
-implemented. For ease of presentation we divide them in three categories.
+implemented. For ease of presentation we divide them into three categories.
 \subsection{Support for Booleans in Zipperposition}
@@
 easy extension of its base superposition calculus to various theories, including
-arithmetic, induction and limited support for higher-order logic
+arithmetic, induction, and limited support for higher-order logic
 \cite{sc-15-simon-phd,sc-supind-17}.
@@
 nested Booleans. They call including the axiom $\iforall
-p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
-explosive behavior of the axiom, they imposed the following two requirements to
+p. \; p \eq \itrue \ior p \ieq \ifalse$  ``a recipe for disaster''. To combat the
+explosive behavior of the axiom, they impose the following two requirements to
 the simplification order $\succ$: $\itrue \succ \ifalse$ and
@@
 these requirements are met, there is no self-paramodulation of the clause
-and only paramodulation possible is from literal $p \eq \itrue$ of the mentioned axiom
+and the only paramodulation possible is from the literal $p \eq \itrue$ of the mentioned axiom
 into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
@@
 $\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
-Booleans  to first-order logic without interpreted Booleans; thus, the symbols $\itrue$ and $\ifalse$, and the type $o$
+Booleans  to first-order logic without interpreted Booleans; thus, the symbols $\itrue$ and $\ifalse$ and the type $o$
 correspond to proxy symbols and types introduced during the translation.
@@
 %
+Let $s \eqneq t$ denote $\eq$ or $\not\eq$.
 We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
 where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
-is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
+is not a logical symbol. The rule \infname{IC}
 is applicable if clause $C = L_1 \llor
@@
 the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
-\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
+\cdots \ior L_n$ where $\tuple{x}{}$ consists of all free variables of $C$.
 Using Boolean extensionality, Zipperposition's clausification
@@
 To combat these issues we
-implemented pragmatic restrictions of the rule: only $f$ which is
+implemented pragmatic restrictions of the rule: only the $f$ which is
 the leftmost outermost (or innermost) eligible subterm will be considered. With
@@
 generates imitations (for $P$) of all elements of $S$. Mode $\emph{Pragmatic}$
-generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
-such that $i\not= j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
+generates imitations of $\inot$, $\itrue$, and $\ifalse$; if there exist indices $i,j$
+such that $i\not= j$ and  $\tau_i = \tau_j$, then it generates $\lam{\tuplen{x}}{x_i
 \ieq x_j}$; if there exist indices $i,j$ such that $i \not= j$, and $\tau_i =
@@
 generates $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ generates
-imitations of $\itrue$, $\ifalse$ and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ which generates imitations
+imitations of $\itrue$, $\ifalse$, and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ generates imitations
 of both $\iforall$ and $\iexists$.
@@
 P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
-added to set of choice functions $\mathit{CF}$ (which initially contains just
+added to the set of choice functions $\mathit{CF}$ (which initially contains just
 $\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
@@
 Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
-arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
+an arbitrary context around $t$, the chosen instantiation for $P$ from the axiom of choice;
 the variable $X$ can later be replaced by imitations of logical symbols to create more
@@
 Zipperposition options, $D$ will either be deleted or kept. Note that $D$
-will not subsume its instances, since the matching algorithm Zipperposition uses favors performance over completeness and is
+will not subsume its instances, as the matching algorithm Zipperposition uses favors performance over completeness and is
 thus too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
@@
 \noindent\looseness=-1
-In each of the rules, $\sigma$ is a most general unifier of the types of $s$ and
-$s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
+In each of the rules, $\sigma$ is an MGU of the types of $s$ and
+$s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,\allowbreak
 (s_n, s'_n)\}$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
-extensional versions of superposition, equality resolution and equality
+extensional versions of superposition, equality resolution, and equality
 factoring (Sect.~\ref{sec:pre:rules}). All side conditions for extensional rules are the same as for
-the standard rules, except that condition that $s$ and $s'$ are unifiable is
-replaced by the condition that at least one $s_i$ is of functional type and that
-$n>0$. By \infname{Abs} we denote the union of
+the standard rules, except that the condition that $s$ and $s'$ are unifiable is
+replaced by the condition that $n>0$ and at least one $s_i$ is of functional type. By \infname{Abs} we denote the union of
 these three rules. We extend \infname{Abs} to support Boolean extensionality by
@@
 definition is also applicable, the newly added one is preferred. In the rest of the
-text with \infname{Abs} we assume the version that supports Boolean extensionality.
+text \infname{Abs} refers to the version that supports Boolean extensionality.
 
@@
 the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
-equal. Without the \infname{Abs} rules Zipperposition relies on
+equal. Without the \infname{Abs} rules Zipperposition relies on the
 \infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
@@
   Zipperposition restricts \infname{AbsSup} by
-  requiring that $s$ and $s'$ are not of function or Boolean types. If the terms are of function type, our experience is
-  that better treatment of function extensionality is to apply fresh free
+  requiring that $s$ and $s'$ are not of function or Boolean type. If the terms are of function type, our experience is
+  that a better treatment of function extensionality is to apply fresh free
   variables (or Skolem terms, depending on the sign
@@
   algorithm described in Sect.~\ref{sec:unif:the-unification-procedure}, and
-  remove the premise after conclusion of \infname{ER} is computed.
+  remove the premise after the conclusion of \infname{ER} has been computed.
 \end{sloppypar}
@@
 %
+Intuitively, it embodies the principle that if two terms are indistinguishable by propositions,
+they must be equal.
 Leibniz equality often appears in TPTP problems  \cite{gs-17-tptp}. Since modern provers have native support
@@
 % \pagebreak[2]
-where $P$ is a free variable, $P$ does not occur in any $s_k^l$ or $t_k^l$, or
+where $P$ is a free variable, that does not occur in any $s_k^l$ or $t_k^l$, or
 in $C$; $\sigma$ is defined as $\{ P \mapsto \lam{\tuplen{x}}{\pmb{\bigvee}_{k=1}^{j}
@@
 \looseness=-1
-where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma = \{ P
+where $P$ is a free variable that does not occur in $t_i$, $\sigma = \{ P
 \mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$, and $\sigma' = \{ P \mapsto
 \lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. \infname{ElimLeibniz} directly applies
-Leibniz equality: When the premise of the Leibniz equality is clausified it
+Leibniz equality: When the premise of Leibniz equality is clausified it
 becomes $P \, X \not\eq \itrue \llor P \, Y \eq \itrue$. The rule then tries to find
-a similar subclause in a clause and then instantiate the clause with a
+a similar two-literal subclause in a clause $C$, and if successful instantiates $C$ with a
 substitution that asserts equality of $X$ and $Y$.
@@
 might improve performance  by generating fewer intermediate clauses.  Coming
-back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor
+back to the example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor
 P \, \cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2
@@
 two terms $\Diamond \, \tuplen{s}$ and $\Diamond \, \tuplen{t}$, where neither
-of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
+of $\tuplen{s}$ nor $\tuplen{t}$ contains duplicates, it performs the
 following steps: First, it removes all terms that appear in both
@@
 with $\cst{q} \, (\cst{f} \, \cst{a})$, and fail to find a unifier. However, by
-removing term  $\cst{q} \, (\cst{f} \, \cst{a})$ from the argument lists, we
+removing the term  $\cst{q} \, (\cst{f} \, \cst{a})$ from the argument lists, we
 will be left with the problem $\cst{p} \, \cst{a} \unif  R \, (\cst{f} \, (\cst{f}
@@
 form for application of the corresponding base rules. In particular,
-for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
+for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$, which
 allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
@@
 The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
-the \infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.
+the \infname{Bool} family of rules helps solve problems that get obfuscated by clausification.
 
@@
 them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
-\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
+\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ consists of all the type and term variables occurring
 freely in $\iexists x.\, g$. The rules described above are as given by Ganzinger and Stuber (adapted to
@@
 occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ replacing them, where
-$\tuplen{X}$ are all free variables of $f$. We keep the count of all nonatomic
+$\tuplen{X}$ consists of all free variables of $f$. We keep the count of all nonatomic
 formulas occurring as either side of a literal. Before applying the \infname{DC}
@@
 $\sigmaterm{g} = f$, for some substitution $\sigma$. This check can have three
-outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ is already
+outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ was already
 introduced for $g$ with the same polarity as $f \, \dot{\eq} \,
@@
 but we need to rely on extensionality reasoning to deal with $\lambda$-abstractions
-whose body has type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
+whose bodies have type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
 implies that $\lam{\tuplen{x}}{f}$ is functional-extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
@@
 
-During most of 2010s, the core calculus of the best performing
-higher-order prover at CASC was tableaux---a striking contrast from first-order part
+During most of the 2010s, the core calculus of the best performing
+higher-order prover at CASC was tableaux---a striking contrast from the first-order part
 of the competition dominated by superposition-based provers. TPTP
@@
 clausification this problem becomes $ \cst{sk}_1 \, (\cst{sk}_2 \, Y) \eq Y   $
-where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, we can use
+where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, we can use the
 \infname{ArgCong} rule \cite{bbtvw-21-sup-lam} which applies
@@
 \eq Y \, W$. Superposition-based higher-order theorem
-provers (such as Leo-III, Vampire and Zipperposition) will split this clause
+provers (such as Leo-III, Vampire, and Zipperposition) will split this clause
 into two clauses $C_1 = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \not\eq \top \llor Y
@@
 if the original clause is split into two as described above, Zipperposition will
-rely on \infname{PI} rule to instantiate $Y$ with imitation of $\inot$ and on
+rely on the \infname{PI} rule to instantiate $Y$ with an imitation of $\inot$ and on
 equality factoring to further instantiate this approximation. These desired
 inferences need to be applied on both new clauses and represent only a fraction
-of inferences that can be done with $C_1$ and $C_2$, reducing the chance of
+of inferences that can be done with $C_1$ and $C_2$, reducing the chance of a
 successful proof attempt. Rule \infname{BoolER} imitates the behavior of
 a tableaux prover: It essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
-(\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $ which makes finding the necessary substitution easy and
+(\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $, which makes finding the necessary substitution easy and
 does not require a clausification step.
@@
 \looseness=-1
-Combining rule \infname{(Bool)ER} with dynamic clausification is very fruitful as
+Combining the rule \infname{(Bool)ER} with dynamic clausification is very fruitful, as
 the benchmark \texttt{SYO033\^{}1} illustrates. This problem contains the
@@
 clausification and \infname{BoolER}, Zipperposition times out in any reasonable
-time limit; with those two options it solves mentioned problems in less than
+time limit; with those two options it solves the mentioned problems in less than
 100\,ms.
@@
 In this problem, all functions $\cst{f}_i$ are of type $o \rightarrow o$, and constant $\cst{a}$ is of type $o$.
-FOOL unfolding of nested Boolean terms will result in exponential blowup in the problem size. However,
+FOOL unfolding of nested Boolean terms will result in an exponential blowup in the problem size. However,
 superposition-based theorem provers are well equipped for this issue: Their clausification algorithms use smart simplifications and
@@
 solve the problem swiftly, using their default modes. However, if the problem is not preprocessed, Zipperposition struggles to prove it using
-\infname{Cases(Simp)} and due to the large number of (redundant) clauses it creates, succeeds only if specific heuristic choices are made.
+\infname{Cases(Simp)}, and due to the large number of (redundant) clauses it creates, succeeds only if specific heuristic choices are made.
 
@@
 To evaluate the usefulness of
-all rules we described above, we enable, disable, or change the
+all rules described above, we enabled, disabled, or changed the
 parameters of a single rule, while keeping all other parameters of b intact. 
-In figures that contain sufficiently different configurations, cells are of the form $n (m)$ where $n$ is the total
+In figures that contain sufficiently many different configurations, cells are of the form $n (m)$ where $n$ is the total
 number of proved problems by a particular configuration and $m$ is the number of
@@
 configurations in the same figure. Intersections of rows and columns
-denote corresponding combination of parameters.
-Result for the base configuration is written in \emph{italics}; the best result
+denote the corresponding combination of parameters.
+The result for the base configuration is written in \emph{italics}; the best result
 is written in \textbf{bold}.
 
-First, we tested different parameters of \infname{Cases} and \infname{CasesSimp}
+First, we tested different parameters of the \infname{Cases} and \infname{CasesSimp}
 rules. In Figure \ref{fig:cases} we report the results. The columns correspond to
-three possible options to choose subterm on which the inference is performed:
-{a} stands for any eligible subterm, {lo} and {li} stands
+three possible options to choose the subterm on which the inference is performed:
+{a} stands for any eligible subterm, {lo} and {li} stand
 for leftmost outermost and leftmost innermost subterms, respectively. The rows
@@
     \end{tabular}}
-    \caption{Effect of the \infname{Cases}(\infname{Simp}) rule on the success rate }
+    \caption{Effect of the \infname{Cases}(\infname{Simp}) rule on success rate }
     \label{fig:cases}
@@
 \looseness=-1
-Second, we evaluated all the modes of \infname{PI} rule with 3 values for
+Second, we evaluated all the modes of the \infname{PI} rule with three values for
 parameter $k$: 1, 2, and~8 (Figure \ref{fig:pi}). The columns denote, from left to
@@
 different values of $k$. The results show that different values for $k$ have a
-modest effect on success rate. The raw data reveal that when we focus our
-attention to configurations with $k=2$, mode \emph{Full} can solve 10 problems
+modest effect on the success rate. The raw data reveal that when we focus our
+attention to configurations with $k=2$, mode \emph{Full} can solve ten problems
 no other mode (including disabling the \infname{PI} rule) can. Modes
-\emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve 2  problems, whereas
+\emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve two  problems whereas
 \emph{Imit}$_{\ior}$ solves one problem uniquely. This result suggests that, even
 though this is not evident from Figure \ref{fig:pi}, the sets of problems solved 
-solved by different modes somewhat differ.
+by different modes somewhat differ.
 
@@
 Leibniz equality on the calculus level: EL stands for \infname{ElimLeibniz},
-whereas EPV denotes \infname{ElimPredVar}; signs $-$ and $+$ denote that rule is
+whereas EPV denotes \infname{ElimPredVar}; signs $-$ and $+$ denote that the corresponding rule is
 removed from or added to configuration b, respectively. Disabling both rules
@@
     \end{tabular}}
-    \caption{Effect of \infname{PI} rule on the success rate}
+    \caption{Effect of the \infname{PI} rule on success rate}
     \label{fig:pi}
@@
 
-Similarly, Figure \ref{fig:bool-rules} shows the merits of including ($+$) or
-excluding ($-$) \infname{BoolER} (BER) and \infname{BoolEF} (BEF) rules. Our
-expectations were that inclusion of those two rules would make a larger impact on
-success rate. It turns out that, in practice, most of the effects of these
-rules could be achieved using a combination of the \infname{PI} rule and
+Similarly, Figure \ref{fig:bool-rules} shows the merits of excluding ($-$) or including ($+$)
+\infname{BoolER} (BER) and \infname{BoolEF} (BEF) rules. Our
+expectations were that inclusion of those two rules would have a significant impact on
+the success rate. It turns out that, in practice, most of the effects of these
+rules could be achieved using a combination of the \infname{PI} rule and the
 superposition rules.
@@
 conclusion $D$ of \infname{Choice}, after it is simplified, makes no difference.
-Counterintuitively, disabling \infname{BoolSimp} rule results in 1640 problems,
-which is only 6 problems short of configuration b. Disabling \infname{Abs} and
+Counterintuitively, disabling the \infname{BoolSimp} rule results in 1640 problems,
+which is only 6 problems short of configuration b. Disabling the \infname{Abs} and
 \infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
@@
 
-Last, we compare our approach to alternatives. Axiomatizing Booleans brings
+Last, we compared our approach to alternatives. Axiomatizing Booleans brings
 Zipperposition down to a grinding halt: only 1106 problems can be solved using
-this mode. On the other hand, preprocessing is fairly competitive: it solves
+this mode. On the other hand, preprocessing is fairly competitive: it solved
 only 8 problems less than the b configuration.
@@
 We compared Zipperposition with all higher-order theorem provers that took part
-in higher-order division of 2019 edition of CASC \cite{gs-19-casc27}: CVC4 1.8 prerelease \cite{cbetal-11-cvc4},
+in higher-order division of the 2019 edition of CASC \cite{gs-19-casc27}: CVC4 1.8 prerelease \cite{cbetal-11-cvc4},
 Leo-III 1.4 \cite{sb-21-leo3}, Satallax 3.4
 \cite{cb-12-satallax}, and Vampire-THF 4.4
-\cite{lkav-13-vampire}. In this part of the evaluation, Zipperposition is ran
-in portfolio mode that runs configurations in different time slices. 
-We set the CPU time limit to 180 s, the time allotted to each prover at 2019 edition of CASC.
+\cite{lkav-13-vampire}. In this part of the evaluation, Zipperposition
+used the portfolio mode that runs configurations in different time slices. 
+We set the CPU time limit to 180 s, the time allotted to each prover at the 2019 edition of CASC.
 
 Leo-III and Satallax are cooperative theorem provers---they periodically invoke
-first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
+first-order provers to finish the proof attempt. Leo-III uses CVC4, E, and
 iProver \cite{kk-08-iprover} as backends, and Satallax uses Ehoh
 % \cite{EHOH CHAPTER}
- as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we run
+ as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we ran
 the cooperative provers in two versions: \emph{uncoop}, which disables backends,
-and \emph{coop} which uses all supported backends.
+and \emph{coop}, which uses all supported backends.
 
@@
 in close second, showing that our approach is a promising basis for further
-extensions. Indeed, as we show in Chapter \ref{ch:ho-techniques}, with
+extensions. Indeed, as will be shown in Chapter \ref{ch:ho-techniques}, with
 fine-tuning of the heuristics and further extensions to the calculus, this
@@
 higher-order ``hammer'' or software verifier frontends and first-order backends.
-Considerable amount of research effort has gone into making the translations of
+A considerable amount of research effort has gone into making the translations of
 higher-order logic as efficient as possible. Descriptions of hammers like
@@
 \looseness=-1
-Established higher-order provers like Leo-III and Satallax perform very
+Established higher-order provers like Leo-III and Satallax
 have been optimized to perform well on TPTP; however, recent evaluations (including the one in Sect.~\ref{sec:ehoh:evaluation}) show that on Sledgehammer
@@
 higher-order provers---treatment of exclusively higher-order issues such as
-extensionality or choice is built into them usually using explosive rules. Those
+extensionality or choice is built into them often using explosive rules. Those
 explosive rules might contribute to their suboptimal performance on mostly
@@
 In contrast, the approach taken in this thesis is to start with a first-order prover and gradually extend it
-with higher-order features. The work performed in the context of Matryoshka project~\cite{matryoshka}, 
-in which I participate, resulted in adding support for
+with higher-order features. The work performed in the context of the Matryoshka project~\cite{matryoshka}, 
+in which I participated, resulted in adding support for
 $\lambda$-free higher-order logic with Booleans to E (Chapter~\ref{ch:ehoh})
 and veriT \cite{cbetal-11-cvc4}, and adding support for Boolean-free higher-order
-logic to Zipperposition. Authors of many all state-of-the-art
+logic to Zipperposition. Many authors of state-of-the-art
 first-order provers have implemented some form of support for higher-order reasoning. This
@@
 before in a modular way. It is based on \lsup{}, a calculus that generalizes the highly successful first-order superposition calculus.
-that incurs around 1\% of overhead on first-order problems compared with classic
+It incurs around 1\% of overhead on first-order problems compared with classic
 superposition \cite{bbtvw-21-sup-lam}.
@@
 previous research efforts that extended first-order provers with complementary
-features of higher-order logic. It also proposes some solutions for the issues that
-emerge with this combination. The implementation shows clear improvement over previous
-techniques and competitive performance.
+features of higher-order logic. It also proposes some solutions for issues that
+emerge with this combination. The implementation shows a clear improvement over previous
+techniques and a competitive performance.
 

### 2022-03-14 ehoh/ehoh.tex
past tense

@@
 \ourpara{Term Orders} Superposition provers rely on term orders to prune the search
-space. The order must be a simplification order that
+space. The order must be a reduction order
 %can be extended to a simplification order that
@@
 % simplification order and the prover may underapproximate it. -JB
-is total on variable-free
-terms. E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
+(Sect.~\ref{sec:pre:order}). E implements both the Knuth--Bendix order (KBO) and the lexicographic path order
 (LPO). KBO is widely regarded as the more
@@
 higher-order logic, the second clause set would be unsatisfiable thanks to
-the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-conversion.)
+the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-reduction.)
 %

### 2022-03-14 introduction/introduction.tex
past tense

@@
 
+\looseness=-1
 These attempts resulted in early higher-order theorem provers. Andrews and
-colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
-allows users to specify proof outlines, based on expansion proofs. Benzm\"uller
-and colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
-resolution which introduced the cooperative paradigm. Provers implementing this paradigm
-invoke first-order backends in regular intervals to finish the proof. The latest
-iteration in the LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
-\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
-to close tableau branches. It won the higher-order division of CASC on eight occasions:
-in 2011 and every year from 2013 to 2019.
+colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which allows
+users to specify proof outlines, based on expansion proofs. Benzm\"uller and
+colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
+resolution which introduced the cooperative paradigm. Provers implementing this
+paradigm invoke first-order backends in regular intervals to finish the proof.
+The latest iteration in the LEO family of provers is Leo-III \cite{sb-21-leo3}.
+Satallax \cite{cb-12-satallax} is based on higher-order analytic tableaux. Such
+representations of logical formulas are tree structures with at each node a
+subformula of the original formula to be proved or refuted. Satallax improves on
+the method by using a SAT solver to close tableau branches.
+It won the higher-order division of CASC on eight occasions: in 2011 and every
+year from 2013 to 2019.
 

### 2022-03-14 pre/pre.tex
past tense

@@
 \looseness=-1
-With $\alpha$-renaming we assume the rule $(\lambda x. \, s) \longrightarrow \lambda
-y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$ and 
-$y$ is not bound in $s$. Application of substitution $\sigma$ to term
-$t$, implicitly $\alpha$-renames bound variables to avoid capture.  For example,
-$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
-$\beta$-reduction rule corresponds to applying arguments to function: $(\lambda
-x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where the bound variables in $s$
-are renamed to avoid capture. Lastly, $\eta$-reduction is defined as $\lambda
-x.\, s \, x \longrightarrow s$ under the condition that $x$ is not loosely
-bound in $x$. We write $s \equi t$ if terms $s$ and $t$ are equal modulo $\alpha\beta\eta$
-converison.
+With $\alpha$-conversion we assume the rule $(\lambda x. \, s) \longrightarrow
+\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$
+and $y$ is not bound in $s$. Application of a substitution to a term,
+implicitly $\alpha$-renames bound variables to avoid capture.  For example, $\{X
+\mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
+$\beta$-reduction rule, roughly speaking, corresponds to passing arguments in a
+function call: $(\lambda x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where
+the bound variables in $s$ are renamed to avoid capture. Lastly,
+$\eta$-reduction is defined as $\lambda x.\, s \, x \longrightarrow s$ under the
+condition that $x$ is not loosely bound in $x$. We write $s \equi t$ if terms
+$s$ and $t$ are equal modulo $\alpha\beta\eta$-conversion. Terms that are
+convertible with respect to any of the rules form equivalence classes. Thus, we
+also say that $\alpha$($\beta\eta$)-convertible terms are
+$\alpha$($\beta\eta$)-equivalent.
 
@@
 formula-building logical connectives are written in boldface.
-
 In standard, non-clausal first-order logic, an atom is either
@@
 heuristics to prune the search space such as term order and selection functions.
+
+This calculus works on problems in clausified form and proves the problem by
+refuting its negation. In practice this means that, to apply superposition to a
+given problem, we first must negate its conjecture and clausify the whole
+problem using some of the clausification algorithms
+\cite{nw-01-small-cnf,rsv-16-vcnf}. The calculus then applies its inference
+rules to the clauses, adding results of the inferences to the working set of
+clauses. As this process is quite prolific it also uses the term order to determine
+if some clauses can be simplified or eliminated from the working set.
+Superposition is refutationally complete: This means that, if results of all
+inferences are computed in a systematic, fair manner, empty clause ($\bot$) will
+eventually be derived for a provable problem. As first-order logic is
+undecidable, no guarantees are given for unprovable problems.
 Before introducing the rules of the calculus, we provide definitions of these and other
@@
 In higher-order logic, unification is performed modulo rules of
-$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
+$\alpha\beta\eta$-conversion. Under these conditions, the uniqueness of the most
 general unifier is not guaranteed. Consider the constraint $X \,
@@
 
-Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what mathematicians do when
-they ``replace equals by equals''. The left premise intuitively states that an
-equation $s \eq t$ holds under condition $C$ (or more precisely, its negation).
-Similarly, the right premise asserts that a (dis)equation holds under a
-condition. The superposition rule simply concatenates the conditions for both
-premises and replaces equals by equals in two main inference terms. For example
-given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and $\neg
-\cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition between their
-last literals results in $\neg
-\cst{even}(X) \llor \neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq
-\cst{b}$.
+Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
+mathematicians do when they ``replace equals by equals''. Name paramodulation is
+also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
+states that an equation $s \eq t$ holds under condition $C$ (or more precisely,
+its negation). Similarly, the right premise asserts that a (dis)equation holds
+under a condition. The superposition rule simply concatenates the conditions for
+both premises and replaces equals by equals in two main inference terms. For
+example given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and
+$\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
+between their last literals results in $\neg \cst{even}(X) \llor \neg
+\cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.
 

### 2022-03-14 ho-techniques/ho-techniques.tex
fix

@@
 Both sets of rules are disabled or replaced with incomplete, ad hoc rules
-described in the previus chapter in most configurations
+described in the previous chapter in most configurations
 of the CASC portfolio.
@@
 closely behind Satallax 3.4 %~and~3.5
-at higher-order division of CASC in 2020.
+at the higher-order division of CASC in 2020.
 
@@
 incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
-well on a wide range of problems.
+well on a wide range of problems. The set of parameters used for baseline configuration in this
+chapter differs from the one in the previous chapter.
 %Note that
@@
 The \osup{} calculus supports \emph{delayed clausification rules} that insert
-problems into the proof state in their original, nonclausified form, and
+the formulas of a problem into the proof state in their original, nonclausified form, and
 clausify them gradually. Delayed clausification allows the prover to analyze the
@@
 set of heuristically chosen terms. This set contains $\lambda$-abstractions
-whose bodies are formulas and that occur in activated clauses; it also containts
+whose bodies are formulas and that occur in activated clauses; it also contains
 \emph{primitive instantiations}---that is, imitations (in the sense of
@@
 quantified variable of $t$'s type has been activated, we remember the clauses
-$\sigmaterm{\{x \mapsto X\}}{\varphi} \eq \itrue\allowbreak \llor C$ and instantiate them when
+$\substterm{\{x \mapsto X\}}{\varphi} \eq \itrue\allowbreak \llor C$ and instantiate them when
 $\mathit{Inst}$ is extended. Conveniently, these instantiated clauses are not
@@
 One of the problems Boolean instantiation helps solve is \texttt{NUM636\^{}2} (a
-re-encoding of \texttt{NUM636\^{}3}, which is disucssed in Example \ref{hot:exa:num636}).
+reencoding of \texttt{NUM636\^{}3}, which is discussed in Example \ref{hot:exa:num636}).
 It conjectures that $\iforall x.\, \cst{s} \, x \ineq x$, where $x$ ranges over

### 2022-03-14 ho-techniques/ho-techniques.tex
more

@@
 contain predicate variables. Therefore, \relax{BI} applies to fewer clauses.
+
+\section{Exploring Boolean Selection Functions}
+\label{sec:ho-tech:bool-select}
+
+As discussed in Sect.~\ref{sec:pre:sup}, superposition calculi are parameterized by a literal selection function and a
+term order that help prune considerable swaths of the search space without
+jeopardizing completeness. The core inferences apply only to a clause's
+\emph{eligible} literals, defined as either the clause's selected
+literals or, if none are selected, the clause's literals that are maximal with
+respect to the term order. To further restrict which terms can be targeted by
+an inference, the \osup{} calculus introduces \emph{Boolean selection
+functions}.
+
+A Boolean selection function chooses \emph{green subterms} of Boolean type
+(different than $\itrue$ or $\ifalse$ and not occurring at either side of a positive
+literal) in a clause and gives rise to a notion of eligibility that considers
+the formula structure. Green subterms correspond to the first-order skeleton of
+a higher-order term; that is, they do not occur in positions under applied
+variables, quantifiers, or $\lambda$-abstractions.
+
+%In particular, green
+%subterms are subterms arising inside first-order-like contexts, and for first
+%other problems, all subterms are green.
+
+\begin{defi}[Green subterms and green positions]
+  \,Green subterms and green positions are defined inductively as
+  follows: $t$ is a green subterm of $t$ at green position $\varepsilon$; if $t$ is a green subterm of $u_i$ at green position $p$
+  and $\cst{f}$ is a constant different from $\iforall$ and $\iexists$, then
+  $t$ is a green subterm of $\cst{f} \, \tuplen{u}$ at green position $i.p$,
+  assuming $i \leq n$.
+\end{defi}
+
+\begin{exa} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
+(\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
+\, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
+\end{exa} 
+%
+Green positions are lifted to clauses as follows: If $p$ is the green position
+of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
+t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
+\osup{} mandates additional restrictions on the Boolean selection function:
+$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
+$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
+variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
+maximal term of the clause, $s$ cannot be selected.
+
+\begin{defi}[Eligibility]
+  \,Given a substitution $\sigma$ and term order $\succ$, we say a literal $l$
+  is (strictly) eligible with respect to $\sigma$ in $C$ if it is selected in
+  $C$ or there are no selected literals and no selected Boolean subterms in
+  $C$ and $\sigmaterm{l}$ is (strictly) maximal in $\sigmacl{C}$ with respect to the
+  term order.
+%
+  The eligible subterms of a clause $C$ with respect to a substitution
+  $\sigma$ are inductively defined as follows:
+  Any subterm selected by the Boolean selection function is eligible.
+  %Any selected subterm is eligible.
+  For a strictly eligible literal $s \eq t$ with $\sigmaterm{t} \not\succ
+  \sigmaterm{s}$, $s$ is eligible. For an eligible literal $s \not\eq t$ with
+  $\sigmaterm{t} \not\succ \sigmaterm{s}$, $s$ is eligible. If a subterm $t$ is eligible
+  and the head of $t$ is not $\ieq$ or $\ineq$, all direct green subterms of
+  $t$ are eligible. If a subterm $t$ is eligible and $t$ is of the form $u
+  \ieq v$ or $u \ineq v$, then $u$ is eligible if $\sigmaterm{v} \not\succ \sigmaterm{u}$
+  and $v$ is eligible if $\sigmaterm{u} \not\succ \sigmaterm{v}$.
+\end{defi}
+The above definitions of green subterms and eligibility were originally introduced
+with \osup{} \cite{bbtv-21-full-ho-sup}.
+The Boolean selection function plays a similar role as the literal
+selection function in standard superposition.
+Literal selection functions eliminate some of the nondeterminism present in the superposition
+calculus by focusing on selected parts of the search space. Boolean selection functions achieve
+the same goal, but in a different context: They eliminate nondeterminism that is not
+present in standard superposition, namely, the choice of subformula on which 
+the Boolean calculus rules are to be applied. As with literal selection functions,
+selecting few (and smaller) subterms can give rise to fewer possible inferences
+and reduce clause proliferation.
+
+%
+This notion of eligibility opens up possibilities for reasoning with
+formulas that are hard to simulate with the existing superposition machinery.
+For example, given a formula $\varphi \iimplies \psi$, selecting the antecedent
+simulates forward reasoning, whereas selecting the consequent simulates backward
+reasoning. The new eligibility also makes it possible to restrict the proof
+search to a small, promising part of a formula. Note that
+literal selection can override Boolean selection: Selecting a literal 
+might make some of its green subterms eligible, regardless
+of Boolean selection.
+
+In our previous work \cite{nbtv-2021-foboolsup}, we left this area of new
+possibilities largely unexplored. We designed simple functions that selected
+smallest, largest, innermost, or outermost terms, but they did not impact
+performance much. Here, we propose alternatives.
+%
+Intuitively, a well-performing literal selection
+function might succeed at taming the combinatorial explosion if the
+selected literal can take part in few inferences
+\cite{hrsv-16-selsel}. However, Boolean selection functions
+introduce another factor to consider:\ the context in which the selected
+subterm occurs. This suggests the following definition:
+
+\begin{defi}[Contextualized Boolean selection function]
+  \label{def:context-bool-sel}
+  \,Let $\mathit{ctx}(C)$ be a function that maps a clause $C$ to a set of green positions
+  $p$ such that $C|_p$ is a selectable Boolean subterm, and let
+  $\vartriangleright$ be a partial order on pairs of terms and green positions.
+  The \emph{context
+  Boolean selection function} $\mathit{Sel\/}_{\mathit{ctx}}^{\,\vartriangleright}(C)$
+  selects all terms $t$ such that $t=C|_p$, $p \in \mathit{ctx}(C)$, and
+  $(t,p)$ is maximal with respect to $\vartriangleright$.
+\end{defi}
+
+In the above definition, the function $\mathit{ctx}$ lets
+us choose the context in which the Boolean subterm appears. Then, among
+the terms in the chosen context, we choose the ones that are maximal with
+respect to $\vartriangleright$.
+
+Ganzinger and Stuber considered Boolean subterm selection for their extension of
+first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus requires
+selection of subterms occurring in negative green positions, defined below.
+
+\begin{defi}[Polarity of green positions]
+ \,Negative and positive green positions in a clause $C = l_1 \llor \cdots \llor
+ l_n$ are defined inductively as follows: For each
+ $1 \leq i \leq n$, the green position $l_i.s$ is positive if $l_i = \poslit{s}$ and negative if $l_i
+ = \neglit{s}$. If $p$ is positive (negative) and $C|_p = s \,
+ \tuplen{t}$ where $s$ is either $\iand$ or $\ior$, then each
+ $p.i, 1 \leq i \leq n$, is positive (negative). If $p$ is positive and $C|_p =
+ \inot \, s$, then $p.1$ is negative; if $p$ is negative and $C|_p = \inot \, s$,
+ then $p.1$ is positive. If $p$ is positive and $C|_p = s \iimplies t$, then
+ $p.1$ is negative and $p.2$ is positive; if $p$ is negative and $C|_p = s
+ \iimplies t$, then $p.1$ is positive and $p.2$ is negative.
+\end{defi}
+Note that the polarity of $p$ is undefined whenever $C|_p$ is not a green Boolean subterm
+or it occurs under a (dis)equivalence or an uninterpreted symbol.
+To assess how the function $\mathit{ctx}$ affects performance, we use the following
+selection functions that consider green positions of selectable Boolean terms:
+%
+% \begin{description}[labelwidth=\widthof{\rm\texttt{Backward}~}]
+%   \item[\rm\texttt{Any}~] select all green positions;
+%   \item[\rm\texttt{Pos}~] select all positive green positions;
+%   \item[\rm\texttt{Neg}~] select all negative green positions;
+%   \item[\rm\texttt{Forward}~] select all green positions $p = q.1$ such that $C|_q = s \iimplies t$;
+%   \item[\rm\texttt{Backward}~] select all green positions $p = q.2$ such that $C|_q = s \iimplies t$;
+%   \item[\rm\texttt{Deep}~] select all green positions of maximal length;
+%   \item[\rm\texttt{Shallow}~] select all green positions of minimal length.
+% \end{description}
+
+\begin{tabular}{ll}
+\texttt{Any} &  select all green positions; \\
+\texttt{Pos} &  select all positive green positions; \\
+\texttt{Neg} &  select all negative green positions; \\
+\texttt{Forward} &  select all green positions $p = q.1$ such that $C|_q = s \iimplies t$; \\
+\texttt{Backward} &  select all green positions $p = q.2$ such that $C|_q = s \iimplies t$; \\
+\texttt{Deep} &  select all green positions of maximal length; \\
+\texttt{Shallow} &  select all green positions of minimal length.   \\
+\end{tabular}
+
+\smallskip
+\looseness=-1
+We also introduce three partial orders for selecting subterms from a given
+context. For all three orders, if exactly one of the subterms has a logical head,
+then the subterm with the nonlogical head is larger, because logical symbols
+are more explosive. Otherwise, the orders use the following criteria:
+%
+% \begin{description}[labelwidth=\widthof{$\vartriangleright_\text{ground}$~}]
+%   \item[\rm$\vartriangleright_\text{ground}$~] If exactly one of the subterms is ground, make the
+%   ground subterm larger; otherwise, if exactly one of the subterms is of the form $s \ieq
+%   t$, make this subterm larger.
+
+%   \item[\rm$\vartriangleright_\text{depth}$~] If one of the subterms has larger subterm depth
+%   (longest valid green subterm position), make this subterm larger; otherwise, if one of the
+%   subterms has less distinct variables, make this subterm larger.
+
+%   \item[\rm$\vartriangleright_\text{def}$~] If exactly one of the subterms is of the form
+%   $\cst{p}\,\tuplen{X}$ where $\tuplen{X}$ is a tuple of free variables, make
+%   the other subterm larger; otherwise, if exactly one of the subterms is of the form $X \,
+%   \tuplen{s}$, make the other subterm larger.
+% \end{description}
+\smallskip
+\begin{tabular}{lp{0.8\linewidth}}
+  $\vartriangleright_\text{ground}$ & If exactly one of the subterms is ground, make the
+  ground subterm larger; otherwise, if exactly one of the subterms is of the form $s \ieq
+  t$, make this subterm larger. \\[2\jot]
+  $\vartriangleright_\text{depth}$ & If one of the subterms has larger subterm depth
+     (longest valid green subterm position), make this subterm larger; otherwise, if one of the
+     subterms has less distinct variables, make this subterm larger. \\[2\jot]
+  $\vartriangleright_\text{def}$ & If exactly one of the subterms is of the form
+  $\cst{p}\,\tuplen{X}$ where $\tuplen{X}$ is a tuple of free variables, make
+  the other subterm larger; otherwise, if exactly one of the subterms is of the form $X \,
+  \tuplen{s}$, make the other subterm larger.
+\end{tabular}
+\smallskip
+
+In case of a tie, the subterm with the smaller syntactic weight is made larger,
+and if both subterms have the same weight, the term that occurs in a position further
+to the left (i.e., that has a lexicographically smaller position) is made larger.
+
+These orders follow the design principle enunciated by Hoder et
+al.~\cite{hrsv-16-selsel} that ground or deep terms and terms with repeated
+variables are ``less unifiable'' with the similar observation for higher-order
+logic that reasoning about interpreted symbols or applied variables is usually
+explosive.
+
+\begin{exa}
+  Selecting the right Boolean subterm can help avoid elaborating
+  high\-er-order inferences. Consider the unsatisfiable clause set consisting of $\poslit{\cst{p} \, (\lambda y.\, X
+  \, (\lambda x.\, x) \, \cst{a}) \iimplies\allowbreak \inot(\cst{p} \, (\lambda y. \allowbreak\, X
+  \, y \, \cst{a}))}$, $\poslit{\cst{p} \, (\lambda y.\, \cst{a})}$, and
+  $\poslit{\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})}$, where superscript $i$ denotes $i$-fold
+  application of a unary term.
+  Note that $\cst{p} \, (\lambda y.\, X
+  \, (\lambda x.\, x) \, \cst{a})$ and $\cst{p} \,
+  (\lambda y.\, \cst{a})$ have infinitely many unifiers of the form $\{ X \mapsto \lambda fx.
+  \, f^i\, (x) \}, i \geq 0$, whereas terms $\cst{p} \, (\lambda y. \, X
+  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have the MGU. 
+  When \texttt{Forward} context selection is enabled, % the subterm
+  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
+  forcing computation of at least 100 unifiers (under the assumption that
+  unifiers are returned in order of increasing $i$) before we get to refute
+  $\inot(\cst{p} \, (\lambda y. y^{100} \, \cst{b}))$.
+  In contrast, \texttt{Backward} context selection allows us to
+  superpose from $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$ into $\cst{p} \, (\lambda y. \, X
+  \, y \, \cst{a})$, avoiding this explosion. 
+\end{exa}
+
+\ourpara{Evaluation and Discussion}
+
+\begin{figure}
+\centering
+\def\arraystretch{1.1}%
+ \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
+  &                                    & \texttt{Any} & \texttt{Pos} & \texttt{Neg}        & \texttt{Forward} & \texttt{Backward} & \texttt{Deep}     & \texttt{Shallow}  \\ \midrule
+  TPTP &
+    $\vartriangleright_\text{ground}$  & 1538         & 1550         & 1547                & 1534             & {\bf 1554}        & 1539              & 1538     \\
+  & $\vartriangleright_\text{depth}$   & 1542         & 1550         & 1528                & 1542             & 1550              & 1547              & 1535     \\
+  & $\vartriangleright_\text{def}$     & 1543         & 1551         & 1540                & 1540             & 1551              & 1545              & 1537  \\ \midrule
+
+%  \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
+  SH &
+    $\vartriangleright_\text{ground}$  & \colalign386 & \colalign379  & \colalign386       & \colalign386     & \colalign379      & \colalign{\bf387} & \colalign{\bf387}     \\
+  & $\vartriangleright_\text{depth}$   & \colalign377 & \colalign376  & \colalign384       & \colalign378     & \colalign376      & \colalign379      & \colalign376      \\
+  & $\vartriangleright_\text{def}$     & \colalign379 & \colalign374  & \colalign{\bf387}  & \colalign379     & \colalign380      & \colalign377      & \colalign381  \\ \bottomrule
+ \end{tabular}
+ \caption{Impact of the Boolean selection function}
+ \label{fig:bool-sel}
+\end{figure}
+
+When the input problem is clausified using immediate clausification, almost
+all Boolean structure is lost. In this case, we expect Boolean selection to
+have a modest effect. To better assess this feature, in this evaluation we use
+\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:ho-tech:formulas} as the baseline
+configuration. To avoid interference of literal and Boolean selection, we
+additionally forbid the literal selection function from selecting a literal if
+it contains a selectable Boolean subterm.
+
+The results of evaluating 21 concrete selection functions obtained
+by instantiating the contextualized Boolean selection function are shown in
+Fig.~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
+while columns denote the function $\mathit{ctx}$.
+
+\looseness=-1
+On TPTP benchmarks, Boolean selection helps tame the explosion caused by
+dynamic clausification used as inference: \NumberOK{All but one} selection functions
+outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
+\texttt{NUM636\^{}2} from Sect.~\ref{sec:ho-tech:formulas}, using Boolean selection
+can reduce the number of given clause loop iterations from
+\NumberOK{294} to \NumberOK{71}.
+
+The results suggest that selection of term context has more impact
+than the partial term order. % The number of proved problems fluctuates more when
+%changing the function $\mathit{ctx}$ than when changing the partial order.
+Also, the best results are obtained when a context more specific than \texttt{Any}
+is chosen. Remarkably, functions using \texttt{Pos} context
+perform better than the ones using \texttt{Neg} context
+on TPTP but the opposite is observed on SH.
+
+Using different Boolean selection functions yields vastly different sets of
+proved problems on TPTP benchmarks: In total, there are \NumberOK{103} problems
+proved by some configuration from Fig.~\ref{fig:bool-sel} but not by
+\relax{DCI}$-$\relax{LA}. However, there are only \NumberOK{16} such SH problems.
+It would seem that the advanced formula reasoning facilitated by the Boolean
+selection formulas is usually not required by Sledgehammer problems.

### 2022-03-14 scratch/paper.tex
more

@@
 
-\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
+\spnewtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
 \spnewtheorem{exa}[theorem]{Example}{\bfseries}{\rmfamily}
@@
 
-\begin{definitionx}[Green subterms and green positions]
+\begin{definition}[Green subterms and green positions]
   \,Green subterms and green positions are defined inductively as
@@
   assuming $i \leq n$.
-\end{definitionx}
+\end{definition}
 
@@
 
-\begin{definitionx}[Eligibility]
+\begin{definition}[Eligibility]
   \,Given a substitution $\sigma$ and term order $\succ$, we say a literal $l$
@@
   and $v$ is eligible if $u\sigma \not\succ v\sigma$.
-\end{definitionx}
+\end{definition}
 
@@
 
-\begin{definitionx}[Contextualized Boolean selection function]
+\begin{defi}[Contextualized Boolean selection function]
   \label{def:context-bool-sel}
@@
   $(t,p)$ is maximal with respect to $\vartriangleright$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}[Polarity of green positions]
+\begin{defi}[Polarity of green positions]
  \,Negative and positive green positions in a clause $C = l_1 \llor \cdots \llor
@@
  \iimplies t$, then $p.1$ is positive and $p.2$ is negative.
-\end{definitionx}
+\end{defi}
 Note that the polarity of $p$ is undefined whenever $C|_p$ is not a green Boolean subterm

### 2022-03-15 ho-techniques/ho-techniques.tex
unif

@@
 Peano-style numbers specified by $\cst{z}$ and $\cst{s}$. The given axioms are
-the induction principle $\forall p.\, p \, \cst{z} \,\iand\, \forall x. \, (p \,
-x \iimplies p \, (\cst{s} \, x)) \iimplies \forall x. \, p\, x$, injectivity
-$\forall x y. \, \cst{s}\,x \ieq \cst{s}\,y \iimplies x \ieq y$, and
-distinctness $\forall x. \, \cst{s}\,x \ineq \cst{z}$. The conjecture is easily
+the induction principle $\iforall p.\, p \, \cst{z} \,\iand\, \iforall x. \, (p \,
+x \iimplies p \, (\cst{s} \, x)) \iimplies \iforall x. \, p\, x$, injectivity
+$\iforall x y. \, \cst{s}\,x \ieq \cst{s}\,y \iimplies x \ieq y$, and
+distinctness $\iforall x. \, \cst{s}\,x \ineq \cst{z}$. The conjecture is easily
 proved if Boolean instantiation is enabled: Even though the conjecture literal
@@
 smallest, largest, innermost, or outermost terms, but they did not impact
-performance much. Here, we propose alternatives.
+performance much. A similar idea has been discussed in the context of \infname{Cases}(\infname{Simp}) rule
+in Sect.~\ref{sect:bool:native}.  Here, we propose alternatives.
 %
@@
 selection formulas is usually not required by Sledgehammer problems.
+
+\section{Enumerating Infinitely Branching Inferences}
+\label{sec:ho-tech:infinite-branching}
+
+\looseness=-1
+As an optimization and to simplify the code, Leo-III
+\cite{as-18-phd} and Vampire 4.4 \cite{br-19-restricted-unif} (which uses
+\confrep{}{\emph{restricted combinatory unification}, }a predecessor of combinatory
+superposition) compute only a finite subset of the possible conclusions for
+inferences that require enumerating a CSU. Not only is this a source of
+incompleteness, but choosing the cardinality of the computed subset is a
+difficult heuristic choice. Small sets can result in missing the necessary unifier, while large sets make the prover spend too long in
+the unification procedure, generate useless clauses, and possibly get sidetracked into the wrong parts of the search space.
+
+We propose a modification to the given clause procedure to seamlessly
+interleave unifier computation and proof state exploration. Given a complete
+unification procedure, which may yield infinite streams of unifiers, our
+modification fairly enumerates all conclusions of inferences relying on
+elements of a CSU. Under some reasonable assumptions, it behaves exactly like
+the standard given clause procedure on purely first-order problems.
+We also describe heuristics that help achieve a similar
+performance as when using incomplete, terminating unification procedures
+without sacrificing completeness.
+\pagebreak[2]
+
+Given that we cannot decide whether there exists a next CSU
+element in a stream of unifiers, the request for the next conclusion might not
+terminate,
+effectively bringing the theorem prover to a halt. Our modified given clause procedure
+expects the unification procedure to return a lazily computed stream
+\cite[Sect.~4.2]{co-1999-funds}, where each element is either $\emptyset$ or
+a singleton set containing a unifier. To avoid getting stuck waiting for a unifier
+that may not exist, the unification procedure should return
+$\emptyset$ after it performs some number of operations without finding a unifier.
+
+The complete unification procedure described in Chapter \ref{ch:unif} returns such a stream. Other procedures such
+as Huet's \cite{gh-75-unification} and Jensen and
+Pietrzykowski's \cite{jp-76-unif} can easily be adapted to meet this
+requirement. Based on the stream of unifiers interspersed with $\emptyset$, we
+can construct a stream of inferences similarly interspersed with $\emptyset$.
+Any finite prefixes of this stream can be computed in finite time.
+
+To support such streams in the given clause procedure, we extend it to
+represent the proof state not only by the active ($\mathcal{A}$) and passive ($\mathcal{P}$) clause
+sets, but also by a priority queue $\mathcal{Q}$ containing the inference streams,
+similar to the ``to do'' set $T$ present in the abstract Zipperposition loop of 
+Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
+Each stream is associated with a weight, and $\mathcal{Q}$ is sorted in order of
+increasing weight\confrep{}{, a low weight corresponding to a high priority}.
+When they introduced \lsup, Bentkamp et al.\
+\cite{bbtvw-21-sup-lam} described an older version of this extension. Here
+we present a newer version in more detail, including heuristics to postpone
+unpromising streams. The pseudocode of the modified procedure is as follows:
+\newcommand\ParamMode{\ensuremath{K_{\mathrm{fair}}}}
+\newcommand\ParamMaxStreams{\ensuremath{K_{\mathrm{best}}}}
+\newcommand\ParamRetry{\ensuremath{K_{\mathrm{retry}}}}
+\newcommand{\assign}[2]{\State \ensuremath{\mathit{#1} \gets #2}}
+\newcommand{\assignSameLine}[2]{\ensuremath{\mathit{#1} \gets #2}}
+\algrenewcommand\algorithmicindent{1em}
+\begin{algorithmic}[0]
+    \Function{ExtractClause}{$Q$, $\mathit{stream}$}
+      \assign{maybe\_clause}{\text{pop and compute the first element of } \mathit{stream} }
+      \If{$\mathit{stream}$ is not empty}
+        \State add $\mathit{stream}$ to $Q$ with an increased weight
+      \EndIf
+      \State \Return $\mathit{maybe\_clause}$
+    \EndFunction
+
+    \vspace{2\jot}
+
+    \Function{HeuristicProbe}{$Q$}
+      \assign{i}{0}
+      \assign{collected\_clauses}{\emptyset}
+      \While{$i < \ParamMaxStreams$ and $Q \not= \emptyset$}
+        \assign{j}{0}
+        \assign{maybe\_clause}{\emptyset}
+        \While{$j < \ParamRetry$ and $Q \not= \emptyset$ and $\mathit{maybe\_clause} = \emptyset$}
+          \assign{stream}{\text{pop the lowest-weight stream in } Q}
+          \assign{maybe\_clause}{\textsc{ExtractClause}(Q, \mathit{stream})}
+          \assign{j}{j+1}
+        \EndWhile
+        \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \mathit{maybe\_clause}}
+        \assign{i}{i+1}
+      \EndWhile
+      \State \Return $\mathit{collected\_clauses}$
+    \EndFunction
+
+  \newpage
+  \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}
+    \assign{collected\_clauses}{\emptyset}
+    \assign{oldest\_streams}{\text{pop } \mathit{num\_oldest} \text{ oldest streams from } Q}
+    \For{$\mathit{stream}$ in $\mathit{oldest\_streams}$}
+      \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \textsc{ExtractClause}(Q, \mathit{stream})}
+    \EndFor
+    \State \Return $\mathit{collected\_clauses}$
+  \EndFunction
+
+  \vspace{2\jot}
+
+  \Function{ForceProbe}{$Q$}
+    \assign{collected\_clauses}{\emptyset}
+    \While {$Q \not= \emptyset$ and $\mathit{collected\_clauses} = \emptyset$}
+      \assign{collected\_clauses}{\textsc{FairProbe}(Q, |Q|)}
+    \EndWhile
+
+    \If{$Q = \emptyset$ and $\mathit{collected\_clauses} = \emptyset$}
+      \assign{status}{\textsf{Satisfiable}}
+      % \Return (\textsf{Satisfiable}, $\mathit{collected\_clauses}$)
+    \Else{}
+      \assign{status}{\textsf{Unknown}}
+      % \Return (\textsf{Unknown}, $\mathit{collected\_clauses}$)
+    \EndIf
+
+    \State \Return $(\mathit{status}, \mathit{collected\_clauses})$
+  \EndFunction
+
+  \vspace{2\jot}
+
+  \Function{GivenClause}{$P$, $A$, $Q$}
+
+  \assign{i}{0}
+  \assign{status}{\textsf{Unknown}}
+  \While{$\mathit{status} = \textsf{Unknown}$}
+    \If{$P = \emptyset$}
+      \assign{(status, \,forced\_clauses)}{\textsc{ForceProbe}(Q)}
+      \assign{P}{P \mathrel\cup \mathit{forced\_clauses}}
+    \Else
+      \assign{given}{\text{pop a chosen clause from }P \text{ and simplify it}}
+      \If{$\mathit{given}$ is the empty clause}
+        \assign{status}{\textsf{Unsatisfiable}}
+      \Else
+        \assign{A}{A \mathrel\cup \{\mathit{given}\}}
+        \For{$\mathit{stream}$ in streams of inferences between $\mathit{given}$ and $\mathit{other} \in A$}
+          \If{$\mathit{stream}$ is not empty} \assign{P}{P \mathrel\cup \textsc{ExtractClause}($Q$, \mathit{stream})}
+          \EndIf
+        \EndFor
+        \assign{i}{i+1}
+        \If{$i \; \mathrm{mod} \; \ParamMode = 0$} \assign{P}{P \mathrel\cup \textsc{FairProbe}(Q, i/\ParamMode )}
+        \Else{} \assign{P}{P \mathrel\cup \textsc{HeuristicProbe}(Q)}
+        \EndIf
+      \EndIf
+    \EndIf
+  \EndWhile
+
+  \State \Return $\mathit{status}$
+
+  \EndFunction
+\end{algorithmic}
+% \caption{Our modified given clause procedure}
+% \label{fig:gc-pseudocode}
+%
+% \end{figure}
+% \vspace{2.5\jot}
+
+Entry point to the above pseudocode is the function \textsc{GivenClause}, called
+with arguments $(\mathcal{P}, \mathcal{A}, \mathcal{Q})$, which are initialized
+as usual: All input clauses are put into $\mathcal{P}$, and $\mathcal{A}$ and
+$\mathcal{Q}$ are empty. In other words, the parameters $P, A$, and $Q$ intuitively
+correspond to $\mathcal{P}, \mathcal{A}$, and $\mathcal{Q}$. Unlike in the
+standard given clause procedure, inference results are represented as clause
+streams. The first element is inserted into $P$, and the rest of the stream is
+stored in $Q$ with some positive integer weight computed from the inference
+rule.
+
+\looseness=-1
+To eventually consider inference conclusions from streams in $Q$ as given
+clauses, we extract elements from, or \emph{probe}, streams and move any obtained
+clauses to $P$. Analogously to the traditional pick--given ratio
+\cite{ss-02-brainiac,mcw-1997-otter}, we use a parameter
+\ParamMode{} (by default, $\ParamMode = 70$) to ensure fairness: Every \ParamMode{}th iteration,
+\textsc{FairProbe} probes an increasing number of oldest streams, which achieves
+dovetailing. In all other iterations, \textsc{HeuristicProbe} attempts to
+extract up to \ParamMaxStreams{}~clauses from the most promising streams (by default,
+$\ParamMaxStreams = 7$).
+In each attempt, the most promising stream in $Q$ is chosen. If its first
+element is $\emptyset$, the rest of the stream is inserted into $Q$ and a new stream is
+chosen. This is repeated until either \ParamRetry{} occurrences of $\emptyset$ have  been
+met (by default, $\ParamRetry = 20$) or the stream yields a singleton. Setting $\ParamRetry > 0$ increases
+the chance that \textsc{HeuristicProbe} will return $\ParamMaxStreams$ clauses, as desired. Finally, if $P$ becomes empty, \textsc{ForceProbe}
+searches relentlessly for a clause in $Q$, as a fallback.
+
+\looseness=-1
+The function \textsc{ExtractClause} extracts an element from a nonempty stream
+not in $Q$ and inserts the remaining
+stream into $Q$ with an increased weight, calculated as follows.
+Let $n$ be the number of times the stream was chosen for
+probing. If probing results in $\emptyset$, the stream's weight is increased by
+$\max\,\{2{,}\; n-16\}$. If probing results in a clause $C$ whose penalty is
+$p$, the stream's weight is increased by $p \cdot \max\,\{1{,}\; n-64\}$. The
+penalty of a clause is a number assigned by Zipperposition based on
+features such as the depth of its derivation and the rules used in it.
+The constants $16$ and $64$ increase the chance that newer clause-producing streams are picked,
+which is desirable because their first clauses are expected to be useful.
+
+\looseness=-1
+All three probing functions are invoked by
+\textsc{GivenClause}, which contains the saturation loop. It differs
+from the standard given clause procedure in three ways:
+First, the proof state includes $Q$ in addition to $P$ and $A$. Second,
+new inferences involving the given clause are added to $Q$ instead of being
+performed immediately. Third, inferences in $Q$ are periodically performed
+lazily to fill $P$.
+
+\newcommand\infstream[1]{[#1]}
+
+\begin{exa} 
+  \begin{sloppypar}
+  In this example we simplify the notation by writing positive predicate literals
+  $\poslit{s}$ as $s$ and negative predicate literals $\neglit{t}$ as $\neg t$.
+  Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
+  \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}),\allowbreak
+  \neg \cst{p} \, (\cst{f}^{100} \, \cst{a})  \}$ and a selection function which
+  selects negative literals. 
+  Let $P \mid A \mid Q$ denote
+  the state of the given clause loop (i.e., the contents of the passive and active set
+  and of the stream queue), and let $\infstream{ a_1, a_2, \ldots }$
+  denote an infinite stream of elements.
+  \end{sloppypar}
+
+  The given clause loop begins in the state $X \, (\cst{f} \, \cst{a}) \not\eq
+  \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}), \neg \cst{p} \,
+  (\cst{f}^{100} \, \cst{a}) \mid \emptyset \mid \emptyset$. If the clause $\neg
+  \cst{p} \, (\cst{f}^{100} \, \cst{a})$ is chosen for processing, since $Q$ is empty
+  and no inferences with the chosen clause are possible, the state becomes $X \,
+  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
+  \cst{a}) \mid \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \emptyset$. When
+  the clause $X \, (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor
+  \cst{p} \, (X \, \cst{a})$ is chosen, a new stream which enumerates results of
+  equality resolution (on its first literal) is created. There are infinitely many
+  conclusions of this inference, since there are infinitely many unifiers for the
+  first literal of the form $\{ X \mapsto \lambda x. \, \cst{f}^i \, x \}$, for
+  $i \geq 0$. Thus, the stream is $\infstream{ \{\cst{p} \, \cst{a}\}, \{\cst{p} \, (\cst{f}\,\cst{a})\}, \ldots }$,
+  possibly with $\emptyset$s interspersed. With the standard given clause procedure,
+  there would have been no way to represent this infinitary result.
+  %The creation of the
+  %inference result stream is the most striking difference between our
+  %and standard given clause procedure.
+
+  When the stream is created, its first element is popped and put into $P$. Then, based on the
+  parameters that control inference stream probing, some number of clauses from
+  the stream are computed and moved to $P$. After two iterations, the state might be
+  $ \cst{p} \,
+  \cst{a}, \cst{p} \, (\cst{f} \, \cst{a}), \allowbreak \cst{p} \, (\cst{f} \, (\cst{f} \, \cst{a}))   \mid X \,
+  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
+  \cst{a}), \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \infstream{ \{\cst{p} \, (\cst{f}^3\,\cst{a})\}, \ldots }$.
+
+  In the next iterations, some clause of the form $\cst{p} \, (\cst{f}^{i} \,
+  \cst{a})$, where $i < 100$, is chosen, but no inferences with it can be
+  performed. Then, the stream created in the second iteration is probed, and its
+  results fill the set $P$. Ultimately, the clause $\cst{p} \, (\cst{f}^{100} \,
+  \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
+\end{exa}
+
+\textsc{GivenClause} eagerly stores the first element of a new inference stream
+in $P$ to imitate the standard given clause procedure. If the underlying
+unification procedure behaves like the standard first-order unification
+algorithm on higher-order logic's first-order fragment, our given clause
+procedure coincides with the standard one. The unification procedure described in Chapter \ref{ch:unif}
+terminates on the first-order and other fragments. To avoid computing
+complicated unifiers eagerly, it immediately returns $\emptyset$ for a problem that does not
+belong to one of the fragments that admit efficient unifier computation.
+
+The design of our given clause procedure was guided by folklore knowledge about
+higher-order theorem proving. First, in our experience most steps in
+long higher-order proofs involve first-order literals. The unification
+procedure and inference scheduling ensure that first-order inference
+conclusions are put in the proof state as early as possible. Second, some
+inference rules are expected to be largely useless. We initialize the stream
+penalty differently for each rule, allowing old streams of more useful
+inferences to be queried before newly added, but potentially less useful
+streams. Finally, if we use a unification procedure that has aggressive
+redundancy elimination, we will often find the necessary unifier within the
+first few unifiers returned. Similarly, if a stream keeps returning
+$\emptyset$, it is likely that it is blocked in a nonterminating computation
+and should be ignored. Our heuristics to increase the stream penalties take
+both observations into account.
+
+\ourpara{Evaluation and Discussion}
+
+\begin{sloppypar}
+
+  The evaluation of our unification algorithm in Sect.~\ref{sec:unif:evaluation}
+  shows that Zipperposition is the only competing
+  higher-order prover that proves all Church numeral problems from the TPTP,
+  never spending more than 5~s on a problem. On these hard
+  unification problems, the stream system allows the prover to explore the proof
+  state lazily.
+\end{sloppypar}
+
+Consider the TPTP problem \verb|NUM800^1|, which requires finding
+a function $F$ such that $F \, \cst{c_1} \, \cst{c_2}
+\ieq \cst{c_2} \iand F \, \cst{c_2} \, \cst{c_3} \ieq \cst{c_6}$, where
+$\cst{c}_n$ abbreviates the Church numeral for~$n$, $\lambda s\, z. \>
+s^n \, z$. To prove
+\confrep{it}{the problem}, it suffices to take $F$ to be the multiplication operator
+$\lambda x \, y \, s \, z. \> x \, (y \, s) \, z$.
+However, this unifier is only one out of many available for each occurrence of
+$F$.
+
+% TODO: If I add a new benchmark set, make sure this sentence is fixed
+\looseness=-1
+For evaluation of our unification algorithm (using somewhat different
+evaluation setup and older version of Zipperposition), we also compared a complete, nonterminating variant of the
+unification procedure and a pragmatic, terminating variant. The
+pragmatic variant was used directly---all the inference conclusions were put
+immediately in $P$, bypassing $Q$. The complete variant, which relies on
+possibly infinite streams and is much more prolific, proved only 15  problems
+less than the most competitive pragmatic variant. Furthermore, it proved 19
+problems not proved by the pragmatic variant.
+%
+This shows that our given clause procedure, with its heuristics, allows the
+prover to defer exploring less promising branches of the unification and uses
+the full power of a complete higher-order unifier search to solve unification
+problems that cannot be proved by a restricted procedure.
+
+\looseness=-1
+The parameters \ParamMode{}, \ParamRetry{}, and \ParamMaxStreams{} can greatly
+influence the behavior of the given clause procedure, even when the same
+unification procedure is used. Figure~\ref{fig:streams}
+presents the effects of these parameters on TPTP and SH.
+Selecting a low number of best clauses seems to
+perform well on both benchmark sets. However, on SH benchmarks, which require
+overwhelmingly first-order unifiers, visiting older streams should be delayed
+a lot.
+
+As with Boolean selection functions, changing these three parameters causes
+a substantial difference in the set of proved problems. For example, the
+configuration that performs the worst on TPTP benchmarks proves \NumberOK{12}
+problems that the configuration performing the best on TPTP cannot prove; moreover, there
+are \NumberOK{29} TPTP problems that are proved by some set of parameters
+other than $\ParamMode=\ParamMaxStreams=16, \ParamRetry=2$. On SH, these
+effects are much weaker; most reasonable combinations
+of parameters perform similarly.
+
+Among the competing higher-order provers, only Satallax uses infinitely
+branching calculus rules. It maintains a queue of ``commands'' that contain
+instructions on how to create a successor state in the tableau. One
+command describes infinite enumeration of all closed terms of a given function
+type. Each execution of this command makes progress in the enumeration. Unlike
+evaluation of streams representing elements of CSU, each command execution
+is guaranteed to make progress in enumerating the next closed functional
+term, so there is no need to ever return $\emptyset$.
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+
+\centering
+\begin{subfigure}[b]{1\textwidth}
+  \centering
+  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
+  &&&&&&& \ParamMode \\[.5\jot]
+  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
+  \cline{3-5}\cline{7-9}\cline{11-13}
+  \\[-1.5\jot]
+  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
+  %                                     2                        16                     256
+  &                         & 2    & 16   & 128  & & 2         & 16   & 128  & & 2    & 16   & 128 \\\midrule
+  & $2$                     & 1643 & 1645 & 1645 & & 1661      & 1661 & 1658 & & 1669 & 1664 & 1664 \\[0.5\jot]
+  $\ParamMaxStreams$ & $16$ & 1647 & 1646 & 1609 & & {\bf1670} & 1654 & 1602 & & 1665 & 1659 & 1597 \\[0.5\jot]
+  & $128$                   & 1646 & 1644 & 1583 & & 1661      & 1656 & 1577 & & 1665 & 1658 & 1576 \\ \bottomrule
+  \end{tabular}
+  \caption{TPTP benchmarks}
+  \label{fig:streams-tptp}
+\end{subfigure}
+\par\bigskip
+\begin{subfigure}[b]{1\textwidth}
+  \centering
+  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
+  &&&&&&& \ParamMode \\[.5\jot]
+  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
+  \cline{3-5}\cline{7-9}\cline{11-13}
+  \\[-1.5\jot]
+  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
+  %                                          2                                               16                                      128
+  &                         & 2            & 16            & 128          & & 2            & 16            & 128          & & 2                  & 16            & 128 \\\midrule
+  & $2$                     & \colalign460 & \colalign455  & \colalign454 & & \colalign465 & \colalign463  & \colalign458 & & \colalign466       & \colalign461  & \colalign461 \\[0.5\jot]
+  $\ParamMaxStreams$ & $16$ & \colalign458 & \colalign453  & \colalign445 & & \colalign464 & \colalign459  & \colalign441 & & \colalign{\bf468}  & \colalign459  & \colalign442 \\[0.5\jot]
+  & $128$                   & \colalign456 & \colalign452  & \colalign430 & & \colalign465 & \colalign458  & \colalign428 & & \colalign{\bf468}  & \colalign459  & \colalign425 \\ \bottomrule
+  \end{tabular}
+  \caption{SH benchmarks}
+  \label{fig:streams-sh}
+\end{subfigure}
+\caption{Impact of the stream enumeration parameter}
+\label{fig:streams}
+\end{figure}

### 2022-03-15 ho-techniques/ho-techniques.tex
more

@@
 \end{figure}
+
+\section{Controlling Prolific Rules}
+\label{sec:ho-tech:explosiveness}
+
+To support higher-order features
+such as function extensionality and quantification over functions,
+many refutationally complete calculi employ highly prolific rules.
+For example, \lsup{} includes a
+\infname{FluidSup} rule \cite{bbtvw-21-sup-lam} that very often applies to two
+clauses if one of them contains a term of the form $F \, \overline{s}_n$,
+where $n > 0$. This rule is inherited in the successor of the calculus,
+\osup{}, that we use in this chapter.
+We describe three mechanisms to keep rules like these under control.
+
+First, \emph{we limit applicability of the prolific rules}. In practice, it
+often suffices to apply prolific higher-order rules only to initial or shallow
+clauses---clauses with a shallow derivation depth. Thus, we added an option to
+forbid the application of a rule if the derivation depth of any premise exceeds a
+limit.
+
+\newcommand\ParamPenaltyIncrease{\ensuremath{K_\mathrm{incr}}}
+Second, \emph{we penalize the streams of expensive inferences}. The weight of
+each stream is given an initial value based on characteristics of the inference
+premises such as their derivation depth. For prolific rules such as
+\infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
+less prolific variants of this rule, such as \infname{DupSup} \cite{bbtvw-21-sup-lam}, are increased by a
+fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).
+
+\looseness=-1
+Third, \emph{we defer the selection of prolific clauses}. To select the given
+clause, most saturating provers evaluate clauses according to some criteria and
+choose the clause with the lowest evaluation. To make this choice efficient,
+passive clauses are organized into a priority queue ordered by their
+evaluations. Like E, Zipper\-position maintains multiple
+queues, ordered by different evaluations, that are visited in a round-robin
+fashion. It also uses E's two-layered evaluation, a variant of which
+has recently been implemented in Vampire~\cite{gs-20-clausesel}.
+%
+The two layers are \emph{clause priority} and \emph{clause weight}. Clauses
+with higher priority are preferred, and the weight is used for tie-breaking.
+Intuitively, the first layer crudely separates clauses into priority classes,
+while the second one uses heuristic weights to prefer clauses within a
+priority class. To control the selection of prolific clauses, we introduce new
+clause priority functions that take into account features specific to
+higher-order clauses.
+
+The first new priority function, \verb|PreferHOSteps| (\verb|PHOS|), assigns a
+higher priority if rules specific to higher-order superposition calculi
+were used in the clause derivation. Since most of the other clause priority
+functions tend to defer higher-order clauses, having a clause queue that prefers
+them might be useful to find some proof more
+efficiently. A simpler function, which prefers clauses containing
+$\lambda$-abstractions, is \verb|PreferLambda| (\verb|PL|).
+
+\verb|PreferHOSteps| separates clauses created using first-
+and higher-order inference rules crudely. However, within higher-order
+inference rules there are the ones which make clauses simpler and are thus more
+preferable. An example of such a rule is
+%
+\[\namedinference{\infname{ArgCong}}{C \lor s \eq t}{C \lor s \, \tuplen{X}
+\eq t \, \tuplen{X}}\]
+%
+where $s$ is of the type $\alpha_1 \rightarrow \cdots \rightarrow \alpha_k
+\rightarrow \beta$, $\beta$ is a base type, $n \leq k$, free variables
+$\tuplen{X}$ are fresh, and literal $s \eq t$ is strictly eligible (for paramodulation). When $n =
+k$, in most cases, the resulting clause has a first-order literal $s \, \tuplen{X}
+\eq t \, \tuplen{X}$ in place of the literal $s \eq t$ of functional type, which usually makes the clause more useful. To
+prefer clauses that are only mildly higher-order, we designed the function
+\verb|PreferEasyHO| (\verb|PEHO|). It prefers clauses that are the result of
+\infname{ArgCong}, have equations between terms of functional type or
+between higher-order patterns, or have literals containing logical symbols, in that
+order of priority.
+
+A higher-order inference that applies a complicated substitution to a clause is
+usually followed by a $\beta\eta$-reduction step. If
+$\beta\eta$-reduction greatly reduces the size of a clause, it is likely
+that this substitution simplifies the clause (e.g., by removing a variable's
+arguments). The new priority function \verb|ByNormalizationFactor| (\verb|BNF|)
+is designed to exploit this observation. It prefers clauses that were produced
+by $\beta\eta$-reduction, and among those it prefers the ones with larger
+size reductions.
+
+Another new priority function is \verb|PreferShallowAppVars| (\verb|PSAV|). This
+prefers clauses with lower depths of the deepest occurrence of an applied
+variable---that is, $C[X \, \cst{a}]$ is preferred over $C[\cst{f}\,(X \,
+\cst{a})]$. The intuition is that applying a substitution to an applied
+variable often reduces the variable to a term with a constant head,
+yielding a less explosive clause, and the gain is greater for variables closer
+to the top level.  Among the functions that rely
+on properties of applied variables, we implemented \verb|PreferDeepAppVars|
+(\verb|PDAV|), which returns the priority opposite of \verb|PSAV|, and
+\verb|ByAppVarNum| (\verb|BAVN|), which prefers clauses with fewer occurrences
+of applied variables.
+
+\ourpara{Evaluation and Discussion}
+
+\begin{figure}[t]
+  \centering
+  % \def\arraystretch{1.1}%
+  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
+            & \texttt{CP}                & \texttt{BAVN}  & \texttt{PL}  & \texttt{PSAV}       & \texttt{PHOS} & \texttt{PEHO}  & \texttt{BNF} & \texttt{PDAV}      \\ \midrule
+   TPTP     & {1635}$^\star$             & {\bf 1640}     & 1604         & {1635}              & 1609          & 1617           & 1575         & 1533 \\[0.5\jot]
+   SH       & {\bf \colalign452}$^\star$ & \colalign451   & \colalign417 & {\colalign450}      & \colalign439  & \colalign407   & \colalign411 & \colalign302 \\ \bottomrule
+  \end{tabular}}
+  \caption{Impact of the priority function}
+  \label{fig:priorities}
+\end{figure}
+\begin{figure}[t]
+  \centering
+  \def\arraystretch{1.1}%
+  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
+              & $\infty$                     & $16$          & $8$           & $4$             & $2$             & $1$      \\ \midrule
+   TPTP       & {\bf 1635}$^\star$           & 1625          & 1632          & 1629            & 1628            & 1618 \\[0.5\jot]
+   SH         & {\bf \colalign452}$^\star$   & \colalign438  & \colalign435  & \colalign439    & \colalign435    & \colalign440 \\ \bottomrule
+
+  \end{tabular}}
+  \caption{Impact of the \infname{FluidSup} weight increment \ParamPenaltyIncrease}
+  \label{fig:penalties}
+\end{figure}
+
+\looseness=-1
+In the base configuration (\emph{base}), Zipperposition visits several clause
+queues. The configuration uses queues that prefer the clauses that stem from
+the conjecture, the ones that have at least one positive literal, the ones that
+have been moved from active to passive set, and so on. One of the queues uses the constant
+priority function \texttt{ConstPrio} (\texttt{CP}), meaning that it assigns the same
+priority to every clause. As this queue is the most often visited one in \emph{base},
+changing its priority function should affect the result noticeably.
+To evaluate the new priority functions, we replaced
+\texttt{CP} with one of the new functions in this queue,
+%the queue ordered by \texttt{CP} with the queue ordered by one of the new
+%functions,
+leaving the clause weight intact. The results are shown in Fig.~\ref{fig:priorities}.
+
+Even though constant priority function achieves remarkable performance, the new priority functions are
+useful additions to the prover's repertoire: \NumberOK{37} additional
+TPTP problems and \NumberOK{17} additional SH problems can be proved when some
+nonconstant priority is used. The generally average-performing \verb|PEHO|
+function can prove \NumberOK{9} problems not proved with any other priority function on TPTP
+(and \NumberOK{1} on SH). Globally, \NumberOK{24} TPTP problems and \NumberOK{6} SH problems
+can be proved exclusively using one particular priority function.
+
+Although it is necessary for refutational completeness, the \infname{FluidSup}
+rule is disabled in \emph{base} because it is so explosive and so seldom useful.
+To test whether increasing inference stream weights makes a difference on the
+success rate, we tried enabling \infname{FluidSup} with different weight
+increments~$\ParamPenaltyIncrease$ for \infname{FluidSup} inference queues. The
+results are shown in Fig.~\ref{fig:penalties}. As expected, using a low
+increment with \infname{FluidSup} is detrimental on TPTP. On this
+benchmark set, \NumberOK{16} additional problems can be proved when
+\infname{FluidSup} is enabled. The penalty mostly affects only proving time: All
+but \NumberOK{2} of these problems were proved by using at least three different
+values of $\ParamPenaltyIncrease$. On SH problems, the best result is obtained
+when the rule is disabled as well. Unexpectedly, the next best result is obtained when
+$\ParamPenaltyIncrease=1$.
+
+\section{Controlling the Use of Backends}
+\label{sec:ho-tech:backends}
+
+\newcommand{\ParamNumClauses}{\ensuremath{K_\mathrm{size}}}
+\newcommand{\ParamTime}{\ensuremath{K_\mathrm{time}}}
+
+Cooperation with efficient off-the-shelf first-order theorem provers is an
+essential feature of higher-order theorem provers such as Leo-III
+\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-2013-satallax}.
+% but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
+Those provers invoke first-order backends repeatedly
+during a proof attempt and spend a substantial amount of time in backend
+collaboration. Since \lsup{} generalizes a highly efficient
+first-order calculus, we expect that future efficient \lsup{}
+implementations will not benefit much from backends.
+Nevertheless, experimental provers such
+as Zipperposition can still gain a lot. We present some
+techniques for controlling the use of backends.
+
+In his thesis \cite[Sect.~6.1]{as-18-phd}, Steen extensively studies
+the effects of using different first-order backends on the performance of
+Leo-III. His results suggest that adding only one backend already substantially
+improves the performance. To reduce the effort required for integrating multiple backends, we chose Ehoh, an extension of E \cite{scv-19-e23} to support \lfsup{},  as our single
+backend. In Chapter \ref{ch:ehoh}, we described Ehoh in detail.
+%formulas occurring as arguments of function symbols.
+On the one hand, Ehoh provides the efficiency of E while easing the translation from full
+higher-order logic---the only missing syntactic feature is
+$\lambda$-abstraction. On the other hand, Ehoh's higher-order reasoning
+capabilities are limited. Its unification algorithm is essentially first-order,
+and it cannot synthesize $\lambda$-abstractions.
+% or instantiate predicate variables.
+
+In a departure from Leo-III and other cooperative provers,
+\confrep{}{instead of regularly invoking the backend, }%
+we invoke \confrep{the backend}{it} at most once during a run of Zipperposition.
+This is because most competitive higher-order provers, including Zipperposition, use a portfolio
+mode in which many configurations are run for a short time, and we want to
+leave enough time for native higher-order reasoning. Moreover, multiple
+backend invocations tend to be wasteful, because currently each invocation
+starts with no knowledge of the previous ones.
+
+\looseness=-1
+Only a carefully chosen subset of the available clauses are translated and sent
+to Ehoh. Let $I$ be the set of \confrep{input clauses}{clauses representing the
+input problem}. Given a proof state, let \confrep{$M = P \cup A$}{$M$ denote the
+union of the current active and passive sets}, and let $M_\mathrm{ho}$ denote
+the subset of $M$ that contains only clauses that were derived using at least
+one \lsup{} rule not present in regular superposition. We order the clauses in
+$M_\mathrm{ho}$ by increasing derivation depth, using syntactic weight to break
+ties. Then we choose all clauses in $I$ and the first \ParamNumClauses~clauses from
+$M_\mathrm{ho}$ for use with the backend reasoner.
+%We include all the clauses in $I$ since they constitute the backbone of the
+%initial problem. -- Tautology. --JB
+We leave out clauses in $M \setminus (I \cup M_\mathrm{ho})$ because Ehoh can rederive them.
+We also expect large clauses with deep derivations to be less useful.
+
+The remaining step is the translation of $\lambda$-abstractions. We implemented two
+translation methods:\ $\lambda$-lifting \cite{tj-1985-lambdalift} and
+$\cst{SKBCI}$ combinators \cite{da-1979-combtrans}. For
+$\cst{SKBCI}$, we omit the combinator definition axioms, because they are
+very explosive \cite{br-20-full-sup-w-combs}. A third mode simply omits clauses
+containing $\lambda$-abstractions.
+
+\ourpara{Evaluation and Discussion}
+  %
+  % \begin{enumerate}
+  %   \item Different points when E is called, for different time periods
+  %   \item Number of clauses from $M_\mathrm{ho}$ that is translated
+  %   \item Different kinds of liftings
+  % \end{enumerate}
+
+  % We evaluated the following parameters of backend collaboration: the moment of
+  % backend invocation, size of $M_\mathrm{ho}$, and $\lambda$-abstraction
+  % translation method.
+  In Zipperposition, we can adjust the CPU time allotted to Ehoh, Ehoh's own
+  parameters, the point when Ehoh is invoked, the number \ParamNumClauses~of selected clauses from
+  $M_\mathrm{ho}$, and the $\lambda$ translation method. We fix the time
+  limit to 3~s, use Ehoh in \emph{autoschedule} mode, and focus on the last three
+  parameters. In \emph{base}, collaboration with Ehoh is
+  disabled (labeled $-$Ehoh).
+  %Parameters are evaluated by enabling the collaboration with E and
+  %varying parameter values.  -- Uninformative. --JB
+
+  \begin{figure}[t]
+    \noindent\hbox{}\hfill
+    \begin{minipage}[t]{.46\linewidth}
+      \centering
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
+            & $-$Ehoh                   & $0.1$              & $0.25$          & $0.5$               & $0.75$      \\ \midrule
+      TPTP & 1635$^\star$               & {\bf 1981}         & 1980            & 1979                & 1972 \\[0.5\jot]
+      SH   & \colalign452$^\star$       & \colalign606       & {\bf \colalign608}    & \colalign600        & \colalign592 \\ \bottomrule
+      \end{tabular}}
+      \caption{Impact of the backend invocation point $\ParamTime$}
+      \label{fig:invocation}
+    \end{minipage}\hfill\hfill
+    \begin{minipage}[t]{.46\textwidth}
+      \centering
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
+           & $-$Ehoh              & lifting              & \textsf{SKBCI}    & omitted \\ \midrule
+      TPTP & 1635$^\star$         & {\bf 1980}           & 1877              & 1866 \\[0.5\jot]
+      SH   & \colalign452$^\star$ & \colalign{\bf 608}   & \colalign577      & \colalign566 \\ \bottomrule
+      \end{tabular}}
+      \caption{Impact of the method used to translate $\lambda$-abstractions}
+      \label{fig:translation}
+    \end{minipage}%
+    \hfill\hbox{}
+  \end{figure}
+
+  \begin{figure}[t]
+    \centering
+    \def\arraystretch{1.1}%
+    \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}}\toprule
+          & $-$Ehoh              & $16$               & $32$               & $64$           & $128$             & $256$         & $512$     \\ \midrule
+     TPTP & 1635$^\star$         & {\bf 1985}         & 1980               & 1978           & 1968              & 1968          & 1919 \\[0.5\jot]
+     SH   & \colalign452$^\star$ & \colalign606       & {\bf \colalign608} & \colalign600   & \colalign598      & \colalign596  & \colalign589 \\ \bottomrule
+    \end{tabular}}
+    \caption{Impact of the number of selected clauses~$\ParamNumClauses$}
+    \label{fig:m-ho-size}
+  \end{figure}
+
+  \looseness=-1
+  Ehoh is invoked after $ \ParamTime \cdot t $ CPU seconds, where $0 \le \ParamTime < 1$ and $t$
+  is the total CPU time allotted to Zipperposition. Figure~\ref{fig:invocation}
+  shows the effect of varying $\ParamTime$ when $\ParamNumClauses = 32$
+  and $\lambda$-lifting is used. The evaluation confirms that using a highly
+  optimized backend such as Ehoh greatly improves the performance of a
+  less optimized prover such as Zipperposition.
+  The figure indicates that it is preferable to invoke the backend early. We
+  have indeed observed that if the backend is invoked late, small clauses with
+  deep derivations tend to be present by then. These clauses might have been
+  used to delete important shallow clauses already. But due to their derivation
+  depth, they will not be translated. In such situations, it is better to
+  invoke the backend before the important clauses are deleted.
+
+  Figure~\ref{fig:translation} quantifies the effects of the three
+  $\lambda$-abstraction translation methods. We fixed $\ParamTime = 0.25$ and
+  $\ParamNumClauses=32$. The clear winner is $\lambda$-lifting.
+  $\cst{SKBCI}$ combinators perform slightly better than omitting clauses
+  containing $\lambda$-abstractions.
+
+  \looseness=-1
+  Figure~\ref{fig:m-ho-size} shows the effect of $\ParamNumClauses$ on
+  performance, with $\ParamTime = 0.25$ and $\lambda$-lifting. Including a
+  small number of higher-order clauses with the lowest weight performs better
+  than including a large number of such clauses.
+
+  \section{Comparison with Other Provers}
+\label{sec:ho-tech:comparison}
+
+%The raw evaluation data for the previous experiments show that
+\looseness=-1
+Different choices of
+parameters lead to noticeably different sets of proved problems. In an attempt
+to use Zipperposition~2 to its full potential, we have created a portfolio mode
+that runs up to 50 configurations in parallel during the allotted time. The portfolio
+was designed to solve as many problems as possible from the TPTP benchmark set. To
+provide some context, we compare Zipperposition~2 with the following versions of
+all higher-order provers that competed at CASC in 2020:\ CVC4 1.9
+\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
+\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
+The provers were run using the same parameters as in CASC, but with updated
+executables.
+Note that
+Vampire's higher-order schedule is optimized for running on a single core.
+We also include Ehoh~2.7, the first version of this prover to syntactically support
+full higher-order logic, including $\lambda$-abstractions.
+Semantically, Ehoh~2.7 is arguably the weakest among the listed provers:
+It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
+$\lambda$-lifting before it applies \relax{\lfsup{}} \cite{bbcw-21-lfho}
+on the preprocessed problem.
+
+We use the same benchmark sets as elsewhere in this chapter. To imitate the
+setup of the 2020 edition of CASC, we use a 120~s wall-clock limit and a 960~s CPU limit.
+We even carried out our evaluation on the 8-core CPU nodes that were used for
+CASC in 2020. We also ran Zipperposition in uncooperative mode, in which its
+collaboration with a backend is disabled. Figure~\ref{fig:other-provers}
+summarizes the results.
+
+\begin{figure}[t]
+  \centering
+  \def\arraystretch{1.1}%
+  \relax{\begin{tabular}{@{}l@{\hskip 2.5em}c@{\hskip 2em}c@{}} \toprule
+                  & TPTP    & SH  \\ \midrule
+   CVC4           & 1816    & 587   \\
+   Ehoh           & 1980    & 676   \\
+   Leo-III        & 2122    & 616  \\
+   Satallax       & 2175    & 588  \\
+   Vampire        & 2072    & 660   \\
+   Zipperposition-uncoop & 2311 & 652 \\
+   Zipperposition & {\bf 2412}    & {\bf 715}  \\ \bottomrule
+  \end{tabular}}
+  \caption{Comparison of competing higher-order theorem provers}
+  \label{fig:other-provers}
+\end{figure}
+
+% Among the cooperative provers, Zipperposition is the one that depends the least
+% on its backend, and its \emph{uncooperative} mode is only \NumberNOK{one}~problem
+% behind Satallax's \emph{cooperative} mode. This confirms our hypothesis that
+% \osup{} is a suitable basis for automatic higher-order reasoning.
+% \confrep{This also}{The increase in performance due to the addition of an efficient backend}
+% suggests that the implementation of this calculus in a modern first-order
+% superposition prover such as E or Vampire would achieve markedly better results.
+% Moreover, we believe that there are still techniques inspired by tableaux, SAT
+% solving, and SMT solving that could be adapted and integrated in saturation
+% provers.
+
+The evaluation results corroborate the CASC results. They also
+show that Zipperposition outperforms all other provers on SH benchmarks. This
+confirms our hypothesis that \osup{} is a suitable basis for automatic
+higher-order reasoning. Further confirmation is provided by the success rate of
+Zipperposition's uncooperative version: Even without backend,
+Zipperposition is substantially better than all other provers on TPTP
+benchmarks, and it matches the performance of the top contenders on SH.
+On the other hand, the increase in performance due to the addition
+of an efficient backend suggests that the implementation of this calculus in a
+modern first-order superposition prover such as E or Vampire would achieve
+even better results.
+
+We believe that there are still techniques inspired by
+tableaux, SAT solving, and SMT solving that could be adapted and integrated in
+saturation provers. In particular, there are still \NumberOK{25} TPTP problems
+and \NumberOK{17} SH problems that can be proved by other provers but not by
+Zipperposition.
+
+\section{Discussion and Conclusion}
+\label{sec:ho-tech:discussion}
+
+%  In this paper, we proposed solutions to issues that arise when implementing a
+%  higher-order prover based on \lsup. Some of these also apply
+%  to combinatory superposition. We also discussed many choices that can be made
+%  during proof search and evaluated some of those choices.
+
+\begin{sloppypar}
+Back in 1994, Kohlhase \cite[Sect.~1.3]{mk-94-hores} was optimistic about the
+future of higher-order automated reasoning:
+%
+\begin{quote}
+  %The author believes that
+  The obstacles to proof search intrinsic
+  to higher-order logic may well be compensated by the greater expressive power
+  of higher-order logic and by the existence of shorter proofs. Thus
+  higher-order automated theorem proving will be practically as feasible as
+  first-order theorem proving is now as soon as the technological backlog is made up.
+\end{quote}
+%
+For higher-order superposition, the backlog consisted of designing calculus
+extensions, heuristics, and algorithms that mitigate its weaknesses. In
+this chapter, we presented such enhancements, justified their design, and
+evaluated them. We explained how each weak point in the
+higher-order proving pipeline could be improved, from preprocessing to reasoning
+about formulas, to delaying unpromising or explosive inferences, to invoking a
+backend. Our evaluation indicates that higher-order superposition is now the
+state of the art in higher-order reasoning.
+\end{sloppypar}
+
+\looseness=-1
+Higher-order extensions of first-order superposition have been
+considered %from a theoretical perspective    -- The point is made later in the same para. --JB
+by Bentkamp et al.\
+\cite{bbtvw-21-sup-lam, bbcw-21-lfho} and Bhayat and Reger
+\cite{br-19-restricted-unif, br-20-full-sup-w-combs}. They introduced proof calculi,
+proved them refutationally complete, and suggested optional rules, but
+they hardly discussed the practical aspects of higher-order superposition. Extensions
+of SMT are discussed by Barbosa et al. \cite{brotb-19-ho-smt}.
+Bachmair and Ganzinger \cite{bg-1992-nonclausal}, Manna and Waldinger
+\cite{mw-1979-nonclausal}, and Murray \cite{nm-1982-nonclausal} have studied
+nonclausal resolution calculi.
+
+In contrast, there is a vast literature on practical aspects of first-order
+reasoning using superposition and related calculi.
+The literature evaluates various procedures and techniques
+\cite{hv-09-unifalgs,rsv-15-playing-with-avatar}, literal and term order selection
+functions \cite{hrsv-16-selsel}, and clause evaluation functions
+\cite{sm-16-clausesel, gs-20-clausesel}, among others. Our work joins the
+select club of publications devoted to practical aspects of higher-order
+reasoning
+\cite{sb-15-beta,wskb-16-effective-norm,fb-2016-internal-guidance-satallax,benzmueller-et-al-05-can-ho-fo-coop}.
+
+%In contrast, higher-order calculi implemented in competitive provers are usually discussed from a theoretical viewpoint.
+
+%\section{Conclusion and Future Work}
+%\label{sec:ho-tech:conclusion}
+
+% We filled in this gap: we describe how
+% We described some ways in which the explosion incurred by
+% \lsup{} can be kept under control. We showed how native
+% support for a Boolean type can dramatically improve a higher-order prover's
+% performance. For many choices that can be made during the higher-order proof
+% search, we described heuristics and evaluated them. We also designed a
+% saturation loop that enumerates infinite sets of higher-order inference
+% conclusions, while maintaining the same behavior as the standard given clause
+% procedure on first-order clauses.
+
+% As a next step, we chose a sub.
+% %%% @PETAR: In E 2.6, basically. --JB
+% % Ehoh \cite{ehoh-section}, the $\lambda$-free higher-order extension of E.
+% We
+% expect the resulting prover to be substantially more efficient than
+% Zipperposition. Moreover, we want to investigate the proofs found by provers
+% such as CVC4 and Satallax but missed by Zipperposition. Finding the reason
+% behind why Zipperposition fails to prove specific problems will likely result in useful new techniques.
+
+The work presented in this chapter proved invaluable to reaching the last stop
+on the road to the implementation of full higher-order logic in a
+state-of-the-art first-order prover. We used the results and experience we
+gained with this work to choose a set of well-performing easy-to-implement
+calculus extensions and heuristics. We implemented them in Ehoh, obtaining a
+prover called $\lambda$E. This new prover substantially increases the
+higher-order reasoning capabilities of Ehoh. Even without all of the techniques
+described in this chapter, it outperforms all higher-order provers except for
+Zipperposition. We give detailed description of $\lambda$E, as well as reasoning
+behind all design decision we took in Chapter \ref{ch:ehoh2}.

### 2022-03-15 satfol/satfol.tex
new bib and paper

@@
 \label{ch:satfol}
+
+\includeversion{noqle}
+\excludeversion{qle}
+\newcommand{\paper}[0]{chapter}
+
+\authors{
+    Joint work with\\
+    Jasmin Blanchette,
+    and Marijn J.H. Heule
+}
+
+\begin{abstract}
+Optimized SAT solvers not only preprocess the clause set, they also transform it
+during solving as inprocessing. Some preprocessing techniques have been
+generalized to first-order logic with equality. In this \paper, we port
+inprocessing techniques to work with superposition, a leading first-order proof
+calculus, and we strengthen known preprocessing techniques. Specifically, we
+look into elimination of hidden literals, variables (predicates), and blocked
+clauses. Our evaluation using the Zipperposition prover confirms that the new
+techniques usefully supplement the existing superposition machinery.
+\end{abstract}
+
+\blfootnote{In this work I was the main designer behind all the presented techniques.
+Marijn Heule did weekly supervision and provided his knowledge of SAT solving to 
+guide the design of all techniques. Jasmin Blanchette found the exact conditions
+under which superposition remains complete when predicate elimination rule is added
+to the calculus. I also implemented and evaluated all the techniques. 
+}
+
+\newpage

### 2022-03-15 scratch/paper.tex
new bib and paper

@@
 \cite{henkin-1950-completeness}), resulting in
-\emph{\lsup} \cite{bbtvw-21-sup-lam}, which we developed
+\emph{\lsup} \cite{bentkamp-et-al-2021-lamsup-journal}, which we developed
 together with Waldmann, as well as \emph{combinatory superposition}
@@
 \ourpara{Zipperposition}
-Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
+Zipperposition \cite{sc-15-simon-phd,bentkamp-et-al-2021-lamsup-journal} is a higher-order
 theorem prover that implements both \lsup{} variants, combinatory
@@
 When they introduced \lsup, Bentkamp et al.\
-\cite{bbtvw-21-sup-lam} described an older version of this extension. Here
+\cite{bentkamp-et-al-2021-lamsup-journal} described an older version of this extension. Here
 we present a newer version in more detail, including heuristics to postpone
@@
 For example, \lsup{} includes a
-\infname{FluidSup} rule \cite{bbtvw-21-sup-lam} that very often applies to two
+\infname{FluidSup} rule \cite{bentkamp-et-al-2021-lamsup-journal} that very often applies to two
 clauses if one of them contains a term of the form $F \, \overline{s}_n$,
@@
 \infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
-less prolific variants of this rule, such as \infname{DupSup} \cite{bbtvw-21-sup-lam}, are increased by a
+less prolific variants of this rule, such as \infname{DupSup} \cite{bentkamp-et-al-2021-lamsup-journal}, are increased by a
 fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).
@@
 
-\section{Discussion and Conclusion}
-\label{sec:ho-tech:discussion}
-
-%  In this paper, we proposed solutions to issues that arise when implementing a
-%  higher-order prover based on \lsup. Some of these also apply
-%  to combinatory superposition. We also discussed many choices that can be made
-%  during proof search and evaluated some of those choices.
+% For VSCode
+% !TEX root = ./conf.tex
+
+%%% TODO @JASMIN @PETAR: The definition of flat L-resolvent is broken because
+%%% of the condition on the variables. --JB
+%%% TODO @JASMIN @PETAR: The definition of "tautology" is probably wrong for
+%%% BCE. Cf. Kiesl et al.
+
+%%% TODO @PETAR: Check @PETAR below. --JB
+
+%\def\orcid#1{} %{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+%TODO: uncomment before submission:
+% \def\orcid#1{\orcidID{#1}}
+
+% \bibliographystyle{splncs04}
+% \usepackage{amsthm}
+% \usepackage{ntheorem}
+\usepackage{flushend}
+\usepackage{cite}
+\usepackage{etoolbox}
+\usepackage{dcolumn}
+\usepackage{algorithmicx}
+\usepackage{regexpatch}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage[mathscr]{eucal}
+\usepackage{prftree}
+\usepackage[utf8]{inputenc}
+% \usepackage[T1]{fontenc}
+% \usepackage[english]{babel}
+\usepackage{enumerate}
+\usepackage{graphicx}
+\usepackage{cite}
+\usepackage{tikz}
+\usepackage{pgfplots} 
+\usepackage{amsmath}
+\usepackage{amssymb}
+%\usepackage{subcaption}
+%\usepackage{caption}
+\usepackage{booktabs}
+\usepackage[nomessages]{fp}
+\usepackage{enumerate}
+%\captionsetup{justification=raggedright,singlelinecheck=false}
+\xpatchcmd{\algorithmic}{\labelsep 0.5em}{\labelsep 3em}{\typeout{Success!}}{\typeout{Oh dear!}}
+
+%% BEGIN Times font etc.
+\usepackage{mathptmx}
+\usepackage[scaled=.82]{beramono}
+\usepackage[scaled=.865]{helvet}
+%\usepackage{avant}
+\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
+%% END Times font etc.
+
+% Used for displaying a sample figure. If possible, figure files should be
+% included in EPS format.
+
+%\usepackage[
+%   a4paper,
+%   pdftex,
+%   pdftitle={SAT-Inspired Eliminations for Superposition},
+%   pdfauthor={Petar Vukmirovi\'c, Jasmin Blanchette, and Marijn J.H. Heule},
+%   pdfkeywords={},
+%   pdfborder={0 0 0},
+%   draft=false,
+%   colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue,
+%   bookmarksnumbered,
+%   bookmarks,
+%   bookmarksdepth=2,
+%   bookmarksopenlevel=2,
+%   bookmarksopen]{hyperref}
+% If you use the hyperref package, please uncomment the following line to
+% display URLs in blue roman font according to Springer's eBook style:
+%\renewcommand\UrlFont{\color{blue}\rmfamily}
+
+\allowdisplaybreaks
+
+% \usepackage[mathlines]{lineno}
+% \linenumbersep=30pt
+% \linenumbers
+
+% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+%   \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+%   \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+%   \renewenvironment{#1}%
+%%      {\linenomath\csname old#1\endcsname}%
+%%      {\csname oldend#1\endcsname\endlinenomath}}% 
+% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+%   \patchAmsMathEnvironmentForLineno{#1}%
+%   \patchAmsMathEnvironmentForLineno{#1*}}%
+% \AtBeginDocument{%
+% \patchBothAmsMathEnvironmentsForLineno{equation}%
+% \patchBothAmsMathEnvironmentsForLineno{align}%
+% \patchBothAmsMathEnvironmentsForLineno{flalign}%
+% \patchBothAmsMathEnvironmentsForLineno{alignat}%
+% \patchBothAmsMathEnvironmentsForLineno{gather}%
+% \patchBothAmsMathEnvironmentsForLineno{multline}%
+% }
+
+% \usepackage{etoolbox} %% <- for \cspreto, \csappto, \patchcmd, \pretocmd, \apptocmd
+
+% %% Patch 'normal' math environments:
+%% \newcommand*\linenomathpatch[1]{%
+%%   \cspreto{#1}{\linenomath}%
+%%   \cspreto{#1*}{\linenomath}%
+%%   \cspreto{end#1}{\endlinenomath}%
+%%   \cspreto{end#1*}{\endlinenomath}%
+% }
+% %% Patch AMS math environments:
+%% \newcommand*\linenomathpatchAMS[1]{%
+%%   \cspreto{#1}{\linenomathAMS}%
+%%   \cspreto{#1*}{\linenomathAMS}%
+%%   \csappto{end#1}{\endlinenomath}%
+%%   \csappto{end#1*}{\endlinenomath}%
+% }
+
+%% %% Definition of \linenomathAMS depends on whether the mathlines option is provided
+%% \expandafter\ifx\linenomath\linenomathWithnumbers
+%%   \let\linenomathAMS\linenomathWithnumbers
+%   %% The following line gets rid of an extra line numbers at the bottom:
+%%   \patchcmd\linenomathAMS{\advance\postdisplaypenalty\linenopenalty}{}{}{}
+% \else
+%%   \let\linenomathAMS\linenomathNonumbers
+% \fi
+
+%% \linenomathpatch{equation}
+%% \linenomathpatchAMS{gather}
+%% \linenomathpatchAMS{multline}
+%% \linenomathpatchAMS{align}
+%% \linenomathpatchAMS{alignat}
+%% \linenomathpatchAMS{flalign}
+
+\title{SAT-Inspired Eliminations for~Superposition}
+
+\author{
+\IEEEauthorblockN{Petar Vukmirovi\'c\textsuperscript{1}\orcid{0000-0001-7049-6847},
+Jasmin Blanchette\textsuperscript{1,2}\orcid{0000-0002-8367-0936},
+Marijn J.H. Heule\textsuperscript{3}\orcid{0000-0002-5587-8801}}\\
+\IEEEauthorblockA{%
+\textsuperscript{1}Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\ 
+\textsuperscript{2}Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France \\
+\textsuperscript{3}Carnegie Mellon University, Pittsburgh, Pennsylvania, United States}
+}
+
+% \author{
+% \IEEEauthorblockN{Petar Vukmirovi\'c\orcid{0000-0001-7049-6847}}
+% \IEEEauthorblockA{Vrije Universiteit Amsterdam}
+% \and
+% \IEEEauthorblockN{Jasmin Blanchette\orcid{0000-0002-8367-0936}}
+% \IEEEauthorblockA{
+% Vrije Universiteit Amsterdam, \\
+% Max-Planck-Institut f\"ur Informatik, \\
+% Universit\'e de Lorraine}
+% \and
+% \IEEEauthorblockN{Marijn J.H. Heule\orcid{0000-0002-5587-8801}}
+% \IEEEauthorblockA{Carnegie Mellon University}
+% }
+
+% mathcal
+\DeclareFontFamily{OT1}{pzc}{}
+\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
+\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}
+%\DeclareMathAlphabet{\mathcalx}{OMS}{zplm}{m}{n}
+
+%\hyphenation{pre-process-ing in-process-ing}
+
+\begin{document}
+
+% \newcommand\ourparagraph[1]{\noindent{\textbf{\upshape#1.}}}
+\newcommand\ourparagraph[1]{\subsection{#1}}
+
+%% TYPESETTING: hacks
+\newcommand\medrightarrow{\mathrel{{{\color{black}\relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\rightarrow}}}
+\newcommand\medleftarrow{\mathrel{{\color{black}\leftarrow}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex\joinrel{{\color{black}\relbar}}}}
+\newcommand\medleftrightarrow{\mathrel{\leftarrow\kern-1.685ex\rightarrow}}
+\newcommand\Medrightarrow{\mathrel{{{\color{black}\Relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\Rightarrow}}}
+\newcommand\Medleftrightarrow{\mathrel{\Leftarrow\kern-1.685ex\Rightarrow}}
+
+\undef\leftrightarrow %% use \medleftrightarrow instead for iff
+
+\newcommand\scif{\quad\text{if~}}
+\newcommand{\HL}{\ensuremath{\mathrm{HL}}}
+\newcommand{\PHL}{\ensuremath{\mathrm{HL_\mathrm{p}}}}
+\newcommand{\cst}[1]{{\mathsf{#1}}}
+\newcommand{\eq}{\mathbin{\approx}}
+\newcommand{\noteq}{\mathbin{\not\approx}}
+\newcommand{\seq}[1]{\ensuremath{(#1)_{n=1}^{\infty}}}
+\newcommand{\impk}[4]{\ensuremath{#1 \impl^{#3}_{#4} #2}}
+\newcommand{\infname}[1]{\textsc{#1}}
+\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
+\newcommand{\namedsimpsc}[4]{\prftree[d][r]{\relax{\infname{#1}\scif $#4$}}{\strut#2}{\strut#3}}
+
+\newcommand{\impl}{\ensuremath{\medrightarrow}}
+% \newcommand{\bigpimpl}[1]{\ensuremath{\overset{#1}{\rightharpoonup}}}
+% \newcommand{\bigfimpl}[1]{\ensuremath{\overset{#1}{\hookrightarrow}}}
+\newcommand{\bigpimpl}[1]{\hookrightarrow_\mathrm{p}}
+\newcommand{\bigfimpl}[1]{\hookrightarrow}
+\newcommand{\eqlit}[2]{\ensuremath{#1 \eq #2}}
+\newcommand{\neqlit}[2]{\ensuremath{#1 \noteq #2}}
+% \newcommand{\eqneqlit}[2]{\ensuremath{#1 \mathrel{\dot{\eq}} #2}}
+\newcommand{\pospredlit}[1]{{#1}}
+\newcommand{\negpredlit}[1]{\neglit{#1}}
+\newcommand{\neglit}[1]{\neg{\kern.2ex #1}}
+\newcommand{\arbpredlit}[1]{(\neg)\kern.2ex\pospredlit{#1}}
+\newcommand{\cor}{\ensuremath{\mathrel{\lor}}}
+\newcommand{\tuple}[2]{\ensuremath{\vec{#1}_{#2}}}
+\newcommand{\tuplen}[1]{\tuple{#1}{n}}
+\newcommand{\tupleempty}[1]{\tuple{#1}{}}
+\newcommand{\flatres}{\mathbin{\smash{\hbox{\Large$\kern-.1ex\rtimes$}}}}  %% (slightly) asymmetric symbol
+\newcommand{\flatresiter}{\mathbin\leadsto}
+\newcommand{\withoutpred}[2]{\,\overline{\!#1}_{\!#2}}
+\newcommand{\withpred}[2]{#1_{#2}}
+\newcommand{\withpredpos}[2]{#1^+_{#2}}
+\newcommand{\withpredneg}[2]{#1^-_{#2}}
+\newcommand{\binset}[1]{#1{\downarrow}_2}
+\newcommand\NumberOK[1]{#1}
+\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
+\newcolumntype{d}{D{.}{.}{0}}
+\newcommand{\confrep}[2]{\begin{conf}#1\end{conf}\begin{rep}#2\end{rep}}
+
+\definecolor{light-gray}{gray}{0.925}
+\newcommand\MAX[1]{\smash{\setlength{\fboxsep}{.3ex}\colorbox{light-gray}{\ensuremath{\vphantom{('q}{#1}}}}}
+
+\newtheorem{theorem}{Theorem}{\bfseries}{\slshape}
+\newtheorem{example}[theorem]{Example}{\bfseries}{\rmfamily}
+\newtheorem{remark}[theorem]{Remark}{\bfseries}{\rmfamily}
+\newtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
+\newtheorem{lemma}[theorem]{Lemma}{\bfseries}{\slshape}
+\newtheorem{corollary}[theorem]{Corollary}{\bfseries}{\slshape}
+% \newtheorem*{proof}{Proof}{\slshape}{\rmfamily}
+
+\newlength{\myskip}
+\setlength{\myskip}{0.6ex}
+% TYPESETTING: when two environments are next to
+% each other, we need to somehow squeeze them
+\newcommand{\removemyskip}{\vspace{-\myskip}}
+
+\newenvironment{theoremx}% environment name
+{% begin code
+  \vspace{\myskip}
+  \begin{theorem}
+}%
+{\end{theorem}
+ \vspace{\myskip}}% end code
+
+\newenvironment{examplex}% environment name
+{% begin code
+  \vspace{\myskip}
+  \begin{example}
+}%
+{\end{example}
+ \vspace{\myskip}}% end code
+
+\newenvironment{remarkx}% environment name
+{% begin code
+  \vspace{\myskip}
+  \begin{remark}
+}%
+{\end{remark}
+ \vspace{\myskip}}% end code
+
+\newenvironment{definitionx}% environment name
+ {% begin code
+   \vspace{\myskip}
+   \begin{definition}
+ }%
+ {\end{definition}
+  \vspace{\myskip}}% end code
+
+\newenvironment{lemmax}% environment name
+{% begin code
+\vspace{\myskip}
+\begin{lemma}
+}%
+{\end{lemma}
+\vspace{\myskip}}% end code
+
+\newenvironment{corollaryx}% environment name
+{% begin code
+\vspace{\myskip}
+\begin{corollary}
+}%
+{\end{corollary}
+\vspace{\myskip}}% end code
+
+\newenvironment{proof}% environment name
+{% begin code
+  \vspace{\myskip}
+  \textit{Proof:}%
+}%
+{\vspace{\myskip}}% end code
+
+% Stolen from Jasmin's paper
+\newcommand{\qed}{\hfill{\mbox{\ $\IEEEQED$}}}
+
+\maketitle
+\setcounter{footnote}{0}
+
+\newcommand{\ourmodel}{\ensuremath{\mathscr{J}}}
+
+\begin{abstract}
+Optimized SAT solvers not only preprocess the clause set, they also transform it
+during solving as inprocessing. Some preprocessing techniques have been
+generalized to first-order logic with equality. In this \paper, we port
+inprocessing techniques to work with superposition, a leading first-order proof
+calculus, and we strengthen known preprocessing techniques. Specifically, we
+look into elimination of hidden literals, variables (predicates), and blocked
+clauses. Our evaluation using the Zipperposition prover confirms that the new
+techniques usefully supplement the existing superposition machinery.
+\end{abstract}
+
+\section{Introduction}
+\label{sec:introduction}
+
+Automated reasoning tools have become much more powerful in the last few
+decades thanks to procedures such as conflict-driven clause learning (CDCL)
+\cite{MSLM09HBSAT} for propositional logic and superposition
+\cite{bg-94-superposition} for first-order logic with equality. However,
+the effectiveness of these procedures crucially depends on how the input
+problem is represented as a clause set. The clause set
+can be optimized beforehand (\emph{preprocessing}) or during the execution of
+the procedure (\emph{inprocessing}). In this \paper, we lift several preprocessing and
+inprocessing techniques from propositional logic to clausal first-order logic
+and demonstrate their usefulness in a superposition prover.
+
+For many years, SAT solvers have used inexpensive clause simplification
+techniques such as hidden literal and hidden tautology elimination
+\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and failed literal
+detection \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
+first-order logic with equality
+(Sect.~\ref{sec:hidden-literal-based-elimination}). Since the generalization
+involves reasoning about infinite sets of literals, we propose restrictions to
+make them usable.
+
+\emph{Variable elimination}, based on Davis--Putnam resolution \cite{dp-60-dp}, has
+been studied in the context of both propositional logic
+\cite{sp-04-niver,cs-00-zres} and quantified Boolean formulas (QBFs)
+\cite{ab-2004-re}. The basic idea is to resolve all clauses with negative
+occurrences of a propositional variable (i.e., a nullary predicate symbol) against
+clauses with positive occurrences and delete the parent clauses. E\'en and
+Biere \cite{eb-2005-satpreprocess} refined the technique to identify a subset
+of clauses that effectively define a variable and use it to further optimize the
+clause set. This latter technique, \emph{variable elimination by substitution},
+has been an important preprocessor component in many SAT solvers since its
+introduction in 2004.
+
+Specializing second-order quantifier elimination
+\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
+\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
+problems, yielding a technique we call \emph{singular predicate elimination}. We
+extend their work along two axes (Sect.~\ref{sec:predicate-elimination}): We
+generalize E\'en and Biere's refinement to first-order logic, resulting in
+\emph{defined predicate elimination}, and explain how both types of predicate
+elimination can be used during the proof search as inprocessing.
+
+\begin{qle}
+It is well known that clauses containing \emph{pure literals}---predicate
+literals whose root symbol occurs only with one polarity in the problem---can
+be eliminated. We propose a generalization of this technique, under the name
+\emph{quasipure literal elimination}, to allow the elimination of some bipolar
+literals as well (Sect.~\ref{sec:quasipure-literal-elimination}).
+\end{qle}
+
+The last technique we study is \emph{blocked clause elimination} (Sect.\
+\ref{sec:satisfiability-by-clause-elimination}). It is used in both SAT
+\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
+logic has produced good results when used as a preprocessor, especially on
+satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
+blocked clause elimination on satisfiable problems, including using it to
+establish equisatisfiability with an empty clause set or as an inprocessing
+rule. Unfortunately, we find that its use as inprocessing can compromise the
+refutational completeness of superposition.
+
+% Preprocessing has been a crucial part of SAT solvers since 2004, when E\'en and
+% Biere introduced bounded variable elimination by substitution~\cite{BVE}.
+% They based it on Davis--Putnam resolution \cite{dp-60-dp}, which
+% replaces all clauses containing a variable $x$ by all nontautological
+% resolvents on $x$, but showed how to exploit definition information
+% extracted form the clause set to add only a subset of the nontautological
+% resolvents. In this way, many variables can be eliminated without
+% growing the clause set. % (in terms of the number of clauses). ---that comes later --JB
+
+% Preprocessing is generally effective, but it can be expensive. As a compromise,
+% many solvers use inprocessing:\ interleaving the decision procedure with
+% preprocessing steps~\cite{mhb-12-inprocessing}. Top-tier SAT solvers, such as CaDiCaL
+% \cite{xxx} and Kissat \cite{xxx} alternate about every second between solving
+% and inprocessing. In contrast to preprocessing, inprocessing can also be used
+% to simplify the clause set after some important deductions have been made.
+
+% Pre-~and inprocessing are successful beyond propositional logic. For example,
+% lifting variable and clause elimination techniques to quantified Boolean
+% formulas (QBF) helps solve many formulas that are too difficult in practice for
+% decision procedures~\cite{bls-11-bloqqer}. This suggests that richer logics, such as
+% first- and higher-order logic, can also benefit from elimination techniques.
+
+% To improve the success rate of superposition provers, we extend several of
+% preprocessing and inprocessing techniques from propositional to first-order logic,
+% including failed literal detection, hidden literal elimination, hidden
+% tautology elimination
+% (Sect.~\ref{sec:hidden-literal-based-elimination}). These
+% techniques are fairly inexpensive in propositional logic but need to be
+% curtailed in first-order logic because they would otherwise involve reasoning
+% about an infinite set of literals.
+
+% Khasidashvili and Korovin \cite{kk-2016-pe-fol} generalized Davis--Putnam
+% resolution to first-order logic with equality. We call their technique
+% \emph{singular predicate elimination}. We also adapt bounded variable
+% elimination by substitution, yielding \emph{defined predicate elimination}
+% (Sect.~\ref{sec:predicate-elimination}). With some care, we can integrate these
+% techniques with superposition without losing refutational completeness.
+
+% Finally, we study a first-order generalization of blocked clause
+% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satisfiability-by-clause-elimination}).
+% %Unlike the other techniques, it weakens the clause set.
+% If all clauses in the current clause set are blocked, this means that it is
+% satisfiable.
+% %It can be used to show
+% %satisfiability by establishing equisatisfiability between the initial clause
+% %set and the empty clause set.
+% We show how this technique can help superposition provers detect
+% satisfiability early instead of diverging. We also show, however, that
+% interleaving the technique with superposition compromises refutational
+% completeness.
+
+All techniques are implemented in the Zipperposition prover
+(Sect.~\ref{sec:implementation}), allowing us to ascertain their usefulness
+(Sect.~\ref{sec:evaluation}). The best configuration solves \NumberOK{160}
+additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
+\cite{gs-17-tptp}. The raw experimental data are publicly available.%
+\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
+More details, including all the proofs, can be found in a
+technical report \cite{our-report}.}{}
+
+\begin{rep}
+This \paper{} expands on our FMCAD 2021 paper \cite{fmcad}. It includes more
+examples, lemmas, and proofs. It also corrects a mistake in the handling of
+variables in Definition~\ref{def:equality-blocked}.
+\end{rep}
+
+\section{Preliminaries}
+\label{sec:preliminaries}
+
+\ourparagraph{Clausal First-Order Logic}
+\looseness=-1
+Our setting is many-sorted, or many-typed, first-order logic \cite{jg-1987-logic-textbook} with
+interpreted equality and a distinguished type (or sort) $o$.
+Each variable $x$ is assigned a non-Boolean
+type, and each symbol $\cst{f}$ is assigned a tuple $(\tau_1, \ldots, \tau_n,
+\tau)$ where $n \geq 0$, $\tau_i$ are non-Boolean types, and $\tau$ is
+the \emph{result type}. We distinguish between
+\emph{predicate symbols}, with $o$ as the result type, and \emph{function symbols}.
+Nullary function symbols are called \emph{constants}. Terms are either
+variables~$x$ or \confrep{well-typed }{}applications $\cst{f}(t_1, \ldots,
+t_n)$\confrep{, or $\cst{f}$ if $n = 0.$}{, of type $\tau$ where $\cst{f}$ is assigned $(\tau_1, \ldots,
+\tau_n, \tau)$, and each $t_i$ is a term of type $\tau_i$.
+The parentheses are omitted if $n = 0$.} A term is
+\emph{ground} if it contains no variables.
+We assume standard definitions and
+notations for positions, subterms, and contexts \cite{bn-98-tr-and-all-that}.
+We abbreviate a vector $(a_1, \ldots, a_n)$ to $\tuplen{a}$ or $\tuple{a}{}$,
+and write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol $\cst{f}$
+(e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$).
+
+An atom is an equation $\eqlit{s}{t}$ corresponding to an unordered pair $\{s,
+t\}$. A literal is an equation $\eqlit{s}{t}$ or a disequation $\neqlit{s}{t}$.
+For every predicate symbol~$\cst{p}$, \confrep{}{the notation }$\cst{p}(\tuple{s}{})$
+abbreviates $\eqlit{\cst{p}(\tuple{s}{})}{\top}$, and
+$\neglit{\cst{p}(\tuple{s}{})}$ abbreviates
+$\neqlit{\cst{p}(\tuple{s}{})}{\top}$, where $\top$ is a distinguished
+constant of type $o$. We distinguish between \emph{predicate literals}
+$\arbpredlit{\cst{p}(\tuple{s}{})}$ and \emph{functional literals} $\eqlit{s}{t}$,
+where $s$ and $t$ are not of type $o$.
+Given a literal $L$, we overload notation and write $\neglit{L}$ to
+denote its complement. A clause $C$ is a multiset of literals, written as $L_1
+\cor \cdots \cor L_n$ and interpreted disjunctively. Clauses are often defined
+as sets of literals, but superposition needs multisets; with multisets,
+an instance $C\sigma$ always has the same number of literals as $C$, a most
+convenient property. Given a clause set $N$, $\binset{N}$ denotes the subset of
+its binary clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.
+
+\begin{rep}
+A substitution $\sigma$ is a well-typed mapping from
+variables to terms such that the set $\{x \mid \sigma(x) \not= x \}$ is finite.
+%Assuming $x_i \not= t_i$,
+Substitutions are written as $\{x_1 \mapsto t_1,
+\ldots,\allowbreak x_n \mapsto t_n\}$ or $\{ \tupleempty{x} \mapsto \tupleempty{t} \}$.%
+\confrep{}{ A substitution is a \emph{variable renaming} if it is a
+bijection from a set of variables to a set of variables.}
+
+We assume the natural extensions of domain, valuation, interpretation and model
+(as defined by Fitting \cite{mf-1996-fol}) from unsorted to many-sorted
+logic.
+Models we consider are \emph{normal}, i.e., they interpret $\eq$ as an
+equality relation. Usual notions of (un)satisfiability and (in)validity are
+assumed. We write $\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$
+satisfies a clause set $N$, for a variable assignment $\xi$. If $\ourmodel$
+is a model of $N$ (i.e., it satisfies it under every variable assignment), 
+we simply write $\ourmodel \models N.$  Abusing notation,
+we write $M \models N$ to denote that $M$ \emph{entails} $N$, i.e., that every
+model of $M$ is a model of $N.$
+
+A canonical model $\ourmodel$ is a normal model such that for every element $d$
+in one of $\ourmodel$'s domains there exists a ground term $t$ such that
+$\ourmodel$ interprets $t$ as $d$. Canonical models generalize Herbrand models
+to first-order logic with equality: If a clause set is satisfiable, then it
+is satisfiable in a canonical model.
+\end{rep}
+
+\ourparagraph{Superposition Provers}
+%
+Superposition \cite{bg-94-superposition} is a calculus for clausal
+first-order logic that extends ordered resolution
+\cite{bg-01-resolution} with equality reasoning. It is
+refutationally complete: Given a finite, unsatisfiable clause set, it will
+eventually derive the empty clause. It is parameterized by a \emph{selection
+function} that influences which of a clause's literals are eligible as the target
+of inferences. Moreover, it is compatible with the \emph{standard redundancy
+criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.
+
+\looseness=-1
+The redundancy criterion relies on an order $\succ$ that
+compares terms, literals, or clauses.
+The order is used to determine whether clauses can be deleted.
+If $N$ is ground, $C$ can be deleted
+if it is entailed by $\prec$-smaller clauses in $N.$ This definition is lifted
+to nonground sets $N$. The criterion can be used to delete a
+clause that is \emph{subsumed} by another clause (e.g.,
+$\cst{\cst{p}(\cst{a})} \cor \cst{q}$ by $\cst{p}(x)$) or
+to \emph{simplify} a clause $C$ into $C'$, which amounts to adding $C'$ and
+then deleting $C$ as redundant with respect to $N \cup \{C'\}$.
+Subsumption and simplification are the main inprocessing mechanisms available
+to superposition provers. Some provers also implement clause splitting
+\cite{riazanov-voronkov-2001,fietzke-weidenbach-2009,av-2014-avatar}.
+
+Superposition provers saturate the input problem with respect to the calculus's
+inference rules using the \emph{given clause procedure}
+\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
+passive set $\mathcalx{P}$ and an active set $\mathcalx{A}$. All clauses start in
+$\mathcalx{P}$. At each iteration of the procedure's main loop, the prover
+chooses a clause $C$ from $\mathcalx{P}$, simplifies it, and moves it to
+$\mathcalx{A}$. Then all inferences between $C$ and active clauses are performed.
+The resulting clauses are again simplified and put in $\mathcalx{P}$.\confrep{}{
+The provers differ in which clauses are used for simplification: Otter-loop
+\cite{mcw-1997-otter} provers use both active and passive clauses whereas
+DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.}
+
+\section{Hidden-Literal-Based Elimination}
+\label{sec:hidden-literal-based-elimination}
+
+In propositional logic, binary clauses from a clause set $N$ can be used to
+efficiently discover literals $L, L'$ for which the implication $L' \impl L$
+is entailed by $N$'s binary clauses---i.e., $\binset{N} \models L' \impl L$.
+Heule et al.\ \cite{hjb-2011-big-simplification} introduced the concept of \emph{hidden
+literals} to capture such implications.
+
+\begin{definitionx}
+   Given a propositional literal $L$ and a propositional clause set~$N$, the
+   set of \emph{propositional hidden literals} for $L$ and~$N$ is
+   $\PHL(L, N) = \{ L' \mid L' \bigpimpl{N}^\ast L \} \setminus \{ L \}$, where $\bigpimpl{N}$ is
+   defined such that $\neglit{L_1} \bigpimpl{N} L_2$ whenever $L_1 \cor L_2 \in N.$
+   Moreover, $\PHL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n
+   \PHL(L_i, N)$.
+\end{definitionx}
+
+Heule et al.\ used a fixpoint computation, but our definition based on the
+reflexive transitive closure is equivalent. Intuitively, a hidden literal can be added to
+or removed from a clause without affecting its semantics in models of~$N.$
+By eliminating hidden literals from $C$, we simplify it. By adding hidden literals to $C$, we might get a tautology $C'$ (i.e., a valid clause:\ $\models
+C'$), meaning that $\binset{N} \models C$, thereby enabling us to delete $C$.
+Note that $\PHL(L,N)$ is finite for a finite~$N.$
+
+\begin{definitionx}
+   \label{def:hle}\label{def:hte}
+   Given $L' \cor L \cor C \in N,$ if $L' \in
+   \PHL(L, N)$, \emph{hidden literal elimination} (HLE) replaces $N$ by $(N
+   \setminus \{L' \cor L \cor C\}) \cup \{L \cor C\}$.
+   Given $C \in N$, $\{L_1,\dots,L_n\} = \PHL(C, N)$, and $C' = C \cor L_1 \cor
+   \cdots \cor L_n$, if $C'$ is a tautology, \emph{hidden
+   tautology elimination} (HTE) replaces $N$ by $N \setminus \{C\}$.
+\end{definitionx}
+
+\begin{theoremx}
+   The result of applying HLE or HTE to a clause set $N$ is
+   equivalent to $N$.
+\end{theoremx}
+
+\begin{proof}
+For HLE, if $L' \in \PHL(L, N)$,  $\binset{N} \models \neglit{L'} \cor L$.
+Then, subsumption resolution yields shortened clause $L \cor C'$ from
+Definition~\ref{def:hle}. For HTE, it can be shown that $N' \models C$ if and
+only if $C \cor L'$, where $L' \in \PHL(C, N)$. By transitivity of equivalence,
+we get the desired result.
+\qed
+\end{proof}
+
+We generalize hidden literals to first-order logic with equality by
+considering substitutivity of variables as well as congruence of equality.
+
+\begin{definitionx}
+\label{def:hl-fo}
+Given a literal $L$ and a clause set $N$,
+the set of \emph{hidden literals} for $L$ and $N$
+is $\HL(L,N) = \{ L' \mid L' \bigfimpl{N}^\ast L \} \setminus \{ L \}$, where
+$\bigfimpl{N}$ is defined so that
+(1)~$\neglit{L'\sigma} \bigfimpl{N} L\sigma$ if
+$L' \cor L \in N$ and $\sigma$ is a substitution;
+(2)~$\eqlit{s}{t} \bigfimpl{N} \eqlit{u[s]}{u[t]}$ for all terms $s, t$ and
+contexts $u[\phantom{\cdot}]$; and
+(3)~$\neqlit{u[s]}{u[t]} \bigfimpl{N} \neqlit{s}{t}$ for all terms $s, t$ and
+contexts $u[\phantom{\cdot}]$.
+Moreover, $\HL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n \HL(L_i, N)$.
+\end{definitionx}
+
+\looseness=-1
+The generalized definition also enjoys the key property that $L'
+\in \HL(L, N)$ implies $\binset{N} \models L' \impl L$. However, $\HL(L, N)$
+may be infinite even for predicate literals; for example,
+$\pospredlit{\cst{p}(\cst{f}^i(x))} \in \HL(\pospredlit{\cst{p}(x)}, \{\pospredlit{\cst{p}(x)} \cor
+\negpredlit{\cst{p}(\cst{f}(x))} \})$ for every $i$.
+
+Based on Definition \ref{def:hl-fo}, we can generalize hidden literal elimination
+and support a related technique:
+%
+\begin{align*}
+   \vcenter{\namedsimp{HLE}{L' \cor L \cor C}{L \cor C}}
+   & \scif {L' \in \HL(L,N)} \\[.5\jot]
+   \vcenter{\namedsimp{\rlap{FLE}\phantom{HLE}}{L \cor C}{C}}
+   & \scif {L',\neglit{L'} \in \HL(\neglit{L}, N)}
+\end{align*}
+%
+Double lines denote \emph{simplification rules}: When the premises appear in
+the clause set, the prover can use the redundancy criterion to replace
+them by the conclusions. The second rule is called \textit{failed literal
+elimination},
+inspired by the SAT technique of asserting $\neglit{L}$ if $L$ is a \textit{failed literal} \cite{jwf-1995-fld}.
+It is easy to see that rule \infname{HLE} is sound.
+From $L' \in \HL(L,N)$ we have $N \models L' \impl L$ (i.e., $\neglit{L'} \cor L$).
+Performing subsumption resolution \cite{bg-01-resolution} between $L'
+\cor L \cor C$ and $\neglit{L'} \cor L$ yields the conclusion, which is
+therefore entailed by $N.$ For \infname{FLE}, the condition $L',
+\neglit{L'} \in \HL(\neglit{L},N)$ means that $\binset{N} \models \{\neglit{L'} \cor
+\neglit{L}{,}\; L' \cor \neglit{L}\} \models \neglit{L}$.
+
+\begin{examplex}
+   \label{example:hle}
+   Consider the clause set $N = \{ \cst{p}(x) \cor \neglit{\cst{p}(\cst{f}(x))}{,}\;
+   \cst{p}(\cst{f}(\cst{f}(x))) \cor \eqlit{\cst{a}}{\cst{b}} \}$ and the clause
+   $C = \neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \cor \cst{p}(x)$. The first
+   clause in $N$ induces $\cst{p}(\cst{f}(x)) \bigfimpl{N} {\cst{p}(x)}$,
+   $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N} \cst{p}(\cst{f}(x))$, and
+   hence $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N}^\ast  \cst{p}(x)$.
+   Together with the second clause in $N$, it can be used to derive $\neqlit{\cst{a}}{\cst{b}}
+   \bigfimpl{N}^\ast  \cst{p}(x)$. Finally, using rule~(3) of Definition
+   \ref{def:hl-fo}, we derive $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})}
+   \bigfimpl{N}^\ast  \cst{p}(x)$---that is,
+   $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \in \HL(\cst{p}(x), N)$. This
+   allows us to remove $C$'s first literal using \infname{HLE}.
+\end{examplex}
+
+Two special cases of \infname{HLE} exploit equality congruence as embodied by
+conditions (2)~and~(3) of Definition~\ref{def:hl-fo} without requiring to
+compute the $\HL$~set:
+\begin{align*}
+  %% TYPESETTING: \,'s below
+  \namedsimp{CongHLE$^+$}{\eqlit{s}{t} \cor \eqlit{u[s]}{u[t]} \cor C}{\eqlit{u[s]}{u[t]} \cor C}\,\, \\[.5\jot]
+  \namedsimp{CongHLE$^-$}{\neqlit{s}{t} \cor \neqlit{u[s]}{u[t]} \cor C}{\neqlit{s}{t} \cor C}
+\end{align*}%
+
+Hidden literals can be combined with unit clauses~$L'$ to remove more literals:
+\[\namedsimpsc{UnitHLE}{L' \quad L \cor C}{L' \quad C}{L'\sigma \in \HL(\neglit{L},N)}
+%\text{~for some~}\sigma -- %%% Not necessary in a rule -- all variables are universally quantified. --JB
+\]
+%
+Given a unit clause $L' \in N$, the rule uses it to discharge $L'\sigma$ in
+$N \models L'\sigma \impl \neglit{L}$. As a result, we have $N \models
+\neglit{L}$, making it possible to remove $L$ from $L \cor C$.
+
+% Kiesl and Suda described first-order generalization of {\it asymmetric literals}
+% \cite{ks-2017-unif-principle}: literal $L\sigma$ is asymmetric for a clause $C$
+% and a clause set $N$ if there exists a clause $D \cor \neglit{L} \in N \setminus
+% \{C \}$ such that $D\sigma \subseteq C$ for some substitution $\sigma$.
+
+\begin{examplex}
+   \label{example:unithle}
+   \looseness=-1
+   Consider the clause set $N = \{\cst{p}(x) \cor \cst{q}(\cst{f}(x)){,}\;
+   \neglit{\cst{q}(\cst{f}(\cst{a}))} \cor \eqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}{,}\;
+   \neqlit{\cst{f}(x)}{\cst{g}(y)}\}$ and the clause $C = \neglit{\cst{p}(\cst{a})} \cor
+   \neglit{\cst{q}(\cst{b})}$. The first clause in $N$ induces
+   $\neglit{\cst{q}(\cst{f}(\cst{a}))} \bigfimpl{N} \cst{p}(\cst{a})$, whereas
+   the second one induces $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})} \bigfimpl{N}
+   \neglit{\cst{q}(\cst{f}(\cst{a}))}$. Thus, we have $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}
+   \bigfimpl{N}^\ast \cst{p}(\cst{a})$---that is, $\neqlit{\cst{f}(\cst{b})}{\cst{f}(\cst{c})} \in
+   \HL(\cst{p}(\cst{a}), N)$. By applying the substitution $\{x \mapsto \cst{b}{,}\allowbreak\; y
+   \mapsto \cst{c}\}$ to the third clause in $N$, we can fulfill the conditions of
+   \infname{UnitHLE} and remove $C$'s first literal.
+\end{examplex}
+
+Next, we generalize hidden tautologies to first-order logic.
+
+\begin{definitionx}
+   A clause $C$ is a \emph{hidden tautology} for a clause set $N$ if there
+   exists a finite set $\{L_1,\dots,L_n\} \subseteq \HL(C,N)$ such that
+   $C \cor L_1 \cor \cdots \cor L_n$ is a tautology.
+\end{definitionx}
+\removemyskip
+\begin{examplex}
+   \label{example:hidden-tautologies-completeness}
+   \looseness=-1
+   In general, hidden tautologies are not redundant and cannot be deleted during saturation.
+   Consider the unsatisfiable set $N = \{ \negpredlit{\cst{a}}{,}\;
+   \negpredlit{\cst{b}}{,}\; \pospredlit{\cst{a}} \cor
+   \pospredlit{\cst{c}}{,}\allowbreak\; \pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}
+   \}$, the order $\cst{a} \prec
+   \cst{b} \prec \cst{c}$, and the empty selection function. The
+   only possible superposition inference from $N$ is between the last two clauses,
+   yielding the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
+   (after simplifying away $\neqlit{\top}{\top}$), which is entailed by the larger
+   clauses $\pospredlit{\cst{a}} \cor \pospredlit{\cst{c}}$ and
+   $\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}$. If this clause is removed,
+   the prover could enter an infinite loop, forever generating and deleting the
+   hidden tautology\confrep{}{ and never getting the opportunity to derive the
+   empty clause}.
+\end{examplex}
+
+\begin{rep}
+   \newcommand{\aset}{\mathcalx{A}}
+   \newcommand{\pset}{\mathcalx{P}}
+
+   In practice, most provers use a variant of the given clause procedure.
+   Removing hidden tautologies breaks the invariant of the procedure that all
+   inferences between clauses in $\aset$ are redundant. The end result is not
+   that the prover diverges, but that it terminates without deriving empty
+   clause.
+
+   To observe this, assume the setting as in Example
+   \ref{example:hidden-tautologies-completeness}, and let $\pset=N$ and
+   $\aset=\emptyset$. After moving the first three clauses from $\pset$ to
+   $\aset$ ($\aset = \{ \negpredlit{\cst{a}}, \negpredlit{\cst{b}}, \pospredlit{\cst{a}} \cor
+   \pospredlit{\cst{c}} \}$, $\pset = \{\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}\}$), 
+   no inferences are possible, and no new clauses are added to $\pset$. After the last clause
+   is moved to $\aset$, the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
+   is produced. If it is deleted, the prover terminates with the unsatisfiable set $\aset$,
+   but does not derive the empty clause. 
+\end{rep}
+
+To delete hidden tautologies during saturation, the prover could check that all
+the relevant clause instances encountered along the computation of $\HL$ are
+$\prec$-smaller than a given hidden tautology. However, this would be expensive
+and seldom succeed, given that superposition creates lots of nonredundant
+hidden tautologies. Instead, we propose to simplify hidden tautologies using the
+following rules:
+%
+%\begin{linenomath*}
+% \begin{align*}
+%    %% TYPESETTING: \,'s below
+%    \namedsimpsc{HTR}{L \cor L' \cor C}{L \cor L'}{\neglit{L'} \in \HL(L, N)} &\\
+%    \namedsimpsc{FLR}{L \cor C}{L}{L',\neglit{L'} \in \HL(L, N)} &
+% \end{align*}
+%
+\[\begin{aligned}[t]
+   \vcenter{\namedsimp{HTR}{L \cor L' \cor C}{L \cor L'}}
+   & \scif {\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot} \\[.5\jot]
+   \vcenter{\namedsimp{\rlap{FLR}\phantom{HTR}}{L \cor C}{L}}
+   & \scif {L',\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot}
+\end{aligned}\]
+%
+
+%\end{linenomath*}
+%
+We call these techniques \emph{hidden tautology reduction} and \emph{failed literal
+reduction}, respectively.
+% \infname{FLD} is inspired by the \confrep{}{eponymous }SAT
+% approach of asserting $L$ if $\neglit{L}$ is determined to be a failed literal
+% \cite{jwf-1995-fld}.
+%
+Both rules are sound.
+%
+As with hidden literals, unit clauses~$L'$ can be exploited:
+\[\namedsimpsc{UnitHTR}{L' \quad L \cor C}{L' \quad L}{L'\sigma \in \HL(L,N) \text{~and~} C \not= \bot}
+%\text{~for some~}\sigma
+\]
+
+We give the simplification rules above\confrep{}{ (for hidden literal
+elimination, hidden tautology reduction,
+failed literal detection, and their variants)} the collective name of
+\emph{hidden-literal-based elimination} (HLBE).
+Yet another use of hidden literals is for \emph{equivalent literal
+substitution} \cite{hjb-2010-cl-elim}: If both $L' \in \HL(L, N)$ and $L
+\in \HL(L', N)$, we can often simplify $L'\sigma$ to $L\sigma$ in $N$ if
+$L'\sigma \succ L\sigma$. We want to investigate this~further.
+
+% \looseness=-1
+% Provers based on the DISCOUNT loop use only active clauses for
+% simplification. If we restrict our attention to a finite subset of hidden
+% literals, we can organize them into data structures that support efficient
+% lookup of a hidden literal. Efficient implementation of these data structures
+% allows to use all passive clauses for simplification, strengthening DISCOUNT
+% loop simplification machinery.
+
+\begin{theoremx}
+The rules \infname{HLE}, \infname{FLE}, \infname{CongHLE$^+$},
+\infname{Cong\allowbreak HLE$^-$}, \infname{UnitHLE}, \infname{HTR},
+\infname{FLR}, and \infname{UnitHTR} are sound simplification rules.
+\end{theoremx}
+
+\begin{rep}
+\begin{proof}
+It is easy to see that the deleted premises are entailed by the conclusions that
+replace them and that the conclusions' instances are $\prec$-smaller than the
+premises' instances, as required by the redundancy criterion. It remains to
+check soundness.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{HLE}:\enskip
+We have $N{,}\> L' \models L$ by the side condition and must show
+$N{,}\> L' \lor L \lor C \models L \lor C$. Let
+$\ourmodel \models N{,}\> L' \lor L \lor C$. If $\ourmodel \models L'$, then
+we also have $\ourmodel \models L$ thanks to the side condition and hence
+$\ourmodel \models L \lor C$. Otherwise, we have $\ourmodel \models L \lor C$,
+which is exactly what we need to show.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{FLE}:\enskip
+We have $N{,}\> L \models L'$ and $N, L \models \lnot L'$ by the side condition.
+If $\ourmodel \models N{,}\> L$, then both $\ourmodel \models L'$ and
+$\ourmodel \models \lnot L'$, an absurdity.
+Otherwise, we have $\ourmodel \models C$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{CongHLE$^+$}, \infname{CongHLE$^-$}:\enskip
+Obvious by congruence of equality.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{UnitHLE}:\enskip
+We have $N{,}\> L \models \lnot L'\sigma$ by the side condition. If $\ourmodel
+\models N{,}\> L$, then $N \models \lnot L'\sigma$. But since $L' \in N$, this is
+an absurdity. Otherwise, we have $\ourmodel \models C$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{HTR}:\enskip
+We have $N{,}\> \lnot L' \models L$ by the side condition. If either $\ourmodel
+\models L$ or $\ourmodel \models L'$, the desired result follows directly.
+Otherwise, from $\ourmodel \models \lnot L'$ we also have $\ourmodel \models L$
+thanks to the side condition, contradicting $\ourmodel \models \lnot L$.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{FLR}:\enskip
+We have $N{,}\> L' \models L$ and $N{,}\> \lnot L' \models L$ by the side condition.
+Hence $N \models L$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{UnitHTR}:\enskip
+We have $N{,}\> L'\sigma \models L$. Since $L' \in N$, we have $N \models L$,
+as desired.
+\qed
+\end{proof}
+\end{rep}
+
+\section{Predicate Elimination}
+\label{sec:predicate-elimination}
+
+For propositional logic, variable elimination \cite{eb-2005-satpreprocess} is
+one of the main preprocessing and inprocessing techniques. Following Gabbay and Ohlbach's ideas
+\cite{go-1992-so-pred-elim}, Khasidashvili and Korovin \cite{kk-2016-pe-fol}
+generalized variable elimination to first-order logic with equality and
+demonstrated that it is effective as a preprocessor. We propose an improvement
+that makes this applicable in more cases and show that, with a minor
+restriction, it can be integrated in a superposition prover
+without compromising its refutational completeness.
+
+\maybesubsection{Singular Predicates}
+
+Khasidashvili and Korovin's preprocessing technique removes singular predicates
+(which they call ``non-self-referential predicates'') from the problem using
+so-called flat resolution.
+
+\begin{definitionx}
+   A predicate symbol is called \emph{singular}\confrep{ (or ``non-self-referential'')}{}
+   for a clause set $N$ if it occurs at most once in every clause contained in
+   $N.$
+\end{definitionx}
+\removemyskip
+\begin{definitionx}
+   \label{def:flat-res}
+   Let $C = \pospredlit{\cst{p}(\tuplen{s})} \cor C'$ and $D =
+   \negpredlit{\cst{p}(\tuplen{t})} \cor D'$ be clauses with no variables in
+   common. The clause $\neqlit{s_1}{t_1} \cor \cdots \cor \neqlit{s_n}{t_n}
+   \cor C' \cor D'$ is a \emph{flat resolvent} of $C$ and $D$ on $\cst{p}$.
+\end{definitionx}
+
+% Predicate elimination saturates with respect to flat resolvents, removing all
+% clauses containing $\cst{p}$. Given two (possibly identical) clause sets $M, N$,
+% the resolved set $M \flatres_{\!\cst{p}} N$, defined below, repeatedly draws
+% one inference partner from $M$ and the other one from $N$.
+
+Given two (possibly identical) clause sets $M, N$, predicate elimination
+iteratively replaces clauses from $N$ containing the symbol $\cst{p}$ with all flat
+resolvents against clauses in $M$. Eventually, it yields a
+set with no occurrences of $\cst{p}$.
+
+\begin{definitionx}
+\label{def:flat-res-set}
+   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
+   $M$. Let $\flatresiter$ be the following relation on clause set pairs and clause sets:
+   \begin{enumerate}
+   \item $(M{,}\> \{\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'\} \uplus N) \flatresiter (M{,}\; N' \cup N)$ if
+   $N'$ is the set that consists of all clauses (up to variable renaming) that are flat resolvents
+   with $\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'$ on $\cst{p}$ and a clause from $M$ as premises.
+   The premises' variables are renamed apart.
+
+   \smallskip
+   \item $(M, N) \flatresiter N$ if $N$ has no occurrences of $\cst{p}$.
+   \end{enumerate}
+   \noindent  The \emph{resolved set} $M \flatres_{\!\cst{p}} N$ is the clause set $N'$ such that $(M,N) \flatresiter^* N'$.
+   % obtained by the following
+   % procedure: Set $N' = N$. Then repeatedly (1)~choose a clause $D \in N'$
+   % containing $\cst{p}$, (2)~compute all flat resolvents involving $D$ and
+   % partners from $M$, renaming variables apart, and (3)~replace $D$ with these
+   % flat resolvents in $N'$. Stop when $N'$ contains no more occurrence of
+   % $\cst{p}$.
+\end{definitionx}
+
+\pagebreak[2]
+
+\begin{rep}
+\begin{lemmax}
+   \label{lem:flat-res-set-termination-confluence}
+   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
+   $M$. The resolved set $N'$ is reached in a finite number of $\flatresiter$ steps,
+   and it is unique up to variable renaming.
+\end{lemmax}
+\begin{proof}
+   To show $\flatresiter$ is terminating we use the following ordinal measure on
+   clause sets: $\nu(\{D_1,\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus
+   \cdots \oplus \omega^{\nu(D_n)}$, where $\nu(D)$ is the number of
+   occurrences of $\cst{p}$ in $D$, $\omega$ is the first infinite ordinal, and
+   $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
+   For every transition
+   $(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have $\nu(N) > \nu(N')$
+   because $\omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'|$.
+   Eventually, a state $(M,N')$ with $\nu(N')=\omega^0\cdot n$ is reached.
+   Then, we apply the second rule of Definition \ref{def:flat-res-set} to obtain the resolved set $N'$.
+
+   To show that $N'$ is unique, i.e., $\flatresiter$ is confluent, it suffices
+   to show (since $\flatresiter$ is terminating and Newmann's lemma applies
+   \cite{bn-98-tr-and-all-that}) that $\flatresiter$ is locally confluent. In
+   other words, whenever $(M,N) \flatresiter (M, N_1)$ and $(M,N) \flatresiter
+   (M, N_2)$ there exists $N'$ such that $(M, N_1) \flatresiter (M, N')$ and
+   $(M, N_2) \flatresiter (M,N')$.
+
+   There are two main sources of nondeterminism of $\flatresiter$: The choice of $C
+   \in N$ and the choice of the literal in $C$ to act on. Let us focus on the choice of
+   $C$ in $N$; the same discussion applies for the choice of literal in $C$.
+
+   Let $N = \{C_1\} \uplus \{C_2\} \uplus N'$, where $C_1$ and $C_2$ are
+   clauses with occurrences of $\cst{p}$. Then, $(M, \{C_1\} \uplus \{ C_2 \cup
+   N' \}) \flatresiter (M, N'_1 \cup \{ C_2 \cup N'\})$ and $(M, \{C_2\} \uplus
+   \{C_1 \cup N' \}) \flatresiter (M, N'_2 \cup \{ C_1 \cup N' \})$ where $N'_1$ and
+   $N'_2$ are sets of corresponding resolvents. Both of $\flatresiter$ steps can
+   be joined (up to variable renaming) to $(M, N'_1 \cup N'_2 \cup N')$, showing
+   that $\flatresiter$ is locally confluent.\qed
+\end{proof}
+\end{rep}
+
+\begin{conf}
+The relation $\flatresiter$ is confluent up to variable renaming. Thanks to
+the singularity constraint on $M$, it
+also terminates on finite sets because the following ordinal measure decreases:
+$\nu(\{D_1,\allowbreak\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus \cdots \oplus
+\omega^{\nu(D_n)}$, where $\nu(D)$ counts the occurrences of $\cst{p}$ in
+$D$, $\omega$ is the first infinite ordinal,
+and $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
+For every transition
+$(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have
+$\nu(\{C\}) = \omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'| = \nu(N')$.
+\end{conf}
+
+Next, it is useful to partition
+clause sets into subsets based on the presence and polarity of a singular
+predicate.
+
+\begin{definitionx}
+   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
+   $N.$ Let $\withpredpos{N}{\cst{p}}$ consist of all clauses of the form $\pospredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
+   let $\withpredneg{N}{\cst{p}}$ consist of all clauses of the form $\negpredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
+   let $\withpred{N}{\cst{p}} = \withpredpos{N}{\cst{p}} \cup \withpredneg{N}{\cst{p}}$,
+   and let $\withoutpred{N}{\cst{p}} = N \setminus \withpred{N}{\cst{p}}$.
+\end{definitionx}
+\removemyskip
+\begin{definitionx}
+   \label{def:pred-elim}
+   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
+   $N.$
+   \emph{Singular predicate elimination} (SPE) of $\cst{p}$ in $N$ replaces $N$ by
+   $\withoutpred{N}{\cst{p}} \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
+\end{definitionx}
+
+The result of SPE is satisfiable if and only if $N$ is satisfiable
+\cite[Theorem~1]{kk-2016-pe-fol}, justifying SPE's use in a preprocessor.
+However, eliminating singular predicates aggressively can dramatically increase
+the number of clauses. To prevent this, Khasidashvili and Korovin suggested to
+replace $N$ by $N'$ only if $\lambda(N') \leq \lambda(N)$
+and $\mu(N') \leq \mu(N)$, where
+$\lambda(N)$ is the number of literals in~$N$ and
+$\mu(N)$ is the sum for all clauses $C \in
+N$ of the square of the number of distinct variables in $C$.
+
+Compared with what modern SAT solvers use, this
+criterion is fairly restrictive. We relax it to make it possible to
+eliminate more predicates, within reason. Let $K_\mathrm{tol} \in \mathbb{N}$ be a tolerance
+parameter. A predicate elimination
+step from $N$ to $N'$ is allowed if
+$\lambda(N') < \lambda(N) + K_\mathrm{tol}$ or
+$\mu(N') < \mu(N)$ or
+$|N'| < |N| + K_\mathrm{tol}$.
+\confrep{}{(A refinement, which we want to try out in future work, would be to gradually
+increment the tolerance $K_\mathrm{tol}$, as is done in some SAT solvers.)}
+
+\maybesubsection{Defined Predicates}
+
+SPE is effective, but an important refinement has not yet been adapted to
+first-order logic:\ variable elimination by substitution. E{\'{e}}n and Biere
+\cite{eb-2005-satpreprocess} discovered that a propositional variable~$\cst{x}$ can be
+eliminated without computing all resolvents if it is expressible as an
+equivalence $\cst{x} \medleftrightarrow \varphi$, where $\varphi$, the ``gate,''
+is an arbitrary formula that does not reference~$\cst{x}$.
+They \confrep{partition}{extract from} a set $N$
+\confrep{}{of propositional clauses }into a definition set $G$, essentially the
+clausification of $\cst{x} \medleftrightarrow \varphi$, and
+$R = \withpred{N}{\cst{p}} \setminus G$, the remaining
+clauses containing~$\cst{p}$. To eliminate $\cst{x}$ from $N$ while
+preserving satisfiability, it suffices to resolve clauses from $G$ against
+clauses from $R$, effectively substituting $\varphi$ for $\cst{x}$ in $R$.
+Crucially, we do not need to resolve pairs of clauses from $G$
+or pairs of clauses from $R$.
+We generalize this idea to first-order logic.
+
+\begin{definitionx}
+   \label{def:definition}
+   \looseness=-1
+   Let $G$ be a clause set, $\cst{p}$ be a predicate symbol, and
+   $\tupleempty{x}$ be distinct variables.
+%
+   The set $G$ is a \emph{definition set} for $\cst{p}$ if (1)~$\cst{p}$
+   is singular for $G$, (2)~$G$ consists of clauses of the form
+   $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming), (3)~the
+   variables in $C'$ are all among $\tupleempty{x}$, (4)~all clauses in $\withpredpos{G}{\cst{p}}
+   \flatres_{\!\cst{p}} \withpredneg{G}{\cst{p}}$ are tautologies, and
+   (5)~$E(\tupleempty{\cst{c}})$ is unsatisfiable, where
+   the \emph{environment} $E(\tupleempty{x})$ consists of all subclauses $C'$ of any
+   $\arbpredlit{\cst{p}(\tupleempty{x})} \mathbin{\cor} C' \in G$ and
+   $\tupleempty{\cst{c}}$ is a tuple of distinct fresh constants substituted
+   in for $\tupleempty{x}$.
+\end{definitionx}
+
+A definition set $G$ corresponds intuitively to a definition by
+cases in mathematics---e.g.,
+\[\cst{p}(\tuple{x}{}) =
+\begin{cases} \top & \mathrm{if}~\varphi(\tuple{x}{}{}) \\[-\jot]
+\bot & \mathrm{if}~\psi(\tuple{x}{})\end{cases}\]
+Part~(4) states that the case conditions are mutually exclusive
+(e.g., $\neglit{\varphi(\tuple{x}{})} \lor \neglit{\psi(\tuple{x}{})}$),
+and part~(5) states that they are exhaustive
+(e.g., $\nexists\tuple{\cst{c}}.\; \neglit{\varphi(\tuple{\cst{c}}{})} \land \neglit{\psi(\tuple{\cst{c}}{})}$).
+%
+Given a quantifier-free formula $\cst{p}(\tuple{x}{}) \medleftrightarrow
+\varphi(\tuple{x}{})$ with distinct variables $\tuple{x}{}$ such that
+$\varphi(\tuple{x}{})$ does not contain $\cst{p}$, any reasonable
+clausification algorithm would produce a definition set for $\cst{p}$.
+
+\begin{examplex}
+   \label{example:tautologies}
+   Given the formula $\cst{p}(x) \medleftrightarrow \cst{q}(x) \mathrel\land
+   (\cst{r}(x) \cor \cst{s}(x))$, a standard clausification algorithm
+   \cite{nw-01-small-cnf} produces $\{ \negpredlit{\cst{p}(x)} \cor
+   \pospredlit{\cst{q}(x)}{,}\; \negpredlit{\cst{p}(x)} \cor
+   \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
+   \pospredlit{\cst{p}(x)} \cor \negpredlit{\cst{q}(x)} \cor
+   \negpredlit{\cst{r}(x)}{,}\; \pospredlit{\cst{p}(x)} \cor
+   \negpredlit{\cst{q}(x)} \cor \negpredlit{\cst{s}(x)}\}$, which qualifies
+   as a definition set for $\cst{p}$.
+\end{examplex}
+
+Definition sets generalize E{\'{e}}n and Biere's
+gates. They can be recognized syntactically for formulas such as
+$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigvee_{\!i} \cst{q}_i(\tuple{s_i}{})$ or
+$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigwedge_i \cst{q}_i(\tuple{s_i}{})$,
+or semantically: Condition~(4) can be checked
+using the congruence closure algorithm, and condition~(5) amounts to
+a propositional unsatisfiability check.
+
+The key result about propositional gates carries over to definition sets.
+
+\begin{definitionx}
+   \looseness=-1
+   Let $N$ be a clause set, $\cst{p}$ be a predicate symbol,
+   $G \subseteq N$ be a definition set
+   for $\cst{p}$, and $R = \withpred{N}{\cst{p}}
+   \setminus G$. \emph{Defined predicate elimination} (DPE) of $\cst{p}$ in $N$ replaces $N$ by
+   $\withoutpred{N}{\cst{p}} \cup
+%%% @PETAR: Double-check the supression of _p below. --JB
+   (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}} \confrep{\withpred{R}{\cst{p}}}{R})$.
+\end{definitionx}
+
+\newcommand{\mm}{\ourmodel}
+\begin{rep}
+   \begin{lemmax}
+   \label{lem:unsat-fresh-consts}
+   Let $N(\tupleempty{x})$ be a clause set such that the variables of all
+   clauses in it are among the argument $n$-tuple $\tupleempty{x}$, and let
+   $\tupleempty{\cst{c}}$ be an $n$-tuple of distinct fresh constants. If
+   $N(\tupleempty{\cst{c}})$ {\upshape(}i.e.,
+   $N(\tupleempty{x})\{\tupleempty{x}\mapsto\tupleempty{\cst{c}}\}${\upshape)} is
+   unsatisfiable, then for every interpretation $\mm$ and valuation $\xi$,
+   $\mm \not\models_\xi N$.
+   \end{lemmax}
+   \begin{proof}
+      We show the contrapositive. Assume that for some $\mm$ and $\xi$,  $\mm \models_\xi N(\tupleempty{x})$.
+      Then let $\ourmodel'$
+      be a model that assigns each $\cst{c}_i$ the interpretation of $x_i$ under $\mm$ and $\xi$, and otherwise
+      coincides with $\ourmodel$. We obtain $\ourmodel' \models N(\tupleempty{\cst{c}})$.\qed
+   \end{proof}
+
+   \begin{lemmax}
+      \label{lem:flat-res-set-step-satisfiability}
+      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }an arbitrary clause
+      set. If $(G, N) \flatresiter (G, N')$ then $G \cup N$ and $G \cup N'$ are
+      equivalent.
+   \end{lemmax}
+   \begin{proof}
+      Since flat resolution is sound, the nontrivial direction is to show that a
+      model $\mm$ of the set $G \cup N'$ is also a model of $G \cup N$. As the only clause in $N
+      \setminus N'$ is $C = \arbpredlit{\cst{p}(\tuple{s}{n})} \cor C'$ on which
+      the $\flatresiter$ step is performed, we must show $\mm \models C$.
+      % Without
+
+      % Let $\mm$ be canonical model of $G \cup N'$ \cite[Theorem~9.5.2]{mf-1996-fol} and let $C
+      % \in N \setminus N'$. We need to show that $\mm \models C$---which is to say,
+      % for all variable interpretation $\xi$, we have $\mm \models_\xi C$. To
+      % lighten notations, we will assume $C$ is ground and ignore $\xi$. Canonical
+      % models make this step possible.
+
+      % First, $C$ must have the form $\arbpredlit{\cst{p}(\tuple{s}{})} \cor C'$,
+      % where $\tuple{s}{}$ is a tuple of ground terms. 
+      Without loss of generality,
+      we assume that the leading literal of $C$ is positive.
+      Towards a contradiction, assume $\xi$ is a valuation such that $\mm \not\models_\xi C$. Then, $\mm \not\models_\xi
+      \pospredlit{\cst{p}(\tuple{s}{n})}$. 
+      % Let $\xi'$ be a valuation that
+      % assigns every instance $D = \pospredlit{\cst{p}(\tupleempty{s})} \cor D'$ of a clause in
+      % $G$, since $\mm \not\models \pospredlit{\cst{p}(\tuple{s}{})}$, we have $\mm
+      % \models D'$.
+      Consider an arbitrary clause $D = \pospredlit{\cst{p}}(\tuplen{x}) \cor D'
+      \in \withpredpos{G}{\cst{p}}$ and a valuation $\xi'$, which assigns each $x_i$ the
+      interpretation of $s_i$ under $\mm$ and $\xi$. As $\mm \not\models_{\xi'}
+      \pospredlit{\cst{p}}(\tuple{x}{n})$ and $\mm \models G$, then $\mm \models_{\xi'} D'$ for every
+      such clause $D$.
+   %
+      However, by part (5) of Definition~\ref{def:definition} and by Lemma
+      \ref{lem:unsat-fresh-consts}, $\mm \not\models_{\xi'} E(\tuplen{x})$, where $E(\tuplen{x})$
+      is the environment associated with the definition set $G$.
+      Therefore, there must exist a clause $D =
+      \negpredlit{\relax{\cst{p}(\tuplen{x})}} \cor D'$ in $\withpredneg{G}{\cst{p}}$ such
+      that $\mm \not\models_\xi D'$.
+
+      Now consider the flat resolvent of $C$ and $D$ on~$\cst{p}$: $R =
+      \neqlit{x_1}{s_1} \cor \cdots \cor \neqlit{x_n}{s_n} \cor C' \cor D'$.
+      Let $\zeta$ be a valuation coinciding with $\xi$ on variables of $C$
+      and with $\xi'$ on $\tuplen{x}$.
+      Clearly, $\mm \not\models_\zeta R$. Yet, $R \in N'$, and as $\mm \models N'$, we reach a contradiction.\qed
+   \end{proof}
+
+   \begin{lemmax}
+      \label{lem:flat-res-set-last-step-satisfiability}
+      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }a clause set with no occurrences of $\cst{p}$.
+      Then $G \cup N$ is satisfiable if and only if $N$ is satisfiable.
+   \end{lemmax}
+   \begin{proof}
+      The nontrivial direction is to show that if $N$ is satisfiable, $G \cup N$ is as well.
+      %
+      Let $\mm$ be a model of $N$. We construct a model $\mm'$ of $G$ over
+      the same universe as $\mm$. For any atom $A$ such that $\cst{p}$ does not
+      occur in $A$ and for every $\xi$, we set $\mm' \models_\xi A$ if and only if
+      $\mm \models_\xi A$. For any clause $\pospredlit{\cst{p}(\tuplen{x})} \cor
+      C' \in G$ and any assignment $\xi$ such that $\mm \not\models_\xi C'$, we
+      define $\mm'$ so that $\mm'\models_\xi \cst{p}(\tuplen{x})$. By
+      construction, $\mm' \models \withpredpos{G}{\cst{p}} \cup N$. It remains to show that $\mm' \models
+      \withpredneg{G}{}$.
+
+      Let $C = \negpredlit{\cst{p}(\tuplen{x})} \cor C' \in G$ and let $\xi$ be
+      an arbitrary assignment. Towards a contradiction, assume
+      $\mm'\not\models_\xi C$, and consequently $\mm' \models_\xi
+      \cst{p}(\tuplen{x})$. By construction of $\mm'$, there exists a clause
+      $\pospredlit{\cst{p}(\tuplen{y})} \cor D' \in G$ and an assignment $\xi'$
+      which assigns each $y_i$ value of $\xi(x_i)$ such that $\mm
+      \not\models_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
+      \neqlit{x_n}{y_n} \lor C' \lor D'$ is a tautology, according to condition
+      (4) of Definition \ref{def:definition}. However, for a valuation that
+      behaves like $\xi$ on $\tupleempty{x}$ and $\xi'$ on $\tupleempty{y}$,
+      $\mm'$ does not satisfy $R \in N$, contradicting our assumption.
+      % @Jasmin: Your trick with var renaming does not work as flat resolvent is
+      % defined on
+      % variable-disjoint clauses.
+
+   \end{proof}
+\end{rep}
+\begin{conf}
+\removemyskip
+\end{conf}
+\begin{theoremx}
+\label{thm:pes-sat-equiv}
+   The result of applying DPE to a clause set $N$ is
+   satisfiable if and only if $N$ is satisfiable.
+\end{theoremx}
+\begin{rep}
+\begin{proof}
+   Let $\cst{p}$ be a predicate symbol and $G \subseteq N$ be the
+   definition set used by DPE, and let $R = \withpred{N}{\cst{p}} \setminus G$.
+%
+   % First, using Lemmas \ref{lem:flat-res-set-termination-confluence} and \ref{lem:flat-res-set-step-satisfiability}
+   % we show that we will eventually obtain a set $R'$
+   % that contains no occurrences of $\cst{p}$ and that preserves satisfiability
+   % of the original $R$. Finally, we will show that we can
+   % omit $G$ in the result without affecting satisfiability.
+
+   Using Lemma
+   \ref{lem:flat-res-set-termination-confluence}, we get that there is a
+   derivation $(G,R) \flatresiter^n (G,R') \flatresiter R'$. Applying Lemma
+   \ref{lem:flat-res-set-step-satisfiability} $n$ times, we get that $G \cup R$
+   is equivalent to $G \cup R'$. Finally, Lemma \ref{lem:flat-res-set-last-step-satisfiability}
+   gives us the desired result.
+   \qed
+\end{proof}
+\end{rep}
+
+Since there will typically be at most only a few defined predicates in the
+problem, it makes sense to fall back on SPE when no definition is found.
+
+\begin{definitionx}
+   \looseness=-1
+   Let $N$ be a clause set and $\cst{p}$ be a predicate symbol. If there exists a
+   definition set $G \subseteq N$ for $\cst{p}$, \emph{portfolio predicate
+   elimination} (PPE) on $\cst{p}$ in $N$ replaces $N$ with
+%%% @PETAR: Double-check the supression of _p below. --JB
+   $\withoutpred{N}{\cst{p}} \cup (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}}
+   \confrep{\withpred{R}{\cst{p}}}{R})$, where $R = \withpred{N}{\cst{p}} \setminus G$.
+   Otherwise, if $\cst{p}$ is singular in $N$, it results in $\withoutpred{N}{\cst{p}}
+   \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
+   In all other cases, it is not applicable.
+\end{definitionx}
+
+\maybesubsection{Refutational Completeness}
+\label{ssec:predicate-elimination-refutational-completeness}
+
+\newcommand\FInf{\mathit{FInf}}
+\newcommand\FLInf{\mathit{FLInf}}
+\newcommand\Red{\mathit{Red}}
+\newcommand\RedI{\Red_\mathrm{I}}
+\newcommand\RedF{\Red_\mathrm{F}}
+\newcommand\LRed{\mathit{LRed}}
+\newcommand\LRedI{\LRed_\mathrm{I}}
+\newcommand\LRedF{\LRed_\mathrm{F}}
+\newcommand\gnd{\mathcalx{G}}
+\newcommand\pow[1]{\mathcalx{pe}(#1)}
+\newcommand\Lab{\mathbf{L}}
+\newcommand\FFLab{{\mathbf{FL}}}
+\newcommand\Inv{\mathit{Inv}}
+
+\newcommand\RedWk{\smash{\Red\Wksup}}
+\newcommand\RedIWk{\smash{\Red_\mathrm{I}\Wksup}}
+\newcommand\RedFWk{\smash{\Red_\mathrm{F}\Wksup}}
+
+\newcommand\Wksym{\flat}
+\newcommand\Wksup{^\Wksym}
+\newcommand\modelsWk{\models\Wksup}
+%\newcommand\modelsWkcapgnd{\models^{\Wksym{}\gnd}_\cap}
+\newcommand\modelsWkcapgnd{\modelsWk}
+\newcommand\LRedWk{\LRed^{\Wksym}}
+\newcommand\LRedWksub{\LRed^{\Wksym,\sqsupset}}
+\newcommand\LRedFWksub{\LRedF^{\Wksym,\sqsupset}}
+\newcommand\LRedIWk{\LRedI^{\Wksym}}
+\newcommand\Wk[1]{#1^\Wksym}
+\newcommand\prop{o}
+\newcommand\true{\top}
+\newcommand\false{\bot}
+
+\looseness=-1
+Hidden-literal-based techniques fit within the traditional framework of
+saturation, because they delete or reduce a clause based on the \emph{presence}
+of other clauses. In contrast, predicate elimination relies on the
+\emph{absence} of clauses from the proof state. We can still integrate it with
+superposition as follows: At every $k$th iteration of the given clause
+procedure, perform predicate elimination on $\mathcalx{A}
+\cup \mathcalx{P}$, and add all new clauses to $\mathcalx{P}$.
+
+One may wonder whether such an approach preserves the refutational
+completeness of the calculus. The answer is no.
+%
+\begin{conf}
+To see why, consider the following \emph{binary splitting} rule
+based on Riazanov and Voronkov \cite{riazanov-voronkov-2001}:
+%
+\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
+%
+Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
+$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
+smaller than the premise, the rule can be applied aggressively as a
+simplification. But notice that the effect of splitting can be undone by
+singular predicate elimination, possibly giving rise to loops
+$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
+This breaks completeness.
+
+Our solution is to curtail the entailment relation used by the redundancy
+criterion to disallow splitting-like simplifications. Weak entailment $\modelsWk$
+is defined via an ad hoc nonclassical logic so that $\{\cst{p} \cor C{,}\;
+\neglit{\cst{p}} \cor C\} \not\modelsWk \{C\}$ and yet $\modelsWk \{\cst{p} \cor
+\neglit{\cst{p}}\}$.
+%
+More precisely, this logic is defined via an
+encoding: $M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$, where
+%
+$\Wk{\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \false$,
+$\Wk{\lnot\,\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \true$, and
+$\Wk{L} = L$ otherwise.
+%
+Moreover, the type $\prop$ may be interpreted as any set of cardinality at least
+2, and $\false$ must be a distinguished symbol interpreted differently
+from~$\top$.
+
+The standard redundancy criterion $\RedWk$ based on $\modelsWk$ supports all the
+familiar deletion and simplification techniques except splitting. Using $\RedWk$
+not only prevents looping, but it also enables the use of
+the given clause procedure, because any redundant inference
+according to $\RedWk$ remains redundant after SPE or DPE. As usual, the devil is
+in the details, and the details are in the report \cite{our-report}.
+\end{conf}
+
+\begin{rep}
+To see why, consider the following \emph{binary splitting} rule
+based on Riazanov and Voronkov \cite{riazanov-voronkov-2001}:
+%
+\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
+%
+Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
+$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
+smaller than the premise, the rule can be applied aggressively as a
+simplification. But notice that the effect of splitting can be undone by
+singular predicate elimination, possibly giving rise to loops
+$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
+Clearly, we need to curtail predicate elimination.
+
+Under which conditions is predicate elimination refutationally complete? To
+answer this question, we employ the saturation framework of Waldmann, Tourret,
+Robillard, and Blanchette \cite{wtrb-20-sat-framework}. Let
+$(\FInf, \Red)$ be the base calculus without predicate elimination---e.g.,
+resolution or superposition inferences together with the standard
+redundancy criterion \cite[Sect.~4.2]{bg-01-resolution}. The
+inference system $\FInf$ is a set of inferences $(C_n,\ldots,C_1,C_0)$, for $n
+\ge 1$, where $C_n,\ldots,C_1$ are the premises and $C_0$ is the conclusion.
+$C_1$ is called the main premise. %; $C_n,\ldots,C_2$ are the side premises.
+The redundancy criterion is a pair $\Red = (\RedI, \RedF)$ where $\RedI$
+determines which inferences can be omitted and $\RedF$ is used to remove
+clauses.
+
+Next, consider an abstract proving process working on a single clause set.
+Let $\rhd_{\Red}$ denote the transition relation that supports (1)~adding
+arbitrary clauses and (2)~removing clauses deemed useless by $\RedF$.
+Typically, the added clauses are the result of performing inferences and are
+entailed by the premises, but other clauses can be added as well. A
+\emph{$\rhd_{\Red}$-derivation} is a finite or infinite sequence of clause sets
+$N_0 \rhd_{\Red} N_1 \rhd_{\Red} \cdots$.
+
+We fix a finite set $\mathbf{P}$ of predicate symbols that may be subjected to
+predicate elimination. These might include all the predicate symbols occurring
+in the input problem, but exclude any symbols introduced by splitting or other
+rules. Given a clause or clause set $N$, we write $\mathbf{P}(N)$ to denote the
+set of all predicate symbols from $\mathbf{P}$ occurring in $N.$
+%
+Let $\rhd_\mathbf{P}$ denote the elimination of a
+singular or defined predicate symbol from $\mathbf{P}$. A \emph{mixed
+derivation} consists of transitions either of the form $N
+\rhd_\mathbf{P} N'$ or of the form $N \rhd_{\Red} N'$ where
+$\mathbf{P}(N) \supseteq \mathbf{P}(N')$.
+Because $\mathbf{P}$ is finite, any mixed derivation consists of at most
+finitely many $\rhd_\mathbf{P}$ transitions. Hence, in any derivation,
+there exists an index~$k$ from which all transitions are standard
+${\rhd}_{\Red}$-transitions.
+
+This suggests the following path to completeness: Pretend that the transitions
+between $N_0$ and $N_k$ are merely preprocessing and start the actual
+derivation at $N_k$. This works at the abstract level of derivations on single
+clause sets. It fails, however, for an actual saturation prover that
+distinguishes between passive and active clauses.
+
+\begin{examplex}
+\label{ex:cex-predelim-sat}
+The counterexample below is based on the given clause prover \textsf{GC}
+from the saturation framework. It shows how predicate elimination can break
+\textsf{GC}'s key invariant, which states that all inferences between active
+clauses are redundant. Breaking the invariant means that the limit might be
+unsaturated, breaking the refutational completeness proof.
+
+We use superposition with the order $\cst{a} \prec
+\cst{b} \prec \cst{c} \prec \cst{d}$ and without selection.
+Assume $\cst{a} \in \mathbf{P}$ and
+suppose we start with the satisfiable clause set %consisting of
+%
+%\begin{linenomath*}
+\begin{align*}
+  & \neglit{\cst{a}} \cor \MAX{\cst{d}}
+  && \neglit{\cst{a}} \cor \MAX{\neglit{\cst{d}}}
+  && \cst{a} \cor \cst{b} \cor \MAX{\cst{c}}
+  && \cst{c} \cor \MAX{\cst{d}}
+  && \cst{b} \cor \MAX{\neglit{\cst{d}}}
+\end{align*}
+%\end{linenomath*}
+%
+where gray boxes mark maximal literals. Suppose the prover makes
+$\cst{c} \cor \cst{d}$ and $\cst{b} \cor \neglit{\cst{d}}$ active. From these
+two clauses, a superposition inference~$\iota$ could derive the conclusion
+$\cst{b} \cor \cst{c}$. However, the three passive clauses are
+$\prec$-smaller than $\iota$'s main premise $\cst{b} \cor \neglit{\cst{d}}$ and
+collectively entail $\iota$'s conclusion. This means that $\iota$ is
+redundant and can be ignored.
+
+If the prover now eliminates the predicate $\cst{a}$ using \infname{SPE}, the
+passive set is reduced to $\{\cst{b} \cor \cst{c} \cor \cst{d}{,}\;
+\cst{b} \cor \cst{c} \cor \neglit{\cst{d}}\}$.
+%
+Either clause is subsumed by an active clause, so the prover deletes it.
+It stops with the active set
+$\{\cst{c} \cor \cst{d}{,}\; \cst{b} \cor \neglit{\cst{d}}\}$, which is
+unsaturated because $\iota$ is no longer redundant. The invariant is broken.
+\end{examplex}
+
+\begin{examplex}
+\label{ex:cex-predelim-unsat}
+In Example~\ref{ex:cex-predelim-sat}, the initial clause set was satisfiable.
+If it is unsatisfiable, we can even lose refutational
+completeness. To see why, we add the unit clauses
+$\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ to the initial clause set of
+Example~\ref{ex:cex-predelim-sat} to make it unsatisfiable. We repeat the same
+steps as above, including the subsumptions at the end,
+yielding the passive set $\{\MAX{\neglit{\cst{b}}}, \MAX{\neglit{\cst{c}}}\}$ and
+the active set $\{\cst{c} \cor \MAX{\cst{d}}{,}\; \cst{b} \cor \MAX{\neglit{\cst{d}}}\}$. Then,
+making $\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ active triggers no
+inferences. The prover stops with an
+unsatisfiable four-clause active set that does not contain the empty clause.
+\end{examplex}
+
+A solution could be to move all active clauses to the passive set at step~$k$
+or later, but this would be costly, since it would force the prover to redo
+inferences whose conclusions might then have to be simplified or subsumed again.
+Instead, we salvage the existing completeness proof for $\mathsf{GC}$, by
+resolving the issues concerning splitting and the $\mathsf{GC}$ invariant. Our
+approach is to weaken the redundancy criterion slightly, enough both to disable
+splitting on $\mathbf{P}$-predicates and to ensure that
+inferences such as $\iota$ in Examples
+\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} are performed. The
+required weakening is so mild that it does not invalidate any practical
+simplification or subsumption techniques we are aware of, except of course
+splitting.
+
+In accordance with the saturation framework, let $\mathbf{F}$ be the set of
+first-order $\mathrm{\Sigma}$-clauses, let $\mathbf{G}$ be its ground subset,
+and let $\gnd$ be a function that maps an $\mathbf{F}$-clause to the set of its
+$\mathbf{G}$-clause instances and analogously for $\mathbf{F}$-inferences. We
+define an extension $\mathbf{G}\Wksup$ of $\mathbf{G}$ for
+$\mathrm{\Sigma}\Wksup$-clauses in an ad hoc nonclassical logic reminiscent
+of paraconsistent logics \cite{carnielli-et-al-2007}. The objective is to
+disallow the entailment that makes splitting and Examples
+\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} possible.
+%
+The signature~$\mathrm{\Sigma}\Wksup$ extends $\mathrm{\Sigma}$ with a
+distinguished predicate symbol $\false$ that is interpreted differently
+from~$\top$. For $\mathrm{\Sigma}\Wksup$, the Boolean type $\prop$ may be
+interpreted as any set of cardinality at least 2.
+
+\begin{definitionx}
+The operator $\Wk{}$ translates $\mathrm{\Sigma}$-literals to
+$\mathrm{\Sigma}\Wksup$-literals as follows, where $\cst{p} \in \mathbf{P}$,
+$\cst{q} \notin \mathbf{P}$, and $s, t$ are non-Boolean terms:
+%
+%\begin{linenomath*}
+\begin{align*}
+  \Wk{\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \false
+& \Wk{\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \eq \true
+& \Wk{(s \eq t)} & = s \eq t \\[-\jot]
+  \Wk{\lnot\,\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \true
+& \Wk{\lnot\,\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \noteq \true
+& \Wk{(s \noteq t)} & = s \noteq t
+\end{align*}
+%\end{linenomath*}
+%
+The operator is lifted elementwise to $\mathbf{G}$-clauses and
+$\mathbf{G}$-clause sets.
+The \emph{weak entailment} $\modelsWk$ over $\mathbf{G}$-clause sets
+is defined via an encoding into $\mathrm{\Sigma}\Wksup$-clauses:
+$M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$.
+The lifting to $\mathbf{F}$-clauses and $\mathbf{F}$-clause set is achieved
+in the standard way via grounding.
+\end{definitionx}
+
+The following property of weak entailment will allow us to eliminate
+$\mathbf{P}$-predicates without losing completeness:
+
+\begin{lemmax}
+\label{lem:p-Wk-entail}
+Let $C$ be a clause that contains the predicate symbol $\cst{p} \in
+\mathbf{P}$ and $D$ be a clause that does not contain $\cst{p}$. If $N
+\cup \{C\} \modelsWk \{D\}$, then $N \modelsWk \{D\}$.
+\end{lemmax}
+
+\begin{proof}
+Suppose $\mathscr{J} \models \Wk{N}$. We will define $\mathscr{J}'$ so that
+$\mathscr{J}' \models \Wk{N} \cup \{\Wk{C}\}$, retrieve $\mathscr{J}'
+\models \Wk{D}$, and then argue that $\mathscr{J} \models \Wk{D}$. We
+take $\mathscr{J}'$ to coincide with $\mathscr{J}$ except that we extend the
+domain for $\prop$ with one fresh value and use this value as the
+interpretation of $\cst{p}(\tuple{t}{})$ for all argument tuples $\tuple{t}{}$.
+This modification makes any $\cst{p}$ literal of $\Wk{C}$ true, and it
+preserves the truth of $\Wk{N}$. By the hypothesis, $\mathscr{J}' \models
+\Wk{D}$. And since $\cst{p}$ does not occur in $D$, we have $\mathscr{J}
+\models \Wk{D}$.\qed
+\end{proof}
+
+Note that the above lemma does not hold for classical entailment $\models$;
+indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
+\{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
+entailment: $\modelsWk \cst{p} \cor \lnot\,\cst{p}$. In fact, all classical
+clausal tautologies are tautologies for $\modelsWk$.
+
+The standard redundancy criterion $\Red$ is obtained by lifting a criterion on
+$\mathbf{G}$-clauses to $\mathbf{F}$-clauses. The same construction can be
+replicated using $\modelsWk$ instead of $\models$, yielding the \emph{weak
+redundancy criterion} $\RedWk$.
+It is easy to check that the usual simplification techniques implemented in
+superposition provers can be justified using $\RedWk$. Specifically,
+this concerns the following rules described by Schulz \cite[Sects.\
+2.3.1~and~2.3.2]{ss-02-brainiac}:\
+deletion of duplicated literals, 
+deletion of resolved literals, 
+syntactic tautology deletion,
+semantic tautology deletion,
+rewriting of negative literals,
+positive simplify-reflect,
+negative simplify-reflect,
+clause subsumption, and
+equality subsumption.
+Moreover, rewriting of positive literals is possible if the
+rewriting clause is smaller than the rewritten
+clause (a condition that is also needed with $\models$ but omitted by Schulz).
+Finally, destructive equality resolution cannot be justified with $\models$,
+let alone $\modelsWk$.
+
+We instantiate the saturation framework with $(\FInf,\RedWk)$ to obtain a
+given clause prover \textsf{GC}. The prover operates on sets of labeled clauses
+$(C, l)$, where $C$ is a standard clause and $l \in \Lab$ is a label.
+The $\mathsf{active}$ label identifies active clauses; all other clauses are
+passive. The prover takes the form of two rules,
+\textsc{Process} and \textsc{Infer}, restricted to prevent the introduction of
+$\mathbf{P}$-predicates. We extend it with a third rule,
+\textsc{PredElim}, for predicate elimination, and call the extended prover
+\textsf{GCP}. The rules are as follows, using again the framework notations:
+
+\begin{enumerate}[\rm\textsc{PredElim}\,]
+\itemsep1\jot
+\item[\rm\textsc{Process}\,]
+  $\mathcalx{N} \cup \mathcalx{M}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \mathcalx{M}'$
+\\
+  where $\mathcalx{M} \subseteq \LRedFWksub(\mathcalx{N} \cup \mathcalx{M}')$,
+  $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$,
+  and \\
+  $\mathbf{P}(\lfloor\mathcalx{M}'\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N} \cup \mathcalx{M}\rfloor)$
+
+\item[\rm\textsc{Infer}\,]
+  $\mathcalx{N} \cup \{(C, l)\}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \{(C, \ensuremath{\mathsf{active}})\} \cup \mathcalx{M}$
+\\
+  where $l \not= \ensuremath{\mathsf{active}}$,
+  $\mathcalx{M}{\downarrow}_\mathsf{active} = \emptyset$, \\
+  $\FInf(\lfloor\mathcalx{N}{\downarrow}_\mathsf{active}\rfloor, \{C\}) \subseteq
+    \RedIWk^{{\cap}\gnd}(\lfloor\mathcalx{N}\rfloor \cup \{C\} \cup \lfloor\mathcalx{M}\rfloor)$,
+  and \\
+  $\mathbf{P}(\lfloor\mathcalx{M}\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N}\rfloor \cup \{C\})$
+
+\item[\rm\textsc{PredElim}\,]
+  $\mathcalx{N} \cup \mathcalx{M}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \mathcalx{M}'$
+\\
+  where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P} \mathcalx{N} \cup \mathcalx{M}'$
+  and $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$
+\end{enumerate}
+
+Here is a summary of the main framework notations:
+%
+\begin{itemize}
+\item The letters $\mathcalx{M}, \mathcalx{N}$ range over sets of labeled clauses.
+  $\mathcalx{M}{\downarrow}_l$ denotes the subset of clauses in $\mathcalx{M}$
+  labeled with $l$. The operator $\lfloor\phantom{\cdot}\rfloor$ erases all
+  labels in a labeled clause or clause set.
+
+\smallskip
+
+\item $\FInf(N)$ denotes the set of all base calculus inferences with premises
+  in $N$, and $\FInf(N, M) = \FInf(N \cup M) \setminus \FInf(N \setminus M)$.
+  The same notations are also available for the straightforward extension $\FLInf$
+  of $\FInf$ with labels.
+
+\smallskip
+
+\item $\LRedWksub$ is the extension of the standard redundancy
+  criterion $\RedWk$ defined using $\modelsWk$ to nonground labeled clauses
+  with subsumption ($\sqsubset$).
+
+\smallskip
+
+\item Given a sequence $(\mathcalx{N}{}_i)_i$, its \emph{limit} (\emph{inferior})
+  is $\mathcalx{N}{}_\infty = \bigcup_i \bigcap_{j\ge i} \mathcalx{N}{}_j$.
+\end{itemize}
+
+The completeness proof follows the invariance-based argument found the
+forthcoming journal submission \cite{waldmann-et-al-202x-article}
+by Waldmann et al.
+
+\begin{lemmax}
+\label{lem:gcp-derivations-are-red-black-derivations}
+Every $\Longrightarrow_\mathsf{GCP}$-derivation is a mixed derivation.
+\end{lemmax}
+
+\begin{proof}
+The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
+al., with adjustments to show that $\cst{P}$-predicates cannot reappear once
+they have disappeared. The case for \textsc{ElimPred} is trivial.\qed
+\end{proof}
+
+Let $\Inv\Wksup_\mathcalx{N}(k)$ denote the condition
+$\textstyle\FLInf(\mathcalx{N}{}_k{\downarrow}_\mathsf{active}) \subseteq
+\LRedIWk(\mathcalx{N}{}_k)$.
+Notice the difference with the definition of the key invariant
+$\Inv_\mathcalx{N}$ in the saturation framework, whose right-hand side is
+$\bigcup_{i=0}^k \RedI^{\Lab}(\mathcalx{N}{}_i)$. We cannot use the
+big union ${\bigcup}$ starting at $i = 0$ because we will need to truncate a
+sequence prefix of an a priori unknown length. The argument will still work
+thanks to monotonicity properties of redundancy criteria.
+
+\begin{lemmax}
+\label{lem:gcp-invar-gcp}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
+If
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$,
+then
+$\Inv\Wksup_\mathcalx{N}(k)$ holds for all indices~$k$.
+\end{lemmax}
+
+\begin{proof}
+The base case is as in Waldmann et al.
+
+For \textsc{Process} and \textsc{Infer}, the proof is essentially as in Waldmann
+et al., except that we also need to show that
+$\LRedIWk(\mathcalx{N}{}_k) \subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
+This is a consequence of
+$\mathcalx{N}{}_k \rhd_{\LRedWksub} \mathcalx{N}{}_{k+1}$ and of
+properties (R2) and (R3) of redundancy criteria.
+
+A new case to consider is that of a \textsc{PredElim} transition
+$\mathcalx{N}{}_{k} \Longrightarrow_\mathsf{GCP} \mathcalx{N}{}_{k+1}$.
+Let $\mathcalx{N}{}_{k} = \mathcalx{N} \cup \mathcalx{M}
+\Longrightarrow_\mathsf{GCP} \mathcalx{N} \cup \mathcalx{M}' = \mathcalx{N}{}_{k+1}$,
+where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P}
+\mathcalx{N} \cap \mathcalx{M}'$ and
+$\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
+We assume without loss of generality that $\mathcalx{M} \cap \mathcalx{M}' =
+\emptyset$.
+Let $\cst{p}$ be the eliminated predicate. Note that $\cst{p}$ occurs in every
+clause in $\mathcalx{M}$ but in none of the clauses in $\mathcalx{N}$ or
+$\mathcalx{M}'$.
+We must show $\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
+\subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
+As for \textsc{Process}, we have the inclusion
+$\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
+\subseteq \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$,
+by the side condition that
+$\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
+Moreover, by the induction hypothesis,
+$\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
+\subseteq \LRedIWk(\mathcalx{N}{}_k)$. Thus,
+$\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
+\subseteq \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
+
+Let $\iota \in \FLInf(\mathcalx{N} {\downarrow}_\mathsf{active})$.
+By the argument above, we have
+$\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
+We must
+show $\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.
+%
+By definition of $\LRedIWk$, it suffices to show that for
+every ground instance $(C_n,\ldots,C_1,C_0)$ of $\iota$, there exists a finite
+clause set $\mathcalx{D} \subseteq \gnd(\mathcalx{N}) \cup \gnd(\mathcalx{M}')$
+that is $\prec$-smaller than $C_1$ and such that
+$\{C_n,\ldots,C_2\} \cup \mathcalx{D} \modelsWk \{C_0\}$.
+Without loss of generality, we assume that $\mathcalx{D}$ is the smallest such
+set.
+
+By definition of $\mathcalx{N}$, $\cst{p}$ cannot occur in $C_0$.
+By Lemma~\ref{lem:p-Wk-entail}, if $\cst{p}$ occurred in $D \in \mathcalx{D}$,
+we could remove $D$, but this would mean $\mathcalx{D}$ is not minimal.
+As a result, $\mathcalx{D}$ cannot contain clauses from $\gnd(\mathcalx{M})$ and
+hence $\mathcalx{D} \subseteq \gnd(\mathcalx{N})$. Thus,
+$\iota \in \LRedIWk(\mathcalx{N})$. By property (R2) of
+redundancy criteria, we have the desired result: $\iota \in
+\LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.\qed
+\end{proof}
+
+\begin{lemmax}
+\label{lem:gcp-invar-infinity}
+Let $(\mathcalx{N}{}_i)_i$ be a
+$\rhd_{\LRedWksub}$-derivation.
+If $\Inv\Wksup_\mathcalx{N}(i)$ holds for all indices~$i$, then
+$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
+\bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
+holds.
+\end{lemmax}
+
+\begin{proof}
+We assume $\iota \in \FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$
+and show $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
+for some arbitrary $\iota$.
+%
+For $\iota$ to belong to $\FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$, all of
+its finitely many premises must be in $\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active}$.
+Therefore, there must exist an index $k$ such that
+$\mathcalx{N}{}_k {\downarrow}_\mathsf{active}$ contains all of them, and therefore
+$\iota \in \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$. Since $\Inv_\mathcalx{N}(k)$
+holds, $\iota \in \LRedIWk(\mathcalx{N}{}_k)$.
+Hence, $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$.\qed
+\end{proof}
+
+\begin{lemmax}
+\label{lem:fair-gcp-derivations}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
+If
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
+and
+$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
+$l \not= \ensuremath{\mathsf{active}}$,
+then there exists an index $k$ such that
+$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
+\end{lemmax}
 
 \begin{sloppypar}
-Back in 1994, Kohlhase \cite[Sect.~1.3]{mk-94-hores} was optimistic about the
-future of higher-order automated reasoning:
-%
-\begin{quote}
-  %The author believes that
-  The obstacles to proof search intrinsic
-  to higher-order logic may well be compensated by the greater expressive power
-  of higher-order logic and by the existence of shorter proofs. Thus
-  higher-order automated theorem proving will be practically as feasible as
-  first-order theorem proving is now as soon as the technological backlog is made up.
-\end{quote}
-%
-For higher-order superposition, the backlog consisted of designing calculus
-extensions, heuristics, and algorithms that mitigate its weaknesses. In
-this article, we presented such enhancements, justified their design, and
-evaluated them. We explained how each weak point in the
-higher-order proving pipeline could be improved, from preprocessing to reasoning
-about formulas, to delaying unpromising or explosive inferences, to invoking a
-backend. Our evaluation indicates that higher-order superposition is now the
-state of the art in higher-order reasoning.
+\begin{proof}
+By Lemma~\ref{lem:gcp-derivations-are-red-black-derivations}, there must exist
+an index $k$ such that the sequence $(\mathcalx{N}{}_{k+i})_i$ is a pure
+$\rhd_{\LRedWksub}$-derivation.
+By Lemma~\ref{lem:gcp-invar-gcp},
+$\Inv\Wksup_\mathcalx{N}(k+i)$ holds for all indices~$i$.
+Hence, by Lemma~\ref{lem:gcp-invar-infinity},
+$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
+\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.
+By the second hypothesis, this inclusion simplifies to
+$\FLInf(\mathcalx{N}{}_\infty) \subseteq
+\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.\qed
+\end{proof}
 \end{sloppypar}
 
+\begin{theoremx}
+\label{thm:gcp-complete}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation
+with
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
+and
+$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
+$l \not= \ensuremath{\mathsf{active}}$.
+If $\lfloor \mathcalx{N}{}_0\rfloor$ is unsatisfiable,
+then
+some $\mathcalx{N}{}_i$
+contains the empty clause with some arbitrary label.
+\end{theoremx}
+
+\begin{proof}
+By Lemma~\ref{lem:fair-gcp-derivations},
+we know that there exists an index $k$ such that
+$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
+Moreover, since
+$\rhd_{\LRedWksub}$ and
+$\rhd_\mathbf{P}$ preserve unsatisfiability
+(by (R1) of redundancy criteria, Khasidashvili and Korovin's Theorem~1, and our
+Theorem~\ref{thm:pes-sat-equiv}), we have that
+$\lfloor \mathcalx{N}{}_k\rfloor$ is unsatisfiable.
+Since the base calculus $(\FLInf,\allowbreak\LRed)$ is assumed to be statically
+refutationally complete with respect to $\models$, the calculus
+$(\FLInf,\allowbreak\LRedWk)$ with a weaker redundancy criterion is also
+statically complete with respect to $\models$, and by the saturation framework,
+$(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
+static and dynamic completeness, we conclude that some $\mathcalx{N}{}_{k+i}$ must
+contain a labeled empty clause.\qed
+\end{proof}
+\end{rep}
+
+\begin{qle}
+\section{Quasipure Literal Elimination}
+\label{sec:quasipure-literal-elimination}
+
+\emph{Pure literal elimination} (PLE) is one of the simplest optimizations
+implemented in SAT solvers. It is a special case of variable elimination: If a
+given variable always occurs with the same polarity in a problem, the solver can
+assign it that polarity without loss of generality, making all the clauses that
+contain it tautologies. PLE consists of recursively deleting all such clauses.
+%
+PLE's generalization to first-order logic considers predicate literals
+$\arbpredlit{\cst{p}(\tuple{s}{})}$, where the arguments $\tuple{s}{}$ are
+ignored by the analysis; only the polarity matters.
+
+\begin{examplex}
+   \label{example:ple}
+   Consider the clause set $N = \{ \cst{p}(x) \cor \cst{q}(\cst{a}, x){,}\allowbreak\;
+   \cst{p}(\cst{f}(x)){,}\allowbreak\; \neglit{\cst{q}(\cst{a}, \cst{a})} \}$.
+   Since $\cst{p}$ occurs only positively in $N$, PLE deletes the two first
+   clauses. At that point, $\cst{q}$ occurs only negatively in the remaining
+   singleton clause set and can be deleted as well. The result is the empty set,
+   which is obviously satisfiable, indicating that $N$ is satisfiable.
+   As model of $N$, we can take $\ourmodel$ such that $\ourmodel$ makes all
+   atoms $\cst{p}(\ldots)$ true and all atoms $\cst{q}(\ldots)$ false.
+\end{examplex}
+
+\begin{examplex}
+   \label{example:qle-i}
+   Consider the clause set $N' = \{
+   \cst{p}(\cst{a}){,}\allowbreak\;
+   \neglit{\cst{p}(x)} \lor \cst{p}(\cst{f}(x))
+    \}$. This time, PLE does not apply because $\cst{p}$ occurs with both
+    polarities. Yet we notice that $\cst{p}$ occurs positively in
+    both clauses, and hence that the same reasoning as in the previous
+    example applies. In particular, $\ourmodel \models N'$.
+\end{examplex}
+
+\begin{examplex}
+   \label{example:qle-ii}
+   Consider the clause set $N'' = \{
+   \cst{p}(\cst{a}){,}\allowbreak\;
+   \cst{q}(x) \lor \cst{p}(\cst{f}(x)){,}\allowbreak\;
+   \neglit{\cst{q}(\cst{f}(\cst{a}))}{,}\allowbreak\;
+   \neglit{\cst{p}(x)} \lor \neglit{\cst{q}(\cst{f}(x))}
+   \}$. This clause set is clearly beyond PLE. Yet each clause contains either
+   a positive $\cst{p}$ literal or a negative $\cst{q}$ literal---any
+   extraneous, wrong-polarity literals are harmless. Thus $\ourmodel \models
+   N''$.
+\end{examplex}
+
+Examples \ref{example:qle-i} and \ref{example:qle-ii} suggest that PLE is
+needlessly restrained, presumably due to its origins in propositional logic. We
+propose a generalization to ``quasipure'' predicates.
+
+\begin{definitionx}
+\label{def:quasipure-set}
+A set $P$ of predicate symbols is \emph{quasipure} with a polarity map $s$ in a
+clause set $N$ if for every symbol $\cst{p} \in P$, every clause in $N$ that
+contains a member of $P$ contains at least one occurrence of some $\cst{q} \in
+P$ with polarity $s_\cst{q} \in \{+, -\}$.
+%
+The set $P$ is \emph{quasipure} in $N$ if there exists a polarity map $s$
+such that $P$ is quasipure with $s$ in $N$.
+\end{definitionx}
+
+In Example~\ref{example:qle-i}, $\{\cst{p}\}$ is quasipure with $s_\cst{p} =
+{+}$. In Example~\ref{example:qle-ii}, $\{\cst{p}, \cst{q}\}$ is quasipure with
+$s_\cst{p} = {+}$ and $s_\cst{q} = {-}$. For this example, it is crucial to
+consider $\cst{p}$ and $\cst{q}$ together; neither of the singletons
+$\{\cst{p}\}$ and $\{\cst{q}\}$ is quasipure.
+
+\begin{definitionx}
+\label{def:quasipure-sym}
+A predicate symbol $\cst{p}$ is \emph{quasipure} with polarity $s_\cst{p} \in
+\{+, -\}$ in a clause set $N$ if there exists a set $P \ni \cst{p}$ of predicate
+symbols and a polarity map $s$ such that $P$ is quasipure with $s$ in $N$.
+%
+The symbol $\cst{p}$ is \emph{quasipure} in $N$ if there exists a polarity
+$s_\cst{p} \in \{+, -\}$ such that $\cst{p}$ is quasipure with $s_\cst{p}$ in
+$N$.
+%
+A literal $L = \arbpredlit{\cst{p}(\ldots)}$ is \emph{quasipure} in $N$ if
+$\cst{p}$ is quasipure with $L$'s polarity in $N$.
+\end{definitionx}
+
+Deleting a clause containing a quasipure literal is monotonic in the sense that
+it might create new opportunities for quasipure literal elimination but never
+ruin existing ones. Therefore, the following nondeterministic procedure yields a
+well-defined result:
+
+\begin{definitionx}
+\label{def:qle}
+Given a clause set, \emph{quasipure literal elimination} (QLE) iteratively
+deletes clauses containing quasipure predicate symbols until no such clauses
+remain.
+\end{definitionx}
+
+The key property of QLE is that it preserves unsatisfiability:
+
+\begin{lemmax}
+Let $C \in N$ be a clause containing a quasipure literal. If $N \setminus \{C\}$
+is satisfiable, then $N$ is satisfiable.
+\end{lemmax}
+
+\begin{proof}
+Let $\ourmodel$ be a model of $N \setminus \{C\}$.
+Let $P$ be the set of predicates and $s$ the polarity map whose existence is
+guaranteed by definition of quasipurity. Let $N_0 \subseteq N \setminus \{C\}$
+be the result of applying QLE on $N$. It is easy to see that $N_0$ contains no
+occurrences of the symbols in $P$. Extend $\ourmodel$ to $\ourmodel'$ by
+making each symbol $\cst{p}$ such that $s_\cst{p} = {+}$ true and
+each symbol $\cst{p}$ such that $s_\cst{p} = {-}$ false. Clearly $\ourmodel'$ is
+a model not only of $N_0$, which contains no $P$ symbols, but also of $N$,
+because each clause in $N \setminus N_0$ contains a quasipure literal, which is
+satisfied by $\ourmodel'$.
+\qed
+\end{proof}
+
+Definition~\ref{def:qle} suggests a naive, nondeterminisitc procedure for
+discovering and eliminating quasipure predicate symbols: Choose a predicate
+symbol $\cst{p}$ and a polarity $s_\cst{p}$, and take $P = \{\cst{p}\}$. If the
+predicate occurs with the wrong polarity in a clause
+$\arbpredlit{\cst{p}(\ldots)} \lor \arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
+\arbpredlit{\cst{q}_n(\ldots)} \lor C$, try to extend the polarity map $s$ and
+the predicate set $P$ with one of the $\arbpredlit{\cst{q}_i}$'s and continue
+recursively with $\cst{q}_i$. In Sect.~\ref{sec:implementation}, we will see a
+more efficient approach based on a SAT encoding.
+\end{qle}
+
+\section{Satisfiability by Clause Elimination}
+\label{sec:satisfiability-by-clause-elimination}
+
 \looseness=-1
-Higher-order extensions of first-order superposition have been
-considered %from a theoretical perspective    -- The point is made later in the same para. --JB
-by Bentkamp et al.\
-\cite{bbtvw-21-sup-lam, bbcw-21-lfho} and Bhayat and Reger
-\cite{br-19-restricted-unif, br-20-full-sup-w-combs}. They introduced proof calculi,
-proved them refutationally complete, and suggested optional rules, but
-they hardly discussed the practical aspects of higher-order superposition. Extensions
-of SMT are discussed by Barbosa et al. \cite{brotb-19-ho-smt}.
-Bachmair and Ganzinger \cite{bg-1992-nonclausal}, Manna and Waldinger
-\cite{mw-1979-nonclausal}, and Murray \cite{nm-1982-nonclausal} have studied
-nonclausal resolution calculi.
-
-In contrast, there is a vast literature on practical aspects of first-order
-reasoning using superposition and related calculi.
-The literature evaluates various procedures and techniques
-\cite{hv-09-unifalgs,rsv-15-playing-with-avatar}, literal and term order selection
-functions \cite{hrsv-16-selsel}, and clause evaluation functions
-\cite{sm-2016-clausesel, gs-20-clausesel}, among others. Our work joins the
-select club of papers devoted to practical aspects of higher-order
-reasoning
-\cite{sb-15-beta,wskb-16-effective-norm,fb-2016-internal-guidance-satallax,benzmueller-et-al-05-can-ho-fo-coop}.
-
-%In contrast, higher-order calculi implemented in competitive provers are usually discussed from a theoretical viewpoint.
-
-%\section{Conclusion and Future Work}
-%\label{sec:ho-tech:conclusion}
-
-% We filled in this gap: we describe how
-% We described some ways in which the explosion incurred by
-% \lsup{} can be kept under control. We showed how native
-% support for a Boolean type can dramatically improve a higher-order prover's
-% performance. For many choices that can be made during the higher-order proof
-% search, we described heuristics and evaluated them. We also designed a
-% saturation loop that enumerates infinite sets of higher-order inference
-% conclusions, while maintaining the same behavior as the standard given clause
-% procedure on first-order clauses.
-
-As a next step, we plan to implement the described techniques in E.
-%%% @PETAR: In E 2.6, basically. --JB
-% Ehoh \cite{ehoh-section}, the $\lambda$-free higher-order extension of E.
-We
-expect the resulting prover to be substantially more efficient than
-Zipperposition. Moreover, we want to investigate the proofs found by provers
-such as CVC4 and Satallax but missed by Zipperposition. Finding the reason
-behind why Zipperposition fails to prove specific problems will likely result in useful new techniques.
+The main approaches to show satisfiability of a first-order problem are to
+produce either a finite Herbrand model or a saturated clause set. Saturations
+rarely occur except for very small problems or within decidable fragments. In
+this section, we explore an alternative approach that establishes satisfiability
+by iteratively removing clauses while preserving
+%satisfiability and
+unsatisfiability,
+until the clause set has been transformed into the empty set. So
+far, this technique has been studied only for QBF~\cite{hsb-14-QRAT}. We show that
+\emph{blocked clause elimination} (BCE)
+% for first-order logic
+can be used for this purpose. It can efficiently solve some problems for which
+the saturated set would be infinite. However, it can break the refutational
+completeness of a saturation prover. We conclude with a procedure that
+transforms a finite Herbrand model into a sequence of clause elimination steps
+ending in the empty clause set, thereby demonstrating the theoretical power of
+clause elimination.
+
+Kiesl et al.\ \cite{ksstb-2017-blockedfol} generalized blocked clause
+elimination to first-order logic. Their generalization uses flat
+$L$-resolvents, an extension of flat resolvents that resolves a single
+literal~$L$ against $m$~literals of the other clause.
+
+\begin{definitionx}
+   Let $C = L \cor C'$ and $D = L_1 \cor \cdots \cor L_m \cor D'$, where
+   (1)~$m \ge\nobreak 1$,
+   (2) the literals $L_i$ are of opposite polarity to $L$,
+   (3) $L$'s atom is $\cst{p}(\tuplen{s})$,
+   (4) $L_i$'s atom is $\cst{p}(\tuple{t_i}{})$ for each~$i$, and
+   (5) $C$ and $D$ have no variables in common.
+   The clause
+   $\bigl(\bigvee_{i=1}^m \bigvee_{j=1}^n
+     \neqlit{s_j}{t_{ij}}\bigr) \cor C' \cor D'$
+   is a \emph{flat $L$-resolvent} of $C$~and~$D$.
+\end{definitionx}
+\removemyskip
+\begin{definitionx}
+\label{def:equality-blocked}
+\begin{conf}%
+   A clause $C = L \cor C'$ is (\emph{equality-})\emph{blocked} by $L$ in a clause
+   set $N$ if all flat $L$-resolvents between $C$ and clauses in $N \setminus \{C\}$ are
+   tautologies.
+\end{conf}%
+\begin{rep}%
+   Let $C = L \cor C'$ be a clause and $N$ be a clause set.
+   Let $N'$ consist of all clauses from $N \setminus \{C\}$ with their variables
+   renamed so that they share no variables with $C$.
+   The clause $C$ is (\emph{equality-})\emph{blocked} by $L$ in the set $N$ if
+   all flat $L$-resolvents between $C$ and clauses in $N'$ are tautologies.
+\end{rep}
+\end{definitionx}
+
+Removing a blocked clause from a set preserves unsatisfiability
+\cite{ksstb-2017-blockedfol}. Kiesl et al.\ evaluated the effect of removing all
+blocked clauses as a preprocessing step and found that it increases
+\confrep{}{the }prover's success rate.
+
+In fact, there exist satisfiable problems that cannot be saturated in finitely
+many steps regardless of the calculus's parameters but that can be reduced to
+an empty, vacuously satisfiable problem through blocked clause elimination.
+
+\begin{examplex}
+   \label{example:infinite-blocked-set}
+   \looseness=-1
+   Consider the clause set $N$ consisting of $C = \pospredlit{\cst{p}(x,x)}$ and
+   $D = \negpredlit{\cst{p}(y_1,y_3)} \cor \pospredlit{\cst{p}(y_1,y_2)} \cor
+   \pospredlit{\cst{p}(y_2,y_3)}$. Note that if no literal is
+   selected, all literals are eligible for superposition. In
+   particular, the superposition of
+   $\pospredlit{\cst{p}(x,x)}$ into $D$'s negative literal
+   eventually needs to be performed regardless of the chosen
+   selection function or term order, with the conclusion $E_1 =
+   \pospredlit{\cst{p}(z_1,z_2)} \cor \pospredlit{\cst{p}(z_2,z_1)}$. Then,
+   superposition of $E_1$ into $D$ yields $E_2 = \pospredlit{\cst{p}(z_1,z_2)}
+   \cor \pospredlit{\cst{p}(z_2,z_3)} \cor \pospredlit{\cst{p}(z_3,z_1)}$.
+   Repeating this process yields infinitely many clauses $E_i =
+   \pospredlit{\cst{p}(z_1,z_2)} \cor \cdots \cor
+   \pospredlit{\cst{p}(z_i,z_{i+1})} \cor \pospredlit{\cst{p}(z_{i+1},z_1)}$
+   that cannot be eliminated using standard redu\-ndancy-based techniques.
+\end{examplex}
+
+In the example above, the clause $D$ is blocked by its second or third
+literal. If we delete $D$, $C$ becomes blocked in turn. Deleting
+$C$ leaves us with the empty set, which is vacuously satisfiable. The example
+suggests that using BCE during saturation might help focus the proof search.
+Indeed, Kiesl et al.\ ended their investigations by asking whether BCE can be
+used as an inprocessing technique in a saturation prover. Unfortunately,
+in general the answer is no\confrep{.}{:}
+
+\begin{examplex}
+   \label{example:bce-completeness}
+   Consider the unsatisfiable set $N=\{ C_1, \ldots,\allowbreak C_6 \}$, where
+%   \begin{linenomath*}
+   \begin{align*}
+      C_1 &= \negpredlit{\cst{c}} \cor \pospredlit{\cst{e}} \cor \MAX{\negpredlit{\cst{a}}} 
+      & C_2 &= \negpredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{e}}}
+      & C_3 &= \pospredlit{\cst{b}} \cor \MAX{\pospredlit{\cst{c}}}
+      \\
+      C_4 &= \negpredlit{\cst{b}} \cor \MAX{\negpredlit{\cst{c}}}
+      & C_5 &= \pospredlit{\cst{a}} \cor \MAX{\pospredlit{\cst{b}}} 
+      & C_6 &= \pospredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{b}}} 
+         &
+   \end{align*}
+%   \end{linenomath*}
+   Assume the simplification ordering $\cst{a} \prec \cst{b} \prec \cst{c} \prec
+   \cst{d} \prec \cst{e}$ and the selection function that chooses the last negative literal of a clause as
+   presented. Gray boxes indicate literals that can take part in superposition
+   inferences. Only two superposition inferences are
+   possible:\ from $C_3$ into $C_4$, yielding the tautology $C_{7} = \pospredlit{\cst{b}} \cor
+   \MAX{\negpredlit{\cst{b}}}$, and from $C_5$ into $C_6$, yielding $C_{8} = \pospredlit{\cst{a}} \cor
+   \MAX{\pospredlit{\cst{c}}}$. Clause $C_{7}$ is
+   clearly redundant, whereas $C_{8}$ is blocked by its first literal. If we
+   allow removing blocked clauses, the prover enters a loop: $C_{8}$ is repeatedly generated and 
+   deleted. Thus, the prover will never generate the
+   empty clause for this unsatisfiable set.
+\end{examplex}
+
+   \newcommand{\aset}{\mathcalx{A}}
+   \newcommand{\pset}{\mathcalx{P}}
+   As with hidden tautologies, removing blocked clauses breaks the invariant of
+   the given clause procedure that all inferences between clauses in $\aset$ are
+   redundant. To see this, assume the setting of Example
+   \ref{example:bce-completeness}, and let $\pset = N$ and $\aset = \emptyset$.
+   Assume $C_1, C_2, C_3$ are moved to the active set. As there are
+   no possible inferences between them, the proof state becomes $\aset = \{C_1,
+   C_2, C_3\}$ and $\pset = \{C_4,C_5,C_6\}$. After $C_4$ is moved to
+   $\aset$, the conclusion $C_7$ is computed, but it is not added to $\pset$ as it
+   is redundant. Moving $C_5$ to $\aset$ produces no new conclusions, but after
+   $C_6$ is moved, $C_{8}$ is produced. However, if we allow eliminating blocked
+   clauses, it will not be added to $\pset$ as it is blocked. The prover then
+   terminates with $\aset=N$ and $\pset=\emptyset$, even though the original set
+   $N$ is unsatisfiable.
+
+Although using BCE as inprocessing breaks the completeness of superposition in
+general, it is conceivable that a well-behaved fragment of BCE might exist.
+This could be investigated further.
+
+% @MARIJN : Can you please motivate this part a bit better and
+%           add the necessary references? --PV
+% In particular, why do we care about short proofs (in FOL)? --JB
+% Short proofs are often easier to generate, process, and understand than long
+% proofs. SAT solvers can often generate short refutational proofs by adding and
+% removing well-chosen clauses using satisfiability-preserving techniques.
+% Similarly, by adding some clauses, we can transform a first-order problem so
+% that all clauses can be removed by BCE. Since the empty clause set is trivially
+% satisfiable, this approach produces easily checkable proofs of satisfiability.
+
+Not only can BCE prevent infinite saturation (Example \ref{example:infinite-blocked-set}),
+but it can also be used to convert a finite Herbrand model
+into a certificate of clause set satisfiability. The certificate uses only
+blocked clause elimination and addition, in conjunction with a transformation
+to reduce the clause set to an empty set. This theoretical result
+explores the relationship between Herbrand models and satisfiability certificates
+based on clause elimination and addition. It is conceivable that it can form
+the basis of an efficient way to certify Herbrand models.
+
+\looseness=-1 
+In propositional logic, \emph{asymmetric literals} can be added to
+or removed from clauses, retaining the equivalence of the resulting clause set with
+the original one. Kiesl and Suda \cite{ks-2017-unif-principle} described an extension of
+this technique to first-order logic.
+Their definition of asymmetric literals can be relaxed to
+allow the addition of more literals, but the resulting set is then only
+equisatisfiable to the original one, not equivalent.
+This in turn allows us to show that a problem is satisfiable by
+reducing it to an empty problem, as is done in some SAT solvers.
+
+For the rest of this section, we work with clausal first-order logic
+\relax{without equality}. We use \relax{Herbrand models} as canonical
+representatives of first-order models, recalling that every
+satisfiable set has a Herbrand model \cite[Sect.~5.4]{mf-1996-fol}.
+
+\begin{definitionx}
+   \looseness=-1
+   A literal $L$ is a \emph{global asymmetric literal} (GAL) for a clause $C$
+   and a clause set $N$ if for every ground instance $C\sigma$ of $C$, there
+   exists a ground instance $D\varrho \cor L'\varrho$ of $D \cor L' \in N \setminus \{C\}$ such that $D\varrho
+   \subseteq C\sigma$ and $\neglit L'\varrho = L\sigma$.
+ \end{definitionx}
+\confrep{\removemyskip}{}
+
+Every asymmetric literal is GAL, but the converse does not hold:
+
+\begin{examplex}
+   \looseness=-1
+   Consider a clause $C = \pospredlit{\cst{p}(x,y)}$ and a clause set $N = \{\pospredlit{\cst{q}} \cor
+  \pospredlit{\cst{p}(\cst{a},\cst{a})}\}$. Then, $\negpredlit{\cst{q}}$ is not an asymmetric literal
+  for $C$ and $N$, but it is a GAL for $C$ and $N$.
+\end{examplex}
+
+Adding and removing GALs preserves and reflects satisfiability:
+
+\begin{theoremx}
+   \label{thm:gal}
+   If $L$ is a GAL for the clause $C$ and the clause set $N$, then the set $(N \setminus
+   \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
+\end{theoremx}
+\begin{rep}%
+\begin{proof}
+   \looseness=-1
+   Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
+   Non-trivial direction is proving that if $N'$ has a model, so does
+   $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
+   $\ourmodel$ satisfies every clause in $N$, with the possible exception of
+   $C$. Assume there exists a ground instance $C\sigma$ falsified by
+   $\ourmodel$. Since $L$ is a GAL for $C$ and $N$, then there exists a ground
+   instance $D\tau \cor L'\tau$ of $D \cor L' \in N'$ and $\neg L'\tau =
+   L\sigma$. If $\ourmodel \models
+   L'\tau$, for $C\sigma \cor L\sigma$ to be satisfied by $\ourmodel$, some
+   literal in $C\sigma$ must be satisfied by $\ourmodel$, contradicting that
+   $C\sigma$ is falsified by $\ourmodel$. If $\ourmodel \not\models L'\tau$
+   then some literal in $D\tau$ must be satisfied. As $D\tau \subseteq C\sigma$,
+   we again get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
+   as needed.\qed
+ \end{proof}
+
+\end{rep}
+\looseness=-1
+For first-order logic without equality, a clause $L \cor C$ is blocked if all its
+$L$-resolvents are tautologies \cite{ksstb-2017-blockedfol}. The $L$-resolvent
+between $L \cor C$ and $\neglit{L_1} \cor \cdots \cor \neglit{L_n} \cor
+D$ is $(C \cor D)\sigma$, where $\sigma$ is the most general unifier
+of the literals $L, L_1, \ldots, L_n$ \cite{bg-01-resolution}.
+Given a Herbrand model $\ourmodel$ of a problem, the following procedure
+removes all clauses while preserving satisfiability:
+
+\begin{enumerate}
+   \item \label{item:step1}
+   \looseness=-1
+   Let $\cst{q}$ be a fresh predicate symbol. For each atom
+   $\cst{p}(\tupleempty{s})$ in the Herbrand universe: If
+   $\ourmodel \models \cst{p}(\tupleempty{s})$, add the clause $\cst{q} \cor
+   \cst{p}(\tupleempty{s})$; otherwise, add %the clause
+   $\cst{q} \cor \neglit{\cst{p}(\tupleempty{s})}$. Adding either clause
+   preserves satisfiability as both are blocked by~$\cst{q}$.
+% \pagebreak[2]
+   \smallskip
+   \item \label{item:step2}
+   Since $\ourmodel$ is a model, for each ground instance $C\sigma$, there exists a clause
+   $\cst{q} \cor L$ with $L \in C\sigma$. We can transform $C \in N$ into
+   $C \cor \neglit{\cst{q}}$, since $\neglit{\cst{q}}$ is a GAL for $C$ and $N$.
+
+   \smallskip
+   \item \label{item:step3}
+   \looseness=-1
+   Consider the clause $\cst{q} \cor L$ added by step \ref{item:step1}. Since
+   $L$ is ground and no clause $\cst{q} \cor \neglit{L}$ was added (since $\ourmodel$ is a
+   model), the only $L$-resolvents are against clauses added by step~\ref{item:step2}.
+   Since all of those clauses contain $\neglit{\cst{q}}$, the resolvents are tautologies. 
+   Thus, each $\cst{q} \cor L$ is blocked and can be removed in turn.
+
+   \smallskip
+   \item The remaining clauses all contain the literal $\neglit{\cst{q}}$. They can be removed
+   by BCE as well.
+\end{enumerate}
+
+The procedure is limited to the first-order logic without equality, since step
+\ref{item:step3} is justified only if $L$ is a predicate literal. (Otherwise,
+$L$ cannot block clause $\cst{q} \cor L$ \cite{ksstb-2017-blockedfol}.) The
+procedure also terminates only for finite Herbrand models.
+
+% \begin{rep}
+
+\begin{examplex}
+   Consider the satisfiable clause set $N = \{\pospredlit{\cst{r}(x)} \cor
+   \pospredlit{\cst{s}(x)}{,}\; \negpredlit{\cst{r}(\cst{a})}{,}\allowbreak\;
+   \negpredlit{\cst{s}(\cst{b})}\}$ and a Herbrand model $\ourmodel$ over
+   $\{\cst{a}, \cst{b}, \cst{r}, \cst{s}\}$ such that
+   $\cst{r}(\cst{b})$ and $\cst{s}(\cst{a})$ are the only true atoms in
+   $\ourmodel$. We show how to remove all clauses in $N$ using $\ourmodel$ 
+   by following the procedure above.
+
+   Let $N_{\ourmodel} = \{ \cst{q} \cor \negpredlit{\cst{r}(\cst{a})}{,}\; \cst{q}
+   \cor \pospredlit{\cst{r}(\cst{b})}{,}\; \cst{q} \cor
+   \pospredlit{\cst{s}(\cst{a})}{,}\; \cst{q} \cor \negpredlit{\cst{s}(\cst{b})}
+   \}$. We set $N \gets N \cup N_{\ourmodel}$. This preserves satisfiability
+   since all clauses in $N_{\ourmodel}$ are blocked. It is easy to check that
+   $\negpredlit{\cst{q}}$ is GAL for every clause in $N \setminus N_{\ourmodel}$. The
+   only substitutions that need to be considered are $\{x \mapsto \cst{a} \}$ and $\{x \mapsto
+   \cst{b} \}$ for $\pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}$.
+   So we set $N \gets \{\negpredlit{\cst{q}} \cor \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
+   \negpredlit{\cst{q}} \cor \negpredlit{\cst{r}(\cst{a})}{,}\;
+   \negpredlit{\cst{q}} \cor \negpredlit{\cst{s}(\cst{b})}\} \cup N_{\ourmodel}$.
+   Clearly, all clauses in $N_{\ourmodel}$ are blocked, so we set $N \gets N \setminus
+   N_{\ourmodel}$. All clauses remaining in $N$ have a literal $\negpredlit{\cst{q}}$
+   and can be removed, leaving $N$ empty as desired.
+\end{examplex}
+
+% The procedure works only for first-order logic without equality, because it
+% relies on adding and removing blocked clauses. In step~\ref{item:step3}, binary
+% clauses $\cst{q} \cor L$ are blocked by $L$. If $L$ were a
+% functional literal, by definition $\cst{q} \cor L$ would not be blocked.
+% \end{rep}
+
+% \looseness=-1
+% For many problems model can be described only using infinite Herbrand universe and the described technique for
+% certifying satisfiability is not applicable. However, in cases of finite
+% Herbrand universes, provers that support GAL addition and blocked clause
+% elimination and addition can certify that a given interpretation is
+% indeed a Herbrand model.
+
+\section{Implementation}
+\label{sec:implementation}
+% {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}
+
+\newcommand{\impmap}{\textit{Imp}}
+
+Hidden-literal-based, predicate, \begin{qle}quasipure, \end{qle}and blocked clause
+elimination all admit efficient implementations in a superposition
+prover. In this section, we describe how to implement the first
+\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
+techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
+All techniques are implemented in the Zipperposition prover
+\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
+of improvements to superposition, but it implements many of the most successful
+heuristics from the E prover \cite{scv-19-e23} and has recently become
+quite competitive \cite{gs-19-casc27}.
+
+\ourparagraph{Hidden-Literal-Based Elimination}
+% 
+% Heule et al.\ \cite{hjb-2010-cl-elim} show how hidden tautologies can be removed
+% without creating the binary implication graph. For each literal~$L$ occurring in
+% the clause set $N$, they compute the set $\PHL(L,N)$. Then for each $\neglit{L}'
+% \in \PHL(L,N)$, they label $L'$ with $L$. A clause $C$ is a hidden tautology if
+% there are two literals $L,L' \in C$ such that $L'$ is labeled with $L$. The
+% described approach can be generalized to first-order logic and used to implement
+% all rules from Sect.~\ref{sec:hidden-literal-based-elimination}.
+% 
+For HLBE, an efficient representation of $\HL(L,N)$ is crucial. Because
+this set may be infinite, we underapproximate it by
+restricting the length of the
+transitive chains via a parameter $K_\mathrm{len}$. Given the current
+clause set $N$, the finite map $\impmap[L']$ associates with each literal $L'$
+a set of pairs $(L,M)$ such that $L' \bigfimpl{N}^k L$, where $k \leq
+K_\mathrm{len}$ and $M$ is the multiset of clauses used to derive $L' \bigfimpl{N}^k L$.
+Moreover, we consider only transitions of type (1) (as per
+Definition~\ref{def:hl-fo}).
+%Intuitively,
+%$\impmap[L']$ contains all literals which can be reached from $L'$ following $k$
+%edges of the generalized binary implication graph.
+The following algorithm
+maintains $\impmap$ dynamically, updating it as the prover derives and deletes
+clauses.
+It depends on the global variable $\impmap$ and the parameters
+$K_\mathrm{len}$ and $K_{\mathrm{imp}}$.
+
+% By choosing a limit $K_\mathrm{len}$ we can create a map $\impmap[L]$, which maps a literal $L$
+% to a finite a set of pairs $(L', M)$ such that
+% $\impk{L}{L'}{i}{M}$ where $i \leq K_\mathrm{len}$. It is important to record clauses used
+% in deriving the implication, as this information is necessary to construct the
+% proof. Mapping $\impmap$ is constructed so that whenever $(L', M) \in
+% \impmap[L]$, then $L \in \HL(L',M)$.
+% Proof by very simple induction
+
+\looseness=-1
+
+\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
+\algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}
+\algrenewcommand\algorithmicindent{0.75em}
+
+%%% TYPESETTING: hack
+\vskip\abovedisplayskip
+\kern-.5\jot
+
+%% \begin{nolinenumbers}%\begin{figure}[h!]
+\begin{algorithmic}[5]
+
+%   \Procedure{DecomposeNegLit}{$c$, $C$}
+%   \State $\vn{res} \gets \{(c, C)\}$
+%   \ForAll{non-empty contexts $u$ such that $c = \neqlit{u[s]}{u[t]}$}
+%       \State $\vn{res} \gets \{ (\neqlit{s}{t}, C) \} \cup \vn{res}$
+%   \EndFor
+%   \State \Return $\vn{res}$
+%   \EndProcedure
+
+%  \kern1\jot
+
+  \Procedure{AddImplication}{$L_\mathrm{a}{,}\> L_\mathrm{c}{,}\> C$}
+    \If{$\impmap[L_\mathrm{a}\sigma] \not = \emptyset$ for some renaming $\sigma$} \label{statement:alpha}
+      \State $(L_\mathrm{a}, L_\mathrm{c}) \gets (L_\mathrm{a}\sigma, L_\mathrm{c}\sigma)$
+    \EndIf
+
+   %  \vspace{0.1em}
+    \If{there are no $L,L',M,\sigma$ such that $(L',M) \in \impmap[L]$,\label{statement:more-general}%
+        \\\noindent\kern3.5ex $L\sigma = L_\mathrm{a}$, and $L'\sigma=L_\mathrm{c}$} 
+      % \State $\vn{Inst} \gets \{(\sigma, M) \mid (L_\mathrm{c}\sigma, M) \in \impmap[L_\mathrm{a}\sigma]\}$
+      \ForAll{$(\sigma, M)$ such that $(L_\mathrm{c}\sigma, M) \in \impmap[L_\mathrm{a}\sigma]$} \label{statement:instances}
+        \State {erase all $(L',M')$ such that $M \subseteq M'$ from $\impmap[L_{\mathrm{a}}\sigma]$}
+      \EndFor
+
+      \ForAll{$L$ such that $(L', M) \in \impmap[L]$\\ \noindent\kern10ex and $L_\mathrm{a}\sigma = L'$ for some $\sigma$} \label{statement:transitivity}
+        \If{$|M| < K_\mathrm{len}$}
+          \State $\impmap[L] \gets \impmap[L] \cup \{(L_\mathrm{c}\sigma, M \uplus \{C\})\} $
+        \EndIf
+      \EndFor
+
+      \ForAll{$L$ such that $\impmap[L] \not= \emptyset$\\\noindent\kern10ex and $L\sigma = L_\mathrm{c}$ for some $\sigma$} \label{statement:triggered}
+         % \If{$|M| < K_\mathrm{len}-1$}
+         \State $\vn{Concl} \gets \{(L'\sigma, M \uplus \{\mathit{C}\}) \mid$%
+         \\\noindent\kern14.7ex $(L',M) \in \impmap[L], |M| < K_\mathrm{len}  \}$
+         \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup  \vn{Concl}$
+         % (L'\sigma, M \cup \{\mathit{C}\}) $
+         % \EndIf
+      \EndFor
+
+      \State $\vn{Congr} \gets \{
+         (\neqlit{s}{t}, \{C\}) \mid \exists u.\, L_\mathrm{c} = \neqlit{u[s]}{u[t]}
+         \}$
+      \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup \{ (L_\mathrm{c},\{C\}) \} \cup \vn{Congr}$ \label{statement:decomposition}
+
+    \EndIf
+
+  \EndProcedure
+
+  \kern1\jot
+
+  \Procedure{TrackClause}{$C$}
+    \If{$C$ = $L_1 \cor L_2$}
+      \State \textsc{AddImplication}($\neglit{L_1}$, $L_2$, $C$)
+      \State \textsc{AddImplication}($\neglit{L_2}$, $L_1$, $C$)
+      \If{$L_2 = \neglit L_1\sigma$ for some nonidempotent $\sigma$}  \label{statement:self-impl}
+      \ForAll{$i \gets 1$  to $K_\mathrm{imp}$}
+         \State $L_2 \gets L_2\sigma$
+         \State \textsc{AddImplication}($\neglit L_1$, $L_2$, $C$)
+      \EndFor
+      \EndIf
+    \EndIf
+  \EndProcedure
+
+  \kern1\jot
+
+  \Procedure{UntrackClause}{$C$}
+   %  \ForAll {$L_\mathrm{a}, L_\mathrm{c}, M$ such that $ (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$}
+   \ForAll{$L_\mathrm{a}, L_\mathrm{c}, M$ such that $(L_\mathrm{c},M) \in \impmap[L_\mathrm{a}]$}
+      \If{$C \in M$}
+         \State erase $(L_\mathrm{c},M)$ from $\impmap[L_\mathrm{a}]$
+      \EndIf
+      % \EndFor
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+%\end{nolinenumbers}
+%\end{figure}
+
+%% TYPESETTING: hack
+\vskip-1.5\jot
+\vskip\belowdisplayskip
+
+The algorithm views a clause $L \cor L'$ as two implications
+$\neglit{L} \impl L'$ and $\neglit{L'} \impl L$.
+%
+It stores only one entry for all literals equal up to variable renaming (line
+\ref{statement:alpha}). Each implication $L_\mathrm{a} \impl L_\mathrm{c}$
+represented by the clause is stored only if its generalization is not present in
+$\impmap$ (line~\ref{statement:more-general}). Conversely, all instances of the
+implication are removed (line~\ref{statement:instances}).
+
+Next, the algorithm
+finds each implication stored in $\impmap$ that can be linked to $L_\mathrm{a} \impl L_\mathrm{c}$: Either
+$L_\mathrm{c}$ becomes the new consequent (line~\ref{statement:transitivity}) or
+$L_\mathrm{a}$ becomes the new antecedent (line~\ref{statement:triggered}). If
+$L_\mathrm{c}$ can be decomposed into $\neqlit{u[s]}{u[t]}$, rule~(3) of
+Definition \ref{def:hl-fo} allows us to store $\neqlit{s}{t}$ in
+$\impmap[L_\mathrm{a}]$ (line~\ref{statement:decomposition}). This is an
+exception to the idea that transitive chains should use only rule~(1). The
+application of rule~(3) does not count toward the bound $K_\mathrm{len}$.
+If $L_\mathrm{a}$ is of the form $\eqlit{u[s]}{u[t]}$, then
+$\impmap$ could be extended so that $\impmap[\eqlit{s}{t}] =
+\impmap[L_\mathrm{a}]$, but this would substantially increase $\impmap$'s
+memory footprint.
+
+In first-order logic, different instances of the same clause can
+be used along a transitive chain. For example, the clause $C =
+\negpredlit{\cst{p}(x)} \cor \pospredlit{\cst{p}(\cst{f}(x))}$ induces $\cst{p}(x) \bigfimpl{M}^i \cst{p}(\cst{f}^i(x))$ for all $i$.
+The algorithm discovers such self-implications (line~\ref{statement:self-impl}): For
+each clause~$C$ of the form $\neglit{L} \cor L\sigma$, where $\sigma$ is
+some nonidempotent substitution, the entires $(L\sigma^2, \{C\}),\allowbreak
+\ldots,\allowbreak (L\sigma^{K_{\mathrm{imp}} + 1}, \{C\})$ are added to
+$\impmap[L]$, where $K_{\mathrm{imp}}$ is a parameter.
+
+\newcommand{\concl}{\ensuremath{\impmap^{-1}}}
+
+\looseness=-1
+To track and untrack clauses efficiently, we implement the mapping $\impmap$ as a
+nonperfect discrimination tree \cite{rsv-2001-term-indexing}. Given a query
+literal $L$, this indexing data structure efficiently finds all literals $L'$ such that
+for some $\sigma$, $L'\sigma = L$ and $\impmap[L'] \not= \emptyset$. We can use
+it to optimize all lookups except the one on line~\ref{statement:transitivity}.
+For this remaining lookup, we add an
+index \concl{} that inverts $\impmap$, i.e., $\concl[L] = \{ L' \mid
+\impmap[L'] = (L,M) \text{ for some } M \}$. To avoid sequentially going through
+all entries in $\impmap$ when the prover deletes them, for each clause $C$
+we keep track of each literal $L$ such that $C$ appears in $\impmap[L]$.
+Finally, we limit the number of entries stored in $\impmap[L]$ -- by default,
+up to 48 pairs in each $\impmap[L]$ are stored.
+
+\newcommand{\unitset}{\ensuremath{\textit{Unit}}}
+\begin{conf} 
+\looseness=-1   
+Rules \infname{HLE} and \infname{HTR} have a simple
+implementation based on $\impmap$ lookups. To implement \infname{UnitHLE} and
+\infname{UnitHTR}, we maintain the index \unitset, containing literals
+$L_\mathrm{c}\sigma$, such that $(L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$ for
+some $M$ and $L_\mathrm{a}$ and $\sigma$ is the most general unifier of $L'$ and
+$L_\mathrm{a}$, for some unit clause $\{ L' \}$. The implementation
+of \infname{FLE} and \infname{FLR} also uses \unitset:
+When $(L',M)$ is added to $\impmap[L]$, we
+check if $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $\neglit{L}$
+is added to $\unitset$.
+\end{conf}
+\begin{rep}
+
+   To implement the \infname{HLE} rule, we use $\impmap[L]$ as
+   follows: Given a clause $C = L \cor L' \cor C'$, if there are two literals
+   $L_1,L_2$ and a substitution $\sigma$ such that $(L_2, M) \in \impmap[L_1]$,
+   $C \not\in M$,  $L_1\sigma = L$, and $L_2\sigma = L'$, we remove $L$ from $C$.
+   Literal $L$ can also be removed if $L_1\sigma = \neglit{L'}$ and $L_2\sigma = \neglit{L}$. 
+   Rule \infname{HTR} is implemented analogously.
+
+% @PETAR: Used to say "UnitHTR rules rely", but there's only one. Double-check. --JB
+   The \infname{UnitHTR} rule relies on maintaining the index $\unitset$, which is
+   built as follows. Whenever the prover derives a unit clause $C = \{ L \}$, we
+   find all entries $L_\mathrm{a}$ in $\impmap$ such that $L_\mathrm{a}$ and $L$
+   are unifiable with the most general unifier $\sigma$. Then, we set $\unitset
+   \gets \unitset \cup \{ (L_\mathrm{c}\sigma, M \cup \{ C \} ) \mid
+   (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}] \}$. Given a clause $L \cor C'$
+   we apply \infname{UnitHLE} by looking for $(L', M) \in \unitset$ such that
+   $L'\sigma = \neglit{L}$, for some substitution $\sigma$; we apply
+   \infname{UnitHTR} by looking for $L'$ such that $L'\sigma = L$.  The sets
+   stored together with literals in $\unitset$ are used for building the proof
+   object and to remove literals from $\unitset$ once a clause from the given
+   set becomes redundant.
+
+   The same data structure is used
+   for supporting \infname{FLE} and \infname{FLR}. When $(L',M)$ is added to $\impmap[L]$, we
+   check whether $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $(\neglit{L}, M \cup M')$
+   is added to $\unitset$.
+\end{rep}
+
+% % Using this algorithm, hidden-literal-based elimination can be implemented
+% % efficiently. We implement them as simplifications, used during saturation.
+% \begin{rep}
+% We implement \infname{ShallowHLE} by checking if there are two literals $L,L'
+% \in C$ such that for some substitution $\sigma$, and literals $L_1$ and $L_2$,
+% $(L_2, M) \in \impmap[L_1]$ and $L_1\sigma = L$, $L_2\sigma = L'$, $L$ can be
+% removed from $C$. When $(L',M)$ is added to $\impmap[L]$, Zipperposition checks
+% if $(\neglit{L'},M) \in \impmap[L]$ for some $M$. If so, it notes that there
+% are complementary literals in $\HL(\neglit{L},M)$. This information is used to
+% implement \infname{FLD}. For \infname{HTR}, a check whether there are $L,L' \in
+% C$ such that $(L_1, M) \in \impmap[L_0]$, where $\neglit{L_0\sigma} = L$ and
+% $L_1\sigma = L'$ is performed. Rule \infname{DeepHLE} is not implemented.
+
+% Rules \infname{UnitHLE} and \infname{UnitHTR} are implemented as well. For
+% \infname{UnitHLE}, given a clause $L \cor C$ we do as follows. First we try to
+% find a literal $L_0$ such that $\neglit{L_0\sigma} = L$ for some $\sigma$ and
+% $(L_1,M) \in \impmap[L_0]$ such that $L_1\sigma$ is subsumed by a unit clause
+% indexed by Zipperposition. If this does not succeed, we try finding a literal
+% $L_0$ such that $L_0\sigma = L$ for some $\sigma$ and $(L_1,M) \in \impmap[L_0]$
+% such that $\neglit{L_1}$ is subsumed by a unit clause. Rule \infname{UnitHTR} is
+% implemented analogously.
+% \end{rep}
+
+In propositional logic, the conventional approach constructs the
+\emph{binary implication graph} for the clause set~$N$
+\cite{hjb-2011-big-simplification}, with edges $(\neglit{L}, L')$ and
+$(\neglit{L'}, L)$ whenever $L \cor L' \in N.$ To avoid traversing the graph
+repeatedly, solvers rely on timestamps to discover connections between
+literals. This relies on syntactic literal comparisons, which is very
+fast in propositional logic but not in first-order logic, because of
+substitutions and congruence.
+
+\ourparagraph{Predicate Elimination}
+% 
+To implement portfolio predicate elimination, we maintain a record for each
+predicate symbol $\cst{p}$ occurring in the problem with the
+following fields:\
+set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
+$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
+once. These records are kept in a priority queue, prioritized by properties such
+as presence of definition sets and number of estimated resolutions. If
+$\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
+eliminate it by removing all the clauses stored in $\cst{p}$'s record from the
+proof state and by adding flat resolvents to the passive set.
+Eliminating a symbol might make another symbol eligible.
+%In this case, the updated record is put on the processing queue.
+
+As an optimization, predicate elimination keeps track only of symbols that appear
+at most $K_\mathrm{occ}$~times in the clause set. For inprocessing, we use
+signals that the prover emits whenever a clause is added to or removed from
+the proof state and update the records. At the beginning of the
+1st, $(K_\mathrm{iter} + 1)$st, $(2 K_\mathrm{iter} + 1)$st, $\dots{}$ iteration of the given clause procedure's
+loop body, predicate elimination is systematically applied to the entire proof
+state. The first application of inprocessing amounts to preprocessing. By
+default, $K_\mathrm{occ} = 512$ and $K_\mathrm{iter}=10$. The same ideas and
+limits apply for blocked clause elimination.
+
+The most important novel aspect of our predicate elimination implementation is
+recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
+as follows:
+\begin{enumerate}
+   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
+   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
+   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise continue.
+
+   \smallskip
+   \item Rename all clauses in $G$ so that their only variables are~$\tupleempty{x}$.
+
+   \smallskip
+   \item\label{step:encoding} Let $\lfloor a \rfloor$ be a function that assigns
+   a propositional variable to each atom $a$. This function is lifted to
+   literals by assigning $\lfloor \neg a \rfloor  = \neg x$, if $\lfloor a
+   \rfloor = x$, and to clauses pointwise. Furthermore,  
+   let $E = \{ \lfloor C' \rfloor \mid
+   \arbpredlit{\cst{p}(\tupleempty{x})} \cor C' \in G \}$. If $E$ is satisfiable,
+   report failure. Else, let $E'$ be the unsatisfiable core of $E$
+   and $G'$ the set of corresponding first-order clauses and continue.
+
+   \smallskip
+   \item If all resolvents in $G'_{\pospredlit{\cst{p}}} \flatres_{\!\cst{p}}
+   G'_{\negpredlit{\cst{p}}}$ are tautologies, then $G'$ is the definition set
+   for symbol $\cst{p}$. Else, report failure.
+\end{enumerate}
+
+The invalidity of set $E$ from step \ref{step:encoding} is checked using a SAT
+solver, which is already integrated in Zipperposition. As modern theorem provers
+(such as E or Vampire) also use SAT solvers, the method can easily be
+implemented.
+
+During experimentation, we noticed that recognizing definitions of symbols that
+occur in the conjecture often harms performance. Thus, Zipperposition
+recognizes definitions only for the remaining symbols.
+
+\begin{qle}
+\ourparagraph{Quasipure Literal Elimination}
+
+A simple and efficient implementation of QLE uses a SAT solver. Let $N$ be a
+finite $\mathrm{\Sigma}$-clause set. Without loss of generality, we can assume
+that $\mathrm{\Sigma}$ is finite. The signature of the SAT problem consists of
+the variables $\cst{p}^+, \cst{p}^-$ for each predicate symbol $\cst{p}$ in
+$\mathrm{\Sigma}$, where $\cst{p}^s$ means ``$\cst{p}$ is quasipure with
+polarity~$s$.'' The SAT problem consists of the following clauses:
+%
+\begin{enumerate}
+\itemsep1\jot
+\item For each clause $\arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
+  \arbpredlit{\cst{q}_n(\ldots)} \allowbreak\lor C$ contained in $N$, where
+  the first $n$~literals have polarities $s_1,\dots,s_n$, respectively,
+  and $C$ consists of functional literals,
+  generate $n$~clauses of the form
+  \[\cst{q}_1^{s_1}\lor\cdots\lor\cst{q}_{i-1}^{s_{i-1}}
+    \lor
+    \neglit{\cst{q}_i^{-s_i}}
+    \lor
+    \cst{q}_{i+1}^{s_{i+1}}\lor\cdots\lor\cst{q}_n^{s_n}
+   \]
+   where $-s$ flips the polarity $s$. Such a clause ensures that whenever a
+   literal $\arbpredlit{\cst{q}_i(\ldots)}$ has the wrong polarity according to
+   the current variable assignment, one of the other literals must be quasipure.
+
+\item For each $\cst{p}$ in $\Sigma$, generate the clause $\lnot \cst{p}^+ \lor
+  \lnot \cst{p}^-$. It ensures that a single polarity is assigned to a quasipure
+  predicate.
+
+\item Generate the clause $\cst{p}_1^+ \lor \cst{p}_1^- \lor \cdots \lor
+  \cst{p}_n^+ \lor \cst{p}_n^-$, where $\{\cst{p}_1,\dots,\cst{p}_n\}$ are the
+  predicate symbols in $\mathrm{\Sigma}$. It tells the SAT solver to look for a
+  nontrivial solution, in which at least one predicate is quasipure.
+\end{enumerate}
+
+From a satisfying assignment, we can easily read off a predicate set $P$ and a
+polarity map $s$. The process can be iterated until we reach a maximal solution,
+at which point the SAT solver returns a verdict of ``unsatisfiable.''
+
+So far, we have been unable to prove that quasipurity is NP-complete. Given the
+nondeterministic nature of the naive procedure, we suspect that it is, but we
+have not found a reduction from 3SAT to quasipurity. Thus, it is unclear whether
+our use of a SAT solver is fully satisfactory from a theoretical point of view,
+even though it works perfectly well in practice.
+\end{qle}
+
+\section{Evaluation}
+\label{sec:evaluation}
+
+% {\bf\large TODO: Evaluate \textit{real} PPE in Figure \ref{fig:preprocess} and discuss SPPE and PPE.
+% References to PPE are broken now. }
+
+\newcommand{\baseres}[0]{7897}
+% \newcommand{\baseresplusfive}[0]{1563}
+
+\newcommand{\resnum}[1]{
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \ifnumcomp{\result}{>}{0}{$+\result$}{$\result$}%
+}
+\newcommand{\resnumphantom}[1]{%
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{$0$}{${+}\result$}}{\phantom{$0$}{${-}\negresult$}}%
+}
+\newcommand{\resnumbold}[1]{
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{${+}\result$}\llap{$\mathbf{\boldsymbol{+}\result}$}}{\phantom{${-}\negresult$}$\llap{\mathbf{\boldsymbol{-}\negresult}}$}%
+}
+\newcommand{\resnumboldphantom}[1]{%
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{$0{+}\result$}\llap{$\mathbf{\phantom{0}{\boldsymbol{+}}\result}$}}{\phantom{$0{-}\negresult$}$\llap{\mathbf{\phantom{0}{\boldsymbol{-}}\negresult}}$}%
+}
+\newcommand{\evaluating}[0]{\textbf{?}}
+
+\looseness=-1
+We measure the impact of our elimination techniques for various values of their
+parameters. As a baseline, we use Zipperposition's first-order portfolio mode,
+which runs the prover in 13~configurations of
+heuristic parameters in consecutive time slices. None of these configurations
+use our new techniques. To evaluate a given parameter value, we fix
+it across all 13~configurations and compare the results with the baseline.
+
+\looseness=-1
+The benchmark set consists of all 13\,495 CNF and FOF TPTP 7.3.0 theorems
+\cite{gs-17-tptp}.
+%
+The experiments were carried out on StarExec servers \cite{sst-14-starexec}
+equipped with Intel Xeon E5-2609 CPUs clocked at 2.40 GHz. The portfolio mode
+uses a single CPU core with a CPU time limit of 180~s. The base
+configuration solves \NumberOK{\baseres} problems. The values in the tables
+indicate the number of problems solved minus \NumberOK{\baseres}. Thus,
+positive numbers indicate gains over the baseline. The best result is shown in
+bold.
+
+\ourparagraph{Hidden-Literal-Based Elimination}
+% 
+The first experiments use all implemented HLBE rules.
+To avoid overburdening Zipperposition, we can enable an option to limit the
+number of tracked clauses for hidden literals. Once the limit has been reached, any
+request for tracking a clause will be rejected until a tracked
+clause is deleted. We can choose which kind of clauses are
+tracked:\ only clauses from the active set $\mathcalx{A}$,
+only clauses from the passive set $\mathcalx{P}$, or both.
+We also vary the maximal implication chain length $K_\mathrm{len}$ and the
+number of computed self-implications $K_\mathrm{imp}$.
+
+In Zipperposition, every lookup for instances or generalizations of
+$\eqlit{s}{t}$ must be done once for each orientation of the equation.
+To avoid this inefficiency, and also because the implementation of hidden literals does not fully
+exploit congruence, we can disable\pagebreak[2] tracking clauses with at least one
+functional literal. Clauses containing functional literals can then still be
+simplified.
+
+Figures~\ref{fig:hle-clauses-neq} and \ref{fig:hle-clauses-eq}
+%~\ref{fig:hle}
+show the results, without and with functional literal tracking enabled,
+for $K_\mathrm{len} = 2$ and $K_\mathrm{imp} = 0$.%
+The columns specify different limits on the number of tracked clauses, with $\infty$
+denoting that no limit is imposed. The rows represent different kinds of tracked clauses.
+The results suggest that tracking functional literals is \NumberOK{not} worth the effort
+but that tracking predicate literals \NumberOK{is}. The best improvement is observed when
+both active and passive clauses are tracked. Normally DISCOUNT-loop
+provers \cite{adf-1995-discount} such as Zipperposition do not simplify
+active clauses using passive clauses, but here we see that this can be effective.
+%
+%--Raw data shows that when up to 250 clauses in either set are
+%--tracked, tracking functional literals helps solve \NumberNOK{5} problems that
+%--cannot be solved by a configuration that tracks only predicate literals or the
+%--base configuration. This suggests that a more optimized indexing implementation
+%--would help our technique reach its full potential.
+Figure~\ref{fig:km-kimp} shows the impact of varying $K_\mathrm{len}$ and
+$K_\mathrm{imp}$, when 500 clauses from the entire proof state are tracked.
+These results suggest that computing long implication chains \NumberOK{is counterproductive.}
+% Furthermore, raw data reveals that there is
+% no gain in this case: not a single problem that configuration with $K_\mathrm{len}=2$ and
+% $K_\mathrm{imp}=0$ cannot solve is solved when $K_\mathrm{imp}K_\mathrm{len}$ is set to 8 and
+% $K_\mathrm{imp}$ to 0.
+
+\begin{figure}[t]%
+\small
+\centering
+\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
+                     & \multicolumn{4}{c}{Tracked clauses} \\
+                     & $250$        & $500$         & $1000$       &$\infty$ \\ \midrule 
+\strut Active       & \resnum{7883}  & \resnum{7881}   & \resnum{7889}  & \resnum{7885}    \\
+\strut Passive      & \resnumphantom{7904}  & \resnum{7907}   & \resnum{7902}  & \resnum{7862}    \\
+\strut Both         & \resnumbold{7909}  & \resnum{7907}   & \resnum{7904}  & \resnum{7852} \\     
+\end{tabular}
+\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when only predicate literals are tracked}
+\label{fig:hle-clauses-neq}
+\end{figure}
+\begin{figure}[t]%
+\small
+\centering
+\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
+                     & \multicolumn{4}{c}{Tracked clauses} \\
+                     & $250$                   & $500$                 & $1000$                    & $\infty$ \\ \midrule 
+\strut Active       & \resnum{7887}            & \resnum{7883}         & \resnumphantom{7889}      & \resnum{7879}    \\
+\strut Passive      & \resnumphantom{7892}     & \resnumphantom{7892}  & \resnum{7883}             & \resnum{7826}    \\
+\strut Both         & \resnumboldphantom{7899} & \resnumphantom{7896}  & \resnumphantom{7889}      & \resnum{7818} \\ 
+\end{tabular}
+\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when all literals are tracked}
+\label{fig:hle-clauses-eq}
+\end{figure}
+
+\begin{figure}[t]
+\small
+\centering
+\def\arraystretch{1.1}%
+\begin{tabular}[t]{l@{\hskip 2em}c@{\hskip 1.5em}c@{\hskip1.5em}c@{\hskip 1.5em}c} 
+                             & \multicolumn{4}{c}{Chain length $K_\mathrm{len}$} \\
+                             & $1$                   & $2$                 & $4$                   & $8$ \\ \midrule 
+   $K_\mathrm{imp}=0$        & \resnum{7906}         & \resnum{7907}       & \resnum{7904}         & \resnum{7902}    \\
+   $K_\mathrm{imp}=1$        & \resnum{7902}         & \resnumbold{7908}       & \resnum{7904}         & \resnum{7901}    \\
+   $K_\mathrm{imp}=2$        & \resnum{7903}         & \resnumbold{7908}       & \resnum{7905}         & \resnum{7905}    \\
+\end{tabular}
+\caption{Impact of the parameters $K_\mathrm{len}$ and $K_\mathrm{imp}$ on HLBE performance}
+\label{fig:km-kimp}
+\end{figure}
+
+\begin{figure}[t!]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
+                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
+                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
+   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
+   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
+   \end{tabular}
+   \caption{Impact of the choice of criterion on predicate elimination performance}
+   \label{fig:criterion}
+\end{figure}
+
+\begin{figure}[t]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
+                         &               &               &                &               &                    & \smash{HLBE} \\[-1pt] %% TYPESETTING
+                         &               &               & \smash{SPE}    &               & \smash{PPE}        & \smash{+PPE} \\[-1pt] %% TYPESETTING
+                         & BCE           & SPE           & +BCE           & PPE           & +BCE               & +BCE           \\ \midrule 
+   \strut Preprocessing  & \resnum{7927} & \resnum{8051} & \resnum{8056}  & \resnum{8057} & \resnumbold{8063}  & \resnum{8059} \\
+   \strut Inprocessing   & \resnum{7849} & \resnum{8037} & \resnum{8024}  & \resnum{8043} & \resnum{8028}      & \resnum{8024}     \\
+   \end{tabular}
+   \caption{Performance of predicate and blocked clause elimination}
+   \label{fig:preprocess}
+\end{figure}
+
+\begin{qle}
+\begingroup
+\begin{figure}[t]
+\color{red}
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
+                         &               &               & \smash{QLE}    \\[-1pt] %% TYPESETTING
+                         & PLE           & QLE           & +PPE           \\ \midrule 
+   \strut Preprocessing  & \resnum{9999} & \resnum{9999} & \resnum{9999} \\
+   \strut Inprocessing   & \resnum{9999} & \resnum{9999} & \resnum{9999} \\
+   \end{tabular}
+   \caption{Performance of pure and quasipure literal elimination
+    (TODO: Compute numbers)}
+   \label{fig:ple-qle}
+\end{figure}
+\end{qle}
+
+\renewcommand{\baseres}{856}
+
+\begin{figure}[t]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1em}c} \\
+                         &              &              &                   &              &              & \smash{HLBE}   \\[-1pt] 
+                         &              &              & \smash{SPE}       &              & \smash{PPE}  & \smash{+PPE}   \\[-1pt] 
+                         & BCE          & SPE          & +BCE              & PPE          & +BCE          & \smash{+BCE}    \\ \midrule 
+   \strut Preprocessing  & \resnum{885} & \resnum{902} & \resnumbold{916}  & \resnum{903} & \resnum{915} & \resnum{911}  \\
+   \end{tabular}
+   \caption{Performance of predicate and blocked clause elimination for establishing satisfiability}
+   \label{fig:bce-sat}
+\end{figure}
+
+\ourparagraph{Predicate and Blocked Clause Elimination}
+%
+% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
+% \textit{Do not forget to mention SPE was already implemented and }
+For defined predicate elimination, the number of resolvents grows
+exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
+computation, we limit the applicability of PPE to proof states for which $\cst{p}$
+is singular. According to our informal experiments, full PPE,
+without this restriction, generally performs less well.
+
+\begin{qle}\color{red} (TODO: Add QLE)\color{black}\end{qle}
+
+Predicate elimination can be done using Khasidashvili and Korovin's
+criterion (K\&K) or using our relaxed criterion with different values of
+$K_\mathrm{tol}$. Figure~\ref{fig:criterion} shows the results
+for SPE and PPE used as preprocessors.
+Our numbers corroborate Khasidashvili and Korovin's findings:
+SPE with K\&K proves \NumberOK{70} more problems than the base, a
+\NumberOK{0.9\%} increase, comparable to the 1.8\% they observe when
+they combine SPE with additional preprocessing.
+Remarkably, the number of additional proved problems more than
+\NumberOK{doubles} when we use our criterion with $K_\mathrm{tol} > 0$, for both SPE and PPE.
+
+\looseness=-1
+Although this is not evident in Figure~\ref{fig:criterion}, varying
+$K_\mathrm{tol}$ substantially changes the set of problems solved. For example,
+when $K_\mathrm{tol}=0$, SPE proves \NumberOK{60} theorems not proved using
+$K_\mathrm{tol}=50$. The effect weakens as $K_\mathrm{tol}$ grows. When
+$K_\mathrm{tol}=100$, SPE proves only \NumberOK{13} problems not found when
+$K_\mathrm{tol}=200$. Similarly, the set of problems proved by SPE and PPE
+differs: When $K_\mathrm{tol}=25$, \NumberOK{14} problems are proved by PPE but
+missed by SPE. Recognizing definition sets is useful:
+PPE \NumberOK{outperforms} SPE regardless of the criterion.
+
+% \looseness=-1 Having implemented those techniques, we are in position to answer
+% two interesting questions: {How helpful are definition clauses in defined
+% predicate elimination (DPE)?} and {What are the effects of performing blocked
+% clause elimination (BCE) and DPE until fixpoint (BD)?}. The second question was
+% also of interest to Kiesl et al.\ \cite{ksstb-2017-blockedfol}, but was left
+% unanswered as they did not implement predicate elimination.
+
+\looseness=-1
+Performing BCE and variable elimination until fixpoint increases the performance
+of SAT solvers~\cite{jbh-10-BCE}. We can check whether the same holds for superposition provers.
+In this experiment, we use the relaxed criterion with
+$K_\mathrm{tol}=25$ and HLBE which tracks up
+to 500 clauses from any clause set, $K_\mathrm{len}=2$, and
+$K_{\mathrm{imp}}=0$. We use each technique as preprocessing and inprocessing.
+
+The results are summarized in Figure~\ref{fig:preprocess}, where
+the +~sign denotes the combination of techniques.
+We confirm the results obtained by Kiesl et al.\ about the performance of BCE
+as preprocessing:
+It helps prove \NumberOK{30} more problems from our benchmark set, increasing the
+success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained Kiesl et al.
+%
+Using BCE as inprocessing, however, hurts performance, presumably because of
+its incompatibility with the redundancy criterion.
+
+\looseness=-1%
+For preprocessing, the combinations SPE+BCE and PPE+BCE performed roughly \NumberOK{on a
+par} with SPE and PPE, respectively. This stands in contrast to the situation
+with SAT solvers, where such a combination usually helps. It is also worth noting
+that the inprocessing techniques never outperform their preprocessing
+counterparts.
+%
+The last column shows that combining HLBE with other elimination techniques
+\NumberOK{overburdens} the prover.
+
+\begin{qle}
+\ourparagraph{Pure and Quasipure Literal Elimination}
+\color{red}
+Pure literal elimination is a known technique, but quasipure literals are new,
+so it is quite interesting to find out whether they bring benefits on
+benchmarks. Figure~\ref{ig:ple-qle} considers PLE, QLE, and a combination
+of QLE and portfolio predicate elimination, applied in that order. All three
+modes are tried both as preprocessing and as inprocessing.
+
+We see TODO.
+\endgroup
+\end{qle}
+
+\ourparagraph{Satisfiability by Blocked Clause Elimination}
+%
+\looseness=-1%
+Kiesl et al.\ found that blocked clause elimination is especially effective on
+satisfiable problems. To corroborate their results and ascertain
+whether a combination of predicate elimination and blocked clause elimination
+increases the success rate, we evaluate BCE on all 2273
+satisfiable or TPTP FOF and CNF problems.
+The hardware and CPU time limits are the same as in the experiments above.
+Figure~\ref{fig:bce-sat} presents the results.
+
+\looseness=-1
+The baseline establishes the satisfiability of \baseres~problems. We consider
+only preprocessing techniques, since BCE compromises refutational
+completeness---a saturation does not guarantee that the original problem was
+satisfiable. We note that recognizing definition sets makes almost no difference on
+satisfiable problems. The sets of problems solved by BCE and PPE
+differ---\NumberOK{30} problems are solved by BCE and not by PPE.
+% Kiesl sees around 2.5%, we see 3.5% with BCE
+
+\begin{rep}
+\section{Discussion and Related Work}
+\label{sec:discussion-and-related-work}
+
+%\textbf{TODO: This section probably needs to be reworked.}
+
+We briefly surveyed related work in Sect.~\ref{sec:introduction}. In this
+section, we give a more detail overview and further discuss connections with
+the related work.
+
+The research presented in this \paper{} is two-pronged. For SAT elimination
+techniques already generalized to preprocess first-order problems, we looked for
+ways to interleave them with the given clause procedure of a superposition
+prover, as inprocessing. For techniques that had not yet been ported to
+first-order logic, we looked for generalizations that allow both preprocessing and
+inprocessing.
+
+Hidden tautology elimination was first described
+by Heule et al.\ \cite{hjb-2010-cl-elim}. Better implementation that also
+supports hidden literal elimination was later described by the same group of
+authors \cite{hjb-2011-big-simplification}. We generalized the underlying
+theoretical concepts to first-order logic, and provided an efficient way to deal
+with the infinite number of hidden literals that arise  with this
+generalization. More efficient graph-based techniques are yet to be explored.
+
+Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
+the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
+\cite{ab-2004-re}. It was generalized to first-order logic (as a
+preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
+predicate elimination. Improvement of variable elimination,
+which uses formula definition information, has been popularized as a preprocessing and
+inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
+\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
+logic and combined it with existing extension. With tolerable restrictions,
+this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
+allowing variable elimination to slightly increase in the clause set size
+improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
+achieving the double the number of additional proofs found compared to more
+restrictive approaches.
+
+Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
+solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
+\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
+preprocessor. We showed that blocked clauses cannot be removed during
+saturation, but that they can be effectively used to show satisfiability of the
+clause set. A combination of blocked clause elimination and
+variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
+improvement when their generalizations are combined.
+
+Our approach is one of many ways to combine ideas from SAT solving and
+first-order proving. Other noteworthy architectures that either incorporate a
+SAT solver or that generalize the CDCL calculus include DPLL($T$) with
+quantifier instantiation
+\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
+DPLL($\mathrm{\Gamma}+T$) \cite{bonacina-et-al-2009},
+labeled splitting \cite{fietzke-weidenbach-2009},
+AVATAR \cite{av-2014-avatar},
+MCSAT \cite{de-moura-jovanovic-2013},
+CDSAT \cite{bonacina-et-al-2017}, and
+SGGS~\cite{bonacina-plaisted-2014}.
+
+% See also \url{https://academic.oup.com/logcom/advance-article-abstract/doi/10.1093/logcom/exaa065/6038918}
+
+% OLD TEXT, might be useful: {\sl A similar issue arises with
+% the related SAT technique of \emph{variable addition} \cite{xxx}, which can be
+% used to replace the six-clause set $\{\cst{b} \cor \cst{e}{,}\; \cst{c} \cor
+% \cst{e}{,}\; \cst{d} \cor \cst{e}{,}\; \cst{b} \cor \cst{f}{,}\; \cst{c} \cor
+% \cst{f}{,}\; \cst{d} \cor \cst{f}\}$ by the five-clause set $\{\cst{b} \cor
+% \cst{a}{,}\; \cst{c} \cor \cst{a}{,}\; \cst{d} \cor \cst{a}{,}\; \cst{e} \cor
+% \neglit{\cst{a}}{,}\; \cst{f} \cor \neglit{\cst{a}}\}$, at the cost of one
+% extra symbol ($\cst{a}$).}
+
+% In QBF predicate elimination happens more often than in SAT. Contrast to what we do in this paper.
+%%% I'm not sure what to do with this comment. For me, QBF isn't some
+%%% intermediate logic between SAT and FOL; it's a freak. We can't
+%%% quantify over Booleans in FOL. I also weakened the conjecture about
+%%% "the richer the logic, the more useful PE becomes" in the intro. Anyway,
+%%% I don't think it belongs in a summary/future-work Conclusion, but feel
+%%% free to speculate in the Discussion. ;) --JB
+\end{rep}
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+We adapted several preprocessing and inprocessing elimination techniques implemented in
+modern SAT solvers so that they work in a superposition prover. This involved lifting
+the techniques to first-order logic with equality but also tailoring them to
+work in tandem with superposition and its redundancy criterion.
+Although SAT solvers and superposition provers embody radically different
+philosophies, we found that the lifted SAT techniques provide valuable optimizations.
+
+We see several avenues for future work. First, the implementation of hidden
+literals could be extended to exploit equality congruence. Second, although
+inprocessing blocked clause elimination is incomplete in general, we hope to achieve
+refutational completeness for a substantial fragment of it. Third, predicate and
+blocked clause elimination, which thrives on the absence of clauses from the
+proof state, could be enhanced by tagging and ignoring generated clauses that
+have not yet been used to subsume or simplify untagged clauses. Fourth,
+predicate and blocked clause elimination could be extended to work with
+functional literals. Fifth, more SAT techniques could be adapted, including
+bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
+addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
+work with other first-order calculi, or generalized
+further to work with higher-order calculi such as combinatory superposition
+\cite{br-20-full-sup-w-combs} and $\lambda$-superposition
+\cite{bentkamp-et-al-2021-lamsup-journal}.
+
+\let\Acksize=\relax
+
+\def\ackname{\Acksize Acknowledgment}
+\ourparagraph{\ackname}
+{\Acksize
+We are grateful to the maintainers of StarExec for letting us use their
+service.
+Uwe Waldmann participated in the search for a counterexample to completeness of
+BCE as inprocessing and confirmed that Example~\ref{example:bce-completeness}
+is correct. He also suggested major simplifications and helped us debug the
+proofs of the claims about predicate elimination.
+Anne Baanen helped us define the nonclassical logic used to disallow
+splitting.
+Ahmed Bhayat, Armin Biere, Mathias Fleury, Benjamin Kiesl, and the anonymous
+reviewers made some useful comments on our manuscript, and Mark Summerfield
+suggested many textual improvements. We thank them all.
+
+Vukmirovi\'c and Blanchette's research has received funding from
+the European Research Council (ERC) under the European Union's Horizon 2020
+research and innovation program (grant agreement No.\ 713999, Matryoshka).
+%
+Blanchette's research has also
+received funding from the Netherlands Organization for Scientific Research (NWO)
+under the Vidi program (project No.\ 016.Vidi.189.037, Lean Forward).
+%
+Heule is supported by the National Science Foundation (NSF) under grant CCF-2015445.
+
+}
+
+\bibliographystyle{IEEEtran}
+% \begin{raggedright}
+\bibliography{bib}
+% \end{raggedright}
+
+\end{document}
 

### 2022-03-15 satfol/satfol.tex
more

@@
 
+\excludeversion{conf}
+\includeversion{rep}
 \includeversion{noqle}
@@
 \newcommand{\paper}[0]{chapter}
+
+\newcommand{\eqlit}[2]{\ensuremath{#1 \eq #2}}
+\newcommand{\neqlit}[2]{\ensuremath{#1 \noteq #2}}
+% \newcommand{\eqneqlit}[2]{\ensuremath{#1 \mathrel{\dot{\eq}} #2}}
+\newcommand{\pospredlit}[1]{{#1}}
+\renewcommand{\neglit}[1]{\neg{\kern.2ex #1}}
+\newcommand{\negpredlit}[1]{\neglit{#1}}
+\newcommand{\arbpredlit}[1]{(\neg)\kern.2ex\pospredlit{#1}}
+\newcommand{\cor}{\ensuremath{\mathrel{\lor}}}
+\newcolumntype{d}{D{.}{.}{0}}
 
@@
 \newpage
+
+\section{Introduction}
+\label{sec:satfol:introduction}
+
+Automated reasoning tools have become much more powerful in the last few
+decades thanks to procedures such as conflict-driven clause learning (CDCL)
+\cite{MSLM09HBSAT} for propositional logic and superposition
+\cite{bg-94-superposition} for first-order logic with equality. However,
+the effectiveness of these procedures crucially depends on how the input
+problem is represented as a clause set. The clause set
+can be optimized beforehand (\emph{preprocessing}) or during the execution of
+the procedure (\emph{inprocessing}). In this \paper, we lift several preprocessing and
+inprocessing techniques from propositional logic to clausal first-order logic
+and demonstrate their usefulness in a superposition prover.
+
+For many years, SAT solvers have used inexpensive clause simplification
+techniques such as \emph{hidden literal} and \emph{hidden tautology elimination}
+\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and \emph{failed literal
+detection} \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
+first-order logic with equality
+(Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). Since the generalization
+involves reasoning about infinite sets of literals, we propose restrictions to
+make them usable.
+
+\emph{Variable elimination}, based on Davis--Putnam resolution \cite{dp-60-dp}, has
+been studied in the context of both propositional logic
+\cite{sp-04-niver,cs-00-zres} and quantified Boolean formulas (QBFs)
+\cite{ab-2004-re}. The basic idea is to resolve all clauses with negative
+occurrences of a propositional variable (i.e., a nullary predicate symbol) against
+clauses with positive occurrences and delete the parent clauses. E\'en and
+Biere \cite{eb-2005-satpreprocess} refined the technique to identify a subset
+of clauses that effectively define a variable and use it to further optimize the
+clause set. This latter technique, \emph{variable elimination by substitution},
+has been an important preprocessor component in many SAT solvers since its
+introduction in 2004.
+
+Specializing second-order quantifier elimination
+\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
+\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
+problems, yielding a technique we call \emph{singular predicate elimination}. We
+extend their work along two axes (Sect.~\ref{sec:satfol:predicate-elimination}): We
+generalize E\'en and Biere's refinement to first-order logic, resulting in
+\emph{defined predicate elimination}, and explain how both types of predicate
+elimination can be used during the proof search as inprocessing.
+
+\begin{qle}
+It is well known that clauses containing \emph{pure literals}---predicate
+literals whose root symbol occurs only with one polarity in the problem---can
+be eliminated. We propose a generalization of this technique, under the name
+\emph{quasipure literal elimination}, to allow the elimination of some bipolar
+literals as well (Sect.~\ref{sec:satfol:quasipure-literal-elimination}).
+\end{qle}
+
+The last technique we study is \emph{blocked clause elimination} (Sect.\
+\ref{sec:satfol:satisfiability-by-clause-elimination}). It is used in both SAT
+\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
+logic has produced good results when used as a preprocessor, especially on
+satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
+blocked clause elimination on satisfiable problems, including using it to
+establish equisatisfiability with an empty clause set or as an inprocessing
+rule. Unfortunately, we find that its use as inprocessing can compromise the
+refutational completeness of superposition.
+
+All techniques are implemented in the Zipperposition prover
+(Sect.~\ref{sec:satfol:implementation}), allowing us to ascertain their usefulness
+(Sect.~\ref{sec:satfol:evaluation}). The best configuration solves \NumberOK{160}
+additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
+\cite{gs-17-tptp}. The raw experimental data are publicly available.%
+\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
+More details, including all the proofs, can be found in a
+technical report \cite{our-report}.}{}
+
+\section{Preliminaries}
+\label{sec:satfol:preliminaries}
+
+% \ourpara{Clausal First-Order Logic}%
+Our setting is many-sorted (i.e., many-typed), first-order
+logic~\cite{jg-1987-logic-textbook} with interpreted equality and a
+distinguished type (or sort) $o$. We introduced this logic in
+Sect.~\ref{sec:pre:fol} and we use the same notation introduced in this section.
+We write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol
+$\cst{f}$ (e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$) and call nullary
+function symbols \emph{constants}.
+
+We assume the same notion of substitution introduced in Sect.~\ref{sec:pre:fol}.
+We extend the notation by writing a substitution $\{x_1 \mapsto t_1,
+\ldots,\allowbreak x_n \mapsto t_n\}$ shortly as $\{ \tupleempty{x} \mapsto \tupleempty{t} \}$.
+A substitution is a \emph{variable renaming} if it is a
+bijection from a set of variables to a set of variables.
+
+As we are working with clausal logic, we assume the clausal structure as defined
+in Sect.~\ref{sec:pre:clauses}. Recall that predicate literals  are encoded
+as (dis)equations with equality. To simplify the notation in this chapter we write
+predicate literals in unencoded form: positive literal $\eqlit{s}{\itrue}$ is written
+as $s$ and negative literal $\neqlit{s}{\top}$ is written as $\negpredlit{s}$.
+Literals $\eqlit{s}{t}$ where neither $s$ nor $t$ have the type $o$ are called \emph{functional}.
+Given a literal $L$, we overload notation and write $\neglit{L}$ to
+denote its complement. Clauses are often defined
+as sets of literals, but superposition needs multisets; with multisets,
+an instance $\sigmacl{C}$ always has the same number of literals as $C$, a most
+convenient property. Given a clause set $N$, $\binset{N}$ denotes the subset of
+its binary clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.
+
+We assume the natural extensions of domain, valuation, interpretation and model
+(as defined by Fitting \cite{mf-1996-fol}) from unsorted to many-sorted logic.
+We also assume the notion of normal models from Sect.~\ref{sec:pre:fol} (with
+corresponding notation) and extend it with the notion of \emph{cannonical
+models}: A canonical model $\ourmodel$ is a normal model such that for every
+element $d$ in one of $\ourmodel$'s domains there exists a ground term $t$ such
+that $\ourmodel$ interprets $t$ as $d$. Canonical models generalize Herbrand
+models \cite{mf-1996-fol} to first-order logic with equality: If a clause set is satisfiable, then
+it is satisfiable in a canonical model.
+
+% \ourpara{Superposition Provers}
+%
+% Superposition \cite{bg-94-superposition} is a calculus for clausal
+% first-order logic that extends ordered resolution
+% \cite{bg-01-resolution} with equality reasoning. It is
+% refutationally complete: Given a finite, unsatisfiable clause set, it will
+% eventually derive the empty clause. It is parameterized by a \emph{selection
+% function} that influences which of a clause's literals are eligible as the target
+% of inferences. Moreover, it is compatible with the \emph{standard redundancy
+% criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.
+
+% \looseness=-1
+% The redundancy criterion relies on an order $\succ$ that
+% compares terms, literals, or clauses.
+% The order is used to determine whether clauses can be deleted.
+% If $N$ is ground, $C$ can be deleted
+% if it is entailed by $\prec$-smaller clauses in $N.$ This definition is lifted
+% to nonground sets $N$. The criterion can be used to delete a
+% clause that is \emph{subsumed} by another clause (e.g.,
+% $\cst{\cst{p}(\cst{a})} \cor \cst{q}$ by $\cst{p}(x)$) or
+% to \emph{simplify} a clause $C$ into $C'$, which amounts to adding $C'$ and
+% then deleting $C$ as redundant with respect to $N \cup \{C'\}$.
+% Subsumption and simplification are the main inprocessing mechanisms available
+% to superposition provers. Some provers also implement clause splitting
+% \cite{riazanov-voronkov-2001,fietzke-weidenbach-2009,av-2014-avatar}.
+
+% Superposition provers saturate the input problem with respect to the calculus's
+% inference rules using the \emph{given clause procedure}
+% \cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
+% passive set $\mathcalx{P}$ and an active set $\mathcalx{A}$. All clauses start in
+% $\mathcalx{P}$. At each iteration of the procedure's main loop, the prover
+% chooses a clause $C$ from $\mathcalx{P}$, simplifies it, and moves it to
+% $\mathcalx{A}$. Then all inferences between $C$ and active clauses are performed.
+% The resulting clauses are again simplified and put in $\mathcalx{P}$.\confrep{}{
+% The provers differ in which clauses are used for simplification: Otter-loop
+% \cite{mcw-1997-otter} provers use both active and passive clauses whereas
+% DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.}
+
+All of the techniques described in this chapter are applied in the context of
+first-order superposition which is introduced in Sect.~\ref{sec:pre:sup}. We
+reuse all the notions introduced in this section and the corresponding notation.

### 2022-03-15 scratch/paper.tex
more

@@
 \section{Introduction}
-\label{sec:ho-tech:intro}
+\label{sec:satfol:ho-tech:intro}
 
@@
 %of which the most successful is
-such as turning them into rewrite rules (Sect.~\ref{sec:ho-tech:preprocessing}).
+such as turning them into rewrite rules (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
 %Interesting patterns can be observed in various higher-order problem encodings.
-%We show how we can exploit these to simplify problems (Sect.~\ref{sec:ho-tech:preprocessing}).
+%We show how we can exploit these to simplify problems (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
 %
@@
 simulate most successful tableau techniques in a saturating prover
-(Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
+(Sect.~\ref{sec:satfol:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
 functions}, a mechanism that allows us to choose on which Boolean subterms
@@
 We implemented some Boolean selection functions and
-evaluated them (Sect.~\ref{sec:ho-tech:bool-select}).
+evaluated them (Sect.~\ref{sec:satfol:ho-tech:bool-select}).
 
@@
 branching inferences with the standard saturation process
-(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover
+(Sect.~\ref{sec:satfol:ho-tech:infinite-branching}). The prover
 retains the same behavior as
@@
 We also propose some heuristics to curb the explosion induced by highly
-prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).
+prolific calculus rules (Sect.~\ref{sec:satfol:ho-tech:explosiveness}).
 
@@
 between allowing native higher-order reasoning and
-delegating reasoning to a backend (Sect.~\ref{sec:ho-tech:backends}).
+delegating reasoning to a backend (Sect.~\ref{sec:satfol:ho-tech:backends}).
 %
@@
 higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
-evaluation than at CASC (Sect.~\ref{sec:ho-tech:comparison}). Our evaluation
+evaluation than at CASC (Sect.~\ref{sec:satfol:ho-tech:comparison}). Our evaluation
 corroborates the competition results.
@@
 \section{Background and Setting}
-\label{sec:ho-tech:background}
+\label{sec:satfol:ho-tech:background}
 
@@
 \section{Preprocessing Higher-Order Problems}
-\label{sec:ho-tech:preprocessing}
+\label{sec:satfol:ho-tech:preprocessing}
 
@@
 \section{Reasoning about Formulas}
-\label{sec:ho-tech:formulas}
+\label{sec:satfol:ho-tech:formulas}
 
@@
 \section{Exploring Boolean Selection Functions}
-\label{sec:ho-tech:bool-select}
+\label{sec:satfol:ho-tech:bool-select}
 
@@
 have a modest effect. To better assess this feature, in this evaluation we use
-\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:ho-tech:formulas} as the baseline
+\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:satfol:ho-tech:formulas} as the baseline
 configuration. To avoid interference of literal and Boolean selection, we
@@
 outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
-\texttt{NUM636\^{}2} from Sect.~\ref{sec:ho-tech:formulas}, using Boolean selection
+\texttt{NUM636\^{}2} from Sect.~\ref{sec:satfol:ho-tech:formulas}, using Boolean selection
 can reduce the number of given clause loop iterations from
@@
 \section{Enumerating Infinitely Branching Inferences}
-\label{sec:ho-tech:infinite-branching}
+\label{sec:satfol:ho-tech:infinite-branching}
 
@@
 \section{Controlling Prolific Rules}
-\label{sec:ho-tech:explosiveness}
+\label{sec:satfol:ho-tech:explosiveness}
 
@@
 \section{Controlling the Use of Backends}
-\label{sec:ho-tech:backends}
+\label{sec:satfol:ho-tech:backends}
 
@@
 \section{Comparison with Other Provers}
-\label{sec:ho-tech:comparison}
+\label{sec:satfol:ho-tech:comparison}
 
@@
 Semantically, E~2.7 is arguably the weakest among the listed provers:
-It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
+It simply performs $o$-RW rewriting described in Sect.~\ref{sec:satfol:ho-tech:preprocessing} followed by
 $\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
@@
 
-% \newcommand\ourparagraph[1]{\noindent{\textbf{\upshape#1.}}}
-\newcommand\ourparagraph[1]{\subsection{#1}}
+% \newcommand\ourpara[1]{\noindent{\textbf{\upshape#1.}}}
+\newcommand\ourpara[1]{\subsection{#1}}
 
@@
 \section{Introduction}
-\label{sec:introduction}
+\label{sec:satfol:introduction}
 
@@
 first-order logic with equality
-(Sect.~\ref{sec:hidden-literal-based-elimination}). Since the generalization
+(Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). Since the generalization
 involves reasoning about infinite sets of literals, we propose restrictions to
@@
 problems, yielding a technique we call \emph{singular predicate elimination}. We
-extend their work along two axes (Sect.~\ref{sec:predicate-elimination}): We
+extend their work along two axes (Sect.~\ref{sec:satfol:predicate-elimination}): We
 generalize E\'en and Biere's refinement to first-order logic, resulting in
@@
 \emph{quasipure literal elimination}, to allow the elimination of some bipolar
-literals as well (Sect.~\ref{sec:quasipure-literal-elimination}).
+literals as well (Sect.~\ref{sec:satfol:quasipure-literal-elimination}).
 \end{qle}
@@
 The last technique we study is \emph{blocked clause elimination} (Sect.\
-\ref{sec:satisfiability-by-clause-elimination}). It is used in both SAT
+\ref{sec:satfol:satisfiability-by-clause-elimination}). It is used in both SAT
 \cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
@@
 % tautology elimination
-% (Sect.~\ref{sec:hidden-literal-based-elimination}). These
+% (Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). These
 % techniques are fairly inexpensive in propositional logic but need to be
@@
 % elimination by substitution, yielding \emph{defined predicate elimination}
-% (Sect.~\ref{sec:predicate-elimination}). With some care, we can integrate these
+% (Sect.~\ref{sec:satfol:predicate-elimination}). With some care, we can integrate these
 % techniques with superposition without losing refutational completeness.
@@
 % Finally, we study a first-order generalization of blocked clause
-% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satisfiability-by-clause-elimination}).
+% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satfol:satisfiability-by-clause-elimination}).
 % %Unlike the other techniques, it weakens the clause set.
@@
 All techniques are implemented in the Zipperposition prover
-(Sect.~\ref{sec:implementation}), allowing us to ascertain their usefulness
-(Sect.~\ref{sec:evaluation}). The best configuration solves \NumberOK{160}
+(Sect.~\ref{sec:satfol:implementation}), allowing us to ascertain their usefulness
+(Sect.~\ref{sec:satfol:evaluation}). The best configuration solves \NumberOK{160}
 additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
@@
 \section{Preliminaries}
-\label{sec:preliminaries}
-
-\ourparagraph{Clausal First-Order Logic}
+\label{sec:satfol:preliminaries}
+
+\ourpara{Clausal First-Order Logic}
 \looseness=-1
@@
 
-\ourparagraph{Superposition Provers}
+\ourpara{Superposition Provers}
 %
@@
 \section{Hidden-Literal-Based Elimination}
-\label{sec:hidden-literal-based-elimination}
+\label{sec:satfol:hidden-literal-based-elimination}
 
@@
 \section{Predicate Elimination}
-\label{sec:predicate-elimination}
+\label{sec:satfol:predicate-elimination}
 
@@
 \section{Quasipure Literal Elimination}
-\label{sec:quasipure-literal-elimination}
+\label{sec:satfol:quasipure-literal-elimination}
 
@@
 the predicate set $P$ with one of the $\arbpredlit{\cst{q}_i}$'s and continue
-recursively with $\cst{q}_i$. In Sect.~\ref{sec:implementation}, we will see a
+recursively with $\cst{q}_i$. In Sect.~\ref{sec:satfol:implementation}, we will see a
 more efficient approach based on a SAT encoding.
@@
 \section{Satisfiability by Clause Elimination}
-\label{sec:satisfiability-by-clause-elimination}
+\label{sec:satfol:satisfiability-by-clause-elimination}
 
@@
 \section{Implementation}
-\label{sec:implementation}
+\label{sec:satfol:implementation}
 % {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}
@@
 
-\ourparagraph{Hidden-Literal-Based Elimination}
+\ourpara{Hidden-Literal-Based Elimination}
 % 
@@
 % described approach can be generalized to first-order logic and used to implement
-% all rules from Sect.~\ref{sec:hidden-literal-based-elimination}.
+% all rules from Sect.~\ref{sec:satfol:hidden-literal-based-elimination}.
 % 
@@
 
-\ourparagraph{Predicate Elimination}
+\ourpara{Predicate Elimination}
 % 
@@
 \begin{qle}
-\ourparagraph{Quasipure Literal Elimination}
+\ourpara{Quasipure Literal Elimination}
 
@@
 \section{Evaluation}
-\label{sec:evaluation}
+\label{sec:satfol:evaluation}
 
@@
 
-\ourparagraph{Predicate and Blocked Clause Elimination}
+\ourpara{Predicate and Blocked Clause Elimination}
 %
@@
 \begin{qle}
-\ourparagraph{Pure and Quasipure Literal Elimination}
+\ourpara{Pure and Quasipure Literal Elimination}
 \color{red}
@@
 
-\ourparagraph{Satisfiability by Blocked Clause Elimination}
+\ourpara{Satisfiability by Blocked Clause Elimination}
 %
@@
 \section{Discussion and Related Work}
-\label{sec:discussion-and-related-work}
+\label{sec:satfol:discussion-and-related-work}
 
@@
 
-We briefly surveyed related work in Sect.~\ref{sec:introduction}. In this
+We briefly surveyed related work in Sect.~\ref{sec:satfol:introduction}. In this
 section, we give a more detail overview and further discuss connections with
@@
 \section{Conclusion}
-\label{sec:conclusion}
+\label{sec:satfol:conclusion}
 
@@
 \def\ackname{\Acksize Acknowledgment}
-\ourparagraph{\ackname}
+\ourpara{\ackname}
 {\Acksize

### 2022-03-16 satfol/satfol.tex
fixes

@@
     Joint work with\\
-    Jasmin Blanchette,
+    Jasmin Blanchette
     and Marijn J.H. Heule
@@
 
+\begin{sloppypar}
     Specializing second-order quantifier elimination
-\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
-\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
+    \cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin~\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
     problems, yielding a technique we call \emph{singular predicate elimination}. We
@@
     elimination can be used during the proof search as inprocessing.    
+\end{sloppypar}
 

### 2022-03-16 satfol/satfol.tex
hle

@@
 \newcolumntype{d}{D{.}{.}{0}}
+\definecolor{light-gray}{gray}{0.925}
+\newcommand\MAX[1]{\smash{\setlength{\fboxsep}{.3ex}\colorbox{light-gray}{\ensuremath{\vphantom{('q}{#1}}}}}
 
@@
 As we are working with clausal logic, we assume the clausal structure as defined
-in Sect.~\ref{sec:pre:clauses}. Recall that predicate literals  are encoded
-as (dis)equations with equality. To simplify the notation in this chapter we write
-predicate literals in unencoded form: positive literal $\eqlit{s}{\itrue}$ is written
-as $s$ and negative literal $\neqlit{s}{\top}$ is written as $\negpredlit{s}$.
-Literals $\eqlit{s}{t}$ where neither $s$ nor $t$ have the type $o$ are called \emph{functional}.
-Given a literal $L$, we overload notation and write $\neglit{L}$ to
-denote its complement. Clauses are often defined
-as sets of literals, but superposition needs multisets; with multisets,
-an instance $\sigmacl{C}$ always has the same number of literals as $C$, a most
-convenient property. Given a clause set $N$, $\binset{N}$ denotes the subset of
-its binary clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.
+in Sect.~\ref{sec:pre:clauses}. In this section we use the uppercase letter $L$
+to denote literals, as this is more customary in the related literature for
+propositional logic. Recall that predicate literals  are encoded as
+(dis)equations with equality. To simplify the notation in this chapter we write
+predicate literals in unencoded form: positive literal $\eqlit{s}{\itrue}$ is
+written as $s$ and negative literal $\neqlit{s}{\top}$ is written as
+$\negpredlit{s}$. Literals $\eqlit{s}{t}$ where neither $s$ nor $t$ have the
+type $o$ are called \emph{functional}. Given a literal $L$, we overload notation
+and write $\neglit{L}$ to denote its complement. Clauses are often defined as
+sets of literals, but superposition needs multisets; with multisets, an instance
+$\sigmacl{C}$ always has the same number of literals as $C$, a most convenient
+property. Given a clause set $N$, $\binset{N}$ denotes the subset of its binary
+clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.
 
@@
 
-% \ourpara{Superposition Provers}
-%
-% Superposition \cite{bg-94-superposition} is a calculus for clausal
-% first-order logic that extends ordered resolution
-% \cite{bg-01-resolution} with equality reasoning. It is
-% refutationally complete: Given a finite, unsatisfiable clause set, it will
-% eventually derive the empty clause. It is parameterized by a \emph{selection
-% function} that influences which of a clause's literals are eligible as the target
-% of inferences. Moreover, it is compatible with the \emph{standard redundancy
-% criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.
-
-% \looseness=-1
-% The redundancy criterion relies on an order $\succ$ that
-% compares terms, literals, or clauses.
-% The order is used to determine whether clauses can be deleted.
-% If $N$ is ground, $C$ can be deleted
-% if it is entailed by $\prec$-smaller clauses in $N.$ This definition is lifted
-% to nonground sets $N$. The criterion can be used to delete a
-% clause that is \emph{subsumed} by another clause (e.g.,
-% $\cst{\cst{p}(\cst{a})} \cor \cst{q}$ by $\cst{p}(x)$) or
-% to \emph{simplify} a clause $C$ into $C'$, which amounts to adding $C'$ and
-% then deleting $C$ as redundant with respect to $N \cup \{C'\}$.
-% Subsumption and simplification are the main inprocessing mechanisms available
-% to superposition provers. Some provers also implement clause splitting
-% \cite{riazanov-voronkov-2001,fietzke-weidenbach-2009,av-2014-avatar}.
-
-% Superposition provers saturate the input problem with respect to the calculus's
-% inference rules using the \emph{given clause procedure}
-% \cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
-% passive set $\mathcalx{P}$ and an active set $\mathcalx{A}$. All clauses start in
-% $\mathcalx{P}$. At each iteration of the procedure's main loop, the prover
-% chooses a clause $C$ from $\mathcalx{P}$, simplifies it, and moves it to
-% $\mathcalx{A}$. Then all inferences between $C$ and active clauses are performed.
-% The resulting clauses are again simplified and put in $\mathcalx{P}$.\confrep{}{
-% The provers differ in which clauses are used for simplification: Otter-loop
-% \cite{mcw-1997-otter} provers use both active and passive clauses whereas
-% DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.}
-
 All of the techniques described in this chapter are applied in the context of
@@
 reuse all the notions introduced in this section and the corresponding notation.
+
+\section{Hidden-Literal-Based Elimination}
+\label{sec:satfol:hidden-literal-based-elimination}
+
+In propositional logic, binary clauses from a clause set $N$ can be used to
+efficiently discover literals $L, L'$ for which the implication $L' \impl L$
+is entailed by $N$'s binary clauses---i.e., $\binset{N} \models L' \impl L$.
+Heule et al.\ \cite{hjb-2011-big-simplification} introduced the concept of \emph{hidden
+literals} to capture such implications.
+
+\begin{defi}
+   Given a propositional literal $L$ and a propositional clause set~$N$, the
+   set of \emph{propositional hidden literals} for $L$ and~$N$ is
+   $\PHL(L, N) = \{ L' \mid L' \bigpimpl{N}^\ast L \} \setminus \{ L \}$, where $\bigpimpl{N}$ is
+   defined such that $\neglit{L_1} \bigpimpl{N} L_2$ whenever $L_1 \cor L_2 \in N.$
+   Moreover, $\PHL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n
+   \PHL(L_i, N)$.
+\end{defi}
+
+Heule et al.\ used a definition based on a fixpoint computation, but our
+definition based on the reflexive transitive closure $\bigpimpl{N}^\ast$ of $\bigpimpl{N}$ is equivalent. Intuitively,
+a hidden literal can be added to or removed from a clause without affecting its
+semantics in models of~$N.$ By eliminating hidden literals from $C$, we simplify
+it. By adding hidden literals to $C$, we might get a tautology $C'$ (i.e., a
+valid clause:\ $\models C'$), meaning that $\binset{N} \models C$, thereby
+enabling us to delete $C$. Note that $\PHL(L,N)$ is finite for a finite~$N.$
+
+\begin{defi}
+   \label{def:hle}\label{def:hte}
+   Given $L' \cor L \cor C \in N,$ if $L' \in
+   \PHL(L, N)$, \emph{hidden literal elimination} (HLE) replaces $N$ by $(N
+   \setminus \{L' \cor L \cor C\}) \cup \{L \cor C\}$.
+   Given $C \in N$, $\{L_1,\dots,L_n\} = \PHL(C, N)$, and $C' = C \cor L_1 \cor
+   \cdots \cor L_n$, if $C'$ is a tautology, \emph{hidden
+   tautology elimination} (HTE) replaces $N$ by $N \setminus \{C\}$.
+\end{defi}
+
+\begin{theorem}
+   The result of applying HLE or HTE to a clause set $N$ is
+   equivalent to $N$.
+\end{theorem}
+
+\begin{proof}
+For HLE, if $L' \in \PHL(L, N)$,  $\binset{N} \models \neglit{L'} \cor L$.
+Then, subsumption resolution \cite{bg-01-resolution} yields shortened clause $L \cor C'$ from
+Definition~\ref{def:hle}. For HTE, it can be shown that $N' \models C$ if and
+only if $C \cor L'$, where $L' \in \PHL(C, N)$. By transitivity of equivalence,
+we get the desired result.
+\end{proof}
+
+We generalize hidden literals to first-order logic with equality by
+considering substitutivity of variables as well as congruence of equality.
+
+\begin{defi}
+\label{def:hl-fo}
+Given a literal $L$ and a clause set $N$,
+the set of \emph{hidden literals} for $L$ and $N$
+is $\HL(L,N) = \{ L' \mid L' \bigfimpl{N}^\ast L \} \setminus \{ L \}$, where
+$\bigfimpl{N}$ is defined so that:
+\begin{enumerate}
+    \item $\sigmaterm{\neglit{L'}} \bigfimpl{N} \sigmaterm{L}$ if
+    $L' \cor L \in N$ and $\sigma$ is a substitution
+    \item $\eqlit{s}{t} \bigfimpl{N} \eqlit{u[s]}{u[t]}$ for all terms $s, t$ and
+    contexts $u[\phantom{\cdot}]$
+    \item $\neqlit{u[s]}{u[t]} \bigfimpl{N} \neqlit{s}{t}$ for all terms $s, t$ and
+    contexts $u[\phantom{\cdot}]$
+\end{enumerate}
+% (1)~;
+% (2)~; and
+% (3)~.
+Moreover, $\HL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n \HL(L_i, N)$.
+\end{defi}
+
+\looseness=-1
+The generalized definition also enjoys the key property that $L'
+\in \HL(L, N)$ implies $\binset{N} \models L' \impl L$. However, $\HL(L, N)$
+may be infinite even for predicate literals; for example,
+$\pospredlit{\cst{p}(\cst{f}^i(x))} \in \HL(\pospredlit{\cst{p}(x)}, \{\pospredlit{\cst{p}(x)} \cor
+\negpredlit{\cst{p}(\cst{f}(x))} \})$ for every $i$.
+
+Based on Definition \ref{def:hl-fo}, we can generalize hidden literal elimination
+and support a related technique:
+\begin{align*}
+    & \namedsimpsc{HLE}{L' \cor L \cor C}{L \cor C}{L' \in \HL(L,N)} \\[.5\jot]
+    & \namedsimpsc{\rlap{FLE}\phantom{HLE}}{L \cor C}{C}{L',\neglit{L'} \in \HL(\neglit{L}, N)}
+ \end{align*}
+Recall that double lines denote simplification rules (Sect.~\ref{sec:pre:sup}).
+The second rule is called \textit{failed literal elimination},
+inspired by the SAT technique of asserting $\neglit{L}$ if $L$ is a \textit{failed literal} \cite{jwf-1995-fld}.
+It is easy to see that rule \infname{HLE} is sound.
+From $L' \in \HL(L,N)$ we have $N \models L' \impl L$ (i.e., $\neglit{L'} \cor L$).
+Performing subsumption resolution \cite{bg-01-resolution} between $L'
+\cor L \cor C$ and $\neglit{L'} \cor L$ yields the conclusion, which is
+therefore entailed by $N.$ For \infname{FLE}, the condition $L',
+\neglit{L'} \in \HL(\neglit{L},N)$ means that $\binset{N} \models \{\neglit{L'} \cor
+\neglit{L}{,}\; L' \cor \neglit{L}\} \models \neglit{L}$.
+
+\begin{exa}
+   \label{example:hle}
+   Consider the clause set $N = \{ \cst{p}(x) \cor \neglit{\cst{p}(\cst{f}(x))}{,}\;
+   \cst{p}(\cst{f}(\cst{f}(x))) \cor \eqlit{\cst{a}}{\cst{b}} \}$ and the clause
+   $C = \neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \cor \cst{p}(x)$. The first
+   clause in $N$ induces $\cst{p}(\cst{f}(x)) \bigfimpl{N} {\cst{p}(x)}$,
+   $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N} \cst{p}(\cst{f}(x))$, and
+   hence $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N}^\ast  \cst{p}(x)$.
+   Together with the second clause in $N$, it can be used to derive $\neqlit{\cst{a}}{\cst{b}}
+   \bigfimpl{N}^\ast  \cst{p}(x)$. Finally, using rule~(3) of Definition
+   \ref{def:hl-fo}, we derive $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})}
+   \bigfimpl{N}^\ast  \cst{p}(x)$---that is,
+   $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \in \HL(\cst{p}(x), N)$. This
+   allows us to remove $C$'s first literal using \infname{HLE}.
+\end{exa}
+
+Two special cases of \infname{HLE} exploit equality congruence as embodied by
+conditions 2 and 3 of Definition~\ref{def:hl-fo} without requiring to
+compute the $\HL$~set:
+\begin{align*}
+  %% TYPESETTING: \,'s below
+  \namedsimp{CongHLE$^+$}{\eqlit{s}{t} \cor \eqlit{u[s]}{u[t]} \cor C}{\eqlit{u[s]}{u[t]} \cor C}\,\, \\[.5\jot]
+  \namedsimp{CongHLE$^-$}{\neqlit{s}{t} \cor \neqlit{u[s]}{u[t]} \cor C}{\neqlit{s}{t} \cor C}
+\end{align*}%
+
+Hidden literals can be combined with unit clauses~$L'$ to remove more literals:
+\[\namedsimpsc{UnitHLE}{L' \quad L \cor C}{L' \quad C}{\sigmaterm{L'} \in \HL(\neglit{L},N)}
+%\text{~for some~}\sigma -- %%% Not necessary in a rule -- all variables are universally quantified. --JB
+\]
+%
+Given a unit clause $L' \in N$, the rule uses it to discharge $\sigmaterm{L'}$ in
+$N \models \sigmaterm{L'} \impl \neglit{L}$. As a result, we have $N \models
+\neglit{L}$, making it possible to remove $L$ from $L \cor C$.
+
+% Kiesl and Suda described first-order generalization of {\it asymmetric literals}
+% \cite{ks-2017-unif-principle}: literal $L\sigma$ is asymmetric for a clause $C$
+% and a clause set $N$ if there exists a clause $D \cor \neglit{L} \in N \setminus
+% \{C \}$ such that $D\sigma \subseteq C$ for some substitution $\sigma$.
+
+\begin{exa}
+   \label{example:unithle}
+   \looseness=-1
+   Consider the clause set $N = \{\cst{p}(x) \cor \cst{q}(\cst{f}(x)){,}\;
+   \neglit{\cst{q}(\cst{f}(\cst{a}))} \cor \eqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}{,}\;
+   \neqlit{\cst{f}(x)}{\cst{g}(y)}\}$ and the clause $C = \neglit{\cst{p}(\cst{a})} \cor
+   \neglit{\cst{q}(\cst{b})}$. The first clause in $N$ induces
+   $\neglit{\cst{q}(\cst{f}(\cst{a}))} \bigfimpl{N} \cst{p}(\cst{a})$, whereas
+   the second one induces $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})} \bigfimpl{N}
+   \neglit{\cst{q}(\cst{f}(\cst{a}))}$. Thus, we have $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}
+   \bigfimpl{N}^\ast \cst{p}(\cst{a})$---that is, $\neqlit{\cst{f}(\cst{b})}{\cst{f}(\cst{c})} \in
+   \HL(\cst{p}(\cst{a}), N)$. By applying the substitution $\{x \mapsto \cst{b}{,}\allowbreak\; y
+   \mapsto \cst{c}\}$ to the third clause in $N$, we can fulfill the conditions of
+   \infname{UnitHLE} and remove $C$'s first literal.
+\end{exa}
+
+Next, we generalize hidden tautologies to first-order logic.
+
+\begin{defi}
+   A clause $C$ is a \emph{hidden tautology} for a clause set $N$ if there
+   exists a finite set $\{L_1,\dots,L_n\} \subseteq \HL(C,N)$ such that
+   $C \cor L_1 \cor \cdots \cor L_n$ is a tautology.
+\end{defi}
+
+\begin{exa}
+   \label{example:hidden-tautologies-completeness}
+   \looseness=-1
+   In general, hidden tautologies are not redundant and cannot be deleted during saturation.
+   Consider the unsatisfiable set $N = \{ \negpredlit{\cst{a}}{,}\;
+   \negpredlit{\cst{b}}{,}\; \pospredlit{\cst{a}} \cor
+   \pospredlit{\cst{c}}{,}\allowbreak\; \pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}
+   \}$, the order $\cst{a} \prec
+   \cst{b} \prec \cst{c}$, and the empty selection function. The
+   only possible superposition inference from $N$ is between the last two clauses,
+   yielding the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
+   (after simplifying away $\neqlit{\top}{\top}$), which is entailed by the larger
+   clauses $\pospredlit{\cst{a}} \cor \pospredlit{\cst{c}}$ and
+   $\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}$. If this clause is removed,
+   the prover could enter an infinite loop, forever generating and deleting the
+   hidden tautology\confrep{}{ and never getting the opportunity to derive the
+   empty clause}.
+\end{exa}
+
+\begin{rep}
+   \newcommand{\aset}{\mathcal{A}}
+   \newcommand{\pset}{\mathcal{P}}
+
+   In practice, most provers use a variant of the given clause procedure.
+   Removing hidden tautologies breaks the invariant of the procedure that all
+   inferences between clauses in $\aset$ are redundant. The end result is not
+   that the prover diverges, but that it terminates without deriving empty
+   clause.
+
+   To observe this, assume the setting as in Example
+   \ref{example:hidden-tautologies-completeness}, and let $\pset=N$ and
+   $\aset=\emptyset$. After moving the first three clauses from $\pset$ to
+   $\aset$ ($\aset = \{ \negpredlit{\cst{a}}, \negpredlit{\cst{b}}, \pospredlit{\cst{a}} \cor
+   \pospredlit{\cst{c}} \}$, $\pset = \{\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}\}$), 
+   no inferences are possible, and no new clauses are added to $\pset$. After the last clause
+   is moved to $\aset$, the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
+   is produced. If it is deleted, the prover terminates with the unsatisfiable set $\aset$,
+   but does not derive the empty clause. 
+\end{rep}
+
+To delete hidden tautologies during saturation, the prover could check that all
+the relevant clause instances encountered along the computation of $\HL$ are
+$\prec$-smaller than a given hidden tautology. However, this would be expensive
+and seldom succeed, given that superposition creates lots of nonredundant
+hidden tautologies. Instead, we propose to simplify hidden tautologies using the
+following rules:
+%
+%\begin{linenomath*}
+% \begin{align*}
+%    %% TYPESETTING: \,'s below
+%    \namedsimpsc{HTR}{L \cor L' \cor C}{L \cor L'}{\neglit{L'} \in \HL(L, N)} &\\
+%    \namedsimpsc{FLR}{L \cor C}{L}{L',\neglit{L'} \in \HL(L, N)} &
+% \end{align*}
+%
+\begin{align*}
+   & \namedsimpsc{HTR}{L \cor L' \cor C}{L \cor L'}{\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot} \\[.5\jot]
+   & \namedsimpsc{\rlap{FLR}\phantom{HTR}}{L \cor C}{L}{L',\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot}
+\end{align*}
+%
+
+%\end{linenomath*}
+%
+We call these techniques \emph{hidden tautology reduction} and \emph{failed literal
+reduction}, respectively.
+% \infname{FLD} is inspired by the \confrep{}{eponymous }SAT
+% approach of asserting $L$ if $\neglit{L}$ is determined to be a failed literal
+% \cite{jwf-1995-fld}.
+%
+Both rules are sound.
+%
+As with hidden literals, unit clauses~$L'$ can be exploited:
+\[\namedsimpsc{UnitHTR}{L' \quad L \cor C}{L' \quad L}{\sigmaterm{L'} \in \HL(L,N) \text{~and~} C \not= \bot}
+%\text{~for some~}\sigma
+\]
+
+We give the simplification rules above\confrep{}{ (for hidden literal
+elimination, hidden tautology reduction,
+failed literal detection, and their variants)} the collective name of
+\emph{hidden-literal-based elimination} (HLBE).
+Yet another use of hidden literals is for \emph{equivalent literal
+substitution} \cite{hjb-2010-cl-elim}: If both $L' \in \HL(L, N)$ and $L
+\in \HL(L', N)$, we can often simplify $\sigmacl{L'}$ to $\sigmacl{L}$ in $N$ if
+$\sigmacl{L'} \succ \sigmacl{L}$. We want to investigate this~further.
+
+% \looseness=-1
+% Provers based on the DISCOUNT loop use only active clauses for
+% simplification. If we restrict our attention to a finite subset of hidden
+% literals, we can organize them into data structures that support efficient
+% lookup of a hidden literal. Efficient implementation of these data structures
+% allows to use all passive clauses for simplification, strengthening DISCOUNT
+% loop simplification machinery.
+
+\begin{theorem}
+\begin{sloppypar}
+The rules \infname{HLE}, \infname{FLE}, \infname{CongHLE$^+$},
+\infname{Cong\allowbreak HLE$^-$}, \infname{UnitHLE}, \infname{HTR},
+\infname{FLR}, and \infname{UnitHTR} are sound simplification rules.
+\end{sloppypar}
+\end{theorem}
+
+\begin{rep}
+\begin{proof}
+It is easy to see that the deleted premises are entailed by the conclusions that
+replace them and that the conclusions' instances are $\prec$-smaller than the
+premises' instances, as required by the redundancy criterion. It remains to
+check soundness.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{HLE}:\enskip
+We have $N{,}\> L' \models L$ by the side condition and must show
+$N{,}\> L' \lor L \lor C \models L \lor C$. Let
+$\ourmodel \models N{,}\> L' \lor L \lor C$. If $\ourmodel \models L'$, then
+we also have $\ourmodel \models L$ thanks to the side condition and hence
+$\ourmodel \models L \lor C$. Otherwise, we have $\ourmodel \models L \lor C$,
+which is exactly what we need to show.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{FLE}:\enskip
+We have $N{,}\> L \models L'$ and $N, L \models \lnot L'$ by the side condition.
+If $\ourmodel \models N{,}\> L$, then both $\ourmodel \models L'$ and
+$\ourmodel \models \lnot L'$, an absurdity.
+Otherwise, we have $\ourmodel \models C$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{CongHLE$^+$}, \infname{CongHLE$^-$}:\enskip
+Obvious by congruence of equality.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{UnitHLE}:\enskip
+We have $N{,}\> L \models \lnot \sigmaterm{L'}$ by the side condition. If $\ourmodel
+\models N{,}\> L$, then $N \models \lnot \sigmaterm{L'}$. But since $L' \in N$, this is
+an absurdity. Otherwise, we have $\ourmodel \models C$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{HTR}:\enskip
+We have $N{,}\> \lnot L' \models L$ by the side condition. If either $\ourmodel
+\models L$ or $\ourmodel \models L'$, the desired result follows directly.
+Otherwise, from $\ourmodel \models \lnot L'$ we also have $\ourmodel \models L$
+thanks to the side condition, contradicting $\ourmodel \models \lnot L$.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{FLR}:\enskip
+We have $N{,}\> L' \models L$ and $N{,}\> \lnot L' \models L$ by the side condition.
+Hence $N \models L$, as desired.
+
+\medskip
+
+\noindent
+\textsc{Case} \infname{UnitHTR}:\enskip
+We have $N{,}\> \sigmaterm{L'} \models L$. Since $L' \in N$, we have $N \models L$,
+as desired.
+\qedhere
+\end{proof}
+\end{rep}

### 2022-03-16 scratch/paper.tex
hle

@@
 % each other, we need to somehow squeeze them
-\newcommand{\removemyskip}{\vspace{-\myskip}}
-
-\newenvironment{theoremx}% environment name
+\newcommand{}{\vspace{-\myskip}}
+
+\newenvironment{theorem}% environment name
 {% begin code
@@
 
-\newenvironment{examplex}% environment name
+\newenvironment{exa}% environment name
 {% begin code
@@
 
-\newenvironment{definitionx}% environment name
+\newenvironment{defi}% environment name
  {% begin code
@@
 
-\begin{definitionx}
+\begin{defi}
    Given a propositional literal $L$ and a propositional clause set~$N$, the
@@
    \PHL(L_i, N)$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    \label{def:hle}\label{def:hte}
@@
    tautology elimination} (HTE) replaces $N$ by $N \setminus \{C\}$.
-\end{definitionx}
-
-\begin{theoremx}
+\end{defi}
+
+\begin{theorem}
    The result of applying HLE or HTE to a clause set $N$ is
    equivalent to $N$.
-\end{theoremx}
+\end{theorem}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 \label{def:hl-fo}
@@
 Moreover, $\HL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n \HL(L_i, N)$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:hle}
@@
    allows us to remove $C$'s first literal using \infname{HLE}.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:unithle}
@@
    \infname{UnitHLE} and remove $C$'s first literal.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    A clause $C$ is a \emph{hidden tautology} for a clause set $N$ if there
@@
    $C \cor L_1 \cor \cdots \cor L_n$ is a tautology.
-\end{definitionx}
-\removemyskip
-\begin{examplex}
+\end{defi}
+
+\begin{exa}
    \label{example:hidden-tautologies-completeness}
@@
    empty clause}.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{theoremx}
+\begin{theorem}
 The rules \infname{HLE}, \infname{FLE}, \infname{CongHLE$^+$},
@@
 \infname{FLR}, and \infname{UnitHTR} are sound simplification rules.
-\end{theoremx}
+\end{theorem}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    A predicate symbol is called \emph{singular}\confrep{ (or ``non-self-referential'')}{}
@@
    $N.$
-\end{definitionx}
-\removemyskip
-\begin{definitionx}
+\end{defi}
+
+\begin{defi}
    \label{def:flat-res}
@@
    \cor C' \cor D'$ is a \emph{flat resolvent} of $C$ and $D$ on $\cst{p}$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 \label{def:flat-res-set}
@@
    % $\cst{p}$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
@@
    and let $\withoutpred{N}{\cst{p}} = N \setminus \withpred{N}{\cst{p}}$.
-\end{definitionx}
-\removemyskip
-\begin{definitionx}
+\end{defi}
+
+\begin{defi}
    \label{def:pred-elim}
@@
    $\withoutpred{N}{\cst{p}} \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    \label{def:definition}
@@
    in for $\tupleempty{x}$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:tautologies}
@@
    as a definition set for $\cst{p}$.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    \looseness=-1
@@
    (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}} \confrep{\withpred{R}{\cst{p}}}{R})$.
-\end{definitionx}
+\end{defi}
 
@@
 \begin{conf}
-\removemyskip
+
 \end{conf}
-\begin{theoremx}
+\begin{theorem}
 \label{thm:pes-sat-equiv}
@@
    satisfiable if and only if $N$ is satisfiable.
-\end{theoremx}
+\end{theorem}
 \begin{rep}
@@
    In all other cases, it is not applicable.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{examplex}
+\begin{exa}
 \label{ex:cex-predelim-sat}
@@
 unsaturated because $\iota$ is no longer redundant. The invariant is broken.
-\end{examplex}
-
-\begin{examplex}
+\end{exa}
+
+\begin{exa}
 \label{ex:cex-predelim-unsat}
@@
 unsatisfiable four-clause active set that does not contain the empty clause.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 The operator $\Wk{}$ translates $\mathrm{\Sigma}$-literals to
@@
 in the standard way via grounding.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{theoremx}
+\begin{theorem}
 \label{thm:gcp-complete}
@@
 contains the empty clause with some arbitrary label.
-\end{theoremx}
+\end{theorem}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:ple}
@@
    atoms $\cst{p}(\ldots)$ true and all atoms $\cst{q}(\ldots)$ false.
-\end{examplex}
-
-\begin{examplex}
+\end{exa}
+
+\begin{exa}
    \label{example:qle-i}
@@
     example applies. In particular, $\ourmodel \models N'$.
-\end{examplex}
-
-\begin{examplex}
+\end{exa}
+
+\begin{exa}
    \label{example:qle-ii}
@@
    N''$.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 \label{def:quasipure-set}
@@
 such that $P$ is quasipure with $s$ in $N$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 \label{def:quasipure-sym}
@@
 $\cst{p}$ is quasipure with $L$'s polarity in $N$.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
 \label{def:qle}
@@
 remain.
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{definitionx}
+\begin{defi}
    Let $C = L \cor C'$ and $D = L_1 \cor \cdots \cor L_m \cor D'$, where
@@
    is a \emph{flat $L$-resolvent} of $C$~and~$D$.
-\end{definitionx}
-\removemyskip
-\begin{definitionx}
+\end{defi}
+
+\begin{defi}
 \label{def:equality-blocked}
@@
 \end{rep}
-\end{definitionx}
+\end{defi}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:infinite-blocked-set}
@@
    that cannot be eliminated using standard redu\-ndancy-based techniques.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \label{example:bce-completeness}
@@
    empty clause for this unsatisfiable set.
-\end{examplex}
+\end{exa}
 
@@
    \subseteq C\sigma$ and $\neglit L'\varrho = L\sigma$.
- \end{definitionx}
-\confrep{\removemyskip}{}
+ \end{defi}
+\confrep{}{}
 
@@
 
-\begin{examplex}
+\begin{exa}
    \looseness=-1
@@
   for $C$ and $N$, but it is a GAL for $C$ and $N$.
-\end{examplex}
+\end{exa}
 
@@
 
-\begin{theoremx}
+\begin{theorem}
    \label{thm:gal}
@@
    \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
-\end{theoremx}
+\end{theorem}
 \begin{rep}%
@@
 
-\begin{examplex}
+\begin{exa}
    Consider the satisfiable clause set $N = \{\pospredlit{\cst{r}(x)} \cor
@@
    and can be removed, leaving $N$ empty as desired.
-\end{examplex}
+\end{exa}
 

### 2022-03-16 satfol/satfol.tex
more

@@
 \begin{proof}
-For HLE, if $L' \in \PHL(L, N)$,  $\binset{N} \models \neglit{L'} \cor L$.
-Then, subsumption resolution \cite{bg-01-resolution} yields shortened clause $L \cor C'$ from
-Definition~\ref{def:hle}. For HTE, it can be shown that $N' \models C$ if and
-only if $C \cor L'$, where $L' \in \PHL(C, N)$. By transitivity of equivalence,
-we get the desired result.
+For HLE, if $L' \in \PHL(L, N)$,  $\binset{N} \models \neglit{L'} \cor L$. Then,
+subsumption resolution (a sound rule that applies resolution followed by
+subsumption \cite{bg-01-resolution}) yields shortened clause $L \cor C'$ from
+Definition~\ref{def:hle}. For HTE, it can be shown that sets $N$ and $N
+\cup \{ C' \} \setminus \{C\}$ are equivalent
+\cite[Sect.~2.1]{hjb-2011-big-simplification}. As clause $C'$ is a tautology,
+$N$ and $N \setminus \{C\}$ are equivalent.\qedhere
 \end{proof}
@@
    Together with the second clause in $N$, it can be used to derive $\neqlit{\cst{a}}{\cst{b}}
-   \bigfimpl{N}^\ast  \cst{p}(x)$. Finally, using rule~(3) of Definition
+   \bigfimpl{N}^\ast  \cst{p}(x)$. Finally, using rule 3 of Definition
    \ref{def:hl-fo}, we derive $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})}

### 2022-03-16 ho-techniques/ho-techniques.tex
more

@@
 \blfootnote{In this work I was the main designer of all presented techniques,
-with the exception of inference streams which were designed by Sophie Tourret
-and Alexander Bentkamp. Alexander Bentkamp and Jasmin Blanchette also discussed many
-of the techniques with me and suggested important updates. Visa Nummelin worked on
-implementation of FOOL preprocessing. Simon Cruanes is the original developer
-of Zipperposition and provided us with invaluable knowledge.
+with the exception of inference streams which were designed by Alexander
+Bentkamp and Sophie Tourret. Alexander Bentkamp and Jasmin Blanchette also
+discussed many of the techniques with me and suggested important updates. Visa
+Nummelin worked on implementation of FOOL preprocessing. Simon Cruanes is the
+original developer of Zipperposition and provided us with invaluable knowledge.
 }
@@
     implementation in the Zipperposition theorem prover. Largely thanks to their use,
-    Zipperposition won the higher-order division of the CASC competition in 2020.  
+    Zipperposition won the higher-order division of the CASC competition in 2020 and 2021.  
   \end{abstract}
@@
 
-The landscape of higher-order proving techniques based on extension of efficient
+The landscape of higher-order proving techniques based on the extension of efficient
 first-order ones has tremendously expanded in the late 2010s and early 2020s. As
-mentioned is Sect.~\ref{sec:pre:ho-sup-calculi} we have implemented three
+mentioned in Sect.~\ref{sec:pre:ho-sup-calculi} we have implemented three
 higher-order calculi---\lfsup{}, \lsup{}, and \osup{}---which extend first-order superposition in a graceful way.
 Bhayat and Reger also gracefully extended superposition to higher-order logic using
-\textsf{SKBCI} combinators \cite{br-20-full-sup-w-combs}. Significant progress has been
+\textsf{SKBCI} combinators \cite{br-20-full-sup-w-combs}, resulting in a calculus called combinatory superposition. Significant progress has been
 made on SMT front as well \cite{brotb-19-ho-smt}.
@@
 In 2019 we tested if the idea of gracefully extending first-order provers to
-higher-order logic really improves the state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
-Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
-calculus to deal with Booleans. It finished third at that year's higher-order
-division of CASC competition \cite{gs-19-casc27}, 12~percentage points behind the
-winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.
+higher-order logic really improves the state of the art for the first time. We
+implemented \lsup{} \cite{bbtvw-21-sup-lam} in Zipperposition~1.5 with basic
+heuristics and rudimentary extensions of the calculus to deal with Booleans. It
+finished third at that year's higher-order division of CASC competition
+\cite{gs-19-casc27}, 12~percentage points behind the winner, the tableau prover
+Satallax 3.4 \cite{cb-2013-satallax}.
 
@@
 \looseness=-1
-The main drawback of both $\lambda$-superposition variants compared with
+The main implementation challenge of both $\lambda$-superposition variants compared with
 combinatory superposition is that they rely on rules that enumerate possibly
@@
 notation for this logic as introduced in Chapter \ref{ch:pre}. Since we are
-working with extensions of superposition, we assume the clausal structure
+working with extensions of superposition, we assume a clausal structure
 (Sect.~\ref{sec:pre:clauses}). Like in the previous chapter, literals
 of clauses can contain arbitrary higher-order terms, including formulas.  
+At CASC, most theorem provers, including Zipperposition, are invoked using a
+sequence of different configurations (possibly in parallel) until either the
+time limit is reached or a proof is found. We call this sequence a \emph{portfolio}.
 
@@
 
-\ourpara{Zipperposition}
-Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
-theorem prover that implements both $\lambda$-superposition variants, combinatory
-superposition, and other superposition-like calculi.
-The prover was conceived as a testbed for rapidly
-experimenting with extensions of first-order superposition, but over time it
-has assimilated many of E's techniques and heuristics and become quite powerful.
-Several of our techniques extend the \emph{given clause procedure}, the standard
-saturation procedure described in Sect.~\ref{sec:pre:saturation}.
-
 \ourpara{Experimental Setup}
@@
 
-The second technique aims at preserving completeness: We can try to force the term order that
-parameterizes superposition to orient as many definitions as possible and rely on
-demodulation to simplify the proof state. Usually, the Knuth--Bendix order (KBO)
-\cite{db-1970-kbo} is used. It compares terms by first comparing their weights,
+The second technique aims at preserving completeness: We can try to choose, as the term order that
+parameterizes superposition, one that orients as many definitions as possible and rely on
+demodulation to simplify the proof state. Usually, an instance of the Knuth--Bendix order (KBO)
+\cite{db-1970-kbo} is used. KBO compares terms by first comparing their weights,
 which is the sum of all the weights assigned to the symbols it contains. Given a
@@
 
-The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
+The results are given in Fig.~\ref{fig:rewrite}. In all the figures in this chapter, each
 cell gives the number of proved problems, and cells marked with $\star$
@@
 Even though this is the laziest approach to clausification, the earlier
-evaluation by Bentkamp et al. \cite{bbtv-21-full-ho-sup} showed that this
+evaluation \cite{bbtv-21-full-ho-sup} showed that this
 approach is inefficient. Thus, we focus only on the outer rules.
@@
 evaluation environment. The fact that \relax{BI} has no effect on SH benchmarks
-is to be expected because Sledgehammer does
-not include lemmas whose name contains the substring \texttt{.induct} and that
+is to be expected because 
+Sledgehammer deliberately tries to exclude induction rules from the problem by
+excluding lemmas whose name contains the substring \texttt{.induct} and that
 contain predicate variables. Therefore, \relax{BI} applies to fewer clauses.
@@
 
-A Boolean selection function chooses \emph{green subterms} of Boolean type
-(different than $\itrue$ or $\ifalse$ and not occurring at either side of a positive
-literal) in a clause and gives rise to a notion of eligibility that considers
-the formula structure. Green subterms correspond to the first-order skeleton of
-a higher-order term; that is, they do not occur in positions under applied
-variables, quantifiers, or $\lambda$-abstractions.
+A Boolean selection function chooses \emph{green subterms} of Boolean type,
+which are different than $\itrue$ or $\ifalse$ and not occurring at either side
+of a positive literal in a clause. It gives rise to a notion of eligibility that
+considers the formula structure. We call terms that can be selected
+\emph{selectable subterms}. Green subterms correspond to the first-order
+skeleton of a higher-order term; that is, they do not occur in positions under
+applied variables, quantifiers, or $\lambda$-abstractions.
 
@@
 \end{defi}
+
+\begin{exa}
+  Consider a clause $\poslit{\cst{p}} \llor \poslit{\cst{q} \, (\cst{p} \ior \cst{a}
+  \ieq \cst{b})}$, literal and Boolean selection functions that select no
+  literals and terms, respectively, and the precedence of symbols $\itrue \prec
+  \cst{a} \prec \cst{b} \prec \cst{c} \prec \cst{p} \prec \cst{q}$. As no
+  literals and subterms are selected, the second literal is eligible, as it is maximal. This makes the term 
+  $\cst{q} \, (\cst{p} \ior \cst{a} \ieq \cst{b})$ eligible. Further, the green subterm
+  $\cst{p} \ior \cst{a} \ieq \cst{b}$ is eligible. Similarly, $\cst{p}$ and $\cst{a} \ieq \cst{b}$
+  are eligible. Lastly, as $\cst{a} \prec \cst{b}$, $\cst{b}$ is eligible.
+\end{exa}
+
 The above definitions of green subterms and eligibility were originally introduced
@@
 simulates forward reasoning, whereas selecting the consequent simulates backward
-reasoning. The new eligibility also makes it possible to restrict the proof
+reasoning. This concept of eligibility also makes it possible to restrict the proof
 search to a small, promising part of a formula. Note that
@@
   selects all terms $t$ such that $t=C|_p$, $p \in \mathit{ctx}(C)$, and
-  $(t,p)$ is maximal with respect to $\vartriangleright$.
+  $(t,p)$ is maximal with respect to $\vartriangleright$, compared with all other pairs $(C|_{p'}, p')$,
+  $p' \neq p$, and $p \in \mathit{ctx}(C)$.
 \end{defi}
@@
 
-Entry point to the above pseudocode is the function \textsc{GivenClause}, called
+The entry point of the above pseudocode is the function \textsc{GivenClause}, called
 with arguments $(\mathcal{P}, \mathcal{A}, \mathcal{Q})$, which are initialized
@@
 new inferences involving the given clause are added to $Q$ instead of being
-performed immediately. Third, inferences in $Q$ are periodically performed
-lazily to fill $P$.
+performed immediately. Third, a number of inferences in $Q$ are periodically computed
+to fill $P$.
 
@@
 and should be ignored. Our heuristics to increase the stream penalties take
-both observations into account.
+these observations into account.
 
@@
 Leo-III. His results suggest that adding only one backend already substantially
-improves the performance. To reduce the effort required for integrating multiple backends, we chose Ehoh, an extension of E \cite{scv-19-e23} to support \lfsup{},  as our single
+improves the performance. To reduce the effort required for integrating multiple backends, we chose Ehoh, an extension of E \cite{scv-19-e23} that supports \lfsup{},  as our single
 backend. In Chapter \ref{ch:ehoh}, we described Ehoh in detail.
@@
 provide some context, we compare Zipperposition~2 with the following versions of
-all higher-order provers that competed at CASC in 2020:\ CVC4 1.9
+all other higher-order provers that competed at CASC in 2020:\ CVC4 1.9
 \cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
@@
 $\lambda$-lifting before it applies \relax{\lfsup{}} \cite{bbcw-21-lfho}
-on the preprocessed problem.
+on the preprocessed problem. It is a conservative extension of Ehoh presented in Chapter~\ref{ch:ehoh},
+and serves as a baseline for comparison with $\lambda$E introduced in Chapter~\ref{ch:ehoh2}.
 
@@
 described in this chapter, it outperforms all higher-order provers except for
-Zipperposition. We give detailed description of $\lambda$E, as well as reasoning
+Zipperposition. We give detailed a description of $\lambda$E, as well as reasoning
 behind all design decision we took in Chapter \ref{ch:ehoh2}.

### 2022-03-16 pre/pre.tex
more
fixes

@@
 \emph{Terms} are the basic building blocks of first-order logic, built inductively as follows. Variables
-$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
+$x,y,\ldots$, assigned types $\tau \in T\setminus \{ o \}$, are terms. If $t_1,\ldots,t_n$ are
 terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
@@
 \section{Superposition}
+\label{sec:pre:sup}
 
@@
 The resulting clauses are again simplified and put in $\mathcal{P}$.
+With \emph{proof state} we assume the pair $(\mathcal{P}, \mathcal{A})$.
 The provers differ in which clauses are used for simplification: Otter-loop
@@
 and \osup{} where the context gives enough information to disambiguate this name.
+
+\subsection{Theorem Provers}
+\label{sec:pre:theorem-provers}
+
+All the techniques described in this thesis are implemented in two superposition
+theorem provers: E \cite{ss-02-brainiac} and Zipperposition
+\cite{sc-15-simon-phd,bbtvw-21-sup-lam}. The latter was used as a testbed for
+prototyping ideas, while the former is the main target of our work as it is more
+efficient.
+
+\ourpara{E} E is a state-of-the-art first-order prover based on superposition.
+In the last decade, together with its derivatives, it has revolved around the
+second place at the first-order division of CASC theorem prover
+competition \cite{ss-96-casc}. It is open-source, written in C (without using any
+external libraries), and designed around efficient algorithms and data structures
+rather than highly optimized codebase. This makes it a good target for implementing
+successful higher-order techniques.
+
+\ourpara{Zipperposition}
+Zipperposition is a higher-order theorem prover that implements \lfsup{},
+\lsup{}, \osup{}, and other superposition-like calculi (including Bhayat and
+Reger's combinatory superposition \cite{br-20-full-sup-w-combs}). The prover was
+conceived as a testbed for rapidly experimenting with extensions of first-order
+superposition, but over time it has assimilated many of E's techniques and
+heuristics and become quite powerful. Still, its first-order performance is a
+far cry from E's. Zipperposition is written in OCaml, and features a modular
+system for adding new rules and techniques to implemented calculi.
+This makes it attractive for fast evaluation of various extensions of
+superposition.

### 2022-03-16 satfol/satfol.tex
dpe

@@
 $\cst{f}$ (e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$) and call nullary
-function symbols \emph{constants}.
+function symbols \emph{constants}. In this chapter, propositional variables
+(corresponding to predicate constants in first-order logic) are written as
+$\cst{x}, \cst{y}, \cst{p}, \cst{q}, \ldots$, while first-order free variables
+are written in lowercase as $x,y,z,\ldots$.
 
@@
 \end{rep}
+
+\section{Predicate Elimination}
+\label{sec:satfol:predicate-elimination}
+
+For propositional logic, variable elimination \cite{eb-2005-satpreprocess} is
+one of the main preprocessing and inprocessing techniques. Following Gabbay and Ohlbach's ideas
+\cite{go-1992-so-pred-elim}, Khasidashvili and Korovin \cite{kk-2016-pe-fol}
+generalized variable elimination to first-order logic with equality and
+demonstrated that it is effective as a preprocessor. We propose an improvement
+that makes this applicable in more cases and show that, with a minor
+restriction, it can be integrated in a superposition prover
+without compromising its refutational completeness.
+
+\subsection{Singular Predicates}
+
+Khasidashvili and Korovin's preprocessing technique removes singular predicates
+(which they call ``non-self-referential predicates'') from the problem using
+so-called flat resolution.
+
+\begin{defi}
+   A predicate symbol is called \emph{singular}\confrep{ (or ``non-self-referential'')}{}
+   for a clause set $N$ if it occurs at most once in every clause contained in
+   $N.$
+\end{defi}
+
+\begin{defi}
+   \label{def:flat-res}
+   Let $C = \pospredlit{\cst{p}(\tuplen{s})} \cor C'$ and $D =
+   \negpredlit{\cst{p}(\tuplen{t})} \cor D'$ be clauses with no variables in
+   common. The clause $\neqlit{s_1}{t_1} \cor \cdots \cor \neqlit{s_n}{t_n}
+   \cor C' \cor D'$ is a \emph{flat resolvent} of $C$ and $D$ on $\cst{p}$.
+\end{defi}
+
+% Predicate elimination saturates with respect to flat resolvents, removing all
+% clauses containing $\cst{p}$. Given two (possibly identical) clause sets $M, N$,
+% the resolved set $M \flatres_{\!\cst{p}} N$, defined below, repeatedly draws
+% one inference partner from $M$ and the other one from $N$.
+
+Given two (possibly identical) clause sets $M, N$, predicate elimination
+iteratively replaces clauses from $N$ containing the symbol $\cst{p}$ with all flat
+resolvents against clauses in $M$. Eventually, it yields a
+set with no occurrences of $\cst{p}$.
+
+\begin{defi}
+\label{def:flat-res-set}
+   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
+   $M$. Let $\flatresiter$ be the following relation on clause set pairs and clause sets:
+   \begin{enumerate}
+   \item $(M{,}\> \{\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'\} \uplus N) \flatresiter (M{,}\; N' \cup N)$ if
+   $N'$ is the set that consists of all clauses (up to variable renaming) that are flat resolvents
+   with $\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'$ on $\cst{p}$ and a clause from $M$ as premises.
+   The premises' variables are renamed apart.
+
+   \smallskip
+   \item $(M, N) \flatresiter N$ if $N$ has no occurrences of $\cst{p}$.
+   \end{enumerate}
+   \noindent  The \emph{resolved set} $M \flatres_{\!\cst{p}} N$ is the clause set $N'$ such that $(M,N) \flatresiter^* N'$.
+   % obtained by the following
+   % procedure: Set $N' = N$. Then repeatedly (1)~choose a clause $D \in N'$
+   % containing $\cst{p}$, (2)~compute all flat resolvents involving $D$ and
+   % partners from $M$, renaming variables apart, and (3)~replace $D$ with these
+   % flat resolvents in $N'$. Stop when $N'$ contains no more occurrence of
+   % $\cst{p}$.
+\end{defi}
+
+\pagebreak[2]
+
+\begin{rep}
+\begin{lemma}
+   \label{lem:flat-res-set-termination-confluence}
+   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
+   $M$. The resolved set $N'$ is reached in a finite number of $\flatresiter$ steps,
+   and it is unique up to variable renaming.
+\end{lemma}
+\begin{proof}
+   To show $\flatresiter$ is terminating we use the following ordinal measure on
+   clause sets: $\nu(\{D_1,\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus
+   \cdots \oplus \omega^{\nu(D_n)}$, where $\nu(D)$ is the number of
+   occurrences of $\cst{p}$ in $D$, $\omega$ is the first infinite ordinal, and
+   $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
+   For every transition
+   $(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have $\nu(\{C\} \cup N) > \nu(N' \cup N)$
+   because $\omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'|$.
+   Eventually, a state $(M,N')$ with $\nu(N')=\omega^0\cdot n$ is reached.
+   Then, we apply the second rule of Definition \ref{def:flat-res-set} to obtain the resolved set $N'$.
+
+   To show that $N'$ is unique, i.e., $\flatresiter$ is confluent, it suffices
+   to show (since $\flatresiter$ is terminating and Newmann's lemma applies
+   \cite{bn-98-tr-and-all-that}) that $\flatresiter$ is locally confluent. In
+   other words, whenever $(M,N) \flatresiter (M, N_1)$ and $(M,N) \flatresiter
+   (M, N_2)$ there exists $N'$ such that $(M, N_1) \flatresiter (M, N')$ and
+   $(M, N_2) \flatresiter (M,N')$.
+
+   There are two main sources of nondeterminism of $\flatresiter$: The choice of $C
+   \in N$ and the choice of the literal in $C$ to act on. Let us focus on the choice of
+   $C$ in $N$; the same discussion applies for the choice of literal in $C$.
+
+   Let $N = \{C_1\} \uplus \{C_2\} \uplus N'$, where $C_1$ and $C_2$ are
+   clauses with occurrences of $\cst{p}$. Then, $(M, \{C_1\} \uplus \{ C_2 \cup
+   N' \}) \flatresiter (M, N'_1 \cup \{ C_2 \cup N'\})$ and $(M, \{C_2\} \uplus
+   \{C_1 \cup N' \}) \flatresiter (M, N'_2 \cup \{ C_1 \cup N' \})$ where $N'_1$ and
+   $N'_2$ are sets of corresponding resolvents. Both of $\flatresiter$ steps can
+   be joined (up to variable renaming) to $(M, N'_1 \cup N'_2 \cup N')$, showing
+   that $\flatresiter$ is locally confluent.
+\end{proof}
+\end{rep}
+
+\begin{conf}
+The relation $\flatresiter$ is confluent up to variable renaming. Thanks to
+the singularity constraint on $M$, it
+also terminates on finite sets because the following ordinal measure decreases:
+$\nu(\{D_1,\allowbreak\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus \cdots \oplus
+\omega^{\nu(D_n)}$, where $\nu(D)$ counts the occurrences of $\cst{p}$ in
+$D$, $\omega$ is the first infinite ordinal,
+and $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
+For every transition
+$(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have
+$\nu(\{C\}) = \omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'| = \nu(N')$.
+\end{conf}
+
+Next, it is useful to partition
+clause sets into subsets based on the presence and polarity of a singular
+predicate.
+
+\begin{defi}
+   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
+   $N.$ Let $\withpredpos{N}{\cst{p}}$ consist of all clauses of the form $\pospredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
+   let $\withpredneg{N}{\cst{p}}$ consist of all clauses of the form $\negpredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
+   let $\withpred{N}{\cst{p}} = \withpredpos{N}{\cst{p}} \cup \withpredneg{N}{\cst{p}}$,
+   and let $\withoutpred{N}{\cst{p}} = N \setminus \withpred{N}{\cst{p}}$.
+\end{defi}
+
+\begin{defi}
+   \label{def:pred-elim}
+   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
+   $N.$
+   \emph{Singular predicate elimination} (SPE) of $\cst{p}$ in $N$ replaces $N$ by
+   $\withoutpred{N}{\cst{p}} \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
+\end{defi}
+
+The result of SPE is satisfiable if and only if $N$ is satisfiable
+\cite[Theorem~1]{kk-2016-pe-fol}, justifying SPE's use in a preprocessor.
+However, eliminating singular predicates aggressively can dramatically increase
+the number of clauses. To prevent this, Khasidashvili and Korovin suggested to
+replace $N$ by $N'$ only if $\lambda(N') \leq \lambda(N)$
+and $\mu(N') \leq \mu(N)$, where
+$\lambda(N)$ is the number of literals in~$N$ and
+$\mu(N)$ is the sum for all clauses $C \in
+N$ of the square of the number of distinct variables in $C$.
+
+Compared with what modern SAT solvers use, this
+criterion is fairly restrictive. We relax it to make it possible to
+eliminate more predicates, within reason. Let $K_\mathrm{tol} \in \mathbb{N}$ be a tolerance
+parameter. A predicate elimination
+step from $N$ to $N'$ is allowed if
+$\lambda(N') < \lambda(N) + K_\mathrm{tol}$ or
+$\mu(N') < \mu(N)$ or
+$|N'| < |N| + K_\mathrm{tol}$.
+\confrep{}{A refinement, which we want to try out in future work, would be to gradually
+increment the tolerance $K_\mathrm{tol}$, as is done in some SAT solvers.}
+
+\subsection{Defined Predicates}
+
+SPE is effective, but an important refinement has not yet been adapted to
+first-order logic:\ variable elimination by substitution. E{\'{e}}n and Biere
+\cite{eb-2005-satpreprocess} discovered that a propositional variable~$\cst{x}$ can be
+eliminated without computing all resolvents if it is expressible as an
+equivalence $\cst{x} \medleftrightarrow \varphi$, where $\varphi$, the ``gate,''
+is an arbitrary formula that does not reference~$\cst{x}$.
+They partition a set $N$
+\confrep{}{of propositional clauses }into a definition set $G$, essentially the
+clausification of $\cst{x} \medleftrightarrow \varphi$, and
+$R = \withpred{N}{\cst{p}} \setminus G$, the remaining
+clauses containing~$\cst{p}$. To eliminate $\cst{x}$ from $N$ while
+preserving satisfiability, it suffices to resolve clauses from $G$ against
+clauses from $R$, effectively substituting $\varphi$ for $\cst{x}$ in $R$.
+Crucially, we do not need to resolve pairs of clauses from $G$
+or pairs of clauses from $R$.
+We generalize this idea to first-order logic.
+
+\begin{defi}
+   \label{def:definition}
+   \looseness=-1
+   Let $G$ be a clause set, $\cst{p}$ be a predicate symbol, and
+   $\tupleempty{x}$ be distinct variables.
+%
+   The set $G$ is a \emph{definition set} for $\cst{p}$ if
+   \begin{enumerate}
+      \item $\cst{p}$ is singular for $G$
+      \item $G$ consists of clauses of the form $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming)
+      \item the variables in $C'$ are all among $\tupleempty{x}$
+      \item all clauses in $\withpredpos{G}{\cst{p}}
+      \flatres_{\!\cst{p}} \withpredneg{G}{\cst{p}}$ are tautologies
+      \item $E(\tupleempty{\cst{c}})$ is unsatisfiable, where
+      the \emph{environment} $E(\tupleempty{x})$ consists of all subclauses $C'$ of any
+      $\arbpredlit{\cst{p}(\tupleempty{x})} \mathbin{\cor} C' \in G$ and
+      $\tupleempty{\cst{c}}$ is a tuple of distinct fresh constants substituted
+      in for $\tupleempty{x}$
+   \end{enumerate}
+\end{defi}
+
+A definition set $G$ corresponds intuitively to a definition by
+cases in mathematics---e.g.,
+\[\cst{p}(\tuple{x}{}) =
+\begin{cases} \top & \mathrm{if}~\varphi(\tuple{x}{}{}) \\[-\jot]
+\bot & \mathrm{if}~\psi(\tuple{x}{})\end{cases}\]
+Part~4 states that the case conditions are mutually exclusive
+(e.g., $\neglit{\varphi(\tuple{x}{})} \lor \neglit{\psi(\tuple{x}{})}$),
+and part~5 states that they are exhaustive
+(e.g., $\nexists\tuple{\cst{c}}{} .\; \neglit{\varphi(\tuple{\cst{c}}{})} \land \neglit{\psi(\tuple{\cst{c}}{})}$).
+%
+Given a quantifier-free formula $\cst{p}(\tuple{x}{}) \medleftrightarrow
+\varphi(\tuple{x}{})$ with distinct variables $\tuple{x}{}$ such that
+$\varphi(\tuple{x}{})$ does not contain $\cst{p}$, any reasonable
+clausification algorithm would produce a definition set for $\cst{p}$.
+
+\begin{exa}
+   \label{example:tautologies}
+   Given the formula $\cst{p}(x) \medleftrightarrow \cst{q}(x) \mathrel\land
+   (\cst{r}(x) \cor \cst{s}(x))$, a standard clausification algorithm
+   \cite{nw-01-small-cnf} produces $\{ \negpredlit{\cst{p}(x)} \cor
+   \pospredlit{\cst{q}(x)}{,}\; \negpredlit{\cst{p}(x)} \cor
+   \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
+   \pospredlit{\cst{p}(x)} \cor \negpredlit{\cst{q}(x)} \cor
+   \negpredlit{\cst{r}(x)}{,}\; \pospredlit{\cst{p}(x)} \cor
+   \negpredlit{\cst{q}(x)} \cor \negpredlit{\cst{s}(x)}\}$, which qualifies
+   as a definition set for $\cst{p}$.
+\end{exa}
+
+Definition sets generalize E{\'{e}}n and Biere's
+gates. They can be recognized syntactically for formulas such as
+$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigvee_{\!i} \cst{q}_i(\tuple{s_i}{})$ or
+$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigwedge_i \cst{q}_i(\tuple{s_i}{})$,
+or semantically: Condition~4 can be checked
+using the congruence closure algorithm, and condition~5 amounts to
+a propositional unsatisfiability check.
+
+The key result about propositional gates carries over to definition sets.
+
+\begin{defi}
+   \looseness=-1
+   Let $N$ be a clause set, $\cst{p}$ be a predicate symbol,
+   $G \subseteq N$ be a definition set
+   for $\cst{p}$, and $R = \withpred{N}{\cst{p}}
+   \setminus G$. \emph{Defined predicate elimination} (DPE) of $\cst{p}$ in $N$ replaces $N$ by
+   $\withoutpred{N}{\cst{p}} \cup
+%%% @PETAR: Double-check the supression of _p below. --JB
+   (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}} \confrep{\withpred{R}{\cst{p}}}{R})$.
+\end{defi}
+
+\newcommand{\mm}{\ourmodel}
+\newcommand{\notmodels}{\nvDash}
+\begin{rep}
+   \begin{lemma}
+   \label{lem:unsat-fresh-consts}
+   Let $N(\tupleempty{x})$ be a clause set such that the variables of all
+   clauses in it are among the argument $n$-tuple $\tupleempty{x}$, and let
+   $\tupleempty{\cst{c}}$ be an $n$-tuple of distinct fresh constants. If
+   $N(\tupleempty{\cst{c}})$ {\upshape(}i.e.,
+   $\substterm{\{\tupleempty{x}\mapsto\tupleempty{\cst{c}}\}}{N(\tupleempty{x})}${\upshape)} is
+   unsatisfiable, then for every interpretation $\mm$ and valuation $\xi$,
+   $\mm \notmodels_\xi N$.
+   \end{lemma}
+   \begin{proof}
+      We show the contrapositive. Assume that for some $\mm$ and $\xi$,  $\mm \models_\xi N(\tupleempty{x})$.
+      Then let $\ourmodel'$
+      be a model that assigns each $\cst{c}_i$ the interpretation of $x_i$ under $\mm$ and $\xi$, and otherwise
+      coincides with $\ourmodel$. We obtain $\ourmodel' \models N(\tupleempty{\cst{c}})$.
+   \end{proof}
+
+   \begin{lemma}
+      \label{lem:flat-res-set-step-satisfiability}
+      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }an arbitrary clause
+      set. If $(G, N) \flatresiter (G, N')$ then $G \cup N$ and $G \cup N'$ are
+      equivalent.
+   \end{lemma}
+   \begin{proof}
+      Since flat resolution is sound, the nontrivial direction is to show that a
+      model $\mm$ of the set $G \cup N'$ is also a model of $G \cup N$. As the only clause in $N
+      \setminus N'$ is $C = \arbpredlit{\cst{p}(\tuple{s}{n})} \cor C'$ on which
+      the $\flatresiter$ step is performed, we must show $\mm \models C$.
+      % Without
+
+      % Let $\mm$ be canonical model of $G \cup N'$ \cite[Theorem~9.5.2]{mf-1996-fol} and let $C
+      % \in N \setminus N'$. We need to show that $\mm \models C$---which is to say,
+      % for all variable interpretation $\xi$, we have $\mm \models_\xi C$. To
+      % lighten notations, we will assume $C$ is ground and ignore $\xi$. Canonical
+      % models make this step possible.
+
+      % First, $C$ must have the form $\arbpredlit{\cst{p}(\tuple{s}{})} \cor C'$,
+      % where $\tuple{s}{}$ is a tuple of ground terms. 
+      Without loss of generality,
+      we assume that the leading literal of $C$ is positive.
+      Towards a contradiction, assume $\xi$ is a valuation such that $\mm \notmodels_\xi C$. Then, $\mm \notmodels_\xi
+      \pospredlit{\cst{p}(\tuple{s}{n})}$. 
+      % Let $\xi'$ be a valuation that
+      % assigns every instance $D = \pospredlit{\cst{p}(\tupleempty{s})} \cor D'$ of a clause in
+      % $G$, since $\mm \notmodels \pospredlit{\cst{p}(\tuple{s}{})}$, we have $\mm
+      % \models D'$.
+      Consider an arbitrary clause $D = \pospredlit{\cst{p}}(\tuplen{x}) \cor D'
+      \in \withpredpos{G}{\cst{p}}$ and a valuation $\xi'$, which assigns each $x_i$ the
+      interpretation of $s_i$ under $\mm$ and $\xi$. As $\mm \notmodels_{\xi'}
+      \pospredlit{\cst{p}}(\tuple{x}{n})$ and $\mm \models G$, then $\mm \models_{\xi'} D'$ for every
+      such clause $D$.
+   %
+      However, by part 5 of Definition~\ref{def:definition} and by Lemma
+      \ref{lem:unsat-fresh-consts}, $\mm \notmodels_{\xi'} E(\tuplen{x})$, where $E(\tuplen{x})$
+      is the environment associated with the definition set $G$.
+      Therefore, there must exist a clause $D =
+      \negpredlit{\relax{\cst{p}(\tuplen{x})}} \cor D'$ in $\withpredneg{G}{\cst{p}}$ such
+      that $\mm \notmodels_\xi D'$.
+
+      Now consider the flat resolvent of $C$ and $D$ on~$\cst{p}$: $R =
+      \neqlit{x_1}{s_1} \cor \cdots \cor \neqlit{x_n}{s_n} \cor C' \cor D'$.
+      Let $\zeta$ be a valuation coinciding with $\xi$ on variables of $C$
+      and with $\xi'$ on $\tuplen{x}$.
+      Clearly, $\mm \notmodels_\zeta R$. Yet, $R \in N'$, and as $\mm \models N'$, we reach a contradiction.
+   \end{proof}
+
+   \begin{lemma}
+      \label{lem:flat-res-set-last-step-satisfiability}
+      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }a clause set with no occurrences of $\cst{p}$.
+      Then $G \cup N$ is satisfiable if and only if $N$ is satisfiable.
+   \end{lemma}
+   \begin{proof}
+      The nontrivial direction is to show that if $N$ is satisfiable, $G \cup N$ is as well.
+      %
+      Let $\mm$ be a model of $N$. We construct a model $\mm'$ of $G$ over
+      the same universe as $\mm$. For any atom $A$ such that $\cst{p}$ does not
+      occur in $A$ and for every $\xi$, we set $\mm' \models_\xi A$ if and only if
+      $\mm \models_\xi A$. For any clause $\pospredlit{\cst{p}(\tuplen{x})} \cor
+      C' \in G$ and any assignment $\xi$ such that $\mm \notmodels_\xi C'$, we
+      define $\mm'$ so that $\mm'\models_\xi \cst{p}(\tuplen{x})$. By
+      construction, $\mm' \models \withpredpos{G}{\cst{p}} \cup N$. It remains to show that $\mm' \models
+      \withpredneg{G}{}$.
+
+      Let $C = \negpredlit{\cst{p}(\tuplen{x})} \cor C' \in G$ and let $\xi$ be
+      an arbitrary assignment. Towards a contradiction, assume
+      $\mm'\notmodels_\xi C$, and consequently $\mm' \models_\xi
+      \cst{p}(\tuplen{x})$. By construction of $\mm'$, there exists a clause
+      $\pospredlit{\cst{p}(\tuplen{y})} \cor D' \in G$ and an assignment $\xi'$
+      which assigns each $y_i$ value of $\xi(x_i)$ such that $\mm
+      \notmodels_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
+      \neqlit{x_n}{y_n} \lor C' \lor D'$ is a tautology, according to condition
+      4 of Definition \ref{def:definition}. However, for a valuation that
+      behaves like $\xi$ on $\tupleempty{x}$ and $\xi'$ on $\tupleempty{y}$,
+      $\mm'$ does not satisfy $R \in N$, contradicting our assumption.
+      % @Jasmin: Your trick with var renaming does not work as flat resolvent is
+      % defined on
+      % variable-disjoint clauses.
+   \qedhere
+   \end{proof}
+\end{rep}
+\begin{conf}
+
+\end{conf}
+\begin{theorem}
+\label{thm:pes-sat-equiv}
+   The result of applying DPE to a clause set $N$ is
+   satisfiable if and only if $N$ is satisfiable.
+\end{theorem}
+\begin{rep}
+\begin{proof}
+   Let $\cst{p}$ be a predicate symbol and $G \subseteq N$ be the
+   definition set used by DPE, and let $R = \withpred{N}{\cst{p}} \setminus G$.
+%
+   % First, using Lemmas \ref{lem:flat-res-set-termination-confluence} and \ref{lem:flat-res-set-step-satisfiability}
+   % we show that we will eventually obtain a set $R'$
+   % that contains no occurrences of $\cst{p}$ and that preserves satisfiability
+   % of the original $R$. Finally, we will show that we can
+   % omit $G$ in the result without affecting satisfiability.
+
+   Using Lemma
+   \ref{lem:flat-res-set-termination-confluence}, we get that there is a
+   derivation $(G,R) \flatresiter^n (G,R') \flatresiter R'$. Applying Lemma
+   \ref{lem:flat-res-set-step-satisfiability} $n$ times, we get that $G \cup R$
+   is equivalent to $G \cup R'$. Finally, Lemma \ref{lem:flat-res-set-last-step-satisfiability}
+   gives us the desired result.
+\end{proof}
+\end{rep}
+
+Since there will typically be at most only a few defined predicates in the
+problem, it makes sense to fall back on SPE when no definition is found.
+
+\begin{defi}
+   \looseness=-1
+   Let $N$ be a clause set and $\cst{p}$ be a predicate symbol. If there exists a
+   definition set $G \subseteq N$ for $\cst{p}$, \emph{portfolio predicate
+   elimination} (PPE) on $\cst{p}$ in $N$ replaces $N$ with
+%%% @PETAR: Double-check the supression of _p below. --JB
+   $\withoutpred{N}{\cst{p}} \cup (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}}
+   \confrep{\withpred{R}{\cst{p}}}{R})$, where $R = \withpred{N}{\cst{p}} \setminus G$.
+   Otherwise, if $\cst{p}$ is singular in $N$, it results in $\withoutpred{N}{\cst{p}}
+   \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
+   In all other cases, it is not applicable.
+\end{defi}

### 2022-03-16 scratch/paper.tex
dpe

@@
 
-\newenvironment{lemmax}% environment name
+\newenvironment{lemma}% environment name
 {% begin code
@@
 
-\maybesubsection{Singular Predicates}
+\subsection{Singular Predicates}
 
@@
 \begin{rep}
-\begin{lemmax}
+\begin{lemma}
    \label{lem:flat-res-set-termination-confluence}
@@
    and it is unique up to variable renaming.
-\end{lemmax}
+\end{lemma}
 \begin{proof}
@@
 
-\maybesubsection{Defined Predicates}
+\subsection{Defined Predicates}
 
@@
 \begin{rep}
-   \begin{lemmax}
+   \begin{lemma}
    \label{lem:unsat-fresh-consts}
@@
    unsatisfiable, then for every interpretation $\mm$ and valuation $\xi$,
-   $\mm \not\models_\xi N$.
-   \end{lemmax}
+   $\mm \notmodels_\xi N$.
+   \end{lemma}
    \begin{proof}
@@
 
-   \begin{lemmax}
+   \begin{lemma}
       \label{lem:flat-res-set-step-satisfiability}
@@
       equivalent.
-   \end{lemmax}
+   \end{lemma}
    \begin{proof}
@@
       we assume that the leading literal of $C$ is positive.
-      Towards a contradiction, assume $\xi$ is a valuation such that $\mm \not\models_\xi C$. Then, $\mm \not\models_\xi
+      Towards a contradiction, assume $\xi$ is a valuation such that $\mm \notmodels_\xi C$. Then, $\mm \notmodels_\xi
       \pospredlit{\cst{p}(\tuple{s}{n})}$. 
@@
       % assigns every instance $D = \pospredlit{\cst{p}(\tupleempty{s})} \cor D'$ of a clause in
-      % $G$, since $\mm \not\models \pospredlit{\cst{p}(\tuple{s}{})}$, we have $\mm
+      % $G$, since $\mm \notmodels \pospredlit{\cst{p}(\tuple{s}{})}$, we have $\mm
       % \models D'$.
@@
       \in \withpredpos{G}{\cst{p}}$ and a valuation $\xi'$, which assigns each $x_i$ the
-      interpretation of $s_i$ under $\mm$ and $\xi$. As $\mm \not\models_{\xi'}
+      interpretation of $s_i$ under $\mm$ and $\xi$. As $\mm \notmodels_{\xi'}
       \pospredlit{\cst{p}}(\tuple{x}{n})$ and $\mm \models G$, then $\mm \models_{\xi'} D'$ for every
@@
       However, by part (5) of Definition~\ref{def:definition} and by Lemma
-      \ref{lem:unsat-fresh-consts}, $\mm \not\models_{\xi'} E(\tuplen{x})$, where $E(\tuplen{x})$
+      \ref{lem:unsat-fresh-consts}, $\mm \notmodels_{\xi'} E(\tuplen{x})$, where $E(\tuplen{x})$
       is the environment associated with the definition set $G$.
@@
       \negpredlit{\relax{\cst{p}(\tuplen{x})}} \cor D'$ in $\withpredneg{G}{\cst{p}}$ such
-      that $\mm \not\models_\xi D'$.
+      that $\mm \notmodels_\xi D'$.
 
@@
       and with $\xi'$ on $\tuplen{x}$.
-      Clearly, $\mm \not\models_\zeta R$. Yet, $R \in N'$, and as $\mm \models N'$, we reach a contradiction.\qed
+      Clearly, $\mm \notmodels_\zeta R$. Yet, $R \in N'$, and as $\mm \models N'$, we reach a contradiction.\qed
    \end{proof}
 
-   \begin{lemmax}
+   \begin{lemma}
       \label{lem:flat-res-set-last-step-satisfiability}
@@
       Then $G \cup N$ is satisfiable if and only if $N$ is satisfiable.
-   \end{lemmax}
+   \end{lemma}
    \begin{proof}
@@
       $\mm \models_\xi A$. For any clause $\pospredlit{\cst{p}(\tuplen{x})} \cor
-      C' \in G$ and any assignment $\xi$ such that $\mm \not\models_\xi C'$, we
+      C' \in G$ and any assignment $\xi$ such that $\mm \notmodels_\xi C'$, we
       define $\mm'$ so that $\mm'\models_\xi \cst{p}(\tuplen{x})$. By
@@
       an arbitrary assignment. Towards a contradiction, assume
-      $\mm'\not\models_\xi C$, and consequently $\mm' \models_\xi
+      $\mm'\notmodels_\xi C$, and consequently $\mm' \models_\xi
       \cst{p}(\tuplen{x})$. By construction of $\mm'$, there exists a clause
@@
       which assigns each $y_i$ value of $\xi(x_i)$ such that $\mm
-      \not\models_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
+      \notmodels_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
       \neqlit{x_n}{y_n} \lor C' \lor D'$ is a tautology, according to condition
@@
 
-\maybesubsection{Refutational Completeness}
+\subsection{Refutational Completeness}
 \label{ssec:predicate-elimination-refutational-completeness}
@@
 is defined via an ad hoc nonclassical logic so that $\{\cst{p} \cor C{,}\;
-\neglit{\cst{p}} \cor C\} \not\modelsWk \{C\}$ and yet $\modelsWk \{\cst{p} \cor
+\neglit{\cst{p}} \cor C\} \notmodelsWk \{C\}$ and yet $\modelsWk \{\cst{p} \cor
 \neglit{\cst{p}}\}$.
@@
 
-\begin{lemmax}
+\begin{lemma}
 \label{lem:p-Wk-entail}
@@
 \cup \{C\} \modelsWk \{D\}$, then $N \modelsWk \{D\}$.
-\end{lemmax}
+\end{lemma}
 
@@
 
-\begin{lemmax}
+\begin{lemma}
 \label{lem:gcp-derivations-are-red-black-derivations}
 Every $\Longrightarrow_\mathsf{GCP}$-derivation is a mixed derivation.
-\end{lemmax}
+\end{lemma}
 
@@
 
-\begin{lemmax}
+\begin{lemma}
 \label{lem:gcp-invar-gcp}
@@
 $\Inv\Wksup_\mathcalx{N}(k)$ holds for all indices~$k$.
-\end{lemmax}
+\end{lemma}
 
@@
 
-\begin{lemmax}
+\begin{lemma}
 \label{lem:gcp-invar-infinity}
@@
 holds.
-\end{lemmax}
+\end{lemma}
 
@@
 
-\begin{lemmax}
+\begin{lemma}
 \label{lem:fair-gcp-derivations}
@@
 $(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
-\end{lemmax}
+\end{lemma}
 
@@
 
-\begin{lemmax}
+\begin{lemma}
 Let $C \in N$ be a clause containing a quasipure literal. If $N \setminus \{C\}$
 is satisfiable, then $N$ is satisfiable.
-\end{lemmax}
+\end{lemma}
 
@@
    literal in $C\sigma$ must be satisfied by $\ourmodel$, contradicting that
-   $C\sigma$ is falsified by $\ourmodel$. If $\ourmodel \not\models L'\tau$
+   $C\sigma$ is falsified by $\ourmodel$. If $\ourmodel \notmodels L'\tau$
    then some literal in $D\tau$ must be satisfied. As $D\tau \subseteq C\sigma$,

### 2022-03-17 bools/bools.tex
new bib and paper
wan

@@
 functional type, whereas Steen relies on the unification rules of Leo-III's
-calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality. As a
+calculus \cite[Sect.~4.3.3]{as-18-phd} to deal with extensionality. As a
 pragmatic extension of the \lsup{} calculus, Bentkamp et al.
@@
 \cst{impl}, \cst{equiv},\allowbreak \cst{forall}, \cst{exists}, \cst{choice}$, and $\cst{eq}$ which 
-correspond to the homologous logical constants from Section \ref{sect:bool:background}. In their
+correspond to the homologous logical constants from Sect.~\ref{sect:bool:background}. In their
 type declarations, $o$ is replaced by $\typ{bool}$.
@@
 What our work misses is an overview of heuristics that can be used to curb the
-explosion incurred by some of the rules described in this paper. In the next chapter,
+explosion incurred by some of the rules described in this chapter. In the next chapter,
 we explore exactly this topic.

### 2022-03-17 ehoh/ehoh.tex
wan

@@
 and its semantics by interpreting the Boolean type $o$ as a domain of
-cardinality~2. Functional extensionality can be obtained by adding suitable
-axioms \cite[Sect.~3.1]{bbcw-21-lfho}.
+cardinality~2. Functional extensionality, the property that two functions are equal if they
+always return the same value when given the same arguments, can be obtained by
+adding suitable axioms can be obtained by adding suitable axioms
+\cite[Sect.~3.1]{bbcw-21-lfho}.
 
@@
 $\sigma$. Applying $\sigma$ to a variable beyond $\sigma$'s domain is the
-identity. We deviated from the view of substitutions in Section \ref{sec:pre:hol}
-as it made proofs in Sections \ref{sec:ehoh:unif-match} and \ref{sec:ehoh:indexing}
+identity. We deviated from the view of substitutions in Sect.~\ref{sec:pre:hol}
+as it made proofs in Sect.~\ref{sec:ehoh:unif-match} and \ref{sec:ehoh:indexing}
 easier. It is easy to check that both views are equivalent. We also consider unification 

### 2022-03-17 ho-techniques/ho-techniques.tex
wan

@@
 discussed many of the techniques with me and suggested important updates. Visa
-Nummelin worked on implementation of FOOL preprocessing. Simon Cruanes is the
+Nummelin worked on the implementation of FOOL preprocessing. Simon Cruanes is the
 original developer of Zipperposition and provided us with invaluable knowledge.
@@
 \textsf{SKBCI} combinators \cite{br-20-full-sup-w-combs}, resulting in a calculus called combinatory superposition. Significant progress has been
-made on SMT front as well \cite{brotb-19-ho-smt}.
-
-In 2019 we tested if the idea of gracefully extending first-order provers to
-higher-order logic really improves the state of the art for the first time. We
+made on the SMT front as well \cite{brotb-19-ho-smt}.
+
+In 2019 we tested for the first time if the idea of gracefully extending first-order provers to
+higher-order logic really improves the state of the art. We
 implemented \lsup{} \cite{bbtvw-21-sup-lam} in Zipperposition~1.5 with basic
@@
 incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
-well on a wide range of problems. The set of parameters used for baseline configuration in this
+well on a wide range of problems. The set of parameters used for the baseline configuration in this
 chapter differs from the one in the previous chapter.
@@
 manifests itself on definitions of nonpredicate symbols. In some cases, it is
-preferable to rely on superposition's term order and powerful simplification
+preferable to rely on superposition's term order and the powerful simplification
 engine to rewrite the proof state rather than to blindly rewrite definitions. On
@@
 
-The results are given in Fig.~\ref{fig:rewrite}. In all the figures in this chapter, each
+The results are given in Figure~\ref{fig:rewrite}. In all the figures in this chapter, each
 cell gives the number of proved problems, and cells marked with $\star$
@@
 no substantial improvement: Looking at raw data, we found only
-\NumberOK{2}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
+\NumberOK{two}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
 feature was used in the proof. For no-RW and no-RW$+$KBO, the
-\NumberOK{2}-problem difference may be just noise. Even though it proves fewer
+\NumberOK{two}-problem difference may be just noise. Even though it proves fewer
 problems, the configuration $o$-RW has some advantages over pre-RW: It proves
-\NumberOK{16} problems that pre-RW does not, \NumberOK{3} of which have a TPTP
+\NumberOK{16} problems that pre-RW does not, \NumberOK{three} of which have a TPTP
 difficulty rating (the ratio of eligible %TPTP
@@
 % \looseness=-1
-Rewriting after clausification avoids getting stuck rewriting parts of the
+Rewriting after clausification avoids getting stuck in rewriting parts of the
 proof state that might not contribute to the proof. In practice, we noticed that
@@
 pre-RW. Moreover, there are \NumberOK{41} problems that can be proved only
-by in-RW but not by any other above described configuration. % from Fig.~\ref{fig:rewrite}.
+by in-RW but not by any other above described configuration. % from Figure~\ref{fig:rewrite}.
 
@@
 %
-Overall, the lightweight AVATAR harms performance, but the sets of
+Overall, lightweight AVATAR harms performance, but the sets of
 problems proved with and without it are vastly different. For example,
@@
 %To assess it, we enabled it in the best configuration from
-Fig.~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
+Figure~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
 Zipperposition proves \NumberOK{1700} TPTP problems and \NumberOK{456} SH
 problems.
-On TPTP, even though $+$\relax{BI} solves \NumberOK{3} problems less
+On TPTP, even though $+$\relax{BI} solves \NumberOK{three} problems less
 than \relax{DCS}$-$LA, it is very useful: \NumberOK{41}
@@
 The $+$\relax{BI} configuration proves \NumberOK{18} TPTP problems no other
-configuration from Fig.~\ref{fig:avatar-clause} can prove. Among these is
+configuration from Figure~\ref{fig:avatar-clause} can prove. Among these is
 \texttt{DAT056\^{}2} (Example~\ref{ex:dat056-2}). In contrast, on SH benchmarks, only
-\NumberOK{6} problems are proved using $+$\relax{BI} and not
+\NumberOK{six} problems are proved using $+$\relax{BI} and not
 \relax{DCS}$-$\relax{LA}. For all these problems, Boolean instantiation does not
@@
 A Boolean selection function chooses \emph{green subterms} of Boolean type,
-which are different than $\itrue$ or $\ifalse$ and not occurring at either side
+which are different than $\itrue$ or $\ifalse$ and do not occur at either side
 of a positive literal in a clause. It gives rise to a notion of eligibility that
@@
 of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
-t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
+t\}$ of a clause $C$, the green position of the same subterm in $C$ is denoted by $l.s.p$.
 \osup{} mandates additional restrictions on the Boolean selection function:
-$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
-$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
-variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
-maximal term of the clause, $s$ cannot be selected.
+$\itrue$, $\ifalse$, and variable-headed terms cannot be selected; for literals
+$s \eq t$, neither $s$ nor $t$ cannot be selected; if $s$ (or $t$) contains a
+variable $X$ as a green subterm and $X\> \tuplen{u}$, with $n \ge 1$, is a
+maximal term of $C$, then $s$ (or $t$) cannot be selected.
 
@@
   $\sigmaterm{t} \not\succ \sigmaterm{s}$, $s$ is eligible. If a subterm $t$ is eligible
-  and the head of $t$ is not $\ieq$ or $\ineq$, all direct green subterms of
+  and the head of $t$ is not $\ieq$ or $\ineq$, all green subterms directly below the head of
   $t$ are eligible. If a subterm $t$ is eligible and $t$ is of the form $u
@@
 smallest, largest, innermost, or outermost terms, but they did not impact
-performance much. A similar idea has been discussed in the context of \infname{Cases}(\infname{Simp}) rule
+performance much. A similar idea has been discussed in the context of the \infname{Cases}(\infname{Simp}) rule
 in Sect.~\ref{sect:bool:native}.  Here, we propose alternatives.
@@
 first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus requires
-selection of subterms occurring in negative green positions, defined below.
+only allows the selection of subterms that occur in negative green positions, defined below.
 
@@
      (longest valid green subterm position), make this subterm larger; otherwise, if one of the
-     subterms has less distinct variables, make this subterm larger. \\[2\jot]
+     subterms has fewer distinct variables, make this subterm larger. \\[2\jot]
   $\vartriangleright_\text{def}$ & If exactly one of the subterms is of the form
@@
   \, f^i\, (x) \}, i \geq 0$, whereas terms $\cst{p} \, (\lambda y. \, X
-  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have the MGU. 
+  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have a finite CSU (in fact an MGU). 
   When \texttt{Forward} context selection is enabled, % the subterm
@@
 by instantiating the contextualized Boolean selection function are shown in
-Fig.~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
+Figure~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
 while columns denote the function $\mathit{ctx}$.
@@
 Also, the best results are obtained when a context more specific than \texttt{Any}
-is chosen. Remarkably, functions using \texttt{Pos} context
-perform better than the ones using \texttt{Neg} context
-on TPTP but the opposite is observed on SH.
+is chosen. Remarkably, functions using the \texttt{Pos} context
+perform better than the ones using the \texttt{Neg} context
+on TPTP, but the opposite is observed on SH.
 
@@
 proved problems on TPTP benchmarks: In total, there are \NumberOK{103} problems
-proved by some configuration from Fig.~\ref{fig:bool-sel} but not by
+proved by some configuration from Figure~\ref{fig:bool-sel} but not by
 \relax{DCI}$-$\relax{LA}. However, there are only \NumberOK{16} such SH problems.
@@
 difficult heuristic choice. Small sets can result in missing the necessary unifier, while large sets make the prover spend too long in
-the unification procedure, generate useless clauses, and possibly get sidetracked into the wrong parts of the search space.
+the unification procedure, generate useless clauses, and possibly get sidetracked into wrong parts of the search space.
 
@@
 Given that we cannot decide whether there exists a next CSU
-element in a stream of unifiers, the request for the next conclusion might not
+element in a stream of unifiers, a request for the next conclusion might not
 terminate,
@@
 clauses, we extract elements from, or \emph{probe}, streams and move any obtained
-clauses to $P$. Analogously to the traditional pick--given ratio
+clauses to $P$. Analogous to the traditional pick--given ratio
 \cite{ss-02-brainiac,mcw-1997-otter}, we use a parameter
@@
 \looseness=-1
-For evaluation of our unification algorithm (using somewhat different
-evaluation setup and older version of Zipperposition), we also compared a complete, nonterminating variant of the
+To evaluate our unification algorithm (using a somewhat different
+evaluation setup and an older version of Zipperposition), we also compared a complete, nonterminating variant of the
 unification procedure and a pragmatic, terminating variant. The
@@
 instructions on how to create a successor state in the tableau. One
-command describes infinite enumeration of all closed terms of a given function
-type. Each execution of this command makes progress in the enumeration. Unlike
-evaluation of streams representing elements of CSU, each command execution
+command describes an infinite enumeration of all closed terms of a given function
+type. Each execution of this command makes progress in the enumeration. In contrast to
+computing inferences using streams representing elements of CSU, each command execution
 is guaranteed to make progress in enumerating the next closed functional
@@
 the conjecture, the ones that have at least one positive literal, the ones that
-have been moved from active to passive set, and so on. One of the queues uses the constant
+have been moved from the active to the passive set, and so on. One of the queues uses the constant
 priority function \texttt{ConstPrio} (\texttt{CP}), meaning that it assigns the same
@@
 %functions,
-leaving the clause weight intact. The results are shown in Fig.~\ref{fig:priorities}.
-
-Even though constant priority function achieves remarkable performance, the new priority functions are
+leaving the clause weight intact. The results are shown in Figure~\ref{fig:priorities}.
+
+Even though the constant priority function achieves a remarkable performance, the new priority functions are
 useful additions to the prover's repertoire: \NumberOK{37} additional
@@
 nonconstant priority is used. The generally average-performing \verb|PEHO|
-function can prove \NumberOK{9} problems not proved with any other priority function on TPTP
-(and \NumberOK{1} on SH). Globally, \NumberOK{24} TPTP problems and \NumberOK{6} SH problems
+function can prove \NumberOK{nine} problems not proved with any other priority function on TPTP
+(and \NumberOK{one} on SH). Globally, \NumberOK{24} TPTP problems and \NumberOK{six} SH problems
 can be proved exclusively using one particular priority function.
@@
 increments~$\ParamPenaltyIncrease$ for \infname{FluidSup} inference queues. The
-results are shown in Fig.~\ref{fig:penalties}. As expected, using a low
+results are shown in Figure~\ref{fig:penalties}. As expected, using a low
 increment with \infname{FluidSup} is detrimental on TPTP. On this
@@
 \infname{FluidSup} is enabled. The penalty mostly affects only proving time: All
-but \NumberOK{2} of these problems were proved by using at least three different
+but \NumberOK{two} of these problems were proved by using at least three different
 values of $\ParamPenaltyIncrease$. On SH problems, the best result is obtained
@@
 they hardly discussed the practical aspects of higher-order superposition. Extensions
-of SMT are discussed by Barbosa et al. \cite{brotb-19-ho-smt}.
+of SMT are discussed by Barbosa et al.\ \cite{brotb-19-ho-smt}.
 Bachmair and Ganzinger \cite{bg-1992-nonclausal}, Manna and Waldinger
@@
 described in this chapter, it outperforms all higher-order provers except for
-Zipperposition. We give detailed a description of $\lambda$E, as well as reasoning
-behind all design decision we took in Chapter \ref{ch:ehoh2}.
+Zipperposition. We give a detailed description of $\lambda$E, as well as the reasoning
+behind all design decisions we took in Chapter \ref{ch:ehoh2}.

### 2022-03-17 pre/pre.tex
wan

@@
 
+\looseness=-1
 {\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
@@
 The composition $\varrho\sigma$ of substitutions is defined by
-$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution $\sigma$ is {\em grounding}
+$\left(\varrho\sigma\right)t=\varrho\left(\sigma (t)\right)$. We call a substitution $\sigma$ {\em grounding}
 if $\sigma(x) \not= x$ implies that $\sigma(x)$ is ground.
@@
 $\upsilon$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau
-\rightarrow \upsilon$. The syntactic distinction between free and bound variables
-gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
-\cite{tn-93-patterns}. Note that there is no requirement that a symbol is not
-applied to terms of Boolean type as in first-order logic. Furthermore, all
-logical connectives are symbols present in the set $\Sigma$. This means that
-higher-order logic does not distinguish between terms and formulas. For example,
-$\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand \cst{q})$ are well-formed
-terms of higher-order logic. For convenience, we still call terms of Boolean
-type \emph{formulas}.
+\rightarrow \upsilon$. The syntactic distinction between free and bound
+variables gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
+\cst{a}}$) \cite{tn-93-patterns}. Note that there is no requirement that a
+symbol is not applied to terms of Boolean type as in first-order logic.
+Furthermore, all logical connectives are symbols present in the set $\Sigma$.
+This means that higher-order logic does not distinguish between terms and
+formulas. For example, $\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand
+\cst{q})$ are well-formed terms of higher-order logic. For convenience, we still
+call terms of Boolean type \emph{formulas}. Our logic also supports Hilbert
+choice operator $\varepsilon$. Intuitively, this operator chooses an arbitrary value that
+satisfies a predicate and it is axiomatized as follows: $\iforall p. \, (\iexists
+x. \, p \, x) \iimplies p \, (\varepsilon \, p)$.
 

### 2022-03-17 scratch/paper.tex
wan

@@
 
-The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
+The results are given in Figure~\ref{fig:rewrite}. In all the figures, each
 cell gives the number of proved problems, and cells marked with $\star$
@@
 %To assess it, we enabled it in the best configuration from
-Fig.~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
+Figure~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
 Zipperposition proves \NumberOK{1700} TPTP problems and \NumberOK{456} SH
@@
 The $+$\relax{BI} configuration proves \NumberOK{18} TPTP problems no other
-configuration from Fig.~\ref{fig:avatar-clause} can prove. Among these is
+configuration from Figure~\ref{fig:avatar-clause} can prove. Among these is
 \texttt{DAT056\^{}2} (Example~\ref{ex:dat056-2}). In contrast, on SH benchmarks, only
@@
 %functions,
-leaving the clause weight intact. The results are shown in Fig.~\ref{fig:priorities}.
+leaving the clause weight intact. The results are shown in Figure~\ref{fig:priorities}.
 

### 2022-03-17 unif/unif.tex
wan

@@
 
-Building on the JP procedure, we designed a new procedure (Section\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Sect.~\ref{sec:unif:proof-of-completeness}).
+Building on the JP procedure, we designed a new procedure (Sect.~\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Sect.~\ref{sec:unif:proof-of-completeness}).
 The new procedure addresses many of the issues that are

### 2022-03-17 satfol/satfol.tex
more

@@
 \end{defi}
+
+\subsection{Refutational Completeness}
+\label{ssec:predicate-elimination-refutational-completeness}
+
+\newcommand\FInf{\mathit{FInf}}
+\newcommand\FLInf{\mathit{FLInf}}
+\newcommand\Red{\mathit{Red}}
+\newcommand\RedI{\Red_\mathrm{I}}
+\newcommand\RedF{\Red_\mathrm{F}}
+\newcommand\LRed{\mathit{LRed}}
+\newcommand\LRedI{\LRed_\mathrm{I}}
+\newcommand\LRedF{\LRed_\mathrm{F}}
+\newcommand\gnd{\mathcalx{G}}
+\newcommand\pow[1]{\mathcalx{pe}(#1)}
+\newcommand\Lab{\mathbf{L}}
+\newcommand\FFLab{{\mathbf{FL}}}
+\newcommand\Inv{\mathit{Inv}}
+
+\newcommand\RedWk{\smash{\Red\Wksup}}
+\newcommand\RedIWk{\smash{\Red_\mathrm{I}\Wksup}}
+\newcommand\RedFWk{\smash{\Red_\mathrm{F}\Wksup}}
+
+\newcommand\Wksym{\flat}
+\newcommand\Wksup{^\Wksym}
+\newcommand\modelsWk{\models\Wksup}
+%\newcommand\modelsWkcapgnd{\models^{\Wksym{}\gnd}_\cap}
+\newcommand\modelsWkcapgnd{\modelsWk}
+\newcommand\LRedWk{\LRed^{\Wksym}}
+\newcommand\LRedWksub{\LRed^{\Wksym,\sqsupset}}
+\newcommand\LRedFWksub{\LRedF^{\Wksym,\sqsupset}}
+\newcommand\LRedIWk{\LRedI^{\Wksym}}
+\newcommand\Wk[1]{#1^\Wksym}
+\newcommand\prop{o}
+\newcommand\true{\top}
+\newcommand\false{\bot}
+
+\looseness=-1
+Hidden-literal-based techniques fit within the traditional framework of
+saturation, because they delete or reduce a clause based on the \emph{presence}
+of other clauses. In contrast, predicate elimination relies on the
+\emph{absence} of clauses from the proof state. We can still integrate it with
+superposition as follows: At every $k$th iteration of the given clause
+procedure, perform predicate elimination on $\mathcal{A}
+\cup \mathcal{P}$, and add all new clauses to $\mathcal{P}$.
+
+One may wonder whether such an approach preserves the refutational
+completeness of the calculus. The answer is no.
+%
+\begin{conf}
+To see why, consider the following \emph{binary splitting} rule
+based on Riazanov and Voronkov \cite{riazanov-voronkov-2001}:
+%
+\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
+%
+Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
+$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
+smaller than the premise, the rule can be applied aggressively as a
+simplification. But notice that the effect of splitting can be undone by
+singular predicate elimination, possibly giving rise to loops
+$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
+This breaks completeness.
+
+Our solution is to curtail the entailment relation used by the redundancy
+criterion to disallow splitting-like simplifications. Weak entailment $\modelsWk$
+is defined via an ad hoc nonclassical logic so that $\{\cst{p} \cor C{,}\;
+\neglit{\cst{p}} \cor C\} \notmodelsWk \{C\}$ and yet $\modelsWk \{\cst{p} \cor
+\neglit{\cst{p}}\}$.
+%
+More precisely, this logic is defined via an
+encoding: $M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$, where
+%
+$\Wk{\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \false$,
+$\Wk{\lnot\,\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \true$, and
+$\Wk{L} = L$ otherwise.
+%
+Moreover, the type $\prop$ may be interpreted as any set of cardinality at least
+2, and $\false$ must be a distinguished symbol interpreted differently
+from~$\top$.
+
+The standard redundancy criterion $\RedWk$ based on $\modelsWk$ supports all the
+familiar deletion and simplification techniques except splitting. Using $\RedWk$
+not only prevents looping, but it also enables the use of
+the given clause procedure, because any redundant inference
+according to $\RedWk$ remains redundant after SPE or DPE. As usual, the devil is
+in the details, and the details are in the report \cite{our-report}.
+\end{conf}
+
+\begin{rep}
+To see why, consider the following \emph{binary splitting} rule
+based on Riazanov and Vo\-ro\-nkov~\cite{riazanov-voronkov-2001}:
+%
+\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
+%
+Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
+$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
+smaller than the premise, the rule can be applied aggressively as a
+simplification. But notice that the effect of splitting can be undone by
+singular predicate elimination, possibly giving rise to loops
+$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
+Clearly, we need to curtail predicate elimination.
+
+Under which conditions is predicate elimination refutationally complete? To
+answer this question, we employ the saturation framework of Waldmann, Tourret,
+Robillard, and Blanchette \cite{wtrb-20-sat-framework}. Let
+$(\FInf, \Red)$ be the base calculus without predicate elimination---e.g.,
+resolution or superposition inferences together with the standard
+redundancy criterion \cite[Sect.~4.2]{bg-01-resolution}. The
+inference system $\FInf$ is a set of inferences $(C_n,\ldots,C_1,C_0)$, for $n
+\ge 1$, where $C_n,\ldots,C_1$ are the premises and $C_0$ is the conclusion.
+$C_1$ is called the main premise. %; $C_n,\ldots,C_2$ are the side premises.
+The redundancy criterion is a pair $\Red = (\RedI, \RedF)$ where $\RedI$
+determines which inferences can be omitted and $\RedF$ is used to remove
+clauses.
+
+Next, consider an abstract proving process working on a single clause set.
+Let $\rhd_{\Red}$ denote the transition relation that supports (1)~adding
+arbitrary clauses and (2)~removing clauses deemed useless by $\RedF$.
+Typically, the added clauses are the result of performing inferences and are
+entailed by the premises, but other clauses can be added as well. A
+\emph{$\rhd_{\Red}$-derivation} is a finite or infinite sequence of clause sets
+$N_0 \rhd_{\Red} N_1 \rhd_{\Red} \cdots$.
+
+We fix a finite set $\mathbf{P}$ of predicate symbols that may be subjected to
+predicate elimination. These might include all the predicate symbols occurring
+in the input problem, but exclude any symbols introduced by splitting or other
+rules. Given a clause or clause set $N$, we write $\mathbf{P}(N)$ to denote the
+set of all predicate symbols from $\mathbf{P}$ occurring in $N.$
+%
+Let $\rhd_\mathbf{P}$ denote the elimination of a
+singular or defined predicate symbol from $\mathbf{P}$. A \emph{mixed
+derivation} consists of transitions either of the form $N
+\rhd_\mathbf{P} N'$ or of the form $N \rhd_{\Red} N'$ where
+$\mathbf{P}(N) \supseteq \mathbf{P}(N')$.
+Because $\mathbf{P}$ is finite, any mixed derivation consists of at most
+finitely many $\rhd_\mathbf{P}$ transitions. Hence, in any derivation,
+there exists an index~$k$ from which all transitions are standard
+${\rhd}_{\Red}$-transitions.
+
+This suggests the following path to completeness: Pretend that the transitions
+between $N_0$ and $N_k$ are merely preprocessing and start the actual
+derivation at $N_k$. This works at the abstract level of derivations on single
+clause sets. It fails, however, for an actual saturation prover that
+distinguishes between passive and active clauses.
+
+\begin{exa}
+\label{ex:cex-predelim-sat}
+The counterexample below is based on the given clause prover \textsf{GC}
+from the saturation framework. It shows how predicate elimination can break
+\textsf{GC}'s key invariant, which states that all inferences between active
+clauses are redundant. Breaking the invariant means that the limit might be
+unsaturated, breaking the refutational completeness proof.
+
+We use superposition with the order $\cst{a} \prec
+\cst{b} \prec \cst{c} \prec \cst{d}$ and without selection.
+Assume $\cst{a} \in \mathbf{P}$ and
+suppose we start with the satisfiable clause set %consisting of
+%
+%\begin{linenomath*}
+\begin{align*}
+  & \neglit{\cst{a}} \cor \MAX{\cst{d}}
+  && \neglit{\cst{a}} \cor \MAX{\neglit{\cst{d}}}
+  && \cst{a} \cor \cst{b} \cor \MAX{\cst{c}}
+  && \cst{c} \cor \MAX{\cst{d}}
+  && \cst{b} \cor \MAX{\neglit{\cst{d}}}
+\end{align*}
+%\end{linenomath*}
+%
+where gray boxes mark maximal (i.e., eligible) literals. Suppose the prover makes
+$\cst{c} \cor \cst{d}$ and $\cst{b} \cor \neglit{\cst{d}}$ active. From these
+two clauses, a superposition inference~$\iota$ could derive the conclusion
+$\cst{b} \cor \cst{c}$. However, the three passive clauses are
+$\prec$-smaller than $\iota$'s main premise $\cst{b} \cor \neglit{\cst{d}}$ and
+collectively entail $\iota$'s conclusion. This means that $\iota$ is
+redundant and can be ignored.
+
+If the prover now eliminates the predicate $\cst{a}$ using \infname{SPE}, the
+passive set is reduced to $\{\cst{b} \cor \cst{c} \cor \cst{d}{,}\;
+\cst{b} \cor \cst{c} \cor \neglit{\cst{d}}\}$.
+%
+Either clause is subsumed by an active clause, so the prover deletes it.
+It stops with the active set
+$\{\cst{c} \cor \cst{d}{,}\; \cst{b} \cor \neglit{\cst{d}}\}$, which is
+unsaturated because $\iota$ is no longer redundant. The invariant is broken.
+\end{exa}
+
+\begin{exa}
+\label{ex:cex-predelim-unsat}
+In Example~\ref{ex:cex-predelim-sat}, the initial clause set was satisfiable.
+If it is unsatisfiable, we can even lose refutational
+completeness. To see why, we add the unit clauses
+$\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ to the initial clause set of
+Example~\ref{ex:cex-predelim-sat} to make it unsatisfiable. We repeat the same
+steps as above, including the subsumptions at the end,
+yielding the passive set $\{\MAX{\neglit{\cst{b}}}, \MAX{\neglit{\cst{c}}}\}$ and
+the active set $\{\cst{c} \cor \MAX{\cst{d}}{,}\; \cst{b} \cor \MAX{\neglit{\cst{d}}}\}$. Then,
+making $\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ active triggers no
+inferences. The prover stops with an
+unsatisfiable four-clause active set that does not contain the empty clause.
+\end{exa}
+
+A solution could be to move all active clauses to the passive set at step~$k$
+or later, but this would be costly, since it would force the prover to redo
+inferences whose conclusions might then have to be simplified or subsumed again.
+Instead, we salvage the existing completeness proof for $\mathsf{GC}$, by
+resolving the issues concerning splitting and the $\mathsf{GC}$ invariant. Our
+approach is to weaken the redundancy criterion slightly, enough both to disable
+splitting on $\mathbf{P}$-predicates and to ensure that
+inferences such as $\iota$ in Examples
+\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} are performed. The
+required weakening is so mild that it does not invalidate any practical
+simplification or subsumption techniques we are aware of, except of course
+splitting.
+
+In accordance with the saturation framework, let $\mathbf{F}$ be the set of
+first-order $\mupSigma$-clauses, let $\mathbf{G}$ be its ground subset,
+and let $\gnd$ be a function that maps an $\mathbf{F}$-clause to the set of its
+$\mathbf{G}$-clause instances and analogously for $\mathbf{F}$-inferences. We
+define an extension $\mathbf{G}\Wksup$ of $\mathbf{G}$ for
+$\mupSigma\Wksup$-clauses in an ad hoc nonclassical logic reminiscent
+of paraconsistent logics \cite{carnielli-et-al-2007}. The objective is to
+disallow the entailment that makes splitting and Examples
+\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} possible.
+%
+The signature~$\mupSigma\Wksup$ extends $\mupSigma$ with a
+distinguished predicate symbol $\false$ that is interpreted differently
+from~$\top$. For $\mupSigma\Wksup$, the Boolean type $\prop$ may be
+interpreted as any set of cardinality at least 2.
+
+\begin{defi}
+The operator $\Wk{}$ translates $\mupSigma$-literals to
+$\mupSigma\Wksup$-literals as follows, where $\cst{p} \in \mathbf{P}$,
+$\cst{q} \notin \mathbf{P}$, and $s, t$ are non-Boolean terms:
+%
+%\begin{linenomath*}
+\begin{align*}
+  \Wk{\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \false
+& \Wk{\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \eq \true
+& \Wk{(s \eq t)} & = s \eq t \\[-\jot]
+  \Wk{\lnot\,\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \true
+& \Wk{\lnot\,\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \noteq \true
+& \Wk{(s \noteq t)} & = s \noteq t
+\end{align*}
+%\end{linenomath*}
+%
+The operator is lifted elementwise to $\mathbf{G}$-clauses and
+$\mathbf{G}$-clause sets.
+The \emph{weak entailment} $\modelsWk$ over $\mathbf{G}$-clause sets
+is defined via an encoding into $\mupSigma\Wksup$-clauses:
+$M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$.
+The lifting to $\mathbf{F}$-clauses and $\mathbf{F}$-clause set is achieved
+in the standard way via grounding.
+\end{defi}
+
+The following property of weak entailment will allow us to eliminate
+$\mathbf{P}$-predicates without losing completeness:
+
+\begin{lemma}
+\label{lem:p-Wk-entail}
+Let $C$ be a clause that contains the predicate symbol $\cst{p} \in
+\mathbf{P}$ and $D$ be a clause that does not contain $\cst{p}$. If $N
+\cup \{C\} \modelsWk \{D\}$, then $N \modelsWk \{D\}$.
+\end{lemma}
+
+\begin{proof}
+Suppose $\mathscr{J} \models \Wk{N}$. We will define $\mathscr{J}'$ so that
+$\mathscr{J}' \models \Wk{N} \cup \{\Wk{C}\}$, retrieve $\mathscr{J}'
+\models \Wk{D}$, and then argue that $\mathscr{J} \models \Wk{D}$. We
+take $\mathscr{J}'$ to coincide with $\mathscr{J}$ except that we extend the
+domain for $\prop$ with one fresh value and use this value as the
+interpretation of $\cst{p}(\tuple{t}{})$ for all argument tuples $\tuple{t}{}$.
+This modification makes any $\cst{p}$ literal of $\Wk{C}$ true, and it
+preserves the truth of $\Wk{N}$. By the hypothesis, $\mathscr{J}' \models
+\Wk{D}$. And since $\cst{p}$ does not occur in $D$, we have $\mathscr{J}
+\models \Wk{D}$.\qed
+\end{proof}
+
+Note that the above lemma does not hold for classical entailment $\models$;
+indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
+\{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
+entailment: $\modelsWk \cst{p} \cor \lnot\,\cst{p}$. In fact, all classical
+clausal tautologies are tautologies for $\modelsWk$.
+
+The standard redundancy criterion $\Red$ is obtained by lifting a criterion on
+$\mathbf{G}$-clauses to $\mathbf{F}$-clauses. The same construction can be
+replicated using $\modelsWk$ instead of $\models$, yielding the \emph{weak
+redundancy criterion} $\RedWk$.
+It is easy to check that the usual simplification techniques implemented in
+superposition provers can be justified using $\RedWk$. Specifically,
+this concerns the following rules described by Schulz \cite[Sects.\
+2.3.1~and~2.3.2]{ss-02-brainiac}:\
+deletion of duplicated literals, 
+deletion of resolved literals, 
+syntactic tautology deletion,
+semantic tautology deletion,
+rewriting of negative literals,
+positive simplify-reflect,
+negative simplify-reflect,
+clause subsumption, and
+equality subsumption.
+Moreover, rewriting of positive literals is possible if the
+rewriting clause is smaller than the rewritten
+clause (a condition that is also needed with $\models$ but omitted by Schulz).
+Finally, destructive equality resolution cannot be justified with $\models$,
+let alone $\modelsWk$.
+
+We instantiate the saturation framework with $(\FInf,\RedWk)$ to obtain a
+given clause prover \textsf{GC}. The prover operates on sets of labeled clauses
+$(C, l)$, where $C$ is a standard clause and $l \in \Lab$ is a label.
+The $\mathsf{active}$ label identifies active clauses; all other clauses are
+passive. The prover takes the form of two rules,
+\textsc{Process} and \textsc{Infer}, restricted to prevent the introduction of
+$\mathbf{P}$-predicates. We extend it with a third rule,
+\textsc{PredElim}, for predicate elimination, and call the extended prover
+\textsf{GCP}. The rules are as follows, using again the framework notations:
+
+\begin{description}[labelwidth=\widthof{\rm\textsc{PredElim}\,}]
+\itemsep1\jot
+\item[\rm\textsc{Process}\,]
+  $\mathcalx{N} \cup \mathcalx{M}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \mathcalx{M}'$
+\\
+  where $\mathcalx{M} \subseteq \LRedFWksub(\mathcalx{N} \cup \mathcalx{M}')$,
+  $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$,
+  and \\
+  $\mathbf{P}(\lfloor\mathcalx{M}'\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N} \cup \mathcalx{M}\rfloor)$
+
+\item[\rm\textsc{Infer}\,]
+  $\mathcalx{N} \cup \{(C, l)\}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \{(C, \ensuremath{\mathsf{active}})\} \cup \mathcalx{M}$
+\\
+  where $l \not= \ensuremath{\mathsf{active}}$,
+  $\mathcalx{M}{\downarrow}_\mathsf{active} = \emptyset$, \\
+  $\FInf(\lfloor\mathcalx{N}{\downarrow}_\mathsf{active}\rfloor, \{C\}) \subseteq
+    \RedIWk^{{\cap}\gnd}(\lfloor\mathcalx{N}\rfloor \cup \{C\} \cup \lfloor\mathcalx{M}\rfloor)$,
+  and \\
+  $\mathbf{P}(\lfloor\mathcalx{M}\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N}\rfloor \cup \{C\})$
+
+\item[\rm\textsc{PredElim}\,]
+  $\mathcalx{N} \cup \mathcalx{M}
+   \,\Longrightarrow_\mathsf{GCP}\,
+   \mathcalx{N} \cup \mathcalx{M}'$
+\\
+  where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P} \mathcalx{N} \cup \mathcalx{M}'$
+  and $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$
+\end{description}
+
+Here is a summary of the main framework notations:
+%
+\begin{itemize}
+\item The letters $\mathcalx{M}, \mathcalx{N}$ range over sets of labeled clauses.
+  $\mathcalx{M}{\downarrow}_l$ denotes the subset of clauses in $\mathcalx{M}$
+  labeled with $l$. The operator $\lfloor\phantom{\cdot}\rfloor$ erases all
+  labels in a labeled clause or clause set.
+
+\smallskip
+
+\item $\FInf(N)$ denotes the set of all base calculus inferences with premises
+  in $N$, and $\FInf(N, M) = \FInf(N \cup M) \setminus \FInf(N \setminus M)$.
+  The same notations are also available for the straightforward extension $\FLInf$
+  of $\FInf$ with labels.
+
+\smallskip
+
+\item $\LRedWksub$ is the extension of the standard redundancy
+  criterion $\RedWk$ defined using $\modelsWk$ to nonground labeled clauses
+  with subsumption ($\sqsubset$).
+
+\smallskip
+
+\item Given a sequence $(\mathcalx{N}{}_i)_i$, its \emph{limit} (\emph{inferior})
+  is $\mathcalx{N}{}_\infty = \bigcup_i \bigcap_{j\ge i} \mathcalx{N}{}_j$.
+\end{itemize}
+
+\begin{sloppypar}
+   The completeness proof follows the invariance-based argument found in the
+   article by Waldmann et al \cite{waldmann-et-al-202x-article}.
+\end{sloppypar}
+
+\begin{lemma}
+\label{lem:gcp-derivations-are-red-black-derivations}
+Every $\Longrightarrow_\mathsf{GCP}$-derivation is a mixed derivation.
+\end{lemma}
+
+\begin{proof}
+The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
+al., with adjustments to show that $\cst{P}$-predicates cannot reappear once
+they have disappeared. The case for \textsc{ElimPred} is trivial.
+\end{proof}
+
+Let $\Inv\Wksup_\mathcalx{N}(k)$ denote the condition
+$\textstyle\FLInf(\mathcalx{N}{}_k{\downarrow}_\mathsf{active}) \subseteq
+\LRedIWk(\mathcalx{N}{}_k)$.
+Notice the difference with the definition of the key invariant
+$\Inv_\mathcalx{N}$ in the saturation framework, whose right-hand side is
+$\bigcup_{i=0}^k \RedI^{\Lab}(\mathcalx{N}{}_i)$. We cannot use the
+big union ${\bigcup}$ starting at $i = 0$ because we will need to truncate a
+sequence prefix of an a priori unknown length. The argument will still work
+thanks to monotonicity properties of redundancy criteria.
+
+\begin{lemma}
+\label{lem:gcp-invar-gcp}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
+If
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$,
+then
+$\Inv\Wksup_\mathcalx{N}(k)$ holds for all indices~$k$.
+\end{lemma}
+
+\begin{proof}
+The base case is as in Waldmann et al.
+
+For \textsc{Process} and \textsc{Infer}, the proof is essentially as in Waldmann
+et al., except that we also need to show that
+$\LRedIWk(\mathcalx{N}{}_k) \subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
+This is a consequence of
+$\mathcalx{N}{}_k \rhd_{\LRedWksub} \mathcalx{N}{}_{k+1}$ and of
+properties (R2) and (R3) of redundancy criteria.
+
+\begin{sloppypar}
+
+   A new case to consider is that of a \textsc{PredElim} transition
+   $\mathcalx{N}{}_{k} \Longrightarrow_\mathsf{GCP} \mathcalx{N}{}_{k+1}$.
+   Let $\mathcalx{N}{}_{k} = \mathcalx{N} \cup \mathcalx{M}
+   \Longrightarrow_\mathsf{GCP} \mathcalx{N} \cup \mathcalx{M}' = \mathcalx{N}{}_{k+1}$,
+   where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P}
+   \mathcalx{N} \cap \mathcalx{M}'$ and
+   $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
+   We assume without loss of generality that $\mathcalx{M} \cap \mathcalx{M}' =
+   \emptyset$.
+   Let $\cst{p}$ be the eliminated predicate. Note that $\cst{p}$ occurs in every
+   clause in $\mathcalx{M}$ but in none of the clauses in $\mathcalx{N}$ or
+   $\mathcalx{M}'$.
+   We must show $\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
+   \subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
+   As for \textsc{Process}, we have the inclusion
+   $\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
+   \subseteq \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$,
+   by the side condition that
+   $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
+   Moreover, by the induction hypothesis,
+   $\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
+   \subseteq \LRedIWk(\mathcalx{N}{}_k)$. Thus,
+   $\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
+   \subseteq \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
+\end{sloppypar}
+
+Let $\iota \in \FLInf(\mathcalx{N} {\downarrow}_\mathsf{active})$.
+By the argument above, we have
+$\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
+We must
+show $\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.
+%
+By definition of $\LRedIWk$, it suffices to show that for
+every ground instance $(C_n,\ldots,C_1,C_0)$ of $\iota$, there exists a finite
+clause set $\mathcalx{D} \subseteq \gnd(\mathcalx{N}) \cup \gnd(\mathcalx{M}')$
+that is $\prec$-smaller than $C_1$ and such that
+$\{C_n,\ldots,C_2\} \cup \mathcalx{D} \modelsWk \{C_0\}$.
+Without loss of generality, we assume that $\mathcalx{D}$ is the smallest such
+set.
+
+By definition of $\mathcalx{N}$, $\cst{p}$ cannot occur in $C_0$.
+By Lemma~\ref{lem:p-Wk-entail}, if $\cst{p}$ occurred in $D \in \mathcalx{D}$,
+we could remove $D$, but this would mean $\mathcalx{D}$ is not minimal.
+As a result, $\mathcalx{D}$ cannot contain clauses from $\gnd(\mathcalx{M})$ and
+hence $\mathcalx{D} \subseteq \gnd(\mathcalx{N})$. Thus,
+$\iota \in \LRedIWk(\mathcalx{N})$. By property (R2) of
+redundancy criteria, we have the desired result: $\iota \in
+\LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.
+\end{proof}
+
+\begin{lemma}
+\label{lem:gcp-invar-infinity}
+Let $(\mathcalx{N}{}_i)_i$ be a
+$\rhd_{\LRedWksub}$-derivation.
+If $\Inv\Wksup_\mathcalx{N}(i)$ holds for all indices~$i$, then
+$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
+\bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
+holds.
+\end{lemma}
+
+\begin{proof}
+We assume $\iota \in \FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$
+and show $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
+for some arbitrary $\iota$.
+%
+For $\iota$ to belong to $\FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$, all of
+its finitely many premises must be in $\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active}$.
+Therefore, there must exist an index $k$ such that
+$\mathcalx{N}{}_k {\downarrow}_\mathsf{active}$ contains all of them, and therefore
+$\iota \in \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$. Since $\Inv_\mathcalx{N}(k)$
+holds, $\iota \in \LRedIWk(\mathcalx{N}{}_k)$.
+Hence, $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$.
+\end{proof}
+
+\begin{lemma}
+\label{lem:fair-gcp-derivations}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
+If
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
+and
+$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
+$l \not= \ensuremath{\mathsf{active}}$,
+then there exists an index $k$ such that
+$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
+\end{lemma}
+
+\begin{sloppypar}
+\begin{proof}
+By Lemma~\ref{lem:gcp-derivations-are-red-black-derivations}, there must exist
+an index $k$ such that the sequence $(\mathcalx{N}{}_{k+i})_i$ is a pure
+$\rhd_{\LRedWksub}$-derivation.
+By Lemma~\ref{lem:gcp-invar-gcp},
+$\Inv\Wksup_\mathcalx{N}(k+i)$ holds for all indices~$i$.
+Hence, by Lemma~\ref{lem:gcp-invar-infinity},
+$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
+\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.
+By the second hypothesis, this inclusion simplifies to
+$\FLInf(\mathcalx{N}{}_\infty) \subseteq
+\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.
+\end{proof}
+\end{sloppypar}
+
+\begin{theorem}
+\label{thm:gcp-complete}
+Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation
+with
+$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
+and
+$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
+$l \not= \ensuremath{\mathsf{active}}$.
+If $\lfloor \mathcalx{N}{}_0\rfloor$ is unsatisfiable,
+then
+some $\mathcalx{N}{}_i$
+contains the empty clause with some arbitrary label.
+\end{theorem}
+
+\begin{proof}
+By Lemma~\ref{lem:fair-gcp-derivations},
+we know that there exists an index $k$ such that
+$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
+Moreover, since
+$\rhd_{\LRedWksub}$ and
+$\rhd_\mathbf{P}$ preserve unsatisfiability
+(by (R1) of redundancy criteria, Khasidashvili and Korovin's Theorem~1, and our
+Theorem~\ref{thm:pes-sat-equiv}), we have that
+$\lfloor \mathcalx{N}{}_k\rfloor$ is unsatisfiable.
+Since the base calculus $(\FLInf,\allowbreak\LRed)$ is assumed to be statically
+refutationally complete with respect to $\models$, the calculus
+$(\FLInf,\allowbreak\LRedWk)$ with a weaker redundancy criterion is also
+statically complete with respect to $\models$, and by the saturation framework,
+$(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
+static and dynamic completeness, we conclude that some $\mathcalx{N}{}_{k+i}$ must
+contain a labeled empty clause.
+\end{proof}
+\end{rep}

### 2022-03-17 ho-techniques/ho-techniques.tex
more

@@
 
-We focus on monomorphic higher-order logic, without the axiom of infinity or the
-axiom of at least two individuals \cite{bm-14-automation-ho}. However, the
-techniques can easily be extended with polymorphism. Indeed, Zipperposition
+We focus on monomorphic higher-order logic, defined in Sect.~\ref{sec:pre:hol}. However, the
+techniques can easily be extended with rank-1 polymorphism \cite{ksr-16-th1}. Indeed, Zipperposition
 already supports some of them polymorphically. Further, we use exactly the same
-notation for this logic as introduced in Chapter \ref{ch:pre}. Since we are
+notation for this logic and superposition calculus as introduced in Chapter \ref{ch:pre}. Since we are
 working with extensions of superposition, we assume a clausal structure
@@
 
+\looseness=-1
 To support such streams in the given clause procedure, we extend it to
@@
 Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
-Each stream is associated with a weight, and $\mathcal{Q}$ is sorted in order of
+Each stream is given a weight, and $\mathcal{Q}$ is sorted in order of
 increasing weight\confrep{}{, a low weight corresponding to a high priority}.
@@
 
-  \newpage
+  \pagebreak[2]
   \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}

### 2022-03-17 pre/pre.tex
more

@@
 
-In this thesis we use classic simply typed monomorphic higher-order logic with
-the choice operator (and some variations of this logic). Assuming a set of
-base types $T$ and a set of symbols $\Sigma$, let us
-define terms and types. Types are either base types $\tau \in T$, or function
-types $\tau \rightarrow \upsilon$ where both $\tau$ and $\upsilon$ are types.
-When we write $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$
-we assume $\upsilon$ to be a base type ($\upsilon \in T$), by convention.
-Each symbol $\cst{f} \in \Sigma$ is assigned a type.
+In this thesis we use classic extensional simply typed monomorphic higher-order
+logic with the choice operator, but without the description operator, the axiom of infinity,
+and the axiom of at least two individuals \cite[Sect.~3.5]{bm-14-automation-ho}.
+This logic corresponds to Henkin's extensional type theory. We additionally use
+some variations of this logic, clearly stating the differences. Assuming a set of
+base types $T$ and a set of symbols $\Sigma$, let us define terms and types.
+Types are either base types $\tau \in T$, or function types $\tau \rightarrow
+\upsilon$ where both $\tau$ and $\upsilon$ are types. When we write $\tau_1
+\rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$ we assume $\upsilon$
+to be a base type ($\upsilon \in T$), by convention. Each symbol $\cst{f} \in
+\Sigma$ is assigned a type.
 

### 2022-03-17 satfol/satfol.tex
more

@@
 \Wk{D}$. And since $\cst{p}$ does not occur in $D$, we have $\mathscr{J}
-\models \Wk{D}$.\qed
+\models \Wk{D}$.
 \end{proof}
@@
 \end{rep}
+
+\section{Satisfiability by Clause Elimination}
+\label{sec:satfol:satisfiability-by-clause-elimination}
+
+\looseness=-1
+The main approaches to show satisfiability of a first-order problem are to
+produce either a finite Herbrand model or a saturated clause set. Saturations
+rarely occur except for very small problems or within decidable fragments. In
+this section, we explore an alternative approach that establishes satisfiability
+by iteratively removing clauses while preserving
+%satisfiability and
+unsatisfiability,
+until the clause set has been transformed into the empty set. So
+far, this technique has been studied only for QBF~\cite{hsb-14-QRAT}. We show that
+\emph{blocked clause elimination} (BCE)
+% for first-order logic
+can be used for this purpose. It can efficiently solve some problems for which
+the saturated set would be infinite. However, it can break the refutational
+completeness of a saturation prover. We conclude with a procedure that
+transforms a finite Herbrand model into a sequence of clause elimination steps
+ending in the empty clause set, thereby demonstrating the theoretical power of
+clause elimination.
+
+Kiesl et al.\ \cite{ksstb-2017-blockedfol} generalized blocked clause
+elimination to first-order logic. Their generalization uses flat
+$L$-resolvents, an extension of flat resolvents that resolves a single
+literal~$L$ against $m$~literals of the other clause.
+
+\begin{defi}
+   Let $C = L \cor C'$ and $D = L_1 \cor \cdots \cor L_m \cor D'$, where
+   \begin{enumerate}
+      \item $m \ge\nobreak 1$
+      \item the literals $L_i$ are of opposite polarity to $L$
+      \item $L$'s atom is $\cst{p}(\tuplen{s})$
+      \item $L_i$'s atom is $\cst{p}(\tuple{t_i}{})$ for each~$i$
+      \item $C$ and $D$ have no variables in common
+   \end{enumerate}
+   The clause
+   $\bigl(\bigvee_{i=1}^m \bigvee_{j=1}^n
+     \neqlit{s_j}{t_{ij}}\bigr) \cor C' \cor D'$
+   is a \emph{flat $L$-resolvent} of $C$~and~$D$.
+\end{defi}
+
+\begin{defi}
+   \label{def:equality-blocked}
+   \begin{conf}%
+      A clause $C = L \cor C'$ is (\emph{equality-})\emph{blocked} by $L$ in a clause
+      set $N$ if all flat $L$-resolvents between $C$ and clauses in $N \setminus \{C\}$ are
+      tautologies.
+   \end{conf}%
+   \begin{rep}%
+      Let $C = L \cor C'$ be a clause and $N$ be a clause set.
+      Let $N'$ consist of all clauses from $N \setminus \{C\}$ with their variables
+      renamed so that they share no variables with $C$.
+      The clause $C$ is (\emph{equality-})\emph{blocked} by $L$ in the set $N$ if
+      all flat $L$-resolvents between $C$ and clauses in $N'$ are tautologies.
+   \end{rep}
+\end{defi}
+
+Removing a blocked clause from a set preserves unsatisfiability
+\cite{ksstb-2017-blockedfol}. Kiesl et al.\ evaluated the effect of removing all
+blocked clauses as a preprocessing step and found that it increases
+\confrep{}{the }prover's success rate.
+
+In fact, there exist satisfiable problems that cannot be saturated in finitely
+many steps regardless of the calculus's parameters but that can be reduced to
+an empty, vacuously satisfiable problem through blocked clause elimination.
+
+\begin{exa}
+   \label{example:infinite-blocked-set}
+   \looseness=-1
+   Consider the clause set $N$ consisting of $C = \pospredlit{\cst{p}(x,x)}$ and
+   $D = \negpredlit{\cst{p}(y_1,y_3)} \cor \pospredlit{\cst{p}(y_1,y_2)} \cor
+   \pospredlit{\cst{p}(y_2,y_3)}$. Note that if no literal is selected, all
+   literals can take part in superposition inferences (i.e., they are eligible), as literals in
+   clause $D$ are $\succ$-uncomparable. In particular, the superposition of
+   $\pospredlit{\cst{p}(x,x)}$ into $D$'s negative literal eventually needs to
+   be performed regardless of the chosen selection function or term order, with
+   the conclusion $E_1 = \pospredlit{\cst{p}(z_1,z_2)} \cor
+   \pospredlit{\cst{p}(z_2,z_1)}$. Then, superposition of $E_1$ into $D$ yields
+   $E_2 = \pospredlit{\cst{p}(z_1,z_2)} \cor \pospredlit{\cst{p}(z_2,z_3)} \cor
+   \pospredlit{\cst{p}(z_3,z_1)}$. Repeating this process yields infinitely many
+   clauses $E_i = \pospredlit{\cst{p}(z_1,z_2)} \cor \cdots \cor
+   \pospredlit{\cst{p}(z_i,z_{i+1})} \cor \pospredlit{\cst{p}(z_{i+1},z_1)}$
+   that cannot be eliminated using standard redu\-ndancy-based techniques.
+\end{exa}
+
+\looseness=-1
+In the example above, the clause $D$ is blocked by its second or third
+literal. If we delete $D$, $C$ becomes blocked in turn. Deleting
+$C$ leaves us with the empty set, which is vacuously satisfiable. The example
+suggests that using BCE during saturation might help focus the proof search.
+Indeed, Kiesl et al.\ ended their investigations by asking whether BCE can be
+used as an inprocessing technique in a saturation prover. Unfortunately,
+in general the answer is no\confrep{.}{:}
+
+\begin{exa}
+   \label{example:bce-completeness}
+   Consider the unsatisfiable set $N=\{ C_1, \ldots,\allowbreak C_6 \}$, where
+%   \begin{linenomath*}
+   \begin{align*}
+      C_1 &= \negpredlit{\cst{c}} \cor \pospredlit{\cst{e}} \cor \MAX{\negpredlit{\cst{a}}} 
+      & C_2 &= \negpredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{e}}}
+      & C_3 &= \pospredlit{\cst{b}} \cor \MAX{\pospredlit{\cst{c}}}
+      \\
+      C_4 &= \negpredlit{\cst{b}} \cor \MAX{\negpredlit{\cst{c}}}
+      & C_5 &= \pospredlit{\cst{a}} \cor \MAX{\pospredlit{\cst{b}}} 
+      & C_6 &= \pospredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{b}}} 
+         &
+   \end{align*}
+%   \end{linenomath*}
+   Assume the simplification ordering $\cst{a} \prec \cst{b} \prec \cst{c} \prec
+   \cst{d} \prec \cst{e}$ and the selection function that chooses the last negative literal of a clause as
+   presented. Gray boxes indicate literals that can take part in superposition
+   inferences. Only two superposition inferences are
+   possible:\ from $C_3$ into $C_4$, yielding the tautology $C_{7} = \pospredlit{\cst{b}} \cor
+   \MAX{\negpredlit{\cst{b}}}$, and from $C_5$ into $C_6$, yielding $C_{8} = \pospredlit{\cst{a}} \cor
+   \MAX{\pospredlit{\cst{c}}}$. Clause $C_{7}$ is
+   clearly redundant, whereas $C_{8}$ is blocked by its first literal. If we
+   allow removing blocked clauses, the prover enters a loop: $C_{8}$ is repeatedly generated and 
+   deleted. Thus, the prover will never generate the
+   empty clause for this unsatisfiable set.
+\end{exa}
+
+   \newcommand{\aset}{\mathcal{A}}
+   \newcommand{\pset}{\mathcal{P}}
+   As with hidden tautologies, removing blocked clauses breaks the invariant of
+   the given clause procedure that all inferences between clauses in $\aset$ are
+   redundant. To see this, assume the setting of Example
+   \ref{example:bce-completeness}, and let $\pset = N$ and $\aset = \emptyset$.
+   Assume $C_1, C_2, C_3$ are moved to the active set. As there are
+   no possible inferences between them, the proof state becomes $\aset = \{C_1,
+   C_2, C_3\}$ and $\pset = \{C_4,C_5,C_6\}$. After $C_4$ is moved to
+   $\aset$, the conclusion $C_7$ is computed, but it is not added to $\pset$ as it
+   is redundant. Moving $C_5$ to $\aset$ produces no new conclusions, but after
+   $C_6$ is moved, $C_{8}$ is produced. However, if we allow eliminating blocked
+   clauses, it will not be added to $\pset$ as it is blocked. The prover then
+   terminates with $\aset=N$ and $\pset=\emptyset$, even though the original set
+   $N$ is unsatisfiable.
+
+Although using BCE as inprocessing breaks the completeness of superposition in
+general, it is conceivable that a well-behaved fragment of BCE might exist.
+This could be investigated further.
+
+Not only can BCE prevent infinite saturation (Example \ref{example:infinite-blocked-set}),
+but it can also be used to convert a finite Herbrand model
+into a certificate of clause set satisfiability. The certificate uses only
+blocked clause elimination and addition, in conjunction with a transformation
+to reduce the clause set to an empty set. This theoretical result
+explores the relationship between Herbrand models and satisfiability certificates
+based on clause elimination and addition. It is conceivable that it can form
+the basis of an efficient way to certify Herbrand models.
+
+In propositional logic, \emph{asymmetric literals} can be added to
+or removed from clauses, retaining the equivalence of the resulting clause set with
+the original one. Kiesl and Suda~\cite{ks-2017-unif-principle} described an extension of
+this technique to first-order logic.
+Their definition of asymmetric literals can be relaxed to
+allow the addition of more literals, but the resulting set is then only
+equisatisfiable to the original one, not equivalent.
+This in turn allows us to show that a problem is satisfiable by
+reducing it to an empty problem, as is done in some SAT solvers.
+
+For the rest of this section, we work with clausal first-order logic
+\relax{without equality}. We use \relax{Herbrand models} as canonical
+representatives of first-order models, recalling that every
+satisfiable set has a Herbrand model \cite[Sect.~5.4]{mf-1996-fol}.
+
+\begin{defi}
+   \looseness=-1
+   A literal $L$ is a \emph{global asymmetric literal} (GAL) for a clause $C$
+   and a clause set $N$ if for every ground instance $\sigmacl{C}$ of $C$, there
+   exists a ground instance $\substterm{\varrho}{D} \cor \substterm{\varrho}{L'}$ of $D \cor L' \in N \setminus \{C\}$ such that $\substterm{\varrho}{D}
+   \subseteq \substterm{\sigma}{C}$ and $\substterm{\varrho}{\neglit L'} = \sigmaterm{L}$.
+ \end{defi}
+\confrep{}{}
+
+Every asymmetric literal is GAL, but the converse does not hold:
+
+\begin{exa}
+   \looseness=-1
+   Consider a clause $C = \pospredlit{\cst{p}(x,y)}$ and a clause set $N = \{\pospredlit{\cst{q}} \cor
+  \pospredlit{\cst{p}(\cst{a},\cst{a})}\}$. Then, $\negpredlit{\cst{q}}$ is not an asymmetric literal
+  for $C$ and $N$, but it is a GAL for $C$ and $N$.
+\end{exa}
+
+Adding and removing GALs preserves and reflects satisfiability:
+
+\begin{theorem}
+   \label{thm:gal}
+   If $L$ is a GAL for the clause $C$ and the clause set $N$, then the set $(N \setminus
+   \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
+\end{theorem}
+\begin{rep}%
+\begin{proof}
+   \looseness=-1
+   Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
+   Non-trivial direction is proving that if $N'$ has a model, so does
+   $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
+   $\ourmodel$ satisfies every clause in $N$, with the possible exception of
+   $C$. Assume that there exists a grounding substitution $\sigma$ such that ground instance $\sigmaterm{C}$ is falsified by
+   $\ourmodel$. Since $L$ is a GAL for $C$ and $N$, then there exists a clause $D \cor L' \in N'$, and a grounding substitution $\tau$ such that
+   $\substcl{\tau}{D} \subseteq \sigmacl{C}$ and $\substcl{\tau}{\neg L'} =
+   \sigmaterm{L}$. If $\ourmodel \models
+   \substterm{\tau}{L'}$, for $\sigmaterm{C} \cor \sigmaterm{L}$ to be satisfied by $\ourmodel$ (as $C \cor L \in N'$), some
+   literal in $\sigmaterm{C}$ must be satisfied by $\ourmodel$, contradicting that
+   $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$
+   then some literal in $\substterm{\tau}{D}$ must be satisfied. As $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
+   we again get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
+   as needed.
+ \end{proof}
+
+\end{rep}
+\looseness=-1
+For first-order logic without equality, a clause $L \cor C$ is blocked if all its
+$L$-resolvents are tautologies \cite{ksstb-2017-blockedfol}. The $L$-resolvent
+between $L \cor C$ and $\neglit{L_1} \cor \cdots \cor \neglit{L_n} \cor
+D$ (which are renamed apart) is $\sigmacl{C \cor D}$, where $\sigma$ is the most general unifier
+of the literals $L, L_1, \ldots, L_n$.
+Given a Herbrand model $\ourmodel$ of a problem, the following procedure
+removes all clauses while preserving satisfiability:
+
+\begin{enumerate}
+   \item \label{item:step1}
+   \looseness=-1
+   Let $\cst{q}$ be a fresh predicate symbol. For each atom
+   $\cst{p}(\tupleempty{s})$ in the Herbrand universe: If
+   $\ourmodel \models \cst{p}(\tupleempty{s})$, add the clause $\cst{q} \cor
+   \cst{p}(\tupleempty{s})$; otherwise, add %the clause
+   $\cst{q} \cor \neglit{\cst{p}(\tupleempty{s})}$. Adding either clause
+   preserves satisfiability as both are blocked by~$\cst{q}$.
+% \pagebreak[2]
+   \smallskip
+   \item \label{item:step2}
+   Since $\ourmodel$ is a model, for each ground instance $\sigmacl{C}$, there exists a clause
+   $\cst{q} \cor L$ with $L \in \sigmacl{C}$. We can transform $C \in N$ into
+   $C \cor \neglit{\cst{q}}$, since $\neglit{\cst{q}}$ is a GAL for $C$ and $N$.
+
+   \smallskip
+   \item \label{item:step3}
+   \looseness=-1
+   Consider the clause $\cst{q} \cor L$ added by step \ref{item:step1}. Since
+   $L$ is ground and no clause $\cst{q} \cor \neglit{L}$ was added (since $\ourmodel$ is a
+   model), the only $L$-resolvents are against clauses added by step~\ref{item:step2}.
+   Since all of those clauses contain $\neglit{\cst{q}}$, the resolvents are tautologies. 
+   Thus, each $\cst{q} \cor L$ is blocked and can be removed in turn.
+
+   \smallskip
+   \item The remaining clauses all contain the literal $\neglit{\cst{q}}$. They can be removed
+   by BCE as well.
+\end{enumerate}
+
+The procedure is limited to the first-order logic without equality, since step
+\ref{item:step3} is justified only if $L$ is a predicate literal. (Otherwise,
+$L$ cannot block clause $\cst{q} \cor L$ \cite{ksstb-2017-blockedfol}.) The
+procedure also terminates only for finite Herbrand models.
+
+% \begin{rep}
+
+\begin{exa}
+   Consider the satisfiable clause set $N = \{\pospredlit{\cst{r}(x)} \cor
+   \pospredlit{\cst{s}(x)}{,}\; \negpredlit{\cst{r}(\cst{a})}{,}\allowbreak\;
+   \negpredlit{\cst{s}(\cst{b})}\}$ and a Herbrand model $\ourmodel$ over
+   $\{\cst{a}, \cst{b}, \cst{r}, \cst{s}\}$ such that
+   $\cst{r}(\cst{b})$ and $\cst{s}(\cst{a})$ are the only true atoms in
+   $\ourmodel$. We show how to remove all clauses in $N$ using $\ourmodel$ 
+   by following the procedure above.
+
+   Let $N_{\ourmodel} = \{ \cst{q} \cor \negpredlit{\cst{r}(\cst{a})}{,}\; \cst{q}
+   \cor \pospredlit{\cst{r}(\cst{b})}{,}\; \cst{q} \cor
+   \pospredlit{\cst{s}(\cst{a})}{,}\; \cst{q} \cor \negpredlit{\cst{s}(\cst{b})}
+   \}$. We set $N \gets N \cup N_{\ourmodel}$. This preserves satisfiability
+   since all clauses in $N_{\ourmodel}$ are blocked. It is easy to check that
+   $\negpredlit{\cst{q}}$ is GAL for every clause in $N \setminus N_{\ourmodel}$. The
+   only substitutions that need to be considered are $\{x \mapsto \cst{a} \}$ and $\{x \mapsto
+   \cst{b} \}$ for $\pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}$.
+   So we set $N \gets \{\negpredlit{\cst{q}} \cor \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
+   \negpredlit{\cst{q}} \cor \negpredlit{\cst{r}(\cst{a})}{,}\;
+   \negpredlit{\cst{q}} \cor \negpredlit{\cst{s}(\cst{b})}\} \cup N_{\ourmodel}$.
+   Clearly, all clauses in $N_{\ourmodel}$ are blocked, so we set $N \gets N \setminus
+   N_{\ourmodel}$. All clauses remaining in $N$ have a literal $\negpredlit{\cst{q}}$
+   and can be removed, leaving $N$ empty as desired.
+\end{exa}
+
+\section{Implementation}
+\label{sec:satfol:implementation}
+% {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}
+
+\newcommand{\impmap}{\textit{Imp}}
+
+Hidden-literal-based, predicate, \begin{qle}quasipure, \end{qle}and blocked clause
+elimination all admit efficient implementations in a superposition
+prover. In this section, we describe how to implement the first
+\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
+techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
+All techniques are implemented in the Zipperposition prover
+\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
+of improvements to superposition, but it implements many of the most successful
+heuristics from the E prover \cite{scv-19-e23} and has recently become
+quite competitive \cite{gs-19-casc27}.
+
+\ourpara{Hidden-Literal-Based Elimination}
+% 
+% Heule et al.\ \cite{hjb-2010-cl-elim} show how hidden tautologies can be removed
+% without creating the binary implication graph. For each literal~$L$ occurring in
+% the clause set $N$, they compute the set $\PHL(L,N)$. Then for each $\neglit{L}'
+% \in \PHL(L,N)$, they label $L'$ with $L$. A clause $C$ is a hidden tautology if
+% there are two literals $L,L' \in C$ such that $L'$ is labeled with $L$. The
+% described approach can be generalized to first-order logic and used to implement
+% all rules from Sect.~\ref{sec:satfol:hidden-literal-based-elimination}.
+% 
+For HLBE, an efficient representation of $\HL(L,N)$ is crucial. Because
+this set may be infinite, we underapproximate it by
+restricting the length of the
+transitive chains via a parameter $K_\mathrm{len}$. Given the current
+clause set $N$, the finite map $\impmap[L']$ associates with each literal $L'$
+a set of pairs $(L,M)$ such that $L' \bigfimpl{N}^k L$, where $k \leq
+K_\mathrm{len}$ and $M$ is the multiset of clauses used to derive $L' \bigfimpl{N}^k L$.
+Moreover, we consider only transitions of type 1 (as per
+Definition~\ref{def:hl-fo}).
+%Intuitively,
+%$\impmap[L']$ contains all literals which can be reached from $L'$ following $k$
+%edges of the generalized binary implication graph.
+The following algorithm
+maintains $\impmap$ dynamically, updating it as the prover derives and deletes
+clauses.
+It depends on the global variable $\impmap$ and the parameters
+$K_\mathrm{len}$ and $K_{\mathrm{imp}}$.
+
+% By choosing a limit $K_\mathrm{len}$ we can create a map $\impmap[L]$, which maps a literal $L$
+% to a finite a set of pairs $(L', M)$ such that
+% $\impk{L}{L'}{i}{M}$ where $i \leq K_\mathrm{len}$. It is important to record clauses used
+% in deriving the implication, as this information is necessary to construct the
+% proof. Mapping $\impmap$ is constructed so that whenever $(L', M) \in
+% \impmap[L]$, then $L \in \HL(L',M)$.
+% Proof by very simple induction
+
+\looseness=-1
+
+\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
+\algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}
+\algrenewcommand\algorithmicindent{0.75em}
+
+%%% TYPESETTING: hack
+% \vskip\abovedisplayskip
+% \kern-.5\jot
+
+%% \begin{nolinenumbers}%\begin{figure}[h!]
+\begin{algorithmic}[5]
+
+%   \Procedure{DecomposeNegLit}{$c$, $C$}
+%   \State $\vn{res} \gets \{(c, C)\}$
+%   \ForAll{non-empty contexts $u$ such that $c = \neqlit{u[s]}{u[t]}$}
+%       \State $\vn{res} \gets \{ (\neqlit{s}{t}, C) \} \cup \vn{res}$
+%   \EndFor
+%   \State \Return $\vn{res}$
+%   \EndProcedure
+
+%  \kern1\jot
+
+  \Procedure{AddImplication}{$L_\mathrm{a}{,}\> L_\mathrm{c}{,}\> C$}
+    \If{$\impmap[\sigmaterm{L_\mathrm{a}}] \not = \emptyset$ for some renaming $\sigma$} \label{statement:alpha}
+      \State $(L_\mathrm{a}, L_\mathrm{c}) \gets (\sigmaterm(L_\mathrm{a}), \sigmaterm{L_\mathrm{c}})$
+    \EndIf
+
+   %  \vspace{0.1em}
+    \If{there are no $L,L',M,\sigma$ such that $(L',M) \in \impmap[L]$,\label{statement:more-general}
+      %   \\\noindent\kern3.5ex 
+        $\sigmaterm{L} = L_\mathrm{a}$, and $\sigmaterm{L'}=L_\mathrm{c}$} 
+      % \State $\vn{Inst} \gets \{(\sigma, M) \mid (L_\mathrm{c}\sigma, M) \in \impmap[L_\mathrm{a}\sigma]\}$
+      \ForAll{$(\sigma, M)$ such that $(\sigmaterm{L_\mathrm{c}}, M) \in \impmap[\sigmaterm{L_\mathrm{a}}]$} \label{statement:instances}
+        \State {erase all $(L',M')$ such that $M \subseteq M'$ from $\impmap[\sigmaterm{L_{\mathrm{a}}}]$}
+      \EndFor
+
+      \ForAll{$L$ such that $(L', M) \in \impmap[L]$ and $\sigmaterm{L_\mathrm{a}} = L'$ for some $\sigma$} \label{statement:transitivity}
+        \If{$|M| < K_\mathrm{len}$}
+          \State $\impmap[L] \gets \impmap[L] \cup \{(\sigmaterm{L_\mathrm{c}}, M \uplus \{C\})\} $
+        \EndIf
+      \EndFor
+
+      \ForAll{$L$ such that $\impmap[L] \not= \emptyset$ and $\sigmaterm{L} = L_\mathrm{c}$ for some $\sigma$} \label{statement:triggered}
+         % \If{$|M| < K_\mathrm{len}-1$}
+         \State $\vn{Concl} \gets \{(\sigmaterm{L'}, M \uplus \{\mathit{C}\}) \mid (L',M) \in \impmap[L], |M| < K_\mathrm{len}  \}$
+         \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup  \vn{Concl}$
+         % (L'\sigma, M \cup \{\mathit{C}\}) $
+         % \EndIf
+      \EndFor
+
+      \State $\vn{Congr} \gets \{
+         (\neqlit{s}{t}, \{C\}) \mid \exists u.\, L_\mathrm{c} = \neqlit{u[s]}{u[t]}
+         \}$
+      \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup \{ (L_\mathrm{c},\{C\}) \} \cup \vn{Congr}$ \label{statement:decomposition}
+
+    \EndIf
+
+  \EndProcedure
+
+  \kern1\jot
+%   \pagebreak[2]
+   \newpage
+
+  \Procedure{TrackClause}{$C$}
+    \If{$C$ = $L_1 \cor L_2$}
+      \State \textsc{AddImplication}($\neglit{L_1}$, $L_2$, $C$)
+      \State \textsc{AddImplication}($\neglit{L_2}$, $L_1$, $C$)
+      \If{$L_2 = \sigmaterm{\neglit {L_1}}$ for some nonidempotent $\sigma$}  \label{statement:self-impl}
+      \ForAll{$i \gets 1$  to $K_\mathrm{imp}$}
+         \State $L_2 \gets \sigmaterm{L_2}$
+         \State \textsc{AddImplication}($\neglit L_1$, $L_2$, $C$)
+      \EndFor
+      \EndIf
+    \EndIf
+  \EndProcedure
+
+  \kern1\jot
+
+  \Procedure{UntrackClause}{$C$}
+   %  \ForAll {$L_\mathrm{a}, L_\mathrm{c}, M$ such that $ (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$}
+   \ForAll{$L_\mathrm{a}, L_\mathrm{c}, M$ such that $(L_\mathrm{c},M) \in \impmap[L_\mathrm{a}]$}
+      \If{$C \in M$}
+         \State erase $(L_\mathrm{c},M)$ from $\impmap[L_\mathrm{a}]$
+      \EndIf
+      % \EndFor
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+%\end{nolinenumbers}
+%\end{figure}
+
+%% TYPESETTING: hack
+\vskip-1.5\jot
+\vskip\belowdisplayskip
+
+The algorithm views a clause $L \cor L'$ as two implications
+$\neglit{L} \impl L'$ and $\neglit{L'} \impl L$.
+%
+It stores only one entry for all literals equal up to variable renaming (line
+\ref{statement:alpha}). Each implication $L_\mathrm{a} \impl L_\mathrm{c}$
+represented by the clause is stored only if its generalization is not present in
+$\impmap$ (line~\ref{statement:more-general}). Conversely, all instances of the
+implication are removed (line~\ref{statement:instances}).
+
+Next, the algorithm
+finds each implication stored in $\impmap$ that can be linked to $L_\mathrm{a} \impl L_\mathrm{c}$: Either
+$L_\mathrm{c}$ becomes the new consequent (line~\ref{statement:transitivity}) or
+$L_\mathrm{a}$ becomes the new antecedent (line~\ref{statement:triggered}). If
+$L_\mathrm{c}$ can be decomposed into $\neqlit{u[s]}{u[t]}$, rule~3 of
+Definition \ref{def:hl-fo} allows us to store $\neqlit{s}{t}$ in
+$\impmap[L_\mathrm{a}]$ (line~\ref{statement:decomposition}). This is an
+exception to the idea that transitive chains should use only rule~1. The
+application of rule~3 does not count toward the bound $K_\mathrm{len}$.
+If $L_\mathrm{a}$ is of the form $\eqlit{u[s]}{u[t]}$, then
+$\impmap$ could be extended so that $\impmap[\eqlit{s}{t}] =
+\impmap[L_\mathrm{a}]$, but this would substantially increase $\impmap$'s
+memory footprint.
+
+In first-order logic, different instances of the same clause can
+be used along a transitive chain. For example, the clause $C =
+\negpredlit{\cst{p}(x)} \cor \pospredlit{\cst{p}(\cst{f}(x))}$ induces $\cst{p}(x) \bigfimpl{M}^i \cst{p}(\cst{f}^i(x))$ for all $i$.
+The algorithm discovers such self-implications (line~\ref{statement:self-impl}): For
+each clause~$C$ of the form $\neglit{L} \cor \sigma(L)$, where $\sigma$ is
+some nonidempotent substitution, the entires $(\sigma^2(L), \{C\}),\allowbreak
+\ldots,\allowbreak (\sigma^{K_{\mathrm{imp}} + 1}(L), \{C\})$ are added to
+$\impmap[L]$, where $K_{\mathrm{imp}}$ is a parameter.
+
+\newcommand{\concl}{\ensuremath{\impmap^{-1}}}
+
+\looseness=-1
+To track and untrack clauses efficiently, we implement the mapping $\impmap$ as a
+nonperfect discrimination tree \cite{rsv-2001-term-indexing}. Given a query
+literal $L$, this indexing data structure efficiently finds all literals $L'$ such that
+for some $\sigma$, $\sigmaterm{L'} = L$ and $\impmap[L'] \not= \emptyset$. We can use
+it to optimize all lookups except the one on line~\ref{statement:transitivity}.
+For this remaining lookup, we add an
+index \concl{} that inverts $\impmap$, i.e., $\concl[L] = \{ L' \mid
+\impmap[L'] = (L,M) \text{ for some } M \}$. To avoid sequentially going through
+all entries in $\impmap$ when the prover deletes them, for each clause $C$
+we keep track of each literal $L$ such that $C$ appears in $\impmap[L]$.
+Finally, we limit the number of entries stored in $\impmap[L]$ -- by default,
+up to 48 pairs in each $\impmap[L]$ are stored.
+
+\newcommand{\unitset}{\ensuremath{\textit{Unit}}}
+\begin{conf} 
+\looseness=-1   
+Rules \infname{HLE} and \infname{HTR} have a simple
+implementation based on $\impmap$ lookups. To implement \infname{UnitHLE} and
+\infname{UnitHTR}, we maintain the index \unitset, containing literals
+$L_\mathrm{c}\sigma$, such that $(L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$ for
+some $M$ and $L_\mathrm{a}$ and $\sigma$ is the most general unifier of $L'$ and
+$L_\mathrm{a}$, for some unit clause $\{ L' \}$. The implementation
+of \infname{FLE} and \infname{FLR} also uses \unitset:
+When $(L',M)$ is added to $\impmap[L]$, we
+check if $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $\neglit{L}$
+is added to $\unitset$.
+\end{conf}
+\begin{rep}
+
+   To implement the \infname{HLE} rule, we use $\impmap[L]$ as
+   follows: Given a clause $C = L \cor L' \cor C'$, if there are two literals
+   $L_1,L_2$ and a substitution $\sigma$ such that $(L_2, M) \in \impmap[L_1]$,
+   $C \not\in M$,  $\sigmaterm{L_1} = L$, and $\sigmaterm{L_2} = L'$, we remove $L$ from $C$.
+   Literal $L$ can also be removed if $\sigmaterm{L_1} = \neglit{L'}$ and $\sigmaterm{L_2} = \neglit{L}$. 
+   Rule \infname{HTR} is implemented analogously.
+
+% @PETAR: Used to say "UnitHTR rules rely", but there's only one. Double-check. --JB
+   The \infname{UnitHTR} rule relies on maintaining the index $\unitset$, which is
+   built as follows. Whenever the prover derives a unit clause $C = \{ L \}$, we
+   find all entries $L_\mathrm{a}$ in $\impmap$ such that $L_\mathrm{a}$ and $L$
+   are unifiable with the most general unifier $\sigma$. Then, we set $\unitset
+   \gets \unitset \cup \{ (\sigmaterm{L_\mathrm{c}}, M \cup \{ C \} ) \mid
+   (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}] \}$. Given a clause $L \cor C'$
+   we apply \infname{UnitHLE} by looking for $(L', M) \in \unitset$ such that
+   $\sigmaterm{L'} = \neglit{L}$, for some substitution $\sigma$; we apply
+   \infname{UnitHTR} by looking for $L'$ such that $\sigmaterm{L'} = L$.  The sets
+   stored together with literals in $\unitset$ are used for building the proof
+   object and to remove literals from $\unitset$ once a clause from the given
+   set becomes redundant.
+
+   The same data structure is used
+   for supporting \infname{FLE} and \infname{FLR}. When $(L',M)$ is added to $\impmap[L]$, we
+   check whether $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $(\neglit{L}, M \cup M')$
+   is added to $\unitset$.
+\end{rep}
+
+% % Using this algorithm, hidden-literal-based elimination can be implemented
+% % efficiently. We implement them as simplifications, used during saturation.
+% \begin{rep}
+% We implement \infname{ShallowHLE} by checking if there are two literals $L,L'
+% \in C$ such that for some substitution $\sigma$, and literals $L_1$ and $L_2$,
+% $(L_2, M) \in \impmap[L_1]$ and $L_1\sigma = L$, $L_2\sigma = L'$, $L$ can be
+% removed from $C$. When $(L',M)$ is added to $\impmap[L]$, Zipperposition checks
+% if $(\neglit{L'},M) \in \impmap[L]$ for some $M$. If so, it notes that there
+% are complementary literals in $\HL(\neglit{L},M)$. This information is used to
+% implement \infname{FLD}. For \infname{HTR}, a check whether there are $L,L' \in
+% C$ such that $(L_1, M) \in \impmap[L_0]$, where $\neglit{L_0\sigma} = L$ and
+% $L_1\sigma = L'$ is performed. Rule \infname{DeepHLE} is not implemented.
+
+% Rules \infname{UnitHLE} and \infname{UnitHTR} are implemented as well. For
+% \infname{UnitHLE}, given a clause $L \cor C$ we do as follows. First we try to
+% find a literal $L_0$ such that $\neglit{L_0\sigma} = L$ for some $\sigma$ and
+% $(L_1,M) \in \impmap[L_0]$ such that $L_1\sigma$ is subsumed by a unit clause
+% indexed by Zipperposition. If this does not succeed, we try finding a literal
+% $L_0$ such that $L_0\sigma = L$ for some $\sigma$ and $(L_1,M) \in \impmap[L_0]$
+% such that $\neglit{L_1}$ is subsumed by a unit clause. Rule \infname{UnitHTR} is
+% implemented analogously.
+% \end{rep}
+
+In propositional logic, the conventional approach constructs the
+\emph{binary implication graph} for the clause set~$N$
+\cite{hjb-2011-big-simplification}, with edges $(\neglit{L}, L')$ and
+$(\neglit{L'}, L)$ whenever $L \cor L' \in N.$ To avoid traversing the graph
+repeatedly, solvers rely on timestamps to discover connections between
+literals. This relies on syntactic literal comparisons, which is very
+fast in propositional logic but not in first-order logic, because of
+substitutions and congruence.
+
+\ourpara{Predicate Elimination}
+% 
+To implement portfolio predicate elimination, we maintain a record for each
+predicate symbol $\cst{p}$ occurring in the problem with the
+following fields:\
+set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
+$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
+once. These records are kept in a priority queue, prioritized by properties such
+as presence of definition sets and number of estimated resolutions. If
+$\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
+eliminate it by removing all the clauses stored in $\cst{p}$'s record from the
+proof state and by adding flat resolvents to the passive set.
+Eliminating a symbol might make another symbol eligible.
+%In this case, the updated record is put on the processing queue.
+
+As an optimization, predicate elimination keeps track only of symbols that appear
+at most $K_\mathrm{occ}$~times in the clause set. For inprocessing, we use
+signals that the prover emits whenever a clause is added to or removed from
+the proof state and update the records. At the beginning of the
+1st, $(K_\mathrm{iter} + 1)$st, $(2 K_\mathrm{iter} + 1)$st, $\dots{}$ iteration of the given clause procedure's
+loop body, predicate elimination is systematically applied to the entire proof
+state. The first application of inprocessing amounts to preprocessing. By
+default, $K_\mathrm{occ} = 512$ and $K_\mathrm{iter}=10$. The same ideas and
+limits apply for blocked clause elimination.
+
+The most important novel aspect of our predicate elimination implementation is
+recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
+as follows:
+\begin{enumerate}
+   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
+   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
+   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise continue.
+
+   \smallskip
+   \item Rename all clauses in $G$ so that their only variables are~$\tupleempty{x}$.
+
+   \smallskip
+   \item\label{step:encoding} Let $\lfloor a \rfloor$ be a function that assigns
+   a propositional variable to each atom $a$. This function is lifted to
+   literals by assigning $\lfloor \neg a \rfloor  = \neg x$, if $\lfloor a
+   \rfloor = x$, and to clauses pointwise. Furthermore,  
+   let $E = \{ \lfloor C' \rfloor \mid
+   \arbpredlit{\cst{p}(\tupleempty{x})} \cor C' \in G \}$. If $E$ is satisfiable,
+   report failure. Else, let $E'$ be the unsatisfiable core of $E$
+   and $G'$ the set of corresponding first-order clauses and continue.
+
+   \smallskip
+   \item If all resolvents in $G'_{\pospredlit{\cst{p}}} \flatres_{\!\cst{p}}
+   G'_{\negpredlit{\cst{p}}}$ are tautologies, then $G'$ is the definition set
+   for symbol $\cst{p}$. Else, report failure.
+\end{enumerate}
+
+The invalidity of set $E$ from step \ref{step:encoding} is checked using a SAT
+solver, which is already integrated in Zipperposition. As modern theorem provers
+(such as E or Vampire) also use SAT solvers, the method can easily be
+implemented.
+
+During experimentation, we noticed that recognizing definitions of symbols that
+occur in the conjecture often harms performance. Thus, Zipperposition
+recognizes definitions only for the remaining symbols.

### 2022-03-17 satfol/satfol.tex
more

@@
 recognizes definitions only for the remaining symbols.
+
+\section{Evaluation}
+\label{sec:satfol:evaluation}
+
+% {\bf\large TODO: Evaluate \textit{real} PPE in Figure \ref{fig:preprocess} and discuss SPPE and PPE.
+% References to PPE are broken now. }
+
+\newcommand{\baseres}[0]{7897}
+% \newcommand{\baseresplusfive}[0]{1563}
+
+\newcommand{\resnum}[1]{
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \ifnumcomp{\result}{>}{0}{$+\result$}{$\result$}%
+}
+\newcommand{\resnumphantom}[1]{%
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{$0$}{${+}\result$}}{\phantom{$0$}{${-}\negresult$}}%
+}
+\newcommand{\resnumbold}[1]{
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{${+}\result$}\llap{$\mathbf{\boldsymbol{+}\result}$}}{\phantom{${-}\negresult$}$\llap{\mathbf{\boldsymbol{-}\negresult}}$}%
+}
+\newcommand{\resnumboldphantom}[1]{%
+   \FPeval{\result}{clip(#1-\baseres)}%
+   \FPeval{\negresult}{clip(-\result)}%
+   \ifnumcomp{\result}{>}{0}{\phantom{$0{+}\result$}\llap{$\mathbf{\phantom{0}{\boldsymbol{+}}\result}$}}{\phantom{$0{-}\negresult$}$\llap{\mathbf{\phantom{0}{\boldsymbol{-}}\negresult}}$}%
+}
+\newcommand{\evaluating}[0]{\textbf{?}}
+
+\looseness=-1
+We measure the impact of our elimination techniques for various values of their
+parameters. As a baseline, we use Zipperposition's first-order portfolio mode,
+which runs the prover in 13~configurations of
+heuristic parameters in consecutive time slices. None of these configurations
+use our new techniques. To evaluate a given parameter value, we fix
+it across all 13~configurations and compare the results with the baseline.
+
+\looseness=-1
+The benchmark set consists of all 13\,495 CNF and FOF TPTP 7.3.0 theorems
+\cite{gs-17-tptp}.
+%
+The experiments were carried out on StarExec servers \cite{sst-14-starexec}
+equipped with Intel Xeon E5-2609 CPUs clocked at 2.40 GHz. The portfolio mode
+uses a single CPU core with a CPU time limit of 180~s. The base
+configuration solves \NumberOK{\baseres} problems. The values in the tables
+indicate the number of problems solved minus \NumberOK{\baseres}. Thus,
+positive numbers indicate gains over the baseline. The best result is shown in
+bold.
+
+\ourpara{Hidden-Literal-Based Elimination}
+% 
+The first experiments use all implemented HLBE rules.
+To avoid overburdening Zipperposition, we can enable an option to limit the
+number of tracked clauses for hidden literals. Once the limit has been reached, any
+request for tracking a clause will be rejected until a tracked
+clause is deleted. We can choose which kind of clauses are
+tracked:\ only clauses from the active set $\mathcal{A}$,
+only clauses from the passive set $\mathcal{P}$, or both.
+We also vary the maximal implication chain length $K_\mathrm{len}$ and the
+number of computed self-implications $K_\mathrm{imp}$.
+
+In Zipperposition, every lookup for instances or generalizations of
+$\eqlit{s}{t}$ must be done once for each orientation of the equation.
+To avoid this inefficiency, and also because the implementation of hidden literals does not fully
+exploit congruence, we can disable\pagebreak[2] tracking clauses with at least one
+functional literal. Clauses containing functional literals can then still be
+simplified.
+
+Figures~\ref{fig:hle-clauses-neq} and \ref{fig:hle-clauses-eq}
+%~\ref{fig:hle}
+show the results, without and with functional literal tracking enabled,
+for $K_\mathrm{len} = 2$ and $K_\mathrm{imp} = 0$.
+The columns specify different limits on the number of tracked clauses, with $\infty$
+denoting that no limit is imposed. The rows represent different kinds of tracked clauses.
+The results suggest that tracking functional literals is \NumberOK{not} worth the effort
+but that tracking predicate literals \NumberOK{is}. The best improvement is observed when
+both active and passive clauses are tracked. Normally DISCOUNT-loop
+provers \cite{adf-1995-discount} such as Zipperposition do not simplify
+active clauses using passive clauses, but here we see that this can be effective.
+%
+%--Raw data shows that when up to 250 clauses in either set are
+%--tracked, tracking functional literals helps solve \NumberNOK{5} problems that
+%--cannot be solved by a configuration that tracks only predicate literals or the
+%--base configuration. This suggests that a more optimized indexing implementation
+%--would help our technique reach its full potential.
+Figure~\ref{fig:km-kimp} shows the impact of varying $K_\mathrm{len}$ and
+$K_\mathrm{imp}$, when 500 clauses from the entire proof state are tracked.
+These results suggest that computing long implication chains \NumberOK{is counterproductive.}
+% Furthermore, raw data reveals that there is
+% no gain in this case: not a single problem that configuration with $K_\mathrm{len}=2$ and
+% $K_\mathrm{imp}=0$ cannot solve is solved when $K_\mathrm{imp}K_\mathrm{len}$ is set to 8 and
+% $K_\mathrm{imp}$ to 0.
+
+\begin{figure}[t]
+   \centering
+   \begin{minipage}{.48\textwidth}
+     \begin{center}
+       \def\arraystretch{1.1}%
+       \relax{\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
+         & \multicolumn{4}{c}{Tracked clauses} \\
+         & $250$        & $500$         & $1000$       &$\infty$ \\ \midrule 
+\strut Active       & \resnum{7883}  & \resnum{7881}   & \resnum{7889}  & \resnum{7885}    \\
+\strut Passive      & \resnumphantom{7904}  & \resnum{7907}   & \resnum{7902}  & \resnum{7862}    \\
+\strut Both         & \resnumbold{7909}  & \resnum{7907}   & \resnum{7904}  & \resnum{7852} \\     
+\end{tabular}}
+\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when only predicate literals are tracked}
+\label{fig:hle-clauses-neq}
+     \end{center}
+   \end{minipage}
+   \hfill
+   \begin{minipage}{.45\textwidth}
+     \begin{center}
+       \def\arraystretch{1.1}%
+       \relax{\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
+         & \multicolumn{4}{c}{Tracked clauses} \\
+         & $250$                   & $500$                 & $1000$                    & $\infty$ \\ \midrule 
+\strut Active       & \resnum{7887}            & \resnum{7883}         & \resnumphantom{7889}      & \resnum{7879}    \\
+\strut Passive      & \resnumphantom{7892}     & \resnumphantom{7892}  & \resnum{7883}             & \resnum{7826}    \\
+\strut Both         & \resnumboldphantom{7899} & \resnumphantom{7896}  & \resnumphantom{7889}      & \resnum{7818} \\ 
+\end{tabular}}
+\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when all literals are tracked}
+\label{fig:hle-clauses-eq}
+     \end{center}
+   \end{minipage}
+ \end{figure}
+
+\begin{figure}[t]
+\small
+\centering
+\def\arraystretch{1.1}%
+\begin{tabular}[t]{l@{\hskip 2em}c@{\hskip 1.5em}c@{\hskip1.5em}c@{\hskip 1.5em}c} 
+                             & \multicolumn{4}{c}{Chain length $K_\mathrm{len}$} \\
+                             & $1$                   & $2$                 & $4$                   & $8$ \\ \midrule 
+   $K_\mathrm{imp}=0$        & \resnum{7906}         & \resnum{7907}       & \resnum{7904}         & \resnum{7902}    \\
+   $K_\mathrm{imp}=1$        & \resnum{7902}         & \resnumbold{7908}       & \resnum{7904}         & \resnum{7901}    \\
+   $K_\mathrm{imp}=2$        & \resnum{7903}         & \resnumbold{7908}       & \resnum{7905}         & \resnum{7905}    \\
+\end{tabular}
+\caption{Impact of the parameters $K_\mathrm{len}$ and $K_\mathrm{imp}$ on HLBE performance}
+\label{fig:km-kimp}
+\end{figure}
+
+\begin{figure}[t!]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
+                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
+                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
+   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
+   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
+   \end{tabular}
+   \caption{Impact of the choice of criterion on predicate elimination performance}
+   \label{fig:criterion}
+\end{figure}
+
+\begin{figure}[t]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
+                         &               &               &                &               &                    & \smash{HLBE} \\[-1pt] %% TYPESETTING
+                         &               &               & \smash{SPE}    &               & \smash{PPE}        & \smash{+PPE} \\[-1pt] %% TYPESETTING
+                         & BCE           & SPE           & +BCE           & PPE           & +BCE               & +BCE           \\ \midrule 
+   \strut Preprocessing  & \resnum{7927} & \resnum{8051} & \resnum{8056}  & \resnum{8057} & \resnumbold{8063}  & \resnum{8059} \\
+   \strut Inprocessing   & \resnum{7849} & \resnum{8037} & \resnum{8024}  & \resnum{8043} & \resnum{8028}      & \resnum{8024}     \\
+   \end{tabular}
+   \caption{Performance of predicate and blocked clause elimination}
+   \label{fig:preprocess}
+\end{figure}
+
+\renewcommand{\baseres}{856}
+
+\begin{figure}[t]
+   \small
+   \centering
+   \def\arraystretch{1.1}%
+   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1em}c} \\
+                         &              &              &                   &              &              & \smash{HLBE}   \\[-1pt] 
+                         &              &              & \smash{SPE}       &              & \smash{PPE}  & \smash{+PPE}   \\[-1pt] 
+                         & BCE          & SPE          & +BCE              & PPE          & +BCE          & \smash{+BCE}    \\ \midrule 
+   \strut Preprocessing  & \resnum{885} & \resnum{902} & \resnumbold{916}  & \resnum{903} & \resnum{915} & \resnum{911}  \\
+   \end{tabular}
+   \caption{Performance of predicate and blocked clause elimination for establishing satisfiability}
+   \label{fig:bce-sat}
+\end{figure}
+
+\ourpara{Predicate and Blocked Clause Elimination}
+%
+% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
+% \textit{Do not forget to mention SPE was already implemented and }
+For defined predicate elimination, the number of resolvents grows
+exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
+computation, we limit the applicability of PPE to proof states for which $\cst{p}$
+is singular. According to our informal experiments, full PPE,
+without this restriction, generally performs less well.
+
+\begin{qle}\color{red} (TODO: Add QLE)\color{black}\end{qle}
+
+Predicate elimination can be done using Khasidashvili and Korovin's
+criterion (K\&K) or using our relaxed criterion with different values of
+$K_\mathrm{tol}$. Figure~\ref{fig:criterion} shows the results
+for SPE and PPE used as preprocessors.
+Our numbers corroborate Khasidashvili and Korovin's findings:
+SPE with K\&K proves \NumberOK{70} more problems than the base, a
+\NumberOK{0.9\%} increase, comparable to the 1.8\% they observe when
+they combine SPE with additional preprocessing.
+Remarkably, the number of additional proved problems more than
+\NumberOK{doubles} when we use our criterion with $K_\mathrm{tol} > 0$, for both SPE and PPE.
+
+\looseness=-1
+Although this is not evident in Figure~\ref{fig:criterion}, varying
+$K_\mathrm{tol}$ substantially changes the set of problems solved. For example,
+when $K_\mathrm{tol}=0$, SPE proves \NumberOK{60} theorems not proved using
+$K_\mathrm{tol}=50$. The effect weakens as $K_\mathrm{tol}$ grows. When
+$K_\mathrm{tol}=100$, SPE proves only \NumberOK{13} problems not found when
+$K_\mathrm{tol}=200$. Similarly, the set of problems proved by SPE and PPE
+differs: When $K_\mathrm{tol}=25$, \NumberOK{14} problems are proved by PPE but
+missed by SPE. Recognizing definition sets is useful:
+PPE \NumberOK{outperforms} SPE regardless of the criterion.
+
+% \looseness=-1 Having implemented those techniques, we are in position to answer
+% two interesting questions: {How helpful are definition clauses in defined
+% predicate elimination (DPE)?} and {What are the effects of performing blocked
+% clause elimination (BCE) and DPE until fixpoint (BD)?}. The second question was
+% also of interest to Kiesl et al.\ \cite{ksstb-2017-blockedfol}, but was left
+% unanswered as they did not implement predicate elimination.
+
+\looseness=-1
+Performing BCE and variable elimination until fixpoint increases the performance
+of SAT solvers~\cite{jbh-10-BCE}. We can check whether the same holds for superposition provers.
+In this experiment, we use the relaxed criterion with
+$K_\mathrm{tol}=25$ and HLBE which tracks up
+to 500 clauses from any clause set, $K_\mathrm{len}=2$, and
+$K_{\mathrm{imp}}=0$. We use each technique as preprocessing and inprocessing.
+
+The results are summarized in Figure~\ref{fig:preprocess}, where
+the +~sign denotes the combination of techniques.
+We confirm the results obtained by Kiesl et al.\ about the performance of BCE
+as preprocessing:
+It helps prove \NumberOK{30} more problems from our benchmark set, increasing the
+success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained Kiesl et al.
+%
+Using BCE as inprocessing, however, hurts performance, presumably because of
+its incompatibility with the redundancy criterion.
+
+\looseness=-1%
+For preprocessing, the combinations SPE+BCE and PPE+BCE performed roughly \NumberOK{on a
+par} with SPE and PPE, respectively. This stands in contrast to the situation
+with SAT solvers, where such a combination usually helps. It is also worth noting
+that the inprocessing techniques never outperform their preprocessing
+counterparts.
+%
+The last column shows that combining HLBE with other elimination techniques
+\NumberOK{overburdens} the prover.
+
+\ourpara{Satisfiability by Blocked Clause Elimination}
+%
+\looseness=-1%
+Kiesl et al.\ found that blocked clause elimination is especially effective on
+satisfiable problems. To corroborate their results and ascertain
+whether a combination of predicate elimination and blocked clause elimination
+increases the success rate, we evaluate BCE on all 2273
+satisfiable or TPTP FOF and CNF problems.
+The hardware and CPU time limits are the same as in the experiments above.
+Figure~\ref{fig:bce-sat} presents the results.
+
+\looseness=-1
+The baseline establishes the satisfiability of \baseres~problems. We consider
+only preprocessing techniques, since BCE compromises refutational
+completeness---a saturation does not guarantee that the original problem was
+satisfiable. We note that recognizing definition sets makes almost no difference on
+satisfiable problems. The sets of problems solved by BCE and PPE
+differ---\NumberOK{30} problems are solved by BCE and not by PPE.
+% Kiesl sees around 2.5%, we see 3.5% with BCE
+
+\section{Discussion and Related Work}
+\label{sec:satfol:discussion-and-related-work}
+
+We briefly surveyed related work in Sect.~\ref{sec:satfol:introduction}. In this
+section, we give a more detailed overview and further discuss connections with
+the related work.
+
+The research presented in this \paper{} is two-pronged. For SAT elimination
+techniques already generalized to preprocess first-order problems, we looked for
+ways to interleave them with the given clause procedure of a superposition
+prover, as inprocessing. For techniques that had not yet been ported to
+first-order logic, we looked for generalizations that allow both preprocessing and
+inprocessing.
+
+Hidden tautology elimination was first described
+by Heule et al.\ \cite{hjb-2010-cl-elim}. Better implementation that also
+supports hidden literal elimination was later described by the same group of
+authors \cite{hjb-2011-big-simplification}. We generalized the underlying
+theoretical concepts to first-order logic, and provided an efficient way to deal
+with the infinite number of hidden literals that arise  with this
+generalization. More efficient graph-based techniques are yet to be explored.
+
+Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
+the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
+\cite{ab-2004-re}. It was generalized to first-order logic (as a
+preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
+predicate elimination. Improvement of variable elimination,
+which uses formula definition information, has been popularized as a preprocessing and
+inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
+\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
+logic and combined it with existing extension. With tolerable restrictions,
+this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
+allowing variable elimination to slightly increase in the clause set size
+improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
+achieving the double the number of additional proofs found compared to more
+restrictive approaches.
+
+Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
+solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
+\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
+preprocessor. We showed that blocked clauses cannot be removed during
+saturation, but that they can be effectively used to show satisfiability of the
+clause set. A combination of blocked clause elimination and
+variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
+improvement when their generalizations are combined.
+
+Our approach is one of many ways to combine ideas from SAT solving and
+first-order proving. Other noteworthy architectures that either incorporate a
+SAT solver or that generalize the CDCL calculus include DPLL($T$) with
+quantifier instantiation
+\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
+DPLL($\mupGamma+T$) \cite{bonacina-et-al-2009},
+labeled splitting \cite{fietzke-weidenbach-2009},
+AVATAR \cite{av-2014-avatar},
+MCSAT \cite{de-moura-jovanovic-2013},
+CDSAT \cite{bonacina-et-al-2017}, and
+SGGS~\cite{bonacina-plaisted-2014}.
+
+\section{Conclusion}
+\label{sec:satfol:conclusion}
+
+We adapted several preprocessing and inprocessing elimination techniques implemented in
+modern SAT solvers so that they work in a superposition prover. This involved lifting
+the techniques to first-order logic with equality but also tailoring them to
+work in tandem with superposition and its redundancy criterion.
+Although SAT solvers and superposition provers embody radically different
+philosophies, we found that the lifted SAT techniques provide valuable optimizations.
+
+\pagebreak[2]
+We see several avenues for future work. First, the implementation of hidden
+literals could be extended to exploit equality congruence. Second, although
+inprocessing blocked clause elimination is incomplete in general, we hope to achieve
+refutational completeness for a substantial fragment of it. Third, predicate and
+blocked clause elimination, which thrives on the absence of clauses from the
+proof state, could be enhanced by tagging and ignoring generated clauses that
+have not yet been used to subsume or simplify untagged clauses. Fourth,
+predicate and blocked clause elimination could be extended to work with
+functional literals. Fifth, more SAT techniques could be adapted, including
+bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
+addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
+work with other first-order calculi, or generalized
+further to work with higher-order calculi such as combinatory superposition
+\cite{br-20-full-sup-w-combs} and \osup{}.

### 2022-03-18 pre/pre.tex
more

@@
 
+Some of the approaches described in this thesis are based on {\em rank-1
+polymorphism} \cite{bp-13-tff1,ksr-16-th1}. This is an extension of simply typed
+logic which adds type variables to the definition of types and type arguments to
+polymorphic constants (Chapter \ref{ch:bools}). It further
+requires that type arguments are instantiated with concrete, non-quantified
+types. In clausal structure it allows (implicit)
+quantification over types only on top-level and not inside literals.
+
 \section{Clausal Forms}
@@
 clauses is different from the formula-building $\ior$. In general,
-formula-building logical connectives are written in boldface.
-In standard, non-clausal first-order logic, an atom is either
-predicate or equational. For uniformity, and to
-stay close to the implementation, we encode predicate atoms as equations with
-$\itrue$. Negative literals are encoded as disequations. For example,
-$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
-$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the notation,
-we sometimes write the predicate literals in non-encoded form.
-
-Applying substitution to a literal is reduced to applying it to both sides of
-the (dis)equa\-tion and it is extended pointwise to clauses. It is sometimes
-also written in postfix notation as $l\sigma$ and $C\sigma$. We say
+formula-building logical connectives are written in boldface. In standard,
+non-clausal first-order logic, an atom is either predicate or equational. For
+uniformity, and to stay close to the implementation, we encode predicate atoms
+as equations with $\itrue$. Negative literals are encoded as disequations. For
+example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
+$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the
+notation, we sometimes write the predicate literals in non-encoded form.
+Applying a substitution to a literal is reduced to applying it to both sides of
+the (dis)equa\-tion and it is extended pointwise to clauses. We say
 $\sigmacl{C}$ is a ground instance of $C$ if $\sigma$ is a grounding
 substitution.
+\pagebreak[2]
 

### 2022-03-18 satfol/satfol.tex
more

@@
 A substitution is a \emph{variable renaming} if it is a
-bijection from a set of variables to a set of variables.
+bijection from a set of variables to a set of variables. Iterated, $i$-fold application 
+of nonidempotent application $\sigma$ is denoted by $\sigma^i$.
 
@@
 element $d$ in one of $\ourmodel$'s domains there exists a ground term $t$ such
-that $\ourmodel$ interprets $t$ as $d$. Canonical models generalize Herbrand
-models \cite{mf-1996-fol} to first-order logic with equality: If a clause set is satisfiable, then
+that $\ourmodel$ interprets $t$ as $d$. 
+We also make use of Herbrand models and Herbrand's theorem \cite[Sect.~5.4]{mf-1996-fol}.
+Canonical models generalize Herbrand
+models to first-order logic with equality: If a clause set is satisfiable, then
 it is satisfiable in a canonical model.

### 2022-03-21 satfol/satfol.tex
jasmin

@@
 
-\blfootnote{In this work I was the main designer behind all the presented techniques.
+\blfootnote{In this work I was the main designer of all the presented techniques.
 Marijn Heule did weekly supervision and provided his knowledge of SAT solving to 
 guide the design of all techniques. Jasmin Blanchette found the exact conditions
-under which superposition remains complete when predicate elimination rule is added
+under which superposition remains complete when the predicate elimination rule is added
 to the calculus. I also implemented and evaluated all the techniques. 
@@
 
-We assume the natural extensions of domain, valuation, interpretation and model
+We assume the natural extensions of domain, valuation, interpretation, and model
 (as defined by Fitting \cite{mf-1996-fol}) from unsorted to many-sorted logic.
@@
 models to first-order logic with equality: If a clause set is satisfiable, then
-it is satisfiable in a canonical model.
+it is satisfiable by a canonical model.
 
@@
       Without loss of generality,
-      we assume that the leading literal of $C$ is positive.
+      we assume that the leading literal of $C$ is positive (i.e., $C$ is of the form $\pospredlit{\cst{p}(\tuple{s}{n})} \cor C'$).
       Towards a contradiction, assume $\xi$ is a valuation such that $\mm \notmodels_\xi C$. Then, $\mm \notmodels_\xi
@@
    The completeness proof follows the invariance-based argument found in the
-   article by Waldmann et al \cite{waldmann-et-al-202x-article}.
+   article by Waldmann et al. \cite{waldmann-et-al-202x-article}.
 \end{sloppypar}
@@
 each clause~$C$ of the form $\neglit{L} \cor \sigma(L)$, where $\sigma$ is
-some nonidempotent substitution, the entires $(\sigma^2(L), \{C\}),\allowbreak
+some nonidempotent substitution, the pairs $(\sigma^2(L), \{C\}),\allowbreak
 \ldots,\allowbreak (\sigma^{K_{\mathrm{imp}} + 1}(L), \{C\})$ are added to

### 2022-03-21 satfol/satfol.tex
cosmetic

@@
 indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
+\{\cst{q}\}$ yet $\{\cst{p} \cor \cst{q}\} \not\models
 \{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
@@
    The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
-al., with adjustments to show that $\cst{P}$-predicates cannot reappear once
-they have disappeared. The case for \textsc{ElimPred} is trivial.
+al., with adjustments to show that $\cst{P}$-predicates cannot appear from
+nowhere. The case for \textsc{ElimPred} is trivial.
 \end{proof}
@@
 $(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
-static and dynamic completeness, we conclude that some $\mathcalx{N}{}_{k+i}$ must
+static and dynamic completeness (Lemmas 10~and~11
+in Waldmann et al.~\cite{waldmann-et-al-202x-article}), we conclude that some $\mathcalx{N}{}_{k+i}$ must
 contain a labeled empty clause.
@@
    Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
-   Non-trivial direction is proving that if $N'$ has a model, so does
+   The nontrivial direction is to prove that if $N'$ has a model, so does
    $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
@@
    $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$
-   then some literal in $\substterm{\tau}{D}$ must be satisfied. As $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
-   we again get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
+   then some literal in $\substterm{\tau}{D}$ must be satisfied. Since $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
+   we get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
    as needed.
@@
 
-Hidden-literal-based, predicate, \begin{qle}quasipure, \end{qle}and blocked clause
+Hidden-literal-based elimination, predicate elimination, \begin{qle}quasipure
+elimination, \end{qle}and blocked clause
 elimination all admit efficient implementations in a superposition
@@
    N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
-   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise continue.
+   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise, continue.
 
@@
 solver, which is already integrated in Zipperposition. As modern theorem provers
-(such as E or Vampire) also use SAT solvers, the method can easily be
+(including E and Vampire) also use SAT solvers, the method can easily be
 implemented.
@@
                            & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
-   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
-   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
+   \infname{SPE} preprocessing  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
+   \infname{PPE} preprocessing  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
    \end{tabular}
@@
 preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
-predicate elimination. Improvement of variable elimination,
-which uses formula definition information, has been popularized as a preprocessing and
+predicate elimination. An improvement of variable elimination,
+that uses formula definition information, has been popularized as a preprocessing and
 inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
 \cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
-logic and combined it with existing extension. With tolerable restrictions,
+logic and combined it with Khasidashvili and Korovin's approach. With tolerable restrictions,
 this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
@@
 clause set. A combination of blocked clause elimination and
-variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
+variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no comparable
 improvement when their generalizations are combined.
 
-Our approach is one of many ways to combine ideas from SAT solving and
+Our general approach is one of many ways to combine ideas from SAT solving and
 first-order proving. Other noteworthy architectures that either incorporate a
@@
 literals could be extended to exploit equality congruence. Second, although
-inprocessing blocked clause elimination is incomplete in general, we hope to achieve
+blocked clause elimination is generally incomplete as an inprocessing technique,
+we hope to achieve
 refutational completeness for a substantial fragment of it. Third, predicate and

### 2022-03-23 pre/pre.tex
jasmin
wan

@@
 decide propositional satisfiability problem (SAT solvers) use more advanced
-approaches such as the CDCL calculus \cite{mss-96-cdcl}. Modern SAT solvers also
-heavily preprocess the problem and continue simplifying it during the proving
-process.
+approaches such as the CDCL calculus \cite{mss-96-cdcl}. Crucial improvement of
+this calculus compared to more naive approaches is that it does not backtrack
+chronologically when it determines that the partial model constructed for a set
+of clauses does not satisfy all clauses. Instead, it generates a \emph{learned
+clause} which explains how this conflicting state was reached, which enables
+smarter backtracking.   Modern SAT solvers also heavily preprocess the problem
+and continue simplifying it during the proving process.
 
@@
 requires that type arguments are instantiated with concrete, non-quantified
-types. In clausal structure it allows (implicit)
+types. In a clausal structure it allows (implicit)
 quantification over types only on top-level and not inside literals.

### 2022-03-23 satfol/satfol.tex
wan

@@
 \begin{abstract}
-Optimized SAT solvers not only preprocess the clause set, they also transform it
-during solving as inprocessing. Some preprocessing techniques have been
-generalized to first-order logic with equality. In this \paper, we port
-inprocessing techniques to work with superposition, a leading first-order proof
-calculus, and we strengthen known preprocessing techniques. Specifically, we
-look into elimination of hidden literals, variables (predicates), and blocked
-clauses. Our evaluation using the Zipperposition prover confirms that the new
-techniques usefully supplement the existing superposition machinery.
+Optimized SAT solvers not only simplify the clause set during preprocessing, but
+they also simplify it during solving. This interleaving of
+simplification and solving is called inprocessing. Some preprocessing
+techniques have been generalized to first-order logic with equality. In this
+\paper, we port inprocessing techniques to work with superposition, a leading
+first-order proof calculus, and we strengthen known preprocessing techniques.
+Specifically, we look into elimination of hidden literals, variables
+(predicates), and blocked clauses. Our evaluation using the Zipperposition
+prover confirms that the new techniques usefully supplement the existing
+superposition machinery.
 \end{abstract}
@@
 
-Automated reasoning tools have become much more powerful in the last few
-decades thanks to procedures such as conflict-driven clause learning (CDCL)
-\cite{MSLM09HBSAT} for propositional logic and superposition
-\cite{bg-94-superposition} for first-order logic with equality. However,
-the effectiveness of these procedures crucially depends on how the input
-problem is represented as a clause set. The clause set
+Automated reasoning tools have become much more powerful in the last few decades
+thanks to procedures such as conflict-driven clause learning (CDCL)
+\cite{MSLM09HBSAT} for propositional logic and superposition for first-order
+logic with equality. However, the effectiveness of these procedures crucially
+depends on how the input problem is represented as a clause set. The clause set
 can be optimized beforehand (\emph{preprocessing}) or during the execution of
-the procedure (\emph{inprocessing}). In this \paper, we lift several preprocessing and
-inprocessing techniques from propositional logic to clausal first-order logic
-and demonstrate their usefulness in a superposition prover.
+the procedure (\emph{inprocessing}). In this \paper, we lift several
+preprocessing and inprocessing techniques from propositional logic to clausal
+first-order logic and demonstrate their usefulness in a superposition prover.
 
@@
 
-All techniques are implemented in the Zipperposition prover
+All techniques have been implemented in the Zipperposition prover
 (Sect.~\ref{sec:satfol:implementation}), allowing us to ascertain their usefulness
@@
 distinguished type (or sort) $o$. We introduced this logic in
-Sect.~\ref{sec:pre:fol} and we use the same notation introduced in this section.
+Sect.~\ref{sec:pre:fol} and here we use the same notation introduced in that section.
 We write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol
@@
 
-We assume the same notion of substitution introduced in Sect.~\ref{sec:pre:fol}.
+We assume the notion of substitution introduced in Sect.~\ref{sec:pre:fol}.
 We extend the notation by writing a substitution $\{x_1 \mapsto t_1,
@@
 bijection from a set of variables to a set of variables. Iterated, $i$-fold application 
-of nonidempotent application $\sigma$ is denoted by $\sigma^i$.
+of a nonidempotent application $\sigma$ is denoted by $\sigma^i$.
 
@@
 We also assume the notion of normal models from Sect.~\ref{sec:pre:fol} (with
-corresponding notation) and extend it with the notion of \emph{cannonical
+corresponding notation) and extend it with the notion of \emph{canonical
 models}: A canonical model $\ourmodel$ is a normal model such that for every
@@
 We also make use of Herbrand models and Herbrand's theorem \cite[Sect.~5.4]{mf-1996-fol}.
-Canonical models generalize Herbrand
-models to first-order logic with equality: If a clause set is satisfiable, then
-it is satisfiable by a canonical model.
+In Herbrand's models the domain consists of all ground terms. These models have a 
+useful property that if a formula is satisfiable in a model, then it is also satisfiable in an
+Herbrand model. Herbrand's theorem states that every unsatisfiable clause set has (propositionally)
+unsatisfiable ground instance.     
+Canonical models generalize Herbrand models to first-order logic with equality.
 
 All of the techniques described in this chapter are applied in the context of
-first-order superposition which is introduced in Sect.~\ref{sec:pre:sup}. We
-reuse all the notions introduced in this section and the corresponding notation.
+first-order superposition which was introduced in Sect.~\ref{sec:pre:sup}. We
+reuse all the notions introduced in this section and the corresponding notations.
 
@@
    $\PHL(L, N) = \{ L' \mid L' \bigpimpl{N}^\ast L \} \setminus \{ L \}$, where $\bigpimpl{N}$ is
-   defined such that $\neglit{L_1} \bigpimpl{N} L_2$ whenever $L_1 \cor L_2 \in N.$
+   defined so that $\neglit{L_1} \bigpimpl{N} L_2$ whenever $L_1 \cor L_2 \in N.$
    Moreover, $\PHL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n
@@
 %
-Given a unit clause $L' \in N$, the rule uses it to discharge $\sigmaterm{L'}$ in
+Given a unit clause $L' \in N$, the rule uses $L'$ to discharge $\sigmaterm{L'}$ in
 $N \models \sigmaterm{L'} \impl \neglit{L}$. As a result, we have $N \models
@@
    inferences between clauses in $\aset$ are redundant. The end result is not
-   that the prover diverges, but that it terminates without deriving empty
+   that the prover diverges, but that it terminates without deriving the empty
    clause.
@@
    other words, whenever $(M,N) \flatresiter (M, N_1)$ and $(M,N) \flatresiter
-   (M, N_2)$ there exists $N'$ such that $(M, N_1) \flatresiter (M, N')$ and
+   (M, N_2)$, there exists $N'$ such that $(M, N_1) \flatresiter (M, N')$ and
    $(M, N_2) \flatresiter (M,N')$.
@@
    There are two main sources of nondeterminism of $\flatresiter$: The choice of $C
-   \in N$ and the choice of the literal in $C$ to act on. Let us focus on the choice of
-   $C$ in $N$; the same discussion applies for the choice of literal in $C$.
+   \in N$ and the choice of a literal in $C$ to act on. Let us focus on the choice of
+   $C$ in $N$; the same discussion applies for the choice of a literal in $C$.
 
@@
    \{C_1 \cup N' \}) \flatresiter (M, N'_2 \cup \{ C_1 \cup N' \})$ where $N'_1$ and
-   $N'_2$ are sets of corresponding resolvents. Both of $\flatresiter$ steps can
+   $N'_2$ are sets of corresponding resolvents. Both $\flatresiter$ steps can
    be joined (up to variable renaming) to $(M, N'_1 \cup N'_2 \cup N')$, showing
@@
 $\mu(N') < \mu(N)$ or
-$|N'| < |N| + K_\mathrm{tol}$.
-\confrep{}{A refinement, which we want to try out in future work, would be to gradually
-increment the tolerance $K_\mathrm{tol}$, as is done in some SAT solvers.}
+$|N'| < |N| + K_\mathrm{tol}$. Intuitively, we allow predicate elimination
+even in the cases in which proof state increases in size, as long as we can control
+this increase with the parameter $K_\mathrm{tol}$. 
+A refinement, which we want to try out in future work, would be to gradually
+increment the tolerance $K_\mathrm{tol}$, as is done in some SAT solvers.
 
@@
       Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }an arbitrary clause
-      set. If $(G, N) \flatresiter (G, N')$ then $G \cup N$ and $G \cup N'$ are
+      set. If $(G, N) \flatresiter (G, N')$, then $G \cup N$ and $G \cup N'$ are
       equivalent.
@@
       \neqlit{x_1}{s_1} \cor \cdots \cor \neqlit{x_n}{s_n} \cor C' \cor D'$.
-      Let $\zeta$ be a valuation coinciding with $\xi$ on variables of $C$
+      Let $\zeta$ be a valuation coinciding with $\xi$ on the variables of $C$
       and with $\xi'$ on $\tuplen{x}$.
@@
       $\pospredlit{\cst{p}(\tuplen{y})} \cor D' \in G$ and an assignment $\xi'$
-      which assigns each $y_i$ value of $\xi(x_i)$ such that $\mm
+      which assigns each $y_i$ the value of $\xi(x_i)$ such that $\mm
       \notmodels_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
@@
 Because $\mathbf{P}$ is finite, any mixed derivation consists of at most
-finitely many $\rhd_\mathbf{P}$ transitions. Hence, in any derivation,
+finitely many $\rhd_\mathbf{P}$-transitions. Hence, in any derivation,
 there exists an index~$k$ from which all transitions are standard
@@
 $\cst{q} \notin \mathbf{P}$, and $s, t$ are non-Boolean terms:
+\pagebreak[2]
 %
@@
 rewriting clause is smaller than the rewritten
-clause (a condition that is also needed with $\models$ but omitted by Schulz).
+clause (a condition that is also needed with $\models$ but omitted by Schulz
+to allow more aggressive application of this useful rule).
 Finally, destructive equality resolution cannot be justified with $\models$,
@@
 sequence prefix of an a priori unknown length. The argument will still work
-thanks to monotonicity properties of redundancy criteria.
+thanks to monotonicity properties of the redundancy criteria.
 
@@
 but it can also be used to convert a finite Herbrand model
-into a certificate of clause set satisfiability. The certificate uses only
+into a certificate of clause set satisfiability (i.e., an object that carries the checkable proof of satisfiability). The certificate uses only
 blocked clause elimination and addition, in conjunction with a transformation
@@
    literal in $\sigmaterm{C}$ must be satisfied by $\ourmodel$, contradicting that
-   $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$
+   $\sigmaterm{C}$ is falsified by $\ourmodel$. If $\ourmodel \notmodels \substterm{\tau}{L'}$,
    then some literal in $\substterm{\tau}{D}$ must be satisfied. Since $\substterm{\tau}{D} \subseteq \sigmaterm{C}$,
@@
 between $L \cor C$ and $\neglit{L_1} \cor \cdots \cor \neglit{L_n} \cor
-D$ (which are renamed apart) is $\sigmacl{C \cor D}$, where $\sigma$ is the most general unifier
+D$ (which are renamed apart) is $\sigmacl{C \cor D}$, where $\sigma$ is the MGU
 of the literals $L, L_1, \ldots, L_n$.
@@
 
-The procedure is limited to the first-order logic without equality, since step
+The procedure is limited to first-order logic without equality, since step
 \ref{item:step3} is justified only if $L$ is a predicate literal. (Otherwise,
@@
 techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
-All techniques are implemented in the Zipperposition prover
-\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
-of improvements to superposition, but it implements many of the most successful
-heuristics from the E prover \cite{scv-19-e23} and has recently become
-quite competitive \cite{gs-19-casc27}.
+All techniques have been implemented in the Zipperposition prover
+\cite{sc-supind-17}.
 
@@
    find all entries $L_\mathrm{a}$ in $\impmap$ such that $L_\mathrm{a}$ and $L$
-   are unifiable with the most general unifier $\sigma$. Then, we set $\unitset
+   are unifiable with the MGU $\sigma$. Then, we set $\unitset
    \gets \unitset \cup \{ (\sigmaterm{L_\mathrm{c}}, M \cup \{ C \} ) \mid
-   (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}] \}$. Given a clause $L \cor C'$
+   (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}] \}$. Given a clause $L \cor C'$,
    we apply \infname{UnitHLE} by looking for $(L', M) \in \unitset$ such that
@@
 following fields:\
-set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
-$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
+the set of definition clauses for $\cst{p}$, the set of nondefinition clauses in which
+$\cst{p}$ occurs once, and the set of clauses in which $\cst{p}$ occurs more than
 once. These records are kept in a priority queue, prioritized by properties such
-as presence of definition sets and number of estimated resolutions. If
+as the presence of definition sets and the number of estimated resolutions. If
 $\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
@@
 recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
-as follows:
+as follows, assuming $\tupleempty{x}$ is a fixed tuple of distinct free variables:
 \begin{enumerate}
-   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
-   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
-   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise, continue.
-
-   \smallskip
+   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{y})} \cor C', C \in
+   N$, no variable repeats in $\tupleempty{y}$, and the variables of $C'$ are among
+   $\tupleempty{y}\}$. If $G$ is empty, report failure; otherwise, continue.
+
+%    \smallskip
    \item Rename all clauses in $G$ so that their only variables are~$\tupleempty{x}$.
 
-   \smallskip
+%    \smallskip
    \item\label{step:encoding} Let $\lfloor a \rfloor$ be a function that assigns
    a propositional variable to each atom $a$. This function is lifted to
-   literals by assigning $\lfloor \neg a \rfloor  = \neg x$, if $\lfloor a
+   literals by assigning $\lfloor \neg a \rfloor  = \neg x$ if $\lfloor a
    \rfloor = x$, and to clauses pointwise. Furthermore,  
@@
 
-   \smallskip
+%    \smallskip
    \item If all resolvents in $G'_{\pospredlit{\cst{p}}} \flatres_{\!\cst{p}}
@@
 It helps prove \NumberOK{30} more problems from our benchmark set, increasing the
-success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained Kiesl et al.
+success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained by Kiesl et al.
 %
@@
 only preprocessing techniques, since BCE compromises refutational
-completeness---a saturation does not guarantee that the original problem was
+completeness---a saturation does not guarantee that the original problem is
 satisfiable. We note that recognizing definition sets makes almost no difference on
@@
 section, we give a more detailed overview and further discuss connections with
-the related work.
+related work.
 
@@
 Hidden tautology elimination was first described
-by Heule et al.\ \cite{hjb-2010-cl-elim}. Better implementation that also
+by Heule et al.\ \cite{hjb-2010-cl-elim}. A better implementation that also
 supports hidden literal elimination was later described by the same group of
@@
 \cite{ab-2004-re}. It was generalized to first-order logic (as a
-preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
+preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol}, yielding a technique called
 predicate elimination. An improvement of variable elimination,
@@
 this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
-allowing variable elimination to slightly increase in the clause set size
+allowing variable elimination to slightly increase the clause set size
 improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
-achieving the double the number of additional proofs found compared to more
+achieving double the number of additional proofs found compared to more
 restrictive approaches.
@@
 refutational completeness for a substantial fragment of it. Third, predicate and
-blocked clause elimination, which thrives on the absence of clauses from the
+blocked clause elimination, which thrive on the absence of clauses from the
 proof state, could be enhanced by tagging and ignoring generated clauses that

### 2022-03-23 ho-techniques/ho-techniques.tex
difficulty rating

@@
 \NumberOK{16} problems that pre-RW does not, \NumberOK{three} of which have a TPTP
-difficulty rating (the ratio of eligible %TPTP
-provers that cannot prove the problem) of~1.
+difficulty rating of~1. Difficulty rating is a number from 0 to 1, proportional to the number
+of state-of-the-art provers that attempted, but failed to solve to problem \cite{gscb-01-evaluating}.
 

### 2022-03-23 pre/pre.tex
difficulty rating

@@
 decide propositional satisfiability problem (SAT solvers) use more advanced
-approaches such as the CDCL calculus \cite{mss-96-cdcl}. Crucial improvement of
+approaches such as the CDCL calculus \cite{mss-96-cdcl}. A crucial improvement of
 this calculus compared to more naive approaches is that it does not backtrack

### 2022-03-23 satfol/satfol.tex
difficulty rating

@@
    The completeness proof follows the invariance-based argument found in the
-   article by Waldmann et al. \cite{waldmann-et-al-202x-article}.
+   article by Waldmann et al. \cite{waldmann-et-al-202x-article} about the saturation framework.
 \end{sloppypar}
@@
 $\textstyle\FLInf(\mathcalx{N}{}_k{\downarrow}_\mathsf{active}) \subseteq
-\LRedIWk(\mathcalx{N}{}_k)$.
+\LRedIWk(\mathcalx{N}{}_k)$. \vspace{0.5em}
 Notice the difference with the definition of the key invariant
@@
 $\{C_n,\ldots,C_2\} \cup \mathcalx{D} \modelsWk \{C_0\}$.
-Without loss of generality, we assume that $\mathcalx{D}$ is the smallest such
-set.
+Without loss of generality, we assume that $\mathcalx{D}$ is the $\prec$-smallest such set; such a set exists because $\prec$ is well founded.
 

### 2022-03-24 introduction/introduction.tex
jasmin
intro

@@
 logical relations such as ``if it rains, the road is muddy'' (or formally
-$\cst{rains} \imp \cst{muddy}$), but also to quantify over objects as in ``all
-people are mortal'' ($\iforall x.\, \cst{human}(x) \imp \cst{mortal}(x)$).
+$\cst{rains} \iimplies \cst{muddy}$), but also to quantify over objects as in ``all
+people are mortal'' ($\iforall x.\, \cst{human}(x) \iimplies \cst{mortal}(x)$).
 
@@
 mostly first-order formulas, are rather large, and mostly require simple
-higher-order proof steps. Thus, from the standpoint of proof assistants the ideal
+higher-order proof steps. An example of a problem that is seemingly simple, but its translation to first-order logic
+is out of reach for even the best of first-order provers is \cite{bbbntvw-2x-mechanical-mathematicians}:
+
+$$ \big(\sum_{i=1}^{n} i^2 + 2i + 1\big) \ieq \big(\sum_{i=1}^{n} i^2\big) +
+\big(\sum_{i=1}^{n} 2i\big) + \big(\sum_{i=1}^{n} 1\big) $$ \noindent One of the
+reasons why this problem is hard is that higher-order proof assistants rely on
+unnamed functions, called \emph{$\lambda$-abstractions}, to represent
+variable-binding mathematical operators such as summations, integrals and
+limits. More precisely, the left-hand side of the above equation is usually
+represented as $\cst{sum}(1, n, \lambda i. \, i\cst{*}i + 2\cst{*}i + 1)$. When
+translated to first-order logic, $\lambda$-abstractions, which are ubiquitous as
+all-purpose binders in higher-order logic, are obfuscated to the point that all
+but the simplest  reasoning about them are out of reach for first-order provers.
+However, for higher-order provers the crux of the proof is matching the axiom
+$\iforall i j f g.\, \cst{sum}(i, j, \lambda x. \, f \, x \, \cst{+} \, g\,x) \ieq
+(\cst{sum}(i, j, \lambda x. \, f \, x ) \, \cst{+} \, \cst{sum}(i, j, \lambda x. \, g
+\, x ) )$ against the goal. This operation is relatively simple and, in general, much more efficient
+than first-order reasoning about translations.
+
+Thus, from the standpoint of proof assistants the ideal
 automatic theorem prover fulfills the following wishlist:
@@
 objects. Thus, higher-order logic allows us to make a statement ``all functions
-with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
+with arguments $0$ and $1$ result in $2$" ($\iforall x. \,  x (\cst{0}, \cst{1}) \eq
 \cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
@@
 present in the theory. Suppose we want to prove there is a
-function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
+function that returns 4 when given 2 and 5 when given 3 ($\iexists x. x(\cst{2})
 \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
@@
 represent statements such as ``what Jasmin says is true'' as $\iforall x. \,
-\cst{says}(\cst{jasmin}, x) \imp x$. The last stop is to support Boolean terms
+\cst{says}(\cst{jasmin}, x) \iimplies x$. The last stop is to support Boolean terms
 as first-class citizens.
@@
   manageable.
+  \pagebreak[2]
 
@@
 \begin{enumerate}[resume]
-  \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
-  Draft paper.
-  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
-  Draft article.
+  % \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
+  % Draft paper.
+  % \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
+  % Draft article.
   \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.

### 2022-03-24 introduction/introduction.tex
prepare ehoh2

@@
 mostly first-order formulas, are rather large, and mostly require simple
-higher-order proof steps. An example of a problem that is seemingly simple, but its translation to first-order logic
+higher-order proof steps. An example of a problem that is seemingly simple, but whose translation to first-order logic
 is out of reach for even the best of first-order provers is \cite{bbbntvw-2x-mechanical-mathematicians}:

### 2022-03-24 scratch/paper.tex
more
prepare ehoh2

--- 2022-03-16 scratch/paper.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is a general template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% Copy it to a new file with a new name and use it as the basis
-% for your article. Delete % signs as needed.
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-\documentclass[smallcondensed,draft]{svjour3}     % onecolumn (ditto)
-%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage[utf8]{inputenc}
+% For VSCode
+% !TEX root = ./conf.tex
+
+%% TACAS: 16 pp excluding bibliography
+
+\pagestyle{plain}
+\PassOptionsToPackage{hyphens}{url}
+
+\usepackage[T1]{fontenc}
 \usepackage[misc]{ifsym}
-\usepackage[final]{graphicx}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{cite}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usepackage{prftree}
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{caption}
-\usepackage{subcaption}
+\usepackage{bm}
+\usepackage{booktabs}
+\usepackage{cite}
+\usepackage{color}
 \usepackage{enumerate}
-\usepackage{stmaryrd}
-\usepackage{booktabs}
+\usepackage{enumitem}
+%\usepackage{stmaryrd}
+%\usepackage{textcomp}
+\usepackage{url}
+%\usepackage{xparse}
+%\usepackage{xpatch}
+%\usepackage{mathtools}
+\usepackage{microtype}
+\usepackage{array}
+%\usepackage{nicefrac}
+%\usepackage{subcaption}
+%\usepackage{tabu}
+\usepackage{prftree}
+%\usepackage{float}
+%\usepackage{bm}
+%\usepackage{graphicx}
+%\usepackage{wrapfig}
+\usepackage{multirow}
+%\usepackage{soul}
+\usepackage{fancyvrb}
+\usepackage{siunitx}
+%\let\vec=\relax
+%\usepackage{fdsymbol}
+%\let\longleadsto=\leadsto
+\usepackage{pifont}
+\usepackage{hyphenat}
 \usepackage{algorithmicx}
+\usepackage{regexpatch}
 \usepackage{algorithm}% http://ctan.org/pkg/algorithms
 \usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\xpatchcmd{\algorithmic}{\labelsep 0.5em}{\labelsep 3em}{\typeout{Success!}}{\typeout{Oh dear!}}
+
+\hyphenation{ab-strac-tion}
+
+%% for algorithms
+%\usepackage[plain]{algorithm}
+%\usepackage[noend]{algpseudocode}
+
 \usepackage[
@@
    pdftex,
-   pdftitle={Making Higher-Order Superposition Work},
-   pdfauthor={Petar Vukmirovi\'c, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret},
+%   pdftitle={Superposition for Lambda-Free Higher-Order Logic},
+%   pdfauthor={Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, and Uwe Waldmann},
    pdfkeywords={},
@@
 
-\usepackage{mathptmx}      % use Times fonts if available on your TeX system
-
-\DeclareSymbolFont{letters}{OML}{txmi}{m}{it} %% for Greek
-
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-\journalname{Journal of Automated Reasoning}
+\usepackage{tikz}
+\usepackage{mdframed}
+%\usepackage{tikz-qtree}
+%\usetikzlibrary{arrows,shapes,positioning,shadows,trees,calc}
+
+%%% BEGIN Times font
+%\usepackage{mathptmx}
+%\usepackage[scaled=.82]{beramono}
+%\usepackage[scaled=.865]{helvet}
+%%\usepackage{avant}
+%\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
+%%% END Times font
 
 \makeatletter
-
-\def\vthinspace{\kern+0.083333em}
-\newcommand\CORR{{(\vthinspace\Letter\vthinspace)}}
-
-\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+\g@addto@macro{\UrlBreaks}{\UrlOrds\do\=\do\_}
+\makeatother
+
+\hyphenation{counter-example counter-examples Isa-belle Ma-try-osh-ka
+  frame-work frame-works non-empty sko-lem-iza-tion super-posi-tion
+  over-approx-imate over-approx-imated auto-schedule}
 
@@
 
-%%% HACK Springer's style to get bold subsection headings --JB
-\def\subsection{\@startsection{subsection}{2}{\z@}%
-    {-21dd plus-8pt minus-4pt}{10.5dd}
-     {\normalsize\bfseries\boldmath}}
-
-\spnewtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
-\spnewtheorem{exa}[theorem]{Example}{\bfseries}{\rmfamily}
-
-\begin{document}
-
-\title{Making Higher-Order Superposition Work}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Petar~Vukmirović~\orcid{0000-0001-7049-6847}\and
-Alexander~Bentkamp~\orcid{0000-0002-7158-3595}\and
-Jasmin~Blanchette~\orcid{0000-0002-8367-0936}\and
-Simon~Cruanes~\orcid{0000-0003-3969-5850}\and
-Visa~Nummelin~\orcid{0000-0003-0078-790X} \and
-Sophie~Tourret~\orcid{0000-0002-6070-796X}}
-
-\authorrunning{Petar Vukmirović et al.} % if too long for running head
-
-\institute{
-  Petar~Vukmirovi\'c \CORR \and
-  Alexander Bentkamp\and
-  Jasmin Blanchette\and
-  Visa Nummelin\at%
-    Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\
-    \email{\{p.vukmirovic,a.bentkamp,j.c.blanchette,visa.nummelin\}@vu.nl}
-  \and
-  Jasmin Blanchette\and Sophie Tourret \at%
-    Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France
-    \\
-    \email{\{jasmin.blanchette,sophie.tourret\}@inria.fr}
-  \and
-  Jasmin Blanchette\and Sophie Tourret \at  Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken,~Germany \\
-  \email{\{jblanche,stourret\}@mpi-inf.mpg.de}
-  \and
-  Simon Cruanes \at Imandra, Austin, Texas, USA \\
-  \email{simon@imandra.ai}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Superposition is among the most successful calculi for first-order logic. Its
-extension to higher-order logic introduces new challenges such as infinitely
-branching inference rules, new possibilities such as reasoning about
-Booleans, and the need to curb the explosion of specific higher-order rules. We
-describe techniques that address these issues and extensively evaluate their
-implementation in the Zipperposition theorem prover. Largely thanks to their use,
-Zipperposition won the higher-order division of the CASC-J10 competition.
-\end{abstract}
-
+\let\labelitemi=\labelitemii %% CHEAT!
+
+%%% BEGIN poor (wo)man's algorithms
+% \newcommand\Function[2]{\textbf{function} \textsc{#1}(#2) \textbf{is} \\}
+% \newcommand\Procedure[2]{\textbf{procedure} \textsc{#1}(#2) \textbf{is} \\}
+% \newcommand\If[1]{\textbf{if} #1 \textbf{then} \\}
+% \newcommand\IfWoThen[1]{\textbf{if} #1 \\}
+% \newcommand\Then[1]{\phantom{\textbf{if}} \quad #1 \textbf{then} \\}
+% \newcommand\Else{\textbf{else} \\}
+% \newcommand\ElsIf[1]{\textbf{else} \textbf{if} #1 \textbf{then} \\}
+% \newcommand\ForTo[2]{\textbf{for} #1 \textbf{to} #2 \textbf{do} \\}
+% \newcommand\Do{\textbf{do} \\}
+% \newcommand\Forever{\textbf{forever} \textbf{do} \\}
+% \newcommand\WhileOfDo[1]{\textbf{while} #1}
+% \newcommand\ForDownto[2]{\textbf{for} #1 \textbf{downto} #2 \textbf{do} \\}
+% \newcommand\While[1]{\textbf{while} #1 \textbf{do} \\}
+% \newcommand\Return{\textbf{return} }
+
+% \newcommand\q{\noindent\hbox{}\quad}
+% \newcommand\qq{\q\q}
+% \newcommand\qqq{\qq\q}
+% \newcommand\qqqq{\qqq\q}
+% \newcommand\qqqqq{\qqqq\q}
+% \newcommand\qqqqqq{\qqqqq\q}
+% \newcommand\qqqqqqq{\qqqqqq\q}
+% \newcommand\qqqqqqqq{\qqqqqqq\q}
+% \newcommand\qqqqqqqqq{\qqqqqqqq\q}
+
+% \newlength\quotexwidth
+% \quotexwidth=\textwidth
+% \addtolength\quotexwidth{-\leftmargin}
+
+\newcommand\NumberOK[1]{#1}
+\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
+
+\newcommand\ourpara[1]{\paragraph{\upshape\bfseries#1.}}
+
+\newenvironment{quotex}
+ {\list{}{\rightmargin0pt}%
+  \item\relax}
+ {\endlist}
+%%% END poor (wo)man's algorithms
+
+\newcommand{\cmark}{\text{\ding{51}}}
+\newcommand{\xmark}{\text{\ding{55}}}
+
+\newcommand\CRS{\confrep{\texttt{ConjectureRelativeSymbol}}{\texttt{CRS}}}
+
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\newcommand\HLINE{\\[.25\jot]\hline\noalign{\vskip.5\jot}}
+\newcommand\infname[1]{\textsc{#1}}
+\newcommand{\infer}[3]{\prftree[r]{{\small\infname{#1}}}{\strut#2}{\strut#3}}
+\newcommand{\inferdouble}[3]{\prftree[d,r]{{\small\infname{#1}}}{\strut#2}{\strut#3}}
+
+\newcommand{\hooklongrightarrow}{\lhook\joinrel\longrightarrow}
+
+\newcommand\Unifarrow{\Longrightarrow}
+\newcommand\Matcharrow{\Longrightarrow}
+\newcommand\Pdtarrow{\leadsto}
+\newcommand\Matchiiarrow{\hooklongrightarrow}
+
+% Most general unifier
+\DeclareMathOperator{\mgu}{mgu}
+
+\def\negvthinspace{\kern-0.083333em}
+\def\vthinspace{\kern+0.083333em}
+\def\vvthinspace{\kern+0.0416667em}
+\def\negvvthinspace{\kern-0.0416667em}
+%\def\hypsep{\hskip0.75em}
+\def\hypsep{\quad}
+\def\rulesep{\qquad}
+
+\spnewtheorem{theoremx}[theorem]{Theorem}{\bfseries}{\slshape}
+\spnewtheorem{corollaryx}[theorem]{Corollary}{\bfseries}{\slshape}
+\spnewtheorem{lemmax}[theorem]{Lemma}{\bfseries}{\slshape}
+
+\newcommand\ehohii{$\lambda$E}
+
+\newcommand{\eq}{\approx}
+\newcommand{\noteq}{\not\eq}
+
+\newcommand\UNIF{\mathrel{\smash{\stackrel{\lower.1ex\hbox{\ensuremath{\scriptscriptstyle ?}}}{=}}}}
+%\newcommand\UNIF{=^?}
+\newcommand\MATCH{\mathrel{{\scriptstyle\lesssim}^{\scriptscriptstyle{?\!}}}}
+
+%\newcommand\UNIF{=^{?\!}}
+%\newcommand\MATCH{\lesssim^{?\!}}
+%\newcommand\MATCH{\mathrel{\smash{\stackrel{\lower.1ex\hbox{\ensuremath{\scriptscriptstyle ?}}}{\lesssim}}}}
+
+\newcommand\Var{\mathcal{V}\negvthinspace\mathit{ar}}
+\newcommand\Term{\mathcal{T}\kern-.7ex\mathit{erm}}
+\newcommand\Gfpf{\mathcal{G}\kern-.3ex\mathit{fpf}}
+\newcommand\GfpfRTL{\Gfpf'}
+\newcommand\Fp{\mathcal{F}\kern-.3ex\mathit{p}}
+\newcommand\FpRTL{\Fp'}
+
+%\renewcommand\AA{\textbf{\textsf{A}}}
+%\newcommand\BB{\textbf{\textsf{B}}}
+%\newcommand\NN{\textbf{\textsf{N}}}
+
+\renewcommand\AA{{\textsf{A}}}
+\newcommand\BB{{\textsf{B}}}
+\newcommand\NN{{\textsf{N}}}
+
+\definecolor{verylightgray}{gray}{0.9}
+\definecolor{pdtgreen}{HTML}{97D077}
+\definecolor{pdtorange}{HTML}{D79B00}
+
+\DeclareRobustCommand{\hlorange}[1]{{\sethlcolor{orange}\hl{#1}}}
+\DeclareRobustCommand{\hlgreen}[1]{{\sethlcolor{green}\hl{#1}}}
+\DeclareRobustCommand{\hlcolor}[2]{{\sethlcolor{#1}\hl{#2}}}
+
+% for Springer theorems
+\spnewtheorem{examplex}{Example}{\bfseries}{\rmfamily}
+\spnewtheorem{definitionx}[theorem]{Definition}{\bfseries}{\rmfamily}
+
+% probably can be set up differently
+%\theoremstyle{definition}
+%\newtheorem{definitionx}{Definition}
+
+%\theoremstyle{definition}
+%\newtheorem{examplex}{Example}
+
+%-- from my thesis
+%\theoremstyle{definition}
+%\newtheorem{remark}{Remark}
+
+\newcommand\afterDot{\,} %%% the Springer style is a bit broken---this makes things a bit more tolerable
+\def\cpp{C\nobreak\raisebox{.1ex}{+}\nobreak\raisebox{.1ex}{+}}
+\newcommand\appE{\ensuremath{\cst{@}{+}\text{E}}}
+
+\newcommand\cst[1]{\mathsf{#1}}
+
+%%%%%%%%%%%%%%%%%%%%%%
+% conf and rep
+\newcommand{\confrep}[2]{\begin{conf}#1\end{conf}\begin{rep}#2\end{rep}}
+\newcommand{\lfhol}[0]{$\lambda$fHOL}
+\newcommand{\appvar}[0]{\ensuremath{@}}
+
+\newcommand{\lam}[2]{\ensuremath{\lambda #1.\> #2}}
+\newcommand{\lamx}[1]{\lam{x}{#1}}
+
+% stolen from FOBOOLSUP paper
 \newcommand{\ifalse}{\pmb\bot}
@@
 \newcommand{\comb}[1]{\ensuremath{\mathsf{#1}}}
-\newcommand{\neglit}[1]{\ensuremath{#1 \eq \ifalse}}
+\newcommand{\neglit}[1]{\ensuremath{#1 \not\eq \itrue}}
 \newcommand{\poslit}[1]{\ensuremath{#1 \eq \itrue}}
-\newcommand{\skbci}{\textsf{SKBCI}}
-\def\negvthinspace{\kern-0.083333em}
-
-\newcommand{\cst}[1]{\ensuremath{\mathsf{#1}}}
-\newcommand{\var}[1]{{\mathit{#1}}}
-\newcommand{\typ}[1]{{\mathit{#1}}}
-\newcommand\foralltynospace[1]{\mathsf{\Pi}#1.}
-\newcommand\forallty[1]{\foralltynospace{#1}\;}
-\newcommand{\typeargs}[1]{{\negvthinspace\langle#1\rangle\negvthinspace}}
-\let\oldDelta=\Delta
-\let\oldSigma=\Sigma
-\renewcommand\Sigma{\mathrm{\oldSigma}}
-\newcommand{\Sigmaty}{\Sigma_\mathsf{ty}}
-\newcommand{\VV}{\mathscr{V}}
-\newcommand{\Vty}{\VV_\mathsf{ty}}
-
-% Equality predicate
-\newcommand{\eq}{\approx}
-\newcommand{\noteq}{\not\eq}
 \newcommand{\lland}{\mathrel\land}
@@
 
-\newcommand{\tuple}[2]{\overline{#1}_{#2}}
-\newcommand{\tuplen}[1]{\overline{#1}_{n}}
-
-\newcommand{\lam}[2]{\ensuremath{\lambda #1.\> #2}}
-\newcommand{\lamx}[1]{\lam{x}{#1}}
-
-% Names of inference rules
-\newcommand{\infname}[1]{\textsc{#1}}
-
-% Inference rule
-\newcommand{\namedinference}[3]{\prftree[r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
-\newcommand{\inference}[2]{\namedinference{}{\strut#1}{\strut#2}}
-
-% Simplification rule
-\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
-\newcommand{\simp}[2]{\namedinference{}{\strut#1}{\strut#2}}
-\newcommand{\lsupraw}{$\lambda$-super\-po\-si\-tion}
-\newcommand{\lsup}{{\color{black}\lsupraw}}
-\newcommand{\lsup}{{\color{black}Boolean-free \lsupraw}}
-\newcommand{\osup}{{\color{black}Boolean \lsupraw}}
-
-\newcommand\NumberOK[1]{#1}
-\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
-\newcommand{\ourpara}[1]{\paragraph{\upshape\bfseries#1}}
-
-\hyphenation{su-per-po-si-tion Zip-per-posi-tion tab-leau tab-leaux Bent-kamp ab-strac-tion
-  Bool-ean}
-
-\newcommand{\confrep}[2]{#2}
+\newcommand{\db}[1]{\ensuremath{\mathbf{#1}}}
+\newcommand{\dbvar}[1]{\ensuremath{\textbf{\textit{#1}}}}
+\newcommand{\internallam}{\ensuremath{\texttt{LAM}}}
+\newcommand{\internalat}{\ensuremath{\texttt{@}}}
+
+\newcommand{\lambdabf}{\pmb{\lambda}} %%% TYPESETTING
+\newcommand{\betabf}{\pmb{\beta}} %%% TYPESETTING
+\newcommand{\etabf}{\pmb{\eta}} %%% TYPESETTING
+
+\begin{document}
+
+\newcommand\CORR{\textsuperscript{(\vthinspace\Letter\vthinspace)}}
+
+\confrep
+{\title{Extending a High-Performance Prover to Higher-Order Logic}}
+{\title{Extending a High-Performance Prover to Higher-Order Logic \\ (Technical Report)}}
+%{\title{Extending a Brainiac Prover to \\ Full Higher-Order Logic}}
+%{\title{Extending a Brainiac Prover to \\ Full Higher-Order Logic\\(Technical Report)}}
+
+\author{Petar Vukmirovi\'c\inst{1}\CORR
+  \and
+  Jasmin Blanchette\inst{1,2}
+  \and
+  Stephan Schulz\inst{3}
+  }
+
+\institute{
+  Vrije Universiteit Amsterdam, Amsterdam, the Netherlands \\
+%  \email{p.vukmirovic@vu.nl}
+   \email{\{p.vukmirovic,j.c.blanchette\}@vu.nl}
+%   \and Max-Planck-Institut f\"ur Informatik, Saarland Informatics Campus, Saarbr\"ucken,~Germany \\
+%   \email{jasmin.blanchette@mpi-inf.mpg.de}
+  \and Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France \\
+    \email{jasmin.blanchette@loria.fr}
+  \and DHBW Stuttgart, Stuttgart, Germany \\
+    \email{stephan.schulz@dhbw-stuttgart.de}
+}
+
+\maketitle
+
+\begin{abstract}%
+  The automatic discharge of tedious subgoals is high on the wishlist of many
+  users of proof assistants. Some proof assistants discharge such goals
+  by translating them to first-order logic and invoking an efficient prover on
+  them, but much is lost in translation. As an alternative,
+  we propose to extend first-order provers with native support for
+  higher-order features. Building on our extension of E to $\lambda$-free
+  higher-order logic, we now extend E to full higher-order logic.
+  The resulting prover is the second strongest on both TPTP benchmarks and
+  benchmarks coming from a proof assistant.
+  %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
+\end{abstract}
+
+\setcounter{footnote}{0}
 
 \section{Introduction}
-\label{sec:satfol:ho-tech:intro}
-
-%In recent decades,
-Superposition-based first-order automatic theorem provers
-have emerged as useful reasoning tools. They dominate at the annual CASC
-\cite{gs-2016-casc} theorem prover competition, having always won the
-first-order theorem division. They are also used as backends to proof assistants
-\cite{ck-18-coqhammer,ku-15-holyhammer,pb-12-sh}, automatic
-higher-order theorem provers \cite{sb-21-leo3}, and software verifiers
-\cite{fp-13-why3}.
-
-The superposition calculus has only recently been extended
-to higher-order logic (more precisely, extensional simple type theory
-\cite{henkin-1950-completeness}), resulting in
-\emph{\lsup} \cite{bentkamp-et-al-2021-lamsup-journal}, which we developed
-together with Waldmann, as well as \emph{combinatory superposition}
-\cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
-calculi do not support an interpreted Boolean type,
-they can be extended by ad hoc rules \cite{our-bool-paper} that support
-most of the Boolean reasoning necessary in practice.
-
-Both higher-order superposition calculi were designed to gracefully
-extend first-order reasoning. As most steps in higher-order
-proofs tend to be essentially first-order, extending the most successful first-order
-calculus to higher-order logic seemed worth trying.
-Our first attempt at testing this idea was in 2019:
-Zipperposition~1.5, based on \lsup, finished third
-in the higher-order theorem division of CASC-27 \cite{gs-19-casc27},
-12~percentage points behind the winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.
-
-Studying the competition results, we found that higher-order tableaux have
-some advantages over higher-order superposition. To bridge the gap, we developed
-techniques and heuristics that simulate tableaux in the
-context of saturation. We implemented them in Zipperposition~2, which took part
-in CASC-J10 \cite{gs-21-cascj10} in 2020. This time, our prover won the division, proving 84\% of
-the problems, a whole 20~percentage points ahead of the runner-up, Satallax
-3.4.
-
-In this article, we describe the main techniques that explain this reversal
-of fortunes. They cover most parts of a modern higher-order theorem prover, from
-preprocessing to additional calculus rules to heuristics to backend
-integration. We use a newer version of Zipperposition, based on a newer
-calculus:\ Instead of \lsup{} augmented with ad hoc Boolean rules,
-we work with \emph{\osup} \cite{bbtv-21-full-ho-sup}, a principled extension of superposition to full
-higher-order logic, including an interpreted Boolean type.
-
-Many higher-order problems extensively use symbol definitions to simplify
-their representation. We describe several ways to exploit the definitions,
-%of which the most successful is
-such as turning them into rewrite rules (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
-%Interesting patterns can be observed in various higher-order problem encodings.
-%We show how we can exploit these to simplify problems (Sect.~\ref{sec:satfol:ho-tech:preprocessing}).
+\label{sec:introduction}
+
+In the last few decades, proof assistants have become indispensible tools for
+developing trustworthy formal proofs. They are used both in academia to verify
+mathematical theories \cite{th-05-flyspeck} and in
+industry to verify the correctness of hardware \cite{ckmg-99-hw-verification}
+and software \cite{xl-09-compcert, gk-09-sel4, rgea-16-certikos}.
 %
-By working on formulas rather than clauses, tableau techniques take a more
-holistic view of a higher-order problem.
-Through its support for delayed clausification and, more generally,
-calculus-level formula manipulation, \osup{} enables us to
-simulate most successful tableau techniques in a saturating prover
-(Sect.~\ref{sec:satfol:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
-functions}, a mechanism that allows us to choose on which Boolean subterms
-to perform inferences first.
-We implemented some Boolean selection functions and
-evaluated them (Sect.~\ref{sec:satfol:ho-tech:bool-select}).
-
-The main drawback of both \lsup{} variants compared with combinatory
-superposition is that they rely on rules that enumerate possibly infinite sets
-of unifiers. We describe a mechanism that interleaves infinitely
-branching inferences with the standard saturation process
-(Sect.~\ref{sec:satfol:ho-tech:infinite-branching}). The prover
-retains the same behavior as
-before on first-order problems, smoothly scaling with
-increasing numbers of higher-order clauses.
+However, due to the lack of strong built-in proof automation, proving
+seemingly simple goals can be a tedious manual task. To mitigate
+this, many proof assistants include a subsystem such as CoqHammer,
+HOL(y)Hammer, or Sledgehammer \cite{bkpu-16-hammering} that translates
+higher-order goals to first-order logic and passes them to efficient
+first-order automatic provers. If a first-order prover succeeds, the proof is
+reconstructed and the goal is closed.
+
+Unfortunately, the translation of higher-order constructs is clumsy and leads
+to poor performance on goals that require higher-order reasoning.
+Using native higher-order provers
+such as %Leo-III \cite{sb-21-leo3} or   --- the reference is about Leo-II and Satallax --JB
+Satallax \cite{cb-12-satallax} as backends
+is not necessarily a solution because they are much less optimized than their
+first-order counterparts \cite{ns-13-leo2sh}. To bridge this gap, in 2016
+we proposed to develop a new generation of higher-order provers that
+%are not built from the ground up, but rather
+extend the arguably most successful first-order calculus, superposition, to
+higher-order logic, starting from a position of strength.
+
+Our research has focused on three milestones:\ supporting $\lambda$-free
+higher-order logic, adding support for $\lambda$-terms, and adding support for
+first-class Boolean terms. In 2019, we extended the state-of-the-art first-order
+prover E \cite{scv-19-e23} with a $\lambda$-free superposition calculus
+\cite{section-ehoh}, obtaining a version of E called Ehoh, as a stepping stone
+towards full higher-order logic. Together with Bentkamp, Tourret, and Waldmann,
+we have since developed calculi, called \emph{$\lambda$-superposition}, corresponding
+to the other two milestones \cite{bbtvw-21-sup-lam,bbtv-21-full-ho-sup} and
+implemented them in the experimental superposition prover Zipperposition
+\cite{sc-15-simon-phd}. This prover is implemented in OCaml and not nearly as
+efficient as E. Nevertheless, it has won the higher-order division of the CASC
+theorem prover competition \cite{gs-21-cascj10} in 2020 and 2021 by a wide
+margin, ending nearly a decade of Satallax domination.
+
+In this \paper, we fulfill a three-year-old promise: We present the extension of
+Ehoh to full higher-order logic (Sect.~\ref{sec:logic}) using incomplete variants
+of $\lambda$-superposition. We call this prover \ehohii.
 %
-We also propose some heuristics to curb the explosion induced by highly
-prolific calculus rules (Sect.~\ref{sec:satfol:ho-tech:explosiveness}).
-
-Using first-order backends to finish the proof is common practice in
-higher-order reasoning. Since \lsup{} coincides with standard
-superposition on first-order clauses, invoking backends may
-seem redundant; yet Zipperposition is nowhere as efficient as E
-\cite{scv-19-e23} or Vampire \cite{lkav-13-vampire}, so invoking a more
-efficient backend does make sense. We describe how to achieve a balance
-between allowing native higher-order reasoning and
-delegating reasoning to a backend (Sect.~\ref{sec:satfol:ho-tech:backends}).
-%
-Finally, we compare Zipperposition~2 with other provers on all monomorphic
-higher-order TPTP benchmarks \cite{gs-17-tptp} to perform a more extensive
-evaluation than at CASC (Sect.~\ref{sec:satfol:ho-tech:comparison}). Our evaluation
-corroborates the competition results.
-
-This article is an extended version of a paper accepted at CADE-28
-\cite{making-ho-work}. Compared with the conference paper, it
-describes a new preprocessing technique, explores the effects of Boolean
-selection functions, evaluates %some previously unevaluated
-more techniques,
-introduces new benchmark sets, and presents more examples.
-
-\section{Background and Setting}
-\label{sec:satfol:ho-tech:background}
-
-We focus on monomorphic higher-order logic, without the axiom of infinity 
-or the axiom of at least two individuals. However, the techniques can easily be
-extended with polymorphism. Indeed, Zipperposition already supports some
-of them polymorphically.
-
-\ourpara{Higher-Order Logic}
-We define terms $s, t, u, v$ inductively as free variables $F, X$, bound
-variables $x, y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak
-\cst{a},\allowbreak \cst{b}, \dotsc$, term applications $s \, t$, and
-$\lambda$-abstractions $\lamx{s}$. The syntactic distinction between free and
-bound variables yields \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
-\cst{a}}$) \cite{tn-93-patterns}.
+The $\lambda$-superposition calculi were
+previously implemented in Zipperposition, and
+extensive experiments with various heuristic choices have been performed
+\cite{section-making-ho-work}. In \ehohii{}'s implementation, we used
+these experiences to choose a set of effective rules that could easily be
+retrofitted into an originally first-order prover. Another principle that guided 
+the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
+do not impact already strong first-order performance of Ehoh.
+
+An simpler approach to support higher-order logic is to implement
+combinatory superposition, which is based on \textsf{SKBCI}-combinators
+and implemented in Vampire \cite{br-20-full-sup-w-combs}.
+These can be supported as a thin layer on top of $\lambda$-free terms. However,
+in unpublished experiments, we found that $\lambda$-superposition performs
+substantially better, corroborating the CASC results, so we
+decided to make a more profound change to Ehoh and implement
+$\lambda$-superposition.
+
+% We
+% also used the experience of fine-tuning the calculi
+% \cite{section-making-ho-work} in Zipperposition to carefully choose which
+% calculus extensions and heuristics to use in E.
+
+One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
+term representation (Sect.~\ref{sec:terms}). Furthermore, Ehoh's main inference
+engine was designed with the assumption that it will be used with
+inferences that compute a most general unifier. We
+implementd a higher-order unification procedure \cite{unif-section}
+that can return multiple unifiers (Sect.~\ref{sec:unif-match-index}) and
+integrated it in the inference engine. Finally, we extended and adapted the
+superposition rule, resulting in an incomplete, pragmatic variant of
+$\lambda$-superposition (Sect.~\ref{sec:calculus}).
+
+% most likely text
+We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
+as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
+(Sect.~\ref{sec:eval}). We found
+that \ehohii{} clearly outperforms Ehoh on all benchmarks. Compared
+with other higher-order provers, it is second
+only to the cooperative version of Zipperposition, which employs Ehoh as a
+backend. A fairer comparison without the backend puts \ehohii{} in the
+first place for both benchmark suites.
+We also compared the performance of \ehohii{} with E on first-order
+%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
+%%% check if you agree. --JB
+problems and found that little overhead has been introduced by the
+extension to higher-order logic.
+
+\kern-1pt %% TYPESETTING hack
+
+\section{Logic} %---0.5p
+\label{sec:logic}
+
+% TO BE CHANGED AS NECESSARY
+Our target logic is monomorphic classical higher-order logic with Hilbert choice.
+The following text is partly based on Section 2 of Vukmirovi\'c et al.~\cite{section-making-ho-work}.
+
+Terms $s, t, u, v$ are inductively defined as free variables $F, X, \ldots$, bound variables $x,
+y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b},
+\dotsc$, applications $s \, t$, and $\lambda$-abstractions $\lamx{s}$.
+The syntactic distinction between free and bound variables
+gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
+\cite{tn-93-patterns}.
 %We shorten iterated application $s \, t_1 \, \cdots \, t_n$ to $s \,
@@
 %x_n. \> s$ to $\lam{\overline{x}_n}{s}$.
+
 We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
-$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. The $n$-fold application of
-a unary term $s$ to a term $t$ is denoted by $s^n \, t$. Every
-$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
+$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$.
+Every $\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
 \overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
-of the term. If the type of a term $t$ is of the form $\tau_1 \to \cdots \to
+of the term. If $s$ is a free variable, we call the term \emph{flex};
+otherwise, the term is \emph{rigid}. If the type of a term is of the form $\tau_1 \to \cdots \to
 \tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
-call $t$ a \emph{predicate}. A term of type $o$ is called a \emph{formula}.
-
-A literal $l$ is an equation $s \eq t$ or a disequation $s \not\eq t$. A clause is
-a finite multiset of literals, interpreted and written disjunctively $l_1 \llor
-\cdots \llor l_n$. Logical symbols that may occur within terms are written in
-boldface: $\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Quantified
-formulas are expressed using (a type-indexed
-family of) constants $\iforall$ and $\iexists$ as $\iforall \,
-(\lambda x. \, t)$ and $\iexists \, (\lambda x. \, t)$, usually abbreviated to
-$\iforall x.\,t$ and $\iexists x.\,t$. Following %the convention of
-\osup{}, predicate literals are encoded as equations with $\itrue$ or
-$\ifalse$: for example, $\cst{even}(x)$ becomes $\poslit{\cst{even}(x)}$, and
-$\neg\,\cst{even}(x)$ becomes $\neglit{\cst{even}(x)}$.
-
-\ourpara{Higher-Order Calculi}
+call it a \emph{predicate}. A literal $l$ is an equation $s \eq t$ or
+a disequation $s \not\eq t$. A clause is a finite multiset of literals,
+interpreted and written disjunctively:\ $l_1 \llor \cdots \llor l_n$.
+Logical symbols that may occur within terms are written in bold:
+$\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Predicate
+literals are encoded as (dis)equations with $\itrue$ based on their sign; for
+example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
+$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.
+
+\section{Terms}
+\label{sec:terms}
+
+Ehoh's term representation is centered around perfect term sharing %or exa
+\cite{ls-01-shared}: Any two structurally identical terms are
+guaranteed to be the same object in memory. This is achieved through term
+\emph{cells}, which represent individual terms. Each cell has (among other fields)
+(1)~\texttt{f\_code}, an integer corresponding to the symbol at the head of the term (negative
+if the head is a free variable, positive otherwise); (2)~\texttt{num\_args},
+corresponding to the number of arguments applied to the head; and (3)~\texttt{args},
+a size-\texttt{num\_args} array of pointers to argument terms. We
+use the first-order notation $\cst{f}(s_1, \ldots, s_n)$ to denote a cell whose
+\texttt{f\_code} corresponds to $\cst{f}$, \texttt{num\_args} equals $n$, and
+\texttt{args} points to the cells for $s_1, \ldots s_n$.
+
+Ehoh represents $\lambda$-free higher-order terms using flattened, spine notation.
+Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$ and $\cst{f} \, \cst{a} \,
+\cst{b}$ are represented by the cells $\cst{f}$,
+$\cst{f}(\cst{a})$, and $\cst{f}(\cst{a}, \cst{b})$, respectively.
+To ensure free variables are
+perfectly shared, Ehoh treats applied free variables differently: arguments are
+not applied directly to a free variable, but using an internal symbol
+\internalat{} of variable arity. For example, the term $X \, \cst{a} \, \cst{b}$ is
+represented by the cell $\internalat(X, \cst{a}, \cst{b})$. This ensures that
+two different occurrences of the free variable $X$ correspond to the same object,
+which makes substitutions more efficient \cite{section-ehoh}.
+%%% @PETAR(read): This tries to explain an invisible problem. Either we need to explain
+%%% it better, or we just omit it (my favorite option). It's old Ehoh stuff anyway. --JB
+%Yet, to
+%efficiently access variable arguments we do not use binary application symbol,
+%but flattened representation.
+
+\ourpara{Representation of $\lambdabf$-Terms} To support the calculus for full
+higher-order logic \cite{bbtv-21-full-ho-sup}, the cell data structure must
+be extended to support the $\lambda$ binder. We use the locally nameless
+representation \cite{ac-12-locally-nameless} for this purpose: De Bruijn indices
+%%% @PETAR(checked): Double-check "possibly loose".
+represent (possibly loose) bound variables, whereas we keep the current
+representation for free variables.
+
+%%% @PETAR(read): Sounds odd---we wouldn't be extending the term representation of an
+%%% already HO prover. --JB
+%Unlike other higher-order provers, most of which are written in functional
+%programming languages,
+Extending the term representation of Ehoh with a new term
+kind involves intricate manipulation of the cell data structure. De Bruijn
+indices must be represented as other cells with either negative or positive
+\texttt{f\_code}. However, this has to be done in such a way that De Bruijn
+index can never be mistaken for a function symbol or a variable.
+
+Other than possibly being instantiated during $\beta$-reduction, De Bruijn
+indices mostly behave as constants. Therefore, we decided to represent De
+Bruijn indices using positive \texttt{f\_code}s: De Bruijn index of value $i$
+will have $i$ as the \texttt{f\_code}. To ensure De Bruijn indices are not
+mistaken for function symbols, we use the \texttt{properties} bitfield of the
+cell, which holds precomputed properties of %shared cell
+%%% @PETAR(checked, you are right): I don't understand why you wrote "shared cell". Please double-check. --JB
+the cell. We introduce the
+property \texttt{IsDBVar} to denote that the cell represents a De
+Bruijn index. Any attempt to create a De Bruijn index is performed through
+a dedicated library function that sets \texttt{IsDBVar} property for every term it
+returns. When given the same De Bruijn index and type, this function is
+guaranteed to always return the same object. Finally, we have guarded all the
+functions and macros that manipulate function codes with the check if the
+property \texttt{IsDBVar} is set. To ensure perfect sharing of every occurrence
+of De Bruijn indices, arguments to De Bruijn indices are applied like for free
+variables, using \internalat{}.
+
+Extending cells to support $\lambda$-abstraction is easier. Each
+$\lambda$-ab\-strac\-tion has the distinguished function code \internallam{} as the head
+symbol and two arguments:\ (1)~a De Bruijn index 0 of the type of abstracted variable;
+(2)~the body %(matrix)
+of the $\lambda$-abstraction. Consider
+the term $\lambda x. \, \lambda y.\, \cst{f}\, x \, x$, where both $x$ and $y$ have
+the type~$\iota$. This term is represented as $\lambda\,\lambda\, \cst{f} \,
+\db{1} \, \db{1}$ in locally nameless representation, where bold numbers
+represent De Bruijn indices. In \ehohii{}, the same term is represented by the cell
+$\internallam(\db{0}, \internallam(\db{0}, \cst{f}(\db{1}, \db{1})))$,
+where all De Bruijn variables have type~$\iota$.
+
+The first argument of $\internallam$ is
+redundant, since it can be deduced from the type of the $\lambda$-abstraction.
+However, basic $\lambda$-term manipulation operations often require access to
+this term. We store it explicitly to avoid creating it repeatedly.
+
+\ourpara{Efficient $\betabf$-Reduction}
+Terms are stored in $\beta\eta$-reduced form. As these two reductions are
+performed very often, they ought to be efficient. Ehoh
+performs $\beta$-reduction by reducing the leftmost outermost $\beta$-redex
+first. To represent $\beta$-redexes, E uses the \internalat{} symbol. Thus,
+the term $(\lambda
+x.\, \lambda y.\,  (x \, y)) \, \cst{f} \, \cst{a}$ is represented by
+$\internalat(\internallam(\db{0}, \internallam(\db{0}, \internalat(\db{1}, \db{0}))),
+\cst{f}, \cst{a})$. Another option would have been to add arguments applied to
+$\lambda$-term directly to the $\lambda$ representation (as in
+$\internallam(\db{0}, \internallam(\db{0}, \allowbreak \internalat(\db{1}, \db{0})),
+\cst{f}, \cst{a})$), but this would break the invariant
+that $\internallam$ has two arguments. Furthermore, replacing free
+variables with $\lambda$-abstractions (e.g., replacing $X$ with $\lambda
+x. \, x$ in $\internalat(X, \cst{a})$) would require additional normalization.
+
+A term can be $\beta$-reduced as follows: When a cell of the
+form $ \internalat(\internallam(\db{0}, s),t)$ is encountered, the field
+\texttt{binding} (normally used to record the substitution for a free variable) of the
+cell $\db{0}$ is set to $t$. Then $s$ is traversed to instantiate every
+loose occurrence of $\db{0}$ in $s$ with \texttt{binding}, whose loosely
+bound De Bruijn indices are shifted by the number of $\lambda$ binders above
+the occurrence of $\db{0}$ in $s$ \cite{fk-01-deBruijn}. Next, the same
+procedure is performed on the resulting term and its subterms, in
+leftmost outermost fashion.
+
+\ehohii{}'s basic $\beta$-normalization mechanism works in this way, but it
+features a few optimizations.
+First, \ehohii{} recognizes terms of the form $(\lam{\overline{x}_n}{s}) \,\overline{t}_n$ 
+and performs parallel replacement of the bound variables $x_i$ with
+$t_i$. Since intermediate terms are not constructed, this reduces the number of
+recursive function calls and calls to the cell allocator.
+
+Second, in line with the gracefulness principle, we want \ehohii{} to incur
+little (if any) overhead on first-order problems and to
+excel on higher-order problems with a large first-order component. If
+$\beta$-reduction is implemented naively, finding a $\beta$-redex involves
+traversing the entire term. On purely first-order terms, $\beta$-reduction
+is then a complete waste of time.
+
+To avoid this, we use Ehoh's perfectly shared
+terms and their \texttt{properties} field.
+%
+We introduce the property \texttt{HasBetaReducibleSubterm} that indicates whether
+%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
+a cell is $\beta$-reducible.
+Whenever a new cell that contains a
+$\beta$-reducible term as a direct subterm is shared, the property is set.
+Setting of the property is inductively continued when further superterms are
+shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
+x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
+$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
+\texttt{HasBetaReducibleSubterm} set.
+%
+When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
+cells with this property set. This further means that on first-order
+subterms, a single bit masking operation is enough to determine that no subterm
+should be visited.
+
+Along similar lines, we added a property \texttt{HasDBSubterm} that
+caches whether the cell contains a De Bruijn subterm. This
+%% @PETAR(you are right): Why "also"? What's the other thing?
+%also
+makes
+instantiating De Bruijn indices during $\beta$-normalization faster, since only the
+subterms that contain De Bruijn indices must be visited. Similarly, some other
+operations such as shifting De Bruijn indices or determining whether a term is closed
+(i.e., it contains no loose bound variables) can be sped up or even avoided
+if the term is first-order.
+
+\ourpara{Efficient $\etabf$-Reduction}
+The term $\lambda x.\, s \, x$ is $\eta$-reduced to $s$ whenever $x$ does not appear
+unbound in $s$. Caching the property of $\eta$-reducibility of the term is not
+as beneficial as the one for $\beta$-reducibility, because checking this
+property at the term's top level is done in $O(|s|)$, compared with constant time
+for $\beta$-reducibility.
+However, we use the observation that a term cannot be $\eta$-reduced if it has
+no $\lambda$-abstraction subterms and introduce a property \texttt{HasLambda}
+that notes the presence of $\lambda$-abstraction in a term. Only
+terms with this property are visited during $\eta$-reduction.
+
+\ehohii{} performs parallel $\eta$-reduction: It recognizes terms of the form
+$\lam{\overline{x}_n}{s \, \overline{x}_n } $ such that none of $x_i$
+occurs unbound in $s$. If done naively, reducing terms of this kind requires up to $n$
+traversals of $s$ to check if each $x_i$ occurs in $s$. In \ehohii{}, exactly one
+traversal of $s$ is required.
+
 \looseness=-1
-The \osup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
-complete inference system and redundancy criterion for higher-order logic with
-rank-1 polymorphism, Hilbert choice, and functional and Boolean extensionality.
-The calculus relies on
-\emph{complete sets of unifiers}
-(\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
-$V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
-that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
-\in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
-for all variables $X \in V$. The set $V$ is used to distinguish
-important variables from auxiliary variables (which may arise in intermediary
-states of the unification procedure). We usually omit it.
-
-Unlike \lsup, this calculus 
-does not require axioms defining the logical symbols to cope with formulas.
-Instead, it includes Boolean inference rules that mimic
-superposition from such axioms into Boolean subterms,
-while avoiding the explosion incurred by adding these axioms to the proof state. It
-also includes rules that simulate Boolean inferences below applied variables.
-Both sets of rules are disabled or replaced with incomplete, ad hoc rules
-described by Vukmirović and Nummelin \cite{our-bool-paper} in most configurations
-of the CASC portfolio.
-A new feature of the calculus that we explore in detail is
-the ability to select Boolean subterms
-to restrict Boolean and superposition inferences.
-
-In contrast to both \lsup{} variants, combinatory superposition can
-avoid enumerating CSUs by
-using a form of first-order unification.
-Essentially, it enumerates higher-order terms
-using rules that instantiate applied variables with partially applied
-combinators from the complete combinator set $\{\cst{S}, \cst{K}, \cst{B},
-\cst{C}, \cst{I}\}$. This calculus is the basis of Vampire~4.5
-\cite{br-20-full-sup-w-combs}, which finished
-closely behind Satallax 3.4 %~and~3.5
-at CASC-J10.
+More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
+\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
+the cell will be of the form
+$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
+where $t$ is not a
+$\lambda$-abstraction. Then \ehohii{} breaks down the body~$t$ into a maximal
+decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
+If $n = 0$, the cell is not $\eta$-reducible.
+Otherwise, $u$ is traversed to determine the
+minimal index~$\dbvar{j}$ of a loose De Bruijn index,
+taking $\dbvar{j} = \infty$ if no such index exists.
+%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
+%bottom-up as innermost can suggest): Check "innermost".
+%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
+%%% binder remains (the one to which the 2's point). --JB
+We can then remove the $k = \min(j,n)$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
+\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
+$t$ by the variant of
+$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
+obtained by shifting the loose De Bruijn indices down by~$k$.
+
+To illustrate this convoluted De Bruijn arithmetic, we consider the
+term $\lambda x. \, \lambda y. \, \lambda z.\, \cst{f} \, x \, x \, y
+\, z$. This term is represented by the cell $\internallam(\db{0},
+\internallam(\db{0}, \internallam(\db{0},\allowbreak \cst{f}(\db{2}, \db{2}, \db{1},
+\db{0}))))$. \ehohii{} splits $\cst{f}(\db{2}, \db{2}, \db{1}, \db{0})$ into
+two parts:\ $u = \cst{f} \, \db{2}$ and the arguments $\db{2}, \db{1},
+\db{0}$. Since the minimal index in $u$ is $\db{2}$, we can
+omit the De Bruijn indices $\db{1}$ and $\db{0}$ and their $\lambda$ binders,
+yielding the $\eta$-reduced cell $\internallam(\db{0}, \cst{f}(\db{0},
+\db{0}))$.
+
+The use of the \texttt{HasLambda} property ensures that $\eta$-reduction is not
+tried on first-order or $\lambda$-free higher-order terms, whereas parallel
+$\eta$-reduction both speeds up $\eta$-reduction and avoids creating a linear
+number of intermediate terms. For finding the minimal loose De Bruijn index,
+optimizations such as the \texttt{HasDBSub\-term} property are used.
+
+\ourpara{Representation of Boolean Terms}
+Ehoh represents Boolean terms using
+cells whose \texttt{f\_code}s correspond to internal codes reserved
+for logical symbols. Quantified formulas are represented by cells in which the
+first argument is the quantified variable and the second one is the body of the
+quantified formula. For example, the term $\iforall x.\, \cst{p} \, x$ corresponds
+to the cell $\iforall(X, \cst{p}(X))$, where $X$ is a regular free variable.
+% as Ehoh has no concept of a bound variable.
+
+This representation is convenient for parsing formulas and clausification, which
+is what Ehoh uses it for, but it
+causes $\alpha$-normalization issues during the actual proof search: In
+full higher-order logic, Booleans can appear as subterms in clauses, as in
+$\cst{q}(X) \llor \cst{p}(\iforall(X, \cst{r}(X)))$;
+instantiating $X$ in the first literal should not influence $X$ in the second
+literal.
+
+To avoid this issue, we use $\lambda$ binders to represent quantified formulas.
+Thus, $\iforall x. \, s$ is represented by $\iforall \, (\lambda x. \, s)$.
+Quantifiers are then unary symbols that do not directly bind
+the variables but delegate this responsibility to a $\lambda$-abstraction.
+%
+Since \ehohii{} represents bound variables using De Bruijn indices, this solves the
+$\alpha$-conversion issues. However, this solution is incompatible with
+thousands of decades-old lines of clausification code that assumes the Ehoh
+representation of
+quantified formulas. Therefore, \ehohii{} converts quantified
+formulas only after clausification, for Boolean terms that appear in a
+higher-order context (e.g., as argument to a function symbol).
+
+\ourpara{New Term Orders}
+The $\lambda$-superposition calculus is parameterized by a term order that is
+used to break symmetries in the search space.
+We implemented the versions of the Knuth--Bendix order (KBO) and lexicographic
+path order (LPO) for higher-order terms with $\lambda$-abstractions described by
+Bentkamp et al.~\cite{bbtv-21-full-ho-sup}. These orders encode
+encode $\lambda$-terms as first-order terms and then invoke the standard KBO or
+LPO. %Trading elegance for efficiency, our
+%implementation does not compute the first-order translation.
+%Rather,
+For efficiency, we implemented separate KBO and
+LPO functions that compute the order directly, intertwining the encoding and
+the order computation.
+%%% @PETAR(yes): Irrelevant detail?
+%We kept the
+%implementation of $\lambda$-free KBO as \ehohii{} can be run in $\lambda$-free mode.
+
+Ehoh cells contain a \verb|binding| field that can be used to store the
+substitution for a free variable. Substitutions can then be applied by following
+the \texttt{binding} pointers, replacing each free variable with its instance.
+Thus, when Ehoh needs to perform a KBO or LPO comparison of an instantiated term,
+it needs only follow the \texttt{binding} pointers.
+In full higher-order logic, however, instantiating a variable can trigger a
+series of $\beta\eta$-reductions,
+changing the shape of the term dramatically. To pevent this, \ehohii{}
+computes the $\beta\eta$-reduced instance of the terms
+before comparing them using KBO or LPO.
+
+\section{Unification, Matching, and Term Indexing}
+\label{sec:unif-match-index}
+% \begin{itemize}
+%   \item Procedure that implements a variant of the algo described in Efficient
+%   HO Unif paper.
+%   \item Two oracles -- fixpoint and pattern unif (solid not implemented as
+%   complexity/gain ratio is quite negative)
+%   \item It is implemented iteratively, with subtle optimizations
+%   \item Its iterator is completely contained -- iterators can be stored and
+%   later waked up similarly to what we do in Zipperposition.
+%   \item Mention that so far the prover fares extremely well with complicated
+%   unification problems and that we do not consider implementing more complex
+%   given clause algo in Ehoh.
+%   \item Matching works only with algorithms that return a single matcher:
+%   pattern matching and LFHOL matching : main optimization from E is preserved.
+%   PDT algorithm also works with both kinds of terms LFHOL and pattern. Check if
+%   something is a pattern is efficient!
+%   \item Fingerprints and feature vectors as in Zipperposition.
+%   \item To avoid huge refactorings iterator is used even if the problem to be
+%   solved is first-order.  
+% \end{itemize}
+
+Standard superposition crucially depends on the concept of a most general unifier
+(MGU).
+In higher-order logic, such a unifier does not in general exist, and the concept
+%%% @PETAR: CSU, infinite
+is replaced by that of a complete set of unifiers (CSU), which may be infinite.
+Vukmirovi\'c et al.\ \cite{unif-section} designed an efficient
+procedure to enumerate a CSU for a term pair. They
+implemented it in Zipperposition, together with some extensions to term indexing.
+In \ehohii{}, we further improve the performance of this procedure by implementing a terminating,
+incomplete variant. We also introduce a new indexing data structure.
+
+\ourpara{The Unification Procedure} 
+% Two main guiding principles behind our
+% unification procedure are laziness and redundancy elimination. The first one is
+% embodied in the fact that the procedure makes sure that it never fully
+% normalizes the terms or fully applies the substitution if that is not necessary.
+% Redundancy elimination is achieved through \emph{oracles}
+% \cite{unif-section}: the procedures that solve the unification
+% problem for a subclass of higher-order terms on which unification is decidable,
+% and for \ehohii{}, unary. If those procedures determine that the given terms
+% belong to their class of terms, they will return the most general unifier,
+% avoiding redundant computations and in some cases even non-termination.
+The unification procedure works by maintaining a list of unification pairs to be solved.
+After choosing a pair, it first normalizes it by $\beta$-reducing and
+instantiating the heads of both terms in the pair. Then, if either head is a
+variable, it computes an appropriate binding for this variable, thereby
+approximating the solution.
+
+Unlike in first-order and $\lambda$-free higher-order unification,
+in full higher-order case there may be many bindings that lead to a solution. To
+reduce this mostly blind guessing of bindings, the procedure features support
+for \emph{oracles} \cite{unif-section}. These are procedures that
+solve the unification problem for a subclass of higher-order terms on which
+unification is decidable and, for \ehohii{}, unary. Oracles help
+increase performance, avoid nontermination, and avoid redundant bindings.
+
+Vukmirovi\'c et al.\ described their procedure as a transition system. In
+\ehohii{}, the procedure is implemented nonrecursively, and the unifiers are
+enumerated using an iterator object that encapsulates the state of the unifier
+search. The iterator consists of five fields:
+(1)~\emph{constraints}, which holds the unification
+constraints;
+(2)~\emph{bt\_state}, a stack that contains information necessary to backtrack
+to a previous state;
+(3)~\emph{branch\_iter}, which stores how far we
+are in exploring different possibilities from the current search node;
+(4)~\emph{steps}, which remembers how many different unification bindings (such as
+imitation, projection, and identification) are applied; and
+(5)~\emph{subst},
+a stack storing the variables bound so far.
+
+\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
+\newcommand{\cn}[1]{\ensuremath{\textsc{#1}}} %const name
+\newcommand{\fc}[1]{\ensuremath{\textsc{#1}}} %function call
+\algrenewcommand\alglinenumber[1]{{\color{gray} \tiny #1}}
+
+The iterator is initialized to hold the original problem in \vn{constraints},
+and all other fields are initially empty. The unifiers are retrieved one by one by
+calling the function \fc{ForwardIter}. It returns \fc{True} if the iterator made
+progress, in which case the unifier can be read via the iterator's
+$\vn{subst}$ field. Otherwise, no more unifiers can be found, and the iterator is
+no longer valid. The function's pseudocode is given below, including two
+auxiliary functions:
+
+\begin{algorithmic}[]
+  \Function{NormalizeHead}{$\vn{t}$}
+    \If{$\vn{t.head} = \internalat \land \vn{t.args}[0].\vn{is\_lambda}()$}
+      \State reduce the top-level $\beta$-redex in $t$
+      \State \Return $\fc{NormalizeHead}(t)$
+    \ElsIf{$\vn{t.head.is\_var()} \land \vn{t.head.binding} \not= \cn{nil}$}
+      \State $\vn{t.head} \gets \vn{t.head.binding}$
+      \State \Return $\fc{NormalizeHead}(t)$
+    \Else{}
+      \State \Return $t$
+    \EndIf
+  \EndFunction
+
+  \vspace{\jot}
+  \Function{BacktrackIter}{$\vn{iter}$}
+    \If{\vn{iter.bt\_state.empty()}}
+      \State clear all fields in \vn{iter}
+      \State \Return \cn{False}
+    \Else
+    \State pop $(\vn{constraints}, \vn{branch\_iter}, \vn{steps}, \vn{subst})$ from $\vn{iter.bt\_state}$
+    \State set the corresponding fields of $\vn{iter}$
+    \State \Return \cn{True}
+    \EndIf
+  \EndFunction
+  \vspace{\jot}
+  \Function{ForwardIter}{$\vn{iter}$}
+  \State $\vn{forward} \gets \neg \vn{iter.constraints.empty()} \lor \fc{BacktrackIter}(\vn{iter})$
+  \While{$\vn{forward} \land \neg \vn{iter.constraints.empty()}$}
+    \State $(\vn{lhs}, \vn{rhs}) \gets $ pop pair from $\vn{iter.constraints}$
+    \State $\vn{lhs} \gets \fc{NormalizeHead}(\vn{lhs})$
+    \State $\vn{rhs} \gets \fc{NormalizeHead}(\vn{rhs})$
+    \State normalize and discard the $\lambda$ prefix of $\vn{lhs}$ and $\vn{rhs}$
+
+    \If{$\neg\vn{lhs.head.is\_var()} \land \vn{rhs.head.is\_var()}$}
+      \State swap $\vn{lhs}$ and $\vn{rhs}$
+    \EndIf
+
+    \If{$\vn{lhs.head.is\_var()}$}
+      \State $\vn{oracle\_res} \gets \fc{Fixpoint}(\vn{lhs, rhs, iter.subst}) $
+      \If{$\vn{oracle\_res} = \cn{NotInFragment}$}
+        \State $\vn{oracle\_res} \gets \fc{Pattern}(\vn{lhs, rhs, iter.subst}) $
+      \EndIf
+
+      \If{$\vn{oracle\_res} = \cn{NotUnifiable}$}
+        \State $\vn{forward} \gets \fc{BacktrackIter(iter)}$
+      \ElsIf{$\vn{oracle\_res} = \cn{NotInFragment}$}
+        \State $\vn{n\_steps}, \vn{n\_branch\_iter}, \vn{n\_binding} \gets$
+        \State $\quad \fc{NextBinding}(\vn{lhs}, \vn{rhs}, \vn{iter.steps}, \vn{iter.branch\_iter})$
+
+        \If{$\vn{n\_branch\_iter} \not= \fc{BindEnd}$ }
+          \State push pair (lhs,rhs) back to \vn{iter.constraints}
+          \State push quadruple $(\vn{iter.constraints}, \vn{n\_branch\_iter}, $
+          \State                  $\quad \vn{iter.steps}, \vn{iter.subst})$ onto $\vn{iter.bt\_state}$
+          \State extend $\vn{iter.subst}$ with $n\_binding$
+          \State $\vn{iter.steps} \gets n\_steps$
+          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
+        \ElsIf{\vn{lhs.head = rhs.head}}
+          \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+          \State $\quad$ and push them to \vn{iter.constrants}
+          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
+        \EndIf
+      \EndIf
+    \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
+      \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+      \State $\quad$ and push them to \vn{iter.constrants}
+    \Else{} $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
+    \EndIf
+  \EndWhile
+  \State \Return $\vn{forward}$
+  \EndFunction
+\end{algorithmic}
 
 \looseness=-1
-A different, very successful calculus is Satallax's SAT-guided tableaux
-\cite{backes-brown-2011}. Satallax was the leading higher-order prover of the
-2010s. Its simple and elegant tableaux avoid deep superposition-style rewriting
-inferences.
-Nevertheless, our working hypothesis for the past six years has been
-that superposition would likely provide a stronger basis for higher-order
-reasoning.
-Other competing higher-order calculi include SMT (implemented in CVC4
-\cite{brotb-19-ho-smt, cbetal-11-cvc4}) and extensional paramodulation (implemented in Leo-III \cite{sb-21-leo3}).
-
-\ourpara{Zipperposition}
-Zipperposition \cite{sc-15-simon-phd,bentkamp-et-al-2021-lamsup-journal} is a higher-order
-theorem prover that implements both \lsup{} variants, combinatory
-superposition, and other superposition-like calculi.
-The prover was conceived as a testbed for rapidly
-experimenting with extensions of first-order superposition, but over time it
-has assimilated many of E's techniques and heuristics and become quite powerful.
-
-Several of our techniques extend the \emph{given clause procedure}, the standard
-saturation procedure pioneered by McCune and Wos \cite[Sect.~2.3]{mcw-1997-otter}. It partitions the proof
-state into a set $P$ of \emph{passive} clauses and a set $A$ of \emph{active}
-clauses. Initially, $P$ contains all input clauses, and $A$ is empty. At each
-iteration, a \emph{given} clause is moved from $P$ to $A$ (i.e., it is
-\emph{activated}), all inferences between it and clauses in $A$ are performed,
-and the conclusions are added to $P$. Because Zipperposition fully simplifies
-clauses only when they are activated, it implements a DISCOUNT-style loop
-\cite{adf-1995-discount}.
-
-\ourpara{Experimental Setup}
-To assess our techniques, we carried out experiments with Zipperposition~2. We
-used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
-TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
-and 1253 Sledgehammer-generated
-monomorphic higher-order problems (labeled \emph{SH}).
-Although some techniques support polymorphism, we
-uniformly used monomorphic benchmarks.
-
-We fixed a \emph{base} configuration
-of Zipperposition parameters as a baseline for all comparisons. This is an
-incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
-well on a wide range of problems.
-%Note that
-%we use a different baseline configuration than in our earlier paper \cite{making-ho-work}.
-%%% That goes without saying. We use a different calculus! Now clarified in intro. --JB
-In each experiment, we varied
-the parameters associated with a specific technique to evaluate it. The
-experiments were run on StarExec Miami \cite{sst-14-starexec} servers, equipped with
-Intel Xeon E5-2620 v4 CPUs clocked at 2.10 GHz. Unless otherwise stated, we used a
-CPU time limit of 15~s, roughly the time each configuration is given in the
-CASC portfolio mode. The raw evaluation results are available online.%
-\footnote{\url{http://doi.org/10.5281/zenodo.5007440}}
-
-\section{Preprocessing Higher-Order Problems}
-\label{sec:satfol:ho-tech:preprocessing}
-
-The TPTP library contains thousands of higher-order problems. Despite their
-diversity, they have a markedly different flavor from the TPTP first-order
-problems. Notably, they extensively use the \verb|definition| role to identify
-universally quantified equations (and equivalences) that define symbols.
+\fc{ForwardIter} begins by backtracking if the previous attempt was successful (i.e., 
+all constraints were solved). If it finds a state from which it can continue,
+it takes term pairs from $\vn{constraints}$ until
+there are no more constraints or it is determined that no unifier exists. The terms
+are normalized by instantiating the head variable with its binding and then
+reducing the potential top-level $\beta$-redex that appears. This instantiation
+and reduction process is repeated
+until there are no more top-level $\beta$-redexes and the head is
+not a variable bound to some term. Then the term with shorter
+$\lambda$ prefix is expanded (only on the top-level) so that both have the
+same length of $\lambda$ prefix. Finally, the $\lambda$ prefix is ignored, and we
+focus only on the body. In this way, we avoid fully substituting
+and normalizing terms and perform just enough operations
+to determine the next step of the procedure.
+
+If either term of the constraint is flex, we first invoke oracles to solve the
+constraint. \ehohii{} implements the most efficient oracles implemented in
+Zipperposition:\ fixpoint and pattern oracle
+\cite[Sect.~6]{unif-section}. An oracle can return three results:
+(1)~there is an MGU for the pair (\cn{Unifiable}), which is recorded in
+\emph{subst}, and the next pair in $\emph{constraints}$ is tried;
+(2)~no MGU exists
+for the pair (\cn{NotUnifiable}), which causes the iterator to backtrack;
+% to previous state;
+(3)~if the pairs do not belong to the subclass that oracle
+can solve (\cn{NotInFragment}), we generate possible variable bindings---that is,
+we guess the approximate form of the solution.
+
+\ehohii{} has a special module that generates bindings (\cn{NextBinding}). This
+module is given the current constraint and the values of \emph{branch\_iter} and
+\emph{steps}, and it either returns the next binding and the new values of
+\emph{branch\_iter} and \emph{steps} or reports that all different variable
+bindings are exhausted. The bindings that \ehohii{}'s unification procedure creates
+are imitation, Huet-style projection, identification, and elimination (one
+argument at a time) \cite[Sect.~3]{unif-section}. A limit on the
+total number of applied binding rules can be set, as well as a limit on the
+number of individual rule applications. The binding module checks whether limits
+are reached using the iterator's \vn{steps} field.
+
+This is the only point in the procedure where the search tree branches an
+different possibilities (bindings) are explored.
+Thus, when \ehohii{} follows the branch indicated by the
+binding module, it records its current state in the backtracking stack. The
+values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
+binding was created, intermediate values that \cn{NextBinding} uses
+to remember how far through bindings it is, and \cn{BindEnd}, which indicates
+that all bindings are exhausted.
+
+If all bindings are exhausted, the procedure checks whether the pair is
+flex--flex and both sides have the same head. If so, the pair is decomposed and
+constraints are derived from the arguments of the pair. Otherwise, the iterator
+backtracks.
+% @PETAR(actually the procedure applies bindings to same-head flex-flex pairs,
+% so the text is OK -- we do get "stuck" enumerating those flex-flex pairs. The
+% limits help here): Maybe explain here that you don't want a pragmatic prover
+% like lambdaE to get stuck in wildly explosive flex-flex pairs? --JB
 %
-Definitions $s \eq t$ (or $(s \iequiv t) \eq \itrue$) can be replaced by rewrite
-rules $s \rightarrow t$,
-using the orientation given in the input problem. If there are multiple
-definitions for the same symbol, only the first one is replaced by a rewrite rule.
-Then, whenever a clause is picked in the given clause procedure, it will be rewritten
-using the collected rules.
-Alternatively, we can rewrite
-the input formulas as a preprocessing step. This ensures that the input
-clauses will be fully simplified when the proving process starts and no
-defined symbols will occur in clauses, which usually helps the heuristics.
-
-Since the TPTP format enforces no constraints on
-definitions, rewriting might diverge. To ensure
-termination, we limit the number of applied rewrite steps. In
-practice, most TPTP problems are well behaved: Only one
-definition is given for each symbol, and the definitions are acyclic.
-
-Turning the defining equations into rewrite rules, unfolding the definitions, and
-$\beta$-reduc\-ing the result can eliminate all of a problem's higher-order features, making
-it amenable to first-order methods. However, this can inflate the problem
-beyond recognition and compromise the refutational completeness of
-superposition.
-
-\begin{exa} 
-  Removing higher-order features of a problem can have adverse effects.
-  Consider the TPTP problem \texttt{NUM636\^{}3}, which defines the predicate $\cst{m}$
-  as $\lambda x.\, \cst{s} \, x \ineq x$ and states its conjecture as $\iforall
-  x.\, \cst{m} \, x $, where $\cst{s}$ is the standard Peano-style natural number
-  successor constructor. When this definition is kept as is, the
-  prover can superpose from either $\cst{m}$ or its definition into the
-  (clausified) induction axiom, which is also given in the problem, and quickly prove
-  the conjecture, without using any advanced inductive reasoning. In contrast,
-  when the definition is
-  unfolded and the problem is $\beta$-reduced, both $\cst{m}$ and the
-  corresponding $\lambda$-abstraction disappear, forcing the prover to guess the
-  correct instantiation for the induction axiom.
-\end{exa}
-
-We describe two techniques to mitigate these issues. The first one is based on the observation that in practice,
-the explosion associated with definition unfolding mostly
-manifests itself on definitions of nonpredicate symbols. In some cases, it is
-preferable to rely on superposition's term order and powerful simplification
-engine to rewrite the proof state rather than to blindly rewrite definitions. On
-the other hand, superposition's reasoning with equivalences is often inadequate
-\cite{bbtv-21-full-ho-sup, gs-05-boolsup}. Thus, it makes sense to treat only
-predicate definitions as rewrite rules.
-
-The second technique aims at preserving completeness: We can try to force the term order that
-parameterizes superposition to orient as many definitions as possible and rely on
-demodulation to simplify the proof state. Usually, the Knuth--Bendix order (KBO)
-\cite{db-1970-kbo} is used. It compares terms by first comparing their weights,
-which is the sum of all the weights assigned to the symbols it contains. Given a
-symbol weight assignment $\mathcal{W}$, we can update it so that it orients
-acyclic definitions from left to right assuming that they are of the form $
-\cst{f} \, \overline{X}_m \eq \lambda \overline{Y}_n. \, t$, where the only free
-variables in $t$ are $\overline{X}_m$, no free variable is repeated or appears
-applied in $t$, and $\cst{f}$ does not occur in $t$. Then we traverse the
-symbols $\cst{f}$ that are defined by such equations following the dependency
-relation, starting with a symbol $\cst{f}$ that does not depend on any other
-defined symbol. For each $\cst{f}$, we set $\mathcal{W}(\cst{f})$ to $w + 1$,
-where $w$ is the maximum weight of the right-hand sides of $\cst{f}$'s
-definitions, computed using $\mathcal{W}$. By construction, for each equation
-the left-hand side is heavier. Thus, the equations are orientable from left to
-right.
-
-\begin{exa} 
-  Many of the problems in the TPTP library's \verb|LCL| category encode modal logic
-  in higher-order logic. More complex modal operators (such as
-  implication and equivalence) are defined in terms of basic connectives (such as negation
-  and disjunction). Some of the definitions present in the problems are
-  $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
-  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
-  (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
-  variables, and basic connectives is 2, we can orient equations using
-  the above described approach as follows. Starting from symbols that do not
-  depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
-  $\mathcal{W}(\cst{mor}) = 17$. Then, we use these values to set
-  $\mathcal{W}(\cst{mimplies}) = 37$. Clearly, these weights
-  enable us to orient all definitions from left to right.
-\end{exa}
-
-% Many higher-order problems, especially those generated from proof assistants,
-% contain hundreds of needless axioms.
-% To filter out axioms that are unlikely to be useful in a proof attempt, many
-% theorem provers rely on the SInE algorithm \cite{hv-2011-sine}. SInE starts with
-% the set of symbols occurring in the conjecture and tries to find axioms that
-% define the properties of these symbols. Then it looks for the definitions of
-% newly found symbols until it reaches a fixpoint. Axioms annotated with
-% \verb|definition| ease this search because they explicitly record the
-% dependency between a symbol and its characterization. To exploit
-% this information, we modified SInE to optionally include the definitions of
-% symbols in the conjecture, regardless of whether they are filtered out or not.
-% Similarly, we implemented mode of SInE which selects only conjecture and axioms
-% annotated with \verb|definition|.
-
-\ourpara{Evaluation and Discussion}
-
-% The \textit{base} configuration treats all axioms annotated with
-% \texttt{definition} as rewrite rules applied as preprocessing.
-% In addition, we tested
-We designed and evaluated the following strategies for handling
-\texttt{definition} axioms:
-
-\begin{enumerate}[\rm no-RW$+$KBO~]
-  \item[\rm pre-RW~] rewrite all definitions as a preprocessing step;
-  \item[\rm in-RW~] rewrite all definitions during the saturation, as an inprocessing step;
-  \item[\rm $o$-RW~] rewrite only predicate definitions, during preprocessing;
-  \item[\rm $o$-RW$+$KBO~] like $o$-RW but with adjusted KBO weights for the remaining
-    definitions;
-  \item[\rm no-RW~] no special treatment of definitions;
-  \item[\rm no-RW$+$KBO~] like $no$-RW but adjusting KBO weights for all definitions.
+If the constraint is rigid--rigid, for unification to succeed the heads of both
+sides of the constraint must be the same. Unification then continues
+with new constraints derived from the arguments. Otherwise, the iterator must be
+backtracked.
+
+\ourpara{Matching} In Ehoh, the matching algorithm is mostly used inside
+simplification rules such as demodulation and subsumption \cite{ss-02-brainiac}.
+As these rules must be efficiently performed, using any complex matching
+algorithm is not a viable option. Instead, we implemented a matching algorithm
+for the pattern class of terms \cite{tn-93-patterns} to complement Ehoh's
+$\lambda$-free higher-order matching algorithm \cite[Sect.~4]{section-ehoh}.
+A term is a \emph{pattern} if each of its free variables either
+has no arguments (as in first-order logic) or is applied to distinct De Bruijn
+indices.
+
+To determine which of the two algorithms to call (pattern or $\lambda$-free), we
+added a cached property \texttt{HasNonPatternVar}, which is set for terms of the
+form $X \,
+\overline{s}_n$ where $n>0$ and either there exists some $s_i$ that is not a De Bruijn
+index or there exist indices $i < j$ such that $s_i = s_j$ is
+a De Bruijn index. This property is propagated to the superterms when they are
+perfectly shared. This allows later checks if a term belongs to the pattern
+class to be performed in constant time.
+
+% When we need to determine if one term matches the other one we first check if
+% any of them has a non-pattern applied variables. If so, $\lambda$-free
+% higher-order matching is tried. This algorithm is adjusted to treat $\lambda$ prefixes
+% as above in unification algorithm (ensuring that $\beta$-redexes can never occur)
+% and that free variables are never bound to terms that have loose bound variables.
+% Else if the terms have only pattern variables,
+% pattern matching algorithm is tried. Note that in this sense, first-order order
+% terms behave as pattern terms and thus pattern matching algorithm is run on
+% them. For example, on terms $(X \, \cst{b}, \cst{f} \, \cst{a} \, \cst{b})$
+% $\lambda$-free matching algorithm is invoked; on $(\cst{f} \, (\lambda x. \,
+% \lambda y.\, X \, x), \cst{f} \, \cst{g})$ pattern matching is tried.
+
+We have modified the $\lambda$-free higher-order matching algorithm to treat
+$\lambda$ prefixes as above in unification procedure---by bringing the prefixes
+to the same length and ignoring them afterwards. This ensures that the
+algorithm will never try to match a free variable to a $\lambda$-abstraction,
+making sure that $\beta$-redexes never appear (as in original $\lambda$-free
+higher-order matching). We also modified the algorithm to ensure that free variables
+are never bound to terms that have loose bound variables. This algorithm cannot
+find many complex matching substitutions (matchers), but it can
+efficiently determine whether two terms are variable renamings of each other
+or whether a simple matcher can be used as in the case of $(X \, (\lambda x. \,x) \,
+\cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
+usually the desired matcher. If this algorithm does not find a matcher and both
+terms are patterns, pattern matching is tried.
+
+\ourpara{Indexing} Ehoh, like other modern theorem provers, efficiently retrieves
+unifiable or matchable pairs of terms using indexing data structures. To find
+terms unifiable with a query term or instances of a query term it uses
+\emph{fingerprint indexing} \cite{ss-12-fp-indexing}. We extended this data
+structure to support (full) higher-order terms in Zipperposition
+\cite[Sect.~6]{unif-section}. We used the same approach in \ehohii{},
+and we extended feature vector indices \cite{ss-2013-feature-vector}
+in the same way.
+
+Ehoh uses \emph{perfect discrimination trees} \cite{mcc-92-pdts} to find
+generalizations of the query term (i.e., terms of which the query term is an instance).
+This data structure is a trie
+that indexes terms by representing them in a serialized, flattened form.
+%\begin{rep}, where each node corresponds to a symbol in a term\end{rep}.
+The left branch from the root in Figure \ref{fig:pdt} shows how the first-order terms
+$\cst{f}\, \cst{a} \, X$ and $\cst{f}\, \cst{a} \, \cst{a}$ are stored.
+In Ehoh, this data structure is extended to support partial application
+and applied variables \cite{section-ehoh}.
+
+\begin{figure}[tb]
+\centering
+\begin{tikzpicture}[level distance=3em, sibling distance=7.5em,
+  %  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
+  %   -| (\tikzchildnode.north)},
+    every node/.style = {shape=rectangle, rounded corners=.75em, minimum size=1.75em, draw, align=center}]
+    \node {\phantom{x}}
+      child { node {$\cst{f}$}
+        child[sibling distance=2.5em] { node %[fill=verylightgray]
+          {$\cst{a}$}
+          child {node {$X$}}
+          child {node {$\cst{a}$}} }
+      }
+      child { node {$\cst{g}$}
+        child {node {$X \, \cst{a} \, \cst{b}$}
+               child {node {$\cst{c}$}}}
+      }
+      child { node {$\cst{h}$}
+        child {node {$\internallam_\iota$}
+          child {node {$\internallam_\iota$}
+            child {node {$X \, \db{0} \, \db{1}$}}
+          }
+      }}
+      ;
+\end{tikzpicture}
+\caption{First-order, $\lambda$-free higher-order, and higher-order pattern
+  terms in a perfect discrimination tree}
+\label{fig:pdt}
+\end{figure}
+
+In \ehohii{}, we extended this structure to support $\lambda$-abstractions and
+the higher-order pattern matching algorithm.
+To this end, we changed the way in which terms are
+serialized. First, we require that all terms are fully $\eta$-expanded (except
+for arguments of variables applied in patterns). Then, when the term is serialized,
+we dedicate a separate node for applied variable terms $X \, \overline{s}_n$,
+instead of the node for $X$ followed by nodes for serialization of arguments
+$\overline{s}_n$. We serialize $\lambda$-abstraction $\lambda x.\, s$ using
+a special node $\internallam_\tau$, where $\tau$ is the type of $x$,
+followed by the serialization of $s$. Other than these two changes, serialization
+remains as in Ehoh, following the gracefulness principle.
+Figure \ref{fig:pdt} shows how $\cst{g} \,
+(X\,\cst{a}\,\cst{b}) \, \cst{c}$ and $\cst{h} \, (\lambda x. \, \lambda y. \, X
+\, y \, x)$ are serialized.
+
+Since the terms are stored in serialized form, it is hard to manipulate
+$\lambda$ prefixes of stored terms during matching. Performing $\eta$-expansion
+when serializing terms makes sure that matchable terms have
+$\lambda$ prefixes of the same length.
+
+We have dedicated separate nodes for applied variables because access to arguments
+of applied variable is necessary for the pattern matching algorithm. Even though
+arguments can be obtained by querying the arity $n$ of the variable and taking
+next $n$ arguments in the serialization, this is both inefficient and inelegant.
+As for De Bruijn indices, we treat them the same as function symbols and assign
+them their own nodes.
+
+\newcommand{\tstack}{\ensuremath{\texttt{term\_stack}}}
+\newcommand{\tproc}{\ensuremath{\texttt{term\_proc}}}
+
+Following the notation from the extension of perfect discrimination trees to
+$\lambda$-free higher-order logic \cite{section-ehoh}, we now describe how enumeration of
+generalizations is performed. To traverse the tree, \ehohii{} begins at the root
+node and maintains two stacks:\ \tstack{} and \tproc{}, where \tstack{} contains the
+subterms of the query term that have to be matched, and \tproc{} contains
+processed terms that are used to backtrack to previous states. Initially,
+\tstack{} contains the query term, the current matching substitution
+$\sigma$ is empty, and the successor node is chosen among the child nodes as
+follows:
+
+\begin{enumerate}
+\item[A.] If the node is labeled with a symbol $\xi$ (where
+  $\xi$ is either a De Bruijn index or a constant) and the top item $t$
+  of \tstack{} is of the form $\xi \, \overline{t}_n$, replace $t$ by
+  $n$~new items $t_1,\dots,t_n$, and push $t$ onto \tproc{}.
+\smallskip
+\item[B.] If the node is labeled with a symbol $\internallam_\tau$ and the top
+  item $t$ of \tstack{} is of the form $\lambda x.\, s$ and the type of $x$ is
+  $\tau$, replace $t$ by $s$, and push $t$ onto \tproc{}.
+\smallskip
+\item[C.] If the node is labeled with a possibly applied variable $X \, \overline{s}_n$
+  (where $n \geq 0$), and the top item of \tstack{} is $t$, the 
+  matching algorithm described above is run on $X \, \overline{s}_n$ and $t$.
+  The algorithm takes into account $\sigma$ built so far and extends it
+  if necessary. If the algorithm succeeds, pop $t$ from \tstack{}, push it onto \tproc{},
+  and save the original value of $\sigma$ in the node.
 \end{enumerate}
 
-The results are given in Fig.~\ref{fig:rewrite}. In all the figures, each
-cell gives the number of proved problems, and cells marked with $\star$
-correspond to the base configuration. The highest number in a category is typeset in
-\relax{bold}. SH benchmarks are not
-included because they do not contain the \texttt{definition} role.
-
-\newcommand{\unknownres}{\ensuremath{{\varnothing}}}
-\newcommand{\colalign}{\phantom{0}}
+Backtracking works in the opposite direction: If the current node is labeled with a De
+Bruijn index or function symbol node of arity $n$, pop $n$ terms from \tstack{} and move
+top of \tproc{} to \tstack{}. If the node is labeled with
+$\internallam_\tau$, pop the top of \tstack{} and move the top of \tproc{} to
+\tstack{}. Finally, if the node is labeled with a possibly applied variable, move top of the
+\tproc{} to \tstack{} and restore the value of $\sigma$.
+
+As an example of how finding a generalization works, consider the following
+states of stacks and substitutions, which emerge when looking for generalizations of 
+$\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}$ in the tree of Figure \ref{fig:pdt}:
+%
+\[\begin{array}{@{}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
+             & \epsilon                                                & \cst{g}                                                 & \cst{g}.(X \, \cst{a} \, \cst{b})                                                          & \cst{g}.(X \, \cst{a} \, \cst{b}).\cst{c} \\[.5\jot]
+  \sigma{:}  & \emptyset                                               & \emptyset                                               & \{X \mapsto \cst{f}\}                                                                      & \{X \mapsto \cst{f}\} \\
+  \tstack{:} & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}, \cst{c}]                & [\cst{c}]                                                                                  & [] \\
+  \tproc{:}  & []                                                      & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{c}{,}\;\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}]
+  \end{array}\]
+%
+
+% \begin{nolinenumbers}%\begin{figure}[h!]
+
+% \end{nolinenumbers}
+
+%   ** follow the pseudocode written on the paper at the VU
+
+% To support multiple unifiers we have changed
+
+% ** Describe we decided to go for depth-bound approach
+
+% ** Describe the algo
+
+% ** Describe the indexing
+
+\section{Preprocessing, Calculus, and Extensions}
+
+\label{sec:calculus}
+% \begin{itemize}  
+  % \item The main goal is simplicity, efficiency and maximal compatibility with
+  %     FO superposition.
+  % \item More or less the same as LFHOL sup, but now computes CSUs instead
+  %     of LFHOL MGUs (and (potentially) uses new orders).
+  % \item Out of Boolean rules we have Boolean simplification and black-box clausification.
+  %     Most importantly -- we have no Boolean hoisting rules : everything is dealt with
+  %     ext-sup, and FOOL clausification! It simply works as good as Boolean hoisting.
+  % \item IMPORTANT: Prefix optimization is disabled, now we rely on ArgCong and NegExt.
+  % \item Many extensions:  ExtSup (With the same indexing), leibniz elimination, predicate enumeration,
+  %     FlexElim, Choice instantiation, recognition of induction.
+  % \end{itemize}
+
+Ehoh's simple $\lambda$-free higher-order calculus performed well on
+Sledgehammer problems and formed a promising stepping stone to full higher-order
+logic \cite{section-ehoh}. When implementing support for full higher-order
+logic, we were guided by efficiency and gracefulness with respect to Ehoh's calculus rather
+than completeness. Whereas Zipperposition provides both complete and incomplete
+modes, \ehohii{} only offers incomplete modes.
+
+\ourpara{Preprocessing} Our experience with Zipperposition showed the importance
+of flexibility in preprocessing the higher-order problems
+\cite{section-making-ho-work}. Therefore, we implemented a flexible
+preprocessing module in \ehohii{}.
+
+To maintain compatibility with Ehoh, \ehohii{} can optionally transform all
+$\lambda$-abstractions into named functions. This process is called
+\emph{$\lambda$-lifting} \cite{rjmh-82-lifting}. \ehohii{} also removes all
+occurrences of Boolean subterms (other than $\itrue, \ifalse$, and free variables)
+in higher-order contexts using a FOOL-like transformation \cite{kotelnikov-16-fool}.
+For example, the formula $\cst{f}(\cst{p} \iand \cst{q}) \ieq \cst{a}$ becomes
+$(\cst{p} \iand \cst{q} \iimplies \cst{f}(\itrue) \ieq \cst{a}) \iand
+(\inot(\cst{p} \iand \cst{q}) \iimplies \cst{f}(\ifalse) \ieq \cst{a})$.
+
+\looseness=-1
+Many TPTP problems use the \texttt{definition} role to denote the definitions of
+symbols. \ehohii{} can treat the definition axioms as rewrite rules, and replace
+all occurrences of defined symbols during preprocessing. Furthermore, during
+SInE \cite{hv-2011-sine} axiom selection, it can always
+include the defined symbol in the trigger relation.
+
+\ourpara{Calculus} \ehohii{} implements the same superposition
+calculus as Ehoh with three important changes. First, wherever Ehoh requires
+the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
+Sect.~\ref{sec:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
+RPO orders designed for $\lambda$-terms.
+% as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.
+
+%TODO: @JASMIN: How do we cite the STTT article?
+Third difference is more subtle. One of the main features of Ehoh is
+\emph{prefix optimization} \cite[Sect.~1]{section-ehoh}: a method which, given a demodulator $s \eq t$, allows
+to replace both applied and unapplied occurrences of $s$ by $t$ 
+by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
+optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
+terms can change drastically, making it much harder to
+track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
+To counterbalance this removal, we introduced
+the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
+in \ehohii{}
+and enabled positive and negative functional extensionality (\infname{PE} and
+\infname{NE}) by default:
+
+\vskip\abovedisplayskip
+
+\noindent\hbox{}\hfill
+  \infer{AC}{s \eq t \llor C}{s \, X \eq t \, X \llor C}%
+\hfill
+  \infer{NE}{s \noteq t \llor C}{s \; (\cst{sk} \; \overline{X}) \noteq t \; (\cst{sk} \; \overline{X}) \llor C}%
+\hfill
+  \infer{PE}{s \; X \eq t \; X \llor C}{s \eq t \llor C}%
+\hfill\hbox{}
+
+\vskip\belowdisplayskip
+
+\noindent
+\infname{AC} and \infname{NE} assume that $s$ and $t$ are of function
+type. In \infname{NE}, $\overline{X}$ denote all the free variables occurring in $s$ and
+$t$, and $\cst{sk}$ is a fresh Skolem symbol of the appropriate type. \infname{PE}
+has a side condition that $X$ may not appear in $s$, $t$, or $C$.
+
+\ourpara{Saturation} E's saturation procedure is based on the assumption that
+each attempt to perform an inference will either result in a single clause or
+fail due to one of the inference side conditions. Unification procedures that
+produce multiple substitutions break this invariant, and the saturation
+procedure needed to be adjusted.
+
+For Zipperposition, Vukmirovi\'c et al.\ developed a variant of the saturation
+procedure that elegantly supports interleaving computation of unifiers and
+scheduling
+inferences to be performed \cite{section-making-ho-work}. Since completeness was
+not one of the \ehohii{} design goals, we did not implement this version of the
+saturation procedure. Instead, in places where previously a single unifier was
+expected, \ehohii{} consumes all elements of the iterator used for enumerating a
+unifier, converting them to clauses.
+
+\ourpara{Reasoning about Formulas} Even though most of the Boolean structure is
+removed during preprocessing, formulas can reappear at the top level of
+clauses during saturation. For example, after replacing $X$ with $\lambda x.
+\, \lambda y.\, x \iand y$, the clause $X \, \cst{p} \, \cst{q} \llor \cst{a}
+\eq \cst{b}$ becomes $(\cst{p} \iand \cst{q}) \llor \cst{a} \eq \cst{b}$. \ehohii{}
+converts every clause of the form $\varphi \llor C$, where $\varphi$ has a logic symbol
+as its head (or it is a (dis)equation between two formulas different than $\itrue$),
+to an explicitly quantified formula. Then, the clausification algorithm
+is invoked on the formula to restore the clausal structure. Zipperposition features
+more dynamic clausification modes, but for simplicity we decided
+not to implement them in \ehohii{}.
+
+The $\lambda$-superposition calculus for full higher-order logic
+\cite{bbtv-21-full-ho-sup} includes many rules that act on Boolean subterms,
+which are necessary for completeness. Other than Boolean simplification rules,
+which use simple tautologies such as $\cst{p} \iand \itrue \iequiv\cst{p}$ to simplify terms, we
+have implemented none of the Boolean rules of this calculus in \ehohii{}. First, we have
+observed that complicated rules such as \infname{FluidBoolHoist} and
+\infname{FluidLoobHoist} are hardly ever useful in practice and usually only
+contribute to an uncontrolled increase in the proof state size. Second, simpler
+rules such as \infname{BoolHoist} can usually be simulated by pragmatic rules
+that perform functional extensionality reasoning, described below.
+
+To make up for excluding Boolean rules, we use an incomplete, but more
+easily controllable and intuitive rule, called \emph{primitive instantiation}. This
+rule instantiates applied predicate variables with approximations of formulas
+that are ground instances of this variable. We use the approximations described
+by Vukmirovi\'c and Nummelin~\cite[Sect.~3.3]{bool-chapter} and
+implemented them in a similar manner.
+
+\ehohii{}'s handling of the Hilbert choice operator is inspired by Leo-III's
+\cite{sb-21-leo3}. \ehohii{}
+recognizes clauses of the form $\neg P \, X \lor P \, (\cst{f} \, P)$ which
+essentially denote that $\cst{f}$ is a choice symbol. Then, when subterm
+$\cst{f} \, s$ is found during saturation, $s$ is used to instantiate the choice
+axiom for $\cst{f}$. 
+%
+Similarly, Leibniz equality \cite{bool-chapter} is eliminated by
+recognizing clauses of the form  $\neg P \, \cst{a} \llor P \, \cst{b} \llor C$. These clauses are then
+instantiated with $P \mapsto \lambda x. \, x \eq \cst{a}$ and $P \mapsto \lambda
+x. \, x\noteq\cst{b}$, which results in $\cst{a} \eq \cst{b} \lor C$.
+
+Finally, \ehohii{} treats induction axioms specially. Just like
+Zipperposition \cite[Sect.~4]{section-making-ho-work}, it abstracts literals
+from the goal clauses and uses these abstractions to instantiate induction
+axioms. Since Zipperposition supports dynamic calculus-level clausification,
+instantiation of induction axioms happens during saturation, when these axioms
+are processed. In \ehohii{}, this instantiation is performed statically,
+immediately after clausification. After \ehohii{} collects all the abstractions,
+it traverses the clauses and instantiates those that have applied variable of
+the same type as the abstraction.
+
+\ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reason about
+functional and Boolean extensionality: It uses \emph{abstracting} rules
+\cite{bbtvw-21-sup-lam} which simulate basic superposition calculus rules,
+but do not require unifiability of the partner terms in the inference. More
+precisely, assume a core inference needs to be performed between two
+$\beta$-reduced terms $u$ and $v$, such that they can be represented as
+$u=C[s_1, \ldots, s_n]$ and $v=C[t_1, \ldots, t_n]$, where $C$ is the most general
+(green \cite{bbtvw-21-sup-lam}) common context of $u$ and $v$, not all of
+$s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
+$s_i$ and $t_i$ are not (applied) free variables, and they are of Boolean or
+function type. Then, the conclusion is formed by taking the conclusion $D$ of
+the core inference rule (which would be created if $s$ and $t$ are unifiable)
+and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.
+
+These rules are particularly useful because \ehohii{} has no rules that dynamically
+process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
+given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
+$\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
+superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
+\iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
+the top level and is further processed by clausification. We noticed that this
+aleviates the need for the other Boolean rules in practice.
+
+\section{Evaluation}
+\label{sec:eval}
+
+In this section, we try to answer two questions about \ehohii{}: \emph{How does
+\ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
+\ehohii{} introduce any overhead compared with Ehoh?} To answer
+these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
+benchmars generated by Sledgehammer (SH) \cite{pb-12-sh}.
+The experiments were carried out on StarExec Miami
+\cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
+2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
+wallclock and 960~s CPU. For SH, we used Sledgehammer's
+default time limit:\ 30~s wallclock and CPU.
+
+\ourpara{Comparison with Other Provers}
+To answer the first question, we let
+\ehohii{} compete with the top contenders in the higher-order division of
+CASC-28:\ cvc5
+\NumberOK{0.0.7},\footnote{\url{https://cvc5.github.io/}}
+Ehoh 2.7
+\cite{section-ehoh}, Leo-III \NumberOK{1.6.6} \cite{sb-21-leo3}, Vampire 4.6
+\cite{br-20-full-sup-w-combs}, and Zipperposition 2.1
+\cite{section-making-ho-work}. We also include Satallax 3.5
+\cite{cb-12-satallax}. We used all \NumberOK{2899} higher-order
+theorems in TPTP 7.5.0 as well as 5000 SH higher-order benchmarks
+originating from the Seventeen benchmark suite \cite{desharnais-et-al-202x}.
+On SH benchmarks,
+cvc5, Ehoh, \ehohii{}, Vampire, and Zipperposition were run using custom schedules provided by
+their developers, optimized for single-core usage and low timeouts.
+Otherwise, we used the corresponding CASC configurations.
+Note that Ehoh 2.7 can parse not only $\lambda$-free but also full higher-order logic,
+which is supported through $\lambda$-lifting.
+We included two versions of Zipperposition:\ \emph{coop} uses Ehoh 2.7 as a backend to finish proof attempts,
+whereas \emph{uncoop} does not use this feature. The results
+are shown in Figure \ref{fig:all-provers}.
+
+Both Ehoh and \ehohii{} were run in the automatic scheduling mode. Compared to Ehoh,
+\ehohii{} features a redesigned module for automatic scheduling, it
+can use multiple CPU cores, and its heuristics have been trained better on higher-order problems.
+
+\ehohii{} dramatically improves the higher-order reasoning capabilities
+compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
+and \NumberOK{3\%} more problems on SH benchmarks, where Ehoh was already very
+successful.
+
+\ehohii{} also outperforms all other higher-order provers other than Zipperposition-coop.
+If Zipperposition's Ehoh backend is disabled,
+\ehohii{} outperforms Zipperposition by a wide margin.
+This comparison is arguably fairer; after all,
+\ehohii{} does not use an older version of Zipperposition
+as a backend.
+%
+These results suggest that \ehohii{} already implements most of the necessary features
+for a high-performance higher-order prover
+but could benefit from the kind of fine-tuning that
+Zipperposition underwent in the last two years.
+
+Remarkably, there is a substantial disparity in the set of problems solved by
+\ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
+\NumberOK{144} SH problems and \NumberOK{24} TPTP problems that Zipperposition-coop
+does not. The lower number of uniquely solved TPTP problems is most likely 
+due to Zipperposition's being heavily optimized on TPTP.
+
+\ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
+in a mode that disables most of the higher-order reasoning. This mode is
+designed for users that are interested only in E's first-order capabilities and
+care a lot about performance.
+To answer the second
+evaluation question, about assessing overhead of \ehohii{},
+we chose all the \NumberOK{1138} unique problems used at CASC
+from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
+both in this first-order (FO) mode and in higher-order (HO) mode.
+
+We fixed a single
+configuration of options, because Ehoh and \ehohii{}'s
+automatic scheduling methods could select different configurations
+and we would not be measuring the overhead but the quality of the chosen
+configurations. We chose the \emph{boa} configuration
+\cite[Sect.~7]{section-ehoh},
+which is the configuration that E~2.2 used the most often in its automatic scheduling mode.
+The results are shown in Figure~\ref{fig:ehoh}.
+
+Counterintuitively, the higher-order versions of both provers outperform
+the first-order counterparts. However, the differerence is so small that it can
+be attributed to the changes to memory layout that affect the order in which
+clauses are chosen. Similar effects are seen when comparing the first-order versions.
+We would expected the less heavily modified version of E, Ehoh, to perform better,
+but due to subtle effects \ehohii{} wins.
 
 \begin{figure}[t]
+  \noindent\hbox{}\hfill
+  \begin{minipage}[ht!]{.66\linewidth}
     \centering
     \def\arraystretch{1.1}%
-  \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
-                & pre-RW                 & in-RW        & $o$-RW               & $o$-RW$+$KBO    & no-RW        & no-RW$ + $KBO  \\ \midrule
-           TPTP & {\bf 1635}$^\star$     & 1619         & 1620                 & 1621            & 1298         & 1296
-           \\ \bottomrule
+    \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c@{\hskip 1em}c} \toprule
+
+                            & TPTP       & SH        \\
+      \midrule
+      cvc5                  & 1931       & 2577      \\
+      Ehoh                  & 2105       & 2611      \\
+      \ehohii{}             & 2527       & 2699      \\
+      Leo-III               & 2282       & 1601      \\
+      Satallax              & 2320       & 1719      \\
+      Vampire               & 2203       & 2240      \\
+      Zipperposition-coop   & {\bf 2583} & {\bf 2754}\\
+      Zipperposition-uncoop & 2483       & 2181      \\
+      \bottomrule
     \end{tabular}}
-       \captionof{figure}{Impact of the definition rewriting method}
-       \label{fig:rewrite}
-     \end{figure}
-
-% {\tt Points to discuss:
-% \begin{itemize}
-%   \item On TPTP problems, there are only 2 problem proved by no-RW$+$KBO but not no-RW --- both after 11s: probably because of some strange effects
-%   \item On TPTP problems, there are 4 problems proved by o-RW$+$KBO but not o-RW --- 2 of them under 1 second: suppressing superposition into rhs of definitions
-%   \item 13 problems (some of them of rating 1) proved by oRW but not regular one
-%   %  zip-jar-hot_defs_oRW(13) : NUM/NUM638^4.p,SEU/SEU800^1.p,NUM/NUM701^4.p,NUM/NUM699^4.p,SEU/SEU603^1.p,NUM/NUM647^4.p,SEU/SEU695^2.p,SEU/SEU630^2.p,NUM/NUM648^4.p,SEU/SEU540^1.p,SYO/SYO531^1.p,SEU/SEU594^1.p,SEU/SEU648^1.p
-%   \item SH problems do not use definition tag, so this is just noise.
-% \end{itemize}
-% }
-
-%%% @PETAR: I find this para very confusing. I don't know when you're talking
-%%% about the old paper's numbers and the present one's. Lert's keep it simple
-%%% and avoid mentions to the old paper. It's subsumed.
-%%%
-%%% Also, careful with words like "counterintuitively". What's counterintuitive
-%%% to you might be intuitive to the reader. E.g. my experience is that hardly
-%%% any well-meant change to a prover makes a big difference on performance,
-%%% so I certainly wasn't surprised.
-%
-%In our earlier paper \cite{making-ho-work}, only an
-%approximation of KBO weight adjustment was implemented: no checks for loops and
-%no topological sorting was performed. For this article, we implemented the
-%feature as described above. Counterintuitively, this did not substantially
-%influence the performance of Zipperposition:
-
-The four configurations in which definitions are treated as rewrite rules
-perform much better than the other two. In contrast, adjusting KBO weights gives
-no substantial improvement: Looking at raw data, we found only
-\NumberOK{2}~problems proved by $o$-RW$+$KBO but not by $o$-RW in which the
-feature was used in the proof. For no-RW and no-RW$+$KBO, the
-\NumberOK{2}-problem difference may be just noise. Even though it proves fewer
-problems, the configuration $o$-RW has some advantages over pre-RW: It proves
-\NumberOK{16} problems that pre-RW does not, \NumberOK{3} of which have a TPTP
-difficulty rating (the ratio of eligible %TPTP
-provers that cannot prove the problem) of~1.
-
-% \looseness=-1
-Rewriting after clausification avoids getting stuck rewriting parts of the
-proof state that might not contribute to the proof. In practice, we noticed that
-rewriting can be so expensive that the prover can spend all
-allotted CPU time in the preprocessing phase. The evaluation results confirm this
-observation: There are \NumberOK{64} problems proved by in-RW but not by
-pre-RW. Moreover, there are \NumberOK{41} problems that can be proved only
-by in-RW but not by any other above described configuration. % from Fig.~\ref{fig:rewrite}.
-
-\section{Reasoning about Formulas}
-\label{sec:satfol:ho-tech:formulas}
-
-Higher-order logic identifies formulas with terms of Boolean type. To prove a problem, we often
-need to instantiate a variable with the right predicate.
-Finding this predicate can be easier if the problem is not clausified.
-Consider the conjecture $\iexists f. \, f \, \cst{p} \, \cst{q} \iequiv \cst{p}
-\iand \cst{q}$. Expressed in this form, the formula is easy to prove by taking
-$f := \lambda x \, y. \> x  \iand y$. By contrast, guessing the right
-instantiation for the negated, clausified form $ \neglit{F \, \cst{p} \,
-\cst{q}} \llor \neglit{\cst{p}} \llor \neglit{\cst{q}},
-\poslit{F \, \cst{p} \, \cst{q}} \llor \poslit{\cst{p}}$, $\poslit{F \, \cst{p}
-\, \cst{q}} \llor \poslit{\cst{q}}$ is more challenging.
-One of the strengths of higher-order tableau provers is that they do not clausify the input
-problem. This might partly explain Satallax's dominance in the THF division of CASC
-competitions until CASC-J10.
-
-% Traditionally, superposition provers clausify the input problem, simplify the
-% initial clause set, and start the saturation loop. We propose to integrate
-% clausification tightly in the saturation loop.
-% %
-% Instead of clausifying the input, we represent the input formula $\varphi$
-% as a higher-order clause $\poslit{\varphi}$. Then we use Vukmirović and Nummelin's
-% \emph{lazy clausification} rules \cite[Sect.~3.4]{our-bool-paper},
-% which extend \lsup{}. These incrementally clausify top-level
-% logical symbols; for example, a clause $C' \llor \neglit{(\cst{p} \iand \cst{q})}$ yields a
-% new clause $C' \llor \neglit{\cst p} \llor \neglit{\cst q}$.
-% Lazy clausification rules admit advanced
-% forms of formula renaming and Skolem symbol reusing that are not available in
-% standard clausification algorithms.
-% \looseness=-1
-The \osup{} calculus supports \emph{delayed clausification rules}
-that insert problems into the proof state in their original,
-nonclausified form, and clausify them gradually. 
-Delayed clausification
-allows the prover to analyze the syntactic structure of formulas during saturation,
-whereas the more traditional approach of \emph{immediate clausification}
-applies a standard clausification algorithm \cite{nw-01-small-cnf}
-both as a preprocessing step and whenever predicate variables are instantiated.
-
-An earlier evaluation of the \osup{} calculus \cite{bbtv-21-full-ho-sup} showed
-that the \emph{outer} variant of delayed clausification substantially increases this calculus's performance.
-The outer variant clausifies
-top-level logical symbols, proceeding from the outside inwards;
-for example, a clause $C \llor \neglit{(\cst{p} \iand \cst{q})}$ is transformed into $C \llor
-\neglit{\cst p} \llor \neglit{\cst q}$. The calculus also supports \emph{inner} delayed
-clausification, which uses only the core calculus rules to clausify problems.
-Even though this is the laziest approach to clausification, the earlier
-evaluation showed that this approach is inefficient. Thus, we focus only on the
-outer rules.
-
-\looseness=-1
-Delayed clausification rules can be used as inference rules (which add conclusions
-to the passive set) or as simplification rules (which delete premises and add
-conclusions to the passive set).
-%
-Inferences give more flexibility, since all
-intermediate clausification states will be stored in the proof state, at the
-cost of producing many clauses. Simplifications produce fewer clauses,
-but risk destroying informative syntactic structure.
-Since clausifying equivalences can destroy a lot of syntactic structure
-\cite{gs-05-boolsup}, we never apply simplifying rules
-on them.
-
-Delayed clausification can interfere with clause splitting techniques.
-Zipperposition supports a lightweight variant of AVATAR \cite{av-2014-avatar},
-an architecture that partitions the search space by
-splitting clauses into variable-disjoint subclauses. This lightweight AVATAR is described
-by Ebner et al.\ \cite[Sect.~7]{2021-ebt-unifying-splitting}. Combining it
-with delayed clausification makes it possible to split a %higher-order
-clause $(\varphi_1 \ior \cdots \ior \varphi_n) \eq \itrue$, where
-the $\varphi_i$'s are arbitrarily complex formulas that share no free
-variables with each other, into clauses $\varphi_i \eq \itrue$.
-%
-To finish the proof, it suffices to derive the empty clause under each assumption
-$\varphi_i \eq \itrue$. Since the split is performed at the formula level, this
-technique resembles tableaux, but it exploits the strengths of superposition,
-such as its powerful redundancy criterion and simplification machinery, to
-close the branches.
-
-\newcommand{\instset}{\ensuremath{\mathit{Inst}}}
-% When delayed clausification is used, the prover can make
-% heuristic choices based on information obtained from the formulas on which
-% clausification rules are applied.
-% In particular, we look for $\lambda$-abstractions whose bodies are formulas
-% in each active clause and store them in a set
-% $\mathit{Inst}$. Optionally, $\mathit{Inst}$ can contain \emph{primitive
-% instantiations} \cite{our-bool-paper}---that is, imitations (in the sense
-% of higher-order unification) of logical symbols that approximate the shape of
-% a formula that can instantiate a predicate variable. These instantiations resemble
-% the ones tableau provers perform during the proof search.
-
-Beyond splitting, interleaving clausification and saturation allows us to simulate another tableau-inspired
-technique. Whenever dynamic clausification substitutes a fresh variable $X$ for
-a predicate variable $x$ in a clause of the form $(\iforall x.\, \varphi) \eq
-\itrue \llor C$, yielding $\varphi\{x \mapsto\nobreak X\}
-\eq \itrue \llor C$, we can create additional clauses in which $x$ is replaced
-with $t \in \instset$, where $\instset$ is a set of heuristically chosen terms.
-This set contains $\lambda$-abstractions whose bodies are formulas and that
-occur in activated clauses, and \emph{primitive instantiations}
-\cite{our-bool-paper}---that is, imitations (in the sense of higher-order
-unification) of logical symbols that approximate the shape of a predicate that
-can instantiate a predicate variable.
-
-Since a new term $t$ can be added to $\mathit{Inst}$ after a clause with a
-quantified variable of $t$'s type has been activated, we 
-remember the clauses $\varphi\{x \mapsto X\} \eq \itrue\allowbreak \llor C$ and instantiate
-them when $\mathit{Inst}$ is extended.
-Conveniently, these instantiated clauses are not recognized as subsumed by
-Zipperposition, which uses an optimized, incomplete higher-order subsumption
-algorithm.
-
-Given a disequation $\cst{f}\,\tuple{s}{n} \not\eq \cst{f}\,\tuple{t}{n}$, the
-\emph{abstraction} of $s_i$ is $\lambda x.\, u \ieq v$, where $u$ is obtained by
-replacing all occurrences of $s_i$ in $\cst{f}\,\tuple{s}{n}$ with $x$ and $v$ is
-obtained by replacing all occurrences of $s_i$ in
-$\cst{f}\,\tuple{t}{n}$ with $x$. For \confrep{}{an equation }$\cst{f}\,\tuple{s}{n} \eq
-\cst{f}\,\tuple{t}{n}$, the analogous abstraction is $\lambda x.\, \inot (u \ieq
-v)$.
-%
-%
-Adding abstractions of the literals occurring in the conjecture to $\mathit{Inst}$ can
-provide useful instantiations for formulas such as induction principles of
-datatypes. As the conjecture is negated\confrep{}{ in refutational theorem proving},
-the equation's polarity is inverted in the
-abstraction.
-
-\begin{exa}
-\label{ex:dat056-2}
-The clausified conjecture of the problem \texttt{DAT056\^{}2}
-\cite{ns-13-leo2sh} from the TPTP library is $\cst{ap} \, \cst{xs}
-\, (\cst{ap} \, \cst{ys} \, \cst{zs}) \not\eq \cst{ap} \, (\cst{ap} \, \cst{xs} \,
-\cst{ys}) \, \cst{zs}$, where $\cst{ap}$ is the \confrep{}{list }append operator defined
-recursively on its first argument and $\cst{xs}$, $\cst{ys}$, and $\cst{zs}$ are
-of list type. Abstracting $\cst{xs}$ from the disequation yields $t = \lambda \mathit{xs}.\, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
-\allowbreak\ieq\allowbreak \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$, which is added
-to $\mathit{Inst}$.
-Included in the problem is the induction axiom
-for the list datatype: $\iforall p. \, p \, \cst{nil} \iand (\iforall x \,
-\mathit{xs}. \, p \, \mathit{xs} \iimplies\allowbreak p \, (\cst{cons} \, x \,
-\mathit{xs})) \iimplies\allowbreak \iforall \mathit{xs}. \, p \, \mathit{xs}$, where
-$\cst{nil}$ and $\cst{cons}$ have the usual meanings.
-Instantiating $p$ with $t$ and
-using the $\cst{ap}$ definition, we can prove
-$\iforall \mathit{xs}. \, \cst{ap} \, \mathit{xs} \, (\cst{ap} \, \cst{ys} \, \cst{zs})
-\ieq \cst{ap} \, (\cst{ap} \, \mathit{xs} \, \cst{ys}) \, \cst{zs}$,
-from which we easily derive a contradiction.
-\end{exa}
-
-\ourpara{Evaluation and Discussion}
-
-\begin{figure}
+    \caption{Comparison of higher-order provers}
+    \label{fig:all-provers}
+  \end{minipage}\hfill\hfill
+  \begin{minipage}[ht!]{.30\linewidth}
     \centering
     \def\arraystretch{1.1}%
-  \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 0.5em}@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
-    & & $+$\relax{LA} & $-$\relax{LA} \\ \midrule
-    TPTP &
-%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{TPTP}}}&
-      \relax{IC}  & 1616  & 1635$^\star$    \\
-    & \relax{DCI} & 1507  & 1532\phantom{$^\star$}    \\
-    & \relax{DCS} & 1668  & {\bf 1703}\phantom{$^\star$} \\ \midrule
-    SH &
-%    \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
-      \relax{IC}  & \colalign425 & \colalign452$^\star$    \\
-    & \relax{DCI} & \colalign362 & \colalign385\phantom{$^\star$}    \\
-    & \relax{DCS} & \colalign441 & \colalign\textbf{457}\phantom{$^\star$} \\ \bottomrule
-  \end{tabular}
-  \caption{Impact of clausification and~lightweight AVATAR}
-  \label{fig:avatar-clause}
+    \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
+                      & TPTP     \\
+      \midrule
+      Ehoh--FO      & 535      \\
+      Ehoh--HO      & 538      \\
+      \ehohii{}--FO & 533      \\
+      \ehohii{}--HO & {\bf 541}\\
+      \bottomrule
+    \end{tabular}}
+    \caption{Evaluation of \ehohii{}'s overhead}
+    \label{fig:ehoh}
+  \end{minipage}\hfill\hfill
 \end{figure}
 
-The base configuration (\emph{base}) uses immediate clausification (\relax{IC}) and 
-disables lightweight AVATAR ($-$\relax{LA}). To test
-the merits of delayed clausification, we vary \emph{base}'s parameters along two axes: We
-choose immediate clausification (\relax{IC}), delayed clausification as inference
-(\relax{DCI}), or delayed clausification as simplification (\relax{DCS}), and we
-either enable ($+$\relax{LA}) or disable ($-$\relax{LA}) lightweight AVATAR.
-Neither of the configurations uses instantiation with terms from $\instset$.
-
-Figure~\ref{fig:avatar-clause} shows that using delayed clausification as
-simplification greatly increases the success rate, regardless of whether
-lightweight AVATAR is used. Using delayed clausification as inference has the
-opposite effect on both problem sets, presumably due to the large number of
-clauses it creates. By manually inspecting the  proofs found by the \relax{DCS}
-configuration, we noticed that a main reason for its success is that it does
-not simplify away equivalences.
+\section{Discussion and Related Work} %---0.5p
+\label{sec:discussion}
+
+On the trajectory to \ehohii{}, we developed, together with colleagues, three
+superposition calculi:\ for $\lambda$-free higher-order logic
+\cite{bbcw-21-lfho}, for a higher-order logic with $\lambda$-abstraction
+but no Booleans \cite{bbtvw-21-sup-lam}, and for full higher-order logic
+\cite{bbtvw-21-sup-lam}. These milestones allowed us to carefully
+estimate how the increased reasoning capabilities of each calculus influence its
+performance.
+
+Extending first-order provers with higher-order reasoning capabilities has been
+tried by other researchers as well. Bhayat and Reger first extended Vampire to
+higher-order logic using combinatory unification \cite{br-20-full-sup-w-combs},
+an incomplete approach, before they designed an implemented a complete
+higher-order superposition calculus based on combinators \cite{br-19-restricted-unif}.
+This calculus is also implemented in Zipperposition. Barbosa et al.\ extended
+the SMT solvers CVC4 (now cvc5) and veriT to higher-order logic in an incomplete
+way \cite{brotb-19-ho-smt}.
+
+Possibly the only actively maintained higher-order prover built from the bottom
+up as a higher-order prover is Leo-III \cite{sb-21-leo3}, as the long-time
+winner of CASC higher-order division, Satallax \cite{cb-12-satallax}, is no
+longer maintained. We give a further overview of other traditional higher-order
+provers and the calculi they are based in the paper about Ehoh
+\cite[Sect.~9]{section-ehoh}.
+
+\section{Conclusion} %---0.5p
+\label{sec:conclusion}
+
+In 2019, the reviewers of our paper introducing Ehoh \cite{section-ehoh} were
+skeptical that extending it with support for full higher-order logic would be
+feasible. One of them wrote:
 %
-Overall, the lightweight AVATAR harms performance, but the sets of
-problems proved with and without it are vastly different. For example,
-the \relax{IC}$+$\relax{LA} configuration proves \NumberOK{38} problems not
-proved by \relax{IC}$-$\relax{LA} (i.e., \emph{base}) on TPTP benchmarks and
-\NumberOK{14} such problems on SH benchmarks.
-
-\looseness=-1
-The Boolean instantiation technique presented above requires delayed
-clausification.
-We assessed it in the best configuration from
-%To assess it, we enabled it in the best configuration from
-Fig.~\ref{fig:avatar-clause}, \relax{DCS}$-$\relax{LA}. With this change ($+$\relax{BI}),
-Zipperposition proves \NumberOK{1700} TPTP problems and \NumberOK{456} SH
+\begin{quote}
+A potential criticism could be that this step from E to Ehoh is just extending
+FOL by those aspects of HOL that are easily in reach with rather straightforward
+extensions (none of the extensions is indeed very complicated), and that the
+difficult challenges of fully supporting HOL have yet to be confronted.
+\end{quote}
+%
+We ended up addressing the theoretical ``difficult challenges'' in other work
+with colleagues. In this \paper, we faced the practical challenges pertaining to the extension
+of Ehoh's data structures and algorithms to support full higher-order logic and
+demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
+that this extension makes \ehohii{} the second best higher-order prover on
+benchmarks coming from interactive theorem proving practice. \ehohii{}'s
+performance lags slightly behind Zipperposition, possibly because Zipperposition
+does not assume a clausal structure and can perform subtle formula-level
+inferences. In the future, we plan to implement the same features in \ehohii{}.
+We have also only started tuning \ehohii{}'s heuristics on higher-order
 problems.
-On TPTP, even though $+$\relax{BI} solves \NumberOK{3} problems less
-than \relax{DCS}$-$LA, it is very useful: \NumberOK{41}
-problems can be proved with $+$\relax{BI} but not with \relax{DCS}$-$\relax{LA}. 
-Conversely, \NumberOK{44} problems are solved with \relax{DCS}$-$\relax{LA},
-but not with $+$\relax{BI}, which suggests that Boolean instantiation can be 
-explosive.
-%
-One of the problems Boolean instantiation helps solve is \texttt{NUM636\^{}2} (a
-re-encoding of \texttt{NUM636\^{}3}).
-It conjectures that $\iforall x.\, \cst{s} \, x \ineq x$, where $x$ ranges over
-Peano-style numbers specified by $\cst{z}$ and $\cst{s}$. The given axioms are
-the induction principle $\forall p.\, p \, \cst{z} \,\iand\, \forall x. \, (p \,
-x \iimplies p \, (\cst{s} \, x)) \iimplies \forall x. \, p\, x$, injectivity
-$\forall x y. \, \cst{s}\,x \ieq \cst{s}\,y \iimplies x \ieq y$, and
-distinctness $\forall x. \, \cst{s}\,x \ineq \cst{z}$. The conjecture is easily
-proved if Boolean instantiation is enabled: Even though the conjecture literal
-cannot be abstracted, instantiating $p$ with the term $\lambda x.\, \cst{s}
-\, x \ineq x$ used in the encoding of the (nonclausified) conjecture leads to a proof in just
-\NumberOK{22} given clause loop iterations. Zipperposition also finds a
-proof using the \relax{DCI}$-$\relax{LA} configuration, but this requires
-\NumberOK{294} iterations.
-
-%Raw data reveals that Boolean instantiation
-The $+$\relax{BI} configuration proves \NumberOK{18} TPTP problems no other
-configuration from Fig.~\ref{fig:avatar-clause} can prove. Among these is
-\texttt{DAT056\^{}2} (Example~\ref{ex:dat056-2}). In contrast, on SH benchmarks, only
-\NumberOK{6} problems are proved using $+$\relax{BI} and not
-\relax{DCS}$-$\relax{LA}. For all these problems, Boolean instantiation does not
-appear in the proof, suggesting that this result is due to the randomness in the
-evaluation environment. The fact that \relax{BI} has no effect on SH benchmarks
-is to be expected because Sledgehammer does
-not include lemmas whose name contains the substring \texttt{.induct} and that
-contain predicate variables. Therefore, \relax{BI} applies to fewer clauses.
-
-\section{Exploring Boolean Selection Functions}
-\label{sec:satfol:ho-tech:bool-select}
-
-Superposition calculi are parameterized by a literal selection function and a
-term order that help prune considerable swaths of the search space without
-jeopardizing completeness. The core inferences apply only to a clause's
-\emph{eligible} literals, defined as either the clause's selected
-literals or, if none are selected, the clause's literals that are maximal with
-respect to the term order. To further restrict which terms can be targeted by
-an inference, the \osup{} calculus introduces \emph{Boolean selection
-functions}.
-
-A Boolean selection function chooses \emph{green subterms} of Boolean type
-(different than $\top$ or $\bot$ and not occurring at either side of a positive
-literal) in a clause and gives rise to a notion of eligibility that considers
-the formula structure. Green subterms correspond to the first-order skeleton of
-a higher-order term; that is, they do not occur in positions under applied
-variables, quantifiers, or $\lambda$-abstractions.
-
-%In particular, green
-%subterms are subterms arising inside first-order-like contexts, and for first
-%other problems, all subterms are green.
-
-\begin{definition}[Green subterms and green positions]
-  \,Green subterms and green positions are defined inductively as
-  follows: $t$ is a green subterm of $t$ at green position $\varepsilon$; if $t$ is a green subterm of $u_i$ at green position $p$
-  and $\cst{f}$ is a constant different from $\iforall$ and $\iexists$, then
-  $t$ is a green subterm of $\cst{f} \, \tuplen{u}$ at green position $i.p$,
-  assuming $i \leq n$.
-\end{definition}
-
-\begin{exa} The green subterms of the term $F \, \cst{a} \, \iand \, \cst{p} \,
-(\iforall \, (\lambda x. \, \cst{q} \, x)) \, \cst{b}$ are the term itself, $F \, \cst{a}$, $\cst{p} \, (\iforall \, (\lambda x. \, \cst{q} \, x))
-\, \cst{b}$, $\iforall \, (\lambda x. \, \cst{q} \, x)$, and $\cst{b}$.
-\end{exa} 
-%
-Green positions are lifted to clauses as follows: If $p$ is the green position
-of a subterm in $s$, and $s$ occurs in a literal $l \in \{s \eq t{,}\; s \not\eq
-t\}$ of $C$, the green position of the same subterm in the clause is denoted by $l.s.p$.
-\osup{} mandates additional restrictions on the Boolean selection function:
-$\itrue$, $\ifalse$ and variable-headed terms cannot be selected; for literals
-$s \eq t$, neither $s$ not $t$ cannot be selected; if a term $s$ contains a
-variable $X$ as a green subterm, and $X\> \tuplen{u}$, with $n \ge 1$, is a
-maximal term of the clause, $s$ cannot be selected.
-
-\begin{definition}[Eligibility]
-  \,Given a substitution $\sigma$ and term order $\succ$, we say a literal $l$
-  is (strictly) eligible with respect to $\sigma$ in $C$ if it is selected in
-  $C$ or there are no selected literals and no selected Boolean subterms in
-  $C$ and $l\sigma$ is (strictly) maximal in $C\sigma$ with respect to the
-  term order.
-%
-  The eligible subterms of a clause $C$ with respect to a substitution
-  $\sigma$ are inductively defined as follows:
-  Any subterm selected by the Boolean selection function is eligible.
-  %Any selected subterm is eligible.
-  For a strictly eligible literal $s \eq t$ with $t\sigma \not\succ
-  s\sigma$, $s$ is eligible. For an eligible literal $s \not\eq t$ with
-  $t\sigma \not\succ s\sigma$, $s$ is eligible. If a subterm $t$ is eligible
-  and the head of $t$ is not $\ieq$ or $\ineq$, all direct green subterms of
-  $t$ are eligible. If a subterm $t$ is eligible and $t$ is of the form $u
-  \ieq v$ or $u \ineq v$, then $u$ is eligible if $v\sigma \not\succ u\sigma$
-  and $v$ is eligible if $u\sigma \not\succ v\sigma$.
-\end{definition}
-
-The above definitions of green subterms and eligibility were originally introduced
-with Boolean $\lambda$-superposition \cite{bbtv-21-full-ho-sup}.
-The Boolean selection function plays a similar role as the literal
-selection function in standard superposition.
-Literal selection functions eliminate some of the nondeterminism present in the superposition
-calculus by focusing on selected parts of the search space. Boolean selection functions achieve
-the same goal, but in a different context: They eliminate nondeterminism that is not
-present in standard superposition, namely, the choice of subformula on which 
-the Boolean calculus rules are to be applied. As with literal selection functions,
-selecting few (and smaller) subterms can give rise to fewer possible inferences
-and reduce clause proliferation.
-
-%
-This notion of eligibility opens up possibilities for reasoning with
-formulas that are hard to simulate with the existing superposition machinery.
-For example, given a formula $\varphi \iimplies \psi$, selecting the antecedent
-simulates forward reasoning, whereas selecting the consequent simulates backward
-reasoning. The new eligibility also makes it possible to restrict the proof
-search to a small, promising part of a formula. Note that
-literal selection can override Boolean selection: Selecting a literal 
-might make some of its green subterms eligible, regardless
-of Boolean selection.
-
-In our previous work \cite{nbtv-2021-foboolsup}, we left this area of new
-possibilities largely unexplored. We designed simple functions that selected
-smallest, largest, innermost, or outermost terms, but they did not impact
-performance much. Here, we propose alternatives.
-%
-Intuitively, a well-performing literal selection
-function might succeed at taming the combinatorial explosion if the
-selected literal can take part in few inferences
-\cite{hrsv-16-selsel}. However, Boolean selection functions
-introduce another factor to consider:\ the context in which the selected
-subterm occurs. This suggests the following definition:
-
-\begin{defi}[Contextualized Boolean selection function]
-  \label{def:context-bool-sel}
-  \,Let $\mathit{ctx}(C)$ be a function that maps a clause $C$ to a set of green positions
-  $p$ such that $C|_p$ is a selectable Boolean subterm, and let
-  $\vartriangleright$ be a partial order on pairs of terms and green positions.
-  The \emph{context
-  Boolean selection function} $\mathit{Sel\/}_{\mathit{ctx}}^{\,\vartriangleright}(C)$
-  selects all terms $t$ such that $t=C|_p$, $p \in \mathit{ctx}(C)$, and
-  $(t,p)$ is maximal with respect to $\vartriangleright$.
-\end{defi}
-
-In the above definition, the function $\mathit{ctx}$ lets
-us choose the context in which the Boolean subterm appears. Then, among
-the terms in the chosen context, we choose the ones that are maximal with
-respect to $\vartriangleright$.
-
-Ganzinger and Stuber considered Boolean subterm selection for their extension of
-first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus requires
-selection of subterms occurring in negative green positions, defined below.
-
-\begin{defi}[Polarity of green positions]
- \,Negative and positive green positions in a clause $C = l_1 \llor \cdots \llor
- l_n$ are defined inductively as follows: For each
- $1 \leq i \leq n$, the green position $l_i.s$ is positive if $l_i = \poslit{s}$ and negative if $l_i
- = \neglit{s}$. If $p$ is positive (negative) and $C|_p = s \,
- \tuplen{t}$ where $s$ is either $\iand$ or $\ior$, then each
- $p.i, 1 \leq i \leq n$, is positive (negative). If $p$ is positive and $C|_p =
- \inot \, s$, then $p.1$ is negative; if $p$ is negative and $C|_p = \inot \, s$,
- then $p.1$ is positive. If $p$ is positive and $C|_p = s \iimplies t$, then
- $p.1$ is negative and $p.2$ is positive; if $p$ is negative and $C|_p = s
- \iimplies t$, then $p.1$ is positive and $p.2$ is negative.
-\end{defi}
-Note that the polarity of $p$ is undefined whenever $C|_p$ is not a green Boolean subterm
-or it occurs under a (dis)equivalence or an uninterpreted symbol.
-To assess how the function $\mathit{ctx}$ affects performance, we use the following
-selection functions that consider green positions of selectable Boolean terms:
-%
-\begin{enumerate}[\rm\texttt{Backward}~]
-  \item[\rm\texttt{Any}~] select all green positions;
-  \item[\rm\texttt{Pos}~] select all positive green positions;
-  \item[\rm\texttt{Neg}~] select all negative green positions;
-  \item[\rm\texttt{Forward}~] select all green positions $p = q.1$ such that $C|_q = s \iimplies t$;
-  \item[\rm\texttt{Backward}~] select all green positions $p = q.2$ such that $C|_q = s \iimplies t$;
-  \item[\rm\texttt{Deep}~] select all green positions of maximal length;
-  \item[\rm\texttt{Shallow}~] select all green positions of minimal length.
-\end{enumerate}
-
-\looseness=-1
-We also introduce three partial orders for selecting subterms from a given
-context. For all three orders, if exactly one of the subterms has a logical head,
-then the subterm with the nonlogical head is larger, because logical symbols
-are more explosive. Otherwise, the orders use the following criteria:
-%
-\begin{enumerate}[$\vartriangleright_\text{ground}$~]
-  \item[$\vartriangleright_\text{ground}$~] If exactly one of the subterms is ground, make the
-  ground subterm larger; otherwise, if exactly one of the subterms is of the form $s \ieq
-  t$, make this subterm larger.
-
-\smallskip
-
-  \item[$\vartriangleright_\text{depth}$~] If one of the subterms has larger subterm depth
-  (longest valid green subterm position), make this subterm larger; otherwise, if one of the
-  subterms has less distinct variables, make this subterm larger.
-
-\smallskip
-
-  \item[$\vartriangleright_\text{def}$~] If exactly one of the subterms is of the form
-  $\cst{p}\,\tuplen{X}$ where $\tuplen{X}$ is a tuple of free variables, make
-  the other subterm larger; otherwise, if exactly one of the subterms is of the form $X \,
-  \tuplen{s}$, make the other subterm larger.
-\end{enumerate}
-
-In case of a tie, the subterm with the smaller syntactic weight is made larger,
-and if both subterms have the same weight, the term that occurs in a position further
-to the left (i.e., that has a lexicographically smaller position) is made larger.
-
-These orders follow the design principle enunciated by Hoder et
-al.~\cite{hrsv-16-selsel} that ground or deep terms and terms with repeated
-variables are ``less unifiable'' with the similar observation for higher-order
-logic that reasoning about interpreted symbols or applied variables is usually
-explosive.
-
-\begin{exa}
-  Selecting the right Boolean subterm can help avoid elaborating
-  high\-er-order inferences. Consider the unsatisfiable clause set consisting of $\cst{p} \, (\lambda y.\, X
-  \, (\lambda x.\, x) \, \cst{a}) \iimplies\allowbreak \inot(\cst{p} \, (\lambda y. \allowbreak\, X
-  \, y \, \cst{a}))$, $\cst{p} \, (\lambda y.\, \cst{a})$, and
-  $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$.
-  Note that $\cst{p} \, (\lambda y.\, X
-  \, (\lambda x.\, x) \, \cst{a})$ and $\cst{p} \,
-  (\lambda y.\, \cst{a})$ have infinitely many unifiers of the form $\{ X \mapsto \lambda fx.
-  \, f^i\, (x) \}, i \geq 0$, whereas $\cst{p} \, (\lambda y. \, X
-  \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have only one unifier. 
-  If \texttt{Forward} context selection is enabled,
-  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
-  forcing computation of at least 100 unifiers (under the assumption that
-  unifiers are returned in order of increasing $i$) before we get to refute
-  $\inot(\cst{p} \, (\lambda y. y^{100} \, \cst{b}))$.
-  In contrast, \texttt{Backward} context selection allows us to
-  superpose from $\cst{p} \, (\lambda y. \, y^{100} \, \cst{b})$ into $\cst{p} \, (\lambda y. \, X
-  \, y \, \cst{a})$, avoiding this explosion. 
-\end{exa}
-
-\ourpara{Evaluation and Discussion}
-
-\begin{figure}
-\centering
-\def\arraystretch{1.1}%
- \begin{tabular}{@{}l@{\hskip 1.5em}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
-  &                                    & \texttt{Any} & \texttt{Pos} & \texttt{Neg}        & \texttt{Forward} & \texttt{Backward} & \texttt{Deep}     & \texttt{Shallow}  \\ \midrule
-  TPTP &
-    $\vartriangleright_\text{ground}$  & 1538         & 1550         & 1547                & 1534             & {\bf 1554}        & 1539              & 1538     \\
-  & $\vartriangleright_\text{depth}$   & 1542         & 1550         & 1528                & 1542             & 1550              & 1547              & 1535     \\
-  & $\vartriangleright_\text{def}$     & 1543         & 1551         & 1540                & 1540             & 1551              & 1545              & 1537  \\ \midrule
-
-%  \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{SH}}} &
-  SH &
-    $\vartriangleright_\text{ground}$  & \colalign386 & \colalign379  & \colalign386       & \colalign386     & \colalign379      & \colalign{\bf387} & \colalign{\bf387}     \\
-  & $\vartriangleright_\text{depth}$   & \colalign377 & \colalign376  & \colalign384       & \colalign378     & \colalign376      & \colalign379      & \colalign376      \\
-  & $\vartriangleright_\text{def}$     & \colalign379 & \colalign374  & \colalign{\bf387}  & \colalign379     & \colalign380      & \colalign377      & \colalign381  \\ \bottomrule
- \end{tabular}
- \caption{Impact of the Boolean selection function}
- \label{fig:bool-sel}
-\end{figure}
-
-When the input problem is clausified using immediate clausification, almost
-all Boolean structure is lost. In this case, we expect Boolean selection to
-have a modest effect. To better assess this feature, in this evaluation we use
-\relax{DCI}$-$\relax{LA} from Sect.~\ref{sec:satfol:ho-tech:formulas} as the baseline
-configuration. To avoid interference of literal and Boolean selection, we
-additionally forbid the literal selection function from selecting a literal if
-it contains a selectable Boolean subterm.
-
-The results of evaluating 21 concrete selection functions obtained
-by instantiating the contextualized Boolean selection function are shown in
-Fig.~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
-while columns denote the function $\mathit{ctx}$.
-
-\looseness=-1
-On TPTP benchmarks, Boolean selection helps tame the explosion caused by
-dynamic clausification used as inference: \NumberOK{All but one} selection functions
-outperform the \relax{DCI}$-$\relax{LA} baseline of \NumberOK{1532} proved problems. Coming back to the problem
-\texttt{NUM636\^{}2} from Sect.~\ref{sec:satfol:ho-tech:formulas}, using Boolean selection
-can reduce the number of given clause loop iterations from
-\NumberOK{294} to \NumberOK{71}.
-
-The results suggest that selection of term context has more impact
-than the partial term order. % The number of proved problems fluctuates more when
-%changing the function $\mathit{ctx}$ than when changing the partial order.
-Also, the best results are obtained when a context more specific than \texttt{Any}
-is chosen. Remarkably, functions using \texttt{Pos} context
-perform better than the ones using \texttt{Neg} context
-on TPTP but the opposite is observed on SH.
-
-Using different Boolean selection functions yields vastly different sets of
-proved problems on TPTP benchmarks: In total, there are \NumberOK{103} problems
-proved by some configuration from Fig.~\ref{fig:bool-sel} but not by
-\relax{DCI}$-$\relax{LA}. However, there are only \NumberOK{16} such SH problems.
-It would seem that the advanced formula reasoning facilitated by the Boolean
-selection formulas is usually not required by Sledgehammer problems.
-
-% {\tt
-%   \begin{itemize}
-%      \item To evaluate bool selection functions we used DCI from above, but
-%            disabled literal selection function from selecting a literal if it
-%            has a selectable Boolean subterm.
-%      \item For TPTP it seems that selection is a good idea: all configurations
-%      are better than naked DCI (only 6 for SH).
-%      \item Overall, it seems that position has bigger influence that the
-%      selection function.
-%      \item On SH there are 14 problems provable by some of the Boolean
-%            selection functions but not naked CSI. Sel3\_deep proves 12 problems
-%            not proved by DCI. Interestingly, the best function Sel3\_neg proves
-%            only 7 problems not proved by DCI.
-%      \item On TPTP much better: 110 problems not proved by DCI but proved by
-%      some of the selection functions. sel3\_pos proves 74 problems not proved
-%      by DCI, sel1 backward 73.
-%      \item TODO: Add any context and maybe $\vartriangleright_4$ which uses weight only
-%      \item What we also observed is that it on TPTP problem it is reasonable to have any selection
-%            function and that already stops the explosion. On SH quality is more important.
-%      \item
-%   \end{itemize}
-% }
-
-\section{Enumerating Infinitely Branching Inferences}
-\label{sec:satfol:ho-tech:infinite-branching}
-
-\looseness=-1
-As an optimization and to simplify the code, Leo-III
-\cite{as-18-phd} and Vampire 4.4 \cite{br-19-restricted-unif} (which uses
-\confrep{}{\emph{restricted combinatory unification}, }a predecessor of combinatory
-superposition) compute only a finite subset of the possible conclusions for
-inferences that require enumerating a CSU. Not only is this a source of
-incompleteness, but choosing the cardinality of the computed subset is a
-difficult heuristic choice. Small sets can result in missing the unifier
-necessary for the proof, whereas large sets make the prover spend too long in
-the unification procedure, generate useless clauses, and possibly get sidetracked into the wrong parts of the search space.
-
-We propose a modification to the given clause procedure to seamlessly
-interleave unifier computation and proof state exploration. Given a complete
-unification procedure, which may yield infinite streams of unifiers, our
-modification fairly enumerates all conclusions of inferences relying on
-elements of a CSU. Under some reasonable assumptions, it behaves exactly like
-the standard given clause procedure on purely first-order problems.
-We also describe heuristics that help achieve a similar
-performance as when using incomplete, terminating unification procedures
-without sacrificing completeness.
-
-Given that we cannot decide whether there exists a next CSU
-element in a stream of unifiers, the request for the next conclusion might not
-terminate,
-effectively bringing the theorem prover to a halt. Our modified given clause procedure
-expects the unification procedure to return a lazily computed stream
-\cite[Sect.~4.2]{co-1999-funds}, where each element is either $\emptyset$ or
-a singleton set containing a unifier. To avoid getting stuck waiting for a unifier
-that may not exist, the unification procedure should return
-$\emptyset$ after it performs some number of operations without finding a unifier.
-
-The complete unification procedure by Vukmirović et al.\
-\cite{our-unif-paper} returns such a stream. Other procedures such
-as Huet's \cite{gh-75-unification} and Jensen and
-Pietrzykowski's \cite{jp-76-unif} can easily be adapted to meet this
-requirement. Based on the stream of unifiers interspersed with $\emptyset$, we
-can construct a stream of inferences similarly interspersed with $\emptyset$.
-Any finite prefixes of this stream can be computed in finite time.
-
-To support such streams in the given clause procedure, we extend it to
-represent the proof state not only by the active ($A$) and passive ($P$) clause
-sets, but also by a priority queue $Q$ containing the inference streams,
-similar to the ``to do'' set $T$ present in the abstract Zipperposition loop of 
-Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
-Each stream is associated with a weight, and $Q$ is sorted in order of
-increasing weight\confrep{}{, a low weight corresponding to a high priority}.
-When they introduced \lsup, Bentkamp et al.\
-\cite{bentkamp-et-al-2021-lamsup-journal} described an older version of this extension. Here
-we present a newer version in more detail, including heuristics to postpone
-unpromising streams. The pseudocode of the modified procedure is as follows:
-\vspace{2.5\jot}
-%
-\newcommand\ParamMode{\ensuremath{K_{\mathrm{fair}}}}
-\newcommand\ParamMaxStreams{\ensuremath{K_{\mathrm{best}}}}
-\newcommand\ParamRetry{\ensuremath{K_{\mathrm{retry}}}}
-\newcommand{\assign}[2]{\State \ensuremath{\mathit{#1} \gets #2}}
-\newcommand{\assignSameLine}[2]{\ensuremath{\mathit{#1} \gets #2}}
-\algrenewcommand\algorithmicindent{1em}
-\begin{algorithmic}[0]
-    \Function{ExtractClause}{$Q$, $\mathit{stream}$}
-      \assign{maybe\_clause}{\text{pop and compute the first element of } \mathit{stream} }
-      \If{$\mathit{stream}$ is not empty}
-        \State add $\mathit{stream}$ to $Q$ with an increased weight
-      \EndIf
-      \State \Return $\mathit{maybe\_clause}$
-    \EndFunction
-
-    \vspace{2\jot}
-
-    \Function{HeuristicProbe}{$Q$}
-      \assign{i}{0}
-      \assign{collected\_clauses}{\emptyset}
-      \While{$i < \ParamMaxStreams$ and $Q \not= \emptyset$}
-        \assign{j}{0}
-        \assign{maybe\_clause}{\emptyset}
-        \While{$j < \ParamRetry$ and $Q \not= \emptyset$ and $\mathit{maybe\_clause} = \emptyset$}
-          \assign{stream}{\text{pop the lowest-weight stream in } Q}
-          \assign{maybe\_clause}{\textsc{ExtractClause}(Q, \mathit{stream})}
-          \assign{j}{j+1}
-        \EndWhile
-        \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \mathit{maybe\_clause}}
-        \assign{i}{i+1}
-      \EndWhile
-      \State \Return $\mathit{collected\_clauses}$
-    \EndFunction
-
-    \vspace{2\jot}
-
-  \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}
-    \assign{collected\_clauses}{\emptyset}
-    \assign{oldest\_streams}{\text{pop } \mathit{num\_oldest} \text{ oldest streams from } Q}
-    \For{$\mathit{stream}$ in $\mathit{oldest\_streams}$}
-      \assign{collected\_clauses}{collected\_clauses \mathrel\cup \textsc{ExtractClause}(Q, \mathit{stream})}
-    \EndFor
-    \State \Return $\mathit{collected\_clauses}$
-  \EndFunction
-
-  \vspace{2\jot}
-
-  \Function{ForceProbe}{$Q$}
-    \assign{collected\_clauses}{\emptyset}
-    \While {$Q \not= \emptyset$ and $\mathit{collected\_clauses} = \emptyset$}
-      \assign{collected\_clauses}{\textsc{FairProbe}(Q, |Q|)}
-    \EndWhile
-
-    \If{$Q = \mathit{collected\_clauses} = \emptyset$}
-      \assign{status}{\textsf{Satisfiable}}
-      % \Return (\textsf{Satisfiable}, $\mathit{collected\_clauses}$)
-    \Else{}
-      \assign{status}{\textsf{Unknown}}
-      % \Return (\textsf{Unknown}, $\mathit{collected\_clauses}$)
-    \EndIf
-
-    \State \Return $(\mathit{status}, \mathit{collected\_clauses})$
-  \EndFunction
-
-  \vspace{2\jot}
-
-  \Function{GivenClause}{$P$, $A$, $Q$}
-
-  \assign{i}{0}
-  \assign{status}{\textsf{Unknown}}
-  \While{$\mathit{status} = \textsf{Unknown}$}
-    \If{$P = \emptyset$}
-      \assign{(status, \,forced\_clauses)}{\textsc{ForceProbe}(Q)}
-      \assign{P}{P \mathrel\cup \mathit{forced\_clauses}}
-    \Else
-      \assign{given}{\text{pop a chosen clause from }P \text{ and simplify it}}
-      \If{$\mathit{given}$ is the empty clause}
-        \assign{status}{\textsf{Unsatisfiable}}
-      \Else
-        \assign{A}{A \mathrel\cup \{\mathit{given}\}}
-        \For{$\mathit{stream}$ in streams of inferences between $\mathit{given}$ and $\mathit{other} \in A$}
-          \If{$\mathit{stream}$ is not empty} \assign{P}{P \mathrel\cup \textsc{ExtractClause}($Q$, \mathit{stream})}
-          \EndIf
-        \EndFor
-        \assign{i}{i+1}
-        \If{$i \; \mathrm{mod} \; \ParamMode = 0$} \assign{P}{P \mathrel\cup \textsc{FairProbe}(Q, i/\ParamMode )}
-        \Else{} \assign{P}{P \mathrel\cup \textsc{HeuristicProbe}(Q)}
-        \EndIf
-      \EndIf
-    \EndIf
-  \EndWhile
-
-  \State \Return $\mathit{status}$
-
-  \EndFunction
-\end{algorithmic}
-% \caption{Our modified given clause procedure}
-% \label{fig:gc-pseudocode}
-%
-% \end{figure}
-\vspace{2.5\jot}
-
-Initially, all input clauses are put into $P$, and $A$ and $Q$ are empty. Unlike
-in the standard given clause procedure, inference results are represented as clause
-streams. The first element is inserted into $P$, and the rest of the stream is
-stored in $Q$ with some positive integer weight computed from the inference rule.
-
-% The functions \textsc{Probe} and \textsc{ForceProbe} extract some conclusions
-% from the inference streams and store them in $P$. \textsc{Probe} heuristically
-% chooses some clauses, whereas \textsc{ForceProbe} is used when $P$ is empty to
-% find one clause if previous probes failed to produce any clauses, as a fallback.
-% By choosing clauses from several streams at the same time,
-% we give more freedom to the prover's existing clause selection heuristics,
-% which select a clause from $P$.
-
-% \looseness=-1
-% \textsc{Probe} has two modes of operation, controlled by a parameter \ParamMode~(by
-% default, $\ParamMode = 70$). In every $\ParamMode$th invocation, \textsc{Probe} extracts
-% conclusions from an increasing number of oldest streams. This amounts to
-% dovetailing, which achieves fairness, analogously to the
-% pick--given ratio \cite{ss-02-brainiac,mcw-1997-otter} in the
-% given clause procedure. For the other
-% invocations, \textsc{Probe} selects the $\ParamMaxStreams$ highest-priority streams (by
-% default, $\ParamMaxStreams = 10$) and extracts elements from, or \emph{probes}, each stream,
-% discarding $\emptyset$ values, until a clause is found or until
-% $\ParamRetry$ (by default, $\ParamRetry=20$) $\emptyset$ values have been discarded.
-% Setting $\ParamRetry>1$ ensures that promising streams are given a fair chance to
-% produce a clause, even if they rely on a complicated unifier. Each
-% probed stream~$S$ is put back in $Q$ with a new priority.%
-
-\looseness=-1
-To eventually consider inference conclusions from streams in $Q$ as given
-clauses, we extract elements from, or \emph{probe}, streams and move any obtained
-clauses to $P$. Analogously to the traditional pick--given ratio
-\cite{ss-02-brainiac,mcw-1997-otter}, we use a parameter
-\ParamMode{} (by default, $\ParamMode = 70$) to ensure fairness: Every \ParamMode{}th iteration,
-\textsc{FairProbe} probes an increasing number of oldest streams, which achieves
-dovetailing. In all other iterations, \textsc{HeuristicProbe} attempts to
-extract up to \ParamMaxStreams{}~clauses from the most promising streams (by default,
-$\ParamMaxStreams = 7$).
-In each attempt, the most promising stream in $Q$ is chosen. If its first
-element is $\emptyset$, the rest of the stream is inserted into $Q$ and a new stream is
-chosen. This is repeated until either \ParamRetry{} occurrences of $\emptyset$ have  been
-met (by default, $\ParamRetry = 20$) or the stream yields a singleton. Setting $\ParamRetry > 0$ increases
-the chance that \textsc{HeuristicProbe} will return $\ParamMaxStreams$ clauses, as desired. Finally, if $P$ becomes empty, \textsc{ForceProbe}
-searches relentlessly for a clause in $Q$, as a fallback.
-
-\looseness=-1
-The function \textsc{ExtractClause} extracts an element from a nonempty stream
-not in $Q$ and inserts the remaining
-stream into $Q$ with an increased weight, calculated as follows.
-Let $n$ be the number of times the stream was chosen for
-probing. If probing results in $\emptyset$, the stream's weight is increased by
-$\max\,\{2{,}\; n-16\}$. If probing results in a clause $C$ whose penalty is
-$p$, the stream's weight is increased by $p \cdot \max\,\{1{,}\; n-64\}$. The
-penalty of a clause is a number assigned by Zipperposition based on
-features such as the depth of its derivation and the rules used in it.
-The constants $16$ and $64$ increase the chance that newer clause-producing streams are picked,
-which is desirable because their first clauses are expected to be useful.
-
-\looseness=-1
-All three probing functions are invoked by
-\textsc{GivenClause}, which contains the saturation loop. It differs
-from the standard given clause procedure in three ways:
-First, the proof state includes $Q$ in addition to $P$ and $A$. Second,
-new inferences involving the given clause are added to $Q$ instead of being
-performed immediately. Third, inferences in $Q$ are periodically performed
-lazily to fill $P$.
-
-\newcommand\infstream[1]{[#1]}
-
-\begin{exa} 
-  \begin{sloppypar}
-  Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
-  \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}),\allowbreak
-  \neg \cst{p} \, (\cst{f}^{100} \, \cst{a})  \}$ and a selection function which
-  selects negative literals.
-  Let $P \mid A \mid Q$ denote
-  the state of the given clause loop (i.e., the contents of the passive and active set
-  and of the stream queue), and let $\infstream{ a_1, a_2, \ldots }$
-  denote an infinite stream of elements.
-  \end{sloppypar}
-
-  The given clause loop begins in the state $X \, (\cst{f} \, \cst{a}) \not\eq
-  \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \, \cst{a}), \neg \cst{p} \,
-  (\cst{f}^{100} \, \cst{a}) \mid \emptyset \mid \emptyset$. If the clause $\neg
-  \cst{p} \, (\cst{f}^{100} \, \cst{a})$ is chosen for processing, since $Q$ is empty
-  and no inferences with the chosen clause are possible, the state becomes $X \,
-  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
-  \cst{a}) \mid \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \emptyset$. When
-  the clause $X \, (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor
-  \cst{p} \, (X \, \cst{a})$ is chosen, a new stream which enumerates results of
-  equality resolution (on its first literal) is created. There are infinitely many
-  conclusions of this inference, since there are infinitely many unifiers for the
-  first literal of the form $\{ X \mapsto \lambda x. \, \cst{f}^i \, x \}$, for
-  $i \geq 0$. Thus, the stream is $\infstream{ \{\cst{p} \, \cst{a}\}, \{\cst{p} \, (\cst{f}\,\cst{a})\}, \ldots }$,
-  possibly with $\emptyset$s interspersed. With the standard given clause procedure,
-  there would have been no way to represent this infinitary result.
-  %The creation of the
-  %inference result stream is the most striking difference between our
-  %and standard given clause procedure.
-
-  When the stream is created, its first element is popped and put into $P$. Then, based on the
-  parameters that control inference stream probing, some number of clauses from
-  the stream are computed and moved to $P$. After two iterations, the state might be
-  $ \cst{p} \,
-  \cst{a}, \cst{p} \, (\cst{f} \, \cst{a}), \cst{p} \, (\cst{f} \, (\cst{f} \, \cst{a}))   \mid X \,
-  (\cst{f} \, \cst{a}) \not\eq \cst{f} \, (X \, \cst{a}) \lor \cst{p} \, (X \,
-  \cst{a}), \neg \cst{p} \, (\cst{f}^{100} \, \cst{a}) \mid \infstream{ \{\cst{p} \, (\cst{f}^3\,\cst{a})\}, \ldots }$.
-
-  In the next iterations, some clause of the form $\cst{p} \, (\cst{f}^{i} \,
-  \cst{a})$, where $i < 100$, is chosen, but no inferences with it can be
-  performed. Then, the stream created in the second iteration is probed, and its
-  results fill the set $P$. Ultimately, the clause $\cst{p} \, (\cst{f}^{100} \,
-  \cst{a})$ is chosen, at which point $\bot$ is quickly derived.
-\end{exa}
-
-\textsc{GivenClause} eagerly stores the first element of a new inference stream
-in $P$ to imitate the standard given clause procedure. If the underlying
-unification procedure behaves like the standard first-order unification
-algorithm on higher-order logic's first-order fragment, our given clause
-procedure coincides with the standard one. The unification procedure by
-Vukmirović et al.\ terminates on the first-order and other fragments
-\cite{tn-93-patterns}. To avoid computing
-complicated unifiers eagerly, it immediately returns $\emptyset$ for a problem that does not
-belong to one of the fragments that admit efficient unifier computation.
-
-The design of our given clause procedure was guided by folklore knowledge about
-higher-order theorem proving. First, in our experience most steps in
-long higher-order proofs involve first-order literals. The unification
-procedure and inference scheduling ensure that first-order inference
-conclusions are put in the proof state as early as possible. Second, some
-inference rules are expected to be largely useless. We initialize the stream
-penalty differently for each rule, allowing old streams of more useful
-inferences to be queried before newly added, but potentially less useful
-streams. Finally, if we use a unification procedure that has aggressive
-redundancy elimination, we will often find the necessary unifier within the
-first few unifiers returned. Similarly, if a stream keeps returning
-$\emptyset$, it is likely that it is blocked in a nonterminating computation
-and should be ignored. Our heuristics to increase the stream penalties take
-both observations into account.
-
-\ourpara{Evaluation and Discussion}
-
-When the unification procedure of Vukmirović et al.\ was implemented  in
-Zipperposition, it was observed that this is the only competing
-higher-order prover that proves all Church numeral problems from the TPTP,
-never spending more than 5~s on a problem \cite{our-unif-paper}. On these hard
-unification problems, the stream system allows the prover to explore the proof
-state lazily.
-
-Consider the TPTP problem \verb|NUM800^1|, which requires finding
-a function $F$ such that $F \, \cst{c_1} \, \cst{c_2}
-\ieq \cst{c_2} \iand F \, \cst{c_2} \, \cst{c_3} \ieq \cst{c_6}$, where
-$\cst{c}_n$ abbreviates the Church numeral for~$n$, $\lambda s\, z. \>
-s^n \, z$. To prove
-\confrep{it}{the problem}, it suffices to take $F$ to be the multiplication operator
-$\lambda x \, y \, s \, z. \> x \, (y \, s) \, z$.
-However, this unifier is only one out of many available for each occurrence of
-$F$.
-
-% TODO: If I add a new benchmark set, make sure this sentence is fixed
-In an independent evaluation setup on a set of 2606 TPTP version 7.2.0
-problems almost identical to the one we use, Vukmirović et al.\
-\cite[Sect.~7]{our-unif-paper}
-compared a complete, nonterminating variant of the unification procedure and a
-pragmatic, terminating variant. The
-pragmatic variant was used directly---all the inference conclusions were put
-immediately in $P$, bypassing $Q$. The complete variant, which relies on
-possibly infinite streams and is much more prolific, proved only 15  problems
-less than the most competitive pragmatic variant. Furthermore, it proved 19
-problems not proved by the pragmatic variant.
-%
-This shows that our given clause procedure, with its heuristics, allows the
-prover to defer exploring less promising branches of the unification and uses
-the full power of a complete higher-order unifier search to solve unification
-problems that cannot be proved by a restricted procedure.
-
-The parameters \ParamMode{}, \ParamRetry{}, and \ParamMaxStreams{} can greatly
-influence the behavior of the given clause procedure, even when the same
-unification procedure is used. Figure~\ref{fig:streams}
-presents the effects of these parameters on TPTP and SH.
-Selecting a low number of best clauses seems to
-perform well on both benchmark sets. However, on SH benchmarks, which require
-overwhelmingly first-order unifiers, visiting older streams should be delayed
-a lot.
-
-As with Boolean selection functions, changing these three parameters causes
-a substantial difference in the set of proved problems. For example, the
-configuration that performs the worst on TPTP benchmarks proves \NumberOK{12}
-problems that the configuration performing the best on TPTP cannot prove; moreover, there
-are \NumberOK{29} TPTP problems that are proved by some set of parameters
-other than $\ParamMode=\ParamMaxStreams=16, \ParamRetry=2$. On SH, these
-effects are much weaker; most reasonable combinations
-of parameters perform similarly.
-
-Among the competing higher-order provers, only Satallax uses infinitely
-branching calculus rules. It maintains a queue of ``commands'' that contain
-instructions on how to create a successor state in the tableau. One
-command describes infinite enumeration of all closed terms of a given function
-type. Each execution of this command makes progress in the enumeration. Unlike
-evaluation of streams representing elements of CSU, each command execution
-is guaranteed to make progress in enumerating the next closed functional
-term, so there is no need to ever return $\emptyset$.
-
-\begin{figure}
-\centering
-\begin{subfigure}[b]{1\textwidth}
-  \centering
-  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
-  &&&&&&& \ParamMode \\[.5\jot]
-  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
-  \cline{3-5}\cline{7-9}\cline{11-13}
-  \\[-1.5\jot]
-  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
-  %                                     2                        16                     256
-  &                         & 2    & 16   & 128  & & 2         & 16   & 128  & & 2    & 16   & 128 \\\midrule
-  & $2$                     & 1643 & 1645 & 1645 & & 1661      & 1661 & 1658 & & 1669 & 1664 & 1664 \\[0.5\jot]
-  $\ParamMaxStreams$ & $16$ & 1647 & 1646 & 1609 & & {\bf1670} & 1654 & 1602 & & 1665 & 1659 & 1597 \\[0.5\jot]
-  & $128$                   & 1646 & 1644 & 1583 & & 1661      & 1656 & 1577 & & 1665 & 1658 & 1576 \\ \bottomrule
-  \end{tabular}
-  \caption{TPTP benchmarks}
-  \label{fig:streams-tptp}
-\end{subfigure}
-\par\bigskip
-\begin{subfigure}[b]{1\textwidth}
-  \centering
-  \begin{tabular}{@{}l@{\kern.5em}l@{\qquad}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}l@{}c@{\kern.75em}c@{\kern.75em}c@{}}\toprule
-  &&&&&&& \ParamMode \\[.5\jot]
-  & & & 2 & & \hbox{\qquad} & & 16 & & \hbox{\qquad} & & 128 & \\[.25\jot]
-  \cline{3-5}\cline{7-9}\cline{11-13}
-  \\[-1.5\jot]
-  &&& \ParamRetry &&&& \ParamRetry &&&& \ParamRetry \\[.5\jot]
-  %                                          2                                               16                                      128
-  &                         & 2            & 16            & 128          & & 2            & 16            & 128          & & 2                  & 16            & 128 \\\midrule
-  & $2$                     & \colalign460 & \colalign455  & \colalign454 & & \colalign465 & \colalign463  & \colalign458 & & \colalign466       & \colalign461  & \colalign461 \\[0.5\jot]
-  $\ParamMaxStreams$ & $16$ & \colalign458 & \colalign453  & \colalign445 & & \colalign464 & \colalign459  & \colalign441 & & \colalign{\bf468}  & \colalign459  & \colalign442 \\[0.5\jot]
-  & $128$                   & \colalign456 & \colalign452  & \colalign430 & & \colalign465 & \colalign458  & \colalign428 & & \colalign{\bf468}  & \colalign459  & \colalign425 \\ \bottomrule
-  \end{tabular}
-  \caption{SH benchmarks}
-  \label{fig:streams-sh}
-\end{subfigure}
-\caption{Impact of the stream enumeration parameter}
-\label{fig:streams}
-\end{figure}
-
-\section{Controlling Prolific Rules}
-\label{sec:satfol:ho-tech:explosiveness}
-
-To support higher-order features
-such as function extensionality and quantification over functions,
-many refutationally complete calculi employ highly prolific rules.
-For example, \lsup{} includes a
-\infname{FluidSup} rule \cite{bentkamp-et-al-2021-lamsup-journal} that very often applies to two
-clauses if one of them contains a term of the form $F \, \overline{s}_n$,
-where $n > 0$.
-We describe three mechanisms to keep rules like these under control.
-
-% \begin{enumerate}
-%   \item Penalize the streams of expensive inferences.
-%   \item Defer selecting the resulting clauses for processing.
-%   \item Restrict the applicability of the rules.
-% \end{enumerate}
-
-% Complete provers that implement \lsup{} %, such as Zipperposition,
-% have two main heuristic choices: which inference stream to query for clauses
-% and which clauses  from the passive set to choose for processing. The first two mechanisms
-% control these two choice points. The third mechanism
-% sacrifices completeness by applying the inference rules only to a subset of chosen clauses.
-
-First, \emph{we limit applicability of the prolific rules}. In practice, it
-often suffices to apply prolific higher-order rules only to initial or shallow
-clauses---clauses with a shallow derivation depth. Thus, we added an option to
-forbid the application of a rule if the derivation depth of any premise exceeds a
-limit.
-
-\newcommand\ParamPenaltyIncrease{\ensuremath{K_\mathrm{incr}}}
-Second, \emph{we penalize the streams of expensive inferences}. The weight of
-each stream is given an initial value based on characteristics of the inference
-premises such as their derivation depth. For prolific rules such as
-\infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
-less prolific variants of this rule, such as \infname{DupSup} \cite{bentkamp-et-al-2021-lamsup-journal}, are increased by a
-fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).
-
-Third, \emph{we defer the selection of prolific clauses}. To select the given
-clause, most saturating provers evaluate clauses according to some criteria and
-choose the clause with the lowest evaluation. To make this choice efficient,
-passive clauses are organized into a priority queue ordered by their
-evaluations. Like E, Zipper\-position maintains multiple
-queues, ordered by different evaluations, that are visited in a round-robin
-fashion. It also uses E's two-layer evaluation functions, a variant of which
-has recently been implemented in Vampire \cite{gs-20-clausesel}.
-%
-The two layers are \emph{clause priority} and \emph{clause weight}. Clauses
-with higher priority are preferred, and the weight is used for tie-breaking.
-Intuitively, the first layer crudely separates clauses into priority classes,
-whereas the second one uses heuristic weights to prefer clauses within a
-priority class. To control the selection of prolific clauses, we introduce new
-clause priority functions that take into account features specific to
-higher-order clauses.
-
-The first new priority function, \verb|PreferHOSteps| (\verb|PHOS|), assigns a
-higher priority if rules specific to higher-order superposition calculi
-were used in the clause derivation. Since most of the other clause priority
-functions tend to defer higher-order clauses, having a clause queue that prefers
-them might be useful to find some proof more
-efficiently. A simpler function, which prefers clauses containing
-$\lambda$-abstractions, is \verb|PreferLambda| (\verb|PL|).
-
-\verb|PreferHOSteps| separates clauses created using first-
-and higher-order inference rules crudely. However, within higher-order
-inference rules there are the ones which make clauses simpler and are thus more
-preferable. An example of such a rule is
-%
-\[\namedinference{\infname{ArgCong}}{C \lor s \eq t}{C \lor s \, \tuplen{X}
-\eq t \, \tuplen{X}}\]
-%
-where $s$ is of the type $\alpha_1 \rightarrow \cdots \rightarrow \alpha_k
-\rightarrow \beta$, $\beta$ is a base type, $n \leq k$, free variables
-$\tuplen{X}$ are fresh, and literal $s \eq t$ is strictly eligible. When $n =
-k$, in most cases, the resulting clause has a first-order literal $s \, \tuplen{X}
-\eq t \, \tuplen{X}$ in place of the literal $s \eq t$ of functional type, which usually makes the clause more useful. To
-prefer clauses that are only mildly higher-order, we designed the function
-\verb|PreferEasyHO| (\verb|PEHO|). It prefers clauses that are the result of
-\infname{ArgCong}, have equations between terms of functional type or
-between higher-order patterns, or have literals containing logical symbols, in that
-order of priority.
-
-A higher-order inference that applies a complicated substitution to a clause is
-usually followed by a $\beta\eta$-normalization step. If
-$\beta\eta$-normalization greatly reduces the size of a clause, it is likely
-that this substitution simplifies the clause (e.g., by removing a variable's
-arguments). The new priority function \verb|ByNormalizationFactor| (\verb|BNF|)
-is designed to exploit this observation. It prefers clauses that were produced
-by $\beta\eta$-normalization, and among those it prefers the ones with larger
-size reductions.
-
-Another new priority function is \verb|PreferShallowAppVars| (\verb|PSAV|). This
-prefers clauses with lower depths of the deepest occurrence of an applied
-variable---that is, $C[X \, \cst{a}]$ is preferred over $C[\cst{f}\,(X \,
-\cst{a})]$. The intuition is that applying a substitution to an applied
-variable often reduces the variable to a term with a constant head,
-yielding a less explosive clause, and the gain is greater for variables closer
-to the top level.  Among the functions that rely
-on properties of applied variables, we implemented \verb|PreferDeepAppVars|
-(\verb|PDAV|), which returns the priority opposite of \verb|PSAV|, and
-\verb|ByAppVarNum| (\verb|BAVN|), which prefers clauses with fewer occurrences
-of applied variables.
-
-\ourpara{Evaluation and Discussion}
-
-\begin{figure}[t]
-  \centering
-  % \def\arraystretch{1.1}%
-  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
-            & \texttt{CP}                & \texttt{BAVN}  & \texttt{PL}  & \texttt{PSAV}       & \texttt{PHOS} & \texttt{PEHO}  & \texttt{BNF} & \texttt{PDAV}      \\ \midrule
-   TPTP     & {1635}$^\star$             & {\bf 1640}     & 1604         & {1635}              & 1609          & 1617           & 1575         & 1533 \\[0.5\jot]
-   SH       & {\bf \colalign452}$^\star$ & \colalign451   & \colalign417 & {\colalign450}      & \colalign439  & \colalign407   & \colalign411 & \colalign302 \\ \bottomrule
-  \end{tabular}}
-  \caption{Impact of the priority function}
-  \label{fig:priorities}
-\end{figure}
-\begin{figure}[t]
-  \centering
-  \def\arraystretch{1.1}%
-  \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}} \toprule
-              & $\infty$                     & $16$          & $8$           & $4$             & $2$             & $1$      \\ \midrule
-   TPTP       & {\bf 1635}$^\star$           & 1625          & 1632          & 1629            & 1628            & 1618 \\[0.5\jot]
-   SH         & {\bf \colalign452}$^\star$   & \colalign438  & \colalign435  & \colalign439    & \colalign435    & \colalign440 \\ \bottomrule
-
-  \end{tabular}}
-  \caption{Impact of the \infname{FluidSup} weight increment \ParamPenaltyIncrease}
-  \label{fig:penalties}
-\end{figure}
-
-In the base configuration (\emph{base}), Zipperposition visits several clause
-queues. The configuration uses queues that prefer the clauses that stem from
-the conjecture, the ones that have at least one positive literal, the ones that
-have been moved from active to passive set, and so on. One of the queues uses the constant
-priority function \texttt{ConstPrio} (\texttt{CP}), meaning that it assigns the same
-priority to every clause. As this queue is the most often visited one in \emph{base},
-changing its priority function should affect the result noticeably.
-To evaluate the new priority functions, we replaced
-\texttt{CP} with one of the new functions in this queue,
-%the queue ordered by \texttt{CP} with the queue ordered by one of the new
-%functions,
-leaving the clause weight intact. The results are shown in Fig.~\ref{fig:priorities}.
-
-Even though constant priority function achieves remarkable performance, the new priority functions are
-useful additions to the prover's repertoire: \NumberOK{37} additional
-TPTP problems and \NumberOK{17} additional SH problems can be proved when some
-nonconstant priority is used. The generally average-performing \verb|PEHO|
-function can prove \NumberOK{9} problems not proved with any other priority function on TPTP
-(and \NumberOK{1} on SH). Globally, \NumberOK{24} TPTP problems and \NumberOK{6} SH problems
-can be proved exclusively using one particular priority function.
-
-% It
-% shows that the expensive priority functions \texttt{PHOS} and \texttt{BNF},
-% which require inspecting the proof of clauses, hardly help. Simple functions
-% such as \texttt{PL} are more effective: Compared with \textit{base}, \texttt{PL}
-% loses \NumberNOK{one} problem overall but proves \NumberNOK{22} new problems.
-
-% {\tt
-%  \begin{itemize}
-
-%   \item Another result which is hard to analyze without looking into differences.
-%   \item On TPTP 32 solutions by using different priorities. On SH 20. Interestingly,
-%   generally bad performing PEHO finds the largest number of solutions not found by CP : 15 (TPTP).
-%   On SH the largest number of unique solutions is by PHOS and PEHO.
-%   \item Another illustration of orthogonality: 25 problems can be proved by only one prio fun (TPTP), 9 on SH.
-
-%  \end{itemize}
-% }
-
-Although it is necessary for refutational completeness, the \infname{FluidSup}
-rule is disabled in \emph{base} because it is so explosive and so seldom useful.
-To test whether increasing inference stream weights makes a difference on the
-success rate, we tried enabling \infname{FluidSup} with different weight
-increments~$\ParamPenaltyIncrease$ for \infname{FluidSup} inference queues. The
-results are shown in Fig.~\ref{fig:penalties}. As expected, using a low
-increment with \infname{FluidSup} is detrimental on TPTP. On this
-benchmark set, \NumberOK{16} additional problems can be proved when
-\infname{FluidSup} is enabled. The penalty mostly affects only proving time: All
-but \NumberOK{2} of these problems were proved by using at least three different
-values of $\ParamPenaltyIncrease$. On SH problems, the best result is obtained
-when the rule is disabled as well. Unexpectedly, the next best result is obtained when
-$\ParamPenaltyIncrease=1$.
-% However, as the column for
-% $\ParamPenaltyIncrease=16$ shows, nor should we use too high an increment, since that
-% delays useful \infname{FluidSup} inferences. Interestingly, even
-% though the configuration with $\ParamPenaltyIncrease=1$ proves the least problems overall, it proves
-% \NumberNOK{7} problems not proved by \emph{base}, which is more than any other
-% configuration we tried.
-
-% {\tt
-% \begin{itemize}
-%   \item Not a single solution found using fluidsup on SH
-%   \item 11 solutions found on TPTP, almost all (9) with at least 3 configurations.
-%         Configurations usually only influence the time necessary to prove.
-%   \item Even though it proves the least problems p1 is necessary in portfolio: with p=1 it takes 0.13 s seconds to prove NUM636\^{}2
-%   and 5.95s with p16 (provable by all ps but not base).
-%   \end{itemize}
-% }
-
-%%% @PETAR: Too little, not informative. What's a "shallow clause"? Not defined yet. --JB
-%In informal experiments, we also tested limiting the applicability
-%of some rules to shallow clauses, but this had no clear impact.
-
-\section{Controlling the Use of Backends}
-\label{sec:satfol:ho-tech:backends}
-
-\newcommand{\ParamNumClauses}{\ensuremath{K_\mathrm{size}}}
-\newcommand{\ParamTime}{\ensuremath{K_\mathrm{time}}}
-
-Cooperation with efficient off-the-shelf first-order theorem provers is an
-essential feature of higher-order theorem provers such as Leo-III
-\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-2013-satallax}.
-% but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
-Those provers invoke first-order backends repeatedly
-during a proof attempt and spend a substantial amount of time in backend
-collaboration. Since \lsup{} generalizes a highly efficient
-first-order calculus, we expect that future efficient \lsup{}
-implementations will not benefit much from backends.
-Nevertheless, experimental provers such
-as Zipperposition can still gain a lot. We present some
-techniques for controlling the use of backends.
-
-In his thesis \cite[Sect.~6.1]{as-18-phd}, Steen extensively studies
-the effects of using different first-order backends on the performance of
-Leo-III. His results suggest that adding only one backend already substantially
-improves the performance. To reduce the effort required for integrating multiple backends, we chose Ehoh \cite{ehoh-section} as our single
-backend. Ehoh is an extension of the highly optimized superposition prover E 2.5
-%%% PETAR: Check 2.5. Later we have 2.6 but that's for Ehoh++.
-with support for higher-order features such as partial
-application, applied variables, and interpreted Booleans.
-%formulas occurring as arguments of function symbols.
-On the one hand, Ehoh provides the efficiency of E while easing the translation from full
-higher-order logic---the only missing syntactic feature is
-$\lambda$-abstraction. On the other hand, Ehoh's higher-order reasoning
-capabilities are limited. Its unification algorithm is essentially first-order,
-and it cannot synthesize $\lambda$-abstractions.
-% or instantiate predicate variables.
-
-In a departure from Leo-III and other cooperative provers,
-\confrep{}{instead of regularly invoking the backend, }%
-we invoke \confrep{the backend}{it} at most once during a run of Zipperposition.
-This is because most competitive higher-order provers, including Zipperposition, use a portfolio
-mode in which many configurations are run for a short time, and we want to
-leave enough time for native higher-order reasoning. Moreover, multiple
-backend invocations tend to be wasteful, because currently each invocation
-starts with no knowledge of the previous ones.
-
-\looseness=-1
-Only a carefully chosen subset of the available clauses are translated and sent
-to Ehoh. Let $I$ be the set of \confrep{input clauses}{clauses representing the
-input problem}. Given a proof state, let \confrep{$M = P \cup A$}{$M$ denote the
-union of the current active and passive sets}, and let $M_\mathrm{ho}$ denote
-the subset of $M$ that contains only clauses that were derived using at least
-one \lsup{} rule not present in regular superposition. We order the clauses in
-$M_\mathrm{ho}$ by increasing derivation depth, using syntactic weight to break
-ties. Then we choose all clauses in $I$ and the first \ParamNumClauses~clauses from
-$M_\mathrm{ho}$ for use with the backend reasoner.
-%We include all the clauses in $I$ since they constitute the backbone of the
-%initial problem. -- Tautology. --JB
-We leave out clauses in $M \setminus (I \cup M_\mathrm{ho})$ because Ehoh can rederive them.
-We also expect large clauses with deep derivations to be less useful.
-
-The remaining step is the translation of $\lambda$-abstractions. We implemented two
-translation methods:\ $\lambda$-lifting \cite{tj-1985-lambdalift} and
-$\cst{SKBCI}$ combinators \cite{da-1979-combtrans}. For
-$\cst{SKBCI}$, we omit the combinator definition axioms, because they are
-very explosive \cite{br-20-full-sup-w-combs}. A third mode simply omits clauses
-containing $\lambda$-abstractions.
-
-\ourpara{Evaluation and Discussion}
-  %
-  % \begin{enumerate}
-  %   \item Different points when E is called, for different time periods
-  %   \item Number of clauses from $M_\mathrm{ho}$ that is translated
-  %   \item Different kinds of liftings
-  % \end{enumerate}
-
-  % We evaluated the following parameters of backend collaboration: the moment of
-  % backend invocation, size of $M_\mathrm{ho}$, and $\lambda$-abstraction
-  % translation method.
-  In Zipperposition, we can adjust the CPU time allotted to Ehoh, Ehoh's own
-  parameters, the point when Ehoh is invoked, the number \ParamNumClauses~of selected clauses from
-  $M_\mathrm{ho}$, and the $\lambda$ translation method. We fix the time
-  limit to 3~s, use Ehoh in \emph{autoschedule} mode, and focus on the last three
-  parameters. In \emph{base}, collaboration with Ehoh is
-  disabled (labeled $-$Ehoh).
-  %Parameters are evaluated by enabling the collaboration with E and
-  %varying parameter values.  -- Uninformative. --JB
-
-  \begin{figure}[t]
-    \noindent\hbox{}\hfill
-    \begin{minipage}[t]{.46\linewidth}
-      \centering
-      \def\arraystretch{1.1}%
-      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
-            & $-$Ehoh                   & $0.1$              & $0.25$          & $0.5$               & $0.75$      \\ \midrule
-      TPTP & 1635$^\star$               & {\bf 1981}         & 1980            & 1979                & 1972 \\[0.5\jot]
-      SH   & \colalign452$^\star$       & \colalign606       & {\bf \colalign608}    & \colalign600        & \colalign592 \\ \bottomrule
-      \end{tabular}}
-      \caption{Impact of the backend invocation point $\ParamTime$}
-      \label{fig:invocation}
-    \end{minipage}\hfill\hfill
-    \begin{minipage}[t]{.46\textwidth}
-      \centering
-      \def\arraystretch{1.1}%
-      \relax{\begin{tabular}{@{}l@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{\hskip 0.75em}c@{}} \toprule
-           & $-$Ehoh              & lifting              & \textsf{SKBCI}    & omitted \\ \midrule
-      TPTP & 1635$^\star$         & {\bf 1980}           & 1877              & 1866 \\[0.5\jot]
-      SH   & \colalign452$^\star$ & \colalign{\bf 608}   & \colalign577      & \colalign566 \\ \bottomrule
-      \end{tabular}}
-      \caption{Impact of the method used to translate $\lambda$-abstractions}
-      \label{fig:translation}
-    \end{minipage}%
-    \hfill\hbox{}
-  \end{figure}
-
-  \begin{figure}[t]
-    \centering
-    \def\arraystretch{1.1}%
-    \relax{\begin{tabular}{@{}l@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{}}\toprule
-          & $-$Ehoh              & $16$               & $32$               & $64$           & $128$             & $256$         & $512$     \\ \midrule
-     TPTP & 1635$^\star$         & {\bf 1985}         & 1980               & 1978           & 1968              & 1968          & 1919 \\[0.5\jot]
-     SH   & \colalign452$^\star$ & \colalign606       & {\bf \colalign608} & \colalign600   & \colalign598      & \colalign596  & \colalign589 \\ \bottomrule
-    \end{tabular}}
-    \caption{Impact of the number of selected clauses~$\ParamNumClauses$}
-    \label{fig:m-ho-size}
-  \end{figure}
-
-  \looseness=-1
-  Ehoh is invoked after $ \ParamTime \cdot t $ CPU seconds, where $0 \le \ParamTime < 1$ and $t$
-  is the total CPU time allotted to Zipperposition. Figure~\ref{fig:invocation}
-  shows the effect of varying $\ParamTime$ when $\ParamNumClauses = 32$
-  and $\lambda$-lifting is used. The evaluation confirms that using a highly
-  optimized backend such as Ehoh greatly improves the performance of a
-  less optimized prover such as Zipperposition.
-  The figure indicates that it is preferable to invoke the backend early. We
-  have indeed observed that if the backend is invoked late, small clauses with
-  deep derivations tend to be present by then. These clauses might have been
-  used to delete important shallow clauses already. But due to their derivation
-  depth, they will not be translated. In such situations, it is better to
-  invoke the backend before the important clauses are deleted.
-
-  Figure~\ref{fig:translation} quantifies the effects of the three
-  $\lambda$-abstraction translation methods. We fixed $\ParamTime = 0.25$ and
-  $\ParamNumClauses=32$. The clear winner is $\lambda$-lifting.
-  $\cst{SKBCI}$ combinators perform slightly better than omitting clauses
-  containing $\lambda$-abstractions.
-
-  \looseness=-1
-  Figure~\ref{fig:m-ho-size} shows the effect of $\ParamNumClauses$ on
-  performance, with $\ParamTime = 0.25$ and $\lambda$-lifting. Including a
-  small number of higher-order clauses with the lowest weight performs better
-  than including a large number of such clauses.
-
-\section{Comparison with Other Provers}
-\label{sec:satfol:ho-tech:comparison}
-
-%The raw evaluation data for the previous experiments show that
-Different choices of
-parameters lead to noticeably different sets of proved problems. In an attempt
-to use Zipperposition~2 to its full potential, we have created a portfolio mode
-that runs up to 50 configurations in parallel during the allotted time. The portfolio
-was designed to solve as many problems as possible from the TPTP benchmark set. To
-provide some context, we compare Zipperposition~2 with the latest versions of
-all higher-order provers that competed at CASC-J10:\ CVC4 1.9
-\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
-\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
-The provers were run using the same parameters as in CASC, but with updated
-executables.
-Note that
-Vampire's higher-order schedule is optimized for running on a single core.
-We also include E~2.7 (more precisely, Ehoh, its higher-order configuration), the first version of this prover to syntactically support
-full higher-order logic, including $\lambda$-abstractions.
-Semantically, E~2.7 is arguably the weakest among the listed provers:
-It simply performs $o$-RW rewriting described in Sect.~\ref{sec:satfol:ho-tech:preprocessing} followed by
-$\lambda$-lifting before it applies \relax{$\lambda$-free superposition} \cite{bbcw-21-lfho}
-(a precursor of all three higher-order superposition calculi) on the preprocessed problem.
-
-We use the same benchmark sets as elsewhere in this article. To imitate the
-CASC-J10 setup, we use a 120~s wall-clock limit and a 960~s CPU limit.
-We even carried out our evaluation on the 8-core CPU nodes that were used for
-CASC-J10. We also ran Zipperposition in uncooperative mode, in which its
-collaboration with a backend is disabled. Figure~\ref{fig:other-provers}
-summarizes the results.
-
-\begin{figure}[t]
-  \centering
-  \def\arraystretch{1.1}%
-  \relax{\begin{tabular}{@{}l@{\hskip 2.5em}c@{\hskip 2em}c@{}} \toprule
-                  & TPTP    & SH  \\ \midrule
-   CVC4           & 1816    & 587   \\
-   E              & 1980    & 676   \\
-   Leo-III        & 2122    & 616  \\
-   Satallax       & 2175    & 588  \\
-   Vampire        & 2072    & 660   \\
-   Zipperposition-uncoop & 2311 & 652 \\
-   Zipperposition & {\bf 2412}    & {\bf 715}  \\ \bottomrule
-  \end{tabular}}
-  \caption{Comparison of competing higher-order theorem provers}
-  \label{fig:other-provers}
-\end{figure}
-
-% Among the cooperative provers, Zipperposition is the one that depends the least
-% on its backend, and its \emph{uncooperative} mode is only \NumberNOK{one}~problem
-% behind Satallax's \emph{cooperative} mode. This confirms our hypothesis that
-% \osup{} is a suitable basis for automatic higher-order reasoning.
-% \confrep{This also}{The increase in performance due to the addition of an efficient backend}
-% suggests that the implementation of this calculus in a modern first-order
-% superposition prover such as E or Vampire would achieve markedly better results.
-% Moreover, we believe that there are still techniques inspired by tableaux, SAT
-% solving, and SMT solving that could be adapted and integrated in saturation
-% provers.
-
-The evaluation results corroborate the CASC results. They also
-show that Zipperposition outperforms all other provers on SH benchmarks. This
-confirms our hypothesis that \lsup{} is a suitable basis for automatic
-higher-order reasoning. Further confirmation is provided by the success rate of
-Zipperposition's uncooperative version: Even without backend,
-Zipperposition is substantially better than all other provers on TPTP
-benchmarks, and it matches the performance of the top contenders on SH.
-On the other hand, the increase in performance due to the addition
-of an efficient backend suggests that the implementation of this calculus in a
-modern first-order superposition prover such as E or Vampire would achieve
-even better results.
-
-We believe that there are still techniques inspired by
-tableaux, SAT solving, and SMT solving that could be adapted and integrated in
-saturation provers. In particular, there are still \NumberOK{25} TPTP problems
-and \NumberOK{17} SH problems that can be proved by other provers but not by
-Zipperposition.
-
-% For VSCode
-% !TEX root = ./conf.tex
-
-%%% TODO @JASMIN @PETAR: The definition of flat L-resolvent is broken because
-%%% of the condition on the variables. --JB
-%%% TODO @JASMIN @PETAR: The definition of "tautology" is probably wrong for
-%%% BCE. Cf. Kiesl et al.
-
-%%% TODO @PETAR: Check @PETAR below. --JB
-
-%\def\orcid#1{} %{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
-%TODO: uncomment before submission:
-% \def\orcid#1{\orcidID{#1}}
-
-% \bibliographystyle{splncs04}
-% \usepackage{amsthm}
-% \usepackage{ntheorem}
-\usepackage{flushend}
-\usepackage{cite}
-\usepackage{etoolbox}
-\usepackage{dcolumn}
-\usepackage{algorithmicx}
-\usepackage{regexpatch}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithms
-\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
-\usepackage[mathscr]{eucal}
-\usepackage{prftree}
-\usepackage[utf8]{inputenc}
-% \usepackage[T1]{fontenc}
-% \usepackage[english]{babel}
-\usepackage{enumerate}
-\usepackage{graphicx}
-\usepackage{cite}
-\usepackage{tikz}
-\usepackage{pgfplots} 
-\usepackage{amsmath}
-\usepackage{amssymb}
-%\usepackage{subcaption}
-%\usepackage{caption}
-\usepackage{booktabs}
-\usepackage[nomessages]{fp}
-\usepackage{enumerate}
-%\captionsetup{justification=raggedright,singlelinecheck=false}
-\xpatchcmd{\algorithmic}{\labelsep 0.5em}{\labelsep 3em}{\typeout{Success!}}{\typeout{Oh dear!}}
-
-%% BEGIN Times font etc.
-\usepackage{mathptmx}
-\usepackage[scaled=.82]{beramono}
-\usepackage[scaled=.865]{helvet}
-%\usepackage{avant}
-\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
-%% END Times font etc.
-
-% Used for displaying a sample figure. If possible, figure files should be
-% included in EPS format.
-
-%\usepackage[
-%   a4paper,
-%   pdftex,
-%   pdftitle={SAT-Inspired Eliminations for Superposition},
-%   pdfauthor={Petar Vukmirovi\'c, Jasmin Blanchette, and Marijn J.H. Heule},
-%   pdfkeywords={},
-%   pdfborder={0 0 0},
-%   draft=false,
-%   colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue,
-%   bookmarksnumbered,
-%   bookmarks,
-%   bookmarksdepth=2,
-%   bookmarksopenlevel=2,
-%   bookmarksopen]{hyperref}
-% If you use the hyperref package, please uncomment the following line to
-% display URLs in blue roman font according to Springer's eBook style:
-%\renewcommand\UrlFont{\color{blue}\rmfamily}
-
-\allowdisplaybreaks
-
-% \usepackage[mathlines]{lineno}
-% \linenumbersep=30pt
-% \linenumbers
-
-% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-%   \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-%   \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-%   \renewenvironment{#1}%
-%%      {\linenomath\csname old#1\endcsname}%
-%%      {\csname oldend#1\endcsname\endlinenomath}}% 
-% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-%   \patchAmsMathEnvironmentForLineno{#1}%
-%   \patchAmsMathEnvironmentForLineno{#1*}}%
-% \AtBeginDocument{%
-% \patchBothAmsMathEnvironmentsForLineno{equation}%
-% \patchBothAmsMathEnvironmentsForLineno{align}%
-% \patchBothAmsMathEnvironmentsForLineno{flalign}%
-% \patchBothAmsMathEnvironmentsForLineno{alignat}%
-% \patchBothAmsMathEnvironmentsForLineno{gather}%
-% \patchBothAmsMathEnvironmentsForLineno{multline}%
-% }
-
-% \usepackage{etoolbox} %% <- for \cspreto, \csappto, \patchcmd, \pretocmd, \apptocmd
-
-% %% Patch 'normal' math environments:
-%% \newcommand*\linenomathpatch[1]{%
-%%   \cspreto{#1}{\linenomath}%
-%%   \cspreto{#1*}{\linenomath}%
-%%   \cspreto{end#1}{\endlinenomath}%
-%%   \cspreto{end#1*}{\endlinenomath}%
-% }
-% %% Patch AMS math environments:
-%% \newcommand*\linenomathpatchAMS[1]{%
-%%   \cspreto{#1}{\linenomathAMS}%
-%%   \cspreto{#1*}{\linenomathAMS}%
-%%   \csappto{end#1}{\endlinenomath}%
-%%   \csappto{end#1*}{\endlinenomath}%
-% }
-
-%% %% Definition of \linenomathAMS depends on whether the mathlines option is provided
-%% \expandafter\ifx\linenomath\linenomathWithnumbers
-%%   \let\linenomathAMS\linenomathWithnumbers
-%   %% The following line gets rid of an extra line numbers at the bottom:
-%%   \patchcmd\linenomathAMS{\advance\postdisplaypenalty\linenopenalty}{}{}{}
-% \else
-%%   \let\linenomathAMS\linenomathNonumbers
-% \fi
-
-%% \linenomathpatch{equation}
-%% \linenomathpatchAMS{gather}
-%% \linenomathpatchAMS{multline}
-%% \linenomathpatchAMS{align}
-%% \linenomathpatchAMS{alignat}
-%% \linenomathpatchAMS{flalign}
-
-\title{SAT-Inspired Eliminations for~Superposition}
-
-\author{
-\IEEEauthorblockN{Petar Vukmirovi\'c\textsuperscript{1}\orcid{0000-0001-7049-6847},
-Jasmin Blanchette\textsuperscript{1,2}\orcid{0000-0002-8367-0936},
-Marijn J.H. Heule\textsuperscript{3}\orcid{0000-0002-5587-8801}}\\
-\IEEEauthorblockA{%
-\textsuperscript{1}Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\ 
-\textsuperscript{2}Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France \\
-\textsuperscript{3}Carnegie Mellon University, Pittsburgh, Pennsylvania, United States}
-}
-
-% \author{
-% \IEEEauthorblockN{Petar Vukmirovi\'c\orcid{0000-0001-7049-6847}}
-% \IEEEauthorblockA{Vrije Universiteit Amsterdam}
-% \and
-% \IEEEauthorblockN{Jasmin Blanchette\orcid{0000-0002-8367-0936}}
-% \IEEEauthorblockA{
-% Vrije Universiteit Amsterdam, \\
-% Max-Planck-Institut f\"ur Informatik, \\
-% Universit\'e de Lorraine}
-% \and
-% \IEEEauthorblockN{Marijn J.H. Heule\orcid{0000-0002-5587-8801}}
-% \IEEEauthorblockA{Carnegie Mellon University}
-% }
-
-% mathcal
-\DeclareFontFamily{OT1}{pzc}{}
-\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
-\DeclareMathAlphabet{\mathcalx}{OT1}{pzc}{m}{it}
-%\DeclareMathAlphabet{\mathcalx}{OMS}{zplm}{m}{n}
-
-%\hyphenation{pre-process-ing in-process-ing}
-
-\begin{document}
-
-% \newcommand\ourpara[1]{\noindent{\textbf{\upshape#1.}}}
-\newcommand\ourpara[1]{\subsection{#1}}
-
-%% TYPESETTING: hacks
-\newcommand\medrightarrow{\mathrel{{{\color{black}\relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\rightarrow}}}
-\newcommand\medleftarrow{\mathrel{{\color{black}\leftarrow}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex\joinrel{{\color{black}\relbar}}}}
-\newcommand\medleftrightarrow{\mathrel{\leftarrow\kern-1.685ex\rightarrow}}
-\newcommand\Medrightarrow{\mathrel{{{\color{black}\Relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\Rightarrow}}}
-\newcommand\Medleftrightarrow{\mathrel{\Leftarrow\kern-1.685ex\Rightarrow}}
-
-\undef\leftrightarrow %% use \medleftrightarrow instead for iff
-
-\newcommand\scif{\quad\text{if~}}
-\newcommand{\HL}{\ensuremath{\mathrm{HL}}}
-\newcommand{\PHL}{\ensuremath{\mathrm{HL_\mathrm{p}}}}
-\newcommand{\cst}[1]{{\mathsf{#1}}}
-\newcommand{\eq}{\mathbin{\approx}}
-\newcommand{\noteq}{\mathbin{\not\approx}}
-\newcommand{\seq}[1]{\ensuremath{(#1)_{n=1}^{\infty}}}
-\newcommand{\impk}[4]{\ensuremath{#1 \impl^{#3}_{#4} #2}}
-\newcommand{\infname}[1]{\textsc{#1}}
-\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
-\newcommand{\namedsimpsc}[4]{\prftree[d][r]{\relax{\infname{#1}\scif $#4$}}{\strut#2}{\strut#3}}
-
-\newcommand{\impl}{\ensuremath{\medrightarrow}}
-% \newcommand{\bigpimpl}[1]{\ensuremath{\overset{#1}{\rightharpoonup}}}
-% \newcommand{\bigfimpl}[1]{\ensuremath{\overset{#1}{\hookrightarrow}}}
-\newcommand{\bigpimpl}[1]{\hookrightarrow_\mathrm{p}}
-\newcommand{\bigfimpl}[1]{\hookrightarrow}
-\newcommand{\eqlit}[2]{\ensuremath{#1 \eq #2}}
-\newcommand{\neqlit}[2]{\ensuremath{#1 \noteq #2}}
-% \newcommand{\eqneqlit}[2]{\ensuremath{#1 \mathrel{\dot{\eq}} #2}}
-\newcommand{\pospredlit}[1]{{#1}}
-\newcommand{\negpredlit}[1]{\neglit{#1}}
-\newcommand{\neglit}[1]{\neg{\kern.2ex #1}}
-\newcommand{\arbpredlit}[1]{(\neg)\kern.2ex\pospredlit{#1}}
-\newcommand{\cor}{\ensuremath{\mathrel{\lor}}}
-\newcommand{\tuple}[2]{\ensuremath{\vec{#1}_{#2}}}
-\newcommand{\tuplen}[1]{\tuple{#1}{n}}
-\newcommand{\tupleempty}[1]{\tuple{#1}{}}
-\newcommand{\flatres}{\mathbin{\smash{\hbox{\Large$\kern-.1ex\rtimes$}}}}  %% (slightly) asymmetric symbol
-\newcommand{\flatresiter}{\mathbin\leadsto}
-\newcommand{\withoutpred}[2]{\,\overline{\!#1}_{\!#2}}
-\newcommand{\withpred}[2]{#1_{#2}}
-\newcommand{\withpredpos}[2]{#1^+_{#2}}
-\newcommand{\withpredneg}[2]{#1^-_{#2}}
-\newcommand{\binset}[1]{#1{\downarrow}_2}
-\newcommand\NumberOK[1]{#1}
-\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
-\newcolumntype{d}{D{.}{.}{0}}
-\newcommand{\confrep}[2]{\begin{conf}#1\end{conf}\begin{rep}#2\end{rep}}
-
-\definecolor{light-gray}{gray}{0.925}
-\newcommand\MAX[1]{\smash{\setlength{\fboxsep}{.3ex}\colorbox{light-gray}{\ensuremath{\vphantom{('q}{#1}}}}}
-
-\newtheorem{theorem}{Theorem}{\bfseries}{\slshape}
-\newtheorem{example}[theorem]{Example}{\bfseries}{\rmfamily}
-\newtheorem{remark}[theorem]{Remark}{\bfseries}{\rmfamily}
-\newtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
-\newtheorem{lemma}[theorem]{Lemma}{\bfseries}{\slshape}
-\newtheorem{corollary}[theorem]{Corollary}{\bfseries}{\slshape}
-% \newtheorem*{proof}{Proof}{\slshape}{\rmfamily}
-
-\newlength{\myskip}
-\setlength{\myskip}{0.6ex}
-% TYPESETTING: when two environments are next to
-% each other, we need to somehow squeeze them
-\newcommand{}{\vspace{-\myskip}}
-
-\newenvironment{theorem}% environment name
-{% begin code
-  \vspace{\myskip}
-  \begin{theorem}
-}%
-{\end{theorem}
- \vspace{\myskip}}% end code
-
-\newenvironment{exa}% environment name
-{% begin code
-  \vspace{\myskip}
-  \begin{example}
-}%
-{\end{example}
- \vspace{\myskip}}% end code
-
-\newenvironment{remarkx}% environment name
-{% begin code
-  \vspace{\myskip}
-  \begin{remark}
-}%
-{\end{remark}
- \vspace{\myskip}}% end code
-
-\newenvironment{defi}% environment name
- {% begin code
-   \vspace{\myskip}
-   \begin{definition}
- }%
- {\end{definition}
-  \vspace{\myskip}}% end code
-
-\newenvironment{lemma}% environment name
-{% begin code
-\vspace{\myskip}
-\begin{lemma}
-}%
-{\end{lemma}
-\vspace{\myskip}}% end code
-
-\newenvironment{corollaryx}% environment name
-{% begin code
-\vspace{\myskip}
-\begin{corollary}
-}%
-{\end{corollary}
-\vspace{\myskip}}% end code
-
-\newenvironment{proof}% environment name
-{% begin code
-  \vspace{\myskip}
-  \textit{Proof:}%
-}%
-{\vspace{\myskip}}% end code
-
-% Stolen from Jasmin's paper
-\newcommand{\qed}{\hfill{\mbox{\ $\IEEEQED$}}}
-
-\maketitle
-\setcounter{footnote}{0}
-
-\newcommand{\ourmodel}{\ensuremath{\mathscr{J}}}
-
-\begin{abstract}
-Optimized SAT solvers not only preprocess the clause set, they also transform it
-during solving as inprocessing. Some preprocessing techniques have been
-generalized to first-order logic with equality. In this \paper, we port
-inprocessing techniques to work with superposition, a leading first-order proof
-calculus, and we strengthen known preprocessing techniques. Specifically, we
-look into elimination of hidden literals, variables (predicates), and blocked
-clauses. Our evaluation using the Zipperposition prover confirms that the new
-techniques usefully supplement the existing superposition machinery.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:satfol:introduction}
-
-Automated reasoning tools have become much more powerful in the last few
-decades thanks to procedures such as conflict-driven clause learning (CDCL)
-\cite{MSLM09HBSAT} for propositional logic and superposition
-\cite{bg-94-superposition} for first-order logic with equality. However,
-the effectiveness of these procedures crucially depends on how the input
-problem is represented as a clause set. The clause set
-can be optimized beforehand (\emph{preprocessing}) or during the execution of
-the procedure (\emph{inprocessing}). In this \paper, we lift several preprocessing and
-inprocessing techniques from propositional logic to clausal first-order logic
-and demonstrate their usefulness in a superposition prover.
-
-For many years, SAT solvers have used inexpensive clause simplification
-techniques such as hidden literal and hidden tautology elimination
-\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and failed literal
-detection \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
-first-order logic with equality
-(Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). Since the generalization
-involves reasoning about infinite sets of literals, we propose restrictions to
-make them usable.
-
-\emph{Variable elimination}, based on Davis--Putnam resolution \cite{dp-60-dp}, has
-been studied in the context of both propositional logic
-\cite{sp-04-niver,cs-00-zres} and quantified Boolean formulas (QBFs)
-\cite{ab-2004-re}. The basic idea is to resolve all clauses with negative
-occurrences of a propositional variable (i.e., a nullary predicate symbol) against
-clauses with positive occurrences and delete the parent clauses. E\'en and
-Biere \cite{eb-2005-satpreprocess} refined the technique to identify a subset
-of clauses that effectively define a variable and use it to further optimize the
-clause set. This latter technique, \emph{variable elimination by substitution},
-has been an important preprocessor component in many SAT solvers since its
-introduction in 2004.
-
-Specializing second-order quantifier elimination
-\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
-\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
-problems, yielding a technique we call \emph{singular predicate elimination}. We
-extend their work along two axes (Sect.~\ref{sec:satfol:predicate-elimination}): We
-generalize E\'en and Biere's refinement to first-order logic, resulting in
-\emph{defined predicate elimination}, and explain how both types of predicate
-elimination can be used during the proof search as inprocessing.
-
-\begin{qle}
-It is well known that clauses containing \emph{pure literals}---predicate
-literals whose root symbol occurs only with one polarity in the problem---can
-be eliminated. We propose a generalization of this technique, under the name
-\emph{quasipure literal elimination}, to allow the elimination of some bipolar
-literals as well (Sect.~\ref{sec:satfol:quasipure-literal-elimination}).
-\end{qle}
-
-The last technique we study is \emph{blocked clause elimination} (Sect.\
-\ref{sec:satfol:satisfiability-by-clause-elimination}). It is used in both SAT
-\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
-logic has produced good results when used as a preprocessor, especially on
-satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
-blocked clause elimination on satisfiable problems, including using it to
-establish equisatisfiability with an empty clause set or as an inprocessing
-rule. Unfortunately, we find that its use as inprocessing can compromise the
-refutational completeness of superposition.
-
-% Preprocessing has been a crucial part of SAT solvers since 2004, when E\'en and
-% Biere introduced bounded variable elimination by substitution~\cite{BVE}.
-% They based it on Davis--Putnam resolution \cite{dp-60-dp}, which
-% replaces all clauses containing a variable $x$ by all nontautological
-% resolvents on $x$, but showed how to exploit definition information
-% extracted form the clause set to add only a subset of the nontautological
-% resolvents. In this way, many variables can be eliminated without
-% growing the clause set. % (in terms of the number of clauses). ---that comes later --JB
-
-% Preprocessing is generally effective, but it can be expensive. As a compromise,
-% many solvers use inprocessing:\ interleaving the decision procedure with
-% preprocessing steps~\cite{mhb-12-inprocessing}. Top-tier SAT solvers, such as CaDiCaL
-% \cite{xxx} and Kissat \cite{xxx} alternate about every second between solving
-% and inprocessing. In contrast to preprocessing, inprocessing can also be used
-% to simplify the clause set after some important deductions have been made.
-
-% Pre-~and inprocessing are successful beyond propositional logic. For example,
-% lifting variable and clause elimination techniques to quantified Boolean
-% formulas (QBF) helps solve many formulas that are too difficult in practice for
-% decision procedures~\cite{bls-11-bloqqer}. This suggests that richer logics, such as
-% first- and higher-order logic, can also benefit from elimination techniques.
-
-% To improve the success rate of superposition provers, we extend several of
-% preprocessing and inprocessing techniques from propositional to first-order logic,
-% including failed literal detection, hidden literal elimination, hidden
-% tautology elimination
-% (Sect.~\ref{sec:satfol:hidden-literal-based-elimination}). These
-% techniques are fairly inexpensive in propositional logic but need to be
-% curtailed in first-order logic because they would otherwise involve reasoning
-% about an infinite set of literals.
-
-% Khasidashvili and Korovin \cite{kk-2016-pe-fol} generalized Davis--Putnam
-% resolution to first-order logic with equality. We call their technique
-% \emph{singular predicate elimination}. We also adapt bounded variable
-% elimination by substitution, yielding \emph{defined predicate elimination}
-% (Sect.~\ref{sec:satfol:predicate-elimination}). With some care, we can integrate these
-% techniques with superposition without losing refutational completeness.
-
-% Finally, we study a first-order generalization of blocked clause
-% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satfol:satisfiability-by-clause-elimination}).
-% %Unlike the other techniques, it weakens the clause set.
-% If all clauses in the current clause set are blocked, this means that it is
-% satisfiable.
-% %It can be used to show
-% %satisfiability by establishing equisatisfiability between the initial clause
-% %set and the empty clause set.
-% We show how this technique can help superposition provers detect
-% satisfiability early instead of diverging. We also show, however, that
-% interleaving the technique with superposition compromises refutational
-% completeness.
-
-All techniques are implemented in the Zipperposition prover
-(Sect.~\ref{sec:satfol:implementation}), allowing us to ascertain their usefulness
-(Sect.~\ref{sec:satfol:evaluation}). The best configuration solves \NumberOK{160}
-additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
-\cite{gs-17-tptp}. The raw experimental data are publicly available.%
-\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
-More details, including all the proofs, can be found in a
-technical report \cite{our-report}.}{}
-
-\begin{rep}
-This \paper{} expands on our FMCAD 2021 paper \cite{fmcad}. It includes more
-examples, lemmas, and proofs. It also corrects a mistake in the handling of
-variables in Definition~\ref{def:equality-blocked}.
-\end{rep}
-
-\section{Preliminaries}
-\label{sec:satfol:preliminaries}
-
-\ourpara{Clausal First-Order Logic}
-\looseness=-1
-Our setting is many-sorted, or many-typed, first-order logic \cite{jg-1987-logic-textbook} with
-interpreted equality and a distinguished type (or sort) $o$.
-Each variable $x$ is assigned a non-Boolean
-type, and each symbol $\cst{f}$ is assigned a tuple $(\tau_1, \ldots, \tau_n,
-\tau)$ where $n \geq 0$, $\tau_i$ are non-Boolean types, and $\tau$ is
-the \emph{result type}. We distinguish between
-\emph{predicate symbols}, with $o$ as the result type, and \emph{function symbols}.
-Nullary function symbols are called \emph{constants}. Terms are either
-variables~$x$ or \confrep{well-typed }{}applications $\cst{f}(t_1, \ldots,
-t_n)$\confrep{, or $\cst{f}$ if $n = 0.$}{, of type $\tau$ where $\cst{f}$ is assigned $(\tau_1, \ldots,
-\tau_n, \tau)$, and each $t_i$ is a term of type $\tau_i$.
-The parentheses are omitted if $n = 0$.} A term is
-\emph{ground} if it contains no variables.
-We assume standard definitions and
-notations for positions, subterms, and contexts \cite{bn-98-tr-and-all-that}.
-We abbreviate a vector $(a_1, \ldots, a_n)$ to $\tuplen{a}$ or $\tuple{a}{}$,
-and write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol $\cst{f}$
-(e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$).
-
-An atom is an equation $\eqlit{s}{t}$ corresponding to an unordered pair $\{s,
-t\}$. A literal is an equation $\eqlit{s}{t}$ or a disequation $\neqlit{s}{t}$.
-For every predicate symbol~$\cst{p}$, \confrep{}{the notation }$\cst{p}(\tuple{s}{})$
-abbreviates $\eqlit{\cst{p}(\tuple{s}{})}{\top}$, and
-$\neglit{\cst{p}(\tuple{s}{})}$ abbreviates
-$\neqlit{\cst{p}(\tuple{s}{})}{\top}$, where $\top$ is a distinguished
-constant of type $o$. We distinguish between \emph{predicate literals}
-$\arbpredlit{\cst{p}(\tuple{s}{})}$ and \emph{functional literals} $\eqlit{s}{t}$,
-where $s$ and $t$ are not of type $o$.
-Given a literal $L$, we overload notation and write $\neglit{L}$ to
-denote its complement. A clause $C$ is a multiset of literals, written as $L_1
-\cor \cdots \cor L_n$ and interpreted disjunctively. Clauses are often defined
-as sets of literals, but superposition needs multisets; with multisets,
-an instance $C\sigma$ always has the same number of literals as $C$, a most
-convenient property. Given a clause set $N$, $\binset{N}$ denotes the subset of
-its binary clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.
-
-\begin{rep}
-A substitution $\sigma$ is a well-typed mapping from
-variables to terms such that the set $\{x \mid \sigma(x) \not= x \}$ is finite.
-%Assuming $x_i \not= t_i$,
-Substitutions are written as $\{x_1 \mapsto t_1,
-\ldots,\allowbreak x_n \mapsto t_n\}$ or $\{ \tupleempty{x} \mapsto \tupleempty{t} \}$.%
-\confrep{}{ A substitution is a \emph{variable renaming} if it is a
-bijection from a set of variables to a set of variables.}
-
-We assume the natural extensions of domain, valuation, interpretation and model
-(as defined by Fitting \cite{mf-1996-fol}) from unsorted to many-sorted
-logic.
-Models we consider are \emph{normal}, i.e., they interpret $\eq$ as an
-equality relation. Usual notions of (un)satisfiability and (in)validity are
-assumed. We write $\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$
-satisfies a clause set $N$, for a variable assignment $\xi$. If $\ourmodel$
-is a model of $N$ (i.e., it satisfies it under every variable assignment), 
-we simply write $\ourmodel \models N.$  Abusing notation,
-we write $M \models N$ to denote that $M$ \emph{entails} $N$, i.e., that every
-model of $M$ is a model of $N.$
-
-A canonical model $\ourmodel$ is a normal model such that for every element $d$
-in one of $\ourmodel$'s domains there exists a ground term $t$ such that
-$\ourmodel$ interprets $t$ as $d$. Canonical models generalize Herbrand models
-to first-order logic with equality: If a clause set is satisfiable, then it
-is satisfiable in a canonical model.
-\end{rep}
-
-\ourpara{Superposition Provers}
-%
-Superposition \cite{bg-94-superposition} is a calculus for clausal
-first-order logic that extends ordered resolution
-\cite{bg-01-resolution} with equality reasoning. It is
-refutationally complete: Given a finite, unsatisfiable clause set, it will
-eventually derive the empty clause. It is parameterized by a \emph{selection
-function} that influences which of a clause's literals are eligible as the target
-of inferences. Moreover, it is compatible with the \emph{standard redundancy
-criterion}, which can be used to delete a clause $C$ while preserving completeness of the calculus.
-
-\looseness=-1
-The redundancy criterion relies on an order $\succ$ that
-compares terms, literals, or clauses.
-The order is used to determine whether clauses can be deleted.
-If $N$ is ground, $C$ can be deleted
-if it is entailed by $\prec$-smaller clauses in $N.$ This definition is lifted
-to nonground sets $N$. The criterion can be used to delete a
-clause that is \emph{subsumed} by another clause (e.g.,
-$\cst{\cst{p}(\cst{a})} \cor \cst{q}$ by $\cst{p}(x)$) or
-to \emph{simplify} a clause $C$ into $C'$, which amounts to adding $C'$ and
-then deleting $C$ as redundant with respect to $N \cup \{C'\}$.
-Subsumption and simplification are the main inprocessing mechanisms available
-to superposition provers. Some provers also implement clause splitting
-\cite{riazanov-voronkov-2001,fietzke-weidenbach-2009,av-2014-avatar}.
-
-Superposition provers saturate the input problem with respect to the calculus's
-inference rules using the \emph{given clause procedure}
-\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
-passive set $\mathcalx{P}$ and an active set $\mathcalx{A}$. All clauses start in
-$\mathcalx{P}$. At each iteration of the procedure's main loop, the prover
-chooses a clause $C$ from $\mathcalx{P}$, simplifies it, and moves it to
-$\mathcalx{A}$. Then all inferences between $C$ and active clauses are performed.
-The resulting clauses are again simplified and put in $\mathcalx{P}$.\confrep{}{
-The provers differ in which clauses are used for simplification: Otter-loop
-\cite{mcw-1997-otter} provers use both active and passive clauses whereas
-DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.}
-
-\section{Hidden-Literal-Based Elimination}
-\label{sec:satfol:hidden-literal-based-elimination}
-
-In propositional logic, binary clauses from a clause set $N$ can be used to
-efficiently discover literals $L, L'$ for which the implication $L' \impl L$
-is entailed by $N$'s binary clauses---i.e., $\binset{N} \models L' \impl L$.
-Heule et al.\ \cite{hjb-2011-big-simplification} introduced the concept of \emph{hidden
-literals} to capture such implications.
-
-\begin{defi}
-   Given a propositional literal $L$ and a propositional clause set~$N$, the
-   set of \emph{propositional hidden literals} for $L$ and~$N$ is
-   $\PHL(L, N) = \{ L' \mid L' \bigpimpl{N}^\ast L \} \setminus \{ L \}$, where $\bigpimpl{N}$ is
-   defined such that $\neglit{L_1} \bigpimpl{N} L_2$ whenever $L_1 \cor L_2 \in N.$
-   Moreover, $\PHL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n
-   \PHL(L_i, N)$.
-\end{defi}
-
-Heule et al.\ used a fixpoint computation, but our definition based on the
-reflexive transitive closure is equivalent. Intuitively, a hidden literal can be added to
-or removed from a clause without affecting its semantics in models of~$N.$
-By eliminating hidden literals from $C$, we simplify it. By adding hidden literals to $C$, we might get a tautology $C'$ (i.e., a valid clause:\ $\models
-C'$), meaning that $\binset{N} \models C$, thereby enabling us to delete $C$.
-Note that $\PHL(L,N)$ is finite for a finite~$N.$
-
-\begin{defi}
-   \label{def:hle}\label{def:hte}
-   Given $L' \cor L \cor C \in N,$ if $L' \in
-   \PHL(L, N)$, \emph{hidden literal elimination} (HLE) replaces $N$ by $(N
-   \setminus \{L' \cor L \cor C\}) \cup \{L \cor C\}$.
-   Given $C \in N$, $\{L_1,\dots,L_n\} = \PHL(C, N)$, and $C' = C \cor L_1 \cor
-   \cdots \cor L_n$, if $C'$ is a tautology, \emph{hidden
-   tautology elimination} (HTE) replaces $N$ by $N \setminus \{C\}$.
-\end{defi}
-
-\begin{theorem}
-   The result of applying HLE or HTE to a clause set $N$ is
-   equivalent to $N$.
-\end{theorem}
-
-\begin{proof}
-For HLE, if $L' \in \PHL(L, N)$,  $\binset{N} \models \neglit{L'} \cor L$.
-Then, subsumption resolution yields shortened clause $L \cor C'$ from
-Definition~\ref{def:hle}. For HTE, it can be shown that $N' \models C$ if and
-only if $C \cor L'$, where $L' \in \PHL(C, N)$. By transitivity of equivalence,
-we get the desired result.
-\qed
-\end{proof}
-
-We generalize hidden literals to first-order logic with equality by
-considering substitutivity of variables as well as congruence of equality.
-
-\begin{defi}
-\label{def:hl-fo}
-Given a literal $L$ and a clause set $N$,
-the set of \emph{hidden literals} for $L$ and $N$
-is $\HL(L,N) = \{ L' \mid L' \bigfimpl{N}^\ast L \} \setminus \{ L \}$, where
-$\bigfimpl{N}$ is defined so that
-(1)~$\neglit{L'\sigma} \bigfimpl{N} L\sigma$ if
-$L' \cor L \in N$ and $\sigma$ is a substitution;
-(2)~$\eqlit{s}{t} \bigfimpl{N} \eqlit{u[s]}{u[t]}$ for all terms $s, t$ and
-contexts $u[\phantom{\cdot}]$; and
-(3)~$\neqlit{u[s]}{u[t]} \bigfimpl{N} \neqlit{s}{t}$ for all terms $s, t$ and
-contexts $u[\phantom{\cdot}]$.
-Moreover, $\HL(L_1 \cor \cdots \cor L_n, N) = \bigcup_{i=1}^n \HL(L_i, N)$.
-\end{defi}
-
-\looseness=-1
-The generalized definition also enjoys the key property that $L'
-\in \HL(L, N)$ implies $\binset{N} \models L' \impl L$. However, $\HL(L, N)$
-may be infinite even for predicate literals; for example,
-$\pospredlit{\cst{p}(\cst{f}^i(x))} \in \HL(\pospredlit{\cst{p}(x)}, \{\pospredlit{\cst{p}(x)} \cor
-\negpredlit{\cst{p}(\cst{f}(x))} \})$ for every $i$.
-
-Based on Definition \ref{def:hl-fo}, we can generalize hidden literal elimination
-and support a related technique:
-%
-\begin{align*}
-   \vcenter{\namedsimp{HLE}{L' \cor L \cor C}{L \cor C}}
-   & \scif {L' \in \HL(L,N)} \\[.5\jot]
-   \vcenter{\namedsimp{\rlap{FLE}\phantom{HLE}}{L \cor C}{C}}
-   & \scif {L',\neglit{L'} \in \HL(\neglit{L}, N)}
-\end{align*}
-%
-Double lines denote \emph{simplification rules}: When the premises appear in
-the clause set, the prover can use the redundancy criterion to replace
-them by the conclusions. The second rule is called \textit{failed literal
-elimination},
-inspired by the SAT technique of asserting $\neglit{L}$ if $L$ is a \textit{failed literal} \cite{jwf-1995-fld}.
-It is easy to see that rule \infname{HLE} is sound.
-From $L' \in \HL(L,N)$ we have $N \models L' \impl L$ (i.e., $\neglit{L'} \cor L$).
-Performing subsumption resolution \cite{bg-01-resolution} between $L'
-\cor L \cor C$ and $\neglit{L'} \cor L$ yields the conclusion, which is
-therefore entailed by $N.$ For \infname{FLE}, the condition $L',
-\neglit{L'} \in \HL(\neglit{L},N)$ means that $\binset{N} \models \{\neglit{L'} \cor
-\neglit{L}{,}\; L' \cor \neglit{L}\} \models \neglit{L}$.
-
-\begin{exa}
-   \label{example:hle}
-   Consider the clause set $N = \{ \cst{p}(x) \cor \neglit{\cst{p}(\cst{f}(x))}{,}\;
-   \cst{p}(\cst{f}(\cst{f}(x))) \cor \eqlit{\cst{a}}{\cst{b}} \}$ and the clause
-   $C = \neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \cor \cst{p}(x)$. The first
-   clause in $N$ induces $\cst{p}(\cst{f}(x)) \bigfimpl{N} {\cst{p}(x)}$,
-   $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N} \cst{p}(\cst{f}(x))$, and
-   hence $\cst{p}(\cst{f}(\cst{f}(x))) \bigfimpl{N}^\ast  \cst{p}(x)$.
-   Together with the second clause in $N$, it can be used to derive $\neqlit{\cst{a}}{\cst{b}}
-   \bigfimpl{N}^\ast  \cst{p}(x)$. Finally, using rule~(3) of Definition
-   \ref{def:hl-fo}, we derive $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})}
-   \bigfimpl{N}^\ast  \cst{p}(x)$---that is,
-   $\neqlit{\cst{f}(\cst{a})}{\cst{f}(\cst{b})} \in \HL(\cst{p}(x), N)$. This
-   allows us to remove $C$'s first literal using \infname{HLE}.
-\end{exa}
-
-Two special cases of \infname{HLE} exploit equality congruence as embodied by
-conditions (2)~and~(3) of Definition~\ref{def:hl-fo} without requiring to
-compute the $\HL$~set:
-\begin{align*}
-  %% TYPESETTING: \,'s below
-  \namedsimp{CongHLE$^+$}{\eqlit{s}{t} \cor \eqlit{u[s]}{u[t]} \cor C}{\eqlit{u[s]}{u[t]} \cor C}\,\, \\[.5\jot]
-  \namedsimp{CongHLE$^-$}{\neqlit{s}{t} \cor \neqlit{u[s]}{u[t]} \cor C}{\neqlit{s}{t} \cor C}
-\end{align*}%
-
-Hidden literals can be combined with unit clauses~$L'$ to remove more literals:
-\[\namedsimpsc{UnitHLE}{L' \quad L \cor C}{L' \quad C}{L'\sigma \in \HL(\neglit{L},N)}
-%\text{~for some~}\sigma -- %%% Not necessary in a rule -- all variables are universally quantified. --JB
-\]
-%
-Given a unit clause $L' \in N$, the rule uses it to discharge $L'\sigma$ in
-$N \models L'\sigma \impl \neglit{L}$. As a result, we have $N \models
-\neglit{L}$, making it possible to remove $L$ from $L \cor C$.
-
-% Kiesl and Suda described first-order generalization of {\it asymmetric literals}
-% \cite{ks-2017-unif-principle}: literal $L\sigma$ is asymmetric for a clause $C$
-% and a clause set $N$ if there exists a clause $D \cor \neglit{L} \in N \setminus
-% \{C \}$ such that $D\sigma \subseteq C$ for some substitution $\sigma$.
-
-\begin{exa}
-   \label{example:unithle}
-   \looseness=-1
-   Consider the clause set $N = \{\cst{p}(x) \cor \cst{q}(\cst{f}(x)){,}\;
-   \neglit{\cst{q}(\cst{f}(\cst{a}))} \cor \eqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}{,}\;
-   \neqlit{\cst{f}(x)}{\cst{g}(y)}\}$ and the clause $C = \neglit{\cst{p}(\cst{a})} \cor
-   \neglit{\cst{q}(\cst{b})}$. The first clause in $N$ induces
-   $\neglit{\cst{q}(\cst{f}(\cst{a}))} \bigfimpl{N} \cst{p}(\cst{a})$, whereas
-   the second one induces $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})} \bigfimpl{N}
-   \neglit{\cst{q}(\cst{f}(\cst{a}))}$. Thus, we have $\neqlit{\cst{f}(\cst{b})}{\cst{g}(\cst{c})}
-   \bigfimpl{N}^\ast \cst{p}(\cst{a})$---that is, $\neqlit{\cst{f}(\cst{b})}{\cst{f}(\cst{c})} \in
-   \HL(\cst{p}(\cst{a}), N)$. By applying the substitution $\{x \mapsto \cst{b}{,}\allowbreak\; y
-   \mapsto \cst{c}\}$ to the third clause in $N$, we can fulfill the conditions of
-   \infname{UnitHLE} and remove $C$'s first literal.
-\end{exa}
-
-Next, we generalize hidden tautologies to first-order logic.
-
-\begin{defi}
-   A clause $C$ is a \emph{hidden tautology} for a clause set $N$ if there
-   exists a finite set $\{L_1,\dots,L_n\} \subseteq \HL(C,N)$ such that
-   $C \cor L_1 \cor \cdots \cor L_n$ is a tautology.
-\end{defi}
-
-\begin{exa}
-   \label{example:hidden-tautologies-completeness}
-   \looseness=-1
-   In general, hidden tautologies are not redundant and cannot be deleted during saturation.
-   Consider the unsatisfiable set $N = \{ \negpredlit{\cst{a}}{,}\;
-   \negpredlit{\cst{b}}{,}\; \pospredlit{\cst{a}} \cor
-   \pospredlit{\cst{c}}{,}\allowbreak\; \pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}
-   \}$, the order $\cst{a} \prec
-   \cst{b} \prec \cst{c}$, and the empty selection function. The
-   only possible superposition inference from $N$ is between the last two clauses,
-   yielding the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
-   (after simplifying away $\neqlit{\top}{\top}$), which is entailed by the larger
-   clauses $\pospredlit{\cst{a}} \cor \pospredlit{\cst{c}}$ and
-   $\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}$. If this clause is removed,
-   the prover could enter an infinite loop, forever generating and deleting the
-   hidden tautology\confrep{}{ and never getting the opportunity to derive the
-   empty clause}.
-\end{exa}
-
-\begin{rep}
-   \newcommand{\aset}{\mathcalx{A}}
-   \newcommand{\pset}{\mathcalx{P}}
-
-   In practice, most provers use a variant of the given clause procedure.
-   Removing hidden tautologies breaks the invariant of the procedure that all
-   inferences between clauses in $\aset$ are redundant. The end result is not
-   that the prover diverges, but that it terminates without deriving empty
-   clause.
-
-   To observe this, assume the setting as in Example
-   \ref{example:hidden-tautologies-completeness}, and let $\pset=N$ and
-   $\aset=\emptyset$. After moving the first three clauses from $\pset$ to
-   $\aset$ ($\aset = \{ \negpredlit{\cst{a}}, \negpredlit{\cst{b}}, \pospredlit{\cst{a}} \cor
-   \pospredlit{\cst{c}} \}$, $\pset = \{\pospredlit{\cst{b}} \cor \negpredlit{\cst{c}}\}$), 
-   no inferences are possible, and no new clauses are added to $\pset$. After the last clause
-   is moved to $\aset$, the hidden tautology $\pospredlit{\cst{a}} \cor \pospredlit{\cst{b}}$
-   is produced. If it is deleted, the prover terminates with the unsatisfiable set $\aset$,
-   but does not derive the empty clause. 
-\end{rep}
-
-To delete hidden tautologies during saturation, the prover could check that all
-the relevant clause instances encountered along the computation of $\HL$ are
-$\prec$-smaller than a given hidden tautology. However, this would be expensive
-and seldom succeed, given that superposition creates lots of nonredundant
-hidden tautologies. Instead, we propose to simplify hidden tautologies using the
-following rules:
-%
-%\begin{linenomath*}
-% \begin{align*}
-%    %% TYPESETTING: \,'s below
-%    \namedsimpsc{HTR}{L \cor L' \cor C}{L \cor L'}{\neglit{L'} \in \HL(L, N)} &\\
-%    \namedsimpsc{FLR}{L \cor C}{L}{L',\neglit{L'} \in \HL(L, N)} &
-% \end{align*}
-%
-\[\begin{aligned}[t]
-   \vcenter{\namedsimp{HTR}{L \cor L' \cor C}{L \cor L'}}
-   & \scif {\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot} \\[.5\jot]
-   \vcenter{\namedsimp{\rlap{FLR}\phantom{HTR}}{L \cor C}{L}}
-   & \scif {L',\neglit{L'} \in \HL(L, N) \text{~and~} C \not= \bot}
-\end{aligned}\]
-%
-
-%\end{linenomath*}
-%
-We call these techniques \emph{hidden tautology reduction} and \emph{failed literal
-reduction}, respectively.
-% \infname{FLD} is inspired by the \confrep{}{eponymous }SAT
-% approach of asserting $L$ if $\neglit{L}$ is determined to be a failed literal
-% \cite{jwf-1995-fld}.
-%
-Both rules are sound.
-%
-As with hidden literals, unit clauses~$L'$ can be exploited:
-\[\namedsimpsc{UnitHTR}{L' \quad L \cor C}{L' \quad L}{L'\sigma \in \HL(L,N) \text{~and~} C \not= \bot}
-%\text{~for some~}\sigma
-\]
-
-We give the simplification rules above\confrep{}{ (for hidden literal
-elimination, hidden tautology reduction,
-failed literal detection, and their variants)} the collective name of
-\emph{hidden-literal-based elimination} (HLBE).
-Yet another use of hidden literals is for \emph{equivalent literal
-substitution} \cite{hjb-2010-cl-elim}: If both $L' \in \HL(L, N)$ and $L
-\in \HL(L', N)$, we can often simplify $L'\sigma$ to $L\sigma$ in $N$ if
-$L'\sigma \succ L\sigma$. We want to investigate this~further.
-
-% \looseness=-1
-% Provers based on the DISCOUNT loop use only active clauses for
-% simplification. If we restrict our attention to a finite subset of hidden
-% literals, we can organize them into data structures that support efficient
-% lookup of a hidden literal. Efficient implementation of these data structures
-% allows to use all passive clauses for simplification, strengthening DISCOUNT
-% loop simplification machinery.
-
-\begin{theorem}
-The rules \infname{HLE}, \infname{FLE}, \infname{CongHLE$^+$},
-\infname{Cong\allowbreak HLE$^-$}, \infname{UnitHLE}, \infname{HTR},
-\infname{FLR}, and \infname{UnitHTR} are sound simplification rules.
-\end{theorem}
-
-\begin{rep}
-\begin{proof}
-It is easy to see that the deleted premises are entailed by the conclusions that
-replace them and that the conclusions' instances are $\prec$-smaller than the
-premises' instances, as required by the redundancy criterion. It remains to
-check soundness.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{HLE}:\enskip
-We have $N{,}\> L' \models L$ by the side condition and must show
-$N{,}\> L' \lor L \lor C \models L \lor C$. Let
-$\ourmodel \models N{,}\> L' \lor L \lor C$. If $\ourmodel \models L'$, then
-we also have $\ourmodel \models L$ thanks to the side condition and hence
-$\ourmodel \models L \lor C$. Otherwise, we have $\ourmodel \models L \lor C$,
-which is exactly what we need to show.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{FLE}:\enskip
-We have $N{,}\> L \models L'$ and $N, L \models \lnot L'$ by the side condition.
-If $\ourmodel \models N{,}\> L$, then both $\ourmodel \models L'$ and
-$\ourmodel \models \lnot L'$, an absurdity.
-Otherwise, we have $\ourmodel \models C$, as desired.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{CongHLE$^+$}, \infname{CongHLE$^-$}:\enskip
-Obvious by congruence of equality.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{UnitHLE}:\enskip
-We have $N{,}\> L \models \lnot L'\sigma$ by the side condition. If $\ourmodel
-\models N{,}\> L$, then $N \models \lnot L'\sigma$. But since $L' \in N$, this is
-an absurdity. Otherwise, we have $\ourmodel \models C$, as desired.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{HTR}:\enskip
-We have $N{,}\> \lnot L' \models L$ by the side condition. If either $\ourmodel
-\models L$ or $\ourmodel \models L'$, the desired result follows directly.
-Otherwise, from $\ourmodel \models \lnot L'$ we also have $\ourmodel \models L$
-thanks to the side condition, contradicting $\ourmodel \models \lnot L$.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{FLR}:\enskip
-We have $N{,}\> L' \models L$ and $N{,}\> \lnot L' \models L$ by the side condition.
-Hence $N \models L$, as desired.
-
-\medskip
-
-\noindent
-\textsc{Case} \infname{UnitHTR}:\enskip
-We have $N{,}\> L'\sigma \models L$. Since $L' \in N$, we have $N \models L$,
-as desired.
-\qed
-\end{proof}
-\end{rep}
-
-\section{Predicate Elimination}
-\label{sec:satfol:predicate-elimination}
-
-For propositional logic, variable elimination \cite{eb-2005-satpreprocess} is
-one of the main preprocessing and inprocessing techniques. Following Gabbay and Ohlbach's ideas
-\cite{go-1992-so-pred-elim}, Khasidashvili and Korovin \cite{kk-2016-pe-fol}
-generalized variable elimination to first-order logic with equality and
-demonstrated that it is effective as a preprocessor. We propose an improvement
-that makes this applicable in more cases and show that, with a minor
-restriction, it can be integrated in a superposition prover
-without compromising its refutational completeness.
-
-\subsection{Singular Predicates}
-
-Khasidashvili and Korovin's preprocessing technique removes singular predicates
-(which they call ``non-self-referential predicates'') from the problem using
-so-called flat resolution.
-
-\begin{defi}
-   A predicate symbol is called \emph{singular}\confrep{ (or ``non-self-referential'')}{}
-   for a clause set $N$ if it occurs at most once in every clause contained in
-   $N.$
-\end{defi}
-
-\begin{defi}
-   \label{def:flat-res}
-   Let $C = \pospredlit{\cst{p}(\tuplen{s})} \cor C'$ and $D =
-   \negpredlit{\cst{p}(\tuplen{t})} \cor D'$ be clauses with no variables in
-   common. The clause $\neqlit{s_1}{t_1} \cor \cdots \cor \neqlit{s_n}{t_n}
-   \cor C' \cor D'$ is a \emph{flat resolvent} of $C$ and $D$ on $\cst{p}$.
-\end{defi}
-
-% Predicate elimination saturates with respect to flat resolvents, removing all
-% clauses containing $\cst{p}$. Given two (possibly identical) clause sets $M, N$,
-% the resolved set $M \flatres_{\!\cst{p}} N$, defined below, repeatedly draws
-% one inference partner from $M$ and the other one from $N$.
-
-Given two (possibly identical) clause sets $M, N$, predicate elimination
-iteratively replaces clauses from $N$ containing the symbol $\cst{p}$ with all flat
-resolvents against clauses in $M$. Eventually, it yields a
-set with no occurrences of $\cst{p}$.
-
-\begin{defi}
-\label{def:flat-res-set}
-   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
-   $M$. Let $\flatresiter$ be the following relation on clause set pairs and clause sets:
-   \begin{enumerate}
-   \item $(M{,}\> \{\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'\} \uplus N) \flatresiter (M{,}\; N' \cup N)$ if
-   $N'$ is the set that consists of all clauses (up to variable renaming) that are flat resolvents
-   with $\arbpredlit{\cst{p}(\tupleempty{s})} \cor C'$ on $\cst{p}$ and a clause from $M$ as premises.
-   The premises' variables are renamed apart.
-
-   \smallskip
-   \item $(M, N) \flatresiter N$ if $N$ has no occurrences of $\cst{p}$.
-   \end{enumerate}
-   \noindent  The \emph{resolved set} $M \flatres_{\!\cst{p}} N$ is the clause set $N'$ such that $(M,N) \flatresiter^* N'$.
-   % obtained by the following
-   % procedure: Set $N' = N$. Then repeatedly (1)~choose a clause $D \in N'$
-   % containing $\cst{p}$, (2)~compute all flat resolvents involving $D$ and
-   % partners from $M$, renaming variables apart, and (3)~replace $D$ with these
-   % flat resolvents in $N'$. Stop when $N'$ contains no more occurrence of
-   % $\cst{p}$.
-\end{defi}
-
-\pagebreak[2]
-
-\begin{rep}
-\begin{lemma}
-   \label{lem:flat-res-set-termination-confluence}
-   Let $M, N$ be clause sets and $\cst{p}$ be a singular predicate for
-   $M$. The resolved set $N'$ is reached in a finite number of $\flatresiter$ steps,
-   and it is unique up to variable renaming.
-\end{lemma}
-\begin{proof}
-   To show $\flatresiter$ is terminating we use the following ordinal measure on
-   clause sets: $\nu(\{D_1,\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus
-   \cdots \oplus \omega^{\nu(D_n)}$, where $\nu(D)$ is the number of
-   occurrences of $\cst{p}$ in $D$, $\omega$ is the first infinite ordinal, and
-   $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
-   For every transition
-   $(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have $\nu(N) > \nu(N')$
-   because $\omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'|$.
-   Eventually, a state $(M,N')$ with $\nu(N')=\omega^0\cdot n$ is reached.
-   Then, we apply the second rule of Definition \ref{def:flat-res-set} to obtain the resolved set $N'$.
-
-   To show that $N'$ is unique, i.e., $\flatresiter$ is confluent, it suffices
-   to show (since $\flatresiter$ is terminating and Newmann's lemma applies
-   \cite{bn-98-tr-and-all-that}) that $\flatresiter$ is locally confluent. In
-   other words, whenever $(M,N) \flatresiter (M, N_1)$ and $(M,N) \flatresiter
-   (M, N_2)$ there exists $N'$ such that $(M, N_1) \flatresiter (M, N')$ and
-   $(M, N_2) \flatresiter (M,N')$.
-
-   There are two main sources of nondeterminism of $\flatresiter$: The choice of $C
-   \in N$ and the choice of the literal in $C$ to act on. Let us focus on the choice of
-   $C$ in $N$; the same discussion applies for the choice of literal in $C$.
-
-   Let $N = \{C_1\} \uplus \{C_2\} \uplus N'$, where $C_1$ and $C_2$ are
-   clauses with occurrences of $\cst{p}$. Then, $(M, \{C_1\} \uplus \{ C_2 \cup
-   N' \}) \flatresiter (M, N'_1 \cup \{ C_2 \cup N'\})$ and $(M, \{C_2\} \uplus
-   \{C_1 \cup N' \}) \flatresiter (M, N'_2 \cup \{ C_1 \cup N' \})$ where $N'_1$ and
-   $N'_2$ are sets of corresponding resolvents. Both of $\flatresiter$ steps can
-   be joined (up to variable renaming) to $(M, N'_1 \cup N'_2 \cup N')$, showing
-   that $\flatresiter$ is locally confluent.\qed
-\end{proof}
-\end{rep}
-
-\begin{conf}
-The relation $\flatresiter$ is confluent up to variable renaming. Thanks to
-the singularity constraint on $M$, it
-also terminates on finite sets because the following ordinal measure decreases:
-$\nu(\{D_1,\allowbreak\dotsc,\allowbreak D_n\}) = \omega^{\nu(D_1)} \oplus \cdots \oplus
-\omega^{\nu(D_n)}$, where $\nu(D)$ counts the occurrences of $\cst{p}$ in
-$D$, $\omega$ is the first infinite ordinal,
-and $\oplus$ is the Hessenberg, or natural, sum, which is commutative.
-For every transition
-$(M, \{C\} \cup N) \flatresiter (M, N' \cup N)$, we have
-$\nu(\{C\}) = \omega^{\nu(C)} > \omega^{\nu(C)-1} \cdot |N'| = \nu(N')$.
-\end{conf}
-
-Next, it is useful to partition
-clause sets into subsets based on the presence and polarity of a singular
-predicate.
-
-\begin{defi}
-   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
-   $N.$ Let $\withpredpos{N}{\cst{p}}$ consist of all clauses of the form $\pospredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
-   let $\withpredneg{N}{\cst{p}}$ consist of all clauses of the form $\negpredlit{\cst{p}(\tuple{s}{})} \cor C' \in N$,
-   let $\withpred{N}{\cst{p}} = \withpredpos{N}{\cst{p}} \cup \withpredneg{N}{\cst{p}}$,
-   and let $\withoutpred{N}{\cst{p}} = N \setminus \withpred{N}{\cst{p}}$.
-\end{defi}
-
-\begin{defi}
-   \label{def:pred-elim}
-   Let $N$ be a clause set and $\cst{p}$ be a singular predicate for
-   $N.$
-   \emph{Singular predicate elimination} (SPE) of $\cst{p}$ in $N$ replaces $N$ by
-   $\withoutpred{N}{\cst{p}} \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
-\end{defi}
-
-The result of SPE is satisfiable if and only if $N$ is satisfiable
-\cite[Theorem~1]{kk-2016-pe-fol}, justifying SPE's use in a preprocessor.
-However, eliminating singular predicates aggressively can dramatically increase
-the number of clauses. To prevent this, Khasidashvili and Korovin suggested to
-replace $N$ by $N'$ only if $\lambda(N') \leq \lambda(N)$
-and $\mu(N') \leq \mu(N)$, where
-$\lambda(N)$ is the number of literals in~$N$ and
-$\mu(N)$ is the sum for all clauses $C \in
-N$ of the square of the number of distinct variables in $C$.
-
-Compared with what modern SAT solvers use, this
-criterion is fairly restrictive. We relax it to make it possible to
-eliminate more predicates, within reason. Let $K_\mathrm{tol} \in \mathbb{N}$ be a tolerance
-parameter. A predicate elimination
-step from $N$ to $N'$ is allowed if
-$\lambda(N') < \lambda(N) + K_\mathrm{tol}$ or
-$\mu(N') < \mu(N)$ or
-$|N'| < |N| + K_\mathrm{tol}$.
-\confrep{}{(A refinement, which we want to try out in future work, would be to gradually
-increment the tolerance $K_\mathrm{tol}$, as is done in some SAT solvers.)}
-
-\subsection{Defined Predicates}
-
-SPE is effective, but an important refinement has not yet been adapted to
-first-order logic:\ variable elimination by substitution. E{\'{e}}n and Biere
-\cite{eb-2005-satpreprocess} discovered that a propositional variable~$\cst{x}$ can be
-eliminated without computing all resolvents if it is expressible as an
-equivalence $\cst{x} \medleftrightarrow \varphi$, where $\varphi$, the ``gate,''
-is an arbitrary formula that does not reference~$\cst{x}$.
-They \confrep{partition}{extract from} a set $N$
-\confrep{}{of propositional clauses }into a definition set $G$, essentially the
-clausification of $\cst{x} \medleftrightarrow \varphi$, and
-$R = \withpred{N}{\cst{p}} \setminus G$, the remaining
-clauses containing~$\cst{p}$. To eliminate $\cst{x}$ from $N$ while
-preserving satisfiability, it suffices to resolve clauses from $G$ against
-clauses from $R$, effectively substituting $\varphi$ for $\cst{x}$ in $R$.
-Crucially, we do not need to resolve pairs of clauses from $G$
-or pairs of clauses from $R$.
-We generalize this idea to first-order logic.
-
-\begin{defi}
-   \label{def:definition}
-   \looseness=-1
-   Let $G$ be a clause set, $\cst{p}$ be a predicate symbol, and
-   $\tupleempty{x}$ be distinct variables.
-%
-   The set $G$ is a \emph{definition set} for $\cst{p}$ if (1)~$\cst{p}$
-   is singular for $G$, (2)~$G$ consists of clauses of the form
-   $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming), (3)~the
-   variables in $C'$ are all among $\tupleempty{x}$, (4)~all clauses in $\withpredpos{G}{\cst{p}}
-   \flatres_{\!\cst{p}} \withpredneg{G}{\cst{p}}$ are tautologies, and
-   (5)~$E(\tupleempty{\cst{c}})$ is unsatisfiable, where
-   the \emph{environment} $E(\tupleempty{x})$ consists of all subclauses $C'$ of any
-   $\arbpredlit{\cst{p}(\tupleempty{x})} \mathbin{\cor} C' \in G$ and
-   $\tupleempty{\cst{c}}$ is a tuple of distinct fresh constants substituted
-   in for $\tupleempty{x}$.
-\end{defi}
-
-A definition set $G$ corresponds intuitively to a definition by
-cases in mathematics---e.g.,
-\[\cst{p}(\tuple{x}{}) =
-\begin{cases} \top & \mathrm{if}~\varphi(\tuple{x}{}{}) \\[-\jot]
-\bot & \mathrm{if}~\psi(\tuple{x}{})\end{cases}\]
-Part~(4) states that the case conditions are mutually exclusive
-(e.g., $\neglit{\varphi(\tuple{x}{})} \lor \neglit{\psi(\tuple{x}{})}$),
-and part~(5) states that they are exhaustive
-(e.g., $\nexists\tuple{\cst{c}}.\; \neglit{\varphi(\tuple{\cst{c}}{})} \land \neglit{\psi(\tuple{\cst{c}}{})}$).
-%
-Given a quantifier-free formula $\cst{p}(\tuple{x}{}) \medleftrightarrow
-\varphi(\tuple{x}{})$ with distinct variables $\tuple{x}{}$ such that
-$\varphi(\tuple{x}{})$ does not contain $\cst{p}$, any reasonable
-clausification algorithm would produce a definition set for $\cst{p}$.
-
-\begin{exa}
-   \label{example:tautologies}
-   Given the formula $\cst{p}(x) \medleftrightarrow \cst{q}(x) \mathrel\land
-   (\cst{r}(x) \cor \cst{s}(x))$, a standard clausification algorithm
-   \cite{nw-01-small-cnf} produces $\{ \negpredlit{\cst{p}(x)} \cor
-   \pospredlit{\cst{q}(x)}{,}\; \negpredlit{\cst{p}(x)} \cor
-   \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
-   \pospredlit{\cst{p}(x)} \cor \negpredlit{\cst{q}(x)} \cor
-   \negpredlit{\cst{r}(x)}{,}\; \pospredlit{\cst{p}(x)} \cor
-   \negpredlit{\cst{q}(x)} \cor \negpredlit{\cst{s}(x)}\}$, which qualifies
-   as a definition set for $\cst{p}$.
-\end{exa}
-
-Definition sets generalize E{\'{e}}n and Biere's
-gates. They can be recognized syntactically for formulas such as
-$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigvee_{\!i} \cst{q}_i(\tuple{s_i}{})$ or
-$\cst{p}(\tuple{x}{}) \medleftrightarrow \bigwedge_i \cst{q}_i(\tuple{s_i}{})$,
-or semantically: Condition~(4) can be checked
-using the congruence closure algorithm, and condition~(5) amounts to
-a propositional unsatisfiability check.
-
-The key result about propositional gates carries over to definition sets.
-
-\begin{defi}
-   \looseness=-1
-   Let $N$ be a clause set, $\cst{p}$ be a predicate symbol,
-   $G \subseteq N$ be a definition set
-   for $\cst{p}$, and $R = \withpred{N}{\cst{p}}
-   \setminus G$. \emph{Defined predicate elimination} (DPE) of $\cst{p}$ in $N$ replaces $N$ by
-   $\withoutpred{N}{\cst{p}} \cup
-%%% @PETAR: Double-check the supression of _p below. --JB
-   (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}} \confrep{\withpred{R}{\cst{p}}}{R})$.
-\end{defi}
-
-\newcommand{\mm}{\ourmodel}
-\begin{rep}
-   \begin{lemma}
-   \label{lem:unsat-fresh-consts}
-   Let $N(\tupleempty{x})$ be a clause set such that the variables of all
-   clauses in it are among the argument $n$-tuple $\tupleempty{x}$, and let
-   $\tupleempty{\cst{c}}$ be an $n$-tuple of distinct fresh constants. If
-   $N(\tupleempty{\cst{c}})$ {\upshape(}i.e.,
-   $N(\tupleempty{x})\{\tupleempty{x}\mapsto\tupleempty{\cst{c}}\}${\upshape)} is
-   unsatisfiable, then for every interpretation $\mm$ and valuation $\xi$,
-   $\mm \notmodels_\xi N$.
-   \end{lemma}
-   \begin{proof}
-      We show the contrapositive. Assume that for some $\mm$ and $\xi$,  $\mm \models_\xi N(\tupleempty{x})$.
-      Then let $\ourmodel'$
-      be a model that assigns each $\cst{c}_i$ the interpretation of $x_i$ under $\mm$ and $\xi$, and otherwise
-      coincides with $\ourmodel$. We obtain $\ourmodel' \models N(\tupleempty{\cst{c}})$.\qed
-   \end{proof}
-
-   \begin{lemma}
-      \label{lem:flat-res-set-step-satisfiability}
-      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }an arbitrary clause
-      set. If $(G, N) \flatresiter (G, N')$ then $G \cup N$ and $G \cup N'$ are
-      equivalent.
-   \end{lemma}
-   \begin{proof}
-      Since flat resolution is sound, the nontrivial direction is to show that a
-      model $\mm$ of the set $G \cup N'$ is also a model of $G \cup N$. As the only clause in $N
-      \setminus N'$ is $C = \arbpredlit{\cst{p}(\tuple{s}{n})} \cor C'$ on which
-      the $\flatresiter$ step is performed, we must show $\mm \models C$.
-      % Without
-
-      % Let $\mm$ be canonical model of $G \cup N'$ \cite[Theorem~9.5.2]{mf-1996-fol} and let $C
-      % \in N \setminus N'$. We need to show that $\mm \models C$---which is to say,
-      % for all variable interpretation $\xi$, we have $\mm \models_\xi C$. To
-      % lighten notations, we will assume $C$ is ground and ignore $\xi$. Canonical
-      % models make this step possible.
-
-      % First, $C$ must have the form $\arbpredlit{\cst{p}(\tuple{s}{})} \cor C'$,
-      % where $\tuple{s}{}$ is a tuple of ground terms. 
-      Without loss of generality,
-      we assume that the leading literal of $C$ is positive.
-      Towards a contradiction, assume $\xi$ is a valuation such that $\mm \notmodels_\xi C$. Then, $\mm \notmodels_\xi
-      \pospredlit{\cst{p}(\tuple{s}{n})}$. 
-      % Let $\xi'$ be a valuation that
-      % assigns every instance $D = \pospredlit{\cst{p}(\tupleempty{s})} \cor D'$ of a clause in
-      % $G$, since $\mm \notmodels \pospredlit{\cst{p}(\tuple{s}{})}$, we have $\mm
-      % \models D'$.
-      Consider an arbitrary clause $D = \pospredlit{\cst{p}}(\tuplen{x}) \cor D'
-      \in \withpredpos{G}{\cst{p}}$ and a valuation $\xi'$, which assigns each $x_i$ the
-      interpretation of $s_i$ under $\mm$ and $\xi$. As $\mm \notmodels_{\xi'}
-      \pospredlit{\cst{p}}(\tuple{x}{n})$ and $\mm \models G$, then $\mm \models_{\xi'} D'$ for every
-      such clause $D$.
-   %
-      However, by part (5) of Definition~\ref{def:definition} and by Lemma
-      \ref{lem:unsat-fresh-consts}, $\mm \notmodels_{\xi'} E(\tuplen{x})$, where $E(\tuplen{x})$
-      is the environment associated with the definition set $G$.
-      Therefore, there must exist a clause $D =
-      \negpredlit{\relax{\cst{p}(\tuplen{x})}} \cor D'$ in $\withpredneg{G}{\cst{p}}$ such
-      that $\mm \notmodels_\xi D'$.
-
-      Now consider the flat resolvent of $C$ and $D$ on~$\cst{p}$: $R =
-      \neqlit{x_1}{s_1} \cor \cdots \cor \neqlit{x_n}{s_n} \cor C' \cor D'$.
-      Let $\zeta$ be a valuation coinciding with $\xi$ on variables of $C$
-      and with $\xi'$ on $\tuplen{x}$.
-      Clearly, $\mm \notmodels_\zeta R$. Yet, $R \in N'$, and as $\mm \models N'$, we reach a contradiction.\qed
-   \end{proof}
-
-   \begin{lemma}
-      \label{lem:flat-res-set-last-step-satisfiability}
-      Let $G$ be a definition set for $\cst{p}$\confrep{,}{} and $N$ \confrep{}{be }a clause set with no occurrences of $\cst{p}$.
-      Then $G \cup N$ is satisfiable if and only if $N$ is satisfiable.
-   \end{lemma}
-   \begin{proof}
-      The nontrivial direction is to show that if $N$ is satisfiable, $G \cup N$ is as well.
-      %
-      Let $\mm$ be a model of $N$. We construct a model $\mm'$ of $G$ over
-      the same universe as $\mm$. For any atom $A$ such that $\cst{p}$ does not
-      occur in $A$ and for every $\xi$, we set $\mm' \models_\xi A$ if and only if
-      $\mm \models_\xi A$. For any clause $\pospredlit{\cst{p}(\tuplen{x})} \cor
-      C' \in G$ and any assignment $\xi$ such that $\mm \notmodels_\xi C'$, we
-      define $\mm'$ so that $\mm'\models_\xi \cst{p}(\tuplen{x})$. By
-      construction, $\mm' \models \withpredpos{G}{\cst{p}} \cup N$. It remains to show that $\mm' \models
-      \withpredneg{G}{}$.
-
-      Let $C = \negpredlit{\cst{p}(\tuplen{x})} \cor C' \in G$ and let $\xi$ be
-      an arbitrary assignment. Towards a contradiction, assume
-      $\mm'\notmodels_\xi C$, and consequently $\mm' \models_\xi
-      \cst{p}(\tuplen{x})$. By construction of $\mm'$, there exists a clause
-      $\pospredlit{\cst{p}(\tuplen{y})} \cor D' \in G$ and an assignment $\xi'$
-      which assigns each $y_i$ value of $\xi(x_i)$ such that $\mm
-      \notmodels_{\xi'} D'$. The resolvent $R = \neqlit{x_1}{y_1} \lor \cdots \lor
-      \neqlit{x_n}{y_n} \lor C' \lor D'$ is a tautology, according to condition
-      (4) of Definition \ref{def:definition}. However, for a valuation that
-      behaves like $\xi$ on $\tupleempty{x}$ and $\xi'$ on $\tupleempty{y}$,
-      $\mm'$ does not satisfy $R \in N$, contradicting our assumption.
-      % @Jasmin: Your trick with var renaming does not work as flat resolvent is
-      % defined on
-      % variable-disjoint clauses.
-
-   \end{proof}
-\end{rep}
-\begin{conf}
-
-\end{conf}
-\begin{theorem}
-\label{thm:pes-sat-equiv}
-   The result of applying DPE to a clause set $N$ is
-   satisfiable if and only if $N$ is satisfiable.
-\end{theorem}
-\begin{rep}
-\begin{proof}
-   Let $\cst{p}$ be a predicate symbol and $G \subseteq N$ be the
-   definition set used by DPE, and let $R = \withpred{N}{\cst{p}} \setminus G$.
-%
-   % First, using Lemmas \ref{lem:flat-res-set-termination-confluence} and \ref{lem:flat-res-set-step-satisfiability}
-   % we show that we will eventually obtain a set $R'$
-   % that contains no occurrences of $\cst{p}$ and that preserves satisfiability
-   % of the original $R$. Finally, we will show that we can
-   % omit $G$ in the result without affecting satisfiability.
-
-   Using Lemma
-   \ref{lem:flat-res-set-termination-confluence}, we get that there is a
-   derivation $(G,R) \flatresiter^n (G,R') \flatresiter R'$. Applying Lemma
-   \ref{lem:flat-res-set-step-satisfiability} $n$ times, we get that $G \cup R$
-   is equivalent to $G \cup R'$. Finally, Lemma \ref{lem:flat-res-set-last-step-satisfiability}
-   gives us the desired result.
-   \qed
-\end{proof}
-\end{rep}
-
-Since there will typically be at most only a few defined predicates in the
-problem, it makes sense to fall back on SPE when no definition is found.
-
-\begin{defi}
-   \looseness=-1
-   Let $N$ be a clause set and $\cst{p}$ be a predicate symbol. If there exists a
-   definition set $G \subseteq N$ for $\cst{p}$, \emph{portfolio predicate
-   elimination} (PPE) on $\cst{p}$ in $N$ replaces $N$ with
-%%% @PETAR: Double-check the supression of _p below. --JB
-   $\withoutpred{N}{\cst{p}} \cup (\confrep{\withpred{G}{\cst{p}}}{G} \flatres_{\!\cst{p}}
-   \confrep{\withpred{R}{\cst{p}}}{R})$, where $R = \withpred{N}{\cst{p}} \setminus G$.
-   Otherwise, if $\cst{p}$ is singular in $N$, it results in $\withoutpred{N}{\cst{p}}
-   \cup (\withpredpos{N}{\cst{p}} \flatres_{\!\cst{p}} \withpredneg{N}{\cst{p}})$.
-   In all other cases, it is not applicable.
-\end{defi}
-
-\subsection{Refutational Completeness}
-\label{ssec:predicate-elimination-refutational-completeness}
-
-\newcommand\FInf{\mathit{FInf}}
-\newcommand\FLInf{\mathit{FLInf}}
-\newcommand\Red{\mathit{Red}}
-\newcommand\RedI{\Red_\mathrm{I}}
-\newcommand\RedF{\Red_\mathrm{F}}
-\newcommand\LRed{\mathit{LRed}}
-\newcommand\LRedI{\LRed_\mathrm{I}}
-\newcommand\LRedF{\LRed_\mathrm{F}}
-\newcommand\gnd{\mathcalx{G}}
-\newcommand\pow[1]{\mathcalx{pe}(#1)}
-\newcommand\Lab{\mathbf{L}}
-\newcommand\FFLab{{\mathbf{FL}}}
-\newcommand\Inv{\mathit{Inv}}
-
-\newcommand\RedWk{\smash{\Red\Wksup}}
-\newcommand\RedIWk{\smash{\Red_\mathrm{I}\Wksup}}
-\newcommand\RedFWk{\smash{\Red_\mathrm{F}\Wksup}}
-
-\newcommand\Wksym{\flat}
-\newcommand\Wksup{^\Wksym}
-\newcommand\modelsWk{\models\Wksup}
-%\newcommand\modelsWkcapgnd{\models^{\Wksym{}\gnd}_\cap}
-\newcommand\modelsWkcapgnd{\modelsWk}
-\newcommand\LRedWk{\LRed^{\Wksym}}
-\newcommand\LRedWksub{\LRed^{\Wksym,\sqsupset}}
-\newcommand\LRedFWksub{\LRedF^{\Wksym,\sqsupset}}
-\newcommand\LRedIWk{\LRedI^{\Wksym}}
-\newcommand\Wk[1]{#1^\Wksym}
-\newcommand\prop{o}
-\newcommand\true{\top}
-\newcommand\false{\bot}
-
-\looseness=-1
-Hidden-literal-based techniques fit within the traditional framework of
-saturation, because they delete or reduce a clause based on the \emph{presence}
-of other clauses. In contrast, predicate elimination relies on the
-\emph{absence} of clauses from the proof state. We can still integrate it with
-superposition as follows: At every $k$th iteration of the given clause
-procedure, perform predicate elimination on $\mathcalx{A}
-\cup \mathcalx{P}$, and add all new clauses to $\mathcalx{P}$.
-
-One may wonder whether such an approach preserves the refutational
-completeness of the calculus. The answer is no.
-%
-\begin{conf}
-To see why, consider the following \emph{binary splitting} rule
-based on Riazanov and Voronkov \cite{riazanov-voronkov-2001}:
-%
-\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
-%
-Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
-$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
-smaller than the premise, the rule can be applied aggressively as a
-simplification. But notice that the effect of splitting can be undone by
-singular predicate elimination, possibly giving rise to loops
-$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
-This breaks completeness.
-
-Our solution is to curtail the entailment relation used by the redundancy
-criterion to disallow splitting-like simplifications. Weak entailment $\modelsWk$
-is defined via an ad hoc nonclassical logic so that $\{\cst{p} \cor C{,}\;
-\neglit{\cst{p}} \cor C\} \notmodelsWk \{C\}$ and yet $\modelsWk \{\cst{p} \cor
-\neglit{\cst{p}}\}$.
-%
-More precisely, this logic is defined via an
-encoding: $M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$, where
-%
-$\Wk{\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \false$,
-$\Wk{\lnot\,\cst{p}(\tuple{t}{})} = \cst{p}(\tuple{t}{}) \noteq \true$, and
-$\Wk{L} = L$ otherwise.
-%
-Moreover, the type $\prop$ may be interpreted as any set of cardinality at least
-2, and $\false$ must be a distinguished symbol interpreted differently
-from~$\top$.
-
-The standard redundancy criterion $\RedWk$ based on $\modelsWk$ supports all the
-familiar deletion and simplification techniques except splitting. Using $\RedWk$
-not only prevents looping, but it also enables the use of
-the given clause procedure, because any redundant inference
-according to $\RedWk$ remains redundant after SPE or DPE. As usual, the devil is
-in the details, and the details are in the report \cite{our-report}.
-\end{conf}
-
-\begin{rep}
-To see why, consider the following \emph{binary splitting} rule
-based on Riazanov and Voronkov \cite{riazanov-voronkov-2001}:
-%
-\[\namedsimp{BS}{C \cor D}{\cst{p} \cor C \quad D \cor \neglit{\cst{p}}}\]
-%
-Provisos: $C$ and $D$ have no free variables in common, $\cst{p}$ is fresh, and
-$\cst{p}$ is $\prec$-smaller than $C$ and $D$. Since the conclusions are
-smaller than the premise, the rule can be applied aggressively as a
-simplification. But notice that the effect of splitting can be undone by
-singular predicate elimination, possibly giving rise to loops
-$\infname{BS}, \infname{SPE}, \infname{BS}, \infname{SPE}, \dotsc$.
-Clearly, we need to curtail predicate elimination.
-
-Under which conditions is predicate elimination refutationally complete? To
-answer this question, we employ the saturation framework of Waldmann, Tourret,
-Robillard, and Blanchette \cite{wtrb-20-sat-framework}. Let
-$(\FInf, \Red)$ be the base calculus without predicate elimination---e.g.,
-resolution or superposition inferences together with the standard
-redundancy criterion \cite[Sect.~4.2]{bg-01-resolution}. The
-inference system $\FInf$ is a set of inferences $(C_n,\ldots,C_1,C_0)$, for $n
-\ge 1$, where $C_n,\ldots,C_1$ are the premises and $C_0$ is the conclusion.
-$C_1$ is called the main premise. %; $C_n,\ldots,C_2$ are the side premises.
-The redundancy criterion is a pair $\Red = (\RedI, \RedF)$ where $\RedI$
-determines which inferences can be omitted and $\RedF$ is used to remove
-clauses.
-
-Next, consider an abstract proving process working on a single clause set.
-Let $\rhd_{\Red}$ denote the transition relation that supports (1)~adding
-arbitrary clauses and (2)~removing clauses deemed useless by $\RedF$.
-Typically, the added clauses are the result of performing inferences and are
-entailed by the premises, but other clauses can be added as well. A
-\emph{$\rhd_{\Red}$-derivation} is a finite or infinite sequence of clause sets
-$N_0 \rhd_{\Red} N_1 \rhd_{\Red} \cdots$.
-
-We fix a finite set $\mathbf{P}$ of predicate symbols that may be subjected to
-predicate elimination. These might include all the predicate symbols occurring
-in the input problem, but exclude any symbols introduced by splitting or other
-rules. Given a clause or clause set $N$, we write $\mathbf{P}(N)$ to denote the
-set of all predicate symbols from $\mathbf{P}$ occurring in $N.$
-%
-Let $\rhd_\mathbf{P}$ denote the elimination of a
-singular or defined predicate symbol from $\mathbf{P}$. A \emph{mixed
-derivation} consists of transitions either of the form $N
-\rhd_\mathbf{P} N'$ or of the form $N \rhd_{\Red} N'$ where
-$\mathbf{P}(N) \supseteq \mathbf{P}(N')$.
-Because $\mathbf{P}$ is finite, any mixed derivation consists of at most
-finitely many $\rhd_\mathbf{P}$ transitions. Hence, in any derivation,
-there exists an index~$k$ from which all transitions are standard
-${\rhd}_{\Red}$-transitions.
-
-This suggests the following path to completeness: Pretend that the transitions
-between $N_0$ and $N_k$ are merely preprocessing and start the actual
-derivation at $N_k$. This works at the abstract level of derivations on single
-clause sets. It fails, however, for an actual saturation prover that
-distinguishes between passive and active clauses.
-
-\begin{exa}
-\label{ex:cex-predelim-sat}
-The counterexample below is based on the given clause prover \textsf{GC}
-from the saturation framework. It shows how predicate elimination can break
-\textsf{GC}'s key invariant, which states that all inferences between active
-clauses are redundant. Breaking the invariant means that the limit might be
-unsaturated, breaking the refutational completeness proof.
-
-We use superposition with the order $\cst{a} \prec
-\cst{b} \prec \cst{c} \prec \cst{d}$ and without selection.
-Assume $\cst{a} \in \mathbf{P}$ and
-suppose we start with the satisfiable clause set %consisting of
-%
-%\begin{linenomath*}
-\begin{align*}
-  & \neglit{\cst{a}} \cor \MAX{\cst{d}}
-  && \neglit{\cst{a}} \cor \MAX{\neglit{\cst{d}}}
-  && \cst{a} \cor \cst{b} \cor \MAX{\cst{c}}
-  && \cst{c} \cor \MAX{\cst{d}}
-  && \cst{b} \cor \MAX{\neglit{\cst{d}}}
-\end{align*}
-%\end{linenomath*}
-%
-where gray boxes mark maximal literals. Suppose the prover makes
-$\cst{c} \cor \cst{d}$ and $\cst{b} \cor \neglit{\cst{d}}$ active. From these
-two clauses, a superposition inference~$\iota$ could derive the conclusion
-$\cst{b} \cor \cst{c}$. However, the three passive clauses are
-$\prec$-smaller than $\iota$'s main premise $\cst{b} \cor \neglit{\cst{d}}$ and
-collectively entail $\iota$'s conclusion. This means that $\iota$ is
-redundant and can be ignored.
-
-If the prover now eliminates the predicate $\cst{a}$ using \infname{SPE}, the
-passive set is reduced to $\{\cst{b} \cor \cst{c} \cor \cst{d}{,}\;
-\cst{b} \cor \cst{c} \cor \neglit{\cst{d}}\}$.
-%
-Either clause is subsumed by an active clause, so the prover deletes it.
-It stops with the active set
-$\{\cst{c} \cor \cst{d}{,}\; \cst{b} \cor \neglit{\cst{d}}\}$, which is
-unsaturated because $\iota$ is no longer redundant. The invariant is broken.
-\end{exa}
-
-\begin{exa}
-\label{ex:cex-predelim-unsat}
-In Example~\ref{ex:cex-predelim-sat}, the initial clause set was satisfiable.
-If it is unsatisfiable, we can even lose refutational
-completeness. To see why, we add the unit clauses
-$\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ to the initial clause set of
-Example~\ref{ex:cex-predelim-sat} to make it unsatisfiable. We repeat the same
-steps as above, including the subsumptions at the end,
-yielding the passive set $\{\MAX{\neglit{\cst{b}}}, \MAX{\neglit{\cst{c}}}\}$ and
-the active set $\{\cst{c} \cor \MAX{\cst{d}}{,}\; \cst{b} \cor \MAX{\neglit{\cst{d}}}\}$. Then,
-making $\neglit{\cst{b}}$ and $\neglit{\cst{c}}$ active triggers no
-inferences. The prover stops with an
-unsatisfiable four-clause active set that does not contain the empty clause.
-\end{exa}
-
-A solution could be to move all active clauses to the passive set at step~$k$
-or later, but this would be costly, since it would force the prover to redo
-inferences whose conclusions might then have to be simplified or subsumed again.
-Instead, we salvage the existing completeness proof for $\mathsf{GC}$, by
-resolving the issues concerning splitting and the $\mathsf{GC}$ invariant. Our
-approach is to weaken the redundancy criterion slightly, enough both to disable
-splitting on $\mathbf{P}$-predicates and to ensure that
-inferences such as $\iota$ in Examples
-\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} are performed. The
-required weakening is so mild that it does not invalidate any practical
-simplification or subsumption techniques we are aware of, except of course
-splitting.
-
-In accordance with the saturation framework, let $\mathbf{F}$ be the set of
-first-order $\mathrm{\Sigma}$-clauses, let $\mathbf{G}$ be its ground subset,
-and let $\gnd$ be a function that maps an $\mathbf{F}$-clause to the set of its
-$\mathbf{G}$-clause instances and analogously for $\mathbf{F}$-inferences. We
-define an extension $\mathbf{G}\Wksup$ of $\mathbf{G}$ for
-$\mathrm{\Sigma}\Wksup$-clauses in an ad hoc nonclassical logic reminiscent
-of paraconsistent logics \cite{carnielli-et-al-2007}. The objective is to
-disallow the entailment that makes splitting and Examples
-\ref{ex:cex-predelim-sat}~and~\ref{ex:cex-predelim-unsat} possible.
-%
-The signature~$\mathrm{\Sigma}\Wksup$ extends $\mathrm{\Sigma}$ with a
-distinguished predicate symbol $\false$ that is interpreted differently
-from~$\top$. For $\mathrm{\Sigma}\Wksup$, the Boolean type $\prop$ may be
-interpreted as any set of cardinality at least 2.
-
-\begin{defi}
-The operator $\Wk{}$ translates $\mathrm{\Sigma}$-literals to
-$\mathrm{\Sigma}\Wksup$-literals as follows, where $\cst{p} \in \mathbf{P}$,
-$\cst{q} \notin \mathbf{P}$, and $s, t$ are non-Boolean terms:
-%
-%\begin{linenomath*}
-\begin{align*}
-  \Wk{\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \false
-& \Wk{\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \eq \true
-& \Wk{(s \eq t)} & = s \eq t \\[-\jot]
-  \Wk{\lnot\,\cst{p}(\tuple{t}{})} & = \cst{p}(\tuple{t}{}) \noteq \true
-& \Wk{\lnot\,\cst{q}(\tuple{t}{})} & = \cst{q}(\tuple{t}{}) \noteq \true
-& \Wk{(s \noteq t)} & = s \noteq t
-\end{align*}
-%\end{linenomath*}
-%
-The operator is lifted elementwise to $\mathbf{G}$-clauses and
-$\mathbf{G}$-clause sets.
-The \emph{weak entailment} $\modelsWk$ over $\mathbf{G}$-clause sets
-is defined via an encoding into $\mathrm{\Sigma}\Wksup$-clauses:
-$M \modelsWk N$ if and only if $\Wk{M} \models \Wk{N}$.
-The lifting to $\mathbf{F}$-clauses and $\mathbf{F}$-clause set is achieved
-in the standard way via grounding.
-\end{defi}
-
-The following property of weak entailment will allow us to eliminate
-$\mathbf{P}$-predicates without losing completeness:
-
-\begin{lemma}
-\label{lem:p-Wk-entail}
-Let $C$ be a clause that contains the predicate symbol $\cst{p} \in
-\mathbf{P}$ and $D$ be a clause that does not contain $\cst{p}$. If $N
-\cup \{C\} \modelsWk \{D\}$, then $N \modelsWk \{D\}$.
-\end{lemma}
-
-\begin{proof}
-Suppose $\mathscr{J} \models \Wk{N}$. We will define $\mathscr{J}'$ so that
-$\mathscr{J}' \models \Wk{N} \cup \{\Wk{C}\}$, retrieve $\mathscr{J}'
-\models \Wk{D}$, and then argue that $\mathscr{J} \models \Wk{D}$. We
-take $\mathscr{J}'$ to coincide with $\mathscr{J}$ except that we extend the
-domain for $\prop$ with one fresh value and use this value as the
-interpretation of $\cst{p}(\tuple{t}{})$ for all argument tuples $\tuple{t}{}$.
-This modification makes any $\cst{p}$ literal of $\Wk{C}$ true, and it
-preserves the truth of $\Wk{N}$. By the hypothesis, $\mathscr{J}' \models
-\Wk{D}$. And since $\cst{p}$ does not occur in $D$, we have $\mathscr{J}
-\models \Wk{D}$.\qed
-\end{proof}
-
-Note that the above lemma does not hold for classical entailment $\models$;
-indeed, $\{\cst{p} \cor \cst{q}{,}\; \neglit{\cst{p}} \cor \cst{q}\} \models
-\{\cst{q}\}$. On the other hand, the law of excluded middle does hold for weak
-entailment: $\modelsWk \cst{p} \cor \lnot\,\cst{p}$. In fact, all classical
-clausal tautologies are tautologies for $\modelsWk$.
-
-The standard redundancy criterion $\Red$ is obtained by lifting a criterion on
-$\mathbf{G}$-clauses to $\mathbf{F}$-clauses. The same construction can be
-replicated using $\modelsWk$ instead of $\models$, yielding the \emph{weak
-redundancy criterion} $\RedWk$.
-It is easy to check that the usual simplification techniques implemented in
-superposition provers can be justified using $\RedWk$. Specifically,
-this concerns the following rules described by Schulz \cite[Sects.\
-2.3.1~and~2.3.2]{ss-02-brainiac}:\
-deletion of duplicated literals, 
-deletion of resolved literals, 
-syntactic tautology deletion,
-semantic tautology deletion,
-rewriting of negative literals,
-positive simplify-reflect,
-negative simplify-reflect,
-clause subsumption, and
-equality subsumption.
-Moreover, rewriting of positive literals is possible if the
-rewriting clause is smaller than the rewritten
-clause (a condition that is also needed with $\models$ but omitted by Schulz).
-Finally, destructive equality resolution cannot be justified with $\models$,
-let alone $\modelsWk$.
-
-We instantiate the saturation framework with $(\FInf,\RedWk)$ to obtain a
-given clause prover \textsf{GC}. The prover operates on sets of labeled clauses
-$(C, l)$, where $C$ is a standard clause and $l \in \Lab$ is a label.
-The $\mathsf{active}$ label identifies active clauses; all other clauses are
-passive. The prover takes the form of two rules,
-\textsc{Process} and \textsc{Infer}, restricted to prevent the introduction of
-$\mathbf{P}$-predicates. We extend it with a third rule,
-\textsc{PredElim}, for predicate elimination, and call the extended prover
-\textsf{GCP}. The rules are as follows, using again the framework notations:
-
-\begin{enumerate}[\rm\textsc{PredElim}\,]
-\itemsep1\jot
-\item[\rm\textsc{Process}\,]
-  $\mathcalx{N} \cup \mathcalx{M}
-   \,\Longrightarrow_\mathsf{GCP}\,
-   \mathcalx{N} \cup \mathcalx{M}'$
-\\
-  where $\mathcalx{M} \subseteq \LRedFWksub(\mathcalx{N} \cup \mathcalx{M}')$,
-  $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$,
-  and \\
-  $\mathbf{P}(\lfloor\mathcalx{M}'\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N} \cup \mathcalx{M}\rfloor)$
-
-\item[\rm\textsc{Infer}\,]
-  $\mathcalx{N} \cup \{(C, l)\}
-   \,\Longrightarrow_\mathsf{GCP}\,
-   \mathcalx{N} \cup \{(C, \ensuremath{\mathsf{active}})\} \cup \mathcalx{M}$
-\\
-  where $l \not= \ensuremath{\mathsf{active}}$,
-  $\mathcalx{M}{\downarrow}_\mathsf{active} = \emptyset$, \\
-  $\FInf(\lfloor\mathcalx{N}{\downarrow}_\mathsf{active}\rfloor, \{C\}) \subseteq
-    \RedIWk^{{\cap}\gnd}(\lfloor\mathcalx{N}\rfloor \cup \{C\} \cup \lfloor\mathcalx{M}\rfloor)$,
-  and \\
-  $\mathbf{P}(\lfloor\mathcalx{M}\rfloor) \subseteq \mathbf{P}(\lfloor\mathcalx{N}\rfloor \cup \{C\})$
-
-\item[\rm\textsc{PredElim}\,]
-  $\mathcalx{N} \cup \mathcalx{M}
-   \,\Longrightarrow_\mathsf{GCP}\,
-   \mathcalx{N} \cup \mathcalx{M}'$
-\\
-  where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P} \mathcalx{N} \cup \mathcalx{M}'$
-  and $\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$
-\end{enumerate}
-
-Here is a summary of the main framework notations:
-%
-\begin{itemize}
-\item The letters $\mathcalx{M}, \mathcalx{N}$ range over sets of labeled clauses.
-  $\mathcalx{M}{\downarrow}_l$ denotes the subset of clauses in $\mathcalx{M}$
-  labeled with $l$. The operator $\lfloor\phantom{\cdot}\rfloor$ erases all
-  labels in a labeled clause or clause set.
-
-\smallskip
-
-\item $\FInf(N)$ denotes the set of all base calculus inferences with premises
-  in $N$, and $\FInf(N, M) = \FInf(N \cup M) \setminus \FInf(N \setminus M)$.
-  The same notations are also available for the straightforward extension $\FLInf$
-  of $\FInf$ with labels.
-
-\smallskip
-
-\item $\LRedWksub$ is the extension of the standard redundancy
-  criterion $\RedWk$ defined using $\modelsWk$ to nonground labeled clauses
-  with subsumption ($\sqsubset$).
-
-\smallskip
-
-\item Given a sequence $(\mathcalx{N}{}_i)_i$, its \emph{limit} (\emph{inferior})
-  is $\mathcalx{N}{}_\infty = \bigcup_i \bigcap_{j\ge i} \mathcalx{N}{}_j$.
-\end{itemize}
-
-The completeness proof follows the invariance-based argument found the
-forthcoming journal submission \cite{waldmann-et-al-202x-article}
-by Waldmann et al.
-
-\begin{lemma}
-\label{lem:gcp-derivations-are-red-black-derivations}
-Every $\Longrightarrow_\mathsf{GCP}$-derivation is a mixed derivation.
-\end{lemma}
-
-\begin{proof}
-The cases for \textsc{Process} and \textsc{Infer} are almost as in Waldmann et
-al., with adjustments to show that $\cst{P}$-predicates cannot reappear once
-they have disappeared. The case for \textsc{ElimPred} is trivial.\qed
-\end{proof}
-
-Let $\Inv\Wksup_\mathcalx{N}(k)$ denote the condition
-$\textstyle\FLInf(\mathcalx{N}{}_k{\downarrow}_\mathsf{active}) \subseteq
-\LRedIWk(\mathcalx{N}{}_k)$.
-Notice the difference with the definition of the key invariant
-$\Inv_\mathcalx{N}$ in the saturation framework, whose right-hand side is
-$\bigcup_{i=0}^k \RedI^{\Lab}(\mathcalx{N}{}_i)$. We cannot use the
-big union ${\bigcup}$ starting at $i = 0$ because we will need to truncate a
-sequence prefix of an a priori unknown length. The argument will still work
-thanks to monotonicity properties of redundancy criteria.
-
-\begin{lemma}
-\label{lem:gcp-invar-gcp}
-Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
-If
-$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$,
-then
-$\Inv\Wksup_\mathcalx{N}(k)$ holds for all indices~$k$.
-\end{lemma}
-
-\begin{proof}
-The base case is as in Waldmann et al.
-
-For \textsc{Process} and \textsc{Infer}, the proof is essentially as in Waldmann
-et al., except that we also need to show that
-$\LRedIWk(\mathcalx{N}{}_k) \subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
-This is a consequence of
-$\mathcalx{N}{}_k \rhd_{\LRedWksub} \mathcalx{N}{}_{k+1}$ and of
-properties (R2) and (R3) of redundancy criteria.
-
-A new case to consider is that of a \textsc{PredElim} transition
-$\mathcalx{N}{}_{k} \Longrightarrow_\mathsf{GCP} \mathcalx{N}{}_{k+1}$.
-Let $\mathcalx{N}{}_{k} = \mathcalx{N} \cup \mathcalx{M}
-\Longrightarrow_\mathsf{GCP} \mathcalx{N} \cup \mathcalx{M}' = \mathcalx{N}{}_{k+1}$,
-where $\mathcalx{N} \cup \mathcalx{M} \rhd_\mathbf{P}
-\mathcalx{N} \cap \mathcalx{M}'$ and
-$\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
-We assume without loss of generality that $\mathcalx{M} \cap \mathcalx{M}' =
-\emptyset$.
-Let $\cst{p}$ be the eliminated predicate. Note that $\cst{p}$ occurs in every
-clause in $\mathcalx{M}$ but in none of the clauses in $\mathcalx{N}$ or
-$\mathcalx{M}'$.
-We must show $\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
-\subseteq \LRedIWk(\mathcalx{N}{}_{k+1})$.
-As for \textsc{Process}, we have the inclusion
-$\FLInf(\mathcalx{N}{}_{k+1} {\downarrow}_\mathsf{active})
-\subseteq \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$,
-by the side condition that
-$\mathcalx{M}'{\downarrow}_\mathsf{active} = \emptyset$.
-Moreover, by the induction hypothesis,
-$\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
-\subseteq \LRedIWk(\mathcalx{N}{}_k)$. Thus,
-$\FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})
-\subseteq \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
-
-Let $\iota \in \FLInf(\mathcalx{N} {\downarrow}_\mathsf{active})$.
-By the argument above, we have
-$\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M})$.
-We must
-show $\iota \in \LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.
-%
-By definition of $\LRedIWk$, it suffices to show that for
-every ground instance $(C_n,\ldots,C_1,C_0)$ of $\iota$, there exists a finite
-clause set $\mathcalx{D} \subseteq \gnd(\mathcalx{N}) \cup \gnd(\mathcalx{M}')$
-that is $\prec$-smaller than $C_1$ and such that
-$\{C_n,\ldots,C_2\} \cup \mathcalx{D} \modelsWk \{C_0\}$.
-Without loss of generality, we assume that $\mathcalx{D}$ is the smallest such
-set.
-
-By definition of $\mathcalx{N}$, $\cst{p}$ cannot occur in $C_0$.
-By Lemma~\ref{lem:p-Wk-entail}, if $\cst{p}$ occurred in $D \in \mathcalx{D}$,
-we could remove $D$, but this would mean $\mathcalx{D}$ is not minimal.
-As a result, $\mathcalx{D}$ cannot contain clauses from $\gnd(\mathcalx{M})$ and
-hence $\mathcalx{D} \subseteq \gnd(\mathcalx{N})$. Thus,
-$\iota \in \LRedIWk(\mathcalx{N})$. By property (R2) of
-redundancy criteria, we have the desired result: $\iota \in
-\LRedIWk(\mathcalx{N} \cup \mathcalx{M}')$.\qed
-\end{proof}
-
-\begin{lemma}
-\label{lem:gcp-invar-infinity}
-Let $(\mathcalx{N}{}_i)_i$ be a
-$\rhd_{\LRedWksub}$-derivation.
-If $\Inv\Wksup_\mathcalx{N}(i)$ holds for all indices~$i$, then
-$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
-\bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
-holds.
-\end{lemma}
-
-\begin{proof}
-We assume $\iota \in \FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$
-and show $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$
-for some arbitrary $\iota$.
-%
-For $\iota$ to belong to $\FLInf(\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active})$, all of
-its finitely many premises must be in $\mathcalx{N}{}_\infty {\downarrow}_\mathsf{active}$.
-Therefore, there must exist an index $k$ such that
-$\mathcalx{N}{}_k {\downarrow}_\mathsf{active}$ contains all of them, and therefore
-$\iota \in \FLInf(\mathcalx{N}{}_k {\downarrow}_\mathsf{active})$. Since $\Inv_\mathcalx{N}(k)$
-holds, $\iota \in \LRedIWk(\mathcalx{N}{}_k)$.
-Hence, $\iota \in \bigcup_i \LRedIWk(\mathcalx{N}{}_i)$.\qed
-\end{proof}
-
-\begin{lemma}
-\label{lem:fair-gcp-derivations}
-Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation.
-If
-$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
-and
-$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
-$l \not= \ensuremath{\mathsf{active}}$,
-then there exists an index $k$ such that
-$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
-\end{lemma}
-
-\begin{sloppypar}
-\begin{proof}
-By Lemma~\ref{lem:gcp-derivations-are-red-black-derivations}, there must exist
-an index $k$ such that the sequence $(\mathcalx{N}{}_{k+i})_i$ is a pure
-$\rhd_{\LRedWksub}$-derivation.
-By Lemma~\ref{lem:gcp-invar-gcp},
-$\Inv\Wksup_\mathcalx{N}(k+i)$ holds for all indices~$i$.
-Hence, by Lemma~\ref{lem:gcp-invar-infinity},
-$\FLInf(\mathcalx{N}{}_\infty{\downarrow}_\mathsf{active}) \subseteq
-\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.
-By the second hypothesis, this inclusion simplifies to
-$\FLInf(\mathcalx{N}{}_\infty) \subseteq
-\bigcup_{i} \LRedIWk(\mathcalx{N}{}_{k+i})$.\qed
-\end{proof}
-\end{sloppypar}
-
-\begin{theorem}
-\label{thm:gcp-complete}
-Let $(\mathcalx{N}{}_i)_i$ be a $\Longrightarrow_\mathsf{GCP}$-derivation
-with
-$\mathcalx{N}{}_0{\downarrow}_\mathsf{active} = \emptyset$
-and
-$\mathcalx{N}{}_\infty{\downarrow}_l = \emptyset$ for every label
-$l \not= \ensuremath{\mathsf{active}}$.
-If $\lfloor \mathcalx{N}{}_0\rfloor$ is unsatisfiable,
-then
-some $\mathcalx{N}{}_i$
-contains the empty clause with some arbitrary label.
-\end{theorem}
-
-\begin{proof}
-By Lemma~\ref{lem:fair-gcp-derivations},
-we know that there exists an index $k$ such that
-$(\mathcalx{N}{}_{k+i})_i$ is a fair $\rhd_{\LRedWksub}$-derivation.
-Moreover, since
-$\rhd_{\LRedWksub}$ and
-$\rhd_\mathbf{P}$ preserve unsatisfiability
-(by (R1) of redundancy criteria, Khasidashvili and Korovin's Theorem~1, and our
-Theorem~\ref{thm:pes-sat-equiv}), we have that
-$\lfloor \mathcalx{N}{}_k\rfloor$ is unsatisfiable.
-Since the base calculus $(\FLInf,\allowbreak\LRed)$ is assumed to be statically
-refutationally complete with respect to $\models$, the calculus
-$(\FLInf,\allowbreak\LRedWk)$ with a weaker redundancy criterion is also
-statically complete with respect to $\models$, and by the saturation framework,
-$(\FLInf,\allowbreak\LRedWksub)$ preserves this. Exploiting the equivalence of
-static and dynamic completeness, we conclude that some $\mathcalx{N}{}_{k+i}$ must
-contain a labeled empty clause.\qed
-\end{proof}
-\end{rep}
-
-\begin{qle}
-\section{Quasipure Literal Elimination}
-\label{sec:satfol:quasipure-literal-elimination}
-
-\emph{Pure literal elimination} (PLE) is one of the simplest optimizations
-implemented in SAT solvers. It is a special case of variable elimination: If a
-given variable always occurs with the same polarity in a problem, the solver can
-assign it that polarity without loss of generality, making all the clauses that
-contain it tautologies. PLE consists of recursively deleting all such clauses.
-%
-PLE's generalization to first-order logic considers predicate literals
-$\arbpredlit{\cst{p}(\tuple{s}{})}$, where the arguments $\tuple{s}{}$ are
-ignored by the analysis; only the polarity matters.
-
-\begin{exa}
-   \label{example:ple}
-   Consider the clause set $N = \{ \cst{p}(x) \cor \cst{q}(\cst{a}, x){,}\allowbreak\;
-   \cst{p}(\cst{f}(x)){,}\allowbreak\; \neglit{\cst{q}(\cst{a}, \cst{a})} \}$.
-   Since $\cst{p}$ occurs only positively in $N$, PLE deletes the two first
-   clauses. At that point, $\cst{q}$ occurs only negatively in the remaining
-   singleton clause set and can be deleted as well. The result is the empty set,
-   which is obviously satisfiable, indicating that $N$ is satisfiable.
-   As model of $N$, we can take $\ourmodel$ such that $\ourmodel$ makes all
-   atoms $\cst{p}(\ldots)$ true and all atoms $\cst{q}(\ldots)$ false.
-\end{exa}
-
-\begin{exa}
-   \label{example:qle-i}
-   Consider the clause set $N' = \{
-   \cst{p}(\cst{a}){,}\allowbreak\;
-   \neglit{\cst{p}(x)} \lor \cst{p}(\cst{f}(x))
-    \}$. This time, PLE does not apply because $\cst{p}$ occurs with both
-    polarities. Yet we notice that $\cst{p}$ occurs positively in
-    both clauses, and hence that the same reasoning as in the previous
-    example applies. In particular, $\ourmodel \models N'$.
-\end{exa}
-
-\begin{exa}
-   \label{example:qle-ii}
-   Consider the clause set $N'' = \{
-   \cst{p}(\cst{a}){,}\allowbreak\;
-   \cst{q}(x) \lor \cst{p}(\cst{f}(x)){,}\allowbreak\;
-   \neglit{\cst{q}(\cst{f}(\cst{a}))}{,}\allowbreak\;
-   \neglit{\cst{p}(x)} \lor \neglit{\cst{q}(\cst{f}(x))}
-   \}$. This clause set is clearly beyond PLE. Yet each clause contains either
-   a positive $\cst{p}$ literal or a negative $\cst{q}$ literal---any
-   extraneous, wrong-polarity literals are harmless. Thus $\ourmodel \models
-   N''$.
-\end{exa}
-
-Examples \ref{example:qle-i} and \ref{example:qle-ii} suggest that PLE is
-needlessly restrained, presumably due to its origins in propositional logic. We
-propose a generalization to ``quasipure'' predicates.
-
-\begin{defi}
-\label{def:quasipure-set}
-A set $P$ of predicate symbols is \emph{quasipure} with a polarity map $s$ in a
-clause set $N$ if for every symbol $\cst{p} \in P$, every clause in $N$ that
-contains a member of $P$ contains at least one occurrence of some $\cst{q} \in
-P$ with polarity $s_\cst{q} \in \{+, -\}$.
-%
-The set $P$ is \emph{quasipure} in $N$ if there exists a polarity map $s$
-such that $P$ is quasipure with $s$ in $N$.
-\end{defi}
-
-In Example~\ref{example:qle-i}, $\{\cst{p}\}$ is quasipure with $s_\cst{p} =
-{+}$. In Example~\ref{example:qle-ii}, $\{\cst{p}, \cst{q}\}$ is quasipure with
-$s_\cst{p} = {+}$ and $s_\cst{q} = {-}$. For this example, it is crucial to
-consider $\cst{p}$ and $\cst{q}$ together; neither of the singletons
-$\{\cst{p}\}$ and $\{\cst{q}\}$ is quasipure.
-
-\begin{defi}
-\label{def:quasipure-sym}
-A predicate symbol $\cst{p}$ is \emph{quasipure} with polarity $s_\cst{p} \in
-\{+, -\}$ in a clause set $N$ if there exists a set $P \ni \cst{p}$ of predicate
-symbols and a polarity map $s$ such that $P$ is quasipure with $s$ in $N$.
-%
-The symbol $\cst{p}$ is \emph{quasipure} in $N$ if there exists a polarity
-$s_\cst{p} \in \{+, -\}$ such that $\cst{p}$ is quasipure with $s_\cst{p}$ in
-$N$.
-%
-A literal $L = \arbpredlit{\cst{p}(\ldots)}$ is \emph{quasipure} in $N$ if
-$\cst{p}$ is quasipure with $L$'s polarity in $N$.
-\end{defi}
-
-Deleting a clause containing a quasipure literal is monotonic in the sense that
-it might create new opportunities for quasipure literal elimination but never
-ruin existing ones. Therefore, the following nondeterministic procedure yields a
-well-defined result:
-
-\begin{defi}
-\label{def:qle}
-Given a clause set, \emph{quasipure literal elimination} (QLE) iteratively
-deletes clauses containing quasipure predicate symbols until no such clauses
-remain.
-\end{defi}
-
-The key property of QLE is that it preserves unsatisfiability:
-
-\begin{lemma}
-Let $C \in N$ be a clause containing a quasipure literal. If $N \setminus \{C\}$
-is satisfiable, then $N$ is satisfiable.
-\end{lemma}
-
-\begin{proof}
-Let $\ourmodel$ be a model of $N \setminus \{C\}$.
-Let $P$ be the set of predicates and $s$ the polarity map whose existence is
-guaranteed by definition of quasipurity. Let $N_0 \subseteq N \setminus \{C\}$
-be the result of applying QLE on $N$. It is easy to see that $N_0$ contains no
-occurrences of the symbols in $P$. Extend $\ourmodel$ to $\ourmodel'$ by
-making each symbol $\cst{p}$ such that $s_\cst{p} = {+}$ true and
-each symbol $\cst{p}$ such that $s_\cst{p} = {-}$ false. Clearly $\ourmodel'$ is
-a model not only of $N_0$, which contains no $P$ symbols, but also of $N$,
-because each clause in $N \setminus N_0$ contains a quasipure literal, which is
-satisfied by $\ourmodel'$.
-\qed
-\end{proof}
-
-Definition~\ref{def:qle} suggests a naive, nondeterminisitc procedure for
-discovering and eliminating quasipure predicate symbols: Choose a predicate
-symbol $\cst{p}$ and a polarity $s_\cst{p}$, and take $P = \{\cst{p}\}$. If the
-predicate occurs with the wrong polarity in a clause
-$\arbpredlit{\cst{p}(\ldots)} \lor \arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
-\arbpredlit{\cst{q}_n(\ldots)} \lor C$, try to extend the polarity map $s$ and
-the predicate set $P$ with one of the $\arbpredlit{\cst{q}_i}$'s and continue
-recursively with $\cst{q}_i$. In Sect.~\ref{sec:satfol:implementation}, we will see a
-more efficient approach based on a SAT encoding.
-\end{qle}
-
-\section{Satisfiability by Clause Elimination}
-\label{sec:satfol:satisfiability-by-clause-elimination}
-
-\looseness=-1
-The main approaches to show satisfiability of a first-order problem are to
-produce either a finite Herbrand model or a saturated clause set. Saturations
-rarely occur except for very small problems or within decidable fragments. In
-this section, we explore an alternative approach that establishes satisfiability
-by iteratively removing clauses while preserving
-%satisfiability and
-unsatisfiability,
-until the clause set has been transformed into the empty set. So
-far, this technique has been studied only for QBF~\cite{hsb-14-QRAT}. We show that
-\emph{blocked clause elimination} (BCE)
-% for first-order logic
-can be used for this purpose. It can efficiently solve some problems for which
-the saturated set would be infinite. However, it can break the refutational
-completeness of a saturation prover. We conclude with a procedure that
-transforms a finite Herbrand model into a sequence of clause elimination steps
-ending in the empty clause set, thereby demonstrating the theoretical power of
-clause elimination.
-
-Kiesl et al.\ \cite{ksstb-2017-blockedfol} generalized blocked clause
-elimination to first-order logic. Their generalization uses flat
-$L$-resolvents, an extension of flat resolvents that resolves a single
-literal~$L$ against $m$~literals of the other clause.
-
-\begin{defi}
-   Let $C = L \cor C'$ and $D = L_1 \cor \cdots \cor L_m \cor D'$, where
-   (1)~$m \ge\nobreak 1$,
-   (2) the literals $L_i$ are of opposite polarity to $L$,
-   (3) $L$'s atom is $\cst{p}(\tuplen{s})$,
-   (4) $L_i$'s atom is $\cst{p}(\tuple{t_i}{})$ for each~$i$, and
-   (5) $C$ and $D$ have no variables in common.
-   The clause
-   $\bigl(\bigvee_{i=1}^m \bigvee_{j=1}^n
-     \neqlit{s_j}{t_{ij}}\bigr) \cor C' \cor D'$
-   is a \emph{flat $L$-resolvent} of $C$~and~$D$.
-\end{defi}
-
-\begin{defi}
-\label{def:equality-blocked}
-\begin{conf}%
-   A clause $C = L \cor C'$ is (\emph{equality-})\emph{blocked} by $L$ in a clause
-   set $N$ if all flat $L$-resolvents between $C$ and clauses in $N \setminus \{C\}$ are
-   tautologies.
-\end{conf}%
-\begin{rep}%
-   Let $C = L \cor C'$ be a clause and $N$ be a clause set.
-   Let $N'$ consist of all clauses from $N \setminus \{C\}$ with their variables
-   renamed so that they share no variables with $C$.
-   The clause $C$ is (\emph{equality-})\emph{blocked} by $L$ in the set $N$ if
-   all flat $L$-resolvents between $C$ and clauses in $N'$ are tautologies.
-\end{rep}
-\end{defi}
-
-Removing a blocked clause from a set preserves unsatisfiability
-\cite{ksstb-2017-blockedfol}. Kiesl et al.\ evaluated the effect of removing all
-blocked clauses as a preprocessing step and found that it increases
-\confrep{}{the }prover's success rate.
-
-In fact, there exist satisfiable problems that cannot be saturated in finitely
-many steps regardless of the calculus's parameters but that can be reduced to
-an empty, vacuously satisfiable problem through blocked clause elimination.
-
-\begin{exa}
-   \label{example:infinite-blocked-set}
-   \looseness=-1
-   Consider the clause set $N$ consisting of $C = \pospredlit{\cst{p}(x,x)}$ and
-   $D = \negpredlit{\cst{p}(y_1,y_3)} \cor \pospredlit{\cst{p}(y_1,y_2)} \cor
-   \pospredlit{\cst{p}(y_2,y_3)}$. Note that if no literal is
-   selected, all literals are eligible for superposition. In
-   particular, the superposition of
-   $\pospredlit{\cst{p}(x,x)}$ into $D$'s negative literal
-   eventually needs to be performed regardless of the chosen
-   selection function or term order, with the conclusion $E_1 =
-   \pospredlit{\cst{p}(z_1,z_2)} \cor \pospredlit{\cst{p}(z_2,z_1)}$. Then,
-   superposition of $E_1$ into $D$ yields $E_2 = \pospredlit{\cst{p}(z_1,z_2)}
-   \cor \pospredlit{\cst{p}(z_2,z_3)} \cor \pospredlit{\cst{p}(z_3,z_1)}$.
-   Repeating this process yields infinitely many clauses $E_i =
-   \pospredlit{\cst{p}(z_1,z_2)} \cor \cdots \cor
-   \pospredlit{\cst{p}(z_i,z_{i+1})} \cor \pospredlit{\cst{p}(z_{i+1},z_1)}$
-   that cannot be eliminated using standard redu\-ndancy-based techniques.
-\end{exa}
-
-In the example above, the clause $D$ is blocked by its second or third
-literal. If we delete $D$, $C$ becomes blocked in turn. Deleting
-$C$ leaves us with the empty set, which is vacuously satisfiable. The example
-suggests that using BCE during saturation might help focus the proof search.
-Indeed, Kiesl et al.\ ended their investigations by asking whether BCE can be
-used as an inprocessing technique in a saturation prover. Unfortunately,
-in general the answer is no\confrep{.}{:}
-
-\begin{exa}
-   \label{example:bce-completeness}
-   Consider the unsatisfiable set $N=\{ C_1, \ldots,\allowbreak C_6 \}$, where
-%   \begin{linenomath*}
-   \begin{align*}
-      C_1 &= \negpredlit{\cst{c}} \cor \pospredlit{\cst{e}} \cor \MAX{\negpredlit{\cst{a}}} 
-      & C_2 &= \negpredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{e}}}
-      & C_3 &= \pospredlit{\cst{b}} \cor \MAX{\pospredlit{\cst{c}}}
-      \\
-      C_4 &= \negpredlit{\cst{b}} \cor \MAX{\negpredlit{\cst{c}}}
-      & C_5 &= \pospredlit{\cst{a}} \cor \MAX{\pospredlit{\cst{b}}} 
-      & C_6 &= \pospredlit{\cst{c}} \cor \MAX{\negpredlit{\cst{b}}} 
-         &
-   \end{align*}
-%   \end{linenomath*}
-   Assume the simplification ordering $\cst{a} \prec \cst{b} \prec \cst{c} \prec
-   \cst{d} \prec \cst{e}$ and the selection function that chooses the last negative literal of a clause as
-   presented. Gray boxes indicate literals that can take part in superposition
-   inferences. Only two superposition inferences are
-   possible:\ from $C_3$ into $C_4$, yielding the tautology $C_{7} = \pospredlit{\cst{b}} \cor
-   \MAX{\negpredlit{\cst{b}}}$, and from $C_5$ into $C_6$, yielding $C_{8} = \pospredlit{\cst{a}} \cor
-   \MAX{\pospredlit{\cst{c}}}$. Clause $C_{7}$ is
-   clearly redundant, whereas $C_{8}$ is blocked by its first literal. If we
-   allow removing blocked clauses, the prover enters a loop: $C_{8}$ is repeatedly generated and 
-   deleted. Thus, the prover will never generate the
-   empty clause for this unsatisfiable set.
-\end{exa}
-
-   \newcommand{\aset}{\mathcalx{A}}
-   \newcommand{\pset}{\mathcalx{P}}
-   As with hidden tautologies, removing blocked clauses breaks the invariant of
-   the given clause procedure that all inferences between clauses in $\aset$ are
-   redundant. To see this, assume the setting of Example
-   \ref{example:bce-completeness}, and let $\pset = N$ and $\aset = \emptyset$.
-   Assume $C_1, C_2, C_3$ are moved to the active set. As there are
-   no possible inferences between them, the proof state becomes $\aset = \{C_1,
-   C_2, C_3\}$ and $\pset = \{C_4,C_5,C_6\}$. After $C_4$ is moved to
-   $\aset$, the conclusion $C_7$ is computed, but it is not added to $\pset$ as it
-   is redundant. Moving $C_5$ to $\aset$ produces no new conclusions, but after
-   $C_6$ is moved, $C_{8}$ is produced. However, if we allow eliminating blocked
-   clauses, it will not be added to $\pset$ as it is blocked. The prover then
-   terminates with $\aset=N$ and $\pset=\emptyset$, even though the original set
-   $N$ is unsatisfiable.
-
-Although using BCE as inprocessing breaks the completeness of superposition in
-general, it is conceivable that a well-behaved fragment of BCE might exist.
-This could be investigated further.
-
-% @MARIJN : Can you please motivate this part a bit better and
-%           add the necessary references? --PV
-% In particular, why do we care about short proofs (in FOL)? --JB
-% Short proofs are often easier to generate, process, and understand than long
-% proofs. SAT solvers can often generate short refutational proofs by adding and
-% removing well-chosen clauses using satisfiability-preserving techniques.
-% Similarly, by adding some clauses, we can transform a first-order problem so
-% that all clauses can be removed by BCE. Since the empty clause set is trivially
-% satisfiable, this approach produces easily checkable proofs of satisfiability.
-
-Not only can BCE prevent infinite saturation (Example \ref{example:infinite-blocked-set}),
-but it can also be used to convert a finite Herbrand model
-into a certificate of clause set satisfiability. The certificate uses only
-blocked clause elimination and addition, in conjunction with a transformation
-to reduce the clause set to an empty set. This theoretical result
-explores the relationship between Herbrand models and satisfiability certificates
-based on clause elimination and addition. It is conceivable that it can form
-the basis of an efficient way to certify Herbrand models.
-
-\looseness=-1 
-In propositional logic, \emph{asymmetric literals} can be added to
-or removed from clauses, retaining the equivalence of the resulting clause set with
-the original one. Kiesl and Suda \cite{ks-2017-unif-principle} described an extension of
-this technique to first-order logic.
-Their definition of asymmetric literals can be relaxed to
-allow the addition of more literals, but the resulting set is then only
-equisatisfiable to the original one, not equivalent.
-This in turn allows us to show that a problem is satisfiable by
-reducing it to an empty problem, as is done in some SAT solvers.
-
-For the rest of this section, we work with clausal first-order logic
-\relax{without equality}. We use \relax{Herbrand models} as canonical
-representatives of first-order models, recalling that every
-satisfiable set has a Herbrand model \cite[Sect.~5.4]{mf-1996-fol}.
-
-\begin{defi}
-   \looseness=-1
-   A literal $L$ is a \emph{global asymmetric literal} (GAL) for a clause $C$
-   and a clause set $N$ if for every ground instance $C\sigma$ of $C$, there
-   exists a ground instance $D\varrho \cor L'\varrho$ of $D \cor L' \in N \setminus \{C\}$ such that $D\varrho
-   \subseteq C\sigma$ and $\neglit L'\varrho = L\sigma$.
- \end{defi}
-\confrep{}{}
-
-Every asymmetric literal is GAL, but the converse does not hold:
-
-\begin{exa}
-   \looseness=-1
-   Consider a clause $C = \pospredlit{\cst{p}(x,y)}$ and a clause set $N = \{\pospredlit{\cst{q}} \cor
-  \pospredlit{\cst{p}(\cst{a},\cst{a})}\}$. Then, $\negpredlit{\cst{q}}$ is not an asymmetric literal
-  for $C$ and $N$, but it is a GAL for $C$ and $N$.
-\end{exa}
-
-Adding and removing GALs preserves and reflects satisfiability:
-
-\begin{theorem}
-   \label{thm:gal}
-   If $L$ is a GAL for the clause $C$ and the clause set $N$, then the set $(N \setminus
-   \{C\}) \cup\allowbreak \{ C \cor\nobreak L \}$ is satisfiable if and only if $N$ is satisfiable.
-\end{theorem}
-\begin{rep}%
-\begin{proof}
-   \looseness=-1
-   Let $N' = N \setminus \{C\} \cup \{ C \cor L \}$.
-   Non-trivial direction is proving that if $N'$ has a model, so does
-   $N.$ If $N'$ has a model, it has an Herbrand model $\ourmodel$. Clearly,
-   $\ourmodel$ satisfies every clause in $N$, with the possible exception of
-   $C$. Assume there exists a ground instance $C\sigma$ falsified by
-   $\ourmodel$. Since $L$ is a GAL for $C$ and $N$, then there exists a ground
-   instance $D\tau \cor L'\tau$ of $D \cor L' \in N'$ and $\neg L'\tau =
-   L\sigma$. If $\ourmodel \models
-   L'\tau$, for $C\sigma \cor L\sigma$ to be satisfied by $\ourmodel$, some
-   literal in $C\sigma$ must be satisfied by $\ourmodel$, contradicting that
-   $C\sigma$ is falsified by $\ourmodel$. If $\ourmodel \notmodels L'\tau$
-   then some literal in $D\tau$ must be satisfied. As $D\tau \subseteq C\sigma$,
-   we again get the same contradiction. Therefore, $\ourmodel$ satisfies $N$,
-   as needed.\qed
- \end{proof}
-
-\end{rep}
-\looseness=-1
-For first-order logic without equality, a clause $L \cor C$ is blocked if all its
-$L$-resolvents are tautologies \cite{ksstb-2017-blockedfol}. The $L$-resolvent
-between $L \cor C$ and $\neglit{L_1} \cor \cdots \cor \neglit{L_n} \cor
-D$ is $(C \cor D)\sigma$, where $\sigma$ is the most general unifier
-of the literals $L, L_1, \ldots, L_n$ \cite{bg-01-resolution}.
-Given a Herbrand model $\ourmodel$ of a problem, the following procedure
-removes all clauses while preserving satisfiability:
-
-\begin{enumerate}
-   \item \label{item:step1}
-   \looseness=-1
-   Let $\cst{q}$ be a fresh predicate symbol. For each atom
-   $\cst{p}(\tupleempty{s})$ in the Herbrand universe: If
-   $\ourmodel \models \cst{p}(\tupleempty{s})$, add the clause $\cst{q} \cor
-   \cst{p}(\tupleempty{s})$; otherwise, add %the clause
-   $\cst{q} \cor \neglit{\cst{p}(\tupleempty{s})}$. Adding either clause
-   preserves satisfiability as both are blocked by~$\cst{q}$.
-% \pagebreak[2]
-   \smallskip
-   \item \label{item:step2}
-   Since $\ourmodel$ is a model, for each ground instance $C\sigma$, there exists a clause
-   $\cst{q} \cor L$ with $L \in C\sigma$. We can transform $C \in N$ into
-   $C \cor \neglit{\cst{q}}$, since $\neglit{\cst{q}}$ is a GAL for $C$ and $N$.
-
-   \smallskip
-   \item \label{item:step3}
-   \looseness=-1
-   Consider the clause $\cst{q} \cor L$ added by step \ref{item:step1}. Since
-   $L$ is ground and no clause $\cst{q} \cor \neglit{L}$ was added (since $\ourmodel$ is a
-   model), the only $L$-resolvents are against clauses added by step~\ref{item:step2}.
-   Since all of those clauses contain $\neglit{\cst{q}}$, the resolvents are tautologies. 
-   Thus, each $\cst{q} \cor L$ is blocked and can be removed in turn.
-
-   \smallskip
-   \item The remaining clauses all contain the literal $\neglit{\cst{q}}$. They can be removed
-   by BCE as well.
-\end{enumerate}
-
-The procedure is limited to the first-order logic without equality, since step
-\ref{item:step3} is justified only if $L$ is a predicate literal. (Otherwise,
-$L$ cannot block clause $\cst{q} \cor L$ \cite{ksstb-2017-blockedfol}.) The
-procedure also terminates only for finite Herbrand models.
-
-% \begin{rep}
-
-\begin{exa}
-   Consider the satisfiable clause set $N = \{\pospredlit{\cst{r}(x)} \cor
-   \pospredlit{\cst{s}(x)}{,}\; \negpredlit{\cst{r}(\cst{a})}{,}\allowbreak\;
-   \negpredlit{\cst{s}(\cst{b})}\}$ and a Herbrand model $\ourmodel$ over
-   $\{\cst{a}, \cst{b}, \cst{r}, \cst{s}\}$ such that
-   $\cst{r}(\cst{b})$ and $\cst{s}(\cst{a})$ are the only true atoms in
-   $\ourmodel$. We show how to remove all clauses in $N$ using $\ourmodel$ 
-   by following the procedure above.
-
-   Let $N_{\ourmodel} = \{ \cst{q} \cor \negpredlit{\cst{r}(\cst{a})}{,}\; \cst{q}
-   \cor \pospredlit{\cst{r}(\cst{b})}{,}\; \cst{q} \cor
-   \pospredlit{\cst{s}(\cst{a})}{,}\; \cst{q} \cor \negpredlit{\cst{s}(\cst{b})}
-   \}$. We set $N \gets N \cup N_{\ourmodel}$. This preserves satisfiability
-   since all clauses in $N_{\ourmodel}$ are blocked. It is easy to check that
-   $\negpredlit{\cst{q}}$ is GAL for every clause in $N \setminus N_{\ourmodel}$. The
-   only substitutions that need to be considered are $\{x \mapsto \cst{a} \}$ and $\{x \mapsto
-   \cst{b} \}$ for $\pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}$.
-   So we set $N \gets \{\negpredlit{\cst{q}} \cor \pospredlit{\cst{r}(x)} \cor \pospredlit{\cst{s}(x)}{,}\;
-   \negpredlit{\cst{q}} \cor \negpredlit{\cst{r}(\cst{a})}{,}\;
-   \negpredlit{\cst{q}} \cor \negpredlit{\cst{s}(\cst{b})}\} \cup N_{\ourmodel}$.
-   Clearly, all clauses in $N_{\ourmodel}$ are blocked, so we set $N \gets N \setminus
-   N_{\ourmodel}$. All clauses remaining in $N$ have a literal $\negpredlit{\cst{q}}$
-   and can be removed, leaving $N$ empty as desired.
-\end{exa}
-
-% The procedure works only for first-order logic without equality, because it
-% relies on adding and removing blocked clauses. In step~\ref{item:step3}, binary
-% clauses $\cst{q} \cor L$ are blocked by $L$. If $L$ were a
-% functional literal, by definition $\cst{q} \cor L$ would not be blocked.
-% \end{rep}
-
-% \looseness=-1
-% For many problems model can be described only using infinite Herbrand universe and the described technique for
-% certifying satisfiability is not applicable. However, in cases of finite
-% Herbrand universes, provers that support GAL addition and blocked clause
-% elimination and addition can certify that a given interpretation is
-% indeed a Herbrand model.
-
-\section{Implementation}
-\label{sec:satfol:implementation}
-% {\large\bf {TODO: Shorten and introduce SPPE here in a sentence}}
-
-\newcommand{\impmap}{\textit{Imp}}
-
-Hidden-literal-based, predicate, \begin{qle}quasipure, \end{qle}and blocked clause
-elimination all admit efficient implementations in a superposition
-prover. In this section, we describe how to implement the first
-\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
-techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
-All techniques are implemented in the Zipperposition prover
-\cite{sc-supind-17}. Zipperposition is designed for fast prototyping
-of improvements to superposition, but it implements many of the most successful
-heuristics from the E prover \cite{scv-19-e23} and has recently become
-quite competitive \cite{gs-19-casc27}.
-
-\ourpara{Hidden-Literal-Based Elimination}
-% 
-% Heule et al.\ \cite{hjb-2010-cl-elim} show how hidden tautologies can be removed
-% without creating the binary implication graph. For each literal~$L$ occurring in
-% the clause set $N$, they compute the set $\PHL(L,N)$. Then for each $\neglit{L}'
-% \in \PHL(L,N)$, they label $L'$ with $L$. A clause $C$ is a hidden tautology if
-% there are two literals $L,L' \in C$ such that $L'$ is labeled with $L$. The
-% described approach can be generalized to first-order logic and used to implement
-% all rules from Sect.~\ref{sec:satfol:hidden-literal-based-elimination}.
-% 
-For HLBE, an efficient representation of $\HL(L,N)$ is crucial. Because
-this set may be infinite, we underapproximate it by
-restricting the length of the
-transitive chains via a parameter $K_\mathrm{len}$. Given the current
-clause set $N$, the finite map $\impmap[L']$ associates with each literal $L'$
-a set of pairs $(L,M)$ such that $L' \bigfimpl{N}^k L$, where $k \leq
-K_\mathrm{len}$ and $M$ is the multiset of clauses used to derive $L' \bigfimpl{N}^k L$.
-Moreover, we consider only transitions of type (1) (as per
-Definition~\ref{def:hl-fo}).
-%Intuitively,
-%$\impmap[L']$ contains all literals which can be reached from $L'$ following $k$
-%edges of the generalized binary implication graph.
-The following algorithm
-maintains $\impmap$ dynamically, updating it as the prover derives and deletes
-clauses.
-It depends on the global variable $\impmap$ and the parameters
-$K_\mathrm{len}$ and $K_{\mathrm{imp}}$.
-
-% By choosing a limit $K_\mathrm{len}$ we can create a map $\impmap[L]$, which maps a literal $L$
-% to a finite a set of pairs $(L', M)$ such that
-% $\impk{L}{L'}{i}{M}$ where $i \leq K_\mathrm{len}$. It is important to record clauses used
-% in deriving the implication, as this information is necessary to construct the
-% proof. Mapping $\impmap$ is constructed so that whenever $(L', M) \in
-% \impmap[L]$, then $L \in \HL(L',M)$.
-% Proof by very simple induction
-
-\looseness=-1
-
-\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
-\algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}
-\algrenewcommand\algorithmicindent{0.75em}
-
-%%% TYPESETTING: hack
-\vskip\abovedisplayskip
-\kern-.5\jot
-
-%% \begin{nolinenumbers}%\begin{figure}[h!]
-\begin{algorithmic}[5]
-
-%   \Procedure{DecomposeNegLit}{$c$, $C$}
-%   \State $\vn{res} \gets \{(c, C)\}$
-%   \ForAll{non-empty contexts $u$ such that $c = \neqlit{u[s]}{u[t]}$}
-%       \State $\vn{res} \gets \{ (\neqlit{s}{t}, C) \} \cup \vn{res}$
-%   \EndFor
-%   \State \Return $\vn{res}$
-%   \EndProcedure
-
-%  \kern1\jot
-
-  \Procedure{AddImplication}{$L_\mathrm{a}{,}\> L_\mathrm{c}{,}\> C$}
-    \If{$\impmap[L_\mathrm{a}\sigma] \not = \emptyset$ for some renaming $\sigma$} \label{statement:alpha}
-      \State $(L_\mathrm{a}, L_\mathrm{c}) \gets (L_\mathrm{a}\sigma, L_\mathrm{c}\sigma)$
-    \EndIf
-
-   %  \vspace{0.1em}
-    \If{there are no $L,L',M,\sigma$ such that $(L',M) \in \impmap[L]$,\label{statement:more-general}%
-        \\\noindent\kern3.5ex $L\sigma = L_\mathrm{a}$, and $L'\sigma=L_\mathrm{c}$} 
-      % \State $\vn{Inst} \gets \{(\sigma, M) \mid (L_\mathrm{c}\sigma, M) \in \impmap[L_\mathrm{a}\sigma]\}$
-      \ForAll{$(\sigma, M)$ such that $(L_\mathrm{c}\sigma, M) \in \impmap[L_\mathrm{a}\sigma]$} \label{statement:instances}
-        \State {erase all $(L',M')$ such that $M \subseteq M'$ from $\impmap[L_{\mathrm{a}}\sigma]$}
-      \EndFor
-
-      \ForAll{$L$ such that $(L', M) \in \impmap[L]$\\ \noindent\kern10ex and $L_\mathrm{a}\sigma = L'$ for some $\sigma$} \label{statement:transitivity}
-        \If{$|M| < K_\mathrm{len}$}
-          \State $\impmap[L] \gets \impmap[L] \cup \{(L_\mathrm{c}\sigma, M \uplus \{C\})\} $
-        \EndIf
-      \EndFor
-
-      \ForAll{$L$ such that $\impmap[L] \not= \emptyset$\\\noindent\kern10ex and $L\sigma = L_\mathrm{c}$ for some $\sigma$} \label{statement:triggered}
-         % \If{$|M| < K_\mathrm{len}-1$}
-         \State $\vn{Concl} \gets \{(L'\sigma, M \uplus \{\mathit{C}\}) \mid$%
-         \\\noindent\kern14.7ex $(L',M) \in \impmap[L], |M| < K_\mathrm{len}  \}$
-         \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup  \vn{Concl}$
-         % (L'\sigma, M \cup \{\mathit{C}\}) $
-         % \EndIf
-      \EndFor
-
-      \State $\vn{Congr} \gets \{
-         (\neqlit{s}{t}, \{C\}) \mid \exists u.\, L_\mathrm{c} = \neqlit{u[s]}{u[t]}
-         \}$
-      \State $\impmap[L_\mathrm{a}] \gets \impmap[L_\mathrm{a}] \cup \{ (L_\mathrm{c},\{C\}) \} \cup \vn{Congr}$ \label{statement:decomposition}
-
-    \EndIf
-
-  \EndProcedure
-
-  \kern1\jot
-
-  \Procedure{TrackClause}{$C$}
-    \If{$C$ = $L_1 \cor L_2$}
-      \State \textsc{AddImplication}($\neglit{L_1}$, $L_2$, $C$)
-      \State \textsc{AddImplication}($\neglit{L_2}$, $L_1$, $C$)
-      \If{$L_2 = \neglit L_1\sigma$ for some nonidempotent $\sigma$}  \label{statement:self-impl}
-      \ForAll{$i \gets 1$  to $K_\mathrm{imp}$}
-         \State $L_2 \gets L_2\sigma$
-         \State \textsc{AddImplication}($\neglit L_1$, $L_2$, $C$)
-      \EndFor
-      \EndIf
-    \EndIf
-  \EndProcedure
-
-  \kern1\jot
-
-  \Procedure{UntrackClause}{$C$}
-   %  \ForAll {$L_\mathrm{a}, L_\mathrm{c}, M$ such that $ (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$}
-   \ForAll{$L_\mathrm{a}, L_\mathrm{c}, M$ such that $(L_\mathrm{c},M) \in \impmap[L_\mathrm{a}]$}
-      \If{$C \in M$}
-         \State erase $(L_\mathrm{c},M)$ from $\impmap[L_\mathrm{a}]$
-      \EndIf
-      % \EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-%\end{nolinenumbers}
-%\end{figure}
-
-%% TYPESETTING: hack
-\vskip-1.5\jot
-\vskip\belowdisplayskip
-
-The algorithm views a clause $L \cor L'$ as two implications
-$\neglit{L} \impl L'$ and $\neglit{L'} \impl L$.
-%
-It stores only one entry for all literals equal up to variable renaming (line
-\ref{statement:alpha}). Each implication $L_\mathrm{a} \impl L_\mathrm{c}$
-represented by the clause is stored only if its generalization is not present in
-$\impmap$ (line~\ref{statement:more-general}). Conversely, all instances of the
-implication are removed (line~\ref{statement:instances}).
-
-Next, the algorithm
-finds each implication stored in $\impmap$ that can be linked to $L_\mathrm{a} \impl L_\mathrm{c}$: Either
-$L_\mathrm{c}$ becomes the new consequent (line~\ref{statement:transitivity}) or
-$L_\mathrm{a}$ becomes the new antecedent (line~\ref{statement:triggered}). If
-$L_\mathrm{c}$ can be decomposed into $\neqlit{u[s]}{u[t]}$, rule~(3) of
-Definition \ref{def:hl-fo} allows us to store $\neqlit{s}{t}$ in
-$\impmap[L_\mathrm{a}]$ (line~\ref{statement:decomposition}). This is an
-exception to the idea that transitive chains should use only rule~(1). The
-application of rule~(3) does not count toward the bound $K_\mathrm{len}$.
-If $L_\mathrm{a}$ is of the form $\eqlit{u[s]}{u[t]}$, then
-$\impmap$ could be extended so that $\impmap[\eqlit{s}{t}] =
-\impmap[L_\mathrm{a}]$, but this would substantially increase $\impmap$'s
-memory footprint.
-
-In first-order logic, different instances of the same clause can
-be used along a transitive chain. For example, the clause $C =
-\negpredlit{\cst{p}(x)} \cor \pospredlit{\cst{p}(\cst{f}(x))}$ induces $\cst{p}(x) \bigfimpl{M}^i \cst{p}(\cst{f}^i(x))$ for all $i$.
-The algorithm discovers such self-implications (line~\ref{statement:self-impl}): For
-each clause~$C$ of the form $\neglit{L} \cor L\sigma$, where $\sigma$ is
-some nonidempotent substitution, the entires $(L\sigma^2, \{C\}),\allowbreak
-\ldots,\allowbreak (L\sigma^{K_{\mathrm{imp}} + 1}, \{C\})$ are added to
-$\impmap[L]$, where $K_{\mathrm{imp}}$ is a parameter.
-
-\newcommand{\concl}{\ensuremath{\impmap^{-1}}}
-
-\looseness=-1
-To track and untrack clauses efficiently, we implement the mapping $\impmap$ as a
-nonperfect discrimination tree \cite{rsv-2001-term-indexing}. Given a query
-literal $L$, this indexing data structure efficiently finds all literals $L'$ such that
-for some $\sigma$, $L'\sigma = L$ and $\impmap[L'] \not= \emptyset$. We can use
-it to optimize all lookups except the one on line~\ref{statement:transitivity}.
-For this remaining lookup, we add an
-index \concl{} that inverts $\impmap$, i.e., $\concl[L] = \{ L' \mid
-\impmap[L'] = (L,M) \text{ for some } M \}$. To avoid sequentially going through
-all entries in $\impmap$ when the prover deletes them, for each clause $C$
-we keep track of each literal $L$ such that $C$ appears in $\impmap[L]$.
-Finally, we limit the number of entries stored in $\impmap[L]$ -- by default,
-up to 48 pairs in each $\impmap[L]$ are stored.
-
-\newcommand{\unitset}{\ensuremath{\textit{Unit}}}
-\begin{conf} 
-\looseness=-1   
-Rules \infname{HLE} and \infname{HTR} have a simple
-implementation based on $\impmap$ lookups. To implement \infname{UnitHLE} and
-\infname{UnitHTR}, we maintain the index \unitset, containing literals
-$L_\mathrm{c}\sigma$, such that $(L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}]$ for
-some $M$ and $L_\mathrm{a}$ and $\sigma$ is the most general unifier of $L'$ and
-$L_\mathrm{a}$, for some unit clause $\{ L' \}$. The implementation
-of \infname{FLE} and \infname{FLR} also uses \unitset:
-When $(L',M)$ is added to $\impmap[L]$, we
-check if $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $\neglit{L}$
-is added to $\unitset$.
-\end{conf}
-\begin{rep}
-
-   To implement the \infname{HLE} rule, we use $\impmap[L]$ as
-   follows: Given a clause $C = L \cor L' \cor C'$, if there are two literals
-   $L_1,L_2$ and a substitution $\sigma$ such that $(L_2, M) \in \impmap[L_1]$,
-   $C \not\in M$,  $L_1\sigma = L$, and $L_2\sigma = L'$, we remove $L$ from $C$.
-   Literal $L$ can also be removed if $L_1\sigma = \neglit{L'}$ and $L_2\sigma = \neglit{L}$. 
-   Rule \infname{HTR} is implemented analogously.
-
-% @PETAR: Used to say "UnitHTR rules rely", but there's only one. Double-check. --JB
-   The \infname{UnitHTR} rule relies on maintaining the index $\unitset$, which is
-   built as follows. Whenever the prover derives a unit clause $C = \{ L \}$, we
-   find all entries $L_\mathrm{a}$ in $\impmap$ such that $L_\mathrm{a}$ and $L$
-   are unifiable with the most general unifier $\sigma$. Then, we set $\unitset
-   \gets \unitset \cup \{ (L_\mathrm{c}\sigma, M \cup \{ C \} ) \mid
-   (L_\mathrm{c}, M) \in \impmap[L_\mathrm{a}] \}$. Given a clause $L \cor C'$
-   we apply \infname{UnitHLE} by looking for $(L', M) \in \unitset$ such that
-   $L'\sigma = \neglit{L}$, for some substitution $\sigma$; we apply
-   \infname{UnitHTR} by looking for $L'$ such that $L'\sigma = L$.  The sets
-   stored together with literals in $\unitset$ are used for building the proof
-   object and to remove literals from $\unitset$ once a clause from the given
-   set becomes redundant.
-
-   The same data structure is used
-   for supporting \infname{FLE} and \infname{FLR}. When $(L',M)$ is added to $\impmap[L]$, we
-   check whether $(\neglit{L'},M') \in \impmap[L]$ for some $M'$. If so, $(\neglit{L}, M \cup M')$
-   is added to $\unitset$.
-\end{rep}
-
-% % Using this algorithm, hidden-literal-based elimination can be implemented
-% % efficiently. We implement them as simplifications, used during saturation.
-% \begin{rep}
-% We implement \infname{ShallowHLE} by checking if there are two literals $L,L'
-% \in C$ such that for some substitution $\sigma$, and literals $L_1$ and $L_2$,
-% $(L_2, M) \in \impmap[L_1]$ and $L_1\sigma = L$, $L_2\sigma = L'$, $L$ can be
-% removed from $C$. When $(L',M)$ is added to $\impmap[L]$, Zipperposition checks
-% if $(\neglit{L'},M) \in \impmap[L]$ for some $M$. If so, it notes that there
-% are complementary literals in $\HL(\neglit{L},M)$. This information is used to
-% implement \infname{FLD}. For \infname{HTR}, a check whether there are $L,L' \in
-% C$ such that $(L_1, M) \in \impmap[L_0]$, where $\neglit{L_0\sigma} = L$ and
-% $L_1\sigma = L'$ is performed. Rule \infname{DeepHLE} is not implemented.
-
-% Rules \infname{UnitHLE} and \infname{UnitHTR} are implemented as well. For
-% \infname{UnitHLE}, given a clause $L \cor C$ we do as follows. First we try to
-% find a literal $L_0$ such that $\neglit{L_0\sigma} = L$ for some $\sigma$ and
-% $(L_1,M) \in \impmap[L_0]$ such that $L_1\sigma$ is subsumed by a unit clause
-% indexed by Zipperposition. If this does not succeed, we try finding a literal
-% $L_0$ such that $L_0\sigma = L$ for some $\sigma$ and $(L_1,M) \in \impmap[L_0]$
-% such that $\neglit{L_1}$ is subsumed by a unit clause. Rule \infname{UnitHTR} is
-% implemented analogously.
-% \end{rep}
-
-In propositional logic, the conventional approach constructs the
-\emph{binary implication graph} for the clause set~$N$
-\cite{hjb-2011-big-simplification}, with edges $(\neglit{L}, L')$ and
-$(\neglit{L'}, L)$ whenever $L \cor L' \in N.$ To avoid traversing the graph
-repeatedly, solvers rely on timestamps to discover connections between
-literals. This relies on syntactic literal comparisons, which is very
-fast in propositional logic but not in first-order logic, because of
-substitutions and congruence.
-
-\ourpara{Predicate Elimination}
-% 
-To implement portfolio predicate elimination, we maintain a record for each
-predicate symbol $\cst{p}$ occurring in the problem with the
-following fields:\
-set of definition clauses for $\cst{p}$, set of nondefinition clauses in which
-$\cst{p}$ occurs once, and set of clauses in which $\cst{p}$ occurs more than
-once. These records are kept in a priority queue, prioritized by properties such
-as presence of definition sets and number of estimated resolutions. If
-$\cst{p}$ is the highest-priority symbol that is eligible for SPE or DPE, we
-eliminate it by removing all the clauses stored in $\cst{p}$'s record from the
-proof state and by adding flat resolvents to the passive set.
-Eliminating a symbol might make another symbol eligible.
-%In this case, the updated record is put on the processing queue.
-
-As an optimization, predicate elimination keeps track only of symbols that appear
-at most $K_\mathrm{occ}$~times in the clause set. For inprocessing, we use
-signals that the prover emits whenever a clause is added to or removed from
-the proof state and update the records. At the beginning of the
-1st, $(K_\mathrm{iter} + 1)$st, $(2 K_\mathrm{iter} + 1)$st, $\dots{}$ iteration of the given clause procedure's
-loop body, predicate elimination is systematically applied to the entire proof
-state. The first application of inprocessing amounts to preprocessing. By
-default, $K_\mathrm{occ} = 512$ and $K_\mathrm{iter}=10$. The same ideas and
-limits apply for blocked clause elimination.
-
-The most important novel aspect of our predicate elimination implementation is
-recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
-as follows:
-\begin{enumerate}
-   \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{x})} \cor C', C \in
-   N$, no variable repeats in $\tupleempty{x}$, and variables of $C'$ are among
-   $\tupleempty{x}\}$. If $G$ is empty, report failure; otherwise continue.
-
-   \smallskip
-   \item Rename all clauses in $G$ so that their only variables are~$\tupleempty{x}$.
-
-   \smallskip
-   \item\label{step:encoding} Let $\lfloor a \rfloor$ be a function that assigns
-   a propositional variable to each atom $a$. This function is lifted to
-   literals by assigning $\lfloor \neg a \rfloor  = \neg x$, if $\lfloor a
-   \rfloor = x$, and to clauses pointwise. Furthermore,  
-   let $E = \{ \lfloor C' \rfloor \mid
-   \arbpredlit{\cst{p}(\tupleempty{x})} \cor C' \in G \}$. If $E$ is satisfiable,
-   report failure. Else, let $E'$ be the unsatisfiable core of $E$
-   and $G'$ the set of corresponding first-order clauses and continue.
-
-   \smallskip
-   \item If all resolvents in $G'_{\pospredlit{\cst{p}}} \flatres_{\!\cst{p}}
-   G'_{\negpredlit{\cst{p}}}$ are tautologies, then $G'$ is the definition set
-   for symbol $\cst{p}$. Else, report failure.
-\end{enumerate}
-
-The invalidity of set $E$ from step \ref{step:encoding} is checked using a SAT
-solver, which is already integrated in Zipperposition. As modern theorem provers
-(such as E or Vampire) also use SAT solvers, the method can easily be
-implemented.
-
-During experimentation, we noticed that recognizing definitions of symbols that
-occur in the conjecture often harms performance. Thus, Zipperposition
-recognizes definitions only for the remaining symbols.
-
-\begin{qle}
-\ourpara{Quasipure Literal Elimination}
-
-A simple and efficient implementation of QLE uses a SAT solver. Let $N$ be a
-finite $\mathrm{\Sigma}$-clause set. Without loss of generality, we can assume
-that $\mathrm{\Sigma}$ is finite. The signature of the SAT problem consists of
-the variables $\cst{p}^+, \cst{p}^-$ for each predicate symbol $\cst{p}$ in
-$\mathrm{\Sigma}$, where $\cst{p}^s$ means ``$\cst{p}$ is quasipure with
-polarity~$s$.'' The SAT problem consists of the following clauses:
-%
-\begin{enumerate}
-\itemsep1\jot
-\item For each clause $\arbpredlit{\cst{q}_1(\ldots)} \lor \cdots \lor
-  \arbpredlit{\cst{q}_n(\ldots)} \allowbreak\lor C$ contained in $N$, where
-  the first $n$~literals have polarities $s_1,\dots,s_n$, respectively,
-  and $C$ consists of functional literals,
-  generate $n$~clauses of the form
-  \[\cst{q}_1^{s_1}\lor\cdots\lor\cst{q}_{i-1}^{s_{i-1}}
-    \lor
-    \neglit{\cst{q}_i^{-s_i}}
-    \lor
-    \cst{q}_{i+1}^{s_{i+1}}\lor\cdots\lor\cst{q}_n^{s_n}
-   \]
-   where $-s$ flips the polarity $s$. Such a clause ensures that whenever a
-   literal $\arbpredlit{\cst{q}_i(\ldots)}$ has the wrong polarity according to
-   the current variable assignment, one of the other literals must be quasipure.
-
-\item For each $\cst{p}$ in $\Sigma$, generate the clause $\lnot \cst{p}^+ \lor
-  \lnot \cst{p}^-$. It ensures that a single polarity is assigned to a quasipure
-  predicate.
-
-\item Generate the clause $\cst{p}_1^+ \lor \cst{p}_1^- \lor \cdots \lor
-  \cst{p}_n^+ \lor \cst{p}_n^-$, where $\{\cst{p}_1,\dots,\cst{p}_n\}$ are the
-  predicate symbols in $\mathrm{\Sigma}$. It tells the SAT solver to look for a
-  nontrivial solution, in which at least one predicate is quasipure.
-\end{enumerate}
-
-From a satisfying assignment, we can easily read off a predicate set $P$ and a
-polarity map $s$. The process can be iterated until we reach a maximal solution,
-at which point the SAT solver returns a verdict of ``unsatisfiable.''
-
-So far, we have been unable to prove that quasipurity is NP-complete. Given the
-nondeterministic nature of the naive procedure, we suspect that it is, but we
-have not found a reduction from 3SAT to quasipurity. Thus, it is unclear whether
-our use of a SAT solver is fully satisfactory from a theoretical point of view,
-even though it works perfectly well in practice.
-\end{qle}
-
-\section{Evaluation}
-\label{sec:satfol:evaluation}
-
-% {\bf\large TODO: Evaluate \textit{real} PPE in Figure \ref{fig:preprocess} and discuss SPPE and PPE.
-% References to PPE are broken now. }
-
-\newcommand{\baseres}[0]{7897}
-% \newcommand{\baseresplusfive}[0]{1563}
-
-\newcommand{\resnum}[1]{
-   \FPeval{\result}{clip(#1-\baseres)}%
-   \ifnumcomp{\result}{>}{0}{$+\result$}{$\result$}%
-}
-\newcommand{\resnumphantom}[1]{%
-   \FPeval{\result}{clip(#1-\baseres)}%
-   \FPeval{\negresult}{clip(-\result)}%
-   \ifnumcomp{\result}{>}{0}{\phantom{$0$}{${+}\result$}}{\phantom{$0$}{${-}\negresult$}}%
-}
-\newcommand{\resnumbold}[1]{
-   \FPeval{\result}{clip(#1-\baseres)}%
-   \FPeval{\negresult}{clip(-\result)}%
-   \ifnumcomp{\result}{>}{0}{\phantom{${+}\result$}\llap{$\mathbf{\boldsymbol{+}\result}$}}{\phantom{${-}\negresult$}$\llap{\mathbf{\boldsymbol{-}\negresult}}$}%
-}
-\newcommand{\resnumboldphantom}[1]{%
-   \FPeval{\result}{clip(#1-\baseres)}%
-   \FPeval{\negresult}{clip(-\result)}%
-   \ifnumcomp{\result}{>}{0}{\phantom{$0{+}\result$}\llap{$\mathbf{\phantom{0}{\boldsymbol{+}}\result}$}}{\phantom{$0{-}\negresult$}$\llap{\mathbf{\phantom{0}{\boldsymbol{-}}\negresult}}$}%
-}
-\newcommand{\evaluating}[0]{\textbf{?}}
-
-\looseness=-1
-We measure the impact of our elimination techniques for various values of their
-parameters. As a baseline, we use Zipperposition's first-order portfolio mode,
-which runs the prover in 13~configurations of
-heuristic parameters in consecutive time slices. None of these configurations
-use our new techniques. To evaluate a given parameter value, we fix
-it across all 13~configurations and compare the results with the baseline.
-
-\looseness=-1
-The benchmark set consists of all 13\,495 CNF and FOF TPTP 7.3.0 theorems
-\cite{gs-17-tptp}.
-%
-The experiments were carried out on StarExec servers \cite{sst-14-starexec}
-equipped with Intel Xeon E5-2609 CPUs clocked at 2.40 GHz. The portfolio mode
-uses a single CPU core with a CPU time limit of 180~s. The base
-configuration solves \NumberOK{\baseres} problems. The values in the tables
-indicate the number of problems solved minus \NumberOK{\baseres}. Thus,
-positive numbers indicate gains over the baseline. The best result is shown in
-bold.
-
-\ourpara{Hidden-Literal-Based Elimination}
-% 
-The first experiments use all implemented HLBE rules.
-To avoid overburdening Zipperposition, we can enable an option to limit the
-number of tracked clauses for hidden literals. Once the limit has been reached, any
-request for tracking a clause will be rejected until a tracked
-clause is deleted. We can choose which kind of clauses are
-tracked:\ only clauses from the active set $\mathcalx{A}$,
-only clauses from the passive set $\mathcalx{P}$, or both.
-We also vary the maximal implication chain length $K_\mathrm{len}$ and the
-number of computed self-implications $K_\mathrm{imp}$.
-
-In Zipperposition, every lookup for instances or generalizations of
-$\eqlit{s}{t}$ must be done once for each orientation of the equation.
-To avoid this inefficiency, and also because the implementation of hidden literals does not fully
-exploit congruence, we can disable\pagebreak[2] tracking clauses with at least one
-functional literal. Clauses containing functional literals can then still be
-simplified.
-
-Figures~\ref{fig:hle-clauses-neq} and \ref{fig:hle-clauses-eq}
-%~\ref{fig:hle}
-show the results, without and with functional literal tracking enabled,
-for $K_\mathrm{len} = 2$ and $K_\mathrm{imp} = 0$.%
-The columns specify different limits on the number of tracked clauses, with $\infty$
-denoting that no limit is imposed. The rows represent different kinds of tracked clauses.
-The results suggest that tracking functional literals is \NumberOK{not} worth the effort
-but that tracking predicate literals \NumberOK{is}. The best improvement is observed when
-both active and passive clauses are tracked. Normally DISCOUNT-loop
-provers \cite{adf-1995-discount} such as Zipperposition do not simplify
-active clauses using passive clauses, but here we see that this can be effective.
-%
-%--Raw data shows that when up to 250 clauses in either set are
-%--tracked, tracking functional literals helps solve \NumberNOK{5} problems that
-%--cannot be solved by a configuration that tracks only predicate literals or the
-%--base configuration. This suggests that a more optimized indexing implementation
-%--would help our technique reach its full potential.
-Figure~\ref{fig:km-kimp} shows the impact of varying $K_\mathrm{len}$ and
-$K_\mathrm{imp}$, when 500 clauses from the entire proof state are tracked.
-These results suggest that computing long implication chains \NumberOK{is counterproductive.}
-% Furthermore, raw data reveals that there is
-% no gain in this case: not a single problem that configuration with $K_\mathrm{len}=2$ and
-% $K_\mathrm{imp}=0$ cannot solve is solved when $K_\mathrm{imp}K_\mathrm{len}$ is set to 8 and
-% $K_\mathrm{imp}$ to 0.
-
-\begin{figure}[t]%
-\small
-\centering
-\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
-                     & \multicolumn{4}{c}{Tracked clauses} \\
-                     & $250$        & $500$         & $1000$       &$\infty$ \\ \midrule 
-\strut Active       & \resnum{7883}  & \resnum{7881}   & \resnum{7889}  & \resnum{7885}    \\
-\strut Passive      & \resnumphantom{7904}  & \resnum{7907}   & \resnum{7902}  & \resnum{7862}    \\
-\strut Both         & \resnumbold{7909}  & \resnum{7907}   & \resnum{7904}  & \resnum{7852} \\     
-\end{tabular}
-\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when only predicate literals are tracked}
-\label{fig:hle-clauses-neq}
-\end{figure}
-\begin{figure}[t]%
-\small
-\centering
-\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
-                     & \multicolumn{4}{c}{Tracked clauses} \\
-                     & $250$                   & $500$                 & $1000$                    & $\infty$ \\ \midrule 
-\strut Active       & \resnum{7887}            & \resnum{7883}         & \resnumphantom{7889}      & \resnum{7879}    \\
-\strut Passive      & \resnumphantom{7892}     & \resnumphantom{7892}  & \resnum{7883}             & \resnum{7826}    \\
-\strut Both         & \resnumboldphantom{7899} & \resnumphantom{7896}  & \resnumphantom{7889}      & \resnum{7818} \\ 
-\end{tabular}
-\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when all literals are tracked}
-\label{fig:hle-clauses-eq}
-\end{figure}
-
-\begin{figure}[t]
-\small
-\centering
-\def\arraystretch{1.1}%
-\begin{tabular}[t]{l@{\hskip 2em}c@{\hskip 1.5em}c@{\hskip1.5em}c@{\hskip 1.5em}c} 
-                             & \multicolumn{4}{c}{Chain length $K_\mathrm{len}$} \\
-                             & $1$                   & $2$                 & $4$                   & $8$ \\ \midrule 
-   $K_\mathrm{imp}=0$        & \resnum{7906}         & \resnum{7907}       & \resnum{7904}         & \resnum{7902}    \\
-   $K_\mathrm{imp}=1$        & \resnum{7902}         & \resnumbold{7908}       & \resnum{7904}         & \resnum{7901}    \\
-   $K_\mathrm{imp}=2$        & \resnum{7903}         & \resnumbold{7908}       & \resnum{7905}         & \resnum{7905}    \\
-\end{tabular}
-\caption{Impact of the parameters $K_\mathrm{len}$ and $K_\mathrm{imp}$ on HLBE performance}
-\label{fig:km-kimp}
-\end{figure}
-
-\begin{figure}[t!]
-   \small
-   \centering
-   \def\arraystretch{1.1}%
-   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
-                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
-                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
-   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
-   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
-   \end{tabular}
-   \caption{Impact of the choice of criterion on predicate elimination performance}
-   \label{fig:criterion}
-\end{figure}
-
-\begin{figure}[t]
-   \small
-   \centering
-   \def\arraystretch{1.1}%
-   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
-                         &               &               &                &               &                    & \smash{HLBE} \\[-1pt] %% TYPESETTING
-                         &               &               & \smash{SPE}    &               & \smash{PPE}        & \smash{+PPE} \\[-1pt] %% TYPESETTING
-                         & BCE           & SPE           & +BCE           & PPE           & +BCE               & +BCE           \\ \midrule 
-   \strut Preprocessing  & \resnum{7927} & \resnum{8051} & \resnum{8056}  & \resnum{8057} & \resnumbold{8063}  & \resnum{8059} \\
-   \strut Inprocessing   & \resnum{7849} & \resnum{8037} & \resnum{8024}  & \resnum{8043} & \resnum{8028}      & \resnum{8024}     \\
-   \end{tabular}
-   \caption{Performance of predicate and blocked clause elimination}
-   \label{fig:preprocess}
-\end{figure}
-
-\begin{qle}
-\begingroup
-\begin{figure}[t]
-\color{red}
-   \small
-   \centering
-   \def\arraystretch{1.1}%
-   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
-                         &               &               & \smash{QLE}    \\[-1pt] %% TYPESETTING
-                         & PLE           & QLE           & +PPE           \\ \midrule 
-   \strut Preprocessing  & \resnum{9999} & \resnum{9999} & \resnum{9999} \\
-   \strut Inprocessing   & \resnum{9999} & \resnum{9999} & \resnum{9999} \\
-   \end{tabular}
-   \caption{Performance of pure and quasipure literal elimination
-    (TODO: Compute numbers)}
-   \label{fig:ple-qle}
-\end{figure}
-\end{qle}
-
-\renewcommand{\baseres}{856}
-
-\begin{figure}[t]
-   \small
-   \centering
-   \def\arraystretch{1.1}%
-   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1em}c} \\
-                         &              &              &                   &              &              & \smash{HLBE}   \\[-1pt] 
-                         &              &              & \smash{SPE}       &              & \smash{PPE}  & \smash{+PPE}   \\[-1pt] 
-                         & BCE          & SPE          & +BCE              & PPE          & +BCE          & \smash{+BCE}    \\ \midrule 
-   \strut Preprocessing  & \resnum{885} & \resnum{902} & \resnumbold{916}  & \resnum{903} & \resnum{915} & \resnum{911}  \\
-   \end{tabular}
-   \caption{Performance of predicate and blocked clause elimination for establishing satisfiability}
-   \label{fig:bce-sat}
-\end{figure}
-
-\ourpara{Predicate and Blocked Clause Elimination}
-%
-% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
-% \textit{Do not forget to mention SPE was already implemented and }
-For defined predicate elimination, the number of resolvents grows
-exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
-computation, we limit the applicability of PPE to proof states for which $\cst{p}$
-is singular. According to our informal experiments, full PPE,
-without this restriction, generally performs less well.
-
-\begin{qle}\color{red} (TODO: Add QLE)\color{black}\end{qle}
-
-Predicate elimination can be done using Khasidashvili and Korovin's
-criterion (K\&K) or using our relaxed criterion with different values of
-$K_\mathrm{tol}$. Figure~\ref{fig:criterion} shows the results
-for SPE and PPE used as preprocessors.
-Our numbers corroborate Khasidashvili and Korovin's findings:
-SPE with K\&K proves \NumberOK{70} more problems than the base, a
-\NumberOK{0.9\%} increase, comparable to the 1.8\% they observe when
-they combine SPE with additional preprocessing.
-Remarkably, the number of additional proved problems more than
-\NumberOK{doubles} when we use our criterion with $K_\mathrm{tol} > 0$, for both SPE and PPE.
-
-\looseness=-1
-Although this is not evident in Figure~\ref{fig:criterion}, varying
-$K_\mathrm{tol}$ substantially changes the set of problems solved. For example,
-when $K_\mathrm{tol}=0$, SPE proves \NumberOK{60} theorems not proved using
-$K_\mathrm{tol}=50$. The effect weakens as $K_\mathrm{tol}$ grows. When
-$K_\mathrm{tol}=100$, SPE proves only \NumberOK{13} problems not found when
-$K_\mathrm{tol}=200$. Similarly, the set of problems proved by SPE and PPE
-differs: When $K_\mathrm{tol}=25$, \NumberOK{14} problems are proved by PPE but
-missed by SPE. Recognizing definition sets is useful:
-PPE \NumberOK{outperforms} SPE regardless of the criterion.
-
-% \looseness=-1 Having implemented those techniques, we are in position to answer
-% two interesting questions: {How helpful are definition clauses in defined
-% predicate elimination (DPE)?} and {What are the effects of performing blocked
-% clause elimination (BCE) and DPE until fixpoint (BD)?}. The second question was
-% also of interest to Kiesl et al.\ \cite{ksstb-2017-blockedfol}, but was left
-% unanswered as they did not implement predicate elimination.
-
-\looseness=-1
-Performing BCE and variable elimination until fixpoint increases the performance
-of SAT solvers~\cite{jbh-10-BCE}. We can check whether the same holds for superposition provers.
-In this experiment, we use the relaxed criterion with
-$K_\mathrm{tol}=25$ and HLBE which tracks up
-to 500 clauses from any clause set, $K_\mathrm{len}=2$, and
-$K_{\mathrm{imp}}=0$. We use each technique as preprocessing and inprocessing.
-
-The results are summarized in Figure~\ref{fig:preprocess}, where
-the +~sign denotes the combination of techniques.
-We confirm the results obtained by Kiesl et al.\ about the performance of BCE
-as preprocessing:
-It helps prove \NumberOK{30} more problems from our benchmark set, increasing the
-success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained Kiesl et al.
-%
-Using BCE as inprocessing, however, hurts performance, presumably because of
-its incompatibility with the redundancy criterion.
-
-\looseness=-1%
-For preprocessing, the combinations SPE+BCE and PPE+BCE performed roughly \NumberOK{on a
-par} with SPE and PPE, respectively. This stands in contrast to the situation
-with SAT solvers, where such a combination usually helps. It is also worth noting
-that the inprocessing techniques never outperform their preprocessing
-counterparts.
-%
-The last column shows that combining HLBE with other elimination techniques
-\NumberOK{overburdens} the prover.
-
-\begin{qle}
-\ourpara{Pure and Quasipure Literal Elimination}
-\color{red}
-Pure literal elimination is a known technique, but quasipure literals are new,
-so it is quite interesting to find out whether they bring benefits on
-benchmarks. Figure~\ref{ig:ple-qle} considers PLE, QLE, and a combination
-of QLE and portfolio predicate elimination, applied in that order. All three
-modes are tried both as preprocessing and as inprocessing.
-
-We see TODO.
-\endgroup
-\end{qle}
-
-\ourpara{Satisfiability by Blocked Clause Elimination}
-%
-\looseness=-1%
-Kiesl et al.\ found that blocked clause elimination is especially effective on
-satisfiable problems. To corroborate their results and ascertain
-whether a combination of predicate elimination and blocked clause elimination
-increases the success rate, we evaluate BCE on all 2273
-satisfiable or TPTP FOF and CNF problems.
-The hardware and CPU time limits are the same as in the experiments above.
-Figure~\ref{fig:bce-sat} presents the results.
-
-\looseness=-1
-The baseline establishes the satisfiability of \baseres~problems. We consider
-only preprocessing techniques, since BCE compromises refutational
-completeness---a saturation does not guarantee that the original problem was
-satisfiable. We note that recognizing definition sets makes almost no difference on
-satisfiable problems. The sets of problems solved by BCE and PPE
-differ---\NumberOK{30} problems are solved by BCE and not by PPE.
-% Kiesl sees around 2.5%, we see 3.5% with BCE
-
-\begin{rep}
-\section{Discussion and Related Work}
-\label{sec:satfol:discussion-and-related-work}
-
-%\textbf{TODO: This section probably needs to be reworked.}
-
-We briefly surveyed related work in Sect.~\ref{sec:satfol:introduction}. In this
-section, we give a more detail overview and further discuss connections with
-the related work.
-
-The research presented in this \paper{} is two-pronged. For SAT elimination
-techniques already generalized to preprocess first-order problems, we looked for
-ways to interleave them with the given clause procedure of a superposition
-prover, as inprocessing. For techniques that had not yet been ported to
-first-order logic, we looked for generalizations that allow both preprocessing and
-inprocessing.
-
-Hidden tautology elimination was first described
-by Heule et al.\ \cite{hjb-2010-cl-elim}. Better implementation that also
-supports hidden literal elimination was later described by the same group of
-authors \cite{hjb-2011-big-simplification}. We generalized the underlying
-theoretical concepts to first-order logic, and provided an efficient way to deal
-with the infinite number of hidden literals that arise  with this
-generalization. More efficient graph-based techniques are yet to be explored.
-
-Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
-the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
-\cite{ab-2004-re}. It was generalized to first-order logic (as a
-preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
-predicate elimination. Improvement of variable elimination,
-which uses formula definition information, has been popularized as a preprocessing and
-inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
-\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
-logic and combined it with existing extension. With tolerable restrictions,
-this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
-allowing variable elimination to slightly increase in the clause set size
-improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
-achieving the double the number of additional proofs found compared to more
-restrictive approaches.
-
-Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
-solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
-\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
-preprocessor. We showed that blocked clauses cannot be removed during
-saturation, but that they can be effectively used to show satisfiability of the
-clause set. A combination of blocked clause elimination and
-variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
-improvement when their generalizations are combined.
-
-Our approach is one of many ways to combine ideas from SAT solving and
-first-order proving. Other noteworthy architectures that either incorporate a
-SAT solver or that generalize the CDCL calculus include DPLL($T$) with
-quantifier instantiation
-\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
-DPLL($\mathrm{\Gamma}+T$) \cite{bonacina-et-al-2009},
-labeled splitting \cite{fietzke-weidenbach-2009},
-AVATAR \cite{av-2014-avatar},
-MCSAT \cite{de-moura-jovanovic-2013},
-CDSAT \cite{bonacina-et-al-2017}, and
-SGGS~\cite{bonacina-plaisted-2014}.
-
-% See also \url{https://academic.oup.com/logcom/advance-article-abstract/doi/10.1093/logcom/exaa065/6038918}
-
-% OLD TEXT, might be useful: {\sl A similar issue arises with
-% the related SAT technique of \emph{variable addition} \cite{xxx}, which can be
-% used to replace the six-clause set $\{\cst{b} \cor \cst{e}{,}\; \cst{c} \cor
-% \cst{e}{,}\; \cst{d} \cor \cst{e}{,}\; \cst{b} \cor \cst{f}{,}\; \cst{c} \cor
-% \cst{f}{,}\; \cst{d} \cor \cst{f}\}$ by the five-clause set $\{\cst{b} \cor
-% \cst{a}{,}\; \cst{c} \cor \cst{a}{,}\; \cst{d} \cor \cst{a}{,}\; \cst{e} \cor
-% \neglit{\cst{a}}{,}\; \cst{f} \cor \neglit{\cst{a}}\}$, at the cost of one
-% extra symbol ($\cst{a}$).}
-
-% In QBF predicate elimination happens more often than in SAT. Contrast to what we do in this paper.
-%%% I'm not sure what to do with this comment. For me, QBF isn't some
-%%% intermediate logic between SAT and FOL; it's a freak. We can't
-%%% quantify over Booleans in FOL. I also weakened the conjecture about
-%%% "the richer the logic, the more useful PE becomes" in the intro. Anyway,
-%%% I don't think it belongs in a summary/future-work Conclusion, but feel
-%%% free to speculate in the Discussion. ;) --JB
-\end{rep}
-
-\section{Conclusion}
-\label{sec:satfol:conclusion}
-
-We adapted several preprocessing and inprocessing elimination techniques implemented in
-modern SAT solvers so that they work in a superposition prover. This involved lifting
-the techniques to first-order logic with equality but also tailoring them to
-work in tandem with superposition and its redundancy criterion.
-Although SAT solvers and superposition provers embody radically different
-philosophies, we found that the lifted SAT techniques provide valuable optimizations.
-
-We see several avenues for future work. First, the implementation of hidden
-literals could be extended to exploit equality congruence. Second, although
-inprocessing blocked clause elimination is incomplete in general, we hope to achieve
-refutational completeness for a substantial fragment of it. Third, predicate and
-blocked clause elimination, which thrives on the absence of clauses from the
-proof state, could be enhanced by tagging and ignoring generated clauses that
-have not yet been used to subsume or simplify untagged clauses. Fourth,
-predicate and blocked clause elimination could be extended to work with
-functional literals. Fifth, more SAT techniques could be adapted, including
-bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
-addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
-work with other first-order calculi, or generalized
-further to work with higher-order calculi such as combinatory superposition
-\cite{br-20-full-sup-w-combs} and $\lambda$-superposition
-\cite{bentkamp-et-al-2021-lamsup-journal}.
-
-\let\Acksize=\relax
+%%% @PETAR(read): By submission time, CASC will already have taken place!
+% We plan to train the prover better. ahead of the next CASC competition.
+
+\let\Acksize=\footnotesize
 
 \def\ackname{\Acksize Acknowledgment}
+
 \ourpara{\ackname}
 {\Acksize
-We are grateful to the maintainers of StarExec for letting us use their
-service.
-Uwe Waldmann participated in the search for a counterexample to completeness of
-BCE as inprocessing and confirmed that Example~\ref{example:bce-completeness}
-is correct. He also suggested major simplifications and helped us debug the
-proofs of the claims about predicate elimination.
-Anne Baanen helped us define the nonclassical logic used to disallow
-splitting.
-Ahmed Bhayat, Armin Biere, Mathias Fleury, Benjamin Kiesl, and the anonymous
-reviewers made some useful comments on our manuscript, and Mark Summerfield
-suggested many textual improvements. We thank them all.
-
-Vukmirovi\'c and Blanchette's research has received funding from
-the European Research Council (ERC) under the European Union's Horizon 2020
-research and innovation program (grant agreement No.\ 713999, Matryoshka).
+Ahmed Bhayat and Martin Suda provided us
+with Sledgehammer-optimized configurations for Vampire. Andrew Reynolds did the
+same for cvc5. We thank all of them.
 %
-Blanchette's research has also
-received funding from the Netherlands Organization for Scientific Research (NWO)
-under the Vidi program (project No.\ 016.Vidi.189.037, Lean Forward).
+This research has received funding from the European Research Council (ERC)
+under the European Union's Horizon 2020 research and innovation program (grant
+agreement No.\ 713999, Matryoshka).
 %
-Heule is supported by the National Science Foundation (NSF) under grant CCF-2015445.
-
+Blanchette has received funding from the Netherlands Organization for
+Scientific Research (NWO) under the Vidi program (project No.\
+016.Vidi.189.037, Lean Forward).
 }
 
-\bibliographystyle{IEEEtran}
-% \begin{raggedright}
+\bibliographystyle{splncs03}
 \bibliography{bib}
-% \end{raggedright}
 
@@
 
-\def\ackname{Acknowledgment} % American English
-\begin{acknowledgements}
-We are grateful to the maintainers of StarExec for letting us use their service.
-%Mathias Fleury helped us setup the ORCID icons.
-Ahmed Bhayat and Giles Reger guided us
-through details of Vampire 4.5. Ahmed Bhayat, Michael F\"arber,
-Mathias Fleury, Predrag Jani\v ci\'c,
-Mark Summerfield, and the anonymous reviewers suggested content, textual, and
-typesetting improvements. We thank them~all.
-
-Vukmirovi\'c, Bentkamp, and Blanchette's research has received funding from
-the European Research Council (ERC) under the European Union's Horizon 2020
-research and innovation program (grant agreement No.\ 713999, Matry\-osh\-ka).
-%
-Blanchette and Nummelin's research has
-received funding from the Netherlands Organization for Scientific Research (NWO)
-under the Vidi program (project No.\ 016.Vidi.189.037, Lean Forward).
-\end{acknowledgements}
-
-% Authors must disclose all relationships or interests that
-% could have direct or potential influence or impart bias on
-% the work:
-%
-% \section*{Conflict of interest}
-%
-% The authors declare that they have no conflict of interest.
-
-% BibTeX users please use one of
-\bibliographystyle{spmpsci}
-
-\bibliography{bib}
-
-\end{document}
-% end of file template.tex
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "conf"
+%%% End:

### 2022-03-24 ehoh2/ehoh2.tex
bools
intro

@@
-\chapter{Extending a High-Performance Prover to Higher-Order Logic}
-\setheader{Extending a High-Performance Prover to Higher-Order Logic}
+\chapter{Extending a Brainiac Prover to Higher-Order Logic}
+\setheader{Extending a Brainiac Prover to Higher-Order Logic}
 \label{ch:ehoh2}
+
+% \renewcommand{chapter}[0]{chapter}
+\newcommand\ehohii{$\lambda$E}
+
+\blfootnote{In this work I designed, implemented and evaluated all changes to
+term representation, algorithms and indexing data structures. Jasmin Blanchette
+did the daily supervision. Stephan Schulz provided
+the necessary E expertise.}
+
+\begin{abstract}%
+    The automatic discharge of tedious subgoals is high on the wishlist of many
+  users of proof assistants. Some proof assistants discharge such goals
+  by translating them to first-order logic and invoking an efficient prover on
+  them, but much is lost in translation. As an alternative,
+  we propose to extend first-order provers with native support for
+  higher-order features. Building on our extension of E to $\lambda$-free
+  higher-order logic, we now extend E to full higher-order logic.
+  The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
+  proof assistant, and the second best on TPTP benchmarks.
+    %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
+\end{abstract}
+
+\newpage
+
+\section{Introduction}
+\label{sec:ehoh2:introduction}
+
+We began this thesis by introducing Ehoh, a rather conservative extension of
+state-of-the-art first-order prover to a fragment of higher-order logic. This
+extension gave us a flavor of the difficulties that we might encounter on the
+way to full higher-order logic. In chapters that precede this one,
+we discussed many ways in which those difficulties can be overcome.
+In this chapter, we fulfill a promise we gave in the beginning of the thesis: We present the extension of
+Ehoh to full higher-order logic (Sect.~\ref{sec:ehoh2:logic}) using incomplete variants
+of $\lambda$-superposition. We call this prover \ehohii.
+
+%
+The $\lambda$-superposition calculi were
+previously implemented in Zipperposition, and
+extensive experiments with various heuristic choices have been performed
+(Chapter \ref{ch:ho-techniques}). In \ehohii{}'s implementation, we used
+these experiences to choose a set of effective rules that could easily be
+retrofitted into an originally first-order prover. Another principle that guided 
+the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
+do not impact the strong first-order performance of E and $\lambda$-free higher-order performance of Ehoh.
+
+% We
+% also used the experience of fine-tuning the calculi
+% \cite{section-making-ho-work} in Zipperposition to carefully choose which
+% calculus extensions and heuristics to use in E.
+
+One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
+term representation (Sect.~\ref{sec:ehoh2:terms}). Furthermore, Ehoh's main inference
+engine was designed with the assumption that it will be used with
+inferences that compute a most general unifier. We
+implementd a higher-order unification procedure (Chapter \ref{ch:unif})
+that can return multiple unifiers (Sect.~\ref{sec:ehoh2:unif-match-index}) and
+integrated it in the inference engine. Finally, we extended and adapted the
+superposition rule, resulting in an incomplete, pragmatic variant of
+$\lambda$-superposition (Sect.~\ref{sec:ehoh2:calculus}).
+
+We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
+as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
+(Sect.~\ref{sec:ehoh2:eval}). We found
+that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on
+proof assistant benchmarks; on TPTP benchmarks it ended up second only 
+to the cooperative version of Zipperposition, which employs Ehoh as a
+backend. An arguably fairer comparison without the backend puts \ehohii{} in the
+first place for both benchmark suites.
+We also compared the performance of \ehohii{} with E on first-order
+%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
+%%% check if you agree. --JB
+problems and found that little overhead has been introduced by the
+extension to higher-order logic.

### 2022-03-24 pre/pre.tex
intro

@@
 \newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
-The {\em literal selection function} is a function that selects a (multi)subset of literals
+The {\em literal selection function} selects a (multi)subset of literals
 from a given (multi)set of literals. Superposition requires that at least one of the selected literals is negative.
+\pagebreak[2]
 
@@
 %
-Note that we use the name $\lambda$-superposition interchangeably for \lsup{}
-and \osup{} where the context gives enough information to disambiguate this name.
+Note that we use the name $\lambda$-superposition when we refer to both \lsup{}
+and \osup{}.
 

### 2022-03-24 scratch/paper.tex
intro

@@
   higher-order logic, we now extend E to full higher-order logic.
-  The resulting prover is the second strongest on both TPTP benchmarks and
-  benchmarks coming from a proof assistant.
+  The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
+  proof assistant, and the second best on TPTP benchmarks.
   %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
@@
 \section{Introduction}
-\label{sec:introduction}
+\label{sec:ehoh2:introduction}
 
@@
 
-In this \paper, we fulfill a three-year-old promise: We present the extension of
-Ehoh to full higher-order logic (Sect.~\ref{sec:logic}) using incomplete variants
+In this chapter, we fulfill a three-year-old promise: We present the extension of
+Ehoh to full higher-order logic (Sect.~\ref{sec:ehoh2:logic}) using incomplete variants
 of $\lambda$-superposition. We call this prover \ehohii.
@@
 the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
-do not impact already strong first-order performance of Ehoh.
-
-An simpler approach to support higher-order logic is to implement
-combinatory superposition, which is based on \textsf{SKBCI}-combinators
-and implemented in Vampire \cite{br-20-full-sup-w-combs}.
-These can be supported as a thin layer on top of $\lambda$-free terms. However,
-in unpublished experiments, we found that $\lambda$-superposition performs
-substantially better, corroborating the CASC results, so we
-decided to make a more profound change to Ehoh and implement
-$\lambda$-superposition.
+do not impact the strong first-order performance of E and Ehoh.
 
@@
 One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
-term representation (Sect.~\ref{sec:terms}). Furthermore, Ehoh's main inference
+term representation (Sect.~\ref{sec:ehoh2:terms}). Furthermore, Ehoh's main inference
 engine was designed with the assumption that it will be used with
@@
 implementd a higher-order unification procedure \cite{unif-section}
-that can return multiple unifiers (Sect.~\ref{sec:unif-match-index}) and
+that can return multiple unifiers (Sect.~\ref{sec:ehoh2:unif-match-index}) and
 integrated it in the inference engine. Finally, we extended and adapted the
 superposition rule, resulting in an incomplete, pragmatic variant of
-$\lambda$-superposition (Sect.~\ref{sec:calculus}).
-
-% most likely text
+$\lambda$-superposition (Sect.~\ref{sec:ehoh2:calculus}).
+
 We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
 as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
-(Sect.~\ref{sec:eval}). We found
-that \ehohii{} clearly outperforms Ehoh on all benchmarks. Compared
-with other higher-order provers, it is second
-only to the cooperative version of Zipperposition, which employs Ehoh as a
-backend. A fairer comparison without the backend puts \ehohii{} in the
+(Sect.~\ref{sec:ehoh2:eval}). We found
+that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on
+proof assistant benchmarks; on TPTP benchmarks it ended up second only to 
+to the cooperative version of Zipperposition, which employs Ehoh as a
+backend. An arguably fairer comparison without the backend puts \ehohii{} in the
 first place for both benchmark suites.
@@
 \section{Logic} %---0.5p
-\label{sec:logic}
+\label{sec:ehoh2:logic}
 
@@
 We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
-$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$.
-Every $\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
+$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
+$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
 \overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
-of the term. If $s$ is a free variable, we call the term \emph{flex};
-otherwise, the term is \emph{rigid}. If the type of a term is of the form $\tau_1 \to \cdots \to
-\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
-call it a \emph{predicate}. A literal $l$ is an equation $s \eq t$ or
-a disequation $s \not\eq t$. A clause is a finite multiset of literals,
+of the term. If $s$ is a free variable, we call the term \emph{flex}; otherwise,
+the term is \emph{rigid}. A term of type $o$, where $o$ is the distinguished Boolean type,
+is called a \emph{formula}. A type whose type is of the form $\tau_1 \to
+\cdots \to \tau_n \to o$ is called
+a \emph{predicate}. Logical symbols are part of the signature and
+may thus occur within terms. We write them in bold: $\pmb{\neg}, \iand, \ior,
+\iimplies, \iequiv, \ieq, \dots$.
+
+On top of the terms, we define some clausal structure, as is customary in
+first-order logic.
+A literal $l$ is an equation $s \eq t$ or a
+disequation $s \not\eq t$. A clause is a finite multiset of literals,
 interpreted and written disjunctively:\ $l_1 \llor \cdots \llor l_n$.
-Logical symbols that may occur within terms are written in bold:
-$\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Predicate
-literals are encoded as (dis)equations with $\itrue$ based on their sign; for
+Notice that the clause-level operators are not set in bold.
+Predicate literals are encoded as (dis)equations with $\itrue$ based on their sign; for
 example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
-$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.
+$\neg\,\cst{even}(x)$ as $\neglit{\cst{even}(x)}$.
 
 \section{Terms}
-\label{sec:terms}
-
-Ehoh's term representation is centered around perfect term sharing %or exa
-\cite{ls-01-shared}: Any two structurally identical terms are
+\label{sec:ehoh2:terms}
+
+\looseness=-1
+E has been designed around perfect term sharing %or exa
+\cite{ls-01-shared}, a design that we carried on to  Ehoh and \ehohii{}: Any two structurally identical terms are
 guaranteed to be the same object in memory. This is achieved through term
@@
 indices mostly behave as constants. Therefore, we decided to represent De
-Bruijn indices using positive \texttt{f\_code}s: De Bruijn index of value $i$
-will have $i$ as the \texttt{f\_code}. To ensure De Bruijn indices are not
+Bruijn indices using positive \texttt{f\_code}s: The De Bruijn index of value $i$
+will have $i$ as the~\verb|f_code|. To ensure De Bruijn indices are not
 mistaken for function symbols, we use the \texttt{properties} bitfield of the
@@
 %
-We introduce the property \texttt{HasBetaReducibleSubterm} that indicates whether
+We introduce the property \texttt{HasBetaReducibleSubterm} which is set if
 %%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
@@
 where $t$ is not a
-$\lambda$-abstraction. Then \ehohii{} breaks down the body~$t$ into a maximal
+$\lambda$-abstraction, and $l$ is the number of $\internallam$ symbols above $t$. Then \ehohii{} breaks down the body~$t$ into a maximal
 decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
@@
 %%% binder remains (the one to which the 2's point). --JB
-We can then remove the $k = \min(j,n)$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
+We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
 \ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
@@
 \section{Unification, Matching, and Term Indexing}
-\label{sec:unif-match-index}
+\label{sec:ehoh2:unif-match-index}
 % \begin{itemize}
@@
 Vukmirovi\'c et al.\ \cite{unif-section} designed an efficient
-procedure to enumerate a CSU for a term pair. They
-implemented it in Zipperposition, together with some extensions to term indexing.
+procedure to enumerate a CSU for a term pair. It is implemented in Zipperposition, together with some extensions to term indexing.
 In \ehohii{}, we further improve the performance of this procedure by implementing a terminating,
@@
 
-\label{sec:calculus}
+\label{sec:ehoh2:calculus}
 % \begin{itemize}  
@@
 the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
-Sect.~\ref{sec:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
-RPO orders designed for $\lambda$-terms.
+Sect.~\ref{sec:ehoh2:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
+LPO orders designed for $\lambda$-terms.
 % as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.
@@
 %TODO: @JASMIN: How do we cite the STTT article?
-Third difference is more subtle. One of the main features of Ehoh is
+The third difference is more subtle. One of the main features of Ehoh is
 \emph{prefix optimization} \cite[Sect.~1]{section-ehoh}: a method which, given a demodulator $s \eq t$, allows
@@
 \section{Evaluation}
-\label{sec:eval}
+\label{sec:ehoh2:eval}
 
@@
 these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
-benchmars generated by Sledgehammer (SH) \cite{pb-12-sh}.
+benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
 The experiments were carried out on StarExec Miami
@@
 2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
-wallclock and 960~s CPU. For SH, we used Sledgehammer's
+wallclock and 960~s CPU. For SH and to answer the other question, we used Sledgehammer's
 default time limit:\ 30~s wallclock and CPU.
@@
 \ehohii{} compete with the top contenders in the higher-order division of
-CASC-28:\ cvc5
-\NumberOK{0.0.7},\footnote{\url{https://cvc5.github.io/}}
+CASC-28:\ cvc5 \NumberOK{0.0.7},\footnote{\url{https://cvc5.github.io/}}
 Ehoh 2.7
@@
 compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
-and \NumberOK{3\%} more problems on SH benchmarks, where Ehoh was already very
+and \NumberOK{6\%} more problems on SH benchmarks, where Ehoh was already very
 successful.
 
-\ehohii{} also outperforms all other higher-order provers other than Zipperposition-coop.
+\ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
+excels on SH benchmarks. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
 If Zipperposition's Ehoh backend is disabled,
@@
 \ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
-\NumberOK{144} SH problems and \NumberOK{24} TPTP problems that Zipperposition-coop
+\NumberNOK{166} SH problems and \NumberNOK{21} TPTP problems that Zipperposition-coop
 does not. The lower number of uniquely solved TPTP problems is most likely 
@@
 evaluation question, about assessing overhead of \ehohii{},
-we chose all the \NumberOK{1138} unique problems used at CASC
+we chose all the \NumberNOK{1138} unique problems used at CASC
 from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
@@
       Ehoh                  & 2105       & 2611      \\
-      \ehohii{}             & 2527       & 2699      \\
+      \ehohii{}             & 2523       & {\bf 2766}\\
       Leo-III               & 2282       & 1601      \\
@@
       Vampire               & 2203       & 2240      \\
-      Zipperposition-coop   & {\bf 2583} & {\bf 2754}\\
+      Zipperposition-coop   & {\bf 2583} & 2754      \\
       Zipperposition-uncoop & 2483       & 2181      \\
@@
       \midrule
-      Ehoh--FO      & 535      \\
-      Ehoh--HO      & 538      \\
-      \ehohii{}--FO & 533      \\
-      \ehohii{}--HO & {\bf 541}\\
+      Ehoh -- FO      & 535      \\
+      Ehoh -- HO      & 538      \\
+      \ehohii{} -- FO & {\bf 539}\\
+      \ehohii{} -- HO & {\bf 539}\\
       \bottomrule
@@
 \section{Discussion and Related Work} %---0.5p
-\label{sec:discussion}
+\label{sec:ehoh2:discussion}
 
@@
 Extending first-order provers with higher-order reasoning capabilities has been
-tried by other researchers as well. Bhayat and Reger first extended Vampire to
-higher-order logic using combinatory unification \cite{br-20-full-sup-w-combs},
-an incomplete approach, before they designed an implemented a complete
-higher-order superposition calculus based on combinators \cite{br-19-restricted-unif}.
-This calculus is also implemented in Zipperposition. Barbosa et al.\ extended
+attempted by other researchers as well.
+Barbosa et al.\ extended
 the SMT solvers CVC4 (now cvc5) and veriT to higher-order logic in an incomplete
 way \cite{brotb-19-ho-smt}.
+Bhayat and Reger first extended Vampire to
+higher-order logic using combinatory unification \cite{br-20-full-sup-w-combs},
+an incomplete approach, before they designed and implemented a complete
+higher-order superposition calculus based on \textsf{SKBCI}-com\-binators \cite{br-19-comb-unif}.
+The advantage is that combinators can be supported as a thin layer on top of
+$\lambda$-free terms.
+This calculus is also implemented in Zipperposition.
+However, in unpublished experiments, we found that $\lambda$-superposition
+performs substantially better, corroborating the CASC results, so we decided to
+make a more profound change to Ehoh and implement $\lambda$-superposition.
 
@@
 \section{Conclusion} %---0.5p
-\label{sec:conclusion}
+\label{sec:ehoh2:conclusion}
 
@@
 We ended up addressing the theoretical ``difficult challenges'' in other work
-with colleagues. In this \paper, we faced the practical challenges pertaining to the extension
+with colleagues. In this chapter, we faced the practical challenges pertaining to the extension
 of Ehoh's data structures and algorithms to support full higher-order logic and
 demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
-that this extension makes \ehohii{} the second best higher-order prover on
-benchmarks coming from interactive theorem proving practice. \ehohii{}'s
-performance lags slightly behind Zipperposition, possibly because Zipperposition
+that this extension makes \ehohii{} the best higher-order prover on
+benchmarks coming from interactive theorem proving practice, which has been our goal since 2017. \ehohii{}'s
+performance lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
 does not assume a clausal structure and can perform subtle formula-level

### 2022-03-25 ehoh2/ehoh2.tex
terms

@@
 
-\section{Introduction}
+\section{Introduction and Background}
 \label{sec:ehoh2:introduction}
 
-We began this thesis by introducing Ehoh, a rather conservative extension of
-state-of-the-art first-order prover to a fragment of higher-order logic. This
+In Chapter~\ref{ch:ehoh} of this thesis we introduced Ehoh, a rather conservative extension of
+state-of-the-art first-order prover to a fragment of higher-order logic devoid of $\lambda$-abstraction. This
 extension gave us a flavor of the difficulties that we might encounter on the
@@
 In this chapter, we fulfill a promise we gave in the beginning of the thesis: We present the extension of
-Ehoh to full higher-order logic (Sect.~\ref{sec:ehoh2:logic}) using incomplete variants
+Ehoh to full higher-order logic using incomplete variants
 of $\lambda$-superposition. We call this prover \ehohii.
@@
 inferences that compute a most general unifier. We
-implementd a higher-order unification procedure (Chapter \ref{ch:unif})
+implemented a higher-order unification procedure (Chapter \ref{ch:unif})
 that can return multiple unifiers (Sect.~\ref{sec:ehoh2:unif-match-index}) and
@@
 extension to higher-order logic.
+
+\ourpara{Background} The logic \ehohii{} targets is the monomorphic higher-order
+logic described in Sect.~\ref{sec:pre:hol}. We reuse all the notions from this
+section, with the corresponding notations. Like in the previous chapter, we
+simplify the notation by writing predicate literals in unencoded form: Positive
+literal $\eqlit{s}{\itrue}$ is written as $s$ and negative literal
+$\neqlit{s}{\top}$ is written as $\negpredlit{s}$.   This chapter discusses
+three tightly related provers E, Ehoh and \ehohii{}, which are disambiguated as
+follows:
+
+\begin{itemize}
+  \item E is a state-of-the-art first-order prover based on superposition. It is described in
+  Sect.~\ref{sec:pre:theorem-provers}.
+  \item Ehoh is an extension of E to support $\lambda$-free higher-order logic. Chapter \ref{ch:ehoh}
+  is dedicated to Ehoh.
+  \item \ehohii{} further builds on Ehoh to support full higher-order logic. It is the latest prover
+  described in this chapter.
+\end{itemize}
+
+\section{Terms}
+\label{sec:ehoh2:terms}
+
+\looseness=-1
+E has been designed around perfect term sharing %or exa
+\cite{ls-01-shared}, a design that we carried on to  Ehoh and \ehohii{}: Any two structurally identical terms are
+guaranteed to be the same object in memory. This is achieved through term
+\emph{cells}, which represent individual terms. Each cell has (among other fields)
+(1)~\texttt{f\_code}, an integer corresponding to the symbol at the head of the term (negative
+if the head is a free variable, positive otherwise); (2)~\texttt{num\_args},
+corresponding to the number of arguments applied to the head; and (3)~\texttt{args},
+a size-\texttt{num\_args} array of pointers to argument terms. We
+use the first-order notation $\cst{f}(s_1, \ldots, s_n)$ to denote a cell whose
+\texttt{f\_code} corresponds to $\cst{f}$, \texttt{num\_args} equals $n$, and
+\texttt{args} points to the cells for $s_1, \ldots s_n$.
+
+Ehoh represents $\lambda$-free higher-order terms using flattened, spine notation
+(Sect.~\ref{sec:ehoh:types-and-terms}).
+Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$ and $\cst{f} \, \cst{a} \,
+\cst{b}$ are represented by the cells $\cst{f}$,
+$\cst{f}(\cst{a})$, and $\cst{f}(\cst{a}, \cst{b})$, respectively.
+To ensure free variables are
+perfectly shared, Ehoh treats applied free variables differently: arguments are
+not applied directly to a free variable, but using an internal symbol
+\internalat{} of variable arity. For example, the term $X \, \cst{a} \, \cst{b}$ is
+represented by the cell $\internalat(X, \cst{a}, \cst{b})$. This ensures that
+two different occurrences of the free variable $X$ correspond to the same object,
+which makes substitutions more efficient.
+%%% @PETAR(read): This tries to explain an invisible problem. Either we need to explain
+%%% it better, or we just omit it (my favorite option). It's old Ehoh stuff anyway. --JB
+%Yet, to
+%efficiently access variable arguments we do not use binary application symbol,
+%but flattened representation.
+
+\ourpara{Representation of $\lambdabf$-Terms} To support full
+higher-order logic, Ehoh's $\lambda$-free cell data structure must
+be extended to support the $\lambda$ binder. We use the locally nameless
+representation \cite{ac-12-locally-nameless} for this purpose: De Bruijn indices
+%%% @PETAR(checked): Double-check "possibly loose".
+represent (possibly loose) bound variables, whereas we keep the current
+representation for free (and applied) variables.
+
+%%% @PETAR(read): Sounds odd---we wouldn't be extending the term representation of an
+%%% already HO prover. --JB
+%Unlike other higher-order provers, most of which are written in functional
+%programming languages,
+Extending the term representation of Ehoh with a new term
+kind involves intricate manipulation of the cell data structure. De Bruijn
+indices must be represented as other cells with either negative or positive
+\texttt{f\_code}. However, this has to be done in such a way that De Bruijn
+index can never be mistaken for a function symbol or a variable.
+
+Other than possibly being instantiated during $\beta$-reduction, De Bruijn
+indices mostly behave as constants. Therefore, we decided to represent De
+Bruijn indices using positive \texttt{f\_code}s: The De Bruijn index of value $i$
+will have $i$ as the~\verb|f_code|. To ensure De Bruijn indices are not
+mistaken for function symbols, we use the \texttt{properties} bitfield of the
+cell, which holds precomputed properties of %shared cell
+%%% @PETAR(checked, you are right): I don't understand why you wrote "shared cell". Please double-check. --JB
+the cell. We introduce the
+property \texttt{IsDBVar} to denote that the cell represents a De
+Bruijn index. Any attempt to create a De Bruijn index is performed through
+a dedicated library function that sets \texttt{IsDBVar} property for every term it
+returns. When given the same De Bruijn index and type, this function is
+guaranteed to always return the same object. Finally, we have guarded all the
+functions and macros that manipulate function codes with the check if the
+property \texttt{IsDBVar} is set. To ensure perfect sharing of every occurrence
+of De Bruijn indices, arguments to De Bruijn indices are applied like for free
+variables, using \internalat{}.
+
+Extending cells to support $\lambda$-abstraction is easier. Each
+$\lambda$-ab\-strac\-tion has the distinguished function code \internallam{} as the head
+symbol and two arguments:\ (1)~a De Bruijn index 0 of the type of abstracted variable;
+(2)~the body %(matrix)
+of the $\lambda$-abstraction. Consider
+the term $\lambda x. \, \lambda y.\, \cst{f}\, x \, x$, where both $x$ and $y$ have
+the type~$\iota$. This term is represented as $\lambda\,\lambda\, \cst{f} \,
+\db{1} \, \db{1}$ in locally nameless representation, where bold numbers
+represent De Bruijn indices. In \ehohii{}, the same term is represented by the cell
+$\internallam(\db{0}, \internallam(\db{0}, \cst{f}(\db{1}, \db{1})))$,
+where all De Bruijn variables have type~$\iota$.
+
+The first argument of $\internallam$ is
+redundant, since it can be deduced from the type of the $\lambda$-abstraction.
+However, basic $\lambda$-term manipulation operations often require access to
+this term. We store it explicitly to avoid creating it repeatedly.
+
+\ourpara{Efficient $\betabf$-Reduction}
+Terms are stored in $\beta\eta$-reduced form. As these two reductions are
+performed very often, they ought to be efficient. \ehohii{}
+performs $\beta$-reduction by reducing the leftmost outermost $\beta$-redex
+first. To represent $\beta$-redexes, it uses the \internalat{} symbol. Thus,
+the term $(\lambda
+x.\, \lambda y.\,  (x \, y)) \, \cst{f} \, \cst{a}$ is represented by
+$\internalat(\internallam(\db{0}, \internallam(\db{0}, \internalat(\db{1}, \db{0}))),
+\cst{f}, \cst{a})$. Another option would have been to add arguments applied to
+$\lambda$-term directly to the $\lambda$ representation (as in
+$\internallam(\db{0}, \internallam(\db{0}, \allowbreak \internalat(\db{1}, \db{0})),
+\cst{f}, \cst{a})$), but this would break the invariant
+that $\internallam$ has two arguments. Furthermore, replacing free
+variables with $\lambda$-abstractions (e.g., replacing $X$ with $\lambda
+x. \, x$ in $\internalat(X, \cst{a})$) would require additional normalization.
+
+A term can be $\beta$-reduced as follows: When a cell of the
+form $ \internalat(\internallam(\db{0}, s),t)$ is encountered, the field
+\texttt{binding} (normally used to record the substitution for a free variable) of the
+cell $\db{0}$ is set to $t$. Then $s$ is traversed to instantiate every
+loose occurrence of $\db{0}$ in $s$ with \texttt{binding}, whose loosely
+bound De Bruijn indices are shifted by the number of $\lambda$ binders above
+the occurrence of $\db{0}$ in $s$ \cite{fk-01-deBruijn}. Next, the same
+procedure is performed on the resulting term and its subterms, in
+leftmost outermost fashion.
+
+\ehohii{}'s basic $\beta$-normalization mechanism works in this way, but it
+features a few optimizations.
+First, \ehohii{} recognizes terms of the form $(\lam{\overline{x}_n}{s}) \,\overline{t}_n$ 
+and performs parallel replacement of the bound variables $x_i$ with
+$t_i$. Since intermediate terms are not constructed, this reduces the number of
+recursive function calls and calls to the cell allocator.
+
+Second, in line with the gracefulness principle, we want \ehohii{} to incur
+little (if any) overhead on first-order problems and to
+excel on higher-order problems with a large first-order component. If
+$\beta$-reduction is implemented naively, finding a $\beta$-redex involves
+traversing the entire term. On purely first-order terms, $\beta$-reduction
+is then a complete waste of time.
+
+To avoid this, we use Ehoh's perfectly shared
+terms and their \texttt{properties} field.
+%
+We introduce the property \texttt{HasBetaReducibleSubterm} which is set if
+%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
+a cell is $\beta$-reducible.
+Whenever a new cell that contains a
+$\beta$-reducible term as a direct subterm is shared, the property is set.
+Setting of the property is inductively continued when further superterms are
+shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
+x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
+$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
+\texttt{HasBetaReducibleSubterm} set.
+%
+When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
+cells with this property set. This further means that on first-order
+subterms, a single bit masking operation is enough to determine that no subterm
+should be visited.
+
+Along similar lines, we added a property \texttt{HasDBSubterm} that
+caches whether the cell contains a De Bruijn subterm. This
+%% @PETAR(you are right): Why "also"? What's the other thing?
+%also
+makes
+instantiating De Bruijn indices during $\beta$-norma\-lization faster, since only the
+subterms that contain De Bruijn indices must be visited. Similarly, some other
+operations such as shifting De Bruijn indices or determining whether a term is closed
+(i.e., it contains no loose bound variables) can be sped up or even avoided
+if the term is first-order.
+
+\ourpara{Efficient $\etabf$-Reduction}
+The term $\lambda x.\, s \, x$ is $\eta$-reduced to $s$ whenever $x$ does not appear
+unbound in $s$. Caching the property of $\eta$-reducibility of the term is not
+as beneficial as the one for $\beta$-reducibility, because checking this
+property at the term's top level is done in $O(|s|)$, compared with constant time
+for $\beta$-reducibility.
+However, we use the observation that a term cannot be $\eta$-reduced if it has
+no $\lambda$-abstraction subterms and introduce a property \texttt{HasLambda}
+that notes the presence of $\lambda$-abstraction in a term. Only
+terms with this property are visited during $\eta$-reduction.
+
+\ehohii{} performs parallel $\eta$-reduction: It recognizes terms of the form
+$\lam{\overline{x}_n}{s \, \overline{x}_n } $ such that none of $x_i$
+occurs unbound in $s$. If done naively, reducing terms of this kind requires up to $n$
+traversals of $s$ to check if each $x_i$ occurs in $s$. In \ehohii{}, exactly one
+traversal of $s$ is required.
+
+\looseness=-1
+More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
+\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
+the cell will be of the form
+$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
+where $t$ is not a
+$\lambda$-abstraction, and $l$ is the number of $\internallam$ symbols above $t$. Then \ehohii{} breaks down the body~$t$ into a maximal
+decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
+If $n = 0$, the cell is not $\eta$-reducible.
+Otherwise, $u$ is traversed to determine the
+minimal index~$\dbvar{j}$ of a loose De Bruijn index,
+taking $\dbvar{j} = \infty$ if no such index exists.
+%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
+%bottom-up as innermost can suggest): Check "innermost".
+%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
+%%% binder remains (the one to which the 2's point). --JB
+We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
+\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
+$t$ by the variant of
+$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
+obtained by shifting the loose De Bruijn indices right by~$k$.
+
+To better understand this convoluted De Bruijn arithmetic, consider the
+term $\lambda x. \, \lambda y. \, \lambda z.\,\allowbreak \cst{f} \, x \, x \, y
+\, z$. This term is represented by the cell $\internallam(\db{0},
+\internallam(\db{0}, \internallam(\db{0},\allowbreak \cst{f}(\db{2}, \db{2}, \db{1},
+\db{0}))))$. \ehohii{} splits $\cst{f}(\db{2}, \db{2}, \db{1}, \db{0})$ into
+two parts:\ $u = \cst{f} \, \db{2}$ and the arguments $\db{2}, \db{1},
+\db{0}$. Since the minimal index in $u$ is $\db{2}$, we can
+omit the De Bruijn indices $\db{1}$ and $\db{0}$ and their $\lambda$ binders,
+yielding the $\eta$-reduced cell $\internallam(\db{0}, \cst{f}(\db{0},
+\db{0}))$.
+
+The use of the \texttt{HasLambda} property ensures that $\eta$-reduction is not
+tried on first-order or $\lambda$-free higher-order terms, whereas parallel
+$\eta$-reduction both speeds up $\eta$-reduction and avoids creating a linear
+number of intermediate terms. For finding the minimal loose De Bruijn index,
+optimizations such as the \texttt{HasDBSub\-term} property are used.
+
+\ourpara{Representation of Boolean Terms}
+Ehoh represents Boolean terms using
+cells whose \texttt{f\_code}s correspond to internal codes reserved
+for logical symbols. Quantified formulas are represented by cells in which the
+first argument is the quantified variable and the second one is the body of the
+quantified formula. For example, the term $\iforall x.\, \cst{p} \, x$ corresponds
+to the cell $\iforall(X, \cst{p}(X))$, where $X$ is a regular free variable.
+% as Ehoh has no concept of a bound variable.
+
+This representation is convenient for parsing formulas and clausification, which
+is what Ehoh uses it for, but it
+causes $\alpha$-normalization issues during the actual proof search: In
+full higher-order logic, Boolean terms can appear as subterms in clauses, as in
+$\cst{q}(X) \llor \cst{p}(\iforall(X, \cst{r}(X)))$;
+instantiating $X$ in the first literal should not influence $X$ in the second
+literal.
+
+To avoid this issue, in \ehohii{} we use $\lambda$ binders to represent quantified formulas.
+Thus, $\iforall x. \, s$ is represented by $\iforall \, (\lambda x. \, s)$.
+Quantifiers are then unary symbols that do not directly bind
+the variables but delegate this responsibility to a $\lambda$-abstraction.
+%
+Since \ehohii{} represents bound variables using De Bruijn indices, this solves the
+$\alpha$-conversion issues. However, this solution is incompatible with
+thousands of decades-old lines of clausification code that assumes the Ehoh
+representation of
+quantified formulas. Therefore, \ehohii{} converts quantified
+formulas only after clausification, for Boolean terms that appear in a
+higher-order context (e.g., as argument to a function symbol).
+
+\ourpara{New Term Orders}
+The $\lambda$-superposition calculus is parameterized by a term order that is
+used to break symmetries in the search space.
+We implemented the versions of the Knuth--Bendix order (KBO) and lexicographic
+path order (LPO) for higher-order terms with $\lambda$-abstractions described by
+Bentkamp et al.~\cite{bbtv-21-full-ho-sup}. These orders encode
+encode $\lambda$-terms as first-order terms and then invoke the standard KBO or
+LPO. %Trading elegance for efficiency, our
+%implementation does not compute the first-order translation.
+%Rather,
+For efficiency, we implemented separate KBO and
+LPO functions that compute the order directly, intertwining the encoding and
+the order computation.
+%%% @PETAR(yes): Irrelevant detail?
+%We kept the
+%implementation of $\lambda$-free KBO as \ehohii{} can be run in $\lambda$-free mode.
+
+Ehoh cells contain a \verb|binding| field that can be used to store the
+substitution for a free variable. Substitutions can then be applied by following
+the \texttt{binding} pointers, replacing each free variable with its instance.
+Thus, when Ehoh needs to perform a KBO or LPO comparison of an instantiated term,
+it needs only follow the \texttt{binding} pointers.
+In full higher-order logic, however, instantiating a variable can trigger a
+series of $\beta\eta$-reductions,
+changing the shape of the term dramatically. To pevent this, \ehohii{}
+computes the $\beta\eta$-reduced instance of the terms
+before comparing them using KBO or LPO.

### 2022-03-25 pre/pre.tex
terms

@@
 disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
-implicitly universally quantified. Note that $\llor$ denoting disjunction in
-clauses is different from the formula-building $\ior$. In general,
-formula-building logical connectives are written in boldface. In standard,
+implicitly universally quantified. Note that clause-level operators ($\llor, \eq, \noteq$) are not typeset in bold
+to distinguish them from term-building symbols ($\ior, \ieq, \ineq, \iand, \iforall, \ldots$) which are typeset in bold. In standard,
 non-clausal first-order logic, an atom is either predicate or equational. For

### 2022-03-25 satfol/satfol.tex
terms

@@
 \newcommand{\paper}[0]{chapter}
-
-\newcommand{\eqlit}[2]{\ensuremath{#1 \eq #2}}
-\newcommand{\neqlit}[2]{\ensuremath{#1 \noteq #2}}
-% \newcommand{\eqneqlit}[2]{\ensuremath{#1 \mathrel{\dot{\eq}} #2}}
-\newcommand{\pospredlit}[1]{{#1}}
-\renewcommand{\neglit}[1]{\neg{\kern.2ex #1}}
-\newcommand{\negpredlit}[1]{\neglit{#1}}
-\newcommand{\arbpredlit}[1]{(\neg)\kern.2ex\pospredlit{#1}}
 \newcommand{\cor}{\ensuremath{\mathrel{\lor}}}
@@
 (dis)equations with equality. To simplify the notation in this chapter we write
-predicate literals in unencoded form: positive literal $\eqlit{s}{\itrue}$ is
+predicate literals in unencoded form: Positive literal $\eqlit{s}{\itrue}$ is
 written as $s$ and negative literal $\neqlit{s}{\top}$ is written as

### 2022-03-25 ehoh2/ehoh2.tex
unif

@@
 before comparing them using KBO or LPO.
+
+\section{Unification, Matching, and Term Indexing}
+\label{sec:ehoh2:unif-match-index}
+
+Standard superposition crucially depends on the concept of a most general
+unifier (MGU). In higher-order logic, such a unifier does not in general exist,
+and the concept
+%%% @PETAR: CSU, infinite
+is replaced by that of a complete set of unifiers (CSU), which may be infinite.
+In Chapter \ref{ch:unif}, we described an efficient procedure to enumerate a CSU
+for a term pair. It is implemented in Zipperposition, together with some
+extensions to term indexing. In \ehohii{}, we further improve the performance of
+this procedure by implementing a terminating, incomplete variant. We also
+introduce a new indexing data structure.
+
+\subsection{The Unification Procedure} 
+% Two main guiding principles behind our
+% unification procedure are laziness and redundancy elimination. The first one is
+% embodied in the fact that the procedure makes sure that it never fully
+% normalizes the terms or fully applies the substitution if that is not necessary.
+% Redundancy elimination is achieved through \emph{oracles}
+% \cite{unif-section}: the procedures that solve the unification
+% problem for a subclass of higher-order terms on which unification is decidable,
+% and for \ehohii{}, unary. If those procedures determine that the given terms
+% belong to their class of terms, they will return the most general unifier,
+% avoiding redundant computations and in some cases even non-termination.
+The unification procedure works by maintaining a list of unification pairs to be solved.
+After choosing a pair, it first normalizes it by $\beta$-reducing and
+instantiating the heads of both terms in the pair. Then, if either head is a
+variable, it computes an appropriate binding for this variable, thereby
+approximating the solution.
+
+Unlike in first-order and $\lambda$-free higher-order unification, in full
+higher-order case there may be many bindings that lead to a solution. To reduce
+this mostly blind guessing of bindings, the procedure features support for
+\emph{oracles} (Sect.~\ref{sec:unif:the-unification-procedure}). These are
+procedures that solve the unification problem for a subclass of higher-order
+terms on which unification is decidable and, in the case of \ehohii{}, unary. Oracles help
+increase performance, avoid nontermination, and avoid redundant bindings.
+
+\looseness=-1
+In Chapter \ref{ch:unif}, the unification procedure is described as a transition system. In
+\ehohii{}, the procedure is implemented nonrecursively, and the unifiers are
+enumerated using an iterator object that encapsulates the unifier
+search state. The iterator consists of five fields:
+\begin{enumerate}
+  \item \vn{constraints}, which holds the unification
+  constraints
+  \item \vn{bt\_state}, a stack that contains information necessary to backtrack
+  to a previous state
+  \item \vn{branch\_iter}, which stores how far we
+  are in exploring different possibilities from the current search node
+  \item \vn{steps}, which remembers how many different unification bindings (such as
+  imitation, projection, and identification) are applied
+  \item \vn{subst},
+  a stack storing the variables bound so far
+\end{enumerate}
+\pagebreak[2]
+\newcommand{\cn}[1]{\ensuremath{\textsc{#1}}} %const name
+\newcommand{\fc}[1]{\ensuremath{\textsc{#1}}} %function call
+\algrenewcommand\alglinenumber[1]{{\color{gray} \tiny #1}}
+
+The iterator is initialized to hold the original problem in \vn{constraints},
+and all other fields are initially empty. The unifiers are retrieved one by one by
+calling the function \fc{ForwardIter}. It returns \fc{True} if the iterator made
+progress, in which case the unifier can be read via the iterator's
+$\vn{subst}$ field. Otherwise, no more unifiers can be found, and the iterator is
+no longer valid. The function's pseudocode is given below, including two
+auxiliary functions \fc{NormalizeHead} and \fc{BacktrackIter}:
+
+\vspace{\jot}
+\begin{algorithmic}[]
+  \Function{ForwardIter}{$\vn{iter}$}
+  \State $\vn{forward} \gets \neg \vn{iter.constraints.empty()} \lor \fc{BacktrackIter}(\vn{iter})$
+  \While{$\vn{forward} \land \neg \vn{iter.constraints.empty()}$}
+    \State $(\vn{lhs}, \vn{rhs}) \gets $ pop pair from $\vn{iter.constraints}$
+    \State $\vn{lhs} \gets \fc{NormalizeHead}(\vn{lhs})$
+    \State $\vn{rhs} \gets \fc{NormalizeHead}(\vn{rhs})$
+    \State normalize and discard the $\lambda$ prefix of $\vn{lhs}$ and $\vn{rhs}$
+
+    \If{$\neg\vn{lhs.head.is\_var()} \land \vn{rhs.head.is\_var()}$}
+      \State swap $\vn{lhs}$ and $\vn{rhs}$
+    \EndIf
+
+    \If{$\vn{lhs.head.is\_var()}$}
+      \State $\vn{oracle\_res} \gets \fc{Fixpoint}(\vn{lhs, rhs, iter.subst}) $
+      \If{$\vn{oracle\_res} = \cn{NotInFragment}$}
+        \State $\vn{oracle\_res} \gets \fc{Pattern}(\vn{lhs, rhs, iter.subst}) $
+      \EndIf
+
+      \If{$\vn{oracle\_res} = \cn{NotUnifiable}$}
+        \State $\vn{forward} \gets \fc{BacktrackIter(iter)}$
+      \ElsIf{$\vn{oracle\_res} = \cn{NotInFragment}$}
+        \State $\vn{n\_steps}, \vn{n\_branch\_iter}, \vn{n\_binding} \gets$
+        \State $\quad \fc{NextBinding}(\vn{lhs}, \vn{rhs}, \vn{iter.steps}, \vn{iter.branch\_iter})$
+
+        \If{$\vn{n\_branch\_iter} \not= \fc{BindEnd}$ }
+          \State push pair $(\vn{lhs},\vn{rhs})$ back to \vn{iter.constraints}
+          \State push quadruple $(\vn{iter.constraints}, \vn{n\_branch\_iter}, $
+          \State                  $\quad \vn{iter.steps}, \vn{iter.subst})$ onto $\vn{iter.bt\_state}$
+          \State extend $\vn{iter.subst}$ with $\vn{n\_binding}$
+          \State $\vn{iter.steps} \gets \vn{n\_steps}$
+          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
+        \ElsIf{\vn{lhs.head = rhs.head}}
+          \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+          \State $\quad$ and push them to \vn{iter.constrants}
+          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
+        \EndIf
+      \EndIf
+    \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
+      \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+      \State $\quad$ and push them to \vn{iter.constrants}
+    \Else{} \State $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
+    \EndIf
+  \EndWhile
+  \State \Return $\vn{forward}$
+  \EndFunction
+
+  \newpage
+
+  \Function{NormalizeHead}{$\vn{t}$}
+    \If{$\vn{t.head} = \internalat \land \vn{t.args}[0].\vn{is\_lambda}()$}
+      \State reduce the top-level $\beta$-redex in $t$
+      \State \Return $\fc{NormalizeHead}(t)$
+    \ElsIf{$\vn{t.head.is\_var()} \land \vn{t.head.binding} \not= \cn{nil}$}
+      \State $\vn{t.head} \gets \vn{t.head.binding}$
+      \State \Return $\fc{NormalizeHead}(t)$
+    \Else{}
+      \State \Return $t$
+    \EndIf
+  \EndFunction
+
+  \vspace{\jot}
+  \Function{BacktrackIter}{$\vn{iter}$}
+    \If{\vn{iter.bt\_state.empty()}}
+      \State clear all fields in \vn{iter}
+      \State \Return \cn{False}
+    \Else
+    \State pop $(\vn{constraints}, \vn{branch\_iter}, \vn{steps}, \vn{subst})$ from $\vn{iter.bt\_state}$
+    \State set the corresponding fields of $\vn{iter}$
+    \State \Return \cn{True}
+    \EndIf
+  \EndFunction
+  \vspace{\jot}
+
+\end{algorithmic}
+
+\looseness=-1
+\fc{ForwardIter} begins by backtracking if the previous attempt was successful (i.e., 
+all constraints were solved). If it finds a state from which it can continue,
+it takes term pairs from $\vn{constraints}$ until
+there are no more constraints or it is determined that no unifier exists. The terms
+are normalized by instantiating the head variable with its binding and then
+reducing the potential top-level $\beta$-redex that appears. This instantiation
+and reduction process is repeated
+until there are no more top-level $\beta$-redexes and the head is
+not a variable bound to some term. Then the term with shorter
+$\lambda$ prefix is expanded (only on the top-level) so that both have the
+same length of $\lambda$ prefix. Finally, the $\lambda$ prefix is ignored, and we
+focus only on the body. In this way, we avoid fully substituting
+and normalizing terms and perform just enough operations
+to determine the next step of the procedure.
+
+If either term of the constraint is flex, we first invoke oracles to solve the
+constraint. \ehohii{} implements the most efficient oracles implemented in
+Zipperposition:\ fixpoint and pattern oracle (Sect.~\ref{sec:unif:implementation}). An oracle can return three results:
+(1)~there is an MGU for the pair (\cn{Unifiable}), which is recorded in
+\emph{subst}, and the next pair in $\emph{constraints}$ is tried;
+(2)~no MGU exists
+for the pair (\cn{NotUnifiable}), which causes the iterator to backtrack;
+% to previous state;
+(3)~if the pairs do not belong to the subclass that oracle
+can solve (\cn{NotInFragment}), we generate possible variable bindings---that is,
+we guess the approximate form of the solution.
+
+\ehohii{} has a special module that generates bindings (\cn{NextBinding}). This
+module is given the current constraint and the values of \emph{branch\_iter} and
+\emph{steps}, and it either returns the next binding and the new values of
+\emph{branch\_iter} and \emph{steps} or reports that all different variable
+bindings are exhausted. The bindings that \ehohii{}'s unification procedure creates
+are imitation, Huet-style projection, identification, and elimination (one
+argument at a time) (Sect.~\ref{sec:unif:the-unification-procedure}). A limit on the
+total number of applied binding rules can be set, as well as a limit on the
+number of individual rule applications. The binding module checks whether limits
+are reached using the iterator's \vn{steps} field.
+
+Computing bindings is the only point in the procedure where the search tree
+branches and different possibilities are explored. Thus, when \ehohii{} follows
+the branch indicated by the binding module, it records the state to which it
+needs to return, should the followed branch be backtracked. Therese are values
+of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
+and the value of the \vn{branch\_iter} that points past the followed branch. The
+values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
+binding was created, intermediate values that \cn{NextBinding} uses to remember
+how far through bindings it is, and \cn{BindEnd}, which indicates that all
+bindings are exhausted.
+
+If all bindings are exhausted, the procedure checks whether the pair is
+flex--flex and both sides have the same head. If so, the pair is decomposed and
+constraints are derived from the arguments of the pair. Otherwise, the iterator
+backtracks.
+% @PETAR(actually the procedure applies bindings to same-head flex-flex pairs,
+% so the text is OK -- we do get "stuck" enumerating those flex-flex pairs. The
+% limits help here): Maybe explain here that you don't want a pragmatic prover
+% like lambdaE to get stuck in wildly explosive flex-flex pairs? --JB
+%
+If the constraint is rigid--rigid, for unification to succeed the heads of both
+sides of the constraint must be the same. Unification then continues
+with new constraints derived from the arguments. Otherwise, the iterator must be
+backtracked.
+
+\ourpara{Matching} In Ehoh, the matching algorithm is mostly used inside
+simplification rules such as demodulation and subsumption \cite{ss-02-brainiac}.
+As these rules must be efficiently performed, using any complex matching
+algorithm is not a viable option. Instead, we implemented a matching algorithm
+for the pattern class of terms \cite{tn-93-patterns} to complement Ehoh's
+$\lambda$-free higher-order matching algorithm \cite[Sect.~4]{section-ehoh}.
+A term is a \emph{pattern} if each of its free variables either
+has no arguments (as in first-order logic) or is applied to distinct De Bruijn
+indices.
+
+To determine which of the two algorithms to call (pattern or $\lambda$-free), we
+added a cached property \texttt{HasNonPatternVar}, which is set for terms of the
+form $X \,
+\overline{s}_n$ where $n>0$ and either there exists some $s_i$ that is not a De Bruijn
+index or there exist indices $i < j$ such that $s_i = s_j$ is
+a De Bruijn index. This property is propagated to the superterms when they are
+perfectly shared. This allows later checks if a term belongs to the pattern
+class to be performed in constant time.
+
+% When we need to determine if one term matches the other one we first check if
+% any of them has a non-pattern applied variables. If so, $\lambda$-free
+% higher-order matching is tried. This algorithm is adjusted to treat $\lambda$ prefixes
+% as above in unification algorithm (ensuring that $\beta$-redexes can never occur)
+% and that free variables are never bound to terms that have loose bound variables.
+% Else if the terms have only pattern variables,
+% pattern matching algorithm is tried. Note that in this sense, first-order order
+% terms behave as pattern terms and thus pattern matching algorithm is run on
+% them. For example, on terms $(X \, \cst{b}, \cst{f} \, \cst{a} \, \cst{b})$
+% $\lambda$-free matching algorithm is invoked; on $(\cst{f} \, (\lambda x. \,
+% \lambda y.\, X \, x), \cst{f} \, \cst{g})$ pattern matching is tried.
+
+We have modified the $\lambda$-free higher-order matching algorithm to treat
+$\lambda$ prefixes as above in unification procedure---by bringing the prefixes
+to the same length and ignoring them afterwards. This ensures that the
+algorithm will never try to match a free variable to a $\lambda$-abstraction,
+making sure that $\beta$-redexes never appear (as in original $\lambda$-free
+higher-order matching). We also modified the algorithm to ensure that free variables
+are never bound to terms that have loose bound variables. This algorithm cannot
+find many complex matching substitutions (matchers), but it can
+efficiently determine whether two terms are variable renamings of each other
+or whether a simple matcher can be used as in the case of $(X \, (\lambda x. \,x) \,
+\cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
+usually the desired matcher. If this algorithm does not find a matcher and both
+terms are patterns, pattern matching is tried.
+
+\ourpara{Indexing} Ehoh, like other modern theorem provers, efficiently retrieves
+unifiable or matchable pairs of terms using indexing data structures. To find
+terms unifiable with a query term or instances of a query term it uses
+\emph{fingerprint indexing} \cite{ss-12-fp-indexing}. We extended this data
+structure to support (full) higher-order terms in Zipperposition
+\cite[Sect.~6]{unif-section}. We used the same approach in \ehohii{},
+and we extended feature vector indices \cite{ss-2013-feature-vector}
+in the same way.
+
+Ehoh uses \emph{perfect discrimination trees} \cite{mcc-92-pdts} to find
+generalizations of the query term (i.e., terms of which the query term is an instance).
+This data structure is a trie
+that indexes terms by representing them in a serialized, flattened form.
+%\begin{rep}, where each node corresponds to a symbol in a term\end{rep}.
+The left branch from the root in Figure \ref{fig:pdt} shows how the first-order terms
+$\cst{f}\, \cst{a} \, X$ and $\cst{f}\, \cst{a} \, \cst{a}$ are stored.
+In Ehoh, this data structure is extended to support partial application
+and applied variables \cite{section-ehoh}.
+
+\begin{figure}[tb]
+\centering
+\begin{tikzpicture}[level distance=3em, sibling distance=7.5em,
+  %  edge from parent path={(\tikzparentnode.south) -- ++(0,-0.5em)
+  %   -| (\tikzchildnode.north)},
+    every node/.style = {shape=rectangle, rounded corners=.75em, minimum size=1.75em, draw, align=center}]
+    \node {\phantom{x}}
+      child { node {$\cst{f}$}
+        child[sibling distance=2.5em] { node %[fill=verylightgray]
+          {$\cst{a}$}
+          child {node {$X$}}
+          child {node {$\cst{a}$}} }
+      }
+      child { node {$\cst{g}$}
+        child {node {$X \, \cst{a} \, \cst{b}$}
+               child {node {$\cst{c}$}}}
+      }
+      child { node {$\cst{h}$}
+        child {node {$\internallam_\iota$}
+          child {node {$\internallam_\iota$}
+            child {node {$X \, \db{0} \, \db{1}$}}
+          }
+      }}
+      ;
+\end{tikzpicture}
+\caption{First-order, $\lambda$-free higher-order, and higher-order pattern
+  terms in a perfect discrimination tree}
+\label{fig:pdt}
+\end{figure}
+
+In \ehohii{}, we extended this structure to support $\lambda$-abstractions and
+the higher-order pattern matching algorithm.
+To this end, we changed the way in which terms are
+serialized. First, we require that all terms are fully $\eta$-expanded (except
+for arguments of variables applied in patterns). Then, when the term is serialized,
+we dedicate a separate node for applied variable terms $X \, \overline{s}_n$,
+instead of the node for $X$ followed by nodes for serialization of arguments
+$\overline{s}_n$. We serialize $\lambda$-abstraction $\lambda x.\, s$ using
+a special node $\internallam_\tau$, where $\tau$ is the type of $x$,
+followed by the serialization of $s$. Other than these two changes, serialization
+remains as in Ehoh, following the gracefulness principle.
+Figure \ref{fig:pdt} shows how $\cst{g} \,
+(X\,\cst{a}\,\cst{b}) \, \cst{c}$ and $\cst{h} \, (\lambda x. \, \lambda y. \, X
+\, y \, x)$ are serialized.
+
+Since the terms are stored in serialized form, it is hard to manipulate
+$\lambda$ prefixes of stored terms during matching. Performing $\eta$-expansion
+when serializing terms makes sure that matchable terms have
+$\lambda$ prefixes of the same length.
+
+We have dedicated separate nodes for applied variables because access to arguments
+of applied variable is necessary for the pattern matching algorithm. Even though
+arguments can be obtained by querying the arity $n$ of the variable and taking
+next $n$ arguments in the serialization, this is both inefficient and inelegant.
+As for De Bruijn indices, we treat them the same as function symbols and assign
+them their own nodes.
+
+\newcommand{\tstack}{\ensuremath{\texttt{term\_stack}}}
+\newcommand{\tproc}{\ensuremath{\texttt{term\_proc}}}
+
+Following the notation from the extension of perfect discrimination trees to
+$\lambda$-free higher-order logic \cite{section-ehoh}, we now describe how enumeration of
+generalizations is performed. To traverse the tree, \ehohii{} begins at the root
+node and maintains two stacks:\ \tstack{} and \tproc{}, where \tstack{} contains the
+subterms of the query term that have to be matched, and \tproc{} contains
+processed terms that are used to backtrack to previous states. Initially,
+\tstack{} contains the query term, the current matching substitution
+$\sigma$ is empty, and the successor node is chosen among the child nodes as
+follows:
+
+\begin{enumerate}
+\item[A.] If the node is labeled with a symbol $\xi$ (where
+  $\xi$ is either a De Bruijn index or a constant) and the top item $t$
+  of \tstack{} is of the form $\xi \, \overline{t}_n$, replace $t$ by
+  $n$~new items $t_1,\dots,t_n$, and push $t$ onto \tproc{}.
+\smallskip
+\item[B.] If the node is labeled with a symbol $\internallam_\tau$ and the top
+  item $t$ of \tstack{} is of the form $\lambda x.\, s$ and the type of $x$ is
+  $\tau$, replace $t$ by $s$, and push $t$ onto \tproc{}.
+\smallskip
+\item[C.] If the node is labeled with a possibly applied variable $X \, \overline{s}_n$
+  (where $n \geq 0$), and the top item of \tstack{} is $t$, the 
+  matching algorithm described above is run on $X \, \overline{s}_n$ and $t$.
+  The algorithm takes into account $\sigma$ built so far and extends it
+  if necessary. If the algorithm succeeds, pop $t$ from \tstack{}, push it onto \tproc{},
+  and save the original value of $\sigma$ in the node.
+\end{enumerate}
+
+Backtracking works in the opposite direction: If the current node is labeled with a De
+Bruijn index or function symbol node of arity $n$, pop $n$ terms from \tstack{} and move
+top of \tproc{} to \tstack{}. If the node is labeled with
+$\internallam_\tau$, pop the top of \tstack{} and move the top of \tproc{} to
+\tstack{}. Finally, if the node is labeled with a possibly applied variable, move top of the
+\tproc{} to \tstack{} and restore the value of $\sigma$.
+
+As an example of how finding a generalization works, consider the following
+states of stacks and substitutions, which emerge when looking for generalizations of 
+$\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}$ in the tree of Figure \ref{fig:pdt}:
+%
+\[\begin{array}{@{}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{\kern1em}l@{}}
+             & \epsilon                                                & \cst{g}                                                 & \cst{g}.(X \, \cst{a} \, \cst{b})                                                          & \cst{g}.(X \, \cst{a} \, \cst{b}).\cst{c} \\[.5\jot]
+  \sigma{:}  & \emptyset                                               & \emptyset                                               & \{X \mapsto \cst{f}\}                                                                      & \{X \mapsto \cst{f}\} \\
+  \tstack{:} & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}, \cst{c}]                & [\cst{c}]                                                                                  & [] \\
+  \tproc{:}  & []                                                      & [\cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}] & [\cst{c}{,}\;\cst{f} \, \cst{a} \, \cst{b}{,}\; \cst{g} \, (\cst{f} \, \cst{a} \, \cst{b}) \, \cst{c}]
+  \end{array}\]

### 2022-03-25 ehoh2/ehoh2.tex
indexing

@@
 the branch indicated by the binding module, it records the state to which it
-needs to return, should the followed branch be backtracked. Therese are values
+needs to return, should the followed branch be backtracked. It saves the values
 of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
@@
 
-\ourpara{Matching} In Ehoh, the matching algorithm is mostly used inside
+\subsection{Matching}
+
+In E, the matching algorithm is mostly used inside
 simplification rules such as demodulation and subsumption \cite{ss-02-brainiac}.
@@
 for the pattern class of terms \cite{tn-93-patterns} to complement Ehoh's
-$\lambda$-free higher-order matching algorithm \cite[Sect.~4]{section-ehoh}.
+$\lambda$-free higher-order matching algorithm (Sect.~\ref{sec:ehoh:unif-match}).
 A term is a \emph{pattern} if each of its free variables either
@@
 
-\ourpara{Indexing} Ehoh, like other modern theorem provers, efficiently retrieves
+\subsection{Indexing} E, like other modern theorem provers, efficiently retrieves
 unifiable or matchable pairs of terms using indexing data structures. To find
@@
 \emph{fingerprint indexing} \cite{ss-12-fp-indexing}. We extended this data
-structure to support (full) higher-order terms in Zipperposition
-\cite[Sect.~6]{unif-section}. We used the same approach in \ehohii{},
+structure to support (full) higher-order terms in Zipperposition (Sect.~\ref{sec:unif:indexing}). 
+We used the same approach in \ehohii{},
 and we extended feature vector indices \cite{ss-2013-feature-vector}
@@
 
-Ehoh uses \emph{perfect discrimination trees} \cite{mcc-92-pdts} to find
+E uses \emph{perfect discrimination trees} \cite{mcc-92-pdts} to find
 generalizations of the query term (i.e., terms of which the query term is an instance).
@@
 In Ehoh, this data structure is extended to support partial application
-and applied variables \cite{section-ehoh}.
+and applied variables (Sect.~\ref{sec:ehoh:indexing}).
 
@@
 
+\looseness=-1
 Following the notation from the extension of perfect discrimination trees to
-$\lambda$-free higher-order logic \cite{section-ehoh}, we now describe how enumeration of
+$\lambda$-free higher-order logic (Sect. \ref{sec:ehoh:indexing}), we now describe how enumeration of
 generalizations is performed. To traverse the tree, \ehohii{} begins at the root

### 2022-03-25 ehoh2/ehoh2.tex
eval

@@
   \end{array}\]
+
+  \section{Preprocessing, Calculus, and Extensions}
+  \label{sec:ehoh2:calculus}
+
+  Ehoh's simple $\lambda$-free higher-order calculus performed well on
+  Sledgehammer problems and formed a promising stepping stone to full higher-order
+  logic (Sect.~\ref{sec:ehoh:evaluation}). When implementing support for full higher-order
+  logic, we were guided by efficiency and gracefulness with respect to Ehoh's calculus rather
+  than completeness. Whereas Zipperposition provides both complete and incomplete
+  modes, \ehohii{} only offers incomplete modes.
+
+  \ourpara{Preprocessing} Our experience with Zipperposition showed the importance
+  of flexibility in preprocessing the higher-order problems
+  (Sect.~\ref{sec:ho-tech:preprocessing}). Therefore, we implemented a flexible
+  preprocessing module in \ehohii{}.
+
+  To maintain compatibility with Ehoh, \ehohii{} can optionally transform all
+  $\lambda$-abstractions into named functions. This process is called
+  \emph{$\lambda$-lifting} \cite{rjmh-82-lifting}. \ehohii{} also removes all
+  occurrences of Boolean subterms (other than $\itrue, \ifalse$, and free variables)
+  in higher-order contexts using a FOOL-like transformation \cite{kotelnikov-16-fool}.
+  For example, the formula $\cst{f}(\cst{p} \iand \cst{q}) \ieq \cst{a}$ becomes
+  $(\cst{p} \iand \cst{q} \iimplies \cst{f}(\itrue) \ieq \cst{a}) \iand
+  (\inot(\cst{p} \iand \cst{q}) \iimplies \cst{f}(\ifalse) \ieq \cst{a})$.
+
+  \looseness=-1
+  Many TPTP problems use the \texttt{definition} role to denote the definitions of
+  symbols. \ehohii{} can treat the definition axioms as rewrite rules, and replace
+  all occurrences of defined symbols during preprocessing. Furthermore, during
+  SInE \cite{hv-2011-sine} axiom selection, it can always
+  include the defined symbol in the trigger relation.
+
+  \ourpara{Calculus} \ehohii{} implements the same superposition
+  calculus as Ehoh with three important changes. First, wherever Ehoh requires
+  the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
+  Sect.~\ref{sec:ehoh2:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
+  LPO orders designed for $\lambda$-terms.
+  % as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.
+
+  %TODO: @JASMIN: How do we cite the STTT article?
+  The third difference is more subtle. One of the main features of Ehoh is
+  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): A method which, given a demodulator $s \eq t$, allows
+  to replace both applied and unapplied occurrences of $s$ by $t$ 
+  by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
+  optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
+  terms can change drastically, making it much harder to
+  track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
+  To counterbalance this removal, we introduced
+  the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
+  in \ehohii{}
+  and enabled positive and negative functional extensionality (\infname{PE} and
+  \infname{NE}) by default:
+
+  \vskip\abovedisplayskip
+
+  \noindent\hbox{}\hfill
+    \namedinference{AC}{s \eq t \llor C}{s \, X \eq t \, X \llor C}%
+  \hfill
+    \namedinference{NE}{s \noteq t \llor C}{s \; (\cst{sk} \; \overline{X}) \noteq t \; (\cst{sk} \; \overline{X}) \llor C}%
+  \hfill
+    \namedinference{PE}{s \; X \eq t \; X \llor C}{s \eq t \llor C}%
+  \hfill\hbox{}
+
+  \vskip\belowdisplayskip
+
+  \noindent
+  \infname{AC} and \infname{NE} assume that $s$ and $t$ are of function
+  type. In \infname{NE}, $\overline{X}$ denote all the free variables occurring in $s$ and
+  $t$, and $\cst{sk}$ is a fresh Skolem symbol of the appropriate type. \infname{PE}
+  has a side condition that $X$ may not appear in $s$, $t$, or $C$.
+
+  \ourpara{Saturation} E's saturation procedure is based on the assumption that
+  each attempt to perform an inference will either result in a single clause or
+  fail due to one of the inference side conditions. Unification procedures that
+  produce multiple substitutions break this invariant, and the saturation
+  procedure needed to be adjusted.
+
+  In Sect.~\ref{sec:ho-tech:infinite-branching}, we described a variant of the saturation
+  procedure that elegantly supports interleaving computation of unifiers and
+  scheduling
+  inferences to be performed. Since completeness was
+  not one of the \ehohii{} design goals, we did not implement this version of the
+  saturation procedure. Instead, in places where previously a single unifier was
+  expected, \ehohii{} consumes all elements of the iterator used for enumerating a
+  unifier, converting them to clauses.
+
+  \ourpara{Reasoning about Formulas} Even though most of the Boolean structure is
+  removed during preprocessing, formulas can reappear at the top level of
+  clauses during saturation. For example, after replacing $X$ with $\lambda x.
+  \, \lambda y.\, x \iand y$, the clause $X \, \cst{p} \, \cst{q} \llor \cst{a}
+  \eq \cst{b}$ becomes $(\cst{p} \iand \cst{q}) \llor \cst{a} \eq \cst{b}$. \ehohii{}
+  converts every clause of the form $\varphi \llor C$, where $\varphi$ has a logic symbol
+  as its head (or it is a (dis)equation between two formulas different than $\itrue$),
+  to an explicitly quantified formula. Then, the clausification algorithm
+  is invoked on the formula to restore the clausal structure. Zipperposition features
+  more dynamic clausification modes, but for simplicity we decided
+  not to implement them in \ehohii{}.
+
+  The \osup{} calculus
+  \cite{bbtv-21-full-ho-sup} includes many rules that act on Boolean subterms,
+  which are necessary for completeness. Other than Boolean simplification rules,
+  which use simple tautologies such as $\cst{p} \iand \itrue \iequiv\cst{p}$ to simplify terms, we
+  have implemented none of the Boolean rules of this calculus in \ehohii{}. First, we have
+  observed that complicated rules such as \infname{FluidBoolHoist} and
+  \infname{FluidLoobHoist} are hardly ever useful in practice and usually only
+  contribute to an uncontrolled increase in the proof state size. Second, simpler
+  rules such as \infname{BoolHoist} can usually be simulated by pragmatic rules
+  that perform Boolean and functional extensionality reasoning, described below.
+
+  To make up for excluding Boolean rules, we use an incomplete, but more
+  easily controllable and intuitive rule, called \emph{primitive instantiation}. This
+  rule instantiates free predicate variables with approximations of formulas
+  that are ground instances of this variable. We described this rule in Sect.~\ref{sect:bool:native}.
+  In \ehohii{} it is implemented in a similar manner.
+
+  \ehohii{}'s handling of the Hilbert choice operator is inspired by Leo-III's
+  \cite{sb-21-leo3}. \ehohii{}
+  recognizes clauses of the form $\neg P \, X \lor P \, (\cst{f} \, P)$ which
+  essentially denote that $\cst{f}$ is a choice symbol. Then, when subterm
+  $\cst{f} \, s$ is found during saturation, $s$ is used to instantiate the choice
+  axiom for $\cst{f}$. 
+  %
+  Similarly, Leibniz equality is eliminated by
+  recognizing clauses of the form  $\neg P \, \cst{a} \llor P \, \cst{b} \llor C$. These clauses are then
+  instantiated with $P \mapsto \lambda x. \, x \eq \cst{a}$ and $P \mapsto \lambda
+  x. \, x\noteq\cst{b}$, which results in $\cst{a} \eq \cst{b} \lor C$. Both rules
+  are described in Sect.~\ref{sect:bool:native} in more detail.
+
+  Finally, \ehohii{} treats induction axioms specially. Just like
+  Zipperposition (Sect.~\ref{sec:ho-tech:formulas}), it abstracts literals
+  from the goal clauses and uses these abstractions to instantiate induction
+  axioms. Since Zipperposition supports dynamic calculus-level clausification,
+  instantiation of induction axioms happens during saturation, when these axioms
+  are processed. In \ehohii{}, this instantiation is performed statically,
+  immediately after clausification. After \ehohii{} collects all the abstractions,
+  it traverses the clauses and instantiates those that have applied variable of
+  the same type as the abstraction.
+
+  \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reason about
+  functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
+  which simulate basic superposition calculus rules,
+  but do not require unifiability of the partner terms in the inference. More
+  precisely, assume a core inference needs to be performed between two
+  $\beta$-reduced terms $u$ and $v$, such that they can be represented as
+  $u=C[s_1, \ldots, s_n]$ and $v=C[t_1, \ldots, t_n]$, where $C$ is the most general
+  common context (more precisely, green common context that does not go into the structure of $\lambda$-abstraction or applied variables \cite{bbtvw-21-sup-lam}) of $u$ and $v$, not all of
+  $s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
+  $s_i$ and $t_i$ are not (applied) free variables, and they are of Boolean or
+  function type. Then, the conclusion is formed by taking the conclusion $D$ of
+  the core inference rule (which would be created if $s$ and $t$ are unifiable)
+  and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.
+
+  These rules are particularly useful because \ehohii{} has no rules that dynamically
+  process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
+  given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
+  $\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
+  superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
+  \iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
+  the top level and is further processed by clausification. We noticed that this
+  alleviates the need for the other Boolean rules in practice.
+
+  \section{Evaluation}
+  \label{sec:ehoh2:eval}
+
+  In this section, we try to answer two questions about \ehohii{}: \emph{How does
+  \ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
+  \ehohii{} introduce any overhead compared with Ehoh?} To answer
+  these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
+  benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
+  The experiments were carried out on StarExec Miami
+  \cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
+  2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
+  wallclock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
+  default time limit:\ 30~s wallclock and CPU.
+
+  \ourpara{Comparison with Other Provers}
+  To answer the first question, we let
+  \ehohii{} compete with the top contenders in the higher-order division of
+  CASC-28:\ cvc5 \NumberOK{0.0.7},\footnote{\url{https://cvc5.github.io/}}
+  Ehoh 2.7, Leo-III \NumberOK{1.6.6} \cite{sb-21-leo3}, Vampire 4.6
+  \cite{br-20-full-sup-w-combs}, and Zipperposition 2.1 (Chapter~\ref{ch:ho-techniques}). 
+  We also include Satallax 3.5
+  \cite{cb-12-satallax}. We used all \NumberOK{2899} higher-order
+  theorems in TPTP 7.5.0 as well as 5000 SH higher-order benchmarks
+  originating from the Seventeen benchmark suite \cite{desharnais-et-al-202x}.
+  On SH benchmarks,
+  cvc5, Ehoh, \ehohii{}, Vampire, and Zipperposition were run using custom schedules provided by
+  their developers, optimized for single-core usage and low timeouts.
+  Otherwise, we used the corresponding CASC configurations.
+  Note that Ehoh 2.7 is an updated version of Ehoh described in Chapter
+  \ref{ch:ehoh}, which can parse not only $\lambda$-free but also full
+  higher-order. It is only a syntactic extension as $\lambda$-abstractions are supported using $\lambda$-lifting \cite{rjmh-82-lifting}.
+  We included two versions of Zipperposition:\ \emph{coop} uses Ehoh 2.7 as a backend to finish proof attempts,
+  whereas \emph{uncoop} does not use this feature. The results
+  are shown in Figure \ref{fig:all-provers}.
+
+  Both Ehoh and \ehohii{} were run in the automatic scheduling mode. Compared to Ehoh,
+  \ehohii{} features a redesigned module for automatic scheduling, it
+  can use multiple CPU cores, and its heuristics have been trained better on higher-order problems.
+
+  \ehohii{} dramatically improves the higher-order reasoning capabilities
+  compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
+  and \NumberOK{6\%} more problems on SH benchmarks, where Ehoh was already very
+  successful.
+
+  \ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
+  excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
+  If Zipperposition's Ehoh backend is disabled,
+  \ehohii{} outperforms Zipperposition by a wide margin.
+  This comparison is arguably fairer; after all,
+  \ehohii{} does not use an older version of Zipperposition
+  as a backend.
+  %
+  These results suggest that \ehohii{} already implements most of the necessary features
+  for a high-performance higher-order prover
+  but could benefit from the kind of fine-tuning that
+  Zipperposition underwent in the last two years.
+
+  Remarkably, there is a substantial disparity in the set of problems solved by
+  \ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
+  \NumberOK{166} SH problems and \NumberOK{21} TPTP problems that Zipperposition-coop
+  does not. The lower number of uniquely solved TPTP problems is most likely 
+  due to Zipperposition's being heavily optimized on TPTP.
+
+  \ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
+  in a mode that disables most of the higher-order reasoning. This mode is
+  designed for users that are interested only in E's first-order capabilities and
+  care a lot about performance.
+  To answer the second
+  evaluation question, about assessing overhead of \ehohii{},
+  we chose all the \NumberOK{1138} unique problems used at CASC
+  from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
+  both in this first-order (FO) mode and in higher-order (HO) mode.
+
+  We fixed a single
+  configuration of options, because Ehoh and \ehohii{}'s
+  automatic scheduling methods could select different configurations
+  and we would not be measuring the overhead but the quality of the chosen
+  configurations. We chose the \emph{boa} configuration (Sect.~\ref{sec:ehoh:heuristics}),
+  which is the configuration that E~2.2 used the most often in its automatic scheduling mode.
+  The results are shown in Figure~\ref{fig:ehoh}.
+
+  Counterintuitively, the higher-order versions of both provers outperform
+  the first-order counterparts. However, the differerence is so small that it can
+  be attributed to the changes to memory layout that affect the order in which
+  clauses are chosen. Similar effects are seen when comparing the first-order versions.
+  We would expected the less heavily modified version of E, Ehoh, to perform better,
+  but due to subtle effects \ehohii{} wins.
+
+  \begin{figure}[t]
+    \noindent\hbox{}\hfill
+    \begin{minipage}[ht!]{.66\linewidth}
+      \centering
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c@{\hskip 1em}c} \toprule
+
+                              & TPTP       & SH        \\
+        \midrule
+        cvc5                  & 1931       & 2577      \\
+        Ehoh                  & 2105       & 2611      \\
+        \ehohii{}             & 2523       & {\bf 2766}\\
+        Leo-III               & 2282       & 1601      \\
+        Satallax              & 2320       & 1719      \\
+        Vampire               & 2203       & 2240      \\
+        Zipperposition-coop   & {\bf 2583} & 2754      \\
+        Zipperposition-uncoop & 2483       & 2181      \\
+        \bottomrule
+      \end{tabular}}
+      \caption{Comparison of higher-order provers}
+      \label{fig:all-provers}
+    \end{minipage}\hfill\hfill
+    \begin{minipage}[ht!]{.30\linewidth}
+      \centering
+      \def\arraystretch{1.1}%
+      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
+                        & TPTP     \\
+        \midrule
+        Ehoh -- FO      & 535      \\
+        Ehoh -- HO      & 538      \\
+        \ehohii{} -- FO & {\bf 539}\\
+        \ehohii{} -- HO & {\bf 539}\\
+        \bottomrule
+      \end{tabular}}
+      \caption{Evaluation of \ehohii{}'s overhead}
+      \label{fig:ehoh}
+    \end{minipage}\hfill\hfill
+  \end{figure}

### 2022-03-25 scratch/paper.tex
unif
eval

@@
 \label{sec:ehoh2:unif-match-index}
-% \begin{itemize}
-%   \item Procedure that implements a variant of the algo described in Efficient
-%   HO Unif paper.
-%   \item Two oracles -- fixpoint and pattern unif (solid not implemented as
-%   complexity/gain ratio is quite negative)
-%   \item It is implemented iteratively, with subtle optimizations
-%   \item Its iterator is completely contained -- iterators can be stored and
-%   later waked up similarly to what we do in Zipperposition.
-%   \item Mention that so far the prover fares extremely well with complicated
-%   unification problems and that we do not consider implementing more complex
-%   given clause algo in Ehoh.
-%   \item Matching works only with algorithms that return a single matcher:
-%   pattern matching and LFHOL matching : main optimization from E is preserved.
-%   PDT algorithm also works with both kinds of terms LFHOL and pattern. Check if
-%   something is a pattern is efficient!
-%   \item Fingerprints and feature vectors as in Zipperposition.
-%   \item To avoid huge refactorings iterator is used even if the problem to be
-%   solved is first-order.  
-% \end{itemize}
 
@@
 \section{Preprocessing, Calculus, and Extensions}
-
 \label{sec:ehoh2:calculus}
-% \begin{itemize}  
-  % \item The main goal is simplicity, efficiency and maximal compatibility with
-  %     FO superposition.
-  % \item More or less the same as LFHOL sup, but now computes CSUs instead
-  %     of LFHOL MGUs (and (potentially) uses new orders).
-  % \item Out of Boolean rules we have Boolean simplification and black-box clausification.
-  %     Most importantly -- we have no Boolean hoisting rules : everything is dealt with
-  %     ext-sup, and FOOL clausification! It simply works as good as Boolean hoisting.
-  % \item IMPORTANT: Prefix optimization is disabled, now we rely on ArgCong and NegExt.
-  % \item Many extensions:  ExtSup (With the same indexing), leibniz elimination, predicate enumeration,
-  %     FlexElim, Choice instantiation, recognition of induction.
-  % \end{itemize}
 
@@
 the top level and is further processed by clausification. We noticed that this
-aleviates the need for the other Boolean rules in practice.
+alleviates the need for the other Boolean rules in practice.
 
@@
 \ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
-\NumberNOK{166} SH problems and \NumberNOK{21} TPTP problems that Zipperposition-coop
+\NumberOK{166} SH problems and \NumberOK{21} TPTP problems that Zipperposition-coop
 does not. The lower number of uniquely solved TPTP problems is most likely 
@@
 evaluation question, about assessing overhead of \ehohii{},
-we chose all the \NumberNOK{1138} unique problems used at CASC
+we chose all the \NumberOK{1138} unique problems used at CASC
 from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}

### 2022-03-25 ehoh2/ehoh2.tex
more

@@
   \end{figure}
+
+  \section{Discussion and Related Work} %---0.5p
+  \label{sec:ehoh2:discussion}
+
+  \looseness=-1
+  On the trajectory to \ehohii{}, we developed, together with colleagues, three
+  superposition calculi:\ \lfsup{} for $\lambda$-free higher-order logic
+  \cite{bbcw-21-lfho}, \lsup{} for a higher-order logic with
+  $\lambda$-abstra\-ction but no Booleans \cite{bbtvw-21-sup-lam}, and \osup{}
+  for full higher-order logic \cite{bbtvw-21-sup-lam}. These milestones allowed
+  us to carefully estimate how the increased reasoning capabilities of each
+  calculus influence its performance. We also developed a complete and efficient
+  unification algorithm (Chapter \ref{ch:unif}) which can easily be customized
+  to trade bits of its completeness for efficiency. Practically oriented work
+  described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques},
+  helped us gauge how much reasoning power we lose by using more intuitive and
+  more easily controllable incomplete rules. Both the theoretical and the
+  practical lines of our work met in this chapter, resulting in a prover that
+  outperforms the competition on the benchmarks from proof assistants.
+
+  Extending first-order provers with higher-order reasoning capabilities has been
+  attempted by other researchers as well.
+  Barbosa et al.\ extended
+  the SMT solvers CVC4 (now cvc5) and veriT to higher-order logic in an incomplete
+  way \cite{brotb-19-ho-smt}.
+  Bhayat and Reger first extended Vampire to
+  higher-order logic using combinatory unification \cite{br-19-restricted-unif},
+  an incomplete approach, before they designed and implemented a complete
+  higher-order superposition calculus based on \textsf{SKBCI}-com\-binators \cite{br-20-full-sup-w-combs}.
+  The advantage is that combinators can be supported as a thin layer on top of
+  $\lambda$-free terms.
+  This calculus is also implemented in Zipperposition.
+  However, in unpublished experiments, we found that $\lambda$-superposition
+  performs substantially better, corroborating the CASC results, so we decided to
+  make a more profound change to Ehoh and implement $\lambda$-superposition.
+  \pagebreak[2]
+
+  Possibly the only actively maintained higher-order prover built from the bottom
+  up as a higher-order prover is Leo-III \cite{sb-21-leo3}, as the long-time
+  winner of CASC higher-order division, Satallax \cite{cb-12-satallax}, is no
+  longer maintained. We give a further overview of other traditional higher-order
+  provers and the calculi they are based in Sect.~\ref{sec:ehoh:discussion-and-related-work}.
+
+  \section{Conclusion} %---0.5p
+  \label{sec:ehoh2:conclusion}
+
+  In 2019, the reviewers of our paper introducing Ehoh \cite{vbss-19-ehoh1} were
+  skeptical that extending it with support for full higher-order logic would be
+  feasible. One of them wrote:
+  %
+  \begin{quote}
+  A potential criticism could be that this step from E to Ehoh is just extending
+  FOL by those aspects of HOL that are easily in reach with rather straightforward
+  extensions (none of the extensions is indeed very complicated), and that the
+  difficult challenges of fully supporting HOL have yet to be confronted.
+  \end{quote}
+  %
+  We ended up addressing the theoretical ``difficult challenges'' in other work
+  with colleagues. In this chapter, we faced the practical challenges pertaining to the extension
+  of Ehoh's data structures and algorithms to support full higher-order logic and
+  demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
+  that this extension makes \ehohii{} the best higher-order prover on
+  benchmarks coming from interactive theorem proving practice, which has been our goal since 2016. \ehohii{}'s
+  performance lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
+  does not assume a clausal structure and can perform subtle formula-level
+  inferences. In the future, we plan to implement the same features in \ehohii{}.
+  We have also only started tuning \ehohii{}'s heuristics on higher-order
+  problems.

### 2022-03-26 ehoh2/ehoh2.tex
more

@@
   The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
-  proof assistant, and the second best on TPTP benchmarks.
+  proof assistant, and the \NumberOK{second strongest} on TPTP benchmarks.
     %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
@@
 (Sect.~\ref{sec:ehoh2:eval}). We found
-that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on
-proof assistant benchmarks; on TPTP benchmarks it ended up second only 
+that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on the
+proof assistant benchmarks; on the TPTP benchmarks it ended up second only 
 to the cooperative version of Zipperposition, which employs Ehoh as a
@@
 %
-We introduce the property \texttt{HasBetaReducibleSubterm} which is set if
+We introduce the property \texttt{HasBetaReducibleSubterm}, which is set if
 %%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
@@
 $u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
-obtained by shifting the loose De Bruijn indices right by~$k$.
+obtained by shifting the loose De Bruijn indices down by~$k$.
 
@@
 the branch indicated by the binding module, it records the state to which it
-needs to return, should the followed branch be backtracked. It saves the values
+needs to return, should the followed branch be backtracked. The state consists of
+the values
 of $\vn{constraints}, \vn{steps}$, and $\vn{subst}$ before the branch is followed
-and the value of the \vn{branch\_iter} that points past the followed branch. The
+and the value of \vn{branch\_iter} that points past the followed branch. The
 values of \vn{branch\_iter} are either \cn{BindBegin}, which denotes that no
@@
   The third difference is more subtle. One of the main features of Ehoh is
-  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): A method which, given a demodulator $s \eq t$, allows
+  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): a method that, given a demodulator $s \eq t$, allows
   to replace both applied and unapplied occurrences of $s$ by $t$ 
@@
   2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
-  wallclock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
-  default time limit:\ 30~s wallclock and CPU.
+  wall-clock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
+  default time limit:\ 30~s wall-clock and CPU.
 
@@
 
-  \ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
+  \ehohii{} was mainly designed as an efficient backed to proof assistants. As such it 
   excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
@@
         \midrule
-        Ehoh -- FO      & 535      \\
-        Ehoh -- HO      & 538      \\
-        \ehohii{} -- FO & {\bf 539}\\
-        \ehohii{} -- HO & {\bf 539}\\
+        Ehoh FO      & 535      \\
+        Ehoh HO      & 538      \\
+        \ehohii{} FO & {\bf 539}\\
+        \ehohii{} HO & {\bf 539}\\
         \bottomrule
@@
   that this extension makes \ehohii{} the best higher-order prover on
-  benchmarks coming from interactive theorem proving practice, which has been our goal since 2016. \ehohii{}'s
-  performance lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
+  benchmarks coming from interactive theorem proving practice, which was our goal. \ehohii{}
+  lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
   does not assume a clausal structure and can perform subtle formula-level

### 2022-03-26 ho-techniques/ho-techniques.tex
biblio

@@
 \cite{gs-19-casc27}, 12~percentage points behind the winner, the tableau prover
-Satallax 3.4 \cite{cb-2013-satallax}.
+Satallax 3.4 \cite{cb-12-satallax}.
 
@@
 essential feature of higher-order theorem provers such as Leo-III
-\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-2013-satallax}.
+\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-12-satallax}.
 % but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
@@
 \cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
-\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
+\cite{cb-12-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
 The provers were run using the same parameters as in CASC, but with updated

### 2022-03-26 introduction/introduction.tex
biblio

@@
 first-order logic remained too enticing. In 1960, Davis and Putnam
-\cite{dp-1960-dpll} described a procedure for checking validity of a
+\cite{dp-60-dp} described a procedure for checking validity of a
 first-order formula. As first-order logic is undecidable, this procedure
@@
   ideas and experimenting with calculi much easier. In 2019 we entered the
-  higher-order division of CASC theorem proving competition with Zipperposition
-  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
+  higher-order division of CASC theorem proving competition~\cite{gs-19-casc27} with Zipperposition. 
+  It finished third, one percent point behind Leo-III
   \cite{sb-21-leo3}  and 12 percent points behind  Satallax

### 2022-03-26 unif/unif.tex
more
biblio

@@
 \label{sec:unif:implementation}
-Zipperposition \cite{sc-15-simon-phd, sc-supind-17} is an
+As described in Sect.~\ref{sec:pre:theorem-provers}, Zipperposition \cite{sc-15-simon-phd, sc-supind-17} is an
 open-source\footnote{\url{https://github.com/sneeuwballen/zipperposition}}
@@
 superposition-based theorem provers.
-It was initially
-designed as a prover for polymorphic first-order logic
+It was initially designed as a prover for rank-1 polymorphic \cite{bp-13-tff1} first-order logic 
 and then extended to higher-order logic.

### 2022-03-28 conclusion/conclusion.tex
concl

@@
 \label{ch:conclusion}
+
+As we bring this thesis to an end, I would like to use these last pages to
+reflect not only on the work I have done in the last four years, but also on the
+many bumps on the road that I did not get to describe in the preceding chapters.
+When it comes to reflections, the  performance art piece Rhythm 10, performed by
+Marina Abramovi\'c,  one of the greatest Yugoslav artists, comes to mind. The
+photograph on the cover of this thesis was taken when it was first performed, in
+Edinburgh, in 1973. Its instructions read:
+
+\begin{quote}
+    \underline{Preparation} \\[1\jot]
+    I place a sheet of white paper on the floor.\\
+    I lay 20 knives of different sizes and shapes on the floor.\\
+    I place 2 tape recorders with microphones on the floor.
+
+    \underline{Performance} \\[1\jot]
+    I turn on the first tape recorder.\\
+    I take the first knife and stab in between the fingers of my left hand as
+    fast as possible.\\
+    Every time I cut myself, I change to a different knife. \\
+    When I've used all of the knives (all the rhythms), I rewind the tape
+    recorder. \\ 
+    I listen to the recording of the first part of the performance. \\ I
+    concentrate. \\
+    I repeat the first part of the performance. \\
+    I take the knives in the same order, follow to the same order, follow to the
+    same rhythm and cut myself in the same places. \\
+    In this performance, the mistakes of time past and time present are
+    synchronized.\\
+    I rewind the second tape recorder and listen to the double rhythm of the
+    knives.\\
+    I leave.
+\end{quote}
+
+Just like this performance, the work that I have done in the Matryoshka project
+was cyclic: It consists of three phases (three stops as we described them in
+Chapter \ref{ch:intro}), each of which has the same parts: theory (calculus and
+algorithms), engineering (implementation) and optimization (heuristics). While
+the calculi were mostly developed by Alexander Bentkamp
+\cite{bbcw-21-lfho,bbtvw-21-sup-lam,bbtv-21-full-ho-sup}, the task of developing
+algorithms for indexing or unification, as well as the tasks of implementing and
+optimizing the implementation were mainly mine.
+
+Structuring the project cyclically turned out to be a great idea. When I first
+started doing the research which will later form the basis of Chapter
+\ref{ch:ehoh}, I was overwhelmed by the scope of the project and felt uncertain
+if E could ever fully support $\lambda$-free higher-order logic. Many of the
+extensions described in the  later chapters were even larger in scope. However,
+as each extension relied on the previous one, I quickly developed a feeling of
+where the possible pitfalls are and how to avoid them. This gave me an invaluable
+source of encouragement, without which I could hardly arrive at the end of the
+project.
+
+The work in Chapter~\ref{ch:ehoh} completed the first cycle. Soon after the
+original paper introducing Ehoh \cite{vbss-19-ehoh1} was published, Ehoh found
+its use as the backend of Sledgehammer and Satallax, showing that even when limited
+to a fragment of higher-order logic, efficient higher-order reasoning is in
+demand.
+
+The goal of the next cycle was to support $\lambda$-abstraction. When we first
+implemented \lsup{}, calculus that achieves this goal, we quickly realized that
+one of its main flaws was that it relies on enumerating full unifiers rather
+than preunifiers, like some older resolution-based higher-order calculi do. This
+is why we focused on developing a full unification procedure that removes the
+redundancy present in other procedures and that implements all the advancements
+and optimizations described in the unification literature after the introduction
+of the most influential full unification procedure, Jensen-Pietrzykowski's
+procedure. We also made sure that this procedure can easily be customized to
+trade bits of its completeness for substantially improved performance. This
+procedure, described in Chapter~\ref{ch:unif}, transformed the \lfsup{} calculus
+into a competitive higher-order calculus.
+
+Even though \lfsup{} became competitive, its implementation in Zipperposition
+still still lagged behind the competition. To investigate why this is the case,
+we manually analyzed hundreds of benchmarks on which the competition proves, but
+on which Zipperposition fails. This analysis forms the basis of work that is
+described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}. The main
+conclusion was that the competition, most notably tableaux-based Satallax,
+reasons better with formulas. Tableaux inferences are more intuitive than the
+ones that superposition or resolution provers perform and are more in line with
+how a mathematician structures a proof. Even though fitting these kind of
+inferences into the superposition context was challenging, it proved very
+successful. After they were implemented in Zipperposition, it took its first
+victory at higher-order division of CASC.
+
+Our extension of \lfsup{} was guided by performance, rather than completeness
+with respect to full higher-order logic. After the first victory, we were left
+wondering what were the kind of problems that our implementation can not
+solve, and started looking for complete full higher-order calculus. At the end,
+we designed \osup{} which gave us the precise answer in the form of problems that
+cannot be solved and the rules which are necessary to solve them.

### 2022-03-28 title/title.tex
more
past tense
new bib and paper
intro
concl

@@
 
-\begin{center}
+% \begin{center}
 
-%% Extra whitespace at the top.
-\vspace*{2\bigskipamount}
+% %% Extra whitespace at the top.
+% \vspace*{2\bigskipamount}
 
-%% Print the title.
-{\makeatletter
-\titlestyle\bfseries\LARGE\@title
-\makeatother}
+% %% Print the title.
+% {\makeatletter
+% \titlestyle\bfseries\LARGE\@title
+% \makeatother}
 
-%% Print the optional subtitle.
-{\makeatletter
-\ifx\@subtitle\undefined\else
-    \bigskip
-    \titlefont\titleshape\Large\@subtitle
-\fi
-\makeatother}
+% %% Print the optional subtitle.
+% {\makeatletter
+% \ifx\@subtitle\undefined\else
+%     \bigskip
+%     \titlefont\titleshape\Large\@subtitle
+% \fi
+% \makeatother}
 
-\end{center}
+% \end{center}
 
@@
 
-op gezag van de Rector Magnificus prof.dr.~V.~Subramaniam,
+op gezag van de Rector Magnificus prof.dr.~J.J.G.~Geurts,
 
@@
     copromotoren: & dr. J. C. Blanchette \\
-                  & dr. Stephan Schulz
+                  & prof.$\,$dr. S. Schulz
 \end{tabular}
@@
 
-\textbf{Should this be filled or skipped as in the case of Alex?}
-
 %% List the committee members, starting with the Rector Magnificus and the
@@
 \begin{tabular}{p{4.5cm}l}
-    Rector Magnificus, & voorzitter \\
-    Prof.\ dr.\ A.\ van Deursen, & Technische Universiteit Delft \\
-    Dr.\ A.E.\ Zaidman, & Technische Universiteit Delft \\
-    Dr.\ ir.\ G.\ Gousios, & Technische Universiteit Delft \\
 
-    \medskip
-    \mbox{\emph{Onafhankelijke leden:}} & \\
-    Prof.\ dr.\ ir.\ G.J.P.M.\ Houben, & Technische Universiteit Delft \\
-    Prof. dr. P. Runeson, & Lund Universitet, Sweden \\
-    Dr.\ Th.\ Zimmermann, & Microsoft Research, \\ &United States of America \\
-    Prof. dr. D. Spinellis, & Athens University of Economics and Business, \\&
-    Greece \\
+prof.$\,$dr. Mark Hoogendoorn & Vrije Universiteit Amsterdam \\
+dr. Konstantin Korovin & The University of Manchester \\
+dr. Ruzica Piskac & Yale University \\
+dr. Alexander Steen & University of Greifswald \\
+dr. Martin Suda & Czech Technical University in Prague \\
+prof.$\,$dr. Geoff Sutcliffe & University of Miami \\
 
-    Prof.\ dr.\ ir.\ E.\ Visser, & Technische Universiteit Delft, reservelid \\ \\
-
-    \multicolumn{2}{l}{Prof. dr. D. Spinellis has contributed to the end
-    phase of writing \ldots} \\
 \end{tabular}
@@
 \medskip
-\noindent The author set this thesis in \LaTeX\xspace using the Minion Pro, Nimbus Sans and Inconsolata fonts.
+\noindent The author set this thesis in \LaTeX\xspace using the Libertinus Serif, Inter and Inconsolata fonts.
 

### 2022-03-29 ehoh2/ehoh2.tex
concl
more

@@
 \newcommand\ehohii{$\lambda$E}
+
+\authors{Joint work with\\
+Jasmin~Blanchette and Stephan~Schulz}
 
@@
 cell $\db{0}$ is set to $t$. Then $s$ is traversed to instantiate every
-loose occurrence of $\db{0}$ in $s$ with \texttt{binding}, whose loosely
+loose occurrence of $\db{0}$ in $s$ with \texttt{binding}, whose loose
 bound De Bruijn indices are shifted by the number of $\lambda$ binders above
@@
   wall-clock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
-  default time limit:\ 30~s wall-clock and CPU.
+  default time limit:\ 30~s wall-clock and CPU.  The raw evaluation data is available online\footnote{\url{https://doi.org/10.5281/zenodo.6389849}}.
 

### 2022-03-29 pre/pre.tex
biblio
more

@@
     chapters. We begin by describing the three logics that we work with in this
-    thesis: propositional logic, monomorphic first-order logic and simply typed
-    higher-order logic. Then, we explain the clausal structure which is the
-    backbone of many calculi for automated provers. We finish with the
-    description of the superposition calculus. As this thesis discusses practical
-    aspects of theorem proving, we define only the fundamental notions, while
-    more advanced notions are intuitively described with references to rigorous
-    definitions. The text of this chapter is
-    partially based on the preliminaries sections of the publications listed in
-    Chapter \ref{ch:intro}.
-
+    thesis: propositional logic, first-order logic and higher-order logic. Then,
+    we explain the clausal structure which is the backbone of many calculi for
+    automated provers. We finish with the description of the superposition
+    calculus. As this thesis discusses practical aspects of theorem proving, we
+    define only the fundamental notions, while more advanced notions are
+    intuitively described with references to rigorous definitions. The text of
+    this chapter is partly based on the preliminaries sections of the
+    publications listed in Chapter \ref{ch:intro}.
 \end{abstract}
@@
 $\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
-for any value of $x$, while the second one requires that there is some $x$ for
+for every value of $x$, while the second one requires that there is some $x$ for
 which $\phi$ holds. As soon as there is a single functional symbol with arity
@@
 Even though we do not formally define the semantics of logic, we assume the natural
-extensions of domain, valuation, interpretation and model (as defined by Fitting
+extensions of domain, valuation, interpretation, and model (as defined by Fitting
 \cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
@@
 (i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
-$\ourmodel \models N.$  Abusing notation, we write $M \models N$ to denote that
+$\ourmodel \models N.$  Overloading notation, we write $M \models N$ to denote that
 $M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$
 
-A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
-the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
-$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
-$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
-position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
-hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
-in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
-We say a term is {\em ground} if it has no variables. By $u[s]$
-we denote that $s$ is a subterm of $u$.
+A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$
+denoting the empty tuple. {\em Subterms} and positions are inductively defined
+as follows. The term $t$ is a subterm of itself at position $\varepsilon$. If
+$s$ is a subterm of $t_i$ at position $p$, then $s$ is a subterm of
+$\cst{f}(\tuplen{t})$ at position $i.p$. A {\em context} is a term with zero or
+more subterms replaced by a hole $\square$. We write $C[\overline{u}_n]$ for the
+term resulting from filling in the holes of a context $C$ with the terms
+$\overline{u}_n$, from left to right. We say a term is {\em ground} if it has no
+variables. By $u[s]$ we denote that $s$ is a subterm of $u$.
 
@@
 Types are either base types $\tau \in T$, or function types $\tau \rightarrow
-\upsilon$ where both $\tau$ and $\upsilon$ are types. When we write $\tau_1
-\rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$ we assume $\upsilon$
-to be a base type ($\upsilon \in T$), by convention. Each symbol $\cst{f} \in
+\upsilon$ where both $\tau$ and $\upsilon$ are types. Each symbol $\cst{f} \in
 \Sigma$ is assigned a type.
@@
 \rightarrow \upsilon$. The syntactic distinction between free and bound
-variables gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
-\cst{a}}$) \cite{tn-93-patterns}. Note that there is no requirement that a
+variables gives rise to \emph{loose bound variables} \cite{tn-93-patterns},
+bound variables that appear without enclosing $\lambda$ binder  (e.g., $y$ in
+$\lamx{y \, \cst{a}}$). Note that there is no requirement that a
 symbol is not applied to terms of Boolean type as in first-order logic.
@@
 \cst{q})$ are well-formed terms of higher-order logic. For convenience, we still
-call terms of Boolean type \emph{formulas}. Our logic also supports Hilbert
+call terms of Boolean type \emph{formulas}. Our logic also supports the Hilbert
 choice operator $\varepsilon$. Intuitively, this operator chooses an arbitrary value that
@@
 With $\alpha$-conversion we assume the rule $(\lambda x. \, s) \longrightarrow
-\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$
+\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur as a loose bound variable in $s$
 and $y$ is not bound in $s$. Application of a substitution to a term,
@@
 $\eta$-reduction is defined as $\lambda x.\, s \, x \longrightarrow s$ under the
-condition that $x$ is not loosely bound in $x$. We write $s \equi t$ if terms
+condition that $x$ is not loose bound in $x$. We write $s \equi t$ if terms
 $s$ and $t$ are equal modulo $\alpha\beta\eta$-conversion. Terms that are
@@
 where $a$ is not an application; we call $a$ the \emph{head} of the term. By
-convention, $a, b$ and $\zeta$ denote heads. If $a$ is a free variable, we call the term
+convention, $a, b$, and $\zeta$ denote heads. If $a$ is a free variable, we call the term
 \emph{flex}; otherwise, the term is \emph{rigid}.
@@
 \, \cst{b}$. We say a term is ground if it has no free variables, and closed
-if it has no loosely bound variables.
+if it has no loose bound variables.
 
@@
 with respect to Henkin semantics \cite{bm-14-automation-ho}. Note that no
-complete calculus exists for higher-order calculi with standard (full)
+complete, consistent calculus can exist for higher-order calculi with standard (full)
 semantics.
@@
 
-The resolution and superposition calculi do not work directly on formulas, but
-on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
-set of formulas, must be transformed into a set of clauses. For all three
-discussed logics, there exists such a transformation that does not affect
-satisfiability of the problem \cite{nw-01-small-cnf}.
+The standard resolution and superposition calculi do not work directly on formulas, but
+on their normal forms, {\em clauses}. Thus, the initial problem, expressed as a
+set of formulas, must be transformed into a set of clauses. For propositional
+and first-order logic, there exists such a transformation that does not affect
+satisfiability of the problem \cite{nw-01-small-cnf}. In higher-order logic,
+formulas are first-class citizens and higher-order calculi can in general work
+with nonnormalized problem, in its original form. However, resolution and
+superposition-based higher-order calculi still perform best when the problem is clausified
+using transformation similar to the one used in the first-order case. We have analyzed
+different approaches to clausification in higher-order context \cite{bbtv-21-full-ho-sup}.
 
@@
 example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
-$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the
-notation, we sometimes write the predicate literals in non-encoded form.
+$\neg\,\cst{even}(x)$ is encoded as $\cst{even}(x) \not\eq \itrue$. To lighten the
+notation, we sometimes write the predicate literals in nonencoded form.
 Applying a substitution to a literal is reduced to applying it to both sides of
 the (dis)equa\-tion and it is extended pointwise to clauses. We say
-$\sigmacl{C}$ is a ground instance of $C$ if $\sigma$ is a grounding
-substitution.
+$\sigmacl{C}$ is a ground instance of $C$ if $\sigmacl{C}$ has no free variables.
 \pagebreak[2]
@@
 rules to the clauses, adding results of the inferences to the working set of
-clauses. As this process is quite prolific it also uses the term order to determine
+clauses. As this process is quite prolific it also uses an order on terms which gives an indication of
+term's ``simplicity'' to determine
 if some clauses can be simplified or eliminated from the working set.
 Superposition is refutationally complete: This means that, if results of all
-inferences are computed in a systematic, fair manner, empty clause ($\bot$) will
+inferences are computed in a systematic, fair manner, the empty clause ($\bot$) will
 eventually be derived for a provable problem. As first-order logic is
@@
 \label{sec:pre:order}
-Superposition calculus is parameterized by a {\em reduction ordering} $\succ$, an
-ordering on terms that has the following properties:
+Superposition calculus is parameterized by a
+term order $\succ$ with the following properties:
 
@@
     \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
-    \textit{Respects contexts} & For all terms $s,t$ and contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
+    \textit{Respects contexts} & For all terms $s,t$ and nonempty contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
     \textit{Ground total} & For all distinct
-    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$. \\
+    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$ \\
     \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
@@
 $S$. To use the  multiset extension, positive literals are represented as $ \{ \{s\},
-\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
+\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$, which makes the negative literals slightly greater than the positive ones.
+Clauses are then represented as 
 multisets of such literals.
@@
 Two orders that are commonly used in superposition theorem proving are the
-Knuth-Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and the
+Knuth--Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and the
 lexicographic path order (LPO) \cite[Sect.~5.4.2]{bn-98-tr-and-all-that}.
@@
 substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
-the most general unifier is unique up to variable renaming. Furthermore,
+the most general unifier is unique up to variable renaming. For example, $\cst{f}(X) \ unif \cst{f}(Y)$ yields the mgu 
+$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. Furthermore,
 first-order logic admits efficient algorithms for computing the most general
@@
 $\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
-that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier
-corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
+that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier,
+when it exists, corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
 the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
-if two higher-order terms are unifiable, but there exist algorithms
-that enumerate all elements of the CSU for unifiable terms.
+if two higher-order terms are unifiable, but there exist procedures
+that enumerate all elements of the CSU for two terms.
 
@@
 ones. For example, if $s$ at $t$ are first-order, then there must exist a singleton
-CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
-unifiers as first-order or higher-order and ensure the context provides enough
-information.
+CSU, i.e., the most general unifier.
 
@@
 
-Before spelling out the inference rules of standard, first-order superposition
+Before spelling out the inference rules of standard, first-order superposition,
 let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
 With $t|_p$ we denote the subterm of $t$ at position $p$ and with $t[p \leftarrow
-t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
-$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
+t']$  we denote the term obtained by replacing this subterm with $t'$. Let \selfun{} be a selection function,
+$\succ$ a term order, $C = l \llor C'$ a clause, and $\sigma$ a
 substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
@@
 Let us give some intuition behind how these rules are applied in practice. The
-\infname{ER} rule is used to replace equality with equality within a clause.
+\infname{ER} rule is used to replace equals by equals within a clause.
 Consider the formula $\iforall x.\,  x \eq \cst{2} \iimplies \cst{even}(x)$. In
@@
 Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
-mathematicians do when they ``replace equals by equals''. Name paramodulation is
+mathematicians do when they replace equals by equals. The name paramodulation is
 also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
-states that an equation $s \eq t$ holds under condition $C$ (or more precisely,
-its negation). Similarly, the right premise asserts that a (dis)equation holds
+states that an equation $s \eq t$ holds under condition $C$. Similarly, the right premise asserts that a (dis)equation holds
 under a condition. The superposition rule simply concatenates the conditions for
@@
 $\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
-between their last literals results in $\neg \cst{even}(X) \llor \neg
+between their last literals results in $\neg \cst{even}(Y) \llor \neg
 \cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.
@@
 The rules described above are used to generate new clauses from already derived
-ones. Even though orderings and selection are used to reduce the search space,
+ones. Even though orders and selection are used to reduce the search space,
 it still grows rather fast. To identify unnecessary clauses in the search
@@
 {\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
-clauses in $N$ such that each clause in this subset is smaller than $C$. More
+clauses in $N$ such that each clause in this subset is $\prec$-smaller than $C$. More
 generally, a clause (ground or not) is redundant in set $N$ if every ground
-instance of $C$ is redundant for some grounding of $N$. Removing redundant
+instance of $C$ is redundant for the grounding of $N$. Removing redundant
 clauses does not affect completeness of superposition.
@@
 simplification rules, which are justified using the redundancy criterion. These
-rules replace the premises with conclusions, and we denote them using double
-bars. As an example of such a rule, let us introduce rewriting (demodulation) of
-negative literals (\infname{RN}):
+rules effectively replace the premises with conclusions, and we denote them
+using double bars. In theory, the rules work by adding conclusions and then
+using them to show that premises are redundant, and can thus be eliminated. Asf
+an example of such a rule, let us introduce rewriting (demodulation) of negative
+literals (\infname{RN}):
 
@@
 The resulting clauses are again simplified and put in $\mathcal{P}$.
-With \emph{proof state} we assume the pair $(\mathcal{P}, \mathcal{A})$.
+We call the pair $(\mathcal{P}, \mathcal{A})$ the \emph{proof state}.
 The provers differ in which clauses are used for simplification: Otter-loop
@@
 %
-Note that we use the name $\lambda$-superposition when we refer to both \lsup{}
+Note that we use the name $\lambda$-superposition for both \lsup{}
 and \osup{}.
@@
 All the techniques described in this thesis are implemented in two superposition
-theorem provers: E \cite{ss-02-brainiac} and Zipperposition
-\cite{sc-15-simon-phd,bbtvw-21-sup-lam}. The latter was used as a testbed for
-prototyping ideas, while the former is the main target of our work as it is more
+theorem provers: Zipperposition
+\cite{sc-15-simon-phd,sc-supind-17} and E \cite{ss-02-brainiac}. The former was used as a testbed for
+prototyping ideas, while the latter is the main target of our work as it is more
 efficient.
-
-\ourpara{E} E is a state-of-the-art first-order prover based on superposition.
-In the last decade, together with its derivatives, it has revolved around the
-second place at the first-order division of CASC theorem prover
-competition \cite{ss-96-casc}. It is open-source, written in C (without using any
-external libraries), and designed around efficient algorithms and data structures
-rather than highly optimized codebase. This makes it a good target for implementing
-successful higher-order techniques.
 
@@
 superposition.
+
+\ourpara{E} E is a state-of-the-art first-order prover based on superposition.
+In the last decade, together with its derivatives, it has revolved around the
+second place at the first-order division of CASC theorem prover
+competition \cite{ss-96-casc}. It is open-source, written in C (without using any
+external libraries), and designed around efficient algorithms and data structures
+rather than highly optimized codebase. This makes it a good target for implementing
+successful higher-order techniques.

### 2022-03-29 pre/pre.tex
db

@@
 
+In practice, many implementations of $\lambda$-terms avoid named representation
+of bound variables described above by using \emph{locally nameless
+representation} \cite{ac-12-locally-nameless}. In this representation, free
+variables retain their names, while abstractions are simply represented by a
+$\lambda$ symbol, without any names. Further, bound variables are represented
+using \emph{De Bruijn indices}. A bound variable with De Bruijn index
+$\dbvar{i}$ is bound by the $i+1$-th enclosing $\lambda$-binder. For example,
+the term $\lambda x. \cst{f} \, (\lambda y z. \cst{g} \, y \, x)$ is represented
+as $\lambda\, \cst{f} \, (\lambda \lambda\, \cst{g} \, \db{0} \, \db{2})$.
+Locally nameless representation reduces the $\alpha$-equivalence test to 
+the syntactic equality test.
+
 \looseness=-1
@@
 The standard resolution and superposition calculi do not work directly on formulas, but
-on their normal forms, {\em clauses}. Thus, the initial problem, expressed as a
+on normal forms, {\em clauses}. Thus, the initial problem, expressed as a
 set of formulas, must be transformed into a set of clauses. For propositional
@@
 superposition-based higher-order calculi still perform best when the problem is clausified
-using transformation similar to the one used in the first-order case. We have analyzed
-different approaches to clausification in higher-order context \cite{bbtv-21-full-ho-sup}.
+using a transformation similar to the one used in the first-order case. We have analyzed
+different approaches to clausification in a higher-order context \cite{bbtv-21-full-ho-sup}.
 

### 2022-03-29 unif/unif.tex
more
db

@@
 parentheses around application of substitutions to terms (or other objects containing terms);
-in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$.
+in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$. When we write $\tau_1
+\rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$ we assume $\upsilon$
+to be a base type, by convention.
 
@@
 & \hotofo{F \, \overline{s}} = F
-&& \hotofo{\dbvar{i} \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
+&& \hotofo{\db{i} \, \overline{s}_n } = \cstdba{i}(\hotofo{s_1}, \ldots, \hotofo{s_n})
 && \hotofo{\cst{f} \, \overline{s}_n } = \cst{f}(\hotofo{s_1}, \ldots, \hotofo{s_n}) &&
@@
 \looseness=-1
-We let $\dbvar{i}$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
+We let $\db{i}$ be a bound variable of type $\alpha$ with De Bruijn index $i$,
 and $\cstdba{i}$ be a fresh constant corresponding to this variable. All constants

### 2022-03-29 bools/bools.tex
more
biblio

@@
 We base our work (and parts of the following text) on Bentkamp et al.'s
-\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. We
+\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. This
+logic supports rank-1 polymorphism, as defined by the TH1 format of TPTP
+\cite{ksr-16-th1}.
+We
 extend the syntax of this logic by adding logical connectives to the signature.
@@
 
-As mentioned in Chapter~\ref{ch:unif}, Zipperposition is an open source prover
+As mentioned in the previous chapter, Zipperposition is an open source prover
 written in OCaml. From its inception, it was designed as a prover that supports
@@
 $\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \ieq \lam{\tuplen{x}}{\ifalse})$, we
-designed the following rule (where $\tuplen{x}$ consists of all loosely bound variables of $f$ ):
+designed the following rule (where $\tuplen{x}$ consists of all loose bound variables of $f$ ):
 %\pagebreak[2]
@@
 rules could be achieved using a combination of the \infname{PI} rule and the
-superposition rules.
+rules of the superposition calculus.
 \pagebreak[2]

### 2022-03-29 conclusion/conclusion.tex
more

@@
 When it comes to reflections, the  performance art piece Rhythm 10, performed by
-Marina Abramovi\'c,  one of the greatest Yugoslav artists, comes to mind. The
-photograph on the cover of this thesis was taken when it was first performed, in
-Edinburgh, in 1973. Its instructions read:
+Marina Abramovi\'c,  one of the greatest Yugoslav artists, comes to my mind. The
+themes that it explores such as taking risks and inevitability of making
+mistakes were the root of fears and doubts that I was slowly overcoming over the
+last four years. The photograph on the cover of this thesis was taken when it
+was first performed, in Edinburgh, in 1973. The instructions for the performance read:
 
@@
 
-Just like this performance, the work that I have done in the Matryoshka project
+The work that I have done for my thesis
 was cyclic: It consists of three phases (three stops as we described them in
@@
 as each extension relied on the previous one, I quickly developed a feeling of
-where the possible pitfalls are and how to avoid them. This gave me an invaluable
-source of encouragement, without which I could hardly arrive at the end of the
-project.
+where the possible pitfalls are and how to avoid them. This gave me an
+invaluable source of encouragement, without which I could hardly arrive at the
+end of the project.
 
@@
 original paper introducing Ehoh \cite{vbss-19-ehoh1} was published, Ehoh found
-its use as the backend of Sledgehammer and Satallax, showing that even when limited
-to a fragment of higher-order logic, efficient higher-order reasoning is in
-demand.
+its use as the backend of Sledgehammer and Satallax, showing that even when
+limited to a fragment of higher-order logic, efficient higher-order reasoning is
+in demand.
 
@@
 trade bits of its completeness for substantially improved performance. This
-procedure, described in Chapter~\ref{ch:unif}, transformed the \lfsup{} calculus
+procedure, described in Chapter~\ref{ch:unif}, transformed the \lsup{} calculus
 into a competitive higher-order calculus.
 
-Even though \lfsup{} became competitive, its implementation in Zipperposition
+Even though \lsup{} became competitive, its implementation in Zipperposition
 still still lagged behind the competition. To investigate why this is the case,
-we manually analyzed hundreds of benchmarks on which the competition proves, but
+we manually analyzed hundreds of benchmarks which the competition proves, but
 on which Zipperposition fails. This analysis forms the basis of work that is
@@
 
-Our extension of \lfsup{} was guided by performance, rather than completeness
+Our extension of \lsup{} was guided by performance, rather than completeness
 with respect to full higher-order logic. After the first victory, we were left
@@
 cannot be solved and the rules which are necessary to solve them.
+
+While preparing for the following year's CASC competition, we found problems
+that occur in practice for which the rules of \osup{} are necessary. However,
+these problems occurred in less than one percent of the whole TPTP library.
+Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
+most of CASC portfolio configurations.
+
+This lead us to take a radically different approach when extending Ehoh to
+\ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
+from assistants, in which hard, hand-crafted mathematical puzzles rarely occur.
+Thus, we conjectured that extending \lfsup{} with the most rudimentary features
+of $\lambda$ superposition such as full higher-order unification and
+$\beta$-reduction-aware term orders will help us prove most of higher-order
+problems without the explosion of complete approaches.
+
+While this proved mostly true, we quickly realized that \ehohii{} should
+assimilate most of incomplete Boolean reasoning techniques described in Chapters
+\ref{ch:bools} and \ref{ch:ho-techniques}. All of them were straightforward to
+implement except for dynamic clausification. It is possible that support for
+this technique does not require profound changes to E(hoh)'s formula treatment,
+but due to the lack of time we did not explore this alley.
+
+This highly pragmatic approach proved successful not only on proof assistant
+benchmarks. \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
+Zipperposition. By manually expecting the TPTP benchmarks that are out of reach
+for \ehohii{}, but within the reach of Zipperposition, we realized that lack of
+dynamic clausification might be the weight necessary to tip the scale in the
+favor of \ehohii{} on TPTP benchmarks.
+
+Lastly,

### 2022-03-29 ehoh2/ehoh2.tex
more

@@
 
-\blfootnote{In this work I designed, implemented and evaluated all changes to
-term representation, algorithms and indexing data structures. Jasmin Blanchette
+\blfootnote{In this work I designed, implemented, and evaluated all changes to
+term representation, algorithms, and indexing data structures. Jasmin Blanchette
 did the daily supervision. Stephan Schulz provided
@@
 engine was designed with the assumption that it will be used with
-inferences that compute a most general unifier. We
+inferences that compute an MGU. We
 implemented a higher-order unification procedure (Chapter \ref{ch:unif})
@@
 integrated it in the inference engine. Finally, we extended and adapted the
-superposition rule, resulting in an incomplete, pragmatic variant of
+rules of superposition calculus, resulting in an incomplete, pragmatic variant of
 $\lambda$-superposition (Sect.~\ref{sec:ehoh2:calculus}).
@@
 to the cooperative version of Zipperposition, which employs Ehoh as a
-backend. An arguably fairer comparison without the backend puts \ehohii{} in the
+backend. An arguably fairer comparison without the backend puts \ehohii{} in
 first place for both benchmark suites.
@@
 $\neqlit{s}{\top}$ is written as $\negpredlit{s}$.   This chapter discusses
-three tightly related provers E, Ehoh and \ehohii{}, which are disambiguated as
+three tightly related provers E, Ehoh, and \ehohii{}, which are disambiguated as
 follows:
@@
 (Sect.~\ref{sec:ehoh:types-and-terms}).
-Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$ and $\cst{f} \, \cst{a} \,
+Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$, and $\cst{f} \, \cst{a} \,
 \cst{b}$ are represented by the cells $\cst{f}$,
@@
 indices must be represented as other cells with either negative or positive
-\texttt{f\_code}. However, this has to be done in such a way that De Bruijn
+\texttt{f\_code}. However, this has to be done in such a way that a De Bruijn
 index can never be mistaken for a function symbol or a variable.
@@
 Bruijn index. Any attempt to create a De Bruijn index is performed through
-a dedicated library function that sets \texttt{IsDBVar} property for every term it
+a dedicated library function that sets the \texttt{IsDBVar} property for every term it
 returns. When given the same De Bruijn index and type, this function is
@@
 $\lambda$-ab\-strac\-tion has the distinguished function code \internallam{} as the head
-symbol and two arguments:\ (1)~a De Bruijn index 0 of the type of abstracted variable;
+symbol and two arguments:\ (1)~a De Bruijn index 0 of the type of the abstracted variable;
 (2)~the body %(matrix)
@@
 x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
-$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
+$\cst{g}\,((\lambda x.\, x)\,\cst{a})$ and $t$ itself have the property
 \texttt{HasBetaReducibleSubterm} set.
@@
 \ourpara{Efficient $\etabf$-Reduction}
-The term $\lambda x.\, s \, x$ is $\eta$-reduced to $s$ whenever $x$ does not appear
-unbound in $s$. Caching the property of $\eta$-reducibility of the term is not
+The term $\lambda x.\, s \, x$ is $\eta$-reduced to $s$ whenever $x$ is not a loose bound variable
+in $s$. Caching the property of $\eta$-reducibility of the term is not
 as beneficial as the one for $\beta$-reducibility, because checking this
@@
 \ehohii{} performs parallel $\eta$-reduction: It recognizes terms of the form
-$\lam{\overline{x}_n}{s \, \overline{x}_n } $ such that none of $x_i$
-occurs unbound in $s$. If done naively, reducing terms of this kind requires up to $n$
+$\lam{\overline{x}_n}{s \, \overline{x}_n } $ such that none of the $x_i$
+occurs loose bound in $s$. If done naively, reducing terms of this kind requires up to $n$
 traversals of $s$ to check if each $x_i$ occurs in $s$. In \ehohii{}, exactly one
@@
 \ourpara{Representation of Boolean Terms}
-Ehoh represents Boolean terms using
-cells whose \texttt{f\_code}s correspond to internal codes reserved
+E represents Boolean terms using
+term cells whose \texttt{f\_code}s correspond to internal codes reserved
 for logical symbols. Quantified formulas are represented by cells in which the
@@
 This representation is convenient for parsing formulas and clausification, which
-is what Ehoh uses it for, but it
+is what E uses it for, but it
 causes $\alpha$-normalization issues during the actual proof search: In
@@
 $\alpha$-conversion issues. However, this solution is incompatible with
-thousands of decades-old lines of clausification code that assumes the Ehoh
+thousands of decades-old lines of clausification code that assumes the E
 representation of
@@
 path order (LPO) for higher-order terms with $\lambda$-abstractions described by
-Bentkamp et al.~\cite{bbtv-21-full-ho-sup}. These orders encode
+Bentkamp et al.~\cite{bbtv-21-full-ho-sup}. These orders
 encode $\lambda$-terms as first-order terms and then invoke the standard KBO or
@@
 series of $\beta\eta$-reductions,
-changing the shape of the term dramatically. To pevent this, \ehohii{}
-computes the $\beta\eta$-reduced instance of the terms
+changing the shape of the term dramatically. To prevent this, \ehohii{}
+computes the $\beta\eta$-reduced instances of the terms
 before comparing them using KBO or LPO.
@@
 
-Standard superposition crucially depends on the concept of a most general
-unifier (MGU). In higher-order logic, such a unifier does not in general exist,
+Standard superposition crucially depends on the concept of an MGU. In higher-order logic, such a unifier does not in general exist,
 and the concept
@@
 
-Unlike in first-order and $\lambda$-free higher-order unification, in full
+Unlike in first-order and $\lambda$-free higher-order unification, in the full
 higher-order case there may be many bindings that lead to a solution. To reduce
@@
     \State $\vn{rhs} \gets \fc{NormalizeHead}(\vn{rhs})$
-    \State normalize and discard the $\lambda$ prefix of $\vn{lhs}$ and $\vn{rhs}$
+    \State normalize and discard the $\lambda$ prefixes of $\vn{lhs}$ and $\vn{rhs}$
 
@@
 not a variable bound to some term. Then the term with shorter
-$\lambda$ prefix is expanded (only on the top-level) so that both have the
-same length of $\lambda$ prefix. Finally, the $\lambda$ prefix is ignored, and we
+$\lambda$ prefix is expanded (only on the top-level) so that $\lambda$ prefixes have the
+same length. Finally, the $\lambda$ prefix is ignored, and we
 focus only on the body. In this way, we avoid fully substituting
@@
 constraint. \ehohii{} implements the most efficient oracles implemented in
-Zipperposition:\ fixpoint and pattern oracle (Sect.~\ref{sec:unif:implementation}). An oracle can return three results:
+Zipperposition:\ fixpoint and pattern (Sect.~\ref{sec:unif:implementation}). An oracle can return three results:
 (1)~there is an MGU for the pair (\cn{Unifiable}), which is recorded in
@@
 %
-If the constraint is rigid--rigid, for unification to succeed the heads of both
+If the constraint is rigid--rigid, for unification to succeed, the heads of both
 sides of the constraint must be the same. Unification then continues
@@
 We have modified the $\lambda$-free higher-order matching algorithm to treat
-$\lambda$ prefixes as above in unification procedure---by bringing the prefixes
+$\lambda$ prefixes as above in the unification procedure---by bringing the prefixes
 to the same length and ignoring them afterwards. This ensures that the
@@
 efficiently determine whether two terms are variable renamings of each other
-or whether a simple matcher can be used as in the case of $(X \, (\lambda x. \,x) \,
+or whether a simple matcher can be used, as in the case of $(X \, (\lambda x. \,x) \,
 \cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
@@
 unifiable or matchable pairs of terms using indexing data structures. To find
-terms unifiable with a query term or instances of a query term it uses
+terms unifiable with a query term or instances of a query term, it uses
 \emph{fingerprint indexing} \cite{ss-12-fp-indexing}. We extended this data
@@
 We have dedicated separate nodes for applied variables because access to arguments
-of applied variable is necessary for the pattern matching algorithm. Even though
+of applied variables is necessary for the pattern matching algorithm. Even though
 arguments can be obtained by querying the arity $n$ of the variable and taking
-next $n$ arguments in the serialization, this is both inefficient and inelegant.
+the next $n$ arguments in the serialization, this is both inefficient and inelegant.
 As for De Bruijn indices, we treat them the same as function symbols and assign
@@
 Backtracking works in the opposite direction: If the current node is labeled with a De
-Bruijn index or function symbol node of arity $n$, pop $n$ terms from \tstack{} and move
+Bruijn index or function symbol node of arity $n$, pop $n$ terms from \tstack{} and move the
 top of \tproc{} to \tstack{}. If the node is labeled with
 $\internallam_\tau$, pop the top of \tstack{} and move the top of \tproc{} to
-\tstack{}. Finally, if the node is labeled with a possibly applied variable, move top of the
+\tstack{}. Finally, if the node is labeled with a possibly applied variable, move the top of the
 \tproc{} to \tstack{} and restore the value of $\sigma$.
@@
   calculus as Ehoh with three important changes. First, wherever Ehoh requires
-  the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
+  the MGU of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
   Sect.~\ref{sec:ehoh2:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
@@
   to replace both applied and unapplied occurrences of $s$ by $t$ 
-  by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
-  optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
+  by traversing only first-order subterms of a rewritable term. In a $\lambda$-free setting this
+  optimization is useful, but in the presence of $\beta\eta$-normalization, shapes of
   terms can change drastically, making it much harder to
@@
   \infname{AC} and \infname{NE} assume that $s$ and $t$ are of function
-  type. In \infname{NE}, $\overline{X}$ denote all the free variables occurring in $s$ and
+  type. In \infname{NE}, $\overline{X}$ denotes all the free variables occurring in $s$ and
   $t$, and $\cst{sk}$ is a fresh Skolem symbol of the appropriate type. \infname{PE}
@@
   In Sect.~\ref{sec:ho-tech:infinite-branching}, we described a variant of the saturation
-  procedure that elegantly supports interleaving computation of unifiers and
+  procedure that elegantly supports interleaving computations of unifiers and
   scheduling
@@
   process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
-  given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
-  $\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
-  superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
+  given the clauses $\cst{f} \, (\cst{p} \iand \cst{q})\allowbreak \eq \cst{a}$ and
+  $\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of the
+  superposition rules \infname{SP} and \infname{SN} (Sect.~\ref{sec:pre:rules}) would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
   \iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
@@
   \cite{br-20-full-sup-w-combs}, and Zipperposition 2.1 (Chapter~\ref{ch:ho-techniques}). 
-  We also include Satallax 3.5
+  We also included Satallax 3.5
   \cite{cb-12-satallax}. We used all \NumberOK{2899} higher-order
@@
 
-  \ehohii{} was mainly designed as an efficient backed to proof assistants. As such it 
+  \ehohii{} was mainly designed as an efficient backend to proof assistants. As such it 
   excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
@@
   we chose all the \NumberOK{1138} unique problems used at CASC
-  from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
+  from 2019 to 2021 in the first-order theorem division and ran Ehoh and \ehohii{}
   both in this first-order (FO) mode and in higher-order (HO) mode.
@@
   We fixed a single
-  configuration of options, because Ehoh and \ehohii{}'s
+  configuration of options, because Ehoh's and \ehohii{}'s
   automatic scheduling methods could select different configurations
@@
   configurations. We chose the \emph{boa} configuration (Sect.~\ref{sec:ehoh:heuristics}),
-  which is the configuration that E~2.2 used the most often in its automatic scheduling mode.
+  which is the configuration that E~2.2 used most often in its automatic scheduling mode.
   The results are shown in Figure~\ref{fig:ehoh}.
@@
   unification algorithm (Chapter \ref{ch:unif}) which can easily be customized
-  to trade bits of its completeness for efficiency. Practically oriented work
+  to trade bits of its completeness for efficiency. Practically oriented work,
   described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques},
@@
   This calculus is also implemented in Zipperposition.
-  However, in unpublished experiments, we found that $\lambda$-superposition
+  However, in informal experiments, we found that $\lambda$-superposition
   performs substantially better, corroborating the CASC results, so we decided to
@@
   winner of CASC higher-order division, Satallax \cite{cb-12-satallax}, is no
-  longer maintained. We give a further overview of other traditional higher-order
-  provers and the calculi they are based in Sect.~\ref{sec:ehoh:discussion-and-related-work}.
+  longer maintained. A further overview of other traditional higher-order provers and the calculi they are based on can be found in Sect.~\ref{sec:ehoh:discussion-and-related-work}.
 

### 2022-03-29 pre/pre.tex
more

@@
 using \emph{De Bruijn indices}. A bound variable with De Bruijn index
-$\dbvar{i}$ is bound by the $i+1$-th enclosing $\lambda$-binder. For example,
-the term $\lambda x. \cst{f} \, (\lambda y z. \cst{g} \, y \, x)$ is represented
+$\dbvar{i}$ is bound by the $i+1$-th enclosing $\lambda$ binder. For example,
+the term $\lambda x. \, \cst{f} \, (\lambda y z. \, \cst{g} \, z \, x)$ is represented
 as $\lambda\, \cst{f} \, (\lambda \lambda\, \cst{g} \, \db{0} \, \db{2})$.
@@
 \medskip
+Note that with \emph{superposition rule} we refer to both \infname{SN} and \infname{SP}
+rules.
 Let us give some intuition behind how these rules are applied in practice. The

### 2022-03-29 conclusion/conclusion.tex
more

@@
 As we bring this thesis to an end, I would like to use these last pages to
-reflect not only on the work I have done in the last four years, but also on the
+reflect not only on the work I did in the last four years, but also on the
 many bumps on the road that I did not get to describe in the preceding chapters.
@@
 
-Structuring the project cyclically turned out to be a great idea. When I first
-started doing the research which will later form the basis of Chapter
-\ref{ch:ehoh}, I was overwhelmed by the scope of the project and felt uncertain
-if E could ever fully support $\lambda$-free higher-order logic. Many of the
-extensions described in the  later chapters were even larger in scope. However,
-as each extension relied on the previous one, I quickly developed a feeling of
-where the possible pitfalls are and how to avoid them. This gave me an
-invaluable source of encouragement, without which I could hardly arrive at the
-end of the project.
+Structuring the project cyclically turned out to be a great idea of my
+supervisor. When I first started doing the research which will later form the
+basis of Chapter \ref{ch:ehoh}, I was overwhelmed by the scope of the project
+and felt uncertain if E could ever fully support $\lambda$-free higher-order
+logic. Many of the extensions described in the  later chapters were even larger
+in scope. However, as each extension relied on the previous one, I quickly
+developed a feeling of where the possible pitfalls are and how to avoid them.
+This gave me an invaluable source of encouragement, without which I could hardly
+arrive at the end of the project.
 
@@
 
+\looseness=-1
 The goal of the next cycle was to support $\lambda$-abstraction. When we first
-implemented \lsup{}, calculus that achieves this goal, we quickly realized that
+implemented \lsup{}, the calculus that achieves this goal, we quickly realized that
 one of its main flaws was that it relies on enumerating full unifiers rather
@@
 is why we focused on developing a full unification procedure that removes the
-redundancy present in other procedures and that implements all the advancements
+redundancy present in other procedures. It also implements many advancements
 and optimizations described in the unification literature after the introduction
 of the most influential full unification procedure, Jensen-Pietrzykowski's
-procedure. We also made sure that this procedure can easily be customized to
+procedure in 1970s. We also made sure that our procedure can easily be customized to
 trade bits of its completeness for substantially improved performance. This
@@
 Even though \lsup{} became competitive, its implementation in Zipperposition
-still still lagged behind the competition. To investigate why this is the case,
+lagged behind the competition. To investigate why this is the case,
 we manually analyzed hundreds of benchmarks which the competition proves, but
@@
 ones that superposition or resolution provers perform and are more in line with
-how a mathematician structures a proof. Even though fitting these kind of
-inferences into the superposition context was challenging, it proved very
+how a mathematician structures a proof. Fitting these kind of
+inferences into the superposition context was challenging, but it proved very
 successful. After they were implemented in Zipperposition, it took its first
@@
 with respect to full higher-order logic. After the first victory, we were left
-wondering what were the kind of problems that our implementation can not
+wondering what were the kind of problems that our implementation cannot
 solve, and started looking for complete full higher-order calculus. At the end,
@@
 that occur in practice for which the rules of \osup{} are necessary. However,
-these problems occurred in less than one percent of the whole TPTP library.
+these problems occur in less than one percent of the whole TPTP library.
 Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
@@
 \ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
-from assistants, in which hard, hand-crafted mathematical puzzles rarely occur.
+from proof assistants, in which hard, hand-crafted mathematical puzzles rarely occur.
 Thus, we conjectured that extending \lfsup{} with the most rudimentary features
@@
 This highly pragmatic approach proved successful not only on proof assistant
-benchmarks. \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
+benchmarks: \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
 Zipperposition. By manually expecting the TPTP benchmarks that are out of reach
@@
 
-Lastly,
+This extension of Ehoh to \ehohii{} finished the third cycle. Inspired by our
+approach of extending techniques designed for weaker logics to work in the
+context of richer logics, we decided to look for ways in which superposition can
+assimilate the most successful techniques of SAT solving. Extension of hidden
+literals was straightforward, but finding the right way to tame their infinite
+nature in the first-order case, as well as the right way to integrate equality
+in their definition was challenging. Predicate elimination posed challenges in
+terms of finding the exact conditions in which it can be integrated with the
+saturation loop. We first observed that blocked clause elimination destroys the
+completeness of superposition calculus when we noticed that Zipperposition does
+not prove some problems when it is enabled. After sifting through tens of
+benchmarks and carefully examining the debugging information we realized that
+this was not due to a bug in the implementation but due to incompatibility of
+blocked clause elimination with redundancy criterion. In the end, we were left
+somewhat disappointed to learn that SAT techniques do not scale well to be used
+as superposition simplification techniques. However, we learned that they are
+very successful as preprocessing techniques.
+
+As it is customary, we will conclude the thesis with some remarks about the future work. \ehohii{} has been brought to the shape
+in which further experimentation with higher-order superposition is possible.
+From the experience gained in Chapters \ref{ch:bools} and
+\ref{ch:ho-techniques}, we can point out dynamic clausification as the main
+feature that is not yet implemented. We plan on implementing it in future.
+
+In Chapter \ref{ch:ho-techniques} we only scratched the surface of finding the
+appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
+reasoning it would be interesting to verify that our heuristics are agnostic of
+the version of implemented higher-order superposition calculus.
+
+Some of the SAT-based techniques described in Chapter~\ref{ch:satfol} are
+implemented in E as well. However, this implementation is still in its early
+phases as it is not properly tested and optimized. Verifying the techniques are
+prover-agnostic would increase the confidence in them. We also started the work of extending
+those techniques to full higher-order logic.

### 2022-03-30 ehoh2/ehoh2.tex
more

@@
 retrofitted into an originally first-order prover. Another principle that guided 
-the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
+the design of \ehohii{} was \emph{gracefulness}: We made sure that our changes
 do not impact the strong first-order performance of E and $\lambda$-free higher-order performance of Ehoh.
@@
 (Sect.~\ref{sec:ehoh2:eval}). We found
-that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on the
+that \ehohii{} clearly outperformed Ehoh on all benchmarks. It outperformed all other higher-order provers on the
 proof assistant benchmarks; on the TPTP benchmarks it ended up second only 
@@
 To ensure free variables are
-perfectly shared, Ehoh treats applied free variables differently: arguments are
+perfectly shared, Ehoh treats applied free variables differently: Arguments are
 not applied directly to a free variable, but using an internal symbol
@@
 kind involves intricate manipulation of the cell data structure. De Bruijn
-indices must be represented as other cells with either negative or positive
+indices must be represented as other cells with either a negative or a positive
 \texttt{f\_code}. However, this has to be done in such a way that a De Bruijn
@@
 x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
-$\cst{g}\,((\lambda x.\, x)\,\cst{a})$ and $t$ itself have the property
+$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
 \texttt{HasBetaReducibleSubterm} set.
@@
         \ElsIf{\vn{lhs.head = rhs.head}}
-          \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
-          \State $\quad$ and push them to \vn{iter.constrants}
+          \State push the constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+                 to \vn{iter.constraints}
           \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
@@
     \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
-      \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
-      \State $\quad$ and push them to \vn{iter.constrants}
+    \State push the constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
+           to \vn{iter.constraints}
     \Else{} \State $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
@@
   \EndFunction
-
-  \newpage
+  \vspace{2\jot}
+  \pagebreak[2]
 
@@
 not a variable bound to some term. Then the term with shorter
-$\lambda$ prefix is expanded (only on the top-level) so that $\lambda$ prefixes have the
+$\lambda$ prefix is expanded (only on the top level) so that $\lambda$ prefixes have the
 same length. Finally, the $\lambda$ prefix is ignored, and we
@@
 to the same length and ignoring them afterwards. This ensures that the
-algorithm will never try to match a free variable to a $\lambda$-abstraction,
+algorithm will never try to match a free variable with a $\lambda$-abstraction,
 making sure that $\beta$-redexes never appear (as in original $\lambda$-free
@@
 
-  \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reason about
+  \begin{sloppypar}
+    \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reasoning about
     functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
@@
     $s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
-  $s_i$ and $t_i$ are not (applied) free variables, and they are of Boolean or
+    $s_i$ and $t_i$ are not (possibly applied) free variables, and they are of Boolean or
     function type. Then, the conclusion is formed by taking the conclusion $D$ of
@@
     and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.
+
+  \end{sloppypar}
 
@@
   Counterintuitively, the higher-order versions of both provers outperform
-  the first-order counterparts. However, the differerence is so small that it can
+  the first-order counterparts. However, the difference is so small that it can
   be attributed to the changes to memory layout that affect the order in which

### 2022-03-30 satfol/satfol.tex
unif
biblio
more

@@
    \looseness=-1
-   Let $G$ be a clause set, $\cst{p}$ be a predicate symbol, and
-   $\tupleempty{x}$ be distinct variables.
+   Let $G$ be a clause set and $\cst{p}$ be a predicate symbol.
 %
@@
       \item $\cst{p}$ is singular for $G$
-      \item $G$ consists of clauses of the form $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming)
-      \item the variables in $C'$ are all among $\tupleempty{x}$
+      \item $G$ consists of clauses of the form
+      $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming),
+      where the variables $\tupleempty{x}$ are distinct
+      \item the
+      variables in $C'$ are all among $\cst{p}$'s arguments $\tupleempty{x}$
       \item all clauses in $\withpredpos{G}{\cst{p}}
@@
 techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
-All techniques have been implemented in the Zipperposition prover
-\cite{sc-supind-17}.
+All techniques have been implemented in the Zipperposition prover.
 
@@
 
-\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
 \algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}

### 2022-03-30 summary/summary.tex
summary_eng

\chapter*{Summary}
\addcontentsline{toc}{chapter}{Summary}
\setheader{Summary}

\looseness=-1
In the last decades, proof assistants were immeasurably useful in formally
proving validity of hard mathematical theories and correctness of
safety-critical software. These tools provide a formal language that allows one
to describe a problem and prove statements about it which are later
automatically checked. To make the checking phase efficient and trustworthy,
proof steps need to be of high granularity. As a consequence, seemingly
simple statements must be proven in minute detail, making the use of proof
assistants very tedious.

In an attempt to automate significant parts of this proving process, assistants
can invoke automatic theorem provers to finish the proof. However, most
assistants are based on higher-order logic, while most automatic theorem provers
are based on first-order logic. This means that the two systems must communicate
through translations, which obfuscate the problem and likely make it more
difficult than it originally was.

In this thesis, we try to bridge this translation gap. We start from E, one of the best
first-order proof assistants backends, based on the superposition calculus, and
gradually extend it to higher-order logic. As this approach is large in scope,
we split it in three parts.

The first part extends E to support a fragment of higher-order logic devoid of
$\lambda$-abstra\-ction, yielding a prover called Ehoh. Even though this extension
is arguably conservative, Ehoh shows stronger performance on proof assistant
benchmarks than E.

The second part concerns adding support for $\lambda$-abstraction. The most
important challenge we faced during this extension is that of higher-order
unification. At the time we started, state-of-the-art procedure
for full higher-order unification was developed in 1970s. We designed a new
procedure inspired by this one that enumerates less redundant unifiers, has many
modern optimizations built-in, and can easily be customized to trade its
completeness for performance. It is implemented in a
prototype prover Zipperposition, a less efficient, but more easily extendible
prover compared to E. Our evaluation shows that our
procedure substantially improves on state-of-the-art.

The last part concerns adding support for
first-class Boolean terms. For this part we took inspiration from traditional
automatic higher-order provers, based on tableaux. We fitted those techniques in
superposition context and implemented them in Zipperposition. They
made Zipperposition the best prover in the higher-order division of the annual CASC
theorem prover competition for two consecutive years.

After testing out our ideas in Zipperposition, we decided to go back to Ehoh and
extend it to full higher-order logic, obtaining a prover called \ehohii{}. We
choose the most successful approaches implemented in Zipperposition that can
easily be ported to \ehohii{}. The results were once again positive: Our
evaluation shows that \ehohii{} is the best prover on proof assistant
benchmarks, and second only to Zipperposition on a standard benchmark set.

We took our idea of porting techniques from weaker to stronger logics one step
back: We explored SAT simplification techniques that can be implemented in the
superposition context. We discovered that while some of them can be used during proving,
most of them work best as preprocessors.

\chapter*{Samenvatting}
\addcontentsline{toc}{chapter}{Samenvatting}
\setheader{Samenvatting}

{\selectlanguage{dutch}

  Samenvatting in het Nederlands.
}

### 2022-03-30 conclusion/conclusion.tex
concl

@@
 
-As we bring this thesis to an end, I would like to use these last pages to
+As this thesis is brought to an end, I would like to use these last pages to
 reflect not only on the work I did in the last four years, but also on the
@@
 \ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
-from proof assistants, in which hard, hand-crafted mathematical puzzles rarely occur.
+from proof assistants. Unlike in TPTP, on these benchmarks hard, hand-crafted mathematical puzzles rarely occur.
 Thus, we conjectured that extending \lfsup{} with the most rudimentary features

### 2022-03-30 conclusion/conclusion.tex
more

@@
 appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
-reasoning it would be interesting to verify that our heuristics are agnostic of
+reasoning it would be interesting to verify that heuristics that are still not ported are agnostic of
 the version of implemented higher-order superposition calculus.

### 2022-03-30 conclusion/conclusion.tex
more

--- 2022-03-29 conclusion/conclusion.tex
@@
 completeness of superposition calculus when we noticed that Zipperposition does
-not prove some problems when it is enabled. After sifting through tens of
+not prove some problems when the technique is enabled. After sifting through tens of
 benchmarks and carefully examining the debugging information we realized that
@@
 blocked clause elimination with redundancy criterion. In the end, we were left
-somewhat disappointed to learn that SAT techniques do not scale well to be used
+somewhat disappointed to learn that SAT techniques do not scale good enough to be used
 as superposition simplification techniques. However, we learned that they are

### 2022-03-30 ehoh2/ehoh2.tex
summary_eng
more

--- 2022-03-29 ehoh2/ehoh2.tex
@@
   $\lambda$-abstra\-ction but no Booleans \cite{bbtvw-21-sup-lam}, and \osup{}
-  for full higher-order logic \cite{bbtvw-21-sup-lam}. These milestones allowed
+  for full higher-order logic \cite{bbtv-21-full-ho-sup}. These milestones allowed
   us to carefully estimate how the increased reasoning capabilities of each
@@
 
-  Possibly the only actively maintained higher-order prover built from the bottom
-  up as a higher-order prover is Leo-III \cite{sb-21-leo3}, as the long-time
-  winner of CASC higher-order division, Satallax \cite{cb-12-satallax}, is no
-  longer maintained. A further overview of other traditional higher-order provers and the calculi they are based on can be found in Sect.~\ref{sec:ehoh:discussion-and-related-work}.
+  Possibly the only actively maintained higher-order prover built from the
+  bottom up as a higher-order prover is Leo-III \cite{sb-21-leo3}, as the
+  long-time winner of CASC higher-order division, Satallax
+  \cite{cb-12-satallax}, is no longer maintained. A further overview of other
+  traditional higher-order provers and the calculi they are based on can be
+  found in Sect.~\ref{sec:ehoh:discussion-and-related-work}.
 

### 2022-03-30 ehoh2/ehoh2.tex

--- 2022-03-30 ehoh2/ehoh2.tex
@@
 % \renewcommand{chapter}[0]{chapter}
-\newcommand\ehohii{$\lambda$E}
 

### 2022-03-30 satfol/satfol.tex

@@
 The most important novel aspect of our predicate elimination implementation is
-recognizing the definition clauses for symbol $\cst{p}$ in a clause set $N$, which is performed
+recognizing the definition clauses for a symbol $\cst{p}$ in a clause set $N$, which is performed
 as follows, assuming $\tupleempty{x}$ is a fixed tuple of distinct free variables:
@@
    \item Let $G = \{ C \mid C = \arbpredlit{\cst{p}(\tupleempty{y})} \cor C', C \in
-   N$, no variable repeats in $\tupleempty{y}$, and the variables of $C'$ are among
+   N$, no variable repeats in $\tupleempty{y}$, and the variables of $C'$ are all among
    $\tupleempty{y}\}$. If $G$ is empty, report failure; otherwise, continue.
@@
    \arbpredlit{\cst{p}(\tupleempty{x})} \cor C' \in G \}$. If $E$ is satisfiable,
-   report failure. Else, let $E'$ be the unsatisfiable core of $E$
+   report failure. Otherwise, let $E'$ be an unsatisfiable core of $E$
    and $G'$ the set of corresponding first-order clauses and continue.
@@
    \item If all resolvents in $G'_{\pospredlit{\cst{p}}} \flatres_{\!\cst{p}}
-   G'_{\negpredlit{\cst{p}}}$ are tautologies, then $G'$ is the definition set
-   for symbol $\cst{p}$. Else, report failure.
+   G'_{\negpredlit{\cst{p}}}$ are tautologies, then $G'$ is a definition set
+   for symbol $\cst{p}$. Otherwise, report failure.
 \end{enumerate}

### 2022-03-30 summary/summary.tex
concl

@@
 proving validity of hard mathematical theories and correctness of
-safety-critical software. These tools provide a formal language that allows one
-to describe a problem and prove statements about it which are later
-automatically checked. To make the checking phase efficient and trustworthy,
-proof steps need to be of high granularity. As a consequence, seemingly
-simple statements must be proven in minute detail, making the use of proof
-assistants very tedious.
+safety-critical software. Using these tools one can formally describe a problem
+and produce machine-checkable proofs for statements about it. To make the
+checking phase efficient and trustworthy, proof steps need to be of high
+granularity. As a consequence, seemingly simple statements must be proven in
+minute detail, making the use of proof assistants very tedious.
 
@@
 extend it to full higher-order logic, obtaining a prover called \ehohii{}. We
-choose the most successful approaches implemented in Zipperposition that can
+chose the most successful approaches implemented in Zipperposition that can
 easily be ported to \ehohii{}. The results were once again positive: Our
@@
 
-  Samenvatting in het Nederlands.
 }

### 2022-03-30 title/title.tex
more

@@
 \begin{tabular}{ll}
-    promotor: & prof.dr. W. J. Fokkink \\
+    promotor: & prof.$\,$dr. W. J. Fokkink \\
     copromotoren: & dr. J. C. Blanchette \\
@@
 
-prof.$\,$dr. Mark Hoogendoorn & Vrije Universiteit Amsterdam \\
-dr. Konstantin Korovin & The University of Manchester \\
-dr. Ruzica Piskac & Yale University \\
-dr. Alexander Steen & University of Greifswald \\
-dr. Martin Suda & Czech Technical University in Prague \\
-prof.$\,$dr. Geoff Sutcliffe & University of Miami \\
+prof.$\,$dr. M. Hoogendoorn & Vrije Universiteit Amsterdam \\
+dr. K. Korovin & The University of Manchester \\
+dr. R. Piskac & Yale University \\
+dr. A. Steen & University of Greifswald \\
+dr. M. Suda & Czech Technical University in Prague \\
+prof.$\,$dr. G. Sutcliffe & University of Miami \\
 

### 2022-03-30 conclusion/conclusion.tex
summary

@@
 
-The work that I have done for my thesis
-was cyclic: It consists of three phases (three stops as we described them in
+The work described in this thesis
+was cyclic: It consists of three phases (three stops as described in
 Chapter \ref{ch:intro}), each of which has the same parts: theory (calculus and
@@
 
-Structuring the project cyclically turned out to be a great idea of my
-supervisor. When I first started doing the research which will later form the
+Structuring the project cyclically turned out to be a great idea of Jasmin
+Blanchette. When I first started doing the research which will later form the
 basis of Chapter \ref{ch:ehoh}, I was overwhelmed by the scope of the project
@@
 described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}. The main
-conclusion was that the competition, most notably tableaux-based Satallax,
+conclusion was that the competition, most notably the tableaux-based Satallax,
 reasons better with formulas. Tableaux inferences are more intuitive than the
@@
 
+After its first victory, Zipperposition was successfully integrated in
+Sledgehammer and it is now part of its default installation. We also noticed
+that problems which only Zipperposition can solve do pop up from time to time in
+practice. Antoine Defourné also integrated Zipperposition in TLA+ proof
+assistant and observed that it helps prove some problems that were previously
+out of reach \cite{ad-21-ho-tlaplus}.
+
 Our extension of \lsup{} was guided by performance, rather than completeness
@@
 solve, and started looking for complete full higher-order calculus. At the end,
-we designed \osup{} which gave us the precise answer in the form of problems that
+we designed \osup{}, which gave us the precise answer in the form of problems that
 cannot be solved and the rules which are necessary to solve them.
@@
 Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
-most of CASC portfolio configurations.
+most CASC portfolio configurations.
 
-This lead us to take a radically different approach when extending Ehoh to
+This led us to take a radically different approach when extending Ehoh to
 \ehohii{}. We envisioned \ehohii{} as a prover that excels on problems coming
@@
 Thus, we conjectured that extending \lfsup{} with the most rudimentary features
-of $\lambda$ superposition such as full higher-order unification and
+of $\lambda$-superposition such as full higher-order unification and
 $\beta$-reduction-aware term orders will help us prove most of higher-order
@@
 \ref{ch:bools} and \ref{ch:ho-techniques}. All of them were straightforward to
-implement except for dynamic clausification. It is possible that support for
-this technique does not require profound changes to E(hoh)'s formula treatment,
-but due to the lack of time we did not explore this alley.
+implement except for dynamic clausification, which we did not implement. It is
+possible that support for this technique does not require profound changes to
+E(hoh)'s formula treatment, but due to the lack of time we did not explore this
+alley.
 
@@
 benchmarks: \ehohii{} also excels on TPTP benchmarks, trailing slightly behind
-Zipperposition. By manually expecting the TPTP benchmarks that are out of reach
+Zipperposition. In the coming months we plan to replace Ehoh by \ehohii{} in Sledgehammer.
+As Ehoh is already one of the most successful Sledgehammer backends,
+this will further improve the efficiency of hundreds of mathematicians and computer scientists.
+By manually expecting the TPTP benchmarks that are out of reach
 for \ehohii{}, but within the reach of Zipperposition, we realized that lack of
-dynamic clausification might be the weight necessary to tip the scale in the
+dynamic clausification might be the weight necessary to tip the scale in
 favor of \ehohii{} on TPTP benchmarks.
@@
 in their definition was challenging. Predicate elimination posed challenges in
-terms of finding the exact conditions in which it can be integrated with the
+terms of finding the exact conditions under which it can be integrated with the
 saturation loop. We first observed that blocked clause elimination destroys the
@@
 not prove some problems when the technique is enabled. After sifting through tens of
-benchmarks and carefully examining the debugging information we realized that
+benchmarks and carefully examining the debugging information, we realized that
 this was not due to a bug in the implementation but due to incompatibility of
@@
 
-As it is customary, we will conclude the thesis with some remarks about the future work. \ehohii{} has been brought to the shape
-in which further experimentation with higher-order superposition is possible.
-From the experience gained in Chapters \ref{ch:bools} and
-\ref{ch:ho-techniques}, we can point out dynamic clausification as the main
-feature that is not yet implemented. We plan on implementing it in future.
+As it is customary, we will conclude the thesis with some remarks about the
+future work. \ehohii{} has been brought to the shape in which further
+experimentation with higher-order superposition is possible. From the experience
+gained in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}, we can point out
+dynamic clausification as the main feature that is not yet implemented. We plan
+on implementing it in future.
 
@@
 appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
-reasoning it would be interesting to verify that heuristics that are still not ported are agnostic of
-the version of implemented higher-order superposition calculus.
+reasoning it would be interesting to verify that heuristics that are still not
+ported are agnostic of the version of implemented higher-order superposition
+calculus.
 
@@
 phases as it is not properly tested and optimized. Verifying the techniques are
-prover-agnostic would increase the confidence in them. We also started the work of extending
-those techniques to full higher-order logic.
+prover-agnostic would increase the confidence in them. We also started the work
+of extending those techniques to full higher-order logic.

### 2022-03-30 summary/summary.tex
summary

@@
 and produce machine-checkable proofs for statements about it. To make the
-checking phase efficient and trustworthy, proof steps need to be of high
+checking phase efficient and trustworthy, proof steps need to be of fine
 granularity. As a consequence, seemingly simple statements must be proven in
@@
 The first part extends E to support a fragment of higher-order logic devoid of
-$\lambda$-abstra\-ction, yielding a prover called Ehoh. Even though this extension
-is arguably conservative, Ehoh shows stronger performance on proof assistant
+$\lambda$-abstra\-ction, yielding a prover called Ehoh. Despite this extension is small in scope, 
+Ehoh shows stronger performance on proof assistant
 benchmarks than E.
@@
 for full higher-order unification was developed in 1970s. We designed a new
-procedure inspired by this one that enumerates less redundant unifiers, has many
+procedure inspired by this one that enumerates fewer redundant unifiers, has many
 modern optimizations built-in, and can easily be customized to trade its
 completeness for performance. It is implemented in a
-prototype prover Zipperposition, a less efficient, but more easily extendible
+prototype prover called Zipperposition, a less efficient, but more easily extendible
 prover compared to E. Our evaluation shows that our
-procedure substantially improves on state-of-the-art.
+procedure substantially improves on the state of the art.
 

### 2022-03-31 acks/acks.tex
intro

@@
 
-Without a doubt, the acknowledgments are the most widely and most
-eagerly read part of any thesis.
+When I downloaded the \LaTeX{} template to write this thesis, the following
+sentence was present as a placeholder:
+\begin{quote}
+    Without a doubt, the acknowledgments are the most widely and most eagerly
+    read part of any thesis.
+\end{quote}
+
+I do not know if that is true in general, but I have to admit I am guilty of
+spending quite some time reading acknowledgments. I find that the reason why we
+spend so much time reading acknowledgments is that we want to peek behind the
+rigorous, formal shape that a PhD thesis is required to take. We also want to
+see what PhD candidates have to say about their personal experiences of
+finishing a PhD.
+
+After reading many preface or acknowledgments chapters I realized that there
+is a discrepancy between what research shows candidate's experience of finishing
+a PhD is and what is actually written in these parts of theses. An
+article by Levecque et al. \cite{labhg-17-phdstudents} states that one in two
+PhD students experiences psychological distress, while one in three is at risk
+of a common psychiatric disorder. In the light of this fact, I want to write
+this chapter not only to the people whose help and work indebted me, but also to
+(future) PhD students that, after looking up a detail in this thesis,
+voyeuristically read this chapter (like I did many times for other theses).
+
+In my first year of PhD, I was overwhelmed with fear and doubts. Did I make the
+right choice by starting a PhD? Am I performing up to expectations? How many
+papers should I publish? Should I read more papers? Did I \emph{actually} learn
+anything? I could go on an on like this.
+
+The fact that I was surrounded by very successful postdocs and PhD students on
+later years of their studies made me feel like I was the only one having this
+problem. To the students reading this chapter, I would very much like to assure
+you you are not.
+
+What helped me go out of the self-doubting loop was actually taking action. From
+the simplest ones, like saying ``I don't understand'' out loud instead of
+silently nodding when I have a trouble following scientific discussion, to the
+ones with more impact on my career like actively looking for projects I could
+join while I did not have much work on my plate. As the years went on I felt
+that I could more openly talk about my insecurities and I found out that many
+people in my situation have them as well.
+
+What made my PhD journey more enjoyable are undoubtedly the people I worked
+with. I want to thank the whole Theoretical Computer Science group at the VU
+which was there with me from the early days of my master studies. They helped me
+organize my master programme so that I follow more theory courses which most
+likely lead  me to start an academic career. I also want to thank other postdocs
+and PhDs that studied during my PhD or are still studying. Most notably, I would
+like to thank Alexander Bentkamp, one of the most gifted mathematicians I know.
+I coauthored most of my papers with him, each of which was a pleasure to work
+on. I would also like to thank all of my collaborators and people that suggested
+improvements to the articles I wrote.
+
+I want to thank my daily supervisor, Jasmin Blanchette. He always says that
+the main product of a PhD is not a thesis but a person you come out as after four
+years. With his guidance, I learned to approach things more rigorously, slower,
+and with more attention. He also helped me tremendously  with my writing and
+presentation skills. Even though he tried very hard, he did not teach me to use 
+articles though, and capitulated by adding them to all of the articles I wrote.
+
+Lastly, I would like to thank Martijn. He was besides me for the last four years
+and helped me overcome many of the doubts and issues I mentioned. He also always
+patiently waits until my medication or second glass of Grüner Veltliner kick in
+(not at the same time). I also want to thank Jelisaveta and Tara, which
+sometimes had to deal with me when medication or wine did not kick in. All my other friends,
+in Serbia or the Netherlands, thank you for your support.
 

### 2022-03-31 conclusion/conclusion.tex
intro

@@
 As Ehoh is already one of the most successful Sledgehammer backends,
-this will further improve the efficiency of hundreds of mathematicians and computer scientists.
+this will further improve the efficiency of hundreds of mathematicians and computer scientists using Isabelle.
 By manually expecting the TPTP benchmarks that are out of reach

### 2022-03-31 introduction/introduction.tex
more
intro

@@
 
+  A common thread for extensions of both E and Zipperposition is that we
+  heavily relied on the availability of established higher-order provers and the
+  TPTP benchmark set. Whenever we noticed that there are problems E and Zipperposition
+  cannot solve, but the competition can we tried to come up for an explanation
+  for this issue. This explanation would usually be in the form of missing
+  inference or heuristic. Then we would try to find more problems that have a
+  similar pattern and see if the developed technique is useful on them. If so,
+  we would make it part of the portfolio of approaches tried on every problem.
+  Ideas for our unification procedure (Chapter \ref{ch:unif}) and techniques
+  described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques} have been developed
+  this way.
+
   After spending so much time and energy on developing higher-order provers, we
@@
 
-As a PhD candidate, I coauthored the following articles and papers, that not included in this thesis:
+As a PhD candidate, I coauthored the following articles and papers, that are not included in this thesis:
 
 \begin{enumerate}[resume]
-  % \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
-  % Draft paper.
+  \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
+  Accepted at ITP 2022.
   % \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.

### 2022-04-11 introduction/introduction.tex
wan greater comments

@@
 
-  A common thread for extensions of both E and Zipperposition is that we
+  A common thread for the extensions of both E and Zipperposition is that we
   heavily relied on the availability of established higher-order provers and the
-  TPTP benchmark set. Whenever we noticed that there are problems E and Zipperposition
-  cannot solve, but the competition can we tried to come up for an explanation
-  for this issue. This explanation would usually be in the form of missing
-  inference or heuristic. Then we would try to find more problems that have a
-  similar pattern and see if the developed technique is useful on them. If so,
-  we would make it part of the portfolio of approaches tried on every problem.
-  Ideas for our unification procedure (Chapter \ref{ch:unif}) and techniques
-  described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques} have been developed
-  this way.
+  TPTP benchmark set. Whenever we noticed that there were problems E and
+  Zipperposition could not solve, but the competition could, we looked for an
+  explanation for this issue. This explanation would usually be in the form of
+  missing inference or heuristic. Then we would try to find more problems that
+  have a similar pattern and see if the developed technique is useful on them.
+  If so, we would make it part of the portfolio of approaches tried on every
+  problem. The ideas for our unification procedure (Chapter \ref{ch:unif}) and
+  the techniques described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}
+  were developed in this way.
 

### 2022-04-11 pre/pre.tex
wan greater comments

@@
 such that $\varrho = \theta\sigma$.
-A \emph{most general unifier} is a
+A \emph{most general unifier} ({MGU}) is a
 unifier $\sigma$, such that for any other unifier $\theta$, there is a
 substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
-the most general unifier is unique up to variable renaming. For example, $\cst{f}(X) \ unif \cst{f}(Y)$ yields the mgu 
-$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. Furthermore,
-first-order logic admits efficient algorithms for computing the most general
-unifier \cite{hv-09-unifalgs}.
+the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
+$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient algorithms for computing the MGU \cite{hv-09-unifalgs}.
 
 In higher-order logic, unification is performed modulo rules of
-$\alpha\beta\eta$-conversion. Under these conditions, the uniqueness of the most
-general unifier is not guaranteed. Consider the constraint $X \,
+$\alpha\beta\eta$-conversion. Under these conditions, the uniqueness of the MGU is not guaranteed. Consider the constraint $X \,
 (\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
@@
 $\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
-that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier,
+that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The MGU,
 when it exists, corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
@@
 ones. For example, if $s$ at $t$ are first-order, then there must exist a singleton
-CSU, i.e., the most general unifier.
+CSU, i.e., the MGU.
 
@@
 \sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.  With
-$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using a horizontal
-line to separate premises and the conclusion, four inference rules of
+$\mgu(s,t)$ we denote the MGU of $s$ and $t$. Using a horizontal
+line to separate premises and the conclusion, the four inference rules of
 superposition are as follows:
@@
 \medskip
-Note that with \emph{superposition rule} we refer to both \infname{SN} and \infname{SP}
-rules.
+We call these four rules \emph{generating rules} as their conclusions are added to
+the set of clauses. In the rest of this thesis we denote generating rules with a
+horizontal bar separating premises and conclusions.
+
 Let us give some intuition behind how these rules are applied in practice. The
@@
 Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
-mathematicians do when they replace equals by equals. The name paramodulation is
-also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
-states that an equation $s \eq t$ holds under condition $C$. Similarly, the right premise asserts that a (dis)equation holds
-under a condition. The superposition rule simply concatenates the conditions for
-both premises and replaces equals by equals in two main inference terms. For
-example given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and
-$\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
-between their last literals results in $\neg \cst{even}(Y) \llor \neg
-\cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.
+mathematicians do when they replace equals by equals. In calculi preceding
+superposition, the name paramodulation is often used for rules that replace
+equals by equals. Note that with \emph{the superposition rule} we refer to both
+\infname{SN} and \infname{SP} rules. The left premise intuitively states that
+an equation $s \eq t$ holds under condition $C$. Similarly, the right premise
+asserts that a (dis)equation holds under a condition. The superposition rule
+simply concatenates the conditions for both premises and replaces equals by
+equals in two main inference terms. For example given clauses $\neg
+\cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and $\neg \cst{even}(\cst{b}) \llor
+\cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition between their last literals
+results in $\neg \cst{even}(Y) \llor \neg \cst{even}(\cst{b}) \llor
+\cst{g}(\cst{a}) \eq \cst{b}$.
 
-% \[
-% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
-% \]
-% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
-% for resolution; \infname{ER} stands for equality resolution
-
-% \[
-% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
-% \]
-% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
-% for resolution; \infname{ER} stands for equality resolution
-
-% \[
-% \namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
-% \]
-% where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
-% $(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
-
-% \[
-% \namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
-% \]
-% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-% $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
-% \infname{SP} stands for positive superposition
-
-% \[
-% \namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
-% \]
-% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
-% $(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
-% \infname{SN} stands for negative superposition
 \medskip
@@
 s$ or tautological clauses are clearly justified by the redundancy criterion.
-Schulz gives an extensive list of such rules \cite{ss-02-brainiac}, including rewriting of positive
-literals which has more side conditions.
+Schulz gives an extensive list of such rules \cite{ss-02-brainiac}. This list
+includes the rule which allows rewriting positive literals but has more side
+conditions than \infname{RN}.
 
@@
 \looseness=-1
-In Chapter \ref{ch:intro} we mention that we identified three steps on the road
-to extending efficient superposition provers to higher-order logic. Together with
-our colleagues we did the necessary theoretical work by extending the superposition
-calculus using the same roadmap:
+In Chapter \ref{ch:intro} we described a roadmap (consisting of three steps) to
+extend an efficient superposition prover to higher-order logic. Together with
+our colleagues we did the necessary theoretical work by extending the
+superposition calculus using the same roadmap:
 
@@
 system for adding new rules and techniques to implemented calculi.
-This makes it attractive for fast evaluation of various extensions of
+This makes it attractive for fast evaluation of various promising extensions of
 superposition.
@@
 rather than highly optimized codebase. This makes it a good target for implementing
-successful higher-order techniques.
+successful higher-order techniques previously evaluated in Zipperposition.

### 2022-04-11 pre/pre.tex
wan smaller comments

@@
     chapters. We begin by describing the three logics that we work with in this
-    thesis: propositional logic, first-order logic and higher-order logic. Then,
+    thesis: propositional logic, first-order logic, and higher-order logic. Then,
     we explain the clausal structure which is the backbone of many calculi for
@@
 However, this simple approach is prohibitively expensive and modern tools that
-decide propositional satisfiability problem (SAT solvers) use more advanced
+decide the propositional satisfiability problem (SAT solvers) use more advanced
 approaches such as the CDCL calculus \cite{mss-96-cdcl}. A crucial improvement of
@@
 \ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and $n$
-is the \emph{arity} of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
+is the \emph{arity} of the symbol $\cst{f}$. If the return type of $\cst{f}$ is $o$ we call
 $\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.
@@
 Terms are used to build \emph{atoms}. An atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
-equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
+equation $s \eq t$ where terms $t$ and $s$ are of the same type (\emph{equational atom}). Atoms are
 combined using logical connectives to build formulas just like in propositional
@@
 variables gives rise to \emph{loose bound variables} \cite{tn-93-patterns},
-bound variables that appear without enclosing $\lambda$ binder  (e.g., $y$ in
+bound variables that appear without an enclosing $\lambda$ binder  (e.g., $y$ in
 $\lamx{y \, \cst{a}}$). Note that there is no requirement that a
@@
 \lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur as a loose bound variable in $s$
-and $y$ is not bound in $s$. Application of a substitution to a term,
+and $y$ is not bound in $s$. Application of a substitution to a term
 implicitly $\alpha$-renames bound variables to avoid capture.  For example, $\{X
@@
 
-In practice, many implementations of $\lambda$-terms avoid named representation
-of bound variables described above by using \emph{locally nameless
+In practice, many implementations of $\lambda$-terms avoid a named representation
+of bound variables described above by using a \emph{locally nameless
 representation} \cite{ac-12-locally-nameless}. In this representation, free
 variables retain their names, while abstractions are simply represented by a
-$\lambda$ symbol, without any names. Further, bound variables are represented
+$\lambda$ symbol, without any names. Further, bound variables can be represented
 using \emph{De Bruijn indices}. A bound variable with De Bruijn index
@@
 as $\lambda\, \cst{f} \, (\lambda \lambda\, \cst{g} \, \db{0} \, \db{2})$.
-Locally nameless representation reduces the $\alpha$-equivalence test to 
+The locally nameless representation reduces the $\alpha$-equivalence test to 
 the syntactic equality test.
@@
 Substitutions map only finitely many variables. Given a substitution $\varrho$,
-which maps $F$ to $s$, with $\varrho\setminus\{F \mapsto s\}$ we denote a
+which maps $F$ to $s$, with $\varrho\setminus\{F \mapsto s\}$ we denote the
 substitution that does not map $F$ and otherwise coincides with $\varrho$. Given
@@
 formulas are first-class citizens and higher-order calculi can in general work
-with nonnormalized problem, in its original form. However, resolution and
+with the nonnormalized problem, in its original form. However, resolution and
 superposition-based higher-order calculi still perform best when the problem is clausified
@@
 rules to the clauses, adding results of the inferences to the working set of
-clauses. As this process is quite prolific it also uses an order on terms which gives an indication of
+clauses. As this process is quite prolific, it also uses an order on terms which gives an indication of a
 term's ``simplicity'' to determine
@@
 substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
-$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
+$\sigma$ is a unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
 We always specify if the constraint should be interpreted as ordered or unordered.
@@
 the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
-$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient algorithms for computing the MGU \cite{hv-09-unifalgs}.
+$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient algorithms for computing MGUs \cite{hv-09-unifalgs}.
 
-In higher-order logic, unification is performed modulo rules of
-$\alpha\beta\eta$-conversion. Under these conditions, the uniqueness of the MGU is not guaranteed. Consider the constraint $X \,
+In higher-order logic, unification is performed modulo
+$\alpha\beta\eta$-equivalence. Under these conditions, the uniqueness of the MGU is not guaranteed. Consider the constraint $X \,
 (\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
@@
 denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
-is more general than the other. To generalize most general
-unifiers to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
+is more general than the other. To generalize MGUs to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
 introduced.
@@
 such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
-$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
+$\rho$ of $E$ there exists a $\sigma \in U$ and a substitution $\theta$ such
 that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The MGU,
@@
 within either positive or negative selected literals. We say \sigmacl{l} is
-eligible for paramodulation if it is positive, nothing is selected and
+eligible for paramodulation if it is positive, nothing is selected, and
 \sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.  With
@@
 superposition, the name paramodulation is often used for rules that replace
-equals by equals. Note that with \emph{the superposition rule} we refer to both
-\infname{SN} and \infname{SP} rules. The left premise intuitively states that
+equals by equals. Note that with \emph{the superposition rule} we will refer to both
+\infname{SN} and \infname{SP}. The left premise intuitively states that
 an equation $s \eq t$ holds under condition $C$. Similarly, the right premise
@@
 simply concatenates the conditions for both premises and replaces equals by
-equals in two main inference terms. For example given clauses $\neg
+equals in two main inference terms. For example, given clauses $\neg
 \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and $\neg \cst{even}(\cst{b}) \llor
@@
 
-Other than the four generating rules, superposition provers also use
-simplification rules, which are justified using the redundancy criterion. These
-rules effectively replace the premises with conclusions, and we denote them
+Next to the four generating rules, superposition provers also use
+simplification rules, which are justified by the redundancy criterion. These
+rules effectively replace the pre\-mises with conclusions, and we denote them
 using double bars. In theory, the rules work by adding conclusions and then
-using them to show that premises are redundant, and can thus be eliminated. Asf
+using them to show that premises are redundant, and can thus be eliminated. As
 an example of such a rule, let us introduce rewriting (demodulation) of negative
@@
 \]
-where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.
+where $\sigma = \mgu(u|_p, s)$ and $\sigma(s) \succ \sigma(t)$.
 
 Simple rules such as removing duplicate literals, literals of the form $s \noteq
-s$ or tautological clauses are clearly justified by the redundancy criterion.
+s$, or tautological clauses are clearly justified by the redundancy criterion.
 Schulz gives an extensive list of such rules \cite{ss-02-brainiac}. This list
@@
 We call the pair $(\mathcal{P}, \mathcal{A})$ the \emph{proof state}.
-The provers differ in which clauses are used for simplification: Otter-loop
+Provers differ in which clauses are used for simplification: Otter-loop
 \cite{mcw-1997-otter} provers use both active and passive clauses whereas
@@
     superposition calculus for higher-order logic devoid of
-    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
+    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called the \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
     and we shortly refer to it as \lfsup{}.
@@
     extended this calculus to support $\lambda$-abstraction
-    \cite{bbtvw-21-sup-lam}. They call this new calculus \emph{Boolean-free $\lambda$-superposition calculus},
+    \cite{bbtvw-21-sup-lam}. They call this new calculus the \emph{Boolean-free $\lambda$-superposition calculus},
     and we shortly refer to it as \lsup{}.
@@
 
-All the techniques described in this thesis are implemented in two superposition
+All the techniques described in this thesis have been implemented in two superposition
 theorem provers: Zipperposition
@@
 \ourpara{Zipperposition}
-Zipperposition is a higher-order theorem prover that implements \lfsup{},
+Zipperposition is a higher-order theorem prover implementing \lfsup{},
 \lsup{}, \osup{}, and other superposition-like calculi (including Bhayat and
@@
 In the last decade, together with its derivatives, it has revolved around the
-second place at the first-order division of CASC theorem prover
+second place at the first-order division of the CASC theorem prover
 competition \cite{ss-96-casc}. It is open-source, written in C (without using any
 external libraries), and designed around efficient algorithms and data structures
-rather than highly optimized codebase. This makes it a good target for implementing
+rather than a highly optimized codebase. This makes it a good target for implementing
 successful higher-order techniques previously evaluated in Zipperposition.

### 2022-04-11 acks/acks.tex
more

@@
 \end{quote}
-
+\noindent
 I do not know if that is true in general, but I have to admit I am guilty of
@@
 problem. To the students reading this chapter, I would very much like to assure
-you you are not.
+you that you are not.
 
@@
 like to thank Alexander Bentkamp, one of the most gifted mathematicians I know.
-I coauthored most of my papers with him, each of which was a pleasure to work
+I coauthored many of my papers with him, each of which was a pleasure to work
 on. I would also like to thank all of my collaborators and people that suggested
@@
 
-I want to thank my daily supervisor, Jasmin Blanchette. He always says that
-the main product of a PhD is not a thesis but a person you come out as after four
+I want to thank my daily supervisor, Jasmin Blanchette. He always says that the
+main product of a PhD is not a thesis but a person you come out as after four
 years. With his guidance, I learned to approach things more rigorously, slower,
@@
 presentation skills. Even though he tried very hard, he did not teach me to use
-articles though, and capitulated by adding them to all of the articles I wrote.
+articles (i.e., English articles ``the'', ``a'', or ``an'') though, and
+capitulated by adding them to many of the articles I wrote.
+
+My other two supervisors also helped me greatly. Wan Fokkink read the thesis
+very carefully and provided very valuable comments. For the last four years he
+made my PhD a smooth sail by solving all the administrative issues on time and
+provided a very nice work environment. Stephan Schulz is a real programming and
+logic wizard. His deep understanding of all things superposition and low-level
+Linux programming helped me not only with my PhD, but I also learned a lot of
+new skills that are undoubtedly useful for my future career.
 

### 2022-04-11 conclusion/conclusion.tex
more

@@
 Marina Abramovi\'c,  one of the greatest Yugoslav artists, comes to my mind. The
-themes that it explores such as taking risks and inevitability of making
+themes that it explores such as taking risks and the inevitability of making
 mistakes were the root of fears and doubts that I was slowly overcoming over the
@@
 Chapter \ref{ch:intro}), each of which has the same parts: theory (calculus and
-algorithms), engineering (implementation) and optimization (heuristics). While
+algorithms), engineering (implementation), and optimization (heuristics). While
 the calculi were mostly developed by Alexander Bentkamp
@@
 algorithms for indexing or unification, as well as the tasks of implementing and
-optimizing the implementation were mainly mine.
+optimizing the implementation, were mainly mine.
 
 Structuring the project cyclically turned out to be a great idea of Jasmin
-Blanchette. When I first started doing the research which will later form the
+Blanchette. When I first started doing the research which would later form the
 basis of Chapter \ref{ch:ehoh}, I was overwhelmed by the scope of the project
@@
 developed a feeling of where the possible pitfalls are and how to avoid them.
-This gave me an invaluable source of encouragement, without which I could hardly
-arrive at the end of the project.
+This gave me an invaluable source of encouragement, without which I could hardly have
+arrived at the end of the project.
 
@@
 The goal of the next cycle was to support $\lambda$-abstraction. When we first
-implemented \lsup{}, the calculus that achieves this goal, we quickly realized that
-one of its main flaws was that it relies on enumerating full unifiers rather
-than preunifiers, like some older resolution-based higher-order calculi do. This
-is why we focused on developing a full unification procedure that removes the
-redundancy present in other procedures. It also implements many advancements
-and optimizations described in the unification literature after the introduction
-of the most influential full unification procedure, Jensen-Pietrzykowski's
-procedure in 1970s. We also made sure that our procedure can easily be customized to
-trade bits of its completeness for substantially improved performance. This
-procedure, described in Chapter~\ref{ch:unif}, transformed the \lsup{} calculus
-into a competitive higher-order calculus.
+implemented \lsup{}, the calculus that achieves this goal, we quickly realized
+that one of its main flaws was that it relies on enumerating full unifiers. In
+comparison, some older resolution-based higher-order calculi enumerate
+preunifiers, which are less explosive and thus more manageable. This is why we
+focused on developing a full unification procedure that removes the redundancy
+present in other procedures. It also implements many advancements and
+optimizations described in the unification literature after the introduction of
+the most influential full unification procedure, Jensen-Pietrzykowski's
+procedure in the 1970s. We also made sure that our procedure can easily be
+customized to trade bits of its completeness for substantially improved
+performance. This procedure, described in Chapter~\ref{ch:unif}, transformed the
+\lsup{} calculus into a competitive higher-order calculus.
 
@@
 we manually analyzed hundreds of benchmarks which the competition proves, but
-on which Zipperposition fails. This analysis forms the basis of work that is
+on which Zipperposition failed. This analysis forms the basis of work that is
 described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}. The main
@@
 successful. After they were implemented in Zipperposition, it took its first
-victory at higher-order division of CASC.
+victory at the higher-order division of CASC.
 
-After its first victory, Zipperposition was successfully integrated in
+After its first victory, Zipperposition was successfully integrated into
 Sledgehammer and it is now part of its default installation. We also noticed
 that problems which only Zipperposition can solve do pop up from time to time in
-practice. Antoine Defourné also integrated Zipperposition in TLA+ proof
+practice. Antoine Defourné also integrated Zipperposition in the TLA+ proof
 assistant and observed that it helps prove some problems that were previously
@@
 with respect to full higher-order logic. After the first victory, we were left
-wondering what were the kind of problems that our implementation cannot
-solve, and started looking for complete full higher-order calculus. At the end,
+wondering what was the kind of problems that our implementation cannot
+solve, and started looking for a complete full higher-order calculus. At the end,
 we designed \osup{}, which gave us the precise answer in the form of problems that
@@
 of $\lambda$-superposition such as full higher-order unification and
-$\beta$-reduction-aware term orders will help us prove most of higher-order
+$\beta$-reduction-aware term orders would help us prove most of higher-order
 problems without the explosion of complete approaches.
@@
 While this proved mostly true, we quickly realized that \ehohii{} should
-assimilate most of incomplete Boolean reasoning techniques described in Chapters
+assimilate most of the incomplete Boolean reasoning techniques described in Chapters
 \ref{ch:bools} and \ref{ch:ho-techniques}. All of them were straightforward to
@@
 possible that support for this technique does not require profound changes to
-E(hoh)'s formula treatment, but due to the lack of time we did not explore this
+E(hoh)'s formula treatment, but due to a lack of time we did not explore this
 alley.
@@
 By manually expecting the TPTP benchmarks that are out of reach
-for \ehohii{}, but within the reach of Zipperposition, we realized that lack of
-dynamic clausification might be the weight necessary to tip the scale in
+for \ehohii{}, but within the reach of Zipperposition, we realized that resolving the lack of
+dynamic clausification might be the weight necessary to tip the scales in
 favor of \ehohii{} on TPTP benchmarks.
@@
 nature in the first-order case, as well as the right way to integrate equality
-in their definition was challenging. Predicate elimination posed challenges in
+in their definition, was challenging. Predicate elimination posed challenges in
 terms of finding the exact conditions under which it can be integrated with the
@@
 benchmarks and carefully examining the debugging information, we realized that
-this was not due to a bug in the implementation but due to incompatibility of
-blocked clause elimination with redundancy criterion. In the end, we were left
+this was not due to a bug in the implementation but due to the incompatibility of
+blocked clause elimination with the redundancy criterion. In the end, we were left
 somewhat disappointed to learn that SAT techniques do not scale good enough to be used
@@
 
-In Chapter \ref{ch:ho-techniques} we only scratched the surface of finding the
+In Chapter \ref{ch:ho-techniques} we only scratched the surface of finding
 appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
-reasoning it would be interesting to verify that heuristics that are still not
+reasoning, it would be interesting to verify whether heuristics that are still not
 ported are agnostic of the version of implemented higher-order superposition
@@
 
-Some of the SAT-based techniques described in Chapter~\ref{ch:satfol} are
+Some of the SAT-based techniques described in Chapter~\ref{ch:satfol} have been
 implemented in E as well. However, this implementation is still in its early
-phases as it is not properly tested and optimized. Verifying the techniques are
+phases as it is not properly tested and optimized. Verifying which techniques are
 prover-agnostic would increase the confidence in them. We also started the work

### 2022-04-11 summary/summary.tex
more

@@
 The first part extends E to support a fragment of higher-order logic devoid of
-$\lambda$-abstra\-ction, yielding a prover called Ehoh. Despite this extension is small in scope, 
-Ehoh shows stronger performance on proof assistant
-benchmarks than E.
+$\lambda$-abstra\-ction, yielding a prover called Ehoh. Despite this extension being small in scope, 
+Ehoh shows a stronger performance on proof assistant benchmarks than E.
 
@@
 important challenge we faced during this extension is that of higher-order
-unification. At the time we started, state-of-the-art procedure
+unification. At the time we started, the state-of-the-art procedure
 for full higher-order unification was developed in 1970s. We designed a new
@@
 completeness for performance. It is implemented in a
-prototype prover called Zipperposition, a less efficient, but more easily extendible
+prototype prover called Zipperposition, a less efficient but more easily extendible
 prover compared to E. Our evaluation shows that our
@@
 first-class Boolean terms. For this part we took inspiration from traditional
-automatic higher-order provers, based on tableaux. We fitted those techniques in
+automatic higher-order provers, based on tableaux. We fitted those techniques in the
 superposition context and implemented them in Zipperposition. They
@@
 chose the most successful approaches implemented in Zipperposition that can
-easily be ported to \ehohii{}. The results were once again positive: Our
+be ported easily to \ehohii{}. The results were once again positive: Our
 evaluation shows that \ehohii{} is the best prover on proof assistant
@@
 
+In conclusion, the work described in this thesis shows that first- and
+higher-order provers are much more alike than previously thought. Furthermore,
+we showed that through carefully designed and tuned extension, first-order
+prover can become an award-winning higher-order prover.
+
 \chapter*{Samenvatting}

### 2022-04-11 pre/pre.tex
more

@@
 Let us give some intuition behind how these rules are applied in practice. The
-\infname{ER} rule is used to replace equals by equals within a clause.
-Consider the formula $\iforall x.\,  x \eq \cst{2} \iimplies \cst{even}(x)$. In
-clausal form it is $X \noteq \cst{2} \llor \cst{even}(X)$. Applying the
-\infname{ER} rule effectively applies the precondition $X \eq \cst{2}$ to the
-conclusion $\cst{even}(X)$ to obtain $\cst{even}(\cst{2})$. The \infname{EF}
-rule can be used to find instances of the clause that can be further simplified.
-Consider the clause $\cst{f}(X) \eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor
-\cst{p}(X, Y)$. Applying \infname{EF} to its first two literals yields $\cst{a}
-\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ (with $\sigma = \{ Y \mapsto X \}$). This clause
-then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ using \infname{ER}.
+\infname{ER} rule is used to establish reflexivity of equality---the property
+that each term is equal to itself. Consider the formula $\iforall x.\,  x \eq
+\cst{2} \iimplies \cst{even}(x)$. In clausal form it is $X \noteq \cst{2} \llor
+\cst{even}(X)$. Applying the \infname{ER} rule effectively applies the
+precondition $X \eq \cst{2}$ to the conclusion $\cst{even}(X)$ to obtain
+$\cst{even}(\cst{2})$.
+
+Similarly to the factoring rule in the standard resolution calculus, the
+\infname{EF} rule can be used to find instances of the clause that duplicate
+some literals.
+Additionally, it takes care of some rarely occurring edge cases that are a consequence of restrictive
+side-conditions of other rules of the calculus. Consider the clause $\cst{f}(X)
+\eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor \cst{p}(X, Y)$. Applying
+\infname{EF} to its first two literals yields $\cst{a} \noteq \cst{a} \llor
+\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ (with $\sigma = \{ Y \mapsto X \}$).
+This clause then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X,
+X)$ using \infname{ER}.
 

### 2022-04-18 introduction/introduction.tex
more

@@
 is run on the translated problem. If the first-order prover finds the proof, the
-proof is reconstructed in the proof assistant. Modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
+proof is reconstructed in the proof assistant. Modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light},
 and Isabelle \cite{lc-88-isabelle} implement the approach using the \emph{hammers} CoqHammer
-\cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
+\cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer}, and Sledgehammer
 \cite{bn-10-sh}, respectively.

### 2022-04-18 pre/pre.tex
more

@@
 undecidable, no guarantees are given for unprovable problems.
-Before introducing the rules of the calculus, we provide definitions of these and other
+Before introducing the rules of the calculus, we provide definitions of the
 background concepts.
@@
 A \emph{most general unifier} ({MGU}) is a
-unifier $\sigma$, such that for any other unifier $\theta$, there is a
-substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
+unifier $\sigma$, such that for any other unifier $\varrho$, $\sigma$ is more general than $\varrho$. In first-order logic,
 the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
@@
 {\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
-clauses in $N$ such that each clause in this subset is $\prec$-smaller than $C$. More
+clauses in $N$ such that each clause in this subset is $\succ$-smaller than $C$. More
 generally, a clause (ground or not) is redundant in set $N$ if every ground
@@
 using double bars. In theory, the rules work by adding conclusions and then
-using them to show that premises are redundant, and can thus be eliminated. As
+using them to show that premises are redundant; after this, premises can be eliminated. As
 an example of such a rule, let us introduce rewriting (demodulation) of negative
@@
 such that $\sigmacl{C} \subseteq D$. Subsumption is one of the most important
-operations within a theorem prover. This rule does not adhere to the redundancy
+operations of a theorem prover. This rule does not adhere to the redundancy
 criterion, but can be justified using other mechanisms \cite{wtrb-20-sat-framework}.

### 2022-04-25 introduction/introduction.tex
more

@@
 \cite{jf-01-modern-logic}. Researchers in that time were mostly interested in
-finding a formal language that is expressive enough to describe complex
+finding a formal language that is expressive enough to describe many
 mathematical theories, yet intuitive and understandable enough to allow
-reasonably simple formal reasoning systems. First-order logic seemed to satisfy
-both requirements.  This formalism not only allows one to model simple
-logical relations such as ``if it rains, the road is muddy'' (or formally
-$\cst{rains} \iimplies \cst{muddy}$), but also to quantify over objects as in ``all
-people are mortal'' ($\iforall x.\, \cst{human}(x) \iimplies \cst{mortal}(x)$).
+reasonably simple formal reasoning systems. Arguably, first-order logic seemed to satisfy
+both requirements.  This formalism not only allows one to model simple logical
+relations such as ``if it rains, the road is muddy'' (or formally $\cst{rains}
+\iimplies \cst{muddy}$), but also to quantify over objects as in ``all people
+are mortal'' ($\iforall x.\, \cst{human}(x) \iimplies \cst{mortal}(x)$).
 
-\paragraph{Automatic Theorem Proving}The initial study of first-order logic and the rules to reason about statements in
-it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
-that there is a system of inference rules (calculus) such that for any valid
-statement a proof of validity can be constructed in this system. Around the same
-time, the first computers were leaving the imagination of the engineers and
-began automating many complicated processes.  Naturally, the question as to
-whether a computer can decide if a first-order statement is valid (i.e., a
-theorem) arose. Church \cite{ac-36-fol-undecidable} and Turing
-\cite{tm-37-undecidable} answered this question negatively in 1936.
+\paragraph{Automatic Theorem Proving}The initial study of first-order logic and
+the rules to reason about statements in it was prolific. In 1930 G\"odel
+\cite{kg-30-completeness-theorem} showed that there is a system of inference
+rules (calculus) such that for any valid statement (i.e., a theorem) a proof can be
+constructed in this system. Around the same time, the first computers were
+leaving the imagination of the engineers and began automating many complicated
+processes.  Naturally, the question as to whether a computer can decide if a
+first-order statement is valid  arose. Church
+\cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered this
+question negatively in 1936.
 
@@
 first-order logic remained too enticing. In 1960, Davis and Putnam
-\cite{dp-60-dp} described a procedure for checking validity of a
-first-order formula. As first-order logic is undecidable, this procedure
-terminated only on valid formulas. Even though it was efficient enough to prove
-only simple formulas, it was an impressive achievement for the time.
+\cite{dp-60-dp} described a procedure for checking validity of a first-order
+formula. This procedure terminated only on valid formulas, as Church and Turing
+proved no procedure can check validity of an arbitrary (possibly invalid) formula
+in finite time. Even though it was efficient enough to prove only simple
+formulas, it was an impressive achievement for the time.
 
@@
 disprove this statement. The first stop is to extend E to support
-quantification over functions that are already present in the theory.
+quantification over functions that are already present in the original problem.
 
 Higher-order logic also supports expressing functions that are not explicitly
-present in the theory. Suppose we want to prove there is a
+present in the original problem. Suppose we want to prove there is a
 function that returns 4 when given 2 and 5 when given 3 ($\iexists x. x(\cst{2})

### 2022-04-25 summary/summary.tex
more

@@
 are based on first-order logic. This means that the two systems must communicate
-through translations, which obfuscate the problem and likely make it more
+through translations, which obfuscate the conjecture being proved and likely make it more
 difficult than it originally was.
@@
 unification. At the time we started, the state-of-the-art procedure
-for full higher-order unification was developed in 1970s. We designed a new
+for full higher-order unification was the one developed in 1970s. We designed a new
 procedure inspired by this one that enumerates fewer redundant unifiers, has many
@@
 
-The last part concerns adding support for
-first-class Boolean terms. For this part we took inspiration from traditional
+The last part concerns adding native support for
+Boolean terms. For this part we took inspiration from traditional
 automatic higher-order provers, based on tableaux. We fitted those techniques in the
@@
 higher-order provers are much more alike than previously thought. Furthermore,
-we showed that through carefully designed and tuned extension, first-order
+we showed that through carefully designed and tuned extension, a first-order
 prover can become an award-winning higher-order prover.

### 2022-04-26 conclusion/conclusion.tex
more
future work

@@
-\chapter{Conclusion}
-\setheader{Conclusion}
+\chapter{Conclusion and Future Work}
+\setheader{Conclusion and Future Work}
 \label{ch:conclusion}
@@
 
-As it is customary, we will conclude the thesis with some remarks about the
-future work. \ehohii{} has been brought to the shape in which further
-experimentation with higher-order superposition is possible. From the experience
-gained in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}, we can point out
-dynamic clausification as the main feature that is not yet implemented. We plan
-on implementing it in future.
+% As it is customary, we will conclude the thesis with some remarks about the
+% future work. \ehohii{} has been brought to the shape in which further
+% experimentation with higher-order superposition is possible. From the experience
+% gained in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}, we can point out
+% dynamic clausification as the main feature that is not yet implemented. It is planned to implement it in the future.
 
-In Chapter \ref{ch:ho-techniques} we only scratched the surface of finding
-appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
-reasoning, it would be interesting to verify whether heuristics that are still not
-ported are agnostic of the version of implemented higher-order superposition
-calculus.
+% In Chapter \ref{ch:ho-techniques} we only scratched the surface of finding
+% appropriate heuristics. Now that \ehohii{} implements more advanced higher-order
+% reasoning, it would be interesting to verify whether heuristics that are still not
+% ported are agnostic of the version of implemented higher-order superposition
+% calculus.
 
-Some of the SAT-based techniques described in Chapter~\ref{ch:satfol} have been
-implemented in E as well. However, this implementation is still in its early
-phases as it is not properly tested and optimized. Verifying which techniques are
-prover-agnostic would increase the confidence in them. We also started the work
-of extending those techniques to full higher-order logic.
+% Some of the SAT-based techniques described in Chapter~\ref{ch:satfol} have been
+% implemented in E as well. However, this implementation is still in its early
+% phases as it is not properly tested and optimized. Verifying which techniques are
+% prover-agnostic would increase the confidence in them. We also started the work
+% of extending those techniques to full higher-order logic.
+
+\section*{Future Work}
+
+\looseness=-1
+Even though the work in this thesis completes the three-step cycle, there are
+still many unexplored alleys for future work. We separate them by the topic
+explored in one of the earlier chapters:
+
+\ourpara{Higher-Order Unification} The complete variant of unification procedure
+described in Chapter \ref{ch:unif} is proved complete with substantial
+restrictions on the order and kind of applied rules. To further
+remove redundancy, we plan to find alternative, less explosive unification rules
+that do not compromise completeness. Furthermore, we want to investigate if
+there are more combinations of unification rules that lead to redundant unifiers
+and find ways to remove those redundancy-inducing combinations.
+
+We also described only one pragmatic variant of the unification procedure. This
+variant was designed after we sifted through hundreds of benchmarks to get the
+taste of what unifiers occur in practice. It is possible that this pragmatic variant
+can be further tweaked to perform more efficiently without loosing many useful unifiers.
+
+Lastly, we fixed the order in which Zipperposition applies different unification
+rules. It would be interesting to see if applying imitation before projection
+(or vice versa) yields performance improvements on some kinds of problems. 
+We also postponed the application of the most explosive rules such as iteration
+to the very end. However, it would not be surprising that there are many hard hand-crafted
+problems on which applying these rules would lead to a proof more quickly.
+
+\ourpara{Pragmatic Techniques and Heuristics } Many of the Boolean reasoning
+techniques described in Chapter \ref{ch:bools} were either inspired by the ones
+implemented in traditional higher-order theorem provers or by the unsolved TPTP
+benchmarks. Since the work present in Chapter \ref{ch:bools} anticipated the
+complete higher-order calculus that could make these techniques obsolete, we
+implemented only the most successful ones that are easy to port to
+saturation-based provers.  Most notably, unlike Satallax
+\cite{cb-2013-satallax}, Zipperposition does not use the fact that there only
+finitely many functions of a type built only using function type constructor
+($\rightarrow$) and Boolean type ($o$). Implementing techniques for efficient
+enumeration of all Boolean functions of a given type could help Zipperposition
+prove problems that are still out of its reach, but can be solved by Satallax.
+
+In Chapter \ref{ch:ho-techniques} we explored how we keep the explosion inherent
+to \osup{} calculus under control. Due to time and resource
+constraints, in the evaluation of our new given-clause loop we left some
+parameters that control which streams are queried fixed. We plan to tune these
+parameters in the future.  Similarly, we provided only some of the clause
+priority functions that rely on higher-order features. We plan to investigate
+more proofs in detail to create new priority functions that more cautiously
+separate less and more useful higher-order inferences. Lastly, we observed that
+use of Ehoh as a backend greatly improves the performance of Zipperposition. We
+plan to further investigate the proofs that Ehoh returns and find out how
+Zipperposition can find them without the use of a backend.
+
+\ourpara{Further Extensions of \ehohii{}} \ehohii{} outperforms other provers on
+problems coming proof assistants. However, its performance on TPTP problems is
+slightly behind Zipperposition. We plan to further investigate the benchmarks on
+which \ehohii{} fails, but are proved by Zipperposition. One of the first
+features we plan implement is dynamic clausification, which can possibly help
+improve \ehohii{}'s performance on first-order problems as well. We also plan to
+experiment with implementing some of the \osup{} rules necessary for
+completeness such as \infname{FluidSup}. Heuristics of \ehohii{} are trained on
+a subset of Sledgehammer and TPTP problems. We plan to use other benchmark sets
+such as GRUNGE to make \ehohii{} efficient on a wider range of higher-order
+benchmarks.
+
+\ourpara{SAT-inspired eliminations for superposition} We already implemented
+some of the techniques described in Chapter~\ref{ch:satfol} in E. However, this
+implementation is not properly tested and optimized. We plan to put finishing
+touches to this implementation and test the effects of implemented preprocessing
+techniques on E's performance. There are more techniques used in SAT solving,
+notably bounded variable addition \cite{mhb-12-reencoding} and blocked clause
+addition \cite{ok-99-er}, that we did not port to first-order logic. We want to
+investigate if these techniques can be lifted to first-order logic. Lastly, we
+already started work on porting existing techniques to higher-order logic.

### 2022-04-26 pre/pre.tex
more

@@
 
-\emph{Atomic formulas} of propositional logic are propositional variables
+\emph{Atomic formulas} (\emph{atoms}) of propositional logic are propositional variables
 $\cst{p}, \cst{q}, \ldots$ and constants $\itrue$ and $\ifalse$. More complex
@@
 
-However, this simple approach is prohibitively expensive and modern tools that
+This simple approach is prohibitively expensive and modern tools that
 decide the propositional satisfiability problem (SAT solvers) use more advanced
@@
 \ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and $n$
-is the \emph{arity} of the symbol $\cst{f}$. If the return type of $\cst{f}$ is $o$ we call
+is the \emph{arity} of the symbol $\cst{f}$. If the return type of $\cst{f}$ is $o$, we call
 $\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.
@@
 is a formula and $x$ is a variable, then $\iforall x.\, \phi$, as well as
-$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
+$\iexists x.\, \phi$ are formulas. Intuitively, the first formula requires that $\phi$ holds
 for every value of $x$, while the second one requires that there is some $x$ for
@@
 inferences are computed in a systematic, fair manner, the empty clause ($\bot$) will
-eventually be derived for a provable problem. As first-order logic is
-undecidable, no guarantees are given for unprovable problems.
+eventually be derived for a provable (i.e., valid) problem. As first-order logic is
+undecidable, no guarantees are given for unprovable (i.e., invalid) problems.
 Before introducing the rules of the calculus, we provide definitions of the

### 2022-04-27 ehoh2/ehoh2.tex
new ehoh2 eval

@@
 %%% check if you agree. --JB
-problems and found that little overhead has been introduced by the
+problems and found that no overhead has been introduced by the
 extension to higher-order logic.
@@
   compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
-  and \NumberOK{6\%} more problems on SH benchmarks, where Ehoh was already very
+  and \NumberOK{7\%} more problems on SH benchmarks, where Ehoh was already very
   successful.
@@
   \ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
-  \NumberOK{166} SH problems and \NumberOK{21} TPTP problems that Zipperposition-coop
+  \NumberOK{181} SH problems and \NumberOK{24} TPTP problems that Zipperposition-coop
   does not. The lower number of uniquely solved TPTP problems is most likely 
@@
         Ehoh                  & 2105       & 2611      \\
-        \ehohii{}             & 2523       & {\bf 2766}\\
+        \ehohii{}             & 2533       & {\bf 2804}\\
         Leo-III               & 2282       & 1601      \\
@@
         Ehoh HO      & 538      \\
-        \ehohii{} FO & {\bf 539}\\
-        \ehohii{} HO & {\bf 539}\\
+        \ehohii{}  FO & 537      \\
+        \ehohii{}  HO & {\bf 541}\\
         \bottomrule

### 2022-04-27 acks/acks.tex
more
wan comments

@@
 PhD students experiences psychological distress, while one in three is at risk
-of a common psychiatric disorder. In the light of this fact, I want to write
+of a common psychiatric disorder. In light of this fact, I want to write
 this chapter not only to the people whose help and work indebted me, but also to
@@
 
-The fact that I was surrounded by very successful postdocs and PhD students on
+The fact that I was surrounded by very successful postdocs and PhD students in
 later years of their studies made me feel like I was the only one having this
@@
 
-What helped me go out of the self-doubting loop was actually taking action. From
+What helped me get out of the self-doubting loop was actually taking action. From
 the simplest ones, like saying ``I don't understand'' out loud instead of
-silently nodding when I have a trouble following scientific discussion, to the
+silently nodding when I had trouble following scientific discussion, to the
 ones with more impact on my career like actively looking for projects I could
@@
 organize my master programme so that I follow more theory courses which most
-likely lead  me to start an academic career. I also want to thank other postdocs
+likely led  me to start an academic career. I also want to thank other postdocs
 and PhDs that studied during my PhD or are still studying. Most notably, I would
@@
 I want to thank my daily supervisor, Jasmin Blanchette. He always says that the
-main product of a PhD is not a thesis but a person you come out as after four
+main product of a PhD is not a thesis but the person you come out as after four
 years. With his guidance, I learned to approach things more rigorously, slower,
@@
 
+Predrag Jani\v{c}i\'{c}, my professor from University of Belgrade who 
+inspired me to deepen my knowledge in various areas of computer science, suggested
+many improvements to the text of this thesis. I thank him as well.
+
 Lastly, I would like to thank Martijn. He was besides me for the last four years
 and helped me overcome many of the doubts and issues I mentioned. He also always
-patiently waits until my medication or second glass of Grüner Veltliner kick in
+patiently waits until my medication or second glass of Grüner Veltliner kicks in
 (not at the same time). I also want to thank Jelisaveta and Tara, which

### 2022-04-27 conclusion/conclusion.tex
more
wan comments

@@
 
-\ourpara{Higher-Order Unification} The complete variant of unification procedure
+\ourpara{Higher-Order Unification} The complete variant of the unification procedure
 described in Chapter \ref{ch:unif} is proved complete with substantial
@@
 We also described only one pragmatic variant of the unification procedure. This
-variant was designed after we sifted through hundreds of benchmarks to get the
+variant was designed after we sifted through hundreds of benchmarks to get a
 taste of what unifiers occur in practice. It is possible that this pragmatic variant
@@
 to the very end. However, it would not be surprising that there are many hard hand-crafted
-problems on which applying these rules would lead to a proof more quickly.
+problems on which applying these rules early on would lead to a proof more quickly.
 
@@
 \cite{cb-2013-satallax}, Zipperposition does not use the fact that there only
-finitely many functions of a type built only using function type constructor
+finitely many functions of a type built using only the function type constructor
 ($\rightarrow$) and Boolean type ($o$). Implementing techniques for efficient
@@
 In Chapter \ref{ch:ho-techniques} we explored how we keep the explosion inherent
-to \osup{} calculus under control. Due to time and resource
-constraints, in the evaluation of our new given-clause loop we left some
-parameters that control which streams are queried fixed. We plan to tune these
+to the \osup{} calculus under control. Due to time and resource
+constraints, in the evaluation of our new given-clause loop we fixed some
+parameters that control which streams are queried. We plan to tune these
 parameters in the future.  Similarly, we provided only some of the clause
@@
 \ourpara{Further Extensions of \ehohii{}} \ehohii{} outperforms other provers on
-problems coming proof assistants. However, its performance on TPTP problems is
+problems coming from proof assistants. However, its performance on TPTP problems is
 slightly behind Zipperposition. We plan to further investigate the benchmarks on
-which \ehohii{} fails, but are proved by Zipperposition. One of the first
+which \ehohii{} fails, but that are proved by Zipperposition. One of the first
 features we plan implement is dynamic clausification, which can possibly help
@@
 experiment with implementing some of the \osup{} rules necessary for
-completeness such as \infname{FluidSup}. Heuristics of \ehohii{} are trained on
+completeness such as \infname{FluidSup}. Heuristics for \ehohii{} are trained on
 a subset of Sledgehammer and TPTP problems. We plan to use other benchmark sets
-such as GRUNGE to make \ehohii{} efficient on a wider range of higher-order
+such as GRUNGE \cite{bgksu-19-grunge} to make \ehohii{} efficient on a wider range of higher-order
 benchmarks.
@@
 addition \cite{ok-99-er}, that we did not port to first-order logic. We want to
-investigate if these techniques can be lifted to first-order logic. Lastly, we
-already started work on porting existing techniques to higher-order logic.
+investigate whether these techniques can be lifted to first-order logic. Lastly,
+we already started work on porting existing techniques to higher-order logic.

### 2022-04-27 pre/pre.tex
teletubbies

@@
 
+\looseness=-1
 Terms are used to build \emph{atoms}. An atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
@@
 
+\looseness=-1
 A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$

### 2022-04-28 conclusion/conclusion.tex
more

--- 2022-04-26 conclusion/conclusion.tex
@@
 
-\ourpara{SAT-inspired eliminations for superposition} We already implemented
+\ourpara{SAT-Inspired Eliminations for Superposition} We already implemented
 some of the techniques described in Chapter~\ref{ch:satfol} in E. However, this

### 2022-04-29 introduction/introduction.tex
teletubbies
more

@@
 statements cannot be automatically proved. This is one of the reasons why large
-verification projects can take years to finish.
+verification projects can take years to finish. Thus, providing some level of proof automation
+can substantially increase productivity of users of proof assistants.
 
@@
   higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
+  The name of the prover is a word play which combines one of the hit songs by the
+  Teletubbies\footnote{\url{https://en.wikipedia.org/wiki/Teletubbies_say_\%22Eh-oh!\%22}}
+  and the fact that our prover is a higher-order (HO) extension of E.
   We had to modify the internal term representation of E to support $\lambda$-free
@@
 
+  \looseness=-1
   We also noticed that Zipperposition's support for first-class Booleans
   was lacking compared to competition. We have closely studied the problems on
-  which Zipperposition fails but are easily solved by other provers, and created
+  which Zipperposition fails but are solved by other provers, and created
   a set of incomplete rules that enhance Zipperposition's Boolean support.

### 2022-07-04 ho-techniques/ho-techniques.tex
from my old laptop

@@
   \, y \, \cst{a})$ and $\cst{p} \, (\lambda y. y^{100} \, \cst{b})$ have a finite CSU (in fact an MGU). 
-  When \texttt{Forward} context selection is enabled, % the subterm
-  $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$ is made the target of superposition inference, 
+  When \texttt{Forward} context selection is enabled, 
+  the target of superposition inference becomes $\cst{p} \, (\lambda y.\, X \, (\lambda x.\, x) \, \cst{a})$, 
   forcing computation of at least 100 unifiers (under the assumption that
@@
 
-    \vspace{2\jot}
+    \vspace{1\jot}
+
+    \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}
+      \assign{collected\_clauses}{\emptyset}
+      \assign{oldest\_streams}{\text{pop } \mathit{num\_oldest} \text{ oldest streams from } Q}
+      \For{$\mathit{stream}$ in $\mathit{oldest\_streams}$}
+        \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \textsc{ExtractClause}(Q, \mathit{stream})}
+      \EndFor
+      \State \Return $\mathit{collected\_clauses}$
+    \EndFunction
+    \vspace{3\jot}
+    \pagebreak[2]
 
@@
 
-  \pagebreak[2]
-  \Function{FairProbe}{$Q$, $\mathit{num\_oldest}$}
-    \assign{collected\_clauses}{\emptyset}
-    \assign{oldest\_streams}{\text{pop } \mathit{num\_oldest} \text{ oldest streams from } Q}
-    \For{$\mathit{stream}$ in $\mathit{oldest\_streams}$}
-      \assign{collected\_clauses}{\mathit{collected\_clauses} \mathrel\cup \textsc{ExtractClause}(Q, \mathit{stream})}
-    \EndFor
-    \State \Return $\mathit{collected\_clauses}$
-  \EndFunction
-
-  \vspace{2\jot}
+  % \pagebreak[2]
+
+  \vspace{1\jot}
 
@@
 
-  \vspace{2\jot}
+  \vspace{1\jot}
 
@@
   \EndWhile
-
   \State \Return $\mathit{status}$

### 2022-07-06 introduction/introduction.tex
more

@@
 mathematical theories, yet intuitive and understandable enough to allow
-reasonably simple formal reasoning systems. Arguably, first-order logic seemed to satisfy
+reasonably simple formal reasoning. First-order logic became a commonly used system that satisfies
 both requirements.  This formalism not only allows one to model simple logical
@@
 Integrating proof automation directly into the proof assistant is not an easy task.
-Most proof assistants are based on variants of higher-order logic, which is more
-expressive than first-order logic, but harder to automate as in general it does not even allow complete
+Most proof assistants are based on variants of higher-order logic, a logic expressive enough to
+describe complex mathematical theories
+out of reach for first-order logic.  However, it is harder to automate as in general it does not even allow complete
 proof calculi. To provide some level of automation, proof assistant developers

### 2022-07-06 ehoh/ehoh.tex
intro
more
martin comments 1

@@
 traversing all subterms of a term.\end{rep}
-Using the optimization, the prover traverses subterms recursively in a
-first-order fashion, considering all the prefixes of a given subterm
-together. %, at little additional cost.
+The prefix optimization allows the prover to traverse subterms recursively in a
+first-order fashion, and simultaneously consider the prefixes of a given subterm
+at almost no additional cost.
 %
@@
 always return the same value when given the same arguments, can be obtained by
-adding suitable axioms can be obtained by adding suitable axioms
+adding suitable axioms
 \cite[Sect.~3.1]{bbcw-21-lfho}.
@@
 \looseness=-1
-In \lfhol{}, a type signature is simply a type~$\tau$, in which the
-type constructor $\to$ can be nested---e.g., $(\iota \to \iota)\allowbreak
+In \lfhol{}, function types are built using the type constructor $\to$, which can be
+arbitrarily nested---e.g., $(\iota \to \iota)\allowbreak
 \to \iota.$
@@
 arguments, %as opposed to the declared arity,
-and rename the field \verb|num_args|\end{rep}. This representation, often called
+and rename the field to \verb|num_args|\end{rep}. This representation, often called
 ``spine notation,'' % \cite{cervesato-pfenning-2003},
@@
 We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
-that generalizes Baader and Nipkow \cite{bn-98-tr-and-all-that}.
+that generalizes the one of Baader and Nipkow \cite{bn-98-tr-and-all-that}.
 A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
@@
   \end{proof}
-\newpage
+
   \begin{lemma}\label{lemma:ehoh:unif-normal-imp-solved}
@@
 
-The superposition rules were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
+\begin{sloppypar}
+  The rules of the superposition calculus were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
   we repeat them with slightly simplified notation, as we do not repeat side conditions:
+\end{sloppypar}
 
@@
 Under context $\cst{f}\>[\phantom{.}]$, Ehoh considers the subterms
-$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; x \; \cst{b}$. It finds the
+$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; \cst{g} \; \cst{b}$. It finds the
 prefix $\cst{g} \; \cst{a}$ of $\cst{g} \; \cst{a} \; \cst{b}$ in the
@@
 and neither $\tuple{x}{K}$ nor $\tuple{y}{K}$ shares variables with $\tuple{x}{J}$.
-For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor a \eq b'$,
+For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor b \eq b'$,
 \infname{IR} can derive the existence of the inverse

### 2022-07-06 pre/pre.tex
martin comments 1

@@
 of (un)satisfiability and (in)validity are assumed. We write
-$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
+$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a formula
 set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
@@
 $s$ is a subterm of $t_i$ at position $p$, then $s$ is a subterm of
-$\cst{f}(\tuplen{t})$ at position $i.p$. A {\em context} is a term with zero or
+$\cst{f}(\tuplen{t})$ at position $i.p$ (with $.$ denoting prepending operation). A {\em context} is a term with zero or
 more subterms replaced by a hole $\square$. We write $C[\overline{u}_n]$ for the
@@
 Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$, results in
-replacing all mapped variables by the corresponding terms.
+replacing all variables of $t$ by the corresponding values of the mapping $\sigma$.
 The composition $\varrho\sigma$ of substitutions is defined by
@@
 \namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
-              {s \eq t \qquad \sigmacl{u[p \leftarrow \sigma(t)] \eq v \llor C \llor D}}
+              {s \eq t \qquad u[p \leftarrow \sigma(t)] \noteq v  \llor D}
 \]
-where $\sigma = \mgu(u|_p, s)$ and $\sigma(s) \succ \sigma(t)$.
+where $\sigma(s) = u|_p$ and $\sigma(s) \succ \sigma(t)$.
 

### 2022-07-07 bools/bools.tex
martin suda comments 2/2

@@
 type constructor $o$ as well as binary constructor $\rightarrow$
-to be in $\Sigmaty$. A type $\tau, \upsilon$ is either a type variable $\alpha \in
+to be in $\Sigmaty$. Types $\tau, \upsilon$ are either type variables $\alpha \in
 \Vty$ or of the form $\kappa(\tau_1, \ldots \tau_n)$ where $\kappa$ is an
@@
 These two rules hoist Boolean subterms $f$ to the literal level; therefore,
-some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
+some conclusions of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
 $f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
@@
 Leibniz equality often appears in TPTP problems  \cite{gs-17-tptp}. Since modern provers have native support
-for equality, it is usually beneficial to recognize and replace occurrences of Leibniz equality.
+for equality, it is usually beneficial to recognize and replace occurrences of Leibniz equality by the natively
+supported one.
 
@@
 which is only 6 problems short of configuration b. Disabling the \infname{Abs} and
-\infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
+\infname{Interpret$\lambda$} rules results in solving 25 and 31 problems fewer,
 respectively. The raw data show that in total, using configurations from Figure
@@
 this mode. On the other hand, preprocessing is fairly competitive: it solved
-only 8 problems less than the b configuration.
+only 8 problems fewer than the b configuration.
 

### 2022-07-07 conclusion/conclusion.tex
martin suda comments 2/2

--- 2022-04-27 conclusion/conclusion.tex
@@
 these problems occur in less than one percent of the whole TPTP library.
-Furthermore, some of \osup{}'s rule are so explosive that they are disabled in
+Furthermore, some of \osup{}'s rules are so explosive that they are disabled in
 most CASC portfolio configurations.
@@
 this will further improve the efficiency of hundreds of mathematicians and computer scientists using Isabelle.
-By manually expecting the TPTP benchmarks that are out of reach
+By manually inspecting the TPTP benchmarks that are out of reach
 for \ehohii{}, but within the reach of Zipperposition, we realized that resolving the lack of
@@
 which \ehohii{} fails, but that are proved by Zipperposition. One of the first
-features we plan implement is dynamic clausification, which can possibly help
+features we plan to implement is dynamic clausification, which can possibly help
 improve \ehohii{}'s performance on first-order problems as well. We also plan to

### 2022-07-07 ehoh2/ehoh2.tex
martin suda comments 2/2

@@
   from 2019 to 2021 in the first-order theorem division and ran Ehoh and \ehohii{}
-  both in this first-order (FO) mode and in higher-order (HO) mode.
+  both in this first-order (FO) mode and in higher-order (HO) mode using 30 s CPU and wallclock timeout.
 

### 2022-07-07 ho-techniques/ho-techniques.tex
martin suda comments 2/2

@@
 \llor \neglit{\cst q}$. The calculus also supports \emph{inner} delayed
-clausification, which uses only the core calculus rules to clausify problems.
+clausification, which does not enforce clausifying logical symbols in top-down direction, 
+and uses only the generating calculus rules to clausify problems.
 Even though this is the laziest approach to clausification, the earlier
@@
 problems.
-On TPTP, even though $+$\relax{BI} solves \NumberOK{three} problems less
+On TPTP, even though $+$\relax{BI} solves \NumberOK{three} problems fewer
 than \relax{DCS}$-$LA, it is very useful: \NumberOK{41}
@@
 $\itrue$, $\ifalse$, and variable-headed terms cannot be selected; for literals
-$s \eq t$, neither $s$ nor $t$ cannot be selected; if $s$ (or $t$) contains a
+$s \eq t$, neither $s$ nor $t$ can be selected; if $s$ (or $t$) contains a
 variable $X$ as a green subterm and $X\> \tuplen{u}$, with $n \ge 1$, is a
@@
 Ganzinger and Stuber considered Boolean subterm selection for their extension of
-first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus requires
+first-order superposition with interpreted Boolean type \cite{gs-05-boolsup}. Unlike our calculus, their calculus
 only allows the selection of subterms that occur in negative green positions, defined below.
@@
   In this example we simplify the notation by writing positive predicate literals
-  $\poslit{s}$ as $s$ and negative predicate literals $\neglit{t}$ as $\neg t$.
+  $\poslit{s}$ as $s$ and negative predicate literals $t \noteq \itrue$ as $\neg t$.
   Consider the unsatisfiable two-clause problem $\{ X \, (\cst{f} \,
@@
 possibly infinite streams and is much more prolific, proved only 15  problems
-less than the most competitive pragmatic variant. Furthermore, it proved 19
+fewer than the most competitive pragmatic variant. Furthermore, it proved 19
 problems not proved by the pragmatic variant.

### 2022-07-07 satfol/satfol.tex
martin suda comments 2/2

@@
 blocked clause elimination on satisfiable problems, including using it to
-establish equisatisfiability with an empty clause set or as an inprocessing
+establish equisatisfiability with the empty clause set or as an inprocessing
 rule. Unfortunately, we find that its use as inprocessing can compromise the
@@
 between $L \cor C$ and $\neglit{L_1} \cor \cdots \cor \neglit{L_n} \cor
-D$ (which are renamed apart) is $\sigmacl{C \cor D}$, where $\sigma$ is the MGU
+D$ (renamed apart) is $\sigmacl{C \cor D}$, where $\sigma$ is the MGU
 of the literals $L, L_1, \ldots, L_n$.
-Given a Herbrand model $\ourmodel$ of a problem, the following procedure
+Given a Herbrand model $\ourmodel$ of a clause set $N$, the following procedure
 removes all clauses while preserving satisfiability:
@@
     \If{$\impmap[\sigmaterm{L_\mathrm{a}}] \not = \emptyset$ for some renaming $\sigma$} \label{statement:alpha}
-      \State $(L_\mathrm{a}, L_\mathrm{c}) \gets (\sigmaterm(L_\mathrm{a}), \sigmaterm{L_\mathrm{c}})$
+      \State $(L_\mathrm{a}, L_\mathrm{c}) \gets (\sigmaterm{L_\mathrm{a}}, \sigmaterm{L_\mathrm{c}})$
     \EndIf

### 2022-07-07 unif/unif.tex
spellcheck
martin suda comments 2/2

@@
 \emph{selected} constraint in $E$. In this chapter, we consider \emph{freshness}
-of the variables with respect to the free variables ocurring in $E$. The transitions, defined as follows, are only
+of the variables with respect to the free variables occurring in $E$. The transitions, defined as follows, are only
 applied if the $\selected{\text{grayed}}$ constraint is selected. 
@@
       where $m \geq n$, $\overline{x}_m \not= \overline{y}_n$, $s$ and $t$ are not $\lambda$-abstractions,
-      $\overline{z}_m$ are bound variables fresh with respect to $s$ and $t$, $s' = \{x_1 \mapsto z_1, \ldots, x_m \mapsto z_m \}$,
-      and $t' = \{y_1 \mapsto z_1, \ldots, y_n \mapsto z_n \}$
+      $\overline{z}_m$ are bound variables fresh with respect to $s$ and $t$, $s' = \{x_1 \mapsto z_1, \ldots, x_m \mapsto z_m \}\,s$,
+      and $t' = \{y_1 \mapsto z_1, \ldots, y_n \mapsto z_n \}\,t$
   % \smallskip
@@
 % Also, the complete variant is sometimes called "full". Maybe we should stick to one of the two?
-% @ALEX: Investingating.
+% @ALEX: Investigating.
 assessing the complete variant and the pragmatic variant, the latter with several

### 2022-07-07 introduction/introduction.tex
reger suda voronkov

@@
 heuristics in first-order theorem proving. For example, Hoder and Voronkov
-evaluated different unification algorithms \cite{hv-09-unifalgs}, while Reger and
+evaluated different unification algorithms \cite{hv-09-unifalgs}, while Reger, Suda, and
 Voronkov evaluated different parameters of a novel prover architecture

### 2022-07-08 introduction/introduction.tex
geoff chapter 1

@@
 There I learned the value of formal, mathematical language and finally
-understood him. Not only does using a rigorous mathematical language avoid the
+understood it. Not only does using a rigorous mathematical language avoid the
 ambiguity of the natural one, it allows us to use the formal rules of reasoning
-to make conclusions from initial arguments in a trustworthy manner.
+to make conclusions from initial premises in a trustworthy manner.
 
@@
 \cite{cc-73-resolution-book}. Despite this progress, reasoning about equality of
-objects was still hard for these methods. This changed with introduction of
+objects was still hard for resolution. Even though some improvements, like
+paramodulation calculus \cite{nr-01-paramodulation}, improved this situation,
+reasoning with equality remained impractical. This changed with introduction of
 superposition~\cite{bg-94-superposition} in the early 1990s.
@@
 assistants are also used to show correctness of complex software such as
-compilers \cite{xl-09-compcert} or operating systems kernels \cite{gk-09-sel4}.
+compilers \cite{xl-09-compcert} and operating systems kernels \cite{gk-09-sel4}.
 This verified software can be used in safety-critical scenarios such as nuclear
-plants or autonomous vehicles.
+plants and autonomous vehicles.
 
@@
 automatic theorem provers through a translation. To bridge this translation gap,
-attempts were made to use higher-order automatic theorem provers instead of
+an attempt was made to use higher-order automatic theorem provers instead of
 first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
-higher-order theorem provers of the time were primarily designed for small and
-tricky problems. On the other hand, problems coming from hammers consist of
+higher-order theorem provers at the time of this attempt were primarily designed for small
+problems requiring complex higher-order reasoning. On the other hand, problems coming from hammers consist of
 mostly first-order formulas, are rather large, and mostly require simple
@@
 all-purpose binders in higher-order logic, are obfuscated to the point that all
-but the simplest  reasoning about them are out of reach for first-order provers.
+but the simplest  reasoning about them is out of reach for first-order provers.
 However, for higher-order provers the crux of the proof is matching the axiom
@@
 \paragraph{Our Approach} To improve the automatic reasoning for higher-order
-logic, we decided not to start from a blank slate and build a new theorem prover.
+logic, my colleagues and me decided not to start from a blank slate and build a new theorem prover.
 Rather, we start from a position of strength---from the state-of-the-art
@@
 
-We do not extend E to higher-order logic in one atomic step. Instead, we extend
+We did not extend E to higher-order logic in one atomic step. Instead, we extended
 it by adding support for features of higher-order logic one by one,
 which gradually increase its reasoning capabilities. With this gradual
-approach we make sure that after every extension, E's
+approach we made sure that after every extension, E's
 efficiency on first-order problems did not decrease. Thus, while trying
-to fulfill the point \ref{it:ho} from the wishlist we make sure that points
-\ref{it:fo} and \ref{it:size} are not affected.
+to fulfill the point \ref{it:ho} from the wishlist we made sure that points
+\ref{it:fo} and \ref{it:size} were not affected.
 
@@
 it treats Boolean terms (formulas) as first-class citizens, enabling us to
-model some concepts in a more natural manner. For example, we can naturally
-represent statements such as ``what Jasmin says is true'' as $\iforall x. \,
-\cst{says}(\cst{jasmin}, x) \iimplies x$. The last stop is to support Boolean terms
-as first-class citizens.
-\pagebreak[2]
+model some concepts in a more natural manner. For example, statements such as
+``what Jasmin says is true'' can be represented naturally as $\iforall x. \,
+\cst{says}(\cst{jasmin}, x) \iimplies x$. The last stop is to support Boolean
+terms as first-class citizens.
 
@@
 
-  Together with Alexander Bentkamp and other colleagues we developed
+  Alexander Bentkamp with the help of other colleagues and me developed
   a complete superposition calculus for the second phase described above
@@
 
-  Inspired by this success, we implemented the complete superposition
-  calculus for higher-order logic with Booleans (i.e., the third phase), that we developed with Bentkamp and other
-  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
-  ways our pragmatic extension was too weak to prove hard theorems of
-  higher-order logic. We further fine-tuned the heuristics and implemented new ways
-  to tame the search space explosion. In 2021 we released Zipperposition 2.1
-  which implemented these new features. It again won
-  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
-  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.
+  Inspired by this success, we implemented the complete superposition calculus
+  for higher-order logic with Booleans (i.e., the third phase), that was
+  developed by Bentkamp with the help of other colleagues and me
+  \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in
+  which ways our pragmatic extension was too weak to prove hard theorems of
+  higher-order logic. We further fine-tuned the heuristics and implemented new
+  ways to tame the search space explosion. In 2021 we released Zipperposition
+  2.1 which implemented these new features. It again won CASC, outperforming
+  Zipperposition 2. It was 16 percent points ahead of the runner-up, Vampire 4.5
+  \cite{lkav-13-vampire}.
 
   Building on all the experience we gathered while working on Zipperposition,
-  we further extended Ehoh to full higher-order logic, obtaining a prover we call
+  we further extended Ehoh to full higher-order logic, obtaining a prover called
   \newehoh{}. We again modified the term representation, implemented new unification and
@@
   TPTP benchmark set. Whenever we noticed that there were problems E and
-  Zipperposition could not solve, but the competition could, we looked for an
+  Zipperposition could not solve, but other competitive provers could, we looked for an
   explanation for this issue. This explanation would usually be in the form of
@@
 prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
-unification and added $\lambda$-terms to one of the best superposition
+unification and added $\lambda$-terms to one of the best first-order
 provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition

### 2022-07-08 summary/summary.tex
geoff chapter 1

@@
 \looseness=-1
-In the last decades, proof assistants were immeasurably useful in formally
+In the last decades, proof assistants have been immeasurably useful in formally
 proving validity of hard mathematical theories and correctness of

### 2022-07-09 ehoh/ehoh.tex
geoff ch 3

@@
 a prover developed primarily by Schulz. It is written in C and offers
-good performance, with more emphasis on ``brainiac''
-heuristics than on raw speed. E regularly scores among the top
+good performance. It gets its ``brainiac'' name by emphasizing 
+intelligent heuristics more than raw speed. E regularly scores among the top
 systems at CASC and is usually the strongest open-source prover in
@@
 extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
-and its semantics by interpreting the Boolean type $o$ as a domain of
-cardinality~2. Functional extensionality, the property that two functions are equal if they
+and its semantics by interpreting the Boolean type $o$ as a two-element domain. Functional extensionality, the property that two functions are equal if they
 always return the same value when given the same arguments, can be obtained by
@@
 %
-In particular,
-despite perfect sharing, their \verb|binding| fields are unconnected, leading
-to inconsistencies.
+In particular, when $x()$'s \verb|binding| field is updated, this does not affect the \verb|binding| of $x(\cst{b})$.
 
@@
 superposition \cite{bbww-17-kbo,bww-17-rpo}. For this
-reason, we focus on KBO.
+reason, in this chapter, we focus on KBO.
 
@@
 The main difference is that when comparing two terms $\cst{f} \; \tuple{s}{m}$
-and $\cst{f} \; \tuple{t}{n}$, because of partial application we may now
-have $m \not= n$; this required changing the implementation to
+and $\cst{f} \; \tuple{t}{n}$, because of partial application it is possible that $m \not= n$; 
+this required changing the implementation to
 perform a length-lexicographic comparison of the tuples $\tuple{s}{m}$ and
@@
   %
-  This algorithm closely follows the Ehoh implementation, abstracting away from
-  complications such as prefix optimization. We assume a flattened
+  An algorithm which closely follows the Ehoh implementation, abstracting away from
+  complications such as prefix optimization is presented below. We assume a flattened
   representation of terms; as in Ehoh, each variable stores the term it is bound
@@
 Matching can be reduced to unification by treating variables in $t$ as
-nullary symbols \cite{bn-98-tr-and-all-that}, but E implements it
+nullary symbols \cite{bn-98-tr-and-all-that}, but E implements matching
 separately.
@@
 \looseness=-1
-When generalizing the data structure to \lfhol, we face a complication
-due to partial application. First-order terms can only be stored in leaf nodes,
-but in Ehoh we must also be able to represent partially applied terms,
-such as $\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that
-$\cst{f}$ is unary and $\cst{g}$ is binary). Conceptually, this can be solved
-by storing a Boolean on each node indicating whether it is an accepting state.
-In the implementation, the change is more subtle, because several parts of E's
-code implicitly assume that only leaf nodes are accepting.
+When generalizing the data structure to \lfhol, we face a complication due to
+partial application. First-order terms can only be stored in leaf nodes, but in
+Ehoh we must also be able to represent partially applied terms, such as
+$\cst{f}$, $\cst{g}$, or $\cst{g}\;\cst{a}$ (assuming, as above, that $\cst{f}$
+is unary and $\cst{g}$ is binary). Conceptually, this can be solved by storing a
+Boolean on each node indicating whether it is an accepting state (i.e., if it
+corresponds to a term in the indexed set). In the implementation, the change is
+more subtle, because several parts of E's code implicitly assume that only leaf
+nodes are accepting.
 
@@
   traversed. It first goes through all child nodes labeled with a variable before
-  possibly visiting the child node labeled with a function symbol. We assume that
-  we can iterate through the children of a node using a function
+  possibly visiting the child node labeled with a function symbol. The children of a node can be iterated through using a function
   \textsc{NextVarChild} that, given a tree node and iterator through children,
@@
 $\sigma$ changes only if the variable~$x$ was previously not bound.
-Instead of creating and storing substitutions explicitly, we simply remember
+Instead of creating and storing substitutions explicitly, the algorithm simply remembers
 whether the variable was bound in this step or not, in the
-$\mathit{var\_unbound}$ tuple component. Then we rely on the label~$x$ of the
+$\mathit{var\_unbound}$ tuple component. Then it relies on the label~$x$ of the
 current node and its \textit{binding} field to carry the substitutions.
-Similarly, since our strategy is to traverse the tree by first visiting the
-variable-labeled child nodes, we need to remember how far we have come with
-this traversal. We store this information in the $\mathit{c\_iter}$ tuple
+Similarly, since the strategy is to traverse the tree by first visiting the
+variable-labeled child nodes, it needs to remember how far it has come with
+this traversal. This information is stored in the $\mathit{c\_iter}$ tuple
 component.
@@
 support extensionality via inference rules in Ehoh. We implemented two well-known
-incomplete rules we had experimented with in the context of Zipperposition.
+incomplete rules my colleagues and me had experimented with in the context of Zipperposition.
 
@@
 
-E's heuristics are largely independent of the logic used and work
+E's heuristics are largely independent of the logic used, and work
 unchanged for Ehoh. Yet, in preliminary experiments, we noticed that E proved
@@
 clauses share no symbols; the connection between them is hidden in the types
-of variables and symbols, which are ignored by the heuristics.
+of variables and symbols that are ignored by the heuristics.
 
@@
 %returns either $\{L_i\}$ for some $i$ or $\emptyset$.
-The most widely used function is probably
+The selection function E uses most often is probably
 \texttt{SelectMaxLComplexAvoidPosPred}, which we abbreviate to
@@
   terms with rigid heads. This makes them prolific in terms of possible
-  inference partners, a behavior we might want to avoid. On the other hand, shorter proofs
-  might be found if we prefer selecting applied variables. To cover both
-  scenarios, we implemented selection functions that prefer or defer selecting
-  applied variables. % and that break ties using \texttt{SelectMLCAPP}.
+  inference partners, a behavior that can lead to creating many unnecessary
+  clauses and should thus be avoided. On the other hand, shorter proofs might be
+  found if we prefer selecting applied variables. To cover both scenarios, we
+  implemented selection functions that prefer or defer selecting applied
+  variables. % and that break ties using \texttt{SelectMLCAPP}.
 
@@
   within a partition.
-  E's main loop visits, in round-robin fashion, a set of priority queues. From a given
+  E's main loop visits a set of priority queues in round-robin fashion. From a given
   queue, the clause with the highest priority and the smallest weight is selected.
@@
   Typically, one of the queues will use the clauses' age as priority, to ensure
-  fairness.
+  fairness (i.e., that each nonredundant clause is eventually chosen).
   \end{rep}
@@
 time. One of them is the configuration that is most often chosen by E's \emph{auto}
-mode. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
+mode when running on TPTP benchmarks. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
 %
@@
 One solution proposed by Miller \cite[Sect.~6]{dm-87-compact-proofs} is to ensure that
-Skolem symbols are always applied to a minimum number of arguments.
+Skolem symbols are always applied to a minimal number of arguments.
 %Accordingly, the version of \lfhol{} introduced by Bentkamp et al.\
@@
 not measured the effect of this choice, but we conjecture it is not substantial.
-Namely, appropriately chosen term order will result in a similar outcome without
+Appropriately chosen term order will result in a similar outcome without
 removing the defined symbol. We will enable definition unfolding again once we
-have added support for $\lambda$-terms.
+have added support for $\lambda$-terms (Chapter \ref{ch:ehoh2}).
 
@@
 unification of a variable with a Boolean constant was disallowed to avoid
-unsoundness. We needed to undo this in Ehoh. Ehoh must also remove trivial
+unsoundness. We needed to undo this in Ehoh. Ehoh also removes trivial
 literals $\ifalse \eq \itrue$ and $\itrue \noteq
@@
 applicative encoding, and compared with other provers?
-To answer the first question, we evaluated each new parameter independently. From
+To answer the first question, we evaluated each new heuristic scheme independently. From
 the empirical results, we derived a new configuration optimized for \lfhol{}.
@@
 value while keeping the other parameters the same as for \emph{boa}.
-\begin{rep}This gives an idea of how each parameter affects overall
+\begin{rep}This gave an idea of how each parameter affected overall
 performance.\end{rep}
@@
 
-Clause selection is the heuristic component that we extended the most. We must
+Clause selection is the heuristic component that we extended the most. We had to
 assess the effect of a new heuristic weight function, a multiplier for the
@@
 To test the effect of the new type-based weight function, we replaced
-\emph{boa}'s queue, which uses
+\emph{boa}'s queue that uses
 \texttt{4.CRS\-Weight(}\ldots\texttt{)}, with the queue ordered by
-\texttt{4.CRSTypeWeight(}\ldots\texttt{)}. We call the original
-heuristic \texttt{W} and the type-aware alternative \texttt{TW}.
-%
-We chose nine values for testing the effect of the applied
+\texttt{4.CRSTypeWeight(}\ldots\texttt{)}. The original
+heuristic is called \texttt{W} and the type-aware alternative \texttt{TW}.
+%
+We intuitively chose nine values for testing the effect of the applied
 variable multiplier \textit{appv\_mult}.
@@
 The genuine higher-order provers have
-the unfair advantage that they can instantiate higher-order
+the advantage that they can instantiate higher-order
 variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
@@
 
-We ran all provers except Satallax (which only supports THF) on first-order
+We ran all provers except Satallax (which supports only THF) on first-order
 benchmarks to measure the overhead introduced by our extensions, as
@@
 Fourth, the new \emph{hoboa} %configuration
-  outperforms \emph{boa} on
+  outperforms \emph{boa} on most
   higher-order problems, suggesting that it could be worthwhile to re-train
@@
 
-  The work described in this chapter required modifying many parts of the E
-  prover. The invariant that variables cannot be applied and that symbols are always
+  The work described in this chapter was large in scope: it required modifying almost all parts of the E
+  prover, from the parser, to the inference engine, to heuristics. 
+  The invariant that variables cannot be applied and that symbols are always
   passed the same number of arguments were entrenched in E's code,
@@
   %% Obvious, since Stephan is a coauthor. --JB
-  we found the generalization manageable and are now in a position to add
+  we found the generalization manageable. The generalization put us in a position to add
   support for $\lambda$-terms and higher-order unification, as discussed in Chapter~\ref{ch:ehoh2}.
@@
   adds $\lambda$-terms and second-order unification to
-  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
+  the first-order prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
   % The approach is pragmatic. , with little emphasis on completeness.
-  Zipperposition, also based on superposition, was extended
+  Zipperposition, based on superposition, was extended
   to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
@@
   In practice, on problems with a large first-order component, translations
-  perform very well compared with the existing native provers
+  perform very well compared with native provers
   \cite{ns-13-leo2sh}. Largely thanks to Sledgehammer, Isabelle often came
@@
   By removing the need for the applicative encoding, our work reduces
-  the translation gap. The encoding buries the \lfhol{} terms' heads under
+  the translation gap. The applicative encoding buries the \lfhol{} terms' heads under
   layers of $\cst{@}$ symbols. %\confrep{}{, which impacts the heuristics that inspect terms}.
@@
 Despite considerable progress since the 1970s, until a few years ago higher-order automated
-reasoning has not assimilated some of the most successful methods for
+reasoning had not assimilated some of the most successful methods for
 first-order logic with equality, such as superposition. We presented a
@@
 In Chapter \ref{ch:ehoh2} we describe the extension of Ehoh to full higher-order logic.
-It was heavily inspired by the techniques implemented in Zipperposition, which made it
-dominate the CASC competition in 2020 and 2021.
+It was heavily inspired by the techniques implemented in Zipperposition, which helped it
+dominate CASC in 2020 and 2021.
 \end{sloppypar}

### 2022-07-09 pre/pre.tex
geoff ch 3

@@
 
-We distinguish a set of base types $T$ which is required to have the Boolean
+We distinguish a set of base types $T$ which is required to include the Boolean
 type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
@@
 term order $\succ$ with the following properties:
-
+%
 \noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
@@
 the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
-$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient algorithms for computing MGUs \cite{hv-09-unifalgs}.
+$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient, linear-time algorithms for computing MGUs \cite{hv-09-unifalgs}.
 
@@
 ones. Even though orders and selection are used to reduce the search space,
-it still grows rather fast. To identify unnecessary clauses in the search
+many unnecessary clauses might be created. To identify unnecessary clauses in the search
 space, superposition features a \emph{redundancy criterion}
@@
 simplification rules, which are justified by the redundancy criterion. These
-rules effectively replace the pre\-mises with conclusions, and we denote them
+rules replace the pre\-mises with conclusions, and we denote them
 using double bars. In theory, the rules work by adding conclusions and then
@@
 In Chapter \ref{ch:intro} we described a roadmap (consisting of three steps) to
-extend an efficient superposition prover to higher-order logic. Together with
-our colleagues we did the necessary theoretical work by extending the
-superposition calculus using the same roadmap:
+extend an efficient superposition prover to higher-order logic. My colleagues
+and me did the necessary theoretical work by extending the superposition calculus
+using the same roadmap:
 
@@
 %
-Note that we use the name $\lambda$-superposition for both \lsup{}
+Note that in this thesis the name $\lambda$-superposition is used for both \lsup{}
 and \osup{}.
@@
 \ourpara{E} E is a state-of-the-art first-order prover based on superposition.
-In the last decade, together with its derivatives, it has revolved around the
+In the last decade, together with its derivatives, it was usually the
 second place at the first-order division of the CASC theorem prover

### 2022-07-11 unif/unif.tex
geoff ch 4

@@
 \cite{tl-16-facunif}, and a new fragment 
-we present in Sect.~\ref{sec:unif:solid-oracle}.
+presented in Sect.~\ref{sec:unif:solid-oracle}.
 %
@@
 problem belongs to a decidable fragment, but we take this idea a step further by
-checking this more efficiently and for every subproblem arising during
+checking this more efficiently, and for every subproblem arising during
 unification.
@@
 
-Last, the JP procedure does not allow to apply substitutions and
-$\beta$-reduce lazily.
+Last, the JP procedure does not allow applying substitutions and
+$\beta$-reducing lazily.
 The rules of simpler procedures (e.g., first-order
-\cite{hv-09-unifalgs} and pattern unification \cite{tn-93-patterns}) depend only on
+\cite{hv-09-unifalgs} and pattern unification \cite{tn-93-patterns}) depend on only
 the heads of the unification pair.
@@
 substitute and
-$\beta$-reduce only until the heads of the current unification
+$\beta$-reduce until only the heads of the current unification
 pair are not mapped by the substitution and are not $\lambda$-abstractions. 
@@
 a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
-Unless otherwise stated, we take unification constraint $s \unif t$ to be an
+Unless otherwise stated, we take the unification constraint $s \unif t$ to be an
 unordered pair of two terms of the same type. To ease notation, we do not write
@@
 
-The procedure requires an infinite supply of fresh free variables. These fresh
-variables must be disjoint from the variables occurring in the initial multiset
-$E$. Whenever a transition $(E,\sigma) \newunifarrow (E',\sigma')$ is made, all
-fresh variables used in $\sigma'$ are removed from the supply and cannot be used
-again as fresh variables.
+The procedure requires an infinite supply of fresh free variables that must be
+disjoint from the variables occurring in the initial multiset $E$. Whenever a
+transition $(E,\sigma) \newunifarrow (E',\sigma')$ is made, all fresh variables
+used in $\sigma'$ are removed from the supply and cannot be used again as fresh
+variables.
 
@@
 
-The form of the rules \unifrulename{OracleSucc} and \unifrulename{Bind} is similar: both
+The forms of the rules \unifrulename{OracleSucc} and \unifrulename{Bind} are similar: both
 extend the current substitution. However,
@@
 to certain classes of terms. On the other hand, \unifrulename{Bind} is applied to
-explore the whole search space for any given constraint. To compute CSU using efficient oracles and to discover
+explore the whole search space for any given constraint. To compute a CSU using efficient oracles and to discover
 failures early, \unifrulename{Bind} is applicable only if \unifrulename{OracleSucc}
@@
 \unifrulename{Dereference} partially normalize and partially apply the constructed
-substitution just enough to ensure that the heads are the ones we would get if
+substitution just enough to ensure that the heads are the ones that would be obtained if
 the substitution was fully applied and the term was fully normalized. 
@@
 input belongs to the appropriate fragment. To check whether a unification
-constraint is inside the fragment, we need to fully apply the substitution and
-$\beta$-normalize the constraint.
+constraint is inside the fragment, the  substitution must be fully applied and
+the constraint must be $\beta$-normalized.
 % To use an off-the-shelf oracle
@@
 % not depend on whether the terms belong to the fragment oracle manages. 
-Most oracles contain a decomposition operation which requires only a partial application of
+Most oracles contain a decomposition operation that requires only a partial application of
 the substitution and only partial $\beta$-normalization. If one of the constraints resulting from decomposition is not in
 the fragment, the original problem is not in the fragment. 
-This allows us to detect that the problem is not in the fragment without fully applying
+This allows detecting that the problem is not in the fragment without fully applying
 the substitution and $\beta$-normalizing.
@@
 
-Our procedure chooses child nodes based only on the heads of the
+Our procedure chooses child nodes based on only the heads of the
 selected unification constraint. In contrast, the JP procedure tracks all the
@@
 variable substitution cannot be integrated in the JP procedure a straightforward
-fashion. Moreover, as it does not feature a rule similar to
+fashion. Moreover, as the JP procedure does not feature a rule similar to
 \unifrulename{Decompose}, it always retraverses the already unified part of the
@@
 substitution guide the choice of applicable transition rule.
-We employ a measure based on values of $E_j$ and $\varrho_j$ that decreases with 
-each application of the rules. Therefore, eventually, we will reach the target substitution
-$\sigma$.
+Finally, a measure based on values of $E_j$ and $\varrho_j$ that decreases with 
+each application of the rules is employed. Therefore, eventually, the target substitution
+$\sigma$ will be reached.
 
@@
   \unif Y \, (\cst{f} \, \cst{a})$. The opponent pair for $\varrho\,X$ and $\varrho\,Y$
-  is $(\cst{f} \, x, y)$ which clearly adheres to the conclusion of Lemma \ref{lem:principal-lemma}.
+  is $(\cst{f} \, x, y)$, which clearly adheres to the conclusion of Lemma \ref{lem:principal-lemma}.
 %
@@
   problem $X \, \cst{a}
-  \unif Y \, (\cst{f} \, \cst{a})$ is $\varrho = \{ X \mapsto \lamx{\cst{c}}, Y \mapsto \lamx{\cst{c}} \}$
+  \unif Y \, (\cst{f} \, \cst{a})$ is $\varrho = \{ X \mapsto \lamx{\cst{c}}, Y \mapsto \lamx{\cst{c}} \}$,
   which has no opponent pairs.
@@
   \end{exa}
-  This definition is slightly too restrictive for our purposes.
+  Definition \ref{unif:def:omega-simple} is slightly too restrictive for our purposes.
   It is unfortunate that condition~1 requires $x$ to be applied to
@@
 
-Prehofer focused on the preunification problem as he remarks that the resulting
+Prehofer focused on the preunification problem, remarking that the resulting
 flex-flex pairs are ``intricate" \cite[Sect.~5.2.2]{cp-95-unifphd}. We discovered that these flex-flex
@@
   the unification problem to solve. Projecting~$F$ onto the first argument will
-  lead to a nonunifiable problem, so we perform imitation of~$\cst{g}$ building a
+  lead to a nonunifiable problem, so the imitation of~$\cst{g}$ is performed, yielding a
   binding $\sigma_1 = \{ F \mapsto \lambda x.\, \cst{g} \, (F_1 \, x) \, (F_2 \, x)
@@
   \, (\cst{f} \, \cst{a}) \unif \cst{a}, \allowbreak F_2 \, (\cst{f} \, \cst{a}) \unif G \,
-  \cst{a}\}$. Again, we can only imitate $\cst{a}$ for $F_1$---building a new 
+  \cst{a}\}$. Again, $\cst{a}$ can only be imitated for $F_1$---building a new 
   binding $\sigma_2 = \{F_1 \mapsto \lambda x.\, \cst{a} \}$. Finally, this
   yields the problem $\{ F_2 \, (\cst{f} \, \cst{a}) \unif G \, \cst{a} \}$.
-  According to Lemma \ref{unif:lemma:solid-vars-diff-mgu}, we find CSUs for the
+  According to Lemma \ref{unif:lemma:solid-vars-diff-mgu}, CSUs for the
   problems $J_1 \, \cst{a} = \cst{f} \, \cst{a}$ and $I_1 \, (\cst{f} \, \cst{a})
-  \unif \cst{a}$ using PT. The latter problem has a singleton CSU $\{I_1 \mapsto \lambda x.\,
+  \unif \cst{a}$ are found using PT. The latter problem has a singleton CSU $\{I_1 \mapsto \lambda x.\,
   \cst{a}\}$, whereas the former has a CSU containing $\{ J_1 \mapsto \lambda x.\,
   \cst{f} \, x \}$ and $\{ J_1 \mapsto \lambda x.\, \cst{f} \, \cst{a} \}$.
-  Combining these solutions, we obtain an MGU $\sigma_3 =
+  Combining these solutions, an MGU $\sigma_3 =
   \{ F_2 \mapsto \lambda x.\, H \, x \, x \, \cst{a},\,
   G   \mapsto \lambda x.\, H \, (\cst{f} \, \cst{a}) \, (\cst{f} \, x) \, x \}$ 
-  for $F_2 \, (\cst{f}\, \cst{a}) \unif G \, \cst{a}$.
+  is obtained for $F_2 \, (\cst{f}\, \cst{a}) \unif G \, \cst{a}$.
   %\begin{equation*}
@@
   %\end{equation*}
-  Finally, we get to the MGU of the original problem $\sigma = \sigma_3 \sigma_2 \sigma_1 = \{ F
+  Finally, the MGU of the original problem is computed: $\sigma = \sigma_3 \sigma_2 \sigma_1 = \{ F
   \mapsto \lambda x.\, \cst{g} \, \cst{a} \, (H \, x \, x \, \cst{a}),\allowbreak G \mapsto
@@
 Small examples that violate conditions of Theorem \ref{unif:thm:oracle} 
-and admit only infinite CSUs can be found easily. The problem $\{ \lambda x.\, F \, (\cst{f} \, x) \unif \lambda x.\, \cst{f} \,
+and admit only infinite CSUs can easily be found. The problem $\{ \lambda x.\, F \, (\cst{f} \, x) \unif \lambda x.\, \cst{f} \,
 (F \, x) \}$ violates variable distinctness and is a well-known example of a problem with only infinite CSUs.
@@
 We used OCaml's functors to create a modular implementation. The core of our
-procedure is implemented in a module which is parameterized by another module
-providing oracles and implementing the \textsf{Bind} step. In this way we can
+procedure is implemented in a module that is parameterized by another module
+providing oracles and implementing the \textsf{Bind} step. In this way we
 obtain the complete or pragmatic procedure and seamlessly integrate
@@
 ignored in the traversal. To ensure we do not remain stuck waiting for a unifier
-from a particular lazy list, the procedure will periodically return an empty
+from a particular lazy list, the procedure periodically returns an empty
 set, indicating that the next lazy list should be probed.
@@
 Zipperposition parameters have been fixed to the values of a variant of a
-well-performing configuration we used for the 2019 installment of the higher-order
-division of CASC theorem proving competition \cite{gs-19-casc27}. The cv
+well-performing configuration we used for the 2019 installment of THF
+division of the CASC theorem proving competition \cite{gs-19-casc27}. The cv
 configuration and all of the pv configurations use only pattern unification as
@@
 $d$, and $e$ are the limits of functional projections, eliminations, imitations,
-and identifications, respectively. Figure~\ref{fig:oracles} summarizes the
+and identifications, respectively. The values for $a,b,c,d,$ and $e$ are chosen following
+the intuition that functional projection and identification are the most explosive
+bindings.  Figure~\ref{fig:oracles} summarizes the
 effects of using different oracles.
@@
 competitive higher-order prover. Shortly after the procedure was implemented, in the 2020 edition of CASC, Zipperposition won
-the higher-order division.
+the THF division.
 

### 2022-07-11 bools/bools.tex
geoff ch 5

@@
 
-In Chapter~\ref{ch:intro} we motivated our work by the need to bridge the gap
+In Chapter~\ref{ch:intro} I motivated the work presented in this thesis by the need to bridge the gap
 between higher-order frontends and first-order backends. This gap is
@@
 requires translation. They described two approaches: One based on a calculus-level
-treatment of Booleans and other, which requires no changes to the calculus,
+treatment of Booleans and another that requires no changes to the calculus,
 based on preprocessing.
@@
 \ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
-arguments of symbols is translated to first-order logic.
+arguments of symbols is translated to first-order logic using a translation they describe.
 
@@
 level, delay clausification of nonatomic literals, and reason about formulas under
-$\lambda$-binders. Even though the rules we use are inspired by
+$\lambda$-binders. Even though the rules are inspired by
 the ones of refutationally complete higher-order provers, we do not guarantee
@@
 \looseness=-1
-We compare our native approach with two alternatives which are based on
+We compare our native approach with two alternatives that are based on
 preprocessing (Sect.~\ref{sect:bool:alternative}). First, we compare it to  an
@@
 \cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. This
-logic supports rank-1 polymorphism, as defined by the TH1 format of TPTP
+logic supports rank-1 polymorphism, as defined by the TH1 format of the TPTP
 \cite{ksr-16-th1}.
@@
 
-As mentioned in the previous chapter, Zipperposition is an open source prover
+As mentioned in Chapter \ref{ch:unif}, Zipperposition is an open source prover
 written in OCaml. From its inception, it was designed as a prover that supports
@@
 $\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
-the input problem, no nested Booleans will be modified or renamed using fresh
+the input problem, no nested Booleans are modified or renamed using fresh
 predicate symbols.
@@
 the need to alter Zipperposition's earlier behavior as little as possible. 
-Namely, negative atoms were not used as literals that can be used
+Namely, negative atoms are not used as literals that can be used
 to paramodulate from, and as such added to the laziness of the superposition calculus.
@@
 added to the set of choice functions $\mathit{CF}$ (which initially contains just
-$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
+$\varepsilon$). Later, elements of $\mathit{CF}$ are used to heuristically
 instantiate the axiom of choice. We reused the method of recognizing choice
@@
 Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
-(\inot (\cst{p} \iand \cst{q}))$. This problem is unsatisfiable, as
+(\inot (\cst{p} \iand \cst{q}))$. This clause is unsatisfiable, as
 the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
@@
   applying \infname{IC} on the top-level Boolean term. Thus, for the following
-  discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
+  discussion, we assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
   and $s'$ have the same head, and a functional or Boolean subterm. To speed up
@@
   clauses where they appear as a head of a term that has a functional or Boolean
-  subterm. This index will be empty for first-order problems, incurring no
+  subterm. This index is empty for first-order problems, incurring no
   overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
@@
 Unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
-will be removed. Namely, Booleans that are below $\lambda$-abstraction
+are removed. Namely, Booleans that are below $\lambda$-abstraction
 and contain $\lambda$-bound variables will not be preprocessed. They cannot be easily hoisted to the level of an atom in which
@@
 clausification this problem becomes $ \cst{sk}_1 \, (\cst{sk}_2 \, Y) \eq Y   $
-where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, we can use the
+where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, the
 \infname{ArgCong} rule \cite{bbtvw-21-sup-lam} which applies
-fresh variable $W$ to both sides of the equation, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
+fresh variable $W$ to both sides of the equation can be used, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
 \eq Y \, W$. Superposition-based higher-order theorem
-provers (such as Leo-III, Vampire, and Zipperposition) will split this clause
+provers (such as Leo-III, Vampire, and Zipperposition) split this clause
 into two clauses $C_1 = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \not\eq \top \llor Y
@@
 $$ \iexists x. \, \iforall y.  \, x \, y \iequiv (\iforall z. \, y \,z)$$ The
-problem is easily solved if we replace the variable $x$ with the constant $\forall$.
+problem is easily solved if the variable $x$ is replaced with the constant $\forall$.
 Moreover, the prover does not have to blindly guess this instantiation. Instead,
@@
 \infname{PI} is used, and if both clauses resulting from clausifying the negated
-conjecture are appropriately instantiated. In contrast, dynamic clausification will
-derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, X) $
+conjecture are appropriately instantiated. In contrast, dynamic clausification
+derives the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, X) $
 from the negated conjecture in three steps. Then, equality resolution results in
@@
 clausification and \infname{BoolER}, Zipperposition times out in any reasonable
-time limit; with those two options it solves the mentioned problems in less than
+time limit; with those two options it solves these two problems in less than
 100\,ms.
@@
 used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
-given in a portfolio mode when participating in CASC. Configuration {b} uses the pragmatic variant
+given in Zipperposition's portfolio mode when participating in CASC. Configuration {b} uses the pragmatic variant
 pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
@@
 different values of $k$. The results show that different values for $k$ have a
-modest effect on the success rate. The raw data reveal that when we focus our
+modest impact on the success rate. The raw data reveal that when we focus our
 attention to configurations with $k=2$, mode \emph{Full} can solve ten problems
@@
 though this is not evident from Figure \ref{fig:pi}, the sets of problems solved 
-by different modes somewhat differ.
+by different modes differ somewhat.
 
@@
 \looseness=-1
-Combining these two rules with dynamic clausification is more useful: When rule
+Combining BER and BEF rules with dynamic clausification is more useful: When rule
 \infname{IC} is replaced by rule \infname{DC}, the success rate increases
@@
 in close second, showing that our approach is a promising basis for further
-extensions. Indeed, as will be shown in Chapter \ref{ch:ho-techniques}, with
+extensions. Indeed, as is shown in Chapter \ref{ch:ho-techniques}, with
 fine-tuning of the heuristics and further extensions to the calculus, this

### 2022-07-12 ho-techniques/ho-techniques.tex
geoff ch 6

@@
 \blfootnote{In this work I was the main designer of all presented techniques,
-with the exception of inference streams which were designed by Alexander
+with the exception of inference streams that were designed by Alexander
 Bentkamp and Sophie Tourret. Alexander Bentkamp and Jasmin Blanchette also
@@
     implementation in the Zipperposition theorem prover. Largely thanks to their use,
-    Zipperposition won the higher-order division of the CASC competition in 2020 and 2021.  
+    Zipperposition won the THF division of the CASC competition in 2020 and 2021.  
   \end{abstract}
@@
 The landscape of higher-order proving techniques based on the extension of efficient
-first-order ones has tremendously expanded in the late 2010s and early 2020s. As
-mentioned in Sect.~\ref{sec:pre:ho-sup-calculi} we have implemented three
-higher-order calculi---\lfsup{}, \lsup{}, and \osup{}---which extend first-order superposition in a graceful way.
+first-order ones has expanded tremendously in the late 2010s and early 2020s. As
+mentioned in Sect.~\ref{sec:pre:ho-sup-calculi}, we have implemented three
+higher-order calculi---\lfsup{}, \lsup{}, and \osup{}---that extend first-order superposition in a graceful way.
 Bhayat and Reger also gracefully extended superposition to higher-order logic using
@@
 heuristics and rudimentary extensions of the calculus to deal with Booleans. It
-finished third at that year's higher-order division of CASC competition
+finished third at that year's THF division of CASC competition
 \cite{gs-19-casc27}, 12~percentage points behind the winner, the tableau prover
@@
 
-Many higher-order problems extensively use symbol definitions to simplify
-their representation. We describe several ways to exploit the definitions,
+Many higher-order problems use symbol definitions extensively to decrease the
+verbosity of representation. We describe several ways to exploit the
+definitions,
 %of which the most successful is
@@
 (Sect.~\ref{sec:ho-tech:formulas}). This calculus also supports \emph{Boolean selection
-functions}, a mechanism that allows us to choose on which Boolean subterms
-to perform inferences first.
+functions}, a mechanism that allows us to choose which Boolean subterms
+to perform inferences on first.
 We implemented some Boolean selection functions and
@@
 working with extensions of superposition, we assume a clausal structure
-(Sect.~\ref{sec:pre:clauses}). Like in the previous chapter, literals
+(Sect.~\ref{sec:pre:clauses}). As in the previous chapter, literals
 of clauses can contain arbitrary higher-order terms, including formulas.  
@@
 sequence of different configurations (possibly in parallel) until either the
-time limit is reached or a proof is found. We call this sequence a \emph{portfolio}.
+time limit is reached or a proof is found. This sequence is usually called a \emph{portfolio}.
 
@@
 used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
-TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
+TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP} in the figures of this chapter)
 and 1253 Sledgehammer-generated
@@
 definitions for the same symbol, only the first one is replaced by a rewrite rule.
-Then, whenever a clause is picked in the given clause procedure, it will be rewritten
+Then, whenever a clause is picked in the given clause procedure, it is rewritten
 using the collected rules.
@@
 
-The second technique aims at preserving completeness: We can try to choose, as the term order that
-parameterizes superposition, one that orients as many definitions as possible and rely on
+The second technique aims at preserving completeness: We can try to choose the term order that
+parameterizes superposition, as one that orients as many definitions as possible, and rely on
 demodulation to simplify the proof state. Usually, an instance of the Knuth--Bendix order (KBO)
@@
   $\cst{mnot} := \lambda p\, x. \, \inot \, p \, x$, $\cst{mor} := \lambda p\, q\, x.
-  \, p \, x \ior q \, x$,  and $\cst{mimplies} = \lambda p\, q. \allowbreak\, \cst{mor} \,
+  \, p \, x \ior q \, x$,  and $\cst{mimplies} := \lambda p\, q. \allowbreak\, \cst{mor} \,
   (\cst{mnot} \, p) \, q$. Assuming that the weight of $\lambda$, bound
   variables, and basic connectives is 2, we can orient equations using
-  the above described approach as follows. Starting from symbols that do not
+  the approach above described as follows. Starting from symbols that do not
   depend on the other ones, we set $\mathcal{W}(\cst{mnot}) = 11$ and
@@
 feature was used in the proof. For no-RW and no-RW$+$KBO, the
-\NumberOK{two}-problem difference may be just noise. Even though it proves fewer
+\NumberOK{two}-problem difference may just be noise. Even though it proves fewer
 problems, the configuration $o$-RW has some advantages over pre-RW: It proves
@@
 
-Higher-order logic identifies formulas with terms of Boolean type. To prove a problem, we often
+Higher-order logic identifies formulas with terms of Boolean type. To prove a conjecture, we often
 need to instantiate a variable with the right predicate.
@@
 by instantiating the contextualized Boolean selection function are shown in
-Figure~\ref{fig:bool-sel}. Rows denote the used partial order $\vartriangleright$,
+Figure~\ref{fig:bool-sel}. Rows denote the partial order $\vartriangleright$ which is used,
 while columns denote the function $\mathit{ctx}$.
@@
 \confrep{}{\emph{restricted combinatory unification}, }a predecessor of combinatory
-superposition) compute only a finite subset of the possible conclusions for
+superposition) compute only a finite subset of the possible conclusions of
 inferences that require enumerating a CSU. Not only is this a source of
@@
 
-The complete unification procedure described in Chapter \ref{ch:unif} returns such a stream. Other procedures such
+The complete unification procedure described in Chapter \ref{ch:unif} returns such a stream. Other procedures, such
 as Huet's \cite{gh-75-unification} and Jensen and
-Pietrzykowski's \cite{jp-76-unif} can easily be adapted to meet this
+Pietrzykowski's \cite{jp-76-unif}, can easily be adapted to meet this
 requirement. Based on the stream of unifiers interspersed with $\emptyset$, we
@@
 the chance that \textsc{HeuristicProbe} will return $\ParamMaxStreams$ clauses, as desired. Finally, if $P$ becomes empty, \textsc{ForceProbe}
-searches relentlessly for a clause in $Q$, as a fallback.
+searches relentlessly for a clause in $Q$, as a fallback. The default values
+for $\ParamMode{}, \ParamMaxStreams{},$ and $\ParamRetry{}$ are chosen by informal
+experiments.
 
@@
 Selecting a low number of best clauses seems to
-perform well on both benchmark sets. However, on SH benchmarks, which require
-overwhelmingly first-order unifiers, visiting older streams should be delayed
+perform well on both benchmark sets. However, on SH benchmarks, which mostly require 
+first-order unifiers, visiting older streams should be delayed
 a lot.
@@
   The figure indicates that it is preferable to invoke the backend early. We
-  have indeed observed that if the backend is invoked late, small clauses with
-  deep derivations tend to be present by then. These clauses might have been
+  have observed that if the backend is invoked late, small clauses with
+  deep derivations tend to be present. These clauses might have been
   used to delete important shallow clauses already. But due to their derivation
-  depth, they will not be translated. In such situations, it is better to
+  depth, they will not be sent to Ehoh. In such situations, it is better to
   invoke the backend before the important clauses are deleted.
@@
 parameters lead to noticeably different sets of proved problems. In an attempt
-to use Zipperposition~2 to its full potential, we have created a portfolio mode
+to use Zipperposition~2 to its full potential, we created a portfolio mode
 that runs up to 50 configurations in parallel during the allotted time. The portfolio
@@
 setup of the 2020 edition of CASC, we use a 120~s wall-clock limit and a 960~s CPU limit.
-We even carried out our evaluation on the 8-core CPU nodes that were used for
+We carried out our evaluation on the 8-core CPU nodes that were used for
 CASC in 2020. We also ran Zipperposition in uncooperative mode, in which its
@@
 modern first-order superposition prover such as E or Vampire would achieve
-even better results.
+even better results. In Chapter \ref{ch:ehoh2} we describe how a version of \osup{}
+is implemented in E.
 
 We believe that there are still techniques inspired by
-tableaux, SAT solving, and SMT solving that could be adapted and integrated in
+tableaux, SAT solving, and SMT solving that can be adapted and integrated in
 saturation provers. In particular, there are still \NumberOK{25} TPTP problems

### 2022-07-13 ehoh2/ehoh2.tex
geoff ch7

@@
   them, but much is lost in translation. As an alternative,
-  we propose to extend first-order provers with native support for
+  we propose to extend a first-order prover with native support for
   higher-order features. Building on our extension of E to $\lambda$-free
-  higher-order logic, we now extend E to full higher-order logic.
+  higher-order logic, we extended E to full higher-order logic.
   The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
@@
 state-of-the-art first-order prover to a fragment of higher-order logic devoid of $\lambda$-abstraction. This
-extension gave us a flavor of the difficulties that we might encounter on the
-way to full higher-order logic. In chapters that precede this one,
+extension gave us a feeling for the difficulties that might be encountered on the
+way to full higher-order logic. In the chapters that precede this one,
 we discussed many ways in which those difficulties can be overcome.
-In this chapter, we fulfill a promise we gave in the beginning of the thesis: We present the extension of
+In this chapter, we fulfill the promise given in the beginning of the thesis: We present the extension of
 Ehoh to full higher-order logic using incomplete variants
-of $\lambda$-superposition. We call this prover \ehohii.
+of $\lambda$-superposition. This prover is called \ehohii.
 
@@
 extensive experiments with various heuristic choices have been performed
-(Chapter \ref{ch:ho-techniques}). In \ehohii{}'s implementation, we used
-these experiences to choose a set of effective rules that could easily be
+(Chapter \ref{ch:ho-techniques}). In \ehohii{}'s implementation we used
+this experience to choose a set of effective rules that could easily be
 retrofitted into an originally first-order prover. Another principle that guided 
@@
 
-\ourpara{Background} The logic \ehohii{} targets is the monomorphic higher-order
+\ourpara{Background} The logic that \ehohii{} targets is the monomorphic higher-order
 logic described in Sect.~\ref{sec:pre:hol}. We reuse all the notions from this
-section, with the corresponding notations. Like in the previous chapter, we
+section, with the corresponding notations. As in the previous chapter, we
 simplify the notation by writing predicate literals in unencoded form: Positive
@@
 \ourpara{Representation of $\lambdabf$-Terms} To support full
-higher-order logic, Ehoh's $\lambda$-free cell data structure must
+higher-order logic, Ehoh's $\lambda$-free cell data structure had to
 be extended to support the $\lambda$ binder. We use the locally nameless
@@
 
-A term can be $\beta$-reduced as follows: When a cell of the
+A term is $\beta$-reduced as follows: When a cell of the
 form $ \internalat(\internallam(\db{0}, s),t)$ is encountered, the field
@@
 
-Second, in line with the gracefulness principle, we want \ehohii{} to incur
+Second, in line with the gracefulness principle, we wanted \ehohii{} to incur
 little (if any) overhead on first-order problems and to
@@
 %%% binder remains (the one to which the 2's point). --JB
-We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
-\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
+Then the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
+\ldots, \internallam(\db{0}, t) \ldots)$ can be removed and replaced %the body
 $t$ by the variant of
@@
     functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
-    which simulate basic superposition calculus rules,
+    that simulate basic superposition calculus rules,
     but do not require unifiability of the partner terms in the inference. More

### 2022-07-16 conclusion/conclusion.tex
geoff ch7-refs

@@
 After its first victory, Zipperposition was successfully integrated into
-Sledgehammer and it is now part of its default installation. We also noticed
+Sledgehammer and it is now part of Sledgehammer's default installation. We also noticed
 that problems which only Zipperposition can solve do pop up from time to time in
@@
 blocked clause elimination with the redundancy criterion. In the end, we were left
-somewhat disappointed to learn that SAT techniques do not scale good enough to be used
+somewhat disappointed to learn that SAT techniques do not scale well enough to be used
 as superposition simplification techniques. However, we learned that they are
@@
 restrictions on the order and kind of applied rules. To further
-remove redundancy, we plan to find alternative, less explosive unification rules
+remove redundancy, we hipe to find alternative, less explosive unification rules
 that do not compromise completeness. Furthermore, we want to investigate if
@@
 We also postponed the application of the most explosive rules such as iteration
-to the very end. However, it would not be surprising that there are many hard hand-crafted
+to the very end. However, it would not be surprising if there are many hard hand-crafted
 problems on which applying these rules early on would lead to a proof more quickly.

### 2022-07-16 satfol/satfol.tex
geoff ch7-refs

@@
 \begin{abstract}
-Optimized SAT solvers not only simplify the clause set during preprocessing, but
+Optimized SAT solvers simplify the clause set not only during preprocessing, but
 they also simplify it during solving. This interleaving of
@@
 \paper, we port inprocessing techniques to work with superposition, a leading
-first-order proof calculus, and we strengthen known preprocessing techniques.
+first-order proof calculus, and strengthen known preprocessing techniques.
 Specifically, we look into elimination of hidden literals, variables
@@
 
-Automated reasoning tools have become much more powerful in the last few decades
+Automated reasoning tools have become much more powerful in the last few decades,
 thanks to procedures such as conflict-driven clause learning (CDCL)
@@
 preprocessing and inprocessing techniques from propositional logic to clausal
-first-order logic and demonstrate their usefulness in a superposition prover.
+first-order logic, and demonstrate their usefulness in a superposition prover.
 
@@
 We also make use of Herbrand models and Herbrand's theorem \cite[Sect.~5.4]{mf-1996-fol}.
-In Herbrand's models the domain consists of all ground terms. These models have a 
-useful property that if a formula is satisfiable in a model, then it is also satisfiable in an
+In Herbrand models the domain consists of all ground terms. These models have a 
+useful property that a formula is satisfiable if and only if it is satisfiable in an
 Herbrand model. Herbrand's theorem states that every unsatisfiable clause set has (propositionally)
@@
 loop body, predicate elimination is systematically applied to the entire proof
-state. The first application of inprocessing amounts to preprocessing. By
-default, $K_\mathrm{occ} = 512$ and $K_\mathrm{iter}=10$. The same ideas and
+state. The first application of inprocessing amounts to preprocessing. After some informal experiments,
+we chose $K_\mathrm{occ} = 512$ and $K_\mathrm{iter}=10$ as default values. The analogous optimization and
 limits apply for blocked clause elimination.
@@
 solver, which is already integrated in Zipperposition. As modern theorem provers
-(including E and Vampire) also use SAT solvers, the method can easily be
-implemented.
+(including E and Vampire) also use SAT solvers, this definition set recognition method can easily be
+implemented in those provers as well.
 
@@
 increases the success rate, we evaluate BCE on all 2273
-satisfiable or TPTP FOF and CNF problems.
+satisfiable TPTP FOF and CNF problems.
 The hardware and CPU time limits are the same as in the experiments above.

### 2022-08-22 summary/summary.tex
more

@@
 are based on first-order logic. This means that the two systems must communicate
-through translations, which obfuscate the conjecture being proved and likely make it more
-difficult than it originally was.
+through translations, which obfuscate the conjecture being proved and likely
+make proving the conjecture more difficult.
 
@@
 gradually extend it to higher-order logic. As this approach is large in scope,
-we split it in three parts.
+we split it into three parts.
 
@@
 procedure inspired by this one that enumerates fewer redundant unifiers, has many
-modern optimizations built-in, and can easily be customized to trade its
+modern optimizations built in, and can easily be customized to trade its
 completeness for performance. It is implemented in a
@@
 
+In de afgelopen decennia zijn bewijsassistenten bijzonder nuttig gebleken in het formeel bewijzen van de validiteit van complexe wiskundige theorie\"en zowel als de correctheid van veiligheidskritieke software.
+Met behulp van deze tools kan men een probleem formeel beschrijven en machinaal controleerbare bewijzen voor uitspraken hierover produceren.
+Om de controle-fase effici\"ent en betrouwbaar te maken, dienen bewijsstappen van een zeer fijne granulariteit te zijn.
+Als gevolg hiervan moeten op het eerste gezicht triviale feiten in uiterst precies detail bewezen worden, wat het werken met bewijsassistenten buitengewoon omslachtig maakt.
+
+In een poging om belangrijke delen van dit bewijsproces te automatiseren, kunnen bewijsassistenten automatische bewijzers aanroepen om een bewijs af te maken.
+De meeste bewijsassistenten zijn echter op hogere-orde logica gebaseerd, terwijl de meeste automatische bewijzers op eerste-orde logica gestoeld zijn.
+Dit betekent dat de twee systemen moeten communiceren door vertalingen, die het probleem dat wordt bewezen vertroebelen en het bewijzen ervan bemoeilijken.
+
+In dit proefschrift proberen we deze vertaalkloof te overbruggen.
+We beginnen bij E, een van de beste eerste-orde bewijsassistenten backends, gebaseerd op de superpositie calculus, en we breiden hem geleidelijk uit tot logica's van hogere orde.
+Omdat deze aanpak een omvangrijke reikwijdte heeft, splitsen we hem in drie delen.
+
+Het eerste deel breidt E uit om een fragment van hogere-orde logica te ondersteunen zonder $\lambda$-abstractie, wat een bewijzer oplevert die Ehoh heet.
+Alhoewel deze uitbreiding klein van omvang is, laat Ehoh betere prestaties zien op benchmarks voor bewijsassistenten dan E.
+
+Het tweede deel betreft het toevoegen van ondersteuning voor $\lambda$-abstractie.
+De belangrijkste uitdaging waarmee we tijdens deze uitbreiding werden geconfronteerd is die van hogere-orde unificatie.
+Op het moment dat we begonnen was de state-of-the-art procedure voor volledige hogere-orde unificatie in de jaren zeventig ontwikkeld.
+We ontwierpen hierop ge\"{\i}nspireerd een procedure die minder redundante unifiers vindt, waarbij ook veel moderne optimalisaties zijn ingebouwd.
+Hij kan verder eenvoudig worden aangepast om niet alle unifiers te vinden, maar sneller te zijn.
+Hij is ge\"{\i}mplementeerd in een prototype prover genaamd Zipperposition, een minder effici\"ente maar gemakkelijker uitbreidbare prover, vergeleken met E.
+Uit onze evaluatie blijkt dat onze procedure een aanzienlijk verbetering oplevert in prestaties.
+
+Het laatste deel betreft het toevoegen van bestaande ondersteuning voor Boolse termen.
+Voor dit deel hebben we ons laten inspireren door traditionele automatische hogere-orde bewijzers, gebaseerd op tableaus.
+Die technieken hebben we in de superpositie context toegepast en in Zipperposition ge\"{\i}mplementeerd.
+Dit maakte Zipperposition in twee opvolgende jaren tot de beste bewijzer in de hogere-orde divisie van de jaarlijkse CASC automatische bewijzers competitie.
+
+Na het testen van onze idee\"en in Zipperposition, besloten we terug te gaan naar Ehoh en die uit te breiden tot volledige logica van hogere orde.
+Dit heeft geleid tot een bewijzer genaamd $\lambda$E.
+We hebben hiervoor de meest succesvolle methoden van Zipperposition gekozen die gemakkelijk kunnen worden overgezet naar $\lambda$E.
+De resultaten waren wederom positief: Onze evaluatie toont aan dat $\lambda$E de beste prover is op een benchmark voor bewijsassistenten, en tweede na Zipperposition op een standaard benchmark verzameling.
+
+We hebben vervolgens een stap terug gezet bij onze intentie om technieken van zwakkere naar sterkere logica's over te hevelen: We onderzochten SAT-vereenvoudi\-gingstechnie\-ken die kunnen worden ge\"{\i}mplementeerd in de superpositie context.
+We ontdekten dat alhoewel sommige ervan kunnen worden gebruikt tijdens het bewijzen, de meeste het beste werken als preprocessors.
+
+Concluderend laat het werk beschreven in dit proefschrift zien dat eerste- en hogere-orde bewijzers veel meer op elkaar lijken dan eerder werd gedacht.
+Verder hebben we laten zien dat door een zorgvuldig ontworpen en afgestemde uitbreiding, een eerste-orde bewijzer een bekroonde bewijzer voor hogere orde kan worden.
 }

### 2022-08-22 title/title.tex
from my old laptop
more

@@
 
-ter verkrijging van de graad van doctor
+ter verkrijging van de graad Doctor
 
-aan Vrije Universiteit Amsterdam,
+aan de Vrije Universiteit Amsterdam,
 
-op gezag van de Rector Magnificus prof.dr.~J.J.G.~Geurts,
+op gezag van de rector magnificus
 
-voorzitter van het College voor Promoties,
+prof.dr. J.J.G. Geurts,
 
@@
 
-op [onbekend]
+ten overstaan van de promotiecommissie van de Faculteit der Bètawetenschappen
+
+op dinsdag 18.10.2022. om 13.15 uur
+
+in een bijeenkomst van de universiteit,
+
+De Boelelaan 1105
 
@@
 \begin{tabular}{@{}p{0.2\textwidth}@{}p{0.8\textwidth}}
-  \textit{Keywords:} &  \\[\medskipamount]
+  \textit{Keywords:} & automatic theorem proving, higher-order logic, unification  \\[\medskipamount]
       \textit{Printed by:} &  \\[\medskipamount]
-      \textit{Cover:} &  \\[\medskipamount]
+      \textit{Cover:} &  Marina Abramović, Rhythm 10 (The Biography), Cibachrome Print,
+                         1997, © Marina Abramović,
+                         Courtesy of the Marina Abramović Archives \\[\medskipamount]
       \textit{Style:} & TU Delft House Style, with modifications by Moritz Beller \\& \url{https://github.com/Inventitech/phd-thesis-template} \\[\medskipamount]
@@
 %\medskip
-\noindent ISBN \ldots
+% \noindent ISBN \ldots
 
 \medskip
-\noindent An electronic version of this dissertation is available at \\
-\url{http://repository.tudelft.nl/}.
+% \noindent An electronic version of this dissertation is available at \\
+% \url{http://repository.tudelft.nl/}.
 

### 2022-08-25 title/title.tex
more

@@
 
-\vspace*{2\bigskipamount}
+VRIJE UNIVERSITEIT
+
+\vspace*{4\bigskipamount}
 
@@
 %\includegraphics{title}
-\vfill
+% \vfill
+\vspace*{5\bigskipamount}
 
@@
 
-{\Large\titlefont\bfseries Proefschrift}
+{ACADEMISCH PROEFSCHRIFT}
 
@@
 
-ter verkrijging van de graad Doctor
+ter verkrijging van de graad Doctor aan
 
-aan de Vrije Universiteit Amsterdam,
+de Vrije Universiteit Amsterdam,
 
@@
 
-ten overstaan van de promotiecommissie van de Faculteit der Bètawetenschappen
+ten overstaan van de promotiecommissie
 
-op dinsdag 18.10.2022. om 13.15 uur
+van de Faculteit der Bètawetenschappen
+
+op dinsdag 18 oktober 2022 om 13.45 uur
 
@@
 
-\bigskip
-\bigskip
+\smallskip
 
-%% Print the full name of the author.
-\makeatletter
-{\Large\titlefont\bfseries\@firstname\ \titleshape{\MakeUppercase{\@lastname}}}
-\makeatother
+Petar Vukmirovi\'c
 
-\bigskip
-\bigskip
+\smallskip
 
-Master of Science in Computer Science, \\
-Vrije Universiteit Amsterdam, Nederland,
-
-geboren te Belgrado, Servi\"e.
+geboren te Belgrado, Servi\"e
 
@@
 %% The following line is dictated by the promotieregelement.
-\noindent Dit proefschrift is goedgekeurd door de
 
@@
 \begin{tabular}{ll}
-    promotor: & prof.$\,$dr. W. J. Fokkink \\
+    promotor: & prof.$\,$dr. W. J. Fokkink \\[4\jot]
     copromotoren: & dr. J. C. Blanchette \\
-                  & prof.$\,$dr. S. Schulz
+                  & prof.$\,$dr. S. Schulz \\[4\jot]
+    promotiecommissie: & prof.$\,$dr. M. Hoogendoorn  \\
+                      & dr. K. Korovin               \\
+                      & dr. R. Piskac                \\
+                      & dr. A. Steen                 \\
+                      & dr. M. Suda                  \\
+                      & prof.$\,$dr. G. Sutcliffe    \\
+
 \end{tabular}
 
-\bigskip
-\noindent Samenstelling promotiecommissie:
+% \bigskip
+% \noindent Promotiecommissie:
 
-%% List the committee members, starting with the Rector Magnificus and the
-%% promotor(s) and ending with the reserve members.
-\medskip\noindent
-\begin{tabular}{p{4.5cm}l}
+% %% List the committee members, starting with the Rector Magnificus and the
+% %% promotor(s) and ending with the reserve members.
+% \medskip\noindent
+% \begin{tabular}{p{4.5cm}l}
 
-prof.$\,$dr. M. Hoogendoorn & Vrije Universiteit Amsterdam \\
-dr. K. Korovin & The University of Manchester \\
-dr. R. Piskac & Yale University \\
-dr. A. Steen & University of Greifswald \\
-dr. M. Suda & Czech Technical University in Prague \\
-prof.$\,$dr. G. Sutcliffe & University of Miami \\
+% prof.$\,$dr. M. Hoogendoorn & Vrije Universiteit Amsterdam \\
+% dr. K. Korovin & The University of Manchester \\
+% dr. R. Piskac & Yale University \\
+% dr. A. Steen & University of Greifswald \\
+% dr. M. Suda & Czech Technical University in Prague \\
+% prof.$\,$dr. G. Sutcliffe & University of Miami \\
 
-\end{tabular}
+% \end{tabular}
 
@@
 
-\medskip
-\medskip
+\newpage
 % TODO Include http://www.win.tue.nl/ipa/?page_id=309
